### 2022-07-18 gtlc.tex
Hello, synthetic world!

\documentclass{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{tikz-cd}
\usepackage{mathpartir}
\usepackage{stmaryrd}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{nonnum-theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{definition}{Definition}[section]
\newtheorem{lemma}{Lemma}[section]

\newcommand{\dyn}{\text{Dyn}}
\newcommand{\dynv}{\text{Dyn}^+}
\newcommand{\dync}{\text{Dyn}^-}
\newcommand{\good}[1]{\text{good}_{#1}}
\newcommand{\prop}{\mathbb P}
\newcommand{\with}{\mathrel{\&}}
\newcommand{\eqv}{\simeq}
\newcommand{\soln}{\text{Sol}}
\newcommand{\police}[1]{\text{police}_{#1}}
\newcommand{\later}{{\blacktriangleright}}
\newcommand{\errp}[1]{\mho_{#1}}
\newcommand{\kerrp}[2]{\mho^{#1}_{#2}}
\newcommand{\err}{\mho}
\newcommand{\errlift}{L_\err}
\newcommand{\ostar}{\circledast}

\newcommand{\ltdynp}[1]{\mathrel{\sqsubseteq_{#1}}}
\newcommand{\ltdynlift}[1]{\mathrel{\sqsubseteq_{\errlift#1}}}
\newcommand{\ltprecp}[1]{\mathrel{\sqsubseteq^{\prec}_{\errlift #1}}}
\newcommand{\ltsuccp}[1]{\mathrel{\sqsubseteq^{\succ}_{#1}}}

\newcommand{\kltp}[2]{\mathrel{\sqsubseteq^{#1}_{#2}}}

\newcommand{\ltgt}[1]{\mathbin{\langle\mspace{-1mu}#1\mspace{-1mu}\rangle}}
\newcommand{\fmap}{\ltgt{\$}}
\newcommand{\monto}{\mathrel{\to_{m}}}
\newcommand{\linto}{\multimap}

\begin{document}

\title{Semantics of Gradual Types in SGDT}
\author{Max S. New}

\maketitle

\section{Goals}

We want to define a denotational semantics for the core language of
GTLC and prove graduality.

Here's roughly the structure required
\begin{verbatim}
record GTLC(U)
  type : U
  tm : type -> U
  _<ty=_ : PreOrder type
  _<tm=_ : PreOrder (Sigma A (tm A))
  impl : (A, M) <tm= (B,N) -> A <ty= B

  dyn : type
  dynTop : {A} -> A <ty= dyn

  ans : type
  yes : tm ans
  no  : tm ans

  arr : type -> type -> type
  lam : (tm A -> tm B) -> tm (arr A B)
  app : tm (arr A B) -> tm A -> tm B
  -- beta, eta

  up : A <ty= B -> tm A -> tm B
  upR : ...

  dn : A <ty= B -> tm B -> tm A

\end{verbatim}

\section{Error predomains and domains}

First, we need to define our monad.
\begin{definition}
  Define the error-lift type as a solution to the guarded recursive
  domain equation:
  \[ \errlift X \cong (X + 1 + \later \errlift X) \]

  We label the three injections as follows:
  \begin{align*}
    \eta &: X \to \errlift X\\
    \err &: \errlift X\\
    \theta &: \later \errlift X \to \errlift X
  \end{align*}

  This carries the structure of a strong monad, with $\eta$ as the
  unit and the join defined by L\"ob induction as follows:
  \begin{align*}
    \mu(\eta x) &= x\\
    \mu(\mho) &= \mho\\
    \mu(\theta(u)) &= \theta (\mu^\later \ostar u)
  \end{align*}
  With map (implying strength) defined as
  \begin{align*}
    \errlift f(\eta x) &= f x\\
    \errlift f(\mho) &= \mho\\
    \errlift f(\theta(u)) &= \theta ((\errlift f)^\later \ostar u)
  \end{align*}
  And the equational laws of a monad can be proven using L\"ob
  induction.
\end{definition}

This can be seen as simply a composition of two monads using a
distributive law: the error monad and the recursive lift monad.
%
However, we present them as one combined monad because in gradual
typing we are interested in a specific ordering relation called the
\emph{error ordering} where $\err$ is the least element. This is
\emph{not} the composition of the lift and error monad, which would
instead have the error as an isolated, maximal element.

We can define this error ordering as a variation on the weak
simulation relations of prior work:
\begin{definition}
  Let $X$ be a type and $\sqsubseteq_X$ a binary relation. We define
  $\ltdynp{\errlift X} : \errlift X \to \errlift X \to \prop$ as follows:
  \begin{align*}
    \err \ltdynlift X d &= \top\\
    \eta x \ltdynlift X \eta y &= x \ltdynp X y\\
    \eta x \ltdynlift X \err &= \bot\\
    \eta x \ltdynlift X \theta t &= \exists{n:\mathbb N}.\exists{y: X}. \theta t = \delta^n (\eta y) \wedge x \ltdynp X y\\
    \theta s \ltdynlift X \eta y &= \exists{n:\mathbb N}.(\theta s = \delta^n \err)\vee(\exists{y: X}. \theta t = \delta^n (\eta y) \wedge x \ltdynp X y)\\
    \theta s \ltdynlift X \err &= \exists{n:\mathbb N}.\theta s = \delta^n \err\\
    \theta s \ltdynlift X \theta t &= \later[l \leftarrow s,r\leftarrow t] s \ltdynlift X t
  \end{align*}
\end{definition}

\begin{lemma}
  When $\ltdynp X$ is reflexive, so is $\ltdynlift X$.
\end{lemma}
\begin{proof}
  By L\"ob induction on the proposition $\forall d: \errlift X. d
  \ltdynlift X d$. Error and return cases are immediate.  We need to
  show
  \[ \theta t \ltdynlift X \theta t = \later[l \leftarrow t,r\leftarrow t] l \ltdynlift X r = \later[l \leftarrow t] l \ltdynlift X l \]
  The inductive hypothesis $h : \later (\forall d: \errlift X. d \ltdynlift X d)$
  We can then conclude
  \[ \text{next}[f \leftarrow h, l \leftarrow t]. h l \]
\end{proof}

Unfortunately, transitivity is not preserved.
%
The problematic case is of the form $\theta s \ltdynlift X \theta t
\ltdynlift X \eta z$. In this case we know $\theta t = \delta^n (\eta
y)$ but we cannot conclude that $\theta s$ must eventually terminate!
In fact this is false in the semantics because it might be that
$\theta s$ is only related to $\theta t$ because we ran out of fuel,
and actually $s$ is a diverging computation.

However, this should support transitive reasoning \emph{in the limit},
i.e., if the equivalence holds for \emph{all} levels of approximation
then eventually if $\theta t$ terminates, at high enough level of
approximation it will no longer be related to the diverging
computation.
%
This reasoning \emph{in the limit} is accomplished analytically by
quantification over all steps.
%
Here we accomplish it by using \emph{clocks} and \emph{clock
quantification}.
%
That is rather than a denotation of a syntactic type being a type it
will be given by a ``clocked'' type:

\begin{definition}
  A \emph{clocked} type $X$ is an element of $K \to U$.

  We call $\forall k:K. X k$, the \emph{limit} of $X$.
\end{definition}

Note that, similar to the interval in cubical type theory, $K$ is not
a type, but only a ``pseudo-type'' that we can quantify over in the
type $\Pi_{k:K} B$.
%
Next, we need a ``clocked'' error ordering associated to every type.
%
This will be assumed to be reflexive at each level of approximation,
but only transitive in the limit.

\begin{definition}
  A \emph{clocked} preorder on a clocked type $X : K \to U$ is a
  clocked relation $\kltp {} X : \forall k. X_k \to X_k \to \prop$ satisfying
  \begin{enumerate}
  \item Clocked Reflexivity $\forall k. \prod_{x:X_k} x \kltp k X x$
  \item Transitivity in the limit $\prod_{x,y,z:\forall k. X_k} x \kltp \omega X y \to y \kltp \omega X z \to x \kltp \omega X z$
  \end{enumerate}

  Where 
  \[ x \kltp \omega X y = \forall k. x_k \kltp k X y_k \]

  A \emph{monotone function} is a function $f : \forall k. X_0 k \to
  Y_0 k$ satisfying monotonicity:
  \[ \forall k.\prod_{x,x':X_0^k} x \kltp k X x' \to f_k x \kltp k Y f_k x' \]
\end{definition}

For any clocked preorder we can construct the delayed preorder by
\begin{definition}
  Define $\later : K \to (K \to U) \to (K \to U)$ by
  \[ (\later X)^k = \later^k X^k \]
  and on preorders by:
  \begin{align*}
    x \kltp k {\later X} y &= \later^k[l\leftarrow x, r\leftarrow y].~l \kltp k {X} r
  \end{align*}

\end{definition}

Then we can define our semantic notions
\begin{definition}
  An \emph{error predomain} $X$ consists of
  \begin{enumerate}
  \item A clocked type $X_0$
  \item A clocked preorder $\kltp {} X$ on $X_0$ 
  \end{enumerate}
  A morphism of error predomains is a monotone function.

  An \emph{error domain} is an error predomain $X$ with
  \begin{enumerate}
  \item A least element $\kerrp {} X : \forall k. X$, i.e., it
    satisfies $\forall k. \prod_{x:X_k} \kerrp k X \kltp k X x$.
  \item A monotone ``thunk'' function $\theta_X : \later X \to X$
  \end{enumerate}
  A morphism of error domains is a morphism of the underlying error
  predomains.

  A \emph{linear} morphism of error domains $S : X \multimap Y$ is a
  morphism of error domains additionally satisfying
  \begin{enumerate}
  \item Error preservation $S^k \err^k_X = \err^k_Y$
  \item Thunk preservation $S^k (\theta^k_X x) = \theta^k_Y (S^k \fmap x)$
  \end{enumerate}
\end{definition}

Finally we can return to our original goal of defining the monad for our semantics!
%

\begin{lemma}
  next and $\delta$ are monotone
\end{lemma}

\begin{definition}
  Let $X$ be an error predomain a binary relation. We define the error
  domain $\errlift X$ as

  TODO: put the ks in the right spots

  \begin{enumerate}
  \item $(\errlift X)_0^k = \mu L. X_0^k + 1 + \later^k L$
  \item Ordering defined by cases:
    \begin{align*}
      \err \kltp k {\errlift X} d &= \top\\
      \eta x \kltp k {\errlift X} \eta y &= x \ltdynp X y\\
      \eta x \kltp k {\errlift X} \err &= \bot\\
      \eta x \kltp k {\errlift X} \theta t &= \exists{n:\mathbb N}.\exists{y: X}. \theta t = \delta^n (\eta y) \wedge x \ltdynp X y\\
      \theta s \kltp k {\errlift X} \eta y &= \exists{n:\mathbb N}.(\theta s = \delta^n \err)\vee(\exists{y: X}. \theta t = \delta^n (\eta y) \wedge x \ltdynp X y)\\
      \theta s \kltp k {\errlift X} \err &= \exists{n:\mathbb N}.\theta s = \delta^n \err\\
      \theta s \kltp k {\errlift X} \theta t &= \later[l \leftarrow s,r\leftarrow t] s \kltp k {\errlift X} t
    \end{align*}
  \item reflexive: as above
  \item TODO: transitive in the limit
  \item TODO: strong monad
  \end{enumerate}
\end{definition}

\begin{lemma}
  $\errlift X$ is the free error domain on $X$, i.e., it is left
  adjoint to the forgetful functor from the category of error domains
  and linear maps to the category of predomains.
\end{lemma}

\begin{lemma}
  The adjunction is monadic: error domains are precisely algebras of
  the free error domain. Is this true?
\end{lemma}

\section{Lazy Semantics}

To define a semantics for lazy gradually typed language, we first need
to define our \emph{universal} error domain. Unless we add in features
like pattern matching on the dynamic type, there is some freedom in
this choice. We present two natural solutions:

The first is a classic-looking dynamic type: the lift of the sum of
the base type and the function type. A notable difference are the
guarded recursive uses.
\[ \dyn_+ \cong \errlift (2 + (\later \dyn_+ \monto \later \dyn_+)) \]
The second is slightly closer to untyped lambda calculus: we take a
product rather than a sum, and only need to lift the base type itself.
\[ \dyn_- \cong \errlift 2 \times (\later \dyn_- \monto \later \dyn_-) \]
The idea here is that any dynamically typed term can be safely applied
to any number of arguments and at some point ``projected'' to return a
boolean, which will actually cause computation to occur as indicated
by the $\errlift$.

To construct these domains we need to be able to solve recursive
domain equations \emph{of error domains}. To do this we simply need to
show that the universe of error predomains is a predomain.

\begin{definition}
  Define $\theta_{\predom} : \later \predom \to \predom$ as follows
  \begin{align*}
    \theta_{\predom}P &= (\later[(X,\ltdynp X) \leftarrow P] X, )
  \end{align*}
\end{definition}

\end{document}

### 2022-07-27 paper.tex
init a paper

\documentclass[acmsmall,screen]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{10.1145/1122445.1122456}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
  June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}

\begin{document}

\title{Denotational Semantics for Gradual Typing in Synthetic Guarded Domain Theory}
\author{Eric Giovannini}
\author{Max S. New}

\begin{abstract}
  We develop a denotational semantics for a gradually typed language
  with effects that is adequate and proves the graduality theorem.
  %
  The denotational semantics is constructed using \emph{synthetic
  guarded domain theory} working in a type theory with a later
  modality and clock quantification.
  %
  This provides a remarkably simple presentation of the semantics,
  where gradual types are interpreted as ordinary types in our ambient
  type theory equipped with an ordinary preorder structure to model
  the error ordering.
  %
  This avoids the complexities of classical domain-theoretic models
  (New and Licata) or logical relations models using explicit
  step-indexing (New and Ahmed).
  %
  In particular, we avoid a major technical complexity of New and
  Ahmed that require two logical relations to prove the graduality
  theorem.

  By working synthetically we can treat the domains in which gradual
  types are interpreted as if they were ordinary sets. This allows us
  to give a ``na\"ive'' presentation of gradual typing where each
  gradual type is modeled as a well-behaved subset of the universal
  domain used to model the dynamic type, and type precision is modeled
  as simply a subset relation.
  %
  This provides a rich semantic setting in which to 
\end{abstract}

\section{Introduction}

% gradual typing, graduality

% synthetic guarded domain theory, denotational semantics therein

% Difficulties in prior semantics

\section{Discussion}

\subsection{Synthetic Ordering}

While the use of synthetic guarded domain theory allows us to very
conveniently work with non-well-founded recursive constructions while
abstracting away the precise details of step-indexing, we do work with
the error ordering in a mostly analytic fashion in that gradual types
are interpreted as sets equipped with an ordering relation, and all
terms must be proven to be monotone.
%
It is possible that a combination of synthetic guarded domain theory
with \emph{directed} type theory would allow for an a synthetic
treatment of the error ordering as well.

\end{document}

### 2022-08-17 denotational-gradual-typing.tex
sketch of a naive view of gradual types

\documentclass{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{tikz-cd}
\usepackage{mathpartir}
\usepackage{stmaryrd}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{nonnum-theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{definition}{Definition}[section]
\newtheorem{lemma}{Lemma}[section]

\newcommand{\dynv}{\text{Dyn}^+}
\newcommand{\dync}{\text{Dyn}^-}
\newcommand{\good}[1]{\text{good}_{#1}}
\newcommand{\prop}{\mathbb P}
\newcommand{\with}{\mathrel{\&}}
\newcommand{\eqv}{\simeq}
\newcommand{\soln}{\text{Sol}}
\newcommand{\police}[1]{\text{police}_{#1}}
\newcommand{\later}{{\blacktriangleright}}

\begin{document}

\title{Gradual Types as Subsets}
\author{Max S. New}

\maketitle

We fix a predomain $\dynv$ with poset structure and a domain $\dync$
with bounded below poset structure as solutions to the domain
equations:

\[ \dynv = 2 + \dync \]
\[ \dync = (\dynv \to \dync) \with L (\dynv_{\mho}) \]

Q: does this actually get us the right order?
Q: what domain-theoretic properties does the order have that matter?

There are several ways to define essentially the same interpretation
of CBV gradually typed lambda calculus with booleans into this
semantics:

\begin{enumerate}
\item Types as ((Co-)Kleisli) co-reflections
\item Types as interior operators
\end{enumerate}

I want to explore one that I previously dismissed as somewhat na\"ive:
value types are (well-behaved) subsets of the universal predomain and
computation types are (well-behaved) quotients of the universal
domain.

\begin{enumerate}
\item A value type $A$ is a \emph{policeable} predicate $\good A :
  \dynv \to \prop$
\item A computation type $B$ is a \emph{envelopable} equivalence
  relation $\eqv B : \text{Equivalence}(\dynv)$
\end{enumerate}

Let's explain what these mean.
%
First, define $\soln A$ to be the set of ``good'' elements of $\dynv$
according to $A$, i.e., $\{ d : \dynv \,|\, \good A d \}$.
%
$\good A$ is \emph{policeable} when there exists a function $\police A
: \dynv \to (\soln A)_\mho$ satisfying two properties:
\begin{enumerate}
\item (monotonicity) $\forall d \sqsubseteq d'. \police A d \sqsubseteq \police A d'$
\item (retraction) $\forall x \in \soln A. \police A x = \eta x$
\item (projection) $\forall d \in \dynv. \police d \sqsubseteq \eta d$
\end{enumerate}

If $\sqsubseteq$ is a poset and $\eta : A \to A_\mho$ is order-reflecting,
then $\police A$ is uniquely determined by these two properties, for
if there were another such function $(\police A)'$ we would have
\[ \]

\subsection{Synthetic Guarded Domain Theory}

We can solve the domain equation using synthetic guarded domain
theory.

\begin{align*}
  \dynv &= \mu (\lambda D^+. 2 + \mu (\lambda D^-. L (1 + D^+) \times (D^+ \to D^-) ))\\
  \dync &= \mu (\lambda D^-. L (1 + \later \dynv) \times (\later \dynv \to D^-) )
\end{align*}

unfolding this gives us

\begin{align*}
  \dynv &\cong 2 + \dync \\
  \dync &= L (1 + \later \dynv) \times (\later \dynv \to \later \dync)\\
  &\cong L (1 + \later \dynv) \times \later (\dynv \to \dync)
\end{align*}

\begin{lemma}{Universal Domain}
  $\dync$ is a domain with algebra
  \[ \theta_\dync = \lambda p: \later \dync. (\pi_1p, ) \]
\end{lemma}

Let's define orderings.
\begin{verbatim}
data _<=_ : Dyn+ -> Dyn+ -> Prop where
  bool : forall b : 2. inl b <= inl b

data _<=_ : Dyn- -> Dyn- -> Prop where

err-ord : BinRel a -> BinRel (a + 1)
err-ord _<=_ (inr *)       y = T
err-ord _<=_ (inl x) (inl y) = x <= y
err-ord _<=_ _ _ = _|_

laterr-ord< BinRel a -> BinRel (L (a + 1))
laterr-ord< _<=_ (done (inr *)) _ = T
laterr-ord< _<=_ (done (inl *)) _ = T

laterr-ord> BinRel a -> BinRel (L (a + 1))

err : Dyn-
err = mu (\e. ((inl (inl *)), \x. e))

\end{verbatim}

Let's define

\end{document}

### 2023-01-12 paper-new/paper.tex
init new paper

\documentclass[acmsmall,screen]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{10.1145/1122445.1122456}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
  June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}

\newcommand{\dyn}{?}
\newcommand{\ltdyn}{\sqsubseteq}

\newcommand{\uarrowl}{\mathrel{\rotatebox[origin=c]{-30}{$\leftarrowtail$}}}
\newcommand{\uarrowr}{\mathrel{\rotatebox[origin=c]{60}{$\leftarrowtail$}}}
\newcommand{\darrowl}{\mathrel{\rotatebox[origin=c]{30}{$\twoheadleftarrow$}}}
\newcommand{\darrowr}{\mathrel{\rotatebox[origin=c]{120}{$\twoheadleftarrow$}}}
\newcommand{\vuarrow}{\mathrel{\rotatebox[origin=c]{-90}{$\leftarrowtail$}}}
\newcommand{\vdarrow}{\mathrel{\rotatebox[origin=c]{90}{$\twoheadleftarrow$}}}
\newcommand{\up}[2]{\langle{#2}\uarrowl{#1}\rangle}
\newcommand{\dn}[2]{\langle{#1}\darrowl{#2}\rangle}

\newcommand{\later}{\vartriangleright}
\newcommand{\type}{\texttt{Type}}
\newcommand{\lob}{\text{L\"{o}b}}
\newcommand{\tick}{\texttt{tick}}

\begin{document}

\title{Denotational Semantics for Gradual Typing in Synthetic Guarded Domain Theory}
\author{Eric Giovannini}
\author{Max S. New}

\begin{abstract}
    We develop a denotational semantics for a gradually typed language
    with effects that is adequate and proves the graduality theorem.
    %
    The denotational semantics is constructed using \emph{synthetic
    guarded domain theory} working in a type theory with a later
    modality and clock quantification.
    %
    This provides a remarkably simple presentation of the semantics,
    where gradual types are interpreted as ordinary types in our ambient
    type theory equipped with an ordinary preorder structure to model
    the error ordering.
    %
    This avoids the complexities of classical domain-theoretic models
    (New and Licata) or logical relations models using explicit
    step-indexing (New and Ahmed).
    %
    In particular, we avoid a major technical complexity of New and
    Ahmed that requires two logical relations to prove the graduality
    theorem.

    By working synthetically we can treat the domains in which gradual
    types are interpreted as if they were ordinary sets. This allows us
    to give a ``na\"ive'' presentation of gradual typing where each
    gradual type is modeled as a well-behaved subset of the universal
    domain used to model the dynamic type, and type precision is modeled
    as simply a subset relation.
    %
  \end{abstract}

  \maketitle

  \section{Introduction}

  % gradual typing, graduality
  \subsection{Gradual Typing and Graduality}
  Gradual typing allows a language to have both statically-typed and dynamically-typed terms;
  the statically-typed terms are type checked at compile time, while type checking for the 
  dynamically-typed terms is deferred to runtime.

  Gradually-typed languages support the smooth migration from dynamic typing
  to static typing, in that the programmer can initially leave off the
  typing annotations and provide them later without altering the meaning of the
  program. Moreover, a key feature of gradually typed languages is that the interaction
  between the static and dynamic components of the codebase should be safe -- i.e., should
  preserve the guarantees made by the static types.
  In other words, in the static portions of the codebase, type soundness must be preserved.

  In a gradually-typed language, the mixing of static and dynamic code is seamless, in that
  the dynamically typed parts are checked at runtime. This type checking occurs at the elimination
  forms of the language (e.g., pattern matching, field reference, etc.).
  Gradual languages are generally elaborated to a \emph{cast calculus}, in which the dynamic
  type checking is made explicit through the insertion of \emph{type casts}.

  ...

  % Up and down casts
  There is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that $A$ is a \emph{more
  precise} type than $B$.
  There a dynamic type $\dyn$ with the property that $A \ltdyn \dyn$ for all $A$.

  If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
  and a term $N$ of type $B$ may be \emph{down}casted to $A$, written $\dn A B N$.
  Upcasts always succeed, while downcasts may fail at runtime.

  % Graduality property
  A key property that any gradually-typed language should satisfy is the \emph{dynamic gradual guarantee},
  originally defined by Siek, Vitousek, Cimini, and Boyland \cite{TODO}.
  New et al refer to this property as \emph{graduality}, by analogy with parametricity.
  Intuitively, graduality says that in going from a dynamic to static style
  should not introduce changes in the meaning of the program.
  More specifically, making the types more precise (replacing $\dyn$ with a more specific type
  such as integers) will either result in the same behavior as the less precise program,
  or cause a dynamic type error.

  % Observational equivalence and approximation
  % TODO

  \subsection{Prior Approach}
  % Proving graduality via LR
  The usual approach to proving graduality uses the technique of \emph{logical relations}.
  Specifially, a logical relation is constructed and showed to be sound with respect to
  observational approximation. Because the dynamic type is modeled as a recursive type,
  the logical relation needs to be paramterized by natural numbers, a so-called
  step-indexed logical relation. Reasoning about step-indexed logical relations
  is tedious and error-prone, and there are several very subtle aspects that must
  be taken into account in the proofs.

  % Difficulties in prior semantics
  % TODO

  % - two separate logical relations for expressions
  % - transitivity

  % synthetic guarded domain theory, denotational semantics therein
  \subsection{Synthetic Guarded Domain Theory}
  One way to avoid the tedious reasoning associated with step-indexing is to work
  axiomatically inside of a logical system that can reason about non-well-founded recursive
  constructions while abstracting away the specific details of step-indexing required
  if we were working analytically.
  The system that proves useful fot this purpose is called \emph{synthetic guarded
  domain theory}, or SGDT for short. We provide a brief overview here, but more
  details can be found in \cite{TODO}.

  In SGDT, there is the notion of a \emph{clock}, an abstract type that tracks the
  passage of time through \emph{ticks}. Given a clock $k$, a tick $t : \tick k$ serves
  as evidence that one unit of time has passed according to the clock $k$.

  SGDT also introduces a modality $\later : \type \to \type$, (pronounced ``later'').
  Given a clock $k$ and type $T$, the type $\later_k T$ represents a value of type
  $T$ one unit of time in the future according to clock $k$.
  The type $\later T$ is represented as a function from ticks of a clock $k$ to $T$.
  There is an axiom called $\lob$-induction which is stated as follows:

  \[
    \lob : \forall T, (\later T \to T) \to T.
  \]

  That is, to construct a term of type $T$, it suffices to assume that we have access to
  such a term ``later'' and use that to help us build a term ``now''.

  In particular, this axiom applies to propositions $P : \texttt{Prop}$.

  \subsection{Overview of Remainder of Paper}
  In Section \ref{sec:GTLC}, we introduce the gradually-typed cast calculus
  for which we will prove graduality. Important here are the notions of syntactic
  type precision and term precision.

  In Section \ref{sec:domain-theory}, we define several fundamental constructions
  internal to SGDT that will be needed when we give a denotational semantics to
  the language. This includes the notion of Predomains as well as the concept
  of EP-Pairs.

  In Section \ref{Semantics}, we define the denotational semantics for the gradually-typed
  language

\section{GTLC}\label{sec:GTLC}

\subsection{Syntax}

\subsection{Type Precision}

\subsection{Term Precision}

\section{Domain-Theoretic Constructions}\label{sec:domain-theory}

\subsection{Predomains}

\subsection{Weak Bisimilarity Relation}

\subsection{EP-Pairs}

\section{Semantics}\label{sec:semantics}

\subsection{Types as Predomains}

\subsection{Terms as Monotone Functions}

\subsection{Type Precision as EP-Pairs}

\subsection{Term Precision}
% Homogeneous vs heterogeneous term precision

\section{Graduality}\label{sec:graduality}

\section{Discussion}\label{sec:discussion}

\subsection{Synthetic Ordering}

While the use of synthetic guarded domain theory allows us to very
conveniently work with non-well-founded recursive constructions while
abstracting away the precise details of step-indexing, we do work with
the error ordering in a mostly analytic fashion in that gradual types
are interpreted as sets equipped with an ordering relation, and all
terms must be proven to be monotone.
%
It is possible that a combination of synthetic guarded domain theory
with \emph{directed} type theory would allow for an a synthetic
treatment of the error ordering as well.

\end{document}

### 2023-02-14 notes/lift-monad-topos-of-trees.tex
notes on lift monad in topos of trees model

\documentclass{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{tikz-cd}
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{parskip}
\usepackage{tikz-cd}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{nonnum-theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{definition}{Definition}[section]
\newtheorem{lemma}{Lemma}[section]

\newcommand{\later}{\vartriangleright\hspace{-2px}}
\newcommand{\nxt}{\texttt{next}}

\newcommand{\calS}{\mathcal{S}}
\newcommand{\gfix}{\texttt{gfix}}

\newcommand{\calU}{\mathcal{U}}
\newcommand{\laterhat}{\widehat{\later}}
\newcommand{\El}{\mathsf{El}}

\newcommand{\lift}{L_\mho}
\newcommand{\lifthat}{\widehat{L_\mho}}

\newcommand{\To}{\Rightarrow}
\newcommand{\calC}{\mathcal{C}}
\newcommand{\Set}{\texttt{Set}}
\newcommand{\Yo}{\mathsf{Yo}}
\newcommand{\Hom}{\mathsf{Hom}}

\newcommand{\inl}{\texttt{inl}}
\newcommand{\inr}{\texttt{inr}}

\begin{document}

\title{The Lift Monad in the Topos of Trees}
\author{Eric Giovannini}

\maketitle

\section{Background}

Recall that the topos of trees $\calS$ is the presheaf category $\Set^{\omega^o}$. 
An object $X$ is a family $\{X_i\}$ of sets indexed by natural numbers, along with
restriction maps $r^X_i \colon X_{i+1} \to X_i$.

A morphism from $\{X_i\}$ to $\{Y_i\}$ is a family of functions $f_i \colon X_i \to Y_i$
that commute with the restriction maps in the obvious way, that is,
$f_i \circ r^X_i = r^Y_i \circ f_{i+1}$.

There is an operator $\later$ (``later''), defined on an object $X$ by
$(\later X)_0 = 1$ and $(\later X)_{i+1} = X_i$. The restriction maps are given by
$r^\later_0 = !$, where $!$ is the unique map into $1$, and $r^\later_{i+1} = r^X_i$.

For each $X$, there is a morphism $\nxt^X \colon X \to \later X$ defined pointwise by
$\nxt^X_0 = !$, and $\nxt^X_{i+1} = r^X_i$.

% Note that $\calS$ is a topos, so it is cartesian closed and has a subobject classifier $\Omega$.

We also have a universe $\calU$ of types, with encodings for operations such as sum types
(written as $\widehat{+}$). There is also an operator $\laterhat \colon \later \calU \to \calU$ such that
$\El(\laterhat(\nxt A)) =\,\, \later \El(A)$, where $\El$ is the type corresponding to the code $A$.

\section{Guarded Fixpoints}

Given a morphism $f \colon \later X \to X$, we want to compute its fixed point $\gfix f : X$.
We define $\gfix \colon (\later X \To X) \to X$ as a morphism in $\calS$, where $(\later X \To X)$
is the exponential object, the object of functions from $\later X$ to $X$.

We first describe $(\later X \To X)$. Since $\calS$ is a presheaf category, we can use the
Yoneda Lemma to describe the exponential objects, as reviewed below.

In general, let $X$ and $Y$ be object in the functor category $\Set^{\calC^o}$, i.e., functors
$X$ and $Y$ from $\calC^o$ to $\Set$.
We have by the Yoneda Lemma that for each object $a$ of $\calC$, the set
$(X \To Y)(a)$ is isomorphic to the set of natural transformations $\Yo(a) \to (X \To Y)$,
i.e., the set of morphisms $\Yo(a) \to (X \To Y)$ in the functor category.
But by the universal property of the exponential, such a morphism is the same
as a morphism (in this case, a natural transformation) $\Yo(a) \times X \to Y$.

So, the set $(X \To Y)(a)$ is isomorphic to the set of natural transformations
$\Yo(a) \times X \to Y$, that is, $(\Hom(-,a) \times X) \to Y$.

% Naturality condition

% Action of the exponential functor on morphisms

With the above, we can compute the definition of $(\later X \To X)$ in $\calS$.
Since $\calC = \omega$ is a preorder, the above result specializes to

\[
    (\later X \To X)_i = \{ f^i \colon \forall j \le i. (\later X)_j \to X_j \}.
\]

(The superscript is included to remind us of which set each function belongs to.)

The naturality condition says that, for $j' \le j$ the following diagram commutes:

% https://q.uiver.app/?q=WzAsNCxbMCwwLCIoXFxsYXRlciBYKV9qIl0sWzIsMCwiWF9qIl0sWzIsMiwiWF97aid9Il0sWzAsMiwiKFxcbGF0ZXIgWClfe2onfSJdLFswLDMsInJfe2ogXFxsZSBqJ31ee1xcbGF0ZXIgWH0iLDJdLFsxLDIsInJfe2ogXFxsZSBqJ31ee1h9Il0sWzMsMiwiZl5pIFxcLCBqJyIsMl0sWzAsMSwiZl5pXFwsIGoiXV0=
\[\begin{tikzcd}
	{(\later X)_j} && {X_j} \\
	\\
	{(\later X)_{j'}} && {X_{j'}}
	\arrow["{r_{j \le j'}^{\later X}}"', from=1-1, to=3-1]
	\arrow["{r_{j \le j'}^{X}}", from=1-3, to=3-3]
	\arrow["{f^i \, j'}"', from=3-1, to=3-3]
	\arrow["{f^i\, j}", from=1-1, to=1-3]
\end{tikzcd}\]

%From now on, we will leave the argument $j$ implicit when applying
%the functions $f^i$ defined above.

Additionally, the restriction maps 
$r_i^{(\later X \To X)} \colon (\later X \To X)_{i+1} \to (\later X \To X)_{i}$
are defined in the obvious way.
Namely, if $f : \forall j \le i+1 . (\later X)_j \to X_j$ then in particular
$f : \forall j \le i . (\later X)_j \to X_j$, so we define

\[
    r_i^{(\later X \To X)} (f) := f.
\]

Note that this is just a special case of the action of the functor $X \To Y$
discussed above on morphisms.

\vspace{4ex}

Now we can proceed to define $\gfix$. We define $\gfix_i \colon (\later X \To X)_i \to X_i$
by induction on $i$. We let

\[
    \gfix_0 (f) = f\, 0\, 1,
\]

since here $f\, 0 \colon (\later X)_0 \to X_0$ and we have $(\later X)_0 = 1$.
For $i \ge 1$, we define

\[
    \gfix_i (f) = f\, i\, (\gfix_{i-1} (f\, (i-1))),
\]

which has type $X_i$ as follows: since $f\, (i-1) \colon ((\later X)_{i-1} \to X_{i-1})$,
then $\gfix_{i-1}(f\, (i-1))$ has type $X_{i-1}$, which is equal to $(\later X)_i$,
and so it is a valid input to $f\, i$.

This completes the definition of $\gfix_i$, and hence of $\gfix$.
We need to verify that $\gfix f = f (\nxt (\gfix f))$, where $f \colon \later X \To X$.
This follows from the definition of $\gfix$ above.
% Details?

\section{The Lift Monad}

For an object $X$, the lift monad $\lift X$ is defined as follows:

\begin{align*}
    &\lift X := \\
    &\quad\quad    \eta \colon X \to \lift X \\
    &\quad\quad    \mho \colon \lift X \\
    &\quad\quad    \theta \colon \later (\lift X) \to \lift X
\end{align*}

In the setting of guarded domain theory, the above translates to
the following guarded fixpoint:

\[
    \lifthat X = \gfix(\lambda Y \colon \later \calU . 
      X \,\, \widehat{+} \,\, 1 \,\, \widehat{+} \,\, \widehat{\later} Y)
\]

With this definition, we can take $\lift X = \El(\lifthat X)$,
and we can show that $\lift X = X + 1 + \later (\lift X)$.
This follows from the fact that $\gfix f = f (\nxt (\gfix f))$ and
the rule $\El(\laterhat(\nxt A)) =\,\, \later \El(A)$.

% TODO be explicit about how this follows from the
% description of gfix we computed above
We can compute $(\lift X)_i$ for each $i$, as follows. Intuitively, we can think of
having a potentially-erroring computation that returns values of type $X$,
and $(\lift X)_i$ represents our view of the behavior of the computation
provided we can watch it for $i+1$ steps.

We have

\[
    (\lift X)_0 = X_0 + 1 + (\later \lift X)_0 = X_0 + 1 + 1.
\]

This represents the behavior of the computation after one step.
The first $1$ represents the case where an error occurred, while
the second $1$ represents the case where the computation took more
than one step, and hence we have run out of time to observe it.

For $i \ge 1$, we have

\begin{align*}
    (\lift X)_i &= X_i + 1 + (\later \lift X)_i \\
                &= X_i + 1 + (\lift X)_{i-1}.
\end{align*}

In other words, when observing a computation for $i+1$ steps,
either (1) we get a value in the first step, or (2) we
error in the first step, or (3) the answer is not ready yet
and we must wait, in which case we will have $i$ steps left to watch
the computation.

For concreteness, when $i = 1$, we have

\[
    (\lift X)_1 = X_1 + 1 + (X_0 + 1 + 1).
\]

Here, the first $1$ represents an error occurring in the first step of the
computation, while the second $1$ represents an error in the second step,
and the third $1$ represents having run out of steps to observe the computation.

\vspace{4ex}

We now discuss each of the constructors of $\lift X$.
First consider the constructor $\eta^X \colon X \To \lift X$.
This is defined component-wise as $\eta^X_i = \lambda x . \inl (\inl x)$
(we assume sums are left-associative).
The constructor $\mho^X \colon 1 \To \lift X$ is defined by
$\mho^X_i = \lambda () . \inl (\inr \, ())$.
Lastly, the constructor $\theta^X \colon (\later (\lift X) \To \lift X)$
is defined by $\theta^X_i = \lambda \tilde{l} . \inr \, \tilde{l}$.

The restriction maps $r^{\lift X}_i \colon (\lift X)_{i+1} \to (\lift X)_i$
are defined by ``truncating'' the observation of the computation from $i+2$
steps to $i+1$ steps, as follows.
If the computation has the form $\eta^X_{i+1}(x)$, then we return $\eta^X_{i}(r^X_i(x))$.
If the computation has the form $\mho^X_{i+1}$, then we return $\mho^X_i$.
Lastly, if the computation has the form $\theta^X_{i+1} (\tilde{l})$, then
we return $\theta^X_0 ()$ if $i = 0$ and 
$(\theta^X_i (r^{\lift X}_{i-1} (\tilde{l})))$ if $i \ge 1$.
% Is this correct?

\subsection{The Diverging Computation}

Consider the ``non-terminating'' computation
$\gfix (\theta) \colon \lift X$. This can be described pointwise
as follows:

\[ (\gfix (\theta))_0 = \inr \, () \colon (\lift X)_0, \]

and for $i \ge 1$,

\[ (\gfix (\theta))_i = \inr ((\gfix (\theta))_{i-1}), \]

where the latter type-checks because $(\gfix (\theta))_{i-1} \colon (\lift X)_{i-1}$.

Unfolding the above, we have

\[
    (\gfix (\theta))_i = \inr (\inr ( \dots ( \inr \, () ) \dots )),
\]

where there are $i+1$ nested occurrences of $\inr$.

So, no matter how many steps we observe this computation for,
it will never return a result within that number of steps.
It will instead always say that it needs more time.

\section{Weak Bisimilarity}

We now define a weak bisimilarity relation ``$\approx$'' on $\lift X$.
This relation is parameterized by an equivalence relation
$\sim$ on $X$. Intuitively, two computations $l$ and $l'$
if they are equivalent ``up to $\theta$''. Specifically,
their underlying results are related by $\sim$ and they only
differ in the number of steps taken to deliver their results.

In the below, we define $\delta \colon \lift X \to \lift X$ to be
$\theta \circ \nxt$.

We define $\approx$ via guarded fixpoint as follows:

\begin{align*}
    &\approx \,\, = \gfix (\lambda \,\, rec \,\, l \,\, l'. \\
    &\quad\quad \text{case } l, l' \text{ of} \\
    &\quad\quad\quad | \quad \mho , \mho     \Rightarrow Unit \\
    &\quad\quad\quad | \quad \eta (x) , \eta (y) \Rightarrow x \sim y \\
    &\quad\quad\quad | \quad \theta (\tilde{x}) , \theta (\tilde{y}) \Rightarrow 
      \laterhat (\lambda t . (rec \,\, t) (\tilde{x} \, t) (\tilde{y} \, t)) \\
    &\quad\quad\quad | \quad \theta (\tilde{x}) , \mho    \Rightarrow 
      \Sigma_{n : \mathbb{N}} . \theta (\tilde{x}) = \delta^n (\mho) \\
    &\quad\quad\quad | \quad \theta (\tilde{x}) , \eta(y)  \Rightarrow 
      \Sigma_{n : \mathbb{N}} . \Sigma_{x : X} . 
        \theta (\tilde{x}) = \delta^n(\eta (x)) \times (x \sim y)\\
    &\quad\quad\quad | \quad \mho , \theta (\tilde{y}) \Rightarrow
      \Sigma_{n : \mathbb{N}} . \theta (\tilde{y}) = \delta^n (\mho) \\
    &\quad\quad\quad | \quad \eta (x) , \theta (\tilde{y}) \Rightarrow
      \Sigma_{n : \mathbb{N}} . \Sigma_{y : X} . 
        \theta (\tilde{y}) = \delta^n(\eta (y)) \times (x \sim y)\\
    &\quad\quad\quad | \quad \text{otherwise} \Rightarrow \bot
    )
\end{align*}

Note that in the case where both variables are $\theta$'s, we have
used tick abstraction so that we can apply the guarded recursive
function $rec$.

\vspace{4ex}

We now consider the claim that for all $l : \lift X$,

\[
    l \approx (\gfix (\theta)).
\]

We claim that this does not hold. Specifically, we claim that

\[
    \mho \not\approx (\gfix (\theta)).
\]

In order for these to be related, according to the defintion of the relation,
(and recalling that $(\gfix (\theta)) = \theta (\nxt (\gfix (\theta)))$),
we would need that $(\gfix (\theta)) = \delta^n (\mho)$ for some $n$.

Suppose this were the case for some natural number $n_0$.
Consider an index $i > n_0 \in \omega$. The left-hand side above is equal to

\[ \inr (\inr ( \dots ( \inr \, () ) \dots )), \]

(where there are $i+1$ occurrences of $\inr$), while the right-hand side will have
as its innermost term an $\inl (\inr \, ())$, representing the error case $\mho$.
Thus, the two terms cannot be equal for any $n$. It follows that

\[
    \mho \not\approx (\gfix (\theta)).
\]

\end{document}

### 2023-03-30 paper-new/paper.tex
work on paper

@@
 \usepackage{tikz-cd}
+\usepackage{enumitem}
+\usepackage{wrapfig}
+\usepackage{fancyvrb}
+\usepackage{comment}
+
+\usepackage{ stmaryrd }
 
@@
 
-\newcommand{\dyn}{?}
-\newcommand{\ltdyn}{\sqsubseteq}
-
-\newcommand{\uarrowl}{\mathrel{\rotatebox[origin=c]{-30}{$\leftarrowtail$}}}
-\newcommand{\uarrowr}{\mathrel{\rotatebox[origin=c]{60}{$\leftarrowtail$}}}
-\newcommand{\darrowl}{\mathrel{\rotatebox[origin=c]{30}{$\twoheadleftarrow$}}}
-\newcommand{\darrowr}{\mathrel{\rotatebox[origin=c]{120}{$\twoheadleftarrow$}}}
-\newcommand{\vuarrow}{\mathrel{\rotatebox[origin=c]{-90}{$\leftarrowtail$}}}
-\newcommand{\vdarrow}{\mathrel{\rotatebox[origin=c]{90}{$\twoheadleftarrow$}}}
-\newcommand{\up}[2]{\langle{#2}\uarrowl{#1}\rangle}
-\newcommand{\dn}[2]{\langle{#1}\darrowl{#2}\rangle}
-
-\newcommand{\later}{\vartriangleright}
-\newcommand{\type}{\texttt{Type}}
-\newcommand{\lob}{\text{L\"{o}b}}
-\newcommand{\tick}{\texttt{tick}}
+\input{defs}
 
@@
 
-  Gradually-typed languages support the smooth migration from dynamic typing
-  to static typing, in that the programmer can initially leave off the
-  typing annotations and provide them later without altering the meaning of the
-  program. Moreover, a key feature of gradually typed languages is that the interaction
+  Gradually-typed languages should satisfy two intuitive properties.
+  First, the interaction
   between the static and dynamic components of the codebase should be safe -- i.e., should
@@
   In other words, in the static portions of the codebase, type soundness must be preserved.
+  Second, gradual langugaes should support the smooth migration from dynamic typing
+  to static typing, in that the programmer can initially leave off the
+  typing annotations and provide them later without altering the meaning of the
+  program.
 
+  % Graduality property
+  Formally speaking, gradually typed languages should satisfy the 
+  \emph{dynamic gradual guarantee}, originally defined by Siek, Vitousek, Cimini,
+  and Boyland \cite{siek_et_al:LIPIcs:2015:5031}.
+  This property is also referred to as \emph{graduality}, by analogy with parametricity.
+  Intuitively, graduality says that in going from a dynamic to static style should not
+  introduce changes in the meaning of the program.
+  More specifically, making the types more precise by adding annotations
+  % (replacing $\dyn$ with a more specific type
+  %such as integers)
+  will either result in the same behavior as the less precise program,
+  or result in a type error.
+
+  \subsection{Current Approaches}
+  Current approaches to proving graduality include the methods of Abstracting Gradual Typing
+  \cite{garcia-clark-tanter2016} and the formal tools of the Gradualier \cite{cimini-siek2016}.
+  These allow the language developer to start with a statically typed langauge and derive a
+  gradually typed language that satisfies the gradual guarantee. The downside is that
+  not all gradually typed languages can be derived from these frameworks, and moreover, in both
+  approaches the semantics is derived from the static type system as opposed to the alternative
+  in which the semantics determines the type checking. Without a clear semantic interpretation of type
+  dynamism, it becomes difficult to extend these techniques to new language features such as polymorphism.
+
+  % Proving graduality via LR
+  New and Ahmed have developed a semantic approach to specifying type dynamism in term of
+  embedding-projection pairs, which allows for a particularly elegant formulation of the
+  gradual guarantee.
+  Moreover, this approach allows for type-based reasoning using $\eta$-equivalences
+
+  The downside of the above approach is that each new language requires a different logical relation
+  to prove graduality. As a result, many developments using this approach require vast effort,
+  with many such papers having 50+ pages of proofs.
+  Our aim here is that by mechanizing a graduality proof in a reusable way,
+  we will provide a framework for other language designers to use to ensure that their languages satsify graduality.
+
+  One approach currently used to prove graduality uses the technique of \emph{logical relations}.
+  Specifially, a logical relation is constructed and shown to be sound with respect to
+  observational approximation. Because the dynamic type is modeled as a non-well-founded
+  recursive type, the logical relation needs to be paramterized by natural numbers to restore well-foundedness.
+  This technique is known as a \emph{step-indexed logical relation} \cite{TODO}.
+  Reasoning about step-indexed logical relations
+  can be tedious and error-prone, and there are some very subtle aspects that must
+  be taken into account in the proofs. Figure \ref{TODO} shows an example of a step-indexed logical
+  relation for the gradually-typed lambda calculus.
+
+  % TODO move to another section?
+  % Difficulties in prior semantics
+  % - two separate logical relations for expressions
+  % - transitivity
+  In particular, the prior approach of New and Ahmed requires two separate logical
+  relations for terms, one in which the steps of the left-hand term are counted,
+  and another in which the steps of the right-hand term are counted \cite{TODO}.
+  Then two terms $M$ and $N$ are related in the ``combined'' logical relation if they are
+  related in both of the one-sided logical relations. Having two separate logical relations
+  complicates the statement of the lemmas used to prove graduality, becasue any statement that
+  involves a term stepping needs to take into account whether we are counting steps on the left
+  or the right. Some of the differences can be abstracted over, but difficulties arise for properties %/results
+  as fundamental and seemingly straightforward as transitivty.
+
+  Specifically, for transitivity, we would like to say that if $M$ is related to $N$ at
+  index $i$ and $N$ is related to $P$ at index $i$, then $M$ is related to $P$ at $i$.
+  But this does not actually hold: we requrie that one of the two pairs of terms
+  be related ``at infinity'', i.e., that they are related at $i$ for all $i \in \mathbb{N}$.
+  Which pair is required to satisfy this depends on which logical relation we are considering,
+  (i.e., is it counting steps on the left or on the right),
+  and so any argument that uses transitivity needs to consider two cases, one
+  where $M$ and $N$ must be shown to be related for all $i$, and another where $N$ and $P$ must
+  be related for all $i$. This may not even be possible to show in some scenarios!
+
+  These complications introduced by step-indexing lead one to wonder whether there is a
+  way of proving graduality without relying on tedious arguments involving natural numbers.
+  An alternative approach, which we investigate in this paper, is provided by
+  \emph{synthetic guarded domain theory}, as discussed below.
+  Synthetic guarded domain theory allows the resulting logical relation to look almost
+  identical to a typical, non-step-indexed logical relation.
+
+  \subsection{Contributions}
+  Our main contribution is a framework in Guarded Cubical Agda for proving graduality of a cast calculus.
+  To demonstrate the feasability and utility of our approach, we have used the framework to prove
+  graduality for the simply-typed gradual lambda calculus. Along the way, we have developed an intensional theory
+  of graduality that is of independent interest.
+
+  \subsection{Proving Graduality in SGDT}
+  % TODO move to technical background
+
+  % Cast calculi
   In a gradually-typed language, the mixing of static and dynamic code is seamless, in that
@@
 
-  ...
-
   % Up and down casts
-  There is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that $A$ is a \emph{more
-  precise} type than $B$.
+  In a cast calculus, there is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that
+  $A$ is a \emph{more precise} type than $B$.
   There a dynamic type $\dyn$ with the property that $A \ltdyn \dyn$ for all $A$.
-
+  %
   If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
@@
   Upcasts always succeed, while downcasts may fail at runtime.
+  %
+  % Syntactic term precision
+  We also have a notion of \emph{syntatcic term precision}.
+  If $A \ltdyn B$, and $M$ and $N$ are terms of type $A$ and $B$ respectively, we write
+  $M \ltdyn N : A \ltdyn B$ to mean that $M$ is more precise than $N$, i.e., $M$ and $N$
+  behave the same except that $M$ may error more.
 
-  % Graduality property
-  A key property that any gradually-typed language should satisfy is the \emph{dynamic gradual guarantee},
-  originally defined by Siek, Vitousek, Cimini, and Boyland \cite{TODO}.
-  New et al refer to this property as \emph{graduality}, by analogy with parametricity.
-  Intuitively, graduality says that in going from a dynamic to static style
-  should not introduce changes in the meaning of the program.
-  More specifically, making the types more precise (replacing $\dyn$ with a more specific type
-  such as integers) will either result in the same behavior as the less precise program,
-  or cause a dynamic type error.
+  % Modeling the dynamic type as a recursive sum type?
+  % Observational equivalence and approximation?
 
-  % Observational equivalence and approximation
-  % TODO
+  In this paper, we will be using SGDT techniques to prove graduality for a particularly
+  simple gradually-typed cast calculus, the gradually-typed lambda calculus. This is just
+  the usual simply-typed lambda calculus with a dynamic type $\dyn$ such that $A \ltdyn\, \dyn$ for all types $A$,
+  as well as upcasts and downcasts between any types $A$ and $B$ such that $A \ltdyn B$.
+  The complete definition will be provided in Section \ref{sec:GTLC}.
 
-  \subsection{Prior Approach}
-  % Proving graduality via LR
-  The usual approach to proving graduality uses the technique of \emph{logical relations}.
-  Specifially, a logical relation is constructed and showed to be sound with respect to
-  observational approximation. Because the dynamic type is modeled as a recursive type,
-  the logical relation needs to be paramterized by natural numbers, a so-called
-  step-indexed logical relation. Reasoning about step-indexed logical relations
-  is tedious and error-prone, and there are several very subtle aspects that must
-  be taken into account in the proofs.
+  Our main theorem is the following:
 
-  % Difficulties in prior semantics
-  % TODO
+  % \begin{theorem}[Graduality]
+  %   If $M \ltdyn N : \nat$, then either:
+  %   \begin{enumerate}
+  %     \item $M = \mho$
+  %     \item $M = N = \zro$
+  %     \item $M = N = \suc n$ for some $n$
+  %     \item $M$ and $N$ diverge
+  %   \end{enumerate}
+  % \end{theorem}
 
-  % - two separate logical relations for expressions
-  % - transitivity
+  \begin{theorem}[Graduality]
+    If $\cdot \vdash M \ltdyn N : \nat$, then
+    \begin{enumerate}
+      \item If $N = \mho$, then $M = \mho$
+      \item If $N = `n$, then $M = \mho$ or $M = `n$
+      \item If $M = V$, then $N = V$
+    \end{enumerate}
+  \end{theorem}
+
+  We also should be able to show that $\mho$, $\zro$, and $\suc\, N$ are not equal.
+
+  Our first step toward proving graduality is to formulate an \emph{intensional} gradual lambda calculus,
+  which we call $\intlc$, in which the computation steps taken by a term are made explicit.
+  The ``normal'' graudal lambda calculus for which we want to prove graduality will be called the
+  \emph{extensional} gradual lambda calculus, denoted $\extlc$. We will define an erasure function
+  $\erase{\cdot} : \intlc \to \extlc$ which takes a program in the intensional lambda calculus
+  and ``forgets'' the syntactic information about the steps to produce a term in the extensional calculus.
+
+  Every term $M_e$ in $\extlc$ will have a corresponding program $M_i$ in $\intlc$ such that
+  $\erase{M_i} = M_e$. Moreover, we will show that if $M_e \ltdyn_e M_e'$ in the extensional theory,
+  then there exists terms $M_i$ and $M_i'$ such that $\erase{M_i}=M_e$, $\erase{M_i'}=M_e'$ and
+  $M_i \ltdyn_i M_i'$ in the intensional theory.
+
+  We formulate and prove an analogous graduality theorem for the intensional lambda calculus.
+  We define an interpretation of the intensional lambda calculus into a model in which we prove
+  various results. Using the observation above, given $M_e \ltdyn M_e' : \nat$, we can find
+  intensional programs $M_i$ and $M_i'$ that erase to them and are such that $M_i \ltdyn M_i'$.
+  We will then apply the intensional graduality theorem to $M_i$ and $M_i'$, and translate the result
+  back to $M_e$ and $M_e'$.
+
+  \subsection{Overview of Remainder of Paper}
+
+  In Section \ref{sec:technical-background}, we provide technical background on gradually typed languages and
+  on synthetic guarded domain theory.
+ % 
+  In Section \ref{sec:GTLC}, we introduce the gradually-typed cast calculus
+  for which we will prove graduality. Important here are the notions of syntactic
+  type precision and term precision. We introduce both the extensional gradual lambda calculus
+  ($\extlc$) and the intensional gradual lambda calculus ($\intlc$).
+%
+  In Section \ref{sec:domain-theory}, we define several fundamental constructions
+  internal to SGDT that will be needed when we give a denotational semantics to
+  our intensional lambda calculus.
+  This includes the notion of Predomains as well as the concept
+  of EP-Pairs.
+%
+  In Section \ref{sec:semantics}, we define the denotational semantics for the
+  intensional gradually-typed lambda calculus using the domain theoretic constructions in the
+  previous section.
+%
+  In Section \ref{sec:graduality}, we outline in more detail the proof of graduality for the
+  extensional gradual lambda calculus.
+%
+  In Section \ref{sec:discussion}, we discuss the benefits and drawbacks to our approach in comparison
+  to the traditional step-indexing approach, as well as possibilities for future work.
+
+\section{Technical Background}\label{sec:technical-background}
 
@@
 if we were working analytically.
-  The system that proves useful fot this purpose is called \emph{synthetic guarded
+The system that proves useful for this purpose is called \emph{synthetic guarded
 domain theory}, or SGDT for short. We provide a brief overview here, but more
@@
 
-  In SGDT, there is the notion of a \emph{clock}, an abstract type that tracks the
-  passage of time through \emph{ticks}. Given a clock $k$, a tick $t : \tick k$ serves
-  as evidence that one unit of time has passed according to the clock $k$.
+SGDT offers a synthetic approach to domain theory that allows for guarded recursion
+to be expressed syntactically via a type constructor $\later : \type \to \type$ 
+(pronounced ``later''). The use of a modality to express guarded recursion
+was introduced by Nakano \cite{TODO}.
 
-  SGDT also introduces a modality $\later : \type \to \type$, (pronounced ``later'').
-  Given a clock $k$ and type $T$, the type $\later_k T$ represents a value of type
-  $T$ one unit of time in the future according to clock $k$.
-  The type $\later T$ is represented as a function from ticks of a clock $k$ to $T$.
-  There is an axiom called $\lob$-induction which is stated as follows:
+Given a type $A$, the type $\later A$ represents an element of type $A$
+that is available one time step later. There is an operator $\nxt : A \to\, \later A$
+that ``delays'' an element available now to make it available later.
+We will use a tilde to denote a term of type $\later A$, e.g., $\tilde{M}$.
+
+% TODO later is an applicative functor, but not a monad
+
+There is a fixpoint operator
 
 \[
-    \lob : \forall T, (\later T \to T) \to T.
+  \fix : \forall T, (\later T \to T) \to T.
 \]
@@
 such a term ``later'' and use that to help us build a term ``now''.
+This operator satisfies the axiom that $\fix f = f (\nxt (\fix f))$.
+In particular, this axiom applies to propositions $P : \texttt{Prop}$; proving
+a statement in this manner is known as $\lob$-induction.
 
-  In particular, this axiom applies to propositions $P : \texttt{Prop}$.
+In SGDT, there is also a new sort called \emph{clocks}. A clock serves as a reference
+relative to which the constructions described above are carried out.
+For instance, given a clock $k$ and type $T$, the type $\later^k T$ represents a value of type
+$T$ one unit of time in the future according to clock $k$.
+If we only ever had one clock, then we would not need to bother defining this notion.
+However, the notion of \emph{clock quantification} is crucial for encoding coinductive types using guarded
+recursion, an idea first introduced by Atkey and McBride \cite{TODO}.
 
-  \subsection{Overview of Remainder of Paper}
-  In Section \ref{sec:GTLC}, we introduce the gradually-typed cast calculus
-  for which we will prove graduality. Important here are the notions of syntactic
-  type precision and term precision.
+% Clocked Cubical Type Theory
+\subsubsection{Ticked Cubical Type Theory}
+% TODO motivation for Clocked Cubical Type Theory, e.g., delayed substitutions
 
-  In Section \ref{sec:domain-theory}, we define several fundamental constructions
-  internal to SGDT that will be needed when we give a denotational semantics to
-  the language. This includes the notion of Predomains as well as the concept
-  of EP-Pairs.
+In Ticked Cubical Type Theory \cite{TODO}, there is an additional sort
+called \emph{ticks}. Given a clock $k$, a tick $t : \tick k$ serves
+as evidence that one unit of time has passed according to the clock $k$.
+The type $\later A$ is represented as a function from ticks of a clock $k$ to $A$.
+The type $A$ is allowed to depend on $t$, in which case we write $\later^k_t A$
+to emphasize the dependence.
 
-  In Section \ref{Semantics}, we define the denotational semantics for the gradually-typed
-  language
+% TODO next as a function that ignores its input tick argument
+
+The rules for tick abstraction and application are similar to those of dependent
+$\Pi$ types. 
+In particular, if we have a term $M$ of type $\later^k A$, and we
+have available in the context a tick $t' : \tick k$, then we can apply the tick to $M$ to get
+a term $M[t'] : A[t'/t]$. We will also write tick application as $M_t$.
+Conversely, if in a context $\Gamma, t : \tick k$ we have that $M$ has type $A$,
+then in context $\Gamma$ we have $\lambda t.M$ has type $\later A$. % TODO dependent version?
+
+The statements in this paper have been formalized in a variant of Agda called
+Guarded Cubical Agda \cite{TODO}, an implementation of Clocked Cubical Type Theory.
+
+% TODO axioms (clock irrelevance, tick irrelevance)
+
+% TODO topos of trees model
+\subsection{The Topos of Trees Model}
+
+The topos of trees $\calS$ is the presheaf category $\Set^{\omega^o}$.
+
+  We assume a universe $\calU$ of types, with encodings for operations such
+  as sum types (written as $\widehat{+}$). There is also an operator 
+  $\laterhat \colon \later \calU \to \calU$ such that 
+  $\El(\laterhat(\nxt A)) =\,\, \later \El(A)$, where $\El$ is the type corresponding
+  to the code $A$.
+
+  An object $X$ is a family $\{X_i\}$ of sets indexed by natural numbers,
+  along with restriction maps $r^X_i \colon X_{i+1} \to X_i$.
+
+  A morphism from $\{X_i\}$ to $\{Y_i\}$ is a family of functions $f_i \colon X_i \to Y_i$
+  that commute with the restriction maps in the obvious way, that is,
+  $f_i \circ r^X_i = r^Y_i \circ f_{i+1}$.
+
+  The type operator $\later$ is defined on an object $X$ by
+  $(\later X)_0 = 1$ and $(\later X)_{i+1} = X_i$.
+  The restriction maps are given by $r^\later_0 =\, !$, where $!$ is the
+  unique map into $1$, and $r^\later_{i+1} = r^X_i$.
+  The morphism $\nxt^X \colon X \to \later X$ is defined pointwise by
+  $\nxt^X_0 =\, !$, and $\nxt^X_{i+1} = r^X_i$.
+
+  Given a morphism $f \colon \later X \to X$, we define $\fix f$ pointwise
+  as $\fix_i(f) = f_{i} \circ \dots \circ f_0$.
+
+  % TODO global elements
 
@@
 
+Here we describe the syntax and typing for the graudally-typed lambda calculus.
+We also give the rules for syntactic type and term precision.
+
+We start with the extensional lambda calculus $\extlc$, and then describe the additions
+necessary for the intensional lambda calculus $\intlc$.
+
 \subsection{Syntax}
+
+\begin{align*}
+  &\text{Types } A, B := \nat, \, \dyn, \, (A \To B) \\
+  &\text{Terms } M, N := \err_A, \, \zro, \,  \suc\, M, \, (\lda{x}{M}), \, (M\, N), \, (\up{A}{B} M), \, (\dn{A}{B} M) \\
+  &\text{Contexts } \Gamma := \cdot, \, (\Gamma, x : A)
+\end{align*}
+
+The typing rules are as expected, with a cast between $A$ to $B$ allowed only when $A \ltdyn B$.
+
+\begin{mathpar}
+  \inferrule*{ }{\hasty \Gamma {\err_A} A}
+
+  \inferrule*{ }{\hasty \Gamma \zro \nat}
+
+  \inferrule {\hasty \Gamma M \nat} {\hasty \Gamma {\suc\, M} \nat}
+
+  \inferrule* 
+    {\hasty {\Gamma, x : A} M B} 
+    {\hasty \Gamma {\lda x M} {A \To B}}
+
+  \inferrule *
+    {\hasty \Gamma M {A \To B} \and \hasty \Gamma N A}
+    {\hasty \Gamma {M \, N} B}
+
+  \inferrule*
+    {A \ltdyn B \and \hasty \Gamma M A}
+    {\hasty \Gamma {\up A B M} B }
+
+  \inferrule*
+    {A \ltdyn B \and \hasty \Gamma M B}
+    {\hasty \Gamma {\dn A B M} A}
+\end{mathpar}
+
+The equational theory is also as expected, with $\beta$ and $\eta$ laws for function type.
 
@@
 
+The rules for type precision are as follows:
+
+\begin{mathpar}
+  \inferrule*[right = \dyn]
+    { }{\dyn \ltdyn \dyn}
+
+  \inferrule*[right = \nat]
+    { }{\nat \ltdyn \nat}
+
+  \inferrule*[right = $Inj_\nat$]
+    { }{\nat \ltdyn \dyn}
+
+  \inferrule*[right = $\To$]
+    {A_i \ltdyn B_i \and A_o \ltdyn B_o }
+    {(A_i \To A_o) \ltdyn (B_i \To B_o)}
+
+  \inferrule*[right=$Inj_{\To}$]
+    {(A_i \to A_o) \ltdyn (\dyn \To \dyn)}
+    {(A_i \to A_o) \ltdyn\, \dyn}
+
+\end{mathpar}
+
+% Type precision derivations
+Note that as a consequence of this presentation of the type precision rules, we
+have that if $A \ltdyn B$, there is a unique precision derivation that witnesses this.
+As in previous work, we go a step farther and make these derivations first-class objects,
+known as \emph{type precision derivations}.
+Specifically, for every $A \ltdyn B$, we have a derivation $d : A \ltdyn B$ that is constructed
+using the rules above. For instance, there is a derivation $\dyn : \dyn \ltdyn \dyn$, and a derivation
+$\nat : \nat \ltdyn \nat$, and if $d_i : A_i \ltdyn B_i$ and $d_o : A_o \ltdyn B_o$, then
+there is a derivation $d_i \To d_o : (A_i \To A_o) \ltdyn (B_i \To B_o)$. Likewise for
+the remaining two rules. The benefit to making these derivations explicit in the syntax is that we
+can perform induction over them.
+Note also that for any type $A$, we use $A$ to denote the reflexivity derivation that $A \ltdyn A$,
+i.e., $A : A \ltdyn A$.
+Finally, observe that for type precision derivations $d : A \ltdyn B$ and $d' : B \ltdyn C$, we
+can define their composition $d' \circ d : A \ltdyn C$. This will be used in the statement
+of transitivity of the term precision relation.
+
 \subsection{Term Precision}
+
+We allow for a \emph{heterogeneous} term precision judgment on terms $M$ of type $A$ and $N$ of type $B$,
+provided that $A \ltdyn B$ holds.
+
+% Type precision contexts
+In order to deal with open terms, we will need the notion of a type precision \emph{context}, which we denote
+$\gamlt$. This is similar to a normal context but instead of mapping variables to types,
+it maps variables $x$ to related types $A \ltdyn B$, where $x$ has type $A$ in the left-hand term
+and $B$ in the right-hand term. We may also write $x : d$ where $d : A \ltdyn B$ to indicate this.
+Another way of thinking of type precision contexts is as a zipped pair of contexts $\Gamma^l, \Gamma^r$
+with the same domain such that $\Gamma^l(x) \ltdyn \Gamma^r(x)$ for each $x$ in the domain.
+
+The rules for term precision come in two forms. We first have the \emph{congruence} rules,
+one for each term constructor. These assert that the term constructors respect term precision.
+The congruence rules are as follows:
+
+\begin{mathpar}
+  \inferrule*[right = $\err$]
+    { \hasty \Gamma M A }
+    {\etmprec {\gamlt} {\err_A} M A}
+
+  \inferrule*[right = Var]
+    { d : A \ltdyn B \and \gamlt(x) = (A, B) } { \etmprec {\gamlt} x x d  }
+
+  \inferrule*[right = Zro]
+    { } {\etmprec \gamlt \zro \zro \nat }
+
+  \inferrule*[right = Suc]
+    { \etmprec \gamlt M N \nat } {\etmprec \gamlt {\suc\, M} {\suc\, N} \nat}
+
+  \inferrule*[right = Lambda]
+    { d_i : A_i \ltdyn B_i \and 
+      d_o : A_o \ltdyn B_o \and 
+      \etmprec {\gamlt , x : d_i} M N {d_o} } 
+    { \etmprec \gamlt {\lda x M} {\lda x N} (d_i \To d_o) }
+
+  \inferrule*[right = App]
+    { d_i : A_i \ltdyn B_i \and
+      d_o : A_o \ltdyn B_o \\\\
+      \etmprec \gamlt {M} {M'} (d_i \To d_o) \and
+      \etmprec \gamlt {N} {N'} d_i
+    } 
+    { \etmprec \gamlt {M\, N} {M'\, N'} {d_o}}
+\end{mathpar}
+
+We then have additional equational axioms, including transitivity, $\beta$ and $\eta$ laws, and
+rules characterizing upcasts as least upper bounds, and downcasts as greatest lower bounds.
+
+We write $M \equidyn N$ to mean that both $M \ltdyn N$ and $N \ltdyn M$.
+
+% TODO
+\begin{mathpar}
+  \inferrule*[right = Transitivity]
+    {\etmprec \gamlt M N d \and
+     \etmprec \gamlt N P {d'} } 
+    {\etmprec \gamlt M P {d' \circ d} }
+
+  \inferrule*[right = $\beta$]
+    { \hasty {\Gamma, x : A_i} M {A_o} \and
+      \hasty {\Gamma} V {A_i} } 
+    { \etmequidyn \gamlt {(\lda x M)\, V} {M[V/x]} {A_o} }
+
+  \inferrule*[right = $\eta$]
+    { \hasty {\Gamma} {V} {A_i \To A_o} } 
+    { \etmequidyn \gamlt {\lda x (V\, x)} V {A_i \To A_o} }
+
+  \inferrule*[right = UpR]
+    { d : A \ltdyn B \and 
+      \hasty \Gamma M A } 
+    { \etmprec \gamlt M {\up A B M} d  }
+
+  \inferrule*[right = UpL]
+    { d : A \ltdyn B \and
+      \etmprec \gamlt M N d } 
+    { \etmprec \gamlt {\up A B M} N B }
+
+  \inferrule*[right = DnL]
+    { d : A \ltdyn B \and 
+      \hasty \Gamma M B } 
+    { \etmprec \gamlt {\dn A B M} M d }
+
+  \inferrule*[right = DnR]
+    { d : A \ltdyn B \and
+      \etmprec \gamlt M N d } 
+    { \etmprec \gamlt M {\dn A B N} A }
+\end{mathpar}
+
+% TODO explain the least upper bound/greatest lower bound rules
+The rules UpR, UpL, DnL, and DnR were introduced in \cite{TODO} as a means
+of cleanly axiomatizing the intended behavior of casts in a way that
+doesn't depend on the specific constructs of the language.
+Intuitively, rule UpR says that the upcast of $M$ is an upper bound for $M$
+in that $M$ may error more, and UpL says that the upcast is the \emph{least}
+such upper bound, in that it errors more than any other upper bound for $M$.
+Conversely, DnL says that the downcast of $M$ is a lower bound, and DnR says
+that it is the \emph{greatest} lower bound.
+% These rules provide a clean axiomatization of the behavior of casts that doesn't
+% depend on the specific constructs of the language.
+
+\subsection{The Intensional Lambda Calculus}
+
+Now that we have described the syntax along with the type and term precision judgments for
+$\extlc$, we can now do the same for $\intlc$. One key difference between the two calculi
+is that we define $\intlc$ using the constructs available to us in the language of synthetic
+guarded domain theory, e.g., we use the $\later$ operator. Whereas when we defined the syntax
+of the extensional lambda calculus we were working in the category of sets, when we define the
+syntax of the intensional lambda calculus we will be working in the topos of trees. 
+% in a presheaf category known as the \emph{topos of trees}.
+
+More specifically, in $\intlc$, we not only have normal terms, but also terms available
+``later'', which we denote by $\tilde{M}$.
+We have a term constructor $\theta_s$ which takes a later-term and turns it into
+a term avaialble now. 
+The typing and precision rules for $\theta_s$ involve the $\later$ operator, as shown below.
+Observe that $\theta_s$ is a syntactic analogue to the $\theta$ constructor of the lifting monad
+that we will define in the section on domain theoretic constructions (Section \ref{sec:domain-theory}),
+but it is important to note that $\theta_s (\tilde{M})$ is an actual term in $\intlc$, whereas the
+$\theta$ constructor is a purely semantic construction. These will be connected when we discuss the
+interpretation of $\intlc$ into the semantic model.
+
+% Most aspects of the two calculi are the same, except that in $\intlc$ we add a new term constructor
+% $\theta_s$ which represents a syntactic way of delaying a term by one computation step.
+
+To better understand this situation, note that $\lda{x}{(\cdot)}$ can be viewed as a function
+(at the level of the ambient type theory) from terms of type $B$ under context $\Gamma, x : A$
+to terms of type $A \To B$ under context $\Gamma$. Similarly, we can view $\theta_s(\cdot)$ as a
+function (in the ambient type theory, which is sythetic guarded domain theory) taking terms
+$\tilde{M}$ of type $A$ avaiable \emph{later} to terms of type $A$ available \emph{now}.
+
+Notice that the notion of a ``term available later'' does not need to be part of the syntax
+of the intensional lambda calculus, because this can be expressed in the ambient theory.
+Similarly, we do not need a syntactic form to delay a term, because we can simply use $\nxt$.
+
+\begin{align*}
+  % &\text{Later Terms } \tilde{M}, \tilde{N}
+  &\text{Terms } M, N := \err_A, \, \dots \, \theta_s (\tilde{M}) \\
+\end{align*}
+
+\begin{mathpar}
+  \inferrule*[]
+    { \later_t (\hasty \Gamma {M_t} A) }
+    { \hasty \Gamma {\theta_s M} {A} }
+\end{mathpar}
+
+\begin{mathpar}
+  \inferrule*[]
+    { \later_t (\itmprec \gamlt {M_t} {N_t} d) }
+    { \itmprec \gamlt {\theta_s M} {\theta_s N} d }
+\end{mathpar}
+
+Recall that $\later_t$ is a dependent form of $\later$ where the arugment is allowed
+to mention $t$. In particular, here we apply the tick $t$ to the later-terms $M$ and $N$ to get 
+``now"-terms $M_t$ and $N_t$.
+
+Formally speaking, the term precision relation must be defined as a guarded fixpoint,
+i.e., we assume that the function is defined later, and use it to construct a definition
+that is defined ``now''. This involves applying a tick to the later-function to shift it
+to a now-function. Indeed, this is what we do in the formal Agda development, but in this
+paper, we will elide these issues as they are not relevant to the main ideas.
+
+% TODO equational theory
 
@@
 
+In this section, we discuss the fundamental objects of the model into which we will embed
+the intensional lambda calculus and inequational theory. It is important to remember that
+the constructions in this section are entirely independent of the syntax described in the
+previous section; the notions defined here exist in their own right as purely mathematical
+constructs. In the next section, we will link the syntax and semantics via an interpretation
+function.
+
+\subsection{The Lift Monad}
+
+When thinking about how to model intensional gradually-typed programs, we must consider
+their possible behaviors. On the one hand, we have \emph{failure}: a program may fail
+at run-time because of a type error. In addition to this, a program may ``think'',
+i.e., take a step of computation. If a program thinks forever, then it never returns a value,
+so we can think of the idea of thinking as a way of intensionally modelling \emph{partiality}.
+
+With this in mind, we can describe a semantic object that models these behaviors: a monad
+for embedding computations that has cases for failure and ``thinking''.
+Previous work has studied such a construct in the setting of the latter only, called the lift
+monad \cite{TODO}; here, we add the additional effect of failure.
+
+For a type $A$, we define the \emph{lift monad with failure} $\li A$, which we will just call
+the \emph{lift monad}, as the following datatype:
+
+\begin{align*}
+  \li A &:= \\
+  &\eta \colon A \to \li A \\
+  &\mho \colon \li A \\
+  &\theta \colon \later (\li A) \to \li A
+\end{align*}
+
+Formally, the lift monad $\li A$ is defined as the solution to the guarded recursive type equation
+
+\[ \li A \cong A + 1 + \later \li A. \]
+
+An element of $\li A$ should be viewed as a computation that can either (1) return a value (via $\eta$),
+(2) raise an error and stop (via $\mho$), or (3) think for a step (via $\theta$).
+
+Notice there is a computation $\fix \theta$ of type $\li A$. This represents a computation
+that thinks forever and never returns a value.
+
+Since we claimed that $\li A$ is a monad, we need to define the monadic operations
+and show that they repect the monadic laws. The return is just $\eta$, and extend
+is defined via by guarded recursion by cases on the input.
+% It is instructive to give at least one example of a use of guarded recursion, so
+% we show below how to define extend:
+% TODO
+%
+%
+Verifying that the monadic laws hold requires \lob-induction and is straightforward.
+
+The lift monad has the following universal property. Let $f$ be a function from $A$ to $B$,
+where $B$ is a $\later$-algebra, i.e., there is $\theta_B \colon \later B \to B$.
+Further suppose that $B$ is also an ``error-algebra'', that is, an algebra of the
+constant functor $1 \colon \text{Type} \to \text{Type}$ mapping all types to Unit.
+This latter statement amounts to saying that there is a map $\text{Unit} \to B$, so $B$ has a
+distinguished ``error element" $\mho_B \colon B$.
+
+Then there is a unique homomorphism of algebras $f' \colon \li A \to B$ such that
+$f' \circ \eta = f$. The function $f'(l)$ is defined via guarded fixpoint by cases on $l$. 
+In the $\mho$ case, we simply return $\mho_B$.
+In the $\theta(\tilde{l})$ case, we will return
+
+\[\theta_B (\lambda t . (f'_t \, \tilde{l}_t)). \]
+
+Recalling that $f'$ is a guaded fixpoint, it is available ``later'' and by
+applying the tick we get a function we can apply ``now''; for the argument,
+we apply the tick to $\tilde{l}$ to get a term of type $\li A$.
+
+\subsubsection{Model-Theoretic Description}
+We can describe the lift monad in the topos of trees model as follows.
+
 \subsection{Predomains}
+
+The next important construction is that of a \emph{predomain}. A predomain is intended to
+model the notion of error ordering that we want terms to have. Thus, we define a predomain $A$
+as a partially-ordered set, which consists of a type which we denote $\ty{A}$ and a reflexive,
+transitive, and antisymmetric relation $\le_P$ on $A$.
+
+For each type we want to represent, we define a predomain for the corresponding semantic
+type. For instance, we define a predomain for natural numbers, a predomain for the
+dynamic type, a predomain for functions, and a predomain for the lift monad. We
+describe each of these below.
+
+We define monotone functions between predomain as expected. Given predomains
+$A$ and $B$, we write $f \colon A \monto B$ to indicate that $f$ is a monotone
+function from $A$ to $B$, i.e, for all $a_1 \le_A a_2$, we have $f(a_1) \le_B f(a_2)$.
+
+% TODO
+\begin{itemize}
+  \item There is a predomain Nat for natural numbers, where the ordering is equality.
+
+  \item There is a predomain Dyn to represent the dynamic type. The underlying type
+  for this predomain is defined to be $\mathbb{N} + \later (Dyn \monto Dyn)$.
+  This definition is valid because the occurrences of Dyn are guarded by the $\later$.
+  The ordering is defined via guarded recursion by cases on the argument, using the
+  ordering on $\mathbb{N}$ and the ordering on monotone functions described below.
+
+  \item For a predomain $A$, there is a predomain $\li A$ that is the ``lift'' of $A$
+  using the lift monad. We use the same notation for $\li A$ when $A$ is a type
+  and $A$ is a predomain, since the context should make clear which one we are referring to.
+  The underling type of $\li A$ is simply $\li \ty{A}$, i.e., the lift of the underlying
+  type of $A$.
+  The ordering of $\li A$ is the ``lock-step error-ordering'' which we describe in \ref{subsec:lock-step}.
+
+  \item For predomains $A_i$ and $A_o$, we form the predomain of monotone functions
+  from $A_i$ to $A_o$, which we denote by $A_i \To A_o$.
+  Two such functions are related
+\end{itemize}
+
+Predomains will form the objects of a structure similar to a double category,
+with horizontal morphisms being monotone funcitons, and vertical morphisms being
+embedding-projection pairs discussed below.
+
+\subsection{Lock-step Error Ordering}\label{subsec:lock-step}
+
+As mentioned, the ordering on the lift of a predomain $A$
+The relation is parameterized by an ordering $\le_A$ on $A$.
+We call this the lock-step error-ordering, the idea being that two computations
+$l$ and $l'$ are related if they are in lock-step with regard to their
+intensional behavior. That is, if $l$ is $\eta x$, then $l'$ should be equal to $\eta y$
+for some $y$ such that $x \le_A y$.
 
@@
 
+\subsection{Combined Ordering}
+
 \subsection{EP-Pairs}
+
+The use of embedding-projection pairs in gradual typing was introduced by New and
+Ahmed \cite{TODO}.
+Here, we want to adapt this notion of embedding-projection pair to the setting
+of intensional denotaional semantics.
 
@@
 
-\subsection{Term Precision}
+\subsection{Term Precision via the Lock-Step Error Ordering}
 % Homogeneous vs heterogeneous term precision
@@
 \section{Graduality}\label{sec:graduality}
+
+\subsection{Outline}
+
+\subsection{Extensional to Intensional}
+
+\subsection{Intensional Results}
+
+\subsection{Adequacy}
+
+\subsection{Putting it Together}
 
@@
 
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{references}
+
 \end{document}

### 2023-03-30 presentation/presentation.tex
check-in reading group presentation

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Beamer Presentation
% LaTeX Template
% Version 2.0 (March 8, 2022)
%
% This template originates from:
% https://www.LaTeXTemplates.com
%
% Author:
% Vel (vel@latextemplates.com)
%
% License:
% CC BY-NC-SA 4.0 (https://creativecommons.org/licenses/by-nc-sa/4.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[
	11pt, % Set the default font size, options include: 8pt, 9pt, 10pt, 11pt, 12pt, 14pt, 17pt, 20pt
	%t, % Uncomment to vertically align all slide content to the top of the slide, rather than the default centered
	%aspectratio=169, % Uncomment to set the aspect ratio to a 16:9 ratio which matches the aspect ratio of 1080p and 4K screens and projectors
]{beamer}

\graphicspath{{Images/}{./}} % Specifies where to look for included images (trailing slash required)

\usepackage{booktabs} % Allows the use of \toprule, \midrule and \bottomrule for better rules in tables
\usepackage{comment}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{stmaryrd}

\input{../paper-new/defs}
%\renewcommand{\ltdyn}{\sqsubseteq \,}

%----------------------------------------------------------------------------------------
%	SELECT LAYOUT THEME
%----------------------------------------------------------------------------------------

% Beamer comes with a number of default layout themes which change the colors and layouts of slides. Below is a list of all themes available, uncomment each in turn to see what they look like.

%\usetheme{default}
%\usetheme{AnnArbor}
%\usetheme{Antibes}
%\usetheme{Bergen}
%\usetheme{Berkeley}
%\usetheme{Berlin}
%\usetheme{Boadilla}
%\usetheme{CambridgeUS}
%\usetheme{Copenhagen}
%\usetheme{Darmstadt}
%\usetheme{Dresden}
%\usetheme{Frankfurt}
%\usetheme{Goettingen}
%\usetheme{Hannover}
%\usetheme{Ilmenau}
%\usetheme{JuanLesPins}
%\usetheme{Luebeck}
\usetheme{Madrid}
%\usetheme{Malmoe}
%\usetheme{Marburg}
%\usetheme{Montpellier}
%\usetheme{PaloAlto}
%\usetheme{Pittsburgh}
%\usetheme{Rochester}
%\usetheme{Singapore}
%\usetheme{Szeged}
%\usetheme{Warsaw}

%----------------------------------------------------------------------------------------
%	SELECT COLOR THEME
%----------------------------------------------------------------------------------------

% Beamer comes with a number of color themes that can be applied to any layout theme to change its colors. Uncomment each of these in turn to see how they change the colors of your selected layout theme.

%\usecolortheme{albatross}
%\usecolortheme{beaver}
%\usecolortheme{beetle}
%\usecolortheme{crane}
%\usecolortheme{dolphin}
%\usecolortheme{dove}
%\usecolortheme{fly}
%\usecolortheme{lily}
%\usecolortheme{monarca}
%\usecolortheme{seagull}
%\usecolortheme{seahorse}
%\usecolortheme{spruce}
%\usecolortheme{whale}
%\usecolortheme{wolverine}

%----------------------------------------------------------------------------------------
%	SELECT FONT THEME & FONTS
%----------------------------------------------------------------------------------------

% Beamer comes with several font themes to easily change the fonts used in various parts of the presentation. Review the comments beside each one to decide if you would like to use it. Note that additional options can be specified for several of these font themes, consult the beamer documentation for more information.

\usefonttheme{default} % Typeset using the default sans serif font
%\usefonttheme{serif} % Typeset using the default serif font (make sure a sans font isn't being set as the default font if you use this option!)
%\usefonttheme{structurebold} % Typeset important structure text (titles, headlines, footlines, sidebar, etc) in bold
%\usefonttheme{structureitalicserif} % Typeset important structure text (titles, headlines, footlines, sidebar, etc) in italic serif
%\usefonttheme{structuresmallcapsserif} % Typeset important structure text (titles, headlines, footlines, sidebar, etc) in small caps serif

%------------------------------------------------

%\usepackage{mathptmx} % Use the Times font for serif text
\usepackage{palatino} % Use the Palatino font for serif text

%\usepackage{helvet} % Use the Helvetica font for sans serif text
\usepackage[default]{opensans} % Use the Open Sans font for sans serif text
%\usepackage[default]{FiraSans} % Use the Fira Sans font for sans serif text
%\usepackage[default]{lato} % Use the Lato font for sans serif text

%----------------------------------------------------------------------------------------
%	SELECT INNER THEME
%----------------------------------------------------------------------------------------

% Inner themes change the styling of internal slide elements, for example: bullet points, blocks, bibliography entries, title pages, theorems, etc. Uncomment each theme in turn to see what changes it makes to your presentation.

%\useinnertheme{default}
\useinnertheme{circles}
%\useinnertheme{rectangles}
%\useinnertheme{rounded}
%\useinnertheme{inmargin}

%----------------------------------------------------------------------------------------
%	SELECT OUTER THEME
%----------------------------------------------------------------------------------------

% Outer themes change the overall layout of slides, such as: header and footer lines, sidebars and slide titles. Uncomment each theme in turn to see what changes it makes to your presentation.

%\useoutertheme{default}
%\useoutertheme{infolines}
%\useoutertheme{miniframes}
%\useoutertheme{smoothbars}
%\useoutertheme{sidebar}
%\useoutertheme{split}
%\useoutertheme{shadow}
%\useoutertheme{tree}
%\useoutertheme{smoothtree}

%\setbeamertemplate{footline} % Uncomment this line to remove the footer line in all slides
%\setbeamertemplate{footline}[page number] % Uncomment this line to replace the footer line in all slides with a simple slide count

%\setbeamertemplate{navigation symbols}{} % Uncomment this line to remove the navigation symbols from the bottom of all slides

% See https://tex.stackexchange.com/a/124271
\setbeamertemplate{bibliography item}{\insertbiblabel}

%----------------------------------------------------------------------------------------
%	PRESENTATION INFORMATION
%----------------------------------------------------------------------------------------

\title[SGDT + GTT]{Synthetic Guarded Domain Theory + \\ Gradual Typing} % The short title in the optional parameter appears at the bottom of every slide, the full title in the main parameter is only on the title page

% \subtitle{Optional Subtitle} % Presentation subtitle, remove this command if a subtitle isn't required

\author[Eric Giovannini and Max New]{Eric Giovannini and Max New} % Presenter name(s), the optional parameter can contain a shortened version to appear on the bottom of every slide, while the main parameter will appear on the title slide

%\institute[U-M]{University of Michigan \\ \smallskip \textit{james@LaTeXTemplates.com}}
\institute[U-M]{University of Michigan}
% Your institution, the optional parameter can be used for the institution shorthand and will appear on the bottom of every slide after author names, while the required parameter is used on the title slide and can include your email address or additional information on separate lines

\date[\today]{MPLSE Reading Group \\ \today} % Presentation date or conference/meeting name, the optional parameter can contain a shortened version to appear on the bottom of every slide, while the required parameter value is output to the title slide

%----------------------------------------------------------------------------------------

\begin{document}

%----------------------------------------------------------------------------------------
%	TITLE SLIDE
%----------------------------------------------------------------------------------------

\begin{frame}
	\titlepage % Output the title slide, automatically created using the text entered in the PRESENTATION INFORMATION block above
\end{frame}

%----------------------------------------------------------------------------------------
%	TABLE OF CONTENTS SLIDE
%----------------------------------------------------------------------------------------

% The table of contents outputs the sections and subsections that appear in your presentation, specified with the standard \section and \subsection commands. You may either display all sections and subsections on one slide with \tableofcontents, or display each section at a time on subsequent slides with \tableofcontents[pausesections]. The latter is useful if you want to step through each section and mention what you will discuss.

\begin{frame}
	\frametitle{Overview} % Slide title, remove this command for no title

	\tableofcontents % Output the table of contents (all sections on one slide)
	%\tableofcontents[pausesections] % Output the table of contents (break sections up across separate slides)
\end{frame}

%----------------------------------------------------------------------------------------
%	PRESENTATION BODY SLIDES
%----------------------------------------------------------------------------------------

\section{Introduction}

\subsection{Gradual Typing}

\begin{frame}
	\frametitle{What is Gradual Typing?}

	Gradually-typed languages combine static and dynamic typing in a single language
	and allow smooth interaction between both typed and untyped code.

	\bigskip

	This allows programmers to get the best of both worlds: they can start off programming
	in an untyped style and later annotate the code with types.

	\bigskip

	Doing so should not alter the semantics of the program!

	\bigskip

	Gradually-typed languages are usually compiled to \textbf{cast calculi} where the casts
	are made explicit.

\end{frame}

\begin{frame}
	\frametitle{Graduality}

	Gradual Guarantee (Siek et al. \cite{SVCB2015}):
	Key property for a language to be considered gradually-typed.

	\smallskip

	Adding type annotations should not change the semantics of the program, except to possibly introduce type errors.

	\medskip

	Conversely: Removing type annotations should not change the behavior of the program.

\end{frame}

\begin{frame}
	\frametitle{Type and Term Precision}

	\textbf{Type Precision}: $A \ltdyn B$ means that $A$ is more precise than $B$,
	or equivalently, $B$ is more dynamic

	Least precise type: $\dyn$ (i.e., $A \ltdyn\, \dyn$ for all $A$)

	\medskip

	\textbf{Term precision}: Extension of type precision to terms

	Intuitively: $M \ltdyn N$ means ``$M$ behaves like $N$, but may error more''

	For each type $A$, there is an error-term $\mho_A$ such that $\mho_A \ltdyn M$ for all $M : A$.

	\medskip

	In the cast calculus, we allow casts between types $A$ and $B$ such that $A \ltdyn B$.

\end{frame}

\begin{frame}
	\frametitle{The Current Approach to Proving Graduality}

	Define a notion of \emph{contextual error approximation} (two programs are equivalent, up to one erroring more than the other)

	\medskip

	Construct a \emph{logical relations model} and show that it is sound with respect to contextual
	error approximation.

	\medskip

	This approach has been utilized by New and Ahmed \cite{NA2018} and New, Licata, and Ahmed \cite{NLA2019}.

\end{frame}

\begin{frame}
	\frametitle{Step Indexing}

	The logical relation must be \emph{step-indexed} in order to deal with issues of non-wellfoundnedness
	i.e. we index the relation by a natural number representing the ``fuel'' we have left to observe
	the expression.
	Whenever a non well-founded operation takes place, we decrement the step-index.

	\medskip

	This has a few downsides: 
	\begin{itemize}
		\item 	Need to keep track of step index throughout the proofs
		\item  	Need two seaprate expression logical relations (one that counts steps on the left,
		and one on the right)
		\item 	Transitivity of the logical relation is not straightforward
	\end{itemize}

\end{frame}

\subsection{SGDT}

\begin{frame}
	\frametitle{What is SGDT?}

	SGDT is a logic/type theory with certain new axioms that internalize the notion of step-indexing.

	\bigskip

	There is an endofunctor $\later \colon \type \to \type$, where $\later A$ represents
	values of type $A$ available one time step later.

	\medskip

	There is a ``delaying'' function $\nxt \colon A \to\, \later A$ that takes a value available
	now and views it as a value available later.

\end{frame}

\begin{frame}
	\frametitle{SGDT: Guarded Fixpoints}

	Fixpoint operator $\fix \colon (\later A \to A) \to A$.

	\medskip

	Idea: to construct an $A$ ``now``, it suffices to assume we have an $A$ ``later`` and use that
	to build an $A$ ``now``.

	\medskip

	When used for propositions, this is called ``\lob-induction''.

	\medskip

	Fix satisfies the following unrolling equation:

	\[ \fix(f) = f (\nxt (\fix(f))) \]

\end{frame}

\begin{frame}
	\frametitle{Clocks and Clock Quantification}

	SGDT comes with a notion of clocks, abstract objects which keep track of time steps.

	\medskip

	The operations above are with repsect to a given clock $\kpa$, e.g, we have $\later^\kpa$.

	\medskip

	The notion of \emph{clock quantification} is crucial for encoding coinductive types using guarded
    recursion, an idea first introduced by Atkey and McBride \cite{AM2013}.

\end{frame}

\begin{frame}
	\frametitle{The Topos of Trees Model}

	The denotational semantics of SGDT is in a category called the \emph{topos of trees}, denoted $\calS = \Set^{\omega^o}$.

	\medskip

	Objects: presheaves over the ordered natural numbers, i.e., families $\{X_i\}$ of sets
	indexed by natural numbers, along with restriction maps $r^X_i \colon X_{i+1} \to X_i$.

	\medskip

	Morphisms $\{X_i\}$ to $\{Y_i\}$: family of functions $f_i \colon X_i \to Y_i$
	that commute with the restriction maps in the obvious way, that is,
	$f_i \circ r^X_i = r^Y_i \circ f_{i+1}$.

	\medskip

	% TODO image

\end{frame}

\begin{frame}
	\frametitle{Denotations of Later, Next, and Fix}

	The type operator $\later$ is defined on an object $X$ by $(\later X)_0 = 1$
	and $(\later X)_{i+1} = X_i$.
    The restriction maps are given by $r^\later_0 =\, !$, where $!$ is the unique map into
	$1$, and $r^\later_{i+1} = r^X_i$.

	\medskip

	The morphism $\nxt^X \colon X \to\, \later X$ is defined pointwise by $\nxt^X_0 =\, !$,
	and $\nxt^X_{i+1} = r^X_i$.

	\medskip

  	Given a morphism $f \colon \later X \to X$, we define $\fix\, f$ pointwise
  	as $\fix_i(f) = f_{i} \circ \dots \circ f_0$.

	\bigskip

	Note that as defined, $\fix$ isn't actually a morphism in $\calS$: what is its source?
	We need an object for functions from $\later X \to X$.
	This is the internal hom $\later X \To X$.

	\medskip

	We can then define $\fix \colon (\later X \To X) \to X$; we omit the details.

\end{frame}

\begin{frame}
	\frametitle{Denotations of Later, Next, and Fix}
	% Need to use ampersand replacement in tikzcd (see https://tex.stackexchange.com/a/220603)

% https://q.uiver.app/?q=WzAsMTQsWzAsMSwiWCJdLFswLDIsIlxcbGF0ZXIgWCJdLFsyLDEsIlhfMCJdLFszLDEsIlhfMSJdLFs0LDEsIlhfMiJdLFs1LDEsIlhfMyJdLFsyLDIsIjEiXSxbMywyLCJYXzAiXSxbNCwyLCJYXzEiXSxbNSwyLCJYXzIiXSxbNiwxLCJcXGRvdHMiXSxbNiwyLCJcXGRvdHMiXSxbMCwwLCJcXHRleHR7SW4gJFxcY2FsUyR9Il0sWzIsMCwiXFx0ZXh0e0luICRcXFNldCR9Il0sWzMsMiwicl5YXzAiLDJdLFs0LDMsInJeWF8xIiwyXSxbNSw0LCJyXlhfMiIsMl0sWzcsNiwiISJdLFs4LDddLFs5LDhdLFswLDEsIlxcbnh0Il0sWzIsNiwiISIsMl0sWzMsNywicl5YXzAiLDJdLFs0LDgsInJeWF8xIiwyXSxbNSw5LCJyXlhfMiIsMl0sWzEwLDVdLFsxMSw5XV0=
\[\begin{tikzcd}[ampersand replacement=\&]
	{\text{In $\calS$}} \&\& {\text{In $\Set$}} \\
	X \&\& {X_0} \& {X_1} \& {X_2} \& {X_3} \& \dots \\
	{\later X} \&\& 1 \& {X_0} \& {X_1} \& {X_2} \& \dots
	\arrow["{r^X_0}"', from=2-4, to=2-3]
	\arrow["{r^X_1}"', from=2-5, to=2-4]
	\arrow["{r^X_2}"', from=2-6, to=2-5]
	\arrow["{!}", from=3-4, to=3-3]
	\arrow[from=3-5, to=3-4]
	\arrow[from=3-6, to=3-5]
	\arrow["\nxt", from=2-1, to=3-1]
	\arrow["{!}"', from=2-3, to=3-3]
	\arrow["{r^X_0}"', from=2-4, to=3-4]
	\arrow["{r^X_1}"', from=2-5, to=3-5]
	\arrow["{r^X_2}"', from=2-6, to=3-6]
	\arrow[from=2-7, to=2-6]
	\arrow[from=3-7, to=3-6]
\end{tikzcd}\]

\end{frame}

\begin{frame}
	\frametitle{Ticked Cubical Type Theory}

	In Ticked Cubical Type Theory \cite{MV2019}, there is an additional sort
    called \emph{ticks}.

	\medskip

	Given a clock $k$, a tick $t : \tick\, k$ serves as evidence that one unit of time has passed according to the clock $k$.

  	\medskip

  	The type $\later^k A$ is represented as a function from ticks of a clock $k$ to $A$.

	\medskip

  	The type $A$ is allowed to depend on $t$, in which case we write $\later^k_t A$
  	to emphasize the dependence.

	\medskip

	The rules for tick abstraction and application are similar to those of dependent
  	$\Pi$ types.

\end{frame}

% ########################################

\section{Graduality for GTLC}
\frame{\tableofcontents[currentsection]}

\subsection{GTLC}

\begin{frame}
	\frametitle{GTLC: Syntax}

	\begin{block}{Syntax}
	\begin{align*}
		\text{Types } A, B &:= \nat, \, \dyn, \, (A \To B) \\
		\text{Terms } M, N &:= \err_A, \, \zro, \,  \suc\, M, \, (\lda{x}{M}), \, (M\, N), \, \\
		& (\up{A}{B} M), \, (\dn{A}{B} M) \\
		\text{Contexts } \Gamma &:= \cdot, \, (\Gamma, x : A)
	\end{align*}
	\end{block}

\end{frame}

\begin{frame}
	\frametitle{GTLC: Typing}

	\begin{mathpar}
		\inferrule*{ }{\hasty \Gamma {\err_A} A}

		\inferrule*{ }{\hasty \Gamma \zro \nat}

		\inferrule {\hasty \Gamma M \nat} {\hasty \Gamma {\suc\, M} \nat}

		\inferrule* 
		  {\hasty {\Gamma, x : A} M B} 
		  {\hasty \Gamma {\lda x M} {A \To B}}

		\inferrule *
		  {\hasty \Gamma M {A \To B} \and \hasty \Gamma N A}
		  {\hasty \Gamma {M \, N} B}

		\inferrule*
		  {A \ltdyn B \and \hasty \Gamma M A}
		  {\hasty \Gamma {\up A B M} B }

		\inferrule*
		  {A \ltdyn B \and \hasty \Gamma M B}
		  {\hasty \Gamma {\dn A B M} A}
	  \end{mathpar}

\end{frame}

\begin{frame}
	\frametitle{GTLC: Type Precision}

	\begin{mathpar}
		\inferrule*[right = \dyn]
		  { }{\dyn \ltdyn \dyn}

		\inferrule*[right = \nat]
		  { }{\nat \ltdyn \nat}

		\inferrule*[right = $Inj_\nat$]
		  { }{\nat \ltdyn \dyn}

		\inferrule*[right = $\To$]
		  {A_i \ltdyn B_i \and A_o \ltdyn B_o }
		  {(A_i \To A_o) \ltdyn (B_i \To B_o)}

		\inferrule*[right=$Inj_{\To}$]
		  {(A_i \to A_o) \ltdyn (\dyn \To \dyn)}
		  {(A_i \to A_o) \ltdyn\, \dyn}

	\end{mathpar}

	\medskip

	\textbf{Precision Derivations}:

	For every $A \ltdyn B$, we have a \emph{type precision derivation} $d : A \ltdyn B$
	that is constructed using the rules above.

	\smallskip

	For any type $A$, we use $A$ to denote the reflexivity derivation that $A \ltdyn A$,
	i.e., $A : A \ltdyn A$.

	\smallskip

	For type precision derivations $d : A \ltdyn B$ and $d' : B \ltdyn C$, we
	can define their composition $d' \circ d : A \ltdyn C$.

\end{frame}

\begin{frame}
	\frametitle{GTLC: Term Precision}

	Three kinds of rules: Congruence, Equational, and Cast Rules

	\bigskip

	\textbf{Congruence rules}: one per term constructor (except for casts)

	Two examples (other rules omitted):

	\begin{mathpar}
	  \inferrule*[right = Var]
      { d : A \ltdyn B \and \gamlt(x) = (A, B) } { \etmprec {\gamlt} x x d  }

	  \inferrule*[right = Lambda]
      { d_i : A_i \ltdyn B_i \and 
        d_o : A_o \ltdyn B_o \and 
        \etmprec {\gamlt , x : d_i} M N {d_o} } 
      { \etmprec \gamlt {\lda x M} {\lda x N} (d_i \To d_o) }
	\end{mathpar}

\end{frame}

\begin{frame}
	\frametitle{GTLC: Term Precision}

	\textbf{Equational Rules}: Transitivity, $\beta$ and $\eta$ laws

	\begin{mathpar}
		\inferrule*[right = Transitivity]
		{\etmprec \gamlt M N d \and
		\etmprec \gamlt N P {d'} } 
		{\etmprec \gamlt M P {d' \circ d} }

		\inferrule*[right = $\eta$]
		{ \hasty {\Gamma} {V} {A_i \To A_o} } 
		{ \etmequidyn \gamlt {\lda x (V\, x)} V {A_i \To A_o} }
	\end{mathpar}

\end{frame}

\begin{frame}
	\frametitle{GTLC: Term Precision}

	\textbf{Cast Rules}
	\begin{mathpar}
		\inferrule*[right = UpR]
		{ d : A \ltdyn B \and 
		  \hasty \Gamma M A } 
		{ \etmprec \gamlt M {\up A B M} d  }

	  \inferrule*[right = UpL]
		{ d : A \ltdyn B \and
		  \etmprec \gamlt M N d } 
		{ \etmprec \gamlt {\up A B M} N B }
	\end{mathpar}

	(The other rules DnL, DnR are dual.)

	\medskip

	The cast rules say that upcasts are least upper bounds, and dually, downcasts
	are greatest lower bounds.

	\medskip

\end{frame}

\begin{frame}
	\frametitle{Graduality for GTLC}

	\begin{theorem}[Graduality at Base Type]
		If $\cdot \vdash M \ltdyn N : \nat$, then
		\begin{enumerate}
		  \item If $N = \mho$, then $M = \mho$
		  \item If $N = V$, then $M = \mho$ or $M = V$, where $V = \zro$ or $V = \suc\, V'$
		  \item If $M = V$, then $N = V$
		\end{enumerate}
	  \end{theorem}

	  \bigskip

	  We also should be able to show that $\mho$, $\zro$, and $\suc\, N$ are not equal.

\end{frame}

\begin{frame}
	\frametitle{Intensional GTLC}

	In addition to the above language, which we call the \emph{extensional GTLC}
	($\extlc$ for short), we formalize the \emph{intensional GTLC} ($\intlc$ for short).

	\smallskip

	$\intlc$ includes syntax to express ``delayed" terms as terms, via the term constructor $\theta_s$ taking a term ``later" to a term ``now''.

\end{frame}

\begin{frame}
	\frametitle{Intensional GTLC}

	\begin{align*}
		% &\text{Later Terms } \tilde{M}, \tilde{N}
		&\text{Terms } M, N := \err_A, \, \dots \, \theta_s (\tilde{M}) \\
	  \end{align*}

	  Typing:
	  \begin{mathpar}
		\inferrule*[]
		  { \later_t (\hasty \Gamma {M_t} A) }
		  { \hasty \Gamma {\theta_s M} {A} }
	  \end{mathpar}

	  Term Precision:
	  \begin{mathpar}
		\inferrule*[]
		  { \later_t (\itmprec \gamlt {M_t} {N_t} d) }
		  { \itmprec \gamlt {\theta_s M} {\theta_s N} d }
	  \end{mathpar}

	  Recall that $\later_t$ is a dependent form of $\later$ where the arugment is allowed
	  to mention $t$. In particular, here we apply the tick $t$ to the later-terms $M$ and $N$ to get 
	  ``now"-terms $M_t$ and $N_t$.

\end{frame}

% #########################################################

\subsection{Domain-Theoretic Constructions}
\frame{\tableofcontents[currentsubsection]}

\begin{frame}
	\frametitle{The Lift Monad}

	Datatype that represents computations that at each step can return a value ($\eta$), terminate with an error ($\mho$), or
	``think'', i.e., defer the result to a later step ($\theta$).

	\begin{definition}[Lift Monad]
		\begin{align*}
			\li A &:= \\
			&\eta \colon A \to \li A \\
			&\mho \colon \li A \\
			&\theta \colon \later (\li A) \to \li A
		\end{align*}
	\end{definition}

	There is a computation $\fix(\theta)$ of type $\li A$; this represents a computation
	that thinks forever and never returns a value.

	\medskip

	Notation: We define $\delta \colon \li A \to \li A$ by $\delta = \theta \circ \nxt$

\end{frame}

\begin{frame}
	\frametitle{Predomains and Monotone Functions}

	A \textbf{predomain} $A$ consists of a type (which we denote $\ty{A}$) and a relation
	$\le_A$ on $A$ that satisfies the axioms of a partial ordering.

	\smallskip

	Since our types have an underlying order structure (representing the error ordering),
	we want to model types as partially-ordered sets in the semantics.

	\bigskip

	Then functions between terms will be modeled as \emph{monotone} functions between
	their corresponding predomains.

	\smallskip

	We write $f \colon A \monto B$ to indicate that $f$ is a monotone
	function from $A$ to $B$, i.e, for all $a_1 \le_A a_2$, we have $f(a_1) \le_B f(a_2)$.

\end{frame}

\begin{frame}
	\frametitle{Predomains}

	We define predomains for natural numbers, the dynamic type (which we denote $D$), and for monotone functions between predomains (which we denote $A_i \To A_o$).

	\medskip

  	For Dyn, the underlying type is defined to be

	\[ \ty{D} = \mathbb{N}\,\, + \later (D \monto D) \]

  	This definition is valid because the occurrences of $D$ are guarded by the $\later$.
  	The ordering is defined via guarded recursion by cases on the argument.

	\medskip

	We also define a predomain for the ``lifting'' of a predomain by the $\li$ monad.
	We denote this by $\li A$.

\end{frame}

\begin{frame}
	\frametitle{Lock-Step Ordering and Weak Bisimilarity}
	For a predomain $A$, the ordering on $\li A$ is called the ``lock-step error ordering'', denoted $l \ltls l'$.

	Intuitively: $l$ is less than $l'$ if they are in lock-step with regard to their
	intensional behavior, up to $l$ erroring.

	\begin{itemize}
		\item 	$\eta\, x \ltls \eta\, y$ if $x \le_A y$.
		\item 	$\mho \ltls l$ for all $l$ 
		\item   $\theta\, \tilde{r} \ltls \theta\, \tilde{r'}$ if
		        $\later_t (\tilde{r}_t \ltls \tilde{r'}_t)$
	\end{itemize}

	We analogously define a lifting of a heterogeneous relation $R$ between $A$ and $B$
	to a relation $L(R)$ between $\li A$ and $\li B$.

\end{frame}

\begin{frame}
	\frametitle{Lock-Step Ordering and Weak Bisimilarity}

	We also define another ordering on $\li A$, called ``weak bisimilarity'',
	written $l \bisim l'$.

	We say $l \bisim l'$ if they are equivalent ``up to delays''.

	\begin{align*}
		&\mho \bisim \mho \\
%
		&\eta\, x \bisim \eta\, y \text{ if } 
		  x \sim_A y \\
%		
		&\theta\, \tilde{x} \bisim \theta\, \tilde{y} \text{ if } 
		  \later_t (\tilde{x}_t \bisim \tilde{y}_t) \\
%	
		&\theta\, \tilde{x} \bisim \mho \text{ if } 
		  \theta\, \tilde{x} = \delta^n(\mho) \text { for some $n$ } \\
%	
		&\theta\, \tilde{x} \bisim \eta\, y \text{ if }
		  (\theta\, \tilde{x} = \delta^n(\eta\, x))
		\text { for some $n$ and $x : \ty{A}$ such that $x \sim_A y$ } \\
%
		&\mho \bisim \theta\, \tilde{y} \text { if } 
		  \theta\, \tilde{y} = \delta^n(\mho) \text { for some $n$ } \\
%	
		&\eta\, x \bisim \theta\, \tilde{y} \text { if }
		  (\theta\, \tilde{y} = \delta^n (\eta\, y))
		\text { for some $n$ and $y : \ty{A}$ such that $x \sim_A y$ }
	\end{align*}

\end{frame}

\begin{frame}
	\frametitle{EP Pairs}

	We will model casts as \emph{EP-pairs}.

	\bigskip

	Given predomains $A$ and $B$, an EP-pair $c : A \ep B$ consists of $\emb{c}{\cdot} \colon A \to B$ and $\proj{c}{\cdot} \colon B \to \li A$, and a monotone relation $R_c$ between $A$ and $B$.

	\medskip

	The relation $R_c$ should be related in a specific way to the embedding and projection functions.

\end{frame}

\begin{frame}
	\frametitle{EP Pairs}

	We have an identity EP-pair $\id : A \ep A$, with the embedding and projection equal to the identity and $\eta$, respectively.

	\medskip

	Recall: $D \cong \mathbb{N}\,\, + \later(D \monto D)$

	\medskip

	We have an EP-pair $\injnat$, where the embedding is just $\inl$ and
	Projection checks if the value of type $D$ is a nat and returns it, otherwise returns $\mho$.

\end{frame}

\begin{frame}
	\frametitle{EP Pairs}

	We have an EP-pair $\injarr \colon (D \to \li D) \ep D$.
	The embedding delays the function and injects into the sum type of $D$:
	$e(f) = \inl(\nxt f)$
	The projection does case analysis on the value of type $D$, and if it
	is a nat, returns $\mho$, otherwise, it it's a delayed function $\tilde{f}$, it returns

	\[ \theta_t (\eta (\tilde{f}_t)). \]

	\medskip

	For EP pairs $c_i : A_i \ep B_i$ and $c_o : A_o \ep B_o$ we have the EP-pair $c_i \To c_o : (A_i \monto A_o) \ep (B_i \monto B_o)$.

	The embedding and projection are defined functorially via the embeddings and projections of the domain and codomain.
	%The relation $R$ is the ``heterogeneous'' relation on functions, i.e.,
	%if $a_i\, R_{c_i}\, b_i$ then $f_1(a_i)\, R_{c_o}\, f_2(b_i)$

\end{frame}

\begin{frame}
	\frametitle{EP Pairs: Semantics}

	We would like the semantic analogues of the cast rules to hold, e.g.,

	\begin{mathpar}
		\inferrule*[right = DnL]
		{ c : A \ep B \and M : \ty{B} } 
		{ {\proj c M} \quad L(R) \quad M }
	\end{mathpar}

	Unfortunately, this does not hold, because the projection function for
	$\injarr$ introduces a $\theta$, and so the LHS and RHS are not in lock-step!

	\medskip

	This problem leaks into the embedding functions as well via functoriality in the $c_i \To c_o$ case.

\end{frame}

\begin{frame}
	\frametitle{Wait functions}

	To remedy this, we associate to each EP pair four ``wait'' functions
	that mirror the structure of the embedding and projection functions for their EP-pair.

	\begin{align*}
		&\wle \colon A \monto A \\
		&\wre \colon A \monto A \\
		&\wlp \colon A \monto \li A \\
		&\wrp \colon A \monto \li A \\
	\end{align*}

	Each wait function appears in one of the four semantic analogues of the cast rules, i.e., the rule above becomes

	\begin{mathpar}
		\inferrule*[right = DnL]
		{ c : A \ep B \and M : \ty{B} } 
		{ {\proj c M} \quad L(R) \quad \wrp(c)(M) }
	\end{mathpar}

\end{frame}

% \subsection{Denotational Semantics for Intensional GTLC}

% \begin{frame}
% 	\frametitle{Interpreting Intensional GTLC}

% \end{frame}

\subsection{Outline of Graduality Proof}
\frame{\tableofcontents[currentsubsection]}

\begin{frame}
	\frametitle{Main Theorem}

	\begin{theorem}[Graduality at Base Type]
		If $\cdot \vdash M_e \ltdyn_e N_e : \nat$, then
		\begin{enumerate}
		  \item If $N_e = \mho$, then $M_e = \mho$
		  \item If $N_e = V$, then $M_e = \mho$ or $M_e = V$, where $V = \zro$ or $V = \suc\, V'$
		  \item If $M_e = V$, then $N_e = V$
		\end{enumerate}
	  \end{theorem}

\end{frame}

\begin{frame}
	\frametitle{Extensional Collapse}

	We define a ``collapse" function $\erase{\cdot} \colon \intlc \to \extlc$
	that ``forgets'' about the intensional delay information, i.e., all occurrences of $\theta_s$ are erased.

	\medskip

	Every term $M_e$ in $\extlc$ will have a corresponding program $M_i$ in $\intlc$ such that $\erase{M_i} = M_e$.

	\medskip

	Moreover, we will show that if $M_e \ltdyn_e M_e'$ in the extensional theory,
	then there exists terms $M_i$ and $M_i'$ such that $\erase{M_i}=M_e$, $\erase{M_i'}=M_e'$ and
	$M_i \ltdyn_i M_i'$ in the intensional theory.

\end{frame}

\begin{frame}
	\frametitle{The Current Picture}

	% https://q.uiver.app/?q=WzAsNCxbMCwwLCJcXGludGxjIl0sWzIsMCwiXFxleHRsYyJdLFsyLDIsIlxcdGV4dHtFeHQuIFNlbS59Il0sWzAsMiwiXFx0ZXh0e0ludC4gU2VtLn0iXSxbMCwxLCJcXGVyYXNle1xcY2RvdH0iXSxbMCwzLCJcXHNlbXtcXGNkb3R9XmkiLDJdLFszLDIsIlxcdGV4dHtjb2xsYXBzZX0iLDJdLFsxLDIsIj8/Il1d
\[\begin{tikzcd}[ampersand replacement=\&]
	\intlc \&\& \extlc \\
	\\
	{\text{Int. Sem.}} \&\& {\text{Ext. Sem.}}
	\arrow["{\erase{\cdot}}", from=1-1, to=1-3]
	\arrow["{\sem{\cdot}^i}"', from=1-1, to=3-1]
	\arrow["{\text{collapse}}"', from=3-1, to=3-3]
	\arrow["{??}", from=1-3, to=3-3]
\end{tikzcd}\]

\end{frame}

% \begin{frame}
% 	\frametitle{Intensional Graduality Results}

% \end{frame}

% \begin{frame}
% 	\frametitle{Intensional Adequacy}

% 	We want to prove that the lock-step and weak bisimilarity relations
% 	are \emph{adequate}, i.e.,

% \end{frame}

\section{Discussion and Lessons Learned}
\frame{\tableofcontents[currentsection]}

\begin{frame}
	\frametitle{Benefits and Drawbacks}

	Positives:
	\begin{itemize}
		\item SGDT handles much of the tedious step-index reasoning
		\item Clarifies the underlying semantic and algebraic structure
	\end{itemize}

	Drawbacks:
	\begin{itemize}
		\item 	Intensional semantics is much more complicated (needed to introduce wait functions)
		\item 	Still need to work ``analytically'' with monotone functions
		\item   Need to do a lot of manual ``unfolding'' of fixpoint definitions in Guarded Cubical Agda
	\end{itemize}

\end{frame}

\begin{frame}[allowframebreaks] % Use [allowframebreaks] to allow automatic splitting across slides if the content is too long
	\frametitle{References}

	\begin{thebibliography}{99} % Beamer does not support BibTeX so references must be inserted manually as below, you may need to use multiple columns and/or reduce the font size further if you have many references
		\footnotesize % Reduce the font size in the bibliography

		\bibitem[1]{AM2013}
			Robert Atkey and Conor McBride.
			\newblock Productive coprogramming with guarded recursion.
			\newblock \emph{ACM SIGPLAN Notices 48, 9 (2013), 197-208. }

		\bibitem[2]{BMSS2012}
			Lars Birkedal, Rasmus Ejlers Møgelberg, Jan Schwinghammer, and Kristian Støvring.			
			\newblock First steps in synthetic guarded domain theory: step-indexing in the topos of trees.
			\newblock \emph{Logical Methods in Computer Science 8, 4 (2012). }

		\bibitem[3]{MV2019}
			Rasmus Ejlers Møgelberg and Niccolò Veltri.
			\newblock Bisimulation as path type for guarded recursive types.
			\newblock \emph{Proc. ACM Program. Lang. 3, POPL Article 4 (January 2019)}

		\bibitem[4]{MP2019}
			Rasmus E Møgelberg and Marco Paviotti.		
			\newblock Denotational semantics of recursive types in synthetic guarded domain theory.
			\newblock \emph{Mathematical Structures in Computer Science 29, 3 (2019), 465-510.  }

		\bibitem[5]{NA2018}
			Max S. New and Amal Ahmed.		
			\newblock Graduality from Embedding-Projection Pairs. 
			\newblock \emph{MProc. ACM Program. Lang. 2, ICFP, Article 73 (September 2018), 30 pages. }

		\bibitem[6]{NLA2019}
			Max S. New, Daniel R. Licata, and Amal Ahmed.
			\newblock Gradual type theory. 
			\newblock \emph{Proc. ACM Program. Lang. 3, POPL, Article 15 (January 2019), 31 pages. }

		\bibitem[7]{SVCB2015}
			Jeremy G. Siek, Michael M. Vitousek, Matteo Cimini, and John Tang Boyland
			\newblock Refined Criteria for Gradual Typing
			\newblock \emph{1st Summit on Advances in Programming Languages (SNAPL 2015). }

	\end{thebibliography}
\end{frame}

\begin{comment}
\section{Text Examples} % Sections are added in order to organize your presentation into discrete blocks, all sections and subsections are automatically output to the table of contents as an overview of the talk but NOT output in the presentation as separate slides

%------------------------------------------------

\subsection{Paragraphs and Lists}

\begin{frame}
	\frametitle{Paragraphs of Text}

	Sed iaculis \alert{dapibus gravida}. Morbi sed tortor erat, nec interdum arcu. Sed id lorem lectus. Quisque viverra augue id sem ornare non aliquam nibh tristique. Aenean in ligula nisl. Nulla sed tellus ipsum. Donec vestibulum ligula non lorem vulputate fermentum accumsan neque mollis.

	\bigskip % Vertical whitespace

	% Quote example
	\begin{quote}
		Sed diam enim, sagittis nec condimentum sit amet, ullamcorper sit amet libero. Aliquam vel dui orci, a porta odio.\\
		--- Someone, somewhere\ldots
	\end{quote}

	\bigskip % Vertical whitespace

	Nullam id suscipit ipsum. Aenean lobortis commodo sem, ut commodo leo gravida vitae. Pellentesque vehicula ante iaculis arcu pretium rutrum eget sit amet purus. Integer ornare nulla quis neque ultrices lobortis.
\end{frame}

%------------------------------------------------

\begin{frame}
	\frametitle{Lists}
	\framesubtitle{Bullet Points and Numbered Lists} % Optional subtitle

	\begin{itemize}
		\item Lorem ipsum dolor sit amet, consectetur adipiscing elit
		\item Aliquam blandit faucibus nisi, sit amet dapibus enim tempus
		\begin{itemize}
			\item Lorem ipsum dolor sit amet, consectetur adipiscing elit
			\item Nam cursus est eget velit posuere pellentesque
		\end{itemize}
		\item Nulla commodo, erat quis gravida posuere, elit lacus lobortis est, quis porttitor odio mauris at libero
	\end{itemize}

	\bigskip % Vertical whitespace

	\begin{enumerate}
		\item Nam cursus est eget velit posuere pellentesque
		\item Vestibulum faucibus velit a augue condimentum quis convallis nulla gravida 
	\end{enumerate}
\end{frame}

%------------------------------------------------

\subsection{Blocks}

\begin{frame}
	\frametitle{Blocks of Highlighted Text}

	\begin{block}{Block Title}
		Lorem ipsum dolor sit amet, consectetur adipiscing elit. Integer lectus nisl, ultricies in feugiat rutrum, porttitor sit amet augue.
	\end{block}

	\begin{exampleblock}{Example Block Title}
		Aliquam ut tortor mauris. Sed volutpat ante purus, quis accumsan.
	\end{exampleblock}

	\begin{alertblock}{Alert Block Title}
		Pellentesque sed tellus purus. Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos.
	\end{alertblock}

	\begin{block}{} % Block without title
		Suspendisse tincidunt sagittis gravida. Curabitur condimentum, enim sed venenatis rutrum, ipsum neque consectetur orci.
	\end{block}
\end{frame}

%------------------------------------------------

\subsection{Columns}

\begin{frame}
	\frametitle{Multiple Columns}
	\framesubtitle{Subtitle} % Optional subtitle

	\begin{columns}[c] % The "c" option specifies centered vertical alignment while the "t" option is used for top vertical alignment
		\begin{column}{0.45\textwidth} % Left column width
			\textbf{Heading}
			\begin{enumerate}
				\item Statement
				\item Explanation
				\item Example
			\end{enumerate}
		\end{column}
		\begin{column}{0.5\textwidth} % Right column width
			Lorem ipsum dolor sit amet, consectetur adipiscing elit. Integer lectus nisl, ultricies in feugiat rutrum, porttitor sit amet augue. Aliquam ut tortor mauris. Sed volutpat ante purus, quis accumsan dolor.
		\end{column}
	\end{columns}
\end{frame}

%------------------------------------------------

\section{Table and Figure Examples}

\subsection{Table}

\begin{frame}
	\frametitle{Table}
	\framesubtitle{Subtitle} % Optional subtitle

	\begin{table}
		\begin{tabular}{l l l}
			\toprule
			\textbf{Treatments} & \textbf{Response 1} & \textbf{Response 2}\\
			\midrule
			Treatment 1 & 0.0003262 & 0.562 \\
			Treatment 2 & 0.0015681 & 0.910 \\
			Treatment 3 & 0.0009271 & 0.296 \\
			\bottomrule
		\end{tabular}
		\caption{Table caption}
	\end{table}
\end{frame}

%------------------------------------------------

\subsection{Figure}

\begin{frame}
	\frametitle{Figure}

	\begin{figure}
		%\includegraphics[width=0.8\linewidth]{creodocs_logo.pdf}
		\caption{Creodocs logo.}
	\end{figure}
\end{frame}

%------------------------------------------------

\section{Mathematics}

\begin{frame}
	\frametitle{Definitions \& Examples}

	\begin{definition}
		A \alert{prime number} is a number that has exactly two divisors.
	\end{definition}

	\smallskip % Vertical whitespace

	\begin{example}
		\begin{itemize}
			\item 2 is prime (two divisors: 1 and 2).
			\item 3 is prime (two divisors: 1 and 3).
			\item 4 is not prime (\alert{three} divisors: 1, 2, and 4).
		\end{itemize}
	\end{example}

	\smallskip % Vertical whitespace

	You can also use the \texttt{theorem}, \texttt{lemma}, \texttt{proof} and \texttt{corollary} environments.
\end{frame}

%------------------------------------------------

\begin{frame}
	\frametitle{Theorem, Corollary \& Proof}

	\begin{theorem}[Mass--energy equivalence]
		$E = mc^2$
	\end{theorem}

	\begin{corollary}
		$x + y = y + x$
	\end{corollary}

	\begin{proof}
		$\omega + \phi = \epsilon$
	\end{proof}
\end{frame}

%------------------------------------------------

\begin{frame}
	\frametitle{Equation}

	\begin{equation}
		\cos^3 \theta =\frac{1}{4}\cos\theta+\frac{3}{4}\cos 3\theta
	\end{equation}
\end{frame}

%------------------------------------------------

\begin{frame}[fragile] % Need to use the fragile option when verbatim is used in the slide
	\frametitle{Verbatim}

	\begin{example}[Theorem Slide Code]
		\begin{verbatim}
			\begin{frame}
				\frametitle{Theorem}
				\begin{theorem}[Mass--energy equivalence]
					$E = mc^2$
				\end{theorem}
		\end{frame}\end{verbatim} % Must be on the same line
	\end{example}
\end{frame}

%------------------------------------------------

\begin{frame}
	Slide without title.
\end{frame}

%------------------------------------------------

\section{Referencing}

\begin{frame}
	\frametitle{Citing References}

	An example of the \texttt{\textbackslash cite} command to cite within the presentation:

	\bigskip % Vertical whitespace

	This statement requires citation \cite{p1,p2}.
\end{frame}

%------------------------------------------------

\begin{frame} % Use [allowframebreaks] to allow automatic splitting across slides if the content is too long
	\frametitle{References}

	\begin{thebibliography}{99} % Beamer does not support BibTeX so references must be inserted manually as below, you may need to use multiple columns and/or reduce the font size further if you have many references
		\footnotesize % Reduce the font size in the bibliography

		\bibitem[Smith, 2022]{p1}
			John Smith (2022)
			\newblock Publication title
			\newblock \emph{Journal Name} 12(3), 45 -- 678.

		\bibitem[Kennedy, 2023]{p2}
			Annabelle Kennedy (2023)
			\newblock Publication title
			\newblock \emph{Journal Name} 12(3), 45 -- 678.
	\end{thebibliography}
\end{frame}

%----------------------------------------------------------------------------------------
%	ACKNOWLEDGMENTS SLIDE
%----------------------------------------------------------------------------------------

\begin{frame}
	\frametitle{Acknowledgements}

	\begin{columns}[t] % The "c" option specifies centered vertical alignment while the "t" option is used for top vertical alignment
		\begin{column}{0.45\textwidth} % Left column width
			\textbf{Smith Lab}
			\begin{itemize}
				\item Alice Smith
				\item Devon Brown
			\end{itemize}
			\textbf{Cook Lab}
			\begin{itemize}
				\item Margaret
				\item Jennifer
				\item Yuan
			\end{itemize}
		\end{column}		
		\begin{column}{0.5\textwidth} % Right column width
			\textbf{Funding}
			\begin{itemize}
				\item British Royal Navy
				\item Norwegian Government
			\end{itemize}
		\end{column}
	\end{columns}
\end{frame}

%----------------------------------------------------------------------------------------
%	CLOSING SLIDE
%----------------------------------------------------------------------------------------

\begin{frame}[plain] % The optional argument 'plain' hides the headline and footline
	\begin{center}
		{\Huge The End}

		\bigskip\bigskip % Vertical whitespace

		{\LARGE Questions? Comments?}
	\end{center}
\end{frame}

%----------------------------------------------------------------------------------------
\end{comment}

\end{document}

### 2023-03-31 paper-new/paper.tex
More work on paper

@@
   % Proving graduality via LR
-  New and Ahmed have developed a semantic approach to specifying type dynamism in term of
-  embedding-projection pairs, which allows for a particularly elegant formulation of the
+  New and Ahmed \cite{new-ahmed2018}
+  have developed a semantic approach to specifying type dynamism in terms of
+  \emph{embedding-projection pairs}, which allows for a particularly elegant formulation of the
   gradual guarantee.
-  Moreover, this approach allows for type-based reasoning using $\eta$-equivalences
+  Moreover, their axiomatic account of program equivalence allows for type-based reasoning
+  about program equivalences.
+%
+  In this approach, a logical relation is constructed and shown to be sound with respect to
+  the notion of observational approximation that specifies when one program is more precise than another.
+  The downside of this approach is that each new language requires a different logical relation
+  to prove graduality. Furthermore, the logical relations tend to be quite complicated due
+  to a technical requirement known as \emph{step-indexing}.
+  As a result, developments using this approach tend to require vast effort, with the
+  corresponding technical reports having 50+ pages of proofs.
 
-  The downside of the above approach is that each new language requires a different logical relation
-  to prove graduality. As a result, many developments using this approach require vast effort,
-  with many such papers having 50+ pages of proofs.
-  Our aim here is that by mechanizing a graduality proof in a reusable way,
-  we will provide a framework for other language designers to use to ensure that their languages satsify graduality.
+  An alternative approach, which we investigate in this paper, is provided by
+  \emph{synthetic guarded domain theory}.
+  The tecnhiques of synthetic guarded domain theory allow us to internalize the
+  step-index reasoning normally required in logical relations proofs of graduality,
+  ultimately allowing us to specify the logical relation in a manner that looks nearly
+  identical to a typical, non-step-indexed logical relation.
 
-  One approach currently used to prove graduality uses the technique of \emph{logical relations}.
-  Specifially, a logical relation is constructed and shown to be sound with respect to
-  observational approximation. Because the dynamic type is modeled as a non-well-founded
-  recursive type, the logical relation needs to be paramterized by natural numbers to restore well-foundedness.
-  This technique is known as a \emph{step-indexed logical relation} \cite{TODO}.
-  Reasoning about step-indexed logical relations
-  can be tedious and error-prone, and there are some very subtle aspects that must
-  be taken into account in the proofs. Figure \ref{TODO} shows an example of a step-indexed logical
-  relation for the gradually-typed lambda calculus.
-
-  % TODO move to another section?
-  % Difficulties in prior semantics
-  % - two separate logical relations for expressions
-  % - transitivity
-  In particular, the prior approach of New and Ahmed requires two separate logical
-  relations for terms, one in which the steps of the left-hand term are counted,
-  and another in which the steps of the right-hand term are counted \cite{TODO}.
-  Then two terms $M$ and $N$ are related in the ``combined'' logical relation if they are
-  related in both of the one-sided logical relations. Having two separate logical relations
-  complicates the statement of the lemmas used to prove graduality, becasue any statement that
-  involves a term stepping needs to take into account whether we are counting steps on the left
-  or the right. Some of the differences can be abstracted over, but difficulties arise for properties %/results
-  as fundamental and seemingly straightforward as transitivty.
-
-  Specifically, for transitivity, we would like to say that if $M$ is related to $N$ at
-  index $i$ and $N$ is related to $P$ at index $i$, then $M$ is related to $P$ at $i$.
-  But this does not actually hold: we requrie that one of the two pairs of terms
-  be related ``at infinity'', i.e., that they are related at $i$ for all $i \in \mathbb{N}$.
-  Which pair is required to satisfy this depends on which logical relation we are considering,
-  (i.e., is it counting steps on the left or on the right),
-  and so any argument that uses transitivity needs to consider two cases, one
-  where $M$ and $N$ must be shown to be related for all $i$, and another where $N$ and $P$ must
-  be related for all $i$. This may not even be possible to show in some scenarios!
-
-  These complications introduced by step-indexing lead one to wonder whether there is a
-  way of proving graduality without relying on tedious arguments involving natural numbers.
-  An alternative approach, which we investigate in this paper, is provided by
-  \emph{synthetic guarded domain theory}, as discussed below.
-  Synthetic guarded domain theory allows the resulting logical relation to look almost
-  identical to a typical, non-step-indexed logical relation.
+  In this paper, we report on work we have done to mechanize graduality proofs
+  using SGDT techniques in Agda.
+  Our hope in this work is that by mechanizing a graduality proof in a reusable way,
+  we will provide a framework for other language designers to use to more easily and
+  conveniently prove that their languages satsify graduality.
 
@@
   To demonstrate the feasability and utility of our approach, we have used the framework to prove
-  graduality for the simply-typed gradual lambda calculus. Along the way, we have developed an intensional theory
+  graduality for the simply-typed gradual lambda calculus.Along the way, we have developed an ``intensional" theory
   of graduality that is of independent interest.
@@
   \subsection{Proving Graduality in SGDT}
-  % TODO move to technical background
+  \textbf{TODO:} This section should probably be moved to after the relevant background has been introduced.
 
-  % Cast calculi
-  In a gradually-typed language, the mixing of static and dynamic code is seamless, in that
-  the dynamically typed parts are checked at runtime. This type checking occurs at the elimination
-  forms of the language (e.g., pattern matching, field reference, etc.).
-  Gradual languages are generally elaborated to a \emph{cast calculus}, in which the dynamic
-  type checking is made explicit through the insertion of \emph{type casts}.
-
-  % Up and down casts
-  In a cast calculus, there is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that
-  $A$ is a \emph{more precise} type than $B$.
-  There a dynamic type $\dyn$ with the property that $A \ltdyn \dyn$ for all $A$.
-  %
-  If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
-  and a term $N$ of type $B$ may be \emph{down}casted to $A$, written $\dn A B N$.
-  Upcasts always succeed, while downcasts may fail at runtime.
-  %
-  % Syntactic term precision
-  We also have a notion of \emph{syntatcic term precision}.
-  If $A \ltdyn B$, and $M$ and $N$ are terms of type $A$ and $B$ respectively, we write
-  $M \ltdyn N : A \ltdyn B$ to mean that $M$ is more precise than $N$, i.e., $M$ and $N$
-  behave the same except that $M$ may error more.
-
-  % Modeling the dynamic type as a recursive sum type?
-  % Observational equivalence and approximation?
-
-  In this paper, we will be using SGDT techniques to prove graduality for a particularly
+  In this paper, we will utilize SGDT techniques to prove graduality for a particularly
   simple gradually-typed cast calculus, the gradually-typed lambda calculus. This is just
@@
   for which we will prove graduality. Important here are the notions of syntactic
-  type precision and term precision. We introduce both the extensional gradual lambda calculus
-  ($\extlc$) and the intensional gradual lambda calculus ($\intlc$).
+  type precision and term precision. We introduce both the \emph{extensional} gradual lambda calculus
+  ($\extlc$) and the \emph{intensional} gradual lambda calculus ($\intlc$).
 %
@@
   In Section \ref{sec:graduality}, we outline in more detail the proof of graduality for the
-  extensional gradual lambda calculus.
+  extensional gradual lambda calculus, which will make use of prove properties we prove about
+   the intensional gradual lambda calculus.
 %
@@
 
+\subsection{Gradual Typing}
+
+% Cast calculi
+In a gradually-typed language, the mixing of static and dynamic code is seamless, in that
+the dynamically typed parts are checked at runtime. This type checking occurs at the elimination
+forms of the language (e.g., pattern matching, field reference, etc.).
+Gradual languages are generally elaborated to a \emph{cast calculus}, in which the dynamic
+type checking is made explicit through the insertion of \emph{type casts}.
+
+% Up and down casts
+In a cast calculus, there is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that
+$A$ is a \emph{more precise} type than $B$.
+There a dynamic type $\dyn$ with the property that $A \ltdyn \dyn$ for all $A$.
+%
+If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
+and a term $N$ of type $B$ may be \emph{down}casted to $A$, written $\dn A B N$.
+Upcasts always succeed, while downcasts may fail at runtime.
+%
+% Syntactic term precision
+We also have a notion of \emph{syntatcic term precision}.
+If $A \ltdyn B$, and $M$ and $N$ are terms of type $A$ and $B$ respectively, we write
+$M \ltdyn N : A \ltdyn B$ to mean that $M$ is more precise than $N$, i.e., $M$ and $N$
+behave the same except that $M$ may error more.
+
+% Modeling the dynamic type as a recursive sum type?
+% Observational equivalence and approximation?
+
 % synthetic guarded domain theory, denotational semantics therein
+
+\subsection{Difficulties in Prior Semantics}
+
+ % TODO move to another section?
+  % Difficulties in prior semantics
+  % - two separate logical relations for expressions
+  % - transitivity
+
+  In this work, we compare our approach to proving graduality to the approach
+  introduced by New and Ahmed \cite{new-ahmed2018} which constructs a step-indexed
+  logical relations model and shows that this model is sound with respect to their
+  notion of contextual error approximation.
+
+  Because the dynamic type is modeled as a non-well-founded
+  recursive type, their logical relation needs to be paramterized by natural numbers
+  to restore well-foundedness. This technique is known as a \emph{step-indexed logical relation}.
+  Reasoning about step-indexed logical relations
+  can be tedious and error-prone, and there are some very subtle aspects that must
+  be taken into account in the proofs. Figure \ref{TODO} shows an example of a step-indexed logical
+  relation for the gradually-typed lambda calculus.
+
+  In particular, the prior approach of New and Ahmed requires two separate logical
+  relations for terms, one in which the steps of the left-hand term are counted,
+  and another in which the steps of the right-hand term are counted.
+  Then two terms $M$ and $N$ are related in the ``combined'' logical relation if they are
+  related in both of the one-sided logical relations. Having two separate logical relations
+  complicates the statement of the lemmas used to prove graduality, becasue any statement that
+  involves a term stepping needs to take into account whether we are counting steps on the left
+  or the right. Some of the differences can be abstracted over, but difficulties arise for properties %/results
+  as fundamental and seemingly straightforward as transitivty.
+
+  Specifically, for transitivity, we would like to say that if $M$ is related to $N$ at
+  index $i$ and $N$ is related to $P$ at index $i$, then $M$ is related to $P$ at $i$.
+  But this does not actually hold: we requrie that one of the two pairs of terms
+  be related ``at infinity'', i.e., that they are related at $i$ for all $i \in \mathbb{N}$.
+  Which pair is required to satisfy this depends on which logical relation we are considering,
+  (i.e., is it counting steps on the left or on the right),
+  and so any argument that uses transitivity needs to consider two cases, one
+  where $M$ and $N$ must be shown to be related for all $i$, and another where $N$ and $P$ must
+  be related for all $i$. This may not even be possible to show in some scenarios!
+
+  % These complications introduced by step-indexing lead one to wonder whether there is a
+  % way of proving graduality without relying on tedious arguments involving natural numbers.
+  % An alternative approach, which we investigate in this paper, is provided by
+  % \emph{synthetic guarded domain theory}, as discussed below.
+  % Synthetic guarded domain theory allows the resulting logical relation to look almost
+  % identical to a typical, non-step-indexed logical relation.
+
 \subsection{Synthetic Guarded Domain Theory}
@@
 domain theory}, or SGDT for short. We provide a brief overview here, but more
-details can be found in \cite{TODO}.
+details can be found in \cite{birkedal-mogelberg-schwinghammer-stovring2011}.
 
@@
 (pronounced ``later''). The use of a modality to express guarded recursion
-was introduced by Nakano \cite{TODO}.
-
+was introduced by Nakano \cite{Nakano2000}.
+%
 Given a type $A$, the type $\later A$ represents an element of type $A$
@@
 
-There is a fixpoint operator
+There is a \emph{guarded fixpoint} operator
 
@@
 
-In SGDT, there is also a new sort called \emph{clocks}. A clock serves as a reference
-relative to which the constructions described above are carried out.
+The operators $\later$, $\next$, and $\fix$ described above can be indexed by objects
+called \emph{clocks}. A clock serves as a reference relative to which steps are counted.
 For instance, given a clock $k$ and type $T$, the type $\later^k T$ represents a value of type
@@
 However, the notion of \emph{clock quantification} is crucial for encoding coinductive types using guarded
-recursion, an idea first introduced by Atkey and McBride \cite{TODO}.
+recursion, an idea first introduced by Atkey and McBride \cite{atkey-mcbride2013}.
 
@@
 \subsubsection{Ticked Cubical Type Theory}
-% TODO motivation for Clocked Cubical Type Theory, e.g., delayed substitutions
+% TODO motivation for Clocked Cubical Type Theory, e.g., delayed substitutions?
 
@@
 
-% TODO next as a function that ignores its input tick argument
+% TODO next as a function that ignores its input tick argument?
 
@@
 
-% TODO axioms (clock irrelevance, tick irrelevance)
+% TODO axioms (clock irrelevance, tick irrelevance)?
 
-% TODO topos of trees model
-\subsection{The Topos of Trees Model}
+\subsubsection{The Topos of Trees Model}
+
+The topos of trees model provides a useful intuition for reasoning in SGDT 
+\cite{birkedal-mogelberg-schwinghammer-stovring2011}.
+This section presupposes knowledge of category theory and can be safely skipped without
+disrupting the continuity.
 
 The topos of trees $\calS$ is the presheaf category $\Set^{\omega^o}$. 
-
+%
 We assume a universe $\calU$ of types, with encodings for operations such
@@
 An object $X$ is a family $\{X_i\}$ of sets indexed by natural numbers,
-  along with restriction maps $r^X_i \colon X_{i+1} \to X_i$.
+along with restriction maps $r^X_i \colon X_{i+1} \to X_i$ (see Figure \ref{fig:topos-of-trees-object}).
+These should be thought of as ``sets changing over time", where $X_i$ is the view of the set if we have $i+1$
+time steps to reason about it.
+We can also think of an ongoing computation, with $X_i$ representing the potential results of the computation
+after it has run for $i+1$ steps.
+
+\begin{figure}
+  % https://q.uiver.app/?q=WzAsNSxbMiwwLCJYXzIiXSxbMCwwLCJYXzAiXSxbMSwwLCJYXzEiXSxbMywwLCJYXzMiXSxbNCwwLCJcXGRvdHMiXSxbMiwxLCJyXlhfMCIsMl0sWzAsMiwicl5YXzEiLDJdLFszLDAsInJeWF8yIiwyXSxbNCwzLCJyXlhfMyIsMl1d
+\[\begin{tikzcd}[ampersand replacement=\&]
+	{X_0} \& {X_1} \& {X_2} \& {X_3} \& \dots
+	\arrow["{r^X_0}"', from=1-2, to=1-1]
+	\arrow["{r^X_1}"', from=1-3, to=1-2]
+	\arrow["{r^X_2}"', from=1-4, to=1-3]
+	\arrow["{r^X_3}"', from=1-5, to=1-4]
+\end{tikzcd}\]
+  \caption{An example of an object in the topos of trees.}
+  \label{fig:topos-of-trees-object}
+\end{figure}
 
@@
 that commute with the restriction maps in the obvious way, that is,
-  $f_i \circ r^X_i = r^Y_i \circ f_{i+1}$.
+$f_i \circ r^X_i = r^Y_i \circ f_{i+1}$ (see Figure \ref{fig:topos-of-trees-morphism}).
+
+\begin{figure}
+% https://q.uiver.app/?q=WzAsMTAsWzIsMCwiWF8yIl0sWzAsMCwiWF8wIl0sWzEsMCwiWF8xIl0sWzMsMCwiWF8zIl0sWzQsMCwiXFxkb3RzIl0sWzAsMSwiWV8wIl0sWzEsMSwiWV8xIl0sWzIsMSwiWV8yIl0sWzMsMSwiWV8zIl0sWzQsMSwiXFxkb3RzIl0sWzIsMSwicl5YXzAiLDJdLFswLDIsInJeWF8xIiwyXSxbMywwLCJyXlhfMiIsMl0sWzQsMywicl5YXzMiLDJdLFs2LDUsInJeWV8wIiwyXSxbNyw2LCJyXllfMSIsMl0sWzgsNywicl5ZXzIiLDJdLFs5LDgsInJeWV8zIiwyXSxbMSw1LCJmXzAiLDJdLFsyLDYsImZfMSIsMl0sWzAsNywiZl8yIiwyXSxbMyw4LCJmXzMiLDJdXQ==
+\[\begin{tikzcd}[ampersand replacement=\&]
+	{X_0} \& {X_1} \& {X_2} \& {X_3} \& \dots \\
+	{Y_0} \& {Y_1} \& {Y_2} \& {Y_3} \& \dots
+	\arrow["{r^X_0}"', from=1-2, to=1-1]
+	\arrow["{r^X_1}"', from=1-3, to=1-2]
+	\arrow["{r^X_2}"', from=1-4, to=1-3]
+	\arrow["{r^X_3}"', from=1-5, to=1-4]
+	\arrow["{r^Y_0}"', from=2-2, to=2-1]
+	\arrow["{r^Y_1}"', from=2-3, to=2-2]
+	\arrow["{r^Y_2}"', from=2-4, to=2-3]
+	\arrow["{r^Y_3}"', from=2-5, to=2-4]
+	\arrow["{f_0}"', from=1-1, to=2-1]
+	\arrow["{f_1}"', from=1-2, to=2-2]
+	\arrow["{f_2}"', from=1-3, to=2-3]
+	\arrow["{f_3}"', from=1-4, to=2-4]
+\end{tikzcd}\]
+  \caption{An example of a morphism in the topos of trees.}
+  \label{fig:topos-of-trees-morphism}
+\end{figure}
 
@@
 The morphism $\nxt^X \colon X \to \later X$ is defined pointwise by
-  $\nxt^X_0 =\, !$, and $\nxt^X_{i+1} = r^X_i$.
+$\nxt^X_0 =\, !$, and $\nxt^X_{i+1} = r^X_i$. It is easily checked that
+this satisfies the commutativity conditions required of a morphism in $\calS$.
+%
+Given a morphism $f \colon \later X \to X$, i.e., a family of functions
+$f_i \colon (\later X)_i \to X_i$ that commute with the restrictions in the appropriate way,
+we define $\fix(f) \colon 1 \to X$ pointwise
+by $\fix(f)_i = f_{i} \circ \dots \circ f_0$.
+This can be visualized as a diagram in the category of sets as shown in
+Figure \ref{fig:topos-of-trees-fix}.
+% Observe that the fixpoint is a \emph{global element} in the topos of trees.
+% Global elements allow us to view the entire computation on a global level.
 
-  Given a morphism $f \colon \later X \to X$, we define $\fix f$ pointwise
-  as $\fix_i(f) = f_{i} \circ \dots \circ f_0$.
+% https://q.uiver.app/?q=WzAsOCxbMSwwLCJYXzAiXSxbMiwwLCJYXzEiXSxbMywwLCJYXzIiXSxbMCwwLCIxIl0sWzAsMSwiWF8wIl0sWzEsMSwiWF8xIl0sWzIsMSwiWF8yIl0sWzMsMSwiWF8zIl0sWzMsNCwiZl8wIl0sWzAsNSwiZl8xIl0sWzEsNiwiZl8yIl0sWzIsNywiZl8zIl0sWzAsMywiISIsMl0sWzEsMCwicl5YXzAiLDJdLFsyLDEsInJeWF8xIiwyXSxbNSw0LCJyXlhfMCJdLFs2LDUsInJeWF8xIl0sWzcsNiwicl5YXzIiXV0=
+% \[\begin{tikzcd}[ampersand replacement=\&]
+% 	1 \& {X_0} \& {X_1} \& {X_2} \\
+% 	{X_0} \& {X_1} \& {X_2} \& {X_3}
+% 	\arrow["{f_0}", from=1-1, to=2-1]
+% 	\arrow["{f_1}", from=1-2, to=2-2]
+% 	\arrow["{f_2}", from=1-3, to=2-3]
+% 	\arrow["{f_3}", from=1-4, to=2-4]
+% 	\arrow["{!}"', from=1-2, to=1-1]
+% 	\arrow["{r^X_0}"', from=1-3, to=1-2]
+% 	\arrow["{r^X_1}"', from=1-4, to=1-3]
+% 	\arrow["{r^X_0}", from=2-2, to=2-1]
+% 	\arrow["{r^X_1}", from=2-3, to=2-2]
+% 	\arrow["{r^X_2}", from=2-4, to=2-3]
+% \end{tikzcd}\]
 
-  % TODO global elements
+% \begin{figure}
+%   % https://q.uiver.app/?q=WzAsMTksWzEsMiwiWF8wIl0sWzIsMywiWF8xIl0sWzMsMSwiMSJdLFswLDEsIjEiXSxbMCwyLCJYXzAiXSxbMSwzLCJYXzEiXSxbMSwxLCIxIl0sWzIsMSwiMSJdLFsyLDIsIlhfMCJdLFsyLDQsIlhfMiJdLFszLDIsIlhfMCJdLFszLDMsIlhfMSJdLFszLDQsIlhfMiJdLFszLDUsIlhfMyJdLFs0LDIsIlxcY2RvdHMiXSxbMCwwLCJcXGZpeChmKV8wIl0sWzEsMCwiXFxmaXgoZilfMSJdLFsyLDAsIlxcZml4KGYpXzIiXSxbMywwLCJcXGZpeChmKV8zIl0sWzMsNCwiZl8wIl0sWzAsNSwiZl8xIl0sWzYsMCwiZl8wIl0sWzcsOCwiZl8wIl0sWzgsMSwiZl8xIl0sWzEsOSwiZl8yIl0sWzIsMTAsImZfMCJdLFsxMCwxMSwiZl8xIl0sWzExLDEyLCJmXzIiXSxbMTIsMTMsImZfMyJdXQ==
+%   \[\begin{tikzcd}[ampersand replacement=\&]
+%     {\fix(f)_0} \& {\fix(f)_1} \& {\fix(f)_2} \& {\fix(f)_3} \\
+%     1 \& 1 \& 1 \& 1 \\
+%     {X_0} \& {X_0} \& {X_0} \& {X_0} \& \cdots \\
+%     \& {X_1} \& {X_1} \& {X_1} \\
+%     \&\& {X_2} \& {X_2} \\
+%     \&\&\& {X_3}
+%     \arrow["{f_0}", from=2-1, to=3-1]
+%     \arrow["{f_1}", from=3-2, to=4-2]
+%     \arrow["{f_0}", from=2-2, to=3-2]
+%     \arrow["{f_0}", from=2-3, to=3-3]
+%     \arrow["{f_1}", from=3-3, to=4-3]
+%     \arrow["{f_2}", from=4-3, to=5-3]
+%     \arrow["{f_0}", from=2-4, to=3-4]
+%     \arrow["{f_1}", from=3-4, to=4-4]
+%     \arrow["{f_2}", from=4-4, to=5-4]
+%     \arrow["{f_3}", from=5-4, to=6-4]
+%   \end{tikzcd}\]
+%   \caption{The first few approximations to the guarded fixpoint of $f$.}
+%   \label{fig:topos-of-trees-fix-approx}
+% \end{figure}
+
+\begin{figure}
+  % https://q.uiver.app/?q=WzAsNixbMywwLCIxIl0sWzAsMiwiWF8wIl0sWzIsMiwiWF8xIl0sWzQsMiwiWF8yIl0sWzYsMiwiWF8zIl0sWzgsMiwiXFxkb3RzIl0sWzIsMSwicl5YXzAiXSxbNCwzLCJyXlhfMiJdLFswLDEsIlxcZml4KGYpXzAiLDFdLFswLDIsIlxcZml4KGYpXzEiLDFdLFswLDMsIlxcZml4KGYpXzIiLDFdLFswLDQsIlxcZml4KGYpXzMiLDFdLFswLDUsIlxcY2RvdHMiLDMseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMywyLCJyXlhfMSJdLFs1LDQsInJeWF8zIl1d
+  \[\begin{tikzcd}[ampersand replacement=\&,column sep=2.25em]
+    \&\&\& 1 \\
+    \\
+    {X_0} \&\& {X_1} \&\& {X_2} \&\& {X_3} \&\& \dots
+    \arrow["{r^X_0}", from=3-3, to=3-1]
+    \arrow["{r^X_2}", from=3-7, to=3-5]
+    \arrow["{\fix(f)_0}"{description}, from=1-4, to=3-1]
+    \arrow["{\fix(f)_1}"{description}, from=1-4, to=3-3]
+    \arrow["{\fix(f)_2}"{description}, from=1-4, to=3-5]
+    \arrow["{\fix(f)_3}"{description}, from=1-4, to=3-7]
+    \arrow["\cdots"{marking}, draw=none, from=1-4, to=3-9]
+    \arrow["{r^X_1}", from=3-5, to=3-3]
+    \arrow["{r^X_3}", from=3-9, to=3-7]
+  \end{tikzcd}\]
+  \caption{The guarded fixpoint of $f$.}
+  \label{fig:topos-of-trees-fix}
+\end{figure}
+
+% TODO global elements?
 
@@
 We also give the rules for syntactic type and term precision.
-
-We start with the extensional lambda calculus $\extlc$, and then describe the additions
-necessary for the intensional lambda calculus $\intlc$.
+We define two separate calculi: the normal gradually-typed lambda calculus, which we
+call the extensional lambda calculus ($\extlc$), as well as an \emph{intensional} lambda calculus
+($\intlc$) whose syntax makes explicit the steps taken by a program.
 
@@
 
+\textbf{TODO: Subject to change!}
+
 Now that we have described the syntax along with the type and term precision judgments for
@@
 Previous work has studied such a construct in the setting of the latter only, called the lift
-monad \cite{TODO}; here, we add the additional effect of failure.
+monad \cite{mogelberg-paviotti2016}; here, we add the additional effect of failure.
 
@@
 (2) raise an error and stop (via $\mho$), or (3) think for a step (via $\theta$).
-
+%
 Notice there is a computation $\fix \theta$ of type $\li A$. This represents a computation
@@
 
+% TODO
+
 \subsection{Predomains}
@@
   \item There is a predomain Dyn to represent the dynamic type. The underlying type
-  for this predomain is defined to be $\mathbb{N} + \later (Dyn \monto Dyn)$.
+  for this predomain is defined by guarded fixpoint to be such that
+  $\ty{\Dyn} \cong \mathbb{N}\, +\, \later (\ty{\Dyn} \monto \ty{\Dyn})$.
   This definition is valid because the occurrences of Dyn are guarded by the $\later$.
@@
   type of $A$.
-  The ordering of $\li A$ is the ``lock-step error-ordering'' which we describe in \ref{subsec:lock-step}.
+  The ordering on $\li A$ is the ``lock-step error-ordering'' which we describe in
+  \ref{subsec:lock-step}.
 
@@
   from $A_i$ to $A_o$, which we denote by $A_i \To A_o$.
-  Two such functions are related
+  The ordering is such that $f$ is below $g$ if for all $a$ in $\ty{A_i}$, we have
+  $f(a)$ is below $g(a)$ in the ordering for $A_o$. 
 \end{itemize}
-
-Predomains will form the objects of a structure similar to a double category,
-with horizontal morphisms being monotone funcitons, and vertical morphisms being
-embedding-projection pairs discussed below.
 
@@
 As mentioned, the ordering on the lift of a predomain $A$
-The relation is parameterized by an ordering $\le_A$ on $A$.
-We call this the lock-step error-ordering, the idea being that two computations
+is called the \emph{lock-step error-ordering}, the idea being that two computations
 $l$ and $l'$ are related if they are in lock-step with regard to their
-intensional behavior. That is, if $l$ is $\eta x$, then $l'$ should be equal to $\eta y$
-for some $y$ such that $x \le_A y$.
+intensional behavior, up to $l$ erroring.
+Formally, we define this ordering as follows:
+
+\begin{itemize}
+  \item 	$\eta\, x \ltls \eta\, y$ if $x \le_A y$.
+  \item 	$\mho \ltls l$ for all $l$ 
+  \item   $\theta\, \tilde{r} \ltls \theta\, \tilde{r'}$ if
+          $\later_t (\tilde{r}_t \ltls \tilde{r'}_t)$
+\end{itemize}
+
+We also define a heterogeneous version of this ordering between the lifts of two
+different predomains $A$ and $B$, parameterized by a relation $R$ between $A$ and $B$.
 
@@
 
-\subsection{Combined Ordering}
+We also define another ordering on $\li A$, called ``weak bisimilarity'',
+written $l \bisim l'$.
+Intuitively, we say $l \bisim l'$ if they are equivalent ``up to delays''.
+We introduce the notation $x \sim_A y$ to mean $x \le_A y$ and $y \le_A x$.
+
+The weak bisimilarity relation is defined by guarded fixpoint as follows:
+
+\begin{align*}
+  &\mho \bisim \mho \\
+%
+  &\eta\, x \bisim \eta\, y \text{ if } 
+    x \sim_A y \\
+%		
+  &\theta\, \tilde{x} \bisim \theta\, \tilde{y} \text{ if } 
+    \later_t (\tilde{x}_t \bisim \tilde{y}_t) \\
+%	
+  &\theta\, \tilde{x} \bisim \mho \text{ if } 
+    \theta\, \tilde{x} = \delta^n(\mho) \text { for some $n$ } \\
+%	
+  &\theta\, \tilde{x} \bisim \eta\, y \text{ if }
+    (\theta\, \tilde{x} = \delta^n(\eta\, x))
+  \text { for some $n$ and $x : \ty{A}$ such that $x \sim_A y$ } \\
+%
+  &\mho \bisim \theta\, \tilde{y} \text { if } 
+    \theta\, \tilde{y} = \delta^n(\mho) \text { for some $n$ } \\
+%	
+  &\eta\, x \bisim \theta\, \tilde{y} \text { if }
+    (\theta\, \tilde{y} = \delta^n (\eta\, y))
+  \text { for some $n$ and $y : \ty{A}$ such that $x \sim_A y$ }
+\end{align*}
 
@@
 
-The use of embedding-projection pairs in gradual typing was introduced by New and
-Ahmed \cite{TODO}.
-Here, we want to adapt this notion of embedding-projection pair to the setting
-of intensional denotaional semantics.
+Here, we adapt the notion of embedding-projection pair as used to study
+gradual typing (\cite{new-ahmed2018}) to the setting of intensional denotaional semantics.
 
@@
 
-\subsection{Types as Predomains}
+\subsection{Relational Semantics}
 
-\subsection{Terms as Monotone Functions}
+\subsubsection{Types as Predomains}
 
-\subsection{Type Precision as EP-Pairs}
+\subsubsection{Terms as Monotone Functions}
 
-\subsection{Term Precision via the Lock-Step Error Ordering}
+\subsubsection{Type Precision as EP-Pairs}
+
+\subsubsection{Term Precision via the Lock-Step Error Ordering}
 % Homogeneous vs heterogeneous term precision
 
+\subsection{Logical Relations Semantics}
+
 \section{Graduality}\label{sec:graduality}
+The main theorem we would like to prove is the following:
+
+\begin{theorem}[Graduality]
+  If $\cdot \vdash M \ltdyn N : \nat$, then
+  \begin{enumerate}
+    \item If $N = \mho$, then $M = \mho$
+    \item If $N = `n$, then $M = \mho$ or $M = `n$
+    \item If $M = V$, then $N = V$
+  \end{enumerate}
+\end{theorem}
 

### 2023-04-14 notes/technical-outline/technical-outline.tex
check in technical notes document

\documentclass{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{tikz-cd}
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{parskip}
\usepackage{tikz-cd}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{nonnum-theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{definition}{Definition}[section]
\newtheorem{lemma}{Lemma}[section]

\input{../../paper-new/defs}

\begin{document}

\title{Technical Outline}
\author{Eric Giovannini}

\maketitle

\section{The Overall Plan}

The goal is to show graduality of the extensional gradual lambda calculus.
The main theorem is the following:

\begin{theorem}[Graduality]
If $M \ltdyn M'$, then

\begin{itemize}
    \item $M \Downarrow$ iff $M' \Downarrow$.
    \item $M \Downarrow v_?$ iff $M' \Downarrow v_?'$, where $v_? = \err$ or $v_? = v_?'$.
\end{itemize}
\end{theorem}

\section{Syntax}

    % TODO details on admissibility: we can translate equations of the original logic
    % into the new one and show that they hold there

    We begin with a gradually-typed lambda calculus $(\extlc)$, which is similar to
    the normal call-by-value gradually-typed lambda calculus, but differs in that it
    is actually a fragment of call-by-push-value specialized such that there are no
    non-trivial computation types. We do this for convenience, as either way
    we would need a distinction between values and effectful terms; the framework of
    of call-by-push-value gives us a convenient langugae to define what we need.

    We then observe that composition of type precision derivations is admissible, 
    as is heterogeneous transitivity for term
    precision (via casts), so it will suffice to consider a new language ($\extlcm$)
    in which we don't have composition of type precision derivations or transitivity
    of term precision.

    We further observe that all casts, except those between $\nat$ and $\dyn$
    and between $\dyn \ra \dyn$ and $\dyn$, are admissible
    (we can define the cast of a function type
    functorially using the casts for its domain and codomain).

    This means it is sufficient to consider a new language ($\extlcmm$) in which
    instead of having arbitrary casts, we have injections from $\nat$ and
    $\dyn \ra \dyn$ into $\dyn$, and case inspections from $\dyn$ to $\nat$ and
    $\dyn$ to $\dyn \ra \dyn$.

    From here, we define an intensional GSTLC, so-named because it makes the
    intensional stepping behavior of programs explicit in the syntax.
    This is acocmplished by adding a syntactic ``later'' type and a
    syntactic $\theta$ that maps terms of type later $A$ to terms of type $A$.

    \subsection{Extensional ($\extlcmm$)}

    \begin{align*}
        &\text{Value Types } A := \nat \alt \dyn \alt (A \ra B) \\
        &\text{Computation Types } B := \Ret A \\
        &\text{Value Contexts } \Gamma := \cdot \alt (\Gamma, x : A) \\
        &\text{Computation Contexts } \Delta := \cdot \alt \hole B \alt \Delta , x : A \\
        &\text{Values } V :=  \zro \alt \suc\, V \alt \injnat V \alt \injarr V \\ 
        &\text{Terms } M, N := \err_B \alt \Ret {V} \alt \bind{x}{M}{N} % \alt \suc\, M, 
          \alt \lda{x}{M} \alt V_f\, V_x \alt
          %\\ & \quad\quad \injnat M \alt \injarr M
          \\ & \quad\quad \casenat{V}{M_{no}}{n}{M_{yes}} 
          \alt \casearr{V}{M_{no}}{f}{M_{yes}}
      \end{align*}

    % TODO is err a value? What about injnat and injarr?
    % TODO should suc be a term and a value? What about injnat and injarr?

    The value typing judgment is written $\hasty{\Gamma}{V}{A}$ and 
    the computation typing judgment is written $\hasty{\Delta}{M}{B}$.

    We define substitution for computation contexts as follows:
    \begin{mathpar}
        \inferrule*
        { \delta : \Delta' \to \Delta \and 
          \hasty{\Delta'|_V}{V}{A}}
        { (\delta , V/x ) \colon \Delta' \to \Delta , x : A }

        \inferrule*
        {}
        {\cdot \colon \cdot \to \cdot}

        \inferrule*
        {\hasty{\Delta'}{M}{B}}
        {M \colon \Delta' \to \hole{B}}
    \end{mathpar}

    The typing rules are as follows:
    \begin{mathpar}
        % Err
        \inferrule*{ }{\hasty {\cdot, \Gamma} {\err_B} B}

        % Zero and suc
        \inferrule*{ }{\hasty \Gamma \zro \nat}

        \inferrule*{\hasty \Gamma M \nat} {\hasty \Gamma {\suc\, M} \nat}

        % Lambda
        \inferrule* 
        {\hasty {\cdot, \Gamma, x : A} M {\Ret A'}} 
        {\hasty \Gamma {\lda x M} {A \ra A'}}

        % App
        \inferrule*
        {\hasty \Gamma {V_f} {A \ra A'} \and \hasty \Gamma {V_x} A}
        {\hasty {\cdot , \Gamma} {M \, N} {\Ret A'}}

        % Ret
        \inferrule*
        {\hasty \Gamma V A}
        {\hasty {\cdot , \Gamma} {\ret V} {\Ret A}}

        % Bind
        \inferrule*
        {\hasty \Delta M {\Ret A} \and \hasty{\cdot , \Delta|_V , x : A}{N}{B} } % Need x : A in context
        {\hasty{\Delta}{\bind{x}{M}{N}}{B}}

        % inj-nat
        \inferrule*
        {\hasty \Gamma M \nat}
        {\hasty \Gamma {\injnat M} \dyn}

        % inj-arr 
        \inferrule*
        {\hasty \Gamma M (\dyn \ra \dyn)}
        {\hasty \Gamma {\injarr M} \dyn}

        % Case nat
        \inferrule*
        {\hasty{\Gamma}{V}{\dyn} \and 
          \hasty{\Delta , x : \nat }{M_{yes}}{B} \and 
          \hasty{\Delta}{M_{no}}{B}}
        {\hasty {\Delta} {\casenat{M}{M_{no}}{n}{M_{yes}}} {B}}

        % Case arr
        \inferrule*
        {\hasty{\Gamma}{V}{\dyn} \and 
          \hasty{\Delta , x : (\dyn \ra \dyn) }{M_{yes}}{B} \and 
          \hasty{\Delta}{M_{no}}{B}}
        {\hasty {\Delta} {\casearr{M}{M_{no}}{f}{M_{yes}}} {B}}
    \end{mathpar}

    The equational theory is as follows:
    \begin{mathpar}
        % Function Beta and Eta
        \inferrule*
        {\hasty {\cdot, \Gamma, x : A} M {\Ret A'} \and \hasty \Gamma V A}
        {(\lda x M)\, V = M[V/x]}

        \inferrule*
        {\hasty \Gamma V {A \ra A}}
        {\Gamma \vdash V = \lda x {V\, x}}

        % Case-nat Beta
        \inferrule*
        {\hasty \Gamma V \nat}
        {\casenat {\injnat {V}} {M_{no}} {n} {M_{yes}} = M_{yes}[V/n]}

        \inferrule*
        {\hasty \Gamma V {\dyn \ra \dyn} }
        {\casenat {\injarr {V}} {M_{no}} {n} {M_{yes}} = M_{no}}

        % Case-nat Eta
        \inferrule*
        {}
        {\Gamma , x :\, \dyn \vdash M = \casenat{x}{M}{n}{M[(\injnat{n}) / x]} }

        % Case-arr Beta

        % Case-arr Eta

        % Ret Beta and Eta
        \inferrule*
        {}
        {\bind{x}{\ret\, V}{N} = N[V/x]}

        \inferrule*
        {\hasty {\hole{\Ret A} , \Gamma} {M} {B}}
        {\hole{\Ret A}, \Gamma \vdash M = \bind{x}{\bullet}{M[\ret\, x]}}

    \end{mathpar}

    Equivalent terms in the equational theory are considered equal.

    We now discuss type and term precision. In our language, we do not have
    ``term precision'' but rather arbitrary monotone relations on types,
    which we denote by $A \rel B$. We have relations on value types, as well
    as on computation types.
    %
    In addition, because we don't have casts in our language, we do not have
    the usual cast rules specifying that casts are least upper bounds and greatest
    lower bounds. Instead, we have rules for our injection and case terms.

    \begin{align*}
        &\text{Value Relations } R := \nat \alt \dyn \alt (R \ra R) \alt \alpha_1 \leq_A \alpha_2 \\
        &\text{Computation Relations } S := \Lift R \\
    \end{align*}

    % We define

    % \begin{mathpar}
    %     \inferrule*
    %     {\alpha_1 : A_1 \mid \alpha_2 : A_2 \vdash R \text{ rel} \and 
    %      \alpha_1' : A_1' \vdash V_1 : A_1 \and
    %      \alpha_2' : A_2' \vdash V_2 : A_2 }
    %     {\alpha_1' : A_1' \mid \alpha_2' : A_2' \vdash R (V_1/\alpha_1, V_2/\alpha_2)}
    % \end{mathpar}

    % TODO term precision

    \subsection{Intensional}
    In the intensional syntax, we add a type constructor for later, as well as a
    syntactic $\theta$ term and a syntactic $\nxt$ term.
    We add rules for each of these, and also modify the rules for inj-arr and
    case-arr, since now
    the function is not $\Dyn \ra \Dyn$ but rather $\later (\Dyn \ra \Dyn)$.

    % TODO show changes

    We define an erasure function from intensional syntax to extensional syntax
    by induction on the intensional types and terms.
    The basic idea is that the syntactic type $\later A$ erases to $A$,
    and $\nxt$ and $\theta$ erase to the identity.

\section{Semantics}
To give a semantics of the extensional, quotiented syntax ($\extlcmm$), we proceed
in a few steps:

\begin{enumerate}
    \item We first give a denotational semantics of the intensional syntax $\intlc$
    using synthetic guarded domain theory, i.e., working internal to the topos of trees.
    The denotation of a value type $A$ will be a poset $\sem{A}$, and the denotation of a computation
    type $\Ret A$ will be the lift $\li \sem{A}$, (i.e., the free error domain on $\sem{A}$). % Ret A is our only computation type
    We refer to this semantics as the \emph{step-indexed semantics}.

    %The denotation of a closed term of type $B$ will be as an element of $\li \sem{B}$.

    \item We then show how to go from the above step-indexed semantics to a set-based
    semantics which is still intensional in that the denotations of terms that differ only in
    their number of ``steps'' will not be equal. The denotations here are a type of
    coinductive ``Machine'' that at each step of computation can either return a result,
    error, or continue running. We call this the \emph{Machine semantics}.
    %
    The passage from step-indexed semantics to the coinductive semantics
    will make use of clock quantification; see below for details.

    \item Finally, we collapse the above intensional Machine-based semantics to an
    extensional semantics, so-called because at this point the information about the
    precise number of steps a Machine has taken has been lost. We care only whether the
    Machine runs to a value or error, or whether it diverges. Of course, we cannot
    in general decide whether a given Machine will halt, so the semantic objects here
    are pairs of a proposition $P$ and a function taking a proof of $P$ to an element of
    the poset.

\end{enumerate}

The benefit to working synthetically in step 1 above is that the construction of the logical
relation that proves canonicity for the intensional syntax can be carried out much like
any normal, non-step-indexed logical relation.

\section{Unary Canonicity}

We first want to show soundness:

\begin{lemma}[Soundness]
    In the Machine semantics, none of the following are equal: $\mho, \zro, \suc M, \omega$,
    where $\omega$ represents a diverging program.
\end{lemma}

Next we use a logical relations argument to establish the following:

\begin{lemma}
    $M_i \Dwn^n v_?$ iff $M_i = \delta^n(\qte{v_?})$
\end{lemma}

We will need the following key property relating erasure to the semantics:

\begin{lemma}
    If $\erase{M_i} = M$ and $\erase{M_i'} = M'$, and $M = M'$, 
    then $| \sem{M_i} | = | \sem{M_i'} |$.
\end{lemma}

The idea of the proof is that since $M_i$ and $M_i'$ have equal erasures
(in the extensional equational theory), we can recover from the proof that their
erasures are equal a series of equational rules (e.g., $\beta$ equality), and
apply the intensional analogues of those rules to $M_i$ to ensure that $M_i$ and $M_i'$
differ syntactically only in their number of $\theta$'s.

\section{Graduality}

\end{document}

### 2023-04-25 paper-new/graduality.tex
work on paper (changes to intro, work on description of syntax)

--- 2023-01-12 paper-new/paper.tex
@@
-\documentclass[acmsmall,screen]{acmart}
-\usepackage{mathpartir}
-\usepackage{tikz-cd}
-
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2018}
-\acmYear{2018}
-\acmDOI{10.1145/1122445.1122456}
-
-%% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
-  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
-\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-  June 03--05, 2018, Woodstock, NY}
-\acmPrice{15.00}
-\acmISBN{978-1-4503-XXXX-X/18/06}
-
-\newcommand{\dyn}{?}
-\newcommand{\ltdyn}{\sqsubseteq}
-
-\newcommand{\uarrowl}{\mathrel{\rotatebox[origin=c]{-30}{$\leftarrowtail$}}}
-\newcommand{\uarrowr}{\mathrel{\rotatebox[origin=c]{60}{$\leftarrowtail$}}}
-\newcommand{\darrowl}{\mathrel{\rotatebox[origin=c]{30}{$\twoheadleftarrow$}}}
-\newcommand{\darrowr}{\mathrel{\rotatebox[origin=c]{120}{$\twoheadleftarrow$}}}
-\newcommand{\vuarrow}{\mathrel{\rotatebox[origin=c]{-90}{$\leftarrowtail$}}}
-\newcommand{\vdarrow}{\mathrel{\rotatebox[origin=c]{90}{$\twoheadleftarrow$}}}
-\newcommand{\up}[2]{\langle{#2}\uarrowl{#1}\rangle}
-\newcommand{\dn}[2]{\langle{#1}\darrowl{#2}\rangle}
-
-\newcommand{\later}{\vartriangleright}
-\newcommand{\type}{\texttt{Type}}
-\newcommand{\lob}{\text{L\"{o}b}}
-\newcommand{\tick}{\texttt{tick}}
-
-\begin{document}
-
-\title{Denotational Semantics for Gradual Typing in Synthetic Guarded Domain Theory}
-\author{Eric Giovannini}
-\author{Max S. New}
-
-\begin{abstract}
-    We develop a denotational semantics for a gradually typed language
-    with effects that is adequate and proves the graduality theorem.
-    %
-    The denotational semantics is constructed using \emph{synthetic
-    guarded domain theory} working in a type theory with a later
-    modality and clock quantification.
-    %
-    This provides a remarkably simple presentation of the semantics,
-    where gradual types are interpreted as ordinary types in our ambient
-    type theory equipped with an ordinary preorder structure to model
-    the error ordering.
-    %
-    This avoids the complexities of classical domain-theoretic models
-    (New and Licata) or logical relations models using explicit
-    step-indexing (New and Ahmed).
-    %
-    In particular, we avoid a major technical complexity of New and
-    Ahmed that requires two logical relations to prove the graduality
-    theorem.
-
-    By working synthetically we can treat the domains in which gradual
-    types are interpreted as if they were ordinary sets. This allows us
-    to give a ``na\"ive'' presentation of gradual typing where each
-    gradual type is modeled as a well-behaved subset of the universal
-    domain used to model the dynamic type, and type precision is modeled
-    as simply a subset relation.
-    %
-  \end{abstract}
-
-  \maketitle
-
-  \section{Introduction}
-
-  % gradual typing, graduality
-  \subsection{Gradual Typing and Graduality}
-  Gradual typing allows a language to have both statically-typed and dynamically-typed terms;
-  the statically-typed terms are type checked at compile time, while type checking for the 
-  dynamically-typed terms is deferred to runtime.
-
-  Gradually-typed languages support the smooth migration from dynamic typing
-  to static typing, in that the programmer can initially leave off the
-  typing annotations and provide them later without altering the meaning of the
-  program. Moreover, a key feature of gradually typed languages is that the interaction
-  between the static and dynamic components of the codebase should be safe -- i.e., should
-  preserve the guarantees made by the static types.
-  In other words, in the static portions of the codebase, type soundness must be preserved.
-
-  In a gradually-typed language, the mixing of static and dynamic code is seamless, in that
-  the dynamically typed parts are checked at runtime. This type checking occurs at the elimination
-  forms of the language (e.g., pattern matching, field reference, etc.).
-  Gradual languages are generally elaborated to a \emph{cast calculus}, in which the dynamic
-  type checking is made explicit through the insertion of \emph{type casts}.
-
-  ...
-
-  % Up and down casts
-  There is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that $A$ is a \emph{more
-  precise} type than $B$.
-  There a dynamic type $\dyn$ with the property that $A \ltdyn \dyn$ for all $A$.
-
-  If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
-  and a term $N$ of type $B$ may be \emph{down}casted to $A$, written $\dn A B N$.
-  Upcasts always succeed, while downcasts may fail at runtime.
-
-  % Graduality property
-  A key property that any gradually-typed language should satisfy is the \emph{dynamic gradual guarantee},
-  originally defined by Siek, Vitousek, Cimini, and Boyland \cite{TODO}.
-  New et al refer to this property as \emph{graduality}, by analogy with parametricity.
-  Intuitively, graduality says that in going from a dynamic to static style
-  should not introduce changes in the meaning of the program.
-  More specifically, making the types more precise (replacing $\dyn$ with a more specific type
-  such as integers) will either result in the same behavior as the less precise program,
-  or cause a dynamic type error.
-
-  % Observational equivalence and approximation
-  % TODO
-
-  \subsection{Prior Approach}
-  % Proving graduality via LR
-  The usual approach to proving graduality uses the technique of \emph{logical relations}.
-  Specifially, a logical relation is constructed and showed to be sound with respect to
-  observational approximation. Because the dynamic type is modeled as a recursive type,
-  the logical relation needs to be paramterized by natural numbers, a so-called
-  step-indexed logical relation. Reasoning about step-indexed logical relations
-  is tedious and error-prone, and there are several very subtle aspects that must
-  be taken into account in the proofs.
-
-  % Difficulties in prior semantics
-  % TODO
-
-  % - two separate logical relations for expressions
-  % - transitivity
-
-  % synthetic guarded domain theory, denotational semantics therein
-  \subsection{Synthetic Guarded Domain Theory}
-  One way to avoid the tedious reasoning associated with step-indexing is to work
-  axiomatically inside of a logical system that can reason about non-well-founded recursive
-  constructions while abstracting away the specific details of step-indexing required
-  if we were working analytically.
-  The system that proves useful fot this purpose is called \emph{synthetic guarded
-  domain theory}, or SGDT for short. We provide a brief overview here, but more
-  details can be found in \cite{TODO}.
-
-  In SGDT, there is the notion of a \emph{clock}, an abstract type that tracks the
-  passage of time through \emph{ticks}. Given a clock $k$, a tick $t : \tick k$ serves
-  as evidence that one unit of time has passed according to the clock $k$.
-
-  SGDT also introduces a modality $\later : \type \to \type$, (pronounced ``later'').
-  Given a clock $k$ and type $T$, the type $\later_k T$ represents a value of type
-  $T$ one unit of time in the future according to clock $k$.
-  The type $\later T$ is represented as a function from ticks of a clock $k$ to $T$.
-  There is an axiom called $\lob$-induction which is stated as follows:
-
-  \[
-    \lob : \forall T, (\later T \to T) \to T.
-  \]
-
-  That is, to construct a term of type $T$, it suffices to assume that we have access to
-  such a term ``later'' and use that to help us build a term ``now''.
-
-  In particular, this axiom applies to propositions $P : \texttt{Prop}$.
-
-  \subsection{Overview of Remainder of Paper}
-  In Section \ref{sec:GTLC}, we introduce the gradually-typed cast calculus
-  for which we will prove graduality. Important here are the notions of syntactic
-  type precision and term precision.
-
-  In Section \ref{sec:domain-theory}, we define several fundamental constructions
-  internal to SGDT that will be needed when we give a denotational semantics to
-  the language. This includes the notion of Predomains as well as the concept
-  of EP-Pairs.
-
-  In Section \ref{Semantics}, we define the denotational semantics for the gradually-typed
-  language
-
-\section{GTLC}\label{sec:GTLC}
-
-\subsection{Syntax}
-
-\subsection{Type Precision}
-
-\subsection{Term Precision}
-
-\section{Domain-Theoretic Constructions}\label{sec:domain-theory}
-
-\subsection{Predomains}
-
-\subsection{Weak Bisimilarity Relation}
-
-\subsection{EP-Pairs}
-
-\section{Semantics}\label{sec:semantics}
-
-\subsection{Types as Predomains}
-
-\subsection{Terms as Monotone Functions}
-
-\subsection{Type Precision as EP-Pairs}
-
-\subsection{Term Precision}
-% Homogeneous vs heterogeneous term precision
+\section{Unary Canonicity}
+Before discussing graduality, we seek to prove its ``unary'' analogue.
+Namely, instead of considering inequality between terms, we start by considering equality.
 
 \section{Graduality}\label{sec:graduality}
+The main theorem we would like to prove is the following:
 
-\section{Discussion}\label{sec:discussion}
-
-\subsection{Synthetic Ordering}
-
-While the use of synthetic guarded domain theory allows us to very
-conveniently work with non-well-founded recursive constructions while
-abstracting away the precise details of step-indexing, we do work with
-the error ordering in a mostly analytic fashion in that gradual types
-are interpreted as sets equipped with an ordering relation, and all
-terms must be proven to be monotone.
-%
-It is possible that a combination of synthetic guarded domain theory
-with \emph{directed} type theory would allow for an a synthetic
-treatment of the error ordering as well.
-
-\end{document}
+\begin{theorem}[Graduality]
+  If $\cdot \vdash M \ltdyn N : \nat$, then
+  \begin{enumerate}
+    \item If $N = \mho$, then $M = \mho$
+    \item If $N = `n$, then $M = \mho$ or $M = `n$
+    \item If $M = V$, then $N = V$
+  \end{enumerate}
+\end{theorem}

### 2023-04-25 paper-new/intro.tex
work on paper (changes to intro, work on description of syntax)

--- 2023-01-12 paper-new/paper.tex
@@
-\documentclass[acmsmall,screen]{acmart}
-\usepackage{mathpartir}
-\usepackage{tikz-cd}
-
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2018}
-\acmYear{2018}
-\acmDOI{10.1145/1122445.1122456}
-
-%% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
-  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
-\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-  June 03--05, 2018, Woodstock, NY}
-\acmPrice{15.00}
-\acmISBN{978-1-4503-XXXX-X/18/06}
-
-\newcommand{\dyn}{?}
-\newcommand{\ltdyn}{\sqsubseteq}
-
-\newcommand{\uarrowl}{\mathrel{\rotatebox[origin=c]{-30}{$\leftarrowtail$}}}
-\newcommand{\uarrowr}{\mathrel{\rotatebox[origin=c]{60}{$\leftarrowtail$}}}
-\newcommand{\darrowl}{\mathrel{\rotatebox[origin=c]{30}{$\twoheadleftarrow$}}}
-\newcommand{\darrowr}{\mathrel{\rotatebox[origin=c]{120}{$\twoheadleftarrow$}}}
-\newcommand{\vuarrow}{\mathrel{\rotatebox[origin=c]{-90}{$\leftarrowtail$}}}
-\newcommand{\vdarrow}{\mathrel{\rotatebox[origin=c]{90}{$\twoheadleftarrow$}}}
-\newcommand{\up}[2]{\langle{#2}\uarrowl{#1}\rangle}
-\newcommand{\dn}[2]{\langle{#1}\darrowl{#2}\rangle}
-
-\newcommand{\later}{\vartriangleright}
-\newcommand{\type}{\texttt{Type}}
-\newcommand{\lob}{\text{L\"{o}b}}
-\newcommand{\tick}{\texttt{tick}}
-
-\begin{document}
-
-\title{Denotational Semantics for Gradual Typing in Synthetic Guarded Domain Theory}
-\author{Eric Giovannini}
-\author{Max S. New}
-
-\begin{abstract}
-    We develop a denotational semantics for a gradually typed language
-    with effects that is adequate and proves the graduality theorem.
-    %
-    The denotational semantics is constructed using \emph{synthetic
-    guarded domain theory} working in a type theory with a later
-    modality and clock quantification.
-    %
-    This provides a remarkably simple presentation of the semantics,
-    where gradual types are interpreted as ordinary types in our ambient
-    type theory equipped with an ordinary preorder structure to model
-    the error ordering.
-    %
-    This avoids the complexities of classical domain-theoretic models
-    (New and Licata) or logical relations models using explicit
-    step-indexing (New and Ahmed).
-    %
-    In particular, we avoid a major technical complexity of New and
-    Ahmed that requires two logical relations to prove the graduality
-    theorem.
-
-    By working synthetically we can treat the domains in which gradual
-    types are interpreted as if they were ordinary sets. This allows us
-    to give a ``na\"ive'' presentation of gradual typing where each
-    gradual type is modeled as a well-behaved subset of the universal
-    domain used to model the dynamic type, and type precision is modeled
-    as simply a subset relation.
-    %
-  \end{abstract}
-
-  \maketitle
-
 \section{Introduction}
@@
 \subsection{Gradual Typing and Graduality}
-  Gradual typing allows a language to have both statically-typed and dynamically-typed terms;
-  the statically-typed terms are type checked at compile time, while type checking for the 
-  dynamically-typed terms is deferred to runtime.
+One of the principal categories on which type systems of programming languages are classified
+is that of static versus dynamic type discipline.
+In static typing, the code is type-checked at compile time, while in a dynamic typing,
+the type checking is deferred to run-time. Both approaches have benefits: with static typing,
+the programmer is assured that if the program passes the type-checker, their program
+is free of type errors. Meanwhile, dynamic typing allows the programmer to rapidly prototype
+their application code without needing to commit to fixed type signatures for their functions.
 
-  Gradually-typed languages support the smooth migration from dynamic typing
+\emph{Gradually-typed languages} \cite{siek-taha06} allow for both disciplines
+to be used in the same codebase, and support interoperability between statically-typed
+and dynamically-typed code.
+This flexibility allows programmers to begin their projects in a dynamic style and
+enjoy the benefits of dynamic typing related to rapid prototyping and easy modification
+while the codebase ``solidifies''. Over time, as parts of the code become more mature
+and the programmer is more certain of what the types should be, the code can be
+\emph{gradually} migrated to a statically typed style without needing to start the project
+over in a completely differnt language.
+
+Gradually-typed languages should satisfy two intuitive properties.
+First, the interaction between the static and dynamic components of the codebase
+should be safe -- i.e., should preserve the guarantees made by the static types.
+In particular, while statically-typed code can error at runtime in a gradually-typed language,
+such an error can always be traced back to a dynamically-typed term that
+violated the typing contract imposed by statically typed code.
+% In other words, in the statically-typed portions of the codebase, type soundness must be preserved.
+Second, gradually-typed langugaes should support the smooth migration from dynamic typing
 to static typing, in that the programmer can initially leave off the
 typing annotations and provide them later without altering the meaning of the
-  program. Moreover, a key feature of gradually typed languages is that the interaction
-  between the static and dynamic components of the codebase should be safe -- i.e., should
-  preserve the guarantees made by the static types.
-  In other words, in the static portions of the codebase, type soundness must be preserved.
-
-  In a gradually-typed language, the mixing of static and dynamic code is seamless, in that
-  the dynamically typed parts are checked at runtime. This type checking occurs at the elimination
-  forms of the language (e.g., pattern matching, field reference, etc.).
-  Gradual languages are generally elaborated to a \emph{cast calculus}, in which the dynamic
-  type checking is made explicit through the insertion of \emph{type casts}.
-
-  ...
-
-  % Up and down casts
-  There is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that $A$ is a \emph{more
-  precise} type than $B$.
-  There a dynamic type $\dyn$ with the property that $A \ltdyn \dyn$ for all $A$.
-
-  If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
-  and a term $N$ of type $B$ may be \emph{down}casted to $A$, written $\dn A B N$.
-  Upcasts always succeed, while downcasts may fail at runtime.
+program.
 
 % Graduality property
-  A key property that any gradually-typed language should satisfy is the \emph{dynamic gradual guarantee},
-  originally defined by Siek, Vitousek, Cimini, and Boyland \cite{TODO}.
-  New et al refer to this property as \emph{graduality}, by analogy with parametricity.
-  Intuitively, graduality says that in going from a dynamic to static style
-  should not introduce changes in the meaning of the program.
-  More specifically, making the types more precise (replacing $\dyn$ with a more specific type
-  such as integers) will either result in the same behavior as the less precise program,
-  or cause a dynamic type error.
+Formally speaking, gradually typed languages should satisfy the 
+\emph{dynamic gradual guarantee}, originally defined by Siek, Vitousek, Cimini,
+and Boyland \cite{siek_et_al:LIPIcs:2015:5031}.
+This property is also referred to as \emph{graduality}, by analogy with parametricity.
+Intuitively, graduality says that going from a dynamic to static style should not
+introduce changes in the meaning of the program.
+More specifically, making the types more precise by adding typing annotations
+% (replacing $\dyn$ with a more specific type
+%such as integers)
+will either result in the same behavior as the original, less precise program,
+or will result in a type error.
 
-  % Observational equivalence and approximation
-  % TODO
+\subsection{Current Approaches to Proving Graduality}
+Current approaches to proving graduality include the methods of Abstracting Gradual Typing
+\cite{garcia-clark-tanter2016} and the formal tools of the Gradualier \cite{cimini-siek2016}.
+These allow the language developer to start with a statically typed langauge and derive a
+gradually typed language that satisfies the gradual guarantee. The downside to these approaches
+is that the semantics of the resulting languages are too lazy: the frameworks consider only
+the $\beta$ rules and not the $\eta$ equalities. Furthermore, while these frameworks do prove
+graduality, they do not show the correctness of the equational theory, which is equally important
+to sound gradual typing. For example, programmers often refactor their code without thinking about
+whether the refactoring has broken the semantics of the program. 
+It is the validity of the laws in the equational theory that guarantees that such refactorings are sound. 
+Similarly, correctness of compiler optimizations rests on the validity of the corresponding equations
+from the equational theory. It is therefore important that the langages that claim to be gradually typed
+have provably correct equational theories.
 
-  \subsection{Prior Approach}
+%The downside is that
+%not all gradually typed languages can be derived from these frameworks, and moreover, in both
+%approaches the semantics is derived from the static type system as opposed to the alternative
+%in which the semantics determines the type checking. Without a clear semantic interpretation of type
+%dynamism, it becomes difficult to extend these techniques to new language features such as polymorphism.
+
 % Proving graduality via LR
-  The usual approach to proving graduality uses the technique of \emph{logical relations}.
-  Specifially, a logical relation is constructed and showed to be sound with respect to
-  observational approximation. Because the dynamic type is modeled as a recursive type,
-  the logical relation needs to be paramterized by natural numbers, a so-called
-  step-indexed logical relation. Reasoning about step-indexed logical relations
-  is tedious and error-prone, and there are several very subtle aspects that must
-  be taken into account in the proofs.
+New and Ahmed \cite{new-ahmed2018}
+have developed a semantic approach to specifying type dynamism in terms of
+\emph{embedding-projection pairs}, which allows for a particularly elegant formulation of the
+gradual guarantee.
+Moreover, their axiomatic account of program equivalence allows for type-based reasoning
+about program equivalences.
+%
+In this approach, a logical relation is constructed and shown to be sound with respect to
+the notion of observational approximation that specifies when one program is more precise than another.
+The downside of this approach is that each new language requires a different logical relation
+to prove graduality. Furthermore, the logical relations tend to be quite complicated due
+to a technical requirement known as \emph{step-indexing}.
+As a result, developments using this approach tend to require vast effort, with the
+corresponding technical reports having 50+ pages of proofs.
 
-  % Difficulties in prior semantics
-  % TODO
+An alternative approach, which we investigate in this paper, is provided by
+\emph{synthetic guarded domain theory}.
+The tecnhiques of synthetic guarded domain theory allow us to internalize the
+step-index reasoning normally required in logical relations proofs of graduality,
+ultimately allowing us to specify the logical relation in a manner that looks nearly
+identical to a typical, non-step-indexed logical relation.
 
-  % - two separate logical relations for expressions
-  % - transitivity
+In this paper, we report on work we have done to mechanize proofs of graduality and
+correctness of equational theories using SGDT techniques in Agda.
+Our goal in this work is to mechanize these proofs in a reusable way,
+thereby providing a framework to use to more easily and
+conveniently prove that existing languages satsify graduality and have
+sound equational theories. Moreover, the aim is for designers of new languages
+to utlize the framework to facilitate the design of new provably-correct
+gradually-typed languages with nontrivial features.
 
-  % synthetic guarded domain theory, denotational semantics therein
-  \subsection{Synthetic Guarded Domain Theory}
-  One way to avoid the tedious reasoning associated with step-indexing is to work
-  axiomatically inside of a logical system that can reason about non-well-founded recursive
-  constructions while abstracting away the specific details of step-indexing required
-  if we were working analytically.
-  The system that proves useful fot this purpose is called \emph{synthetic guarded
-  domain theory}, or SGDT for short. We provide a brief overview here, but more
-  details can be found in \cite{TODO}.
+\subsection{Proving Graduality in SGDT}
+\textbf{TODO:} This section should probably be moved to after the relevant background has been introduced.
 
-  In SGDT, there is the notion of a \emph{clock}, an abstract type that tracks the
-  passage of time through \emph{ticks}. Given a clock $k$, a tick $t : \tick k$ serves
-  as evidence that one unit of time has passed according to the clock $k$.
+% TODO introduce the idea of cast calculus and explicit casts?
 
-  SGDT also introduces a modality $\later : \type \to \type$, (pronounced ``later'').
-  Given a clock $k$ and type $T$, the type $\later_k T$ represents a value of type
-  $T$ one unit of time in the future according to clock $k$.
-  The type $\later T$ is represented as a function from ticks of a clock $k$ to $T$.
-  There is an axiom called $\lob$-induction which is stated as follows:
+In this paper, we will utilize SGDT techniques to prove graduality for a particularly
+simple gradually-typed cast calculus, the gradually-typed lambda calculus.
+This is the usual simply-typed lambda calculus with a dynamic type $\dyn$ such that
+$A \ltdyn\, \dyn$ for all types $A$, as well as upcasts and downcasts between any types
+$A$ and $B$ such that $A \ltdyn B$. The complete definition will be provided in
+Section \ref{sec:GTLC}.
+The graduality theorem is shown below.
 
-  \[
-    \lob : \forall T, (\later T \to T) \to T.
-  \]
+% \begin{theorem}[Graduality]
+%   If $M \ltdyn N : \nat$, then either:
+%   \begin{enumerate}
+%     \item $M = \mho$
+%     \item $M = N = \zro$
+%     \item $M = N = \suc n$ for some $n$
+%     \item $M$ and $N$ diverge
+%   \end{enumerate}
+% \end{theorem}
 
-  That is, to construct a term of type $T$, it suffices to assume that we have access to
-  such a term ``later'' and use that to help us build a term ``now''.
+\begin{theorem}[Graduality]
+  If $\cdot \vdash M \ltdyn N : \nat$, then
+  \begin{enumerate}
+    \item $M \Downarrow$ iff $N \Downarrow$
+    \item If $M \Downarrow v_?$ and $N \Downarrow v'_?$ then either $v_? = \mho$, or $v_? = v'_?$.
+  \end{enumerate}
+\end{theorem}
 
-  In particular, this axiom applies to propositions $P : \texttt{Prop}$.
+Details can be found in later sections, but we provide a brief explanation of the terminology and notation:
+
+\begin{itemize}
+  \item $M \ltdyn N : \nat$ means $M$ and $N$ are terms of type $\nat$ such that
+  $M$ is ``syntactically more precise'' than $N$, or equivalently, $N$ is 
+  ``more dynamic'' than $M$. Intuitively this means that $M$ and $N$ are the
+  same except that in some places where $M$ has explicit typing annotations,
+  $N$ has $\dyn$ instead.
+  \item $\cdot \Downarrow$ is a relation on terms that is defined such that $M \Downarrow$ means
+  that $M$ terminates, either with a run-time error or a value $n$ of type $\nat$.
+  \item $\mho$ is a syntactic representation of a run-time type error, which
+  happens, for example, when a programmer tries to call a function with a value whose type
+  is found to be incompatible with the argument type of the function.
+  \item $v_?$ is shorthand for the syntactic representation of a term that is either equal to
+  $\mho$, or equal to the syntactic representation of a value $n$ of type $\nat$.
+\end{itemize}
+
+% We also should be able to show that $\mho$, $\zro$, and $\suc\, N$ are not equal.
+
+Our first step toward proving graduality is to formulate an \emph{step-sensitive}, or \emph{intensional},
+gradual lambda calculus, which we call $\intlc$, in which the computation steps taken by a term are made explicit.
+The ``normal'' gradual lambda calculus for which we want to prove graduality will be called the
+\emph{step-insensitive}, or \emph{extensional}, gradual lambda calculus, denoted $\extlc$.
+We will define an erasure function $\erase{\cdot} : \intlc \to \extlc$ which takes a program in the intensional lambda calculus
+and ``forgets'' the syntactic information about the steps to produce a term in the extensional calculus.
+
+Every term $M_e$ in $\extlc$ will have a corresponding program $M_i$ in $\intlc$ such that
+$\erase{M_i} = M_e$. Moreover, we will show that if $M_e \ltdyn_e M_e'$ in the extensional theory,
+then there exists terms $M_i$ and $M_i'$ such that $\erase{M_i}=M_e$, $\erase{M_i'}=M_e'$ and
+$M_i \ltdyn_i M_i'$ in the intensional theory.
+
+We formulate and prove an analogous graduality theorem for the intensional lambda calculus.
+We define an interpretation of the intensional lambda calculus into a model in which we prove
+various results. Using the observation above, given $M_e \ltdyn M_e' : \nat$, we can find
+intensional programs $M_i$ and $M_i'$ that erase to them and are such that $M_i \ltdyn M_i'$.
+We will then apply the intensional graduality theorem to $M_i$ and $M_i'$, and translate the result
+back to $M_e$ and $M_e'$.
+
+\subsection{Contributions}
+Our main contribution is a reusable framework in Guarded Cubical Agda for developing machine-checked proofs
+of graduality of a cast calculus.
+To demonstrate the feasability and utility of our approach, we have used the framework to prove
+graduality for the simply-typed gradual lambda calculus. Along the way, we have developed an ``intensional" theory
+of graduality that is of independent interest.
 
 \subsection{Overview of Remainder of Paper}
+
+In Section \ref{sec:technical-background}, we provide technical background on gradually typed languages and
+on synthetic guarded domain theory.
+% 
 In Section \ref{sec:GTLC}, we introduce the gradually-typed cast calculus
 for which we will prove graduality. Important here are the notions of syntactic
-  type precision and term precision.
-
+type precision and term precision. We introduce both the \emph{extensional} gradual lambda calculus
+($\extlc$) and the \emph{intensional} gradual lambda calculus ($\intlc$).
+%
 In Section \ref{sec:domain-theory}, we define several fundamental constructions
 internal to SGDT that will be needed when we give a denotational semantics to
-  the language. This includes the notion of Predomains as well as the concept
+our intensional lambda calculus.
+This includes the notion of Predomains as well as the concept
 of EP-Pairs.
-
-  In Section \ref{Semantics}, we define the denotational semantics for the gradually-typed
-  language
-
-\section{GTLC}\label{sec:GTLC}
-
-\subsection{Syntax}
-
-\subsection{Type Precision}
-
-\subsection{Term Precision}
-
-\section{Domain-Theoretic Constructions}\label{sec:domain-theory}
-
-\subsection{Predomains}
-
-\subsection{Weak Bisimilarity Relation}
-
-\subsection{EP-Pairs}
-
-\section{Semantics}\label{sec:semantics}
-
-\subsection{Types as Predomains}
-
-\subsection{Terms as Monotone Functions}
-
-\subsection{Type Precision as EP-Pairs}
-
-\subsection{Term Precision}
-% Homogeneous vs heterogeneous term precision
-
-\section{Graduality}\label{sec:graduality}
-
-\section{Discussion}\label{sec:discussion}
-
-\subsection{Synthetic Ordering}
-
-While the use of synthetic guarded domain theory allows us to very
-conveniently work with non-well-founded recursive constructions while
-abstracting away the precise details of step-indexing, we do work with
-the error ordering in a mostly analytic fashion in that gradual types
-are interpreted as sets equipped with an ordering relation, and all
-terms must be proven to be monotone.
 %
-It is possible that a combination of synthetic guarded domain theory
-with \emph{directed} type theory would allow for an a synthetic
-treatment of the error ordering as well.
-
-\end{document}
+In Section \ref{sec:semantics}, we define the denotational semantics for the
+intensional gradually-typed lambda calculus using the domain theoretic constructions in the
+previous section.
+%
+In Section \ref{sec:graduality}, we outline in more detail the proof of graduality for the
+extensional gradual lambda calculus, which will make use of prove properties we prove about
+ the intensional gradual lambda calculus.
+%
+In Section \ref{sec:discussion}, we discuss the benefits and drawbacks to our approach in comparison
+to the traditional step-indexing approach, as well as possibilities for future work.

### 2023-04-25 paper-new/paper.tex
work on paper (changes to intro, work on description of syntax)

--- 2023-01-12 paper-new/paper.tex
@@
 \begin{abstract}
-    We develop a denotational semantics for a gradually typed language
-    with effects that is adequate and proves the graduality theorem.
+    We develop a denotational semantics for a simple gradually typed language
+    that is adequate and proves the graduality theorem.
     %
@@
 
-  \section{Introduction}
+\input{intro}
 
-  % gradual typing, graduality
-  \subsection{Gradual Typing and Graduality}
-  Gradual typing allows a language to have both statically-typed and dynamically-typed terms;
-  the statically-typed terms are type checked at compile time, while type checking for the 
-  dynamically-typed terms is deferred to runtime.
+\input{technical-background}
 
-  Gradually-typed languages support the smooth migration from dynamic typing
-  to static typing, in that the programmer can initially leave off the
-  typing annotations and provide them later without altering the meaning of the
-  program. Moreover, a key feature of gradually typed languages is that the interaction
-  between the static and dynamic components of the codebase should be safe -- i.e., should
-  preserve the guarantees made by the static types.
-  In other words, in the static portions of the codebase, type soundness must be preserved.
+\input{syntax}
 
-  In a gradually-typed language, the mixing of static and dynamic code is seamless, in that
-  the dynamically typed parts are checked at runtime. This type checking occurs at the elimination
-  forms of the language (e.g., pattern matching, field reference, etc.).
-  Gradual languages are generally elaborated to a \emph{cast calculus}, in which the dynamic
-  type checking is made explicit through the insertion of \emph{type casts}.
+\input{semantics}
 
-  ...
-
-  % Up and down casts
-  There is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that $A$ is a \emph{more
-  precise} type than $B$.
-  There a dynamic type $\dyn$ with the property that $A \ltdyn \dyn$ for all $A$.
-
-  If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
-  and a term $N$ of type $B$ may be \emph{down}casted to $A$, written $\dn A B N$.
-  Upcasts always succeed, while downcasts may fail at runtime.
-
-  % Graduality property
-  A key property that any gradually-typed language should satisfy is the \emph{dynamic gradual guarantee},
-  originally defined by Siek, Vitousek, Cimini, and Boyland \cite{TODO}.
-  New et al refer to this property as \emph{graduality}, by analogy with parametricity.
-  Intuitively, graduality says that in going from a dynamic to static style
-  should not introduce changes in the meaning of the program.
-  More specifically, making the types more precise (replacing $\dyn$ with a more specific type
-  such as integers) will either result in the same behavior as the less precise program,
-  or cause a dynamic type error.
-
-  % Observational equivalence and approximation
-  % TODO
-
-  \subsection{Prior Approach}
-  % Proving graduality via LR
-  The usual approach to proving graduality uses the technique of \emph{logical relations}.
-  Specifially, a logical relation is constructed and showed to be sound with respect to
-  observational approximation. Because the dynamic type is modeled as a recursive type,
-  the logical relation needs to be paramterized by natural numbers, a so-called
-  step-indexed logical relation. Reasoning about step-indexed logical relations
-  is tedious and error-prone, and there are several very subtle aspects that must
-  be taken into account in the proofs.
-
-  % Difficulties in prior semantics
-  % TODO
-
-  % - two separate logical relations for expressions
-  % - transitivity
-
-  % synthetic guarded domain theory, denotational semantics therein
-  \subsection{Synthetic Guarded Domain Theory}
-  One way to avoid the tedious reasoning associated with step-indexing is to work
-  axiomatically inside of a logical system that can reason about non-well-founded recursive
-  constructions while abstracting away the specific details of step-indexing required
-  if we were working analytically.
-  The system that proves useful fot this purpose is called \emph{synthetic guarded
-  domain theory}, or SGDT for short. We provide a brief overview here, but more
-  details can be found in \cite{TODO}.
-
-  In SGDT, there is the notion of a \emph{clock}, an abstract type that tracks the
-  passage of time through \emph{ticks}. Given a clock $k$, a tick $t : \tick k$ serves
-  as evidence that one unit of time has passed according to the clock $k$.
-
-  SGDT also introduces a modality $\later : \type \to \type$, (pronounced ``later'').
-  Given a clock $k$ and type $T$, the type $\later_k T$ represents a value of type
-  $T$ one unit of time in the future according to clock $k$.
-  The type $\later T$ is represented as a function from ticks of a clock $k$ to $T$.
-  There is an axiom called $\lob$-induction which is stated as follows:
-
-  \[
-    \lob : \forall T, (\later T \to T) \to T.
-  \]
-
-  That is, to construct a term of type $T$, it suffices to assume that we have access to
-  such a term ``later'' and use that to help us build a term ``now''.
-
-  In particular, this axiom applies to propositions $P : \texttt{Prop}$.
-
-  \subsection{Overview of Remainder of Paper}
-  In Section \ref{sec:GTLC}, we introduce the gradually-typed cast calculus
-  for which we will prove graduality. Important here are the notions of syntactic
-  type precision and term precision.
-
-  In Section \ref{sec:domain-theory}, we define several fundamental constructions
-  internal to SGDT that will be needed when we give a denotational semantics to
-  the language. This includes the notion of Predomains as well as the concept
-  of EP-Pairs.
-
-  In Section \ref{Semantics}, we define the denotational semantics for the gradually-typed
-  language
-
-\section{GTLC}\label{sec:GTLC}
-
-\subsection{Syntax}
-
-\subsection{Type Precision}
-
-\subsection{Term Precision}
-
-\section{Domain-Theoretic Constructions}\label{sec:domain-theory}
-
-\subsection{Predomains}
-
-\subsection{Weak Bisimilarity Relation}
-
-\subsection{EP-Pairs}
-
-\section{Semantics}\label{sec:semantics}
-
-\subsection{Types as Predomains}
-
-\subsection{Terms as Monotone Functions}
-
-\subsection{Type Precision as EP-Pairs}
-
-\subsection{Term Precision}
-% Homogeneous vs heterogeneous term precision
-
-\section{Graduality}\label{sec:graduality}
+\input{graduality}
 
@@
 
-\subsection{Synthetic Ordering}
+% \subsection{Synthetic Ordering}
 
-While the use of synthetic guarded domain theory allows us to very
-conveniently work with non-well-founded recursive constructions while
-abstracting away the precise details of step-indexing, we do work with
-the error ordering in a mostly analytic fashion in that gradual types
-are interpreted as sets equipped with an ordering relation, and all
-terms must be proven to be monotone.
-%
-It is possible that a combination of synthetic guarded domain theory
-with \emph{directed} type theory would allow for an a synthetic
-treatment of the error ordering as well.
+% While the use of synthetic guarded domain theory allows us to very
+% conveniently work with non-well-founded recursive constructions while
+% abstracting away the precise details of step-indexing, we do work with
+% the error ordering in a mostly analytic fashion in that gradual types
+% are interpreted as sets equipped with an ordering relation, and all
+% terms must be proven to be monotone.
+% %
+% It is possible that a combination of synthetic guarded domain theory
+% with \emph{directed} type theory would allow for an a synthetic
+% treatment of the error ordering as well.
+
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{references}
 

### 2023-04-25 paper-new/semantics.tex
work on paper (changes to intro, work on description of syntax)

\section{Domain-Theoretic Constructions}\label{sec:domain-theory}

In this section, we discuss the fundamental objects of the model into which we will embed
the intensional lambda calculus and inequational theory. It is important to remember that
the constructions in this section are entirely independent of the syntax described in the
previous section; the notions defined here exist in their own right as purely mathematical
constructs. In the next section, we will link the syntax and semantics via an interpretation
function.

\subsection{The Lift Monad}

When thinking about how to model intensional gradually-typed programs, we must consider
their possible behaviors. On the one hand, we have \emph{failure}: a program may fail
at run-time because of a type error. In addition to this, a program may ``think'',
i.e., take a step of computation. If a program thinks forever, then it never returns a value,
so we can think of the idea of thinking as a way of intensionally modelling \emph{partiality}.

With this in mind, we can describe a semantic object that models these behaviors: a monad
for embedding computations that has cases for failure and ``thinking''.
Previous work has studied such a construct in the setting of the latter, called the lift
monad \cite{mogelberg-paviotti2016}; here, we augment it with the additional effect of failure.

For a type $A$, we define the \emph{lift monad with failure} $\li A$, which we will just call
the \emph{lift monad}, as the following datatype:

\begin{align*}
  \li A &:= \\
  &\eta \colon A \to \li A \\
  &\mho \colon \li A \\
  &\theta \colon \later (\li A) \to \li A
\end{align*}

Unless otherwise mentioned, all constructs involving $\later$ or $\fix$
are understood to be with repsect to a fixed clock $k$. So for the above, we really have for each
clock $k$ a type $\li^k A$ with respect to that clock.

Formally, the lift monad $\li A$ is defined as the solution to the guarded recursive type equation

\[ \li A \cong A + 1 + \later \li A. \]

An element of $\li A$ should be viewed as a computation that can either (1) return a value (via $\eta$),
(2) raise an error and stop (via $\mho$), or (3) think for a step (via $\theta$).
%
Notice there is a computation $\fix \theta$ of type $\li A$. This represents a computation
that thinks forever and never returns a value.

Since we claimed that $\li A$ is a monad, we need to define the monadic operations
and show that they repect the monadic laws. The return is just $\eta$, and extend
is defined via by guarded recursion by cases on the input.
% It is instructive to give at least one example of a use of guarded recursion, so
% we show below how to define extend:
% TODO
%
%
Verifying that the monadic laws hold requires \lob-induction and is straightforward.

The lift monad has the following universal property. Let $f$ be a function from $A$ to $B$,
where $B$ is a $\later$-algebra, i.e., there is $\theta_B \colon \later B \to B$.
Further suppose that $B$ is also an ``error-algebra'', that is, an algebra of the
constant functor $1 \colon \text{Type} \to \text{Type}$ mapping all types to Unit.
This latter statement amounts to saying that there is a map $\text{Unit} \to B$, so $B$ has a
distinguished ``error element" $\mho_B \colon B$.

Then there is a unique homomorphism of algebras $f' \colon \li A \to B$ such that
$f' \circ \eta = f$. The function $f'(l)$ is defined via guarded fixpoint by cases on $l$. 
In the $\mho$ case, we simply return $\mho_B$.
In the $\theta(\tilde{l})$ case, we will return

\[\theta_B (\lambda t . (f'_t \, \tilde{l}_t)). \]

Recalling that $f'$ is a guaded fixpoint, it is available ``later'' and by
applying the tick we get a function we can apply ``now''; for the argument,
we apply the tick to $\tilde{l}$ to get a term of type $\li A$.

%\subsubsection{Model-Theoretic Description}
%We can describe the lift monad in the topos of trees model as follows.

\subsection{Predomains}

The next important construction is that of a \emph{predomain}. A predomain is intended to
model the notion of error ordering that we want terms to have. Thus, we define a predomain $A$
as a partially-ordered set, which consists of a type which we denote $\ty{A}$ and a reflexive,
transitive, and antisymmetric relation $\le_P$ on $A$.

For each type we want to represent, we define a predomain for the corresponding semantic
type. For instance, we define a predomain for natural numbers, a predomain for the
dynamic type, a predomain for functions, and a predomain for the lift monad. We
describe each of these below.

We define monotone functions between predomain as expected. Given predomains
$A$ and $B$, we write $f \colon A \monto B$ to indicate that $f$ is a monotone
function from $A$ to $B$, i.e, for all $a_1 \le_A a_2$, we have $f(a_1) \le_B f(a_2)$.

\begin{itemize}
  \item There is a predomain Nat for natural numbers, where the ordering is equality.

  \item There is a predomain Dyn to represent the dynamic type. The underlying type
  for this predomain is defined by guarded fixpoint to be such that
  $\ty{\Dyn} \cong \mathbb{N}\, +\, \later (\ty{\Dyn} \monto \ty{\Dyn})$.
  This definition is valid because the occurrences of Dyn are guarded by the $\later$.
  The ordering is defined via guarded recursion by cases on the argument, using the
  ordering on $\mathbb{N}$ and the ordering on monotone functions described below.

  \item For a predomain $A$, there is a predomain $\li A$ for the ``lift'' of $A$
  using the lift monad. We use the same notation for $\li A$ when $A$ is a type
  and $A$ is a predomain, since the context should make clear which one we are referring to.
  The underling type of $\li A$ is simply $\li \ty{A}$, i.e., the lift of the underlying
  type of $A$.
  The ordering on $\li A$ is the ``step-sensitive error-ordering'' which we describe in
  \ref{subsec:lock-step}.

  \item For predomains $A_i$ and $A_o$, we form the predomain of monotone functions
  from $A_i$ to $A_o$, which we denote by $A_i \To A_o$.
  The ordering is such that $f$ is below $g$ if for all $a$ in $\ty{A_i}$, we have
  $f(a)$ is below $g(a)$ in the ordering for $A_o$. 
\end{itemize}

\subsection{Step-Sensitive Error Ordering}\label{subsec:lock-step}

As mentioned, the ordering on the lift of a predomain $A$ is called the
\emph{step-sensitive error-ordering} (also called ``lock-step error ordering''),
the idea being that two computations $l$ and $l'$ are related if they are in
lock-step with regard to their intensional behavior, up to $l$ erroring.
Formally, we define this ordering as follows:

\begin{itemize}
  \item 	$\eta\, x \ltls \eta\, y$ if $x \le_A y$.
  \item 	$\mho \ltls l$ for all $l$ 
  \item   $\theta\, \tilde{r} \ltls \theta\, \tilde{r'}$ if
          $\later_t (\tilde{r}_t \ltls \tilde{r'}_t)$
\end{itemize}

We also define a heterogeneous version of this ordering between the lifts of two
different predomains $A$ and $B$, parameterized by a relation $R$ between $A$ and $B$.

\subsection{Step-Insensitive Relation}

We define another ordering on $\li A$, called the ``step-insensitive ordering'' or
``weak bisimilarity'', written $l \bisim l'$.
Intuitively, we say $l \bisim l'$ if they are equivalent ``up to delays''.
We introduce the notation $x \sim_A y$ to mean $x \le_A y$ and $y \le_A x$.

The weak bisimilarity relation is defined by guarded fixpoint as follows:

\begin{align*}
  &\mho \bisim \mho \\
%
  &\eta\, x \bisim \eta\, y \text{ if } 
    x \sim_A y \\
%		
  &\theta\, \tilde{x} \bisim \theta\, \tilde{y} \text{ if } 
    \later_t (\tilde{x}_t \bisim \tilde{y}_t) \\
%	
  &\theta\, \tilde{x} \bisim \mho \text{ if } 
    \theta\, \tilde{x} = \delta^n(\mho) \text { for some $n$ } \\
%	
  &\theta\, \tilde{x} \bisim \eta\, y \text{ if }
    (\theta\, \tilde{x} = \delta^n(\eta\, x))
  \text { for some $n$ and $x : \ty{A}$ such that $x \sim_A y$ } \\
%
  &\mho \bisim \theta\, \tilde{y} \text { if } 
    \theta\, \tilde{y} = \delta^n(\mho) \text { for some $n$ } \\
%	
  &\eta\, x \bisim \theta\, \tilde{y} \text { if }
    (\theta\, \tilde{y} = \delta^n (\eta\, y))
  \text { for some $n$ and $y : \ty{A}$ such that $x \sim_A y$ }
\end{align*}

\subsection{Error Domains}

\subsection{Globalization}

Recall that in the above definitions, any occurrences of $\later$ were with
repsect to a fixed clock $k$. Intuitively, this corresponds to a step-indexed set.
It will be necessary to consider the ``globalization'' of these definitions,
i.e., the ``global'' behavior of the type over all potential time steps.
This is accomplished in the type theory by \emph{clock quantification} \cite{atkey-mcbride2013},
whereby given a type $X$ parameterized by a clock $k$, we consider the type
$\forall k. X[k]$. This corresponds to leaving the step-indexed world and passing to
the usual semantics in the category of sets.

\section{Semantics}\label{sec:semantics}

\subsection{Relational Semantics}

\subsubsection{Term Precision via the Step-Sensitive Error Ordering}
% Homogeneous vs heterogeneous term precision

% \subsection{Logical Relations Semantics}

### 2023-04-25 paper-new/technical-background.tex
work on paper (changes to intro, work on description of syntax)

\section{Technical Background}\label{sec:technical-background}

\subsection{Gradual Typing}

% Cast calculi
In a gradually-typed language, the mixing of static and dynamic code is seamless, in that
the dynamically typed parts are checked at runtime. This type checking occurs at the elimination
forms of the language (e.g., pattern matching, field reference, etc.).
Gradual languages are generally elaborated to a \emph{cast calculus}, in which the dynamic
type checking is made explicit through the insertion of \emph{type casts}.

% Up and down casts
In a cast calculus, there is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that
$A$ is a \emph{more precise} type than $B$.
There a dynamic type $\dyn$ with the property that $A \ltdyn\, \dyn$ for all $A$.
%
If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
and a term $N$ of type $B$ may be \emph{down}casted to $A$, written $\dn A B N$.
Upcasts always succeed, while downcasts may fail at runtime.
%
% Syntactic term precision
We also have a notion of \emph{syntactic term precision}.
If $A \ltdyn B$, and $M$ and $N$ are terms of type $A$ and $B$ respectively, we write
$M \ltdyn N : A \ltdyn B$ to mean that $M$ is more precise than $N$, i.e., $M$ and $N$
behave the same except that $M$ may error more.

% Modeling the dynamic type as a recursive sum type?
% Observational equivalence and approximation?

% synthetic guarded domain theory, denotational semantics therein

\subsection{Difficulties in Prior Semantics}
  % Difficulties in prior semantics

  In this work, we compare our approach to proving graduality to the approach
  introduced by New and Ahmed \cite{new-ahmed2018} which constructs a step-indexed
  logical relations model and shows that this model is sound with respect to their
  notion of contextual error approximation.

  Because the dynamic type is modeled as a non-well-founded
  recursive type, their logical relation needs to be paramterized by natural numbers
  to restore well-foundedness. This technique is known as a \emph{step-indexed logical relation}.
  Reasoning about step-indexed logical relations
  can be tedious and error-prone, and there are some very subtle aspects that must
  be taken into account in the proofs. Figure \ref{TODO} shows an example of a step-indexed logical
  relation for the gradually-typed lambda calculus.

  In particular, the prior approach of New and Ahmed requires two separate logical
  relations for terms, one in which the steps of the left-hand term are counted,
  and another in which the steps of the right-hand term are counted.
  Then two terms $M$ and $N$ are related in the ``combined'' logical relation if they are
  related in both of the one-sided logical relations. Having two separate logical relations
  complicates the statement of the lemmas used to prove graduality, becasue any statement that
  involves a term stepping needs to take into account whether we are counting steps on the left
  or the right. Some of the differences can be abstracted over, but difficulties arise for properties %/results
  as fundamental and seemingly straightforward as transitivty.

  Specifically, for transitivity, we would like to say that if $M$ is related to $N$ at
  index $i$ and $N$ is related to $P$ at index $i$, then $M$ is related to $P$ at $i$.
  But this does not actually hold: we requrie that one of the two pairs of terms
  be related ``at infinity'', i.e., that they are related at $i$ for all $i \in \mathbb{N}$.
  Which pair is required to satisfy this depends on which logical relation we are considering,
  (i.e., is it counting steps on the left or on the right),
  and so any argument that uses transitivity needs to consider two cases, one
  where $M$ and $N$ must be shown to be related for all $i$, and another where $N$ and $P$ must
  be related for all $i$. % This may not even be possible to show in some scenarios!

  % These complications introduced by step-indexing lead one to wonder whether there is a
  % way of proving graduality without relying on tedious arguments involving natural numbers.
  % An alternative approach, which we investigate in this paper, is provided by
  % \emph{synthetic guarded domain theory}, as discussed below.
  % Synthetic guarded domain theory allows the resulting logical relation to look almost
  % identical to a typical, non-step-indexed logical relation.

\subsection{Synthetic Guarded Domain Theory}
One way to avoid the tedious reasoning associated with step-indexing is to work
axiomatically inside of a logical system that can reason about non-well-founded recursive
constructions while abstracting away the specific details of step-indexing required
if we were working analytically.
The system that proves useful for this purpose is called \emph{synthetic guarded
domain theory}, or SGDT for short. We provide a brief overview here, but more
details can be found in \cite{birkedal-mogelberg-schwinghammer-stovring2011}.

SGDT offers a synthetic approach to domain theory that allows for guarded recursion
to be expressed syntactically via a type constructor $\later : \type \to \type$ 
(pronounced ``later''). The use of a modality to express guarded recursion
was introduced by Nakano \cite{Nakano2000}.
%
Given a type $A$, the type $\later A$ represents an element of type $A$
that is available one time step later. There is an operator $\nxt : A \to\, \later A$
that ``delays'' an element available now to make it available later.
We will use a tilde to denote a term of type $\later A$, e.g., $\tilde{M}$.

% TODO later is an applicative functor, but not a monad

There is a \emph{guarded fixpoint} operator

\[
  \fix : \forall T, (\later T \to T) \to T.
\]

That is, to construct a term of type $T$, it suffices to assume that we have access to
such a term ``later'' and use that to help us build a term ``now''.
This operator satisfies the axiom that $\fix f = f (\nxt (\fix f))$.
In particular, this axiom applies to propositions $P : \texttt{Prop}$; proving
a statement in this manner is known as $\lob$-induction.

The operators $\later$, $\next$, and $\fix$ described above can be indexed by objects
called \emph{clocks}. A clock serves as a reference relative to which steps are counted.
For instance, given a clock $k$ and type $T$, the type $\later^k T$ represents a value of type
$T$ one unit of time in the future according to clock $k$.
If we only ever had one clock, then we would not need to bother defining this notion.
However, the notion of \emph{clock quantification} is crucial for encoding coinductive types using guarded
recursion, an idea first introduced by Atkey and McBride \cite{atkey-mcbride2013}.

% Clocked Cubical Type Theory
\subsubsection{Ticked Cubical Type Theory}
% TODO motivation for Clocked Cubical Type Theory, e.g., delayed substitutions?

In Ticked Cubical Type Theory \cite{TODO}, there is an additional sort
called \emph{ticks}. Given a clock $k$, a tick $t : \tick k$ serves
as evidence that one unit of time has passed according to the clock $k$.
The type $\later A$ is represented as a function from ticks of a clock $k$ to $A$.
The type $A$ is allowed to depend on $t$, in which case we write $\later^k_t A$
to emphasize the dependence.

% TODO next as a function that ignores its input tick argument?

The rules for tick abstraction and application are similar to those of dependent
$\Pi$ types. 
In particular, if we have a term $M$ of type $\later^k A$, and we
have available in the context a tick $t' : \tick k$, then we can apply the tick to $M$ to get
a term $M[t'] : A[t'/t]$. We will also write tick application as $M_t$.
Conversely, if in a context $\Gamma, t : \tick k$ we have that $M$ has type $A$,
then in context $\Gamma$ we have $\lambda t.M$ has type $\later A$. % TODO dependent version?

The statements in this paper have been formalized in a variant of Agda called
Guarded Cubical Agda \cite{TODO}, an implementation of Clocked Cubical Type Theory.

% TODO axioms (clock irrelevance, tick irrelevance)?

\subsubsection{The Topos of Trees Model}

The topos of trees model provides a useful intuition for reasoning in SGDT 
\cite{birkedal-mogelberg-schwinghammer-stovring2011}.
This section presupposes knowledge of category theory and can be safely skipped without
disrupting the continuity.

The topos of trees $\calS$ is the presheaf category $\Set^{\omega^o}$. 
%
We assume a universe $\calU$ of types, with encodings for operations such
as sum types (written as $\widehat{+}$). There is also an operator 
$\laterhat \colon \later \calU \to \calU$ such that 
$\El(\laterhat(\nxt A)) =\,\, \later \El(A)$, where $\El$ is the type corresponding
to the code $A$.

An object $X$ is a family $\{X_i\}$ of sets indexed by natural numbers,
along with restriction maps $r^X_i \colon X_{i+1} \to X_i$ (see Figure \ref{fig:topos-of-trees-object}).
These should be thought of as ``sets changing over time", where $X_i$ is the view of the set if we have $i+1$
time steps to reason about it.
We can also think of an ongoing computation, with $X_i$ representing the potential results of the computation
after it has run for $i+1$ steps.

\begin{figure}
  % https://q.uiver.app/?q=WzAsNSxbMiwwLCJYXzIiXSxbMCwwLCJYXzAiXSxbMSwwLCJYXzEiXSxbMywwLCJYXzMiXSxbNCwwLCJcXGRvdHMiXSxbMiwxLCJyXlhfMCIsMl0sWzAsMiwicl5YXzEiLDJdLFszLDAsInJeWF8yIiwyXSxbNCwzLCJyXlhfMyIsMl1d
\[\begin{tikzcd}[ampersand replacement=\&]
	{X_0} \& {X_1} \& {X_2} \& {X_3} \& \dots
	\arrow["{r^X_0}"', from=1-2, to=1-1]
	\arrow["{r^X_1}"', from=1-3, to=1-2]
	\arrow["{r^X_2}"', from=1-4, to=1-3]
	\arrow["{r^X_3}"', from=1-5, to=1-4]
\end{tikzcd}\]
  \caption{An example of an object in the topos of trees.}
  \label{fig:topos-of-trees-object}
\end{figure}

A morphism from $\{X_i\}$ to $\{Y_i\}$ is a family of functions $f_i \colon X_i \to Y_i$
that commute with the restriction maps in the obvious way, that is,
$f_i \circ r^X_i = r^Y_i \circ f_{i+1}$ (see Figure \ref{fig:topos-of-trees-morphism}).

\begin{figure}
% https://q.uiver.app/?q=WzAsMTAsWzIsMCwiWF8yIl0sWzAsMCwiWF8wIl0sWzEsMCwiWF8xIl0sWzMsMCwiWF8zIl0sWzQsMCwiXFxkb3RzIl0sWzAsMSwiWV8wIl0sWzEsMSwiWV8xIl0sWzIsMSwiWV8yIl0sWzMsMSwiWV8zIl0sWzQsMSwiXFxkb3RzIl0sWzIsMSwicl5YXzAiLDJdLFswLDIsInJeWF8xIiwyXSxbMywwLCJyXlhfMiIsMl0sWzQsMywicl5YXzMiLDJdLFs2LDUsInJeWV8wIiwyXSxbNyw2LCJyXllfMSIsMl0sWzgsNywicl5ZXzIiLDJdLFs5LDgsInJeWV8zIiwyXSxbMSw1LCJmXzAiLDJdLFsyLDYsImZfMSIsMl0sWzAsNywiZl8yIiwyXSxbMyw4LCJmXzMiLDJdXQ==
\[\begin{tikzcd}[ampersand replacement=\&]
	{X_0} \& {X_1} \& {X_2} \& {X_3} \& \dots \\
	{Y_0} \& {Y_1} \& {Y_2} \& {Y_3} \& \dots
	\arrow["{r^X_0}"', from=1-2, to=1-1]
	\arrow["{r^X_1}"', from=1-3, to=1-2]
	\arrow["{r^X_2}"', from=1-4, to=1-3]
	\arrow["{r^X_3}"', from=1-5, to=1-4]
	\arrow["{r^Y_0}"', from=2-2, to=2-1]
	\arrow["{r^Y_1}"', from=2-3, to=2-2]
	\arrow["{r^Y_2}"', from=2-4, to=2-3]
	\arrow["{r^Y_3}"', from=2-5, to=2-4]
	\arrow["{f_0}"', from=1-1, to=2-1]
	\arrow["{f_1}"', from=1-2, to=2-2]
	\arrow["{f_2}"', from=1-3, to=2-3]
	\arrow["{f_3}"', from=1-4, to=2-4]
\end{tikzcd}\]
  \caption{An example of a morphism in the topos of trees.}
  \label{fig:topos-of-trees-morphism}
\end{figure}

The type operator $\later$ is defined on an object $X$ by
$(\later X)_0 = 1$ and $(\later X)_{i+1} = X_i$.
The restriction maps are given by $r^\later_0 =\, !$, where $!$ is the
unique map into $1$, and $r^\later_{i+1} = r^X_i$.
The morphism $\nxt^X \colon X \to \later X$ is defined pointwise by
$\nxt^X_0 =\, !$, and $\nxt^X_{i+1} = r^X_i$. It is easily checked that
this satisfies the commutativity conditions required of a morphism in $\calS$.
%
Given a morphism $f \colon \later X \to X$, i.e., a family of functions
$f_i \colon (\later X)_i \to X_i$ that commute with the restrictions in the appropriate way,
we define $\fix(f) \colon 1 \to X$ pointwise
by $\fix(f)_i = f_{i} \circ \dots \circ f_0$.
This can be visualized as a diagram in the category of sets as shown in
Figure \ref{fig:topos-of-trees-fix}.
% Observe that the fixpoint is a \emph{global element} in the topos of trees.
% Global elements allow us to view the entire computation on a global level.

% https://q.uiver.app/?q=WzAsOCxbMSwwLCJYXzAiXSxbMiwwLCJYXzEiXSxbMywwLCJYXzIiXSxbMCwwLCIxIl0sWzAsMSwiWF8wIl0sWzEsMSwiWF8xIl0sWzIsMSwiWF8yIl0sWzMsMSwiWF8zIl0sWzMsNCwiZl8wIl0sWzAsNSwiZl8xIl0sWzEsNiwiZl8yIl0sWzIsNywiZl8zIl0sWzAsMywiISIsMl0sWzEsMCwicl5YXzAiLDJdLFsyLDEsInJeWF8xIiwyXSxbNSw0LCJyXlhfMCJdLFs2LDUsInJeWF8xIl0sWzcsNiwicl5YXzIiXV0=
% \[\begin{tikzcd}[ampersand replacement=\&]
% 	1 \& {X_0} \& {X_1} \& {X_2} \\
% 	{X_0} \& {X_1} \& {X_2} \& {X_3}
% 	\arrow["{f_0}", from=1-1, to=2-1]
% 	\arrow["{f_1}", from=1-2, to=2-2]
% 	\arrow["{f_2}", from=1-3, to=2-3]
% 	\arrow["{f_3}", from=1-4, to=2-4]
% 	\arrow["{!}"', from=1-2, to=1-1]
% 	\arrow["{r^X_0}"', from=1-3, to=1-2]
% 	\arrow["{r^X_1}"', from=1-4, to=1-3]
% 	\arrow["{r^X_0}", from=2-2, to=2-1]
% 	\arrow["{r^X_1}", from=2-3, to=2-2]
% 	\arrow["{r^X_2}", from=2-4, to=2-3]
% \end{tikzcd}\]

% \begin{figure}
%   % https://q.uiver.app/?q=WzAsMTksWzEsMiwiWF8wIl0sWzIsMywiWF8xIl0sWzMsMSwiMSJdLFswLDEsIjEiXSxbMCwyLCJYXzAiXSxbMSwzLCJYXzEiXSxbMSwxLCIxIl0sWzIsMSwiMSJdLFsyLDIsIlhfMCJdLFsyLDQsIlhfMiJdLFszLDIsIlhfMCJdLFszLDMsIlhfMSJdLFszLDQsIlhfMiJdLFszLDUsIlhfMyJdLFs0LDIsIlxcY2RvdHMiXSxbMCwwLCJcXGZpeChmKV8wIl0sWzEsMCwiXFxmaXgoZilfMSJdLFsyLDAsIlxcZml4KGYpXzIiXSxbMywwLCJcXGZpeChmKV8zIl0sWzMsNCwiZl8wIl0sWzAsNSwiZl8xIl0sWzYsMCwiZl8wIl0sWzcsOCwiZl8wIl0sWzgsMSwiZl8xIl0sWzEsOSwiZl8yIl0sWzIsMTAsImZfMCJdLFsxMCwxMSwiZl8xIl0sWzExLDEyLCJmXzIiXSxbMTIsMTMsImZfMyJdXQ==
%   \[\begin{tikzcd}[ampersand replacement=\&]
%     {\fix(f)_0} \& {\fix(f)_1} \& {\fix(f)_2} \& {\fix(f)_3} \\
%     1 \& 1 \& 1 \& 1 \\
%     {X_0} \& {X_0} \& {X_0} \& {X_0} \& \cdots \\
%     \& {X_1} \& {X_1} \& {X_1} \\
%     \&\& {X_2} \& {X_2} \\
%     \&\&\& {X_3}
%     \arrow["{f_0}", from=2-1, to=3-1]
%     \arrow["{f_1}", from=3-2, to=4-2]
%     \arrow["{f_0}", from=2-2, to=3-2]
%     \arrow["{f_0}", from=2-3, to=3-3]
%     \arrow["{f_1}", from=3-3, to=4-3]
%     \arrow["{f_2}", from=4-3, to=5-3]
%     \arrow["{f_0}", from=2-4, to=3-4]
%     \arrow["{f_1}", from=3-4, to=4-4]
%     \arrow["{f_2}", from=4-4, to=5-4]
%     \arrow["{f_3}", from=5-4, to=6-4]
%   \end{tikzcd}\]
%   \caption{The first few approximations to the guarded fixpoint of $f$.}
%   \label{fig:topos-of-trees-fix-approx}
% \end{figure}

\begin{figure}
  % https://q.uiver.app/?q=WzAsNixbMywwLCIxIl0sWzAsMiwiWF8wIl0sWzIsMiwiWF8xIl0sWzQsMiwiWF8yIl0sWzYsMiwiWF8zIl0sWzgsMiwiXFxkb3RzIl0sWzIsMSwicl5YXzAiXSxbNCwzLCJyXlhfMiJdLFswLDEsIlxcZml4KGYpXzAiLDFdLFswLDIsIlxcZml4KGYpXzEiLDFdLFswLDMsIlxcZml4KGYpXzIiLDFdLFswLDQsIlxcZml4KGYpXzMiLDFdLFswLDUsIlxcY2RvdHMiLDMseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMywyLCJyXlhfMSJdLFs1LDQsInJeWF8zIl1d
  \[\begin{tikzcd}[ampersand replacement=\&,column sep=2.25em]
    \&\&\& 1 \\
    \\
    {X_0} \&\& {X_1} \&\& {X_2} \&\& {X_3} \&\& \dots
    \arrow["{r^X_0}", from=3-3, to=3-1]
    \arrow["{r^X_2}", from=3-7, to=3-5]
    \arrow["{\fix(f)_0}"{description}, from=1-4, to=3-1]
    \arrow["{\fix(f)_1}"{description}, from=1-4, to=3-3]
    \arrow["{\fix(f)_2}"{description}, from=1-4, to=3-5]
    \arrow["{\fix(f)_3}"{description}, from=1-4, to=3-7]
    \arrow["\cdots"{marking}, draw=none, from=1-4, to=3-9]
    \arrow["{r^X_1}", from=3-5, to=3-3]
    \arrow["{r^X_3}", from=3-9, to=3-7]
  \end{tikzcd}\]
  \caption{The guarded fixpoint of $f$.}
  \label{fig:topos-of-trees-fix}
\end{figure}

% TODO global elements?

### 2023-04-25 paper-new/intro.tex
changes to wording

@@
 \subsection{Gradual Typing and Graduality}
-One of the principal categories on which type systems of programming languages are classified
-is that of static versus dynamic type discipline.
-In static typing, the code is type-checked at compile time, while in a dynamic typing,
+One of the principal ways of categorizing type systems of programming languages is
+whether they are \emph{static} or \emph{dynamic}.
+In static typing, the code is type-checked at compile time, while in dynamic typing,
 the type checking is deferred to run-time. Both approaches have benefits: with static typing,
@@
 
-\emph{Gradually-typed languages} \cite{siek-taha06} allow for both disciplines
-to be used in the same codebase, and support interoperability between statically-typed
+\emph{Gradually-typed languages} \cite{siek-taha06} allow for both static and dynamic typing disciplines
+to be used in the same codebase, and support smooth interoperability between statically-typed
 and dynamically-typed code.
@@
 the $\beta$ rules and not the $\eta$ equalities. Furthermore, while these frameworks do prove
-graduality, they do not show the correctness of the equational theory, which is equally important
-to sound gradual typing. For example, programmers often refactor their code without thinking about
-whether the refactoring has broken the semantics of the program. 
+graduality of the resulting languages, they do not show the correctness of the equational theory,
+which is equally important to sound gradual typing. For example, programmers often refactor their
+code without thinking about whether the refactoring has broken the semantics of the program. 
 It is the validity of the laws in the equational theory that guarantees that such refactorings are sound. 
@@
 \subsection{Proving Graduality in SGDT}
-\textbf{TODO:} This section should probably be moved to after the relevant background has been introduced.
+\textbf{TODO:} This section should probably be moved to after the relevant background has been introduced?
 

### 2023-06-04 paper.tex
some writing

@@
 \usepackage{mathpartir}
+\usepackage{stmaryrd}
 \usepackage{tikz-cd}
@@
 
-\title{Denotational Semantics for Gradual Typing in Synthetic Guarded Domain Theory}
+\newcommand{\sem}[1]{\llbracket{}{#1}\rrbracket{}}
+\newcommand{\opsemGMNV}[4]{#1 \vDash #2 \Downarrow^{#3} #4}
+\newcommand{\opsemMV}[2]{\opsemGMNV \gamma {#1} n {#2}}
+\newcommand{\opsemMNV}[3]{\opsemGMNV \gamma {#1} {#2} {#3}}
+
+\title{Compositional Mechanized Metatheory for Gradual Typing using Synthetic Guarded Domain Theory}
 \author{Eric Giovannini}
@@
 \begin{abstract}
-  We develop a denotational semantics for a gradually typed language
-  with effects that is adequate and proves the graduality theorem.
-  %
-  The denotational semantics is constructed using \emph{synthetic
-  guarded domain theory} working in a type theory with a later
-  modality and clock quantification.
-  %
-  This provides a remarkably simple presentation of the semantics,
-  where gradual types are interpreted as ordinary types in our ambient
-  type theory equipped with an ordinary preorder structure to model
-  the error ordering.
-  %
-  This avoids the complexities of classical domain-theoretic models
-  (New and Licata) or logical relations models using explicit
-  step-indexing (New and Ahmed).
-  %
-  In particular, we avoid a major technical complexity of New and
-  Ahmed that require two logical relations to prove the graduality
-  theorem.
+  We develop a semantics for a gradually typed lambda calculus with
+  effects that and prove the validity of type-based equational
+  reasoning as well as the graduality theorem. Improving on prior
+  work, we use a compositional denotational semantics in category of
+  preordered sets and monotone functions to develop the theory
+  modularly. This means that nearly all theorems and constructions
+  come from general facts about the category of preordered sets and
+  almost no metatheory is specific to gradually typed languages.
 
-  By working synthetically we can treat the domains in which gradual
-  types are interpreted as if they were ordinary sets. This allows us
-  to give a ``na\"ive'' presentation of gradual typing where each
-  gradual type is modeled as a well-behaved subset of the universal
-  domain used to model the dynamic type, and type precision is modeled
-  as simply a subset relation.
-  %
-  This provides a rich semantic setting in which to 
+  The main challenge is in developing a suitable model for the dynamic
+  type, which requires solving a domain equation. We avoid the
+  complexities of both traditional order-theoretic domain theory and
+  explicit step-indexed models by using the recently developed
+  technique of synthetic guarded domain theory in a guarded type
+  theory. This makes constructing solutions to guarded domain
+  equations trivial, and allows us to carry out the equivalent of a
+  step-indexed logical relation without using explicit step-indexing
+  while still maintaining the full expressivity of dependent type
+  theory as our metatheoretic setting.
+
+  Finally, the entire semantics has been mechanized in the proof
+  assistant Guarded Cubical Agda, with much of the semantics entirely
+  in reusable extensions to the standard category theory library. This
+  provides a foundation for reusable modular mechanized metatheory of
+  gradual typing and other effectful denotational semantics.
 \end{abstract}
@@
 \section{Introduction}
+
+Gradually typed languages provide a unified setting in which dynamic
+and statically typed programming styles can be accommodated along with
+free interoperability between static and dynamically typed components.
+
+The metatheory of gradually typed languages is considered especially
+challenging, as it inherently involves (1) general recursion and at
+least an error effect distinct from non-termination (2) the
+verification of type-based equational reasoning to demonstrate the
+soundness of static typing and (3) proving graduality, a relational
+property of program semantics that establishes that increasing the
+amount of static typing in a program results in a refinement of
+program behavior. Most prior work to this metatheory relies on purely
+operational approaches: an operational semantics for the language is
+developed and a simulation-based argument is given for
+graduality. These have the advantage of being fairly elementary and so
+amenable to direct mechanization in a proof assistant but at the cost
+of lacking reusability. Furthermore, equational aspects of the
+language are for the most part ignored. A second line of prior work is
+to develop generic operational frameworks in which the above can be
+carried out: the gradualizer as well as Abstracting Gradual
+Typing. These are reusable techniques that apply across languages but
+still do not address equational theory and . Finally a third line of
+prior work gives a semantic theory of gradual typing based on the idea
+that casts form embedding-projection pairs in a 2-categorical
+semantics. This approach has been formalized in two ways: one using
+classical denotational semantics based on complete partial orders
+(CPOs) and another based on step-indexed logical relations over an
+untyped operational semantics. This approach has similar benefits of
+abstraction to the AGT framework, but suffers from a lack of
+reusability. First, classical domain theory is believed to be unable
+to model higher-order references, a feature of all practical gradually
+typed languages. Second, the proofs based on step-indexed logical
+relations, while known to scale to higher-order references are
+relatively unstructured and complex. While inspired by the
+embedding-projection pair semantics, several components of the proof
+that work in the CPO-based semantics break down in the step-indexed
+setting for unclear reasons. This work also tends to have a high level
+of tedious proof overhead in practice as evidenced that most papers
+using this approach include tech reports of repetitive paper proofs
+that typically dwarf the size of the paper itself.
+
+In this current work it is our goal to combine the benefits of the
+previous approaches: we desire the compositionality and reusability of
+the domain theoretic semantics, the genericity of abstracting gradual
+typing and the gradualizer, the easy mechanizability of the elementary
+operational techniques as well as the generality of step-indexed
+approaches.
+
+\section{Call-by-value Gradual Cast Calculus: Syntax and Axioms}
+
+\subsection{}
+
+\section{A Compositional ``Operational'' Semantics}
+
+We start by giving a concrete description of the semantics our
+denotational approach provides in the style of a monadic
+interpreter.
+
+The values of a syntactic type $A$ in the GCC are interpreted as
+elements of a set $\sem A$: values of number type are interpreted as
+numbers, values of function type are interpreted as functions with a
+monadic output and value of the dynamic type are interpreted in a complex
+set that contains numbers and functions as subsets.
+
+For now for concreteness we will work with a simple monad that
+supports only errors and divergence with observable computational
+steps. This can be described as the composition of two monads $T X =
+L(X_\mho)$ where $LX$ is Capretta's coinductive delay monad and
+$X_\mho$ is simply the maybe monad where we call the additional
+element $\mho$ to connote that it is intended to represent a runtime
+error.
+%
+Capretta's delay monad is a coinductive set defined as the greatest
+solution to the equation $L X \cong \textrm{Done}(X)+
+\textrm{Running}(LX)$. That is an element is either done with a
+produced $X$ or takes an observable computational step. Taking the
+greatest fixed point rather than least fixed point allows for the
+possibility of a program that runs forever.
+%
+We think of this as a computational process that requires external
+``fuel'' to execute: either the process is complete and we receive a
+final value or the computation requires an additional fuel to continue
+executing.
+%
+
+For each computation $\Gamma \vdash M : A$ we define a 4-place
+relation $\opsemGMNV \gamma M n V$ where $\gamma \in
+\sem{\Gamma}$, $n \in \mathbb N$ and $V \in \sem{A}$. Further this
+relation is a decidable partial function when considering $M,\gamma,n$
+as inputs and $V$ as an output and is a partial function when
+considering $M,\gamma$ as inputs and $n$ as an output.
+
+\begin{mathpar}
+  \opsemMNV {\texttt{zro}} 0 0
+
+  \inferrule
+      {\opsemMNV M n m}
+      {\opsemMNV {\texttt{suc}(M)} n {m+1}}
+
+  \inferrule
+  {R(V_i,n,V_o) \iff \opsemGMNV {(\gamma, x\mapsto V_i)} M n {V_o}}
+  {\opsemMNV {\lambda x. M} 0 R}
+\end{mathpar}
+As well as two
+
+of values, computations, upcasts and downcasts in GCC mutually recursively
+
+%% This can be equivalently described without
+%% mention of coinduction as follows. First, let $X_\ohm$ be the maybe
+%% monad but where we call the additional element $\ohm$ to connote that
+%% the extra element represents divergence rather than error. Then an
+%% element of the delay monad $l \in L X$ is a function $l : \mathbb N
+%% \to X_\ohm$ that is defined on at most one input. The intuition is
+%% that an element $l$ represents a computational procedure that when
+%% executed may possibly diverge or run in some finite number of steps to
+%% a value in $X$. Applying $l$ to a number $n$ is asking the question
+
+%% : an element of $L_\mho X$ is a
+%% function $\mathbb N \to X_{\ohm}$
+
+\subsection{Abstracting GCC Semantics}
 

### 2023-09-15 paper-new/paper.tex
some writing

@@
 
-\title{Denotational Semantics for Gradual Typing in Synthetic Guarded Domain Theory}
+\title{Mechanized Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory}
 \author{Eric Giovannini}
@@
 \begin{abstract}
-    We develop a denotational semantics for a simple gradually typed language
-    that is adequate and proves the graduality theorem.
-    %
-    The denotational semantics is constructed using \emph{synthetic
-    guarded domain theory} working in a type theory with a later
-    modality and clock quantification.
-    %
-    This provides a remarkably simple presentation of the semantics,
-    where gradual types are interpreted as ordinary types in our ambient
-    type theory equipped with an ordinary preorder structure to model
-    the error ordering.
-    %
-    This avoids the complexities of classical domain-theoretic models
-    (New and Licata) or logical relations models using explicit
-    step-indexing (New and Ahmed).
-    %
-    In particular, we avoid a major technical complexity of New and
-    Ahmed that requires two logical relations to prove the graduality
-    theorem.
+  Gradually typed programming languages, which allow for soundly
+  mixing static and dynamically typed programming styles, present a
+  strong challenge for metatheorists. Even the simplest sound
+  gradually typed languages feature at least recursion and errors,
+  with realistic languages featuring furthermore runtime allocation of
+  memory locations and dynamic type tags. Further, the desired
+  metatheoretic properties of gradually typed languages have become
+  increasingly sophisticated: validity of typed based equational
+  reasoning as well as the relational property known as the gradual
+  guarantee or graduality. Many recent works have tackled verifying
+  these properties, but the resulting mathematical developments are
+  highly repetitive and tedious, with few reusable theorems persisting
+  across different developments.
 
-    By working synthetically we can treat the domains in which gradual
-    types are interpreted as if they were ordinary sets. This allows us
-    to give a ``na\"ive'' presentation of gradual typing where each
-    gradual type is modeled as a well-behaved subset of the universal
-    domain used to model the dynamic type, and type precision is modeled
-    as simply a subset relation.
-    %
+  In this work, we present a new denotational account of gradual
+  typing semantics developed using guarded domain theory. Guarded
+  domain theory combines the expressive power of step-indexed logical
+  relations for modeling recursive features with the modularity and
+  reusability of denotational semantics. Further, recent extensions to
+  cubical Agda mean that synthetic guarded domain theory is readily
+  mechanized in a proof assistant. We demonstrate the feasibility of
+  this approach with a model of gradually typed lambda calculus and
+  prove the validity of beta-eta equality and the graduality theorem
+  for the denotational model. This model should provide the basis for
+  a reusable mathematical theory of gradually typed program semantics.
+%%     We develop a denotational semantics for a simple gradually typed language
+%%     that is adequate and proves the graduality theorem.
+%%     %
+%%     The denotational semantics is constructed using \emph{synthetic
+%%     guarded domain theory} working in a type theory with a later
+%%     modality and clock quantification.
+%%     %
+%%     This provides a remarkably simple presentation of the semantics,
+%%     where gradual types are interpreted as ordinary types in our ambient
+%%     type theory equipped with an ordinary preorder structure to model
+%%     the error ordering.
+%%     %
+%%     This avoids the complexities of classical domain-theoretic models
+%%     (New and Licata) or logical relations models using explicit
+%%     step-indexing (New and Ahmed).
+%%     %
+%%     In particular, we avoid a major technical complexity of New and
+%%     Ahmed that requires two logical relations to prove the graduality
+%%     theorem.
+
+%%     By working synthetically we can treat the domains in which gradual
+%%     types are interpreted as if they were ordinary sets. This allows us
+%%     to give a ``na\"ive'' presentation of gradual typing where each
+%%     gradual type is modeled as a well-behaved subset of the universal
+%%     domain used to model the dynamic type, and type precision is modeled
+%%     as simply a subset relation.
+%%     %
   \end{abstract}
@@
 \maketitle
+
+% Outline
+
+% 1. Intro: What do we want out of gradually typed languages and why
+%    is it hard to prove? Explanation: Gradual Typing inherently
+%    involves recursive types, multiple effects, relational properties.
+%    We argue that the increasing complexity of the metatheory of
+%    gradual typing makes it a good candidate for
+
+% 2. Extensional Dream Semantics: double categorical
+
+% 3. The Problem with Step-indexing
+
+% 4. A Compromise
+
+% 5. Formalization in Guarded Cubical Agda
 

### 2023-09-15 paper-new/semantics.tex
some writing

@@
+\section{Denotational Semantics}
+
+First, we define a denotational semantics of types and terms of the
+cast calculus by giving a standard monadic denotational semantics in
+the cartesian closed category of preorders and monotone functions,
+extended to model the primitives of gradual typing: the dynamic type,
+errors and type casts. The most interesting part of this semantics is
+the construction of the monad and the dynamic type.
+
 \section{Domain-Theoretic Constructions}\label{sec:domain-theory}

### 2023-09-15 paper-new/technical-background.tex
some writing

@@
 % synthetic guarded domain theory, denotational semantics therein
+
+\subsection{Operational Reduction Proofs}
+
+\subsection{Classical Domain Models}
+
+New and Licata \cite{newlicata} developed an axiomatic account of the
+graduality relation on cast calculus terms and gave a denotational
+model of this calculus using classical domain theory based on
+$\omega$-CPOs. This semantics has scaled up to an analysis of a
+dependently typed gradual calculus in \cite{asdf}. This meets our
+criterion of being a reusable mathematical theory, as general semantic
+theorems about gradual domains can be developed independent of any
+particular syntax and then reused in many different denotational
+models. However, it is widely believed that such classical domain
+theoretic techniques cannot be extended to model higher-order store, a
+standard feature of realistic gradually typed languages such as Typed
+Racket. Thus if we want a reusable mathematical theory of gradual
+typing that can scale to realistic programming languages, we need to
+look elsewhere to so-called ``step-indexed'' techniques.
+
+A series of works \cite{newahmed,newlicataahmed,newjamnerahmed}
+developed step-indexed logical relations models of gradually typed
+languages based on operational semantics. Unlike, classical domain
+theory, such step-indexed techniques are capable of modeling
+higher-order store and runtime-extensible dynamic types
+\cite{amalsthesis,nonpmetricparamorsomething,newjamnerahmed}. However
+their proof developments are highly repetitive and technical, with
+each development formulating a logical relation from first-principles
+and proving many of the same tedious lemmas without reusable
+mathematical abstractions. Our goal in the current work is to extract
+these reusable mathematical principles from these tedious models to
+make formalization of realistic gradual languages tractible.
 

### 2023-11-06 paper-new/semantics.tex
check in semantics file before making changes

@@
 In this section, we discuss the fundamental objects of the model into which we will embed
-the intensional lambda calculus and inequational theory. It is important to remember that
+the step-sensitive lambda calculus $\intlc$ and inequational theory. It is important to remember that
 the constructions in this section are entirely independent of the syntax described in the
 previous section; the notions defined here exist in their own right as purely mathematical
-constructs. In the next section, we will link the syntax and semantics via an interpretation
+constructs. In the next section, we will link the syntax and semantics via a semantic interpretation
 function.
@@
 
-When thinking about how to model intensional gradually-typed programs, we must consider
+When thinking about how to model intensional gradually-typed programs, we should consider
 their possible behaviors. On the one hand, we have \emph{failure}: a program may fail
@@
 Unless otherwise mentioned, all constructs involving $\later$ or $\fix$
-are understood to be with repsect to a fixed clock $k$. So for the above, we really have for each
+are understood to be with respect to a fixed clock $k$. So for the above, we really have for each
 clock $k$ a type $\li^k A$ with respect to that clock.
@@
 Since we claimed that $\li A$ is a monad, we need to define the monadic operations
-and show that they repect the monadic laws. The return is just $\eta$, and extend
+and show that they respect the monadic laws. The return is just $\eta$, and extend
 is defined via by guarded recursion by cases on the input.
@@
 
+\begin{comment}
+% Since this mentions the ordering on B, it should be introduced after introducing predomains.
 The lift monad has the following universal property. Let $f$ be a function from $A$ to $B$,
 where $B$ is a $\later$-algebra, i.e., there is $\theta_B \colon \later B \to B$.
-Further suppose that $B$ is also an ``error-algebra'', that is, an algebra of the
-constant functor $1 \colon \text{Type} \to \text{Type}$ mapping all types to Unit.
-This latter statement amounts to saying that there is a map $\text{Unit} \to B$, so $B$ has a
-distinguished ``error element" $\mho_B \colon B$.
+Further suppose that $B$ is also an ``error-algebra'', that is, there is an error element
+$\mho_B$ such that $\mho_B \le_B y$ for all $y \in B$.
+
+% Further suppose that $B$ is also an algebra of the
+% constant functor $1 \colon \text{Type} \to \text{Type}$ mapping all types to Unit.
+% This latter statement amounts to saying that there is a map $\text{Unit} \to B$, so $B$ has a
+% distinguished ``error element" $\mho_B \colon B$ such that $\mho_B \le_B y$ for all $y \in B$.
 
@@
 
-Recalling that $f'$ is a guaded fixpoint, it is available ``later'' and by
+Recalling that $f'$ is a guarded fixpoint, it is available ``later'' and by
 applying the tick we get a function we can apply ``now''; for the argument,
 we apply the tick to $\tilde{l}$ to get a term of type $\li A$.
+\end{comment}
 
@@
 
-\subsection{Predomains}
+\subsection{Predomains}\label{sec:predomains}
 
@@
 
+We define monotone functions between predomain as expected. Given predomains
+$A$ and $B$, we write $f \colon A_i \monto A_o$ to indicate that $f$ is a monotone
+function from $A_i$ to $A_o$, i.e, for all $a_1 \le_{A_i} a_2$, we have $f(a_1) \le_{A_o} f(a_2)$.
+We also define an ordering on monotone functions as
+$f \le g$ if for all $a$ in $\ty{A_i}$, we have $f(a) \le_{A_o} g(a)$.
+
 For each type we want to represent, we define a predomain for the corresponding semantic
@@
 
-We define monotone functions between predomain as expected. Given predomains
-$A$ and $B$, we write $f \colon A \monto B$ to indicate that $f$ is a monotone
-function from $A$ to $B$, i.e, for all $a_1 \le_A a_2$, we have $f(a_1) \le_B f(a_2)$.
+\begin{itemize}
+  \item There is a predomain $\Nat$ for natural numbers, where the ordering is equality.
 
-\begin{itemize}
-  \item There is a predomain Nat for natural numbers, where the ordering is equality.
-
-  \item There is a predomain Dyn to represent the dynamic type. The underlying type
+  \item There is a predomain $\Dyn$ to represent the dynamic type. The underlying type
   for this predomain is defined by guarded fixpoint to be such that
-  $\ty{\Dyn} \cong \mathbb{N}\, +\, \later (\ty{\Dyn} \monto \ty{\Dyn})$.
+  $\ty{\Dyn} \cong \mathbb{N}\, +\, \later (\ty{\Dyn} \monto \li \ty{\Dyn})$.
   This definition is valid because the occurrences of Dyn are guarded by the $\later$.
   The ordering is defined via guarded recursion by cases on the argument, using the
-  ordering on $\mathbb{N}$ and the ordering on monotone functions described below.
+  ordering on $\mathbb{N}$ and the ordering on monotone functions described above.
 
@@
   from $A_i$ to $A_o$, which we denote by $A_i \To A_o$.
-  The ordering is such that $f$ is below $g$ if for all $a$ in $\ty{A_i}$, we have
-  $f(a)$ is below $g(a)$ in the ordering for $A_o$. 
+
+  \item Given a preomain $A$, we can form the predomain $\later A$ whose underlying
+  type is $\later \ty{A}$. We define $\tilde{x} \le_{\later A} \tilde{y}$ to be
+  $\later_t (\tilde{x}_t \le_A \tilde{y}_t)$.
 \end{itemize}
@@
 
-\subsection{Step-Insensitive Relation}
+\subsection{Step-Insensitive Bisimilarity Relation}
 
-We define another ordering on $\li A$, called the ``step-insensitive ordering'' or
-``weak bisimilarity'', written $l \bisim l'$.
+We define another ordering on $\li A$, called ``step-insensitive bisimilarity"
+or ``weak bisimilarity" written $l \bisim l'$.
 Intuitively, we say $l \bisim l'$ if they are equivalent ``up to delays''.
 We introduce the notation $x \sim_A y$ to mean $x \le_A y$ and $y \le_A x$.
-
-The weak bisimilarity relation is defined by guarded fixpoint as follows:
+% TODO if A is a poset, then we can just say that x = y
+%
+The step-insensitive bisimilarity relation is defined by guarded fixpoint as follows:
 
@@
 
+When both sides are $\eta$, then we ensure that the underlying values are related.
+When one side is a $\theta$ and the other is $\eta x$ (i.e., one side steps),
+we stipulate that the $\theta$-term runs to $\eta y$ where $x$ is related to $y$.
+Similarly when one side is $\theta$ and the other $\mho$.
+If both sides step, then we allow one time step to pass and compare the resulting terms.
+In this way, the definition captures the intuition of terms being equivalent up to
+delays.
+
+It can be shown (by \lob-induction) that the step-sensitive relation is symmetric.
+However, it can also be shown that this relation is \emph{not} transitive:
+One can prove within Clocked Cubical Type Theory
+that if this relation were transitive, then in fact it would be trivial in that
+$l \bisim l'$ for all $l, l'$.
+This issue will be resolved when we consider the relation's \emph{globalization}.
+
 \subsection{Error Domains}
 
-\subsection{Globalization}
+While value types will be interpreted as predomains, we also need a semantics
+for computation types. This will be in the form of \emph{error domains}, of which the
+Lift monad is a prototypical example. For a fixed clock $k$, an error domain $A$
+consists of a predomain (which we also denote by $A$ when there is no risk of confusion),
+along with a bottom element $\mho_A$ and a $\later$-algebra $\theta_A \colon \later^k A \monto A$.
+
+% TODO function space as error domain?
+
+\subsection{Globalization}\label{sec:globalization}
 
 Recall that in the above definitions, any occurrences of $\later$ were with
-repsect to a fixed clock $k$. Intuitively, this corresponds to a step-indexed set.
+respect to a fixed clock $k$. Intuitively, this corresponds to a step-indexed set.
 It will be necessary to consider the ``globalization'' of these definitions,
@@
 
-\subsection{Relational Semantics}
+\subsection{Step-indexed Semantics}
 
-\subsubsection{Term Precision via the Step-Sensitive Error Ordering}
+We give a semantics to the step-sensitive lambda calculus $\intlc$ we defined
+in Section \ref{sec:step-sensitive-lc}.
+%
+Much of the semantics is similar to a normal call-by-value denotational semantics;
+we highlight the differences.
+Recall that we will interpret value types as predomains, and computation types
+as error domains. Value type contexts $\Gamma = x_1 \colon A_1, \dots, x_n \colon A_n$
+will be interpreted as the product $\sem{A_1} \times \cdots \times \sem{A_n}$, and
+computation type contexts $\Delta = \Delta_\Sigma , \Delta|_V$ will be interpreted as a pair
+$(\delta_\Sigma, \delta_V)$ where $\delta_\Sigma$ is either empty or $\sem{B}$.
+
+The semantics of the dynamic type $\dyn$ is the predomain $\Dyn$ introduced in Section
+\ref{sec:predomains}.
+%
+The interpretation of a value $\hasty {\Gamma} V A$ will be a monotone function from
+$\sem{\Gamma}$ to $\sem{A}$. Likewise, a term $\hasty {\Delta} M {\Ret{A}}$ will be interpreted
+as a monotone function from $\sem{\Delta}$ to $\sem{\Ret{A}} = \li \sem{A}$.
+
+Recall that $\Dyn$ is isomorphic to $\Nat\, + \later (\Dyn \monto \li \Dyn)$.
+Thus, the semantics of $\injnat{\cdot}$ and $\injarr{\cdot}$ are simply the
+injections $\inl$ and $\inr$.
+
+The interpretation of $\lda{x}{M}$ works as follows. Recall by the typing rule for
+lambda that $\hasty {\cdot, \Gamma, x : A_i} M {\Ret {A_o}}$, so the interpretation of $M$
+has type $\{*\} \times (\sem{\Gamma} \times \sem{A_i})$ to $\sem{A_o}$.
+The interpretation of lambda is thus a function (in the ambient type theory) that takes
+a value $a$ representing the argument and applies it (along with $\gamma$) as argument to
+the interpretation of $M$.
+%
+The interpretation of bind and of application both make use the monadic extend function on $\li A$.
+%
+The interpretation of case-nat and case-arrow is simply a case inspection on the
+interpretation of the scrutinee, which has type $\Dyn$.
+
+\vspace{2ex}
+
+\noindent Types:
+\begin{align*}
+  \sem{\nat} &= \Nat \\
+  \sem{\dyn} &= \Dyn \\
+  \sem{A \ra A'} &= \sem{A} \To \sem{A'} \\
+  \sem{\later A} &=\, \later \sem{A} \\
+  \sem{\Ret A} &= \li \sem{A}
+\end{align*}
+
+% Contexts:
+
+% TODO check these, especially the semantics of bind, case-nat, and case-arr
+% with respect to their context argument
+\noindent Values and terms:
+\begin{align*}
+  \sem{\zro}         &= \lambda \gamma . 0 \\
+  \sem{\suc\, V}     &= \lambda \gamma . (\sem{V}\, \gamma) + 1 \\
+  \sem{x \in \Gamma} &= \lambda \gamma . \gamma(x) \\
+  \sem{\lda{x}{M}}   &= \lambda \gamma . \lambda a . \sem{M}\, (*,\, (\gamma , a))  \\
+  \sem{\injnat{V_n}} &= \lambda \gamma . \inl\, (\sem{V_n}\, \gamma) \\
+  \sem{\injarr{V_f}} &= \lambda \gamma . \inr\, (\sem{V_f}\, \gamma) \\[2ex]
+  \sem{\nxt\, V}     &= \lambda \gamma . \nxt (\sem{V}\, \gamma) \\
+  \sem{\theta}       &= \lambda \gamma . \theta \\
+%
+  \sem{\err_B}         &= \lambda \delta . \mho \\
+  \sem{\ret\, V}       &= \lambda \gamma . \eta\, \sem{V} \\
+  \sem{\bind{x}{M}{N}} &= \lambda \delta . \ext {(\lambda x . \sem{N}\, (\delta, x))} {\sem{M}\, \delta} \\
+  \sem{V_f\, V_x}      &= \lambda \gamma . \ext {(\lambda f . (\ext {f} {\sem{V_x}\, \gamma}))} {(\sem{V_f}\, \gamma)} \\
+  \sem{\casenat{V}{M_{no}}{n}{M_{yes}}}         &= 
+    \lambda \delta . \text{case $(\sem{V}\, \delta)$ of} \\ 
+    &\quad\quad\quad\quad \alt \inl(n) \to \sem{M_{yes}}(n) \\
+    &\quad\quad\quad\quad \alt \inr(\tilde{f}) \to \sem{M_{no}} \\
+  \sem{\casearr{V}{M_{no}}{\tilde{f}}{M_{yes}}} &= 
+    \lambda \delta . \text{case $(\sem{V}\, \delta)$ of} \\ 
+    &\quad\quad\quad\quad \alt \inl(n) \to \sem{M_{no}} \\
+    &\quad\quad\quad\quad \alt \inr(\tilde{f}) \to \sem{M_{yes}}(\tilde{f})
+\end{align*}
+
+% TODO
+% \noindent Relations:
+% \begin{align*}
+% %
+% \end{align*}
+
+\begin{comment}
+\subsection{Global Semantics}
+
+Having defined the above step-indexed semantics, we now pass to a ``global''
+semantics that does not involve any step-indexing. The resulting semantics is still
+intensional in that terms that produce the same value in a different number of steps
+will be distinct.
+We define
+
+\[ \semgl{\cdot} = \forall (\kpa \colon \Clock) .\, \sem{\cdot}[\kpa]. \]
+
+Note that for a term $M$ of type $\Ret{A}$, the semantics has type
+$\sem{M} \colon \sem{\Delta} \monto \sem{\Ret{A}} = \sem{\Delta} \monto \li \sem{A}$.
+In the case where $\Delta$ is the empty context, i.e., when $M$ is a closed term,
+then this is equivalent to $\li \sem{A}$.
+Then the global semantics in this case is $\forall \kappa . \liclk {\kappa} \sem{A}$.
+We can show in Clocked Cubical Type Theory this type satisfies a coinductive unfolding property
+
+\[ \forall \kappa . \li \sem{A} \cong \sem{A} + 1\, + (\forall \kappa.\li \sem{A}). \]
+\end{comment}
+
+% Machines
+
+% We then define a relation $\Dwn^n$ between terms of type $T$ and $\Machine {\sem{T}}$ by
+
+% \subsection{Extensional Collapse}
+
+% \subsection{Relational Semantics}
+
+% \subsubsection{Term Precision via the Step-Sensitive Error Ordering}
 % Homogeneous vs heterogeneous term precision

### 2023-11-16 paper-new/graduality.tex
begin rearranging: discuss term syn/sem and then ordering syn/sem

@@
-\section{Unary Canonicity}
-Before discussing graduality, we seek to prove its ``unary'' analogue.
-Namely, instead of considering inequality between terms, we start by considering equality.
-
 \section{Graduality}\label{sec:graduality}
+Now we move to the proof of graduality.
 The main theorem we would like to prove is the following:
 
+% TODO use the etmprec notation
 \begin{theorem}[Graduality]
@@
   \begin{enumerate}
-    \item If $N = \mho$, then $M = \mho$
-    \item If $N = `n$, then $M = \mho$ or $M = `n$
-    \item If $M = V$, then $N = V$
+    \item $M \Downarrow$ iff $N \Downarrow$
+    \item If $M \Downarrow v_?$ and $N \Downarrow v'_?$ then either $v_? = \mho$, or $v_? = v'_?$.
   \end{enumerate}
 \end{theorem}
+
+\subsection{Proof Overview}
+
+We split the proof into three main steps. The first concerns the translation
+of the surface syntax $\extlc$ to the intensional syntax $\intlc$.
+
+\begin{lemma}[]
+  If $\etmprec{\gamlt}{M}{N}{c}$, then there exist $M'$ and $N'$ such that
+
+  \[ i(M) \synbisim M' \ltdyn_i N' \synbisim i(N). \]
+\end{lemma}
+
+Next we show that the notions of intensional syntactic term precision and
+syntactic bisimilarity imply their corresponding semantic counterparts:
+
+\begin{lemma}[]
+  If $\itmprec{\gamlt}{M}{N}{c}$, then $\sem{M} \semlt {\sem{N}}$
+\end{lemma}
+
+\begin{lemma}
+  If $M \synbisim N$, then $\sem{M} \bisim \sem{N}$.
+\end{lemma}
+
+Finally, we need an adequacy result stating that the semantic notions of error
+ordering and bisimilarity for natural numbers imply that the values of the
+lift monad that they relate have the expected behavior.
+
+\begin{lemma}
+  Let $l, l' : \li \Nat$.
+  If $l \semlt l'$, then either (1) $l = \delta^n(\mho)$ for some $n$, or (2) 
+  $l = l' = \delta^n(\eta m)$ for some $n$ and $m$, or (3) $l$ and $l'$ fail to terminate.
+\end{lemma}
+
+\begin{lemma}
+  If $l \bisim l'$, then either (1) $l = \delta^n(\mho)$ and $l' = \delta^{n'}(\mho)$ for some $n$ and $n'$,
+  (2) $l = \delta^n(\eta m)$ and $l' = \delta^{n'}(\eta m)$, or $l$ and $l'$ fail to terminate.
+\end{lemma}

### 2023-11-16 paper-new/intro.tex
begin rearranging: discuss term syn/sem and then ordering syn/sem

@@
 the programmer is assured that if the program passes the type-checker, their program
-is free of type errors. Meanwhile, dynamic typing allows the programmer to rapidly prototype
+is free of type errors, and moreover, the equational theory implies that program optimizations are valid.
+Meanwhile, dynamic typing allows the programmer to rapidly prototype
 their application code without needing to commit to fixed type signatures for their functions.
 
-\emph{Gradually-typed languages} \cite{siek-taha06} allow for both static and dynamic typing disciplines
-to be used in the same codebase, and support smooth interoperability between statically-typed
-and dynamically-typed code.
+\emph{Gradually-typed languages} \cite{siek-taha06, tobin-hochstadt06} allow
+for both static and dynamic typing disciplines to be used in the same codebase,
+and support smooth interoperability between statically-typed and dynamically-typed code.
 This flexibility allows programmers to begin their projects in a dynamic style and
@@
 and the programmer is more certain of what the types should be, the code can be
-\emph{gradually} migrated to a statically typed style without needing to start the project
-over in a completely differnt language.
+\emph{gradually} migrated to a statically typed style without needing to
+rewrite the project in a completely different language.
 
-Gradually-typed languages should satisfy two intuitive properties.
+%Gradually-typed languages should satisfy two intuitive properties.
+The following two properties have been identified as useful for gradually typed languages.
 First, the interaction between the static and dynamic components of the codebase
-should be safe -- i.e., should preserve the guarantees made by the static types.
+should preserve the guarantees made by the static types.
 In particular, while statically-typed code can error at runtime in a gradually-typed language,
@@
 % In other words, in the statically-typed portions of the codebase, type soundness must be preserved.
-Second, gradually-typed langugaes should support the smooth migration from dynamic typing
+Second, gradually-typed languages should support the smooth migration from dynamic typing
 to static typing, in that the programmer can initially leave off the
@@
 and Boyland \cite{siek_et_al:LIPIcs:2015:5031}.
-This property is also referred to as \emph{graduality}, by analogy with parametricity.
+This property is also referred to as \emph{graduality} \cite{new-ahmed2018}, by analogy with parametricity.
 Intuitively, graduality says that going from a dynamic to static style should not
@@
 Current approaches to proving graduality include the methods of Abstracting Gradual Typing
-\cite{garcia-clark-tanter2016} and the formal tools of the Gradualier \cite{cimini-siek2016}.
-These allow the language developer to start with a statically typed langauge and derive a
-gradually typed language that satisfies the gradual guarantee. The downside to these approaches
-is that the semantics of the resulting languages are too lazy: the frameworks consider only
-the $\beta$ rules and not the $\eta$ equalities. Furthermore, while these frameworks do prove
+\cite{garcia-clark-tanter2016} and the formal tools of the Gradualizer \cite{cimini-siek2016}.
+These allow the language developer to start with a statically typed language and derive a
+gradually typed language that satisfies the gradual guarantee. The main downside to
+these approaches lies in their inflexibility: since the process in entirely mechanical,
+the language designer must adhere to the predefined framework.
+Many gradually typed languages do not fit into either framework, e.g., Typed Racket
+\cite{tobin-hochstadt06, tobin-hochstadt08}.
+%
+Furthermore, while these frameworks do prove
 graduality of the resulting languages, they do not show the correctness of the equational theory,
 which is equally important to sound gradual typing. For example, programmers often refactor their
-code without thinking about whether the refactoring has broken the semantics of the program. 
-It is the validity of the laws in the equational theory that guarantees that such refactorings are sound. 
+code, and in so doing they rely implicitly on the validity of the laws in the equational theory.
 Similarly, correctness of compiler optimizations rests on the validity of the corresponding equations
-from the equational theory. It is therefore important that the langages that claim to be gradually typed
+from the equational theory. It is therefore important that the languages that claim to be gradually typed
 have provably correct equational theories.
+
+% The approaches are too inflexible... the fact that the resulting semantics are too lazy
+% is a consequence of that inflexibility.
+% The validity of the equational theory captures the programmer's intuitive thinking when they refactor their code
 
@@
 %
-In this approach, a logical relation is constructed and shown to be sound with respect to
-the notion of observational approximation that specifies when one program is more precise than another.
+In this approach, a logical relation is constructed and used to show that the equational theory
+is sound with respect to the operational semantics.
+%and shown to be sound with respect to
+%the notion of observational approximation that specifies when one program is more precise than another.
 The downside of this approach is that each new language requires a different logical relation
 to prove graduality. Furthermore, the logical relations tend to be quite complicated due
-to a technical requirement known as \emph{step-indexing}.
+to a technical requirement known as \emph{step-indexing}, where the stepping behavior of terms
+must be accounted for in the logical relation.
 As a result, developments using this approach tend to require vast effort, with the
 corresponding technical reports having 50+ pages of proofs.
+%
+Additionally, while the axioms of gradual type theory are compositional at a ``global'' level,
+they do not compose in the step-indexed setting. One of the main goals of the present work
+is to formulate a composable theory of gradual typing in a setting where the stepping behavior
+is tracked.
 
@@
 \emph{synthetic guarded domain theory}.
-The tecnhiques of synthetic guarded domain theory allow us to internalize the
+The techniques of synthetic guarded domain theory allow us to internalize the
 step-index reasoning normally required in logical relations proofs of graduality,
@@
 
-In this paper, we report on work we have done to mechanize proofs of graduality and
-correctness of equational theories using SGDT techniques in Agda.
-Our goal in this work is to mechanize these proofs in a reusable way,
+In this paper, we report on work we have done towards mechanizing proofs of graduality
+and soundness of the equational theory of cast calculi using the techniques of SGDT.
+We take a step toward mechanization in the Agda proof assistant by describing a compositional
+denotational semantics for gradual typing in a setting where the steps taken by terms are tracked.
+
+Our longer-term goal is to mechanize these proofs in a reusable way,
 thereby providing a framework to use to more easily and
-conveniently prove that existing languages satsify graduality and have
+conveniently prove that existing languages satisfy graduality and have
 sound equational theories. Moreover, the aim is for designers of new languages
-to utlize the framework to facilitate the design of new provably-correct
-gradually-typed languages with nontrivial features.
+to utilize the framework to facilitate the design of new provably-correct
+gradually-typed languages with more complex features.
 
-\subsection{Proving Graduality in SGDT}
-\textbf{TODO:} This section should probably be moved to after the relevant background has been introduced?
+\subsection{Contributions}
+Our main contribution is a compositional denotational semantics for step-aware gradual typing,
+analogous to Gradual Type Theory but now in the ``intensional" setting.
+In parallel, we are developing a reusable framework in
+Guarded Cubical Agda for developing machine-checked proofs of graduality of a cast calculus.
+To demonstrate the feasibility and utility of our approach, we are applying the framework
+to prove graduality for the simply-typed gradual lambda calculus.
+% Along the way, we have developed an ``intensional" theory of graduality that is of
+% independent interest.
+
+\subsection{Overview of Remainder of Paper}
+
+In Section \ref{sec:overview}, we give an overview of the gradually-typed lambda
+calculus and the graduality theorem.
+%
+In Section \ref{sec:technical-background}, we provide technical background on gradually typed languages and
+on synthetic guarded domain theory.
+% 
+In Section \ref{sec:gtlc-terms}, we introduce the gradually-typed cast calculus
+for which we will prove graduality. We give a semantics to the terms
+using the tools of guarded type theory.
+% Important here are the notions of syntactic
+% type precision and term precision. For reasons we describe below, we
+% introduce two related calculi, one in which the stepping behavior of terms is
+% implicit (an \emph{extensional} calculus, $\extlc$), and another where this behavior
+% is made explicit (an \emph{intensional} calculus, $\intlc$).
+%
+
+In Section \ref{sec:gtlc-precision}, we define the term precision ordering
+and describe our approach to assigning a denotational semantics to this ordering.
+This approach builds on the semantics constructed in the previous section,
+but extending it to the term ordering presents new challenges.
+
+% In Section \ref{sec:domain-theory}, we define several fundamental constructions
+% internal to SGDT that will be needed when we give a denotational semantics to
+% the gradual lambda calculus.
+%This includes the Lift monad, predomains and error domains.
+%
+% In Section \ref{sec:semantics}, we define the denotational semantics for the
+% intensional gradually-typed lambda calculus using the domain theoretic
+% constructions defined in the previous section.
+%
+In Section \ref{sec:graduality}, we outline in more detail the proof of graduality for the
+extensional gradual lambda calculus.
+%
+In Section \ref{sec:discussion}, we discuss the benefits and drawbacks to our approach in comparison
+to the traditional step-indexing approach, as well as possibilities for future work.
+
+\section{Overview}\label{sec:overview}
+
+% This section used to be part of the intro.
+% \subsection{Proving Graduality in SGDT}
+% TODO This section should probably be moved to after the relevant background has been introduced?
 
@@
   \begin{enumerate}
-    \item $M \Downarrow$ iff $N \Downarrow$
+    \item $M \Downarrow$ iff $N \Downarrow$.
     \item If $M \Downarrow v_?$ and $N \Downarrow v'_?$ then either $v_? = \mho$, or $v_? = v'_?$.
@@
   same except that in some places where $M$ has explicit typing annotations,
-  $N$ has $\dyn$ instead.
+  $N$ has $\dyn$ instead. The relation $\ltdyn$ on terms is called ``term precision''.
+  Term precision is intended to model the situation where a term $M$ behaves the
+  same as another term $N$ except that $M$ may error more than $N$.
+  Term precision is defined by a set of axioms that capture this intuitive notion.
+  This will be described in more detail in Section \ref{sec:GTLC}.
+
   \item $\cdot \Downarrow$ is a relation on terms that is defined such that $M \Downarrow$ means
   that $M$ terminates, either with a run-time error or a value $n$ of type $\nat$.
+
   \item $\mho$ is a syntactic representation of a run-time type error, which
@@
   is found to be incompatible with the argument type of the function.
+
   \item $v_?$ is shorthand for the syntactic representation of a term that is either equal to
@@
 
-Our first step toward proving graduality is to formulate an \emph{step-sensitive}, or \emph{intensional},
-gradual lambda calculus, which we call $\intlc$, in which the computation steps taken by a term are made explicit.
+To prove graduality and validate the equational theory, we construct a model of the types
+and terms and show that all of the axioms for term precision and for equality of terms
+hold in this model. Modeling the dynamic type presents a challenge in the presence of a
+language with functions: we want the dynamic type to represent a sum of all possible types
+in the language, so we write down an recursive equation that the semantic object modeling
+dynamic type should satisfy. When the language includes function types, this equation involves a
+negative occurrence of the variable for which we are solving, and so the equation 
+does not have inductive or coinductive solutions.
+%
+To model the dynamic type, we therefore use guarded recursion to define a suitable
+semantic object that satisfies the unfolding isomorphism expected of the dynamic type.
+The key is that we do not actually get an exact solution to the equation in the style
+of traditional domain theory; rather, we get a ``guarded'' solution that holds ``up to a time step''.
+%
+That is, we introduce a notion of ``time'' and in the equation for the dynamic type,
+we guard the negative occurrences of the variable by a special operator that
+specifies that its argument is available ``later''.
+%This can be seen as a logic that internalizes the notion of step-indexing.
+See Section \ref{sec:technical-background} for more details on guarded type theory.
+
+At a high level, the key parts of our proof are as follows:
+
+% TODO revise this
+\begin{itemize}
+  \item Our first step toward proving graduality is to formulate a \emph{step-sensitive},
+  or \emph{intensional}, gradual lambda calculus, which we call $\intlc$, in which the
+  computation steps taken by a term are made explicit.
   The ``normal'' gradual lambda calculus for which we want to prove graduality will be called the
-\emph{step-insensitive}, or \emph{extensional}, gradual lambda calculus, denoted $\extlc$.
-We will define an erasure function $\erase{\cdot} : \intlc \to \extlc$ which takes a program in the intensional lambda calculus
-and ``forgets'' the syntactic information about the steps to produce a term in the extensional calculus.
+  \emph{surface}, \emph{step-insensitive}, or \emph{extensional}, gradual lambda calculus,
+  denoted $\extlc$.
 
-Every term $M_e$ in $\extlc$ will have a corresponding program $M_i$ in $\intlc$ such that
-$\erase{M_i} = M_e$. Moreover, we will show that if $M_e \ltdyn_e M_e'$ in the extensional theory,
-then there exists terms $M_i$ and $M_i'$ such that $\erase{M_i}=M_e$, $\erase{M_i'}=M_e'$ and
-$M_i \ltdyn_i M_i'$ in the intensional theory.
+  \item We define a translation from the surface syntax to the intensional syntax, and
+  prove a theorem relating the term precision in the surface to term precision in the
+  intensional syntax.
 
-We formulate and prove an analogous graduality theorem for the intensional lambda calculus.
-We define an interpretation of the intensional lambda calculus into a model in which we prove
-various results. Using the observation above, given $M_e \ltdyn M_e' : \nat$, we can find
-intensional programs $M_i$ and $M_i'$ that erase to them and are such that $M_i \ltdyn M_i'$.
-We will then apply the intensional graduality theorem to $M_i$ and $M_i'$, and translate the result
-back to $M_e$ and $M_e'$.
+  \item We define a semantics for the intensional syntax in guarded type theory, for both the
+  terms and for the term precision ordering $\ltdyn$.
 
-\subsection{Contributions}
-Our main contribution is a reusable framework in Guarded Cubical Agda for developing machine-checked proofs
-of graduality of a cast calculus.
-To demonstrate the feasability and utility of our approach, we have used the framework to prove
-graduality for the simply-typed gradual lambda calculus. Along the way, we have developed an ``intensional" theory
-of graduality that is of independent interest.
-
-\subsection{Overview of Remainder of Paper}
-
-In Section \ref{sec:technical-background}, we provide technical background on gradually typed languages and
-on synthetic guarded domain theory.
-% 
-In Section \ref{sec:GTLC}, we introduce the gradually-typed cast calculus
-for which we will prove graduality. Important here are the notions of syntactic
-type precision and term precision. We introduce both the \emph{extensional} gradual lambda calculus
-($\extlc$) and the \emph{intensional} gradual lambda calculus ($\intlc$).
-%
-In Section \ref{sec:domain-theory}, we define several fundamental constructions
-internal to SGDT that will be needed when we give a denotational semantics to
-our intensional lambda calculus.
-This includes the notion of Predomains as well as the concept
-of EP-Pairs.
-%
-In Section \ref{sec:semantics}, we define the denotational semantics for the
-intensional gradually-typed lambda calculus using the domain theoretic constructions in the
-previous section.
-%
-In Section \ref{sec:graduality}, we outline in more detail the proof of graduality for the
-extensional gradual lambda calculus, which will make use of prove properties we prove about
- the intensional gradual lambda calculus.
-%
-In Section \ref{sec:discussion}, we discuss the benefits and drawbacks to our approach in comparison
-to the traditional step-indexing approach, as well as possibilities for future work.
+\end{itemize}

### 2023-11-16 paper-new/ordering.tex
begin rearranging: discuss term syn/sem and then ordering syn/sem

\section{Extending the Semantics to Precision}\label{sec:gtlc-precision}

In this section, we extend the set-theoretic semantics for terms given in
the previous section to a semantics for the type and term precision relations
of the gradually-typed lambda calculus. We first introduce the type and term precision
relations, then show how to give them a semantics using SGDT.

% TODO mention intensional syntax

\subsection{Term Precision for GTLC}

% ---------------------------------------------------------------------------------------
% ---------------------------------------------------------------------------------------

%\subsubsection{Term Precision}\label{sec:term-precision}

We allow for a \emph{heterogeneous} term precision judgment on values $V$ of type
$A$ and $V'$ of type $A'$ provided that $A \ltdyn A'$ holds. Likewise, for producers,
if $M$ has type $A$ and $M'$ has type $A'$, we can form the judgment that $M \ltdyn M'$.
We use the same notation for the precision relation on both values and producers.

% Type precision contexts
In order to deal with open terms, we will need the notion of a type precision \emph{context}, which we denote
$\gamlt$. This is similar to a normal context but instead of mapping variables to types,
it maps variables $x$ to related types $A \ltdyn A'$, where $x$ has type $A$ in the left-hand term
and $A'$ in the right-hand term. We may also write $x : d$ where $d : A \ltdyn A'$ to indicate this.

% An equivalent way of thinking of type precision contexts is as a pair of ``normal" typing
% contexts $\Gamma, \Gamma'$ with the same domain such that $\Gamma(x) \ltdyn \Gamma'(x)$ for
% each $x$ in the domain.
% We will write $\gamlt : \Gamma \ltdyn \Gamma'$ when we want to emphasize the pair of contexts.
% Conversely, if we are given $\gamlt$, we write $\gamlt_l$ and $\gamlt_r$ for the normal typing contexts on each side.

An equivalent way of thinking of a type precision context $\gamlt$ is as a
pair of ``normal" typing contexts, $\gamlt_l$ and $\gamlt_r$, with the same
domain and such that $\gamlt_l(x) \ltdyn \gamlt_r(x)$ for each $x$ in the domain.
We will write $\gamlt : \gamlt_l \ltdyn \gamlt_r$ when we want to emphasize the pair of contexts.

As with type precision derivations, we write $\Gamma$ to mean the ``reflexivity" type precision context
$\Gamma : \Gamma \ltdyn \Gamma$.
Concretely, this consists of reflexivity type precision derivations $\Gamma(x) \ltdyn \Gamma(x)$ for
each $x$ in the domain of $\Gamma$.

Furthermore, we write $\gamlt_1 \relcomp \gamlt_2$ to denote the ``composition'' of $\gamlt_1$ and $\gamlt_2$
--- that is, the precision context whose value at $x$ is the type precision derivation
$\gamlt_1(x) \relcomp \gamlt_2(x)$. This of course assumes that each of the type precision
derivations is composable, i.e., that the RHS of $\gamlt_1(x)$ is the same as the left-hand side of $\gamlt_2(x)$.

% We define the same for computation type precision contexts $\deltalt_1$ and $\deltalt_2$,
% provided that both the computation type precision contexts have the same ``shape'', which is defined as
% (1) either the stoup is empty in both, or the stoup has a hole in both, say $\hole{d}$ and $\hole{d'}$
% where $d$ and $d'$ are composable, and (2) their value type precision contexts are composable as described above.

The rules for term precision come in two forms. We first have the \emph{congruence} rules,
one for each term constructor. These assert that the term constructors respect term precision.
The congruence rules are as follows:

\begin{mathpar}

  \inferrule*[right = Var]
    { c : A \ltdyn B \and \gamlt(x) = (A, B) } 
    { \etmprec {\gamlt} x x c }

  \inferrule*[right = Zro]
    { } {\etmprec \gamlt \zro \zro \nat }

  \inferrule*[right = Suc]
    { \etmprec \gamlt V {V'} \nat } {\etmprec \gamlt {\suc\, V} {\suc\, V'} \nat}

  \inferrule*[right = MatchNat]
  {\etmprec \gamlt V {V'} \nat \and 
    \etmprec \deltalt M {M'} d \and \etmprec {\deltalt, n : \nat} {N} {N'} d}
  {\etmprec \deltalt {\matchnat {V} {M} {n} {N}} {\matchnat {V'} {M'} {n} {N'}} d}

  \inferrule*[right = Lambda]
    { c_i : A_i \ltdyn A'_i \and 
      c_o : A_o \ltdyn A'_o \and 
      \etmprec {\gamlt, x : c_i} {M} {M'} {c_o} } 
    { \etmprec \gamlt {\lda x M} {\lda x {M'}} {(c_i \ra c_o)} }

  \inferrule*[right = App]
    { c_i : A_i \ltdyn A'_i \and
      c_o : A_o \ltdyn A'_o \\\\
      \etmprec \gamlt {V_f} {V_f'} {(c_i \ra c_o)} \and
      \etmprec \gamlt {V_x} {V_x'} {c_i}
    } 
    { \etmprec {\gamlt} {V_f\, V_x} {V_f'\, V_x'} {{c_o}}}

  \inferrule*[right = Ret]
    {\etmprec {\gamlt} V {V'} c}
    {\etmprec {\gamlt} {\ret\, V} {\ret\, V'} {c}}

  \inferrule*[right = Bind]
    {\etmprec {\gamlt} {M} {M'} {c} \and 
     \etmprec {\gamlt, x : c} {N} {N'} {d} }
    {\etmprec {\gamlt} {\bind {x} {M} {N}} {\bind {x} {M'} {N'}} {d}}
\end{mathpar}

We then have additional equational axioms, including $\beta$ and $\eta$ laws, and
rules characterizing upcasts as least upper bounds, and downcasts as greatest lower bounds.
For the sake of familiarity, we formulate the cast rules using arbitrary casts; later we
will state the analogous versions for the version of the calculus without arbitrary casts.

We write $M \equidyn N$ to mean that both $M \ltdyn N$ and $N \ltdyn M$.

\begin{mathpar}
  \inferrule*[right = $\err$]
    {\phasty {\Gamma} M B }
    {\etmprec {\Gamma} {\err_B} M B}

  \inferrule*[right = $\beta$-fun]
    { \phasty {\Gamma, x : A_i} M {A_o} \and
      \vhasty {\Gamma} V {A_i} } 
    { \etmequidyn {\Gamma} {(\lda x M)\, V} {M[V/x]} {A_o} }

  \inferrule*[right = $\eta$-fun]
    { \vhasty {\Gamma} {V} {A_i \ra A_o} } 
    { \etmequidyn \Gamma {\lda x (V\, x)} V {A_i \ra A_o} }

  \inferrule*[right = UpR]
    { c : A \ltdyn B \and d : B \ltdyn C \and 
      \etmprec {\gamlt} {M} {N} {c} } 
    { \etmprec {\gamlt} {M} {\up {B} {C} N} {c \circ d}  }

  \inferrule*[right = UpL]
    { c : A \ltdyn B \and d : B \ltdyn C \and
      \etmprec {\gamlt} {M} {N} {c \circ d} } 
    { \etmprec {\gamlt} {\up {A} {B} M} {N} {d} }

  \inferrule*[right = DnL]
    { c : A \ltdyn B \and d : B \ltdyn C \and
      \etmprec {\gamlt} {M} {N} {d} } 
    { \etmprec {\gamlt} {\dn {A} {B} M} {N} {c \circ d} }

  \inferrule*[right = DnR]
    { c : A \ltdyn B \and d : B \ltdyn C \and
      \etmprec {\gamlt} {M} {N} {c \circ d} } 
    { \etmprec {\gamlt} {M} {\dn {B} {C} N} {c} }
\end{mathpar}

% TODO explain the least upper bound/greatest lower bound rules
The rules UpR, UpL, DnL, and DnR were introduced in \cite{new-licata18} as a means
of cleanly axiomatizing the intended behavior of casts in a way that
doesn't depend on the specific constructs of the language.
Intuitively, rule UpR says that the upcast of $M$ is an upper bound for $M$
in that $M$ may error more, and UpL says that the upcast is the \emph{least}
such upper bound, in that it errors more than any other upper bound for $M$.
Conversely, DnL says that the downcast of $M$ is a lower bound, and DnR says
that it is the \emph{greatest} lower bound.
% These rules provide a clean axiomatization of the behavior of casts that doesn't
% depend on the specific constructs of the language.

\subsection{Semantics for Precision}

As a first attempt at giving a semantics to the ordering, we could try to model types as
sets equipped with an ordering that models term precision. Since term precision is reflexive
and transitive, and since we identify terms that are equi-precise, we choose to model types
as partially-ordered sets. We model the term precision ordering $M \ltdyn N : A \ltdyn B$ as an
ordering relation between the posets denoted by $A$ and $B$.

However, it turns out that modeling term precision by a relation defined by guarded fixpoint
is not as straightforward as one might hope.
A first attempt might be to define an ordering $\semltbad$ between $\li X$ and $\li Y$
that allows for computations that may take different numbers of steps to be related.
The relation is parameterized by a relation $\le$ between $X$ and $Y$, and is defined
by guarded fixpoint as follows:
% simultaneously captures the notions of error approximation and equivalence up to stepping behavior:

\begin{align*}
  &\eta\, x \semltbad \eta\, y \text{ if } 
    x \semlt y \\
%		
  &\mho \semltbad l \\
%
  &\theta\, \tilde{l} \semltbad \theta\, \tilde{l'} \text{ if } 
    \later_t (\tilde{l}_t \semltbad \tilde{l'}_t) \\
%	
  &\theta\, \tilde{l} \semltbad \mho \text{ if } 
    \theta\, \tilde{l} = \delta^n(\mho) \text { for some $n$ } \\
%	
  &\theta\, \tilde{l} \semltbad \eta\, y \text{ if }
    (\theta\, \tilde{l} = \delta^n(\eta\, x))
  \text { for some $n$ and $x : \ty{X}$ such that $x \le y$ } \\
%
  &\mho \semltbad \theta\, \tilde{l'} \text { if } 
    \theta\, \tilde{l'} = \delta^n(\mho) \text { for some $n$ } \\
%	
  &\eta\, x \semltbad \theta\, \tilde{l'} \text { if }
    (\theta\, \tilde{l'} = \delta^n (\eta\, y))
  \text { for some $n$ and $y : \ty{Y}$ such that $x \le y$ }
\end{align*}

Two computations that immediately return $(\eta)$ are related if the underlying
values are related in the underlying ordering. 
%
The computation that errors $(\mho)$ is below everything else.
%
If both sides step (i.e., both sides are $\theta$),
then we allow one time step to pass and compare the resulting terms.
(This is where use the relation defined ``later''.)
%
Lastly, if one side steps and the other returns a value, the side that steps should
terminate with a value in some finite number of steps $n$, and that value should
be related to the value returned by the other side.
Likewise, if one side steps and the other errors, then the side that steps
should terminate with error.

The problem with this definition is that the resulting relation is \emph{provably} not
transitive: it can be shown (in Clocked Cubical Type Theory) that if $R$ is a
relation on $\li X$ satisfying three specific properties, one of which is
transitivity, then that relation is trivial.
(The other two properties are that the relation is a congruence with respect to $\theta$,
and that the relation is closed under delays $\delta = \theta \circ \nxt$ on either side.)
Since the above relation \emph{does} satisfy the other two properties, we conclude
that it must not be transitive.

%But having a non-transitive relation to model term precision presents a problem
%for...

We are therefore led to wonder whether we can formulate a version of the relation
that \emph{is} transitive.
It turns out that we can, by sacrificing another of the three properties from
the above lemma. Namely, we give up on closure under delays. Doing so, we end up
with a \emph{lock-step} error ordering, where, roughly speaking, in order for
computations to be related, they must have the same stepping behavior.
%
We then formulate a separate relation, \emph{weak bisimilarity}, that relates computations
that are extensionally equal and may only differ in their stepping behavior.

% As a result, we instead separate the semantics of term precision into two relations:
% an intensional, step-sensitive \emph{error ordering} and a \emph{bisimilarity relation}.

\subsubsection{Double Posets}\label{sec:predomains}

As discussed above, there are two relations that we would like to define
in the semantics: a step-sensitive error ordering, and weak bisimilarity of computations.
%
The semantic objects that interpret our types should therefore be equipped with
two relations. We call these objects ``double posets''.
A double poset $A$ is a set with two relations: an partial order $\semlt_A$ on $A$, and
a reflexive, symmetric relation $\bisim_A$ on $A$.
We write the underling set of $A$ as $\ty{A}$.

We define morphisms of double posets as functions that preserve both
the ordering and the bisimilarity relation. Given double posets
$A$ and $B$, we write $f \colon A \monto B$ to indicate that $f$ is a morphism
from $A$ to $B$, i.e, the following hold:
(1) for all $a_1 \semlt_A a_2$, we have $f(a_1) \semlt_{B} f(a_2)$, and
(2) for all $a_1 \bisim_A a_2$, we have $f(a_1) \bisim_{B} f(a_2)$.

%%%%% RESUME HERE

We define an ordering on morphisms of double posets as
$f \le g$ if for all $a$ in $\ty{A_i}$, we have $f(a) \le_{A_o} g(a)$,
and similarly bisimilarity extends to morphisms via
$f \bisim g$ if for all $a$ in $\ty{A_i}$, we have $f(a) \bisim_{A_o} g(a)$.

For each type we want to represent, we define a double poset for the corresponding semantic
type. For instance, we define a double poset for natural numbers, for the
dynamic type, for functions, and for the lift monad. We
describe each of these below.

\begin{itemize}
  \item There is a double poset $\Nat$ for natural numbers, where the ordering and the
  bisimilarity relations are both equality.

  % TODO explain that there is a theta operator for posets?
  \item There is a double poset $\Dyn$ to represent the dynamic type. The underlying type
  for this double poset is defined by guarded fixpoint to be such that
  $\ty{\Dyn} \cong \mathbb{N}\, +\, \later (\ty{\Dyn} \monto \li \ty{\Dyn})$.
  This definition is valid because the occurrences of Dyn are guarded by the $\later$.
  The ordering is defined via guarded recursion by cases on the argument, using the
  ordering on $\mathbb{N}$ and the ordering on monotone functions described above.

  \item For a double poset $A$, there is a double poset $\li A$ for the ``lift'' of $A$
  using the lift monad. We use the same notation for $\li A$ when $A$ is a type
  and $A$ is a double poset, since the context should make clear which one we are referring to.
  The underling type of $\li A$ is simply $\li \ty{A}$, i.e., the lift of the underlying
  type of $A$.
  The ordering on $\li A$ is the ``lock-step error-ordering'' which we describe in
  \ref{subsec:lock-step}. The bismilarity relation is the ``weak bisimilarity''
  described in Section \ref{}

  \item For double posets $A_i$ and $A_o$, we form the double poset of monotone functions
  from $A_i$ to $A_o$, which we denote by $A_i \To A_o$.

  \item Given a double poset $A$, we can form the double poset $\later A$ whose underlying
  type is $\later \ty{A}$. We define $\tilde{x} \le_{\later A} \tilde{y}$ to be
  $\later_t (\tilde{x}_t \le_A \tilde{y}_t)$.
\end{itemize}

\subsubsection{Step-Sensitive Error Ordering}\label{subsec:lock-step}

As mentioned, the ordering on the lift of a double poset $A$ is called the
\emph{step-sensitive error-ordering} (also called ``lock-step error ordering''),
the idea being that two computations $l$ and $l'$ are related if they are in
lock-step with regard to their intensional behavior, up to $l$ erroring.
Formally, we define this ordering as follows:

\begin{itemize}
  \item 	$\eta\, x \ltls \eta\, y$ if $x \le_A y$.
  \item 	$\mho \ltls l$ for all $l$ 
  \item   $\theta\, \tilde{r} \ltls \theta\, \tilde{r'}$ if
          $\later_t (\tilde{r}_t \ltls \tilde{r'}_t)$
\end{itemize}

We also define a heterogeneous version of this ordering between the lifts of two
different double posets $A$ and $B$, parameterized by a relation $R$ between $A$ and $B$.

\subsubsection{Weak Bisimilarity Relation}

For a double poset $A$, we define a relation on $\li A$, called ``weak bisimilarity",
written $l \bisim l'$. Intuitively, we say $l \bisim l'$ if they are equivalent
``up to delays''.
% We introduce the notation $x \sim_A y$ to mean $x \le_A y$ and $y \le_A x$.
% TODO if A is a poset, then we can just say that x = y
%
The weak bisimilarity relation is defined by guarded fixpoint as follows:

\begin{align*}
  &\mho \bisim \mho \\
%
  &\eta\, x \bisim \eta\, y \text{ if } 
    x \bisim_A y \\
%		
  &\theta\, \tilde{x} \bisim \theta\, \tilde{y} \text{ if } 
    \later_t (\tilde{x}_t \bisim \tilde{y}_t) \\
%	
  &\theta\, \tilde{x} \bisim \mho \text{ if } 
    \theta\, \tilde{x} = \delta^n(\mho) \text { for some $n$ } \\
%	
  &\theta\, \tilde{x} \bisim \eta\, y \text{ if }
    (\theta\, \tilde{x} = \delta^n(\eta\, x))
  \text { for some $n$ and $x : \ty{A}$ such that $x \sim_A y$ } \\
%
  &\mho \bisim \theta\, \tilde{y} \text { if } 
    \theta\, \tilde{y} = \delta^n(\mho) \text { for some $n$ } \\
%	
  &\eta\, x \bisim \theta\, \tilde{y} \text { if }
    (\theta\, \tilde{y} = \delta^n (\eta\, y))
  \text { for some $n$ and $y : \ty{A}$ such that $x \sim_A y$ }
\end{align*}

When both sides are $\eta$, then we ensure that the underlying values are bisimilar
in the underlying bisimilarity relation on $A$.
When one side is a $\theta$ and the other is $\eta x$ (i.e., one side steps),
we stipulate that the $\theta$-term runs to $\eta y$ where $x$ is related to $y$.
Similarly when one side is $\theta$ and the other $\mho$.
If both sides step, then we allow one time step to pass and compare the resulting terms.
In this way, the definition captures the intuition of terms being equivalent up to
delays.

It can be shown (by \lob-induction) that the step-sensitive relation is symmetric.
However, it can also be shown that this relation is \emph{not} transitive:
The argument is the same as that used to show that the step-insensitive error
ordering $\semltbad$ described above is not transitive. Namely, we show that
if it were transitive, then it would have to be trivial in that $l \bisim l'$ for all $l, l'$.
that if this relation were transitive, then in fact it would be trivial in that
%This issue will be resolved when we consider the relation's \emph{globalization}.

\subsection{The Cast Rules}

Unfortunately, the four cast rules defined above do not hold in
the intensional setting where we are tracking the steps taken by terms.
The source of the problem is that the downcast from the dynamic type to
a function involves a delay, i.e., a $\theta$.
So in order to keep the other term in lock-step, we need to insert a ``delay"
that is extensionally equivalent to the identity function.
More concretely, consider a simplified version of the DnL rule shown below:

\begin{mathpar}
  \inferrule*{M \ltdyn_i N : B}
             {\dn c M \ltdyn_i N : c}
\end{mathpar}

If $c$ is inj-arr, then when we downcast $M$ from $dyn$ to $\dyn \ra \dyn$,
semantically this will involve a $\theta$ because the value of type $dyn$
in the semantics will contain a \emph{later} function $\tilde{f}$.
Thus, in order for the right-hand side to be related to the downcast,
we need to insert a delay on the right.
%
The need for delays affects the cast rules involving upcasts as well, because
the upcast for functions involves a downcast on the domain:

\[ \up{c_i \ra c_o}{M} \equiv \lambda (x : B_i). \up{c_o}(M\, (\dn {c_i} x)). \]

Thus, the correct versions of the cast rules involve delays on the side that was not casted.

% Delays for function types and for inj-arr(c)

\subsubsection{Perturbations}

We can describe precisely how the delays are inserted for any type precision
derivation $c$.

To do so, we first define simultaneously an inductive type of \emph{perturbations}
for embeddings $\perte$ and for projections $\pertp$ by the following rules:

\begin{mathpar}

\inferrule{}{\id : \perte A}

\inferrule{}{\id : \pertp A}

\inferrule
  {\delta_c : \pertp A \and \delta_d : \perte B}
  {\delta_c \ra \delta_d : \perte (A \ra B)}

\inferrule
  {\delta_c : \perte A \and \delta_d : \pertp B}
  {\delta_c \ra \delta_d : \pertp (A \ra B)}

\inferrule
  {\delta_\nat : \perte \nat \and \delta_f : \perte (\dyn \ra \dyn)}
  {\pertdyn{\delta_\nat}{\delta_f} : \perte \dyn}

\inferrule
  {\delta_\nat : \pertp \nat \and \delta_f : \pertp (\dyn \ra \dyn)}
  {\pertdyn{\delta_\nat}{\delta_f} : \pertp \dyn}

\end{mathpar}

The structure of embedding perturbations is designed to follow the structure
of the corresponding embeddings, and likewise for the projection perturbations.
Thus, in the function case, an embedding perturbation consists of a \emph{projection}
perturbation for the domain and an \emph{embedding} perturbation for the codomain.
The opposite holds for the projection perturbation for functions.

Another way in which the two kinds of perturbations differ is that there is an additional
projection perturbation for delaying $\delaypert{\delta}$.
This corresponds to the actual delay term $\delta = \theta \circ \nxt$ in the semantics,
and it is the generator/source of all non-trivial perturbations.

Given a perturbation $\delta$, we can turn it into a term, which we also write as
$\delta$ unless there is opportunity for confusion.

### 2023-11-16 paper-new/technical-background.tex
begin rearranging: discuss term syn/sem and then ordering syn/sem

@@
 
-New and Licata \cite{newlicata} developed an axiomatic account of the
+New and Licata \cite{new-licata18} developed an axiomatic account of the
 graduality relation on cast calculus terms and gave a denotational
@@
 
-A series of works \cite{newahmed,newlicataahmed,newjamnerahmed}
+A series of works \cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
 developed step-indexed logical relations models of gradually typed
@@
 higher-order store and runtime-extensible dynamic types
-\cite{amalsthesis,nonpmetricparamorsomething,newjamnerahmed}. However
+\cite{amalsthesis,nonpmetricparamorsomething,new-jamner-ahmed19}. However
 their proof developments are highly repetitive and technical, with
@@
   to restore well-foundedness. This technique is known as a \emph{step-indexed logical relation}.
+  Intuitively, step-indexing makes the steps taken by terms observable.
+  %
   Reasoning about step-indexed logical relations
@@
   related in both of the one-sided logical relations. Having two separate logical relations
-  complicates the statement of the lemmas used to prove graduality, becasue any statement that
+  complicates the statement of the lemmas used to prove graduality, because any statement that
   involves a term stepping needs to take into account whether we are counting steps on the left
   or the right. Some of the differences can be abstracted over, but difficulties arise for properties %/results
-  as fundamental and seemingly straightforward as transitivty.
+  as fundamental and seemingly straightforward as transitivity.
 
@@
   index $i$ and $N$ is related to $P$ at index $i$, then $M$ is related to $P$ at $i$.
-  But this does not actually hold: we requrie that one of the two pairs of terms
+  But this does not actually hold: we require that one of the two pairs of terms
   be related ``at infinity'', i.e., that they are related at $i$ for all $i \in \mathbb{N}$.
@@
   be related for all $i$. % This may not even be possible to show in some scenarios!
+
+  \begin{comment}
+  As another example, the axioms that specify the behavior of casts do not hold in the
+  step-indexed setting. Consider, for example, the ``lower bound'' rule for downcasting:
+
+  \begin{mathpar}
+    \inferrule*
+    {\gamlt \vdash M \ltdyn N : dyn}
+    {\gamlt \vdash \dn{\dyn \ra \dyn}{\dyn} M \ltdyn N}
+  \end{mathpar}
+
+  In the language of the step-indexed logical relation used in prior work, this would
+  take the form
+
+  \begin{mathpar}
+    \inferrule*
+    {(M, N) \in \mathcal{E}^{\sim}_{i}\sem{\dyn}}
+    {(\dn{\dyn \ra \dyn}{\dyn} M, N) \in \mathcal{E}^{\sim}_{i}\sem{\injarr{}}}
+  \end{mathpar}
+
+  where $\sim$ stands for $\mathrel{\preceq}$ or $\mathrel{\succeq}$, i.e.,
+  counting steps on the left or right respectively.
+  %
+  To show this, we would use the fact that the left-hand side steps and
+  apply an anti-reduction lemma, showing that the term to which the LHS steps
+  is related to the RHS where our fuel is now
+
+  The left-hand side steps to a case inspection on $M$,
+  where we unfold the recursive $\dyn$ type into a sum type and see whether the result
+  is a function type.
+
+  One way around these difficulties is to demand that the rules only hold
+  ``extensionally'', i.e., we quantify universally over the step-index and
+  reason about the ``global'' behavior of terms for all possible step indices.
+  This is the approach taken in prior work.
+\end{comment}
 
@@
 
-In Ticked Cubical Type Theory \cite{TODO}, there is an additional sort
-called \emph{ticks}. Given a clock $k$, a tick $t : \tick k$ serves
-as evidence that one unit of time has passed according to the clock $k$.
+Ticked Cubical Type Theory \cite{mogelberg-veltri2019} is an extension of Cubical
+Type Theory \cite{CohenCoquandHuberMortberg2017}
+in which there is an additional sort called \emph{ticks}. Given a clock $k$,
+a tick $t : \tick\, k$ serves as evidence that one unit of time has passed according to the clock $k$.
 The type $\later A$ is represented as a function from ticks of a clock $k$ to $A$.
@@
 In particular, if we have a term $M$ of type $\later^k A$, and we
-have available in the context a tick $t' : \tick k$, then we can apply the tick to $M$ to get
+have available in the context a tick $t' : \tick\, k$, then we can apply the tick to $M$ to get
 a term $M[t'] : A[t'/t]$. We will also write tick application as $M_t$.
-Conversely, if in a context $\Gamma, t : \tick k$ we have that $M$ has type $A$,
+Conversely, if in a context $\Gamma, t : \tick\, k$ we have that $M$ has type $A$,
 then in context $\Gamma$ we have $\lambda t.M$ has type $\later A$. % TODO dependent version?
@@
 The statements in this paper have been formalized in a variant of Agda called
-Guarded Cubical Agda \cite{TODO}, an implementation of Clocked Cubical Type Theory.
+Guarded Cubical Agda \cite{veltri-vezzosi2020}, an implementation of Clocked Cubical Type Theory.
 

### 2023-11-16 paper-new/terms.tex
begin rearranging: discuss term syn/sem and then ordering syn/sem

\section{A Simple Denotational Semantics for the Terms of GTLC}\label{sec:gtlc-terms}

In this section, we introduce the term syntax for the gradually-typed
lambda calculus (GTLC) and give a set-theoretic denotational semantics
using tools from SGDT.

In the following section, we will extend the denotational semantics to accommodate
the type and term precision orderings.

\subsection{Syntax}\label{sec:term-syntax}

Our syntax is based on fine-grained call by value, and as such it has
separate value and producer terms and typing judgments for each.

% Given a term $M$ of type $A$, the term $\bind{x}{M}{N}$ should be thought of as
% running $M$ to a value $V$ and then continuing as $N$, with $V$ in place of $x$.

\begin{align*}
  &\text{Types } A := \nat \alt \,\dyn \alt (A \ra A') \\
  &\text{Contexts } \Gamma := \cdot \alt (\Gamma, x : A) \\
  &\text{Values } V :=  \zro \alt \suc\, V \alt \lda{x}{M} \alt \up{A}{B} V \\ 
  &\text{Producers } M, N := \err_B \alt \matchnat {V} {M} {n} {M'} \\ 
  &\quad\quad \alt \ret {V} \alt \bind{x}{M}{N} \alt V_f\, V_x \alt \dn{A}{B} M 
\end{align*}

The value typing judgment is written $\vhasty{\Gamma}{V}{A}$ and 
the producer typing judgment is written $\phasty{\Gamma}{M}{A}$.

The typing rules are as expected, with a cast between $A$ to $B$ allowed only when $A \ltdyn B$.
The precise rules for $A \ltdyn B$ will be given below.
Notice that the upcast of a value is a value, since it always succeeds, while the downcast
of a value is a producer, since it may fail.

\begin{mathpar}
    % Var
    \inferrule*{ }{\vhasty {\cdot, \Gamma, x : A, \Gamma'} x A}

    % Err
    \inferrule*{ }{\phasty {\cdot, \Gamma} {\err_A} A}

    % Zero and suc
    \inferrule*{ }{\vhasty \Gamma \zro \nat}

    \inferrule*{\vhasty \Gamma V \nat} {\vhasty \Gamma {\suc\, V} \nat}

    % Match-nat
    \inferrule*
    {\vhasty \Gamma V \nat \and 
     \phasty \Delta M A \and \phasty {\Gamma, n : \nat} {M'} A}
    {\phasty \Gamma {\matchnat {V} {M} {n} {M'}} A}

    % Lambda
    \inferrule* 
    {\phasty {\Gamma, x : A} M {A'}} 
    {\vhasty \Gamma {\lda x M} {A \ra A'}}

    % App
    \inferrule*
    {\vhasty \Gamma {V_f} {A \ra A'} \and \vhasty \Gamma {V_x} A}
    {\phasty {\Gamma} {V_f \, V_x} {A'}}

    % Ret
    \inferrule*
    {\vhasty \Gamma V A}
    {\phasty {\Gamma} {\ret\, V} {A}}

    % Bind
    \inferrule*
    {\phasty \Gamma M {A} \and \phasty{\Gamma, x : A}{N}{B} } % Need x : A in context
    {\phasty {\Gamma} {\bind{x}{M}{N}} {B}}

    % Upcast
    \inferrule*
    {A \ltdyn A' \and \vhasty \Gamma V A}
    {\vhasty \Gamma {\up A {A'} V} {A'} }

    \inferrule* % TODO is this correct?
    {A \ltdyn A' \and \phasty {\Gamma} {M} {A'}}
    {\phasty {\Gamma} {\dn A {A'} M} {A}}

\end{mathpar}

In the equational theory, we have $\beta$ and $\eta$ laws for function type,
as well a $\beta$ and $\eta$ law for bind.

\begin{mathpar}
  % Function Beta and Eta
  \inferrule*
  {\phasty {\Gamma, x : A} M {B} \and \vhasty \Gamma V A}
  {(\lda x M)\, V = M[V/x]}

  \inferrule*
  {\vhasty \Gamma V {A \ra A}}
  {\Gamma \vdash V = \lda x {V\, x}}

  % Ret Beta and Eta
  \inferrule*
  {}
  {(\bind{x}{\ret\, V}{N}) = N[V/x]}

  \inferrule*
  {\phasty {\Gamma} {M} {B}}
  {\bind{x}{M}{\ret x} = M}

  % Match-nat Beta
  \inferrule*
  {\phasty \Delta M A \and \phasty {\Gamma, n : \nat} {M'} A}
  {\matchnat{\zro}{M}{n}{M'} = M}

  \inferrule*
  {\vhasty \Gamma V \nat \and 
   \phasty \Gamma M B \and \phasty {\Gamma, n : \nat} {M'} B}
  {\matchnat{\suc\, V}{M}{n}{M'} = M'}

  % Match-nat Eta
  % This doesn't build in substitution
  \inferrule*
  {\hasty {\Gamma , x : \nat} M A}
  {M = \matchnat{x} {M[\zro / x]} {n} {M[(\suc\, n) / x]}}

\end{mathpar}

\subsubsection{Type Precision}\label{sec:type-precision}

The type precision rules specify what it means for a type $A$ to be more precise than $A'$.
We have reflexivity rules for $\dyn$ and $\nat$, as well as rules that $\nat$ is more precise than $\dyn$
and $\dyn \ra \dyn$ is more precise than $\dyn$.
We also have a congruence rule for function types stating that given $A_i \ltdyn A'_i$ and $A_o \ltdyn A'_o$, we can prove
$A_i \ra A_o \ltdyn A'_i \ra A'_o$. Note that precision is covariant in both the domain and codomain.
Finally, we can lift a relation on value types $A \ltdyn A'$ to a relation $\Ret A \ltdyn \Ret A'$ on
computation types.

\begin{mathpar}
  \inferrule*[right = \dyn]
    { }{\dyn \ltdyn\, \dyn}

  \inferrule*[right = \nat]
    { }{\nat \ltdyn \nat}

  \inferrule*[right = $\ra$]
    {A_i \ltdyn A'_i \and A_o \ltdyn A'_o }
    {(A_i \ra A_o) \ltdyn (A'_i \ra A'_o)}

  \inferrule*[right = $\textsf{Inj}_\nat$]
    { }{\nat \ltdyn\, \dyn}

  \inferrule*[right = $\textsf{Inj}_{\ra}$]
    { }
    {(\dyn \ra \dyn) \ltdyn\, \dyn}

  \inferrule*[right = $\injarr{}$]
    {(R \ra S) \ltdyn\, (\dyn \ra \dyn)}
    {(R \ra S) \ltdyn\, \dyn}

\end{mathpar}

We can prove that transitivity of type precision is admissible, i.e.,
if $A \ltdyn B$ and $B \ltdyn C$, then $A \ltdyn C$.

% Type precision derivations
Note that as a consequence of this presentation of the type precision rules, we
have that if $A \ltdyn A'$, there is a unique precision derivation that witnesses this.
As in previous work, we go a step farther and make these derivations first-class objects,
known as \emph{type precision derivations}.
Specifically, for every $A \ltdyn A'$, we have a derivation $c : A \ltdyn A'$ that is constructed
using the rules above. For instance, there is a derivation $\dyn : \dyn \ltdyn \dyn$, and a derivation
$\nat : \nat \ltdyn \nat$, and if $c_i : A_i \ltdyn A_i$ and $c_o : A_o \ltdyn A'_o$, then
there is a derivation $c_i \ra c_o : (A_i \ra A_o) \ltdyn (A'_i \ra A'_o)$. Likewise for
the remaining rules. The benefit to making these derivations explicit in the syntax is that we
can perform induction over them.
Note also that for any type $A$, we use $A$ to denote the reflexivity derivation that $A \ltdyn A$,
i.e., $A : A \ltdyn A$.
Finally, observe that for type precision derivations $c : A \ltdyn A'$ and $c' : A' \ltdyn A''$, we
can define their composition $c \relcomp c' : A \ltdyn A''$.
This notion will be used below in the statement of transitivity of the term precision relation.

\subsection{Removing Casts as Primitives}

% We now observe that all casts, except those between $\nat$ and $\dyn$
% and between $\dyn \ra \dyn$ and $\dyn$, are admissible, in the sense that
% we can start from $\extlcm$, remove casts except the aforementioned ones,
% and in the resulting language we will be able to derive the other casts.

We now observe that all casts, except those between $\nat$ and $\dyn$
and between $\dyn \ra \dyn$ and $\dyn$, are admissible.
That is, consider a new language ($\extlcprime$) in which
instead of having arbitrary casts, we have injections from $\nat$ and
$\dyn \ra \dyn$ into $\dyn$, and a case inspection on $\dyn$.
We claim that in $\extlcprime$, all of the casts present in $\extlc$ are derivable.
It will suffice to verify that casts for function type are derivable.
This holds because function casts are constructed inductively from the casts
of their domain and codomain. The base case is one of the casts involving $\nat$
or $\dyn \ra \dyn$ which are present in $\extlcprime$ as injections and case inspections.

The resulting calculus $\extlcprime$ now lacks arbitrary casts as a primitive notion:

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TODO update

\begin{align*}
  &\text{Types } A := \nat \alt \dyn \alt (A \ra A') \\
  &\text{Contexts } \Gamma := \cdot \alt (\Gamma, x : A) \\
  &\text{Values } V :=  \zro \alt \suc\, V \alt \lda{x}{M} \alt \injnat V \alt \injarr V \\ 
  &\text{Producers } M, N := \err_B \alt \ret {V} \alt \bind{x}{M}{N}
    \alt V_f\, V_x \alt
    \\ & \quad\quad \casenat{V}{M_{no}}{n}{M_{yes}} 
    \alt \casearr{V}{M_{no}}{f}{M_{yes}}
\end{align*}

% New rules
Figure \ref{fig:extlc-minus-minus-typing} shows the new typing rules,
and Figure \ref{fig:extlc-minus-minus-eqns} shows the equational rules
for case-nat (the rules for case-arrow are analogous).

\begin{figure}
  \begin{mathpar}
      % inj-nat
      \inferrule*
      {\hasty \Gamma M \nat}
      {\hasty \Gamma {\injnat M} \dyn}

      % inj-arr 
      \inferrule*
      {\hasty \Gamma M (\dyn \ra \dyn)}
      {\hasty \Gamma {\injarr M} \dyn}

      % Case nat
      \inferrule*
      {\hasty{\Delta|_V}{V}{\dyn} \and 
        \hasty{\Delta , x : \nat }{M_{yes}}{B} \and 
        \hasty{\Delta}{M_{no}}{B}}
      {\hasty {\Delta} {\casenat{V}{M_{no}}{n}{M_{yes}}} {B}}

      % Case arr
      \inferrule*
      {\hasty{\Delta|_V}{V}{\dyn} \and 
        \hasty{\Delta , x : (\dyn \ra \dyn) }{M_{yes}}{B} \and 
        \hasty{\Delta}{M_{no}}{B}}
      {\hasty {\Delta} {\casearr{V}{M_{no}}{f}{M_{yes}}} {B}}
  \end{mathpar}
  \caption{New typing rules for $\extlcmm$.}
  \label{fig:extlc-minus-minus-typing}
\end{figure}

\begin{figure}
  \begin{mathpar}
     % Case-nat Beta
     \inferrule*
     {\hasty \Gamma V \nat}
     {\casenat {\injnat {V}} {M_{no}} {n} {M_{yes}} = M_{yes}[V/n]}

     \inferrule*
     {\hasty \Gamma V {\dyn \ra \dyn} }
     {\casenat {\injarr {V}} {M_{no}} {n} {M_{yes}} = M_{no}}

     % Case-nat Eta
     \inferrule*
     {}
     {\Gamma , x :\, \dyn \vdash M = \casenat{x}{M}{n}{M[(\injnat{n}) / x]} }

     % Case-arr Beta

     % Case-arr Eta

  \end{mathpar}
  \caption{New equational rules for $\extlcprime$ (rules for case-arrow are analogous
           and hence are omitted).}
  \label{fig:extlc-minus-minus-eqns}
\end{figure}

\section{Term Semantics}\label{sec:term-semantics}

\subsection{Domain-Theoretic Constructions}\label{sec:domain-theory}

In this section, we discuss the fundamental objects of the model into which we will embed
the gradual lambda calculus.
It is important to remember that the constructions in this section are entirely
independent of the syntax described in the previous section; the notions defined 
here exist in their own right as purely mathematical constructs.
In Section \ref{sec:term-interpretation}, we will link the syntax and semantics
via a semantic interpretation function.

\subsection{Modeling the Dynamic Type}

When modeling the dynamic type $\dyn$, we need a semantic object $D$ that satisfies the
isomorphism

\[ D \cong \Nat + (D \to (D + 1)). \]

where the $D + 1$ represents the fact that in the function case, the function may return an error.
Unfortunately, this equation does not have inductive or coinductive solutions. The usual way of
dealing with such equations is via domain theory, by which we can obtain an exact solution.
However, the heavy machinery of domain theory can be difficult for language designers to learn
and apply in the mechanized setting.
Instead, we will leverage the tools of guarded type theory, considering instead the following
similar-looking equation:

\[ D \cong \Nat + \later (D \to (D + 1)). \]

Since the negative occurrence of $D$ is guarded under a later, this equation has a (guarded) solution.
Specifically, we consider the following function $f$ of type
$\later \type \to \type$:

\[ \lambda (D' : \later \type) . \Nat + \later_t (D'_t \to (D'_t + 1)). \]

(Recall that the tick $t : \tick$ is evidence that time has passed, and since
$D'$ has type $\later \type$, i.e. $\tick \to \type$, then $D'_t$ has type $\type$.)

Then we define

\[ D = \fix f. \]

% TODO explain better
As it turns out, this definition is not quite correct; let us try to understand why.
The price we pay for such a simple solution to the above equation is that we have
introduced a notion of ``time'', i.e., we must now consider the stepping behavior of terms.
% Therefore, the semantics of terms will need to allow for terms that potentially
% do not terminate. This is accomplished using an extension of the guarded lift monad,
% which we describe in the next section.
Thus, in the equation for the semantics of $\dyn$, the result of the function should be
a computation that may return a value, error, \emph{or} take an observable step.
We model such computations using an extension of the so-called guarded lift monad
\cite{mogelberg-paviotti2016} which we describe in the next section.
We will then discuss the correct definition of the semantics of the dynamic type.

\subsubsection{The Lift + Error Monad}

% TODO ensure the previous section flows into this one
When thinking about how to model gradually-typed programs where we track the steps they take,
we should consider their possible behaviors. On the one hand, we have \emph{failure}: a program may fail
at run-time because of a type error. In addition to this, a program may ``think'',
i.e., take a step of computation. If a program thinks forever, then it never returns a value,
so we can think of the idea of thinking as a way of intensionally modelling \emph{partiality}.

With this in mind, we can describe a semantic object that models these behaviors: a monad
for embedding computations that has cases for failure and ``thinking''.
Previous work has studied such a construct in the setting of the latter, called the lift
monad \cite{mogelberg-paviotti2016}; here, we augment it with the additional effect of failure.

For a type $A$, we define the \emph{lift monad with failure} $\li A$, which we will just call
the \emph{lift monad}, as the following datatype:

\begin{align*}
  \li A &:= \\
  &\eta \colon A \to \li A \\
  &\mho \colon \li A \\
  &\theta \colon \later (\li A) \to \li A
\end{align*}

Unless otherwise mentioned, all constructs involving $\later$ or $\fix$
are understood to be with respect to a fixed clock $k$. So for the above, we really have for each
clock $k$ a type $\li^k A$ with respect to that clock.

Formally, the lift monad $\li A$ is defined as the solution to the guarded recursive type equation

\[ \li A \cong A + 1 + \later \li A. \]

An element of $\li A$ should be viewed as a computation that can either (1) return a value (via $\eta$),
(2) raise an error and stop (via $\mho$), or (3) think for a step (via $\theta$).
%
Notice there is a computation $\fix \theta$ of type $\li A$. This represents a computation
that runs forever and never returns a value.

Since we claimed that $\li A$ is a monad, we need to define the monadic operations
and show that they respect the monadic laws. The return is just $\eta$, and extend
is defined via guarded recursion by cases on the input.
% It is instructive to give at least one example of a use of guarded recursion, so
% we show below how to define extend:
% TODO
%
%
Verifying that the monadic laws hold uses \lob-induction and is straightforward.

%\subsubsection{Model-Theoretic Description}
%We can describe the lift monad in the topos of trees model as follows.

\subsubsection{Revisiting the Dynamic Type}
Now we can state the correct definition of the semantics for the dynamic type.
The set $D$ is defined to be the solution to the guarded equation

\[ D \cong \Nat + \later (D \to \textcolor{red}{\li} D). \]

\subsection{Interpretation}\label{sec:term-interpretation}

We can now give a semantics to the gradual lambda calculus we defined
in Section \ref{sec:step-sensitive-lc}.
%
Much of the semantics is similar to a normal call-by-value denotational semantics;
we highlight the differences.
We will interpret types as sets, and terms as functions.
Contexts $\Gamma = x_1 \colon A_1, \dots, x_n \colon A_n$
will be interpreted as the product $\sem{A_1} \times \cdots \times \sem{A_n}$.

The semantics of the dynamic type $\dyn$ is the set $\Dyn$ introduced in Section
\ref{sec:predomains}.
%
The interpretation of a value $\vhasty {\Gamma} V A$ will be a monotone function from
$\sem{\Gamma}$ to $\sem{A}$. Likewise, a term $\phasty {\Gamma} M {{A}}$ will be interpreted
as a monotone function from $\sem{\Gamma}$ to $\li \sem{A}$.

Recall that $\Dyn$ is isomorphic to $\Nat\, + \later (\Dyn \to \li \Dyn)$.
Thus, the semantics of $\injnat{\cdot}$ is simply $\inl$ and the semantics
of $\injarr{\cdot}$ is simply $\inr \circ \next$.
The semantics of case inspection on dyn performs a case analysis on the sum.

The interpretation of $\lda{x}{M}$ works as follows. Recall by the typing rule for
lambda that $\phasty {\Gamma, x : A_i} M {{A_o}}$, so the interpretation of $M$
has type $(\sem{\Gamma} \times \sem{A_i})$ to $\li \sem{A_o}$.
The interpretation of lambda is thus a function (in the ambient type theory) that takes
a value $a$ representing the argument and applies it (along with $\gamma$) as argument to
the interpretation of $M$.
%
The interpretation of bind and of application both make use the monadic extend function on $\li A$.
%
The interpretation of case-nat and case-arrow is simply a case inspection on the
interpretation of the scrutinee, which has type $\Dyn$.

\vspace{2ex}

\noindent Types:
\begin{align*}
  \sem{\nat} &= \Nat \\
  \sem{\dyn} &= \Dyn \\
  \sem{A \ra A'} &= \sem{A} \To \li \sem{A'} \\
\end{align*}

% Contexts:

% TODO check these, especially the semantics of bind, case-nat, and case-arr
% with respect to their context argument
\noindent Values and terms:
\begin{align*}
  \sem{\zro}         &= \lambda \gamma . 0 \\
  \sem{\suc\, V}     &= \lambda \gamma . (\sem{V}\, \gamma) + 1 \\
  \sem{x \in \Gamma} &= \lambda \gamma . \gamma(x) \\
  \sem{\lda{x}{M}}   &= \lambda \gamma . \lambda a . \sem{M}\, (*,\, (\gamma , a))  \\
  \sem{\injnat{V_n}} &= \lambda \gamma . \inl\, (\sem{V_n}\, \gamma) \\
  \sem{\injarr{V_f}} &= \lambda \gamma . \inr\, (\sem{V_f}\, \gamma) \\[2ex]
  \sem{\nxt\, V}     &= \lambda \gamma . \nxt (\sem{V}\, \gamma) \\
  \sem{\theta}       &= \lambda \gamma . \theta \\
%
  \sem{\err_B}         &= \lambda \delta . \mho \\
  \sem{\ret\, V}       &= \lambda \gamma . \eta\, \sem{V} \\
  \sem{\bind{x}{M}{N}} &= \lambda \delta . \ext {(\lambda x . \sem{N}\, (\delta, x))} {\sem{M}\, \delta} \\
  \sem{V_f\, V_x}      &= \lambda \gamma . \ext {(\lambda f . (\ext {f} {\sem{V_x}\, \gamma}))} {(\sem{V_f}\, \gamma)} \\
  \sem{\casenat{V}{M_{no}}{n}{M_{yes}}}         &= 
    \lambda \delta . \text{case $(\sem{V}\, \delta)$ of} \\ 
    &\quad\quad\quad\quad \alt \inl(n) \to \sem{M_{yes}}(n) \\
    &\quad\quad\quad\quad \alt \inr(\tilde{f}) \to \sem{M_{no}} \\
  \sem{\casearr{V}{M_{no}}{\tilde{f}}{M_{yes}}} &= 
    \lambda \delta . \text{case $(\sem{V}\, \delta)$ of} \\ 
    &\quad\quad\quad\quad \alt \inl(n) \to \sem{M_{no}} \\
    &\quad\quad\quad\quad \alt \inr(\tilde{f}) \to \sem{M_{yes}}(\tilde{f})
\end{align*}

### 2023-11-17 paper-new/denotational-model.tex
begin description of denotational model for intensional gradual typing

\section{A Denotational Model for Intensional Gradual Typing}

In this section we model intensional gradual typing in a suitable double category.

We construct a (thin) double category $\mathsf{IGTT}$ as follows.

% First, we fix a set $D$ with a partial order $\le_D$ and bisimilarity relation $\bisim_D$.
% This is intended to model the dynamic type. Now we define the double category:

\begin{itemize}
  \item \textbf{Objects}: An object consists of the following data:
    \begin{itemize}
        \item A double poset $X$, i.e., a set $X$ equipped with a partial order $\le_X$
        and a reflexive, symmetric ``bisimilarity'' relation $\bisim_X$.
        \item Two commutative monoids of perturbations $P_V$ and $P_C$ with homomorphisms
        \begin{align*}
        \ptb_V &: P_V \to \{ f : X \to_m X \mid f \bisim \id \} \\
        \ptb_C &: P_C \to \{ f : X \to_m \li X \mid f \bisim \eta \}
        \end{align*}
        (where composition in the latter monoid of functions is given by Kleisli composition).

    \end{itemize}

  \item \textbf{Vertical arrows}: An vertical arrow from $(X, P_V^X, P_C^X)$ to $(Y, P_V^Y, P_C^Y)$ is
  a function $f : X \to Y$ that is \emph{monotone} (preserves ordering) and preserves the bisimilarity relation.
  % that preserves ordering and bisimilarity.

  \item \textbf{Horizontal arrows}: A horizontal arrow from $(X, P_V^X, P_C^X)$ to $(Y, P_V^Y, P_C^Y)$
  consists of:
  \begin{itemize}
    \item A relation $R : X \nrightarrow Y$ that is antitone with respect to $\le_X$ and
    monotone with respect to $\le_Y$.
    \item An embedding $e_{XY} : X \to_m Y$ preserving ordering and bisimilarity.
    \item A projection $p_{XY} : Y \to_m \li X$ preserving ordering and bisimilarity.
  \end{itemize}

  such that (1) $R$ is \emph{quasi-representable} by $e_{XY}$ and $p_{XY}$, and 
  (2) $R$ satisfies the \emph{push-pull} property.

  \vspace{3ex}
  The former means that there are distinguished elements $\delta^{l,e} \in P_V^X$, $\delta^{r,e} \in P_V^Y$, 
  $\delta^{l,p} \in P_C^X$ and $\delta^{r,p} \in P_C^Y$ such that the following squares commute:

  \begin{center}
    \begin{tabular}{ c | c } 
        \hline
        \hspace{3em} 
        % UpL
        % https://q.uiver.app/#q=WzAsNCxbMCwwLCJYIl0sWzAsMSwiWSJdLFsxLDAsIlkiXSxbMSwxLCJZIl0sWzAsMiwiUiIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEsMywiXFxsZV9ZIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJlX3tYWX0iLDJdLFsyLDMsIlxccHRiX1ZeWShcXGRlbHRhXntyLGV9KSJdLFs2LDcsIlxcbHRkeW4iLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
        \begin{tikzcd}[ampersand replacement=\&]
            X \& Y \\
            Y \& Y
            \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
            \arrow["{\le_Y}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
            \arrow[""{name=0, anchor=center, inner sep=0}, "{e_{XY}}"', from=1-1, to=2-1]
            \arrow[""{name=1, anchor=center, inner sep=0}, "{\ptb_V^Y(\delta^{r,e})}", from=1-2, to=2-2]
            \arrow["\ltdyn"{description}, draw=none, from=0, to=1]
        \end{tikzcd} & 
        % UpR
        % https://q.uiver.app/#q=WzAsNCxbMCwwLCJYIl0sWzAsMSwiWCJdLFsxLDAsIlgiXSxbMSwxLCJZIl0sWzAsMSwiXFxwdGJfVl5YKFxcZGVsdGFee2wsZX0pIiwyXSxbMiwzLCJlX3tYWX0iXSxbMSwzLCJSIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwyLCJcXGxlX1giLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs0LDUsIlxcbHRkeW4iLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
        \begin{tikzcd}[ampersand replacement=\&]
            X \& X \\
            X \& Y
            \arrow[""{name=0, anchor=center, inner sep=0}, "{\ptb_V^X(\delta^{l,e})}"', from=1-1, to=2-1]
            \arrow[""{name=1, anchor=center, inner sep=0}, "{e_{XY}}", from=1-2, to=2-2]
            \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
            \arrow["{\le_X}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
            \arrow["\ltdyn"{description}, draw=none, from=0, to=1]
        \end{tikzcd} \\
        \hline
        % DnR
        % https://q.uiver.app/#q=WzAsNCxbMCwwLCJYIl0sWzEsMCwiWSJdLFswLDEsIkxYIl0sWzEsMSwiTFgiXSxbMCwyLCJcXHB0Yl9DXlgoXFxkZWx0YV57bCxwfSkiLDJdLFsxLDMsInBfe1hZfSJdLFswLDEsIlIiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsyLDMsIlxcbGVfe0xYfSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzQsNSwiXFxsdGR5biIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
        \begin{tikzcd}[ampersand replacement=\&]
            X \& Y \\
            LX \& LX
            \arrow[""{name=0, anchor=center, inner sep=0}, "{\ptb_C^X(\delta^{l,p})}"', from=1-1, to=2-1]
            \arrow[""{name=1, anchor=center, inner sep=0}, "{p_{XY}}", from=1-2, to=2-2]
            \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
            \arrow["{\le_{LX}}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
            \arrow["\ltdyn"{description}, draw=none, from=0, to=1]
        \end{tikzcd} & 
            \hspace{3em} 
            % DnL
            % https://q.uiver.app/#q=WzAsNCxbMCwwLCJZIl0sWzEsMCwiWSJdLFswLDEsIkxYIl0sWzEsMSwiTFkiXSxbMCwxLCJcXGxlX1kiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDIsInBfe1hZfSIsMl0sWzEsMywiXFxwdGJfQ15ZKFxcZGVsdGFee3IscH0pIl0sWzIsMywiTFIiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs1LDYsIlxcbHRkeW4iLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
            \begin{tikzcd}[ampersand replacement=\&]
                Y \& Y \\
                LX \& LY
                \arrow["{\le_Y}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
                \arrow[""{name=0, anchor=center, inner sep=0}, "{p_{XY}}"', from=1-1, to=2-1]
                \arrow[""{name=1, anchor=center, inner sep=0}, "{\ptb_C^Y(\delta^{r,p})}", from=1-2, to=2-2]
                \arrow["LR"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
                \arrow["\ltdyn"{description}, draw=none, from=0, to=1]
            \end{tikzcd} \\ 
        \hline
    \end{tabular}
    \end{center}
    (Here, $LR$ is the lock step lifting of the relation $R$.)

    \vspace{3ex}

    The push-pull property is defined as follows:
    \begin{itemize}
        \item Given any perturbation $\delta_X \in P_V^X$, we can \emph{push} it forward along $R$ to a
        perturbation $\push(\delta_X) \in P_V^Y$, such that $\ptb_V^X(\delta_X) \le \ptb_V^Y(\push(\delta_X))$.

        \item Conversely, given any perturbation $\delta_Y \in P_V^Y$, we can \emph{pull} it back along $R$
        to a perturbation $\pull(\delta_Y) \in P_V^X$, such that $\ptb_V^X(\pull(\delta_Y)) \le \ptb_V^Y(\delta_Y)$.

        \item Likewise, we can push any perturbation $\delta_X \in P_C^X$ along $LR$
        to get a perturbation $\push(\delta_X) \in P_C^Y$ such that
        $\ptb_C^X(\delta_X) \le \ptb_C^Y(\push(\delta_X))$.

        \item And similarly, we can pull a perturbation in $P_C^Y$ along $LR$ to a perturbation in $P_C^X$
        satisfying the analogous property.
    \end{itemize}

    \textbf{TODO: push and pull might need to be monoid homomorphisms}

  \item \textbf{Two-cells}: Let $f : W \to X$ and $g : Y \to Z$ and let $R : W \nrightarrow Y$ and 
  $S : X \nrightarrow Z$. We define $f \le g$ to mean for all $(w, y) \in R$, we have
  $(f(w), g(y)) \in S$. This is depicted in the square below:

  % https://q.uiver.app/#q=WzAsNCxbMCwwLCJXIl0sWzAsMSwiWCJdLFsxLDAsIlkiXSxbMSwxLCJaIl0sWzAsMiwiUiIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEsMywiUyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMSwiZiIsMl0sWzIsMywiZyJdLFs0LDUsIlxcc3FzdWJzZXRlcSIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
\[\begin{tikzcd}[ampersand replacement=\&]
	W \& Y \\
	X \& Z
	\arrow[""{name=0, anchor=center, inner sep=0}, "R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow[""{name=1, anchor=center, inner sep=0}, "S"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
	\arrow["f"', from=1-1, to=2-1]
	\arrow["g", from=1-2, to=2-2]
	\arrow["\sqsubseteq"{description}, draw=none, from=0, to=1]
\end{tikzcd}\]

\end{itemize}

The category satisfies the following additional properties:
\begin{itemize}
    \item \emph{Existence of Dyn}: There is an object $D$ with the property that for any
    object $X$, there is a horizontal arrow $X \nrightarrow D$.
    The underlying double poset is defined by guarded recursion as the solution to
    \[ D \cong \mathbb{N}\, + \later \hspace{-0.5ex} (D \to_m \li D). \]

    \textbf{TODO: define the perturbations for Dyn and show there is a horizontal arrow $X \nrightarrow D$ for all $X$.}

    \item \emph{Thinness}: There is at most one two-cell for any given square.

    % \item \emph{Push-Pull}: Let $X$ and $Y$ be objects, and let $R : X \nrightarrow Y$.
    % \begin{itemize}
    %     \item Given any perturbation $\delta_X \in P_V^X$, we can \emph{push} it forward along $R$ to a
    %     perturbation $\push(\delta_X) \in P_V^Y$, such that $\ptb_V^X(\delta_X) \le \ptb_V^Y(\push(\delta_X))$.

    %     \item Conversely, given any perturbation $\delta_Y \in P_V^Y$, we can \emph{pull} it back along $R$
    %     to a perturbation $\pull(\delta_Y) \in P_V^X$, such that $\ptb_V^X(\pull(\delta_Y)) \le \ptb_V^Y(\delta_Y)$.

    %     \item Likewise, we can push any perturbation $\delta_X \in P_C^X$ along $L\, R$
    %     (the lock-step lifting of the relation $R$) to get a perturbation $\push(\delta_X) \in P_C^Y$ such that
    %     $\ptb_C^X(\delta_X) \le \ptb_C^Y(\push(\delta_X))$.

    %     \item And similarly we can pull a perturbation in $P_C^Y$ to a perturbation in $P_C^X$
    %     satisfying the analogous property.
    % \end{itemize}
\end{itemize}

% Composability of embedding and projections

We need to verify that this forms a thin double category. 
\begin{itemize}
    \item \emph{Horizontal identity morphism}: 
    Let $X$ be an object. We take $R$ to be $\le_X$ (the ordering relation on $X$),
    which is trivially antitone and monotone with respect to itself.
    We let $e_{XX} = \id$ and $p_{XX} = \eta$. These clearly preserve the
    ordering and bisimilarity.

    \vspace{3ex}

    We first need to show that $R$ is quasi-representable.
    We prove the UpR rule; the others are similar.
    We need to specify a distinguished element $\delta^{l,e} \in P_V^X$ such that
    the following square commutes:

    % https://q.uiver.app/#q=WzAsNCxbMCwwLCJYIl0sWzAsMSwiWCJdLFsxLDAsIlgiXSxbMSwxLCJYIl0sWzAsMSwiXFxwdGJfVl5YKFxcZGVsdGFee2wsZX0pIiwyXSxbMiwzLCJlX3tYWH0gPSBcXGlkIl0sWzAsMiwiXFxsZV9YIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJcXGxlX1giLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
    \[\begin{tikzcd}[ampersand replacement=\&]
        X \& X \\
        X \& X
        \arrow["{\ptb_V^X(\delta^{l,e})}"', from=1-1, to=2-1]
        \arrow["{e_{XX} = \id}", from=1-2, to=2-2]
        \arrow["{\le_X}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
        \arrow["{\le_X}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
    \end{tikzcd}\]

    Taking $\delta^{l,e} = \id$ (the identity of the monoid), we observe that since
    $\ptb_V^X$ is a homomorphism of monoids, we have $\ptb_V^X(\id) = \id$.
    Now it is clear that the above square commutes.

    \vspace{3ex}

    We also need to show that $R$ satisfies the four push-pull properties.
    We show one; the others are similar. Let $\delta_X \in P_V^X$.
    We need to define $\push(\delta_X) \in P_V^X$ such that the following square commutes:

    % https://q.uiver.app/#q=WzAsNCxbMCwwLCJYIl0sWzAsMSwiWCJdLFsxLDAsIlgiXSxbMSwxLCJYIl0sWzAsMiwiXFxsZV9YIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJcXGxlX1giLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxccHRiX1ZeWChcXGRlbHRhX1gpIiwyXSxbMiwzLCJcXHB0Yl9WXlgoXFxwdXNoKFxcZGVsdGFfWCkpIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiZGFzaGVkIn19fV1d
    \[\begin{tikzcd}[ampersand replacement=\&]
	X \& X \\
	X \& X
	\arrow["{\le_X}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow["{\le_X}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
	\arrow["{\ptb_V^X(\delta_X)}"', from=1-1, to=2-1]
	\arrow["{\ptb_V^X(\push(\delta_X))}", dashed, from=1-2, to=2-2]
    \end{tikzcd}\]

    Let $\push(\delta_X) = \delta_X$. We need to show that for $x \le_X x'$,
    we have $\ptb_V^X(\delta_X)(x) \le \ptb_V^X(\delta_X)(x')$, which holds because
    $\ptb_V^X(\delta_X)$ is monotone with respect to $\le_X$.

    \item \emph{Horizontal composition}:

    Let $R : X \nrightarrow Y$ and $S : Y \nrightarrow Z$.
    We define
    $e_{XZ} = e_{YZ} \circ e_{XY}$ and $p_{XZ} = \ext{p_{XY}}{} \circ p_{YZ}$.
    We define the distinguished perturbations in the representability rules as follows:

    \begin{align*}
        \delta_{R\circ S}^{l,e} &= \pull_R(\delta_S^{l,e}) \cdot \delta_R^{l,e} \\
        \delta_{R\circ S}^{r,e} &= \delta_S^{r,e} \cdot \push_S(\delta_R^{r,e}) \\
        \delta_{R\circ S}^{l,p} &= \delta_R^{l,p} \cdot \pull_{LR}(\delta_S^{l,p}) \\
        \delta_{R\circ S}^{r,p} &= \push_{LS}(\delta_R^{r,p}) \cdot \delta_S^{r,p}
    \end{align*}
    where $\cdot$ denotes composition in the appropriate monoid of perturbations.

    We can then show that the four quasi-representability rules are valid with these definitions.
    \textbf{TODO show one or two of the cases}

    \vspace{3ex}

    We also need to show that the push-pull rules hold of the composition $R \circ S$.
    This follows from the fact that they hold for both $R$ and $S$.
    Specifically, we define

    \begin{align*}
        \push_{R \circ S}(\delta^X) &= \push_S(\push_R(\delta^X)) \\
        \pull_{R \circ S}(\delta^Z) &= \pull_R(\pull_S(\delta^Z)) \\
        \push_{L(R \circ S)}(\delta^X) &= \push_{LS}(\push_{LR}(\delta^X)) \\
        \pull_{L(R \circ S)}(\delta^Z) &= \pull_{LR}(\pull_{LS}(\delta^Z))
    \end{align*}

    Then we can verify that the relevant push-pull inequalities hold using the above definitions.
    \textbf{TODO maybe show one of the cases}

    \item \emph{Identity two-cells}:
    The horizontal identity two-cells have the form

    % https://q.uiver.app/#q=WzAsNCxbMCwwLCJYIl0sWzAsMSwiWSJdLFsxLDAsIlgiXSxbMSwxLCJZIl0sWzAsMSwiZiIsMl0sWzIsMywiZiJdLFswLDIsIlxcbGVfWCJdLFsxLDMsIlxcbGVfWSIsMl0sWzQsNSwiXFxsdGR5biIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
    \[\begin{tikzcd}[ampersand replacement=\&]
        X \& X \\
        Y \& Y
        \arrow[""{name=0, anchor=center, inner sep=0}, "f"', from=1-1, to=2-1]
        \arrow[""{name=1, anchor=center, inner sep=0}, "f", from=1-2, to=2-2]
        \arrow["{\le_X}", from=1-1, to=1-2]
        \arrow["{\le_Y}"', from=2-1, to=2-2]
        \arrow["\ltdyn"{description}, draw=none, from=0, to=1]
    \end{tikzcd}\]

    This square commutes because $f$ is monotone with respect to the ordering relation.

    The vertical two-cells have the form

    % https://q.uiver.app/#q=WzAsNCxbMCwwLCJYIl0sWzAsMSwiWCJdLFsxLDAsIlkiXSxbMSwxLCJZIl0sWzAsMiwiUiIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEsMywiUiIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMSwiXFxpZF9YIiwyXSxbMiwzLCJcXGlkX1kiXSxbNiw3LCJcXGx0ZHluIiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
    \[\begin{tikzcd}[ampersand replacement=\&]
        X \& Y \\
        X \& Y
        \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
        \arrow["R", "\shortmid"{marking}, no head, from=2-1, to=2-2]
        \arrow[""{name=0, anchor=center, inner sep=0}, "{\id_X}"', from=1-1, to=2-1]
        \arrow[""{name=1, anchor=center, inner sep=0}, "{\id_Y}", from=1-2, to=2-2]
        \arrow["\ltdyn"{description}, draw=none, from=0, to=1],
    \end{tikzcd}\]

    which commutes trivially.

    \item \emph{Composition of two-cells}:
    Two-cells compose vertically and horizontally, which follows from the definition
    definition of a two-cell in this category.

    \textbf{TODO: elaborate}
\end{itemize}

\vspace{3ex}

\textbf{Kleisli internal hom: TODO}

% The arrow => takes a value type and an algebra and constructs an algebra

### 2023-11-28 paper-new/intro.tex
changes to layout and prose of paper

@@
 \subsection{Gradual Typing and Graduality}
-One of the principal ways of categorizing type systems of programming languages is
-whether they are \emph{static} or \emph{dynamic}.
-In static typing, the code is type-checked at compile time, while in dynamic typing,
-the type checking is deferred to run-time. Both approaches have benefits: with static typing,
-the programmer is assured that if the program passes the type-checker, their program
-is free of type errors, and moreover, the equational theory implies that program optimizations are valid.
+In modern programming language design, there is a tension between \emph{static} typing
+and \emph{dynamic} typing disciplines.
+With static typing, the code is type-checked at compile time, while in dynamic typing,
+the type checking is deferred to run-time. Both approaches have benefits: with static 
+typing, the programmer is assured that if the program passes the type-checker, their
+program is free of type errors, and moreover, soundness of the equational theory implies
+that program optimizations are valid.
 Meanwhile, dynamic typing allows the programmer to rapidly prototype
-their application code without needing to commit to fixed type signatures for their functions.
+their application code without needing to commit to fixed type signatures for their
+functions. Unfortunately, most languages choose between static or dynamic typing.
+As a result, programmers that initially write their code in a dynamically typed language
+in order to benefit from faster prototyping and development time will need to rewrite the
+entire codebase in a static language if they would like to receive the benefits of static
+typing once their codebase has matured.
 
-\emph{Gradually-typed languages} \cite{siek-taha06, tobin-hochstadt06} allow
-for both static and dynamic typing disciplines to be used in the same codebase,
-and support smooth interoperability between statically-typed and dynamically-typed code.
+\emph{Gradually-typed languages} \cite{siek-taha06, tobin-hochstadt06} seek to resolve this tension
+by allowing for both static and dynamic typing disciplines to be used in the same codebase,
+and by supporting smooth interoperability between statically-typed and dynamically-typed code.
 This flexibility allows programmers to begin their projects in a dynamic style and
@@
 %Gradually-typed languages should satisfy two intuitive properties.
-The following two properties have been identified as useful for gradually typed languages.
-First, the interaction between the static and dynamic components of the codebase
+% The following two properties have been identified as useful for gradually typed languages.
+
+In order for this to work as expected, gradually-typed languages should allow for
+different parts of the codebase to be in different places along the spectrum from
+dynamic to static, and allow for those different parts to interact with one another.
+Moreover, gradually-typed languages should support the smooth migration from
+dynamic typing to static typing, in that the programmer can initially leave off the
+typing annotations and provide them later without altering the meaning of the program.
+
+% Sound gradual typing
+Furthermore, the parts of the program that are written in a dynamic style should soundly
+interoperate with the parts that are written in a static style.
+That is, the interaction between the static and dynamic components of the codebase
 should preserve the guarantees made by the static types.
@@
 violated the typing contract imposed by statically typed code.
-% In other words, in the statically-typed portions of the codebase, type soundness must be preserved.
-Second, gradually-typed languages should support the smooth migration from dynamic typing
-to static typing, in that the programmer can initially leave off the
-typing annotations and provide them later without altering the meaning of the
-program.
+
+% Moreover, gradually-typed languages should allow for
+% different parts of the codebase to be in different places along the spectrum from
+% dynamic to static, and allow for those different parts to interact with one another.
+% In a \emph{sound} gradually-typed language,
+% this interaction should respect the guarantees made by the static types.
 
@@
 and Boyland \cite{siek_et_al:LIPIcs:2015:5031}.
-This property is also referred to as \emph{graduality} \cite{new-ahmed2018}, by analogy with parametricity.
+This property is also referred to as \emph{graduality} \cite{new-ahmed2018},
+by analogy with parametricity.
 Intuitively, graduality says that going from a dynamic to static style should not
@@
 More specifically, making the types more precise by adding typing annotations
-% (replacing $\dyn$ with a more specific type
-%such as integers)
 will either result in the same behavior as the original, less precise program,
@@
 \subsection{Current Approaches to Proving Graduality}
-Current approaches to proving graduality include the methods of Abstracting Gradual Typing
+
+\subsubsection{From Static to Gradual}
+
+Current approaches to constructing languages that satisfy the graduality property
+include the methods of Abstracting Gradual Typing
 \cite{garcia-clark-tanter2016} and the formal tools of the Gradualizer \cite{cimini-siek2016}.
@@
 %
-Furthermore, while these frameworks do prove
-graduality of the resulting languages, they do not show the correctness of the equational theory,
-which is equally important to sound gradual typing. For example, programmers often refactor their
-code, and in so doing they rely implicitly on the validity of the laws in the equational theory.
-Similarly, correctness of compiler optimizations rests on the validity of the corresponding equations
-from the equational theory. It is therefore important that the languages that claim to be gradually typed
-have provably correct equational theories.
+Furthermore, while these frameworks do prove graduality of the resulting languages,
+they do not show the correctness of the equational theory, which is equally important
+to sound gradual typing.
+For example, programmers often refactor their code, and in so doing they rely
+implicitly on the validity of the laws in the equational theory.
+Similarly, correctness of compiler optimizations rests on the validity of the
+corresponding equations from the equational theory. It is therefore important
+that the languages that claim to be gradually typed have provably correct
+equational theories.
 
@@
 
-% Proving graduality via LR
-New and Ahmed \cite{new-ahmed2018}
-have developed a semantic approach to specifying type dynamism in terms of
-\emph{embedding-projection pairs}, which allows for a particularly elegant formulation of the
-gradual guarantee.
-Moreover, their axiomatic account of program equivalence allows for type-based reasoning
-about program equivalences.
-%
-In this approach, a logical relation is constructed and used to show that the equational theory
-is sound with respect to the operational semantics.
-%and shown to be sound with respect to
-%the notion of observational approximation that specifies when one program is more precise than another.
+% [Eric] I moved the next two paragraphs from the technical background section
+% to here in the intro.
+\subsubsection{Classical Domain Models}
+
+New and Licata \cite{new-licata18} developed an axiomatic account of the
+graduality relation on cast calculus terms and gave a denotational
+model of this calculus using classical domain theory based on
+$\omega$-CPOs. This semantics has scaled up to an analysis of a
+dependently typed gradual calculus in \cite{asdf}. This meets our
+criterion of being a reusable mathematical theory, as general semantic
+theorems about gradual domains can be developed independent of any
+particular syntax and then reused in many different denotational
+models. However, it is widely believed that such classical domain
+theoretic techniques cannot be extended to model higher-order store, a
+standard feature of realistic gradually typed languages such as Typed
+Racket. Thus if we want a reusable mathematical theory of gradual
+typing that can scale to realistic programming languages, we need to
+look elsewhere to so-called ``step-indexed'' techniques.
+
+\subsubsection{Operational Reduction Proofs}
+
+A series of works \cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
+developed step-indexed logical relations models of gradually typed
+languages based on operational semantics. Unlike classical domain
+theory, such step-indexed techniques are capable of modeling
+higher-order store and runtime-extensible dynamic types
+\cite{amalsthesis,nonpmetricparamorsomething,new-jamner-ahmed19}. However,
+their proof developments are highly repetitive and technical, with
+each development formulating a logical relation from first-principles
+and proving many of the same tedious lemmas without reusable
+mathematical abstractions. Our goal in the current work is to extract
+these reusable mathematical principles from these tedious models to
+make formalization of realistic gradual languages tractible.
+
+% Alternative phrasing:
+\begin{comment}
+\subsubsection{Embedding-Projection Pairs}
+
+The series of works by New, Licata, and Ahmed \cite{new-licata18,new-ahmed2018,new-licata-ahmed2019}
+develop an axiomatic account of gradual typing involving \emph{embedding-projection pairs}.
+This allows for a particularly elegant formulation of the gradual guarantee.
+Moreover, their axiomatic account of program equivalence allows for type-based reasoning about program equivalences.
+
+In \cite{new-licata18}, New and Licata construct a denotational model of the axioms of
+gradual typing using tools from classical domain theory. The benefit to this approach is that it is a reusable mathematical theory:
+general semantic theorems about gradual domains can be developed independent of any particular syntax and then reused in many different denotational models.
+Unfortunately, however, it is unclear how to extend these domain-theoretic techniques to incorporate more advanced language features such as higher-order store, a standard feature of realistic gradually typed languages such as Typed
+Racket.
+Thus, if we want a reusable mathematical theory of gradual typing that can scale to realistic programming languages, we need to look to different techniques.
+
+In \cite{new-ahmed2018,new-licata-ahmed2019}, a more traditional operational semantics for gradual typing is derived from the axioms.
+% The axioms are then proven to be sound with respect to this operational semantics by constructing a logical relation.
+A logical relations model is constructed and used to prove that the axioms are sound with respect to the operational semantics.
 The downside of this approach is that each new language requires a different logical relation
-to prove graduality. Furthermore, the logical relations tend to be quite complicated due
-to a technical requirement known as \emph{step-indexing}, where the stepping behavior of terms
-must be accounted for in the logical relation.
+to prove graduality, even though the logical relations for different languages end up sharing many commonalities.
+Furthermore, the logical relations tend to be quite complicated due to a technical requirement known as \emph{step-indexing},
+where the stepping behavior of terms must be accounted for in the logical relation.
 As a result, developments using this approach tend to require vast effort, with the
 corresponding technical reports having 50+ pages of proofs.
-%
-Additionally, while the axioms of gradual type theory are compositional at a ``global'' level,
-they do not compose in the step-indexed setting. One of the main goals of the present work
-is to formulate a composable theory of gradual typing in a setting where the stepping behavior
-is tracked.
+
+% In this approach, a logical relation is constructed and used to show that the equational theory
+% is sound with respect to the operational semantics.
+
+% Additionally, while the axioms of gradual type theory are compositional at a ``global'' level,
+% they do not compose in the step-indexed setting. One of the main goals of the present work
+% is to formulate a composable theory of gradual typing in a setting where the stepping behavior
+% is tracked.
+\end{comment}
 
@@
 The techniques of synthetic guarded domain theory allow us to internalize the
-step-index reasoning normally required in logical relations proofs of graduality,
+step-indexed reasoning normally required in logical relations proofs of graduality,
 ultimately allowing us to specify the logical relation in a manner that looks nearly
@@
 
-Our longer-term goal is to mechanize these proofs in a reusable way,
-thereby providing a framework to use to more easily and
-conveniently prove that existing languages satisfy graduality and have
-sound equational theories. Moreover, the aim is for designers of new languages
-to utilize the framework to facilitate the design of new provably-correct
-gradually-typed languages with more complex features.
+Our longer-term goal is to mechanize these proofs in a reusable way, thereby providing a framework
+to use to more easily and conveniently prove that existing languages satisfy graduality and have
+sound equational theories. Moreover, the aim is for designers of new languages to utilize the
+framework to facilitate the design of new provably-correct gradually-typed languages with more
+complex features.
 
@@
 
-\subsection{Overview of Remainder of Paper}
+\subsection{Outline of Remainder of Paper}
 
-In Section \ref{sec:overview}, we give an overview of the gradually-typed lambda
-calculus and the graduality theorem.
+% In Section \ref{sec:overview}, we give an overview of the gradually-typed lambda
+% calculus and the graduality theorem.
 %
-In Section \ref{sec:technical-background}, we provide technical background on gradually typed languages and
-on synthetic guarded domain theory.
+In Section \ref{sec:technical-background}, we provide technical background on synthetic guarded domain theory.
 % 
-In Section \ref{sec:gtlc-terms}, we introduce the gradually-typed cast calculus
-for which we will prove graduality. We give a semantics to the terms
-using the tools of guarded type theory.
+In Section \ref{sec:gtlc-terms}, we formally introduce the gradually-typed cast calculus
+for which we will prove graduality. We give a semantics to the terms using
+synthetic guarded domain theory.
 % Important here are the notions of syntactic
@@
 
-In Section \ref{sec:gtlc-precision}, we define the term precision ordering
-and describe our approach to assigning a denotational semantics to this ordering.
+In Section \ref{sec:gtlc-precision}, we define the term precision ordering for
+the gradually-typed lambda calculus and describe our approach to assigning a
+denotational semantics to this ordering.
 This approach builds on the semantics constructed in the previous section,
@@
 In Section \ref{sec:graduality}, we outline in more detail the proof of graduality for the
-extensional gradual lambda calculus.
+gradual lambda calculus.
 %
@@
 
-\section{Overview}\label{sec:overview}
+\subsection{Overview of Results}\label{sec:overview}
 
@@
 simple gradually-typed cast calculus, the gradually-typed lambda calculus.
-This is the usual simply-typed lambda calculus with a dynamic type $\dyn$ such that
-$A \ltdyn\, \dyn$ for all types $A$, as well as upcasts and downcasts between any types
-$A$ and $B$ such that $A \ltdyn B$. The complete definition will be provided in
-Section \ref{sec:GTLC}.
-The graduality theorem is shown below.
+Before stating the graduality theorem, we introduce some definitions related to gradual typing.
+
+% Cast calculi
+% TODO move this to earlier?
+Gradually typed languages are generally elaborated to a \emph{cast calculus}, in which run-time type checking
+that is made explicit. Elaboration inserts \emph{type casts} at the boundaries between static and dynamic code.
+In particular, cast insertion happens at the elimination forms of the language (e.g., pattern matching, field reference, etc.).
+The original gradually typed language that is elaborated to a cast calculus is called the \emph{surface language}.
+
+% Up and down casts
+In a cast calculus, there is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that
+$A$ is a \emph{more precise} type than $B$.
+There a distinguished type $\dyn$, the \emph{dynamic type}, with the property that $A \ltdyn\, \dyn$ for all $A$.
+The parts of the code that are dynamically typed will be given type $\dyn$ in the cast calculus.
+%
+If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
+and a term $N$ of type $B$ may be \emph{down}casted to $A$, written $\dn A B N$.
+Upcasts always succeed, while downcasts may fail at runtime, resulting in a type error.
+Cast calculi have a term $\mho$ representing a run-time type error for any type $A$.
+
+% Syntactic term precision
+We also have a notion of \emph{syntactic term precision}.
+If $A \ltdyn B$, and $M$ and $N$ are terms of type $A$ and $B$ respectively, we write
+$M \ltdyn N : A \ltdyn B$ to mean that
+$M$ is ``syntactically more precise'' than $N$, or equivalently, $N$ is 
+``more dynamic'' than $M$. Intuitively, this models the situation where $M$ and $N$
+behave the same except that $M$ may error more.
+Term precision is defined by a set of axioms that capture this intuitive notion.
+The specific rules for term precision depend on the specifics of the language, but
+many of the rules can be derived from the typing rules in a straightforward way.
+% Cast calculi also have a term $\mho$ representing a run-time type error for any type $A$,
+% and since
+Since $\mho$ always errors, there is a term precision rule stating that $\mho \ltdyn M$ for all $M$.
+
+% GTLC
+The gradually-typed lambda calculus is the usual simply-typed lambda calculus with a dynamic
+type $\dyn$ such that $A \ltdyn\, \dyn$ for all types $A$, as well as upcasts and downcasts
+between any types $A$ and $B$ such that $A \ltdyn B$. The complete definition will be provided in
+Section \ref{sec:gtlc-precision}.
+
+With these definitions, we can state the graduality theorem for the gradually-typed lambda calculus.
 
@@
 \begin{theorem}[Graduality]
-  If $\cdot \vdash M \ltdyn N : \nat$, then
-  \begin{enumerate}
-    \item $M \Downarrow$ iff $N \Downarrow$.
-    \item If $M \Downarrow v_?$ and $N \Downarrow v'_?$ then either $v_? = \mho$, or $v_? = v'_?$.
-  \end{enumerate}
+  Let $\cdot \vdash M \ltdyn N : \nat$. 
+  If $M \Downarrow v_?$ and $N \Downarrow v'_?$, then either $v_? = \mho$, or $v_? = v'_?$.
 \end{theorem}
 
-Details can be found in later sections, but we provide a brief explanation of the terminology and notation:
+Note:
+\begin{itemize}
 
-\begin{itemize}
-  \item $M \ltdyn N : \nat$ means $M$ and $N$ are terms of type $\nat$ such that
-  $M$ is ``syntactically more precise'' than $N$, or equivalently, $N$ is 
-  ``more dynamic'' than $M$. Intuitively this means that $M$ and $N$ are the
-  same except that in some places where $M$ has explicit typing annotations,
-  $N$ has $\dyn$ instead. The relation $\ltdyn$ on terms is called ``term precision''.
-  Term precision is intended to model the situation where a term $M$ behaves the
-  same as another term $N$ except that $M$ may error more than $N$.
-  Term precision is defined by a set of axioms that capture this intuitive notion.
-  This will be described in more detail in Section \ref{sec:GTLC}.
-
-  \item $\cdot \Downarrow$ is a relation on terms that is defined such that $M \Downarrow$ means
+  \item $\cdot \Downarrow$ is a relation between terms that is defined such that $M \Downarrow$ means
   that $M$ terminates, either with a run-time error or a value $n$ of type $\nat$.
-
-  \item $\mho$ is a syntactic representation of a run-time type error, which
-  happens, for example, when a programmer tries to call a function with a value whose type
-  is found to be incompatible with the argument type of the function.
 

### 2023-11-28 paper-new/ordering.tex
changes to layout and prose of paper

@@
 
-\subsection{Term Precision for GTLC}
+\subsection{Term Precision for GTLC}\label{sec:gtlc-term-precision-axioms}
 
@@
   \inferrule*{M \ltdyn_i N : B}
-             {\dn c M \ltdyn_i N : c}
+             {\dnc{c}{M} \ltdyn_i N : c}
 \end{mathpar}
 
-If $c$ is inj-arr, then when we downcast $M$ from $dyn$ to $\dyn \ra \dyn$,
+If $c$ is inj-arr, then when we downcast $M$ from $dyn$ to $\dyntodyn$,
 semantically this will involve a $\theta$ because the value of type $dyn$
@@
 
-\[ \up{c_i \ra c_o}{M} \equiv \lambda (x : B_i). \up{c_o}(M\, (\dn {c_i} x)). \]
+\[ \up{A_i \ra A_o}{B_i \ra B_o}{M} \equiv \lambda (x : B_i). \up{A_o}{B_o}(M\, (\dn {A_i}{B_i} x)). \]
 
@@
 \inferrule
-  {\delta_\nat : \perte \nat \and \delta_f : \perte (\dyn \ra \dyn)}
+  {\delta_\nat : \perte \nat \and \delta_f : \perte (\dyntodyn)}
   {\pertdyn{\delta_\nat}{\delta_f} : \perte \dyn}
@@
 \inferrule
-  {\delta_\nat : \pertp \nat \and \delta_f : \pertp (\dyn \ra \dyn)}
+  {\delta_\nat : \pertp \nat \and \delta_f : \pertp (\dyntodyn)}
   {\pertdyn{\delta_\nat}{\delta_f} : \pertp \dyn}

### 2023-11-28 paper-new/technical-background.tex
changes to layout and prose of paper

@@
 \section{Technical Background}\label{sec:technical-background}
-
-\subsection{Gradual Typing}
-
-% Cast calculi
-In a gradually-typed language, the mixing of static and dynamic code is seamless, in that
-the dynamically typed parts are checked at runtime. This type checking occurs at the elimination
-forms of the language (e.g., pattern matching, field reference, etc.).
-Gradual languages are generally elaborated to a \emph{cast calculus}, in which the dynamic
-type checking is made explicit through the insertion of \emph{type casts}.
-
-% Up and down casts
-In a cast calculus, there is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that
-$A$ is a \emph{more precise} type than $B$.
-There a dynamic type $\dyn$ with the property that $A \ltdyn\, \dyn$ for all $A$.
-%
-If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
-and a term $N$ of type $B$ may be \emph{down}casted to $A$, written $\dn A B N$.
-Upcasts always succeed, while downcasts may fail at runtime.
-%
-% Syntactic term precision
-We also have a notion of \emph{syntactic term precision}.
-If $A \ltdyn B$, and $M$ and $N$ are terms of type $A$ and $B$ respectively, we write
-$M \ltdyn N : A \ltdyn B$ to mean that $M$ is more precise than $N$, i.e., $M$ and $N$
-behave the same except that $M$ may error more.
 
@@
 % synthetic guarded domain theory, denotational semantics therein
-
-\subsection{Operational Reduction Proofs}
-
-\subsection{Classical Domain Models}
-
-New and Licata \cite{new-licata18} developed an axiomatic account of the
-graduality relation on cast calculus terms and gave a denotational
-model of this calculus using classical domain theory based on
-$\omega$-CPOs. This semantics has scaled up to an analysis of a
-dependently typed gradual calculus in \cite{asdf}. This meets our
-criterion of being a reusable mathematical theory, as general semantic
-theorems about gradual domains can be developed independent of any
-particular syntax and then reused in many different denotational
-models. However, it is widely believed that such classical domain
-theoretic techniques cannot be extended to model higher-order store, a
-standard feature of realistic gradually typed languages such as Typed
-Racket. Thus if we want a reusable mathematical theory of gradual
-typing that can scale to realistic programming languages, we need to
-look elsewhere to so-called ``step-indexed'' techniques.
-
-A series of works \cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
-developed step-indexed logical relations models of gradually typed
-languages based on operational semantics. Unlike, classical domain
-theory, such step-indexed techniques are capable of modeling
-higher-order store and runtime-extensible dynamic types
-\cite{amalsthesis,nonpmetricparamorsomething,new-jamner-ahmed19}. However
-their proof developments are highly repetitive and technical, with
-each development formulating a logical relation from first-principles
-and proving many of the same tedious lemmas without reusable
-mathematical abstractions. Our goal in the current work is to extract
-these reusable mathematical principles from these tedious models to
-make formalization of realistic gradual languages tractible.
 
@@
   can be tedious and error-prone, and there are some very subtle aspects that must
-  be taken into account in the proofs. Figure \ref{TODO} shows an example of a step-indexed logical
-  relation for the gradually-typed lambda calculus.
-
+  be taken into account in the proofs.
+  %
   In particular, the prior approach of New and Ahmed requires two separate logical

### 2023-11-28 paper-new/terms.tex
merge two cases(casenat/arr) into one casedyn
changes to layout and prose of paper

@@
 We have reflexivity rules for $\dyn$ and $\nat$, as well as rules that $\nat$ is more precise than $\dyn$
-and $\dyn \ra \dyn$ is more precise than $\dyn$.
+and $\dyntodyn$ is more precise than $\dyn$.
 We also have a congruence rule for function types stating that given $A_i \ltdyn A'_i$ and $A_o \ltdyn A'_o$, we can prove
@@
     { }
-    {(\dyn \ra \dyn) \ltdyn\, \dyn}
+    {(\dyntodyn) \ltdyn\, \dyn}
 
   \inferrule*[right = $\injarr{}$]
-    {(R \ra S) \ltdyn\, (\dyn \ra \dyn)}
+    {(R \ra S) \ltdyn\, (\dyntodyn)}
     {(R \ra S) \ltdyn\, \dyn}
@@
 % We now observe that all casts, except those between $\nat$ and $\dyn$
-% and between $\dyn \ra \dyn$ and $\dyn$, are admissible, in the sense that
+% and between $\dyntodyn$ and $\dyn$, are admissible, in the sense that
 % we can start from $\extlcm$, remove casts except the aforementioned ones,
@@
 We now observe that all casts, except those between $\nat$ and $\dyn$
-and between $\dyn \ra \dyn$ and $\dyn$, are admissible.
+and between $\dyntodyn$ and $\dyn$, are admissible.
 That is, consider a new language ($\extlcprime$) in which
 instead of having arbitrary casts, we have injections from $\nat$ and
-$\dyn \ra \dyn$ into $\dyn$, and a case inspection on $\dyn$.
+$\dyntodyn$ into $\dyn$, and a case inspection on $\dyn$.
 We claim that in $\extlcprime$, all of the casts present in $\extlc$ are derivable.
@@
 of their domain and codomain. The base case is one of the casts involving $\nat$
-or $\dyn \ra \dyn$ which are present in $\extlcprime$ as injections and case inspections.
+or $\dyntodyn$ which are present in $\extlcprime$ as injections and case inspections.
 
@@
       \inferrule*
-      {\hasty \Gamma M (\dyn \ra \dyn)}
+      {\hasty \Gamma M (\dyntodyn)}
       {\hasty \Gamma {\injarr M} \dyn}
 
-      % Case nat
+      % Case dyn
       \inferrule*
       {\hasty{\Delta|_V}{V}{\dyn} \and
-        \hasty{\Delta , x : \nat }{M_{yes}}{B} \and 
-        \hasty{\Delta}{M_{no}}{B}}
-      {\hasty {\Delta} {\casenat{V}{M_{no}}{n}{M_{yes}}} {B}}
-
-      % Case arr
-      \inferrule*
-      {\hasty{\Delta|_V}{V}{\dyn} \and 
-        \hasty{\Delta , x : (\dyn \ra \dyn) }{M_{yes}}{B} \and 
-        \hasty{\Delta}{M_{no}}{B}}
-      {\hasty {\Delta} {\casearr{V}{M_{no}}{f}{M_{yes}}} {B}}
+        \hasty{\Delta , x : \nat }{M_{nat}}{B} \and 
+        \hasty{\Delta , x : (\dyntodyn) }{M_{fun}}{B}
+      }
+      {\hasty {\Delta} {\casedyn{V}{n}{M_{nat}}{f}{M_{fun}}} {B}}
   \end{mathpar}
@@
   \begin{mathpar}
-     % Case-nat Beta
+     % Case-dyn Beta
      \inferrule*
      {\hasty \Gamma V \nat}
-     {\casenat {\injnat {V}} {M_{no}} {n} {M_{yes}} = M_{yes}[V/n]}
+     {\casedyn {\injnat {V}} {n} {M_{nat}} {f} {M_{fun}} = M_{nat}[V/n]}
 
      \inferrule*
-     {\hasty \Gamma V {\dyn \ra \dyn} }
-     {\casenat {\injarr {V}} {M_{no}} {n} {M_{yes}} = M_{no}}
+     {\hasty \Gamma V {\dyntodyn} }
+     {\casedyn {\injarr {V}} {n} {M_{nat}} {f} {M_{fun}} = M_{fun}[V/f]}
 
-     % Case-nat Eta
+     % Case-dyn Eta
      \inferrule*
      {}
-     {\Gamma , x :\, \dyn \vdash M = \casenat{x}{M}{n}{M[(\injnat{n}) / x]} }
-
-     % Case-arr Beta
-
-     % Case-arr Eta
+     {\Gamma , x :\, \dyn \vdash M = \casedyn{x}{n}{M[(\injnat{n}) / x]}{f}{M[(\injarr{f}) / x]} }
 
@@
 Thus, the semantics of $\injnat{\cdot}$ is simply $\inl$ and the semantics
-of $\injarr{\cdot}$ is simply $\inr \circ \next$.
+of $\injarr{\cdot}$ is simply $\inr \circ \nxt$.
 The semantics of case inspection on dyn performs a case analysis on the sum.
@@
   \sem{\bind{x}{M}{N}} &= \lambda \delta . \ext {(\lambda x . \sem{N}\, (\delta, x))} {\sem{M}\, \delta} \\
-  \sem{V_f\, V_x}      &= \lambda \gamma . \ext {(\lambda f . (\ext {f} {\sem{V_x}\, \gamma}))} {(\sem{V_f}\, \gamma)} \\
-  \sem{\casenat{V}{M_{no}}{n}{M_{yes}}}         &= 
+  \sem{V_f\, V_x}      &= \lambda \gamma . {({(\sem{V_f}\, \gamma)} \, {(\sem{V_x}\, \gamma)})} \\
+  \sem{\casedyn{V}{n}{M_{nat}}{\tilde{f}}{M_{fun}}} &=
     \lambda \delta . \text{case $(\sem{V}\, \delta)$ of} \\ 
-    &\quad\quad\quad\quad \alt \inl(n) \to \sem{M_{yes}}(n) \\
-    &\quad\quad\quad\quad \alt \inr(\tilde{f}) \to \sem{M_{no}} \\
-  \sem{\casearr{V}{M_{no}}{\tilde{f}}{M_{yes}}} &= 
-    \lambda \delta . \text{case $(\sem{V}\, \delta)$ of} \\ 
-    &\quad\quad\quad\quad \alt \inl(n) \to \sem{M_{no}} \\
-    &\quad\quad\quad\quad \alt \inr(\tilde{f}) \to \sem{M_{yes}}(\tilde{f})
+    &\quad\quad\quad\quad \alt \inl(n) \to \sem{M_{nat}}(n) \\
+    &\quad\quad\quad\quad \alt \inr(\tilde{f}) \to \sem{M_{fun}}(\tilde{f})
 \end{align*}

### 2023-11-29 paper-new/categorical-model.tex
work on abstract categorical model

\section{Abstract Categorical Models of Graduality}

First, what is a categorical model of call-by-push-value? We will use
the following notion as our basic structure\footnote{There are models
that are closer to syntax e.g., which distinguish between value types
and contexts but this suffices for all of the models we consider}:
\begin{enumerate}
\item A cartesian category $\mathcal V$
\item A category $\mathcal E$
\item An action of $\mathcal V^{op}$ (with the $\mathcal V$ cartesian
  product as monoidal structure) on $\mathcal E$. We write this as
  exponentiation $B^A$.

  This means we have 
  \[ \alpha : B^{A_1 \times A_2} \cong (B^{A_1})^{A^2} \]
  and
  \[ i : B^1 \cong B \]
  satisfying coherence isomorphisms
\item A functor $U : \mathcal E \to \mathcal V$ that ``preserves
  powering'' in that every $U(B^A)$ is an exponential of $UB$ by $A$
  and that $U\alpha$ and $Ui$ are mapped to the canonical isomorphisms
  for exponentials.
\item A left adjoint $F \dashv U$
\item Distributive finite coproducts in $\mathcal V$
\end{enumerate}

\begin{example}
  Given a strong monad $T$ on a bicartesian closed category $\mathcal
  V$, we can extend this to a CBPV model by defining $\mathcal E$ to
  be the category $\mathcal V^T$ of algebras
\end{example}

The above definition can interpreted in any compact closed equipment
(if someone were to figure out a definition for a compact closed
equipment, that is,\ldots). Then we can get a model of a form of GTT
by taking a CBPV object in the equipment of \emph{reflexive graph
categories}. Since reflexive graphs form a topos we can get at this by
interpreting the above definition \emph{internally} to the topos of
reflexive graphs. Essentially what this means is that everything above
has a ``vertex'' component and an ``edge'' component, so we get a
cartesian category $\mathcal V_v$ which we think of as the value types
and pure functions but we also get a cartesian category $\mathcal V_e$
which we think of as the ``value edges'' and ``squares''.

We will work in ``locally thin'' models where there is at most one
square with a given boundary.

In addition to the ordinary universal properties above, when working
with reflexive graph models we also have access to new notions of
universal property that relate the ``function'' morphisms to the
``edges''.

Let $f : \mathcal V_v(A_i, A_o)$ and $R : \mathcal V_e(A_o,A')$ then a
\emph{left restriction} of $R$ along $f$ consists of
\begin{enumerate}
\item $R' : \mathcal V_e(A_i,A')$
\item $S \Rightarrow_{g,h} R'$ naturally isomorphic to $S \Rightarrow_{fg,h}R$
\end{enumerate}
In relational semantics, $R(a_o,a')$ is a relation and the left
restriction along $f$ is the relation $R(f(a_i), a')$.

Then we can formulate our value edges as follows:
\begin{enumerate}
\item We have a thin subcategory $\mathcal V_u$ of $\mathcal V_v$, of ``upcasts''
\item $\mathcal V$ has all left restrictions of edges along upcasts
\end{enumerate}

And our computation edges as follows:
\begin{enumerate}
\item We have a thin subcategory $\mathcal E_d$ of $\mathcal E_v$ of ``downcasts''
\item $\mathcal E$ has all right restrictions of edges along downcasts
\end{enumerate}

We also want something like
\[ F_c : \mathcal V_u^{op} \to \mathcal E_d \]
\[ U_c : \mathcal E_d^{op} \to \mathcal V_u \]
which ensures that if $R$ is a value edge equivalent to $A(u,-)$ then
\[ F(R) = F(A(u,-)) = (F A)(-,F u) \]

### 2023-12-09 paper-new/categorical-model.tex
more work on abstract categorical model

@@
 has a ``vertex'' component and an ``edge'' component, so we get a
-cartesian category $\mathcal V_v$ which we think of as the value types
-and pure functions but we also get a cartesian category $\mathcal V_e$
+cartesian category $\mathcal V_f$ which we think of as the value types
+and pure functions but we also get a cartesian category $\mathcal V_{sq}$
 which we think of as the ``value edges'' and ``squares''.
+
+More formally, for the values, we have
+
+\begin{enumerate}
+  \item Cartesian categories $\mathcal V_f$ and $\mathcal V_e$ with the same set of objects,
+  namely, the value types.
+  The morphisms of $\mathcal V_f$ will be called \emph{(pure) functions}, and the
+  morphisms of $\mathcal V_e$ will be called \emph{edges} or \emph{relations}.
+
+  \item A cartesian category $\mathcal V_{sq}$ whose objects are the morphisms of $\mathcal V_e$
+  and whose morphisms are \emph{commuting squares}.
+
+  \item Functors $s, t : \mathcal V_{sq} \to \mathcal V_f$ and $i : \mathcal V_f \to \mathcal V_{sq}$.
+\end{enumerate}
+
+Likewise, we have an analogous definition for computations.
 
@@
 
-Let $f : \mathcal V_v(A_i, A_o)$ and $R : \mathcal V_e(A_o,A')$ then a
-\emph{left restriction} of $R$ along $f$ consists of
+In particular, let $c \in \mathcal V_e(A, A')$ and $d \in \mathcal V_e(A',A'')$,
+and let $f \in \mathcal V_f(A, A')$ and $g \in \mathcal V_f(A', A'')$.
+We say that the pair $(c,d)$ is \emph{left-representable by} $(f, g)$
+if the following two squares commute:
+
+\begin{center}
+  \begin{tabular}{ m{14em} m{14em} } 
+    % UpL
+    \begin{tikzcd}[ampersand replacement=\&]
+      A \& {A'} \& {A''} \\
+      {A'} \&\& {A''}
+      \arrow["f"', from=1-1, to=2-1]
+      %
+      \arrow[from=1-3, to=2-3, Rightarrow, no head]
+      %
+      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+      \arrow["d", "\shortmid"{marking}, no head, from=1-2, to=1-3]
+      %
+      \arrow["d"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
+    \end{tikzcd}
+    &
+    % UpR
+    \begin{tikzcd}[ampersand replacement=\&]
+      A \&\& {A'} \\
+      {A} \& {A'} \& {A''}
+      \arrow[from=1-1, to=2-1, Rightarrow, no head]
+      %
+      \arrow["g", from=1-3, to=2-3]
+      %
+      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-3]
+      %
+      \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+      \arrow["d"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
+    \end{tikzcd}
+  \end{tabular}
+\end{center}
+
+Dually, let $c \in \mathcal E_e(B, B')$ and $d \in \mathcal E_e(B', B'')$,
+and let $f \in \mathcal E_f(B'', B')$ and $g \in \mathcal E_f(B', B)$.
+We say the pair $(c,d)$ is \emph{right-representable by} $(f,g)$ if the
+following two squares commute:
+
+\begin{center}
+  \begin{tabular}{ m{14em} m{14em} } 
+    % DnR
+    \begin{tikzcd}[ampersand replacement=\&]
+      {B} \& {B'} \& {B''} \\
+      {B} \&\& {B'}
+      \arrow[from=1-1, to=2-1, Rightarrow, no head]
+      %
+      \arrow["f", from=1-3, to=2-3]
+      %
+      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+      \arrow["d", "\shortmid"{marking}, no head, from=1-2, to=1-3]
+      %
+      \arrow["c"', from=2-1, to=2-3, no head]
+    \end{tikzcd}
+    &
+    % DnL
+    \begin{tikzcd}[ampersand replacement=\&]
+      {B'} \&\& {B''} \\
+      {B} \& {B'} \& {B''}
+      \arrow["g"', from=1-1, to=2-1]
+      %
+      \arrow[from=1-3, to=2-3, Rightarrow, no head]
+      %
+      \arrow["d", from=1-1, to=1-3, no head]
+      %
+      \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+      \arrow["d"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
+    \end{tikzcd}
+  \end{tabular}
+\end{center}
+
+Then we formulate the relationship between value edges and functions as follows:
 \begin{enumerate}
-\item $R' : \mathcal V_e(A_i,A')$
-\item $S \Rightarrow_{g,h} R'$ naturally isomorphic to $S \Rightarrow_{fg,h}R$
-\end{enumerate}
-In relational semantics, $R(a_o,a')$ is a relation and the left
-restriction along $f$ is the relation $R(f(a_i), a')$.
+  \item There is a functor $\upf : \mathcal V_e \to \mathcal V_f$ that is the
+  identity on objects, such that $\upf_*(\mathcal V_e)$, the essential image of
+  $\mathcal V_e$ under $\upf$, is thin.
+  The objects of $\upf_*(\mathcal V_e)$ are the objects of $\mathcal V_f$, and
+  the hom-set
+  $\upf_*(\mathcal V_e)(A, A') = \{ f \in \mathcal V_f(A,A') \mid \exists R \in \mathcal V_e(A,A'). \upf(R) = f \}$.
 
-Then we can formulate our value edges as follows:
-\begin{enumerate}
-\item We have a thin subcategory $\mathcal V_u$ of $\mathcal V_v$, of ``upcasts''
-\item $\mathcal V$ has all left restrictions of edges along upcasts
+  \item Every pair of morphisms $(c,d) \in \mathcal V_e(A, A') \times \mathcal V_e(A', A'')$ is
+  left-representable by $(\upf(c), \upf(d))$.
 \end{enumerate}
 
-And our computation edges as follows:
+And likewise for computations:
 \begin{enumerate}
-\item We have a thin subcategory $\mathcal E_d$ of $\mathcal E_v$ of ``downcasts''
-\item $\mathcal E$ has all right restrictions of edges along downcasts
+  \item There is a functor $\dnf : \mathcal E_e^{op} \to \mathcal V_f$ that is
+  the identity on objects, such that the essential image of $\mathcal E_e$ under
+  $\dnf$ is thin.
+
+  \item Every pair of morphisms $(c,d) \in \mathcal E_e(B, B') \times \mathcal E_e(B',B'')$
+  is right-representable by $(\dnf(d), \dnf(c))$.
 \end{enumerate}
 
+We also require that the category $\mathcal V_e$ of relations is thin ``up to an
+identity square'', i.e., for any $R, R' \in \mathcal V_e(A, A')$ we have that the
+following square commutes:
+
+\[\begin{tikzcd}[ampersand replacement=\&]
+  A \& {A'} \\
+  A \& {A'}
+  \arrow[from=1-1, to=2-1, Rightarrow, no head]
+  \arrow[from=1-2, to=2-2, Rightarrow, no head]
+  \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+  \arrow["R'"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+\end{tikzcd}\]
+
+\textbf{TODO: do we still need this?}
 We also want something like
@@
 \[ F(R) = F(A(u,-)) = (F A)(-,F u) \]
+
+\subsection{Intensional Model}
+
+An intensional model of gradual typing is defined similarly to an extensional model,
+but unlike in the extensional model, here we work internally to the category of
+categories, giving us a double category.
+
+In particular, as before, we have cartesian categories $\mathcal V_f$,
+$\mathcal V_e$, and $\mathcal V_{sq}$, in addition to $\mathcal E_f$,
+$\mathcal E_e$, and $\mathcal E_{sq}$.
+But we now have horizontal composition of squares as well.
+
+Working intensionally means we need to take into consideration the steps
+taken by terms. One consequence of this is that the squares in the representable
+properties must now involve a notion of ``delay" or ``perturbation'' in order to
+keep the function morphisms on each side in lock-step. Intuitively, the perturbations
+have no effect other than to cause the function to which they are applied to ``wait''
+in a specific manner.
+We formalize this notion by requiring that for each object $A : \mathcal V_v$,
+there is a monoid of \emph{perturbations} $P^V_A$  and a monoid homomorphism
+$\ptb^V_A : P^V_A \to \mathcal V_v(A,A)$.
+Similarly, for each $B : \mathcal E_v$ there is a monoid $P^C_B$ and a
+homomorphism $\ptb^C_B : P^C_B \to \mathcal E_v(B,B)$.
+If $\delta \in P^V_A$, we will sometimes omit the homomorphism $\ptb^V_A$ and simply write
+$\delta$ to refer to the morphism $\ptb^V_A(\delta) \in \mathcal V_v(A,A)$, and likewise
+for computation perturbations. The context will make clear whether we are referring
+to an element of the perturbation monoid or the corresponding morphism.
+
+The perturbations must satisfy a property that we call the ``push-pull'' property,
+which is formulated as follows. Let $c \in \mathcal V_e(A, A')$.
+Given a perturbation $\delta \in P^V_A$, there is a corresponding perturbation
+$\push_c(\delta) \in P^V_{A'}$. % making the following square commute:
+%
+Likewise, given $\delta' \in P^V_{A'}$ there is a perturbation $\pull_c(\delta') \in P^V_A$.
+% making the following square commute:
+
+Moreover, the following squares commute:
+
+\begin{center}
+  \begin{tabular}{ m{9em} m{9em} } 
+    \begin{tikzcd}[ampersand replacement=\&]
+      A \& {A'} \\
+      A \& {A'}
+      \arrow["\delta"', from=1-1, to=2-1]
+      \arrow["{\push_c(\delta)}", from=1-2, to=2-2]
+      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+      \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+    \end{tikzcd}
+    &
+    \begin{tikzcd}[ampersand replacement=\&]
+      A \& {A'} \\
+      A \& {A'}
+      \arrow["{\pull_c(\delta')}"', from=1-1, to=2-1]
+      \arrow["{\delta'}", from=1-2, to=2-2]
+      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+      \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+    \end{tikzcd}
+  \end{tabular}
+\end{center}
+
+The analogous property should also hold for computation relations and perturbations.
+
+A second consequence of working intensionally is that we need a way to specify
+that two morphisms are the same ``up to delay'', i.e., they differ only in that
+one may wait more than the other.
+%
+To this end, for any pair of objects $A$ and $A'$, in $\mathcal V_f$, we require that
+there is a reflexive, symmetric relation $\bisim_{A,A'}$ on the hom-set $\mathcal V_f(A, A')$,
+called the \emph{weak bisimilarity} relation.
+Similarly for the computation category: there is a relation $\bisim_{B,B'}$
+defined on each hom-set $\mathcal E_v(B, B')$.
+
+We require that all perturbations be weakly bisimilar to the identity function,
+capturing the notion that they have no effect other than to delay.
+
+The weak bisimilarity relation should respect composition:
+if $f \bisim_{A,A'} f'$ and $g \bisim_{A',A''} g'$, then
+$g \circ f \bisim_{A,A''} g' \circ f'$, and likewise for computations.
+
+% We similarly have thin subcategories $\mathcal V_u$ and $\mathcal E_d$ of
+% upcasts and downcasts. The relation between function morphisms and edges
+% is as follows.
+
+As is the case in the extensional model, there is a relationship between
+function morphisms and relation morphisms, but as mentioned above, now there
+are perturbations involved in order to keep both morphisms in lock-step.
+The precise definitions are as follows.
+
+Let $c \in \mathcal V_e(A, A')$. We say that $c$ is \emph{quasi-left-representable by}
+$f \in \mathcal V_f(A, A')$ if there exist perturbations $\delta_c^{l,e} \in P^V_A$ and
+$\delta_c^{r,e} \in P^V_{A'}$ such that the following squares commute:
+
+\begin{center}
+  \begin{tabular}{ m{7em} m{7em} } 
+    % UpL
+    \begin{tikzcd}[ampersand replacement=\&]
+      A \& {A'} \\
+      {A'} \& {A'}
+      \arrow["f"', from=1-1, to=2-1]
+      \arrow["\delta_c^{r,e}", from=1-2, to=2-2]
+      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+      \arrow[from=2-1, to=2-2, Rightarrow, no head]
+    \end{tikzcd}
+    &
+    % UpR
+    \begin{tikzcd}[ampersand replacement=\&]
+      A \& {A} \\
+      {A} \& {A'}
+      \arrow["\delta_c^{l,e}"', from=1-1, to=2-1]
+      \arrow["f", from=1-2, to=2-2]
+      \arrow[from=1-1, to=1-2, Rightarrow, no head]
+      \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+    \end{tikzcd}
+  \end{tabular}
+\end{center}
+
+Similarly, let $d \in \mathcal E_e(B, B')$. We say that $d$ is
+\emph{quasi-right-representable by} $f \in \mathcal E_f(B', B)$
+if there exist perturbations $\delta_d^{l,p} \in P^C_B$ and
+$\delta_d^{r,p} \in P^C{B'}$ such that the following squares commute:
+
+\begin{center}
+  \begin{tabular}{ m{7em} m{7em} } 
+    % DnR
+    \begin{tikzcd}[ampersand replacement=\&]
+      {B} \& {B'} \\
+      {B} \& {B}
+      \arrow["\delta_d^{l,p}"', from=1-1, to=2-1]
+      \arrow["g", from=1-2, to=2-2]
+      \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+      \arrow[from=2-1, to=2-2, Rightarrow, no head]
+    \end{tikzcd}
+    &
+    % DnL
+    \begin{tikzcd}[ampersand replacement=\&]
+      {B'} \& {B'} \\
+      {B} \& {B'}
+      \arrow["g"', from=1-1, to=2-1]
+      \arrow["\delta_d^{r,p}", from=1-2, to=2-2]
+      \arrow[from=1-1, to=1-2, Rightarrow, no head]
+      \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+    \end{tikzcd}
+  \end{tabular}
+\end{center}
+
+Besides the perturbations, one other difference between the extensional
+and intensional versions of the representability axioms is that in the
+extensional setting, the rules build in the notion of composition, whereas
+their intensional counterparts do not.
+In the extensional setting, we do not have horizontal composition of squares, which
+is required to derive the versions of the rules that build in composition
+from the versions that do not.
+In the intensional setting, we do have horizontal composition of squares,
+so we can take the simpler versions as primitive and derive the ones
+involving composition (though we must be careful about the perturbations
+when doing so!).
+
+\subsection{Constructing an Extensional Model from an Intensional Model}

### 2024-01-13 paper-new/technical-background.tex
remove section on topos of trees model

@@
 % TODO axioms (clock irrelevance, tick irrelevance)?
-
-\subsubsection{The Topos of Trees Model}
-
-The topos of trees model provides a useful intuition for reasoning in SGDT 
-\cite{birkedal-mogelberg-schwinghammer-stovring2011}.
-This section presupposes knowledge of category theory and can be safely skipped without
-disrupting the continuity.
-
-The topos of trees $\calS$ is the presheaf category $\Set^{\omega^o}$. 
-%
-We assume a universe $\calU$ of types, with encodings for operations such
-as sum types (written as $\widehat{+}$). There is also an operator 
-$\laterhat \colon \later \calU \to \calU$ such that 
-$\El(\laterhat(\nxt A)) =\,\, \later \El(A)$, where $\El$ is the type corresponding
-to the code $A$.
-
-An object $X$ is a family $\{X_i\}$ of sets indexed by natural numbers,
-along with restriction maps $r^X_i \colon X_{i+1} \to X_i$ (see Figure \ref{fig:topos-of-trees-object}).
-These should be thought of as ``sets changing over time", where $X_i$ is the view of the set if we have $i+1$
-time steps to reason about it.
-We can also think of an ongoing computation, with $X_i$ representing the potential results of the computation
-after it has run for $i+1$ steps.
-
-\begin{figure}
-  % https://q.uiver.app/?q=WzAsNSxbMiwwLCJYXzIiXSxbMCwwLCJYXzAiXSxbMSwwLCJYXzEiXSxbMywwLCJYXzMiXSxbNCwwLCJcXGRvdHMiXSxbMiwxLCJyXlhfMCIsMl0sWzAsMiwicl5YXzEiLDJdLFszLDAsInJeWF8yIiwyXSxbNCwzLCJyXlhfMyIsMl1d
-\[\begin{tikzcd}[ampersand replacement=\&]
-	{X_0} \& {X_1} \& {X_2} \& {X_3} \& \dots
-	\arrow["{r^X_0}"', from=1-2, to=1-1]
-	\arrow["{r^X_1}"', from=1-3, to=1-2]
-	\arrow["{r^X_2}"', from=1-4, to=1-3]
-	\arrow["{r^X_3}"', from=1-5, to=1-4]
-\end{tikzcd}\]
-  \caption{An example of an object in the topos of trees.}
-  \label{fig:topos-of-trees-object}
-\end{figure}
-
-A morphism from $\{X_i\}$ to $\{Y_i\}$ is a family of functions $f_i \colon X_i \to Y_i$
-that commute with the restriction maps in the obvious way, that is,
-$f_i \circ r^X_i = r^Y_i \circ f_{i+1}$ (see Figure \ref{fig:topos-of-trees-morphism}).
-
-\begin{figure}
-% https://q.uiver.app/?q=WzAsMTAsWzIsMCwiWF8yIl0sWzAsMCwiWF8wIl0sWzEsMCwiWF8xIl0sWzMsMCwiWF8zIl0sWzQsMCwiXFxkb3RzIl0sWzAsMSwiWV8wIl0sWzEsMSwiWV8xIl0sWzIsMSwiWV8yIl0sWzMsMSwiWV8zIl0sWzQsMSwiXFxkb3RzIl0sWzIsMSwicl5YXzAiLDJdLFswLDIsInJeWF8xIiwyXSxbMywwLCJyXlhfMiIsMl0sWzQsMywicl5YXzMiLDJdLFs2LDUsInJeWV8wIiwyXSxbNyw2LCJyXllfMSIsMl0sWzgsNywicl5ZXzIiLDJdLFs5LDgsInJeWV8zIiwyXSxbMSw1LCJmXzAiLDJdLFsyLDYsImZfMSIsMl0sWzAsNywiZl8yIiwyXSxbMyw4LCJmXzMiLDJdXQ==
-\[\begin{tikzcd}[ampersand replacement=\&]
-	{X_0} \& {X_1} \& {X_2} \& {X_3} \& \dots \\
-	{Y_0} \& {Y_1} \& {Y_2} \& {Y_3} \& \dots
-	\arrow["{r^X_0}"', from=1-2, to=1-1]
-	\arrow["{r^X_1}"', from=1-3, to=1-2]
-	\arrow["{r^X_2}"', from=1-4, to=1-3]
-	\arrow["{r^X_3}"', from=1-5, to=1-4]
-	\arrow["{r^Y_0}"', from=2-2, to=2-1]
-	\arrow["{r^Y_1}"', from=2-3, to=2-2]
-	\arrow["{r^Y_2}"', from=2-4, to=2-3]
-	\arrow["{r^Y_3}"', from=2-5, to=2-4]
-	\arrow["{f_0}"', from=1-1, to=2-1]
-	\arrow["{f_1}"', from=1-2, to=2-2]
-	\arrow["{f_2}"', from=1-3, to=2-3]
-	\arrow["{f_3}"', from=1-4, to=2-4]
-\end{tikzcd}\]
-  \caption{An example of a morphism in the topos of trees.}
-  \label{fig:topos-of-trees-morphism}
-\end{figure}
-
-The type operator $\later$ is defined on an object $X$ by
-$(\later X)_0 = 1$ and $(\later X)_{i+1} = X_i$.
-The restriction maps are given by $r^\later_0 =\, !$, where $!$ is the
-unique map into $1$, and $r^\later_{i+1} = r^X_i$.
-The morphism $\nxt^X \colon X \to \later X$ is defined pointwise by
-$\nxt^X_0 =\, !$, and $\nxt^X_{i+1} = r^X_i$. It is easily checked that
-this satisfies the commutativity conditions required of a morphism in $\calS$.
-%
-Given a morphism $f \colon \later X \to X$, i.e., a family of functions
-$f_i \colon (\later X)_i \to X_i$ that commute with the restrictions in the appropriate way,
-we define $\fix(f) \colon 1 \to X$ pointwise
-by $\fix(f)_i = f_{i} \circ \dots \circ f_0$.
-This can be visualized as a diagram in the category of sets as shown in
-Figure \ref{fig:topos-of-trees-fix}.
-% Observe that the fixpoint is a \emph{global element} in the topos of trees.
-% Global elements allow us to view the entire computation on a global level.
-
-% https://q.uiver.app/?q=WzAsOCxbMSwwLCJYXzAiXSxbMiwwLCJYXzEiXSxbMywwLCJYXzIiXSxbMCwwLCIxIl0sWzAsMSwiWF8wIl0sWzEsMSwiWF8xIl0sWzIsMSwiWF8yIl0sWzMsMSwiWF8zIl0sWzMsNCwiZl8wIl0sWzAsNSwiZl8xIl0sWzEsNiwiZl8yIl0sWzIsNywiZl8zIl0sWzAsMywiISIsMl0sWzEsMCwicl5YXzAiLDJdLFsyLDEsInJeWF8xIiwyXSxbNSw0LCJyXlhfMCJdLFs2LDUsInJeWF8xIl0sWzcsNiwicl5YXzIiXV0=
-% \[\begin{tikzcd}[ampersand replacement=\&]
-% 	1 \& {X_0} \& {X_1} \& {X_2} \\
-% 	{X_0} \& {X_1} \& {X_2} \& {X_3}
-% 	\arrow["{f_0}", from=1-1, to=2-1]
-% 	\arrow["{f_1}", from=1-2, to=2-2]
-% 	\arrow["{f_2}", from=1-3, to=2-3]
-% 	\arrow["{f_3}", from=1-4, to=2-4]
-% 	\arrow["{!}"', from=1-2, to=1-1]
-% 	\arrow["{r^X_0}"', from=1-3, to=1-2]
-% 	\arrow["{r^X_1}"', from=1-4, to=1-3]
-% 	\arrow["{r^X_0}", from=2-2, to=2-1]
-% 	\arrow["{r^X_1}", from=2-3, to=2-2]
-% 	\arrow["{r^X_2}", from=2-4, to=2-3]
-% \end{tikzcd}\]
-
-% \begin{figure}
-%   % https://q.uiver.app/?q=WzAsMTksWzEsMiwiWF8wIl0sWzIsMywiWF8xIl0sWzMsMSwiMSJdLFswLDEsIjEiXSxbMCwyLCJYXzAiXSxbMSwzLCJYXzEiXSxbMSwxLCIxIl0sWzIsMSwiMSJdLFsyLDIsIlhfMCJdLFsyLDQsIlhfMiJdLFszLDIsIlhfMCJdLFszLDMsIlhfMSJdLFszLDQsIlhfMiJdLFszLDUsIlhfMyJdLFs0LDIsIlxcY2RvdHMiXSxbMCwwLCJcXGZpeChmKV8wIl0sWzEsMCwiXFxmaXgoZilfMSJdLFsyLDAsIlxcZml4KGYpXzIiXSxbMywwLCJcXGZpeChmKV8zIl0sWzMsNCwiZl8wIl0sWzAsNSwiZl8xIl0sWzYsMCwiZl8wIl0sWzcsOCwiZl8wIl0sWzgsMSwiZl8xIl0sWzEsOSwiZl8yIl0sWzIsMTAsImZfMCJdLFsxMCwxMSwiZl8xIl0sWzExLDEyLCJmXzIiXSxbMTIsMTMsImZfMyJdXQ==
-%   \[\begin{tikzcd}[ampersand replacement=\&]
-%     {\fix(f)_0} \& {\fix(f)_1} \& {\fix(f)_2} \& {\fix(f)_3} \\
-%     1 \& 1 \& 1 \& 1 \\
-%     {X_0} \& {X_0} \& {X_0} \& {X_0} \& \cdots \\
-%     \& {X_1} \& {X_1} \& {X_1} \\
-%     \&\& {X_2} \& {X_2} \\
-%     \&\&\& {X_3}
-%     \arrow["{f_0}", from=2-1, to=3-1]
-%     \arrow["{f_1}", from=3-2, to=4-2]
-%     \arrow["{f_0}", from=2-2, to=3-2]
-%     \arrow["{f_0}", from=2-3, to=3-3]
-%     \arrow["{f_1}", from=3-3, to=4-3]
-%     \arrow["{f_2}", from=4-3, to=5-3]
-%     \arrow["{f_0}", from=2-4, to=3-4]
-%     \arrow["{f_1}", from=3-4, to=4-4]
-%     \arrow["{f_2}", from=4-4, to=5-4]
-%     \arrow["{f_3}", from=5-4, to=6-4]
-%   \end{tikzcd}\]
-%   \caption{The first few approximations to the guarded fixpoint of $f$.}
-%   \label{fig:topos-of-trees-fix-approx}
-% \end{figure}
-
-\begin{figure}
-  % https://q.uiver.app/?q=WzAsNixbMywwLCIxIl0sWzAsMiwiWF8wIl0sWzIsMiwiWF8xIl0sWzQsMiwiWF8yIl0sWzYsMiwiWF8zIl0sWzgsMiwiXFxkb3RzIl0sWzIsMSwicl5YXzAiXSxbNCwzLCJyXlhfMiJdLFswLDEsIlxcZml4KGYpXzAiLDFdLFswLDIsIlxcZml4KGYpXzEiLDFdLFswLDMsIlxcZml4KGYpXzIiLDFdLFswLDQsIlxcZml4KGYpXzMiLDFdLFswLDUsIlxcY2RvdHMiLDMseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMywyLCJyXlhfMSJdLFs1LDQsInJeWF8zIl1d
-  \[\begin{tikzcd}[ampersand replacement=\&,column sep=2.25em]
-    \&\&\& 1 \\
-    \\
-    {X_0} \&\& {X_1} \&\& {X_2} \&\& {X_3} \&\& \dots
-    \arrow["{r^X_0}", from=3-3, to=3-1]
-    \arrow["{r^X_2}", from=3-7, to=3-5]
-    \arrow["{\fix(f)_0}"{description}, from=1-4, to=3-1]
-    \arrow["{\fix(f)_1}"{description}, from=1-4, to=3-3]
-    \arrow["{\fix(f)_2}"{description}, from=1-4, to=3-5]
-    \arrow["{\fix(f)_3}"{description}, from=1-4, to=3-7]
-    \arrow["\cdots"{marking}, draw=none, from=1-4, to=3-9]
-    \arrow["{r^X_1}", from=3-5, to=3-3]
-    \arrow["{r^X_3}", from=3-9, to=3-7]
-  \end{tikzcd}\]
-  \caption{The guarded fixpoint of $f$.}
-  \label{fig:topos-of-trees-fix}
-\end{figure}
-
-% TODO global elements?

### 2024-01-13 paper-new/terms.tex
changes to term syntax + semantics

@@
 
-In the following section, we will extend the denotational semantics to accommodate
-the type and term precision orderings.
+In Section \ref{sec:gtlc-precision}, we will discuss how to extend the denotational
+semantics to accommodate the type and term precision orderings.
 
@@
 
+\begin{comment}
 \subsection{Removing Casts as Primitives}
@@
 
-\section{Term Semantics}\label{sec:term-semantics}
+\end{comment}
 
-\subsection{Domain-Theoretic Constructions}\label{sec:domain-theory}
+% \section{Term Semantics}\label{sec:term-semantics}
+
+\subsection{Semantic Constructions}\label{sec:domain-theory}
 
 In this section, we discuss the fundamental objects of the model into which we will embed
-the gradual lambda calculus.
+the terms of the gradually-typed lambda calculus.
 It is important to remember that the constructions in this section are entirely
@@
 % TODO explain better
-As it turns out, this definition is not quite correct; let us try to understand why.
-The price we pay for such a simple solution to the above equation is that we have
-introduced a notion of ``time'', i.e., we must now consider the stepping behavior of terms.
+As it turns out, this definition is not quite correct, as it doesn't provide a way to
+model functions that are potentially non-terminating.
+Another way to think about this is that by using guarded recursion to solve the
+equation for the dynamic type, the solution to the equation involves a notion of
+``time" or ``steps".
+So, in addition to returning a value or erroring, programs may now take one or
+more observable steps of computation, and this possibility must be reflected in
+in the equation for the dynamic type.
+
 % Therefore, the semantics of terms will need to allow for terms that potentially
@@
 % which we describe in the next section.
-Thus, in the equation for the semantics of $\dyn$, the result of the function should be
-a computation that may return a value, error, \emph{or} take an observable step.
+More specifically, in the equation for the semantics of $\dyn$, the result of the
+function should be a computation that may return a value, error, \emph{or} take an observable step.
 We model such computations using an extension of the so-called guarded lift monad
 \cite{mogelberg-paviotti2016} which we describe in the next section.
-We will then discuss the correct definition of the semantics of the dynamic type.
+We will then use this to give the correct definition of the semantics of the dynamic type.
 
@@
 we should consider their possible behaviors. On the one hand, we have \emph{failure}: a program may fail
-at run-time because of a type error. In addition to this, a program may ``think'',
-i.e., take a step of computation. If a program thinks forever, then it never returns a value,
-so we can think of the idea of thinking as a way of intensionally modelling \emph{partiality}.
+at run-time because of a type error. In addition to this, a program may take one or more steps of computation.
+If a program steps forever, then it never returns a value,
+so we can think of the idea of stepping as a way of intensionally modelling \emph{partiality}.
 
 With this in mind, we can describe a semantic object that models these behaviors: a monad
-for embedding computations that has cases for failure and ``thinking''.
+for embedding computations that has cases for failure and ``stepping''.
 Previous work has studied such a construct in the setting of the latter, called the lift
@@
 An element of $\li A$ should be viewed as a computation that can either (1) return a value (via $\eta$),
-(2) raise an error and stop (via $\mho$), or (3) think for a step (via $\theta$).
+(2) raise an error and stop (via $\mho$), or (3) take a step (via $\theta$).
 %
@@
 
-We can now give a semantics to the gradual lambda calculus we defined
-in Section \ref{sec:step-sensitive-lc}.
+We can now give a semantics to the \emph{terms} of the gradual lambda calculus we defined
+above. The full definition is given in Figure \ref{fig:term-semantics}.
 %
-Much of the semantics is similar to a normal call-by-value denotational semantics;
-we highlight the differences.
+Much of the semantics is similar to a normal call-by-value denotational semantics:
 We will interpret types as sets, and terms as functions.
@@
 %
-The interpretation of a value $\vhasty {\Gamma} V A$ will be a monotone function from
+The interpretation of a value $\vhasty {\Gamma} V A$ will be a function from
 $\sem{\Gamma}$ to $\sem{A}$. Likewise, a term $\phasty {\Gamma} M {{A}}$ will be interpreted
-as a monotone function from $\sem{\Gamma}$ to $\li \sem{A}$.
+as a function from $\sem{\Gamma}$ to $\li \sem{A}$.
 
@@
 %
-The interpretation of bind and of application both make use the monadic extend function on $\li A$.
+The interpretation of bind makes use the monadic extend function on $\li A$.
 %
@@
 
+\begin{figure*}
   \noindent Types:
@@
     \sem{\injarr{V_f}} &= \lambda \gamma . \inr\, (\sem{V_f}\, \gamma) \\[2ex]
-  \sem{\nxt\, V}     &= \lambda \gamma . \nxt (\sem{V}\, \gamma) \\
-  \sem{\theta}       &= \lambda \gamma . \theta \\
+    % \sem{\nxt\, V}     &= \lambda \gamma . \nxt (\sem{V}\, \gamma) \\
+    % \sem{\theta}       &= \lambda \gamma . \theta \\
   %
@@
   \end{align*}
+
+  \caption{Term semantics for the gradually-typed lambda calculus.}
+  \label{fig:term-semantics}
+\end{figure*}

### 2024-01-13 paper-new/ordering.tex
changes to term precision syntax + semantics

@@
 one for each term constructor. These assert that the term constructors respect term precision.
-The congruence rules are as follows:
+The congruence rules are shown in Figure \ref{fig:term-prec-congruence-rules}.
 
+\begin{figure*}
 \begin{mathpar}
@@
 \end{mathpar}
+\caption{Term precision: congruence rules.}
+\label{fig:term-prec-congruence-rules}
+\end{figure*}
 
-We then have additional equational axioms, including $\beta$ and $\eta$ laws, and
-rules characterizing upcasts as least upper bounds, and downcasts as greatest lower bounds.
-For the sake of familiarity, we formulate the cast rules using arbitrary casts; later we
-will state the analogous versions for the version of the calculus without arbitrary casts.
+We then have additional equational axioms as shown in Figure \ref{fig:term-prec-additional-rules}.
+In particular, we have a rule stating that the error term is the bottom element,
+$\beta$ and $\eta$ laws as order-equivalences, and lastly rules characterizing upcasts
+as least upper bounds, and downcasts as greatest lower bounds.
+% For the sake of familiarity, we formulate the cast rules using arbitrary casts; later we
+% will state the analogous versions for the version of the calculus without arbitrary casts.
 
@@
 
+\begin{figure*}
 \begin{mathpar}
@@
 \end{mathpar}
+\caption{Term precision: additional rules.}
+\label{fig:term-prec-additional-rules}
+\end{figure*}
 
 % TODO explain the least upper bound/greatest lower bound rules
-The rules UpR, UpL, DnL, and DnR were introduced in \cite{new-licata18} as a means
+The rules \textsc{UpR}, \textsc{UpL}, \textsc{DnL}, and \textsc{DnR}
+were introduced in \cite{new-licata18} as a means
 of cleanly axiomatizing the intended behavior of casts in a way that
 doesn't depend on the specific constructs of the language.
-Intuitively, rule UpR says that the upcast of $M$ is an upper bound for $M$
-in that $M$ may error more, and UpL says that the upcast is the \emph{least}
+Intuitively, rule \textsc{UpR} says that the upcast of $M$ is an upper bound for $M$
+in that $M$ may error more, and \textsc{UpL} says that the upcast is the \emph{least}
 such upper bound, in that it errors more than any other upper bound for $M$.
-Conversely, DnL says that the downcast of $M$ is a lower bound, and DnR says
+Conversely, \textsc{DnL} says that the downcast of $M$ is a lower bound, and \textsc{DnR} says
 that it is the \emph{greatest} lower bound.
@@
 Thus, in order for the right-hand side to be related to the downcast,
-we need to insert a delay on the right.
+the right-hand side must be of the form $\theta(\dots)$, and the argument to
+$\theta$ on the LHS must be related to the argument of $\theta$ on the RHS.
+We can accomplish this by inserting a ``delay" on the right, i.e., wrapping the
+RHS in $\delta = \theta \circ \nxt$.
+% Maybe show the correct form of the rule in the special case $c = inj-arr$.
 %
@@
 
-% Delays for function types and for inj-arr(c)
+The specific manner in which these delays must be inserted for a given type precision
+derivation $c$ is best examined in a more abstract setting, as it is independent of the
+particular concrete model under consideration.
+Thus, we postpone this discussion to the section on the abstract intensional categorical models of
+gradual typing (Section \ref{sec:abstract-intensional-models}).
 
+\begin{comment}
 \subsubsection{Perturbations}
@@
 $\delta$ unless there is opportunity for confusion.
+
+% Delays for function types and for inj-arr(c)
+
+\end{comment}

### 2024-01-13 paper-new/categorical-model.tex
work on section on abstract categorical models

@@
 \item An action of $\mathcal V^{op}$ (with the $\mathcal V$ cartesian
-  product as monoidal structure) on $\mathcal E$. We write this as
-  exponentiation $B^A$.
+  product as monoidal structure) on $\mathcal E$. We write this with
+  an arrow $A \arr B$.
 
   This means we have 
-  \[ \alpha : B^{A_1 \times A_2} \cong (B^{A_1})^{A^2} \]
+  \[ \alpha : {A_1 \times A_2} \arr B \cong A_2 \arr (A_1 \arr B) \]
   and
-  \[ i : B^1 \cong B \]
-  satisfying coherence isomorphisms
+  \[ i : 1 \arr B \cong B \]
+  satisfying coherence isomorphisms.
 \item A functor $U : \mathcal E \to \mathcal V$ that ``preserves
-  powering'' in that every $U(B^A)$ is an exponential of $UB$ by $A$
+  powering'' in that every $U(A \arr B)$ is an exponential of $UB$ by $A$
   and that $U\alpha$ and $Ui$ are mapped to the canonical isomorphisms
@@
 
-The above definition can interpreted in any compact closed equipment
-(if someone were to figure out a definition for a compact closed
-equipment, that is,\ldots). Then we can get a model of a form of GTT
-by taking a CBPV object in the equipment of \emph{reflexive graph
-categories}. Since reflexive graphs form a topos we can get at this by
-interpreting the above definition \emph{internally} to the topos of
-reflexive graphs. Essentially what this means is that everything above
-has a ``vertex'' component and an ``edge'' component, so we get a
-cartesian category $\mathcal V_f$ which we think of as the value types
-and pure functions but we also get a cartesian category $\mathcal V_{sq}$
-which we think of as the ``value edges'' and ``squares''.
-
-More formally, for the values, we have
+% morphisms of CBPV models
+Given call-by-push-value models
+$\mathcal M_1 = (\mathcal V_1, \mathcal E_1, \arr_1, U_1, F_1)$ and
+$\mathcal M_2 = (\mathcal V_2, \mathcal E_2, \arr_2, U_2, F_2)$,
+A morphism $G$ from $M_1$ to $M_2$ is given by a pair of functors
+$G_{\mathcal{V}}: V_1 \to V_2$ and $G_{\mathcal{E}} : E_1 \to E_2$ such that
 
 \begin{enumerate}
-  \item Cartesian categories $\mathcal V_f$ and $\mathcal V_e$ with the same set of objects,
-  namely, the value types.
-  The morphisms of $\mathcal V_f$ will be called \emph{(pure) functions}, and the
-  morphisms of $\mathcal V_e$ will be called \emph{edges} or \emph{relations}.
+  \item $G_{\mathcal{E}} \circ F_1 = F_2 \circ G_{\mathcal{V}}$
+  \item $G_{\mathcal{V}} \circ U_1 = U_2 \circ G_{\mathcal{E}}$
+  \item $G_{\mathcal{E}}(A \arr_1 B) = G_{\mathcal{V}}(A) \arr_2 G_{\mathcal{E}}(B)$
+\end{enumerate}
 
-  \item A cartesian category $\mathcal V_{sq}$ whose objects are the morphisms of $\mathcal V_e$
-  and whose morphisms are \emph{commuting squares}.
+% TODO what about the product and coproducts in V?
+% Do we need that G_V(A \times_1 A') = G_V(A) \times_2 G_V(A')
+% and likewise for coproducts?
 
-  \item Functors $s, t : \mathcal V_{sq} \to \mathcal V_f$ and $i : \mathcal V_f \to \mathcal V_{sq}$.
+With these definitions, we can define a category whose objects are CBPV
+models, and whose arrows are CBPV morphisms.
+This category will serve as the setting in which we formulate the definition
+of a model of extensional gradual typing.
+
+\subsection{Extensional Models of Gradual Typing}
+
+An model $\mathcal{M}$ of extensional gradual typing consists of:
+\begin{itemize}
+  \item CBPV models $\mathcal M_f$ and $\mathcal M_{sq}$
+  \item CBPV morphisms $r : \mathcal M_f \to \mathcal M_{sq}$ and 
+  $s, t : \mathcal M_{sq} \to \mathcal M_f$
+\end{itemize}
+
+satisfying certain additional conditions that will be described below.
+
+% % https://q.uiver.app/#q=WzAsMixbMCwxLCJcXG1hdGhjYWx7TX1fe3NxfSJdLFswLDAsIlxcbWF0aGNhbHtNfV9mIl0sWzEsMCwiciJdLFswLDEsInMiLDAseyJjdXJ2ZSI6LTJ9XSxbMCwxLCJ0IiwyLHsiY3VydmUiOjJ9XV0=
+% \[\begin{tikzcd}[ampersand replacement=\&]
+% 	{\mathcal{M}_f} \\
+% 	{\mathcal{M}_{sq}}
+% 	\arrow["r", from=1-1, to=2-1]
+% 	\arrow["s", curve={height=-12pt}, from=2-1, to=1-1]
+% 	\arrow["t"', curve={height=12pt}, from=2-1, to=1-1]
+% \end{tikzcd}\]
+
+Spelling this out in light of the above definitions, we see that this is
+equivalent to the following in the category $\textbf{Cat}$:
+
+% https://q.uiver.app/#q=WzAsNCxbMCwyLCJcXHZzcSJdLFsyLDIsIlxcZXNxIl0sWzAsMCwiXFx2ZiJdLFsyLDAsIlxcZWYiXSxbMiwzLCJcXEZmIiwwLHsiY3VydmUiOi0yfV0sWzMsMiwiXFxVZiIsMCx7ImN1cnZlIjotMn1dLFswLDEsIlxcRnNxIiwwLHsiY3VydmUiOi0yfV0sWzEsMCwiXFxVc3EiLDAseyJjdXJ2ZSI6LTJ9XSxbMiwwLCJcXHJ2Il0sWzAsMiwiXFxzdiIsMCx7ImN1cnZlIjotMn1dLFsyLDAsIlxcdHYiLDAseyJjdXJ2ZSI6LTJ9XSxbMSwzLCJcXHNlIiwwLHsiY3VydmUiOi0yfV0sWzMsMSwiXFx0ZSIsMCx7ImN1cnZlIjotMn1dLFszLDEsIlxccmUiXSxbNCw1LCJcXGJvdCIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs2LDcsIlxcYm90IiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+\[\begin{tikzcd}[ampersand replacement=\&]
+	\vf \&\& \ef \\
+	\\
+	\vsq \&\& \esq
+	\arrow[""{name=0, anchor=center, inner sep=0}, "\Ff", curve={height=-12pt}, from=1-1, to=1-3]
+	\arrow[""{name=1, anchor=center, inner sep=0}, "\Uf", curve={height=-12pt}, from=1-3, to=1-1]
+	\arrow[""{name=2, anchor=center, inner sep=0}, "\Fsq", curve={height=-12pt}, from=3-1, to=3-3]
+	\arrow[""{name=3, anchor=center, inner sep=0}, "\Usq", curve={height=-12pt}, from=3-3, to=3-1]
+	\arrow["\rv", from=1-1, to=3-1]
+	\arrow["\sv", curve={height=-12pt}, from=3-1, to=1-1]
+	\arrow["\tv", curve={height=-12pt}, from=1-1, to=3-1]
+	\arrow["\se", curve={height=-12pt}, from=3-3, to=1-3]
+	\arrow["\te", curve={height=-12pt}, from=1-3, to=3-3]
+	\arrow["\re", from=1-3, to=3-3]
+	\arrow["\bot"{description}, draw=none, from=0, to=1]
+	\arrow["\bot"{description}, draw=none, from=2, to=3]
+\end{tikzcd}\]
+
+% The above definition can be interpreted in any compact closed equipment
+% (if someone were to figure out a definition for a compact closed
+% equipment, that is,\ldots). Then we can get a model of a form of GTT
+% by taking a CBPV object in the equipment of \emph{reflexive graph
+% categories}. Since reflexive graphs form a topos we can get at this by
+% interpreting the above definition \emph{internally} to the topos of
+% reflexive graphs. Essentially what this means is that everything above
+% has a ``vertex'' component and an ``edge'' component, so we get a
+% cartesian category $\mathcal V_f$ which we think of as the value types
+% and pure functions but we also get a cartesian category $\mathcal V_{sq}$
+% which we think of as the ``value edges'' and ``squares''.
+
+That is, for the values, we have
+
+\begin{enumerate}
+  \item A cartesian category $\mathcal V_f$.
+  The objects of $\mathcal V_f$ will be called \emph{value types}.
+  The morphisms of $\mathcal V_f$ will be called \emph{(pure) functions}.
+
+  \item A cartesian category $\mathcal V_{sq}$. 
+  The objects of $\mathcal V_{sq}$ will be called \emph{value edges} or
+  \emph{value relations}, and the morphisms are \emph{commuting squares}.
+
+  \item Functors $\sv, \tv : \mathcal V_{sq} \to \mathcal V_f$ and 
+  $\rv : \mathcal V_f \to \mathcal V_{sq}$.
 \end{enumerate}
@@
 
+We write $c : A \rel A'$ to mean that $c \in \ob(\vsq)$ such that 
+$\sv(c) = A$ and $\tv(A) = A'$.
+Likewise, let $c_i : A_i \rel A_i'$ and $c_o : A_o \rel A_o'$,
+and let $f \in \vf(A_i, A_o)$ and $f' \in \vf(A_i', A_o')$.
+The judgment $\beta : f \ltdyn_{c_o}^{c_i} f'$ is defined to mean
+
+\begin{enumerate}
+  \item $\beta \in \vsq(c_i, c_o)$
+  \item $\sv(\beta) = f$
+  \item $\tv(\beta) = f'$
+\end{enumerate}
+
+(Recall that $\sv$ and $\tv$ are functors, so in addition to acting on
+the objects of $\vsq$ they also act on morphisms.)
+
+Picorially, this is depicted as a commuting square:
+
+% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBX2kiXSxbMSwwLCJBX2knIl0sWzAsMSwiQV9vIl0sWzEsMSwiQV9vJyJdLFswLDIsImYiLDJdLFsxLDMsImYnIl0sWzAsMSwiY19pIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMiwzLCJjX28iLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs0LDUsIlxcYWxwaGEiLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
+\[\begin{tikzcd}[ampersand replacement=\&]
+	{A_i} \& {A_i'} \\
+	{A_o} \& {A_o'}
+	\arrow[""{name=0, anchor=center, inner sep=0}, "f"', from=1-1, to=2-1]
+	\arrow[""{name=1, anchor=center, inner sep=0}, "{f'}", from=1-2, to=2-2]
+	\arrow["{c_i}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+	\arrow["{c_o}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+	\arrow["\alpha"{description}, draw=none, from=0, to=1]
+\end{tikzcd}\]
+
+Composition of squares $\beta : f \ltdyn_{c_2}^{c_1} g$ and $\beta' : f' \ltdyn_{c_3}^{c_2} g'$
+corresponds to ``stacking'' the square for $\beta'$ below the square for $\beta$.
+Fuctoriality of $s$ and $t$ ensure that the left and right sides of the resulting square are as expected,
+i.e., we get $\beta' \circ \beta : f' \circ f \ltdyn_{c_3}^{c_1} g' \circ g$.
+
+% Fuctoriality of $s$ and $t$ ensure that we can ``vertically" compose 
+% $\beta : f \ltdyn_{c_2}^{c_1} g$ and $\beta' : f' \ltdyn_{c_3}^{c_2} g'$
+% to obtain $\beta' \circ \beta : f' \circ f \ltdyn_{c_3}^{c_1} g' \circ g$.
+% Pictorially, this is represented by ``stacking'' the square for
+% $\beta'$ below the square for $\beta$.
+
+All of the above holds in an analogous manner for the computations.
+
 We will work in ``locally thin'' models where there is at most one
 square with a given boundary.
+That is, if $\beta, \beta' : f \ltdyn_{c_o}^{c_i} f'$, then $\beta = \beta'$.
 
-In addition to the ordinary universal properties above, when working
-with reflexive graph models we also have access to new notions of
-universal property that relate the ``function'' morphisms to the
-``edges''.
+We also have a ``horizontal" composition operation on value edges and on computation edges.
+That is, let 
+%
+\[ X = \{ (c, c') \in \ob(\vsq) \times \ob(\vsq) \mid \tv(c) = \sv(c') \}. \]
+%
+There is an operation $\comp : X \to \ob(\vsq)$ such that $\sv(c \comp c') = \sv(c)$
+and $\tv(c \comp c') = \tv(c')$. Likewise for computations.
+%
+Importantly, we emphasize that this composition is \emph{not} a functor: we only
+require that it act on \emph{objects} of $\vsq$ (i.e. edges) and not on the morphisms
+(i.e. the squares). Intuivitely, this has to do with the fact that in the extensional
+setting, the semantic term precision function is \emph{not} transitive.
 
-In particular, let $c \in \mathcal V_e(A, A')$ and $d \in \mathcal V_e(A',A'')$,
+With this definition, is easily shown that there is a category $\ve$ of value relations,
+whose objects are the objects of $\vf$, and such that $\ve(A, A')$ is the set of objects
+$c$ of $\vsq$ whose source is $A$ and whose target is $A'$.
+Composition of morphisms is defined using the above operation $\comp$.
+Similarly, we have a category $\ee$ of computation relations.
+
+We also require that the category $\ve$ of relations is thin ``up to an
+identity square'', i.e., for any $c, c' \in \ve(A, A')$ we have that the
+following square commutes:
+
+\[\begin{tikzcd}[ampersand replacement=\&]
+  A \& {A'} \\
+  A \& {A'}
+  \arrow[from=1-1, to=2-1, Rightarrow, no head]
+  \arrow[from=1-2, to=2-2, Rightarrow, no head]
+  \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+  \arrow["c'"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+\end{tikzcd}\]
+
+% In addition to the ordinary universal properties above, when working
+% with reflexive graph models we also have access to new notions of
+% universal property that relate the ``function'' morphisms to the
+% ``edges''.
+
+Finally, we specify the abstract behavior of casts.
+In brief, given $c : A \rel A'$, an upcast will be a morphism $\upf\, c$ in $\vf(A, A')$
+such that $\upf\, c$ ``represents $c$" (we will define this shortly). % such that $c$ is \emph{left-representable} by $\up c$.
+In reality, it is slightly more complicated, since we need to build composition of edges
+into the definition, but the idea is similar.
+
+In particular, let $c : A \rel A'$ and $d : A' \rel A''$,
 and let $f \in \mathcal V_f(A, A')$ and $g \in \mathcal V_f(A', A'')$.
@@
 
-Then we formulate the relationship between value edges and functions as follows:
+Then we formulate the relationship between value relation morphisms and
+function morphisms as follows:
 \begin{enumerate}
@@
   the hom-set
-  $\upf_*(\mathcal V_e)(A, A') = \{ f \in \mathcal V_f(A,A') \mid \exists R \in \mathcal V_e(A,A'). \upf(R) = f \}$.
+  $\upf_*(\mathcal V_e)(A, A') = \{ f \in \mathcal V_f(A,A') \mid \exists c \in \mathcal V_e(A,A'). \upf(c) = f \}$.
 
@@
 
-We also require that the category $\mathcal V_e$ of relations is thin ``up to an
-identity square'', i.e., for any $R, R' \in \mathcal V_e(A, A')$ we have that the
-following square commutes:
-
-\[\begin{tikzcd}[ampersand replacement=\&]
-  A \& {A'} \\
-  A \& {A'}
-  \arrow[from=1-1, to=2-1, Rightarrow, no head]
-  \arrow[from=1-2, to=2-2, Rightarrow, no head]
-  \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-  \arrow["R'"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-\end{tikzcd}\]
-
 \textbf{TODO: do we still need this?}
@@
 
-\subsection{Intensional Model}
+In summary, an extensional model consists of:
+\begin{enumerate}
+  \item CBPV models $\mathcal M_f$ and $\mathcal M_{sq}$
+  \item CBPV morphisms $r : \mathcal M_f \to \mathcal M_{sq}$ and $s, t : \mathcal M_{sq} \to \mathcal M_f$
+  \item Thinness: there is at most one square with a given boundary
+  \item A ``horizontal composition" operation on value relations and on computation relations
+  (from which we can define the categories $\ve$ and $\ee$ of value types/relations and computation types/relations, repsectively).
+  \item The categories $\ve$ and $\ee$ are thin up to an identity square
+  \item A functor $\upf : \mathcal V_e \to \mathcal V_f$ that is the identity on objects,
+  such that $\upf_*(\mathcal V_e)$, the essential image of $\mathcal V_e$ under $\upf$, is thin.
+  \item Every pair of morphisms $(c,d) \in \mathcal V_e(A, A') \times \mathcal V_e(A', A'')$ is
+  left-representable by $(\upf(c), \upf(d))$.
+  \item A functor $\dnf : \mathcal E_e^{op} \to \mathcal V_f$ that is
+  the identity on objects, such that the essential image of $\mathcal E_e$ under $\dnf$ is thin.
+  \item Every pair of morphisms $(c,d) \in \mathcal E_e(B, B') \times \mathcal E_e(B',B'')$
+  is right-representable by $(\dnf(d), \dnf(c))$.
+\end{enumerate}
+
+\subsection{Intensional Models}\label{sec:abstract-intensional-models}
 
 An intensional model of gradual typing is defined similarly to an extensional model,
-but unlike in the extensional model, here we work internally to the category of
-categories, giving us a double category.
+with two key differences that will be discussed below.
+%
+The starting point is similar to that of the extensional model: an intensional model
+will be given by a diagram in the category of CBPV objects, satisfying
+additional properties.
+%
+This time, however, since we are working intensionally, the semantic denotation of
+term precision \emph{is} transitive, so we \emph{do} have a horizontal composition
+operation on squares. Compare this to the extensional case, where we could only
+compose \emph{relations} horizontally, not squares.
+What this means is that we can define a functor for composition of value relations
+and squares, and a functor for composition of computation relations and squares.
 
-In particular, as before, we have cartesian categories $\mathcal V_f$,
-$\mathcal V_e$, and $\mathcal V_{sq}$, in addition to $\mathcal E_f$,
-$\mathcal E_e$, and $\mathcal E_{sq}$.
-But we now have horizontal composition of squares as well.
+We can specify this elegantly as a category internal to the category of CBPV models.
+In particular, as in the extensional case, there is a CBPV model of ``objects''
+$\mathcal M_f$ and a CBPV model of ``arrows'' $\mathcal M_{sq}$. 
+There are CBPV morphisms
+$r : \mathcal M_f \to \mathcal M_{sq}$ and $s, t : \mathcal M_{sq} \to \mathcal M_f$,
+just as before.
+%
+But now, we also have a CBPV morphism $m$ from the pullback 
+$\mathcal M_{sq} \times_{s = t} \mathcal M_{sq}$ to $M_{sq}$, i.e., ``composition of arrows".
+In particular, this consists of a functor 
+$m_{\mathcal{V}} : \vsq \times_{\sv = \tv} \vsq \to \vsq$ for composition of value
+relations/squares, and a functor $m_{\mathcal{E}} : \esq \times_{\se = \te} \esq \to \esq$
+for composition of computation relations/squares.
+
+% If we spell this all out explicitly, we end up with a definition similar to the
+% one for the extensional case, but now with the addition of a functor $m_{\mathcal{V}}$ for
+% composition of value relations/squares and a functor $m_{\mathcal{E}}$ for
+% composition of computation relations/squares.
+
+For the sake of ease of reference, we recap the definition of a step-1 model:
+
+\begin{definition}
+  A \emph{step-1} model of intensional gradual typing consists of a category internal to the
+  category of CBPV models.
+\end{definition}
+
+% In particular, as before, we have cartesian categories $\mathcal V_f$,
+% $\mathcal V_e$, and $\mathcal V_{sq}$, in addition to $\mathcal E_f$,
+% $\mathcal E_e$, and $\mathcal E_{sq}$.
+% But we now have horizontal composition of squares as well.
+
+\subsubsection{Bisimilarity}
 
 Working intensionally means we need to take into consideration the steps
-taken by terms. One consequence of this is that the squares in the representable
+taken by terms. One consequence of this is that we need a way to specify
+that two morphisms are the same ``up to delay'', i.e., they differ only in that
+one may wait more than the other.
+
+In particular, for any pair of objects $A$ and $A'$, in $\vf$,
+we require that there is a reflexive, symmetric relation $\bisim_{A,A'}$ on the
+hom-set $\vf(A, A')$, called the \emph{weak bisimilarity} relation.
+Similarly for the computation category: there is a reflexive, symmetric relation
+$\bisim_{B,B'}$ defined on each hom-set $\ef(B, B')$.
+%
+Additionally, the weak bisimilarity relation should respect composition:
+if $f \bisim_{A,A'} f'$ and $g \bisim_{A',A''} g'$, then
+$g \circ f \bisim_{A,A''} g' \circ f'$, and likewise for computations.
+
+We can specify all of this abstractly via categories $\vsim$ and $\esim$ along with
+functors $\rvsim : \vf \to \vsim$ and $\svsim, \tvsim : \vsim \to \vf$,
+and likewise for computations.
+Since bisimilarity of morphisms $f$ and $f'$ requires that they share source and target,
+we require that $\svsim$ and $\tvsim$ agree on objects and likewise for $\sesim and \tesim$.
+Thus, the objects of $\vsim$ are identified with $\ob(\vf)$.
+The morphisms of $\vsim$ are ``bisimilarity proofs'', analogous to the commuting squares of $\vsq$.
+
+There is also a ``symmetry'' endofunctor $\text{sym}_{\mathcal{V}}^\bisim : \vsim \to \vsim$
+such that $\svsim \circ \text{sym}_{\mathcal{V}}^\bisim = \tvsim$
+and $\tvsim \circ \text{sym}_{\mathcal{V}}^\bisim = \svsim$,
+and $\text{sym}_{\mathcal{V}}^\bisim \circ \text{sym}_{\mathcal{V}}^\bisim$ is the identity.
+Likewise there is a symmetry endofunctor $\text{sym}_{\mathcal{E}}^\bisim : \esim \to \esim$.
+
+% TODO explain what this means in terms of bisimilarity "squares"?
+
+% Spelling this all out concretely, for any pair...
+
+\begin{definition}
+A \emph{step-2 intensional model} consists of all the data of a step-1 intensional model along
+with the additional categories and functors for bisimiarity described above.
+\end{definition}
+
+\subsubsection{Perturbations}
+
+A second consequence of working intensionally is that the squares in the representable
 properties must now involve a notion of ``delay" or ``perturbation'' in order to
@@
 in a specific manner.
-We formalize this notion by requiring that for each object $A : \mathcal V_v$,
+We formalize this notion by requiring that for each object $A$ in $\vf$,
 there is a monoid of \emph{perturbations} $P^V_A$ and a monoid homomorphism
-$\ptb^V_A : P^V_A \to \mathcal V_v(A,A)$.
-Similarly, for each $B : \mathcal E_v$ there is a monoid $P^C_B$ and a
-homomorphism $\ptb^C_B : P^C_B \to \mathcal E_v(B,B)$.
-If $\delta \in P^V_A$, we will sometimes omit the homomorphism $\ptb^V_A$ and simply write
-$\delta$ to refer to the morphism $\ptb^V_A(\delta) \in \mathcal V_v(A,A)$, and likewise
+$\ptbv_A : \pv_A \to \vf(A,A)$.
+Similarly, for each $B : \ef$ there is a monoid $P^C_B$ and a
+homomorphism $\ptbe_B : \pe_B \to \ef(B,B)$.
+If $\delta \in P^V_A$, we will sometimes omit the homomorphism $\ptbv_A$ and simply write
+$\delta$ to refer to the morphism $\ptbv_A(\delta) \in \vf(A,A)$, and likewise
 for computation perturbations. The context will make clear whether we are referring
@@
 
-The perturbations must satisfy a property that we call the ``push-pull'' property,
-which is formulated as follows. Let $c \in \mathcal V_e(A, A')$.
-Given a perturbation $\delta \in P^V_A$, there is a corresponding perturbation
-$\push_c(\delta) \in P^V_{A'}$. % making the following square commute:
+The perturbations must be preserved by $\times$, $\arr$, $U$, and $F$.
+
+For reasons that will be made clear in the next section, perturbations must also satisfy a property that we call the ``push-pull'' property,
+which is formulated as follows. Let $c : A \rel A'$.
+Given a perturbation $\delta \in \pv_A$, there is a corresponding perturbation
+$\push_c(\delta) \in \pv_{A'}$. % making the following square commute:
 %
-Likewise, given $\delta' \in P^V_{A'}$ there is a perturbation $\pull_c(\delta') \in P^V_A$.
+Likewise, given $\delta' \in \pv_{A'}$ there is a perturbation $\pull_c(\delta') \in \pv_A$.
 % making the following square commute:
 
-Moreover, the following squares commute:
+Moreover, push-pull states that the following squares must commute:
 
@@
 
-A second consequence of working intensionally is that we need a way to specify
-that two morphisms are the same ``up to delay'', i.e., they differ only in that
-one may wait more than the other.
-%
-To this end, for any pair of objects $A$ and $A'$, in $\mathcal V_f$, we require that
-there is a reflexive, symmetric relation $\bisim_{A,A'}$ on the hom-set $\mathcal V_f(A, A')$,
-called the \emph{weak bisimilarity} relation.
-Similarly for the computation category: there is a relation $\bisim_{B,B'}$
-defined on each hom-set $\mathcal E_v(B, B')$.
+Lastly, we require that all perturbations be weakly bisimilar to the identity morphism,
+capturing the notion that they have no effect other than to delay. We observe that
+the set of endomorphisms $f$ such that $f$ is weakly bisimilar to the identity
+forms a monoid under composition.
 
-We require that all perturbations be weakly bisimilar to the identity function,
-capturing the notion that they have no effect other than to delay.
+This is summarized below:
 
-The weak bisimilarity relation should respect composition:
-if $f \bisim_{A,A'} f'$ and $g \bisim_{A',A''} g'$, then
-$g \circ f \bisim_{A,A''} g' \circ f'$, and likewise for computations.
+\begin{definition}
+  A \emph{step-3} model of intensional gradual typing consists of all the data of a step-2 model plus:
+  \begin{enumerate}
+    \item $\pv_A$ and a monoid homomorphism 
+      \[ \ptbv_A : \pv_A \to \{ f \in \vf(A,A) \mid f \bisim \id \} \]
+    \item $\pe_B$ and a monoid homomorphism 
+      \[ \ptbe_B : \pe_B \to \{ g \in \ef(B,B) \mid g \bisim \id \} \]
+    \item The functors $\times$, $\arr$, $U$, and $F$ preserve perturbations.
+    \item The push-pull property holds for all $c : A \rel A'$ and all $d : B \rel B'$.
+  \end{enumerate}
+\end{definition}
+
+\subsubsection{Behavior of Casts}
 
@@
 As is the case in the extensional model, there is a relationship between
-function morphisms and relation morphisms, but as mentioned above, now there
-are perturbations involved in order to keep both morphisms in lock-step.
+vertical (i.e., function) morphisms and horizontal (i.e., relation) morphisms,
+but as mentioned above, now there
+are perturbations involved in order to keep both sides ``in lock-step".
 The precise definitions are as follows.
 
-Let $c \in \mathcal V_e(A, A')$. We say that $c$ is \emph{quasi-left-representable by}
-$f \in \mathcal V_f(A, A')$ if there exist perturbations $\delta_c^{l,e} \in P^V_A$ and
-$\delta_c^{r,e} \in P^V_{A'}$ such that the following squares commute:
+Let $c : A \rel A'$. We say that $c$ is \emph{quasi-left-representable by}
+$f \in \vf(A, A')$ if there are perturbations $\delta_c^{l,e} \in \pv_A$ and
+$\delta_c^{r,e} \in \pv_{A'}$ such that the following squares commute:
 
@@
 
-Similarly, let $d \in \mathcal E_e(B, B')$. We say that $d$ is
-\emph{quasi-right-representable by} $f \in \mathcal E_f(B', B)$
-if there exist perturbations $\delta_d^{l,p} \in P^C_B$ and
-$\delta_d^{r,p} \in P^C{B'}$ such that the following squares commute:
+Similarly, let $d : B \rel B'$. We say that $d$ is
+\emph{quasi-right-representable by} $f \in \ef(B', B)$
+if there exist perturbations $\delta_d^{l,p} \in \pe_B$ and
+$\delta_d^{r,p} \in \pe_{B'}$ such that the following squares commute:
 
@@
 
-\subsection{Constructing an Extensional Model from an Intensional Model}
+\begin{definition}
+  A \emph{step-4 intensional model}, or simply an \emph{intensional model}, consists
+  of all the data of a step-3 intensional model, with the following additional data:
+  \begin{enumerate}
+    \item Functors $\upf : \ve \to \vf$ and $\dnf : \ee^{op} \to \ef$ % TODO: image is thin?
+    \item Every value edge $c : A \rel A'$ is quasi-left-representable by $\upf(c)$ and
+    every computation edge $d : B \rel B'$ is quasi-right-representable by $\dnf(d)$.
+  \end{enumerate}
+\end{definition}

### 2024-01-13 paper-new/ordering.tex
rename "double poset" to "predomain"

@@
 
-\subsubsection{Double Posets}\label{sec:predomains}
+\subsubsection{Predomains}\label{sec:predomains}
 
@@
 The semantic objects that interpret our types should therefore be equipped with
-two relations. We call these objects ``double posets''.
-A double poset $A$ is a set with two relations: an partial order $\semlt_A$ on $A$, and
+two relations. We call these objects ``predomains''.
+A predomain $A$ is a set with two relations: an partial order $\semlt_A$ on $A$, and
 a reflexive, symmetric relation $\bisim_A$ on $A$.
@@
 
-We define morphisms of double posets as functions that preserve both
-the ordering and the bisimilarity relation. Given double posets
+We define morphisms of predomains as functions that preserve both
+the ordering and the bisimilarity relation. Given predomains
 $A$ and $B$, we write $f \colon A \monto B$ to indicate that $f$ is a morphism
@@
 
-We define an ordering on morphisms of double posets as
+We define an ordering on morphisms of predomains as
 $f \le g$ if for all $a$ in $\ty{A_i}$, we have $f(a) \le_{A_o} g(a)$,
@@
 
-For each type we want to represent, we define a double poset for the corresponding semantic
-type. For instance, we define a double poset for natural numbers, for the
+For each type we want to represent, we define a predomain for the corresponding semantic
+type. For instance, we define a predomain for natural numbers, for the
 dynamic type, for functions, and for the lift monad. We
@@
 \begin{itemize}
-  \item There is a double poset $\Nat$ for natural numbers, where the ordering and the
+  \item There is a predomain $\Nat$ for natural numbers, where the ordering and the
   bisimilarity relations are both equality.
@@
   % TODO explain that there is a theta operator for posets?
-  \item There is a double poset $\Dyn$ to represent the dynamic type. The underlying type
-  for this double poset is defined by guarded fixpoint to be such that
+  \item There is a predomain $\Dyn$ to represent the dynamic type. The underlying type
+  for this predomain is defined by guarded fixpoint to be such that
   $\ty{\Dyn} \cong \mathbb{N}\, +\, \later (\ty{\Dyn} \monto \li \ty{\Dyn})$.
@@
 
-  \item For a double poset $A$, there is a double poset $\li A$ for the ``lift'' of $A$
+  \item For a predomain $A$, there is a predomain $\li A$ for the ``lift'' of $A$
   using the lift monad. We use the same notation for $\li A$ when $A$ is a type
-  and $A$ is a double poset, since the context should make clear which one we are referring to.
+  and $A$ is a predomain, since the context should make clear which one we are referring to.
   The underling type of $\li A$ is simply $\li \ty{A}$, i.e., the lift of the underlying
@@
 
-  \item For double posets $A_i$ and $A_o$, we form the double poset of monotone functions
+  \item For predomains $A_i$ and $A_o$, we form the predomain of monotone functions
   from $A_i$ to $A_o$, which we denote by $A_i \To A_o$.
 
-  \item Given a double poset $A$, we can form the double poset $\later A$ whose underlying
+  \item Given a predomain $A$, we can form the predomain $\later A$ whose underlying
   type is $\later \ty{A}$. We define $\tilde{x} \le_{\later A} \tilde{y}$ to be
@@
 
-As mentioned, the ordering on the lift of a double poset $A$ is called the
+As mentioned, the ordering on the lift of a predomain $A$ is called the
 \emph{step-sensitive error-ordering} (also called ``lock-step error ordering''),
@@
 We also define a heterogeneous version of this ordering between the lifts of two
-different double posets $A$ and $B$, parameterized by a relation $R$ between $A$ and $B$.
+different predomains $A$ and $B$, parameterized by a relation $R$ between $A$ and $B$.
 
@@
 
-For a double poset $A$, we define a relation on $\li A$, called ``weak bisimilarity",
+For a predomain $A$, we define a relation on $\li A$, called ``weak bisimilarity",
 written $l \bisim l'$. Intuitively, we say $l \bisim l'$ if they are equivalent

### 2024-01-18 paper-new/ordering.tex
changes to ordering.tex

@@
 sets equipped with an ordering that models term precision. Since term precision is reflexive
-and transitive, and since we identify terms that are equi-precise, we choose to model types
-as partially-ordered sets. We model the term precision ordering $M \ltdyn N : A \ltdyn B$ as an
-ordering relation between the posets denoted by $A$ and $B$.
+and transitive, and since we identify terms that are equi-precise, we could model types
+as partially-ordered sets and terms $\Gamma \vdash M : B$ as monotone functions.
+We could then model the term precision ordering $M \ltdyn N : A \ltdyn B$ as an
+ordering relation between the monotone functions denoted by $M$ and $N$.
 
-However, it turns out that modeling term precision by a relation defined by guarded fixpoint
-is not as straightforward as one might hope.
+However, it turns out that modeling term precision is not as straightforward as one might hope.
 A first attempt might be to define an ordering $\semltbad$ between $\li X$ and $\li Y$
@@
 transitivity, then that relation is trivial.
-(The other two properties are that the relation is a congruence with respect to $\theta$,
-and that the relation is closed under delays $\delta = \theta \circ \nxt$ on either side.)
+The other two properties are that the relation is a congruence with respect to $\theta$,
+and that the relation is closed under delays $\delta = \theta \circ \nxt$ on either side.
 Since the above relation \emph{does} satisfy the other two properties, we conclude
@@
 It turns out that we can, by sacrificing another of the three properties from
-the above lemma. Namely, we give up on closure under delays. Doing so, we end up
+the above result. Namely, we give up on closure under delays. Doing so, we end up
 with a \emph{lock-step} error ordering, where, roughly speaking, in order for
@@
 that are extensionally equal and may only differ in their stepping behavior.
+
+Finally, the semantics of term precision will be a combination of these two relations.
+% TODO Define it here?
 
@@
   The ordering on $\li A$ is the ``lock-step error-ordering'' which we describe in
-  \ref{subsec:lock-step}. The bismilarity relation is the ``weak bisimilarity''
+  \ref{sec:lock-step}. The bismilarity relation is the ``weak bisimilarity''
   described in Section \ref{}
@@
 
-\subsubsection{Step-Sensitive Error Ordering}\label{subsec:lock-step}
+\subsubsection{Lock-Step Error Ordering}\label{sec:lock-step}
 
 As mentioned, the ordering on the lift of a predomain $A$ is called the
-\emph{step-sensitive error-ordering} (also called ``lock-step error ordering''),
+\emph{lock-step error-ordering} (also called the ``step-sensitive error ordering''),
 the idea being that two computations $l$ and $l'$ are related if they are in
@@
 
-We also define a heterogeneous version of this ordering between the lifts of two
+We similarly define a heterogeneous version of this ordering between the lifts of two
 different predomains $A$ and $B$, parameterized by a relation $R$ between $A$ and $B$.
 
-\subsubsection{Weak Bisimilarity Relation}
+\subsubsection{Weak Bisimilarity Relation}\label{sec:weak-bisimilarity}
 
@@
 
+% To manage this construction, we break it down into multiple steps and do it modularly
+
+% We're not proving that the term model satisfies graduality 
+% (ex. the exponential in the term model includes all functions, whereas the
+% predomain model only includes monotone functions)
+
 \begin{comment}

### 2024-01-18 paper-new/technical-background.tex
changes to technical background

@@
 
-\subsection{Synthetic Guarded Domain Theory}
+\subsection{Synthetic Guarded Domain Theory}\label{sec:sgdt}
 One way to avoid the tedious reasoning associated with step-indexing is to work
@@
 
+Most of the developments in this paper will take place in the context of a single clock $k$,
+but in Section \ref{TODO}, we will need to make use of clock quantification.
+
 % Clocked Cubical Type Theory
@@
 
-The statements in this paper have been formalized in a variant of Agda called
-Guarded Cubical Agda \cite{veltri-vezzosi2020}, an implementation of Clocked Cubical Type Theory.
+% TODO mention Agda implementation of Clocked Cubical Type Theory?
+%The statements in this paper have been formalized in a variant of Agda called
+%Guarded Cubical Agda \cite{veltri-vezzosi2020}, an implementation of Clocked Cubical Type Theory.
 

### 2024-01-18 paper-new/categorical-model.tex
begin discussion of abstract model constructions

@@
-\section{Abstract Categorical Models of Graduality}
+\section{Abstract Categorical Models of Graduality}\label{sec:abstract-models}
 
@@
 
-Likewise, we have an analogous definition for computations.
+Likewise, we have the analogous definitions for computations.
 
 We write $c : A \rel A'$ to mean that $c \in \ob(\vsq)$ such that 
-$\sv(c) = A$ and $\tv(A) = A'$.
+$\sv(c) = A$ and $\tv(c) = A'$.
+%
 Likewise, let $c_i : A_i \rel A_i'$ and $c_o : A_o \rel A_o'$,
 and let $f \in \vf(A_i, A_o)$ and $f' \in \vf(A_i', A_o')$.
-The judgment $\beta : f \ltdyn_{c_o}^{c_i} f'$ is defined to mean
+The notation $\beta : f \ltdyn_{c_o}^{c_i} f'$ is defined to mean
 
@@
 
+% Note: When the identity of the square $\beta$ is not important, we may omit it
+% and write $f \ltdyn_{c_o}^{c_i} f'$. In this case the meaning is that there exists
+% a square $\beta : f \ltdyn_{c_o}^{c_i} f'$.
+
 Composition of squares $\beta : f \ltdyn_{c_2}^{c_1} g$ and $\beta' : f' \ltdyn_{c_3}^{c_2} g'$
@@
 That is, if $\beta, \beta' : f \ltdyn_{c_o}^{c_i} f'$, then $\beta = \beta'$.
+Thus, we may unambiguously omit the identity of the square, i.e., we may write
+$f \ltdyn_{c_o}^{c_i} f'$.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Intensional Models}\label{sec:abstract-intensional-models}
@@
 An intensional model of gradual typing is defined similarly to an extensional model,
-with two key differences that will be discussed below.
+with a few key differences that will be discussed below.
 %
@@
 for composition of computation relations/squares.
+Furthermore, $s \circ m = s \circ \pi_1$ and $t \circ m = t \circ \pi_2$.
 
@@
 
-For the sake of ease of reference, we recap the definition of a step-1 model:
+For the sake of ease of reference, we recap the definition of a step-0 model:
 
 \begin{definition}
-  A \emph{step-1} model of intensional gradual typing consists of a category internal to the
+  A \emph{step-0} model of intensional gradual typing consists of a category internal to the
   category of CBPV models.
@@
 
-\subsubsection{Bisimilarity}
+\subsubsection{Bisimilarity}\label{sec:abstract-model-bisimilarity}
 
@@
 Since bisimilarity of morphisms $f$ and $f'$ requires that they share source and target,
-we require that $\svsim$ and $\tvsim$ agree on objects and likewise for $\sesim and \tesim$.
+we require that $\svsim$ and $\tvsim$ agree on objects and likewise for $\sesim$ and $\tesim$.
 Thus, the objects of $\vsim$ are identified with $\ob(\vf)$.
@@
 % TODO explain what this means in terms of bisimilarity "squares"?
+In this setting, we write $\refl_A : A \bisim A$ to mean that $\refl_A \in \ob(\vsim)$,
+such that $\svsim(\refl_A) = A = \tvsim(\refl_A)$.
+Let $f, f' \in \vf(A_i, A_o)$.
+The judgment $\gamma : f \bisim_{A_i, A_o} f'$ is defined to mean:
+
+\begin{enumerate}
+  \item $\gamma \in \vsim(\refl_{A_i}, \refl_{A_o})$
+  \item $\svsim(\gamma) = f$
+  \item $\tvsim(\gamma) = f'$
+\end{enumerate}
 
@@
 
-\begin{definition}
-A \emph{step-2 intensional model} consists of all the data of a step-1 intensional model along
+\begin{definition}\label{def:step-1-model}
+A \emph{step-1 intensional model} consists of all the data of a step-0 intensional model along
 with the additional categories and functors for bisimiarity described above.
@@
 
-\subsubsection{Perturbations}
+\subsubsection{Perturbations}\label{sec:abstract-model-perturbations}
 
@@
 
-\begin{definition}
-  A \emph{step-3} model of intensional gradual typing consists of all the data of a step-2 model plus:
+\begin{definition}\label{def:step-2-model}
+  A \emph{step-2} model of intensional gradual typing consists of all the data of a step-1 model plus:
   \begin{enumerate}
@@
 
-Let $c : A \rel A'$. We say that $c$ is \emph{quasi-left-representable by}
+\begin{definition}\label{def:quasi-left-representable}
+Let $c : A \rel A'$ be a value relation. We say that $c$ is \emph{quasi-left-representable by}
 $f \in \vf(A, A')$ if there are perturbations $\delta_c^{l,e} \in \pv_A$ and
@@
 \end{center}
+\end{definition}
 
-Similarly, let $d : B \rel B'$. We say that $d$ is
+\begin{definition}\label{def:quasi-right-representable}
+Let $d : B \rel B'$ be a computation relation. We say that $d$ is
 \emph{quasi-right-representable by} $f \in \ef(B', B)$
@@
 \end{center}
+\end{definition}
 
@@
 
-\begin{definition}
-  A \emph{step-4 intensional model}, or simply an \emph{intensional model}, consists
-  of all the data of a step-3 intensional model, with the following additional data:
+\begin{definition}\label{def:step-3-model}
+  A \emph{step-3 intensional model} consists
+  of all the data of a step-2 intensional model, with the following additional data:
   \begin{enumerate}
@@
 \end{definition}
+
+\subsubsection{The Dynamic Type}
+
+Now we can discuss what it means for an intensional model to model the dynamic type.
+
+\begin{definition}\label{def:step-4-model}
+  A \emph{step-4 intensional model}, or simply an \emph{intensional model}, consists
+  of all the data of a step-3 intensional model, along with 
+  %a distinguished value object $D \in \ob(\vf)$ such that:
+  %
+  \begin{enumerate}
+    \item A distinguished value object $D \in \ob(\vf)$.
+    (By definition of a step-3 model, this means there is also a monoid $\pv_D$ of
+    perturbations and a homomorphism $\ptbv_D$.)
+    \item For each value type $A$, there is a value relation $\text{inj}_A : A \rel D$.
+    (By definition of a step-3 model, this relation satisfies the push-pull property and is
+    quasi-left-representable.)
+    \item If $c : A \rel A'$, then $\text{inj}_{A} = c \comp \text{inj}_{A'}$.
+  \end{enumerate}
+\end{definition}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Constructing an Extensional Model}\label{sec:extensional-model-construction}
+
+In the previous section, we have given the definition of an intensional model
+of gradual typing as a series of steps with each definition building on the previous one.
+%
+Here, we discuss how to construct an extensional model from an intensional model.
+We do so in several phases, beginning with a step-1 intensional model
+and ending with an extensional model.
+Moreover, this construction is \emph{modular}, in that each phase of the
+construction does not depend on the details of the previous ones.
+However, this process cannot proceed in isolation: some phases require
+additional inputs. We will make clear what data must be supplied to each phase.
+
+% Going from a step-1 intensional model $\mathcal{M}_1$ to a step-2 intensional model
+% requires defining a bisimilarity relation on the value morphisms and on
+% the computation morphisms of $\mathcal{M}_1$.
+% In particular, we need for every $A$ and $A'$ a reflexive, symmetric relation 
+% $\bisim_{A,A'}$ on $\vf(A, A')$ and for every $B$ and $B'$ a reflexive, symmetric relation
+% $\bisim_{B,B'}$ on $\ef(B, B')$. Moreover, the relation must respect composition of morphisms.
+% This phase of the construction is quite simple, because it consists entirely of extra structure
+% supplied as input to the construction.
+
+\subsubsection{Adding Perturbations}
+
+Suppose we have a \hyperref[def:step-1-model]{step-1 intensional model} $\mathcal{M}$.
+Recall that a step-1 intensional model consists of a step-0 model (i.e., a
+category internal to the category of CBPV models), along with the necessary
+categories and functors for bisimilarity as discussed in Section
+\ref{sec:abstract-model-bisimilarity}.
+Further, recall that a \hyperref[def:step-2-model]{step-2 model} has everything
+a step-1 model has, with the addition of perturbation monoids $\pv_A$ for all
+$A$ and $\pe_B$ for all $B$.
+% These perturbations must be interpretable as endomorphisms via homomorphisms
+% $\ptb_A$ and $\ptb_B$, 
+These perturbations must be interpretable via homomorphisms
+$\ptb_A$ and $\ptb_B$ as endomorphisms bisimilar to the identity,
+and moreover, the push-pull property must hold for all
+value relations $c$ and all computation relations $d$.
+
+% TODO do we really need all of this data?
+What ``external" data do we need to define such a model?
+Suppose we are given, for each value type $A$, a monoid $\pv_A$ with homomorphism
+$\ptb_A$ into the endomorphisms on $A$ bisimilar to the identity, and likewise, for
+each computation type $B$, a monoid $\pv_B$ and homomorphism $\ptb_B$.
+Furthermore, suppose we are given, for each value type $A$, a distinguished endomorphism
+$\delta_A \in \ef(FA, FA)$ such that $\delta_A \bisim \id_{FA}$.
+Finally, suppose that each value relation $c : A \rel A'$ satisfies the push-pull property
+and likewise for each computation relation $d : B \rel B'$.
+
+Then we claim that from the above data, we can construct a step-2 intensional model.
+For the proof, see Lemma \ref{lem:step-1-model-to-step-2-model} in the Appendix.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsubsection{Adding Quasi-Representability}
+
+Now suppose we have a step-2 intensional model $\mathcal{M}$.
+We will describe how to construct a \hyperref[def:step-3-model]{step-3 intensional model} $\mathcal{M'}$.
+We begin with a definition.
+
+\begin{definition}[representation structure]
+Let $c : A \rel A'$ be a value relation. A \emph{left-representation structure} $\rho^L_c$ for $c$ consists of
+a value morphism $u \in \vf(A, A')$ such that $c$ is quasi-left-representable by $u$ (see Definition \ref{def:quasi-left-representable}).
+
+Likewise, let $d : B \rel B'$. A \emph{right-representation structure} $\rho^R_d$ for $d$ consists of
+a computation morphism $d \in \ef(B', B)$ such that $d$ is quasi-right-representable by $d$ (see Definition \ref{def:quasi-right-representable}).
+\end{definition}
+
+(Notice that the direction of the morphism is opposite in the definition of
+right-representation structure.)
+
+We now describe the construction of a step-3 model $\mathcal M'$.
+The objects of $\mathcal M'$ are defined to be the same as the objects of the
+step-2 model $\mathcal M$.
+The value and computation morphisms in $\mathcal M'$ are the same as those of
+$\mathcal M$.
+A value relation is defined to be a pair $(c, \rho^L_c)$ with $c$ a value
+relation in $\mathcal M$ and $\rho^L_c$ a left-representation structure for $c$.
+Likewise, a computation relation is defined to be a pair $(d, \rho^R_d)$ with
+$d$ a computation relation in $\mathcal M$ and $\rho^R_d$ a right-representation
+structure for $d$.
+Lastly, morphisms of value relations (i.e., value squares) are defined by simply
+ignoring the representation structure. That is, a morphism of value relations
+$\alpha \in \vsq'((c, \rho^L_c), (c' \rho^L_{c'}))$ is simply a morphism of value
+relations in $\vsq(c, c')$. Likewise for computations.
+
+Now we define the functors $F$, $U$, and $\arr$.
+On objects, the behavior is the same as the respective functors in $\mathcal M$.
+For relations, we define 
+$\Fsq' (c, \rho^L_c) = (\Fsq c, \rho^R_{F c})$ and
+$\Usq' (d, \rho^R_d) = (\Usq d, \rho^L_{U d})$.
+We define $(c, \rho^L_c) \arr (d, \rho^R_d) = (c \arr d, \rho^R_{c \arr d})$.
+
+We now verify that the construction meets the requirements of a step-3 model.
+First, we check that composition of value relations (resp. computation relations)
+is well-defined. This might seem obvious, but there is actually a subtlety here:
+what should the representation structure be for the composition?
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsubsection{Defining an Extensional Model}
+
+Finally, suppose $\mathcal M$ is a step-3 intensional model.
+We now describe how to build an extensional model.
+
+The idea is to define an extensional model whose squares are the ``bisimilarity-closure''
+of the squares of the provided intensional model $\mathcal M$.
+
+The categories $\vf$, $\ef$ are the same as those of $\mathcal M$.
+Additionally, the objects of $\vsq$ and $\esq$, i.e., the value and
+computation relations, are the same.
+The difference arises in the \emph{morphisms} of $\vsq$ and $\esq$,
+i.e., the commuting squares. In particular, a morphism
+$\alpha_e \in \vsq'(c_i, c_o)$ with source $f$ and target $g$ is given by:
+\begin{itemize}
+  \item a morphism $f' \in \vf(A_i, A_o)$ with $f \bisim f'$.
+  \item a morphism $g' \in \vf(A_i', A_o')$ with $g \bisim g'$.
+  \item a square $\alpha_i \in \vsq(c_i, c_o)$ with source $f'$ and target $g'$.
+\end{itemize}
+
+Using our existing notation, we say that $f \le_{c_o}^{c_i} g$ if there exist $f'$ and $g'$ such that
+
+\[ f \bisim_{A_i,A_o} f' \ltdyn_{c_o}^{c_i} g' \bisim_{A_i',A_o'} g. \]
+
+We make the analogous construction for the computation squares.
+
+Next, we check that the requirements of an extensional model are satisfied.
+In particular, we need to verify the representability properties.
+We define functors $\upf$ and $\dnf$

### 2024-01-18 paper-new/concrete-model.tex
begin section on concrete model

\section{Constructing a Concrete Model}

In this section, we build a concrete extensional model of gradual typing.
We begin by defining a step-1 intensional model, and then
apply the abstract constructions outlined in the previous section
to obtain an extensional model.

We begin with some definitions. These concepts were first introduced in the section
on the term semantics (Section \ref{sec:domain-theory}), but for completeness we repeat them here.
% for ease of reference when instantiating the definition of step-1 model.

\begin{definition}
A \textbf{predomain} $A$ consists of a set $A$ along with two relations:
\begin{itemize}
    \item A partial order $\le_A$.
    \item A reflexive, symmetric ``bisimilarity'' relation $\bisim_A$.
\end{itemize}
\end{definition}

% TODO: later of a predomain
Given a predomain $A$, we can form the predomain $\later A$.
The underlying set is $\later |A|$ and the relation is defined in the obvious way,
i.e., $\tilde{x} \le_{\later A} \tilde{x'}$ iff $\later_t(\tilde{x}_t \le_A \tilde{x'}_t)$.

Morphisms of predomains are functions between the underlying sets that preserve the ordering
and the bisimilarity relation. More formally:
%
\begin{definition}
Let $A$ and $A'$ be predomains.
A morphism $f : A \to A'$ is a function between the underlying sets such that for all $x, x'$,
if $x \le_A x'$, then $f(x) \le f(x')$, and if $x \bisim_A x'$, then $f(x) \bisim_{A'} f(x')$.
\end{definition}

\begin{definition}
An \textbf{error domain} $B$ consists of a predomain $B$ along with the following data:
\begin{itemize}
    \item A distinguished ``error" element $\mho_B \in B$
    \item A morphism of predomains $\theta_B \colon \later B \to B$
\end{itemize}
\end{definition}

Morphisms of error domains are morphisms of the underlying predomains that preserve the
algebraic structure. More formally:
%
\begin{definition}
Let $B$ and $B'$ be error domains.
A morphism $\phi : B \wand B'$ is a morphism between the underlying predomains such that:
\begin{enumerate}
    \item $\phi(\mho_B) = \mho_{B'}$
    \item $\phi(\theta_B(\tilde{x})) = \theta_{B'}(\lambda t. \phi(\tilde{x}_t))$
\end{enumerate}
\end{definition}

We define a (monotone) relation on predomains $A$ and $A'$ to be a relation on the
underlying sets that is downward-closed under $\le_A$ and upward-closed under $\le_{A'}$.
More formally:

\begin{definition}
Let $A$ and $A'$ be predomains. A \emph{predomain relation} between $A$ and $A'$
is a relation $R$ between the underlying sets such that:
\begin{enumerate}
    \item (Downward closure): For all $x_1, x_2 \in A$ and $y \in A'$,
    if $x_1 \le_A x_2$ and $x_2 \mathbin{R} y$, then $x_1 \mathbin{R} y$.
    \item (Upward closure): For all $x \in A$ and $y_1, y_2 \in A'$,
    if $x \mathbin{R} y_1$ and $y_1 \le_{A'} y_2$, then $x \mathbin{R} y_2$.
\end{enumerate}
\end{definition}

Composition of relations on predomains is the usual relational composition.

Similarly, we define a (monotone) relation on error domains to be a relation on the
underlying predomains that respects error and preserves $\theta$.
%
\begin{definition}
    Let $B$ and $B'$ be error domains. An \emph{error domain relation} between
    $B$ and $B'$ is a relation $R$ between the underlying predomains such that
    \begin{enumerate}
       \item (Respects error): For all $y \in B'$, we have $\mho_B \mathbin{R} y$.
       \item (Preserves $\theta$): For all $\tilde{x}$ in $\later B$ and $\tilde{y} \in \later B'$,
       if 
       \[ \later_t( \tilde{x}_t \mathbin{R} \tilde{y}_t ), \]
       then
       \[ \theta_B(\tilde{x}) \mathbin{R} \theta_{B'}(\tilde{y}). \]
    \end{enumerate}
\end{definition}

We define composition of error domain relations $R$ on $B_1$ and $B_2$ and $S$
on $B_2$ and $B_3$ to be the least relation containing $R$ and $S$ that respects
error and preserves $\theta$.
Specifically, it is defined inductively by the following rules:

\begin{mathpar}
    \inferrule*[right = Comp]
    {b_1 \mathbin{R} b_2 \and b_2 \mathbin{S} b_3}
    {b_1 \mathbin{R \relcomp S} b_3}

    \inferrule*[right = PresErr]
    { }
    {\mho_{B_1} \mathbin{R \relcomp S} b_3}

    \inferrule*[right = PresTheta]
    {\later_t( \tilde{b_1} \mathbin{R \relcomp S} \tilde{b_3} ) }
    {\theta_{B_1}(\tilde{b_1}) \mathbin{R \relcomp S} \theta_{B_3}(\tilde{b_3}) }
\end{mathpar}

% We note that this composition has the following universal property.

We now describe the ``commuting squares".
Suppose we are given predomains $A_i, A_o, A_i'$, and $A_o'$,
relations $R_i$ and $R_o$, and morphisms $f, f'$ as shown below.

% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBX2kiXSxbMSwwLCJBX2knIl0sWzAsMSwiQV9vIl0sWzEsMSwiQV9vJyJdLFswLDIsImYiLDJdLFsxLDMsImYnIl0sWzAsMSwiUl9pIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMiwzLCJSX28iLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs0LDUsIlxcbHRkeW4iLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
\[\begin{tikzcd}[ampersand replacement=\&]
	{A_i} \& {A_i'} \\
	{A_o} \& {A_o'}
	\arrow[""{name=0, anchor=center, inner sep=0}, "f"', from=1-1, to=2-1]
	\arrow[""{name=1, anchor=center, inner sep=0}, "{f'}", from=1-2, to=2-2]
	\arrow["{R_i}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow["{R_o}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
	\arrow["\ltdyn"{description}, draw=none, from=0, to=1]
\end{tikzcd}\]

We say that the above square commutes, written $f \le f'$, if for all
$x \in A_i$ and $x' \in A_i'$ with $x \mathbin{R_i} x'$, we have
$f(x) \mathbin{R_o} f'(x')$.

We make the analogous definition for error domains.

% Lift monad
The guarded lift monad $\li$ takes a predomain $A$ to the error domain $\li A$;
the precise definition is given in Section \ref{sec:lift-monad}.
The partial order $\le_{\li A}$ is the lock-step error ordering defined in Section
\ref{sec:lock-step}, and the bisimilarity relation $\bisim_{\li A}$ is the weak bisimilarity
relation defined in Section \ref{sec:weak-bisimilarity}.
We define the action of $\li$ on a relation $R$ between $A$ and $A'$ to be the
``heterogeneous" version of the lock-step error lifting of $R$ (see Section \ref{sec:lift-monad}).

% TODO action of \li on commuting squares

There is a functor $U$ from error domains to predomains that on objects simply takes the
underlying predomain, and on morphisms takes the underlying morphism of predomains.
%
% TODO check this
One can verify that there is a free-forgetful adjunction between the functors $\li$ and $U$;
this holds because $\li A$ is the initial/free error- and later-algebra on the predomain $A$,
i.e., error domain morphisms from $\li A$ to $B$ are in one-to-one correspondence with
predomain morphisms from $A$ to $UB$.

% internal hom for predomains and error domains
Given predomains $A$ and $A'$, we can form the predomain of
predomain morphisms from $A$ to $A'$, denoted $A \To A'$.
\begin{itemize}
    % Should we give the definition involving x and x'?
    \item The ordering is defined by $f \le_{A \To A'} f'$ iff for all
    $x \in A$, we have $f(x) \le_{A'} f'(x)$.
    \item The bisimilarity relation is defined by $f \bisim_{A \To A'} f'$ iff
    for all $x, x' \in A$ with $x \bisim_{A} x'$, we have $f(x) \bisim_{A'} f'(x')$. 
\end{itemize}

Given a predomain $A$ and error domain $B$, we define $A \arr B := A \To UB$.
We note that $A \arr B$ carries a natural error domain structure
(in the below, the lambda is a meta-theoretic notation):
\begin{itemize}
    \item The error is given by $\lambda x . \mho_B$
    \item The $\theta$ operation is defined by
      \[ \theta_{A \arr B}(\tilde{f}) = \lambda x . \theta_B(\lambda t . \tilde{f}_t(x)). \]
\end{itemize}

It is easily verified that $A \arr B$ is an exponential of $UB$ by $A$
in the category of predomains and their morphisms.

Lastly, given a relation of predomains $R$ between $A$ and $A'$, and a relation
of error domains $S$ between $B$ and $B'$, we can form a relation $R \arr S$
between $A \arr B$ and $A' \arr B'$ in the obvious way, i.e., $f \in A \arr B$
is related to $g \in A' \arr B'$ iff for all $x \in A$ and $x' \in A'$ with
$x \mathrel{R} x'$, we have $f(x) \mathrel{S} g(x')$.

With all of the above data, we can form a step-1 intensional model of gradual typing
(See Definition \ref{def:step-1-model}).

\subsection{The Dynamic Type}

% dynamic type
The underlying set of the dynamic type will be defined using guarded recursion as the
solution to the equation

\[ D \cong \mathbb{N} + \later (D \to \li D). \]

The perturbation monoid for $D$ is defined inductively as the free monoid $M$ equipped
with an operation $\To : M \times M \to M$, i.e., we take the set defined inductively by

\begin{mathpar}
    \inferrule*[]
    {}
    {e \in M}

    \inferrule*[]
    {p \in M \and p' \in M}
    {p \odot p' \in M}

    \inferrule*[]
    {p \in M \and p' \in M}
    {p \To p' \in M}
\end{mathpar}

and quotient by the necessary monoid equations.

% perturbations and quasi-representability

\subsection{Obtaining an Extensional Model}

Now that we have defined an intensional model, we can apply
the abstract construction introduced in Section \ref{sec:extensional-model-construction}.
Doing so, we obtain an extensional model of gradual typing.

### 2024-01-18 paper-new/discussion.tex
begin discussion section

--- 2022-07-27 paper.tex
@@
-\documentclass[acmsmall,screen]{acmart}
-\usepackage{mathpartir}
-\usepackage{tikz-cd}
+\section{Discussion}
 
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2018}
-\acmYear{2018}
-\acmDOI{10.1145/1122445.1122456}
+\subsection{Related Work}
+% Discuss Joey Eremondi's thesis on gradual dependent types
 
-%% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
-  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
-\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-  June 03--05, 2018, Woodstock, NY}
-\acmPrice{15.00}
-\acmISBN{978-1-4503-XXXX-X/18/06}
+\subsection{Mechanization}
+% Discuss Guarded Cubical Agda and mechanization efforts
 
-\begin{document}
-
-\title{Denotational Semantics for Gradual Typing in Synthetic Guarded Domain Theory}
-\author{Eric Giovannini}
-\author{Max S. New}
-
-\begin{abstract}
-  We develop a denotational semantics for a gradually typed language
-  with effects that is adequate and proves the graduality theorem.
-  %
-  The denotational semantics is constructed using \emph{synthetic
-  guarded domain theory} working in a type theory with a later
-  modality and clock quantification.
-  %
-  This provides a remarkably simple presentation of the semantics,
-  where gradual types are interpreted as ordinary types in our ambient
-  type theory equipped with an ordinary preorder structure to model
-  the error ordering.
-  %
-  This avoids the complexities of classical domain-theoretic models
-  (New and Licata) or logical relations models using explicit
-  step-indexing (New and Ahmed).
-  %
-  In particular, we avoid a major technical complexity of New and
-  Ahmed that require two logical relations to prove the graduality
-  theorem.
-
-  By working synthetically we can treat the domains in which gradual
-  types are interpreted as if they were ordinary sets. This allows us
-  to give a ``na\"ive'' presentation of gradual typing where each
-  gradual type is modeled as a well-behaved subset of the universal
-  domain used to model the dynamic type, and type precision is modeled
-  as simply a subset relation.
-  %
-  This provides a rich semantic setting in which to 
-\end{abstract}
-
-\section{Introduction}
-
-% gradual typing, graduality
-
-% synthetic guarded domain theory, denotational semantics therein
-
-% Difficulties in prior semantics
-
-\section{Discussion}
+\subsection{Benefits and Drawbacks of the Synthetic Approach}
 
@@
 
-\end{document}
+\subsection{Future Work}
+
+% Cite GrEff paper

### 2024-01-18 paper-new/appendix.tex
begin appendix

\section{Details of the Construction of an Extensional Model}

In Section \ref{sec:extensional-model-construction}, we outline the construction
of an extensional model of gradual typing starting from a step-1 intensional model.
In this section, we provide the details for each of the constructions mentioned there.

\begin{lemma}\label{lem:step-1-model-to-step-2-model}
Let $\mathcal M$ be a \hyperref[def:step-1-model]{step-1 intensional model}.
Suppose we are given the following data:

\begin{enumerate}
    \item For each value type $A$, a monoid $\pv_A$ and homomorphism 
    \[ \ptbv_A : \pv_A \to \{ f \in \vf(A,A) \mid f \bisim \id \} \]

    \item For each computation type $B$, a monoid $\pv_B$ and homomorphism
    \[ \ptbe_B : \pe_B \to \{ g \in \ef(B,B) \mid g \bisim \id \} \]

    \item For each value type $A$, a distinguished endomorphism
    $\delta_A \in \ef(FA, FA)$ such that $\delta_A \bisim \id_{FA}$.
\end{enumerate}

Then we can construct a \hyperref[def:step-2-model]{step-2 intensional model}.
\end{lemma}
\begin{proof}
    Write 
    %
    \[ \mathcal M = (\vf, \vsq, \ef, \esq, \Ff, \Fsq, \Uf, \Usq, \arrf, \arrsq). \] 
    %
    Define a step-2 model as follows:
    \begin{itemize}
      \item Value objects are tuples of an object $A$ in $\vf$ along with the monoid
      $\pv_A$ and homomorphism $\ptbv_A$:
      $\ob(\vf') = \{ (A, \pv_A, \ptbv_A) \mid A \in \ob(\vf) \}$.

      \item Morphisms are given by morphisms of the underlying objects in $\vf$, i.e.,
       $\vf'((A, \pv_A, \ptbv_A), (A', \pv_{A'}, \ptbv_{A'})) = \vf(A, A')$.

      \item Computation objects are tuples 
      $\ob(\ef') = \{ (B, \pe_B, \ptbe_B) \mid B \in \ob(\ef) \}$.

      \item Computation morphisms are $\ef'((B, \pv_B, \ptbv_B), (B', \pv_{B'}, \ptbv_{B'})) = \ef(B, B')$.

      \item The objects $\vsq'$ and $\esq'$ are the same as those of $\vsq$ and $\esq$.

      \item The morphisms of $\vsq'$ and $\esq'$ are the same as those of $\vsq$ and $\esq$.
    %   \item $\ob(\vsq') = \ob(\vsq)$
    %   \item $\ob(\esq') = \ob(\esq)$
    %   \item $\vsq'(c, c') = \vsq(c, c')$
    %   \item $\esq'(d, d') = \esq(d, d')$

      % Functors \times, +, F, U, arrow
      \item We define $F$ on objects by $F (A, \pv_A, \ptbv_A) = (FA, (1 + \pv_A), h_F)$
      where $1$ is the trivial monoid, $+$ is the coproduct in the category of monoids, and $h_F$ is the homomorphism defined as follows:

      \item We define $U$ on objects by $U (B, \pe_B, \ptbe_B) = (UB, \pe_B, h_U)$
      where $h_U(p_B) = U(\ptbe_B(p_B))$.

      \item We define $(A, \pv_A, \ptbv_A) \arr (B, \pe_B, \ptbe_B) = (A \arr B, \pv_A \times \pe_B, h_\arr)$
      where $\times$ is the product in the category of monoids, and $h_\arr$ is defined by 
      $h_\arr(p_A, p_B) = \ptbv_A(p_A) \arr \ptbe_B(p_B)$.
    \end{itemize}
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{lemma}\label{lem:step-2-model-to-step-3-model}
    Let $\mathcal M$ be a \hyperref[def:step-2-model]{step-2 intensional model}.
    Suppose we are given the following data:

    Then we can construct a \hyperref[def:step-3-model]{step-3 intensional model}.
\end{lemma}
\begin{proof}

\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{lemma}\label{lem:step-4-model-to-extensional-model}
  Let $\mathcal M$ be a \hyperref[def:step-4-model]{step-4 intensional model}.
  Then we can define an extensional model.
\end{lemma}
\begin{proof}

  % More formally, we define an extensional model $\mathcal M_e$ as follows.
  % \begin{itemize}
  %   \item 
  % \end{itemize}
\end{proof}

\section{Adequacy}\label{sec:appendix-adequacy}

In this section, we show an adequacy result for the extensional model of GTT we obtained by
applying the abstract construction introduced in Section
\ref{sec:extensional-model-construction} to the concrete model

First we establish some notation. Fix a morphism $f : 1 \to \li \Nat \cong \li \Nat$.
We write that $f \da n$ to mean that there exists $m$ such that $f = \delta^m(\eta n)$
and $f \da \mho$ to mean that there exists $m$ such that $f = \delta^m(\mho)$.

Recall that $\ltls$ denotes the relation on value morphisms defined as the bisimilarity-closure
of the intensional error-ordering on morphisms.
More concretely, we have $f \ltls g$ iff there exists $f'$ and $g'$ with

\[ f \bisim f' \le g' \bisim g. \]

The result we would like to show is as follows:
\begin{lemma}
If $f \ltls g : \li \Nat$, then:
\begin{itemize}
  \item If $f \da n$ then $g \da n$.
  \item If $g \da \mho$ then $f \da \mho$.
  \item If $g \da n$ then $f \da n$.
\end{itemize}
\end{lemma}

Unfortunately, this result is actually not provable!
Roughly speaking, the issue is that this is a ``global'' result, and it is not possible
to prove such results inside of the guarded setting. 
In particular, if we tried to prove a result such as the above in the guarded setting,
we would run into a problem where we would have a natural number ``stuck'' under a $\later$
with no way to get at the underlying number.

Thus, to prove our adequacy result, we need to leave the guarded setting and pass back
to the normal set-theoretic world.
As mentioned in the Technical Background section (Section \ref{sec:sgdt}), we can do this
using \emph{clock quantification}.

Recall that all of the constructions we have made in SGDT take place in the context of a clock $k$.
All of our uses of the later modality and guarded recursion happen with respect to this clock.
For example, consider the definition of the lift monad by guarded recursion in Section \ref{TODO}.
% We define the lift monad $\li^k X$ as the guarded fixpoint of $\lambda \tilde{T}. X + 1 + \later^k_t (\tilde{T}_t)$.
We can view this definition as being parameterized by a clock $k$: $\li^k : \type \to \type$.
Then for $X$ satisfying a certain technical requirement, we can define the ``global lift'' monad as $\li^{gl} X = \forall k. \li^k X$.

It can be shown that the global lift monad is isomorphic to the so-called Delay monad of Capretta \cite{TODO}.

% We have been writing the type as $\li X$, but it is perhaps more accurate to write it as $\li^k X$ to
% emphasize that the construction is parameterized by a clock $k$.

### 2024-01-19 paper-new/intro.tex
add a citation in the intro
some intro work and add macros for comments

@@
 their application code without needing to commit to fixed type signatures for their
-functions. Unfortunately, most languages choose between static or dynamic typing.
-As a result, programmers that initially write their code in a dynamically typed language
-in order to benefit from faster prototyping and development time will need to rewrite the
-entire codebase in a static language if they would like to receive the benefits of static
+functions. Most languages choose between static or dynamic typing and as a result, programmers that initially write their code in a dynamically typed language
+in order to benefit from faster prototyping and development time need to rewrite the
+some or all of their codebase in a static language if they would like to receive the benefits of static
 typing once their codebase has matured.
@@
 typing annotations and provide them later without altering the meaning of the program.
-
 % Sound gradual typing
@@
 % Graduality property
-Formally speaking, gradually typed languages should satisfy the 
+One of the fundamental theorems for gradually typed languages is 
 \emph{dynamic gradual guarantee}, originally defined by Siek, Vitousek, Cimini,
-and Boyland \cite{siek_et_al:LIPIcs:2015:5031}.
-This property is also referred to as \emph{graduality} \cite{new-ahmed2018},
+and Boyland \cite{siek_et_al:LIPIcs:2015:5031}, also called \emph{graduality} \cite{new-ahmed2018},
 by analogy with parametricity.
-Intuitively, graduality says that going from a dynamic to static style should not
-introduce changes in the meaning of the program.
-More specifically, making the types more precise by adding typing annotations
-will either result in the same behavior as the original, less precise program,
-or will result in a type error.
+Informally, graduality says that going from a dynamic to static style should only allow for the introduction of static or dynamic type errors, and not otherwise change the meaning of the program.
+%
+This is a way to capture programmer intuition that increasing type
+precision corresponds to a generalized form assertions at runtime
+only, and so they can trust that prior observed behavior of their
+dynamically typed code will remain unchanged as long as it satisfies
+the new stricter typing discipline.
+
+Additionally, gradually typed languages should offer some of the
+benefits of static typing. While classical type soundness, that
+well-typed programs are free from runtime errors, is not compatible
+with runtime type errors, we can instead focus on \emph{type-based
+reasoning}. For instance, while dynamically typed $\lambda$ calculi
+only satisfy $\beta$ equality for their type formers, statically typed
+$\lambda$ calculi additionally satisfy type-dependent $\eta$
+properties that ensure that functions are determined by their behavior
+under application and that pattern matching on data types
+is safe and exhaustive.
+
+Our goal in this work is to provide a reusable semantic framework for
+gradually typed languages that can be used to prove the aforementioned
+properties: graduality and type-based reasoning. \max{TODO: say more here}
 
@@
 An alternative approach, which we investigate in this paper, is provided by
-\emph{synthetic guarded domain theory}.
+\emph{synthetic guarded domain theory}\cite{birkedal-mogelberg-schwinghammer-stovring2011}.
 The techniques of synthetic guarded domain theory allow us to internalize the

### 2024-01-19 paper-new/appendix.tex
separately functorial actions of CBPV type constructors on Kleisli arrows

@@
+\section{Call-by-push-value}
+
+In CBPV models, all the type constructors are interpreted as functors:
+\begin{enumerate}
+\item $\to : \op\calV \times \calE \to \calE$
+\item $\times : \calV \times \calV \to \calV$
+\item $F : \calV \to \calE$
+\item $U : \calE \to \calV$
+\end{enumerate}
+That is, they all have functorial actions on \emph{pure} morphisms of
+value types and \emph{linear} morphisms of computation types.
+%
+We use these functorial actions extensively in the construction of
+casts and their corresponding perturbations. But when defining
+downcasts of value types and upcasts of computation types, we
+additionally need a second functorial action of these categories:
+functoriality in \emph{impure} morphisms of value types and
+\emph{non-linear} morphisms of computation types. These notions of
+morphism are given by the \emph{Kleisli} categories $\calVk$ and
+$\calEk$ which have value types and computation types as objects but
+morphisms are defined as
+\[ \calVk(A,A') = \calE(F A, FA')\]
+\[ \calEk(B,B') = \calV(U B, U B')\]
+with composition given by composition in $\calE/\calV$.  That is we
+need to define a second functorial action, that agrees with the above
+on objects for these Kleisli categories:
+\begin{enumerate}
+\item $\tok : \op\calVk \square \calEk \to \calEk$
+\item $\timesk : \calVk \square \calVk \to \calVk$
+\item $\Fk : \calVk \to \calEk$
+\item $\Uk : \calEk \to \calVk$
+\end{enumerate}
+Note that rather than the product of categories we use the ``funny
+tensor product'' $\square$. This is because the action on
+impure/non-linear morphisms for $\tok/\timesk$ do not satisfy ``joint
+functoriality'' but instead only ``separate functoriality'', meaning
+we give rather than an action on morphisms in both categories
+simultaneously instead an action on each argument categories morphisms
+with the object in the other category fixed. The existence of these
+functorial actions for $\tok$ and $\timesk$ is reliant on the
+\emph{strength} of the adjunction. We describe them using the internal
+language of CBPV in order to more easily verify their
+existence/functoriality:
+\begin{enumerate}
+\item For $\tok$ we define for $l : \calE(F A,F A')$ and $B \in \calE$ the morphism $l \tok B : \calV(U(A' \to B),U(A\to B))$ as
+  \[ t:U(A'\to B) \vdash l \tok B = \{ \lambda x. x' \leftarrow l\,[\ret x]; ! t x'\} : U(A \to B) \]
+  and for $A \in \calV$ and $f : \calV(UB,UB')$ we define $A \tok f : \calV(U(A \to B),U(A\to B'))$ as
+  \[ t : U(A \to B) \vdash A \tok f = \{ \lambda x. !f[\{ ! t x \}]\} \]
+\item For $\timesk$ we define for $l : \calE(F A_1,FA_2)$ and $A' \in \calV$ the morphism $l \timesk A_2$ as
+  \[ \bullet : F(A_1\times A_2) \vdash l \timesk A_2 = (x_1,x_2) \leftarrow \bullet; x_1' \leftarrow l[\ret x_1]; \ret (x_1',x_2) : F(A_1'\times A_2)\]
+  and $A_1 \timesk l$ is defined symmetrically.
+\item For $\Uk$ we need to define for $f : \calV(UB,UB')$ a morphism $\Uk f : \calE(FUB,FUB')$. This is simply given by the functorial action of $F$: $\Uk f = F(f)$
+\item Similarly $\Fk l = Ul$
+\end{enumerate}
+
+Functoriality in each argument is easily established, meaning for
+example for the function type is functorial in each argument:
+\begin{enumerate}
+\item $(l \circ l') \tok B = (l' \tok B) \circ (l \tok B)$
+\item $\id \tok B = \id$
+\item $A \tok (f \circ f') = (A \tok f) \circ (A \tok f)$
+\item $A \tok \id = \id$
+\end{enumerate}
+
 \section{Details of the Construction of an Extensional Model}

### 2024-01-19 paper-new/categorical-model.tex
update definition of abstract model

@@
 
+Lastly, we require that for any value object $A$, the hom-set $\ef(FA, FA)$ contains a
+distinguished morphism $\delta_A$, such that $\delta_A \bisim_{FA, FA} \id_{FA}$.
+
 % Spelling this all out concretely, for any pair...
@@
 A \emph{step-1 intensional model} consists of all the data of a step-0 intensional model along
-with the additional categories and functors for bisimiarity described above.
+with the additional categories and functors for bisimiarity described above, as  well as
+the existence of a distinguised computation morphism $\delta_A \bisim \id_{FA}$ for each $A$.
 \end{definition}
@@
 to an element of the perturbation monoid or the corresponding morphism.
+
+We require that $\delta_A \in \pe_{FA}$ for all $A$, where $\delta_A$ is the distinguished
+morphism that is required to be present in every hom-set $\ef(FA, FA)$ per the definition
+of a step-1 model.
 
@@
 Now we can discuss what it means for an intensional model to model the dynamic type.
+This applies to any of the above abstract model definitions, i.e., steps 0-3.
 
 \begin{definition}\label{def:step-4-model}
-  A \emph{step-4 intensional model}, or simply an \emph{intensional model}, consists
-  of all the data of a step-3 intensional model, along with 
+  A \emph{step-$i$ intensional model with dyn} is a step-$i$ model $\mathcal M$ such that:
   %a distinguished value object $D \in \ob(\vf)$ such that:
@@
   \begin{enumerate}
-    \item A distinguished value object $D \in \ob(\vf)$.
-    (By definition of a step-3 model, this means there is also a monoid $\pv_D$ of
-    perturbations and a homomorphism $\ptbv_D$.)
+    \item There is a distinguished value object $D \in \ob(\vf)$.
+
     \item For each value type $A$, there is a value relation $\text{inj}_A : A \rel D$.
-    (By definition of a step-3 model, this relation satisfies the push-pull property and is
-    quasi-left-representable.)
+
     \item If $c : A \rel A'$, then $\text{inj}_{A} = c \comp \text{inj}_{A'}$.
@@
 \end{definition}
+
+% (By definition of a step-3 model, this relation satisfies the push-pull property and is
+% quasi-left-representable.)
+
+% (By definition of a step-3 model, this means there is also a monoid $\pv_D$ of
+% perturbations and a homomorphism $\ptbv_D$.)
 
@@
 %
-Here, we discuss how to construct an extensional model from an intensional model.
-We do so in several phases, beginning with a step-1 intensional model
+Here, we discuss how to construct an extensional model from an intensional model with dyn.
+We do so in several phases, beginning with a step-1 intensional model with dyn
 and ending with an extensional model.
@@
 construction does not depend on the details of the previous ones.
-However, this process cannot proceed in isolation: some phases require
-additional inputs. We will make clear what data must be supplied to each phase.
-
-% Going from a step-1 intensional model $\mathcal{M}_1$ to a step-2 intensional model
-% requires defining a bisimilarity relation on the value morphisms and on
-% the computation morphisms of $\mathcal{M}_1$.
-% In particular, we need for every $A$ and $A'$ a reflexive, symmetric relation 
-% $\bisim_{A,A'}$ on $\vf(A, A')$ and for every $B$ and $B'$ a reflexive, symmetric relation
-% $\bisim_{B,B'}$ on $\ef(B, B')$. Moreover, the relation must respect composition of morphisms.
-% This phase of the construction is quite simple, because it consists entirely of extra structure
-% supplied as input to the construction.
+% However, this process cannot proceed in isolation: some phases require
+% additional inputs. We will make clear what data must be supplied to each phase.
 
@@
 
-Suppose we have a \hyperref[def:step-1-model]{step-1 intensional model} $\mathcal{M}$.
+Suppose we have a \hyperref[def:step-1-model]{step-1 intensional model} $\mathcal{M}$ with dyn.
 Recall that a step-1 intensional model consists of a step-0 model (i.e., a
@@
 $A$ and $\pe_B$ for all $B$.
-% These perturbations must be interpretable as endomorphisms via homomorphisms
-% $\ptb_A$ and $\ptb_B$, 
 These perturbations must be interpretable via homomorphisms
@@
 We will describe how to construct a \hyperref[def:step-3-model]{step-3 intensional model} $\mathcal{M'}$.
-We begin with a definition.
-
-\begin{definition}[representation structure]
-Let $c : A \rel A'$ be a value relation. A \emph{left-representation structure} $\rho^L_c$ for $c$ consists of
-a value morphism $u \in \vf(A, A')$ such that $c$ is quasi-left-representable by $u$ (see Definition \ref{def:quasi-left-representable}).
-
-Likewise, let $d : B \rel B'$. A \emph{right-representation structure} $\rho^R_d$ for $d$ consists of
-a computation morphism $d \in \ef(B', B)$ such that $d$ is quasi-right-representable by $d$ (see Definition \ref{def:quasi-right-representable}).
-\end{definition}
-
-(Notice that the direction of the morphism is opposite in the definition of
-right-representation structure.)
 
 We now describe the construction of a step-3 model $\mathcal M'$.
-The objects of $\mathcal M'$ are defined to be the same as the objects of the
-step-2 model $\mathcal M$.
-The value and computation morphisms in $\mathcal M'$ are the same as those of
-$\mathcal M$.
-A value relation is defined to be a pair $(c, \rho^L_c)$ with $c$ a value
-relation in $\mathcal M$ and $\rho^L_c$ a left-representation structure for $c$.
-Likewise, a computation relation is defined to be a pair $(d, \rho^R_d)$ with
-$d$ a computation relation in $\mathcal M$ and $\rho^R_d$ a right-representation
-structure for $d$.
-Lastly, morphisms of value relations (i.e., value squares) are defined by simply
-ignoring the representation structure. That is, a morphism of value relations
-$\alpha \in \vsq'((c, \rho^L_c), (c' \rho^L_{c'}))$ is simply a morphism of value
-relations in $\vsq(c, c')$. Likewise for computations.
-
-Now we define the functors $F$, $U$, and $\arr$.
-On objects, the behavior is the same as the respective functors in $\mathcal M$.
-For relations, we define 
-$\Fsq' (c, \rho^L_c) = (\Fsq c, \rho^R_{F c})$ and
-$\Usq' (d, \rho^R_d) = (\Usq d, \rho^L_{U d})$.
-We define $(c, \rho^L_c) \arr (d, \rho^R_d) = (c \arr d, \rho^R_{c \arr d})$.
-
-We now verify that the construction meets the requirements of a step-3 model.
-First, we check that composition of value relations (resp. computation relations)
-is well-defined. This might seem obvious, but there is actually a subtlety here:
-what should the representation structure be for the composition?
 

### 2024-01-19 paper-new/appendix.tex
work on abstract model constructions

@@
 \begin{lemma}\label{lem:step-1-model-to-step-2-model}
-Let $\mathcal M$ be a \hyperref[def:step-1-model]{step-1 intensional model}.
-Suppose we are given the following data:
+Let $\mathcal M$ be a \hyperref[def:step-1-model]{step-1 intensional model} with dyn.
 
-\begin{enumerate}
-    \item For each value type $A$, a monoid $\pv_A$ and homomorphism 
-    \[ \ptbv_A : \pv_A \to \{ f \in \vf(A,A) \mid f \bisim \id \} \]
-
-    \item For each computation type $B$, a monoid $\pv_B$ and homomorphism
-    \[ \ptbe_B : \pe_B \to \{ g \in \ef(B,B) \mid g \bisim \id \} \]
-
-    \item For each value type $A$, a distinguished endomorphism
-    $\delta_A \in \ef(FA, FA)$ such that $\delta_A \bisim \id_{FA}$.
-\end{enumerate}
-
-Then we can construct a \hyperref[def:step-2-model]{step-2 intensional model}.
+Then we can construct a \hyperref[def:step-2-model]{step-2 intensional model} with dyn.
 \end{lemma}
 \begin{proof}
-    Write 
+    % Write 
+    % %
+    % \[ \mathcal M = (\vf, \vsq, \ef, \esq, \Ff, \Fsq, \Uf, \Usq, \arrf, \arrsq). \] 
+    % %
+
+    Define a step-2 model $\mathcal M'$ as follows:
+    \begin{itemize}
+      \item Value objects are tuples consisting of:
+      \begin{itemize}
+        \item A value object $A$ in $\vf$ 
+        \item A monoid of ``pure'' perturbations $P_A$ 
+        \item A homomorphism of monoids $\ptb_A : P_A \to \{ f \in \vf(A, A) \mid f \bisim \id_A \}$
+        \item A monoid of ``impure'' perturbations $P^K_A$ that contains a distinguished element $\delta^*$
+        \item A homomorphism of monoids $\ptbk_A : P^K_A \to \{ \phi \in \ef(FA, FA) \mid \phi \bisim \id_{FA} \}$
+        such that $\ptbk_A(\delta^*) = \delta_A^*$
+      \end{itemize}
+
+      \item Computation objects are tuples consisting of:
+      \begin{itemize}
+        \item A computation object $B$ in $\ef$
+        \item A monoid of ``pure'' perturbations $P_B$
+        \item A homomorphism of monoids $\ptb_B : P_B \to \{ \phi \in \ef(B, B) \mid \phi \bisim \id_B \}$
+        \item A monoid of ``impure'' perturbations $P^K_B$
+        \item A homomorphism of monoids $\ptbk_B : P^K_B \to \{ g \in \vf(UB, UB) \mid g \bisim \id_{UB} \}$.
+      \end{itemize}
+
+      \item Morphisms are given by morphisms of the underlying objects in $\vf$ and $\ef$, respectively
+      %, i.e.,
+      % \[ \vf'((A, P_A, \ptb_A, P^K_A, \ptbk_A), (A', P_{A'}, \ptb_{A'}, P^K_{A'}, \ptbk_{A'})) = \vf(A, A') \]
       %
-    \[ \mathcal M = (\vf, \vsq, \ef, \esq, \Ff, \Fsq, \Uf, \Usq, \arrf, \arrsq). \] 
-    %
-    Define a step-2 model as follows:
+      % and likewise for computations.
+
+    \end{itemize}
+
+    Before introducing the relations, we make a definition.
+
+    \begin{definition}[push-pull structure]
+      Let $c : A \rel A'$ be a value relation of $\mathcal M$. A \emph{value push-pull structure} $\piv_c$ for $c$ consists of:
       \begin{itemize}
-      \item Value objects are tuples of an object $A$ in $\vf$ along with the monoid
-      $\pv_A$ and homomorphism $\ptbv_A$:
-      $\ob(\vf') = \{ (A, \pv_A, \ptbv_A) \mid A \in \ob(\vf) \}$.
+        \item A function $\push : P_A \to P_{A'}$ 
+              such that for all $\delta^l \in P_A$ we have $\delta^l \ltdyn_c^c \push(\delta^l)$.
+        \item A function $\push^K : P^K_A \to P^K_{A'}$ 
+              such that for all $\delta^K_l \in P^K_A$ we have $\delta^K_l \ltdyn_{Fc}^{Fc} \push(\delta^K_l)$.
+        \item A function $\pull : P_{A'} \to P_A$
+              such that for all $\delta_r \in P_{A'}$ we have $\pull(\delta^r) \ltdyn_{c}^c \delta^r$.
+        \item A function $\pull^K : P^K_{A'} \to P^K_A$
+              such that for all $\delta^K_r \in P^K_{A'}$ we have $\pull(\delta^K_r) \ltdyn_{Fc}^{Fc} \delta^K_r$.
+      \end{itemize}
 
-      \item Morphisms are given by morphisms of the underlying objects in $\vf$, i.e.,
-       $\vf'((A, \pv_A, \ptbv_A), (A', \pv_{A'}, \ptbv_{A'})) = \vf(A, A')$.
+      For $d : B \rel B'$ a computation relation, we define a \emph{computation push-pull structure} $\pie_d$ for $d$
+      in an analogous manner.
+    \end{definition}
 
-      \item Computation objects are tuples 
-      $\ob(\ef') = \{ (B, \pe_B, \ptbe_B) \mid B \in \ob(\ef) \}$.
+    Now we continue with the description of the construction:
+    \begin{itemize}
 
-      \item Computation morphisms are $\ef'((B, \pv_B, \ptbv_B), (B', \pv_{B'}, \ptbv_{B'})) = \ef(B, B')$.
+      \item The objects of $\vsq'$ (i.e., the value relations) are pairs consisting of:
+      \begin{itemize}
+        \item A value relation $c \in \vsq$
+        \item A push-pull structure $\piv_c$ for $c$
+      \end{itemize}
 
-      \item The objects $\vsq'$ and $\esq'$ are the same as those of $\vsq$ and $\esq$.
+      The objects of $\esq'$ are defined analogously.
 
-      \item The morphisms of $\vsq'$ and $\esq'$ are the same as those of $\vsq$ and $\esq$.
-    %   \item $\ob(\vsq') = \ob(\vsq)$
-    %   \item $\ob(\esq') = \ob(\esq)$
-    %   \item $\vsq'(c, c') = \vsq(c, c')$
-    %   \item $\esq'(d, d') = \esq(d, d')$
+      \item The morphisms of $\vsq'$ and $\esq'$ are given by the morphisms of $\vsq$ and $\esq$.
 
       % Functors \times, +, F, U, arrow
-      \item We define $F$ on objects by $F (A, \pv_A, \ptbv_A) = (FA, (1 + \pv_A), h_F)$
-      where $1$ is the trivial monoid, $+$ is the coproduct in the category of monoids, and $h_F$ is the homomorphism defined as follows:
 
-      \item We define $U$ on objects by $U (B, \pe_B, \ptbe_B) = (UB, \pe_B, h_U)$
-      where $h_U(p_B) = U(\ptbe_B(p_B))$.
+      % \item We define $F$ on objects by $F (A, \pv_A, \ptbv_A) = (FA, (1 + \pv_A), h_F)$
+      % where $1$ is the trivial monoid, $+$ is the coproduct in the category of monoids, and $h_F$ is the homomorphism defined as follows:
 
-      \item We define $(A, \pv_A, \ptbv_A) \arr (B, \pe_B, \ptbe_B) = (A \arr B, \pv_A \times \pe_B, h_\arr)$
-      where $\times$ is the product in the category of monoids, and $h_\arr$ is defined by 
-      $h_\arr(p_A, p_B) = \ptbv_A(p_A) \arr \ptbe_B(p_B)$.
+      % \item We define $U$ on objects by $U (B, \pe_B, \ptbe_B) = (UB, \pe_B, h_U)$
+      % where $h_U(p_B) = U(\ptbe_B(p_B))$.
+
+      % \item We define $(A, \pv_A, \ptbv_A) \arr (B, \pe_B, \ptbe_B) = (A \arr B, \pv_A \times \pe_B, h_\arr)$
+      % where $\times$ is the product in the category of monoids, and $h_\arr$ is defined by 
+      % $h_\arr(p_A, p_B) = \ptbv_A(p_A) \arr \ptbe_B(p_B)$.
     \end{itemize}
@@
   Let $\mathcal M$ be a \hyperref[def:step-2-model]{step-2 intensional model}.
-    Suppose we are given the following data:
 
@@
 \begin{proof}
+  Write 
+  %
+  \[ \mathcal M = (\vf, \vsq, \ef, \esq, \Ff, \Fsq, \Uf, \Usq, \arrf, \arrsq). \] 
+  %
 
+  We begin with a definition.
+
+  \begin{definition}[representation structure]
+  Let $c : A \rel A'$ be a value relation. A \emph{left-representation structure} $\rho^L_c$ for $c$ consists of
+  a value morphism $e_c \in \vf(A, A')$ such that $c$ is quasi-left-representable by $e_c$ (see Definition \ref{def:quasi-left-representable}).
+
+  Likewise, let $d : B \rel B'$. A \emph{right-representation structure} $\rho^R_d$ for $d$ consists of
+  a computation morphism $p_d \in \ef(B', B)$ such that $d$ is quasi-right-representable by $p_d$ (see Definition \ref{def:quasi-right-representable}).
+  \end{definition}
+
+  (Notice that the direction of the morphism is opposite in the definition of right-representation structure.)
+
+  We define a step-3 model $\mathcal M'$ as follows:
+  \begin{itemize}
+    \item The objects of $\mathcal M'$ are defined to be the same as the objects of $\mathcal M$.
+    \item The value and computation morphisms in $\mathcal M'$ are the same as those of $\mathcal M$.
+    \item A value relation is defined to be a tuple $(c, \rho^L_c, \rho^R_{Fc})$ where:
+    \begin{itemize}
+      \item $c$ is a value relation in $\mathcal M$, and 
+      \item $\rho^L_c$ a left-representation structure for $c$, and 
+      \item $\rho^R_{Fc}$ a right-representation structure for $Fc$.
+    \end{itemize}
+    \item Likewise, a computation relation is defined to be a tuple $(d, \rho^R_d, \rho^L_{Ud})$ with
+    \begin{itemize}
+      \item $d$ a computation relation in $\mathcal M$
+      \item $\rho^R_d$ a right-representation structure for $d$
+      \item $\rho^L_{Ud}$ a left-representation structure for $Ud$
+    \end{itemize}
+    \item Morphisms of value relations (i.e., the value squares) are defined by simply
+    ignoring the representation structures. That is, a morphism of value relations
+    $\alpha \in \vsq'((c, \rho^L_c, \rho^R_{Fc}), (c' \rho^L_{c'}, \rho^R_{Fc'}))$ is simply a morphism of value
+    relations in $\vsq(c, c')$. Likewise for computations.
+  \end{itemize}
 \end{proof}
+
+% Now we define the functors $F$, $U$, $\times$, and $\arr$.
+
+% On objects, the behavior is the same as the respective functors in $\mathcal M$.
+
+% For relations, we define 
+% $\Fsq' (c, \rho^L_c, \rho^R_{Fc}) = (\Fsq c, \rho^R_{Fc}, UF(\rho^L_c))$ and
+% $\Usq' (d, \rho^R_d, \rho^L_{Ud}) = (\Usq d, \rho^L_{Ud}, FU(\rho^R_d))$.
+
+% We define $(c, \rho^L_c) \arr (d, \rho^R_d) = (c \arr d, \rho^R_{c \arr d})$.
+
+% We now verify that the construction meets the requirements of a step-3 model.
+% First, we check that composition of value relations (resp. computation relations)
+% is well-defined.
 

### 2024-01-19 paper-new/categorical-model.tex
work on abstract model constructions

@@
 Lastly, we require that for any value object $A$, the hom-set $\ef(FA, FA)$ contains a
-distinguished morphism $\delta_A$, such that $\delta_A \bisim_{FA, FA} \id_{FA}$.
+distinguished morphism $\delta_A^*$, such that $\delta_A^* \bisim_{FA, FA} \id_{FA}$.
 
@@
 with the additional categories and functors for bisimiarity described above, as  well as
-the existence of a distinguised computation morphism $\delta_A \bisim \id_{FA}$ for each $A$.
+the existence of a distinguised computation morphism $\delta_A^* \bisim \id_{FA}$ for each $A$.
 \end{definition}
@@
 
-We require that $\delta_A \in \pe_{FA}$ for all $A$, where $\delta_A$ is the distinguished
+We require that $\delta_A^* \in \pe_{FA}$ for all $A$, where $\delta_A^*$ is the distinguished
 morphism that is required to be present in every hom-set $\ef(FA, FA)$ per the definition

### 2024-01-19 paper-new/appendix.tex
describe Kleisli action on squares

@@
 \end{enumerate}
+
+Finally, note that all of these constructions lift to squares in a
+double CBPV model since the squares themselves form a CBPV model and
+the projection functions preserve CBPV structure. For instance, given a square
+$\alpha : l \ltdyn_{F c_o}^{F c_i} l'$ and a horizontal morphism $d : B \rel B'$ of appropriate type, we get a square
+\[ \alpha \tok d : l \tok B \ltdyn_{U(c_o \to d)}^{U(c_i \to d)} l' \tok B' \]
 

### 2024-01-22 paper-new/appendix.tex
match Eric's notation for linear morphisms

--- 2024-01-19 paper-new/appendix.tex
@@
 \begin{enumerate}
-\item For $\tok$ we define for $l : \calE(F A,F A')$ and $B \in \calE$ the morphism $l \tok B : \calV(U(A' \to B),U(A\to B))$ as
-  \[ t:U(A'\to B) \vdash l \tok B = \{ \lambda x. x' \leftarrow l\,[\ret x]; ! t x'\} : U(A \to B) \]
+\item For $\tok$ we define for $\phi : \calE(F A,F A')$ and $B \in \calE$ the morphism $\phi \tok B : \calV(U(A' \to B),U(A\to B))$ as
+  \[ t:U(A'\to B) \vdash \phi \tok B = \{ \lambda x. x' \leftarrow \phi\,[\ret x]; ! t x'\} : U(A \to B) \]
   and for $A \in \calV$ and $f : \calV(UB,UB')$ we define $A \tok f : \calV(U(A \to B),U(A\to B'))$ as
   \[ t : U(A \to B) \vdash A \tok f = \{ \lambda x. !f[\{ ! t x \}]\} \]
-\item For $\timesk$ we define for $l : \calE(F A_1,FA_2)$ and $A' \in \calV$ the morphism $l \timesk A_2$ as
-  \[ \bullet : F(A_1\times A_2) \vdash l \timesk A_2 = (x_1,x_2) \leftarrow \bullet; x_1' \leftarrow l[\ret x_1]; \ret (x_1',x_2) : F(A_1'\times A_2)\]
-  and $A_1 \timesk l$ is defined symmetrically.
+\item For $\timesk$ we define for $\phi : \calE(F A_1,FA_2)$ and $A' \in \calV$ the morphism $\phi \timesk A_2$ as
+  \[ \bullet : F(A_1\times A_2) \vdash \phi \timesk A_2 = (x_1,x_2) \leftarrow \bullet; x_1' \leftarrow \phi[\ret x_1]; \ret (x_1',x_2) : F(A_1'\times A_2)\]
+  and $A_1 \timesk \phi$ is defined symmetrically.
 \item For $\Uk$ we need to define for $f : \calV(UB,UB')$ a morphism $\Uk f : \calE(FUB,FUB')$. This is simply given by the functorial action of $F$: $\Uk f = F(f)$
-\item Similarly $\Fk l = Ul$
+\item Similarly $\Fk \phi = U\phi$
 \end{enumerate}
@@
 \begin{enumerate}
-\item $(l \circ l') \tok B = (l' \tok B) \circ (l \tok B)$
+\item $(\phi \circ \phi') \tok B = (\phi' \tok B) \circ (\phi \tok B)$
 \item $\id \tok B = \id$
@@
 \end{enumerate}
+
+Finally, note that all of these constructions lift to squares in a
+double CBPV model since the squares themselves form a CBPV model and
+the projection functions preserve CBPV structure. For instance, given a square
+$\alpha : \phi \ltdyn_{F c_o}^{F c_i} \phi'$ and a horizontal morphism $d : B \rel B'$ of appropriate type, we get a square
+\[ \alpha \tok d : \phi \tok B \ltdyn_{U(c_o \to d)}^{U(c_i \to d)} \phi' \tok B' \]
 

### 2024-01-23 paper-new/appendix.tex
sketchy stuff in the appendix on syntax and update to the extensional model

@@
+\section{Gradual Typing Syntax}
+
+\max{decide if we want to include $\times$ or not}
+
+\begin{align*} % TODO is hole a term?
+  &\text{Types } A := \dyn \alt \nat \alt A \times A \alt (A \ra A') \\
+  &\text{Value Contexts } \Gamma := \cdot \alt (\Gamma, x : A) \\
+  &\text{Terms } M, N := \upc c M \alt \dnc c M
+  &\quad\quad \alt \ret {V} \alt \bind{x}{M}{N} \alt V_f\, V_x \alt \dn{A}{B} M 
+\end{align*}
+
+The type precision derivations $c : A \ltdyn A'$ is inductively defined by
+reflexivity, transitivity, congruence for $\ra$ and $\times$, and generators
+$\textsf{Inj}_\ra : (D \ra D) \ltdyn D$ and $\textsf{Inj}_{\text{nat}}
+: \nat \ltdyn D$.
+%
+We define equivalence of type precision derivations to be inductively generated by congruence for all constructors, category laws for reflexivity and transitivity as well as functoriality laws for $\ra$ and $\times$ congruence
+\[ (c_i \ra c_o)(c_i' \ra c_o) = (c_ic_i') \ra (c_oc_o') \]
+\[ (c_1 \times c_2)(c_1' \times c_2') = (c_1c_1') \times (c_2c_2') \]
+
+\begin{theorem}
+  \begin{enumerate}
+  \item For every $A$, there is a derivation $c : A \ltdyn D$
+  \item Any two derivations $c,c' : A \ltdyn A'$ of the same precision
+    are equivalent.
+  \end{enumerate}
+\end{theorem}
+\begin{proof}
+  \begin{enumerate}
+  \item See prior work \cite{prior-work}
+  \item We show this by showing that derivations have a canonical
+    form.
+
+    The following presentation of precision derivations has unique derivations
+    \begin{mathpar}
+      \inferrule{}{\textrm{refl}(D) : D \ltdyn D}
+      \inferrule{}{\textsf{Inj}_{\text{nat}} : \nat \ltdyn D}
+      \inferrule{}{\textrm{refl}(\nat) : \nat \ltdyn \nat}
+      \inferrule{c : A_i \ra A_o \ltdyn D\ra D}{c(\textsf{Inj}_{\text{arr}}) : A_i \ra A_o \ltdyn \nat}
+      \inferrule{c : A_i \ltdyn A_i' \and d : A_o \ltdyn A_o'}{c \ra d : A_i \ra A_o \ltdyn A_i'\ra A_o'}
+      %% \inferrule{A_1 \times A_2 \ltdyn D\times D}{A_1 \times A_2 \ltdyn \nat}
+      %% \inferrule{A_1 \ltdyn A_1' \and A_2 \ltdyn A_2'}{A_1 \times A_2 \ltdyn A_1'\times A_2'}
+    \end{mathpar}
+    Since it satisfies reflexivity, cut-elimination and congruence, it
+    is a model of the original theory. Since it is a sub-theory of the
+    original theory, it is equivalent.
+  \end{enumerate}
+\end{proof}
+
+The gradually typed lambda calculus we consider is call-by-value
+gradual lambda calculus with $\ra$ and $\nat$ as the only base types.
+Casts are generated by the rule
+\begin{mathpar}
+  \inferrule
+  {\Gamma \vdash M : A \and c : A \ltdyn A'}
+  {\Gamma \vdash \upc c M : A'}
+
+  \inferrule
+  {\Gamma \vdash M : A' \and c : A \ltdyn A'}
+  {\Gamma \vdash \dnc c M : A}
+\end{mathpar}
+
+Type precision is a binary relation on typed terms. The original
+gradual guarantee rules are as follows:
+\begin{mathpar}
+  \inferrule
+  {\Gamma^\ltdyn \vdash M \ltdyn M' : c \and
+    c : A \ltdyn A'\and
+    c' : A \ltdyn A_2'
+  }
+  {\Gamma^\ltdyn \vdash M \ltdyn (M :: A_2') : c'}
+
+  \inferrule
+  {\Gamma^\ltdyn \vdash M \ltdyn M' : c \and
+    c : A \ltdyn A'\and
+    c' : A_2 \ltdyn A'
+  }
+  {\Gamma^\ltdyn \vdash (M :: A_2) \ltdyn M' : c'}
+\end{mathpar}
+These two rules are admissible from the following principles:
+
+%% \begin{mathpar}
+%%   \inferrule
+%%   {}
+%%   {\dnc {\injarr{}} \upc {\injarr{}} M \equidyn M}
+
+%%   %% \inferrule
+%%   %% {}
+%%   %% {}
+%% \end{mathpar}
+
 \section{Call-by-push-value}

### 2024-01-23 paper-new/categorical-model.tex
sketchy stuff in the appendix on syntax and update to the extensional model

@@
 
-In particular, let $c : A \rel A'$ and $d : A' \rel A''$,
-and let $f \in \mathcal V_f(A, A')$ and $g \in \mathcal V_f(A', A'')$.
-We say that the pair $(c,d)$ is \emph{left-representable by} $(f, g)$
-if the following two squares commute:
-
+In particular, let $c : A \rel A'$ and $f \in \mathcal V_f(A, A')$. We
+say that $c$ is \emph{left-representable by} $f$ if for any $c_l : A_l
+\rel A$ and $c_r : A' \rel A_r$ the following squares commute\max{is ``commute'' good terminology?}:
 \begin{center}
@@
     \begin{tikzcd}[ampersand replacement=\&]
-      A \& {A'} \& {A''} \\
-      {A'} \&\& {A''}
+      A \& {A'} \& {A_r} \\
+      {A'} \&\& {A_r}
       \arrow["f"', from=1-1, to=2-1]
@@
       \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-      \arrow["d", "\shortmid"{marking}, no head, from=1-2, to=1-3]
+      \arrow["c_r", "\shortmid"{marking}, no head, from=1-2, to=1-3]
       %
-      \arrow["d"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
+      \arrow["c_r"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
     \end{tikzcd}
@@
     \begin{tikzcd}[ampersand replacement=\&]
-      A \&\& {A'} \\
-      {A} \& {A'} \& {A''}
+      A_l \&\& {A} \\
+      {A_l} \& {A} \& {A'}
       \arrow[from=1-1, to=2-1, Rightarrow, no head]
       %
-      \arrow["g", from=1-3, to=2-3]
+      \arrow["f", from=1-3, to=2-3]
       %
-      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-3]
+      \arrow["c_l", "\shortmid"{marking}, no head, from=1-1, to=1-3]
       %
-      \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-      \arrow["d"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
+      \arrow["c_l"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+      \arrow["c"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
     \end{tikzcd}
@@
 
-Dually, let $c \in \mathcal E_e(B, B')$ and $d \in \mathcal E_e(B', B'')$,
-and let $f \in \mathcal E_f(B'', B')$ and $g \in \mathcal E_f(B', B)$.
-We say the pair $(c,d)$ is \emph{right-representable by} $(f,g)$ if the
-following two squares commute:
+Dually, let $d \in B \rel B'$ and $\phi \in \mathcal E_f(B', B)$.  We
+say that $d$ is \emph{right-representable by} $\phi$ if for any $d_l :
+B_l \rel B$ and $d_r : B' \rel B_r$ the following two squares commute:
 
@@
     \begin{tikzcd}[ampersand replacement=\&]
-      {B} \& {B'} \& {B''} \\
-      {B} \&\& {B'}
+      {B_l} \& {B} \& {B'} \\
+      {B_l} \&\& {B}
       \arrow[from=1-1, to=2-1, Rightarrow, no head]
       %
-      \arrow["f", from=1-3, to=2-3]
+      \arrow["\phi", from=1-3, to=2-3]
       %
-      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+      \arrow["d_l", "\shortmid"{marking}, no head, from=1-1, to=1-2]
       \arrow["d", "\shortmid"{marking}, no head, from=1-2, to=1-3]
       %
-      \arrow["c"', from=2-1, to=2-3, no head]
+      \arrow["d_l"', from=2-1, to=2-3, no head]
     \end{tikzcd}
@@
     \begin{tikzcd}[ampersand replacement=\&]
-      {B'} \&\& {B''} \\
-      {B} \& {B'} \& {B''}
-      \arrow["g"', from=1-1, to=2-1]
+      {B'} \&\& {B_r} \\
+      {B} \& {B'} \& {B_r}
+      \arrow["\phi"', from=1-1, to=2-1]
       %
@@
       %
-      \arrow["d", from=1-1, to=1-3, no head]
+      \arrow["d_r", from=1-1, to=1-3, no head]
       %
-      \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-      \arrow["d"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
+      \arrow["d"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+      \arrow["d_r"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
     \end{tikzcd}
@@
 
-Then we formulate the relationship between value relation morphisms and
+Then we formulate the relationship between relation morphisms and
 function morphisms as follows:
 \begin{enumerate}
-  \item There is a functor $\upf : \mathcal V_e \to \mathcal V_f$ that is the
-  identity on objects, such that $\upf_*(\mathcal V_e)$, the essential image of
-  $\mathcal V_e$ under $\upf$, is thin.
-  The objects of $\upf_*(\mathcal V_e)$ are the objects of $\mathcal V_f$, and
-  the hom-set
-  $\upf_*(\mathcal V_e)(A, A') = \{ f \in \mathcal V_f(A,A') \mid \exists c \in \mathcal V_e(A,A'). \upf(c) = f \}$.
-
-  \item Every pair of morphisms $(c,d) \in \mathcal V_e(A, A') \times \mathcal V_e(A', A'')$ is
-  left-representable by $(\upf(c), \upf(d))$.
+\item There is an identity-on-objects functor $\upf : \mathcal V_e \to
+  \mathcal V_f$ such that every $c$ is left-representable by $\upf(c)$.
+\item There is an identity-on-objects functor $\dnf : \mathcal
+  E_e^{op} \to \mathcal \mathcal E_f$ such that every $d$ is
+  right-representable by $\dnf(d)$.
 \end{enumerate}
 
-And likewise for computations:
-\begin{enumerate}
-  \item There is a functor $\dnf : \mathcal E_e^{op} \to \mathcal V_f$ that is
-  the identity on objects, such that the essential image of $\mathcal E_e$ under
-  $\dnf$ is thin.
-
-  \item Every pair of morphisms $(c,d) \in \mathcal E_e(B, B') \times \mathcal E_e(B',B'')$
-  is right-representable by $(\dnf(d), \dnf(c))$.
-\end{enumerate}
-
-\textbf{TODO: do we still need this?}
-We also want something like
-\[ F_c : \mathcal V_u^{op} \to \mathcal E_d \]
-\[ U_c : \mathcal E_d^{op} \to \mathcal V_u \]
-which ensures that if $R$ is a value edge equivalent to $A(u,-)$ then
-\[ F(R) = F(A(u,-)) = (F A)(-,F u) \]
+%% \textbf{TODO: do we still need this?}
+%% We also want something like
+%% \[ F_c : \mathcal V_u^{op} \to \mathcal E_d \]
+%% \[ U_c : \mathcal E_d^{op} \to \mathcal V_u \]
+%% which ensures that if $R$ is a value edge equivalent to $A(u,-)$ then
+%% \[ F(R) = F(A(u,-)) = (F A)(-,F u) \]
 
@@
 \begin{enumerate}
-  \item CBPV models $\mathcal M_f$ and $\mathcal M_{sq}$
-  \item CBPV morphisms $r : \mathcal M_f \to \mathcal M_{sq}$ and $s, t : \mathcal M_{sq} \to \mathcal M_f$
-  \item Thinness: there is at most one square with a given boundary
-  \item A ``horizontal composition" operation on value relations and on computation relations
-  (from which we can define the categories $\ve$ and $\ee$ of value types/relations and computation types/relations, repsectively).
-  \item The categories $\ve$ and $\ee$ are thin up to an identity square
-  \item A functor $\upf : \mathcal V_e \to \mathcal V_f$ that is the identity on objects,
-  such that $\upf_*(\mathcal V_e)$, the essential image of $\mathcal V_e$ under $\upf$, is thin.
-  \item Every pair of morphisms $(c,d) \in \mathcal V_e(A, A') \times \mathcal V_e(A', A'')$ is
-  left-representable by $(\upf(c), \upf(d))$.
-  \item A functor $\dnf : \mathcal E_e^{op} \to \mathcal V_f$ that is
-  the identity on objects, such that the essential image of $\mathcal E_e$ under $\dnf$ is thin.
-  \item Every pair of morphisms $(c,d) \in \mathcal E_e(B, B') \times \mathcal E_e(B',B'')$
-  is right-representable by $(\dnf(d), \dnf(c))$.
+  \item A CBPV model internal to reflexive graphs.
+  \item Composition and identity on value and computation relations that form a category.
+  \item An identity-on-objects functor $\upf : \mathcal V_e \to \mathcal V_f$ taking each value relation to a morphism that left-represents it.
+  \item An identity-on-objects functor $\dnf : \mathcal E_e^{op} \to \mathcal E_f$ taking each computation relation to a morphism that right-represents it.
+  \item The CBPV connectives $U,F,\times,\to$ are all \emph{covariant} functorial on relations
+    \begin{itemize}
+    \item $U(dd') = U(d)U(d')$
+    \item $F(cc'') = F(c)F(c')$
+    \item $(cc') \to (dd') = (c \to d)(c' \to d')$
+    \item $(c_1c_1') \times (c_2c_2') = (c_1 \times c_2)(c_1'\times c_2')$
+    \end{itemize}
+  \item A distinguished value type $D$ with distinguished relation
+    $\injarr{}: U(D \to F D) \rel D$ satisfying $\dnc {\injarr{}}F(\upc{\injarr{}}) = \id$
 \end{enumerate}

### 2024-01-24 paper-new/categorical-model.tex
updates to abstract categorical model section

@@
     \item $U(dd') = U(d)U(d')$
-    \item $F(cc'') = F(c)F(c')$
+    \item $F(cc') = F(c)F(c')$
     \item $(cc') \to (dd') = (c \to d)(c' \to d')$
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsubsection{Bisimilarity}\label{sec:abstract-model-bisimilarity}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsubsection{Perturbations}\label{sec:abstract-model-perturbations}
@@
 We formalize this notion by requiring that for each object $A$ in $\vf$,
-there is a monoid of \emph{perturbations} $P^V_A$ and a monoid homomorphism
-$\ptbv_A : \pv_A \to \vf(A,A)$.
-Similarly, for each $B : \ef$ there is a monoid $P^C_B$ and a
-homomorphism $\ptbe_B : \pe_B \to \ef(B,B)$.
-If $\delta \in P^V_A$, we will sometimes omit the homomorphism $\ptbv_A$ and simply write
-$\delta$ to refer to the morphism $\ptbv_A(\delta) \in \vf(A,A)$, and likewise
-for computation perturbations. The context will make clear whether we are referring
-to an element of the perturbation monoid or the corresponding morphism.
+there is a monoid of \emph{value perturbations} $\pv_A$ that is a submonoid of
+$\{ f \in \vf(A,A) \mid f \bisim \id \}$.
+Similarly, for each $B : \ef$ there is a monoid $\pe_B$ of
+\emph{computation perturbations} that is a submonoid of 
+$\{ g \in \ef(B,B) \mid g \bisim \id \}$.
+
+% If $\delta \in P^V_A$, we will sometimes omit the homomorphism $\ptbv_A$ and simply write
+% $\delta$ to refer to the morphism $\ptbv_A(\delta) \in \vf(A,A)$, and likewise
+% for computation perturbations. The context will make clear whether we are referring
+% to an element of the perturbation monoid or the corresponding morphism.
+
+Note that we require that all perturbations be weakly bisimilar to the identity morphism,
+capturing the notion that they have no effect other than to delay.
+% We observe that
+% the set of endomorphisms $f$ such that $f$ is weakly bisimilar to the identity
+% forms a monoid under composition.
 
@@
 
-The perturbations must be preserved by $\times$, $\arr$, $U$, and $F$.
+The perturbations must be preserved by $\times$ $\timesk$, $\arr$, $\tok$, $U$, $\Uk$, $F$, and $\Fk$.
 
-For reasons that will be made clear in the next section, perturbations must also satisfy a property that we call the ``push-pull'' property,
+For reasons that will be made clear in the next section, perturbations must also satisfy
+a property that we call the ``push-pull'' property,
 which is formulated as follows. Let $c : A \rel A'$.
@@
 
-Lastly, we require that all perturbations be weakly bisimilar to the identity morphism,
-capturing the notion that they have no effect other than to delay. We observe that
-the set of endomorphisms $f$ such that $f$ is weakly bisimilar to the identity
-forms a monoid under composition.
-
 This is summarized below:
@@
   \begin{enumerate}
-    \item $\pv_A$ and a monoid homomorphism 
-      \[ \ptbv_A : \pv_A \to \{ f \in \vf(A,A) \mid f \bisim \id \} \]
-    \item $\pe_B$ and a monoid homomorphism 
-      \[ \ptbe_B : \pe_B \to \{ g \in \ef(B,B) \mid g \bisim \id \} \]
-    \item The functors $\times$, $\arr$, $U$, and $F$ preserve perturbations.
+    \item For each value type $A$, there is a monoid $\pv_A$ that is a submonoid of
+    $\{ f \in \vf(A,A) \mid f \bisim \id \}$.
+    \item For each computation type $B$, there is a monoid $\pe_B$ that is a submonoid of
+    $\{ g \in \ef(B,B) \mid g \bisim \id \}$.
+    \item For all $A$, the distinguished endomorphism $\delta_A^*$ is in $\pe_{FA}$.
+    % \item $\pv_A$ and a monoid homomorphism 
+    %   \[ \ptbv_A : \pv_A \to \{ f \in \vf(A,A) \mid f \bisim \id \} \]
+    % \item $\pe_B$ and a monoid homomorphism
+    %   \[ \ptbe_B : \pe_B \to \{ g \in \ef(B,B) \mid g \bisim \id \} \]
+    \item The functors $\times, \timesk$, $\arr, \tok$, $U, \Uk$, $F, \Fk$ preserve perturbations.
     \item The push-pull property holds for all $c : A \rel A'$ and all $d : B \rel B'$.
@@
 \end{definition}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+% TODO: Give these squares names
 \begin{definition}\label{def:quasi-left-representable}
@@
 \end{center}
+
+We call the first square $\upl$ and the second square $\upr$.
+
 \end{definition}
@@
 \end{center}
+
+We call the first square $\dnr$ and the second square $\dnl$.
+
 \end{definition}
@@
 so we can take the simpler versions as primitive and derive the ones
-involving composition (though we must be careful about the perturbations
-when doing so!).
+involving composition.
+
+Lastly, we require that the model satisfy a weak version of functoriality for 
+the CBPV connectives $U,F,\times,\to$. 
+First, we will need a definition:
+
+\begin{definition}
+  Let $c, c' : A \rel A'$. We say that $c$ and $c'$ are \emph{quasi-order-equivalent},
+  written $c \qordeq c'$, if there exist perturbations $\delta^l_1, \delta^l_2 \in \pv_A$ and 
+  $\delta^r_1, \delta^r_2 \in \pv_{A'}$ such that the following two squares exist:
+
+  \begin{center}
+    \begin{tabular}{ m{9em} m{9em} } 
+      \begin{tikzcd}[ampersand replacement=\&]
+        A \& {A'} \\
+        A \& {A'}
+        \arrow["\delta^l_1"', from=1-1, to=2-1]
+        \arrow["\delta^r_1", from=1-2, to=2-2]
+        \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+        \arrow["c'"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+      \end{tikzcd}
+      &
+      \begin{tikzcd}[ampersand replacement=\&]
+        A \& {A'} \\
+        A \& {A'}
+        \arrow["\delta^l_2"', from=1-1, to=2-1]
+        \arrow["\delta^r_2", from=1-2, to=2-2]
+        \arrow["c'", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+        \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+      \end{tikzcd}
+    \end{tabular}
+  \end{center}
+
+  We make the analogous definition for computation relations $d, d' : B \rel B'$.
+\end{definition}
+
+We require that the CBPV connectives $U,F,\times,\to$ are \emph{quasi-functorial} on relations,
+which we specify as follows:
+\begin{itemize}
+  \item $U(d \comp d') \qordeq U(d)U(d')$
+  \item $F(c \comp c'') \qordeq F(c)F(c')$
+  \item $(cc') \to (dd') \qordeq (c \to d)(c' \to d')$
+  \item $(c_1c_1') \times (c_2c_2') \qordeq (c_1 \times c_2)(c_1'\times c_2')$
+\end{itemize}
+
+We summarize the requirements of a step-3 model below:
 
@@
   A \emph{step-3 intensional model} consists
-  of all the data of a step-2 intensional model, with the following additional data:
+  of all the data of a step-2 intensional model, such that additionally:
   \begin{enumerate}
-    \item Functors $\upf : \ve \to \vf$ and $\dnf : \ee^{op} \to \ef$ % TODO: image is thin?
+    \item There are functors $\upf : \ve \to \vf$ and $\dnf : \ee^{op} \to \ef$ % TODO: image is thin?
     \item Every value edge $c : A \rel A'$ is quasi-left-representable by $\upf(c)$ and
     every computation edge $d : B \rel B'$ is quasi-right-representable by $\dnf(d)$.
+    \item The CBPV connectives $U,F,\times,\to$ are quasi-functorial on relations.
   \end{enumerate}
 \end{definition}
+
+% Want: U d \comp U d' = U(d \comp d')
+%       F c \comp F c' = F(c \comp c')
+% Add requirement: Either the model is functorial with respect to up/downcasts or with repsect to relations
+% 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 Now we can discuss what it means for an intensional model to model the dynamic type.
-This applies to any of the above abstract model definitions, i.e., steps 0-3.
+% This applies to any of the above abstract model definitions, i.e., steps 0-3.
 
 \begin{definition}\label{def:step-4-model}
-  A \emph{step-$i$ intensional model with dyn} is a step-$i$ model $\mathcal M$ such that:
+  % A \emph{step-$i$ intensional model with dyn} is a step-$i$ model $\mathcal M$ such that:
+  A step-4 intensional model is a step-3 intensional model $\mathcal M$ such that:
   %a distinguished value object $D \in \ob(\vf)$ such that:
@@
 
-    \item If $c : A \rel A'$, then $\text{inj}_{A} = c \comp \text{inj}_{A'}$.
+    %\item \eric{Do we need this?} If $c : A \rel A'$, then $\text{inj}_{A} = c \comp \text{inj}_{A'}$.
   \end{enumerate}
@@
 
-Suppose we have a \hyperref[def:step-1-model]{step-1 intensional model} $\mathcal{M}$ with dyn.
+Suppose we have a \hyperref[def:step-1-model]{step-1 intensional model} $\mathcal{M}$.
 Recall that a step-1 intensional model consists of a step-0 model (i.e., a
@@
 $A$ and $\pe_B$ for all $B$.
-These perturbations must be interpretable via homomorphisms
-$\ptb_A$ and $\ptb_B$ as endomorphisms bisimilar to the identity,
-and moreover, the push-pull property must hold for all
+Moreover, the push-pull property must hold for all
 value relations $c$ and all computation relations $d$.
 
-% TODO do we really need all of this data?
-What ``external" data do we need to define such a model?
-Suppose we are given, for each value type $A$, a monoid $\pv_A$ with homomorphism
-$\ptb_A$ into the endomorphisms on $A$ bisimilar to the identity, and likewise, for
-each computation type $B$, a monoid $\pv_B$ and homomorphism $\ptb_B$.
-Furthermore, suppose we are given, for each value type $A$, a distinguished endomorphism
-$\delta_A \in \ef(FA, FA)$ such that $\delta_A \bisim \id_{FA}$.
-Finally, suppose that each value relation $c : A \rel A'$ satisfies the push-pull property
-and likewise for each computation relation $d : B \rel B'$.
-
-Then we claim that from the above data, we can construct a step-2 intensional model.
+We claim that from a step-1 model, we can construct a step-2 model. 
+% TODO give overview
 For the proof, see Lemma \ref{lem:step-1-model-to-step-2-model} in the Appendix.
@@
 Now suppose we have a step-2 intensional model $\mathcal{M}$.
-We will describe how to construct a \hyperref[def:step-3-model]{step-3 intensional model} $\mathcal{M'}$.
+We claim that we can construct a \hyperref[def:step-3-model]{step-3 intensional model} $\mathcal{M'}$.
 
-We now describe the construction of a step-3 model $\mathcal M'$.
+% TODO give overview
+
+For the proof, see Lemma \ref{lem:step-2-model-to-step-3-model} in the Appendix.
 
@@
 
-\subsubsection{Defining an Extensional Model}
+\subsubsection{Defining an Extensional Model}\label{sec:extensional-model-definition}
 
-Finally, suppose $\mathcal M$ is a step-3 intensional model.
+Finally, suppose $\mathcal M$ is a \hyperref[def:step-4-model]{step-4 intensional model}
+(i.e., a step-3 model with an interpretation of the dynamic type).
 We now describe how to build an extensional model.
@@
 
-Next, we check that the requirements of an extensional model are satisfied.
-In particular, we need to verify the representability properties.
-We define functors $\upf$ and $\dnf$
+The proof that this indeed defines an extensional model is given in the Appendix 
+(see Section \ref{sec:extensional-construction-appendix}).
+
+% Next, we check that the requirements of an extensional model are satisfied.
+% In particular, we need to verify the representability properties.
+% We define functors $\upf$ and $\dnf$

### 2024-01-24 paper-new/paper.tex
begin rearranging: discuss term syn/sem and then ordering syn/sem
begin description of denotational model for intensional gradual typing
work on abstract categorical model
lics format
lics review format
begin section on concrete model
begin discussion section
begin appendix
some intro work and add macros for comments
correct title/authors
updates to abstract categorical model section

@@
-\documentclass[acmsmall,screen]{acmart}
+\documentclass[sigconf,anonymous,review,screen,9pt]{acmart}
+\let\Bbbk\relax
+
+\usepackage{quiver}
 \usepackage{mathpartir}
-\usepackage{tikz-cd}
+% \usepackage{tikz-cd}
 \usepackage{enumitem}
@@
 \usepackage{comment}
-
-\usepackage{ stmaryrd }
+\usepackage{array}
 
@@
 
-\title{Mechanized Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory}
+\title{Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory}
 \author{Eric Giovannini}
+\affiliation{
+  \department{Electrical Engineering and Computer Science}
+  \institution{University of Michigan}
+  \country{USA}
+}
+\email{ericgio@umich.edu}
+
+\author{Tingting Ding}
+\affiliation{
+  \department{Electrical Engineering and Computer Science}
+  \institution{University of Michigan}
+  \country{USA}
+}
+\email{tingtind@umich.edu}
+
 \author{Max S. New}
+\affiliation{
+  \department{Electrical Engineering and Computer Science}
+  \institution{University of Michigan}
+  \country{USA}
+}
+\email{maxsnew@umich.edu}
 
@@
   metatheoretic properties of gradually typed languages have become
-  increasingly sophisticated: validity of typed based equational
+  increasingly sophisticated: validity of type based equational
   reasoning as well as the relational property known as the gradual
@@
 
+\newif\ifdraft
+\drafttrue
+\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max}: #1]}\fi}
+\newcommand{\eric}[1]{\ifdraft{\color{orange}[{\bf Eric}: #1]}\fi}
+\newcommand{\tingting}[1]{\ifdraft{\color{red}[{\bf Tingting}: #1]}\fi}
+
 \input{intro}
@@
 
-\input{syntax}
+% \input{syntax}
 
-\input{semantics}
+% \input{semantics}
+
+\input{terms}
+
+\input{ordering}
 
@@
 
-\section{Discussion}\label{sec:discussion}
+\input{categorical-model}
+
+%\input{denotational-model}
+\input{concrete-model}
+
+\input{discussion}
+
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{references}
+
+\appendix
+\input{appendix}
+
+% \section{Discussion}\label{sec:discussion}
 
@@
 
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{references}
-
 \end{document}

### 2024-01-24 paper-new/appendix.tex
work on proofs in appendix

@@
 
+\subsection{Constructing a Model with Perturbations}
+
+The goal of this section is to prove the following lemma:
+
 \begin{lemma}\label{lem:step-1-model-to-step-2-model}
-Let $\mathcal M$ be a \hyperref[def:step-1-model]{step-1 intensional model} with dyn.
+Let $\mathcal M$ be a \hyperref[def:step-1-model]{step-1 intensional model}.
 
-Then we can construct a \hyperref[def:step-2-model]{step-2 intensional model} with dyn.
+Then we can construct a \hyperref[def:step-2-model]{step-2 intensional model}.
+\end{lemma}
+
+We begin with a definition and some lemmas that will be useful in the construction
+of the model.
+
+\begin{definition}
+  Let $c : A \rel A'$ be a value relation of $\mathcal M$.
+  Let $P_A$ be a monoid of perturbations on $A$ and $P_{A'}$ a monoid of perturbations on $A'$.
+  A \emph{push-pull structure} $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$ consists of:
+  \begin{itemize}
+    \item A function $\push : P_A \to P_{A'}$ 
+    such that for all $\delta^l \in P_A$ we have $\delta^l \ltdyn_c^c \push(\delta^l)$.
+    \item A function $\pull : P_{A'} \to P_A$
+    such that for all $\delta_r \in P_{A'}$ we have $\pull(\delta^r) \ltdyn_{c}^c \delta^r$.
+  \end{itemize}
+
+  We define a push-pull structure $\pie_d$ for $d : B \rel B'$ with respect to perturbation monoids
+  $P_B$ and $P_{B'}$ in an analogous manner.
+
+  % \eric{do we need that $\delta^*$ is preserved by push and pull?}
+  % We stipulate the additional requirement that if $\delta_A^*$ is in $P_B$ 
+  % (where $\delta_A^*$ is the distinguished perturbation required to be in every $\ef(FA, FA)$),
+  % then $\push(\delta_A^*) = \delta_{A'}^*$ and likewise $\pull(\delta_{A'}^*) = \delta_A^*$.
+\end{definition}
+
+%%%%%%%%%%%%%%%
+% Composition %
+%%%%%%%%%%%%%%%
+\begin{lemma}
+  Let $c : A \rel A'$ and $c' : A' \rel A''$ be value relations,
+  and let $P_A, P_{A'}$ and $P_{A''}$ be monoids of perturbations.
+  Given a push-pull structure $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$, and
+  and a push-pull structure $\piv_{c'}$ for $c'$ with respect to $P_{A'}$ and $P_{A''}$,
+  we can define a push-pull structure $\piv_{c \comp c'}$ for $c \comp c'$.
+
+  Likewise, we can define a push-pull structure for the composition of computation relations.
 \end{lemma}
 \begin{proof}
+  We define $\piv_{c \comp c'}$ as the following push-pull structure:
+
+  \begin{itemize}
+    \item $\push_{c \comp c'} = \push_{c'} \circ \push_{c}$
+    \item $\pull_{c \comp c'} = \pull_{c} \circ \pull_{c'}$
+  \end{itemize}
+
+  We observe that the required squares exist for both push and pull.
+  In particular, for push we have that $\delta^l \ltdyn_c^c \push_c(\delta^l)$
+  using the push property for $c$, and then using the push property for $c'$ we have
+  $\push_c(\delta^l) \ltdyn_{c'}^{c'} \push_{c'}(\push_c(\delta^l))$.
+  We can then compose these squares horizontally to obtain the desired square.
+  The pull property follows similarly.
+
+  The push-pull structure for the composition of computation relations is defined analogously.
+\end{proof}
+
+%%%%%%%%%%%%%
+% UF and FU %
+%%%%%%%%%%%%%
+\begin{lemma}
+  Let $c : A \rel A'$ and let $P_A$ and $P_{A'}$ be monoids of perturbations.
+  Given a push-pull structure $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$,
+  we can define a push-pull structure $\pie_{UF(c)}$ for $UF(c)$ with respect to $UF(P_{A})$
+  and $UF(P_{A'})$.
+  % Eric: Are we assuming here that F and U preserve perturbations?
+\end{lemma}
+\begin{proof}
+  Follows by the functorial action of $UF$ and $FU$ on perturations and squares.
+\end{proof}
+
+%%%%%%%%%%%%
+% Products %
+%%%%%%%%%%%%
+\begin{lemma}
+  Let $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$, and let $P_{A_1}$,
+  $P_{A_2}$, $P_{A_1'}$ and $P_{A_2'}$ be monoids of perturbations.
+  Given push-pull structures 
+  $\piv_{c_1}$ with respect to $P_{A_1}$ and $P_{A_1'}$, and 
+  $\piv_{c_2}$ with respect to $P_{A_2}$ and $P_{A_2'}$,
+  we can define a push-pull structure $\piv_{c_1 \times c_2}$
+  with respect to $P_{A_1} \times P_{A_2}$ and $P_{A_1'} \times P_{A_2'}$.
+
+  Likewise, given push-pull structures 
+  $\pie_{Fc_1}$ with respect to $F(P_{A_1})$ and $F(P_{A_1'})$, and
+  $\pie_{Fc_2}$ with respect to $F(P_{A_2})$ and $F(P_{A_2'})$,
+  we can define a push-pull structure $\pie_{F(c_1 \times c_2)}$
+  with respect to $F(P_{A_1} \times P_{A_2})$ and $F(P_{A_1'} \times P_{A_2'})$.
+\end{lemma}
+\begin{proof}
+We define the push function for $\piv_{c_1 \times c_2}$ by 
+$\push(\delta_1, \delta_2) = (\push_{c_1}(\delta_1), \push_{c_2}(\delta_2))$ and likewise for $\pull$.
+The push-pull property holds because it holds for each component.
+\end{proof}
+
+%%%%%%%%%
+% Arrow %
+%%%%%%%%%
+\begin{lemma}
+  Let $c : A \rel A'$ and $d : B \rel B'$.
+  Given push-pull structures $\piv_c$ for $c$ and $\pie_d$ for $d$,
+  we can define a push-pull structure $\pie_{c \arr d}$ for $c \arr d$.
+
+  Given push-pull structures $\pie_{Fc}$ for $Fc$ and $\piv_{Ud}$ for $Ud$,
+  we can define a push-pull structure $\piv_{U(c \arr d)}$ for $U(c \arr d)$
+  with respect to the monoids $(P_{FA})^{op} \oplus P_B$
+\end{lemma}
+\begin{proof}
+  The former follows from the functorial action of $\arr$ on perturbations and squares,
+  and the latter uses the functorial action of $\tok$ on perturbations and squares.
+\end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%
+% Model Construction %
+%%%%%%%%%%%%%%%%%%%%%%
+We now proceed with the construction of the model:
     % Write 
@@
     \begin{itemize}
-      \item Value objects are tuples consisting of:
+      \item Value objects are triples consisting of:
       \begin{itemize}
         \item A value object $A$ in $\vf$ 
-        \item A monoid of ``pure'' perturbations $P_A$ 
-        \item A homomorphism of monoids $\ptb_A : P_A \to \{ f \in \vf(A, A) \mid f \bisim \id_A \}$
-        \item A monoid of ``impure'' perturbations $P^K_A$ that contains a distinguished element $\delta^*$
-        \item A homomorphism of monoids $\ptbk_A : P^K_A \to \{ \phi \in \ef(FA, FA) \mid \phi \bisim \id_{FA} \}$
-        such that $\ptbk_A(\delta^*) = \delta_A^*$
+        \item A monoid $P_A$ of ``pure'' perturbations that is a submonoid of 
+        $\{ f \in \vf(A, A) \mid f \bisim \id_A \}$
+        \item A monoid $P^K_A$ of ``impure'' or ``Kleisli'' perturbations that is a submonoid of
+        $\{ \phi \in \ef(FA, FA) \mid \phi \bisim \id_{FA} \}$ such that $P^K_A$ contains
+        $\delta_A^*$
       \end{itemize}
 
-      \item Computation objects are tuples consisting of:
+      \item Computation objects are triples consisting of:
       \begin{itemize}
         \item A computation object $B$ in $\ef$
-        \item A monoid of ``pure'' perturbations $P_B$
-        \item A homomorphism of monoids $\ptb_B : P_B \to \{ \phi \in \ef(B, B) \mid \phi \bisim \id_B \}$
-        \item A monoid of ``impure'' perturbations $P^K_B$
-        \item A homomorphism of monoids $\ptbk_B : P^K_B \to \{ g \in \vf(UB, UB) \mid g \bisim \id_{UB} \}$.
+        \item A monoid $P_B$ of ``pure'' perturbations that is a submonoid of
+        $\{ \phi \in \ef(B, B) \mid \phi \bisim \id_B \}$
+        \item A monoid $P^K_B$ of ``impure'' or ``Kleisli'' perturbations that is a submonoid
+        of $\{ g \in \vf(UB, UB) \mid g \bisim \id_{UB} \}$
       \end{itemize}
 
-      \item Morphisms are given by morphisms of the underlying objects in $\vf$ and $\ef$, respectively
+      \item Morphisms are given by morphisms of the underlying objects in $\vf$ and $\ef$, respectively.
       %, i.e.,
@@
 
-    Before introducing the relations, we make a definition.
+    % Before introducing the relations, we make a definition.
 
-    \begin{definition}[push-pull structure]
-      Let $c : A \rel A'$ be a value relation of $\mathcal M$. A \emph{value push-pull structure} $\piv_c$ for $c$ consists of:
+    % \begin{definition}[push-pull structure]
+    %   Let $c : A \rel A'$ be a value relation of $\mathcal M$. A \emph{value push-pull structure} $\piv_c$ for $c$ consists of:
+    %   \begin{itemize}
+    %     \item A function $\push : P_A \to P_{A'}$ 
+    %           such that for all $\delta^l \in P_A$ we have $\delta^l \ltdyn_c^c \push(\delta^l)$.
+    %     \item A function $\push^K : P^K_A \to P^K_{A'}$ 
+    %           such that for all $\delta^K_l \in P^K_A$ we have $\delta^K_l \ltdyn_{Fc}^{Fc} \push(\delta^K_l)$.
+    %     \item A function $\pull : P_{A'} \to P_A$
+    %           such that for all $\delta_r \in P_{A'}$ we have $\pull(\delta^r) \ltdyn_{c}^c \delta^r$.
+    %     \item A function $\pull^K : P^K_{A'} \to P^K_A$
+    %           such that for all $\delta^K_r \in P^K_{A'}$ we have $\pull(\delta^K_r) \ltdyn_{Fc}^{Fc} \delta^K_r$.
+    %   \end{itemize}
+
+    %   For $d : B \rel B'$ a computation relation, we define a \emph{computation push-pull structure} $\pie_d$ for $d$
+    %   in an analogous manner.
+    % \end{definition}
+
+    % Now we continue with the description of the construction:
     \begin{itemize}
-        \item A function $\push : P_A \to P_{A'}$ 
-              such that for all $\delta^l \in P_A$ we have $\delta^l \ltdyn_c^c \push(\delta^l)$.
-        \item A function $\push^K : P^K_A \to P^K_{A'}$ 
-              such that for all $\delta^K_l \in P^K_A$ we have $\delta^K_l \ltdyn_{Fc}^{Fc} \push(\delta^K_l)$.
-        \item A function $\pull : P_{A'} \to P_A$
-              such that for all $\delta_r \in P_{A'}$ we have $\pull(\delta^r) \ltdyn_{c}^c \delta^r$.
-        \item A function $\pull^K : P^K_{A'} \to P^K_A$
-              such that for all $\delta^K_r \in P^K_{A'}$ we have $\pull(\delta^K_r) \ltdyn_{Fc}^{Fc} \delta^K_r$.
+
+      \item The objects of $\vsq'$ (i.e., the value relations) are triples consisting of:
+      \begin{itemize}
+        \item A value relation $c \in \vsq$.
+        \item A push-pull structure $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$.
+        \item A push-pull structure $\pie_{Fc}$ for $Fc$ with respect to $P^K_A$ and $P^K_{A'}$.
       \end{itemize}
 
-      For $d : B \rel B'$ a computation relation, we define a \emph{computation push-pull structure} $\pie_d$ for $d$
-      in an analogous manner.
-    \end{definition}
-
-    Now we continue with the description of the construction:
+      The objects of $\esq'$ are defined analogously, i.e.,
       \begin{itemize}
-
-      \item The objects of $\vsq'$ (i.e., the value relations) are pairs consisting of:
-      \begin{itemize}
-        \item A value relation $c \in \vsq$
-        \item A push-pull structure $\piv_c$ for $c$
+        \item A computation relation $d \in \esq$.
+        \item A push-pull structure $\pie_d$ for $d$ with respect to $P_B$ and $P_{B'}$.
+        \item A push-pull structure $\piv_{Ud}$ for $Ud$ with respect to $P^K_B$ and $P^K_{B'}$.
       \end{itemize}
-
-      The objects of $\esq'$ are defined analogously.
 
@@
 
+      \item We define composition of relations $(c, \piv_c, \pie_{Fc})$ and 
+      $(c', \piv_{c'}, \pie_{Fc'})$ as $(c \comp c', \piv_{c \comp c'}, \pie_{Fc \comp Fc'})$
+      and likewise for computation relations.
+
+    \end{itemize}
+
       % Functors \times, +, F, U, arrow
+      Now we define the actions of the functors:
+      \begin{itemize}
+
+      % \item We begin with the functor $m$ that composes relations and squares horizontally.
+      % Given value relations $(c ,\piv_c, \pie_{Fc})$ and $(c', \piv_{c'}, \pie_{Fc'})$ where $c : A \rel A'$ and $c' : A' \rel A''$,
+      % we define their composition to be the triple $(c \comp c', \piv_{c \comp c'})$ where we define
+      % $\piv_{c \comp c'}$ as the following push-pull structure:
+
+      % \begin{itemize}
+      %   \item $\push_{c \comp c'} = \push_{c'} \circ \push_{c}$
+      %   \item $\push^K_{c \comp c'} = \push^K_{c'} \circ \push^K_{c}$
+      %   \item $\pull_{c \comp c'} = \pull_{c} \circ \pull_{c'}$
+      %   \item $\pull^K_{c \comp c'} = \pull^K_{c} \circ \pull^K_{c'}$
+      % \end{itemize}
+
+      % We observe that each of the components satisfies the commuting necessary square.
+
+      \item We define $\times$ on objects by
+
+      \[ (A_1, P_{A_1}, P^K_{A_1}) \times (A_2, P_{A_2}, P^K_{A_2}) =
+        (A_1 \times A_2, P_{A_1} \times P_{A_2}, P^K_{A_1} \oplus P^K{A_2}). \]
+
+      where $\oplus$ is the coproduct in the category of monoids.
+
+      We define $\times$ on relations by
+
+      \[ (c_1, \piv_{c_1}, \pie_{Fc_1}), (c_2 \piv_{c_2}, \pie_{Fc_2}) = 
+        (c_1 \times c_2, \piv_{c_1 \times c_2}, \pie_{F(c_1 \times c_2)}). \]
+
+      \item We define $F$ on objects by
+
+      \[ F(A, P_A, P^K_A) = (FA, P^K_A, UF(P_A)). \]
+
+      We define $F$ on relations by
+
+      \[ F(c, \piv_c, \pie_{Fc}) = (Fc, \pie_{Fc}, \piv_{UF(c)}). \]
+
+      \item We define $U$ on objects by
+
+      \[ U(B, P_B, P^K_B) = (UB, P^K_B, FU(P_B). )\]
+
+      We define $U$ on relations by
+
+      \[ U(d, \pie_d, \piv_{Ud}) = (Ud, \piv_{Ud}, \pie_{FU(d)})  \]
+
+      We define $\arr$ on objects by
+
+      \[ (A, P_A, P^K_A) \arr (B, P_B, P^K_B) = 
+        (A \arr B, (P_A)^{op} \times P_B, (P^K_A)^{op} \oplus P^K_B). \]
+
+      We define $\arr$ on relations by
+
+      \[ (c, \piv_c, \pie_{Fc}) \arr (d, \pie_d, \piv_{Ud}) =
+        (c \arr d, \pie_{c \arr d}, \piv_{U(c \arr d)}). \]
 
@@
     \end{itemize}
-\end{proof}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Constructing a Model with Quasi-Representable Relations}
+
+The goal of this section is to prove the following lemma:
 
@@
 \end{lemma}
-\begin{proof}
-  Write 
-  %
-  \[ \mathcal M = (\vf, \vsq, \ef, \esq, \Ff, \Fsq, \Uf, \Usq, \arrf, \arrsq). \] 
-  %
 
-  We begin with a definition.
+Before proceeding with the proof, we begin with a definition.
 
 \begin{definition}[representation structure]
-  Let $c : A \rel A'$ be a value relation. A \emph{left-representation structure} $\rho^L_c$ for $c$ consists of
-  a value morphism $e_c \in \vf(A, A')$ such that $c$ is quasi-left-representable by $e_c$ (see Definition \ref{def:quasi-left-representable}).
+Let $c : A \rel A'$ be a value relation. A \emph{left-representation structure}
+$\rho^L_c$ for $c$ consists of a value morphism $e_c \in \vf(A, A')$ such that
+$c$ is quasi-left-representable by $e_c$ (see Definition \ref{def:quasi-left-representable}).
 
-  Likewise, let $d : B \rel B'$. A \emph{right-representation structure} $\rho^R_d$ for $d$ consists of
-  a computation morphism $p_d \in \ef(B', B)$ such that $d$ is quasi-right-representable by $p_d$ (see Definition \ref{def:quasi-right-representable}).
+Likewise, let $d : B \rel B'$ be a computation relation. A \emph{right-representation structure}
+$\rho^R_d$ for $d$ consists of a computation morphism $p_d \in \ef(B', B)$
+such that $d$ is quasi-right-representable by $p_d$ (see Definition
+\ref{def:quasi-right-representable}).
 \end{definition}
 
-  (Notice that the direction of the morphism is opposite in the definition of right-representation structure.)
+(Notice that the direction of the morphism is opposite in the definition of
+right-representation structure.)
+
+% TODO establish the notation
+
+%%%%%%%%%%%%%%%
+% Composition %
+%%%%%%%%%%%%%%%
+\begin{lemma}\label{lem:representation-comp}
+  In the below, let $c : A \rel A'$ and $c' : A' \rel A''$ and $d : B \rel B'$ and $d' : B' \rel B''$.
+  \begin{enumerate}
+    \item Given left-representation structures $\rho^L_c$ for $c$ and $\rho^L_{c'}$ for $c'$,
+    we can define a left-representation structure for the composition $c \comp c'$
+    \item Given right-representation structures $\rho^R_d$ for $d$ and $\rho^R_{d'}$ for $d'$,
+    we can define a right-representation structure for the composition $d \comp d'$
+    \item Given right-representation structures $\rho^R_{Fc}$ for $Fc$ and $\rho^R_{Fc'}$ for $Fc'$,
+    we can define a right-representation structure $\rho^R_{F(c \comp c')}$ for $F(c \comp c')$.
+    \item Given left-representation structures $\rho^L_{Ud}$ for $Ud$ and $\rho^L_{Ud'}$ for $Ud'$,
+    we can define a left-representation structure $\rho^L_{U(d \comp d')}$ for $U(d \comp d')$.
+  \end{enumerate}
+  \eric{Do we need parts 3 and 4?}
+
+\end{lemma}
+\begin{proof}
+  \begin{enumerate}
+
+    % 1.
+    \item We define $\rho^L_{c \comp c'}$ as follows. In the definitions of the perturbations,
+    we make use of the fact that $c$ and $c'$ satisfy the push-pull property.
+    \begin{itemize}
+      \item $e_{c \comp c'} = e_{c'} \circ e_c$
+      \item $\delre_{c \comp c'} = \delre_{c'} \circ \push_{c'}(\delre_c)$
+      \item $\delle_{c \comp c'} = \pull_c(\delle_{c'}) \circ \delle_c$
+      \item $\upl$ is the following square: \input{squares/UpL-comp}
+
+      The square $(*)$ exists by the push-pull property for $c'$, and the square $(**)$ exists
+      because $r$ is a unit for horizontal composition, so $r(A') \comp c' = c'$, and so this
+      is simply the identity square $\id_{c'} \in \vsq(c', c')$.
+
+      \item $\upr$ is the following square: \input{squares/UpR-comp}
+    \end{itemize}
+
+    % 2.
+    \item We define $\rho^R_{d \comp d'}$ as follows:
+    \begin{itemize}
+      \item $p_{d \comp d'} = p_d \circ p_{d'}$
+      \item $\dellp_{d \comp d'} = \dellp_d \circ \pull_d(\dellp_{d'})$
+      \item $\delrp_{d \comp d'} = \push_{d'}(\delrp_d) \circ \delrp_{d'}$
+      \item $\dnr$ is the following square: \input{squares/DnR-comp}
+
+      Here the square $(*)$ exists by the push-pull property for $d$.
+
+      \item $\dnl$ is the following square: \input{squares/DnL-comp}
+    \end{itemize}
+
+    % 3.
+    \item We define $\rho^R_{F(c \comp c')}$ as follows:
+    \begin{itemize}
+      \item $p_{F(c \comp c')} = p_{Fc} \circ p_{Fc'}$ % : FA'' \to FA
+      \item $\dellp_{F(c \comp c')} = $
+      \item $\delrp_{F(c \comp c')} = $
+    \end{itemize}
+
+    % 4.
+    \item We define $\rho^L{(U(d \comp d'))}$ as follows:
+
+  \end{enumerate}
+\end{proof}
+
+% Want to show: U(d \comp d') is weakly equivalent to U(d) \comp U(d').
+% This holds because both are quasi-representable by the same projection
+% \begin{lemma}
+%   Let $\mathcal M$ be a step-3 intensional model, and let
+%   $d : B \rel B'$ and $d' : B' \rel B''$.
+%   Let $\rho_\text{comp}$ be an arbitrary right-representation for $d \comp d'$.
+%   Then the projection $p_{d \circ d'}$ is weakly equivalent to
+% \end{lemma}
+
+%%%%%%%%%%%%%
+% UF and FU %
+%%%%%%%%%%%%%
+\begin{lemma}\label{lem:representation-UF-FU}
+  Let $c : A \rel A'$, and let $\rho^L_c$ be a left-representation structure for $c$.
+  Then we can define a left-representation structure $\rho^L_{UF(c)}$ for $UF(c)$.
+
+  Similarly, let $d : B \rel B'$ and let $\rho^R_d$ be a right-representation structure for $d$.
+  Then we can define a right-representation structure $\rho^R_{FU(d)}$ for $FU(d)$.
+\end{lemma}
+\begin{proof}
+  We define $\rho^L_{UF(c)}$ as follows:
+  \begin{itemize}
+    \item $e_{UF(c)} = UF(e_c)$
+    \item $\delre_{UF(c)} = UF(\delre_c)$ (which is in the monoid of perturbations of $UF(A')$ because the
+    perturbation monoids are closed under the actions of the functors $F$ and $U$)
+    \item $\delle_{UF(c)} = UF(\delle_c)$
+    \item We get the two commuting squares by the functorial action of $UF$ on the two squares for $c$,
+    i.e., $\upr_{UF(c)} = UF(\upr_c)$
+  \end{itemize}
+
+  We define $\rho^R_{FU(d)}$ in a similar manner.
+\end{proof}
+
+%%%%%%%%%%%%
+% Products %
+%%%%%%%%%%%%
+\begin{lemma}\label{lem:representation-product}
+  Let $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$.
+  Let $\rho^L_{c_1}$ be a left-representation structure for $c_1$, and
+  let $\rho^L_{c_2}$ be a left-representation structure for $c_2$.
+  Then we can define a left-representation structure for $c_1 \times c_2$.
+
+  Likewise, let $\rho^R_{Fc_1}$ and $\rho^R_{Fc_2}$ be right-representation
+  structures for $Fc_1$ and $Fc_2$ respectively.
+  Then we can deinfe a right-representation structure for $F(c_1 \times c_2)$.
+\end{lemma}
+\begin{proof}
+  We define $\rho^L_{c_1 \times c_2}$ as follows:
+  \begin{itemize}
+    \item $e_{c_1 \times c_2} = e_{c_1} \times e_{c_2}$
+    \item $\delre_{c_1 \times c_2} = \delre_{c_1} \times \delre_{c_2}$ and
+    likewise for $\delle_{c_1 \times c_2}$
+    \item We get the commuting squares via the functorial action of $\times$
+    on the corresponding squares for $c_1$ and $c_2$.
+  \end{itemize}
+
+  % TODO check this
+  We define $\rho^R_{F(c_1 \times c_2)}$ as follows:
+  \begin{itemize}
+    \item $p_{F(c_1 \times c_2)} = (p_{Fc_1} \timesk A_2) \circ (A_1' \timesk p_{Fc_2})$
+    \item $\dellp_{F(c_1 \times c_2)} = (\dellp_{Fc_1} \timesk A_2) \circ (A_1 \timesk \dellp_{Fc_2})$
+    \item $\delrp_{F(c_1 \times c_2)} = (\delrp_{Fc_1} \timesk A_2') \circ (A_1' \timesk \delrp_{Fc_2})$
+    \item The commuting squares are obtained via the functorial action of $\timesk$ on the squares for $Fc_1$ and $Fc_2$.
+    % TODO elaborate?
+  \end{itemize}
+\end{proof}
+
+%%%%%%%%%
+% Arrow %
+%%%%%%%%%
+\begin{lemma}\label{lem:representation-arrow}
+  Let $c : A \rel A'$ and $d : B \rel B'$.
+  Let $\rho^L_c$ be a left-representation structure for $c$,
+  and let $\rho^R_d$ be a right-representation structure for $d$.
+  Then we can define a right-representation structure for $c \arr d$.
+
+  Likewise, let $\rho^R_{Fc}$ be a right-representation structure for $Fc$,
+  and let $\rho^L_{Ud}$ be a left-representation structure for $Ud$.
+  Then we can define a left-representation structure for $U(c \arr d)$.
+\end{lemma}
+\begin{proof}
+  We define $\rho^R_{c \arr d}$ as follows:
+  \begin{itemize}
+    \item $p_{c \arr d} = e_c \arr p_d \in \ef(A' \arr B', A \arr B)$ 
+    (using the functorial action of $\arr$ on morphisms).
+    \item $\dellp_{c \arr d} = \delle_c \arr \dellp_d$
+    \item $\delrp_{c \arr d} = \delre_c \arr \delrp_d$
+    \item The squares $\dnr$ and $\dnl$ are obtained via the functorial action of $\arr$, i.e., we define
+
+    \[ \dnr_{c \arr d} = \upr_{c} \arr \dnr_{d} : 
+      (\delle_c \arr \dellp_d) \ltdyn_{r(A \arr B)}^{c \arr d} (e_c \arr p_d),  \]
+
+    and
+
+    \[ \dnl_{c \arr d} = \upl_{c} \arr \dnl_{d}. \]
+
+  \end{itemize}
+
+  % TODO check this
+  We define $\rho^L_{U(c \arr d)}$ as follows:
+  \begin{itemize}
+    \item $e_{U(c \arr d)} = (p_{Fc} \tok B') \circ (A \tok e_{Ud})$
+    \item $\delre_{U(c \arr d)} = (\delrp_{Fc} \tok B') \circ (A' \tok \delre_{Ud})$
+    \item $\delle_{U(c \arr d)} = (\dellp_{Fc} \tok B) \circ (A \tok \delle_{Ud})$
+    \item The squares $\upl$ and $\upr$ are obtained via the functorial action of $\tok$.
+    For instance, $\upl$ is given by the following square:
+
+    % https://q.uiver.app/#q=WzAsNixbMCwwLCJVKEEgXFx0byBCKSJdLFsyLDAsIlUoQScgXFx0byBCJykiXSxbMiwxLCJVKEEnIFxcdG8gQicpIl0sWzIsMiwiVShBJyBcXHRvIEInKSJdLFswLDEsIlUoQSBcXHRvIEInKSJdLFswLDIsIlUoQScgXFx0byBCJykiXSxbMCwxLCJVKGMgXFx0byBkKSJdLFs0LDIsIlUoYyBcXHRvIHIoQicpKSJdLFs1LDMsIlUocihBJykgXFx0byByKEInKSkiXSxbNCw1LCJwX3tGY30gXFx0b2sgQiciLDJdLFsxLDIsIkEnIFxcdG9rIFxcZGVscmVfe1VkfSJdLFsyLDMsIlxcZGVscnBfe0ZjfSBcXHRvayBCJyJdLFswLDQsIkEgXFx0b2sgZV97VWR9IiwyXSxbMTIsMTAsIlxcaWRfe0ZjfSBcXHRvayBcXHVwbF97VWR9IiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzksMTEsIlxcZG5sX3tGY30gXFx0b2sgXFxpZF97cihCJyl9IiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+    \[\begin{tikzcd}[ampersand replacement=\&,row sep=large]
+      {U(A \to B)} \&\& {U(A' \to B')} \\
+      {U(A \to B')} \&\& {U(A' \to B')} \\
+      {U(A' \to B')} \&\& {U(A' \to B')}
+      \arrow["{U(c \to d)}", from=1-1, to=1-3]
+      \arrow["{U(c \to r(B'))}", from=2-1, to=2-3]
+      \arrow["{U(r(A') \to r(B'))}", from=3-1, to=3-3]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "{p_{Fc} \tok B'}"', from=2-1, to=3-1]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "{A' \tok \delre_{Ud}}", from=1-3, to=2-3]
+      \arrow[""{name=2, anchor=center, inner sep=0}, "{\delrp_{Fc} \tok B'}", from=2-3, to=3-3]
+      \arrow[""{name=3, anchor=center, inner sep=0}, "{A \tok e_{Ud}}"', from=1-1, to=2-1]
+      \arrow["{\id_{Fc} \tok \upl_{Ud}}"{description}, draw=none, from=3, to=1]
+      \arrow["{\dnl_{Fc} \tok \id_{r(B')}}"{description}, draw=none, from=0, to=2]
+    \end{tikzcd}\]
+  \end{itemize}
+
+  The construction of $\upr$ is similar.
+\end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%
+% Model Construction %
+%%%%%%%%%%%%%%%%%%%%%%
+Now we can give the proof of the main lemma:
+  % Write 
+  % %
+  % \[ \mathcal M = (\vf, \vsq, \ef, \esq, \Ff, \Fsq, \Uf, \Usq, \arrf, \arrsq). \] 
+  % %
 
@@
     \item The value and computation morphisms in $\mathcal M'$ are the same as those of $\mathcal M$.
-    \item A value relation is defined to be a tuple $(c, \rho^L_c, \rho^R_{Fc})$ where:
+    \item A value relation is defined to be a tuple $(c, \rho^L_c, \rho^R_{Fc})$ with
     \begin{itemize}
-      \item $c$ is a value relation in $\mathcal M$, and 
+      \item $c$ a value relation in $\mathcal M$,
       \item $\rho^L_c$ a left-representation structure for $c$, and
-      \item $\rho^R_{Fc}$ a right-representation structure for $Fc$.
+      \item $\rho^R_{Fc}$ a right-representation structure for $Fc$
     \end{itemize}
@@
     \begin{itemize}
-      \item $d$ a computation relation in $\mathcal M$
-      \item $\rho^R_d$ a right-representation structure for $d$
-      \item $\rho^L_{Ud}$ a left-representation structure for $Ud$
+      \item $d$ a computation relation in $\mathcal M$,
+      \item $\rho^R_d$ a right-representation structure for $d$, and
+      \item $\rho^L_{Ud}$ a left-representation structure for $Ud$.
     \end{itemize}
@@
   \end{itemize}
-\end{proof}
 
-% Now we define the functors $F$, $U$, $\times$, and $\arr$.
+We define the functor $m$ which composes relations and squares horizontally as follows:
+Let $c : A \rel A'$ and $c' : A' \rel A''$. We let
 
-% On objects, the behavior is the same as the respective functors in $\mathcal M$.
+\begin{align*} 
+  m ((c, \rho^L_c, \rho^R_{Fc}), (c', \rho^L_{c'}, \rho^R_{Fc'})) = \\
+  (c \comp c', \rho^L_{c \comp c'}, \rho^R_{F(c \comp c')}). 
+\end{align*}
 
-% For relations, we define 
-% $\Fsq' (c, \rho^L_c, \rho^R_{Fc}) = (\Fsq c, \rho^R_{Fc}, UF(\rho^L_c))$ and
-% $\Usq' (d, \rho^R_d, \rho^L_{Ud}) = (\Usq d, \rho^L_{Ud}, FU(\rho^R_d))$.
+Now we define the functors $F$, $U$, $\times$, and $\arr$.
+On objects, the behavior is the same as the respective functors in $\mathcal M$.
+For relations, we define
+
+\[ \Fsq^{\mathcal M'} (c, \rho^L_c, \rho^R_{Fc}) = 
+  (\Fsq^M c, \rho^R_{Fc}, \rho^L_{UF(c)}), \]
+and
+\[ \Usq^{\mathcal M'} (d, \rho^R_d, \rho^L_{Ud}) = 
+  (\Usq^M d, \rho^L_{Ud}, \rho^R_{FU(d)}), \]
+
+where $\rho^L_{UF(c)}$ and $\rho^R_{FU(d)}$ are as defined in the proof of Lemma
+\ref{lem:representation-UF-FU}.
+
+We define \[ (c_1, \rho^L_{c_1}, \rho^R_{Fc_1}) \times (c_2, \rho^L_{c_2}, \rho^R_{Fc_2}) = 
+             (c_1 \times c_2, \rho^L_{c_1 \times c_2}, \rho^R_{F(c_1 \times c_2)}), \]
+
+where $\rho^L_{c_1 \times c_2}$ and $\rho^R_{F(c_1 \times c_2)}$ are as defined in the
+proof of Lemma \ref{lem:representation-product}.
+
+Lastly, we define
+
+\[ (c, \rho^L_c, \rho^R_{Fc}) \arr (d, \rho^R_d, \rho^L_{Ud}) =
+    (c \arr d, \rho^R_{c \arr d}, \rho^L_{U(c \arr d)}), \]
+
+where $\rho^R_{c \arr d}$ and $\rho^L_{U(c \arr d)}$ are as defined in the proof
+of Lemma \ref{lem:representation-arrow}.
 
@@
 
-% We now verify that the construction meets the requirements of a step-3 model.
-% First, we check that composition of value relations (resp. computation relations)
-% is well-defined.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Defining an Extensional Model}\label{sec:extensional-construction-appendix}
+
+We aim to prove the following lemma:
 
@@
 
-  % More formally, we define an extensional model $\mathcal M_e$ as follows.
-  % \begin{itemize}
-  %   \item 
-  % \end{itemize}
+  Recall the extensional model defined in Section \ref{sec:extensional-model-definition}.
+
+  We first establish the representability properties for this model.
+  We show the left-representability squares; the right-representability squares are dual.
+
+  \begin{itemize}
+    \item We have the square
+
+    % https://q.uiver.app/#q=WzAsNixbMCwwLCJBIl0sWzAsMSwiQSciXSxbNCwxLCJBX3IiXSxbNCwwLCJBX3IiXSxbMiwwLCJBJyJdLFsyLDEsIkEnIl0sWzAsNCwiYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzQsMywiY19yIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJlX2MiLDIseyJjdXJ2ZSI6MX1dLFszLDIsIlxccHVzaF97Y19yfShcXGRlbHJlX2MpIiwyLHsiY3VydmUiOjF9XSxbMSw1LCJyKEEnKSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzQsNSwiXFxkZWxyZV9jIiwyXSxbNSwyLCJjX3IiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsImVfYyIsMCx7ImN1cnZlIjotMX1dLFszLDIsIlxcaWQiLDAseyJjdXJ2ZSI6LTF9XSxbOSwxNCwiXFxiaXNpbSIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs4LDEzLCJcXGJpc2ltIiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzE2LDExLCIiLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJsZXZlbCI6Miwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+    \[\begin{tikzcd}[ampersand replacement=\&]
+      A \&\& {A'} \&\& {A_r} \\
+      {A'} \&\& {A'} \&\& {A_r}
+      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-3]
+      \arrow["{c_r}", "\shortmid"{marking}, no head, from=1-3, to=1-5]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "{e_c}"', curve={height=6pt}, from=1-1, to=2-1]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "{\push_{c_r}(\delre_c)}"', curve={height=6pt}, from=1-5, to=2-5]
+      \arrow["{r(A')}"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
+      \arrow[""{name=2, anchor=center, inner sep=0}, "{\delre_c}"', from=1-3, to=2-3]
+      \arrow["{c_r}"', "\shortmid"{marking}, no head, from=2-3, to=2-5]
+      \arrow[""{name=3, anchor=center, inner sep=0}, "{e_c}", curve={height=-6pt}, from=1-1, to=2-1]
+      \arrow[""{name=4, anchor=center, inner sep=0}, "\id", curve={height=-6pt}, from=1-5, to=2-5]
+      \arrow["\bisim"{description}, draw=none, from=1, to=4]
+      \arrow[""{name=5, anchor=center, inner sep=0}, "\bisim"{description}, draw=none, from=0, to=3]
+      \arrow[draw=none, from=5, to=2]
+    \end{tikzcd}\]
+
+    \item We have the square
+
+    % https://q.uiver.app/#q=WzAsNixbMCwwLCJBX2wiXSxbMCwxLCJBX2wiXSxbMiwxLCJBIl0sWzQsMSwiQSciXSxbMiwwLCJBIl0sWzQsMCwiQSJdLFswLDQsImNfbCJdLFs0LDUsInIoQSkiXSxbMSwyLCJjX2wiLDJdLFsyLDMsImMiLDJdLFswLDEsIlxcaWQiLDIseyJjdXJ2ZSI6MX1dLFs1LDMsImVfYyIsMix7ImN1cnZlIjoxfV0sWzQsMiwiXFxkZWxsZV9jIl0sWzAsMSwiXFxwdWxsX3tjX2x9KFxcZGVsbGVfYykiLDAseyJjdXJ2ZSI6LTF9XSxbNSwzLCJlX2MiLDAseyJjdXJ2ZSI6LTF9XSxbMTAsMTMsIlxcYmlzaW0iLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMTEsMTQsIlxcYmlzaW0iLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
+    \[\begin{tikzcd}[ampersand replacement=\&]
+      {A_l} \&\& A \&\& A \\
+      {A_l} \&\& A \&\& {A'}
+      \arrow["{c_l}", from=1-1, to=1-3]
+      \arrow["{r(A)}", from=1-3, to=1-5]
+      \arrow["{c_l}"', from=2-1, to=2-3]
+      \arrow["c"', from=2-3, to=2-5]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "\id"', curve={height=6pt}, from=1-1, to=2-1]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "{e_c}"', curve={height=6pt}, from=1-5, to=2-5]
+      \arrow["{\delle_c}", from=1-3, to=2-3]
+      \arrow[""{name=2, anchor=center, inner sep=0}, "{\pull_{c_l}(\delle_c)}", curve={height=-6pt}, from=1-1, to=2-1]
+      \arrow[""{name=3, anchor=center, inner sep=0}, "{e_c}", curve={height=-6pt}, from=1-5, to=2-5]
+      \arrow["\bisim"{description}, draw=none, from=0, to=2]
+      \arrow["\bisim"{description}, draw=none, from=1, to=3]
+    \end{tikzcd}\]
+
+\end{itemize}
+
 \end{proof}
@@
 
-First we establish some notation. Fix a morphism $f : 1 \to \li \Nat \cong \li \Nat$.
+First we establish some notation. Fix a morphism $f : 1 \to \li \mathbb{N} \cong \li \mathbb{N}$.
 We write that $f \da n$ to mean that there exists $m$ such that $f = \delta^m(\eta n)$
@@
 of the intensional error-ordering on morphisms.
-More concretely, we have $f \ltls g$ iff there exists $f'$ and $g'$ with
+That is, we have $f \ltls g$ iff there exists $f'$ and $g'$ with
 
@@
 \begin{lemma}
-If $f \ltls g : \li \Nat$, then:
+If $f \ltls g : \li \mathbb{N}$, then:
 \begin{itemize}
@@
 
-Unfortunately, this result is actually not provable!
+Unfortunately, this is actually not provable!
 Roughly speaking, the issue is that this is a ``global'' result, and it is not possible
 to prove such results inside of the guarded setting. 
-In particular, if we tried to prove a result such as the above in the guarded setting,
-we would run into a problem where we would have a natural number ``stuck'' under a $\later$
-with no way to get at the underlying number.
+In particular, if we tried to prove a result such as the above in the guarded
+setting, we would run into a problem where we would have a natural number
+``stuck'' under a $\later$, with no way to get out the underlying number.
 
 Thus, to prove our adequacy result, we need to leave the guarded setting and pass back
-to the normal set-theoretic world.
+to the more familiar, set-theoretic world with no internal notion of step-indexing.
 As mentioned in the Technical Background section (Section \ref{sec:sgdt}), we can do this
@@
 Recall that all of the constructions we have made in SGDT take place in the context of a clock $k$.
-All of our uses of the later modality and guarded recursion happen with respect to this clock.
+All of our uses of the later modality and guarded recursion have taken place with respect to this clock.
 For example, consider the definition of the lift monad by guarded recursion in Section \ref{TODO}.
@@
 We can view this definition as being parameterized by a clock $k$: $\li^k : \type \to \type$.
-Then for $X$ satisfying a certain technical requirement, we can define the ``global lift'' monad as $\li^{gl} X = \forall k. \li^k X$.
+Then for $X$ satisfying a certain technical requirement known as \emph{clock-irrelevance},
+we can define the ``global lift'' monad as $\li^{gl} X := \forall k. \li^k X$.
+
+Similarly, we can define a ``global'' version of the lock-step error ordering and the
+weak bisimilarity relation on morphisms.
 
 It can be shown that the global lift monad is isomorphic to the so-called Delay monad of Capretta \cite{TODO}.
+Recall that the delay monad $\text{Delay}(X)$ is defined as the coinductive solution to the equation $\text{Delay}(X) \cong X + \text{Delay}(X)$.
+It can be shown that $\li^{gl} X \cong \text{Delay}(X + 1)$ for $X$ a clock-irrelevant type.
+
+Moreover, we claim that modulo this isomorphism, the global version of the extensional error ordering
 

### 2024-01-24 paper-new/appendix.tex
mostly writing updates

@@
 \end{mathpar}
+Where the cast $M :: A_2$ is defined to be
+\[ \dnc {dyn(A_2)}{\upc {dyn(A)} M} \]
+
 These two rules are admissible from the following principles:
+%% \begin{mathpar}
+%% \end{mathpar}
 
-%% \begin{mathpar}
-%%   \inferrule
-%%   {}
-%%   {\dnc {\injarr{}} \upc {\injarr{}} M \equidyn M}
+For the first rule, we first prove that $\dnc {dyn(A_2)}\upc {dyn(A)} M = \dnc {c'}\upc{c} M$
+\begin{align*}
+  \dnc {dyn(A_2)}\upc {dyn(A)} M
+  &= \dnc {c \,\textrm{dyn}(A')}\upc{c' \,\textrm{dyn}(A')} M \tag{All derivations are equal}\\
+  &= \dnc {c}\dnc {\textrm{dyn}(A')}\upc{\textrm{dyn}(A')}\upc{c'} M\tag{functoriality}\\
+  &= \dnc {c}\upc{c'} M\tag{retraction}\\
+\end{align*}
 
-%%   %% \inferrule
-%%   %% {}
-%%   %% {}
-%% \end{mathpar}
+Then the rest follows by the up/dn rules above and the fact that precision derivations are all equal.
+\begin{mathpar}
+  \inferrule*
+  {c' = \textrm{dyn}(A_2)\max{todo}}
+  {\dnc {c'}\upc {c} M \ltdyn M' : c'}
+\end{mathpar}
+
+Thus the following properties are sufficient to provide an extensional
+model of gradual typing without requiring transitivity of term
+precision:
+\begin{enumerate}
+\item Every precision is representable in the above sense,
+\item The association of casts to precision is functorial
+\item Type constructors are covariant functorial with respect to
+  relational identity and composition
+\end{enumerate}
 

### 2024-01-24 paper-new/categorical-model.tex
mostly writing updates

@@
 \begin{enumerate}
-  \item A CBPV model internal to reflexive graphs.
-  \item Composition and identity on value and computation relations that form a category.
+  \item A reflexive graph internal to CBPV models.
+  \item Composition of value and computation relations that form a category with the reflexive relations as identity.
   \item An identity-on-objects functor $\upf : \mathcal V_e \to \mathcal V_f$ taking each value relation to a morphism that left-represents it.
   \item An identity-on-objects functor $\dnf : \mathcal E_e^{op} \to \mathcal E_f$ taking each computation relation to a morphism that right-represents it.
-  \item The CBPV connectives $U,F,\times,\to$ are all \emph{covariant} functorial on relations
+  \item The CBPV connectives $U,F,\times,\to$ are all \emph{covariant} functorial on relations\footnote{the reflexive graph structure already requires that these functors preserve identity relations}
     \begin{itemize}
@@
 \end{enumerate}
+
+\subsection{Stuff about Intensional Models}
 

### 2024-01-24 paper-new/discussion.tex
mostly writing updates

@@
 % Discuss Joey Eremondi's thesis on gradual dependent types
+
+% Discuss Jeremy Siek's work on graduality in Agda
 
@@
 
-While the use of synthetic guarded domain theory allows us to very
-conveniently work with non-well-founded recursive constructions while
-abstracting away the precise details of step-indexing, we do work with
-the error ordering in a mostly analytic fashion in that gradual types
-are interpreted as sets equipped with an ordering relation, and all
-terms must be proven to be monotone.
+A key to managing the complexity of our concrete construction is in
+using a \emph{synthetic} approach to step-indexing rather than working
+analytically with presheaves. This has helped immensely in our ongoing
+mechanization in cubical Agda as it sidesteps the need to formalize
+these constructions internally. 
 %
-It is possible that a combination of synthetic guarded domain theory
-with \emph{directed} type theory would allow for an a synthetic
-treatment of the error ordering as well.
+However, there are other aspects of the model, the bisimilarity and
+the monotonicity, which are treated analytically and are .
+%
+It may be possible to utilize further synthetic techniques to reduce
+this burden as well, and have all type intrinsically carry a notion of
+bisimilarity and ordering relation, and all constructions to
+automatically preserve them.
+%
+A synthetic approach to ordering is common in (non-guarded) synthetic
+domain theory and has also been used for synthetic reasoning for cost
+models \cite{synthetic-domain-theory,decalf}.
 

### 2024-01-24 paper-new/intro.tex
mostly writing updates

@@
 
-\subsection{Current Approaches to Proving Graduality}
+\subsection{Limitations of Prior Work}
+
+We give an overview of current approaches to proving graduality of
+languages and why they do not meet our criteria of a reusable semantic
+framework.
 
@@
 
-Current approaches to constructing languages that satisfy the graduality property
-include the methods of Abstracting Gradual Typing
-\cite{garcia-clark-tanter2016} and the formal tools of the Gradualizer \cite{cimini-siek2016}.
-These allow the language developer to start with a statically typed language and derive a
-gradually typed language that satisfies the gradual guarantee. The main downside to
-these approaches lies in their inflexibility: since the process in entirely mechanical,
-the language designer must adhere to the predefined framework.
-Many gradually typed languages do not fit into either framework, e.g., Typed Racket
-\cite{tobin-hochstadt06, tobin-hochstadt08}.
+Current approaches to constructing languages that satisfy the
+graduality property include the methods of Abstracting Gradual Typing
+\cite{garcia-clark-tanter2016} and the formal tools of the Gradualizer
+\cite{cimini-siek2016}.  These allow the language developer to start
+with a statically typed language and derive a gradually typed language
+that satisfies the gradual guarantee. The main downside to these
+approaches lies in their inflexibility: since the process in entirely
+mechanical, the language designer must adhere to the predefined
+framework.  Many gradually typed languages do not fit into either
+framework, e.g., Typed Racket \cite{tobin-hochstadt06,
+  tobin-hochstadt08} and the semantics produced is not always the
+desired one.
 %
-Furthermore, while these frameworks do prove graduality of the resulting languages,
-they do not show the correctness of the equational theory, which is equally important
-to sound gradual typing.
-For example, programmers often refactor their code, and in so doing they rely
-implicitly on the validity of the laws in the equational theory.
-Similarly, correctness of compiler optimizations rests on the validity of the
-corresponding equations from the equational theory. It is therefore important
-that the languages that claim to be gradually typed have provably correct
-equational theories.
+Furthermore, while these frameworks do prove graduality of the
+resulting languages, they do not show the correctness of the
+equational theory, which is equally important to sound gradual typing.
+
+%% For example, programmers often refactor their code, and in so doing they rely
+%% implicitly on the validity of the laws in the equational theory.
+%% Similarly, correctness of compiler optimizations rests on the validity of the
+%% corresponding equations from the equational theory. It is therefore important
+%% that the languages that claim to be gradually typed have provably correct
+%% equational theories.
 
@@
 % to here in the intro.
-\subsubsection{Classical Domain Models}
+\subsubsection{Double Categorical Semantics}
 
-New and Licata \cite{new-licata18} developed an axiomatic account of the
-graduality relation on cast calculus terms and gave a denotational
-model of this calculus using classical domain theory based on
-$\omega$-CPOs. This semantics has scaled up to an analysis of a
-dependently typed gradual calculus in \cite{asdf}. This meets our
-criterion of being a reusable mathematical theory, as general semantic
-theorems about gradual domains can be developed independent of any
-particular syntax and then reused in many different denotational
-models. However, it is widely believed that such classical domain
-theoretic techniques cannot be extended to model higher-order store, a
-standard feature of realistic gradually typed languages such as Typed
-Racket. Thus if we want a reusable mathematical theory of gradual
-typing that can scale to realistic programming languages, we need to
-look elsewhere to so-called ``step-indexed'' techniques.
+New and Licata \cite{new-licata18} developed an axiomatic account of
+the graduality relation on a call-by-name cast calculus terms and
+showed that the graduality proof could be modeled using semantics in
+certain kinds of \emph{double categories}, categories internal to the
+category of categories. A double category extends a category with a
+second notion of morphism, often a notion of ``relation'' to be paired
+with the notion of functional morphism, as well as a notion of
+functional morphisms preserving relations. In gradual typing the
+notion of relation models type precision and the squares model the
+term precision relation. This approach was influenced by Ma and
+Reynolds semantics of parametricity using reflexive graph categories:
+reflexive graph categories are essentially double categories without a
+notion of relational composition. In addition to capturing the notions
+of type and term precision, the double categorical approach allows for
+a \emph{universal property} for casts: upcasts are the
+\emph{universal} way to turn a relation arrow into a function in a
+forward direction and downcasts are the dual universal arrow.  Later,
+New, Licata and Ahmed \cite{new-licata-ahmed19} extended this
+axiomatic treatment from call-by-name to call-by-value as well by
+giving an axiomatic theory of type and term precision in
+call-by-push-value. This left implicit any connection to a ``double
+call-by-push-value'', which we make explicit in
+Section~\ref{sec:cbpv}.
 
-\subsubsection{Operational Reduction Proofs}
+With this notion of abstract categorical model in hand, denotational
+semantics is then the work of constructing concrete models that
+exhibit the categorical construction. New and Licata
+\cite{new-licata18} present such a model using categories of
+$\omega$-CPOs, and this model was extended by Lennon-Bertrand,
+Maillard, Tabareau and Tanter to prove graduality of a gradual
+dependently typed calculus $\textrm{CastCIC}^{\mathcalG}$. This
+domain-theoretic approach meets our criteria of being a semantic
+framework for proving graduality, but suffers from the limitations of
+classical domain theory: the inability to model viciously
+self-referential structures such as higher-order extensible state and
+similar features such as runtime-extensible dynamic types. Since these
+features are quite common in dynamically typed languages, we seek a
+new denotational framework that can model these type system features.
 
+The standard alternative to domain theory that scales to essentially
+arbitrary self-referential definitions is \emph{step-indexing} or its
+synthetic form of \emph{guarded recursion}.
+\max{todo: adapt the next paragraph to fit in with this one}
 A series of works \cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
@@
 make formalization of realistic gradual languages tractible.
+
+%% \subsubsection{Classical Domain Models}
+
+%%  and gave a denotational
+%% model of this calculus using classical domain theory based on
+%% $\omega$-CPOs. This semantics has scaled up to an analysis of a
+%% dependently typed gradual calculus in \cite{asdf}. This meets our
+%% criterion of being a reusable mathematical theory, as general semantic
+%% theorems about gradual domains can be developed independent of any
+%% particular syntax and then reused in many different denotational
+%% models. However, it is widely believed that such classical domain
+%% theoretic techniques cannot be extended to model higher-order store, a
+%% standard feature of realistic gradually typed languages such as Typed
+%% Racket. Thus if we want a reusable mathematical theory of gradual
+%% typing that can scale to realistic programming languages, we need to
+%% look elsewhere to so-called ``step-indexed'' techniques.
+
+%% \subsubsection{Operational Reduction Proofs}
 
@@
 \subsection{Contributions}
+
+The main contribution of this work is a categorical and denotational
+semantics for gradually typed langauges that models not just the term
+language but the graduality property as well.
+\begin{enumerate}
+\item First, we give a rational reconstruction \max{todo: more shit.}
+\end{enumerate}
+
 Our main contribution is a compositional denotational semantics for step-aware gradual typing,
+
 analogous to Gradual Type Theory but now in the ``intensional" setting.

### 2024-01-25 paper-new/categorical-model.tex
strict and lax homomorphisms of CBPV models

@@
 \item A left adjoint $F \dashv U$
-\item Distributive finite coproducts in $\mathcal V$
 \end{enumerate}
@@
   V$, we can extend this to a CBPV model by defining $\mathcal E$ to
-  be the category $\mathcal V^T$ of algebras
+  be the category $\mathcal V^T$ of algebras, defining $A \to B$ as
+  the powering of algebras, $F$ as the free algebra and $U$ as the
+  underlying object functor.
 \end{example}
 
+There are two relevant notions of \emph{morphism} of CBPV models:
+\emph{strict} and \emph{lax}.
 % morphisms of CBPV models
@@
 $\mathcal M_2 = (\mathcal V_2, \mathcal E_2, \arr_2, U_2, F_2)$,
-A morphism $G$ from $M_1$ to $M_2$ is given by a pair of functors
-$G_{\mathcal{V}}: V_1 \to V_2$ and $G_{\mathcal{E}} : E_1 \to E_2$ such that
-
+A \emph{strict} morphism $G$ from $M_1$ to $M_2$ is given by a pair of functors
+$G_{\mathcal{V}}: V_1 \to V_2$ and $G_{\mathcal{E}} : E_1 \to E_2$
+that strictly presere the constructors:
 \begin{enumerate}
@@
   \item $G_{\mathcal{E}}(A \arr_1 B) = G_{\mathcal{V}}(A) \arr_2 G_{\mathcal{E}}(B)$
+  \item $G_{\mathcal{V}}(A_1 \times_1 A_2) = G_{\mathcal{V}}(A_1) \times_2 G_{\mathcal{V}}(A_2)$
+  \item $G_{\mathcal{V}}(1_1) = 1_2$
 \end{enumerate}
+As well as strictly preserving the corresponding universal morphisms
+and coherence isomorphisms.
 
+A lax morphism instead preserves these only up to transformation
+\begin{enumerate}
+  \item $G_{\mathcal{E}} \circ F_1 \Rightarrow F_2 \circ G_{\mathcal{V}}$
+  \item $G_{\mathcal{V}} \circ U_1 \Rightarrow U_2 \circ G_{\mathcal{E}}$
+  \item $G_{\mathcal{E}}(A \arr_1 B) \Rightarrow G_{\mathcal{V}}(A) \arr_2 G_{\mathcal{E}}(B)$
+  \item $G_{\mathcal{V}}(A_1 \times_1 A_2) \Rightarrow G_{\mathcal{V}}(A_1) \times_2 G_{\mathcal{V}}(A_2)$
+  \item $G_{\mathcal{V}}(1_1) \Rightarrow 1_2$
+\end{enumerate}
+Additionally a lax morphism should have a relationship between these
+transformations and the universal morphisms, but we will only consider
+lax morphisms of thin categories, where such conditions hold
+trivially.
 % TODO what about the product and coproducts in V?
@@
 
-With these definitions, we can define a category whose objects are CBPV
-models, and whose arrows are CBPV morphisms.
-This category will serve as the setting in which we formulate the definition
-of a model of extensional gradual typing.
+With these definitions, we can define categories $\textrm{CBPV}_s$ and
+$\textrm{CBPV}_l$ whose objects are CBPV models, and whose arrows are
+CBPV morphisms.  This category will serve as the setting in which we
+formulate the definition of a model of extensional gradual typing.
 

### 2024-01-25 paper-new/discussion.tex
discussion on adgda work

@@
 % Discuss Guarded Cubical Agda and mechanization efforts
+We will use Guarded Cubical Agda to prove graduality in the syntax of 
+GTLC, which involves the construction of the abstract model described in 
+\ref{sec:concrete-model} and the extensional model with external dynamic 
+type. This part is still work-in-progress.
+
+% step-1
+Currently, we have defined the step-1 intensional model in Cubical 
+Agda. For value types, we constructed predomain which holds ordering and 
+bisimilarity, and morphisms between predmains. For computation types, we 
+constructed error domain which contains its underlying predomain, 
+elements of $\mho$ and $\theta$ and morphisms between error domains. We 
+also defined two functors $U$ and $F$ which support conversions between 
+predomains and error domains.
+
+% step-2 
+Then we plan to construct the step-2 intensional model. Besides all the 
+data in step-1, we need to include perturbations on value and computation 
+types, functors $\times$, $\arr$, $U$, and $F$ that preserve 
+perturbations and push/pull properties for all morphisms on value and 
+computation types. Notice that for any object $A$ which has value type, 
+we will take not only the monoid of perturbations $P^V_A$ and the monoid 
+homomorphism $\ptbv_A : \pv_A \to \vf(A,A)$ on itself, but also $P^C_{F A}
+$ and $\ptbe_{F A} : \pe_{F A} \to \ef(F A,F A)$ on $F A$, which have 
+computation types. Similarly, for any computation object $B$, we will 
+construct the perturbations on $U B$ besides the monoid $P^C_B$ and 
+monoid homomorphism $\ptbe_B : \pe_B \to \ef(B,B)$. Also, for functors 
+that preserves perturbations, we need to include the ones in the context 
+of Kleisli category.
+
+% step-3
+In the step-3 intensional model, we will enhance it with 
+quasi-representability. For any value relation $c : A \rel A'$, we need 
+to show that there exists a left-representation structure for $c$ and a 
+right-representation structure for $F\ c$. Correspondingly, for any 
+computation relation $d : B \rel B'$, we will show there exists a 
+right-representation structure for $d$ and a left-representation 
+structure for $U\ d$. As we define the quasi-representability for value 
+and computation relation, we will construct the quasi-representability on 
+the function and product of the relation, which makes it necessary to 
+have the dual version of quasi-representability.
+
+% step-4 construct a concrete dynamic type and apply it to the abstract model
+After defining the abstract model and its interface, we will model GTLC 
+by providing explicit construction triples of dynamic type at each step, 
+which includes defining Dyn as a predomain, its pure and Kleisli 
+perturbation monoids, push/pull property for pure and Kleisli 
+perturbation, as well as quasi-representability. The 
+quasi-representability involves explicit rules which show that Nat is 
+more precise than Dyn (Inj-Nat) and Dyn $\to$ Dyn is more precise than 
+Dyn (Inj-Arr). Currently, we have formalized the concrete construction of 
+Dyn in Cubical Agda and it was more challenging than expected because we 
+define Dyn using the technique of fixed point, which means that every 
+time we analyze the case inside of Dyn, we need to unfold it and add 
+corresponding proof.
+
+% adequacy
+Besides the abstract model and its concrete construction on dynamic type, 
+we will also formalize the adequacy result in \ref{sec:appendix-adequacy}, 
+which involves clock quantification of the lift monad, the weak bisim 
+relation, and the lock-step error ordering. In order to prove adequacy, 
+we will first prove that the global lift of X is isomorphic to Delay(1 + X)
+whether X is clock-irrelevant or not. Then, we aim to prove the equivalence 
+between the global lock-step error ordering and the error ordering observed 
+in Delay(1 + X) and equivalence between the global weak bisimilarity 
+relation and the weak bisimilarity relation on Delay(1 + X). We have 
+finished some prerequisite proofs on clock quantification and postulated 
+some theorems on clock globalization.
 

### 2024-01-25 paper-new/appendix.tex
fix build issues, progress on CBPV semantics

@@
 \section{Call-by-push-value}
+
+\subsection{Morphisms of CBPV Models}
+
+There are two relevant notions of \emph{morphism} of CBPV models:
+\emph{strict} and \emph{lax}.
+% morphisms of CBPV models
+Given call-by-push-value models
+$\mathcal M_1 = (\mathcal V_1, \mathcal E_1, \arr_1, U_1, F_1)$ and
+$\mathcal M_2 = (\mathcal V_2, \mathcal E_2, \arr_2, U_2, F_2)$,
+A \emph{strict} morphism $G$ from $M_1$ to $M_2$ is given by a pair of functors
+$G_{\mathcal{V}}: V_1 \to V_2$ and $G_{\mathcal{E}} : E_1 \to E_2$
+that strictly presere the constructors:
+\begin{enumerate}
+  \item $G_{\mathcal{E}} \circ F_1 = F_2 \circ G_{\mathcal{V}}$
+  \item $G_{\mathcal{V}} \circ U_1 = U_2 \circ G_{\mathcal{E}}$
+  \item $G_{\mathcal{E}}(A \arr_1 B) = G_{\mathcal{V}}(A) \arr_2 G_{\mathcal{E}}(B)$
+  \item $G_{\mathcal{V}}(A_1 \times_1 A_2) = G_{\mathcal{V}}(A_1) \times_2 G_{\mathcal{V}}(A_2)$
+  \item $G_{\mathcal{V}}(1_1) = 1_2$
+\end{enumerate}
+As well as strictly preserving the corresponding universal morphisms
+and coherence isomorphisms.
+
+A lax morphism instead preserves these only up to transformation
+\begin{enumerate}
+  \item $G_{\mathcal{E}} \circ F_1 \Rightarrow F_2 \circ G_{\mathcal{V}}$
+  \item $G_{\mathcal{V}} \circ U_1 \Rightarrow U_2 \circ G_{\mathcal{E}}$
+  \item $G_{\mathcal{E}}(A \arr_1 B) \Rightarrow G_{\mathcal{V}}(A) \arr_2 G_{\mathcal{E}}(B)$
+  \item $G_{\mathcal{V}}(A_1 \times_1 A_2) \Rightarrow G_{\mathcal{V}}(A_1) \times_2 G_{\mathcal{V}}(A_2)$
+  \item $G_{\mathcal{V}}(1_1) \Rightarrow 1_2$
+\end{enumerate}
+Additionally a lax morphism should have a relationship between these
+transformations and the universal morphisms, but we will only consider
+lax morphisms of thin categories, where such conditions hold
+trivially.
+
+\subsection{Kleisli Actions of CBPV Type Constructors}
 

### 2024-01-25 paper-new/intro.tex
fix build issues, progress on CBPV semantics

@@
 Maillard, Tabareau and Tanter to prove graduality of a gradual
-dependently typed calculus $\textrm{CastCIC}^{\mathcalG}$. This
+dependently typed calculus $\textrm{CastCIC}^{\mathcal G}$. This
 domain-theoretic approach meets our criteria of being a semantic

### 2024-01-26 paper-new/extensional-model.tex
prose on cbpv models

\section{What is a Model of Gradual Typing?}

In order to organize our construction of denotational models we first
develop sufficient \emph{abstract} categorical semantics of gradually
typed languages. We start by modeling the type and term structure of
gradual typing and then extend this to type and term precision.
%
Gradually typed languages inherently involve computational effects of
errors and non-termination and typically in practice many others such
as mutable state and I/O.
%
To model this cleanly categorically, we follow New, Licata and Ahmed's
GTT calculus and base our models off of Levy's Call-by-push-value
(CBPV) calculus which is a standard model of effectful programming
\cite{levycbpv}.
%
There are several notions of model of CBPV from the literature with
varying requirements of which connectives are present
\cite{levy,fioremm,eec}, we will use a variant which models precisely
the connectives we require and no more
($1,\times,F,U,\to$)\footnote{It is essential in this case that we do
not require a cartesian closed category of values as there is no way
to implement casts for an exponential in general.}.

\begin{enumerate}
\item A cartesian category $\mathcal V$
\item A category $\mathcal E$
\item An action of $\mathcal V^{op}$ (with the $\mathcal V$ cartesian
  product as monoidal structure) on $\mathcal E$. We write this with
  an arrow $A \arr B$.

  This means we have 
  \[ \alpha : {A_1 \times A_2} \arr B \cong A_2 \arr (A_1 \arr B) \]
  and
  \[ i : 1 \arr B \cong B \]
  satisfying coherence isomorphisms.
\item A functor $U : \mathcal E \to \mathcal V$ that ``preserves
  powering'' in that every $U(A \arr B)$ is an exponential of $UB$ by $A$
  and that $U\alpha$ and $Ui$ are mapped to the canonical isomorphisms
  for exponentials.
\item A left adjoint $F \dashv U$
\end{enumerate}

\begin{example}
  Given a strong monad $T$ on a bicartesian closed category $\mathcal
  V$, we can extend this to a CBPV model by defining $\mathcal E$ to
  be the category $\mathcal V^T$ of algebras of the monad, defining $A
  \to B$ as the powering of algebras, $F$ as the free algebra and $U$
  as the underlying object functor.
\end{example}

We can then model CBV terms and types in a straightforward adaptation
of Levy's interpretation of CBV in CBPV. We interpret types $A$ as
objects $A \in \mathcal V$ and CBV terms $\Gamma \vdash M : A$ as
morphism of any of the equivalent forms $\mathcal E(F(\times\Gamma),
F(A)) \cong \mathcal V(\times\Gamma \vdash UF(A)) \cong \mathcal
E(F(1), \Gamma \to F(A))$. The most interesting type translation is
the CBV function type: $A \ra A' = U(A \to F A')$.
%
Such a model validates all type-based equational reasoning, i.e.,
$\beta\eta$ equality, and models the introduction and elimination
rules for CBV.
%
Thus a CBPV model is sufficient to interpret the CBV term language. We
will require additional structure to interpret the precision and type
casts.

New and Licata modeled call-by-name gradual typing using \emph{double
categories}, which are defined to be categories internal to the
category of categories. That is, a double category $\mathcal C$
consists of a category $\mathcal C_o$ of ``objects and function
morphisms'' and a category $\mathcal C_{sq}$ of ``relation morphisms
and squares'' with functors (reflexive relation) $r : C_o \to C_{sq}$
and (source and target) $s,t : C_{sq} \to C_o$ satisfying $sr = tr =
\id$ as well as a composition operation $c : C_{sq} \times_{s,t}
C_{sq} \to C_{sq}$ respecting source and target. This models an
abstract notion of functions and relations. For notation, we write
function morphisms as $f : A \to B$ and relation morphisms as $c : A
\rel B$ where $c \in C_{sq}$ and $s(c) = A$ and $t(c) = B$. Finally a
morphism $\alpha$ from $c$ to $d$ with $s(\alpha) = f$ and $s(\beta) =
g$ is visualized as
% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBIl0sWzEsMCwiQiJdLFswLDEsIkEnIl0sWzEsMSwiQiciXSxbMCwyLCJmIiwyXSxbMSwzLCJnIl0sWzAsMSwiYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzIsMywiZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}[ampersand replacement=\&]
	A \& B \\
	{A'} \& {B'}
	\arrow["f"', from=1-1, to=2-1]
	\arrow["g", from=1-2, to=2-2]
	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow["d"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
\end{tikzcd}\]
And is thought of as an abstraction of the notion of relatedness of
functions: functions take related inputs to related outputs. The
composition operations and functoriality give us a notion of
composition of relations as well as functions and vertical and
horizontal composition of squares. In this work we will be chiefly
interested in \emph{locally thin} double categories, that is, double
categories where there is at most one square for any $f,c,g,d$. In
this case we use the notation $f \leq_{c,d} g$ to mean that a square
like the above exists.

New, Licata and Ahmed \cite{newlicataahmed} extended the axiomatic
syntax to call-by-push-value but did not analyze the structure
categorically. We fill in this missing analysis now: a model of the
congruence rules of their system can be given by a locally thin
``double CBPV model'', which we define as a category internal to the
category of CBPV models and \emph{strict} homomorphisms of CBPV
models\footnote{it may be possible to also define this as a notion of
CBPV model internal to some structured $2$-category of categories, but
the authors are not aware of any such definition of an internal CBPV
model}. A strict homomorphism of CBPV models from $(\mathcal
V,\mathcal E,\ldots)$ to $(\mathcal V', \mathcal E',\ldots)$ consists
of functors $G_v : \mathcal V \to \mathcal V'$ and $G_e : \mathcal E
\to \mathcal E'$ that strictly preserve all CBPV constructions, see
the appendix for a more detailed definition.
Some of that data can be visualized as follows:
% https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXHZzcSJdLFsyLDAsIlxcZXNxIl0sWzAsMiwiXFx2ZiJdLFsyLDIsIlxcZWYiXSxbMiwzLCJcXEZmIiwwLHsiY3VydmUiOi0yfV0sWzMsMiwiXFxVZiIsMCx7ImN1cnZlIjotMn1dLFswLDEsIlxcRnNxIiwwLHsiY3VydmUiOi0yfV0sWzEsMCwiXFxVc3EiLDAseyJjdXJ2ZSI6LTJ9XSxbMiwwLCJcXHJ2Il0sWzAsMiwiXFxzdiIsMCx7ImN1cnZlIjotMn1dLFswLDIsIlxcdHYiLDIseyJjdXJ2ZSI6Mn1dLFsxLDMsIlxcc2UiLDAseyJjdXJ2ZSI6LTJ9XSxbMSwzLCJcXHRlIiwyLHsiY3VydmUiOjJ9XSxbMywxLCJcXHJlIl0sWzQsNSwiXFxib3QiLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNiw3LCJcXGJvdCIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
\[\begin{tikzcd}[ampersand replacement=\&]
	\vsq \&\& \esq \\
	\\
	\vf \&\& \ef
	\arrow[""{name=0, anchor=center, inner sep=0}, "\Ff", curve={height=-12pt}, from=3-1, to=3-3]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\Uf", curve={height=-12pt}, from=3-3, to=3-1]
	\arrow[""{name=2, anchor=center, inner sep=0}, "\Fsq", curve={height=-12pt}, from=1-1, to=1-3]
	\arrow[""{name=3, anchor=center, inner sep=0}, "\Usq", curve={height=-12pt}, from=1-3, to=1-1]
	\arrow["\rv", from=3-1, to=1-1]
	\arrow["\sv", curve={height=-12pt}, from=1-1, to=3-1]
	\arrow["\tv"', curve={height=12pt}, from=1-1, to=3-1]
	\arrow["\se", curve={height=-12pt}, from=1-3, to=3-3]
	\arrow["\te"', curve={height=12pt}, from=1-3, to=3-3]
	\arrow["\re", from=3-3, to=1-3]
	\arrow["\bot"{description}, draw=none, from=0, to=1]
	\arrow["\bot"{description}, draw=none, from=2, to=3]
\end{tikzcd}\]
Type precision $A \ltdyn A'$ is interpreted as a relation morphism
$c_A : A \rel A'$ in $\mathcal V_{sq}$, and term precision $\Gamma
\ltdyn \Gamma' \vdash M \ltdyn M' : A \ltdyn A'$ is interpreted as a
square $M \ltdyn_{c_\Gamma,UF c_A} M'$. The fact that $t,r$ and the
composition are all given by strict CBPV homomorphisms says that all
the type constructors lift to precision (monotonicity of type
constructors) as well as all term constructors (congruence). Further,
$r$ and composition being strict homomorphisms implies that all type
constructors strictly preserve the identity relation (identity
extension) and composition.

Next, to model type casts, their model further requires that every
value relation $c : A \rel A'$ is \emph{left representable} by a
function $u_c : A \to A'$ and every computation relation $d : B \rel
B'$ is \emph{right representable} by a function $d_c : B' \to B$. In a locally thin double category, these are defined as follows:
\begin{definition}
  $c : A \rel B$ is left representable by $f : A \to B$ if $f \ltsq{c}{r(B)} \id$ and $\id \ltsq{r(A)}{c} f $.

  Dually, $c : A \rel B$ is right representable by $g : B \to A$ if
  $\id \ltsq{c}{r(A)} g$ and $g \ltsq{r(B)}{c} \id$.
\end{definition}
These rules are sufficient to model the UpL/UpR/DnL/DnR rules for
casts. Additionally, since representable morphisms compose and so the
compositionality of casts comes for free. However, the retraction
property must be added as an additional axiom to the model.  Finally,
the dynamic type can be modeled as an arbitrary value type $D$ with
arbitrary relations $\nat \rel D$ and $D \ra D \rel D$ and $D \times D
\rel D$ (or whatever basic type cases are required).

\begin{example}
  (Adapted from \cite{new-licata}): Define a double CBPV model where
  $\mathcal V$ is the category of predomain preorders: sets with an
  $\omega$-CPO structure $\leq$ as well as a poset structure
  $\ltdyn$. Functional morphisms are given by $\leq$-continuous and
  $\ltdyn$-monotone functions. Then define $\mathcal E$ to be the
  category of pointed domain preorders which are domain preorders with
  least elements $\bot$ for $\leq$ and $\mho$ for $\ltdyn$ such that
  $\bot$ is $\ltdyn$-maximal, and morphisms are as before but preserve
  $\bot$ and $\mho$. This can be extended to a CBPV model with
  forgetful functor $U : \mathcal E \to \mathcal V$. $D$ can be
  defined by solving a domain equation.

  This can be extended to a double CBPV model by defining a value
  relation $A \rel A'$ to be a $\ltdyn$-\emph{embedding}: a morphism
  $e : A \to A'$ that is injective and such that $F e : F A \to F A'$
  has a right adjoint (with respect to $\ltdyn$) and a square $f
  \ltsq{e}{e'} f' = f \circ e \ltdyn e' \circ f$. Similarly
  computation relations $B \rel B'$ are defined to be
  \emph{projections}: morphisms $p : B' \to B$ that are surjective and
  $Up$ has a left adjoint, with squares defined similarly. A suitable
  dynamic type can be constructed by solving a domain equation $D
  \cong \nat + U(D \to FD) + (D \times D)$.
\end{example}

\subsection{Weakening the Double Category Semantics}

While the double categorical semantics can be satisfied with classical
domain theoretic models, there are obstructions to developing a
semantics based on \emph{guarded} recursion. While guarded type theory
makes construction of arbitrary guarded recursive definitions
possible, it comes at a significant cost to reasoning: unfolding of
recursive definitions is explicit. In a denotational semantics, this
means that non-well-founded recursion must perform a kind of
\emph{observable step}. This is a difficulty in proving graduality,
which is a property that is oblivious to the number of steps that a
program takes. Therefore to prove graduality we must impose a kind of
weak bisimilarity relation on our programs to reason about
step-independent relational properties. However, here we arrive at the
fundamental issue with double categorical semantics using
representable relations:

\begin{enumerate}
\item When reasoning up to weak bisimilarity, transitive reasoning is
  not possible, and so horizontal pasting of squares is not a valid
  reasoning principle.
\item When reasoning with observable computation steps, certain casts
  take observable steps, and can no longer be modeled using
  representable morphisms.
\end{enumerate}

Our solution to this dilemma comes in two parts. Since graduality
ignores computational steps, the syntactic theory of graduality must
be modeled up to weak bisimilarity, where transitive reasoning is not
valid. For this purpose we develop a notion of \emph{extensional}
model which weakens from double categories to a reflexive graph
categories, dropping the operation of horizontal pasting of squares,
but still maintaining a form of representability of
relations.

However, transitive reasoning is essential for compositional semantic
constructions, and so we need to work with an intermediate notion of
\emph{intensional} model which is based on double categories, but
where representability has to be weakened to
``quasi-representability'', a kind of representability \emph{up to}
observable steps. To reason up to observable steps without using weak
bisimilarity, we develop a notion we call \emph{perturbations},
certain terms that are bisimilar to the identity but can be
manipulated explicitly in constructions. We then show that the
compositional construction of casts from domain theoretic models can
be adapted to this guarded setting by incorporating some explicit
manipulation of perturbations. Finally, we show that taking an
``extensional collapse'' by bisimilarity provides a model of the
extensional theory, which can then be used to model the global
graduality property.

### 2024-01-26 paper-new/extensional-model.tex
adjust outline

@@
 
-New and Licata modeled call-by-name gradual typing using \emph{double
-categories}, which are defined to be categories internal to the
-category of categories. That is, a double category $\mathcal C$
-consists of a category $\mathcal C_o$ of ``objects and function
-morphisms'' and a category $\mathcal C_{sq}$ of ``relation morphisms
-and squares'' with functors (reflexive relation) $r : C_o \to C_{sq}$
-and (source and target) $s,t : C_{sq} \to C_o$ satisfying $sr = tr =
-\id$ as well as a composition operation $c : C_{sq} \times_{s,t}
-C_{sq} \to C_{sq}$ respecting source and target. This models an
-abstract notion of functions and relations. For notation, we write
-function morphisms as $f : A \to B$ and relation morphisms as $c : A
-\rel B$ where $c \in C_{sq}$ and $s(c) = A$ and $t(c) = B$. Finally a
-morphism $\alpha$ from $c$ to $d$ with $s(\alpha) = f$ and $s(\beta) =
-g$ is visualized as
+\subsection{Double Categorical Semantics of Graduality}
+
+New and Licata modeled the graduality and type casts for call-by-name
+gradual typing using \emph{double categories}, which are defined to be
+categories internal to the category of categories. That is, a double
+category $\mathcal C$ consists of a category $\mathcal C_o$ of
+``objects and function morphisms'' and a category $\mathcal C_{sq}$ of
+``relation morphisms and squares'' with functors (reflexive relation)
+$r : C_o \to C_{sq}$ and (source and target) $s,t : C_{sq} \to C_o$
+satisfying $sr = tr = \id$ as well as a composition operation $c :
+C_{sq} \times_{s,t} C_{sq} \to C_{sq}$ respecting source and
+target. This models an abstract notion of functions and relations. For
+notation, we write function morphisms as $f : A \to B$ and relation
+morphisms as $c : A \rel B$ where $c \in C_{sq}$ and $s(c) = A$ and
+$t(c) = B$. Finally a morphism $\alpha$ from $c$ to $d$ with
+$s(\alpha) = f$ and $s(\beta) = g$ is visualized as
 % https://q.uiver.app/#q=WzAsNCxbMCwwLCJBIl0sWzEsMCwiQiJdLFswLDEsIkEnIl0sWzEsMSwiQiciXSxbMCwyLCJmIiwyXSxbMSwzLCJnIl0sWzAsMSwiYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzIsMywiZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d

### 2024-01-26 paper-new/intro.tex
adjust outline

@@
 arbitrary self-referential definitions is \emph{step-indexing} or its
-synthetic form of \emph{guarded recursion}.
-\max{todo: adapt the next paragraph to fit in with this one}
-A series of works \cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
+synthetic form of \emph{guarded recursion}.  \max{todo: adapt the next
+  paragraph to fit in with this one} A series of works
+\cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
 developed step-indexed logical relations models of gradually typed
@@
 mathematical abstractions. Our goal in the current work is to extract
-these reusable mathematical principles from these tedious models to
-make formalization of realistic gradual languages tractible.
+these reusable mathematical principles from these explicit
+step-indexed to make formalization of realistic gradual languages
+tractible.
 
@@
 \subsection{Outline of Remainder of Paper}
+
+\max{this is entirely out of date, update later}
 

### 2024-01-26 paper-new/terms.tex
add some prose in terms.tex
adjust outline

@@
 \section{A Simple Denotational Semantics for the Terms of GTLC}\label{sec:gtlc-terms}
+
+\max{Eric should make this a subsection of the next section}
 
@@
 lambda calculus (GTLC) and give a set-theoretic denotational semantics
-using tools from SGDT.
+using tools from SGDT. This serves two purposes: First, it is a simple setting
+in which to employ the tools of SGDT.
+Second, constructing this semantic model establishes the validity of the
+beta and eta principles for the gradually-typed lambda calculus.
 
@@
 
-\subsubsection{The Lift + Error Monad}
+\subsubsection{The Lift + Error Monad}\label{sec:lift-monad}
 

### 2024-01-26 paper-new/intro.tex
more intro

@@
 
+\max{the stuff below is out of date}
 An alternative approach, which we investigate in this paper, is provided by
@@
 \begin{enumerate}
-\item First, we give a rational reconstruction \max{todo: more shit.}
+\item First, we give a simple abstract categorical model of GTT using CBPV double categories.
+\item Next, we modify this semantics to develop reflexive graph- and
+  double categorical models that abstract over the details of
+  step-indexed models, and provide a method for constructing such models.
+\item We instantiate the abstract construction to provide a concrete
+  semantics in informal guarded type theory which supports the
+  addition of arbitrary algebraic effects.
+\item We prove that the resulting denotational model is
+  \emph{adequate} for the graduality property: a closed term precision
+  $M \ltdyn N : \nat$ has the expected semantics, that $M$ errors or
+  $M$ and $N$ have the same extensional behavior.
 \end{enumerate}

### 2024-01-26 paper-new/appendix.tex
updates to appendix

@@
 %%%%%%%%%%%%%%%
-\begin{lemma}
+\begin{lemma}\label{lem:push-pull-comp}
   Let $c : A \rel A'$ and $c' : A' \rel A''$ be value relations,
@@
 
-%%%%%%%%%%%%%
-% UF and FU %
-%%%%%%%%%%%%%
-\begin{lemma}
+%%%%%%%%%%%
+% U and F %
+%%%%%%%%%%%
+\begin{lemma}\label{lem:push-pull-U-F}
   Let $c : A \rel A'$ and let $P_A$ and $P_{A'}$ be monoids of perturbations.
   Given a push-pull structure $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$,
-  we can define a push-pull structure $\pie_{UF(c)}$ for $UF(c)$ with respect to $UF(P_{A})$
-  and $UF(P_{A'})$.
-  % Eric: Are we assuming here that F and U preserve perturbations?
+  we can define a push-pull structure $\pie_{Fc}$ for $F(c)$ with respect to
+  $\mathbb{N} \times P_{A}$ and $\mathbb{N} \times P_{A'}$.
+
+  Likewise, from a push-pull structure for $d$ with respect to $P_B$ and $P_{B'}$,
+  we can define a push-pull structure for $Ud$ with respect to $\mathbb{N} \times P_B$
+  and $\mathbb{N} \times P_{B'}$.
 \end{lemma}
 \begin{proof}
-  Follows by the functorial action of $UF$ and $FU$ on perturations and squares.
+
+  We define $\push_{Fc} : \mathbb{N} \times P_{A} \to \mathbb{N} \times P_{A'}$.
+  Since the product of commutative monoids coincides with the coproduct, it
+  suffices by the universal property of the coproduct to consider the two cases separately.
+
+  Given $n \in \mathbb{N}$ we define the push to be $n$.
+  Then the required square is $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$ which exists by
+  the definition of a step-1 model.
+  On the other hand, given $a \in P_{A}$, we define the push to be $\push_c(a)$
+  and then the required exists by the push property for $c$.
+  Defining $\pull_{Fc}$ is done similarly.
+
+  The proof for $Ud$ is analogous.
 \end{proof}
@@
 %%%%%%%%%%%%
-\begin{lemma}
+\begin{lemma}\label{lem:push-pull-times}
   Let $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$, and let $P_{A_1}$,
@@
 
-  Likewise, given push-pull structures 
-  $\pie_{Fc_1}$ with respect to $F(P_{A_1})$ and $F(P_{A_1'})$, and
-  $\pie_{Fc_2}$ with respect to $F(P_{A_2})$ and $F(P_{A_2'})$,
-  we can define a push-pull structure $\pie_{F(c_1 \times c_2)}$
-  with respect to $F(P_{A_1} \times P_{A_2})$ and $F(P_{A_1'} \times P_{A_2'})$.
+  % Likewise, given push-pull structures 
+  % $\pie_{Fc_1}$ with respect to $F(P_{A_1})$ and $F(P_{A_1'})$, and
+  % $\pie_{Fc_2}$ with respect to $F(P_{A_2})$ and $F(P_{A_2'})$,
+  % we can define a push-pull structure $\pie_{F(c_1 \times c_2)}$
+  % with respect to $F(P_{A_1} \times P_{A_2})$ and $F(P_{A_1'} \times P_{A_2'})$.
 \end{lemma}
@@
 %%%%%%%%%
-\begin{lemma}
+\begin{lemma}\label{lem:push-pull-arrow}
   Let $c : A \rel A'$ and $d : B \rel B'$.
@@
 
-  Given push-pull structures $\pie_{Fc}$ for $Fc$ and $\piv_{Ud}$ for $Ud$,
-  we can define a push-pull structure $\piv_{U(c \arr d)}$ for $U(c \arr d)$
-  with respect to the monoids $(P_{FA})^{op} \oplus P_B$
+  % Given push-pull structures $\pie_{Fc}$ for $Fc$ and $\piv_{Ud}$ for $Ud$,
+  % we can define a push-pull structure $\piv_{U(c \arr d)}$ for $U(c \arr d)$
+  % with respect to the monoids $(P_{FA})^{op} \oplus P_B$
 \end{lemma}
 \begin{proof}
-  The former follows from the functorial action of $\arr$ on perturbations and squares,
-  and the latter uses the functorial action of $\tok$ on perturbations and squares.
+  Follows from the functorial action of $\arr$ on perturbations and squares.
+  % and the latter uses the functorial action of $\tok$ on perturbations and squares.
 \end{proof}
@@
 We now proceed with the construction of the model:
-    % Write 
-    % %
-    % \[ \mathcal M = (\vf, \vsq, \ef, \esq, \Ff, \Fsq, \Uf, \Usq, \arrf, \arrsq). \] 
-    % %
 
@@
     \begin{itemize}
-      \item Value objects are triples consisting of:
+      \item Value objects are pairs consisting of:
       \begin{itemize}
-        \item A value object $A$ in $\vf$ 
-        \item A monoid $P_A$ of ``pure'' perturbations that is a submonoid of 
-        $\{ f \in \vf(A, A) \mid f \bisim \id_A \}$
-        \item A monoid $P^K_A$ of ``impure'' or ``Kleisli'' perturbations that is a submonoid of
-        $\{ \phi \in \ef(FA, FA) \mid \phi \bisim \id_{FA} \}$ such that $P^K_A$ contains
-        $\delta_A^*$
+        \item A value object $A$ in $\vf$.
+        \item A commutative monoid $P_A$ of perturbations
+        and a monoid homomorphism $\ptb_A : P_A \to \{ f \in \vf(A, A) \mid f \bisim \id_A \}$.
+
       \end{itemize}
 
-      \item Computation objects are triples consisting of:
+      \item Computation objects are pairs consisting of:
       \begin{itemize}
-        \item A computation object $B$ in $\ef$
-        \item A monoid $P_B$ of ``pure'' perturbations that is a submonoid of
-        $\{ \phi \in \ef(B, B) \mid \phi \bisim \id_B \}$
-        \item A monoid $P^K_B$ of ``impure'' or ``Kleisli'' perturbations that is a submonoid
-        of $\{ g \in \vf(UB, UB) \mid g \bisim \id_{UB} \}$
+        \item A computation object $B$ in $\ef$.
+        \item A commutative monoid $P_B$ of perturbations and a monoid homomorphism
+        $\ptb_B : P_B \to \{ \phi \in \ef(B, B) \mid \phi \bisim \id_B \}$.
       \end{itemize}
@@
 
-    \end{itemize}
-
-    % Before introducing the relations, we make a definition.
-
-    % \begin{definition}[push-pull structure]
-    %   Let $c : A \rel A'$ be a value relation of $\mathcal M$. A \emph{value push-pull structure} $\piv_c$ for $c$ consists of:
-    %   \begin{itemize}
-    %     \item A function $\push : P_A \to P_{A'}$ 
-    %           such that for all $\delta^l \in P_A$ we have $\delta^l \ltdyn_c^c \push(\delta^l)$.
-    %     \item A function $\push^K : P^K_A \to P^K_{A'}$ 
-    %           such that for all $\delta^K_l \in P^K_A$ we have $\delta^K_l \ltdyn_{Fc}^{Fc} \push(\delta^K_l)$.
-    %     \item A function $\pull : P_{A'} \to P_A$
-    %           such that for all $\delta_r \in P_{A'}$ we have $\pull(\delta^r) \ltdyn_{c}^c \delta^r$.
-    %     \item A function $\pull^K : P^K_{A'} \to P^K_A$
-    %           such that for all $\delta^K_r \in P^K_{A'}$ we have $\pull(\delta^K_r) \ltdyn_{Fc}^{Fc} \delta^K_r$.
-    %   \end{itemize}
-
-    %   For $d : B \rel B'$ a computation relation, we define a \emph{computation push-pull structure} $\pie_d$ for $d$
-    %   in an analogous manner.
-    % \end{definition}
-
-    % Now we continue with the description of the construction:
-    \begin{itemize}
-
-      \item The objects of $\vsq'$ (i.e., the value relations) are triples consisting of:
+      \item The objects of $\vsq'$ (i.e., the value relations) are pairs consisting of:
       \begin{itemize}
@@
         \item A push-pull structure $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$.
-        \item A push-pull structure $\pie_{Fc}$ for $Fc$ with respect to $P^K_A$ and $P^K_{A'}$.
+        %\item A push-pull structure $\pie_{Fc}$ for $Fc$ with respect to $P^K_A$ and $P^K_{A'}$.
       \end{itemize}
@@
         \item A push-pull structure $\pie_d$ for $d$ with respect to $P_B$ and $P_{B'}$.
-        \item A push-pull structure $\piv_{Ud}$ for $Ud$ with respect to $P^K_B$ and $P^K_{B'}$.
+        %\item A push-pull structure $\piv_{Ud}$ for $Ud$ with respect to $P^K_B$ and $P^K_{B'}$.
       \end{itemize}
@@
 
-      \item We define composition of relations $(c, \piv_c, \pie_{Fc})$ and 
-      $(c', \piv_{c'}, \pie_{Fc'})$ as $(c \comp c', \piv_{c \comp c'}, \pie_{Fc \comp Fc'})$
-      and likewise for computation relations.
+      \item We define composition of relations $(c, \piv_c)$ and $(c', \piv_{c'})$
+      as $(c \comp c', \piv_{c \comp c'})$ where $\piv_{c \comp c'}$ is as in Lemma
+      \ref{lem:push-pull-comp}, and likewise for computation relations.
 
@@
 
-      % \item We begin with the functor $m$ that composes relations and squares horizontally.
-      % Given value relations $(c ,\piv_c, \pie_{Fc})$ and $(c', \piv_{c'}, \pie_{Fc'})$ where $c : A \rel A'$ and $c' : A' \rel A''$,
-      % we define their composition to be the triple $(c \comp c', \piv_{c \comp c'})$ where we define
-      % $\piv_{c \comp c'}$ as the following push-pull structure:
-
-      % \begin{itemize}
-      %   \item $\push_{c \comp c'} = \push_{c'} \circ \push_{c}$
-      %   \item $\push^K_{c \comp c'} = \push^K_{c'} \circ \push^K_{c}$
-      %   \item $\pull_{c \comp c'} = \pull_{c} \circ \pull_{c'}$
-      %   \item $\pull^K_{c \comp c'} = \pull^K_{c} \circ \pull^K_{c'}$
-      % \end{itemize}
-
-      % We observe that each of the components satisfies the commuting necessary square.
-
       \item We define $\times$ on objects by
 
-      \[ (A_1, P_{A_1}, P^K_{A_1}) \times (A_2, P_{A_2}, P^K_{A_2}) =
-        (A_1 \times A_2, P_{A_1} \times P_{A_2}, P^K_{A_1} \oplus P^K{A_2}). \]
+      \[ (A_1, P_{A_1}) \times (A_2, P_{A_2}) =
+        (A_1 \times A_2, P_{A_1} \times P_{A_2}). \]
 
@@
 
-      We define $\times$ on relations by
+      Using Lemma \ref{lem:push-pull-times}, we define $\times$ on relations by
 
-      \[ (c_1, \piv_{c_1}, \pie_{Fc_1}), (c_2 \piv_{c_2}, \pie_{Fc_2}) = 
-        (c_1 \times c_2, \piv_{c_1 \times c_2}, \pie_{F(c_1 \times c_2)}). \]
+      \[ (c_1, \piv_{c_1}), (c_2 \piv_{c_2}) = 
+        (c_1 \times c_2, \piv_{c_1 \times c_2}). \]
 
@@
 
-      \[ F(A, P_A, P^K_A) = (FA, P^K_A, UF(P_A)). \]
+      \[ F(A, P_A, \ptb_A) = (FA, \mathbb{N} \times P_A, \ptb_{FA}), \]
 
-      We define $F$ on relations by
+      where we define $\ptb_{FA}(n, a) = (\delta_{FA}^*)^n \circ F(\ptb_A(a))$.
 
-      \[ F(c, \piv_c, \pie_{Fc}) = (Fc, \pie_{Fc}, \piv_{UF(c)}). \]
+      Using Lemma \ref{lem:push-pull-U-F}, we define $F$ on relations by
+
+      \[ F(c, \piv_c) = (Fc, \pie_{Fc}). \]
 
@@
 
-      \[ U(B, P_B, P^K_B) = (UB, P^K_B, FU(P_B). )\]
+      \[ U(B, P_B, \ptb_B) = (UB, \mathbb{N} \times P_B, \ptb_{UB}), \]
 
-      We define $U$ on relations by
+      where we define $\ptb_{UB}(n, b) = (\delta_{UB}^*)^n \circ U(\ptb_B(b))$.
 
-      \[ U(d, \pie_d, \piv_{Ud}) = (Ud, \piv_{Ud}, \pie_{FU(d)})  \]
+      Using Lemma \ref{lem:push-pull-U-F}, we define $U$ on relations by
+
+      \[ U(d, \pie_d) = (Ud, \piv_{Ud}).  \]
 
@@
 
-      \[ (A, P_A, P^K_A) \arr (B, P_B, P^K_B) = 
-        (A \arr B, (P_A)^{op} \times P_B, (P^K_A)^{op} \oplus P^K_B). \]
+      \[ (A, P_A, \ptb_A) \arr (B, P_B, \ptb_B) = 
+        (A \arr B, P_A \times P_B, \ptb_{A \arr B}), \]
 
-      We define $\arr$ on relations by
+      where we define $\ptb_{A \arr B}(a, b) = \ptb_A(a) \arr \ptb_B(b)$.
 
-      \[ (c, \piv_c, \pie_{Fc}) \arr (d, \pie_d, \piv_{Ud}) =
-        (c \arr d, \pie_{c \arr d}, \piv_{U(c \arr d)}). \]
+      Using Lemma \ref{lem:push-pull-arrow}, we define $\arr$ on relations by
 
-      % \item We define $F$ on objects by $F (A, \pv_A, \ptbv_A) = (FA, (1 + \pv_A), h_F)$
-      % where $1$ is the trivial monoid, $+$ is the coproduct in the category of monoids, and $h_F$ is the homomorphism defined as follows:
+      \[ (c, \piv_c) \arr (d, \pie_d) =
+        (c \arr d, \pie_{c \arr d}). \]
 
-      % \item We define $U$ on objects by $U (B, \pe_B, \ptbe_B) = (UB, \pe_B, h_U)$
-      % where $h_U(p_B) = U(\ptbe_B(p_B))$.
+    \end{itemize}
 
-      % \item We define $(A, \pv_A, \ptbv_A) \arr (B, \pe_B, \ptbe_B) = (A \arr B, \pv_A \times \pe_B, h_\arr)$
-      % where $\times$ is the product in the category of monoids, and $h_\arr$ is defined by 
-      % $h_\arr(p_A, p_B) = \ptbv_A(p_A) \arr \ptbe_B(p_B)$.
-    \end{itemize}
+    \eric{Make sure this part is correct}
+    We define a ``Kleisli arrow'' operation taking a perturbation in $P_{UB}$
+    to a perturbation in $P_{U(A \to B)}$, as follows:
+    Given a perturbation $u = (n, b) \in P_{UB}$ we define $\id \tok u$ to be 
+    $(n, \id_{P_A}, b) \in P_{U(A \to B)}$.
+    We need to show that
+%
+    \[ \ptb_{U(A \arr B)}(\id \tok (n, b)) = \id \tok \ptb_{UB}(n, b). \]
+%
+    Since the product and coproduct coincide in the category of commutative monoids,
+    and since $\mathbb{N}$ is generated by the element $1$,
+    it suffices by the universal property of the coproduct to show
+%
+    \[ \ptb_{U(A \arr B)}(\id \tok (1, \id_{P_B})) = \id \tok \ptb_{UB}(1, \id_{P_B}), \]
+%
+    and
+%
+    \[ \ptb_{U(A \arr B)}(\id \tok (0, b)) = \id \tok \ptb_{UB}(0, b). \]
+
+    For the former, the LHS is
+%
+    \begin{align*}
+      \ptb_{U(A \arr B)}&(\id \tok (1, \id_{P_B})) \\ 
+      &= \ptb_{U(A \arr B)}(1, \id_{P_A}, \id_{P_B}) \\ 
+      &= \delta_{U(A \arr B)}^* \circ U(\ptb_A(\id_{P_A}) \arr \ptb_B(\id_{P_B})) \\
+      %&= \delta_{U(A \arr B)}^* \circ U(\id_A \arr \id_B) \\
+      &= \delta_{U(A \arr B)}^*
+    \end{align*}
+%
+    and we need to show that this is equal to 
+    $\id \tok \ptb_{UB}(1, \id_{P_B}) = \id \tok \delta_{UB}^*$.
+    But this holds because $\delta_{UB}^*$ commutes with all computation morphisms,
+    in particular, with bind and application.
+
+    For the latter, we need to show
+%
+    \[ U(\id \arr \ptb_B(b)) = \id \tok U(\ptb_B(b)), \]
+%
+    which holds by definition of $\tok$.
+
+    Given a perturbation $q = (n, a) \in P_{FA}$ we define
+    $(n, a) \tok \id = (n, a, \id_{P_B}) \in P_{U(A \to B)}$.
+    We need to show that
+  %
+    \[ \ptb_{U(A \arr B)}((n, a) \tok \id) = \ptb_{FA}(n, a) \tok \id. \]
+  %
+    By similar reasoning to the above, it will suffice to show that
+
+    \[ \delta_{U(A \arr B)}^* = \delta_{FA}^* \tok \id \]
+
+    And this follows from the fact that $\delta_{FA}^*$ commutes with all computation morphisms.
+
+    We can similarly construct the Kleisli product operation on perturbations.
+    Given a perturbation $(n, a_1, a_2) \in P_{F(A_1 \times A_2)}$ we define
+    $(n, a_1, a_2) \timesk \id = (n, a_1, \id_{P_{A_2}})$,
+    and likewise we define $\id \timesk (n, a_1, a_2) = (n, \id_{P_{A_1}}, a_2)$.
+
+% We have $\ptb_{U(A \arr B)}(\id \tok (n, b)) = \ptb_{U(A \arr B)}(n, \id, b)$, 
+% which is equal to $\delta_{U(A \arr B)}^n (\id \to \ptb_B(b))$
 
@@
   $c$ is quasi-left-representable by $e_c$ (see Definition \ref{def:quasi-left-representable}).
+  We similarly define a {right-representation structure} $\rho^R_c$ for $c$ to consist of a morphism
+  $p_c \in \vf{A', A}$ such that $c$ is quasi-right-representable by $p_c$ (see Definition \ref{def:quasi-right-representable}).
+  % (Notice that the direction of the morphism is opposite in the definition of
+  % right-representation structure.)
 
-Likewise, let $d : B \rel B'$ be a computation relation. A \emph{right-representation structure}
-$\rho^R_d$ for $d$ consists of a computation morphism $p_d \in \ef(B', B)$
-such that $d$ is quasi-right-representable by $p_d$ (see Definition
-\ref{def:quasi-right-representable}).
+  Likewise, let $d : B \rel B'$ be a computation relation. Left- and right-representation
+  structures for $d$ are defined in a similar manner, except the representing morphisms
+  are now computation morphisms rather than value morphisms.
+
+  % Likewise, let $d : B \rel B'$ be a computation relation. A \emph{right-representation structure}
+  % $\rho^R_d$ for $d$ consists of a computation morphism $p_d \in \ef(B', B)$
+  % such that $d$ is quasi-right-representable by $p_d$ (see Definition
+  % \ref{def:quasi-right-representable}).
 \end{definition}
 
-(Notice that the direction of the morphism is opposite in the definition of
-right-representation structure.)
+% TODO establish the notation
 
-% TODO establish the notation
+\begin{definition}
+  Let $c, c' : A \rel A'$. We say that $c$ and $c'$ are \emph{quasi-order-equivalent},
+  written $c \qordeq c'$, if there exist perturbations $\delta^l_1, \delta^l_2 \in \pv_A$ and 
+  $\delta^r_1, \delta^r_2 \in \pv_{A'}$ such that the following two squares exist:
+
+  \begin{center}
+    \begin{tabular}{ m{9em} m{9em} } 
+      \begin{tikzcd}[ampersand replacement=\&]
+        A \& {A'} \\
+        A \& {A'}
+        \arrow["\delta^l_1"', from=1-1, to=2-1]
+        \arrow["\delta^r_1", from=1-2, to=2-2]
+        \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+        \arrow["c'"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+      \end{tikzcd}
+      &
+      \begin{tikzcd}[ampersand replacement=\&]
+        A \& {A'} \\
+        A \& {A'}
+        \arrow["\delta^l_2"', from=1-1, to=2-1]
+        \arrow["\delta^r_2", from=1-2, to=2-2]
+        \arrow["c'", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+        \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+      \end{tikzcd}
+    \end{tabular}
+  \end{center}
+
+  We make the analogous definition for computation relations $d, d' : B \rel B'$.
+\end{definition}
+
+\begin{lemma}\label{lem:left-rep-by-same-morphism}
+  Let $c, c' : A \rel A'$. If $c$ and $c'$ are both quasi-left-representable
+  by the same $f$, then $c \qordeq c'$.
+\end{lemma}
+\begin{proof}
+  By $\upr$ for $c'$, there exists a perturbation $\delle_{c'}$ and a square
+  $\upr_{c'} : \delle_{c'} \ltdyn_{c'}^{r(A)} f$.
+
+  By $\upl$ for $c$, there exists a perturbation $\delre_c$ and a square 
+  $\upl_c : f \ltdyn_{r(A')}^{c} \delre_c$.
+
+  Composing these horizontally we get the following square:
+
+  % https://q.uiver.app/#q=WzAsNixbMCwwLCJBIl0sWzEsMCwiQSJdLFsyLDAsIkEnIl0sWzAsMSwiQSJdLFsxLDEsIkEnIl0sWzIsMSwiQSciXSxbMCwxLCJyKEEpIl0sWzEsMiwiYyJdLFszLDQsImMnIiwyXSxbNCw1LCJyKEEnKSIsMl0sWzAsMywiXFxkZWxsZV97Yyd9IiwyXSxbMSw0LCJmIiwyXSxbMiw1LCJcXGRlbHJlX2MiXV0=
+  \[\begin{tikzcd}[ampersand replacement=\&]
+    A \& A \& {A'} \\
+    A \& {A'} \& {A'}
+    \arrow["{r(A)}", from=1-1, to=1-2]
+    \arrow["c", from=1-2, to=1-3]
+    \arrow["{c'}"', from=2-1, to=2-2]
+    \arrow["{r(A')}"', from=2-2, to=2-3]
+    \arrow["{\delle_{c'}}"', from=1-1, to=2-1]
+    \arrow["f"', from=1-2, to=2-2]
+    \arrow["{\delre_c}", from=1-3, to=2-3]
+  \end{tikzcd}\]
+
+  And since $r(A) \comp c = c$ and $c' \comp r(A') = c'$, we are finished.
+
+  The other square (i.e., with $c'$ on top) is constructed in an analogous manner.
+\end{proof}
 
@@
     % 3.
-    \item We define $\rho^R_{F(c \comp c')}$ as follows:
-    \begin{itemize}
-      \item $p_{F(c \comp c')} = p_{Fc} \circ p_{Fc'}$ % : FA'' \to FA
-      \item $\dellp_{F(c \comp c')} = $
-      \item $\delrp_{F(c \comp c')} = $
-    \end{itemize}
+    \item We define $\rho^R_{F(c \comp c')}$ as follows.
+    % First, we observe that by Lemma \ref{lem:left-rep-by-same-morphism},
+    % $F(c \comp c')$ and $Fc \comp Fc'$ are quasi-order-equivalent,
+    % as both are left-represented by $Fe_c' \circ Fe_c$.
+
+    First, we claim that $F(c \comp c')$ and $Fc \comp Fc'$ are
+    both left-represented by $Fe_{c'} \circ Fe_c$.
+    Indeed, we have by part (1) that $e_c' \circ e_c$ left-represents $c \comp c'$,
+    and then since $F$ preserves perturbations and squares it follows that
+    $F(e_c' \circ e_c) = Fe_{c'} \circ Fe_c$ quasi-left-represents $F(c \comp c')$.
+    On the other hand, we also know that $Fe_c$ quasi-left-represents $Fc$ and $Fe_{c'}$
+    left-represents $Fc'$, so their composition quasi-left-represents $Fc \comp Fc'$.
+    Thus, there is a square $\alpha$ of the form
+
+    % https://q.uiver.app/#q=WzAsNSxbMCwwLCJGQSJdLFsyLDAsIkZBJyciXSxbMCwxLCJGQSJdLFsxLDEsIkZBJyJdLFsyLDEsIkZBJyciXSxbMCwxLCJGKGMgXFxjb21wIGMnKSIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzIsMywiRmMiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFszLDQsIkZjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMiwiXFxkZWx0YV5sIiwyXSxbMSw0LCJcXGRlbHRhXnIiXSxbOCw5LCJcXGFscGhhIiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+    \[\begin{tikzcd}[ampersand replacement=\&]
+      FA \&\& {FA''} \\
+      FA \& {FA'} \& {FA''}
+      \arrow["{F(c \comp c')}", "\shortmid"{marking}, no head, from=1-1, to=1-3]
+      \arrow["Fc", "\shortmid"{marking}, no head, from=2-1, to=2-2]
+      \arrow["{Fc'}", "\shortmid"{marking}, no head, from=2-2, to=2-3]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "{\delta^l}"', from=1-1, to=2-1]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "{\delta^r}", from=1-3, to=2-3]
+      \arrow["\alpha"{description}, draw=none, from=0, to=1]
+    \end{tikzcd}\]
+
+    for some perturbations $\delta^l$ and $\delta^r$.
+    We define the projection $p_{F(c \comp c')}$ to be $p_{Fc \comp Fc'} \circ \delta^r$.
+    We define $\dellp_{F(c \comp c')}$ to be $\dellp_{Fc \comp Fc'} \circ \delta^l$
+
+    Then we can build the square for $\dnr$ by pasting the square $\alpha$ on top of
+    the $\dnr$ square for the composition $Fc \comp Fc'$, as shown below:
+
+    % https://q.uiver.app/#q=WzAsNyxbMCwwLCJGQSJdLFsyLDAsIkZBJyciXSxbMCwxLCJGQSJdLFsxLDEsIkZBJyJdLFsyLDEsIkZBJyciXSxbMCwyLCJGQSJdLFsyLDIsIkZBIl0sWzAsMSwiRihjIFxcY29tcCBjJykiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsyLDMsIkZjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMyw0LCJGYyciLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs1LDYsInIoRkEpIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwyLCJcXGRlbHRhXmwiLDJdLFsyLDUsIlxcZGVsbHBfe0ZjIFxcY29tcCBGYyd9IiwyXSxbMSw0LCJcXGRlbHRhXnIiXSxbNCw2LCJwX3tGYyBcXGNvbXAgRmMnfSJdLFsxMSwxMywiXFxhbHBoYSIsMyx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxMiwxNCwiXFxkbnJfe0ZjIFxcY29tcCBGYyd9IiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+    \[\begin{tikzcd}[ampersand replacement=\&]
+      FA \&\& {FA''} \\
+      FA \& {FA'} \& {FA''} \\
+      FA \&\& FA
+      \arrow["{F(c \comp c')}", "\shortmid"{marking}, no head, from=1-1, to=1-3]
+      \arrow["Fc", "\shortmid"{marking}, no head, from=2-1, to=2-2]
+      \arrow["{Fc'}", "\shortmid"{marking}, no head, from=2-2, to=2-3]
+      \arrow["{r(FA)}"', "\shortmid"{marking}, no head, from=3-1, to=3-3]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "{\delta^l}"', from=1-1, to=2-1]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "{\dellp_{Fc \comp Fc'}}"', from=2-1, to=3-1]
+      \arrow[""{name=2, anchor=center, inner sep=0}, "{\delta^r}", from=1-3, to=2-3]
+      \arrow[""{name=3, anchor=center, inner sep=0}, "{p_{Fc \comp Fc'}}", from=2-3, to=3-3]
+      \arrow["\alpha"{marking, allow upside down}, draw=none, from=0, to=2]
+      \arrow["{\dnr_{Fc \comp Fc'}}"{description}, draw=none, from=1, to=3]
+    \end{tikzcd}\]
+
+    We define $\delrp_{F(c \comp c')}$ to be $\delrp_{Fc \comp Fc'} \circ \delta^r$.
+    For $\dnl$, we paste the identity square $\delta^r \ltdyn \delta^r$ on top of
+    the $\dnl$ square for the composition $Fc \comp Fc'$, and below that we paste
+    the square $\id \ltdyn_{F(c \comp c')}^{Fc \comp Fc'} \id$ which we get from
+    the lax property of $F$.
+
+    % https://q.uiver.app/#q=WzAsOSxbMCwwLCJGQScnIl0sWzIsMCwiRkEnJyJdLFswLDEsIkZBJyciXSxbMiwxLCJGQScnIl0sWzAsMiwiRkEiXSxbMiwyLCJGQScnIl0sWzAsMywiRkEiXSxbMiwzLCJGQScnIl0sWzEsMiwiRkEnIl0sWzAsMSwicihGQScnKSIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzIsMywicihGQScnKSIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzYsNywiRihjIFxcY29tcCBjJykiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDIsIlxcZGVsdGFeciIsMl0sWzIsNCwicF97RmMgXFxjb21wIEZjJ30iLDJdLFs0LDYsIlxcaWQiLDJdLFsxLDMsIlxcZGVsdGFeciJdLFszLDUsIlxcZGVscnBfe0ZjIFxcY29tcCBGYyd9Il0sWzUsNywiXFxpZCJdLFs0LDgsIkZjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbOCw1LCJGYyciLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxMiwxNSwicihcXGRlbHRhXnIpIiwzLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEzLDE2LCJcXGRubF97RmMgXFxjb21wIEZjJ30iLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
+    \[\begin{tikzcd}[ampersand replacement=\&]
+      {FA''} \&\& {FA''} \\
+      {FA''} \&\& {FA''} \\
+      FA \& {FA'} \& {FA''} \\
+      FA \&\& {FA''}
+      \arrow["{r(FA'')}", "\shortmid"{marking}, no head, from=1-1, to=1-3]
+      \arrow["{r(FA'')}", "\shortmid"{marking}, no head, from=2-1, to=2-3]
+      \arrow["{F(c \comp c')}"', "\shortmid"{marking}, no head, from=4-1, to=4-3]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "{\delta^r}"', from=1-1, to=2-1]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "{p_{Fc \comp Fc'}}"', from=2-1, to=3-1]
+      \arrow["\id"', from=3-1, to=4-1]
+      \arrow[""{name=2, anchor=center, inner sep=0}, "{\delta^r}", from=1-3, to=2-3]
+      \arrow[""{name=3, anchor=center, inner sep=0}, "{\delrp_{Fc \comp Fc'}}", from=2-3, to=3-3]
+      \arrow["\id", from=3-3, to=4-3]
+      \arrow["Fc", "\shortmid"{marking}, no head, from=3-1, to=3-2]
+      \arrow["{Fc'}", "\shortmid"{marking}, no head, from=3-2, to=3-3]
+      \arrow["{r(\delta^r)}"{marking, allow upside down}, draw=none, from=0, to=2]
+      \arrow["{\dnl_{Fc \comp Fc'}}"{description}, draw=none, from=1, to=3]
+    \end{tikzcd}\]
 
     % 4.
-    \item We define $\rho^L{(U(d \comp d'))}$ as follows:
+    \item We define $\rho^L{(U(d \comp d'))}$ in an analogous manner to the above.
 
@@
 Similarly, we can define a ``global'' version of the lock-step error ordering and the
-weak bisimilarity relation on morphisms.
+weak bisimilarity relation on elements of the global lift; the former is defined by
 
-It can be shown that the global lift monad is isomorphic to the so-called Delay monad of Capretta \cite{TODO}.
-Recall that the delay monad $\text{Delay}(X)$ is defined as the coinductive solution to the equation $\text{Delay}(X) \cong X + \text{Delay}(X)$.
-It can be shown that $\li^{gl} X \cong \text{Delay}(X + 1)$ for $X$ a clock-irrelevant type.
+\[ x \le^{gl} y := \forall k. x[k] \le y[k], \]
 
-Moreover, we claim that modulo this isomorphism, the global version of the extensional error ordering
+and the latter is defined by
+
+\[ x \bisim^{gl} y := \forall k. x[k] \bisim y[k]. \]
+
+It can be shown that there is an isomorphism between the global lift monad and the
+so-called Delay monad of Capretta \cite{lmcs:2265}.
+Recall that the delay monad $\text{Delay}(X)$ is defined as the coinductive solution
+to the equation
+
+\[ \text{Delay}(X) \cong X + \text{Delay}(X). \]
+
+It can be shown that for a clock-irrelevant type $X$, $\li^{gl} X$ is a final
+coalgebra of the functor $F(Y) = X + 1 + Y$ (For example, this follows from Theorem 4.3 in
+\cite{kristensen-mogelberg-vezzosi2022}.) Since $\text{Delay}(X + 1)$ is also a final coalgebra
+of this functor, then we have $\li^{gl} X \cong \text{Delay}(X + 1)$.
+
+We can define a ``lock-step error ordering'' on $\text{Delay}(X + 1)$
+
+And we similarly define a ``weak bisimilarity'' on $\text{Delay}(X + 1)$
+
+We claim that modulo the above isomorphism, the ``global" version of the
+extensional error ordering on $\li^{gl} X$ is equivalent to
 

### 2024-01-26 paper-new/categorical-model.tex
changes to categorical model

@@
 
-Lastly, we require that for any value object $A$, the hom-set $\ef(FA, FA)$ contains a
-distinguished morphism $\delta_A^*$, such that $\delta_A^* \bisim_{FA, FA} \id_{FA}$.
+% Spelling this all out concretely, for any pair...
 
-% Spelling this all out concretely, for any pair...
+% Lastly, we require that for any value object $A$, the hom-set $\ef(FA, FA)$ contains a
+% distinguished morphism $\delta_{FA}^*$, such that $\delta_A^* \bisim_{FA, FA} \id_{FA}$.
+% Moreover, we require that these morphisms are related in that for any
+% $c : A \rel A'$, we have a square $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$.
+
+Lastly, we require that for any computation object $B$, the hom-set $\vf(UB, UB)$ contains a
+distinguished morphism $\delta_{UB}^*$, such that $\delta_{UB}^* \bisim_{UB, UB} \id_{UB}$.
+Moreover, we require that these morphisms are related in that for any
+$d : B \rel B'$, we have a square $\delta_{UB}^* \ltdyn_{Ud}^{Ud} \delta_{UB'}^*$.
+We also require that these morphisms commute with computation morphisms, in the sense
+that for any $\phi \in \ef(B, B')$ we have $U\phi \circ \delta_{UB_1}^* = \delta_{UB_2} \circ U\phi$.
+
+Given the existence of the morphisms $\delta_{UB}^*$, we can define computation morphisms
+$\delta_{FA}^* \in \ef(FA, FA)$ for all $A$ using the
 
@@
 A \emph{step-1 intensional model} consists of all the data of a step-0 intensional model along
-with the additional categories and functors for bisimiarity described above, as  well as
-the existence of a distinguised computation morphism $\delta_A^* \bisim \id_{FA}$ for each $A$.
+with:
+\begin{itemize}
+  \item The categories and functors for bisimiarity described above.
+  \item The existence of a distinguished value morphism $\delta_{UB}^* \bisim \id_{UB}$ for each $B$.
+  \item A square $\delta_{UB}^* \ltdyn_{Ud}^{Ud} \delta_{UB'}^*$ for all $d : B \rel B'$.
+  \item The commutativity condition $U\phi \circ \delta_{UB_1}^* = \delta_{UB_2} \circ U\phi$ for any $\phi \in \ef(B, B')$.
+\end{itemize}
+% We also require the existence of a distinguised computation morphism $\delta_{FA}^* \bisim \id_{FA}$ for each $A$,
+% and a square $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$ for all $c : A \rel A'$.
 \end{definition}
@@
 
-We require that $\delta_A^* \in \pe_{FA}$ for all $A$, where $\delta_A^*$ is the distinguished
+We require that $\delta_{FA}^* \in \pe_{FA}$ for all $A$, where $\delta_{FA}^*$ is the distinguished
 morphism that is required to be present in every hom-set $\ef(FA, FA)$ per the definition
@@
     $\{ g \in \ef(B,B) \mid g \bisim \id \}$.
-    \item For all $A$, the distinguished endomorphism $\delta_A^*$ is in $\pe_{FA}$.
+    \item For all $A$, the distinguished endomorphism $\delta_{FA}^*$ is in $\pe_{FA}$.
     % \item $\pv_A$ and a monoid homomorphism 
@@
 are perturbations involved in order to keep both sides ``in lock-step".
+We begin with a step-2 intensional model as defined in the previous section,
+and provide the additional conditions that axiomatize the behavior of casts.
 The precise definitions are as follows.
 
-% TODO: Give these squares names
+First, let $\mathcal M$ be any double category with a notion of perturbations,
+i.e., for any object $X$ in $\mathcal M$ there is a monoid $P_X$ which is a
+submonoid of the endomorphisms on $X$.
+
 \begin{definition}\label{def:quasi-left-representable}
-Let $c : A \rel A'$ be a value relation. We say that $c$ is \emph{quasi-left-representable by}
-$f \in \vf(A, A')$ if there are perturbations $\delta_c^{l,e} \in \pv_A$ and
-$\delta_c^{r,e} \in \pv_{A'}$ such that the following squares commute:
+  Let $R$ be a horizontal morphism in $\mathcal M$ between objects $X$ and $Y$.
+  We say that $R$ is quasi-left-representable by a vertical morphism $f$ in
+  $\mathcal M(X, Y)$ if there are perturbations $\delle_R \in P_X$ and
+  $\delre_R \in P_Y$ such that the following squares commute:
 
@@
       \begin{tikzcd}[ampersand replacement=\&]
-      A \& {A'} \\
-      {A'} \& {A'}
+        X \& {Y} \\
+        {Y} \& {Y}
         \arrow["f"', from=1-1, to=2-1]
-      \arrow["\delta_c^{r,e}", from=1-2, to=2-2]
-      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+        \arrow["\delta_R^{r,e}", from=1-2, to=2-2]
+        \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
         \arrow[from=2-1, to=2-2, Rightarrow, no head]
@@
       \begin{tikzcd}[ampersand replacement=\&]
-      A \& {A} \\
-      {A} \& {A'}
-      \arrow["\delta_c^{l,e}"', from=1-1, to=2-1]
+        X \& {X} \\
+        {X} \& {Y}
+        \arrow["\delta_R^{l,e}"', from=1-1, to=2-1]
         \arrow["f", from=1-2, to=2-2]
         \arrow[from=1-1, to=1-2, Rightarrow, no head]
-      \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+        \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
       \end{tikzcd}
@@
 \begin{definition}\label{def:quasi-right-representable}
-Let $d : B \rel B'$ be a computation relation. We say that $d$ is
-\emph{quasi-right-representable by} $f \in \ef(B', B)$
-if there exist perturbations $\delta_d^{l,p} \in \pe_B$ and
-$\delta_d^{r,p} \in \pe_{B'}$ such that the following squares commute:
+  Let $R$ be a horizontal morphism between $X$ and $Y$. We say that $R$ is
+  \emph{quasi-right-representable by} $f \in \mathcal M(Y, X)$
+  if there exist perturbations $\dellp_R \in P_X$ and
+  $\delrp_R \in P_Y$ such that the following squares commute:
 
@@
       \begin{tikzcd}[ampersand replacement=\&]
-      {B} \& {B'} \\
-      {B} \& {B}
-      \arrow["\delta_d^{l,p}"', from=1-1, to=2-1]
-      \arrow["g", from=1-2, to=2-2]
+        {X} \& {Y} \\
+        {X} \& {X}
+        \arrow["\delta_R^{l,p}"', from=1-1, to=2-1]
+        \arrow["f", from=1-2, to=2-2]
         \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
@@
       \begin{tikzcd}[ampersand replacement=\&]
-      {B'} \& {B'} \\
-      {B} \& {B'}
-      \arrow["g"', from=1-1, to=2-1]
-      \arrow["\delta_d^{r,p}", from=1-2, to=2-2]
+        {Y} \& {Y} \\
+        {X} \& {Y}
+        \arrow["f"', from=1-1, to=2-1]
+        \arrow["\delta_R^{r,p}", from=1-2, to=2-2]
         \arrow[from=1-1, to=1-2, Rightarrow, no head]
@@
   \end{definition}
+
+% TODO: Give these squares names
+% \begin{definition}\label{def:quasi-left-representable}
+% Let $c : A \rel A'$ be a value relation. We say that $c$ is \emph{quasi-left-representable by}
+% $f \in \vf(A, A')$ if there are perturbations $\delta_c^{l,e} \in \pv_A$ and
+% $\delta_c^{r,e} \in \pv_{A'}$ such that the following squares commute:
+
+% \begin{center}
+%   \begin{tabular}{ m{7em} m{7em} } 
+%     % UpL
+%     \begin{tikzcd}[ampersand replacement=\&]
+%       A \& {A'} \\
+%       {A'} \& {A'}
+%       \arrow["f"', from=1-1, to=2-1]
+%       \arrow["\delta_c^{r,e}", from=1-2, to=2-2]
+%       \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+%       \arrow[from=2-1, to=2-2, Rightarrow, no head]
+%     \end{tikzcd}
+%     &
+%     % UpR
+%     \begin{tikzcd}[ampersand replacement=\&]
+%       A \& {A} \\
+%       {A} \& {A'}
+%       \arrow["\delta_c^{l,e}"', from=1-1, to=2-1]
+%       \arrow["f", from=1-2, to=2-2]
+%       \arrow[from=1-1, to=1-2, Rightarrow, no head]
+%       \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+%     \end{tikzcd}
+%   \end{tabular}
+% \end{center}
+
+% We call the first square $\upl$ and the second square $\upr$.
+
+% \end{definition}
+
+% \begin{definition}\label{def:quasi-right-representable}
+% Let $d : B \rel B'$ be a computation relation. We say that $d$ is
+% \emph{quasi-right-representable by} $f \in \ef(B', B)$
+% if there exist perturbations $\delta_d^{l,p} \in \pe_B$ and
+% $\delta_d^{r,p} \in \pe_{B'}$ such that the following squares commute:
+
+% \begin{center}
+%   \begin{tabular}{ m{7em} m{7em} } 
+%     % DnR
+%     \begin{tikzcd}[ampersand replacement=\&]
+%       {B} \& {B'} \\
+%       {B} \& {B}
+%       \arrow["\delta_d^{l,p}"', from=1-1, to=2-1]
+%       \arrow["g", from=1-2, to=2-2]
+%       \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+%       \arrow[from=2-1, to=2-2, Rightarrow, no head]
+%     \end{tikzcd}
+%     &
+%     % DnL
+%     \begin{tikzcd}[ampersand replacement=\&]
+%       {B'} \& {B'} \\
+%       {B} \& {B'}
+%       \arrow["g"', from=1-1, to=2-1]
+%       \arrow["\delta_d^{r,p}", from=1-2, to=2-2]
+%       \arrow[from=1-1, to=1-2, Rightarrow, no head]
+%       \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+%     \end{tikzcd}
+%   \end{tabular}
+% \end{center}
+
+% We call the first square $\dnr$ and the second square $\dnl$.
+
+% \end{definition}
+
+With these definitions, we return to the more specific setting of a step-2
+intensional model $\mathcal M$ and specify the new requirements for relations.
+We require that there are functors $\upf : \ve \to \vf$ and $\dnf : \ee^{op} \to \ef$
+Every value edge $c : A \rel A'$ must be quasi-left-representable by $\upf(c)$,
+and every computation edge $d : B \rel B'$ is quasi-right-representable by $\dnf(d)$.
 
@@
 
+\begin{comment}
 Lastly, we require that the model satisfy a weak version of functoriality for 
@@
 \end{itemize}
+\end{comment}
 
@@
     every computation edge $d : B \rel B'$ is quasi-right-representable by $\dnf(d)$.
-    \item The CBPV connectives $U,F,\times,\to$ are quasi-functorial on relations.
+    \item \eric{Do we need the quasi-functoriality of CBPV connectives on relations?}
+    % \item The CBPV connectives $U,F,\times,\to$ are quasi-functorial on relations.
   \end{enumerate}
@@
 %
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\subsubsection{Adding Perturbations}
+\subsubsection{Adding Perturbations}\label{sec:constructing-perturbations}
 
@@
 We claim that from a step-1 model, we can construct a step-2 model. 
-% TODO give overview
+\eric{TODO give overview}
 For the proof, see Lemma \ref{lem:step-1-model-to-step-2-model} in the Appendix.
@@
 
-% TODO give overview
+\eric{TODO give overview}
 

### 2024-01-26 paper-new/discussion.tex
revise discussion

@@
 \ref{sec:concrete-model} and the extensional model with external dynamic 
-type. This part is still work-in-progress.
+type. We also plan to formalize the adequacy result in \ref{sec:appendix-adequacy}.
 
@@
 Then we plan to construct the step-2 intensional model. Besides all the 
-data in step-1, we need to include perturbations on value and computation 
-types, functors $\times$, $\arr$, $U$, and $F$ that preserve 
+data in step-1, we need to include perturbations, functors $\times$, $\arr$, $U$, and $F$ that preserve 
 perturbations and push/pull properties for all morphisms on value and 
@@
 that preserves perturbations, we need to include the ones in the context 
-of Kleisli category.
+of Kleisli category. For this part, we need to define the perturbation on 
+not only the objects itself, but also the global lift and delay of objects, 
+which requires us to provide each piece of supporting constructor. This step 
+and futher steps towards to the model construction are still 
+work-in-progress, but once it's finished, we will provide a complete 
+framework which takes formalization on an explicit type and obtains an 
+extensional model.
 
@@
 Dyn in Cubical Agda and it was more challenging than expected because we 
-define Dyn using the technique of fixed point, which means that every 
-time we analyze the case inside of Dyn, we need to unfold it and add 
-corresponding proof.
+define Dyn using the technique of guarded recursion and fixed point, which 
+means that every time we analyze the case inside of Dyn, we need to unfold 
+it and add corresponding proof.
 

### 2024-01-26 paper-new/concrete-model.tex
work on concrete model

@@
 
-We begin with some definitions. These concepts were first introduced in the section
-on the term semantics (Section \ref{sec:domain-theory}), but for completeness we repeat them here.
-% for ease of reference when instantiating the definition of step-1 model.
+We begin with some definitions:
 
@@
 i.e., $\tilde{x} \le_{\later A} \tilde{x'}$ iff $\later_t(\tilde{x}_t \le_A \tilde{x'}_t)$.
+Likewise for bisimilarity.
+
+We also give a predomain structure to the natural numbers $\mathbb{N}$, where both the
+ordering and the bisimiarity relation are equality.
 
@@
 
+\subsection{Guarded Lift Monad}\label{sec:guarded-lift-monad}
+
 % Lift monad
-The guarded lift monad $\li$ takes a predomain $A$ to the error domain $\li A$;
-the precise definition is given in Section \ref{sec:lift-monad}.
-The partial order $\le_{\li A}$ is the lock-step error ordering defined in Section
-\ref{sec:lock-step}, and the bisimilarity relation $\bisim_{\li A}$ is the weak bisimilarity
-relation defined in Section \ref{sec:weak-bisimilarity}.
-We define the action of $\li$ on a relation $R$ between $A$ and $A'$ to be the
-``heterogeneous" version of the lock-step error lifting of $R$ (see Section \ref{sec:lift-monad}).
+The guarded error-lift monad $\li$ takes a predomain $A$ to the error domain $\li A$.
+It is defined as follows:
+
+\begin{align*}
+  \li A &:= \\
+  &\eta \colon A \to \li A \\
+  &\mho \colon \li A \\
+  &\theta \colon \later (\li A) \to \li A
+\end{align*}
+
+Formally, the lift monad $\li A$ is defined as the solution to the guarded recursive type equation
+
+\[ \li A \cong A + 1 + \later \li A. \]
+
+This captures the intuition that a program may either return a value,
+fail at run-time, or take one or more observable steps of computation.
+Previous work has studied such a similar construct, called the guarded lift
+monad \cite{mogelberg-paviotti2016}; our version here our version augments it with the notion of error.
+
+Since we claimed that $\li A$ is a monad, we need to define the monadic operations
+and show that they respect the monadic laws. The return is just $\eta$, and the monadic extend
+is defined via guarded recursion by cases on the input.
+Verifying that the monadic laws hold uses \lob-induction and is straightforward.
+
+% \eric{Check}
+% We can also show that $\li A$ is the free error- and -later algebra on $A$, in that
+% for any morphism of predomains $f : A \to UB$, there is a unique morphism of error domains
+% $f^* : \li A \to \li A$ extending $f$.
+
+There is a functor $U$ from error domains to predomains that on objects simply returns the
+underlying predomain, and on morphisms returns the underlying morphism of predomains.
+%
+% TODO check this
+
+It is easily verified that $\li A$ is the free error- and later-algebra on the predomain $A$,
+so we have that $\li$ is left-adjoint to $U$.
+
+% i.e., error domain morphisms from $\li A$ to $B$ are in one-to-one correspondence with
+% predomain morphisms from $A$ to $UB$.
+
+% We define $\delta : U(\li A) \To U(\li A)$ by $\delta(x) = \theta(\nxt x)$.
+% We define $\hat{\delta} : \li A \arr \li A$ by $\hat{\delta} = \ext{(\delta \circ \eta)}{}$.
+% Note that by definition of $\text{ext}$, we have that $\hat{\delta}$ is a morphism of error domains.
+
+%\subsubsection{Lock-Step Error Ordering}\label{sec:lock-step}
+
+The partial order $\le_{\li A}$ is the lock-step error ordering defined by guarded recursion as follows:
+
+\begin{itemize}
+    \item 	$\eta\, x \ltls \eta\, y$ if $x \le_A y$.
+    \item 	$\mho \ltls l$ for all $l$ 
+    \item   $\theta\, \tilde{r} \ltls \theta\, \tilde{r'}$ if
+            $\later_t (\tilde{r}_t \ltls \tilde{r'}_t)$
+  \end{itemize}
+
+The idea is that two computations $l$ and $l'$ are related if they are in
+lock-step with regard to their intensional behavior, up to $l$ erroring.
+
+Given a relation $R : A \rel A'$, we define in an analogous manner a heterogeneous
+version of the lock-step error ordering between $\li R : \li A \rel \li A'$.
+% We define the action of $\li$ on a relation $R$ between $A$ and $A'$ to be the
+% ``heterogeneous" version of the lock-step error ordering.
 
@@
 
-There is a functor $U$ from error domains to predomains that on objects simply takes the
-underlying predomain, and on morphisms takes the underlying morphism of predomains.
+%\subsubsection{Weak Bisimilarity}\label{sec:weak-bisimilarity}
+
+For a predomain $A$, we define a relation on $\li A$, called ``weak bisimilarity",
+written $l \bisim l'$. Intuitively, we say $l \bisim l'$ if they are equivalent ``up to delay''.
+The weak bisimilarity relation is defined by guarded recursion as follows:
+
+\begin{align*}
+  &\mho \bisim \mho \\
 %
-% TODO check this
-One can verify that there is a free-forgetful adjunction between the functors $\li$ and $U$;
-this holds because $\li A$ is the initial/free error- and later-algebra on the predomain $A$,
-i.e., error domain morphisms from $\li A$ to $B$ are in one-to-one correspondence with
-predomain morphisms from $A$ to $UB$.
+  &\eta\, x \bisim \eta\, y \text{ if } 
+    x \bisim_A y \\
+%		
+  &\theta\, \tilde{x} \bisim \theta\, \tilde{y} \text{ if } 
+    \later_t (\tilde{x}_t \bisim \tilde{y}_t) \\
+%	
+  &\theta\, \tilde{x} \bisim \mho \text{ if } 
+    \theta\, \tilde{x} = \delta^n(\mho) \text { for some $n$ } \\
+%	
+  &\theta\, \tilde{x} \bisim \eta\, y \text{ if }
+    (\theta\, \tilde{x} = \delta^n(\eta\, x))
+  \text { for some $n$ and $x : \ty{A}$ such that $x \bisim_A y$ } \\
+%
+  &\mho \bisim \theta\, \tilde{y} \text { if } 
+    \theta\, \tilde{y} = \delta^n(\mho) \text { for some $n$ } \\
+%	
+  &\eta\, x \bisim \theta\, \tilde{y} \text { if }
+    (\theta\, \tilde{y} = \delta^n (\eta\, y))
+  \text { for some $n$ and $y : \ty{A}$ such that $x \bisim_A y$ }
+\end{align*}
+
+When both sides are $\eta$, then we ensure that the underlying values are related
+by the bisimilarity relation on $A$.
+When one side is a $\theta$ and the other is $\eta x$ (i.e., one side steps),
+we stipulate that the $\theta$-term runs to $\eta y$ where $x$ is bisimilar to $y$.
+Similarly when one side is $\theta$ and the other $\mho$.
+If both sides step, then we allow one time step to pass and compare the resulting terms.
+In this way, the definition captures the intuition of terms being equivalent up to
+delays.
+
+It can be shown (by \lob-induction) that the step-sensitive relation is symmetric.
+However, it can also be shown that this relation is \emph{not} transitive:
+The argument is the same as that used to show that the step-insensitive error
+ordering $\semltbad$ described above is not transitive. Namely, we show that
+if it were transitive, then it would have to be trivial in that $l \bisim l'$ for all $l, l'$.
+that if this relation were transitive, then it would relate all values of type $\li A$.
 
@@
 
-Given a predomain $A$ and error domain $B$, we define $A \arr B := A \To UB$.
-We note that $A \arr B$ carries a natural error domain structure
+Given $f : A_1' \to A_1$ and $g : A_2 \to A_2'$ we define the predomain morphism
+$f \To g : (A_1 \To A_2) \to (A_1' \To A_2')$ by $\lambda h. \lambda x'. g(h(f(x')))$.
+
+% TODO: include this?
+% The monadic extension operation $\ext{\cdot}{} : (A_1 \To U (\li A_2)) \To (\li A_1 \To U(\li A_2))$
+% is a morphism of predomains from $A_1 \To U(\li A_2)$ to $U(\li A_1) \To U(\li A_2)$, i.e.,
+% it preserves the ordering and bisimilarity relations.
+
+% Given a predomain $A$ and error domain $B$, we define $A \arr B := A \To UB$.
+We note that $A \To UB$ carries a natural error domain structure
 (in the below, the lambda is a meta-theoretic notation):
@@
     \item The $\theta$ operation is defined by
-      \[ \theta_{A \arr B}(\tilde{f}) = \lambda x . \theta_B(\lambda t . \tilde{f}_t(x)). \]
+      \[ \theta_{A \To UB}(\tilde{f}) = \lambda x . \theta_B(\lambda t . \tilde{f}_t(x)). \]
 \end{itemize}
+
+Given a predomain $A$ and error domain $B$, we define
+$A \arr B$ to be the error domain such that $U(A \arr B) = A \To UB$,
+and whose error and $\theta$ operations are as defined above.
+We can define the functorial action of $\arr$ on morphisms
+$f \arr \phi$ in the obvious way.
 
@@
 Lastly, given a relation of predomains $R$ between $A$ and $A'$, and a relation
-of error domains $S$ between $B$ and $B'$, we can form a relation $R \arr S$
+of error domains $S$ between $B$ and $B'$, we define the relation $R \arr S$
 between $A \arr B$ and $A' \arr B'$ in the obvious way, i.e., $f \in A \arr B$
@@
 $x \mathrel{R} x'$, we have $f(x) \mathrel{S} g(x')$.
+%
+One can verify that this relation is indeed a relation of error domains
+in that it respects error and preserves $\theta$.
 
@@
 
-% dynamic type
-The underlying set of the dynamic type will be defined using guarded recursion as the
-solution to the equation
+The predomain representing the dynamic type will be defined using guarded recursion
+as the solution to the equation
 
-\[ D \cong \mathbb{N} + \later (D \to \li D). \]
+\[ D \cong \mathbb{N}\, + (D \times D)\, + \laterhs U(D \arr FD). \]
 
-The perturbation monoid for $D$ is defined inductively as the free monoid $M$ equipped
-with an operation $\To : M \times M \to M$, i.e., we take the set defined inductively by
+% Note that the operators in the above equation are all combinators for predomains, so
+% this also defines the ordering and the bisimilarity relation for $D$.
 
-\begin{mathpar}
-    \inferrule*[]
-    {}
-    {e \in M}
+For the sake of clarity, we name the ``constructors" $\text{nat}$, $\text{times}$,
+and $\text{fun}$, respectively.
 
-    \inferrule*[]
-    {p \in M \and p' \in M}
-    {p \odot p' \in M}
+We define $e_\mathbb{N} : \mathbb{N} \to D$ to be the injection into the first
+component of the sum, and $e_\times : D \times D \to D$ to be the injection into
+the second component of the sum, and $e_\to : U(D \arr F D)$ to be the morphism
+$\nxt$ followed by the injection into the third component of the sum.
 
-    \inferrule*[]
-    {p \in M \and p' \in M}
-    {p \To p' \in M}
-\end{mathpar}
+Explicitly, the ordering on $D$ is given by:
 
-and quotient by the necessary monoid equations.
+\begin{align*}
+    \tnat(n) \le \tnat(n') 
+        &\iff n = n' \\
+    \ttimes (d_1, d_2) \le \ttimes (d_1', d_2')
+        &\iff d_1 \le d_2 \text{ and } d_1' \le d_2'\\
+    \tfun(\tilde{f}) \le \tfun(\tilde{f'}) 
+        &\iff \later_t(\tilde{f}_t \le \tilde{f'}_t)
+\end{align*}
+
+We define a relation $\inat : \mathbb{N} \rel D$ by
+$(n, d) \in \inat$ iff $e_\mathbb{N} \le_D d$.
+We similarly define $\itimes : D \times D \rel D$ by
+$((d_1, d_2), d) \in \itimes$ iff $e_\times(d_1, d_2) \le_D d$,
+and we define $\text{inj}_\to : U(D \arr F D) \rel D$ by
+$(f, d) \in \iarr$ iff $e_\to(f) \le_D d$.
+
+Now we define the perturbations for $D$.
+Recall from our construction of a model with perturbations
+(Section \ref{sec:constructing-perturbations}) that for each value
+type $A$ we associate a commutative monoid $P_A$ of perturbations
+and a homomorphism into the monoid of endomorphisms bisimilar to the identity.
+
+We define the perturbations for $D$ via least-fixpoint in the category of
+commutative monoids as
+
+\[ P_D \cong (P_D \times P_D) \times (P_D \times (\mathbb{N} \times P_D)), \]
+
+where $\oplus$ is the product (also the coproduct) of commutative monoids.
+
+We now explain how to interpret these perturbations as endomorphisms.
+We define $\ptb_D : P_D \to \{ f : D \to D \mid f \bisim \id \}$ as in the below.
+Note the use of the functorial action of $\arr$ on morphisms.
+% via the universal property of the coproduct of monoids, giving a case for each of the generators.
+% In the below, note the use of the functorial action of $\arr$ on morphisms.
+
+%
+% \[ P_D \cong (P_D \times P_D) \times (P_D \times (\mathbb{N} \times P_D)), \]
+%
+
+$\ptb_D(p_1, p_2, p_3, n, p_4) = \lambda d.\text{case $d$ of}$
+\begin{align*}
+    &\alt \tnat(m) \mapsto \tnat(m) \\
+    &\alt \ttimes(d_1, d_2) \mapsto {\ttimes(\ptb_D(p_1)(d_1), \ptb_D(p_2)(d_2))} \\
+    &\alt \tfun(\tilde{f}) \mapsto {\tfun(\lambda t. (\ptb_D(p_3) \arr (\delta^n \circ F(\ptb_D(p_4))))(\tilde{f}_t))} \\
+\end{align*}
+
+One can verify that this forms a homomorphism from $P_D \to \{ f : D \to D : f \bisim \id \}$.
+For this it is easiest to use the fact that the product coindices with the coproduct
+in the category of commutative monoids, and thus it suffices to check the condition for each summand separately.
+
+We claim that the three relations $\inat$, $\itimes$, and $\iarr$
+%and their lifted versions 
+satisfy the push-pull property.
+As an illustrative case, we establish the pull property for the relation $\iarr$.
+
+We define $\pull_{\iarr} : P_D \to P_{U(D \arr FD)}$ by cases as
+
+\[ \pull_{\iarr}(p_1, p_2, p_3, n, p_4) = (0, p_3, n, p_4) \]
+
+(recall that $P_{UB} = \mathbb{N} \times P_B$, and $P_{A \arr B} = P_A \times P_B$,
+so $P_{U(D \arr FD)}) = \mathbb{N} \times P_D \times \mathbb{N} \times P_D$.
+
+Now let $(f, d) \in \iarr$. This means that $d$ must be of the form $\tfun{\tilde{f}}$ where
+Then we have $\ptb_{U(D \arr FD)}(0, p_3, n, p_4) = $
+
+\eric{Details}
+
+\eric{Quasi-representability}
+
+    % \item $\ptb_D(1)$
+    % \item $\ptb_D(\delta^K_D)$ is defined similarly to the previous but has
+    % \[ \id \arr i^K(\delta^K_D) \] instead.
+    % \item $\ptb_D(\delta^K_D) = \lambda d.\text{case $d$ of}$
+    %   \begin{align*} 
+    %     &\alt \ttimes(d_1, d_2) \mapsto {\ttimes(i^K(\delta^K_D)(d_1), d_2)} \\
+    %     &\alt d' \to d'
+    %   \end{align*}
+    % \item $\ptb_D(\delta^K_D)$ is defined similarly to the previous but has 
+    % \[ (d_1, \ptb_D(\delta^K_D)(d_2)) \] instead.
+
+\begin{comment}
+Before defining the perturbations, recall from our concrete construction
+(Section \ref{TODO}) that to each value type $A$ we associate a monoid $P_A$
+of pure perturbations and a monoid $P^K_{A}$ of Kleisli perturbations.
+
+We define simultaneously the pure and Kleisli perturbation monoids for $D$,
+denoted $P_D$ and $P^K_D$ respectively.
+We define $P_D$ to be the monoid such that
+
+\[ P_D \cong ((P^K_D)^{op} \times P_D) \oplus (P_D \times P_D), \]
+
+and we define $P^K_D$ to be the monoid such that
+
+\[ P^K_D \cong \mathbb{N} \oplus ((P_D)^{op} \oplus P^K_D) \oplus (P^K_D \oplus P^K_D), \]
+
+where we use $\mathbb{N}$ because it is the free monoid on one generator.
+
+Now we explain how to interpret these monoids as submonoids of endomorphisms.
+We define $i : P_D \to \{ f : D \to D \mid f \bisim \id \}$
+and $i^K : P^K_D \to \{ \phi : FD \to FD \mid \phi \bisim \id \}$ mutually
+via the universal property of the coproduct of monoids, giving a case for each of the generators.
+In the below, note the use of the functorial action of $\arr$ on morphisms.
+
+\begin{itemize}
+    \item $i(\delta^K_D, \delta_D) = \lambda d.\text{case $d$ of}$
+    \begin{align*}
+        &\alt \tfun(\tilde{f}) \mapsto 
+          \tfun(\lambda t. (i^K(\delta^K_D) \to i(\delta_D))(\tilde{f}_t)) \\
+        &\alt d' \to d'
+    \end{align*}
+
+    \item $i(\delta_D^1, \delta_D^2) = \lambda d.\text{case $d$ of}$ 
+    \begin{align*}
+        &\alt \ttimes(d_1, d_2) \mapsto
+           \ttimes(i(\delta_D^1)(d_1), i(\delta_D^2)(d_2)) \\
+        &\alt d' \to d'
+    \end{align*}
+\end{itemize}
+
+We define $i^K(\delta^D)$ by
+%
+%  P^K_D \cong \mathbb{N} \oplus ((P_D)^{op} \oplus P^K_D) \oplus (P^K_D \oplus P^K_D)
+%
+\begin{itemize}
+    \item $i^K(1) = \hat{\delta}$
+    \item $i^K(\delta_D) = F(\lambda d.\text{case $d$ of}$
+      \begin{align*} 
+        &\alt \tfun(\tilde{f}) \mapsto {\tfun(\lambda t. (i(\delta_D) \arr \id)(\tilde{f}_t))} \\
+        &\alt d' \to d')
+      \end{align*}
+    \item $i^K(\delta^K_D)$ is defined similarly to the previous but has
+    \[ \id \arr i^K(\delta^K_D) \] instead.
+    \item $i^K(\delta^K_D) = F(\lambda d.\text{case $d$ of}$
+      \begin{align*} 
+        &\alt \ttimes(d_1, d_2) \mapsto {\ttimes(i^K(\delta^K_D)(d_1), d_2)} \\
+        &\alt d' \to d')
+      \end{align*}
+    \item $i^K(\delta^K_D)$ is defined similarly to the previous but has 
+    \[ (d_1, i^K(\delta^K_D)(d_2)) \] instead.
+\end{itemize}
 
 % perturbations and quasi-representability
+We claim that the three relations $\inat$, $\itimes$, and $\iarr$ and their lifted
+versions satisfy the push-pull property.
+As an illustrative case, we establish the pull property for the $\li \iarr$.
+We define $\pull : P^K_D \to P^K_{U(D \arr FD)}$ by cases as
 
-\subsection{Obtaining an Extensional Model}
+\begin{align*}
+ \pull(1) = \hat{\delta}
+\end{align*}
 
-Now that we have defined an intensional model, we can apply
+We need to check that the relevant squares commute for all $\delta \in P^K_D$.
+It suffices by the universal property of the coproduct of monoids to ensure that
+this holds for the generators.
+
+We claim that the relations $\inat$, $\itimes$, and $\iarr$ are quasi-left-representable,
+and that their lifts are quasi-right-representable.
+Indeed, since the relations are functional, it is easy to see that they are left-representable
+where the perturbations are simply the identity.
+
+The most interesting case is the right-quasi-representability of $\li(\iarr)$.
+We define the projection as follows:
+
+\begin{align*}
+\end{align*}
+
+And for the perturbations, we take
+
+% The perturbation monoid for $D$ is defined inductively as the free monoid $M$ equipped
+% with an operation $\To : M \times M \to M$, i.e., we take the set defined inductively by
+
+% \begin{mathpar}
+%     \inferrule*[]
+%     {}
+%     {e \in M}
+
+%     \inferrule*[]
+%     {p \in M \and p' \in M}
+%     {p \odot p' \in M}
+
+%     \inferrule*[]
+%     {p \in M \and p' \in M}
+%     {p \To p' \in M}
+% \end{mathpar}
+
+% and quotient by the necessary monoid equations.
+
+\end{comment}
+
+% \subsection{Obtaining an Extensional Model}
+
+Now that we have defined an intensional model with an interpetation for the dynamic type, we can apply
 the abstract construction introduced in Section \ref{sec:extensional-model-construction}.
-Doing so, we obtain an extensional model of gradual typing.
+Doing so, we obtain an extensional model of gradual typing, where the squares are given by the
+``bisimilarity closure'' of the intensional error ordering.

### 2024-01-26 paper-new/categorical-model.tex
updates mostly to extensional model

@@
-\section{Abstract Categorical Models of Graduality}\label{sec:abstract-models}
+\section{Revised Categorical Models of Graduality}\label{sec:abstract-models}
 
-First, what is a categorical model of call-by-push-value? We will use
-the following notion as our basic structure\footnote{There are models
-that are closer to syntax e.g., which distinguish between value types
-and contexts but this suffices for all of the models we consider}:
-\begin{enumerate}
-\item A cartesian category $\mathcal V$
-\item A category $\mathcal E$
-\item An action of $\mathcal V^{op}$ (with the $\mathcal V$ cartesian
-  product as monoidal structure) on $\mathcal E$. We write this with
-  an arrow $A \arr B$.
-
-  This means we have 
-  \[ \alpha : {A_1 \times A_2} \arr B \cong A_2 \arr (A_1 \arr B) \]
-  and
-  \[ i : 1 \arr B \cong B \]
-  satisfying coherence isomorphisms.
-\item A functor $U : \mathcal E \to \mathcal V$ that ``preserves
-  powering'' in that every $U(A \arr B)$ is an exponential of $UB$ by $A$
-  and that $U\alpha$ and $Ui$ are mapped to the canonical isomorphisms
-  for exponentials.
-\item A left adjoint $F \dashv U$
-\end{enumerate}
-
-\begin{example}
-  Given a strong monad $T$ on a bicartesian closed category $\mathcal
-  V$, we can extend this to a CBPV model by defining $\mathcal E$ to
-  be the category $\mathcal V^T$ of algebras, defining $A \to B$ as
-  the powering of algebras, $F$ as the free algebra and $U$ as the
-  underlying object functor.
-\end{example}
-
-There are two relevant notions of \emph{morphism} of CBPV models:
-\emph{strict} and \emph{lax}.
-% morphisms of CBPV models
-Given call-by-push-value models
-$\mathcal M_1 = (\mathcal V_1, \mathcal E_1, \arr_1, U_1, F_1)$ and
-$\mathcal M_2 = (\mathcal V_2, \mathcal E_2, \arr_2, U_2, F_2)$,
-A \emph{strict} morphism $G$ from $M_1$ to $M_2$ is given by a pair of functors
-$G_{\mathcal{V}}: V_1 \to V_2$ and $G_{\mathcal{E}} : E_1 \to E_2$
-that strictly presere the constructors:
-\begin{enumerate}
-  \item $G_{\mathcal{E}} \circ F_1 = F_2 \circ G_{\mathcal{V}}$
-  \item $G_{\mathcal{V}} \circ U_1 = U_2 \circ G_{\mathcal{E}}$
-  \item $G_{\mathcal{E}}(A \arr_1 B) = G_{\mathcal{V}}(A) \arr_2 G_{\mathcal{E}}(B)$
-  \item $G_{\mathcal{V}}(A_1 \times_1 A_2) = G_{\mathcal{V}}(A_1) \times_2 G_{\mathcal{V}}(A_2)$
-  \item $G_{\mathcal{V}}(1_1) = 1_2$
-\end{enumerate}
-As well as strictly preserving the corresponding universal morphisms
-and coherence isomorphisms.
-
-A lax morphism instead preserves these only up to transformation
-\begin{enumerate}
-  \item $G_{\mathcal{E}} \circ F_1 \Rightarrow F_2 \circ G_{\mathcal{V}}$
-  \item $G_{\mathcal{V}} \circ U_1 \Rightarrow U_2 \circ G_{\mathcal{E}}$
-  \item $G_{\mathcal{E}}(A \arr_1 B) \Rightarrow G_{\mathcal{V}}(A) \arr_2 G_{\mathcal{E}}(B)$
-  \item $G_{\mathcal{V}}(A_1 \times_1 A_2) \Rightarrow G_{\mathcal{V}}(A_1) \times_2 G_{\mathcal{V}}(A_2)$
-  \item $G_{\mathcal{V}}(1_1) \Rightarrow 1_2$
-\end{enumerate}
-Additionally a lax morphism should have a relationship between these
-transformations and the universal morphisms, but we will only consider
-lax morphisms of thin categories, where such conditions hold
-trivially.
-% TODO what about the product and coproducts in V?
-% Do we need that G_V(A \times_1 A') = G_V(A) \times_2 G_V(A')
-% and likewise for coproducts?
-
-With these definitions, we can define categories $\textrm{CBPV}_s$ and
-$\textrm{CBPV}_l$ whose objects are CBPV models, and whose arrows are
-CBPV morphisms.  This category will serve as the setting in which we
-formulate the definition of a model of extensional gradual typing.
+Next, we develop our appropriate weakened notions of models of
+graduality, which we divide into \emph{extensional} models, where
+ordering is up to weak bisimilarity, and \emph{intensional} models,
+where ordering relates terms that considers computational steps to be
+observable. We develop the notion of intensional model of gradual
+typing in stages and show how to develop one from a base model of
+effectful functions and relations.
 
@@
 
-An model $\mathcal{M}$ of extensional gradual typing consists of:
+Since we lack transitivity of ordering when reasoning in guarded type
+theory, our weakened notion of extensional model is based on reflexive
+graph categories rather than double categories. This means we lose the
+reasoning principle of horizontal pasting of squares. We will still
+require a notion of \emph{composition} of relations, to model the
+transitivity of type precision. We note that without horizontal
+pasting of squares, the notion of left/right representability of
+squares is not sufficient to interpret the cast rules of gradual
+typing. Instead we generalize it slightly as follows.
+
+Let $c : A \rel A'$ and $f : A \to A'$ in a reflexive graph category
+with composition of relations. We say that $c$ is \emph{universally
+left-representable by} $f$ if for any $c_l : A_l \rel A$ and $c_r : A'
+\rel A_r$ we have $f \ltsq{cc_r}{c_r} \id$ and $\id \ltsq{c_l}{c_lc}
+f$. Dually, let $d : B \rel B'$ and $g : B' \to B$.  We say that $d$
+is \emph{universally right-representable by} $g$ if for any $d_l : B_l
+\rel B$ and $d_r : B' \rel B_r$ we have $\id \ltsq{d_ld}{d_l} \phi$
+and $\phi \ltsq{d_r}{dd_r}\id$. In a reflexive graph category these
+are stronger than the definitions of left/right representability since
+we can pick $c_l,c_r,d_l,d_r$ to be the reflexive relations. In a
+double category these are equivalent, but the equivalence uses
+horizontal pasting.
+
+Additionally, while in the presence of horizontal pasting
+compositionality of representable morphisms is automatic, without this
+principle we must require it explicitly. Finally, we explicitly add in
+a requirement that the type constructors are functorial in the
+
+In summary, an extensional model consists of:
+\begin{enumerate}
+  \item A locally thin reflexive graph internal to CBPV models.
+  \item Composition of value and computation relations that form a category with the reflexive relations as identity.
+  \item An identity-on-objects functor $\upf : \mathcal V_e \to \mathcal V_f$ taking each value relation to a morphism that universally left-represents it.
+  \item An identity-on-objects functor $\dnf : \mathcal E_e^{op} \to \mathcal E_f$ taking each computation relation to a morphism that universally right-represents it.
+  \item The CBPV connectives $U,F,\times,\to$ are all \emph{covariant} functorial on relations\footnote{the reflexive graph structure already requires that these functors preserve identity relations}
     \begin{itemize}
-  \item CBPV models $\mathcal M_f$ and $\mathcal M_{sq}$
-  \item CBPV morphisms $r : \mathcal M_f \to \mathcal M_{sq}$ and 
-  $s, t : \mathcal M_{sq} \to \mathcal M_f$
+    \item $U(dd') \equidyn U(d)U(d')$
+    \item $F(cc') \equidyn F(c)F(c')$
+    \item $(cc') \to (dd') \equidyn (c \to d)(c' \to d')$
+    \item $(c_1c_1') \times (c_2c_2') \equidyn (c_1 \times c_2)(c_1'\times c_2')$
     \end{itemize}
+    where $c \equidyn c'$ means $\id \ltsq{c}{c'}\id$ and $\id \ltsq{c'}{c} \id$.
+  \item Distinguished value type $\nat$ with morphisms $z : \mathcal
+    V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
+  \item Distinguished value types $D$ with distinguished relations
+    $\iarr{}: U(D \to F D) \rel D$ and $\inat : \nat \rel D$ and $\itimes : D \times D \rel D$
+    each satisfying the retraction property $\dnc {\injarr{}}F(\upc{\injarr{}}) \equidyn \id$.
+\end{enumerate}
 
-satisfying certain additional conditions that will be described below.
+%% %% A model $\mathcal{M}$ of extensional gradual typing consists of the
+%% %% following:
+%% \begin{itemize}
+%% \item A reflexive graph internal to the category of CBPV models and
+%%   strict morphisms that is locally thin: there is at most one morphism
+%%   in $\mathcal V_{sq}$ for any given domain, codomain, source and
+%%   target, and as well for $\mathcal E$.
+%% \item 
+%% \item A natural transformation $\mho : 1 \Rightarrow U$ satisfying $\mho \circ ! \ltsq{r(A)}{r(B)} f$ for every $f$.
+%% \item Such that all relation value morphisms are \emph{universally}
+%%   left representable and all relation computation morphisms are
+%%   universally right representable.
+%% \end{itemize}
 
-% % https://q.uiver.app/#q=WzAsMixbMCwxLCJcXG1hdGhjYWx7TX1fe3NxfSJdLFswLDAsIlxcbWF0aGNhbHtNfV9mIl0sWzEsMCwiciJdLFswLDEsInMiLDAseyJjdXJ2ZSI6LTJ9XSxbMCwxLCJ0IiwyLHsiY3VydmUiOjJ9XV0=
-% \[\begin{tikzcd}[ampersand replacement=\&]
-% 	{\mathcal{M}_f} \\
-% 	{\mathcal{M}_{sq}}
-% 	\arrow["r", from=1-1, to=2-1]
-% 	\arrow["s", curve={height=-12pt}, from=2-1, to=1-1]
-% 	\arrow["t"', curve={height=12pt}, from=2-1, to=1-1]
-% \end{tikzcd}\]
+%% satisfying certain additional conditions that will be described below.
 
-Spelling this out in light of the above definitions, we see that this is
-equivalent to the following in the category $\textbf{Cat}$:
+%% % % https://q.uiver.app/#q=WzAsMixbMCwxLCJcXG1hdGhjYWx7TX1fe3NxfSJdLFswLDAsIlxcbWF0aGNhbHtNfV9mIl0sWzEsMCwiciJdLFswLDEsInMiLDAseyJjdXJ2ZSI6LTJ9XSxbMCwxLCJ0IiwyLHsiY3VydmUiOjJ9XV0=
+%% % \[\begin{tikzcd}[ampersand replacement=\&]
+%% % 	{\mathcal{M}_f} \\
+%% % 	{\mathcal{M}_{sq}}
+%% % 	\arrow["r", from=1-1, to=2-1]
+%% % 	\arrow["s", curve={height=-12pt}, from=2-1, to=1-1]
+%% % 	\arrow["t"', curve={height=12pt}, from=2-1, to=1-1]
+%% % \end{tikzcd}\]
 
-% https://q.uiver.app/#q=WzAsNCxbMCwyLCJcXHZzcSJdLFsyLDIsIlxcZXNxIl0sWzAsMCwiXFx2ZiJdLFsyLDAsIlxcZWYiXSxbMiwzLCJcXEZmIiwwLHsiY3VydmUiOi0yfV0sWzMsMiwiXFxVZiIsMCx7ImN1cnZlIjotMn1dLFswLDEsIlxcRnNxIiwwLHsiY3VydmUiOi0yfV0sWzEsMCwiXFxVc3EiLDAseyJjdXJ2ZSI6LTJ9XSxbMiwwLCJcXHJ2Il0sWzAsMiwiXFxzdiIsMCx7ImN1cnZlIjotMn1dLFsyLDAsIlxcdHYiLDAseyJjdXJ2ZSI6LTJ9XSxbMSwzLCJcXHNlIiwwLHsiY3VydmUiOi0yfV0sWzMsMSwiXFx0ZSIsMCx7ImN1cnZlIjotMn1dLFszLDEsIlxccmUiXSxbNCw1LCJcXGJvdCIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs2LDcsIlxcYm90IiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
-\[\begin{tikzcd}[ampersand replacement=\&]
-	\vf \&\& \ef \\
-	\\
-	\vsq \&\& \esq
-	\arrow[""{name=0, anchor=center, inner sep=0}, "\Ff", curve={height=-12pt}, from=1-1, to=1-3]
-	\arrow[""{name=1, anchor=center, inner sep=0}, "\Uf", curve={height=-12pt}, from=1-3, to=1-1]
-	\arrow[""{name=2, anchor=center, inner sep=0}, "\Fsq", curve={height=-12pt}, from=3-1, to=3-3]
-	\arrow[""{name=3, anchor=center, inner sep=0}, "\Usq", curve={height=-12pt}, from=3-3, to=3-1]
-	\arrow["\rv", from=1-1, to=3-1]
-	\arrow["\sv", curve={height=-12pt}, from=3-1, to=1-1]
-	\arrow["\tv", curve={height=-12pt}, from=1-1, to=3-1]
-	\arrow["\se", curve={height=-12pt}, from=3-3, to=1-3]
-	\arrow["\te", curve={height=-12pt}, from=1-3, to=3-3]
-	\arrow["\re", from=1-3, to=3-3]
-	\arrow["\bot"{description}, draw=none, from=0, to=1]
-	\arrow["\bot"{description}, draw=none, from=2, to=3]
-\end{tikzcd}\]
+%% Spelling this out in light of the above definitions, we see that this is
+%% equivalent to the following in the category $\textbf{Cat}$:
+
+%% % https://q.uiver.app/#q=WzAsNCxbMCwyLCJcXHZzcSJdLFsyLDIsIlxcZXNxIl0sWzAsMCwiXFx2ZiJdLFsyLDAsIlxcZWYiXSxbMiwzLCJcXEZmIiwwLHsiY3VydmUiOi0yfV0sWzMsMiwiXFxVZiIsMCx7ImN1cnZlIjotMn1dLFswLDEsIlxcRnNxIiwwLHsiY3VydmUiOi0yfV0sWzEsMCwiXFxVc3EiLDAseyJjdXJ2ZSI6LTJ9XSxbMiwwLCJcXHJ2Il0sWzAsMiwiXFxzdiIsMCx7ImN1cnZlIjotMn1dLFsyLDAsIlxcdHYiLDAseyJjdXJ2ZSI6LTJ9XSxbMSwzLCJcXHNlIiwwLHsiY3VydmUiOi0yfV0sWzMsMSwiXFx0ZSIsMCx7ImN1cnZlIjotMn1dLFszLDEsIlxccmUiXSxbNCw1LCJcXGJvdCIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs2LDcsIlxcYm90IiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+%% \[\begin{tikzcd}[ampersand replacement=\&]
+%% 	\vf \&\& \ef \\
+%% 	\\
+%% 	\vsq \&\& \esq
+%% 	\arrow[""{name=0, anchor=center, inner sep=0}, "\Ff", curve={height=-12pt}, from=1-1, to=1-3]
+%% 	\arrow[""{name=1, anchor=center, inner sep=0}, "\Uf", curve={height=-12pt}, from=1-3, to=1-1]
+%% 	\arrow[""{name=2, anchor=center, inner sep=0}, "\Fsq", curve={height=-12pt}, from=3-1, to=3-3]
+%% 	\arrow[""{name=3, anchor=center, inner sep=0}, "\Usq", curve={height=-12pt}, from=3-3, to=3-1]
+%% 	\arrow["\rv", from=1-1, to=3-1]
+%% 	\arrow["\sv", curve={height=-12pt}, from=3-1, to=1-1]
+%% 	\arrow["\tv", curve={height=-12pt}, from=1-1, to=3-1]
+%% 	\arrow["\se", curve={height=-12pt}, from=3-3, to=1-3]
+%% 	\arrow["\te", curve={height=-12pt}, from=1-3, to=3-3]
+%% 	\arrow["\re", from=1-3, to=3-3]
+%% 	\arrow["\bot"{description}, draw=none, from=0, to=1]
+%% 	\arrow["\bot"{description}, draw=none, from=2, to=3]
+%% \end{tikzcd}\]
 
@@
 
-That is, for the values, we have
+%% That is, for the values, we have
 
-\begin{enumerate}
-  \item A cartesian category $\mathcal V_f$.
-  The objects of $\mathcal V_f$ will be called \emph{value types}.
-  The morphisms of $\mathcal V_f$ will be called \emph{(pure) functions}.
+%% \begin{enumerate}
+%%   \item A cartesian category $\mathcal V_f$.
+%%   The objects of $\mathcal V_f$ will be called \emph{value types}.
+%%   The morphisms of $\mathcal V_f$ will be called \emph{(pure) functions}.
 
-  \item A cartesian category $\mathcal V_{sq}$. 
-  The objects of $\mathcal V_{sq}$ will be called \emph{value edges} or
-  \emph{value relations}, and the morphisms are \emph{commuting squares}.
+%%   \item A cartesian category $\mathcal V_{sq}$. 
+%%   The objects of $\mathcal V_{sq}$ will be called \emph{value edges} or
+%%   \emph{value relations}, and the morphisms are \emph{commuting squares}.
 
-  \item Functors $\sv, \tv : \mathcal V_{sq} \to \mathcal V_f$ and 
-  $\rv : \mathcal V_f \to \mathcal V_{sq}$.
-\end{enumerate}
+%%   \item Functors $\sv, \tv : \mathcal V_{sq} \to \mathcal V_f$ and 
+%%   $\rv : \mathcal V_f \to \mathcal V_{sq}$.
+%% \end{enumerate}
 
-Likewise, we have the analogous definitions for computations.
+%% Likewise, we have the analogous definitions for computations.
 
-We write $c : A \rel A'$ to mean that $c \in \ob(\vsq)$ such that 
-$\sv(c) = A$ and $\tv(c) = A'$.
-%
-Likewise, let $c_i : A_i \rel A_i'$ and $c_o : A_o \rel A_o'$,
-and let $f \in \vf(A_i, A_o)$ and $f' \in \vf(A_i', A_o')$.
-The notation $\beta : f \ltdyn_{c_o}^{c_i} f'$ is defined to mean
+%% We write $c : A \rel A'$ to mean that $c \in \ob(\vsq)$ such that 
+%% $\sv(c) = A$ and $\tv(c) = A'$.
+%% %
+%% Likewise, let $c_i : A_i \rel A_i'$ and $c_o : A_o \rel A_o'$,
+%% and let $f \in \vf(A_i, A_o)$ and $f' \in \vf(A_i', A_o')$.
+%% The notation $\beta : f \ltdyn_{c_o}^{c_i} f'$ is defined to mean
 
-\begin{enumerate}
-  \item $\beta \in \vsq(c_i, c_o)$
-  \item $\sv(\beta) = f$
-  \item $\tv(\beta) = f'$
-\end{enumerate}
+%% \begin{enumerate}
+%%   \item $\beta \in \vsq(c_i, c_o)$
+%%   \item $\sv(\beta) = f$
+%%   \item $\tv(\beta) = f'$
+%% \end{enumerate}
 
-(Recall that $\sv$ and $\tv$ are functors, so in addition to acting on
-the objects of $\vsq$ they also act on morphisms.)
+%% (Recall that $\sv$ and $\tv$ are functors, so in addition to acting on
+%% the objects of $\vsq$ they also act on morphisms.)
 
-Picorially, this is depicted as a commuting square:
+%% Picorially, this is depicted as a commuting square:
 
-% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBX2kiXSxbMSwwLCJBX2knIl0sWzAsMSwiQV9vIl0sWzEsMSwiQV9vJyJdLFswLDIsImYiLDJdLFsxLDMsImYnIl0sWzAsMSwiY19pIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMiwzLCJjX28iLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs0LDUsIlxcYWxwaGEiLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
-\[\begin{tikzcd}[ampersand replacement=\&]
-	{A_i} \& {A_i'} \\
-	{A_o} \& {A_o'}
-	\arrow[""{name=0, anchor=center, inner sep=0}, "f"', from=1-1, to=2-1]
-	\arrow[""{name=1, anchor=center, inner sep=0}, "{f'}", from=1-2, to=2-2]
-	\arrow["{c_i}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-	\arrow["{c_o}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-	\arrow["\alpha"{description}, draw=none, from=0, to=1]
-\end{tikzcd}\]
+%% % https://q.uiver.app/#q=WzAsNCxbMCwwLCJBX2kiXSxbMSwwLCJBX2knIl0sWzAsMSwiQV9vIl0sWzEsMSwiQV9vJyJdLFswLDIsImYiLDJdLFsxLDMsImYnIl0sWzAsMSwiY19pIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMiwzLCJjX28iLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs0LDUsIlxcYWxwaGEiLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
+%% \[\begin{tikzcd}[ampersand replacement=\&]
+%% 	{A_i} \& {A_i'} \\
+%% 	{A_o} \& {A_o'}
+%% 	\arrow[""{name=0, anchor=center, inner sep=0}, "f"', from=1-1, to=2-1]
+%% 	\arrow[""{name=1, anchor=center, inner sep=0}, "{f'}", from=1-2, to=2-2]
+%% 	\arrow["{c_i}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+%% 	\arrow["{c_o}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+%% 	\arrow["\alpha"{description}, draw=none, from=0, to=1]
+%% \end{tikzcd}\]
 
-% Note: When the identity of the square $\beta$ is not important, we may omit it
-% and write $f \ltdyn_{c_o}^{c_i} f'$. In this case the meaning is that there exists
-% a square $\beta : f \ltdyn_{c_o}^{c_i} f'$.
+%% % Note: When the identity of the square $\beta$ is not important, we may omit it
+%% % and write $f \ltdyn_{c_o}^{c_i} f'$. In this case the meaning is that there exists
+%% % a square $\beta : f \ltdyn_{c_o}^{c_i} f'$.
 
-Composition of squares $\beta : f \ltdyn_{c_2}^{c_1} g$ and $\beta' : f' \ltdyn_{c_3}^{c_2} g'$
-corresponds to ``stacking'' the square for $\beta'$ below the square for $\beta$.
-Fuctoriality of $s$ and $t$ ensure that the left and right sides of the resulting square are as expected,
-i.e., we get $\beta' \circ \beta : f' \circ f \ltdyn_{c_3}^{c_1} g' \circ g$.
+%% Composition of squares $\beta : f \ltdyn_{c_2}^{c_1} g$ and $\beta' : f' \ltdyn_{c_3}^{c_2} g'$
+%% corresponds to ``stacking'' the square for $\beta'$ below the square for $\beta$.
+%% Fuctoriality of $s$ and $t$ ensure that the left and right sides of the resulting square are as expected,
+%% i.e., we get $\beta' \circ \beta : f' \circ f \ltdyn_{c_3}^{c_1} g' \circ g$.
 
-% Fuctoriality of $s$ and $t$ ensure that we can ``vertically" compose 
-% $\beta : f \ltdyn_{c_2}^{c_1} g$ and $\beta' : f' \ltdyn_{c_3}^{c_2} g'$
-% to obtain $\beta' \circ \beta : f' \circ f \ltdyn_{c_3}^{c_1} g' \circ g$.
-% Pictorially, this is represented by ``stacking'' the square for
-% $\beta'$ below the square for $\beta$.
+%% % Fuctoriality of $s$ and $t$ ensure that we can ``vertically" compose 
+%% % $\beta : f \ltdyn_{c_2}^{c_1} g$ and $\beta' : f' \ltdyn_{c_3}^{c_2} g'$
+%% % to obtain $\beta' \circ \beta : f' \circ f \ltdyn_{c_3}^{c_1} g' \circ g$.
+%% % Pictorially, this is represented by ``stacking'' the square for
+%% % $\beta'$ below the square for $\beta$.
 
-All of the above holds in an analogous manner for the computations.
+%% All of the above holds in an analogous manner for the computations.
 
-We will work in ``locally thin'' models where there is at most one
-square with a given boundary.
-That is, if $\beta, \beta' : f \ltdyn_{c_o}^{c_i} f'$, then $\beta = \beta'$.
-Thus, we may unambiguously omit the identity of the square, i.e., we may write
-$f \ltdyn_{c_o}^{c_i} f'$.
+%% We will work in ``locally thin'' models where there is at most one
+%% square with a given boundary.
+%% That is, if $\beta, \beta' : f \ltdyn_{c_o}^{c_i} f'$, then $\beta = \beta'$.
+%% Thus, we may unambiguously omit the identity of the square, i.e., we may write
+%% $f \ltdyn_{c_o}^{c_i} f'$.
 
-We also have a ``horizontal" composition operation on value edges and on computation edges.
-That is, let 
-%
-\[ X = \{ (c, c') \in \ob(\vsq) \times \ob(\vsq) \mid \tv(c) = \sv(c') \}. \]
-%
-There is an operation $\comp : X \to \ob(\vsq)$ such that $\sv(c \comp c') = \sv(c)$
-and $\tv(c \comp c') = \tv(c')$. Likewise for computations.
-%
-Importantly, we emphasize that this composition is \emph{not} a functor: we only
-require that it act on \emph{objects} of $\vsq$ (i.e. edges) and not on the morphisms
-(i.e. the squares). Intuivitely, this has to do with the fact that in the extensional
-setting, the semantic term precision function is \emph{not} transitive.
+%% We also have a ``horizontal" composition operation on value edges and on computation edges.
+%% That is, let 
+%% %
+%% \[ X = \{ (c, c') \in \ob(\vsq) \times \ob(\vsq) \mid \tv(c) = \sv(c') \}. \]
+%% %
+%% There is an operation $\comp : X \to \ob(\vsq)$ such that $\sv(c \comp c') = \sv(c)$
+%% and $\tv(c \comp c') = \tv(c')$. Likewise for computations.
+%% %
+%% Importantly, we emphasize that this composition is \emph{not} a functor: we only
+%% require that it act on \emph{objects} of $\vsq$ (i.e. edges) and not on the morphisms
+%% (i.e. the squares). Intuivitely, this has to do with the fact that in the extensional
+%% setting, the semantic term precision function is \emph{not} transitive.
 
-With this definition, is easily shown that there is a category $\ve$ of value relations,
-whose objects are the objects of $\vf$, and such that $\ve(A, A')$ is the set of objects
-$c$ of $\vsq$ whose source is $A$ and whose target is $A'$.
-Composition of morphisms is defined using the above operation $\comp$.
-Similarly, we have a category $\ee$ of computation relations.
+%% With this definition, is easily shown that there is a category $\ve$ of value relations,
+%% whose objects are the objects of $\vf$, and such that $\ve(A, A')$ is the set of objects
+%% $c$ of $\vsq$ whose source is $A$ and whose target is $A'$.
+%% Composition of morphisms is defined using the above operation $\comp$.
+%% Similarly, we have a category $\ee$ of computation relations.
 
-We also require that the category $\ve$ of relations is thin ``up to an
-identity square'', i.e., for any $c, c' \in \ve(A, A')$ we have that the
-following square commutes:
+%% We also require that the category $\ve$ of relations is thin ``up to an
+%% identity square'', i.e., for any $c, c' \in \ve(A, A')$ we have that the
+%% following square commutes:
 
-\[\begin{tikzcd}[ampersand replacement=\&]
-  A \& {A'} \\
-  A \& {A'}
-  \arrow[from=1-1, to=2-1, Rightarrow, no head]
-  \arrow[from=1-2, to=2-2, Rightarrow, no head]
-  \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-  \arrow["c'"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-\end{tikzcd}\]
+%% \[\begin{tikzcd}[ampersand replacement=\&]
+%%   A \& {A'} \\
+%%   A \& {A'}
+%%   \arrow[from=1-1, to=2-1, Rightarrow, no head]
+%%   \arrow[from=1-2, to=2-2, Rightarrow, no head]
+%%   \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+%%   \arrow["c'"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+%% \end{tikzcd}\]
 
@@
 
-Finally, we specify the abstract behavior of casts.
-In brief, given $c : A \rel A'$, an upcast will be a morphism $\upf\, c$ in $\vf(A, A')$
-such that $\upf\, c$ ``represents $c$" (we will define this shortly). % such that $c$ is \emph{left-representable} by $\up c$.
-In reality, it is slightly more complicated, since we need to build composition of edges
-into the definition, but the idea is similar.
-
-In particular, let $c : A \rel A'$ and $f \in \mathcal V_f(A, A')$. We
-say that $c$ is \emph{left-representable by} $f$ if for any $c_l : A_l
-\rel A$ and $c_r : A' \rel A_r$ the following squares commute\max{is ``commute'' good terminology?}:
-\begin{center}
-  \begin{tabular}{ m{14em} m{14em} } 
-    % UpL
-    \begin{tikzcd}[ampersand replacement=\&]
-      A \& {A'} \& {A_r} \\
-      {A'} \&\& {A_r}
-      \arrow["f"', from=1-1, to=2-1]
-      %
-      \arrow[from=1-3, to=2-3, Rightarrow, no head]
-      %
-      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-      \arrow["c_r", "\shortmid"{marking}, no head, from=1-2, to=1-3]
-      %
-      \arrow["c_r"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
-    \end{tikzcd}
-    &
-    % UpR
-    \begin{tikzcd}[ampersand replacement=\&]
-      A_l \&\& {A} \\
-      {A_l} \& {A} \& {A'}
-      \arrow[from=1-1, to=2-1, Rightarrow, no head]
-      %
-      \arrow["f", from=1-3, to=2-3]
-      %
-      \arrow["c_l", "\shortmid"{marking}, no head, from=1-1, to=1-3]
-      %
-      \arrow["c_l"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-      \arrow["c"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
-    \end{tikzcd}
-  \end{tabular}
-\end{center}
-
-Dually, let $d \in B \rel B'$ and $\phi \in \mathcal E_f(B', B)$.  We
-say that $d$ is \emph{right-representable by} $\phi$ if for any $d_l :
-B_l \rel B$ and $d_r : B' \rel B_r$ the following two squares commute:
-
-\begin{center}
-  \begin{tabular}{ m{14em} m{14em} } 
-    % DnR
-    \begin{tikzcd}[ampersand replacement=\&]
-      {B_l} \& {B} \& {B'} \\
-      {B_l} \&\& {B}
-      \arrow[from=1-1, to=2-1, Rightarrow, no head]
-      %
-      \arrow["\phi", from=1-3, to=2-3]
-      %
-      \arrow["d_l", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-      \arrow["d", "\shortmid"{marking}, no head, from=1-2, to=1-3]
-      %
-      \arrow["d_l"', from=2-1, to=2-3, no head]
-    \end{tikzcd}
-    &
-    % DnL
-    \begin{tikzcd}[ampersand replacement=\&]
-      {B'} \&\& {B_r} \\
-      {B} \& {B'} \& {B_r}
-      \arrow["\phi"', from=1-1, to=2-1]
-      %
-      \arrow[from=1-3, to=2-3, Rightarrow, no head]
-      %
-      \arrow["d_r", from=1-1, to=1-3, no head]
-      %
-      \arrow["d"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-      \arrow["d_r"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
-    \end{tikzcd}
-  \end{tabular}
-\end{center}
-
-Then we formulate the relationship between relation morphisms and
-function morphisms as follows:
-\begin{enumerate}
-\item There is an identity-on-objects functor $\upf : \mathcal V_e \to
-  \mathcal V_f$ such that every $c$ is left-representable by $\upf(c)$.
-\item There is an identity-on-objects functor $\dnf : \mathcal
-  E_e^{op} \to \mathcal \mathcal E_f$ such that every $d$ is
-  right-representable by $\dnf(d)$.
-\end{enumerate}
+%% Then we formulate the relationship between relation morphisms and
+%% function morphisms as follows:
+%% \begin{enumerate}
+%% \item There is an identity-on-objects functor $\upf : \mathcal V_e \to
+%%   \mathcal V_f$ such that every $c$ is left-representable by $\upf(c)$.
+%% \item There is an identity-on-objects functor $\dnf : \mathcal
+%%   E_e^{op} \to \mathcal \mathcal E_f$ such that every $d$ is
+%%   right-representable by $\dnf(d)$.
+%% \end{enumerate}
 
@@
 %% \[ F(R) = F(A(u,-)) = (F A)(-,F u) \]
-
-In summary, an extensional model consists of:
-\begin{enumerate}
-  \item A reflexive graph internal to CBPV models.
-  \item Composition of value and computation relations that form a category with the reflexive relations as identity.
-  \item An identity-on-objects functor $\upf : \mathcal V_e \to \mathcal V_f$ taking each value relation to a morphism that left-represents it.
-  \item An identity-on-objects functor $\dnf : \mathcal E_e^{op} \to \mathcal E_f$ taking each computation relation to a morphism that right-represents it.
-  \item The CBPV connectives $U,F,\times,\to$ are all \emph{covariant} functorial on relations\footnote{the reflexive graph structure already requires that these functors preserve identity relations}
-    \begin{itemize}
-    \item $U(dd') = U(d)U(d')$
-    \item $F(cc') = F(c)F(c')$
-    \item $(cc') \to (dd') = (c \to d)(c' \to d')$
-    \item $(c_1c_1') \times (c_2c_2') = (c_1 \times c_2)(c_1'\times c_2')$
-    \end{itemize}
-  \item A distinguished value type $D$ with distinguished relation
-    $\injarr{}: U(D \to F D) \rel D$ satisfying $\dnc {\injarr{}}F(\upc{\injarr{}}) = \id$
-\end{enumerate}
-
-\subsection{Stuff about Intensional Models}
 

### 2024-01-26 paper-new/extensional-model.tex
updates mostly to extensional model

@@
-\section{What is a Model of Gradual Typing?}
+\section{Idealized Double Categorical Models of Graduality}
+\label{sec:cbpv}
 
@@
 \end{example}
+
+To additionally model the error terms, we add a requirement that there
+is a natural transformation $\mho : 1 \to U$. The naturality
+requirement encodes that strict morphisms (e.g., the denotations of
+evaluation contexts) preserve errors.
 
@@
 compositionality of casts comes for free. However, the retraction
-property must be added as an additional axiom to the model.  Finally,
-the dynamic type can be modeled as an arbitrary value type $D$ with
+property must be added as an additional axiom to the model.  To model
+the error being a least element we add the requirement that $\mho
+\circ ! \ltsq{r(A)}{r(UB)} f$ holds for all $f : \mathcal V(A,B)$.
+Finally, the dynamic type can be modeled as an arbitrary value type $D$ with
 arbitrary relations $\nat \rel D$ and $D \ra D \rel D$ and $D \times D

### 2024-01-26 paper-new/intro.tex
updates mostly to extensional model

@@
 notion of relation models type precision and the squares model the
-term precision relation. This approach was influenced by Ma and
-Reynolds semantics of parametricity using reflexive graph categories:
-reflexive graph categories are essentially double categories without a
-notion of relational composition. In addition to capturing the notions
-of type and term precision, the double categorical approach allows for
-a \emph{universal property} for casts: upcasts are the
-\emph{universal} way to turn a relation arrow into a function in a
-forward direction and downcasts are the dual universal arrow.  Later,
-New, Licata and Ahmed \cite{new-licata-ahmed19} extended this
-axiomatic treatment from call-by-name to call-by-value as well by
-giving an axiomatic theory of type and term precision in
-call-by-push-value. This left implicit any connection to a ``double
-call-by-push-value'', which we make explicit in
-Section~\ref{sec:cbpv}.
+term precision relation. This approach was influenced by the semantics
+of parametricity using reflexive graph categories
+\cite{ma-reynolds,dunphythesis,reynoldsprogramme}: reflexive graph
+categories are essentially double categories without a notion of
+relational composition. In addition to capturing the notions of type
+and term precision, the double categorical approach allows for a
+\emph{universal property} for casts: upcasts are the \emph{universal}
+way to turn a relation arrow into a function in a forward direction
+and downcasts are the dual universal arrow.  Later, New, Licata and
+Ahmed \cite{new-licata-ahmed19} extended this axiomatic treatment from
+call-by-name to call-by-value as well by giving an axiomatic theory of
+type and term precision in call-by-push-value. This left implicit any
+connection to a ``double call-by-push-value'', which we make explicit
+in Section~\ref{sec:cbpv}.
 

### 2024-01-26 paper-new/categorical-model.tex
more syntax and sketch semantics in extensional model

@@
 squares is not sufficient to interpret the cast rules of gradual
-typing. Instead we generalize it slightly as follows.
+typing. Instead we generalize the notion of representability to match
+the syntactic rules in Section~\ref{sec:GTLC}.
 
@@
   \item A locally thin reflexive graph internal to CBPV models.
-  \item Composition of value and computation relations that form a category with the reflexive relations as identity.
-  \item An identity-on-objects functor $\upf : \mathcal V_e \to \mathcal V_f$ taking each value relation to a morphism that universally left-represents it.
-  \item An identity-on-objects functor $\dnf : \mathcal E_e^{op} \to \mathcal E_f$ taking each computation relation to a morphism that universally right-represents it.
+  \item Composition of value and computation relations that form a category with the reflexive relations as identity. Call these categories $\mathcal V_r,\mathcal E_r$
+  \item An identity-on-objects functor $\upf : \mathcal V_r \to \mathcal V_f$ taking each value relation to a morphism that universally left-represents it.
+  \item An identity-on-objects functor $\dnf : \mathcal E_r^{op} \to \mathcal E_f$ taking each computation relation to a morphism that universally right-represents it.
   \item The CBPV connectives $U,F,\times,\to$ are all \emph{covariant} functorial on relations\footnote{the reflexive graph structure already requires that these functors preserve identity relations}
@@
 \end{enumerate}
+
+This mathematical structure is sufficient to interpret the theory of
+gradual typing in Section~\ref{sec:gtlc}.
+\begin{definition}
+  Given any extensional gradual typing model, we interpret
+  \begin{enumerate}
+  \item Each type $A$ as a value type, interpreting the base types and
+    $\times$ as their semantic analogues and $\sem {A \ra A'}$ as $U(\sem{A} \to
+    F(\sem{A'}))$.
+  \item Every type precision derivation $c : A \ltdyn A'$ is
+    interpreted as a relation morphism $\sem{c} : \sem{A} \rel \sem{A'}$ in the obvious
+    way. Every equivalence axiom $c \equiv c'$ implies that $\sem{c} \equidyn \sem{c'}$.
+  \item Every term $\Gamma \vdash M : A$ is interpreted as a morphism
+    $\sem{M} : \mathcal V_f (\times\sem{\Gamma},UF\sem{A})$. Upcasts are interpreted as
+    $UF(u_{\sem{c}})$ and downcasts as $Ud_{F\sem{c}}$.
+  \item If $\Delta \vdash M \ltdyn M' : c$ then $\sem{M}
+    \ltsq{\times\sem{\Delta}}{\sem{c}} \sem{M'}$ holds.
+  \end{enumerate}
+\end{definition}
 

### 2024-01-26 paper-new/categorical-model.tex
errors in the extensional model

@@
   \item An identity-on-objects functor $\dnf : \mathcal E_r^{op} \to \mathcal E_f$ taking each computation relation to a morphism that universally right-represents it.
-  \item The CBPV connectives $U,F,\times,\to$ are all \emph{covariant} functorial on relations\footnote{the reflexive graph structure already requires that these functors preserve identity relations}
+  \item The CBPV connectives $U,F,\times,\to$ are all \emph{covariant} functorial on relations up to equivalence\footnote{the reflexive graph structure already requires that these functors preserve identity relations}
     \begin{itemize}
@@
     where $c \equidyn c'$ means $\id \ltsq{c}{c'}\id$ and $\id \ltsq{c'}{c} \id$.
+  \item A natural transformation $\mho : 1 \Rightarrow U$ such that
+    $\mho \circ ! \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$
   \item Distinguished value type $\nat$ with morphisms $z : \mathcal

### 2024-01-26 paper-new/discussion.tex
updates mostly to extensional model
update discussion

@@
 % Discuss Joey Eremondi's thesis on gradual dependent types
+Eremondi \cite{Eremondi_2023} uses guarded type theory to
+define a syntactic model for a gradually-typed source
+language with dependent types. By working in guarded type theory, they are
+able to give an exact semantics to non-terminating computations in a language
+which is logically consistent, allowing for metatheoretic results about the
+source language to be proven.
+%
+Similarly to our approach, they define a guarded lift monad to model potentially-
+nonterminating computations and use guarded recursion to model the dynamic type.
+However, they do not give a denotational semantics to term precision and it is unclear
+how to prove the gradual guarantee in this setting.
+The work includes a formalization of the syntactic model in Guarded Cubical Agda.
 
@@
 
-\subsection{Benefits and Drawbacks of the Synthetic Approach}
+\subsection{Comparison to Explicit Step-Indexing}
 

### 2024-01-26 paper-new/discussion.tex
discuss Siek-Chen

@@
 % Discuss Jeremy Siek's work on graduality in Agda
+Siek and Chen \cite{siekchen2021} give a proof in Agda of graduality
+for an operational semantics. While they do not use the Guarded
+Cubical extension, they do use a guarded logic of propositions
+shallowly embedded in Agda. Our denotational approach requires full
+guarded type theory not just guarded logic. An advantage of the
+denotational approach is that it easily validates equational
+reasoning, not just graduality, and it is completely independent of
+any particular syntax of gradual typing.
 

### 2024-01-27 paper-new/categorical-model.tex
update some definitions in the abstract model

@@
 
-We can specify this elegantly as a category internal to the category of CBPV models.
+We can specify this elegantly as a category internal to the category of CBPV models
+and lax morphisms, where we require that the reflexivity, source, and target morphisms
+are strict.
 In particular, as in the extensional case, there is a CBPV model of ``objects''
@@
 
+As in the extensional model, we also require the existence of a natural transformation
+$\mho : 1 \Rightarrow U$ such that $\mho \circ ! \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$,
+and a distinguished value type $\nat$ with morphisms $z : \mathcal V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
+% Lastly, we require the existence of a morphism $\mho_B \in \vf(\Gamma, UB)$ for all $B$
+% satisfying the same ordering and commutativity properties required in the extensional model.
+
 % If we spell this all out explicitly, we end up with a definition similar to the
@@
 \begin{definition}
-  A \emph{step-0} model of intensional gradual typing consists of a category internal to the
-  category of CBPV models.
+  A \emph{step-0} model of intensional gradual typing consists of:
+  \begin{itemize}
+    \item A category internal to the category of CBPV models and lax morphisms,
+    where we require that the morphisms $r$, $s$, and $t$ are strict.
+    \item A natural transformation $\mho : 1 \Rightarrow U$ such that $\mho \circ ! \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$.
+    \item A value type $\nat$ with morphisms $z : \mathcal V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
+  \end{itemize}
+
 \end{definition}
@@
 
-Given the existence of the morphisms $\delta_{UB}^*$, we can define computation morphisms
-$\delta_{FA}^* \in \ef(FA, FA)$ for all $A$ using the
+Given the existence of the morphisms $\delta_{UB}^*$, we can define a computation morphism
+$\delta_{FA}^* \in \ef(FA, FA)$ for all $A$ by composing the unit $\eta_A \in \vf(A, UFA)$ with
+the morphism $\delta_{UFA}^* \in \vf(UFA, UFA)$, and then by the adjunction
+we get a computation morphism $\delta_{FA}^ \in \ef(FA, FA)$. Moreover, we get a square
+$\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$ for all $c : A \rel A$.
 
@@
 We formalize this notion by requiring that for each object $A$ in $\vf$,
-there is a monoid of \emph{value perturbations} $\pv_A$ that is a submonoid of
-$\{ f \in \vf(A,A) \mid f \bisim \id \}$.
+there is a monoid of \emph{value perturbations} $P_A$ and a homomorphism of monoids
+$\ptb_A : P_A \to \{ f \in \vf(A,A) \mid f \bisim \id \}$.
 Similarly, for each $B : \ef$ there is a monoid $\pe_B$ of
-\emph{computation perturbations} that is a submonoid of 
-$\{ g \in \ef(B,B) \mid g \bisim \id \}$.
+\emph{computation perturbations} and a homomorphism of monoids 
+$\ptb_B : P_B \to \{ g \in \ef(B,B) \mid g \bisim \id \}$.
 
@@
 
-We require that $\delta_{FA}^* \in \pe_{FA}$ for all $A$, where $\delta_{FA}^*$ is the distinguished
-morphism that is required to be present in every hom-set $\ef(FA, FA)$ per the definition
+We will slightly abuse notation and refer to an endomorphism $f \in \vf(A, A)$ as being ``in''
+the monoid of perturbations, but which we actually mean there is an element $p \in P_A$
+that is mapped to $f$ under the homomorphism.
+
+We require that $\delta_{UB}^* \in P_{UB}$ for all $A$, where $\delta_{UB}^*$ is the distinguished
+morphism that is required to be present in every hom-set $\vf(UB, UB)$ per the definition
 of a step-1 model.
@@
 
-For reasons that will be made clear in the next section, perturbations must also satisfy
-a property that we call the ``push-pull'' property,
+Perturbations must also satisfy a property that we call the ``push-pull'' property,
 which is formulated as follows. Let $c : A \rel A'$.
-Given a perturbation $\delta \in \pv_A$, there is a corresponding perturbation
-$\push_c(\delta) \in \pv_{A'}$. % making the following square commute:
+Given a perturbation $\delta \in P_A$, there is a corresponding perturbation
+$\push_c(\delta) \in P_{A'}$. % making the following square commute:
 %
-Likewise, given $\delta' \in \pv_{A'}$ there is a perturbation $\pull_c(\delta') \in \pv_A$.
+Likewise, given $\delta' \in P_{A'}$ there is a perturbation $\pull_c(\delta') \in P_A$.
 % making the following square commute:
@@
   \begin{enumerate}
-    \item For each value type $A$, there is a monoid $\pv_A$ that is a submonoid of
-    $\{ f \in \vf(A,A) \mid f \bisim \id \}$.
-    \item For each computation type $B$, there is a monoid $\pe_B$ that is a submonoid of
-    $\{ g \in \ef(B,B) \mid g \bisim \id \}$.
-    \item For all $A$, the distinguished endomorphism $\delta_{FA}^*$ is in $\pe_{FA}$.
+    \item For each value type $A$, there is a monoid $P_A$ and a homomorphism of monoids
+    $\ptb_A : P_A \to \{ f \in \vf(A,A) \mid f \bisim \id \}$.
+    \item For each computation type $B$, there is a monoid $\pe_B$ and a homomorphism of monoids
+    $\ptb_B : P_B \to \{ g \in \ef(B,B) \mid g \bisim \id \}$.
+    \item For all $B$, the distinguished endomorphism $\delta_{UB}^*$ is in $P_{UB}$.
     % \item $\pv_A$ and a monoid homomorphism 
@@
 First, let $\mathcal M$ be any double category with a notion of perturbations,
-i.e., for any object $X$ in $\mathcal M$ there is a monoid $P_X$ which is a
-submonoid of the endomorphisms on $X$.
+i.e., for any object $X$ in $\mathcal M$ there is a monoid $P_X$ with a monoid
+homomorphism into the endomorphisms on $X$.
 
@@
 
-\begin{comment}
 Lastly, we require that the model satisfy a weak version of functoriality for 
@@
 
-\begin{definition}
+\begin{definition}[quasi-order-equivalence]\label{def:quasi-order-equivalent}
   Let $c, c' : A \rel A'$. We say that $c$ and $c'$ are \emph{quasi-order-equivalent},
@@
 \end{itemize}
-\end{comment}
 
@@
     every computation edge $d : B \rel B'$ is quasi-right-representable by $\dnf(d)$.
-    \item \eric{Do we need the quasi-functoriality of CBPV connectives on relations?}
-    % \item The CBPV connectives $U,F,\times,\to$ are quasi-functorial on relations.
+    \item The CBPV connectives $U,F,\times,\to$ are quasi-functorial on relations.
   \end{enumerate}
@@
     \item There is a distinguished value object $D \in \ob(\vf)$.
+    \item There are distinguished value relations 
+    $\iarr{}: U(D \to F D) \rel D$ and $\inat : \nat \rel D$ and $\itimes : D \times D \rel D$
+    each satisfying the retraction property up to bisimilarity.
+    %  $\dnc {\injarr{}}F(\upc{\injarr{}}) \equidyn \id$.
 
-    \item For each value type $A$, there is a value relation $\text{inj}_A : A \rel D$.
+    %\item For each value type $A$, there is a value relation $\text{inj}_A : A \rel D$.
 
@@
 
-\subsubsection{Defining an Extensional Model}\label{sec:extensional-model-definition}
+\subsubsection{Constructing an Extensional Model}\label{sec:extensional-model-definition}
 

### 2024-01-27 paper-new/discussion.tex
fix trailing sentence
work on discussion

@@
 \subsection{Related Work}
+\max{review what I wrote, and discuss Jeremy's work}
 % Discuss Joey Eremondi's thesis on gradual dependent types
@@
 \subsection{Mechanization}
+\max{review}
 % Discuss Guarded Cubical Agda and mechanization efforts
-We will use Guarded Cubical Agda to prove graduality in the syntax of 
-GTLC, which involves the construction of the abstract model described in 
-\ref{sec:concrete-model} and the extensional model with external dynamic 
-type. We also plan to formalize the adequacy result in \ref{sec:appendix-adequacy}.
+In parallel with developing the theory discussed in this paper,
+we are mechanizing our results and developing a reusable framework
+for proving graduality in Guarded Cubical Agda \cite{veltri-vezzosi2020}.
 
-% step-1
-Currently, we have defined the step-1 intensional model in Cubical 
-Agda. For value types, we constructed predomain which holds ordering and 
-bisimilarity, and morphisms between predmains. For computation types, we 
-constructed error domain which contains its underlying predomain, 
-elements of $\mho$ and $\theta$ and morphisms between error domains. We 
-also defined two functors $U$ and $F$ which support conversions between 
-predomains and error domains.
+As of this writing, the work is in progress, but we have
+constructed most parts of the concrete model discussed in
+Section \ref{sec:concrete-model}. For instance, we have defined
+types for predomains, error domains, and their morphisms and relations,
+and we used the guarded features to define the guarded lift + error monad
+and the dynamic type.
+
+We plan to formalize the construction of perturbations and 
+quasi-representable relations, but we have yet to decide
+whether to follow the approach we take in this work and define
+the abstract notion of intensional model and formalize the constructions in that setting,
+and then apply those abstract constructions to the concrete model.
+Alternatively, it may be better from a mechanization standpoint
+to carry out those abstract constructions explicitly in the concrete model,
+i.e., our representation of objects in the concrete model of predomains
+would include a field for the perturbations and our notion of relations
+would include fields for the push-pull property and quasi-representability.
+We leave this investigation to future work.
+
+Lastly, we would like to formalize the adequacy result discussed in
+Section \ref{sec:adequacy}. This will involve adding axioms about
+clock quantification as well as about the clock-irrelevance of
+booleans and natural numbers, since as of this writing these
+axioms are not built-in to Guarded Cubical Agda. As an experiment, we have 
+formalized some basic results involving clocks as part of our development.
+
+\begin{comment}
+% prove graduality in the syntax of 
+% GTLC, which involves the construction of the abstract model described in 
+% \ref{sec:concrete-model} and the extensional model with external dynamic 
+% type. We also plan to formalize the adequacy result in \ref{sec:appendix-adequacy}.
 
@@
 some theorems on clock globalization.
+\end{comment}
 
 \subsection{Comparison to Explicit Step-Indexing}
+\max{check/revise/delete if it's not important}
+Working internally to guarded type theory reduces the overhead of needing to
+carry around the step-indices in the proofs as is required when using explicit
+step-indexing. Additionally, the logical relations constructed to prove
+graduality suffer from technical complications of requiring two separate
+expression relations, one that counts steps on the left and the other on the right.
+There is no analogue of this in our approach.
+However, the price we pay for this simplicity is that it is unclear
+whether using our approach one can obtain an extensional model that models transitivity
+of term precision, which is attainable using the traditional step-indexing approach.
 
 \subsection{Synthetic Ordering}
-
+\max{revise (fix citations) or remove if it's no longer relevant}
 A key to managing the complexity of our concrete construction is in
@@
 However, there are other aspects of the model, the bisimilarity and
-the monotonicity, which are treated analytically and are .
+the monotonicity, which are treated analytically and are similarly
+tedious.
 %
@@
 \subsection{Future Work}
+\max{review and change if needed}
+We would like to apply our approach to give a denotational semantics to gradually-typed
+languages with algebraic effects, building on prior work on gradual typing for effect handlers
+\cite{greff}. In particular, that work proves graduality via a complicated step-indexed logical relation,
+and we hope to prove their results by building a denotational model for GrEff.
+This would serve as a step towards applying our techniques to prove graduality for languages
+with other advanced features.
 
-% Cite GrEff paper
+The extensional model we construct differs from the usual notion of extensional
+model considered in prior work on gradual typing in that it lacks horizontal composition of squares.
+We would like to clarify the relationship between our notion of model and prior extensional models,
+with the aim of determining whether our approach could allow for the construction of such a model.

### 2024-01-27 paper-new/discussion.tex
discussion revisions

@@
 \subsection{Related Work}
-\max{review what I wrote, and discuss Jeremy's work}
 % Discuss Joey Eremondi's thesis on gradual dependent types
@@
 \subsection{Mechanization}
-\max{review}
 % Discuss Guarded Cubical Agda and mechanization efforts
@@
 for proving graduality in Guarded Cubical Agda \cite{veltri-vezzosi2020}.
-
+%
 As of this writing, the work is in progress, but we have
@@
 
-Lastly, we would like to formalize the adequacy result discussed in
+Lastly, we plan to formalize the adequacy result discussed in
 Section \ref{sec:adequacy}. This will involve adding axioms about
@@
 \subsection{Comparison to Explicit Step-Indexing}
-\max{check/revise/delete if it's not important}
-Working internally to guarded type theory reduces the overhead of needing to
-carry around the step-indices in the proofs as is required when using explicit
-step-indexing. Additionally, the logical relations constructed to prove
-graduality suffer from technical complications of requiring two separate
-expression relations, one that counts steps on the left and the other on the right.
-There is no analogue of this in our approach.
-However, the price we pay for this simplicity is that it is unclear
-whether using our approach one can obtain an extensional model that models transitivity
-of term precision, which is attainable using the traditional step-indexing approach.
+
+Working internally to guarded type theory reduces the overhead of
+needing to carry around the step-indices in the proofs as is required
+when using explicit step-indexing. Additionally, the logical relations
+constructed to prove graduality in prior work \cite{newahmed,gtt,greff} suffer
+from technical complications of requiring two separate expression
+relations, one that counts steps on the left and the other on the
+right, and there is no analogue of this in our approach. However,
+using two expression relations allows some but not all transitivite
+reasoning of term precision to be recovered. In the future we aim to
+explore if this approach is feasible in guarded semantics.
 
 \subsection{Synthetic Ordering}
-\max{revise (fix citations) or remove if it's no longer relevant}
 A key to managing the complexity of our concrete construction is in
@@
 \subsection{Future Work}
-\max{review and change if needed}
-We would like to apply our approach to give a denotational semantics to gradually-typed
-languages with algebraic effects, building on prior work on gradual typing for effect handlers
-\cite{greff}. In particular, that work proves graduality via a complicated step-indexed logical relation,
-and we hope to prove their results by building a denotational model for GrEff.
-This would serve as a step towards applying our techniques to prove graduality for languages
-with other advanced features.
 
-The extensional model we construct differs from the usual notion of extensional
-model considered in prior work on gradual typing in that it lacks horizontal composition of squares.
-We would like to clarify the relationship between our notion of model and prior extensional models,
-with the aim of determining whether our approach could allow for the construction of such a model.
+In the future, we plan to apply our approach to give a denotational
+semantics for languages that feature higher-order state or
+runtime-extensible dynamic typing \cite{guarded-higherorderstate} as
+well as richer type disciplines such as gradual dependent types and
+effect systems.
+
+%% to gradually-typed
+%% languages with algebraic effects, building on prior work on gradual typing for effect handlers
+%% \cite{greff}. In particular, that work proves graduality via a complicated step-indexed logical relation,
+%% and we hope to prove their results by building a denotational model for GrEff.
+%% This would serve as a step towards applying our techniques to prove graduality for languages
+%% with other advanced features.
+
+%% The extensional model we construct differs from the usual notion of extensional
+%% model considered in prior work on gradual typing in that it lacks horizontal composition of squares.
+%% We would like to clarify the relationship between our notion of model and prior extensional models,
+%% with the aim of determining whether our approach could allow for the construction of such a model.

### 2024-01-27 paper-new/appendix.tex
work on proof of adequacy and move it to concrete model; work on dyn in concrete model

@@
 
-  We define $\push_{Fc} : \mathbb{N} \times P_{A} \to \mathbb{N} \times P_{A'}$.
-  Since the product of commutative monoids coincides with the coproduct, it
-  suffices by the universal property of the coproduct to consider the two cases separately.
-
-  Given $n \in \mathbb{N}$ we define the push to be $n$.
-  Then the required square is $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$ which exists by
-  the definition of a step-1 model.
-  On the other hand, given $a \in P_{A}$, we define the push to be $\push_c(a)$
-  and then the required exists by the push property for $c$.
-  Defining $\pull_{Fc}$ is done similarly.
+  We define $\push_{Fc} : \mathbb{N} \times P_{A} \to \mathbb{N} \times P_{A'}$ by
+
+  \[ \push_{Fc}(n, a) = (n, \push_c(a)). \]
+
+  Then we need to build the following square:
+
+  % https://q.uiver.app/#q=WzAsNCxbMCwwLCJGQSJdLFsxLDAsIkZBJyJdLFswLDEsIkZBIl0sWzEsMSwiRkEnIl0sWzIsMywiRmMiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIkZjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwyLCIoXFxkZWx0YV97RkF9XiopXm4gXFxjaXJjIFxccHRiX0EoYSkiLDJdLFsxLDMsIihcXGRlbHRhX3tGQSd9XiopXm4gXFxjaXJjIFxccHRiX3tBJ30oXFxwdXNoX2MoYSkpIl1d
+  \[\begin{tikzcd}[ampersand replacement=\&]
+    FA \& {FA'} \\
+    FA \& {FA'}
+    \arrow["Fc", "\shortmid"{marking}, no head, from=2-1, to=2-2]
+    \arrow["Fc", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+    \arrow["{(\delta_{FA}^*)^n \circ \ptb_A(a)}"', from=1-1, to=2-1]
+    \arrow["{(\delta_{FA'}^*)^n \circ \ptb_{A'}(\push_c(a))}", from=1-2, to=2-2]
+  \end{tikzcd}\]
+
+  This is obtained by composing the square $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$
+  with the square corresponding to the push property for $B$
+
+  % Given $n \in \mathbb{N}$ we define the push to be $n$.
+  % Then the required square is $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$,
+  % which exists by the definition of a step-1 model.
+  % On the other hand, given $a \in P_{A}$, we define the push to be $\push_c(a)$
+  % and then the required exists by the push property for $c$.
+  % Defining $\pull_{Fc}$ is symmetric.
 
@@
 We define the push function for $\piv_{c_1 \times c_2}$ by 
-$\push(\delta_1, \delta_2) = (\push_{c_1}(\delta_1), \push_{c_2}(\delta_2))$ and likewise for $\pull$.
-The push-pull property holds because it holds for each component.
+$\push(a_1, a_2) = (\push_{c_1}(a_1), \push_{c_2}(a_2))$ and likewise for $\pull$.
+The push property holds because it holds for each component.
+The pull property is established similarly.
 \end{proof}
@@
 
-      \[ (A_1, P_{A_1}) \times (A_2, P_{A_2}) =
-        (A_1 \times A_2, P_{A_1} \times P_{A_2}). \]
-
-      where $\oplus$ is the coproduct in the category of monoids.
+      \[ (A_1, P_{A_1}, \ptb_{A_1}) \times (A_2, P_{A_2}, \ptb_{A_2}) =
+        (A_1 \times A_2, P_{A_1} \times P_{A_2}, \ptb_{A_1 \times A_2}) \]
+
+      where $\ptb_{A_1 \times A_2}(p_1, p_2) = \ptb_{A_1}(p_1) \times \ptb_{A_2}(p_2)$.
 
@@
       \[ (A, P_A, \ptb_A) \arr (B, P_B, \ptb_B) = 
-        (A \arr B, P_A \times P_B, \ptb_{A \arr B}), \]
+        (A \arr B, P_A^{op} \times P_B, \ptb_{A \arr B}), \]
 
@@
 
-    \eric{Make sure this part is correct}
     We define a ``Kleisli arrow'' operation taking a perturbation in $P_{UB}$
@@
 %
-    Since the product and coproduct coincide in the category of commutative monoids,
-    and since $\mathbb{N}$ is generated by the element $1$,
-    it suffices by the universal property of the coproduct to show
+    That is, we need to show that
 %
-    \[ \ptb_{U(A \arr B)}(\id \tok (1, \id_{P_B})) = \id \tok \ptb_{UB}(1, \id_{P_B}), \]
+    \[ (\delta_{U(A \arr B)}^*)^n \circ U(\id_{A} \arr \ptb_B(b)) = 
+        \id \tok (\delta_{UB}^*)^n \circ U(\ptb_B(b)) \]
 %
-    and
+    This follows from the fact that $\delta_{UB}^*$ commutes with computation morphisms.
 %
-    \[ \ptb_{U(A \arr B)}(\id \tok (0, b)) = \id \tok \ptb_{UB}(0, b). \]
-
-    For the former, the LHS is
-%
-    \begin{align*}
-      \ptb_{U(A \arr B)}&(\id \tok (1, \id_{P_B})) \\ 
-      &= \ptb_{U(A \arr B)}(1, \id_{P_A}, \id_{P_B}) \\ 
-      &= \delta_{U(A \arr B)}^* \circ U(\ptb_A(\id_{P_A}) \arr \ptb_B(\id_{P_B})) \\
-      %&= \delta_{U(A \arr B)}^* \circ U(\id_A \arr \id_B) \\
-      &= \delta_{U(A \arr B)}^*
-    \end{align*}
-%
-    and we need to show that this is equal to 
-    $\id \tok \ptb_{UB}(1, \id_{P_B}) = \id \tok \delta_{UB}^*$.
-    But this holds because $\delta_{UB}^*$ commutes with all computation morphisms,
-    in particular, with bind and application.
-
-    For the latter, we need to show
-%
-    \[ U(\id \arr \ptb_B(b)) = \id \tok U(\ptb_B(b)), \]
-%
-    which holds by definition of $\tok$.
 
@@
   %
-    By similar reasoning to the above, it will suffice to show that
-
-    \[ \delta_{U(A \arr B)}^* = \delta_{FA}^* \tok \id \]
-
-    And this follows from the fact that $\delta_{FA}^*$ commutes with all computation morphisms.
-
-    We can similarly construct the Kleisli product operation on perturbations.
+    By similar reasoning to the above, this holds because $\delta_{FA}^*$ commutes with all computation morphisms.
+
+    We can similarly construct a Kleisli product operation on perturbations.
     Given a perturbation $(n, a_1, a_2) \in P_{F(A_1 \times A_2)}$ we define
@@
 
-\begin{definition}
-  Let $c, c' : A \rel A'$. We say that $c$ and $c'$ are \emph{quasi-order-equivalent},
-  written $c \qordeq c'$, if there exist perturbations $\delta^l_1, \delta^l_2 \in \pv_A$ and 
-  $\delta^r_1, \delta^r_2 \in \pv_{A'}$ such that the following two squares exist:
-
-  \begin{center}
-    \begin{tabular}{ m{9em} m{9em} } 
-      \begin{tikzcd}[ampersand replacement=\&]
-        A \& {A'} \\
-        A \& {A'}
-        \arrow["\delta^l_1"', from=1-1, to=2-1]
-        \arrow["\delta^r_1", from=1-2, to=2-2]
-        \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-        \arrow["c'"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-      \end{tikzcd}
-      &
-      \begin{tikzcd}[ampersand replacement=\&]
-        A \& {A'} \\
-        A \& {A'}
-        \arrow["\delta^l_2"', from=1-1, to=2-1]
-        \arrow["\delta^r_2", from=1-2, to=2-2]
-        \arrow["c'", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-        \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-      \end{tikzcd}
-    \end{tabular}
-  \end{center}
-
-  We make the analogous definition for computation relations $d, d' : B \rel B'$.
-\end{definition}
+Recall the notion of quasi-order-equivalence as defined in Definition \ref{def:quasi-order-equivalent}.
+The following lemma will be useful in showing that two relations are quasi-order-equivalent.
 
@@
   by the same $f$, then $c \qordeq c'$.
+
+  Dually, if $d$ and $d'$ are both quasi-right-representable by the same $\phi$, then
+  $d \qordeq d'$.
 \end{lemma}
@@
   \end{enumerate}
-  \eric{Do we need parts 3 and 4?}
 
@@
     First, we claim that $F(c \comp c')$ and $Fc \comp Fc'$ are
-    both left-represented by $Fe_{c'} \circ Fe_c$.
+    both quasi-left-represented by $Fe_{c'} \circ Fe_c$.
     Indeed, we have by part (1) that $e_c' \circ e_c$ left-represents $c \comp c'$,
@@
     left-represents $Fc'$, so their composition quasi-left-represents $Fc \comp Fc'$.
-    Thus, there is a square $\alpha$ of the form
+    Thus, by Lemma \ref{lem:left-rep-by-same-morphism}, there is a square $\alpha$ of the form
 
@@
     \item The commuting squares are obtained via the functorial action of $\timesk$ on the squares for $Fc_1$ and $Fc_2$.
-    % TODO elaborate?
   \end{itemize}
@@
 
-\subsection{Defining an Extensional Model}\label{sec:extensional-construction-appendix}
+\subsection{Constructing an Extensional Model}\label{sec:extensional-construction-appendix}
 
@@
 
-  Recall the extensional model defined in Section \ref{sec:extensional-model-definition}.
-
+  Recall the extensional model construction outlined in Section \ref{sec:extensional-model-definition}.
   We first establish the representability properties for this model.
@@
 
-\end{proof}
-
-\section{Adequacy}\label{sec:appendix-adequacy}
-
-In this section, we show an adequacy result for the extensional model of GTT we obtained by
-applying the abstract construction introduced in Section
-\ref{sec:extensional-model-construction} to the concrete model
-
-First we establish some notation. Fix a morphism $f : 1 \to \li \mathbb{N} \cong \li \mathbb{N}$.
-We write that $f \da n$ to mean that there exists $m$ such that $f = \delta^m(\eta n)$
-and $f \da \mho$ to mean that there exists $m$ such that $f = \delta^m(\mho)$.
-
-Recall that $\ltls$ denotes the relation on value morphisms defined as the bisimilarity-closure
-of the intensional error-ordering on morphisms.
-That is, we have $f \ltls g$ iff there exists $f'$ and $g'$ with
-
-\[ f \bisim f' \le g' \bisim g. \]
-
-The result we would like to show is as follows:
-\begin{lemma}
-If $f \ltls g : \li \mathbb{N}$, then:
-\begin{itemize}
-  \item If $f \da n$ then $g \da n$.
-  \item If $g \da \mho$ then $f \da \mho$.
-  \item If $g \da n$ then $f \da n$.
-\end{itemize}
-\end{lemma}
-
-Unfortunately, this is actually not provable!
-Roughly speaking, the issue is that this is a ``global'' result, and it is not possible
-to prove such results inside of the guarded setting. 
-In particular, if we tried to prove a result such as the above in the guarded
-setting, we would run into a problem where we would have a natural number
-``stuck'' under a $\later$, with no way to get out the underlying number.
-
-Thus, to prove our adequacy result, we need to leave the guarded setting and pass back
-to the more familiar, set-theoretic world with no internal notion of step-indexing.
-As mentioned in the Technical Background section (Section \ref{sec:sgdt}), we can do this
-using \emph{clock quantification}.
-
-Recall that all of the constructions we have made in SGDT take place in the context of a clock $k$.
-All of our uses of the later modality and guarded recursion have taken place with respect to this clock.
-For example, consider the definition of the lift monad by guarded recursion in Section \ref{TODO}.
-% We define the lift monad $\li^k X$ as the guarded fixpoint of $\lambda \tilde{T}. X + 1 + \later^k_t (\tilde{T}_t)$.
-We can view this definition as being parameterized by a clock $k$: $\li^k : \type \to \type$.
-Then for $X$ satisfying a certain technical requirement known as \emph{clock-irrelevance},
-we can define the ``global lift'' monad as $\li^{gl} X := \forall k. \li^k X$.
-
-Similarly, we can define a ``global'' version of the lock-step error ordering and the
-weak bisimilarity relation on elements of the global lift; the former is defined by
-
-\[ x \le^{gl} y := \forall k. x[k] \le y[k], \]
-
-and the latter is defined by
-
-\[ x \bisim^{gl} y := \forall k. x[k] \bisim y[k]. \]
-
-It can be shown that there is an isomorphism between the global lift monad and the
-so-called Delay monad of Capretta \cite{lmcs:2265}.
-Recall that the delay monad $\text{Delay}(X)$ is defined as the coinductive solution
-to the equation
-
-\[ \text{Delay}(X) \cong X + \text{Delay}(X). \]
-
-It can be shown that for a clock-irrelevant type $X$, $\li^{gl} X$ is a final
-coalgebra of the functor $F(Y) = X + 1 + Y$ (For example, this follows from Theorem 4.3 in
-\cite{kristensen-mogelberg-vezzosi2022}.) Since $\text{Delay}(X + 1)$ is also a final coalgebra
-of this functor, then we have $\li^{gl} X \cong \text{Delay}(X + 1)$.
-
-We can define a ``lock-step error ordering'' on $\text{Delay}(X + 1)$
-
-And we similarly define a ``weak bisimilarity'' on $\text{Delay}(X + 1)$
-
-We claim that modulo the above isomorphism, the ``global" version of the
-extensional error ordering on $\li^{gl} X$ is equivalent to
-
-% We have been writing the type as $\li X$, but it is perhaps more accurate to write it as $\li^k X$ to
-% emphasize that the construction is parameterized by a clock $k$.
+We observe that the functoriality of the CBPV connectives on relations up to
+order-equivalence is a direct consequence of the functoriality of the CBPV connectives
+on relations up to quasi-order-equivalence in the step-4 intensional model, since the
+perturbations on both sides of the square are by definition bisimilar to the identity.
+
+Likewise, the retraction properties for the relations
+$\iarr{}: U(D \to F D) \rel D$, $\inat : \nat \rel D$, and $\itimes : D \times D \rel D$
+hold because they held up to delays in the intensional model, and the delays
+disappear in the extensional model construction.
+
+\end{proof}

### 2024-01-27 paper-new/concrete-model.tex
work on proof of adequacy and move it to concrete model; work on dyn in concrete model

@@
-\section{Constructing a Concrete Model}
+\section{Constructing a Concrete Model}\label{sec:concrete-model}
 
@@
 The partial order $\le_{\li A}$ is the lock-step error ordering defined by guarded recursion as follows:
-
+%
 \begin{itemize}
-    \item 	$\eta\, x \ltls \eta\, y$ if $x \le_A y$.
-    \item 	$\mho \ltls l$ for all $l$ 
-    \item   $\theta\, \tilde{r} \ltls \theta\, \tilde{r'}$ if
-            $\later_t (\tilde{r}_t \ltls \tilde{r'}_t)$
+    \item 	$\eta\, x \le_{\li A} \eta\, y$ if $x \le_A y$.
+    \item 	$\mho \le_{\li A} l$ for all $l$ 
+    \item   $\theta\, \tilde{r} \le_{\li A} \theta\, \tilde{r'}$ if
+            $\later_t (\tilde{r}_t \le_{\li A} \tilde{r'}_t)$
   \end{itemize}
-
+%
 The idea is that two computations $l$ and $l'$ are related if they are in
@@
 The weak bisimilarity relation is defined by guarded recursion as follows:
-
+%
 \begin{align*}
@@
 \end{align*}
-
+%
 When both sides are $\eta$, then we ensure that the underlying values are related
@@
 (Section \ref{sec:constructing-perturbations}) that for each value
-type $A$ we associate a commutative monoid $P_A$ of perturbations
-and a homomorphism into the monoid of endomorphisms bisimilar to the identity.
+type $A$ we associate a monoid $P_A$ of perturbations
+and a homomorphism into the monoid of endomorphisms bisimilar to the identity,
+and likewise for computation types.
 
-We define the perturbations for $D$ via least-fixpoint in the category of
-commutative monoids as
+We define the perturbations for $D$ via least-fixpoint in the category of monoids as
 
-\[ P_D \cong (P_D \times P_D) \times (P_D \times (\mathbb{N} \times P_D)), \]
+\[ P_D \cong (P_{D \times D}) \times P_{U(D \to FD)}. \]
 
-where $\oplus$ is the product (also the coproduct) of commutative monoids.
+Unfolding these definitions explicitly, this is
+
+\[ P_D \cong (P_D \times P_D) \times (\mathbb{N} \times P_D^{op} \times \mathbb{N} \times P_D). \]
 
 We now explain how to interpret these perturbations as endomorphisms.
-We define $\ptb_D : P_D \to \{ f : D \to D \mid f \bisim \id \}$ as in the below.
-Note the use of the functorial action of $\arr$ on morphisms.
+We define $\ptb_D : P_D \to \{ f : D \to D \mid f \bisim \id \}$ below,
 % via the universal property of the coproduct of monoids, giving a case for each of the generators.
@@
 
-$\ptb_D(p_1, p_2, p_3, n, p_4) = \lambda d.\text{case $d$ of}$
+$\ptb_D((p_{\text{times}}), p_{\text{fun}}) = \lambda d.\text{case $d$ of}$
 \begin{align*}
     &\alt \tnat(m) \mapsto \tnat(m) \\
-    &\alt \ttimes(d_1, d_2) \mapsto {\ttimes(\ptb_D(p_1)(d_1), \ptb_D(p_2)(d_2))} \\
-    &\alt \tfun(\tilde{f}) \mapsto {\tfun(\lambda t. (\ptb_D(p_3) \arr (\delta^n \circ F(\ptb_D(p_4))))(\tilde{f}_t))} \\
+    &\alt \ttimes(d_1, d_2) \mapsto {\ttimes(\ptb_{D \times D}(p_\text{times})(d_1, d_2))} \\
+    &\alt \tfun(\tilde{f}) \mapsto {\tfun(\lambda t. \ptb_{U(D \to FD)}()(\tilde{f}_t))} \\
 \end{align*}
-
-One can verify that this forms a homomorphism from $P_D \to \{ f : D \to D : f \bisim \id \}$.
-For this it is easiest to use the fact that the product coindices with the coproduct
-in the category of commutative monoids, and thus it suffices to check the condition for each summand separately.
-
-We claim that the three relations $\inat$, $\itimes$, and $\iarr$
-%and their lifted versions 
-satisfy the push-pull property.
-As an illustrative case, we establish the pull property for the relation $\iarr$.
-
-We define $\pull_{\iarr} : P_D \to P_{U(D \arr FD)}$ by cases as
-
-\[ \pull_{\iarr}(p_1, p_2, p_3, n, p_4) = (0, p_3, n, p_4) \]
-
-(recall that $P_{UB} = \mathbb{N} \times P_B$, and $P_{A \arr B} = P_A \times P_B$,
-so $P_{U(D \arr FD)}) = \mathbb{N} \times P_D \times \mathbb{N} \times P_D$.
-
-Now let $(f, d) \in \iarr$. This means that $d$ must be of the form $\tfun{\tilde{f}}$ where
-Then we have $\ptb_{U(D \arr FD)}(0, p_3, n, p_4) = $
-
-\eric{Details}
-
-\eric{Quasi-representability}
 
@@
     % \[ (d_1, \ptb_D(\delta^K_D)(d_2)) \] instead.
+
+One can verify that this forms a homomorphism from $P_D \to \{ f : D \to D : f \bisim \id \}$.
+
+We claim that the three relations $\inat$, $\itimes$, and $\iarr$
+%and their lifted versions 
+satisfy the push-pull property.
+As an illustrative case, we establish the push-pull property for the relation $\iarr$.
+
+We define $\pull_{\iarr} : P_D \to P_{U(D \arr FD)}$ by
+
+\[ \pull_{\iarr}(p_{\text{times}}, p_{\text{fun}}) = p_{\text{fun}}, \]
+% (recall that $P_{U(D \arr FD)} = \mathbb{N} \times P_D^{op} \times \mathbb{N} \times P_D$).
+
+i.e., we simply forget the other perturbation.
+
+We define $\push_{\iarr} : P_{U(D \arr FD)} \to P_D$ by
+
+\[ \push{\iarr}(p_{\text{fun}}) = (\id, p_{\text{fun}}), \]
+
+Showing that the relevant squares commute is straightforward.
+
+% To do so, consider an arbitrary perturbation $p$ on $D$. 
+% Let $(f, d) \in \iarr$. This means that $d$ must be of the form $\tfun{\tilde{f}}$. 
+% When we interpret the perturbation on $D$, obtaining and endomorphism to which we then apply $d$,
+% we will be perform the action that was performed by the perturbation on the other side and thus
+% we will be done by our assumption that $f$ is related to $d$.
+
+We next claim that the relations $\inat$, $\itimes$, and $\iarr$ are quasi-left-representable,
+and that their lifts are quasi-right-representable.
+Indeed, since the relations are functional, it is easy to see that they are quasi-left-representable
+where the perturbations are taken to be the identity.
+
+For quasi-right-representability, the most interesting case is $\li(\iarr)$.
+Defining the projection $p_{\iarr} : FD \to FU(D \to FD)$ is equivalent to defining
+$p' : D \to UFU(D \to FD)$. We define
+
+$p' = \lambda d.\text{case $d$ of}$
+\begin{align*}
+    &\alt \tnat(m) \mapsto \mho \\
+    &\alt \ttimes(d_1, d_2) \mapsto \mho \\
+    &\alt \tfun(\tilde{f}) \mapsto \theta (\lambda t. \eta(\tilde{f}_t)).
+\end{align*}
+
+We define $\dellp_D = \delrp_D = \theta \circ \nxt$.
+Then it is easy to show that the squares for $\dnl$ and $\dnr$ commute.
+% Then for $\dnr$, we need to show that if $(f, d) \in \iarr$ then
+
+% TODO retraction property
+It is also straightforward to establish the retraction property for
+each of these three relations. In the case of $\iarr$, we will
+have that the property holds up-to a delay $\theta \circ \nxt$.
 
@@
 
-Now that we have defined an intensional model with an interpetation for the dynamic type, we can apply
-the abstract construction introduced in Section \ref{sec:extensional-model-construction}.
+Now that we have defined an intensional model with an interpretation for the dynamic type, we can apply
+the abstract constructions introduced in Section \ref{sec:extensional-model-construction}.
 Doing so, we obtain an extensional model of gradual typing, where the squares are given by the
 ``bisimilarity closure'' of the intensional error ordering.
+
+\subsection{Adequacy}\label{sec:adequacy}
+
+In this section, we prove an adequacy result for the concrete extensional model of GTT we obtained in the previous section.
+applying the abstract constructions introduced in Section
+\ref{sec:extensional-model-construction} to the concrete model built in the previous section.
+%\ref{sec:concrete-model}.
+
+First we establish some notation. Fix a morphism $f : 1 \to \li \mathbb{N} \cong \li \mathbb{N}$.
+We write that $f \da n$ to mean that there exists $m$ such that $f = \delta^m(\eta n)$
+and $f \da \mho$ to mean that there exists $m$ such that $f = \delta^m(\mho)$.
+
+Recall that $\ltls$ denotes the relation on value morphisms defined as the bisimilarity-closure
+of the intensional error-ordering on morphisms.
+That is, we have $f \ltls g$ iff there exists $f'$ and $g'$ with
+%
+\[ f \bisim f' \le g' \bisim g. \]
+%
+The result we would like to show is as follows:
+\begin{lemma}
+If $f \ltls g : \li \mathbb{N}$, then:
+\begin{itemize}
+  \item If $f \da n$ then $g \da n$.
+  \item If $g \da \mho$ then $f \da \mho$.
+  \item If $g \da n$ then $f \da n$.
+\end{itemize}
+\end{lemma}
+%
+Unfortunately, this is actually not provable!
+Roughly speaking, the issue is that this is a ``global'' result, and it is not possible
+to prove such results inside of the guarded setting. 
+In particular, if we tried to prove the above result in the guarded
+setting, we would run into a problem where we would have a natural number
+``stuck'' under a $\later$, with no way to get out the underlying number.
+
+Thus, to prove our adequacy result, we need to leave the guarded setting and pass back
+to the more familiar, set-theoretic world with no internal notion of step-indexing.
+As mentioned in the Technical Background section (Section \ref{sec:sgdt}), we can do this
+using \emph{clock quantification}.
+
+Recall that all of the constructions we have made in SGDT take place in the context of a clock $k$.
+All of our uses of the later modality and guarded recursion have taken place with respect to this clock.
+For example, recall the definition of the lift monad by guarded recursion.
+% We define the lift monad $\li^k X$ as the guarded fixpoint of $\lambda \tilde{T}. X + 1 + \later^k_t (\tilde{T}_t)$.
+We can view this definition as being parameterized by a clock $k$: $\li^k : \type \to \type$.
+Then for $X$ satisfying a certain technical requirement known as \emph{clock-irrelevance},
+\footnote{A type $X$ is clock-irrelevant if there is an isomorphism $\forall k.X \cong X$.}
+we can define the ``global lift'' monad as $\li^{gl} X := \forall k. \li^k X$.
+
+It can be shown that there is an isomorphism between the global lift monad and the
+delay monad of Capretta \cite{lmcs:2265}.
+Recall that, given a type $X$, the delay monad $\text{Delay}(X)$ is defined as the coinductive
+type generated by 
+$\tnow : X \to \delay(X)$ and $\tlater : \delay(X) \to \delay(X)$.
+
+% solution to the equation
+
+% \[ \text{Delay}(X) \cong X + \text{Delay}(X). \]
+
+It can be shown that for a clock-irrelevant type $X$, $\li^{gl} X$ is a final
+coalgebra of the functor $F(Y) = X + 1 + Y$ (For example, this follows from Theorem 4.3 in
+\cite{kristensen-mogelberg-vezzosi2022}.) 
+\footnote{The proof relies on the existence of an operation  
+$\mathsf{force} : \forall k. \later^k A \to \forall k. X$ that
+allows us to eliminate the later operator under a clock quantifier.
+This must be added as an axiom in guarded type theory.}
+Since $\delay(X + 1)$ is also a final coalgebra
+of this functor, then we have $\li^{gl} X \cong \delay(X + 1)$.
+
+Given a predomain $X$ on a clock-irrelevant type, we can define a
+``global'' version of the lock-step error ordering and the
+weak bisimilarity relation on elements of the global lift; the former is defined by
+%
+\[ x \le^{gl}_X y := \forall k. x[k] \le y[k], \]
+%
+and the latter is defined by
+%
+\[ x \bisim^{gl}_X y := \forall k. x[k] \bisim y[k]. \]
+%
+On the other hand, we can define coinductively a ``lock-step error ordering"
+relation on $\delay(X + 1)$:
+%
+\begin{mathpar}
+  \inferrule*[]
+  { }
+  {\tnow (\inr\, 1) \ledelay d}
+
+  \inferrule*[]
+  {x_1 \le_X x_2}
+  {\tnow (\inl\, x_1) \ledelay \tnow (\inl\, x_2)}
+
+  \inferrule*[]
+  {d_1 \ledelay d_2}
+  {\tlater\, d_1 \ledelay \tlater\, d_2}
+\end{mathpar}
+%
+And we similarly define by coinduction a ``weak bisimilarity'' relation on $\delay(X + 1)$, which uses
+a relation $d \Da x_?$ between $\delay(X+1)$ and $X+1$ that is defined as 
+$d \Da x_? := \Sigma_{m \in \mathbb{N}} x = \tlater^m(\tnow\, x_?)$.
+Then weak bisimilarity is defined by the rules
+%
+\begin{mathpar}
+  \inferrule*[]
+  {x_? \bisim_{X + 1} y_?}
+  {\tnow\, x_? \bisimdelay \tnow\, y_? }
+
+  \inferrule*[]
+  {d_1 \Da x_? \and x_? \bisim_{X + 1} y'_?}
+  {\tlater\, d_1 \bisimdelay \tnow\, y'_? }
+
+  \inferrule*[]
+  {d_2 \Da y_? \and x_? \bisim_{X + 1} y'_?}
+  {\tnow\, x_? \bisimdelay \tlater\, d_2}
+
+  \inferrule*[]
+  {d_1 \bisimdelay d_2}
+  {\tlater\, d_1 \bisimdelay \tlater\, d_2 }
+
+  % \inferrule*[]
+  % {d_1 \Da x_? \and d_2 \Da y_? \and x_? \bisim_{X + 1} y_?}
+  % {d_1 \bisimdelay d_2}
+
+  % \inferrule*[]
+  % {d_1 \bisimdelay d_2}
+  % {\tlater d_1 \bisimdelay \tlater d_2 }
+
+\end{mathpar}
+%
+Note the similarity of these definitions to the corresponding guarded definitions.
+By adapting the aforementioned theorem to the setting of inductively-defined relations,
+we can show that both the global lock-step error ordering and the global weak bisimilarity
+admit coinductive definitions. In particular, modulo the above isomorphism
+between $\li^{gl} X$ and $\delay(X+1)$, the global version of the lock-step
+error ordering is equivalent to the lock-step error ordering on $\delay(X + 1)$,
+and likewise, the global version of the weak bisimilarity relation is equivalent to the
+weak bisimilarity relation on $\delay(X + 1)$.
+
+This implies that the global version of the extensional term precision semantics for
+$\li^{gl} X$ agrees with the corresponding notion for $\delay(X + 1)$.
+Then adequacy follows by proving the corresponding
+result for $\delay(X + 1)$ which in turn follows from the definitions of the relations.
+
+% We have been writing the type as $\li X$, but it is perhaps more accurate to write it as $\li^k X$ to
+% emphasize that the construction is parameterized by a clock $k$.
+
+% Need : nat is clock irrelevant, as well as the inputs and outputs of effects
+% Axioms about forcing clock
+% Adapt prior argument to get that the defining of the global bisim
+% and global lock-step error ordering are coinductive

### 2024-01-27 paper-new/discussion.tex
fix a lot of references

@@
 % Discuss Jeremy Siek's work on graduality in Agda
-Siek and Chen \cite{siekchen2021} give a proof in Agda of graduality
+Siek and Chen \cite{siek-chen2021} give a proof in Agda of graduality
 for an operational semantics. While they do not use the Guarded
@@
 when using explicit step-indexing. Additionally, the logical relations
-constructed to prove graduality in prior work \cite{newahmed,gtt,greff} suffer
+constructed to prove graduality in prior work \cite{new-ahmed2018,new-licata-ahmed2019,new-giovannini-licata-2022} suffer
 from technical complications of requiring two separate expression
@@
 domain theory and has also been used for synthetic reasoning for cost
-models \cite{synthetic-domain-theory,decalf}.
+models \cite{fiore_1997,GrodinNSH24}.
 
@@
 semantics for languages that feature higher-order state or
-runtime-extensible dynamic typing \cite{guarded-higherorderstate} as
-well as richer type disciplines such as gradual dependent types and
-effect systems.
+runtime-extensible dynamic typing
+\cite{DBLP:journals/corr/abs-2210-02169} as well as richer type
+disciplines such as gradual dependent types and effect systems.
 

### 2024-01-27 paper-new/extensional-model.tex
discuss Siek-Chen
fix a lot of references

@@
 (CBPV) calculus which is a standard model of effectful programming
-\cite{levycbpv}.
+\cite{levy99}.
 %
@@
 varying requirements of which connectives are present
-\cite{levy,fioremm,eec}, we will use a variant which models precisely
+\cite{levy99,cfm2016,eec}, we will use a variant which models precisely
 the connectives we require and no more
@@
 
-New, Licata and Ahmed \cite{newlicataahmed} extended the axiomatic
+New, Licata and Ahmed \cite{new-licata-ahmed2019} extended the axiomatic
 syntax to call-by-push-value but did not analyze the structure
@@
 \to \mathcal E'$ that strictly preserve all CBPV constructions, see
-the appendix for a more detailed definition.
-Some of that data can be visualized as follows:
+the appendix for a more detailed definition. We call this a strict morphism in contrast to a \emph{lax} morphism, which only preserves CBPV constructions up to transformation.
+Some of the data of a double CBPV model can be visualized as follows:
 % https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXHZzcSJdLFsyLDAsIlxcZXNxIl0sWzAsMiwiXFx2ZiJdLFsyLDIsIlxcZWYiXSxbMiwzLCJcXEZmIiwwLHsiY3VydmUiOi0yfV0sWzMsMiwiXFxVZiIsMCx7ImN1cnZlIjotMn1dLFswLDEsIlxcRnNxIiwwLHsiY3VydmUiOi0yfV0sWzEsMCwiXFxVc3EiLDAseyJjdXJ2ZSI6LTJ9XSxbMiwwLCJcXHJ2Il0sWzAsMiwiXFxzdiIsMCx7ImN1cnZlIjotMn1dLFswLDIsIlxcdHYiLDIseyJjdXJ2ZSI6Mn1dLFsxLDMsIlxcc2UiLDAseyJjdXJ2ZSI6LTJ9XSxbMSwzLCJcXHRlIiwyLHsiY3VydmUiOjJ9XSxbMywxLCJcXHJlIl0sWzQsNSwiXFxib3QiLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNiw3LCJcXGJvdCIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
@@
 \begin{example}
-  (Adapted from \cite{new-licata}): Define a double CBPV model where
+  (Adapted from \cite{new-licata18}): Define a double CBPV model where
   $\mathcal V$ is the category of predomain preorders: sets with an

### 2024-01-27 paper-new/intro.tex
fix a lot of references

@@
 and downcasts are the dual universal arrow.  Later, New, Licata and
-Ahmed \cite{new-licata-ahmed19} extended this axiomatic treatment from
+Ahmed \cite{new-licata-ahmed2019} extended this axiomatic treatment from
 call-by-name to call-by-value as well by giving an axiomatic theory of
@@
 higher-order store and runtime-extensible dynamic types
-\cite{amalsthesis,nonpmetricparamorsomething,new-jamner-ahmed19}. However,
+\cite{appelmcallester01,ahmed06,neis09,new-jamner-ahmed19}. However,
 their proof developments are highly repetitive and technical, with

### 2024-01-27 paper-new/appendix.tex
fix a lot of references
max appendix fix

@@
 
-\max{decide if we want to include $\times$ or not}
-
-\begin{align*} % TODO is hole a term?
-  &\text{Types } A := \dyn \alt \nat \alt A \times A \alt (A \ra A') \\
-  &\text{Value Contexts } \Gamma := \cdot \alt (\Gamma, x : A) \\
-  &\text{Terms } M, N := \upc c M \alt \dnc c M
-  &\quad\quad \alt \ret {V} \alt \bind{x}{M}{N} \alt V_f\, V_x \alt \dn{A}{B} M 
-\end{align*}
-
-The type precision derivations $c : A \ltdyn A'$ is inductively defined by
-reflexivity, transitivity, congruence for $\ra$ and $\times$, and generators
-$\textsf{Inj}_\ra : (D \ra D) \ltdyn D$ and $\textsf{Inj}_{\text{nat}}
-: \nat \ltdyn D$.
-%
-We define equivalence of type precision derivations to be inductively generated by congruence for all constructors, category laws for reflexivity and transitivity as well as functoriality laws for $\ra$ and $\times$ congruence
-\[ (c_i \ra c_o)(c_i' \ra c_o) = (c_ic_i') \ra (c_oc_o') \]
-\[ (c_1 \times c_2)(c_1' \times c_2') = (c_1c_1') \times (c_2c_2') \]
+\newcommand{\dynof}[1]{\textrm{dyn}(#1)}
+\begin{theorem}
+  For every $A$, there is a derivation $\dynof A : A \ltdyn D$
+\end{theorem}
+\begin{proof}
+  By induction on $A$:
+  \begin{enumerate}
+  \item $A = \dyn$, then $r(\dyn) : \dyn \ltdyn \dyn$
+  \item $A = \nat$, then $\inat : \nat \ltdyn \dyn$
+  \item $A = A_1 \ra A_2$ then $(\dynof {A_1} \ra \dynof {A_2})\iarr : A_1 \ra A_2 \ltdyn \dyn$ 
+  \item $A = A_1 \times A_2$ then $(\dynof {A_1} \times \dynof{A_2})\itimes : A_1 \times A_2 \ltdyn \dyn$ 
+  \end{enumerate}
+\end{proof}
 
 \begin{theorem}
-  \begin{enumerate}
-  \item For every $A$, there is a derivation $c : A \ltdyn D$
-  \item Any two derivations $c,c' : A \ltdyn A'$ of the same precision
-    are equivalent.
-  \end{enumerate}
+  For any two derivations $c,c' : A \ltdyn A'$ of the same precision
+  $c \equiv c'$
 \end{theorem}
@@
   \begin{enumerate}
-  \item See prior work \cite{prior-work}
   \item We show this by showing that derivations have a canonical
@@
     \begin{mathpar}
-      \inferrule{}{\textrm{refl}(D) : D \ltdyn D}
-      \inferrule{}{\textsf{Inj}_{\text{nat}} : \nat \ltdyn D}
-      \inferrule{}{\textrm{refl}(\nat) : \nat \ltdyn \nat}
-      \inferrule{c : A_i \ra A_o \ltdyn D\ra D}{c(\textsf{Inj}_{\text{arr}}) : A_i \ra A_o \ltdyn \nat}
+      \inferrule{}{\textrm{refl}(D) : D \ltdyn D}\and
+      \inferrule{}{\textsf{Inj}_{\text{nat}} : \nat \ltdyn D}\and
+      \inferrule{}{\textrm{refl}(\nat) : \nat \ltdyn \nat}\and
+      \inferrule{c : A_i \ra A_o \ltdyn D\ra D}{c(\textsf{Inj}_{\text{arr}}) : A_i \ra A_o \ltdyn \nat}\and
       \inferrule{c : A_i \ltdyn A_i' \and d : A_o \ltdyn A_o'}{c \ra d : A_i \ra A_o \ltdyn A_i'\ra A_o'}
@@
 \end{proof}
-
-The gradually typed lambda calculus we consider is call-by-value
-gradual lambda calculus with $\ra$ and $\nat$ as the only base types.
-Casts are generated by the rule
-\begin{mathpar}
-  \inferrule
-  {\Gamma \vdash M : A \and c : A \ltdyn A'}
-  {\Gamma \vdash \upc c M : A'}
-
-  \inferrule
-  {\Gamma \vdash M : A' \and c : A \ltdyn A'}
-  {\Gamma \vdash \dnc c M : A}
-\end{mathpar}
 
@@
 
-These two rules are admissible from the following principles:
-%% \begin{mathpar}
-%% \end{mathpar}
+These two rules are admissible in our presentation.
 
@@
   \dnc {dyn(A_2)}\upc {dyn(A)} M
-  &= \dnc {c \,\textrm{dyn}(A')}\upc{c' \,\textrm{dyn}(A')} M \tag{All derivations are equal}\\
-  &= \dnc {c}\dnc {\textrm{dyn}(A')}\upc{\textrm{dyn}(A')}\upc{c'} M\tag{functoriality}\\
+  &= \dnc {(c \dynof {A'})}\upc{(c'\dynof{A'})} M \tag{All derivations are equal}\\
+  &= \dnc {c}\dnc {\dynof{A'}}\upc{\dynof{A'}}\upc{c'} M\tag{functoriality}\\
   &= \dnc {c}\upc{c'} M\tag{retraction}\\
@@
 
-Then the rest follows by the up/dn rules above and the fact that precision derivations are all equal.
-\begin{mathpar}
-  \inferrule*
-  {c' = \textrm{dyn}(A_2)\max{todo}}
-  {\dnc {c'}\upc {c} M \ltdyn M' : c'}
-\end{mathpar}
+Then the rest follows by the up/dn rules above and the fact that
+precision derivations are all equal.
 

### 2024-01-27 paper-new/intro.tex
Revise sections 1 and 2

@@
 gradually typed languages that can be used to prove the aforementioned
-properties: graduality and type-based reasoning. \max{TODO: say more here}
+properties: graduality and type-based reasoning.
 
@@
 arbitrary self-referential definitions is \emph{step-indexing} or its
-synthetic form of \emph{guarded recursion}.  \max{todo: adapt the next
-  paragraph to fit in with this one} A series of works
+synthetic form of \emph{guarded recursion}. A series of works
 \cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
@@
 
-\max{the stuff below is out of date}
-An alternative approach, which we investigate in this paper, is provided by
-\emph{synthetic guarded domain theory}\cite{birkedal-mogelberg-schwinghammer-stovring2011}.
-The techniques of synthetic guarded domain theory allow us to internalize the
-step-indexed reasoning normally required in logical relations proofs of graduality,
-ultimately allowing us to specify the logical relation in a manner that looks nearly
-identical to a typical, non-step-indexed logical relation.
+An alternative approach, which we
+investigate in this paper, is provided by \emph{synthetic guarded
+domain theory}\cite{birkedal-mogelberg-schwinghammer-stovring2011}.
+The techniques of synthetic guarded domain theory allow us to
+internalize the step-indexed reasoning normally required in logical
+relations proofs of graduality, ultimately allowing us to specify the
+logical relation in a manner that looks nearly identical to a typical,
+non-step-indexed logical relation. In fact, guarded domain theory goes
+further, allowing us to define step-indexed \emph{denotational
+semantics} not just step-indexed relations, just as easily as
+constructing an ordinary set-theoretic semantics.
 
-In this paper, we report on work we have done towards mechanizing proofs of graduality
-and soundness of the equational theory of cast calculi using the techniques of SGDT.
-We take a step toward mechanization in the Agda proof assistant by describing a compositional
-denotational semantics for gradual typing in a setting where the steps taken by terms are tracked.
-
-Our longer-term goal is to mechanize these proofs in a reusable way, thereby providing a framework
-to use to more easily and conveniently prove that existing languages satisfy graduality and have
-sound equational theories. Moreover, the aim is for designers of new languages to utilize the
-framework to facilitate the design of new provably-correct gradually-typed languages with more
-complex features.
+In this paper, we develop an adequate denotational semantics that
+satisfies graduality and soundness of the equational theory of cast
+calculi using the techniques of SGDT.  Our longer-term goal is to
+mechanize these proofs in a reusable way in the Guarded Cubical Agda
+proof assistant, thereby providing a framework to use to more easily
+and conveniently prove that existing languages satisfy graduality and
+have sound equational theories. Moreover, the aim is for designers of
+new languages to utilize the framework to facilitate the design of new
+provably-correct gradually-typed languages with more complex features.
 
@@
 \item We instantiate the abstract construction to provide a concrete
-  semantics in informal guarded type theory which supports the
-  addition of arbitrary algebraic effects.
+  semantics in informal guarded type theory.
 \item We prove that the resulting denotational model is
@@
 
-Our main contribution is a compositional denotational semantics for step-aware gradual typing,
+%% \subsection{Outline of Remainder of Paper}
 
-analogous to Gradual Type Theory but now in the ``intensional" setting.
-In parallel, we are developing a reusable framework in
-Guarded Cubical Agda for developing machine-checked proofs of graduality of a cast calculus.
-To demonstrate the feasibility and utility of our approach, we are applying the framework
-to prove graduality for the simply-typed gradual lambda calculus.
-% Along the way, we have developed an ``intensional" theory of graduality that is of
-% independent interest.
+%% \max{this is entirely out of date, update later}
 
-\subsection{Outline of Remainder of Paper}
+%% % In Section \ref{sec:overview}, we give an overview of the gradually-typed lambda
+%% % calculus and the graduality theorem.
+%% %
+%% In Section \ref{sec:technical-background}, we provide technical background on synthetic guarded domain theory.
+%% % 
+%% In Section \ref{sec:gtlc-terms}, we formally introduce the gradually-typed cast calculus
+%% for which we will prove graduality. We give a semantics to the terms using
+%% synthetic guarded domain theory.
+%% % Important here are the notions of syntactic
+%% % type precision and term precision. For reasons we describe below, we
+%% % introduce two related calculi, one in which the stepping behavior of terms is
+%% % implicit (an \emph{extensional} calculus, $\extlc$), and another where this behavior
+%% % is made explicit (an \emph{intensional} calculus, $\intlc$).
+%% %
 
-\max{this is entirely out of date, update later}
+%% In Section \ref{sec:gtlc-precision}, we define the term precision ordering for
+%% the gradually-typed lambda calculus and describe our approach to assigning a
+%% denotational semantics to this ordering.
+%% This approach builds on the semantics constructed in the previous section,
+%% but extending it to the term ordering presents new challenges.
 
-% In Section \ref{sec:overview}, we give an overview of the gradually-typed lambda
-% calculus and the graduality theorem.
-%
-In Section \ref{sec:technical-background}, we provide technical background on synthetic guarded domain theory.
-% 
-In Section \ref{sec:gtlc-terms}, we formally introduce the gradually-typed cast calculus
-for which we will prove graduality. We give a semantics to the terms using
-synthetic guarded domain theory.
-% Important here are the notions of syntactic
-% type precision and term precision. For reasons we describe below, we
-% introduce two related calculi, one in which the stepping behavior of terms is
-% implicit (an \emph{extensional} calculus, $\extlc$), and another where this behavior
-% is made explicit (an \emph{intensional} calculus, $\intlc$).
-%
+%% % In Section \ref{sec:domain-theory}, we define several fundamental constructions
+%% % internal to SGDT that will be needed when we give a denotational semantics to
+%% % the gradual lambda calculus.
+%% %This includes the Lift monad, predomains and error domains.
+%% %
+%% % In Section \ref{sec:semantics}, we define the denotational semantics for the
+%% % intensional gradually-typed lambda calculus using the domain theoretic
+%% % constructions defined in the previous section.
+%% %
+%% In Section \ref{sec:graduality}, we outline in more detail the proof of graduality for the
+%% gradual lambda calculus.
+%% %
+%% In Section \ref{sec:discussion}, we discuss the benefits and drawbacks to our approach in comparison
+%% to the traditional step-indexing approach, as well as possibilities for future work.
 
-In Section \ref{sec:gtlc-precision}, we define the term precision ordering for
-the gradually-typed lambda calculus and describe our approach to assigning a
-denotational semantics to this ordering.
-This approach builds on the semantics constructed in the previous section,
-but extending it to the term ordering presents new challenges.
+%% \subsection{Overview of Results}\label{sec:overview}
 
-% In Section \ref{sec:domain-theory}, we define several fundamental constructions
-% internal to SGDT that will be needed when we give a denotational semantics to
-% the gradual lambda calculus.
-%This includes the Lift monad, predomains and error domains.
-%
-% In Section \ref{sec:semantics}, we define the denotational semantics for the
-% intensional gradually-typed lambda calculus using the domain theoretic
-% constructions defined in the previous section.
-%
-In Section \ref{sec:graduality}, we outline in more detail the proof of graduality for the
-gradual lambda calculus.
-%
-In Section \ref{sec:discussion}, we discuss the benefits and drawbacks to our approach in comparison
-to the traditional step-indexing approach, as well as possibilities for future work.
+%% % This section used to be part of the intro.
+%% % \subsection{Proving Graduality in SGDT}
+%% % TODO This section should probably be moved to after the relevant background has been introduced?
 
-\subsection{Overview of Results}\label{sec:overview}
+%% % TODO introduce the idea of cast calculus and explicit casts?
 
-% This section used to be part of the intro.
-% \subsection{Proving Graduality in SGDT}
-% TODO This section should probably be moved to after the relevant background has been introduced?
+%% In this paper, we will utilize SGDT techniques to prove graduality for a particularly
+%% simple gradually-typed cast calculus, the gradually-typed lambda calculus.
+%% Before stating the graduality theorem, we introduce some definitions related to gradual typing.
 
-% TODO introduce the idea of cast calculus and explicit casts?
+%% % Cast calculi
+%% % TODO move this to earlier?
+%% Gradually typed languages are generally elaborated to a \emph{cast calculus}, in which run-time type checking
+%% that is made explicit. Elaboration inserts \emph{type casts} at the boundaries between static and dynamic code.
+%% In particular, cast insertion happens at the elimination forms of the language (e.g., pattern matching, field reference, etc.).
+%% The original gradually typed language that is elaborated to a cast calculus is called the \emph{surface language}.
 
-In this paper, we will utilize SGDT techniques to prove graduality for a particularly
-simple gradually-typed cast calculus, the gradually-typed lambda calculus.
-Before stating the graduality theorem, we introduce some definitions related to gradual typing.
+%% % Up and down casts
+%% In a cast calculus, there is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that
+%% $A$ is a \emph{more precise} type than $B$.
+%% There a distinguished type $\dyn$, the \emph{dynamic type}, with the property that $A \ltdyn\, \dyn$ for all $A$.
+%% The parts of the code that are dynamically typed will be given type $\dyn$ in the cast calculus.
+%% %
+%% If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
+%% and a term $N$ of type $B$ may be \emph{down}casted to $A$, written $\dn A B N$.
+%% Upcasts always succeed, while downcasts may fail at runtime, resulting in a type error.
+%% Cast calculi have a term $\mho$ representing a run-time type error for any type $A$.
 
-% Cast calculi
-% TODO move this to earlier?
-Gradually typed languages are generally elaborated to a \emph{cast calculus}, in which run-time type checking
-that is made explicit. Elaboration inserts \emph{type casts} at the boundaries between static and dynamic code.
-In particular, cast insertion happens at the elimination forms of the language (e.g., pattern matching, field reference, etc.).
-The original gradually typed language that is elaborated to a cast calculus is called the \emph{surface language}.
+%% % Syntactic term precision
+%% We also have a notion of \emph{syntactic term precision}.
+%% If $A \ltdyn B$, and $M$ and $N$ are terms of type $A$ and $B$ respectively, we write
+%% $M \ltdyn N : A \ltdyn B$ to mean that
+%% $M$ is ``syntactically more precise'' than $N$, or equivalently, $N$ is 
+%% ``more dynamic'' than $M$. Intuitively, this models the situation where $M$ and $N$
+%% behave the same except that $M$ may error more.
+%% Term precision is defined by a set of axioms that capture this intuitive notion.
+%% The specific rules for term precision depend on the specifics of the language, but
+%% many of the rules can be derived from the typing rules in a straightforward way.
+%% % Cast calculi also have a term $\mho$ representing a run-time type error for any type $A$,
+%% % and since
+%% Since $\mho$ always errors, there is a term precision rule stating that $\mho \ltdyn M$ for all $M$.
 
-% Up and down casts
-In a cast calculus, there is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that
-$A$ is a \emph{more precise} type than $B$.
-There a distinguished type $\dyn$, the \emph{dynamic type}, with the property that $A \ltdyn\, \dyn$ for all $A$.
-The parts of the code that are dynamically typed will be given type $\dyn$ in the cast calculus.
-%
-If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
-and a term $N$ of type $B$ may be \emph{down}casted to $A$, written $\dn A B N$.
-Upcasts always succeed, while downcasts may fail at runtime, resulting in a type error.
-Cast calculi have a term $\mho$ representing a run-time type error for any type $A$.
+%% % GTLC
+%% The gradually-typed lambda calculus is the usual simply-typed lambda calculus with a dynamic
+%% type $\dyn$ such that $A \ltdyn\, \dyn$ for all types $A$, as well as upcasts and downcasts
+%% between any types $A$ and $B$ such that $A \ltdyn B$. The complete definition will be provided in
+%% Section \ref{sec:gtlc-precision}.
 
-% Syntactic term precision
-We also have a notion of \emph{syntactic term precision}.
-If $A \ltdyn B$, and $M$ and $N$ are terms of type $A$ and $B$ respectively, we write
-$M \ltdyn N : A \ltdyn B$ to mean that
-$M$ is ``syntactically more precise'' than $N$, or equivalently, $N$ is 
-``more dynamic'' than $M$. Intuitively, this models the situation where $M$ and $N$
-behave the same except that $M$ may error more.
-Term precision is defined by a set of axioms that capture this intuitive notion.
-The specific rules for term precision depend on the specifics of the language, but
-many of the rules can be derived from the typing rules in a straightforward way.
-% Cast calculi also have a term $\mho$ representing a run-time type error for any type $A$,
-% and since
-Since $\mho$ always errors, there is a term precision rule stating that $\mho \ltdyn M$ for all $M$.
+%% With these definitions, we can state the graduality theorem for the gradually-typed lambda calculus.
 
-% GTLC
-The gradually-typed lambda calculus is the usual simply-typed lambda calculus with a dynamic
-type $\dyn$ such that $A \ltdyn\, \dyn$ for all types $A$, as well as upcasts and downcasts
-between any types $A$ and $B$ such that $A \ltdyn B$. The complete definition will be provided in
-Section \ref{sec:gtlc-precision}.
+%% % \begin{theorem}[Graduality]
+%% %   If $M \ltdyn N : \nat$, then either:
+%% %   \begin{enumerate}
+%% %     \item $M = \mho$
+%% %     \item $M = N = \zro$
+%% %     \item $M = N = \suc n$ for some $n$
+%% %     \item $M$ and $N$ diverge
+%% %   \end{enumerate}
+%% % \end{theorem}
 
-With these definitions, we can state the graduality theorem for the gradually-typed lambda calculus.
+%% \begin{theorem}[Graduality]
+%%   Let $\cdot \vdash M \ltdyn N : \nat$. 
+%%   If $M \Downarrow v_?$ and $N \Downarrow v'_?$, then either $v_? = \mho$, or $v_? = v'_?$.
+%% \end{theorem}
 
-% \begin{theorem}[Graduality]
-%   If $M \ltdyn N : \nat$, then either:
-%   \begin{enumerate}
-%     \item $M = \mho$
-%     \item $M = N = \zro$
-%     \item $M = N = \suc n$ for some $n$
-%     \item $M$ and $N$ diverge
-%   \end{enumerate}
-% \end{theorem}
+%% Note:
+%% \begin{itemize}
 
-\begin{theorem}[Graduality]
-  Let $\cdot \vdash M \ltdyn N : \nat$. 
-  If $M \Downarrow v_?$ and $N \Downarrow v'_?$, then either $v_? = \mho$, or $v_? = v'_?$.
-\end{theorem}
+%%   \item $\cdot \Downarrow$ is a relation between terms that is defined such that $M \Downarrow$ means
+%%   that $M$ terminates, either with a run-time error or a value $n$ of type $\nat$.
 
-Note:
-\begin{itemize}
+%%   \item $v_?$ is shorthand for the syntactic representation of a term that is either equal to
+%%   $\mho$, or equal to the syntactic representation of a value $n$ of type $\nat$.
+%% \end{itemize}
 
-  \item $\cdot \Downarrow$ is a relation between terms that is defined such that $M \Downarrow$ means
-  that $M$ terminates, either with a run-time error or a value $n$ of type $\nat$.
+%% % We also should be able to show that $\mho$, $\zro$, and $\suc\, N$ are not equal.
 
-  \item $v_?$ is shorthand for the syntactic representation of a term that is either equal to
-  $\mho$, or equal to the syntactic representation of a value $n$ of type $\nat$.
-\end{itemize}
+%% To prove graduality and validate the equational theory, we construct a model of the types
+%% and terms and show that all of the axioms for term precision and for equality of terms
+%% hold in this model. Modeling the dynamic type presents a challenge in the presence of a
+%% language with functions: we want the dynamic type to represent a sum of all possible types
+%% in the language, so we write down an recursive equation that the semantic object modeling
+%% dynamic type should satisfy. When the language includes function types, this equation involves a
+%% negative occurrence of the variable for which we are solving, and so the equation 
+%% does not have inductive or coinductive solutions.
+%% %
+%% To model the dynamic type, we therefore use guarded recursion to define a suitable
+%% semantic object that satisfies the unfolding isomorphism expected of the dynamic type.
+%% The key is that we do not actually get an exact solution to the equation in the style
+%% of traditional domain theory; rather, we get a ``guarded'' solution that holds ``up to a time step''.
+%% %
+%% That is, we introduce a notion of ``time'' and in the equation for the dynamic type,
+%% we guard the negative occurrences of the variable by a special operator that
+%% specifies that its argument is available ``later''.
+%% %This can be seen as a logic that internalizes the notion of step-indexing.
+%% See Section \ref{sec:technical-background} for more details on guarded type theory.
 
-% We also should be able to show that $\mho$, $\zro$, and $\suc\, N$ are not equal.
+%% At a high level, the key parts of our proof are as follows:
 
-To prove graduality and validate the equational theory, we construct a model of the types
-and terms and show that all of the axioms for term precision and for equality of terms
-hold in this model. Modeling the dynamic type presents a challenge in the presence of a
-language with functions: we want the dynamic type to represent a sum of all possible types
-in the language, so we write down an recursive equation that the semantic object modeling
-dynamic type should satisfy. When the language includes function types, this equation involves a
-negative occurrence of the variable for which we are solving, and so the equation 
-does not have inductive or coinductive solutions.
-%
-To model the dynamic type, we therefore use guarded recursion to define a suitable
-semantic object that satisfies the unfolding isomorphism expected of the dynamic type.
-The key is that we do not actually get an exact solution to the equation in the style
-of traditional domain theory; rather, we get a ``guarded'' solution that holds ``up to a time step''.
-%
-That is, we introduce a notion of ``time'' and in the equation for the dynamic type,
-we guard the negative occurrences of the variable by a special operator that
-specifies that its argument is available ``later''.
-%This can be seen as a logic that internalizes the notion of step-indexing.
-See Section \ref{sec:technical-background} for more details on guarded type theory.
+%% % TODO revise this
+%% \begin{itemize}
+%%   \item Our first step toward proving graduality is to formulate a \emph{step-sensitive},
+%%   or \emph{intensional}, gradual lambda calculus, which we call $\intlc$, in which the
+%%   computation steps taken by a term are made explicit.
+%%   The ``normal'' gradual lambda calculus for which we want to prove graduality will be called the
+%%   \emph{surface}, \emph{step-insensitive}, or \emph{extensional}, gradual lambda calculus,
+%%   denoted $\extlc$.
 
-At a high level, the key parts of our proof are as follows:
+%%   \item We define a translation from the surface syntax to the intensional syntax, and
+%%   prove a theorem relating the term precision in the surface to term precision in the
+%%   intensional syntax.
 
-% TODO revise this
-\begin{itemize}
-  \item Our first step toward proving graduality is to formulate a \emph{step-sensitive},
-  or \emph{intensional}, gradual lambda calculus, which we call $\intlc$, in which the
-  computation steps taken by a term are made explicit.
-  The ``normal'' gradual lambda calculus for which we want to prove graduality will be called the
-  \emph{surface}, \emph{step-insensitive}, or \emph{extensional}, gradual lambda calculus,
-  denoted $\extlc$.
+%%   \item We define a semantics for the intensional syntax in guarded type theory, for both the
+%%   terms and for the term precision ordering $\ltdyn$.
 
-  \item We define a translation from the surface syntax to the intensional syntax, and
-  prove a theorem relating the term precision in the surface to term precision in the
-  intensional syntax.
-
-  \item We define a semantics for the intensional syntax in guarded type theory, for both the
-  terms and for the term precision ordering $\ltdyn$.
-
-\end{itemize}
+%% \end{itemize}

### 2024-01-27 paper-new/technical-background.tex
Revise sections 1 and 2

@@
-\section{Technical Background}\label{sec:technical-background}
+\section{Background on Guarded Domain Theory}\label{sec:technical-background}
 
@@
 
-\subsection{Difficulties in Prior Semantics}
-  % Difficulties in prior semantics
+%% \subsection{Difficulties in Prior Semantics}
+%%   % Difficulties in prior semantics
 
-  In this work, we compare our approach to proving graduality to the approach
-  introduced by New and Ahmed \cite{new-ahmed2018} which constructs a step-indexed
-  logical relations model and shows that this model is sound with respect to their
-  notion of contextual error approximation.
+%%   In this work, we compare our approach to proving graduality to the approach
+%%   introduced by New and Ahmed \cite{new-ahmed2018} which constructs a step-indexed
+%%   logical relations model and shows that this model is sound with respect to their
+%%   notion of contextual error approximation.
 
-  Because the dynamic type is modeled as a non-well-founded
-  recursive type, their logical relation needs to be paramterized by natural numbers
-  to restore well-foundedness. This technique is known as a \emph{step-indexed logical relation}.
-  Intuitively, step-indexing makes the steps taken by terms observable.
-  %
-  Reasoning about step-indexed logical relations
-  can be tedious and error-prone, and there are some very subtle aspects that must
-  be taken into account in the proofs.
-  %
-  In particular, the prior approach of New and Ahmed requires two separate logical
-  relations for terms, one in which the steps of the left-hand term are counted,
-  and another in which the steps of the right-hand term are counted.
-  Then two terms $M$ and $N$ are related in the ``combined'' logical relation if they are
-  related in both of the one-sided logical relations. Having two separate logical relations
-  complicates the statement of the lemmas used to prove graduality, because any statement that
-  involves a term stepping needs to take into account whether we are counting steps on the left
-  or the right. Some of the differences can be abstracted over, but difficulties arise for properties %/results
-  as fundamental and seemingly straightforward as transitivity.
+%%   Because the dynamic type is modeled as a non-well-founded
+%%   recursive type, their logical relation needs to be paramterized by natural numbers
+%%   to restore well-foundedness. This technique is known as a \emph{step-indexed logical relation}.
+%%   Intuitively, step-indexing makes the steps taken by terms observable.
+%%   %
+%%   Reasoning about step-indexed logical relations
+%%   can be tedious and error-prone, and there are some very subtle aspects that must
+%%   be taken into account in the proofs.
+%%   %
+%%   In particular, the prior approach of New and Ahmed requires two separate logical
+%%   relations for terms, one in which the steps of the left-hand term are counted,
+%%   and another in which the steps of the right-hand term are counted.
+%%   Then two terms $M$ and $N$ are related in the ``combined'' logical relation if they are
+%%   related in both of the one-sided logical relations. Having two separate logical relations
+%%   complicates the statement of the lemmas used to prove graduality, because any statement that
+%%   involves a term stepping needs to take into account whether we are counting steps on the left
+%%   or the right. Some of the differences can be abstracted over, but difficulties arise for properties %/results
+%%   as fundamental and seemingly straightforward as transitivity.
 
-  Specifically, for transitivity, we would like to say that if $M$ is related to $N$ at
-  index $i$ and $N$ is related to $P$ at index $i$, then $M$ is related to $P$ at $i$.
-  But this does not actually hold: we require that one of the two pairs of terms
-  be related ``at infinity'', i.e., that they are related at $i$ for all $i \in \mathbb{N}$.
-  Which pair is required to satisfy this depends on which logical relation we are considering,
-  (i.e., is it counting steps on the left or on the right),
-  and so any argument that uses transitivity needs to consider two cases, one
-  where $M$ and $N$ must be shown to be related for all $i$, and another where $N$ and $P$ must
-  be related for all $i$. % This may not even be possible to show in some scenarios!
+%%   Specifically, for transitivity, we would like to say that if $M$ is related to $N$ at
+%%   index $i$ and $N$ is related to $P$ at index $i$, then $M$ is related to $P$ at $i$.
+%%   But this does not actually hold: we require that one of the two pairs of terms
+%%   be related ``at infinity'', i.e., that they are related at $i$ for all $i \in \mathbb{N}$.
+%%   Which pair is required to satisfy this depends on which logical relation we are considering,
+%%   (i.e., is it counting steps on the left or on the right),
+%%   and so any argument that uses transitivity needs to consider two cases, one
+%%   where $M$ and $N$ must be shown to be related for all $i$, and another where $N$ and $P$ must
+%%   be related for all $i$. % This may not even be possible to show in some scenarios!
 
-  \begin{comment}
-  As another example, the axioms that specify the behavior of casts do not hold in the
-  step-indexed setting. Consider, for example, the ``lower bound'' rule for downcasting:
+%%   \begin{comment}
+%%   As another example, the axioms that specify the behavior of casts do not hold in the
+%%   step-indexed setting. Consider, for example, the ``lower bound'' rule for downcasting:
 
-  \begin{mathpar}
-    \inferrule*
-    {\gamlt \vdash M \ltdyn N : dyn}
-    {\gamlt \vdash \dn{\dyn \ra \dyn}{\dyn} M \ltdyn N}
-  \end{mathpar}
+%%   \begin{mathpar}
+%%     \inferrule*
+%%     {\gamlt \vdash M \ltdyn N : dyn}
+%%     {\gamlt \vdash \dn{\dyn \ra \dyn}{\dyn} M \ltdyn N}
+%%   \end{mathpar}
 
-  In the language of the step-indexed logical relation used in prior work, this would
-  take the form
+%%   In the language of the step-indexed logical relation used in prior work, this would
+%%   take the form
 
-  \begin{mathpar}
-    \inferrule*
-    {(M, N) \in \mathcal{E}^{\sim}_{i}\sem{\dyn}}
-    {(\dn{\dyn \ra \dyn}{\dyn} M, N) \in \mathcal{E}^{\sim}_{i}\sem{\injarr{}}}
-  \end{mathpar}
+%%   \begin{mathpar}
+%%     \inferrule*
+%%     {(M, N) \in \mathcal{E}^{\sim}_{i}\sem{\dyn}}
+%%     {(\dn{\dyn \ra \dyn}{\dyn} M, N) \in \mathcal{E}^{\sim}_{i}\sem{\injarr{}}}
+%%   \end{mathpar}
 
-  where $\sim$ stands for $\mathrel{\preceq}$ or $\mathrel{\succeq}$, i.e.,
-  counting steps on the left or right respectively.
-  %
-  To show this, we would use the fact that the left-hand side steps and
-  apply an anti-reduction lemma, showing that the term to which the LHS steps
-  is related to the RHS where our fuel is now
+%%   where $\sim$ stands for $\mathrel{\preceq}$ or $\mathrel{\succeq}$, i.e.,
+%%   counting steps on the left or right respectively.
+%%   %
+%%   To show this, we would use the fact that the left-hand side steps and
+%%   apply an anti-reduction lemma, showing that the term to which the LHS steps
+%%   is related to the RHS where our fuel is now
 
-  The left-hand side steps to a case inspection on $M$,
-  where we unfold the recursive $\dyn$ type into a sum type and see whether the result
-  is a function type.
+%%   The left-hand side steps to a case inspection on $M$,
+%%   where we unfold the recursive $\dyn$ type into a sum type and see whether the result
+%%   is a function type.
 
-  One way around these difficulties is to demand that the rules only hold
-  ``extensionally'', i.e., we quantify universally over the step-index and
-  reason about the ``global'' behavior of terms for all possible step indices.
-  This is the approach taken in prior work.
-\end{comment}
+%%   One way around these difficulties is to demand that the rules only hold
+%%   ``extensionally'', i.e., we quantify universally over the step-index and
+%%   reason about the ``global'' behavior of terms for all possible step indices.
+%%   This is the approach taken in prior work.
+%% \end{comment}
 
-  % These complications introduced by step-indexing lead one to wonder whether there is a
-  % way of proving graduality without relying on tedious arguments involving natural numbers.
-  % An alternative approach, which we investigate in this paper, is provided by
-  % \emph{synthetic guarded domain theory}, as discussed below.
-  % Synthetic guarded domain theory allows the resulting logical relation to look almost
-  % identical to a typical, non-step-indexed logical relation.
+%%   % These complications introduced by step-indexing lead one to wonder whether there is a
+%%   % way of proving graduality without relying on tedious arguments involving natural numbers.
+%%   % An alternative approach, which we investigate in this paper, is provided by
+%%   % \emph{synthetic guarded domain theory}, as discussed below.
+%%   % Synthetic guarded domain theory allows the resulting logical relation to look almost
+%%   % identical to a typical, non-step-indexed logical relation.
 
-\subsection{Synthetic Guarded Domain Theory}\label{sec:sgdt}
+%% \subsection{Synthetic Guarded Domain Theory}\label{sec:sgdt}
 One way to avoid the tedious reasoning associated with step-indexing is to work
@@
 % Clocked Cubical Type Theory
-\subsubsection{Ticked Cubical Type Theory}
+\subsection{Ticked Cubical Type Theory}
 % TODO motivation for Clocked Cubical Type Theory, e.g., delayed substitutions?

### 2024-01-27 paper-new/appendix.tex
editing

@@
 
-  This is obtained by composing the square $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$
-  with the square corresponding to the push property for $B$
+  This is obtained by pasting $n$ copies of the square $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$
+  on top of the square corresponding to the push property for $c$.
 
@@
         \item A value object $A$ in $\vf$.
-        \item A commutative monoid $P_A$ of perturbations
+        \item A monoid $P_A$ of perturbations
         and a monoid homomorphism $\ptb_A : P_A \to \{ f \in \vf(A, A) \mid f \bisim \id_A \}$.
@@
         \item A computation object $B$ in $\ef$.
-        \item A commutative monoid $P_B$ of perturbations and a monoid homomorphism
+        \item A monoid $P_B$ of perturbations and a monoid homomorphism
         $\ptb_B : P_B \to \{ \phi \in \ef(B, B) \mid \phi \bisim \id_B \}$.
@@
 
-      \item The objects of $\vsq'$ (i.e., the value relations) are pairs consisting of:
+      \item Given objects $(A, P_A, \ptb_A)$ and $(A', P_{A'}, \ptb_{A'})$ a value relation
+      is a pair consisting of
       \begin{itemize}
@@
 
-      The objects of $\esq'$ are defined analogously, i.e.,
+      \item Similarly, a computation relation between $(B, P_B, \ptb_B)$ and $(B', P_{B'}, \ptb_{B'})$
+      consists of
       \begin{itemize}
@@
 
-      \item The morphisms of $\vsq'$ and $\esq'$ are given by the morphisms of $\vsq$ and $\esq$.
+      \item The squares are the same as the squares of the original model $\mathcal M$
+      %morphisms of $\vsq'$ and $\esq'$ are given by the morphisms of $\vsq$ and $\esq$.
 
@@
 
-    We define a ``Kleisli arrow'' operation taking a perturbation in $P_{UB}$
-    to a perturbation in $P_{U(A \to B)}$, as follows:
+    We define a Kleisli arrow operation on perturbations that takes a perturbation
+    in $P_{UB}$ to a perturbation in $P_{U(A \to B)}$, as follows:
     Given a perturbation $u = (n, b) \in P_{UB}$ we define $\id \tok u$ to be 
@@
 %
-    This follows from the fact that $\delta_{UB}^*$ commutes with computation morphisms.
+    This can be checked by unfolding the definition of $\tok$; the proof uses the fact that
+    $\delta_{UB}^*$ commutes with computation morphisms.
 %
@@
   %
-    By similar reasoning to the above, this holds because $\delta_{FA}^*$ commutes with all computation morphisms.
+    By similar reasoning to the above, this holds because $\delta_{FA}^*$
+    commutes with computation morphisms.
 
@@
 
-\begin{definition}[representation structure]
+\begin{definition}[representation structure]\label{def:representation-structure}
   Let $c : A \rel A'$ be a value relation. A \emph{left-representation structure}
@@
   We similarly define a {right-representation structure} $\rho^R_c$ for $c$ to consist of a morphism
-  $p_c \in \vf{A', A}$ such that $c$ is quasi-right-representable by $p_c$ (see Definition \ref{def:quasi-right-representable}).
+  $p_c \in \vf(A', A)$ such that $c$ is quasi-right-representable by $p_c$ (see Definition \ref{def:quasi-right-representable}).
   % (Notice that the direction of the morphism is opposite in the definition of
@@
     the square $\id \ltdyn_{F(c \comp c')}^{Fc \comp Fc'} \id$ which we get from
-    the lax property of $F$.
+    the fact that $F$ is lax.
 
@@
   structures for $Fc_1$ and $Fc_2$ respectively.
-  Then we can deinfe a right-representation structure for $F(c_1 \times c_2)$.
+  Then we can define a right-representation structure for $F(c_1 \times c_2)$.
 \end{lemma}
@@
 
-We define the functor $m$ which composes relations and squares horizontally as follows:
-Let $c : A \rel A'$ and $c' : A' \rel A''$. We let
-
+We define horizontal composition of relations and squares as follows:
+Let $c : A \rel A'$ and $c' : A' \rel A''$. We define
+%
 \begin{align*} 
-  m ((c, \rho^L_c, \rho^R_{Fc}), (c', \rho^L_{c'}, \rho^R_{Fc'})) = \\
-  (c \comp c', \rho^L_{c \comp c'}, \rho^R_{F(c \comp c')}). 
+  ((c, \rho^L_c, &\rho^R_{Fc}) \comp (c', \rho^L_{c'}, \rho^R_{Fc'})) =
+  (c \comp c', \rho^L_{c \comp c'}, \rho^R_{F(c \comp c')}), 
 \end{align*}
+%
+and
+%
+\begin{align*}
+  ((d, \rho^R_d, &\rho^L_{Ud}) \comp (d', \rho^R_{d'}, \rho^L_{Ud'})) =
+  (d \comp d', \rho^R_{d \comp d'}, \rho^L_{U(d \comp d')}),
+\end{align*}
+%
+where $\rho^L_{c \comp c'}, \rho^R_{F(c \comp c')}, \rho^R_{d \comp d'},$
+and $\rho^L_{U(d \comp d')}$ are as defined in Lemma \ref{lem:representation-comp}.
 
@@
 
-\[ \Fsq^{\mathcal M'} (c, \rho^L_c, \rho^R_{Fc}) = 
-  (\Fsq^M c, \rho^R_{Fc}, \rho^L_{UF(c)}), \]
+\[ F(c, \rho^L_c, \rho^R_{Fc}) = 
+  (F c, \rho^R_{Fc}, \rho^L_{UF(c)}), \]
 and
-\[ \Usq^{\mathcal M'} (d, \rho^R_d, \rho^L_{Ud}) = 
-  (\Usq^M d, \rho^L_{Ud}, \rho^R_{FU(d)}), \]
+\[ U(d, \rho^R_d, \rho^L_{Ud}) = 
+  (U d, \rho^L_{Ud}, \rho^R_{FU(d)}), \]
 
@@
 % We define $(c, \rho^L_c) \arr (d, \rho^R_d) = (c \arr d, \rho^R_{c \arr d})$.
+
+We now establish the quasi-order-equivalence for the functors.
+We already showed that $U(d \comp d') \qordeq U(d)U(d')$ and $F(c \comp c'') \qordeq F(c)F(c')$
+in the proof of Lemma \ref{lem:representation-comp}.
+The other two properties
+$(cc') \to (dd') \qordeq (c \to d)(c' \to d')$ and
+$(c_1c_1') \times (c_2c_2') \qordeq (c_1 \times c_2)(c_1'\times c_2')$
+are proved similarly, noting in both cases that the both relations are
+quasi-represented by the same morphism.
 

### 2024-01-27 paper-new/categorical-model.tex
work on proof of adequacy and move it to concrete model; work on dyn in concrete model
editing

@@
 This mathematical structure is sufficient to interpret the theory of
-gradual typing in Section~\ref{sec:gtlc}.
+gradual typing in Section~\ref{sec:GTLC}.
 \begin{definition}
@@
 
-We can specify this elegantly as a category internal to the category of CBPV models
+We can specify this compactly as a category internal to the category of CBPV models
 and lax morphisms, where we require that the reflexivity, source, and target morphisms
@@
 For the sake of ease of reference, we recap the definition of a step-0 model:
-
+%
 \begin{definition}
@@
 
-% TODO explain what this means in terms of bisimilarity "squares"?
 In this setting, we write $\refl_A : A \bisim A$ to mean that $\refl_A \in \ob(\vsim)$,
@@
 that for any $\phi \in \ef(B, B')$ we have $U\phi \circ \delta_{UB_1}^* = \delta_{UB_2} \circ U\phi$.
-
+%
 Given the existence of the morphisms $\delta_{UB}^*$, we can define a computation morphism
@@
 We will slightly abuse notation and refer to an endomorphism $f \in \vf(A, A)$ as being ``in''
-the monoid of perturbations, but which we actually mean there is an element $p \in P_A$
+the monoid of perturbations, by which we actually mean there is an element $p \in P_A$
 that is mapped to $f$ under the homomorphism.
 
-We require that $\delta_{UB}^* \in P_{UB}$ for all $A$, where $\delta_{UB}^*$ is the distinguished
+We require that $\delta_{UB}^* \in P_{UB}$ for all $B$, where $\delta_{UB}^*$ is the distinguished
 morphism that is required to be present in every hom-set $\vf(UB, UB)$ per the definition
@@
 We claim that from a step-1 model, we can construct a step-2 model. 
-\eric{TODO give overview}
-For the proof, see Lemma \ref{lem:step-1-model-to-step-2-model} in the Appendix.
+The value objects of the model are defined to be triples $(A, P_A, \ptb_A)$ where $A$ is a
+value object in $\mathcal{M}$, $P_A$ is a monoid and $\ptb_A$ is a homomorphism of monoids
+from $P_A$ to the endomorphisms on $A$ that are bisimilar to the identity.
+Likewise, computation objects are triples $(B, P_B, \ptb_B)$.
+The morphisms are the same as the morphisms of $\mathcal{M}$.
+A value relation between $(A, P_A, \ptb_A)$ and $(A', P_{A'}, \ptb_{A'})$ is given by
+a pair of a relation in $c$ and a \emph{push-pull structure} $\Pi_c$ specifying that
+that $c$ satisfies the push-pull property. Computation relations are defined analogously.
+The squares are the same as those of $\mathcal{M}$.
+We define the action of the functor $F$ on objects as $F(A, P_A, \ptb_A) = 
+(FA, \mathbb{N} \times P_A, \ptb_{FA})$ where $\ptb_{FA}(n, a) = (\delta_{FA}^*)^n \circ F(\ptb_A(a))$
+(i.e., we use the distinguished delay morphism $\delta_{FA}^*$).
+The action of $U$ on objects is defined similarly, where the perturbations
+are defined to be $\mathbb{N} \times P_B$.
+
+For the full details of the construction, see Lemma \ref{lem:step-1-model-to-step-2-model} in the Appendix.
 
@@
 We claim that we can construct a \hyperref[def:step-3-model]{step-3 intensional model} $\mathcal{M'}$.
+In the construction, the objects and morphisms are the same as those of $\mathcal M$,
+and a value relation between $A$ and $A'$ consists of a triple $(c, \rho^L_c, \rho^R_{Fc})$
+where $c : A \rel A'$ is a relation in $\mathcal{M}$,
+$\rho^L_c$ is a quasi-\emph{left}-representation for $c$ 
+(i.e., an embedding $e_c$, perturbations $\delre_c \in P_{A'}$ and $\delle_c \in P_A$, 
+and the two relevant squares), and similarly $\rho^R_{Fc}$ is
+a quasi-\emph{right}-representation for $Fc$.
+Computation relations are triples $(d, \rho^R_d, \rho^L_{Ud})$ where $d : B \rel B'$,
+$\rho^R_d$ is a quasi-right-representation for $d$ and $\rho^L{Ud}$ is a quasi-left-representation for $Ud$.
+The value and computation squares are the same as those of $\mathcal{M}$.
+We then define composition of relations and the action of the functors $F$, $U$, $\times$, and $\arr$.
 
-\eric{TODO give overview}
-
-For the proof, see Lemma \ref{lem:step-2-model-to-step-3-model} in the Appendix.
+For the full details of the construction, see Lemma \ref{lem:step-2-model-to-step-3-model} in the Appendix.
 
@@
 
-Using our existing notation, we say that $f \le_{c_o}^{c_i} g$ if there exist $f'$ and $g'$ such that
+Using our existing notation, we say that $f \ltls_{c_o}^{c_i} g$ if there exist $f'$ and $g'$ such that
 

### 2024-01-27 paper-new/concrete-model.tex
small changes in adequacy proof
editing

@@
 to obtain an extensional model.
-
 We begin with some definitions:
-
 \begin{definition}
@@
 Likewise for bisimilarity.
-
+%
 We also give a predomain structure to the natural numbers $\mathbb{N}$, where both the
 ordering and the bisimiarity relation are equality.
-
+%
 Morphisms of predomains are functions between the underlying sets that preserve the ordering
@@
 \end{definition}
-
+%
 \begin{definition}
@@
 \end{definition}
-
+%
+For an error domain $B$, we define the predomain morphism $\delta_B := \theta_B \circ \nxt$.
+%
 Morphisms of error domains are morphisms of the underlying predomains that preserve the
@@
 \end{definition}
-
+%
 We define a (monotone) relation on predomains $A$ and $A'$ to be a relation on the
@@
 More formally:
-
+%
 \begin{definition}
@@
 \end{definition}
-
+%
 Composition of relations on predomains is the usual relational composition.
-
 Similarly, we define a (monotone) relation on error domains to be a relation on the
@@
        \item (Preserves $\theta$): For all $\tilde{x}$ in $\later B$ and $\tilde{y} \in \later B'$,
-       if 
-       \[ \later_t( \tilde{x}_t \mathbin{R} \tilde{y}_t ), \]
-       then
-       \[ \theta_B(\tilde{x}) \mathbin{R} \theta_{B'}(\tilde{y}). \]
+       if $\later_t( \tilde{x}_t \mathbin{R} \tilde{y}_t )$ then
+       $\theta_B(\tilde{x}) \mathbin{R} \theta_{B'}(\tilde{y})$.
     \end{enumerate}
 \end{definition}
-
+%
 We define composition of error domain relations $R$ on $B_1$ and $B_2$ and $S$
@@
 Specifically, it is defined inductively by the following rules:
-
+%
 \begin{mathpar}
@@
 \end{mathpar}
-
+%
 % We note that this composition has the following universal property.
-
+%
 We now describe the ``commuting squares".
 Suppose we are given predomains $A_i, A_o, A_i'$, and $A_o'$,
-relations $R_i$ and $R_o$, and morphisms $f, f'$ as shown below.
-
-% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBX2kiXSxbMSwwLCJBX2knIl0sWzAsMSwiQV9vIl0sWzEsMSwiQV9vJyJdLFswLDIsImYiLDJdLFsxLDMsImYnIl0sWzAsMSwiUl9pIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMiwzLCJSX28iLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs0LDUsIlxcbHRkeW4iLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
-\[\begin{tikzcd}[ampersand replacement=\&]
-	{A_i} \& {A_i'} \\
-	{A_o} \& {A_o'}
-	\arrow[""{name=0, anchor=center, inner sep=0}, "f"', from=1-1, to=2-1]
-	\arrow[""{name=1, anchor=center, inner sep=0}, "{f'}", from=1-2, to=2-2]
-	\arrow["{R_i}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-	\arrow["{R_o}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-	\arrow["\ltdyn"{description}, draw=none, from=0, to=1]
-\end{tikzcd}\]
-
-We say that the above square commutes, written $f \le f'$, if for all
+relations $R_i : A_i \rel A_i'$ and $R_o : A_o \rel A_o'$, and morphisms $f : A_i \to A_o, f' : A_i' \to A_o'$
+Given a square with these morphisms and relations, we say that the square commutes, written $f \le f'$, if for all
 $x \in A_i$ and $x' \in A_i'$ with $x \mathbin{R_i} x'$, we have
 $f(x) \mathbin{R_o} f'(x')$.
-
 We make the analogous definition for error domains.
@@
 The guarded error-lift monad $\li$ takes a predomain $A$ to the error domain $\li A$.
-It is defined as follows:
-
+It is defined \footnote{Formally, the lift monad $\li A$ is defined as the solution to
+the guarded recursive type equation $\li A \cong A + 1 + \later \li A$.} as follows :
+%
 \begin{align*}
-  \li A &:= \\
-  &\eta \colon A \to \li A \\
-  &\mho \colon \li A \\
-  &\theta \colon \later (\li A) \to \li A
+  \li A :=
+  \alt \eta \colon A \to \li A
+  \alt \mho \colon \li A
+  \alt \theta \colon \later (\li A) \to \li A
 \end{align*}
-
-Formally, the lift monad $\li A$ is defined as the solution to the guarded recursive type equation
-
-\[ \li A \cong A + 1 + \later \li A. \]
+%
 
@@
 monad \cite{mogelberg-paviotti2016}; our version here our version augments it with the notion of error.
-
 Since we claimed that $\li A$ is a monad, we need to define the monadic operations
@@
 underlying predomain, and on morphisms returns the underlying morphism of predomains.
-%
-% TODO check this
-
 It is easily verified that $\li A$ is the free error- and later-algebra on the predomain $A$,
 so we have that $\li$ is left-adjoint to $U$.
-
 % i.e., error domain morphisms from $\li A$ to $B$ are in one-to-one correspondence with
@@
 lock-step with regard to their intensional behavior, up to $l$ erroring.
-
 Given a relation $R : A \rel A'$, we define in an analogous manner a heterogeneous
@@
 %\subsubsection{Weak Bisimilarity}\label{sec:weak-bisimilarity}
-
+%
 For a predomain $A$, we define a relation on $\li A$, called ``weak bisimilarity",
@@
     (\theta\, \tilde{x} = \delta^n(\eta\, x))
-  \text { for some $n$ and $x : \ty{A}$ such that $x \bisim_A y$ } \\
+  \text { for some $n$ and $x \in A$ such that $x \bisim_A y$ } \\
 %
@@
     (\theta\, \tilde{y} = \delta^n (\eta\, y))
-  \text { for some $n$ and $y : \ty{A}$ such that $x \bisim_A y$ }
+  \text { for some $n$ and $y \in A$ such that $x \bisim_A y$ }
 \end{align*}
@@
 
-It can be shown (by \lob-induction) that the step-sensitive relation is symmetric.
-However, it can also be shown that this relation is \emph{not} transitive:
-The argument is the same as that used to show that the step-insensitive error
-ordering $\semltbad$ described above is not transitive. Namely, we show that
-if it were transitive, then it would have to be trivial in that $l \bisim l'$ for all $l, l'$.
-that if this relation were transitive, then it would relate all values of type $\li A$.
+% It can be shown (by \lob-induction) that the step-sensitive relation is symmetric.
+% However, it can also be shown that this relation is \emph{not} transitive:
+% The argument is the same as that used to show that the step-insensitive error
+% ordering $\semltbad$ described above is not transitive. Namely, we show that
+% if it were transitive, then it would have to be trivial in that $l \bisim l'$ for all $l, l'$.
+% that if this relation were transitive, then it would relate all values of type $\li A$.
 
@@
 \end{itemize}
-
+%
 Given $f : A_1' \to A_1$ and $g : A_2 \to A_2'$ we define the predomain morphism
@@
 $f \arr \phi$ in the obvious way.
-
+%
 It is easily verified that $A \arr B$ is an exponential of $UB$ by $A$
@@
 \subsection{The Dynamic Type}
-
 The predomain representing the dynamic type will be defined using guarded recursion
 as the solution to the equation
-
+\footnote{In this section, we write $F$ instead of $\li$ so that the notation follows that of the abstract model section.}
+%
 \[ D \cong \mathbb{N}\, + (D \times D)\, + \laterhs U(D \arr FD). \]
-
+%
 % Note that the operators in the above equation are all combinators for predomains, so
@@
 and $\text{fun}$, respectively.
-
+%
 We define $e_\mathbb{N} : \mathbb{N} \to D$ to be the injection into the first
@@
 Explicitly, the ordering on $D$ is given by:
-
+%
 \begin{align*}
@@
 \end{align*}
-
+%
 We define a relation $\inat : \mathbb{N} \rel D$ by
@@
 and likewise for computation types.
-
 We define the perturbations for $D$ via least-fixpoint in the category of monoids as
-
+%
 \[ P_D \cong (P_{D \times D}) \times P_{U(D \to FD)}. \]
-
-Unfolding these definitions explicitly, this is
-
+%
+Unfolding the above, the above equation is the same as
+%
 \[ P_D \cong (P_D \times P_D) \times (\mathbb{N} \times P_D^{op} \times \mathbb{N} \times P_D). \]
-
+%
 We now explain how to interpret these perturbations as endomorphisms.
@@
 One can verify that this forms a homomorphism from $P_D \to \{ f : D \to D : f \bisim \id \}$.
-
 We claim that the three relations $\inat$, $\itimes$, and $\iarr$
@@
 As an illustrative case, we establish the push-pull property for the relation $\iarr$.
-
 We define $\pull_{\iarr} : P_D \to P_{U(D \arr FD)}$ by
-
+%
 \[ \pull_{\iarr}(p_{\text{times}}, p_{\text{fun}}) = p_{\text{fun}}, \]
 % (recall that $P_{U(D \arr FD)} = \mathbb{N} \times P_D^{op} \times \mathbb{N} \times P_D$).
-
+%
 i.e., we simply forget the other perturbation.
-
+%
 We define $\push_{\iarr} : P_{U(D \arr FD)} \to P_D$ by
-
+%
 \[ \push{\iarr}(p_{\text{fun}}) = (\id, p_{\text{fun}}), \]
-
+%
 Showing that the relevant squares commute is straightforward.
-
+%
 % To do so, consider an arbitrary perturbation $p$ on $D$. 
@@
 
-We define $\dellp_D = \delrp_D = \theta \circ \nxt$.
-Then it is easy to show that the squares for $\dnl$ and $\dnr$ commute.
+We define $\dellp_D = \delrp_D = \delta_{FD}$.
+Then it is easy to show using the definition of $\iarr$ that the squares for $\dnl$ and $\dnr$ commute.
 % Then for $\dnr$, we need to show that if $(f, d) \in \iarr$ then
 
-% TODO retraction property
 It is also straightforward to establish the retraction property for
-each of these three relations. In the case of $\iarr$, we will
-have that the property holds up-to a delay $\theta \circ \nxt$.
-
-\begin{comment}
-Before defining the perturbations, recall from our concrete construction
-(Section \ref{TODO}) that to each value type $A$ we associate a monoid $P_A$
-of pure perturbations and a monoid $P^K_{A}$ of Kleisli perturbations.
-
-We define simultaneously the pure and Kleisli perturbation monoids for $D$,
-denoted $P_D$ and $P^K_D$ respectively.
-We define $P_D$ to be the monoid such that
-
-\[ P_D \cong ((P^K_D)^{op} \times P_D) \oplus (P_D \times P_D), \]
-
-and we define $P^K_D$ to be the monoid such that
-
-\[ P^K_D \cong \mathbb{N} \oplus ((P_D)^{op} \oplus P^K_D) \oplus (P^K_D \oplus P^K_D), \]
-
-where we use $\mathbb{N}$ because it is the free monoid on one generator.
-
-Now we explain how to interpret these monoids as submonoids of endomorphisms.
-We define $i : P_D \to \{ f : D \to D \mid f \bisim \id \}$
-and $i^K : P^K_D \to \{ \phi : FD \to FD \mid \phi \bisim \id \}$ mutually
-via the universal property of the coproduct of monoids, giving a case for each of the generators.
-In the below, note the use of the functorial action of $\arr$ on morphisms.
-
-\begin{itemize}
-    \item $i(\delta^K_D, \delta_D) = \lambda d.\text{case $d$ of}$
-    \begin{align*}
-        &\alt \tfun(\tilde{f}) \mapsto 
-          \tfun(\lambda t. (i^K(\delta^K_D) \to i(\delta_D))(\tilde{f}_t)) \\
-        &\alt d' \to d'
-    \end{align*}
-
-    \item $i(\delta_D^1, \delta_D^2) = \lambda d.\text{case $d$ of}$ 
-    \begin{align*}
-        &\alt \ttimes(d_1, d_2) \mapsto
-           \ttimes(i(\delta_D^1)(d_1), i(\delta_D^2)(d_2)) \\
-        &\alt d' \to d'
-    \end{align*}
-\end{itemize}
-
-We define $i^K(\delta^D)$ by
-%
-%  P^K_D \cong \mathbb{N} \oplus ((P_D)^{op} \oplus P^K_D) \oplus (P^K_D \oplus P^K_D)
-%
-\begin{itemize}
-    \item $i^K(1) = \hat{\delta}$
-    \item $i^K(\delta_D) = F(\lambda d.\text{case $d$ of}$
-      \begin{align*} 
-        &\alt \tfun(\tilde{f}) \mapsto {\tfun(\lambda t. (i(\delta_D) \arr \id)(\tilde{f}_t))} \\
-        &\alt d' \to d')
-      \end{align*}
-    \item $i^K(\delta^K_D)$ is defined similarly to the previous but has
-    \[ \id \arr i^K(\delta^K_D) \] instead.
-    \item $i^K(\delta^K_D) = F(\lambda d.\text{case $d$ of}$
-      \begin{align*} 
-        &\alt \ttimes(d_1, d_2) \mapsto {\ttimes(i^K(\delta^K_D)(d_1), d_2)} \\
-        &\alt d' \to d')
-      \end{align*}
-    \item $i^K(\delta^K_D)$ is defined similarly to the previous but has 
-    \[ (d_1, i^K(\delta^K_D)(d_2)) \] instead.
-\end{itemize}
-
-% perturbations and quasi-representability
-We claim that the three relations $\inat$, $\itimes$, and $\iarr$ and their lifted
-versions satisfy the push-pull property.
-As an illustrative case, we establish the pull property for the $\li \iarr$.
-We define $\pull : P^K_D \to P^K_{U(D \arr FD)}$ by cases as
-
-\begin{align*}
- \pull(1) = \hat{\delta}
-\end{align*}
-
-We need to check that the relevant squares commute for all $\delta \in P^K_D$.
-It suffices by the universal property of the coproduct of monoids to ensure that
-this holds for the generators.
-
-We claim that the relations $\inat$, $\itimes$, and $\iarr$ are quasi-left-representable,
-and that their lifts are quasi-right-representable.
-Indeed, since the relations are functional, it is easy to see that they are left-representable
-where the perturbations are simply the identity.
-
-The most interesting case is the right-quasi-representability of $\li(\iarr)$.
-We define the projection as follows:
-
-\begin{align*}
-\end{align*}
-
-And for the perturbations, we take
-
-% The perturbation monoid for $D$ is defined inductively as the free monoid $M$ equipped
-% with an operation $\To : M \times M \to M$, i.e., we take the set defined inductively by
-
-% \begin{mathpar}
-%     \inferrule*[]
-%     {}
-%     {e \in M}
-
-%     \inferrule*[]
-%     {p \in M \and p' \in M}
-%     {p \odot p' \in M}
-
-%     \inferrule*[]
-%     {p \in M \and p' \in M}
-%     {p \To p' \in M}
-% \end{mathpar}
-
-% and quotient by the necessary monoid equations.
-
-\end{comment}
+each of these three relations. In the case of $\iarr$, we
+have that the property holds up to a delay.
 
@@
 First we establish some notation. Fix a morphism $f : 1 \to \li \mathbb{N} \cong \li \mathbb{N}$.
-We write that $f \da n$ to mean that there exists $m$ such that $f = \delta^m(\eta n)$
+We write $f \da n$ to mean that there exists $m$ such that $f = \delta^m(\eta n)$
 and $f \da \mho$ to mean that there exists $m$ such that $f = \delta^m(\mho)$.
@@
 %
-\[ f \bisim f' \le g' \bisim g. \]
+\[ f \bisim_{\li \mathbb{N}} f' \le_{\li \mathbb{N}} g' \bisim_{\li \mathbb{N}} g. \]
 %
-The result we would like to show is as follows:
+Here $\le_{\li \mathbb{N}}$ is the lock-step error ordering, and
+$\bisim_{\li \mathbb{N}}$ is weak bisimilarity.
+First observe that in this ordering, the semantics of error is not equivalent to
+the semantics of the diverging term.
+The main result we would like to show is as follows:
 \begin{lemma}
@@
   \item If $g \da \mho$ then $f \da \mho$.
-  \item If $g \da n$ then $f \da n$.
+  \item If $g \da n$ then $f \da n$ or $f \da \mho$.
 \end{itemize}
@@
 ``stuck'' under a $\later$, with no way to get out the underlying number.
-
+%
 Thus, to prove our adequacy result, we need to leave the guarded setting and pass back
 to the more familiar, set-theoretic world with no internal notion of step-indexing.
-As mentioned in the Technical Background section (Section \ref{sec:sgdt}), we can do this
-using \emph{clock quantification}.
-
+We can do this using a process known as \emph{clock quantification}.
 Recall that all of the constructions we have made in SGDT take place in the context of a clock $k$.
@@
 a relation $d \Da x_?$ between $\delay(X+1)$ and $X+1$ that is defined as 
-$d \Da x_? := \Sigma_{m \in \mathbb{N}} x = \tlater^m(\tnow\, x_?)$.
+$d \Da x_? := \Sigma_{m \in \mathbb{N}} d = \tlater^m(\tnow\, x_?)$.
 Then weak bisimilarity is defined by the rules
@@
   \inferrule*[]
-  {d_1 \Da x_? \and x_? \bisim_{X + 1} y'_?}
-  {\tlater\, d_1 \bisimdelay \tnow\, y'_? }
+  {d_1 \Da x_? \and x_? \bisim_{X + 1} y_?}
+  {\tlater\, d_1 \bisimdelay \tnow\, y_? }
 
   \inferrule*[]
-  {d_2 \Da y_? \and x_? \bisim_{X + 1} y'_?}
+  {d_2 \Da y_? \and x_? \bisim_{X + 1} y_?}
   {\tnow\, x_? \bisimdelay \tlater\, d_2}

### 2024-01-27 paper-new/technical-background.tex
editing

@@
 
-The operators $\later$, $\next$, and $\fix$ described above can be indexed by objects
+The operators $\later$, $\nxt$, and $\fix$ described above can be indexed by objects
 called \emph{clocks}. A clock serves as a reference relative to which steps are counted.
@@
 Most of the developments in this paper will take place in the context of a single clock $k$,
-but in Section \ref{TODO}, we will need to make use of clock quantification.
+but later on, we will need to make use of clock quantification.
 

### 2024-01-27 paper-new/categorical-model.tex
squeeze some space

@@
   \item Composition of value and computation relations that form a category with the reflexive relations as identity. Call these categories $\mathcal V_r,\mathcal E_r$
-  \item An identity-on-objects functor $\upf : \mathcal V_r \to \mathcal V_f$ taking each value relation to a morphism that universally left-represents it.
-  \item An identity-on-objects functor $\dnf : \mathcal E_r^{op} \to \mathcal E_f$ taking each computation relation to a morphism that universally right-represents it.
-  \item The CBPV connectives $U,F,\times,\to$ are all \emph{covariant} functorial on relations up to equivalence\footnote{the reflexive graph structure already requires that these functors preserve identity relations}
-    \begin{itemize}
-    \item $U(dd') \equidyn U(d)U(d')$
-    \item $F(cc') \equidyn F(c)F(c')$
-    \item $(cc') \to (dd') \equidyn (c \to d)(c' \to d')$
-    \item $(c_1c_1') \times (c_2c_2') \equidyn (c_1 \times c_2)(c_1'\times c_2')$
-    \end{itemize}
+  \item Identity-on-objects functors $\upf : \mathcal V_r \to \mathcal V_f$ and $\dnf : \mathcal E_r^{op} \to \mathcal E_f$ such that every $\upf c$ universally left-represents $c$ and every $\dnf d$ universally represents $d$.
+  \item The CBPV connectives $U,F,\times,\to$ are all \emph{covariant} functorial on relations up to equivalence: $U(dd') \equidyn U(d)U(d')$ etc.\footnote{the reflexive graph structure already requires that these functors preserve identity relations}
+    %% \begin{itemize}
+    %% \item $U(dd') \equidyn U(d)U(d')$
+    %% \item $F(cc') \equidyn F(c)F(c')$
+    %% \item $(cc') \to (dd') \equidyn (c \to d)(c' \to d')$
+    %% \item $(c_1c_1') \times (c_2c_2') \equidyn (c_1 \times c_2)(c_1'\times c_2')$
+    %% \end{itemize}
     where $c \equidyn c'$ means $\id \ltsq{c}{c'}\id$ and $\id \ltsq{c'}{c} \id$.
@@
     F(\sem{A'}))$.
-  \item Every type precision derivation $c : A \ltdyn A'$ is
-    interpreted as a relation morphism $\sem{c} : \sem{A} \rel \sem{A'}$ in the obvious
+  \item Type precision derivations $c : A \ltdyn A'$ are
+    interpreted as relation morphisms $\sem{c} : \sem{A} \rel \sem{A'}$ in the obvious
     way. Every equivalence axiom $c \equiv c'$ implies that $\sem{c} \equidyn \sem{c'}$.

### 2024-01-27 paper-new/extensional-model.tex
more space squeeze

@@
 \begin{enumerate}
-\item A cartesian category $\mathcal V$
-\item A category $\mathcal E$
+\item A cartesian category $\mathcal V$ and a category $\mathcal E$.
 \item An action of $\mathcal V^{op}$ (with the $\mathcal V$ cartesian
@@
   an arrow $A \arr B$.
-
-  This means we have 
-  \[ \alpha : {A_1 \times A_2} \arr B \cong A_2 \arr (A_1 \arr B) \]
-  and
-  \[ i : 1 \arr B \cong B \]
-  satisfying coherence isomorphisms.
-\item A functor $U : \mathcal E \to \mathcal V$ that ``preserves
+  This means we have natural isomorphisms
+  $\alpha : {A_1 \times A_2} \arr B \cong A_2 \arr (A_1 \arr B)$ and $i : 1 \arr B \cong B$ satisfying pentagon and triangle identities\cite{action}.
+\item $F \dashv U$ where $U : \mathcal E \to \mathcal V$ such that $U$ ``preserves
   powering'' in that every $U(A \arr B)$ is an exponential of $UB$ by $A$
@@
   for exponentials.
-\item A left adjoint $F \dashv U$
 \end{enumerate}

### 2024-01-27 paper-new/categorical-model.tex
more squeezing

@@
 %
-The starting point is similar to that of the extensional model: an intensional model
-will be given by a diagram in the category of CBPV objects, satisfying
-additional properties.
-%
+The starting point is similar to that of the extensional model.
 This time, however, since we are working intensionally, the semantic denotation of
@@
 compose \emph{relations} horizontally, not squares.
-What this means is that we can define a functor for composition of value relations
-and squares, and a functor for composition of computation relations and squares.
-
+% What this means is that we can define a functor for composition of value relations
+% and squares, and a functor for composition of computation relations and squares.
 We can specify this compactly as a category internal to the category of CBPV models
@@
 $\mathcal M_f$ and a CBPV model of ``arrows'' $\mathcal M_{sq}$. 
-There are CBPV morphisms
+There are strict CBPV morphisms
 $r : \mathcal M_f \to \mathcal M_{sq}$ and $s, t : \mathcal M_{sq} \to \mathcal M_f$,
@@
 
-For the sake of ease of reference, we recap the definition of a step-0 model:
-%
-\begin{definition}
-  A \emph{step-0} model of intensional gradual typing consists of:
-  \begin{itemize}
-    \item A category internal to the category of CBPV models and lax morphisms,
-    where we require that the morphisms $r$, $s$, and $t$ are strict.
-    \item A natural transformation $\mho : 1 \Rightarrow U$ such that $\mho \circ ! \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$.
-    \item A value type $\nat$ with morphisms $z : \mathcal V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
-  \end{itemize}
+% For the sake of ease of reference, we recap the definition of a step-0 model:
+% %
+% \begin{definition}
+%   A \emph{step-0} model of intensional gradual typing consists of:
+%   \begin{itemize}
+%     \item A category internal to the category of CBPV models and lax morphisms,
+%     where we require that the morphisms $r$, $s$, and $t$ are strict.
+%     \item A natural transformation $\mho : 1 \Rightarrow U$ such that $\mho \circ ! \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$.
+%     \item A value type $\nat$ with morphisms $z : \mathcal V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
+%   \end{itemize}
 
-\end{definition}
+% \end{definition}
 
@@
 
-We can specify all of this abstractly via categories $\vsim$ and $\esim$ along with
-functors $\rvsim : \vf \to \vsim$ and $\svsim, \tvsim : \vsim \to \vf$,
-and likewise for computations.
-Since bisimilarity of morphisms $f$ and $f'$ requires that they share source and target,
-we require that $\svsim$ and $\tvsim$ agree on objects and likewise for $\sesim$ and $\tesim$.
-Thus, the objects of $\vsim$ are identified with $\ob(\vf)$.
-The morphisms of $\vsim$ are ``bisimilarity proofs'', analogous to the commuting squares of $\vsq$.
+% We can specify all of this abstractly via categories $\vsim$ and $\esim$ along with
+% functors $\rvsim : \vf \to \vsim$ and $\svsim, \tvsim : \vsim \to \vf$,
+% and likewise for computations.
+% Since bisimilarity of morphisms $f$ and $f'$ requires that they share source and target,
+% we require that $\svsim$ and $\tvsim$ agree on objects and likewise for $\sesim$ and $\tesim$.
+% Thus, the objects of $\vsim$ are identified with $\ob(\vf)$.
+% The morphisms of $\vsim$ are ``bisimilarity proofs'', analogous to the commuting squares of $\vsq$.
 
-There is also a ``symmetry'' endofunctor $\text{sym}_{\mathcal{V}}^\bisim : \vsim \to \vsim$
-such that $\svsim \circ \text{sym}_{\mathcal{V}}^\bisim = \tvsim$
-and $\tvsim \circ \text{sym}_{\mathcal{V}}^\bisim = \svsim$,
-and $\text{sym}_{\mathcal{V}}^\bisim \circ \text{sym}_{\mathcal{V}}^\bisim$ is the identity.
-Likewise there is a symmetry endofunctor $\text{sym}_{\mathcal{E}}^\bisim : \esim \to \esim$.
+% There is also a ``symmetry'' endofunctor $\text{sym}_{\mathcal{V}}^\bisim : \vsim \to \vsim$
+% such that $\svsim \circ \text{sym}_{\mathcal{V}}^\bisim = \tvsim$
+% and $\tvsim \circ \text{sym}_{\mathcal{V}}^\bisim = \svsim$,
+% and $\text{sym}_{\mathcal{V}}^\bisim \circ \text{sym}_{\mathcal{V}}^\bisim$ is the identity.
+% Likewise there is a symmetry endofunctor $\text{sym}_{\mathcal{E}}^\bisim : \esim \to \esim$.
 
-In this setting, we write $\refl_A : A \bisim A$ to mean that $\refl_A \in \ob(\vsim)$,
-such that $\svsim(\refl_A) = A = \tvsim(\refl_A)$.
-Let $f, f' \in \vf(A_i, A_o)$.
-The judgment $\gamma : f \bisim_{A_i, A_o} f'$ is defined to mean:
+% In this setting, we write $\refl_A : A \bisim A$ to mean that $\refl_A \in \ob(\vsim)$,
+% such that $\svsim(\refl_A) = A = \tvsim(\refl_A)$.
+% Let $f, f' \in \vf(A_i, A_o)$.
+% The judgment $\gamma : f \bisim_{A_i, A_o} f'$ is defined to mean:
 
-\begin{enumerate}
-  \item $\gamma \in \vsim(\refl_{A_i}, \refl_{A_o})$
-  \item $\svsim(\gamma) = f$
-  \item $\tvsim(\gamma) = f'$
-\end{enumerate}
+% \begin{enumerate}
+%   \item $\gamma \in \vsim(\refl_{A_i}, \refl_{A_o})$
+%   \item $\svsim(\gamma) = f$
+%   \item $\tvsim(\gamma) = f'$
+% \end{enumerate}
 
@@
 \begin{itemize}
-  \item The categories and functors for bisimiarity described above.
+  \item The existence of a reflexive, symmetric relation $\bisim$ on the hom-sets $\vf(A,A')$ and $\ef(B,B')$
+  such that $\bisim$ preserves composition.
   \item The existence of a distinguished value morphism $\delta_{UB}^* \bisim \id_{UB}$ for each $B$.
@@
 The perturbations must be preserved by $\times$ $\timesk$, $\arr$, $\tok$, $U$, $\Uk$, $F$, and $\Fk$.
-
+%
 Perturbations must also satisfy a property that we call the ``push-pull'' property,
@@
 % making the following square commute:
+%
+Moreover, push-pull states that there are squares 
+$\delta \ltsq{c}{c} \push_c(\delta)$ and
+$\pull_c(\delta') \ltsq{c}{c} \delta'$.
 
-Moreover, push-pull states that the following squares must commute:
-
-\begin{center}
-  \begin{tabular}{ m{9em} m{9em} } 
-    \begin{tikzcd}[ampersand replacement=\&]
-      A \& {A'} \\
-      A \& {A'}
-      \arrow["\delta"', from=1-1, to=2-1]
-      \arrow["{\push_c(\delta)}", from=1-2, to=2-2]
-      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-      \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-    \end{tikzcd}
-    &
-    \begin{tikzcd}[ampersand replacement=\&]
-      A \& {A'} \\
-      A \& {A'}
-      \arrow["{\pull_c(\delta')}"', from=1-1, to=2-1]
-      \arrow["{\delta'}", from=1-2, to=2-2]
-      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-      \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-    \end{tikzcd}
-  \end{tabular}
-\end{center}
+% \begin{center}
+%   \begin{tabular}{ m{9em} m{9em} } 
+%     \begin{tikzcd}[ampersand replacement=\&]
+%       A \& {A'} \\
+%       A \& {A'}
+%       \arrow["\delta"', from=1-1, to=2-1]
+%       \arrow["{\push_c(\delta)}", from=1-2, to=2-2]
+%       \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+%       \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+%     \end{tikzcd}
+%     &
+%     \begin{tikzcd}[ampersand replacement=\&]
+%       A \& {A'} \\
+%       A \& {A'}
+%       \arrow["{\pull_c(\delta')}"', from=1-1, to=2-1]
+%       \arrow["{\delta'}", from=1-2, to=2-2]
+%       \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+%       \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+%     \end{tikzcd}
+%   \end{tabular}
+% \end{center}
 
 The analogous property should also hold for computation relations and perturbations.
-
-This is summarized below:
-
+The requirements are summarized below:
+%
 \begin{definition}\label{def:step-2-model}
@@
 homomorphism into the endomorphisms on $X$.
-
+%
 \begin{definition}\label{def:quasi-left-representable}
@@
   $\mathcal M(X, Y)$ if there are perturbations $\delle_R \in P_X$ and
-  $\delre_R \in P_Y$ such that the following squares commute:
-
-  \begin{center}
-    \begin{tabular}{ m{7em} m{7em} } 
-      % UpL
-      \begin{tikzcd}[ampersand replacement=\&]
-        X \& {Y} \\
-        {Y} \& {Y}
-        \arrow["f"', from=1-1, to=2-1]
-        \arrow["\delta_R^{r,e}", from=1-2, to=2-2]
-        \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-        \arrow[from=2-1, to=2-2, Rightarrow, no head]
-      \end{tikzcd}
-      &
-      % UpR
-      \begin{tikzcd}[ampersand replacement=\&]
-        X \& {X} \\
-        {X} \& {Y}
-        \arrow["\delta_R^{l,e}"', from=1-1, to=2-1]
-        \arrow["f", from=1-2, to=2-2]
-        \arrow[from=1-1, to=1-2, Rightarrow, no head]
-        \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-      \end{tikzcd}
-    \end{tabular}
-  \end{center}
-
-  We call the first square $\upl$ and the second square $\upr$.
-
+  $\delre_R \in P_Y$ such that there is a square
+  $\upl : f \ltsq{R}{r(Y)} \delre_R$ and a square 
+  $\upr : \delle_R \ltsq{r(X)}{R}$
+%
+%   \begin{center}
+%     \begin{tabular}{ m{7em} m{7em} } 
+%       % UpL
+%       \begin{tikzcd}[ampersand replacement=\&]
+%         X \& {Y} \\
+%         {Y} \& {Y}
+%         \arrow["f"', from=1-1, to=2-1]
+%         \arrow["\delta_R^{r,e}", from=1-2, to=2-2]
+%         \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+%         \arrow[from=2-1, to=2-2, Rightarrow, no head]
+%       \end{tikzcd}
+%       &
+%       % UpR
+%       \begin{tikzcd}[ampersand replacement=\&]
+%         X \& {X} \\
+%         {X} \& {Y}
+%         \arrow["\delta_R^{l,e}"', from=1-1, to=2-1]
+%         \arrow["f", from=1-2, to=2-2]
+%         \arrow[from=1-1, to=1-2, Rightarrow, no head]
+%         \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+%       \end{tikzcd}
+%     \end{tabular}
+%   \end{center}
+% %
+%   We call the first square $\upl$ and the second square $\upr$. 
+%
 \end{definition}
-
+%
 \begin{definition}\label{def:quasi-right-representable}
@@
   if there exist perturbations $\dellp_R \in P_X$ and
-  $\delrp_R \in P_Y$ such that the following squares commute:
-
-  \begin{center}
-    \begin{tabular}{ m{7em} m{7em} } 
-      % DnR
-      \begin{tikzcd}[ampersand replacement=\&]
-        {X} \& {Y} \\
-        {X} \& {X}
-        \arrow["\delta_R^{l,p}"', from=1-1, to=2-1]
-        \arrow["f", from=1-2, to=2-2]
-        \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-        \arrow[from=2-1, to=2-2, Rightarrow, no head]
-      \end{tikzcd}
-      &
-      % DnL
-      \begin{tikzcd}[ampersand replacement=\&]
-        {Y} \& {Y} \\
-        {X} \& {Y}
-        \arrow["f"', from=1-1, to=2-1]
-        \arrow["\delta_R^{r,p}", from=1-2, to=2-2]
-        \arrow[from=1-1, to=1-2, Rightarrow, no head]
-        \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-      \end{tikzcd}
-    \end{tabular}
-  \end{center}
-
-  We call the first square $\dnr$ and the second square $\dnl$.
-
+  $\delrp_R \in P_Y$ such that we have a square 
+  $\dnr : \dellp_R \ltsq{R}{r(X)} f$
+  and a square $\dnl : f \ltsq{r(Y)}{R} \delrp_R$.
+  %
+  % \begin{center}
+  %   \begin{tabular}{ m{7em} m{7em} } 
+  %     % DnR
+  %     \begin{tikzcd}[ampersand replacement=\&]
+  %       {X} \& {Y} \\
+  %       {X} \& {X}
+  %       \arrow["\delta_R^{l,p}"', from=1-1, to=2-1]
+  %       \arrow["f", from=1-2, to=2-2]
+  %       \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+  %       \arrow[from=2-1, to=2-2, Rightarrow, no head]
+  %     \end{tikzcd}
+  %     &
+  %     % DnL
+  %     \begin{tikzcd}[ampersand replacement=\&]
+  %       {Y} \& {Y} \\
+  %       {X} \& {Y}
+  %       \arrow["f"', from=1-1, to=2-1]
+  %       \arrow["\delta_R^{r,p}", from=1-2, to=2-2]
+  %       \arrow[from=1-1, to=1-2, Rightarrow, no head]
+  %       \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+  %     \end{tikzcd}
+  %   \end{tabular}
+  % \end{center}
+  % %
+  % We call the first square $\dnr$ and the second square $\dnl$.
+  %
   \end{definition}
@@
 First, we will need a definition:
-
+%
 \begin{definition}[quasi-order-equivalence]\label{def:quasi-order-equivalent}
@@
   written $c \qordeq c'$, if there exist perturbations $\delta^l_1, \delta^l_2 \in \pv_A$ and 
-  $\delta^r_1, \delta^r_2 \in \pv_{A'}$ such that the following two squares exist:
-
-  \begin{center}
-    \begin{tabular}{ m{9em} m{9em} } 
-      \begin{tikzcd}[ampersand replacement=\&]
-        A \& {A'} \\
-        A \& {A'}
-        \arrow["\delta^l_1"', from=1-1, to=2-1]
-        \arrow["\delta^r_1", from=1-2, to=2-2]
-        \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-        \arrow["c'"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-      \end{tikzcd}
-      &
-      \begin{tikzcd}[ampersand replacement=\&]
-        A \& {A'} \\
-        A \& {A'}
-        \arrow["\delta^l_2"', from=1-1, to=2-1]
-        \arrow["\delta^r_2", from=1-2, to=2-2]
-        \arrow["c'", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-        \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-      \end{tikzcd}
-    \end{tabular}
-  \end{center}
-
+  $\delta^r_1, \delta^r_2 \in \pv_{A'}$ such that there is a square
+  $\delta^l_1 \ltsq{c}{c'} \delta^r_1$ and a square $\delta^l_2 \ltsq{c'}{c} \delta^r_2$.
+%
+  % \begin{center}
+  %   \begin{tabular}{ m{9em} m{9em} } 
+  %     \begin{tikzcd}[ampersand replacement=\&]
+  %       A \& {A'} \\
+  %       A \& {A'}
+  %       \arrow["\delta^l_1"', from=1-1, to=2-1]
+  %       \arrow["\delta^r_1", from=1-2, to=2-2]
+  %       \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+  %       \arrow["c'"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+  %     \end{tikzcd}
+  %     &
+  %     \begin{tikzcd}[ampersand replacement=\&]
+  %       A \& {A'} \\
+  %       A \& {A'}
+  %       \arrow["\delta^l_2"', from=1-1, to=2-1]
+  %       \arrow["\delta^r_2", from=1-2, to=2-2]
+  %       \arrow["c'", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+  %       \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+  %     \end{tikzcd}
+  %   \end{tabular}
+  % \end{center}
+%
   We make the analogous definition for computation relations $d, d' : B \rel B'$.
 \end{definition}
-
+%
 We require that the CBPV connectives $U,F,\times,\to$ are \emph{quasi-functorial} on relations,
@@
 We summarize the requirements of a step-3 model below:
-
+%
 \begin{definition}\label{def:step-3-model}

### 2024-01-27 paper-new/concrete-model.tex
more squeezing
12 pages

@@
 %
-\[ P_D \cong (P_{D \times D}) \times P_{U(D \to FD)}. \]
+\( P_D \cong (P_{D \times D}) \times P_{U(D \to FD)}. \)
 %
-Unfolding the above, the above equation is the same as
+Unfolding the definitions, this is the same as
 %
-\[ P_D \cong (P_D \times P_D) \times (\mathbb{N} \times P_D^{op} \times \mathbb{N} \times P_D). \]
+\( P_D \cong (P_D \times P_D) \times (\mathbb{N} \times P_D^{op} \times \mathbb{N} \times P_D). \)
 %
@@
 % In the below, note the use of the functorial action of $\arr$ on morphisms.
-
 %
@@
 %
-
-$\ptb_D((p_{\text{times}}), p_{\text{fun}}) = \lambda d.\text{case $d$ of}$
 \begin{align*}
+ \ptb_D(p_{\text{times}}, p_{\text{fun}}) &= \lambda d.\text{case $d$ of}  \\
  &\alt \tnat(m) \mapsto \tnat(m) \\
@@
 \end{align*}
-
     % \item $\ptb_D(1)$
@@
     % \[ (d_1, \ptb_D(\delta^K_D)(d_2)) \] instead.
-
-One can verify that this forms a homomorphism from $P_D \to \{ f : D \to D : f \bisim \id \}$.
+One can verify that this defines a homomorphism from $P_D \to \{ f : D \to D : f \bisim \id \}$.
 We claim that the three relations $\inat$, $\itimes$, and $\iarr$
@@
 As an illustrative case, we establish the push-pull property for the relation $\iarr$.
-We define $\pull_{\iarr} : P_D \to P_{U(D \arr FD)}$ by
-%
-\[ \pull_{\iarr}(p_{\text{times}}, p_{\text{fun}}) = p_{\text{fun}}, \]
+We define $\pull_{\iarr} : P_D \to P_{U(D \arr FD)}$ by\( \pull_{\iarr}(p_{\text{times}}, p_{\text{fun}}) = p_{\text{fun}}, \)
 % (recall that $P_{U(D \arr FD)} = \mathbb{N} \times P_D^{op} \times \mathbb{N} \times P_D$).
@@
 %
-We define $\push_{\iarr} : P_{U(D \arr FD)} \to P_D$ by
-%
-\[ \push{\iarr}(p_{\text{fun}}) = (\id, p_{\text{fun}}), \]
+We define $\push_{\iarr} : P_{U(D \arr FD)} \to P_D$ by \( \push{\iarr}(p_{\text{fun}}) = (\id, p_{\text{fun}}) \).
 %
@@
 where the perturbations are taken to be the identity.
-
+%
 For quasi-right-representability, the most interesting case is $\li(\iarr)$.
@@
 $p' : D \to UFU(D \to FD)$. We define
-
-$p' = \lambda d.\text{case $d$ of}$
 \begin{align*}
-    &\alt \tnat(m) \mapsto \mho \\
+ p' = \lambda d.\text{case $d$ of}   &\alt \tnat(m) \mapsto \mho \\
     &\alt \ttimes(d_1, d_2) \mapsto \mho \\
@@
 \end{align*}
-
+%
 We define $\dellp_D = \delrp_D = \delta_{FD}$.
@@
 % Then for $\dnr$, we need to show that if $(f, d) \in \iarr$ then
-
+%
 It is also straightforward to establish the retraction property for
@@
 %
-\[ x \le^{gl}_X y := \forall k. x[k] \le y[k], \]
+\( x \le^{gl}_X y := \forall k. x[k] \le y[k], \)
 %
@@
 %
-\[ x \bisim^{gl}_X y := \forall k. x[k] \bisim y[k]. \]
+\( x \bisim^{gl}_X y := \forall k. x[k] \bisim y[k]. \)
 %

### 2024-03-23 paper-new/extensional-model.tex
more work on response

@@
 To additionally model the error terms, we add a requirement that there
-is a natural transformation $\mho : 1 \to U$. The naturality
+is a natural transformation $\mho : 1 \to U B$. The naturality
 requirement encodes that strict morphisms (e.g., the denotations of
@@
 We can then model CBV terms and types in a straightforward adaptation
-of Levy's interpretation of CBV in CBPV. We interpret types $A$ as
-objects $A \in \mathcal V$ and CBV terms $\Gamma \vdash M : A$ as
-morphism of any of the equivalent forms $\mathcal E(F(\times\Gamma),
-F(A)) \cong \mathcal V(\times\Gamma \vdash UF(A)) \cong \mathcal
-E(F(1), \Gamma \to F(A))$. The most interesting type translation is
-the CBV function type: $A \ra A' = U(A \to F A')$.
+of Levy's interpretation of CBV in CBPV. We interpret types $A$ and
+contexts $\Gamma$ as objects $A,\Gamma \in \mathcal V$ and CBV terms
+$\Gamma \vdash M : A$ as morphisms of any of the equivalent forms
+$\mathcal E(F\Gamma, FA) \cong \mathcal V(\Gamma \vdash UFA) \cong
+\mathcal E(F(1), \Gamma \to F(A))$. The most interesting type
+translation is the CBV function type: $A \ra A' = U(A \to F A')$.
 %

### 2024-05-23 paper-new/categorical-model.tex
several fixes to paper

@@
 \rel A_r$ we have $f \ltsq{cc_r}{c_r} \id$ and $\id \ltsq{c_l}{c_lc}
-f$. Dually, let $d : B \rel B'$ and $g : B' \to B$.  We say that $d$
-is \emph{universally right-representable by} $g$ if for any $d_l : B_l
+f$. Dually, let $d : B \rel B'$ and $\phi : B' \to B$.  We say that $d$
+is \emph{universally right-representable by} $\phi$ if for any $d_l : B_l
 \rel B$ and $d_r : B' \rel B_r$ we have $\id \ltsq{d_ld}{d_l} \phi$
@@
 principle we must require it explicitly. Finally, we explicitly add in
-a requirement that the type constructors are functorial in the
+a requirement that the type constructors are functorial on relations
+up to order equivalence.
 
@@
   \item Composition of value and computation relations that form a category with the reflexive relations as identity. Call these categories $\mathcal V_r,\mathcal E_r$
-  \item Identity-on-objects functors $\upf : \mathcal V_r \to \mathcal V_f$ and $\dnf : \mathcal E_r^{op} \to \mathcal E_f$ such that every $\upf c$ universally left-represents $c$ and every $\dnf d$ universally represents $d$.
+  \item Identity-on-objects functors $\upf : \mathcal V_r \to \mathcal V_f$ and $\dnf : \mathcal E_r^{op} \to \mathcal E_f$ such that every $\upf\, c$ universally left-represents $c$ and every $\dnf\, d$ universally right-represents $d$.
   \item The CBPV connectives $U,F,\times,\to$ are all \emph{covariant} functorial on relations up to equivalence: $U(dd') \equidyn U(d)U(d')$ etc.\footnote{the reflexive graph structure already requires that these functors preserve identity relations}
@@
   \item A natural transformation $\mho : 1 \Rightarrow U$ such that
-    $\mho \circ ! \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$
+    $\mho \circ {!} \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$
   \item Distinguished value type $\nat$ with morphisms $z : \mathcal
@@
     $\iarr{}: U(D \to F D) \rel D$ and $\inat : \nat \rel D$ and $\itimes : D \times D \rel D$
-    each satisfying the retraction property $\dnc {\injarr{}}F(\upc{\injarr{}}) \equidyn \id$.
+    each satisfying the retraction property $\dnc {\iarr{}}F(\upc{\iarr{}}) \equidyn \id$.
 \end{enumerate}
@@
     $\sem{M} : \mathcal V_f (\times\sem{\Gamma},UF\sem{A})$. Upcasts are interpreted as
-    $UF(u_{\sem{c}})$ and downcasts as $Ud_{F\sem{c}}$.
+    $UF(\upf\, {\sem{c}})$ and downcasts as $U (\dnf\, {(F\sem{c})})$.
   \item If $\Delta \vdash M \ltdyn M' : c$ then $\sem{M}
@@
 with a few key differences that will be discussed below.
+To aid in understanding the definition of an intensional model, we break the
+definition into five steps, with each one building on the previous.
+The last step will be the definition we have in mind when we refer to an intensional model.
+
+% This also facilitates the construction of an intensional model by breaking it down
+% into a few modular constructions between each subsequent phase.
+
 %
@@
 But now, we also have a CBPV morphism $m$ from the pullback 
-$\mathcal M_{sq} \times_{s = t} \mathcal M_{sq}$ to $M_{sq}$, i.e., ``composition of arrows".
+$\mathcal M_{sq} \times_{s = t} \mathcal M_{sq}$ to $\mathcal M_{sq}$, i.e., ``composition of arrows".
 In particular, this consists of a functor 
@@
 As in the extensional model, we also require the existence of a natural transformation
-$\mho : 1 \Rightarrow U$ such that $\mho \circ ! \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$,
+$\mho : 1 \Rightarrow U$ such that $\mho \circ {!} \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$,
 and a distinguished value type $\nat$ with morphisms $z : \mathcal V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
@@
 % composition of computation relations/squares.
+
+We call a model satisfying these properties a \emph{step-0 intensional model}.
 
@@
   \begin{enumerate}
-    \item For each value type $A$, there is a monoid $P_A$ and a homomorphism of monoids
-    $\ptb_A : P_A \to \{ f \in \vf(A,A) \mid f \bisim \id \}$.
+    \item For each value type $A$, there is a monoid $\pv_A$ and a homomorphism of monoids
+    $\ptb_A : \pv_A \to \{ f \in \vf(A,A) \mid f \bisim \id \}$.
     \item For each computation type $B$, there is a monoid $\pe_B$ and a homomorphism of monoids
-    $\ptb_B : P_B \to \{ g \in \ef(B,B) \mid g \bisim \id \}$.
-    \item For all $B$, the distinguished endomorphism $\delta_{UB}^*$ is in $P_{UB}$.
+    $\ptb_B : \pe_B \to \{ g \in \ef(B,B) \mid g \bisim \id \}$.
+    \item For all $B$, the distinguished endomorphism $\delta_{UB}^*$ is in $\pv_{UB}$.
     % \item $\pv_A$ and a monoid homomorphism 
@@
   Let $R$ be a horizontal morphism in $\mathcal M$ between objects $X$ and $Y$.
-  We say that $R$ is quasi-left-representable by a vertical morphism $f$ in
+  We say that $R$ is \emph{quasi-left-representable by} a vertical morphism $f$ in
   $\mathcal M(X, Y)$ if there are perturbations $\delle_R \in P_X$ and
@@
   $\upl : f \ltsq{R}{r(Y)} \delre_R$ and a square 
-  $\upr : \delle_R \ltsq{r(X)}{R}$
+  $\upr : \delle_R \ltsq{r(X)}{R} f$.
 %
@@
 We require that the CBPV connectives $U,F,\times,\to$ are \emph{quasi-functorial} on relations,
-which we specify as follows:
+which we specify as follows, with $c \comp c'$ denoting the relational composition of $c$ and $c'$:
 \begin{itemize}
-  \item $U(d \comp d') \qordeq U(d)U(d')$
-  \item $F(c \comp c'') \qordeq F(c)F(c')$
-  \item $(cc') \to (dd') \qordeq (c \to d)(c' \to d')$
-  \item $(c_1c_1') \times (c_2c_2') \qordeq (c_1 \times c_2)(c_1'\times c_2')$
+  \item $U(d \comp d') \qordeq U(d) \comp U(d')$
+  \item $F(c \comp c') \qordeq F(c) \comp F(c')$
+  \item $(c \comp c') \to (d \comp d') \qordeq (c \to d) \comp (c' \to d')$
+  \item $(c_1 \comp c_1') \times (c_2 \comp c_2') \qordeq (c_1 \times c_2) \comp (c_1'\times c_2')$
 \end{itemize}
@@
   \begin{enumerate}
-    \item There are functors $\upf : \ve \to \vf$ and $\dnf : \ee^{op} \to \ef$ % TODO: image is thin?
+    \item There are functors $\upf : \vr \to \vf$ and $\dnf : \er^{op} \to \ef$ % TODO: image is thin?
     \item Every value edge $c : A \rel A'$ is quasi-left-representable by $\upf(c)$ and
@@
 %
-Here, we discuss how to construct an extensional model from an intensional model with dyn.
+Here, we discuss how to construct an extensional model from a step-4 intensional model.
 We do so in several phases, beginning with a step-1 intensional model with dyn
@@
 Computation relations are triples $(d, \rho^R_d, \rho^L_{Ud})$ where $d : B \rel B'$,
-$\rho^R_d$ is a quasi-right-representation for $d$ and $\rho^L{Ud}$ is a quasi-left-representation for $Ud$.
+$\rho^R_d$ is a quasi-right-representation for $d$ and $\rho^L_{Ud}$ is a quasi-left-representation for $Ud$.
 The value and computation squares are the same as those of $\mathcal{M}$.

### 2024-05-23 paper-new/concrete-model.tex
several fixes to paper

@@
 We also give a predomain structure to the natural numbers $\mathbb{N}$, where both the
-ordering and the bisimiarity relation are equality.
+ordering and the bisimilarity relation are equality.
 %
@@
 Previous work has studied such a similar construct, called the guarded lift
-monad \cite{mogelberg-paviotti2016}; our version here our version augments it with the notion of error.
+monad \cite{mogelberg-paviotti2016}; our version here augments it with the notion of error.
 Since we claimed that $\li A$ is a monad, we need to define the monadic operations
@@
 We define a relation $\inat : \mathbb{N} \rel D$ by
-$(n, d) \in \inat$ iff $e_\mathbb{N} \le_D d$.
+$(n, d) \in \inat$ iff $e_\mathbb{N}(n) \le_D d$.
 We similarly define $\itimes : D \times D \rel D$ by

### 2024-05-23 paper-new/extensional-model.tex
several fixes to paper

@@
 
-In order to organize our construction of denotational models we first
+In order to organize our construction of denotational models, we first
 develop sufficient \emph{abstract} categorical semantics of gradually
@@
 To model this cleanly categorically, we follow New, Licata and Ahmed's
-GTT calculus and base our models off of Levy's Call-by-push-value
+GTT calculus and base our models on Levy's Call-by-push-value
 (CBPV) calculus which is a standard model of effectful programming
@@
 varying requirements of which connectives are present
-\cite{levy99,cfm2016,eec}, we will use a variant which models precisely
+\cite{levy99,cfm2016,eec}. We will use a variant which models precisely
 the connectives we require and no more
@@
   be the category $\mathcal V^T$ of algebras of the monad, defining $A
-  \to B$ as the powering of algebras, $F$ as the free algebra and $U$
+  \to B$ as the powering of algebras, $F$ as the free algebra functor and $U$
   as the underlying object functor.
@@
 To additionally model the error terms, we add a requirement that there
-is a natural transformation $\mho : 1 \to U B$. The naturality
+is a natural transformation $\mho : 1 \Rightarrow U$. The naturality
 requirement encodes that strict morphisms (e.g., the denotations of
@@
 $\Gamma \vdash M : A$ as morphisms of any of the equivalent forms
-$\mathcal E(F\Gamma, FA) \cong \mathcal V(\Gamma \vdash UFA) \cong
+$\mathcal E(F\Gamma, FA) \cong \mathcal V(\Gamma , UFA) \cong
 \mathcal E(F(1), \Gamma \to F(A))$. The most interesting type
@@
 
-New and Licata modeled the graduality and type casts for call-by-name
+New and Licata \cite{new-licata18} modeled the graduality and type casts for call-by-name
 gradual typing using \emph{double categories}, which are defined to be
@@
 morphisms as $c : A \rel B$ where $c \in C_{sq}$ and $s(c) = A$ and
-$t(c) = B$. Finally a morphism $\alpha$ from $c$ to $d$ with
+$t(c) = B$. Finally, a morphism $\alpha$ from $c$ to $d$ with
 $s(\alpha) = f$ and $s(\beta) = g$ is visualized as
@@
 \end{tikzcd}\]
-And is thought of as an abstraction of the notion of relatedness of
+Such a morphism is thought of as an abstraction of the notion of relatedness of
 functions: functions take related inputs to related outputs. The
@@
 syntax to call-by-push-value but did not analyze the structure
-categorically. We fill in this missing analysis now: a model of the
-congruence rules of their system can be given by a locally thin
+categorically.
+We fill in this missing analysis now, extending the double categorical semantics
+given in \cite{new-licata18} from cartesian closed categories to CBPV models.
+A model of the congruence rules of the CBPV system can be given by a locally thin
 ``double CBPV model'', which we define as a category internal to the
@@
 \to \mathcal E'$ that strictly preserve all CBPV constructions, see
-the appendix for a more detailed definition. We call this a strict morphism in contrast to a \emph{lax} morphism, which only preserves CBPV constructions up to transformation.
+the appendix for a more detailed definition. We call this a strict morphism
+in contrast to a \emph{lax} morphism, which only preserves CBPV constructions
+up to transformation.
 Some of the data of a double CBPV model can be visualized as follows:
@@
 the error being a least element we add the requirement that $\mho
-\circ ! \ltsq{r(A)}{r(UB)} f$ holds for all $f : \mathcal V(A,B)$.
+\circ {!} \ltsq{r(A)}{r(UB)} f$ holds for all $f : \mathcal V(A,B)$.
 Finally, the dynamic type can be modeled as an arbitrary value type $D$ with

### 2024-05-23 paper-new/intro.tex
several fixes to paper

@@
 functions. Most languages choose between static or dynamic typing and as a result, programmers that initially write their code in a dynamically typed language
-in order to benefit from faster prototyping and development time need to rewrite the
+in order to benefit from faster prototyping and development time need to rewrite
 some or all of their codebase in a static language if they would like to receive the benefits of static
@@
 mechanize these proofs in a reusable way in the Guarded Cubical Agda
-proof assistant, thereby providing a framework to use to more easily
+proof assistant, thereby providing a framework to more easily
 and conveniently prove that existing languages satisfy graduality and
@@
 \begin{enumerate}
-\item First, we give a simple abstract categorical model of GTT using CBPV double categories.
+\item First, we give a simple abstract categorical model of Gradual Type Theory using CBPV double categories.
 \item Next, we modify this semantics to develop reflexive graph- and

### 2024-05-23 paper-new/technical-background.tex
several fixes to paper

@@
 
-Ticked Cubical Type Theory \cite{mogelberg-veltri2019} is an extension of Cubical
-Type Theory \cite{CohenCoquandHuberMortberg2017}
-in which there is an additional sort called \emph{ticks}. Given a clock $k$,
-a tick $t : \tick\, k$ serves as evidence that one unit of time has passed according to the clock $k$.
-The type $\later A$ is represented as a function from ticks of a clock $k$ to $A$.
-The type $A$ is allowed to depend on $t$, in which case we write $\later^k_t A$
-to emphasize the dependence.
+Ticked Cubical Type Theory \cite{mogelberg-veltri2019} is an extension of
+Cubical Type Theory \cite{CohenCoquandHuberMortberg2017} that has clocks as well
+as an additional sort called \emph{ticks}. Ticks were originally introduced in
+\cite{bahr-grathwohl-bugge-mogelberg2017}. Given a clock $k$, a tick $t :
+\tick\, k$ serves as evidence that one unit of time has passed according to the
+clock $k$. In Ticked Cubical Type Theory, the type $\later^k A$ is the type of
+dependent functions from ticks of the clock $k$ to $A$. The type $A$ is allowed
+to depend on $t$, in which case we write $\later^k_t A$ to emphasize the
+dependence.
 
@@
 
-The rules for tick abstraction and application are similar to those of dependent
-$\Pi$ types. 
-In particular, if we have a term $M$ of type $\later^k A$, and we
-have available in the context a tick $t' : \tick\, k$, then we can apply the tick to $M$ to get
-a term $M[t'] : A[t'/t]$. We will also write tick application as $M_t$.
-Conversely, if in a context $\Gamma, t : \tick\, k$ we have that $M$ has type $A$,
-then in context $\Gamma$ we have $\lambda t.M$ has type $\later A$. % TODO dependent version?
+The rules for tick abstraction and application are similar to those of ordinary
+$\Pi$ types. In particular, if in a context $\Gamma, t : \tick\, k$ we have that $M$
+has type $A$, then in context $\Gamma$ we have $\lambda t.M$ has type $\later^k
+A$. % TODO dependent version?
+%
+Conversely, if we have a term $M$ of type $\later^k A$, and we have available in
+the context a tick $t' : \tick\, k$, then we can apply the tick to $M$ to get a
+term $M[t'] : A[t'/t]$. % TODO restriction on tick application
+%
+For the sake of brevity, we will also write tick application as $M_t$.
 

### 2024-05-23 paper-new/concrete-term-model.tex
section in paper on concrete term model

\section{A Concrete Term Model}\label{sec:concrete-term-model}

In this section, we use SGDT to give a set-theoretic denotational semantics to
the terms of the gradually-typed lambda calculus introduced in Section
\ref{sec:GTLC}. This serves two purposes: First, it is a simple setting in which
to employ the tools of SGDT. Second, constructing this semantic model
establishes the validity of the beta and eta principles for the gradually-typed
lambda calculus.
%
In Section \ref{sec:towards-relational-model}, we will discuss how to extend the
denotational semantics to accommodate the type and term precision orderings.

%\subsection{Semantic Constructions}\label{sec:domain-theory}

We begin by discussing the objects of the model into which we will embed
the types and terms of the gradually-typed lambda calculus. Recall that we are
working in an ambient guarded type theory (see Section
\ref{sec:technical-background}) for details. The constructions in this section
will make significant use of the primitives of guarded type theory.

\subsection{Value and Computation Types}

% TODO there aren't actually computation types in the syntax. They
% arise because we are embedding call-by-value into CBPV.

% TODO should we introduce the lift monad first, and then introduce error
% domains as algebras of the lift monad?

We will give an embedding of our call-by-value syntax into a call-by-push-value
model based on a free-forgetful adjunction $F \dashv U$. Value types will denote
sets. Computation types will denote sets with additional algebraic structure
corresponding to the possible behaviors of a term. Recall that we are modelling
gradually-typed programs. Thus, one such behavior is \emph{failure}: a program
may fail at run-time because of a type error. In addition to this, a program may
fail to terminate. This is possible because the dynamic type includes function
types and so is recursive. We will discuss the dynamic type further in a
subsequent section, but for now it suffices to know that we can express a
diverging computation in our syntax. Therefore, it will be necessary to track
the ``steps'' taken by a term; we can regard this as a way of intensionally
modelling \emph{partiality}.

The semantic objects modelling computations will be sets $B$ that are
equipped with a distinguished error element $\mho_B$ and a map $\theta_B \colon
\laterhs B \to B$. We call such sets \emph{error domains}. A \emph{morphism of
error domains} from $B_1$ to $B_2$ is a function $\phi$ on the underlying sets
that respects the error and $\theta$ maps, i.e., for which 
%
(1) $\phi(\mho_{B_1}) = \mho_{B_2}$ and
%
(2) $\phi(\theta_{B_1}(\tilde{x})) = \theta_{B_2}(\later (\phi(\tilde{x})))$
(using the functorial action of $\later$).

\subsection{The Lift + Error Monad}\label{sec:lift-monad}

% When thinking about how to model gradually-typed programs semantically, we
% should consider their possible behaviors. On the one hand, we have
% \emph{failure}: a program may fail at run-time because of a type error. In
% addition to this, a program may fail to terminate. This is possible because the
% dynamic type includes function types and so is recursive. We will discuss the
% dynamic type further in the next section, but for now it suffices to know that
% we can express a diverging computation in our syntax. Therefore, it will be
% useful to track the ``steps'' taken by a term; we can regard this as a way of
% intensionally modelling \emph{partiality}.

% With this in mind, we can describe a semantic object that models these
% behaviors: a monad for embedding computations that has cases for failure and
% ``stepping''. Previous work has studied such a construct in the setting of the
% latter, called the lift monad \cite{mogelberg-paviotti2016}; here, we augment it
% with the additional effect of failure.

We now describe the monad that decomposes into the free-forgetful adjunction
between sets and error domains. We base the construction on the \emph{guarded
lift monad} described in \cite{mogelberg-paviotti2016}. Here, we augment the
guarded lift monad to accommodate the additional effect of failure. For a type
$A$, we define the \emph{guarded lift monad with failure} $\li A$, which we will
just call the \emph{guarded lift monad} or simply the \emph{lift monad}.

\begin{align*}
  \li A &:= \\
  &\eta \colon A \to \li A \\
  &\mho \colon \li A \\
  &\theta \colon \laterhs (\li A) \to \li A
\end{align*}

Unless otherwise mentioned, all constructs involving $\later$ or $\fix$ are
understood to be with respect to a fixed clock $k$. So for the above, we really
have for each clock $k$ a type $\li^k A$ with respect to that clock.
%
Formally, the lift monad $\li A$ is defined as the solution to the guarded
recursive type equation
%
\[ \li A \cong A + 1\, + \laterhs \li A. \]
%
An element of $\li A$ should be viewed as a computation that can either (1)
return a value (via $\eta$), (2) raise an error and stop (via $\mho$), or (3)
take a step (via $\theta$).
%
Notice there is a computation $\fix\, \theta$ of type $\li A$. This represents a
computation that runs forever and never returns a value.

Since we claimed that $\li A$ is a monad, we need to define the monadic
operations and show that they respect the monadic laws. The return is just
$\eta$, and the monadic extension is defined via guarded recursion by cases on
the input.
% It is instructive to give at least one example of a use of guarded recursion, so
% we show below how to define extend:
% TODO
%
%
Verifying that the monadic laws hold uses \lob-induction and is straightforward.

% TODO mention that error domains are algebras of the lift monad?
We observe that the guarded lift monad constructs the free error domain on a set
$A$, i.e., we have $\li = UF$ where $F : \Set \to \errdom$ and $U : \errdom \to
\Set$ and $F \dashv U$. Given a set $A$ and an error domain $B$, the set of
functions $A \to UB$ is naturally isomorphic to the set of error domain
morphisms $FA \to B$.

% The function type A \ra A' will be modeled as \sem{A} \to \li \sem{A'}

\subsection{Modeling the Dynamic Type}\label{sec:dynamic-type}

We discuss how to model the dynamic type $\dyn$.
% For simplicity, we omit the
% case of products in the dynamic type, as from a semantic viewpoint it is the
% function case that is problematic.
We will define a semantic object $D$ that satisfies the isomorphism

% TODO make sure we deal with the product correctly
\[ D \cong \Nat + (D \times D) + (D \to \li D). \]

where the $\li D$ represents the fact that in the case where the value of type
$\dyn$ is a function, the function when called may return an error or take a
step of computation. Unfortunately, this equation does not have inductive or
coinductive solutions. The usual way of dealing with such equations is via
domain theory, by which we can obtain an exact solution. However, the machinery
of domain theory can be difficult for language designers to learn and apply in
the mechanized setting. Instead, we will leverage the tools of guarded type
theory. Consider the following similar-looking equation:
%
\[ D \cong \Nat + (D \times D)\, + \laterhs (D \to \li D). \]
%
Since the negative occurrence of $D$ is guarded under a later, this equation has
a solution constructed via a combination of least fixed-point and guarded
recursion. Specifically, consider the parameterized inductive type
%
\[ D'\, X := \mu T. \Nat + (T \times T) + X. \]
%
Now consider the function $f$ defined by
%
\[ \lambda (\tilde{D} \colon \laterhs \type) . D' (\later_t (\tilde{D}_t \to (\li (\tilde{D}_t)))). \]
%
Recall that the tick $t : \tick$ is evidence that time has passed, and since
$\tilde{D}$ has type $\later \type$, i.e. $\tick \to \type$, then $\tilde{D}_t$
has type $\type$.

Now, define 
\[ D := \fix f. \]
%
% TODO refer to technical background?
Observe that the unfolding equality of the guarded fixpoint implies that $D = D'
(\later D \to \li D)$, and by unfolding $D'$ we see that this is equal to $\Nat
+ (D \times D) + \later (D \to \li D)$.

\subsection{Term Semantics}\label{sec:term-interpretation}

We can now give a semantics to the terms of the gradually-typed lambda calculus.
The full definition is given in Figure \ref{fig:term-semantics}.
%
Much of the semantics is similar to a normal call-by-value denotational
semantics: We interpret types as sets, and terms as functions. Contexts $\Gamma
= x_1 \colon A_1, \dots, x_n \colon A_n$ is interpreted as the product
$\sem{A_1} \times \cdots \times \sem{A_n}$. The semantics of the dynamic type
$\dyn$ is the set $D$ introduced in Section \ref{sec:dynamic-type}. The product
type $A \times A'$ is interpreted as the product of the denotations of $A$ and
$A'$. The function type $A \ra A'$ is interpreted as the set of functions from
$\sem{A}$ to $\li (\sem {A'})$.

The interpretation of a value $\hasty {\Gamma} V A$ is a function from
$\sem{\Gamma}$ to $\sem{A}$. Likewise, a term $\hasty {\Gamma} M {{A}}$ is
interpreted as a function from $\sem{\Gamma}$ to $\li \sem{A}$.

Upcasts are pure, so an upcast $\upc c$ where $c : A \ltdyn A'$ denotes an
ordinary function from $\sem{A}$ to $\sem{A'}$. Downcasts are effectful, so a
downcast $\dnc c$ is interpreted as a morphism of error domains between
$F\sem{A'}$ and $F\sem{A}$, which by definition of the adjunction is
equivalently an ordinary function between $\sem{A'}$ and $UF\sem{A}$, i.e.,
$\sem{A'} \to \li \sem{A}$.

The semantics of the casts for a type precision derivation $c$ are defined by
induction on $c$. The interesting cases are $\inat$, $\itimes$, and $\iarr$.
Recall that $\Dyn$ is isomorphic to $\Nat\, + (\Dyn \times \Dyn) + \laterhs
(\Dyn \to \li \Dyn)$ (say that the sum is left-associative).
%
For $\inat$, the upcast is simply $\inl \circ \inl$. The downcast has type $\Dyn
\to \li \Nat$ and performs a case inspection on the sum type. If it's a natural
number $n$, then we return $\eta\, n$. Otherwise, we return $\mho$, modelling
the fact that a run-time error occurs.
%
For $\itimes$, the upcast is $\inr \circ \inl$, and the downcast performs a case
inspection analogously to the natural number downcast.
%
% TODO explain this further
For $\iarr$, the upcast is given by $\inr \circ \nxt$. The downcast performs a
case inspection, and in the $\inr$ case, it uses a $\theta$ in order to gain
access to the function under the $\later$.
%
% TODO write this out explicitly?

% TODO function casts

% The interpretation of $\lda{x}{M}$ works as follows. Recall by the typing rule
% for lambda that $\phasty {\Gamma, x : A_i} M {{A_o}}$, so the interpretation of
% $M$ has type $(\sem{\Gamma} \times \sem{A_i})$ to $\li \sem{A_o}$. The
% interpretation of lambda is thus a function (in the ambient type theory) that
% takes a value $a$ representing the argument and applies it (along with $\gamma$)
% as argument to the interpretation of $M$.

\vspace{2ex}

\begin{figure*}
  \noindent Types:
  \begin{align*}
    \sem{\nat} &= \Nat \\
    \sem{\dyn} &= \Dyn \\
    \sem{A \times A'} &= \sem{A} \times \sem{A'} \\ 
    \sem{A \ra A'} &= \sem{A} \To \li \sem{A'} \\
  \end{align*}

  % Contexts:

  % TODO check these
  \noindent Values and terms:
  \begin{align*}
    \sem{\zro}         &= \lambda \gamma . 0 \\
    \sem{\suc\, V}     &= \lambda \gamma . (\sem{V}\, \gamma) + 1 \\
    \sem{x \in \Gamma} &= \lambda \gamma . \gamma(x) \\
    \sem{\lda{x}{M}}   &= \lambda \gamma . \lambda a . \sem{M}\, (*,\, (\gamma , a))  \\
    \sem{V_f\, V_x}    &= \lambda \gamma . {({(\sem{V_f}\, \gamma)} \, {(\sem{V_x}\, \gamma)})} \\
    \sem{\err_B}       &= \lambda \gamma . \mho \\
    %
    % upcasts
    \sem{\upc{\inat{V_n}}} &= \lambda \gamma . \inl\, \inl\, (\sem{V_n}\, \gamma) \\
    \sem{\upc{\itimes{V_1, V_2}}} &= \lambda \gamma . \inl\, \inr\, (\sem{V_1}\, \gamma, \sem{V_2}\, \gamma) \\
    \sem{\upc{\iarr{V_f}}} &= \lambda \gamma . \inr\, \nxt (\sem{V_f}\, \gamma) \\[2ex]
    %
    % downcasts
    \sem{\dnc{\inat V_d}} &=
      \lambda \delta . \text{case $(\sem{V_d}\, \gamma)$ of}
      \{ \inl\, (\inl\, n) \to \eta\, n
         \alt \text{otherwise} \to \mho \} \\
    \sem{\dnc{\itimes V_d}} &=
      \lambda \delta . \text{case $(\sem{V_d}\, \gamma)$ of}
      \{ \inl\, (\inr\, (d_1, d_2)) \to \eta\, (d_1, d_2)
         \alt \text{otherwise} \to \mho \} \\
    \sem{\dnc{\iarr V_d}} &=
      \lambda \delta . \text{case $(\sem{V_d}\, \gamma)$ of}
      \{ \inr\, \tilde{f} \to \theta\, (\lambda t. \eta (\tilde{f}_t))
         \alt \text{otherwise} \to \mho \} \\
    %
    % TODO product and function
    %
    % \sem{\casedyn{V}{n} {M_{nat}} {\tilde{f}}{M_{fun}}} &=
    % \lambda \delta . \text{case $(\sem{V}\, \delta)$ of} \\ 
    % &\quad\quad\quad\quad \alt \inl(n) \to \sem{M_{nat}}(n) \\
    % &\quad\quad\quad\quad \alt \inr(\tilde{f}) \to \sem{M_{fun}}(\tilde{f})
    %
    %
    % \sem{\ret\, V}       &= \lambda \gamma . \eta\, \sem{V} \\
    % \sem{\bind{x}{M}{N}} &= \lambda \delta . \ext {(\lambda x . \sem{N}\, (\delta, x))} {\sem{M}\, \delta} \\
  \end{align*}

  \caption{Term semantics for the gradually-typed lambda calculus.}
  \label{fig:term-semantics}
\end{figure*}

### 2024-06-13 paper-new/intro.tex
[paper] add outline of remainder of paper to intro

@@
 
-%% \subsubsection{Classical Domain Models}
-
-%%  and gave a denotational
-%% model of this calculus using classical domain theory based on
-%% $\omega$-CPOs. This semantics has scaled up to an analysis of a
-%% dependently typed gradual calculus in \cite{asdf}. This meets our
-%% criterion of being a reusable mathematical theory, as general semantic
-%% theorems about gradual domains can be developed independent of any
-%% particular syntax and then reused in many different denotational
-%% models. However, it is widely believed that such classical domain
-%% theoretic techniques cannot be extended to model higher-order store, a
-%% standard feature of realistic gradually typed languages such as Typed
-%% Racket. Thus if we want a reusable mathematical theory of gradual
-%% typing that can scale to realistic programming languages, we need to
-%% look elsewhere to so-called ``step-indexed'' techniques.
-
-%% \subsubsection{Operational Reduction Proofs}
-
 % Alternative phrasing:
@@
 
-An alternative approach, which we
-investigate in this paper, is provided by \emph{synthetic guarded
-domain theory}\cite{birkedal-mogelberg-schwinghammer-stovring2011}.
-The techniques of synthetic guarded domain theory allow us to
-internalize the step-indexed reasoning normally required in logical
-relations proofs of graduality, ultimately allowing us to specify the
-logical relation in a manner that looks nearly identical to a typical,
-non-step-indexed logical relation. In fact, guarded domain theory goes
-further, allowing us to define step-indexed \emph{denotational
-semantics} not just step-indexed relations, just as easily as
-constructing an ordinary set-theoretic semantics.
+An alternative approach, which we investigate in this paper, is provided by
+\emph{synthetic guarded domain
+theory}\cite{birkedal-mogelberg-schwinghammer-stovring2011}, abbreviated as
+SGDT. The techniques of SGDT allow us to internalize the step-indexed reasoning
+normally required in logical relations proofs of graduality, ultimately allowing
+us to specify the logical relation in a manner that looks nearly identical to a
+typical, non-step-indexed logical relation. In fact, guarded domain theory goes
+further, allowing us to define step-indexed \emph{denotational semantics} not
+just step-indexed relations, just as easily as constructing an ordinary
+set-theoretic semantics.
 
-In this paper, we develop an adequate denotational semantics that
-satisfies graduality and soundness of the equational theory of cast
-calculi using the techniques of SGDT.  Our longer-term goal is to
-mechanize these proofs in a reusable way in the Guarded Cubical Agda
-proof assistant, thereby providing a framework to more easily
-and conveniently prove that existing languages satisfy graduality and
-have sound equational theories. Moreover, the aim is for designers of
-new languages to utilize the framework to facilitate the design of new
-provably-correct gradually-typed languages with more complex features.
+In this paper, we develop an adequate denotational semantics that satisfies
+graduality and soundness of the equational theory of cast calculi using the
+techniques of SGDT.
+
+\eric{This is out of date: the mechanization is now an artifact of the current paper.}
+Our longer-term goal is to mechanize these proofs in a reusable way in the
+Guarded Cubical Agda \cite{veltri-vezzosi2020} proof assistant, thereby
+providing a framework to more easily and conveniently prove that existing
+languages satisfy graduality and have sound equational theories.
+
+Moreover, the aim is for designers of new languages to utilize the framework to
+facilitate the design of new provably-correct gradually-typed languages with
+more complex features.
 
@@
 
-The main contribution of this work is a categorical and denotational
-semantics for gradually typed langauges that models not just the term
-language but the graduality property as well.
+The main contribution of this work is a categorical and denotational semantics
+for gradually typed langauges that models not just the term language but the
+graduality property as well.
 \begin{enumerate}
-\item First, we give a simple abstract categorical model of Gradual Type Theory using CBPV double categories.
-\item Next, we modify this semantics to develop reflexive graph- and
-  double categorical models that abstract over the details of
-  step-indexed models, and provide a method for constructing such models.
-\item We instantiate the abstract construction to provide a concrete
-  semantics in informal guarded type theory.
-\item We prove that the resulting denotational model is
-  \emph{adequate} for the graduality property: a closed term precision
-  $M \ltdyn N : \nat$ has the expected semantics, that $M$ errors or
-  $M$ and $N$ have the same extensional behavior.
+\item First, we give a simple abstract categorical model of Gradual Type Theory
+using CBPV double categories.
+\item Next, we modify this semantics to develop reflexive graph- and double
+  categorical models that abstract over the details of step-indexed models, and
+  provide a method for constructing such models.
+\item We instantiate the abstract construction to provide a concrete semantics
+  in informal guarded type theory.
+\item We prove that the resulting denotational model is \emph{adequate} for the
+  graduality property: a closed term precision $M \ltdyn N : \nat$ has the
+  expected semantics, that $M$ errors or $M$ and $N$ have the same extensional
+  behavior.
 \end{enumerate}
+
+% TODO is it okay to refer to our cast calculus as *the* gradually-typed lambda
+% calculus?
+\eric{Where should we put the description of the big-step semantics and adequacy
+of the semantics? I can put it in the outline of the remainder of the paper, or
+I can postpone it until the subsequent section on the syntax of gradual typing.
+Adequacy seems to be better introduced there since at that point we have
+introduced the notion of term precision.}
+
+The remainder of the paper is organized as follows. In Section
+\ref{sec:technical-background}, we provide technical background on synthetic
+guarded domain theory and some important details about the specific variant of
+SGDT we employ as our ambient language.
+%
+In Section \ref{sec:GTLC}, we describe the syntactic theory of a standard cast
+calculus for gradual typing, which we also refer to as the \emph{gradually-typed
+lambda calculus}. We do this to establish a common syntax that will be
+interpreted into the models we subsequently construct.
+%
+In Section \ref{sec:cbpv}, we develop \emph{abstract} categorical
+semantics of gradually typed languages with the goal of organizing our
+construction of denotational models of gradual typing.
+%
+In Sections \ref{sec:concrete-term-model} and
+\ref{sec:towards-relational-model}, we set out to instantiate the abstract
+categorical semantics by constructing a concrete model in a suitable double
+category. First, in Section \ref{sec:concrete-term-model}, we give a concrete
+model for the \emph{terms} of the gradually-typed lambda calculus. This involves
+using the tools of SGDT to solve a domain equation to give the semantics of the
+dynamic type. We explain how to \emph{globalize} this model to obtain a model in
+the usual (non-guarded) dependent type theory. In the end, we obtain a
+\emph{big-step semantics} for closed terms of type $\nat$:
+%
+\[ \sem{\cdot} \colon (\cdot \vdash M : \nat) \to (\mathbb{N} \to (\sem{nat} + 1)) \] 
+%
+where a closed term $M$ of type $\nat$ denotes a \emph{function} from
+$\mathbb{N}$ to $\sem{nat} + 1$ which given input $i$ returns $\inl\, m$ if $M$
+terminates with value $m$ in $i$ or fewer steps, and returns $\inr\, *$ if $M$
+fails to terminate in $i$ or fewer steps. This is already a significant result,
+as it serves to verify the $\beta$ and $\eta$ laws of the equational theory.
+%
+Then in Section \ref{sec:towards-relational-model}, we attempt to extend the
+methodology of the previous section to give a denotational semantics that models
+the relational theory of the gradually-typed lambda calculus in addition to the
+terms. Interesting issues arise here arising from the \emph{intensional} nature
+of SGDT, that is, the fact that \emph{steps are observable}. We then develop
+novel solutions to address these issues. We do not complete the construction of
+the concrete relational model in this section. Instead, we choose to reformulate
+the definition of abstract categorical model to take into account the solutions
+to the issues discussed here.
+
+% Interesting issues arise here, and the novel solutions we employ to
+% address these issues inform the definition of the abstract models introduced in
+% the subsequent section.
+
+With the lessons from the concrete setting at our disposal, in Section
+\ref{sec:abstract-models}, we describe revised categorical models of gradual
+typing. We actually introduce two separate notions of model: \emph{intensional},
+where the stepping behavior of terms is observable, and \emph{extensional},
+where we consider terms that differ only in their stepping behavior to be equal.
+The extensional model is what we will ultimately use to interpret the syntax and
+relational axioms of gradually-typed languages. The principal benefit of the
+notion of intensional model is that it admits more compositional reasoning. It
+therefore serves as a convenient intermediate setting that enjoys many of the
+benefits of the usual double-categorical models of gradual typing.
+% We give the definition of intensional model as a sequence of steps each
+% building on the previous. 
+We then outline how to build an extensional model in a sequence of steps beginning
+with a significantly simpler ``input'' model. The construction proceeds in
+steps, first building up an intensional model, and then from there extracting an
+extensional model.
+%We are able to leverage the compositional reasoning afforded by the intensional
+%model to more easily meet the requirements of an extensional model.
+This gives us a more tractable way to construct models.
+Moreover, the abstract definitions and constructions in the definitions make no
+mention of SGDT, so that the places where SGDT techniques are employed are
+conveniently limited to the data passed as input to the model construction.
+%
+%In addition to organizing and facilitating the construction of a model, the abstract constructions are described independently of SGDT
+%
+%we also isolate the parts of the definition the notion of model without any explicit mention of SGDT.
+%
+Then in Section \ref{sec:concrete-model} we return to the task of defining a
+concrete model. Equipped now with the abstract constructions of the previous
+section, we provide only the data needed as an input to the construction. Many
+of the relevant definitions have already been introduced in Section
+\ref{sec:towards-relational-model} and can be provided to the construction
+as-is; the remainder of the necessary data is defined in this section.
+%
+We end the section by sketching a proof of the \emph{adequacy} of our concrete
+model with respect to graduality.
+%
+Finally, in Section \ref{sec:discussion} we conclude with discussion: comparison
+to related work, the mechanization of our concrete model in Guarded Cubical
+Agda, and directions for future research.
 

### 2024-06-13 paper-new/technical-background.tex
[paper] Update background on Ticked Cubical Type Theory

@@
-\section{Background on Guarded Domain Theory}\label{sec:technical-background}
+\section{Background on Synthetic Guarded Domain Theory}\label{sec:technical-background}
 
@@
 There is a \emph{guarded fixpoint} operator
-
+%
 \[
@@
 \]
-
+%
 That is, to construct a term of type $T$, it suffices to assume that we have access to
@@
 
+% TODO include a figure with some of the rules for ticks
+
 The rules for tick abstraction and application are similar to those of ordinary
-$\Pi$ types. In particular, if in a context $\Gamma, t : \tick\, k$ we have that $M$
-has type $A$, then in context $\Gamma$ we have $\lambda t.M$ has type $\later^k
-A$. % TODO dependent version?
+$\Pi$ types. A context now consists of ordinary variables $x : A$ as well as
+tick variables $t : \tick\, k$. The presence of the tick variable $t$ in context
+$\Gamma, (t : \tick\, k), \Gamma'$ intuitively means that the values of the
+variables in $\Gamma$ arrive ``first'', then one time step occurs on $k$, and
+then the values of the variables in $\Gamma'$ arrive.
+
+The abstraction rule for ticks states that if in context $\Gamma, t : \tick\, k$
+the term $M$ has type $A$, then in context $\Gamma$ the term $\lambda t.M$ has
+type $\later^k_t A$. % TODO dependent version?
 %
@@
 the context a tick $t' : \tick\, k$, then we can apply the tick to $M$ to get a
-term $M[t'] : A[t'/t]$. % TODO restriction on tick application
+term $M[t'] : A[t'/t]$. However, there is an important restriction on when we
+are allowed to apply ticks. In order to apply $M$ to tick $t$, $M$ must be
+well-typed in the prefix of the context occurring before the tick $t$. That is,
+all variables mentioned in $M$ must be available before $t$. This ensures that
+we cannot, for example, define a term of type $\later \laterhs A \to\, \laterhs
+A$ via repeated tick application.
+% TODO restriction on tick application
 %

### 2024-06-13 paper-new/concrete-term-model.tex
[paper] work on concrete term model section

@@
 
-In this section, we use SGDT to give a set-theoretic denotational semantics to
-the terms of the gradually-typed lambda calculus introduced in Section
-\ref{sec:GTLC}. This serves two purposes: First, it is a simple setting in which
-to employ the tools of SGDT. Second, constructing this semantic model
-establishes the validity of the beta and eta principles for the gradually-typed
-lambda calculus.
+\eric{Wording?}
+The principal goal of the next two sections is to ground the refined notions of
+abstract model of gradual typing described in Section \ref{sec:abstract-models}
+in a concrete setting. We aim to introduce the key issues whose resolutions in
+the concrete setting inspire the refined definitions of abstract model.
+
+In the present section, we use the techniques of SGDT to give a set-theoretic
+denotational semantics to the terms of the gradually-typed lambda calculus
+introduced in Section \ref{sec:GTLC}. This serves two important purposes: First,
+it acts as a warm-up for employing the tools of SGDT to define a denotational
+semantics. Second, constructing this semantic model establishes the validity of
+the $\beta$ and $\eta$ principles for the gradually-typed lambda calculus. It
+also acts as a stepping stone into the next section, where we extend the
+techniques employed here to accommodate the type and term precision orderings of
+the gradually-typed lambda calculus.
 %
-In Section \ref{sec:towards-relational-model}, we will discuss how to extend the
-denotational semantics to accommodate the type and term precision orderings.
+% In Section \ref{sec:towards-relational-model}, we will discuss how to extend the
+% denotational semantics to accommodate the type and term precision orderings.
+
+%necessitate the abstract components that are found in the definition of the model.
 
@@
 
-We begin by discussing the objects of the model into which we will embed
-the types and terms of the gradually-typed lambda calculus. Recall that we are
-working in an ambient guarded type theory (see Section
+We begin by discussing the objects of the model into which we will embed the
+types and terms of the gradually-typed lambda calculus. Recall that we are
+working informally in a guarded type theory with clocks and ticks (see Section
 \ref{sec:technical-background}) for details. The constructions in this section
@@
 
-The semantic objects modelling computations will be sets $B$ that are
-equipped with a distinguished error element $\mho_B$ and a map $\theta_B \colon
-\laterhs B \to B$. We call such sets \emph{error domains}. A \emph{morphism of
-error domains} from $B_1$ to $B_2$ is a function $\phi$ on the underlying sets
-that respects the error and $\theta$ maps, i.e., for which 
+The semantic objects modelling computations will be sets $B$ that are equipped
+with a distinguished error element $\mho_B$ and a map $\theta_B \colon \laterhs
+B \to B$. We call such sets \emph{error domains}. A \emph{morphism of error
+domains} from $B_1$ to $B_2$ is a function $\phi$ on the underlying sets that
+respects the error and $\theta$ maps, i.e., for which 
 %
@@
 just call the \emph{guarded lift monad} or simply the \emph{lift monad}.
-
+%
 \begin{align*}
-  \li A &:= \\
-  &\eta \colon A \to \li A \\
-  &\mho \colon \li A \\
-  &\theta \colon \laterhs (\li A) \to \li A
+  \li A :=\quad
+  &\alt \eta \colon A \to \li A \\
+  &\alt \mho \colon \li A \\
+  &\alt \theta \colon \laterhs (\li A) \to \li A
 \end{align*}
-
+%
 Unless otherwise mentioned, all constructs involving $\later$ or $\fix$ are
@@
 %
-\[ \lambda (\tilde{D} \colon \laterhs \type) . D' (\later_t (\tilde{D}_t \to (\li (\tilde{D}_t)))). \]
+\[ \lambda (\tilde{D} \colon \laterhs \type) . D' (\laterhs_t (\tilde{D}_t \to (\li (\tilde{D}_t)))). \]
 %
@@
 has type $\type$.
-
-Now, define 
+%
+Finally, define 
 \[ D := \fix f. \]
@@
 Observe that the unfolding equality of the guarded fixpoint implies that $D = D'
-(\later D \to \li D)$, and by unfolding $D'$ we see that this is equal to $\Nat
-+ (D \times D) + \later (D \to \li D)$.
+(\later D \to \li D)$, and by the definition of $D'$ as a least fixpoint we see
+that this is equal to $\Nat + (D \times D) + \later (D \to \li D)$.
 
@@
 We can now give a semantics to the terms of the gradually-typed lambda calculus.
-The full definition is given in Figure \ref{fig:term-semantics}.
+The definition of the interpretation of types and cast terms is given in Figure
+\ref{fig:term-semantics}.
 %
@@
 semantics: We interpret types as sets, and terms as functions. Contexts $\Gamma
-= x_1 \colon A_1, \dots, x_n \colon A_n$ is interpreted as the product
+= x_1 \colon A_1, \dots, x_n \colon A_n$ are interpreted as the product
 $\sem{A_1} \times \cdots \times \sem{A_n}$. The semantics of the dynamic type
@@
 induction on $c$. The interesting cases are $\inat$, $\itimes$, and $\iarr$.
-Recall that $\Dyn$ is isomorphic to $\Nat\, + (\Dyn \times \Dyn) + \laterhs
-(\Dyn \to \li \Dyn)$ (say that the sum is left-associative).
+Recall that $D$ is isomorphic to $\Nat\, + (D \times D)\, + \laterhs (D \to \li
+D)$ (say that the sum is left-associative).
 %
-For $\inat$, the upcast is simply $\inl \circ \inl$. The downcast has type $\Dyn
-\to \li \Nat$ and performs a case inspection on the sum type. If it's a natural
+For $\inat$, the upcast is simply $\inl \circ \inl$. The downcast has type $D
+\to \li D$ and performs a case inspection on the sum type. If it is a natural
 number $n$, then we return $\eta\, n$. Otherwise, we return $\mho$, modelling
@@
 
-% TODO function casts
+% TODO should we include function casts?
 
@@
 
-\vspace{2ex}
-
 \begin{figure*}
@@
     \sem{\nat} &= \Nat \\
-    \sem{\dyn} &= \Dyn \\
+    \sem{\dyn} &= D \\
     \sem{A \times A'} &= \sem{A} \times \sem{A'} \\ 
@@
   % TODO check these
-  \noindent Values and terms:
+  \noindent Casts:
   \begin{align*}
-    \sem{\zro}         &= \lambda \gamma . 0 \\
-    \sem{\suc\, V}     &= \lambda \gamma . (\sem{V}\, \gamma) + 1 \\
-    \sem{x \in \Gamma} &= \lambda \gamma . \gamma(x) \\
-    \sem{\lda{x}{M}}   &= \lambda \gamma . \lambda a . \sem{M}\, (*,\, (\gamma , a))  \\
-    \sem{V_f\, V_x}    &= \lambda \gamma . {({(\sem{V_f}\, \gamma)} \, {(\sem{V_x}\, \gamma)})} \\
-    \sem{\err_B}       &= \lambda \gamma . \mho \\
+    % \sem{\zro}         &= \lambda \gamma . 0 \\
+    % \sem{\suc\, V}     &= \lambda \gamma . (\sem{V}\, \gamma) + 1 \\
+    % \sem{x \in \Gamma} &= \lambda \gamma . \gamma(x) \\
+    % \sem{\lda{x}{M}}   &= \lambda \gamma . \lambda a . \sem{M}\, (*,\, (\gamma , a))  \\
+    % \sem{V_f\, V_x}    &= \lambda \gamma . {({(\sem{V_f}\, \gamma)} \, {(\sem{V_x}\, \gamma)})} \\
+    % \sem{\err_B}       &= \lambda \gamma . \mho \\
     %
@@
 \end{figure*}
+
+\eric{TODO: big-step semantics via clock quantification}
+
+\subsection{A Big-Step Term Semantics}

### 2024-06-13 paper-new/concrete-relational-model.tex
[paper] new section on extending the concrete term model to a relational model

\section{Towards a Full Relational Model}\label{sec:towards-relational-model}

\eric{Wording?}
In this section, we build on the ideas used in defining a term model with the
goal of defining a model of the full gradually-typed lambda calculus, i.e., one
that has constructs that model type and term precision.
%
% The purpose of this section is to motivate the need for many of the
% seemingly complicated features of the abstract model introduced in
% Section \ref{sec:abstract-models}.

We will not give the complete description of the concrete model in this section;
rather we seek to develop just enough to motivate the need for the non-standard
aspects in the definitions of abstract model introduced in Section
\ref{sec:abstract-models}. After defining the notion of abstract model and
describing a modular way of constructing such a model, we will return in Section
\ref{sec:concrete-model} to the complete construction of a concrete model.

\subsection{A First Attempt}

% TODO should we frame this in terms of the idealized categorical models
% i.e. say what the vertical and horizontal morphisms and squares are
% introduced previously? What is the proper denotation of values versus terms?
% We don't distinguish between value types and computation types in the syntax

As a first attempt at giving a semantics to the term precision relation, we
could try to model types as sets equipped with an ordering that models term
precision. Since term precision is reflexive and transitive, and since we
identify terms that are equi-precise, we could model value types as
partially-ordered sets and values $\Gamma \vdash V : A$ as \emph{monotone}
functions. Analogously, every error domain is now equipped with a partial
ordering for which the error element is the bottom, and the map $\theta_B
\laterhs B \to B$ is now required to be monotone. 
%
\footnote{This requires us to define a notion of ``later'' for partially-ordered
sets $X$, but this can be done. The elements are simply ``later''-elements of
$X$, and the ordering is the obvious one, i.e., two later-elements are related
if after one time step, the elements they step to are related.}
%
% \eric{What about computation
% types? We need to extend the error domains mentioned in the term model section
% to also have an ordering such that the error element is the least element and
% the theta morphism is monotone. Is this best introduced here, or should we wait
% until after we have introduced the revised notions of abstract model?}
%
We model type precision $c : A \ltdyn B$ as a \emph{montone relation}, i.e., a
relation $R$ that is upward-closed under the relation on $B$ and downward-closed
under the relation on $A$. We could then model the term precision ordering $M
\ltdyn N : A \ltdyn B$ as an ordering relation between the monotone functions
denoted by $M$ and $N$.

However, it turns out that modeling term precision is not as straightforward as
one might hope. A first attempt might be to define an ordering $\semltbad$
between $\li X$ and $\li Y$ that allows for computations that may take different
numbers of steps to be related. We call this the \emph{step-insensitive error
ordering}. The ordering is parameterized by an ordering relation $\le$ between
$X$ and $Y$. The definition is by guarded recursion and is shown in Figure
\ref{fig:step-insensitive-error-ordering}.

\begin{figure*}
    \begin{align*}
        &\eta\, x \semltbad \eta\, y \text{ if } 
            x \semlt y \\
        %		
        &\mho \semltbad l \\
        %
        &\theta\, \tilde{l} \semltbad \theta\, \tilde{l'} \text{ if } 
            \later_t (\tilde{l}_t \semltbad \tilde{l'}_t) \\
        %	
        &\theta\, \tilde{l} \semltbad \mho \text{ if } 
            \theta\, \tilde{l} = \delta^n(\mho) \text { for some $n$ } \\
        %	
        &\theta\, \tilde{l} \semltbad \eta\, y \text{ if }
            (\theta\, \tilde{l} = \delta^n(\eta\, x))
        \text { for some $n$ and $x : \ty{X}$ such that $x \le y$ } \\
        %
        &\mho \semltbad \theta\, \tilde{l'} \text { if } 
            \theta\, \tilde{l'} = \delta^n(\mho) \text { for some $n$ } \\
        %	
        &\eta\, x \semltbad \theta\, \tilde{l'} \text { if }
            (\theta\, \tilde{l'} = \delta^n (\eta\, y))
        \text { for some $n$ and $y : \ty{Y}$ such that $x \le y$ }
    \end{align*}
    \caption{The step-insensitive error ordering.}
    \label{fig:step-insensitive-error-ordering}
\end{figure*}

Two computations that immediately return $(\eta)$ are related if the returned
values are related in the underlying ordering. The computation that errors
$(\mho)$ is the least term in the ordering. If both sides step (i.e., both sides
are $\theta$), then we allow one time step to pass and compare the resulting
terms (this is where use the relation defined ``later'' and is why we employ
guarded recursion to define the relation).
%
Lastly, if one side steps and the other immediately returns a value, then in
order for these terms to be related, the side that steps must terminate with a
value in some finite number of steps $n$, and that value must be related to
the value returned by the other side. Likewise, if the LHS steps and the RHS
immediately errors, then the LHS must eventually terminate with error.

\subsection{Problem: Failure of Transitivity}

The problem with the above definition is that the relation is \emph{not}
transitive, which we explain below. We begin with a general ``no-go'' theorem
about relations defined on the guarded lift monad:

\begin{theorem}[No-go theorem for relations on $\li X$]\label{thm:no-go}
    Let $R$ be a relation on the guarded lift monad satisfying the following
    three properties:
    \begin{enumerate}
        \item Transitivity

        \item Congruence with respect to $\theta$: If $\later_t (\tilde{x}_t
        \binrel{R} \tilde{y}_t)$, then $\theta(\tilde{x}) \binrel{R}
        \theta(\tilde{y})$

        \item Closure under delay on both sides: If $x \binrel{R} y$, then $x
        \binrel{R} \delta y$ and $\delta x  \binrel{R} y$
    \end{enumerate}

    Then $R$ is trivial, i.e., $x \binrel{R} y$ for all $x : \li X$ and $y : \li
    Y$.

\end{theorem}

This theorem follows from two ``one-sided'' lemmas where we assume that the
relation $R$ is closed under delay on the left (resp. right) and show that
$(\fix\, \theta) \binrel{R} y$ for all $y$ (resp. $x \binrel{R} (\fix\, \theta)$
for all $x$). These lemmas in turn are proved by \lob-induction.

The theorem implies that the relation $\semltbad$ defined on $\li X \times \li
Y$ is \emph{not} transitive whenever the underlying relation $\semlt$ on $X$ and
$Y$ is not trivial, i.e., when there are $x_0$ and $y_0$ that are not related in
the relation $\semlt$ on $X$ and $Y$. To show this, first observe that
$\semltbad$ satisfies properties (2) and (3) in the above theorem. Congruence
with respect to $\theta$ is immediate from the definition, while closure under
delay can be proved using \lob-induction.
% TODO the proof of this uses tick irrelevance
Then since $x_0$ is not related to $y_0$, we have that $\eta\, x_0$ is not
related to $\eta\, y_0$ in $\semltbad$, i.e., $\semltbad$ is non-trivial. But if
it were transitive, then by the theorem, it would be trivial, a contradiction.
Thus, we are able to refute transitivity of $\semltbad$ provided the underlying
relation is non-trivial.

% It can be shown (in Clocked Cubical Type
% Theory) that if $R$ is a relation on $\li X$ satisfying three specific
% properties, one of which is transitivity, then that relation is trivial. The
% other two properties are that the relation is a congruence with respect to
% $\theta$, and that the relation is closed under delays $\delta = \theta \circ
% \nxt$ on either side. Since the above relation \emph{does} satisfy the other
% two properties, we conclude that it must not be transitive.

\eric{I'm not sure that this is a satisfactory explanation...}
At this point, we could give up on transitivity and instead of requiring types
to be modeled by partially-ordered sets, we could model types as sets equipped
with a reflexive, anti-symmetric ordering relation. However, this is not
convenient for carrying out the constructions necessary for proving graduality.
For example, in order to model the UpL/UpR/DnL/DnR rules for casts, in the
presence of transitivity it suffices to model simpler rules that do not build in
composition. We can then use transitivity to derive the actual, more complicated
rules we need to show.
% For example, in order to model the UpL/UpR/DnL/DnR rules for casts, prior work
% has taken as primitive simpler rules that do not involve composition, and then
% uses transitivity to establish the more complicated versions that have
% composition ``built in''.
Without transitivity, however, this is not possible, and so we must instead take as
primitive the more complicated versions of the cast rules. Then in the model, we
must verify that those rules hold ``from scratch''. So, although this approach
would suffice for proving graduality, it lacks the compositionality that we seek
in a reusable framework for the semantics of gradual typing.

%Because it is convenient to make use of transitive reasoning in proving graduality, ...
We are therefore led to wonder whether we can formulate an ordering
relation on $\li X \times \li Y$ that \emph{is} transitive. The theorem above
tells us, however, that any such relation must either not be a congruence with
respect to $\theta$, or must not be closed under delay.
\footnote{Technically, we only need to give up closure under delay on \emph{one}
side, as the above theorem requires $R$ to be closed on both sides. We could
define two separate ``step-semi-sensitive'' relations on $\li X \times \li Y$,
each closed under delays on the left and right respectively. The denotation of
term precision would then involve the conjunction of those relations. We do not
pursue this approach in this paper, because (1) it is not clear if the
step-semi-sensitive relations would be transitive, and (2) we would need to
reason about the two ordering relations separately for every pair of terms.}
%
We cannot forego the property of being a $\theta$-congruence, as without this we
would not be able to prove properties of the relation using \lob-induction,
e.g., that the monadic bind for the guarded lift monad is a \emph{monotone}
function with respect to the ordering.
%and moreover this would mean that $\theta$ would not be a monotone function.

Thus, we choose to sacrifice closure under delays. That is, we will define an
ordering relation that requires terms to be in ``lock-step''. In order for two
computations to be related in this ordering, they must have the same stepping
behavior (or the left-hand side must be an error). To deal with terms that take
differing numbers of steps, we then define a separate relation called \emph{weak
bisimilarity} that relates terms that are extensionally equal and may differ by
a finite number of delays. Then, the semantics of the error ordering for the
guarded lift monad will involve a combination of these two relations: the
``closure'' of the lock-step error ordering under weak bisimilarity on both
sides.
% Although the combined relation will not be transitive (for the same reason that
% $\semltbad$ is not transitive), this...
This decomposition has the advantage that we can recover some transitive
reasoning and push much of the reasoning about stepping to the margins of the
development.

% Doing so, we define a \emph{lock-step} error ordering, where roughly speaking,
% in order for computations to be related, they must have the same stepping
% behavior. We then formulate a separate relation, \emph{weak bisimilarity}, that
% relates computations that are extensionally equal and may differ only in their
% stepping behavior. % up to a finite number of \delta's
% Finally, the semantics of term precision will involve a combination of these two
% relations, a sort of closure of the lock-step ordering under weak bisimilarity
% on both sides. This decomposition has the advantage that we can recover some
% transitive reasoning and push the parts involving stepping to the margins of the
% development.

\subsection{Solution: Splitting into Two Relations}

As mentioned, to recover transitive reasoning we will decompose the error
ordering on the guarded lift monad into two separate relations. The first is the
\emph{lock-step error-ordering} (also called the ``step-sensitive error
ordering''), the idea being that two computations $l$ and $l'$ are related if
they are in lock-step with regard to their intensional behavior, up to $l$
erroring. This will be the partial ordering with which we equip the guarded lift monad.
Figure \ref{fig:lock-step-error-ordering} gives the definition.

\begin{figure}
    \fbox{$l_1 \ltls l_2$}
    \begin{align*}
        &\eta\, x \ltls \eta\, y \text{ if } 
            x \mathbin{R} y \\
        %		
        &\mho \ltls l' \\
        %
        &\theta\, \tilde{l} \ltls \theta\, \tilde{l'} \text{ if } 
            \later_t (\tilde{l}_t \ltls \tilde{l'}_t)
    \end{align*}
    \caption{The lock-step error ordering.}
    \label{fig:lock-step-error-ordering}
\end{figure}

When both sides are $\eta$, then we check that the returned values are related
in $R$. The error term $\mho$ is below everything. Lastly, if both sides step
(i.e., are a $\theta$) then we compare the resulting computations one time step
later.
%
It is straightforward to prove using \lob-induction that this relation is
reflexive, transitive and anti-symmetric given that the underyling relation $R$
has those properties. The lock-step ordering is therefore the partial ordering
we will associate with $\li X$.
%
We similarly define a heterogeneous version of this ordering between the lifts
of two different types $X$ and $Y$, parameterized by a relation $R$ between $X$
and $Y$.

The second relation we define on the guarded lift monad is \emph{weak bisimilarity}.
% For a type $X$, we define a relation on $\li X$, called ``weak bisimilarity",
% written $l \bisim l'$. 
Intuitively, we say $l \bisim l'$ if they are the same ``up to delays''.
%
Weak bisimilarity on $\li X$ is parameterized by an underlying relation $R$ on $X$
and is defined by guarded fixpoint as shown in Figure \ref{fig:weak-bisimilarity}.

\begin{figure}
    \fbox{$l_1 \bisim l_2$}
    \begin{align*}
        &\mho \bisim \mho \\
      %
        &\eta\, x \bisim \eta\, y \text{ if } 
          x \mathbin{R} y \\
      %		
        &\theta\, \tilde{x} \bisim \theta\, \tilde{y} \text{ if } 
          \later_t (\tilde{x}_t \bisim \tilde{y}_t) \\
      %	
        &\theta\, \tilde{x} \bisim \mho \text{ if } 
          \theta\, \tilde{x} = \delta^n(\mho) \text { for some $n$ } \\
      %	
        &\theta\, \tilde{x} \bisim \eta\, y \text{ if }
          (\theta\, \tilde{x} = \delta^n(\eta\, x))
        \text { for some $n$ and $x : X$ such that $x \mathbin{R} y$ } \\
      %
        &\mho \bisim \theta\, \tilde{y} \text { if } 
          \theta\, \tilde{y} = \delta^n(\mho) \text { for some $n$ } \\
      %	
        &\eta\, x \bisim \theta\, \tilde{y} \text { if }
          (\theta\, \tilde{y} = \delta^n (\eta\, y))
        \text { for some $n$ and $y : X$ such that $x \mathbin{R} y$ }
      \end{align*}
    \caption{The weak bisimilarity relation.}
    \label{fig:weak-bisimilarity}
\end{figure}

When both sides are $\eta$, then we ensure that the underlying values are
bisimilar in the underlying bisimilarity relation on $A$. When one side is a
$\theta$ and the other is $\eta x$ (i.e., one side steps), we stipulate that the
$\theta$-term runs to $\eta y$ where $x$ is related to $y$. When one side is
$\theta$ and the other $\mho$, we require that the $\theta$-term runs to $\mho$.
If both sides step, then we allow one time step to pass and compare the
resulting terms. This definition captures the intuition of two terms being
related up to delays.

It can be shown (by \lob-induction) that weak bisimilarity is reflexive and
symmetric. We can also show that this relation is \emph{not} transitive. The
argument is the same as that used to show that the step-insensitive error
ordering $\semltbad$ described above is not transitive. In particular, we
observe that weak bisimilarity is a congruence with respect to $\theta$, and we
can show by $\lob$-induction that it is closed under delay on both sides. Thus,
we may appeal to the no-go theorem as before.
% We appeal to the no-go theorem given above

% This issue of non-transitivity will be resolved when we consider the relation's \emph{globalization}.

Now we can define the denotation of term precision for closed terms of type
$\nat$ as follows:
%
\[ l_1 \leq l_2 := 
  \Sigma_{l_1', l_2' \in \li \mathbb{N}}. l_1 \bisim l_1' \ltls l_2' \bisim l_2. \]

\eric{Does this belong in the next section?}
Importantly, we note that as with the original ordering $\semltbad$ defined at
the beginning of this section, the relation just defined is not transitive
(again as a result of the above no-go theorem). It may therefore seem that we
have not solved the original issue we faced. In a sense, this is true. Indeed,
we will need to weaken our abstract notion of categorical model to no longer
require transitivity of term precision. On the other hand, we now have the
ability to apply transitive reasoning when working with the lock-step ordering.
This allows us to carry out complex proofs compositionally, a key goal of our
framework.

\subsection{Failure of the Cast Rules}

We have seen how we can recover transitive reasoning by working with the
lock-step error ordering. However, we cannot completely isolate the reasoning
about steps from the reasoning about the ordering.
% Unfortunately, by defining the partial order on the guarded lift monad to be
% the lock-step error ordering, we introduce a new issue.
In particular, the cast rules required for proving graduality no longer hold if
interpreted using the lock-step ordering. The source of the problem is that the
downcast from the dynamic type to a function takes a step, i.e., involves a
$\theta$. Consider a simplified version of the DnL rule shown below (see Section
\ref{sec:GTLC} for background):
%
\begin{mathpar}
  \inferrule*{c : A \ltdyn B \and M \ltdyn_i N : B}
             {\dnc{c}{M} \ltdyn_i N : c}
\end{mathpar}
%
Suppose that $c$ is $\iarr \colon \dyntodyn \ltdyn\, \dyn$, and recall the
definition of the semantics of a downcast given in Section
\ref{sec:term-interpretation}. When we downcast $M$ from $dyn$ to $\dyntodyn$,
semantically this will involve a $\theta$ because the value of type $D$ in the
model will be a \emph{later} function $\tilde{f}$.
%
Thus, in order for the downcast to be related to the term $N$ in the lock-step
error ordering on $\li (D \to \li D)$, the right-hand side must be of the form
$\theta(\dots)$, and moreover, after one time step, the argument to $\theta$ on
the LHS must be related to the argument of $\theta$ on the RHS. We can achieve
this by inserting a ``delay'' on the right, i.e., wrapping the RHS in $\delta =
\theta \circ \nxt$. Notice that this delay function is weakly bisimlar to the
identity, and thus it will disappear when we apply the bisimilarity-closure
construction described above. That is, this delay makes no difference in the
extensional setting, but its presence is crucial in the intensional setting.

% TODO Maybe show the correct form of the rule in the special case $c = inj-arr$.

\subsection{Extending the Model to Higher-Order Types}

% At this point, we could carry out the full construction of a concrete relational
% model with these additional features.

So far, we have been focusing on closed terms of base type, for which the above
techniques give a semantic interpretation of term precision. We now consider how
to extend these constructions to \emph{all} types. In particular, in order to
model higher-order data types (i.e., functions) we need to equip \emph{every}
semantic object with not only a partial ordering relation but also a
``bisimilarity'' relation that is reflexive and symmetric. We must similarly
equip every object with a structured set of delays that can be inserted to
ensure the appropriate cast rules hold at all higher-order types. For example,
the upcast for a derivation $c \ra d$ involves the downcast corresponding to $c$
and an upcast corresponding to $d$. We must be able to insert delays in a
functorial manner to mimic the structure of the casts.

For the sake of reusability and modularity, rather than carry out this
construction in the concrete setting developed here, we will instead return to
the abstract setting and adjust our definition of model to account for these
requirements. We will break the construction into smaller steps and isolate the
pieces that require the techniques of SGDT from those that do not. Then with
this framework at our disposal, we will return to the construction of a concrete
model in Section \ref{sec:concrete-model}.
% , taking as a starting point the definitions introduced in the current section.

% Thus in the next section we define revised notions of a model of
% gradual typing based on the lessons learned in the present section.

### 2024-07-05 paper-new/appendix.tex
[paper] small correction in appendix

@@
 
-  % https://q.uiver.app/#q=WzAsNCxbMCwwLCJGQSJdLFsxLDAsIkZBJyJdLFswLDEsIkZBIl0sWzEsMSwiRkEnIl0sWzIsMywiRmMiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIkZjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwyLCIoXFxkZWx0YV97RkF9XiopXm4gXFxjaXJjIFxccHRiX0EoYSkiLDJdLFsxLDMsIihcXGRlbHRhX3tGQSd9XiopXm4gXFxjaXJjIFxccHRiX3tBJ30oXFxwdXNoX2MoYSkpIl1d
+  % https://q.uiver.app/#q=WzAsNCxbMCwwLCJGQSJdLFsxLDAsIkZBJyJdLFswLDEsIkZBIl0sWzEsMSwiRkEnIl0sWzIsMywiRmMiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIkZjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwyLCIoXFxkZWx0YV97RkF9XiopXm4gXFxjaXJjIEYoXFxwdGJfQShhKSkiLDJdLFsxLDMsIihcXGRlbHRhX3tGQSd9XiopXm4gXFxjaXJjIEYoXFxwdGJfe0EnfShcXHB1c2hfYyhhKSkpIl1d
   \[\begin{tikzcd}[ampersand replacement=\&]
@@
     \arrow["Fc", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-    \arrow["{(\delta_{FA}^*)^n \circ \ptb_A(a)}"', from=1-1, to=2-1]
-    \arrow["{(\delta_{FA'}^*)^n \circ \ptb_{A'}(\push_c(a))}", from=1-2, to=2-2]
+    \arrow["{(\delta_{FA}^*)^n \circ F(\ptb_A(a))}"', from=1-1, to=2-1]
+    \arrow["{(\delta_{FA'}^*)^n \circ F(\ptb_{A'}(\push_c(a)))}", from=1-2, to=2-2]
   \end{tikzcd}\]
@@
   This is obtained by pasting $n$ copies of the square $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$
-  on top of the square corresponding to the push property for $c$.
+  on top of the square obtained by lifting via $F$ the square corresponding to the push property for $c$.
 

### 2024-07-05 paper-new/intro.tex
[paper] some minor fixes

@@
 %
-\[ \sem{\cdot} \colon (\cdot \vdash M : \nat) \to (\mathbb{N} \to (\sem{nat} + 1)) \] 
+\[ \sem{\cdot} \colon (\cdot \vdash M : \nat) \to (\mathbb{N} \to (\sem{\nat} + 1)) \] 
 %
 where a closed term $M$ of type $\nat$ denotes a \emph{function} from
-$\mathbb{N}$ to $\sem{nat} + 1$ which given input $i$ returns $\inl\, m$ if $M$
+$\mathbb{N}$ to $\sem{\nat} + 1$ which given input $i$ returns $\inl\, m$ if $M$
 terminates with value $m$ in $i$ or fewer steps, and returns $\inr\, *$ if $M$

### 2024-07-05 paper-new/technical-background.tex
[paper] some minor fixes

@@
 SGDT offers a synthetic approach to domain theory that allows for guarded recursion
-to be expressed syntactically via a type constructor $\later : \type \to \type$ 
+to be expressed syntactically via a type constructor $\later \colon \type \to \type$ 
 (pronounced ``later''). The use of a modality to express guarded recursion

### 2024-07-06 paper-new/concrete-term-model.tex
[paper] changes to concrete term model section

@@
 in a concrete setting. We aim to introduce the key issues whose resolutions in
-the concrete setting inspire the refined definitions of abstract model.
-
+the concrete setting will inform the refined definitions of abstract model.
+%
 In the present section, we use the techniques of SGDT to give a set-theoretic
@@
 % denotational semantics to accommodate the type and term precision orderings. 
-
+%
 %necessitate the abstract components that are found in the definition of the model.
-
+%
 %\subsection{Semantic Constructions}\label{sec:domain-theory}
-
+%
 We begin by discussing the objects of the model into which we will embed the
@@
 computation that runs forever and never returns a value.
-
+%
 Since we claimed that $\li A$ is a monad, we need to define the monadic
@@
 Verifying that the monadic laws hold uses \lob-induction and is straightforward.
-
+%
 % TODO mention that error domains are algebras of the lift monad?
@@
 
-We discuss how to model the dynamic type $\dyn$.
+We now discuss how to model the dynamic type $\dyn$.
 % For simplicity, we omit the
@@
 % function case that is problematic.
-We will define a semantic object $D$ that satisfies the isomorphism
-
+We will define a set $D$ that satisfies the isomorphism
+%
 % TODO make sure we deal with the product correctly
 \[ D \cong \Nat + (D \times D) + (D \to \li D). \]
-
+%
 where the $\li D$ represents the fact that in the case where the value of type
@@
 % TODO refer to technical background?
-Observe that the unfolding equality of the guarded fixpoint implies that $D = D'
-(\later D \to \li D)$, and by the definition of $D'$ as a least fixpoint we see
-that this is equal to $\Nat + (D \times D) + \later (D \to \li D)$.
+Observe that the unfolding equality characterizing the guarded fixpoint implies
+that $D = D' (\later D \to \li D)$, and by the definition of $D'$ as a least
+fixpoint we see that this is equal to $\Nat + (D \times D) + \later (D \to \li
+D)$.
 
@@
 
-We can now give a semantics to the terms of the gradually-typed lambda calculus.
-The definition of the interpretation of types and cast terms is given in Figure
-\ref{fig:term-semantics}.
+We can now give a semantics to the types and terms of the gradually-typed lambda
+calculus. 
 %
 Much of the semantics is similar to a normal call-by-value denotational
-semantics: We interpret types as sets, and terms as functions. Contexts $\Gamma
-= x_1 \colon A_1, \dots, x_n \colon A_n$ are interpreted as the product
-$\sem{A_1} \times \cdots \times \sem{A_n}$. The semantics of the dynamic type
-$\dyn$ is the set $D$ introduced in Section \ref{sec:dynamic-type}. The product
-type $A \times A'$ is interpreted as the product of the denotations of $A$ and
-$A'$. The function type $A \ra A'$ is interpreted as the set of functions from
-$\sem{A}$ to $\li (\sem {A'})$.
-
+semantics, so we focus only on the cast semantics. We interpret types as sets;
+the interpretation of types is given in Figure \ref{fig:type-interpretation}.
+Contexts $\Gamma = x_1 \colon A_1, \dots, x_n \colon A_n$ are interpreted as the
+product $\sem{A_1} \times \cdots \times \sem{A_n}$. The semantics of the dynamic
+type $\dyn$ is the set $D$ introduced in Section \ref{sec:dynamic-type}. The
+product type $A \times A'$ is interpreted as the Cartesian product of the
+denotations of $A$ and $A'$. The function type $A \ra A'$ is interpreted as the
+set of functions from $\sem{A}$ to $\li (\sem {A'})$.
+%
 The interpretation of a value $\hasty {\Gamma} V A$ is a function from
@@
 
-Upcasts are pure, so an upcast $\upc c$ where $c : A \ltdyn A'$ denotes an
-ordinary function from $\sem{A}$ to $\sem{A'}$. Downcasts are effectful, so a
-downcast $\dnc c$ is interpreted as a morphism of error domains between
-$F\sem{A'}$ and $F\sem{A}$, which by definition of the adjunction is
-equivalently an ordinary function between $\sem{A'}$ and $UF\sem{A}$, i.e.,
-$\sem{A'} \to \li \sem{A}$.
+We now discuss the semantics of the cast terms, shown in Figure
+\ref{fig:term-semantics}. Upcasts are pure, so the upcast ${\upc c}$, where $c :
+A \ltdyn A'$, denotes an ordinary function from $\sem{A}$ to $\sem{A'}$.
+Downcasts are effectful, so a downcast ${\dnc c}$ is interpreted as a morphism
+of error domains between $F\sem{A'}$ and $F\sem{A}$. By definition of the
+adjunction this is equivalent to an ordinary function between $\sem{A'}$ and
+$UF\sem{A}$, i.e., $\sem{A'} \to \li \sem{A}$.
 
-The semantics of the casts for a type precision derivation $c$ are defined by
-induction on $c$. The interesting cases are $\inat$, $\itimes$, and $\iarr$.
-Recall that $D$ is isomorphic to $\Nat\, + (D \times D)\, + \laterhs (D \to \li
-D)$ (say that the sum is left-associative).
+Recall the definition of type precision derivations given in Figure
+\ref{fig:typrec}. The semantics of the up- and downcasts for a type precision
+derivation $c$ are defined by induction on $c$. For $r(A)$ the up- and downcasts
+are simply the identity function. For the composition $c \comp c'$ we simply
+compose the cast for $c$ and the cast for $c'$ as functions. For $c_i \ra c_o$
+we apply the casts for $c_i$ and $c_o$ in the domain and codomain respectively.
+More concretely, for the upcast we are given a function $V_f : A_i \to \li A_o$
+and we must define a function $A_i' \to \li A_o'$. The function first downcasts
+its argument according to $c_i$, resulting in an element of $\li A_i$. It then
+applies $\ext{V_f}{}$ to this value to obtain an element of $\li A_o$. Finally, it
+maps the upcast of $c_o$ over the lift monad, obtaining an element of $\li A_o'$
+as needed. For the downcast from $A_i' \to \li A_o'$ to $A_i \to \li A_o$, the
+resulting function first applies the upcast by $c_i$ to its argument, then
+applies the original function, and finally applies the downcast by $c_o$ to the
+result. Likewise, for $c_1 \times c_2$, we apply the cast for $c_1$ on the left
+and the cast for $c_2$ on the right.
+
+The ``base cases" for the casts are  $\inat$, $\itimes$, and $\iarr$. Recall
+that $D$ is isomorphic to $\Nat\, + (D \times D)\, + \laterhs (D \to \li D)$
+(say that the sum is left-associative).
 %
@@
 %
-For $\itimes$, the upcast is $\inr \circ \inl$, and the downcast performs a case
+For $\itimes$, the upcast is $\inl \circ \inr$, and the downcast performs a case
 inspection analogously to the natural number downcast.
@@
 access to the function under the $\later$.
+
 %
@@
 
-% The interpretation of $\lda{x}{M}$ works as follows. Recall by the typing rule
-% for lambda that $\phasty {\Gamma, x : A_i} M {{A_o}}$, so the interpretation of
-% $M$ has type $(\sem{\Gamma} \times \sem{A_i})$ to $\li \sem{A_o}$. The
-% interpretation of lambda is thus a function (in the ambient type theory) that
-% takes a value $a$ representing the argument and applies it (along with $\gamma$)
-% as argument to the interpretation of $M$.
-
 \begin{figure*}
-  \noindent Types:
   \begin{align*}
@@
   \end{align*}
+  \caption{Interpretation of types}
+  \label{fig:type-interpretation}
+\end{figure*}
 
-  % Contexts:
+\begin{figure*}
+  \eric{Check this}
 
   % TODO check these
-  \noindent Casts:
   \begin{align*}
@@
     % upcasts
-    \sem{\upc{\inat{V_n}}} &= \lambda \gamma . \inl\, \inl\, (\sem{V_n}\, \gamma) \\
-    \sem{\upc{\itimes{V_1, V_2}}} &= \lambda \gamma . \inl\, \inr\, (\sem{V_1}\, \gamma, \sem{V_2}\, \gamma) \\
-    \sem{\upc{\iarr{V_f}}} &= \lambda \gamma . \inr\, \nxt (\sem{V_f}\, \gamma) \\[2ex]
+    \sem{\upc{r(A)}} &= \id \\
+    \sem{\upc{c \comp c'}(V)} &= \sem{\upc{c'}} \circ \sem{\upc{c}} \\
+    \sem{\upc{c_i \ra c_o}} &=
+      \lambda V_f . \text{map } \upc{c_o} \circ \ext{V_f}{} \circ \dnc{c_i} \\
+    \sem{\upc{c_1 \times c_2}} &= 
+      \lambda (V_1, V_2) . ((\sem{\upc{c_1} V_1}) , 
+                            (\sem{\upc{c_2} V_2})) \\
+    \sem{\upc{\inat}} &= 
+      \lambda V_n . \inl\, \inl\, V_n \\
+    \sem{\upc{\itimes}} &= 
+      \lambda (V_1, V_2) . \inl\, \inr\, (V_1 , V_2) \\
+    \sem{\upc{\iarr}} &= 
+      \lambda V_f . \inr\, \nxt V_f \\[2ex]
     %
     % downcasts
-    \sem{\dnc{\inat V_d}} &=
-      \lambda \delta . \text{case $(\sem{V_d}\, \gamma)$ of}
+    \sem{\dnc{\inat}} &=
+      \lambda V_d . \text{case $({V_d})$ of}
       \{ \inl\, (\inl\, n) \to \eta\, n
          \alt \text{otherwise} \to \mho \} \\
-    \sem{\dnc{\itimes V_d}} &=
-      \lambda \delta . \text{case $(\sem{V_d}\, \gamma)$ of}
+    \sem{\dnc{\itimes}} &=
+      \lambda V_d . \text{case $({V_d})$ of}
       \{ \inl\, (\inr\, (d_1, d_2)) \to \eta\, (d_1, d_2)
          \alt \text{otherwise} \to \mho \} \\
-    \sem{\dnc{\iarr V_d}} &=
-      \lambda \delta . \text{case $(\sem{V_d}\, \gamma)$ of}
+    \sem{\dnc{\iarr}} &=
+      \lambda V_d . \text{case $({V_d})$ of}
       \{ \inr\, \tilde{f} \to \theta\, (\lambda t. \eta (\tilde{f}_t))
          \alt \text{otherwise} \to \mho \} \\
-    %
-    % TODO product and function
-    %
-    % \sem{\casedyn{V}{n} {M_{nat}} {\tilde{f}}{M_{fun}}} &=
-    % \lambda \delta . \text{case $(\sem{V}\, \delta)$ of} \\ 
-    % &\quad\quad\quad\quad \alt \inl(n) \to \sem{M_{nat}}(n) \\
-    % &\quad\quad\quad\quad \alt \inr(\tilde{f}) \to \sem{M_{fun}}(\tilde{f})
-    %
-    %
     % \sem{\ret\, V}       &= \lambda \gamma . \eta\, \sem{V} \\
@@
 
-  \caption{Term semantics for the gradually-typed lambda calculus.}
+  \caption{Semantics of casts.}
   \label{fig:term-semantics}
 \end{figure*}
-
-\eric{TODO: big-step semantics via clock quantification}
-
-\subsection{A Big-Step Term Semantics}

### 2024-07-06 paper-new/discussion.tex
several fixes to paper
[paper] add outline of remainder of paper to intro
[paper] update discussion of mechanization

@@
-\section{Discussion}
+\section{Discussion}\label{sec:discussion}
 
@@
 % Discuss Guarded Cubical Agda and mechanization efforts
-In parallel with developing the theory discussed in this paper,
-we are mechanizing our results and developing a reusable framework
-for proving graduality in Guarded Cubical Agda \cite{veltri-vezzosi2020}.
+In parallel with developing the theory discussed in this paper, we have
+developed a partial formalization of our results in Guarded Cubical Agda
+\cite{veltri-vezzosi2020}.
 %
-As of this writing, the work is in progress, but we have
-constructed most parts of the concrete model discussed in
-Section \ref{sec:concrete-model}. For instance, we have defined
-types for predomains, error domains, and their morphisms and relations,
-and we used the guarded features to define the guarded lift + error monad
-and the dynamic type.
+We have formalized the key aspects in the definition of the concrete model
+discussed in Section \ref{sec:concrete-model}. For instance, we have defined
+types for predomains, error domains, and their morphisms and relations, and we
+used the guarded features to define the guarded lift + error monad and the
+dynamic type.
 
-We plan to formalize the construction of perturbations and 
-quasi-representable relations, but we have yet to decide
-whether to follow the approach we take in this work and define
-the abstract notion of intensional model and formalize the constructions in that setting,
-and then apply those abstract constructions to the concrete model.
-Alternatively, it may be better from a mechanization standpoint
-to carry out those abstract constructions explicitly in the concrete model,
-i.e., our representation of objects in the concrete model of predomains
-would include a field for the perturbations and our notion of relations
-would include fields for the push-pull property and quasi-representability.
-We leave this investigation to future work.
+Building on these definitions, we formalized the notion of perturbations and
+push-pull structures as well as quasi-representable relations, culminating in
+the definition of value and computation types as predomains (resp. error
+domains) equipped with a monoid of perturbations with an interpretation
+homomorphism into the endomorphisms bisimilar to the identity. We constructed
+the predomain for the dynamic type via mixed induction and guarded recursion,
+defined the three relations $\inat$, $\itimes$, and $\iarr$, and proved that they are
+quasi-representable.
 
-Lastly, we plan to formalize the adequacy result discussed in
-Section \ref{sec:adequacy}. This will involve adding axioms about
-clock quantification as well as about the clock-irrelevance of
-booleans and natural numbers, since as of this writing these
-axioms are not built-in to Guarded Cubical Agda. As an experiment, we have 
-formalized some basic results involving clocks as part of our development.
+% We plan to formalize the construction of perturbations and 
+% quasi-representable relations, but we have yet to decide
+% whether to follow the approach we take in this work and define
+% the abstract notion of intensional model and formalize the constructions in that setting,
+% and then apply those abstract constructions to the concrete model.
+% Alternatively, it may be better from a mechanization standpoint
+% to carry out those abstract constructions explicitly in the concrete model,
+% i.e., our representation of objects in the concrete model of predomains
+% would include a field for the perturbations and our notion of relations
+% would include fields for the push-pull property and quasi-representability.
+% We leave this investigation to future work.
+
+Lastly, we have formalized the adequacy result discussed in Section
+\ref{sec:adequacy}. This required us to add axioms about clock quantification as
+well as axioms asserting the clock-irrelevance of booleans and natural numbers,
+since as of this writing these axioms are not built-in to Guarded Cubical Agda.
+The mechanization of adequacy also required us to formalize some essential
+lemmas involving clocks and clock-irrelevance; we may later refactor these as
+part of a ``standard library'' for Guarded Cubical Agda.
 
@@
 right, and there is no analogue of this in our approach. However,
-using two expression relations allows some but not all transitivite
+using two expression relations allows some but not all transitive
 reasoning of term precision to be recovered. In the future we aim to

### 2024-07-07 paper-new/categorical-model.tex
[paper] move and edit prose on weakening the double categorical semantics

@@
 
-Next, we develop our appropriate weakened notions of models of
-graduality, which we divide into \emph{extensional} models, where
-ordering is up to weak bisimilarity, and \emph{intensional} models,
-where ordering relates terms that considers computational steps to be
-observable. We develop the notion of intensional model of gradual
-typing in stages and show how to develop one from a base model of
-effectful functions and relations.
+\eric{This section may need to be revised in light of the previous sections on the concrete model.}
+
+\subsection{Weakening the Double Category Semantics}
+
+While the double categorical semantics introduced in Section \ref{sec:cbpv} can
+be satisfied with classical domain theoretic models, we have seen in the
+previous section that there are obstructions to developing a concrete model in
+\emph{guarded} type theory. While guarded type theory makes the construction of
+arbitrary guarded recursive definitions possible, it comes at a significant cost
+to reasoning: unfolding of recursive definitions is explicit. In a denotational
+semantics, this means that non-well-founded recursion must perform a kind of
+\emph{observable step}. This presents a difficulty in proving graduality, which
+is a property that is oblivious to the number of steps that a program takes.
+
+Therefore, to prove graduality, we must be able to relate terms that take
+different numbers of steps. In the previous section we have examined two ways of
+doing this: in the first approach by defining an ordering $\semltbad$ that is
+closed under delays, and in the second by decomposing this ordering into a
+lock-step error ordering and a weak bisimilarity relation. But either way, we
+cannot escape the confines of Theorem \ref{thm:no-go}, which tells us that
+neither of these fomulations is transitive.
+
+With this fundamental constraint in mind, we develop an abstract notion of
+\emph{extensional} model where intuitively, ordering is up to weak bisimilarity.
+Transitivity of term precision does not hold here, so the definition of
+extensional model weakens the idealized notion of model from double categories
+to reflexive graph categories. In particular, the definition drops the operation
+of horizontal pasting of squares, but it maintains a form of representability of
+relations crucial for graduality. An extensional model is sufficient for
+interpreting the terms and inequational theory of gradual typing. Unfortunately,
+as was mentioned in the previous section, constructing such a model directly is
+difficult: since we lack transitvity, we must take as primitive more complicated
+versions of the cast rules and verify them ``from scratch'' in the model.
+
+However, the story does not end here. The approach of decomposing the ordering
+into the lock-step ordering and weak bisimilarity has the advantage that we can
+employ transitive reasoning when working with the lock-step ordering. The caveat
+is that we can no longer model casts as representable morphisms when reasoning
+with observable computation steps.
+%
+To axiomatize this situation abstractly, we introduce an intermediate notion of
+\emph{intensional} model, where intuitively, the ordering considers computational
+steps to be observable. Intensional models are based on double categories, but
+representability must be weakened to ``quasi-representability'', a kind of
+representability \emph{up to} observable steps. To reason up to observable steps
+without using weak bisimilarity, we develop a more refined notion we call
+\emph{perturbations}, certain terms that are bisimilar to the identity but can
+be manipulated explicitly in constructions. We then show that the compositional
+construction of casts from domain theoretic models can be adapted to this
+intensional setting by incorporating some explicit manipulation of
+perturbations. Finally, we show that taking an ``extensional collapse'' by weak
+bisimilarity provides a model of the extensional theory, which can then be used
+to model the global graduality property.
+
+% However, here we arrive at the fundamental issue with double categorical
+% semantics using representable relations:
+
+% \begin{enumerate}
+% \item When reasoning up to weak bisimilarity, transitive reasoning is not
+%   possible, and so horizontal pasting of squares is not a valid reasoning
+%   principle.
+% \item When reasoning with observable computation steps, certain casts take
+%   observable steps, and can no longer be modeled using representable morphisms.
+% \end{enumerate}
+
+% Our solution to this dilemma comes in two parts. Since graduality ignores
+% computational steps, the syntactic theory of graduality must be modeled up to
+% weak bisimilarity, where transitive reasoning is not valid. For this purpose we
+% develop a notion of \emph{extensional} model which weakens from double
+% categories to reflexive graph categories, dropping the operation of horizontal
+% pasting of squares, but still maintaining a form of representability of
+% relations.
+
+\begin{comment}
+In the previous section, we saw what goes wrong when one tries to define a
+partial ordering relation on the guarded lift monad that is closed under delays.
+In particular, such a relation is not transitive. The solution we adopted was to
+give up on closure under delay, taking the partial ordering on the guarded lift
+monad to be the lock-step error ordering, 
+%to give up on closure under delay and define two separate relations, a lock-step error ordering and a weak bisimilarity relation. 
+The denotation of term precision for closed terms was then defined to be the
+closure of the lock-step ordering under weak bisimilarity. With the lock-step
+error ordering as our poset structure, we encountered complications when
+attempting to model the cast rules of gradual typing.
+% Furthermore, we saw that defining the partial ordering on the guarded lift
+% monad as the lock-step error ordering introduces complications for modeling
+% the cast rules of gradual typing.
+The basic issue was that casts may involve ``stepping'' (i.e., a cast may
+introduce a $\theta$ constructor), and in such cases we need to insert delays
+artificially in order to keep both sides of the desired inequality in lock-step.
+\end{comment}
+
+In the remainder of this section, we define the notions of extensional and
+intensional models. We develop the notion of intensional model in stages and show how
+to develop one from a base model of effectful functions and relations.
+
+% In the remainder of this section, we develop our appropriate weakened notions of
+% models of graduality, which we divide into \emph{extensional} models, where
+% ordering is up to weak bisimilarity, and \emph{intensional} models, where the
+% ordering considers computational steps to be observable. 
+% %Our ultimate notion of abstract model of gradual typing will be the extensional model.
+% We develop the notion of intensional model of gradual typing in stages and show
+% how to develop one from a base model of effectful functions and relations.
 
@@
 
-Since we lack transitivity of ordering when reasoning in guarded type
+Since we lack transitivity of ordering when reasoning extensionally in guarded type
 theory, our weakened notion of extensional model is based on reflexive
@@
 squares is not sufficient to interpret the cast rules of gradual
-typing. Instead we generalize the notion of representability to match
+typing. Instead, we generalize the notion of representability to match
 the syntactic rules in Section~\ref{sec:GTLC}.
@@
 the morphism $\delta_{UFA}^* \in \vf(UFA, UFA)$, and then by the adjunction
-we get a computation morphism $\delta_{FA}^ \in \ef(FA, FA)$. Moreover, we get a square
+we get a computation morphism $\delta_{FA}^* \in \ef(FA, FA)$. Moreover, we get a square
 $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$ for all $c : A \rel A$.
@@
 which is formulated as follows. Let $c : A \rel A'$.
-Given a perturbation $\delta \in P_A$, there is a corresponding perturbation
+Given a perturbation $\delta \in P_A$, we require a corresponding perturbation
 $\push_c(\delta) \in P_{A'}$. % making the following square commute:
 %
-Likewise, given $\delta' \in P_{A'}$ there is a perturbation $\pull_c(\delta') \in P_A$.
+Likewise, given $\delta' \in P_{A'}$ we require a perturbation $\pull_c(\delta') \in P_A$.
 % making the following square commute:

### 2024-07-07 paper-new/extensional-model.tex
[paper] move and edit prose on weakening the double categorical semantics

@@
 \end{example}
-
-\subsection{Weakening the Double Category Semantics}
-
-While the double categorical semantics can be satisfied with classical
-domain theoretic models, there are obstructions to developing a
-semantics based on \emph{guarded} recursion. While guarded type theory
-makes construction of arbitrary guarded recursive definitions
-possible, it comes at a significant cost to reasoning: unfolding of
-recursive definitions is explicit. In a denotational semantics, this
-means that non-well-founded recursion must perform a kind of
-\emph{observable step}. This is a difficulty in proving graduality,
-which is a property that is oblivious to the number of steps that a
-program takes. Therefore to prove graduality we must impose a kind of
-weak bisimilarity relation on our programs to reason about
-step-independent relational properties. However, here we arrive at the
-fundamental issue with double categorical semantics using
-representable relations:
-
-\begin{enumerate}
-\item When reasoning up to weak bisimilarity, transitive reasoning is
-  not possible, and so horizontal pasting of squares is not a valid
-  reasoning principle.
-\item When reasoning with observable computation steps, certain casts
-  take observable steps, and can no longer be modeled using
-  representable morphisms.
-\end{enumerate}
-
-Our solution to this dilemma comes in two parts. Since graduality
-ignores computational steps, the syntactic theory of graduality must
-be modeled up to weak bisimilarity, where transitive reasoning is not
-valid. For this purpose we develop a notion of \emph{extensional}
-model which weakens from double categories to a reflexive graph
-categories, dropping the operation of horizontal pasting of squares,
-but still maintaining a form of representability of
-relations.
-
-However, transitive reasoning is essential for compositional semantic
-constructions, and so we need to work with an intermediate notion of
-\emph{intensional} model which is based on double categories, but
-where representability has to be weakened to
-``quasi-representability'', a kind of representability \emph{up to}
-observable steps. To reason up to observable steps without using weak
-bisimilarity, we develop a notion we call \emph{perturbations},
-certain terms that are bisimilar to the identity but can be
-manipulated explicitly in constructions. We then show that the
-compositional construction of casts from domain theoretic models can
-be adapted to this guarded setting by incorporating some explicit
-manipulation of perturbations. Finally, we show that taking an
-``extensional collapse'' by bisimilarity provides a model of the
-extensional theory, which can then be used to model the global
-graduality property.

### 2024-07-07 paper-new/concrete-model.tex
[paper] edit section on construction of concrete model to remove definitions that were moved to earlier section

@@
 
-In this section, we build a concrete extensional model of gradual typing.
-We begin by defining a step-1 intensional model, and then
-apply the abstract constructions outlined in the previous section
-to obtain an extensional model.
-We begin with some definitions:
+In this section, we return to the construction of a concrete model of gradual
+typing begun in Section \ref{sec:towards-relational-model}. We begin by constructing
+a step-1 intensional model, and then apply the abstract constructions outlined
+in the previous section to obtain an extensional model. We begin with some
+definitions:
 \begin{definition}
@@
 % TODO: later of a predomain
-Given a predomain $A$, we can form the predomain $\later A$.
-The underlying set is $\later |A|$ and the relation is defined in the obvious way,
-i.e., $\tilde{x} \le_{\later A} \tilde{x'}$ iff $\later_t(\tilde{x}_t \le_A \tilde{x'}_t)$.
+Given a predomain $A$, we can form the predomain $\later A$. The underlying set
+is $\later |A|$ and the relation is defined in the obvious way, i.e., $\tilde{x}
+\le_{\later A} \tilde{x'}$ iff $\later_t(\tilde{x}_t \le_A \tilde{x'}_t)$.
 Likewise for bisimilarity.
 %
-We also give a predomain structure to the natural numbers $\mathbb{N}$, where both the
-ordering and the bisimilarity relation are equality.
+We also give a predomain structure to the natural numbers $\mathbb{N}$, where
+both the ordering and the bisimilarity relation are equality.
 %
-Morphisms of predomains are functions between the underlying sets that preserve the ordering
-and the bisimilarity relation. More formally:
+Morphisms of predomains are functions between the underlying sets that preserve
+the ordering and the bisimilarity relation. More formally:
 %
@@
 %
-For an error domain $B$, we define the predomain morphism $\delta_B := \theta_B \circ \nxt$.
+For an error domain $B$, we define the predomain morphism $\delta_B := \theta_B
+\circ \nxt$.
 %
-Morphisms of error domains are morphisms of the underlying predomains that preserve the
-algebraic structure. More formally:
+Morphisms of error domains are morphisms of the underlying predomains that
+preserve the algebraic structure. More formally:
 %
@@
 %
-We define a (monotone) relation on predomains $A$ and $A'$ to be a relation on the
-underlying sets that is downward-closed under $\le_A$ and upward-closed under $\le_{A'}$.
-More formally:
+We define a (monotone) relation on predomains $A$ and $A'$ to be a relation on
+the underlying sets that is downward-closed under $\le_A$ and upward-closed
+under $\le_{A'}$. More formally:
 %
@@
 %
-Composition of relations on predomains is the usual relational composition.
-Similarly, we define a (monotone) relation on error domains to be a relation on the
-underlying predomains that respects error and preserves $\theta$.
+Composition of relations on predomains is the usual relational composition
+(which is truncated to be propositional). Similarly, we define a (monotone)
+relation on error domains to be a relation on the underlying predomains that
+respects error and preserves $\theta$:
 %
@@
 %
-We define composition of error domain relations $R$ on $B_1$ and $B_2$ and $S$
-on $B_2$ and $B_3$ to be the least relation containing $R$ and $S$ that respects
-error and preserves $\theta$.
+Because of a technical limitation involving the interaction between
+propositional truncation and the later modality, we cannot define the
+composition of error domain relations as simply the composition of their
+underlying predomain relations. Instead, we define composition of error domain
+relations $R$ on $B_1$ and $B_2$ and $S$ on $B_2$ and $B_3$ to be the least
+relation containing $R$ and $S$ that is downward closed under $\le_{B_1}$,
+upward-closed under $\le_{B_3}$, respects error, and preserves $\theta$.
 Specifically, it is defined inductively by the following rules:
@@
     {b_1 \mathbin{R \relcomp S} b_3}
+
+    \inferrule*[right = DnClosed]
+    {b_1' \le_{B_1} b_1 \and b_1 \mathbin{R \relcomp S} b_3}
+    {b_1' \mathbin{R \relcomp S} b_3}
+
+    \inferrule*[right = UpClosed]
+    {b_1 \mathbin{R \relcomp S} b_3 \and b_3 \le_{B_3} b_3'}
+    {b_1 \mathbin{R \relcomp S} b_3'}
 
@@
 %
-We now describe the ``commuting squares".
-Suppose we are given predomains $A_i, A_o, A_i'$, and $A_o'$,
-relations $R_i : A_i \rel A_i'$ and $R_o : A_o \rel A_o'$, and morphisms $f : A_i \to A_o, f' : A_i' \to A_o'$
-Given a square with these morphisms and relations, we say that the square commutes, written $f \le f'$, if for all
-$x \in A_i$ and $x' \in A_i'$ with $x \mathbin{R_i} x'$, we have
-$f(x) \mathbin{R_o} f'(x')$.
-We make the analogous definition for error domains.
+We now describe the squares. Suppose we are given predomains $A_i,
+A_o, A_i'$, and $A_o'$, relations $R_i : A_i \rel A_i'$ and $R_o : A_o \rel
+A_o'$, and morphisms $f : A_i \to A_o, f' : A_i' \to A_o'$ Given a square with
+these morphisms and relations, we say that the square commutes, written $f \le
+f'$, if for all $x \in A_i$ and $x' \in A_i'$ with $x \mathbin{R_i} x'$, we have
+$f(x) \mathbin{R_o} f'(x')$. We make the analogous definition for error domains.
 
-\subsection{Guarded Lift Monad}\label{sec:guarded-lift-monad}
+\subsection{Guarded Lift Monad and Free-Forgetful Adjunction}\label{sec:guarded-lift-monad}
 
 % Lift monad
-The guarded error-lift monad $\li$ takes a predomain $A$ to the error domain $\li A$.
-It is defined \footnote{Formally, the lift monad $\li A$ is defined as the solution to
-the guarded recursive type equation $\li A \cong A + 1 + \later \li A$.} as follows :
-%
-\begin{align*}
-  \li A :=
-  \alt \eta \colon A \to \li A
-  \alt \mho \colon \li A
-  \alt \theta \colon \later (\li A) \to \li A
-\end{align*}
-%
+The guarded error-lift monad $\li$ takes a predomain $A$ to the error domain
+$\li A$. A similar construct was already introduced in Section
+$\ref{sec:lift-monad}$; here we extend it to a monad on \emph{predomains} rather
+than on sets. Given a predomain $A$, the underlying set of $\li A$ is simply the
+guarded lift monad on sets applied to the underlying set of $A$. Recall that
+this has three constructors $\eta : A \to \li A$, $\mho : \li A$, and $\theta :
+\laterhs (\li A) \to \li A$.
 
-This captures the intuition that a program may either return a value,
-fail at run-time, or take one or more observable steps of computation.
-Previous work has studied such a similar construct, called the guarded lift
-monad \cite{mogelberg-paviotti2016}; our version here augments it with the notion of error.
-Since we claimed that $\li A$ is a monad, we need to define the monadic operations
-and show that they respect the monadic laws. The return is just $\eta$, and the monadic extend
-is defined via guarded recursion by cases on the input.
-Verifying that the monadic laws hold uses \lob-induction and is straightforward.
+The ordering relation $\le_{\li A}$ is defined to be the \emph{lock-step error
+ordering} defined in Section \ref{sec:lock-step-and-weak-bisim}. The
+bisimilarity relation $\bisim_{\li A}$ is defined to be the \emph{weak
+bisimilarity relation} also defined in Section
+\ref{sec:lock-step-and-weak-bisim}. With these relations on $\li
+A$, the constructors $\eta$ and $\theta$ are in fact morphisms of predomains,
+i.e., they are monotone and preserve bisimilarity.
 
-% \eric{Check}
-% We can also show that $\li A$ is the free error- and -later algebra on $A$, in that
-% for any morphism of predomains $f : A \to UB$, there is a unique morphism of error domains
-% $f^* : \li A \to \li A$ extending $f$.
+The monad $\li$ decomposes into a free-forgetful adjunction $F \dashv U$. Given
+a predomain $A$, we define the error domain $FA$ whose underlying predomain is
+$\li A$, whose error element is $\mho$, and whose $\theta$ map is the
+constructor $\theta$. The functor $U$ from error domains to predomains simply
+returns the underlying predomain. On morphisms, it returns the underlying
+morphism of predomains.
 
-There is a functor $U$ from error domains to predomains that on objects simply returns the
-underlying predomain, and on morphisms returns the underlying morphism of predomains.
-It is easily verified that $\li A$ is the free error- and later-algebra on the predomain $A$,
-so we have that $\li$ is left-adjoint to $U$.
+The monadic bind operation takes a predomain morphism $f : A \to UB$ to an error
+domain morphism $FA \to B$; with this we define the action of $F$ on morphisms:
+Given $f : A \to A'$ we observe that $\eta \circ f : A \to UFA'$ and so we define
+$Ff = \ext{\eta \circ f}{} : FA \to FA'$.
+
+It is easily verified that $\li A$ is the free error- and later-algebra on the
+predomain $A$, so we have that $\li$ is left-adjoint to $U$.
+
 % i.e., error domain morphisms from $\li A$ to $B$ are in one-to-one correspondence with
@@
 
-The partial order $\le_{\li A}$ is the lock-step error ordering defined by guarded recursion as follows:
-%
-\begin{itemize}
-    \item 	$\eta\, x \le_{\li A} \eta\, y$ if $x \le_A y$.
-    \item 	$\mho \le_{\li A} l$ for all $l$ 
-    \item   $\theta\, \tilde{r} \le_{\li A} \theta\, \tilde{r'}$ if
-            $\later_t (\tilde{r}_t \le_{\li A} \tilde{r'}_t)$
-  \end{itemize}
-%
-The idea is that two computations $l$ and $l'$ are related if they are in
-lock-step with regard to their intensional behavior, up to $l$ erroring.
-Given a relation $R : A \rel A'$, we define in an analogous manner a heterogeneous
-version of the lock-step error ordering between $\li R : \li A \rel \li A'$.
 % We define the action of $\li$ on a relation $R$ between $A$ and $A'$ to be the
@@
 
-%\subsubsection{Weak Bisimilarity}\label{sec:weak-bisimilarity}
-%
-For a predomain $A$, we define a relation on $\li A$, called ``weak bisimilarity",
-written $l \bisim l'$. Intuitively, we say $l \bisim l'$ if they are equivalent ``up to delay''.
-The weak bisimilarity relation is defined by guarded recursion as follows:
-%
-\begin{align*}
-  &\mho \bisim \mho \\
-%
-  &\eta\, x \bisim \eta\, y \text{ if } 
-    x \bisim_A y \\
-%		
-  &\theta\, \tilde{x} \bisim \theta\, \tilde{y} \text{ if } 
-    \later_t (\tilde{x}_t \bisim \tilde{y}_t) \\
-%	
-  &\theta\, \tilde{x} \bisim \mho \text{ if } 
-    \theta\, \tilde{x} = \delta^n(\mho) \text { for some $n$ } \\
-%	
-  &\theta\, \tilde{x} \bisim \eta\, y \text{ if }
-    (\theta\, \tilde{x} = \delta^n(\eta\, x))
-  \text { for some $n$ and $x \in A$ such that $x \bisim_A y$ } \\
-%
-  &\mho \bisim \theta\, \tilde{y} \text { if } 
-    \theta\, \tilde{y} = \delta^n(\mho) \text { for some $n$ } \\
-%	
-  &\eta\, x \bisim \theta\, \tilde{y} \text { if }
-    (\theta\, \tilde{y} = \delta^n (\eta\, y))
-  \text { for some $n$ and $y \in A$ such that $x \bisim_A y$ }
-\end{align*}
-%
-When both sides are $\eta$, then we ensure that the underlying values are related
-by the bisimilarity relation on $A$.
-When one side is a $\theta$ and the other is $\eta x$ (i.e., one side steps),
-we stipulate that the $\theta$-term runs to $\eta y$ where $x$ is bisimilar to $y$.
-Similarly when one side is $\theta$ and the other $\mho$.
-If both sides step, then we allow one time step to pass and compare the resulting terms.
-In this way, the definition captures the intuition of terms being equivalent up to
-delays.
-
-% It can be shown (by \lob-induction) that the step-sensitive relation is symmetric.
-% However, it can also be shown that this relation is \emph{not} transitive:
-% The argument is the same as that used to show that the step-insensitive error
-% ordering $\semltbad$ described above is not transitive. Namely, we show that
-% if it were transitive, then it would have to be trivial in that $l \bisim l'$ for all $l, l'$.
-% that if this relation were transitive, then it would relate all values of type $\li A$.
-
+\subsection{Other Constructions}
 % internal hom for predomains and error domains
@@
     &\alt \ttimes(d_1, d_2) \mapsto {\ttimes(\ptb_{D \times D}(p_\text{times})(d_1, d_2))} \\
-    &\alt \tfun(\tilde{f}) \mapsto {\tfun(\lambda t. \ptb_{U(D \to FD)}()(\tilde{f}_t))} \\
+    &\alt \tfun(\tilde{f}) \mapsto {\tfun(\lambda t. \ptb_{U(D \to FD)}()(\tilde{f}_t))}
 \end{align*}
@@
 As an illustrative case, we establish the push-pull property for the relation $\iarr$.
-We define $\pull_{\iarr} : P_D \to P_{U(D \arr FD)}$ by\( \pull_{\iarr}(p_{\text{times}}, p_{\text{fun}}) = p_{\text{fun}}, \)
+We define $\pull_{\iarr} : P_D \to P_{U(D \arr FD)}$ by \( \pull_{\iarr}(p_{\text{times}}, p_{\text{fun}}) = p_{\text{fun}}, \)
 % (recall that $P_{U(D \arr FD)} = \mathbb{N} \times P_D^{op} \times \mathbb{N} \times P_D$).
@@
 %
-We define $\push_{\iarr} : P_{U(D \arr FD)} \to P_D$ by \( \push{\iarr}(p_{\text{fun}}) = (\id, p_{\text{fun}}) \).
+We define $\push_{\iarr} : P_{U(D \arr FD)} \to P_D$ by \( \push_{\iarr}(p_{\text{fun}}) = (\id, p_{\text{fun}}) \).
 %

### 2024-07-07 paper-new/concrete-relational-model.tex
[paper] slight wording change in section on concrete relational model

@@
 
+% In the previous section, we gave a denotational semantics to the terms of the
+% gradually-typed lambda calculus using the tools of SGDT. We defined a big-step
+% semantics for closed terms of type $\nat$, as functions from $\mathbb{N} \to
+% ((\mathbb{N} + 1) + 1)$.
+
 \eric{Wording?}
-In this section, we build on the ideas used in defining a term model with the
-goal of defining a model of the full gradually-typed lambda calculus, i.e., one
-that has constructs that model type and term precision.
+In the previous section, we gave a denotational semantics to the terms of the
+gradually-typed lambda calculus using the tools of SGDT. In this section, we
+build on the ideas used in defining a term model with the goal of defining a
+model of the full gradually-typed lambda calculus, i.e., one that has constructs
+that model type and term precision.
 %
@@
 
-\subsection{Solution: Splitting into Two Relations}
+\subsection{Solution: Splitting into Two Relations}\label{sec:lock-step-and-weak-bisim}
 

### 2024-07-07 paper-new/concrete-term-model.tex
[paper] move technical background on SGDT; edit section on concrete term model

@@
 \eric{Wording?}
-The principal goal of the next two sections is to ground the refined notions of
-abstract model of gradual typing described in Section \ref{sec:abstract-models}
-in a concrete setting. We aim to introduce the key issues whose resolutions in
-the concrete setting will inform the refined definitions of abstract model.
 %
-In the present section, we use the techniques of SGDT to give a set-theoretic
-denotational semantics to the terms of the gradually-typed lambda calculus
-introduced in Section \ref{sec:GTLC}. This serves two important purposes: First,
-it acts as a warm-up for employing the tools of SGDT to define a denotational
-semantics. Second, constructing this semantic model establishes the validity of
-the $\beta$ and $\eta$ principles for the gradually-typed lambda calculus. It
-also acts as a stepping stone into the next section, where we extend the
-techniques employed here to accommodate the type and term precision orderings of
-the gradually-typed lambda calculus.
+In the present section, we use the techniques of guarded type theory to give a
+set-theoretic denotational semantics to the terms of the gradually-typed lambda
+calculus introduced in Section \ref{sec:GTLC}. This serves two important
+purposes: First, it acts as a warm-up for employing guarded type theory to
+define a denotational semantics. This acts as a stepping stone into the next
+section, where we extend the techniques employed here to accommodate the type
+and term precision orderings of the gradually-typed lambda calculus. Second,
+constructing this semantic model establishes the validity of the $\beta$ and
+$\eta$ principles for the gradually-typed lambda calculus. 
 %
@@
 %necessitate the abstract components that are found in the definition of the model.
+
+\subsection{Guarded Type Theory}
+
+We begin with a brief overview of synthetic guarded domain theory, which we will
+use to construct the model. \emph{Synthetic guarded domain theory}, or SGDT for
+short, is an axiomatic framework in which we can reason about non-well-founded
+recursive constructions while abstracting away the specific details of
+step-indexing that we would need to track if we were working analytically. This
+allows us to avoid the tedious reasoning associated with traditional
+step-indexing techniques. We provide a brief overview here; more details can be
+found in \cite{birkedal-mogelberg-schwinghammer-stovring2011}.
+
+SGDT offers a synthetic approach to domain theory that allows for guarded
+recursion to be expressed syntactically via a type constructor $\later \colon
+\type \to \type$ (pronounced ``later''). The use of a modality to express
+guarded recursion was introduced by Nakano \cite{Nakano2000}.
 %
-%\subsection{Semantic Constructions}\label{sec:domain-theory}
+Given a type $A$, the type $\later A$ represents an element of type $A$ that is
+available one time step later. There is an operator $\nxt : A \to\, \later A$
+that ``delays'' an element available now to make it available later. We will use
+a tilde to denote a term of type $\later A$, e.g., $\tilde{M}$.
+
+% TODO later is an applicative functor, but not a monad
+
+There is a \emph{guarded fixpoint} operator
 %
-We begin by discussing the objects of the model into which we will embed the
-types and terms of the gradually-typed lambda calculus. Recall that we are
-working informally in a guarded type theory with clocks and ticks (see Section
-\ref{sec:technical-background}) for details. The constructions in this section
-will make significant use of the primitives of guarded type theory.
+\[
+  \fix : \forall T, (\later T \to T) \to T.
+\]
+%
+That is, to construct a term of type $T$, it suffices to assume that we have
+access to such a term ``later'' and use that to help us build a term ``now''.
+This operator satisfies the axiom that $\fix f = f (\nxt (\fix f))$. In
+particular, this axiom applies to propositions $P : \texttt{Prop}$; proving a
+statement in this manner is known as $\lob$-induction.
 
-\subsection{Value and Computation Types}
+% Clocked Cubical Type Theory
+\subsubsection{Ticked Cubical Type Theory}
+% TODO motivation for Clocked Cubical Type Theory, e.g., delayed substitutions?
+
+Ticked Cubical Type Theory \cite{mogelberg-veltri2019} is an extension of
+Cubical Type Theory \cite{CohenCoquandHuberMortberg2017} that has an additional
+sort called \emph{ticks}. Ticks were originally introduced in
+\cite{bahr-grathwohl-bugge-mogelberg2017}. A tick $t : \tick$ serves as evidence
+that one unit of time has passed. In Ticked Cubical Type Theory, the type
+$\later A$ is the type of dependent functions from ticks to $A$. The type $A$ is
+allowed to depend on $t$, in which case we write $\later_t A$ to emphasize the
+dependence.
+
+% TODO next as a function that ignores its input tick argument?
+
+% TODO include a figure with some of the rules for ticks
+
+The rules for tick abstraction and application are similar to those of ordinary
+$\Pi$ types. A context now consists of ordinary variables $x : A$ as well as
+tick variables $t : \tick$. The presence of the tick variable $t$ in context
+$\Gamma, (t : \tick), \Gamma'$ intuitively means that the values of the
+variables in $\Gamma$ arrive ``first'', then one time step occurs, and then the
+values of the variables in $\Gamma'$ arrive.
+%
+The abstraction rule for ticks states that if in context $\Gamma, t : \tick$
+the term $M$ has type $A$, then in context $\Gamma$ the term $\lambda t.M$ has
+type $\later_t A$.
+%
+Conversely, if we have a term $M$ of type $\later A$, and we have available in
+the context a tick $t' : \tick$, then we can apply the tick to $M$ to get a
+term $M[t'] : A[t'/t]$. However, there is an important restriction on when we
+are allowed to apply ticks. In order to apply $M$ to tick $t$, $M$ must be
+well-typed in the prefix of the context occurring before the tick $t$. That is,
+all variables mentioned in $M$ must be available before $t$. This ensures that
+we cannot, for example, define a term of type $\later \laterhs A \to\, \laterhs
+A$ via repeated tick application.
+% TODO restriction on tick application
+%
+For the sake of brevity, we will also write tick application as $M_t$.
+
+\subsection{A Call-by-push-value Model}
 
@@
 
-% TODO should we introduce the lift monad first, and then introduce error
-% domains as algebras of the lift monad?
+For the sake of modularity and ease of description, we will formulate the model
+for the gradually-typed lambda calculus using the \emph{call-by-push-value}
+paradigm \cite{levy99}. In a call-by-push-value model, there is a
+category $\calV$ of \emph{value types} and a category $\calE$ of
+\emph{computation types}. Value types represent static values, while computation
+types represent effectful terms. We denote value types by $A, A', A_1, A_i,
+A_o$, etc, and computation types by $B, B', B_1, B_i, B_o$, etc. There is a pair
+of adjoint functors $F : \calV \to \calE$ and $U : \calE \to \calV$ with $F
+\dashv U$, and a functor $\arr : \calV^{op} \times \calE \to \calE$ such that
+$U(A \arr B) \cong A \To UB$ where $\To$ denotes the internal hom in $\calV$.
 
-We will give an embedding of our call-by-value syntax into a call-by-push-value
-model based on a free-forgetful adjunction $F \dashv U$. Value types will denote
-sets. Computation types will denote sets with additional algebraic structure
-corresponding to the possible behaviors of a term. Recall that we are modelling
-gradually-typed programs. Thus, one such behavior is \emph{failure}: a program
-may fail at run-time because of a type error. In addition to this, a program may
-fail to terminate. This is possible because the dynamic type includes function
-types and so is recursive. We will discuss the dynamic type further in a
-subsequent section, but for now it suffices to know that we can express a
-diverging computation in our syntax. Therefore, it will be necessary to track
-the ``steps'' taken by a term; we can regard this as a way of intensionally
-modelling \emph{partiality}.
+We embed our call-by-value gradually-typed cast calculus into a
+call-by-push-value model as follows. The value types will be sets, and
+computation types will be sets with additional algebraic structure corresponding
+to the possible behaviors of a gradually-typed program. One such behavior is
+\emph{failure}: a program may fail at run-time because of a type error. In
+addition to this, a program may fail to terminate. This is possible because the
+dynamic type includes function types and so is recursive. We will discuss the
+dynamic type further in a subsequent section, but for now it suffices to know
+that we can express a diverging computation in our syntax. Therefore, it will be
+necessary to track the ``steps'' taken by a term; we can regard this as a way of
+intensionally modelling \emph{partiality}.
 
-The semantic objects modelling computations will be sets $B$ that are equipped
-with a distinguished error element $\mho_B$ and a map $\theta_B \colon \laterhs
-B \to B$. We call such sets \emph{error domains}. A \emph{morphism of error
-domains} from $B_1$ to $B_2$ is a function $\phi$ on the underlying sets that
-respects the error and $\theta$ maps, i.e., for which 
+More concretely, the computation types are defined to be sets $B$ that are
+equipped with a distinguished error element $\mho_B$ and a map $\theta_B \colon
+\laterhs B \to B$. We call such sets \emph{error domains}. A \emph{morphism of
+error domains} from $B_1$ to $B_2$ is a function $\phi$ on the underlying sets
+that respects the error and $\theta$ maps, i.e., for which 
 %
@@
 
+The functor $U : \errdom \to \Set$ simply takes the underlying set of the error
+domain. The functor $F : \Set \to \errdom$ is defined to be the left adjoint to
+$U$ and constructs the \emph{free error domain} on a set $A$.
+%
+The functor $\arr : \Set^{op} \times \errdom \to \errdom$ is defined on objects
+as follows. Given a set $A$ and error domain $B$, $A \arr B$ is the error domain
+whose underlying set is the set of functions $A \to UB$. The error element is
+the constant error function, i.e., $\lambda x. \mho_B$. The map $\theta :
+(\laterhs (A \to UB)) \to (A \to UB)$ is defined by 
+$\theta(\tilde{f}) = \lambda x . \theta_B (\tilde{f}_t)$.
+%
+Given a morphism $f : A_o \to A_i$ and $\phi : B_i \to B_o$, the morphism $f
+\arr \phi : (A_i \arr B_i) \to (A_o \arr B_o)$ is given by pre-composing by $f$
+and post-composing by $\phi$.
+
 \subsection{The Lift + Error Monad}\label{sec:lift-monad}
 
-% When thinking about how to model gradually-typed programs semantically, we
-% should consider their possible behaviors. On the one hand, we have
-% \emph{failure}: a program may fail at run-time because of a type error. In
-% addition to this, a program may fail to terminate. This is possible because the
-% dynamic type includes function types and so is recursive. We will discuss the
-% dynamic type further in the next section, but for now it suffices to know that
-% we can express a diverging computation in our syntax. Therefore, it will be
-% useful to track the ``steps'' taken by a term; we can regard this as a way of
-% intensionally modelling \emph{partiality}.
-
-% With this in mind, we can describe a semantic object that models these
-% behaviors: a monad for embedding computations that has cases for failure and
-% ``stepping''. Previous work has studied such a construct in the setting of the
-% latter, called the lift monad \cite{mogelberg-paviotti2016}; here, we augment it
-% with the additional effect of failure.
-
-We now describe the monad that decomposes into the free-forgetful adjunction
+We describe a monad that decomposes into the free-forgetful adjunction
 between sets and error domains. We base the construction on the \emph{guarded
@@
 operations and show that they respect the monadic laws. The return is just
-$\eta$, and the monadic extension is defined via guarded recursion by cases on
-the input.
+$\eta$, and given $f : A \to \li A'$, the monadic extension $\text{ext}(f)
+\colon (\li A \to \li A')$ is defined via guarded recursion by cases on the
+input value of type $\li A$.
 % It is instructive to give at least one example of a use of guarded recursion, so
@@
 % TODO mention that error domains are algebras of the lift monad?
-We observe that the guarded lift monad constructs the free error domain on a set
-$A$, i.e., we have $\li = UF$ where $F : \Set \to \errdom$ and $U : \errdom \to
-\Set$ and $F \dashv U$. Given a set $A$ and an error domain $B$, the set of
-functions $A \to UB$ is naturally isomorphic to the set of error domain
-morphisms $FA \to B$.
+We observe that the guarded lift monad applied to a set $A$ is the underlying
+set of the free error domain on $A$ $A$, i.e., we have $\li = UF$ where $F :
+\Set \to \errdom$ and $U : \errdom \to \Set$ and $F \dashv U$. Given a set $A$
+and an error domain $B$, the set of functions $A \to UB$ is naturally isomorphic
+to the set of error domain morphisms $FA \to B$.
 
@@
 We now discuss how to model the dynamic type $\dyn$.
-% For simplicity, we omit the
-% case of products in the dynamic type, as from a semantic viewpoint it is the
-% function case that is problematic.
-We will define a set $D$ that satisfies the isomorphism
+The goal is to define a set $D$ that satisfies the isomorphism
 %
-% TODO make sure we deal with the product correctly
 \[ D \cong \Nat + (D \times D) + (D \to \li D). \]
 %
-where the $\li D$ represents the fact that in the case where the value of type
-$\dyn$ is a function, the function when called may return an error or take a
-step of computation. Unfortunately, this equation does not have inductive or
+where the use of $\li D$ in the codomain corresponds to the fact that when a
+value of type $\dyn$ is a function, the function when called may return an error
+or take a step of computation. This equation does not have inductive or
 coinductive solutions. The usual way of dealing with such equations is via
-domain theory, by which we can obtain an exact solution. However, the machinery
-of domain theory can be difficult for language designers to learn and apply in
-the mechanized setting. Instead, we will leverage the tools of guarded type
-theory. Consider the following similar-looking equation:
+domain theory, by which we can obtain an exact solution. However, by using
+guarded recursion we can solve the equation and obtain a guarded solution in a
+process analogous to giving a normal set-theoretic denotational semantics.
+Consider the following similar-looking equation:
 %
@@
 %
-% TODO refer to technical background?
 Observe that the unfolding equality characterizing the guarded fixpoint implies
@@
 adjunction this is equivalent to an ordinary function between $\sem{A'}$ and
-$UF\sem{A}$, i.e., $\sem{A'} \to \li \sem{A}$.
+$UF\sem{A}$, i.e., $\sem{A'} \to \li \sem{A}$. We use $\text{ext}$ to go from
+the former to the latter in the definitions of the downcasts.
+
+% The upcast for $c_i \ra c_o$ and the downcast for $c_1 \times c_2$ make use of
+% \emph{Kleisli functors} $\tok$ and $\timesk$.
+
+\eric{Introduce Kleisli categories and functors}
 
@@
 \ref{fig:typrec}. The semantics of the up- and downcasts for a type precision
-derivation $c$ are defined by induction on $c$. For $r(A)$ the up- and downcasts
-are simply the identity function. For the composition $c \comp c'$ we simply
-compose the cast for $c$ and the cast for $c'$ as functions. For $c_i \ra c_o$
-we apply the casts for $c_i$ and $c_o$ in the domain and codomain respectively.
-More concretely, for the upcast we are given a function $V_f : A_i \to \li A_o$
-and we must define a function $A_i' \to \li A_o'$. The function first downcasts
-its argument according to $c_i$, resulting in an element of $\li A_i$. It then
-applies $\ext{V_f}{}$ to this value to obtain an element of $\li A_o$. Finally, it
-maps the upcast of $c_o$ over the lift monad, obtaining an element of $\li A_o'$
-as needed. For the downcast from $A_i' \to \li A_o'$ to $A_i \to \li A_o$, the
-resulting function first applies the upcast by $c_i$ to its argument, then
-applies the original function, and finally applies the downcast by $c_o$ to the
-result. Likewise, for $c_1 \times c_2$, we apply the cast for $c_1$ on the left
-and the cast for $c_2$ on the right.
+derivation $c$ are defined by induction on $c$.
+%
+For $r(A)$ the up- and downcasts are simply the identity function. 
+%
+For the composition $c \comp c'$ we compose the cast for $c$ and the cast for
+$c'$ as functions.
+%
+For $c_i \ra c_o$ we apply the casts for $c_i$ and $c_o$ in the domain and
+codomain respectively. More concretely, for the upcast we are given a function
+$V_f : A_i \to \li A_o$ and we must define a function $A_i' \to \li A_o'$. The
+function first downcasts its argument according to $c_i$, resulting in an
+element of $\li A_i$. It then applies $\ext{V_f}{}$ to this value to obtain an
+element of $\li A_o$. Finally, it applies the upcast of $c_o$ via the functorial
+action of the lift monad, obtaining an element of $\li A_o'$ as needed. For the
+downcast from $A_i' \to \li A_o'$ to $A_i \to \li A_o$, the resulting function
+first applies the upcast by $c_i$ to its argument, then applies the original
+function, and finally applies the downcast by $c_o$ to the result.
+%
+Likewise, in the upcast for $c_1 \times c_2$, we apply the cast for $c_1$ on the
+left and the cast for $c_2$ on the right.
 
-The ``base cases" for the casts are  $\inat$, $\itimes$, and $\iarr$. Recall
+The ``base cases'' for the casts are $\inat$, $\itimes$, and $\iarr$. Recall
 that $D$ is isomorphic to $\Nat\, + (D \times D)\, + \laterhs (D \to \li D)$
@@
     \sem{\upc{r(A)}} &= \id \\
-    \sem{\upc{c \comp c'}(V)} &= \sem{\upc{c'}} \circ \sem{\upc{c}} \\
-    \sem{\upc{c_i \ra c_o}} &=
+    \sem{\upc{(c \comp c')}} &= \sem{\upc{c'}} \circ \sem{\upc{c}} \\
+    \sem{\upc{(c_i \ra c_o)}} &=
       \lambda V_f . \text{map } \upc{c_o} \circ \ext{V_f}{} \circ \dnc{c_i} \\
-    \sem{\upc{c_1 \times c_2}} &= 
+    \sem{\upc{(c_1 \times c_2)}} &= 
       \lambda (V_1, V_2) . ((\sem{\upc{c_1} V_1}) , 
@@
     % downcasts
-    \sem{\dnc{\inat}} &=
+    \sem{\dnc{r(A)}} &= \id \\
+    \sem{\dnc{(c \comp c')}} &= \sem{\dnc{c}} \circ \sem{\dnc{c'}} \\
+    \sem{\dnc{(c_i \ra c_o)}} &=
+     \text{ext}(\eta(\lambda V_f . \dnc{c_o} \circ V_f \circ \upc{c_i})) \\
+    \sem{\dnc{(c_1 \times c_2)}} &= (\dnc{c_1} \timesk A_2) \circ (A_1' \timesk \dnc{c_2}) \\
+    \sem{\dnc{\inat}} &= \text{ext}(
       \lambda V_d . \text{case $({V_d})$ of}
       \{ \inl\, (\inl\, n) \to \eta\, n
-         \alt \text{otherwise} \to \mho \} \\
-    \sem{\dnc{\itimes}} &=
+         \alt \text{otherwise} \to \mho \}) \\
+    \sem{\dnc{\itimes}} &= \text{ext}(
       \lambda V_d . \text{case $({V_d})$ of}
       \{ \inl\, (\inr\, (d_1, d_2)) \to \eta\, (d_1, d_2)
-         \alt \text{otherwise} \to \mho \} \\
-    \sem{\dnc{\iarr}} &=
+         \alt \text{otherwise} \to \mho \}) \\
+    \sem{\dnc{\iarr}} &= \text{ext}(
       \lambda V_d . \text{case $({V_d})$ of}
       \{ \inr\, \tilde{f} \to \theta\, (\lambda t. \eta (\tilde{f}_t))
-         \alt \text{otherwise} \to \mho \} \\
+         \alt \text{otherwise} \to \mho \}) \\
     % \sem{\ret\, V}       &= \lambda \gamma . \eta\, \sem{V} \\

### 2024-07-08 paper-new/concrete-relational-model.tex
[paper] revise first section on relational model

@@
 
-\eric{Wording?}
-In the previous section, we gave a denotational semantics to the terms of the
-gradually-typed lambda calculus using the tools of SGDT. In this section, we
-build on the ideas used in defining a term model with the goal of defining a
-model of the full gradually-typed lambda calculus, i.e., one that has constructs
-that model type and term precision.
+\eric{Review this section}
+
+In the previous section, we gave a denotational call-by-push-value model for the
+terms of the gradually-typed lambda calculus using the tools of SGDT. In this
+section, we build on the ideas used in defining a term model with the goal of
+defining a model of the full gradually-typed lambda calculus, i.e., one that has
+constructs that model type and term precision.
+
+Recall that the cast calculus introduced in Section \ref{sec:GTLC} has not only
+terms but also an \emph{inequational theory} with axioms for type precision and
+term precision that constitute an abstract formulation of the graduality
+property. Thus, our goal now is to build a denotational model interpreting these
+axioms.
 %
-% The purpose of this section is to motivate the need for many of the
-% seemingly complicated features of the abstract model introduced in
-% Section \ref{sec:abstract-models}.
 
-We will not give the complete description of the concrete model in this section;
-rather we seek to develop just enough to motivate the need for the non-standard
-aspects in the definitions of abstract model introduced in Section
-\ref{sec:abstract-models}. After defining the notion of abstract model and
-describing a modular way of constructing such a model, we will return in Section
-\ref{sec:concrete-model} to the complete construction of a concrete model.
+\subsection{Modeling Type and Term Precision}
 
-\subsection{A First Attempt}
+% We will model type precision $c : A \ltdyn A'$ as a relation between the sets
+% $\sem{A}$ and $\sem{A'}$.
 
-% TODO should we frame this in terms of the idealized categorical models
-% i.e. say what the vertical and horizontal morphisms and squares are
-% introduced previously? What is the proper denotation of values versus terms?
-% We don't distinguish between value types and computation types in the syntax
-
-As a first attempt at giving a semantics to the term precision relation, we
-could try to model types as sets equipped with an ordering that models term
-precision. Since term precision is reflexive and transitive, and since we
-identify terms that are equi-precise, we could model value types as
+To model term precision, we begin by equipping the denotation of every type with
+an ordering relation. Since term precision is reflexive and transitive, and
+since we identify terms that are equi-precise, we model value types as
 partially-ordered sets and values $\Gamma \vdash V : A$ as \emph{monotone}
 functions. Analogously, every error domain is now equipped with a partial
-ordering for which the error element is the bottom, and the map $\theta_B
-\laterhs B \to B$ is now required to be monotone. 
+ordering for which the error element is the bottom element, and the map
+$\theta_B \laterhs B \to B$ is now required to be monotone. 
 %
 \footnote{This requires us to define a notion of ``later'' for partially-ordered
-sets $X$, but this can be done. The elements are simply ``later''-elements of
-$X$, and the ordering is the obvious one, i.e., two later-elements are related
-if after one time step, the elements they step to are related.}
+sets $X$. The elements are simply ``later''-elements of $X$, and the ordering is
+the obvious one, i.e., two later-elements are related if after one time step,
+the elements they step to are related.}
+
+We model type precision $c : A \ltdyn A'$ as a \emph{monotone relation}, i.e., a
+relation $c$ that is upward-closed under the relation on $A'$ and
+downward-closed under the relation on $A$. The relation on the poset $A$ is
+denoted $r(A)$.
+% This extends to products $c_1 \times c_2$ in the obvious way.
 %
-% \eric{What about computation
-% types? We need to extend the error domains mentioned in the term model section
-% to also have an ordering such that the error element is the least element and
-% the theta morphism is monotone. Is this best introduced here, or should we wait
-% until after we have introduced the revised notions of abstract model?}
+Likewise, we define a notion of \emph{computation relation} on computation types
+$B$ and $B'$. A computation relation $d : B \rel B'$ is a monotone relation on
+the underlying sets that preserves error and respects $\theta$. That is, $\mho_B
+\mathbin{d} y$ for all $y : B'$ and if $\laterhs_t (\tilde{x}_t \mathbin{d}
+\tilde{y}_t)$, then $\theta_B (\tilde{x}) \mathbin{d} \theta_{B'} (\tilde{y})$.
+The functors $U$ and $F$ introduced in the previous section extend to relations,
+The functor $U$ simply takes the underlying monotone relation of a computation
+relation; we will give the definition of $F$ on relations later in this section.
+
+Before moving on, we introduce the notion of \emph{square}, which specifies
+a relation between two monotone functions. The definition is as follows. Let
+$A_i, A_o, A_i', A_o'$ be partially-ordered sets. Let $c_i : A_i \rel A_i'$ and
+$c_o : A_o \rel A_o'$ be monotone relations, and let $f : A_i \to A_o$ and $g :
+A_i' \to A_o'$ be monotone functions. We say that $f \ltsq{c_i}{c_o} g$ if for
+all $x : A_i$ and $y : A_i'$ with $x \binrel{c_i} y$, we have $f(x) \binrel{c_o}
+g(y)$. We visualize this situation as the following square:
+
+% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBX2kiXSxbMCwxLCJBX28iXSxbMSwwLCJBX2knIl0sWzEsMSwiQV9vJyJdLFsyLDMsImciXSxbMCwxLCJmIiwyXSxbMCwyLCJjX2kiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxLDMsImNfbyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+\[\begin{tikzcd}[ampersand replacement=\&]
+	{A_i} \& {A_i'} \\
+	{A_o} \& {A_o'}
+	\arrow["{c_i}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+	\arrow["f"', from=1-1, to=2-1]
+	\arrow["g", from=1-2, to=2-2]
+	\arrow["{c_o}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+\end{tikzcd}\]
+
+We note that for any $c$, there is a ``vertical identity'' square $\id
+\ltsq{c}{c} \id$, as can be seen by unfolding the definition. Similarly, for any
+montone function $f : A_i \to A_o$, there is a ``horizontal identity'' square $f
+\ltsq{r(A_i)}{r(A_o)}$ by the definition of monotone function.
+
+The cast rules UpL/UpR/DnL/DnR specify a relationship between the semantics of
+type precision and the corresponding casts (see Section \ref{sec:GTLC}). In
+particular, they specify that the relations are \emph{representable} by the
+casts. The UpL and UpR rules refer to upcasts, and since upcasts are pure, these
+rules apply to \emph{value} relations $c$. The DnL and DnR rules refer to
+downcasts, and since downcasts are effectful, these rules apply to
+\emph{computation} relations $d$.
+
+As an example, consider the UpL rule. It states that if $c : A_1 \ltdyn A_2$ and
+$c' : A_2 \ltdyn A_3$ and $x : A_1$ is related to $y : A_3$ via the composition
+$c \comp c'$, then the upcast $\upc{c} x$ is related to $y$ via $c'$. This is
+equivalent to the existence of the following square:
+
+% https://q.uiver.app/#q=WzAsNSxbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMyw0LCJjJyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMywiXFx1cGN7Y30iLDJdLFsyLDQsIlxcaWQiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+\[\begin{tikzcd}[ampersand replacement=\&]
+	{A_1} \& {A_2} \& {A_3} \\
+	{A_2} \&\& {A_3}
+	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+	\arrow["{\upc{c}}"', from=1-1, to=2-1]
+	\arrow["{c'}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
+	\arrow["\id", from=1-3, to=2-3]
+	\arrow["{c'}"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
+\end{tikzcd}\]
+
+The three squares corresponding to the other rules are analogous. Observe that
+although this is really a fact about the relationship between $c$ and $\upc{c}$,
+we must universally quantify over all relations $c'$. This means we cannot
+require the existence of this square as part of the definition of a monotone
+relation, as it is self-referential. This would seem to imply that we cannot
+give a syntax-independent semantic model for gradual typing. However, the key
+observation is that through compositionality, we can \emph{derive} the above
+squares from simpler ones that do not involve composition of relations. For
+example, below is the simpler square corresponding to UpL:
+
+% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzIiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzIiXSxbMCwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJyKEFfMikiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxcdXBje2N9IiwyXSxbMiwzLCJcXGlkIl1d
+\[\begin{tikzcd}[ampersand replacement=\&]
+	{A_1} \& {A_2} \\
+	{A_2} \& {A_2}
+	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+	\arrow["{\upc{c}}"', from=1-1, to=2-1]
+	\arrow["\id", from=1-2, to=2-2]
+	\arrow["{r(A_2)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+\end{tikzcd}\]
+
+We can use this to derive the original UpL square by composing this square with
+the identity square for $c'$:
+
+% https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMSwxLCJBXzIiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzMsNSwicihBXzIpIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNSw0LCJjJyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMywiXFx1cGN7Y30iLDJdLFsyLDQsIlxcaWQiXSxbMSw1LCJcXGlkIiwyXV0=
+\[\begin{tikzcd}[ampersand replacement=\&]
+	{A_1} \& {A_2} \& {A_3} \\
+	{A_2} \& {A_2} \& {A_3}
+	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+	\arrow["{\upc{c}}"', from=1-1, to=2-1]
+	\arrow["{c'}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
+	\arrow["\id"', from=1-2, to=2-2]
+	\arrow["\id", from=1-3, to=2-3]
+	\arrow["{r(A_2)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+	\arrow["{c'}"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
+\end{tikzcd}\]
+
+The existence of these simpler squares for a relation can be stated without
+reference to other relations. Thus we can require that these squares exist in
+the definition of relation without running into the circularity issue we faced
+with the original rules.
+
+Finally, we model the term precision ordering $\Gamma \ltdyn \Gamma' \vdash M \ltdyn
+N : A \ltdyn A'$ as an ordering relation between the monotone functions denoted
+by $M$ and $N$, as shown below:
+
+% https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXEdhbW1hIl0sWzAsMSwiVUYgQSJdLFsxLDAsIlxcR2FtbWEnIl0sWzEsMSwiVUZBJyJdLFsyLDMsIlxcc2Vte059Il0sWzAsMSwiXFxzZW17TX0iLDJdLFswLDIsIlxcR2FtbWFee1xcbHRkeW59IiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJVRmMiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
+\[\begin{tikzcd}[ampersand replacement=\&]
+	\Gamma \& {\Gamma'} \\
+	{UF A} \& {UFA'}
+	\arrow["{\Gamma^{\ltdyn}}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+	\arrow["{\sem{M}}"', from=1-1, to=2-1]
+	\arrow["{\sem{N}}", from=1-2, to=2-2]
+	\arrow["UFc"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+\end{tikzcd}\]
+
+where $\Gamma = \{A_1, A_2, \dots\}$ and $\Gamma' = \{A_1' , A_2', \dots\}$ and for
+each $j$, $c_j : A_j \ltdyn A_j'$, and the relation $\Gamma^{\ltdyn}$ is simply
+the product of the $c_i$.
+
+It remains to define $Fc$, i.e., the action of $F$ on relations.
 %
-We model type precision $c : A \ltdyn B$ as a \emph{montone relation}, i.e., a
-relation $R$ that is upward-closed under the relation on $B$ and downward-closed
-under the relation on $A$. We could then model the term precision ordering $M
-\ltdyn N : A \ltdyn B$ as an ordering relation between the monotone functions
-denoted by $M$ and $N$.
-
-However, it turns out that modeling term precision is not as straightforward as
-one might hope. A first attempt might be to define an ordering $\semltbad$
-between $\li X$ and $\li Y$ that allows for computations that may take different
-numbers of steps to be related. We call this the \emph{step-insensitive error
-ordering}. The ordering is parameterized by an ordering relation $\le$ between
-$X$ and $Y$. The definition is by guarded recursion and is shown in Figure
-\ref{fig:step-insensitive-error-ordering}.
+% However, lifting a relation between $A$ and $A'$ to a relation between $\li A$
+% and $\li A'$ proves problematic if we allow computations that take different
+% numbers of steps to be related. To illustrate the issue, let us define an
+% ordering $\semltbad$ between $\li A$ and $\li A'$; we call this the
+% \emph{step-insensitive error ordering}. 
+%
+We note that the graduality property and the axioms of the inequational theory
+are independent of the intensional stepping behavior of terms, so our ultimate
+notion that interprets term precision will need to be oblivious to stepping as
+well.
+%
+To that end, let us define a \emph{step-insensitive error ordering} $\semltbad$
+between $\li A$ and $\li A'$; The ordering is parameterized by an ordering
+relation $\le$ between $A$ and $A'$. The definition is by guarded recursion and
+is shown in Figure \ref{fig:step-insensitive-error-ordering}. Recall that
+$\delta : \li A \to \li A$ is defined by $\delta = \theta_A \circ \nxt$.
 
@@
             (\theta\, \tilde{l} = \delta^n(\eta\, x))
-        \text { for some $n$ and $x : \ty{X}$ such that $x \le y$ } \\
+        \text { for some $n$ and $x : A$ such that $x \le y$ } \\
         %
-        &\mho \semltbad \theta\, \tilde{l'} \text { if } 
-            \theta\, \tilde{l'} = \delta^n(\mho) \text { for some $n$ } \\
+        &\theta\, \tilde{l} \semltbad \mho \text { if } 
+            \theta\, \tilde{l} = \delta^n(\mho) \text { for some $n$ } \\
         %	
@@
             (\theta\, \tilde{l'} = \delta^n (\eta\, y))
-        \text { for some $n$ and $y : \ty{Y}$ such that $x \le y$ }
+        \text { for some $n$ and $y : A'$ such that $x \le y$ }
     \end{align*}
@@
 order for these terms to be related, the side that steps must terminate with a
-value in some finite number of steps $n$, and that value must be related to
-the value returned by the other side. Likewise, if the LHS steps and the RHS
-immediately errors, then the LHS must eventually terminate with error.
+value in some finite number of steps $n$, and that value must be related to the
+value returned by the other side. Likewise, if the LHS steps and the RHS
+immediately errors, then in order to be related, the LHS must eventually
+terminate with error.
 
@@
 
-The problem with the above definition is that the relation is \emph{not}
-transitive, which we explain below. We begin with a general ``no-go'' theorem
-about relations defined on the guarded lift monad:
+The key issue with the above ordering is it is \emph{not} transitive, as we will
+show below. We begin with a general ``no-go'' theorem about relations defined on
+the guarded lift monad:
 
@@
 
-% It can be shown (in Clocked Cubical Type
-% Theory) that if $R$ is a relation on $\li X$ satisfying three specific
-% properties, one of which is transitivity, then that relation is trivial. The
-% other two properties are that the relation is a congruence with respect to
-% $\theta$, and that the relation is closed under delays $\delta = \theta \circ
-% \nxt$ on either side. Since the above relation \emph{does} satisfy the other
-% two properties, we conclude that it must not be transitive.
+\eric{Read this}
+The lack of transitivity presents a major barrier towards giving a compositional
+semantics to gradual typing: if our relations are not transitive, then we cannot
+compose squares horizontally. But we have argued above that horizontal
+composition is essential for modelling the axioms of gradual typing in a
+syntax-independent manner.
 
-\eric{I'm not sure that this is a satisfactory explanation...}
-At this point, we could give up on transitivity and instead of requiring types
-to be modeled by partially-ordered sets, we could model types as sets equipped
-with a reflexive, anti-symmetric ordering relation. However, this is not
-convenient for carrying out the constructions necessary for proving graduality.
-For example, in order to model the UpL/UpR/DnL/DnR rules for casts, in the
-presence of transitivity it suffices to model simpler rules that do not build in
-composition. We can then use transitivity to derive the actual, more complicated
-rules we need to show.
-% For example, in order to model the UpL/UpR/DnL/DnR rules for casts, prior work
-% has taken as primitive simpler rules that do not involve composition, and then
-% uses transitivity to establish the more complicated versions that have
-% composition ``built in''.
-Without transitivity, however, this is not possible, and so we must instead take as
-primitive the more complicated versions of the cast rules. Then in the model, we
-must verify that those rules hold ``from scratch''. So, although this approach
-would suffice for proving graduality, it lacks the compositionality that we seek
-in a reusable framework for the semantics of gradual typing.
+% In particular, to model the UpL/UpR/DnL/DnR rules
+% for casts in the presence of transitivity, it is sufficient to establish the
+% validity of simpler rules that do not ``build in'' composition. We can then use
+% transitivity to derive the original versions of the rules. On the other hand,
+% without transitivity we must instead validate the cast rules in the model ``from
+% scratch''. In fact, we cannot even define the semantic interpretation of type
+% precision in a syntax-independent manner. The issue is that in the definition of
+% relation, the requirement that it be representable now involves quantifying over
+% all other relations, which is circular.
+
+% So, although this approach would suffice for proving graduality, it lacks the
+% compositionality that we seek in a reusable framework for the semantics of
+% gradual typing.
 
 %Because it is convenient to make use of transitive reasoning in proving graduality, ...
-We are therefore led to wonder whether we can formulate an ordering
-relation on $\li X \times \li Y$ that \emph{is} transitive. The theorem above
-tells us, however, that any such relation must either not be a congruence with
-respect to $\theta$, or must not be closed under delay.
+We therefore seek to formulate an ordering relation on $\li A \times \li A'$
+that \emph{is} transitive. The theorem above tells us, however, that any such
+relation must either not be a congruence with respect to $\theta$, or must not
+be closed under delay.
 \footnote{Technically, we only need to give up closure under delay on \emph{one}
 side, as the above theorem requires $R$ to be closed on both sides. We could
-define two separate ``step-semi-sensitive'' relations on $\li X \times \li Y$,
+define two separate ``step-semi-sensitive'' relations on $\li A \times \li A'$,
 each closed under delays on the left and right respectively. The denotation of
@@
 would not be able to prove properties of the relation using \lob-induction,
-e.g., that the monadic bind for the guarded lift monad is a \emph{monotone}
-function with respect to the ordering.
+e.g., that $\ext{f}{}$ is a \emph{monotone} function with respect to the
+ordering.
 %and moreover this would mean that $\theta$ would not be a monotone function.
@@
 
-\eric{Does this belong in the next section?}
 Importantly, we note that as with the original ordering $\semltbad$ defined at
@@
 (again as a result of the above no-go theorem). It may therefore seem that we
-have not solved the original issue we faced. In a sense, this is true. Indeed,
-we will need to weaken our abstract notion of categorical model to no longer
-require transitivity of term precision. On the other hand, we now have the
-ability to apply transitive reasoning when working with the lock-step ordering.
-This allows us to carry out complex proofs compositionally, a key goal of our
-framework.
+have not solved the original issue we faced. In a sense, this is true. The model
+of gradual typing that we end up with will not support compositional reasoning.
+However, by decomposing the denotation of term precision in the above manner, we
+can employ compositional reasoning provided we work only with the
+\emph{lock-step error ordering}. Thus, the tentative plan going forward will be
+to carry out the proofs compositionally using the lock-step ordering. We will
+then apply the closure under weak bisimilarity in the denotation of term
+precision given above to handle the aspects involving stepping.
 
-\subsection{Failure of the Cast Rules}
+% ...to carry out the proofs using the lock-step ordering and handle the stepping
+% behavior separately via weak bisimilarity. In the end we combine everything
+% together using the above denotation for term precision
+
+% Indeed, we will need to weaken our abstract notion of categorical model to no
+% longer require transitivity of term precision. On the other hand, we now have
+% the ability to apply transitive reasoning when working with the lock-step
+% ordering. This allows us to carry out complex proofs compositionally, a key goal
+% of our framework.
+
+\subsection{Adjusting the Cast Rules}
 
@@
 downcast from the dynamic type to a function takes a step, i.e., involves a
-$\theta$. Consider a simplified version of the DnL rule shown below (see Section
-\ref{sec:GTLC} for background):
+$\theta$. Consider the DnL rule shown below:
 %
 \begin{mathpar}
-  \inferrule*{c : A \ltdyn B \and M \ltdyn_i N : B}
-             {\dnc{c}{M} \ltdyn_i N : c}
+  \inferrule*{c : A \ltdyn A' \and M \ltdyn N : A'}
+             {\dnc{c}{M} \ltdyn N : c}
 \end{mathpar}
@@
 definition of the semantics of a downcast given in Section
-\ref{sec:term-interpretation}. When we downcast $M$ from $dyn$ to $\dyntodyn$,
-semantically this will involve a $\theta$ because the value of type $D$ in the
-model will be a \emph{later} function $\tilde{f}$.
+\ref{sec:term-interpretation}. The DnL rule corresponds to the following square:
+
+% https://q.uiver.app/#q=WzAsNCxbMCwwLCJGRCJdLFswLDEsIkZVKEQgXFxhcnIgRkQpIl0sWzEsMCwiRkQiXSxbMSwxLCJGRCJdLFswLDIsInIoRkQpIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJGXFxpYXJyIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJcXGRuY3tcXGlhcnJ9IiwyXSxbMiwzLCJcXGlkIl1d
+\[\begin{tikzcd}[ampersand replacement=\&]
+	FD \& FD \\
+	{FU(D \arr FD)} \& FD
+	\arrow["{r(FD)}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+	\arrow["{\dnc{\iarr}}"', from=1-1, to=2-1]
+	\arrow["\id", from=1-2, to=2-2]
+	\arrow["F\iarr"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+\end{tikzcd}\]
+
+The downcast on the left will insert a $\theta$ in the case where the value of
+type $D$ is a later-function $\tilde{f}$.
 %
-Thus, in order for the downcast to be related to the term $N$ in the lock-step
-error ordering on $\li (D \to \li D)$, the right-hand side must be of the form
-$\theta(\dots)$, and moreover, after one time step, the argument to $\theta$ on
-the LHS must be related to the argument of $\theta$ on the RHS. We can achieve
-this by inserting a ``delay'' on the right, i.e., wrapping the RHS in $\delta =
-\theta \circ \nxt$. Notice that this delay function is weakly bisimlar to the
-identity, and thus it will disappear when we apply the bisimilarity-closure
-construction described above. That is, this delay makes no difference in the
-extensional setting, but its presence is crucial in the intensional setting.
+Thus, in order for the downcast to be related to the RHS in the lock-step error
+ordering on $\li (D \to \li D)$, the RHS must be of the form $\theta(\dots)$,
+and moreover, after one time step, the argument to $\theta$ on the LHS must be
+related to the argument of $\theta$ on the RHS. As it stands now, this need not
+be the case. Thus, we conclude that the DnL rule does not in general hold under
+the lock-step error ordering. However, observe that we can remedy this
+particular situation by inserting a ``delay'' on the right, i.e., wrapping the
+RHS in $\delta = \theta \circ \nxt$. Notice that this delay function is weakly
+bisimlar to the identity, and thus it will disappear when we apply the
+bisimilarity-closure construction described above. That is, this delay makes no
+difference in the extensional setting, but its presence is crucial in the
+intensional setting.
 
@@
 and an upcast corresponding to $d$. We must be able to insert delays in a
-functorial manner to mimic the structure of the casts.
+functorial manner to mimic the structure of the casts. In the next section, we
+will make the necessary definitions and describe the relevant constructions.
 
-For the sake of reusability and modularity, rather than carry out this
-construction in the concrete setting developed here, we will instead return to
-the abstract setting and adjust our definition of model to account for these
-requirements. We will break the construction into smaller steps and isolate the
-pieces that require the techniques of SGDT from those that do not. Then with
-this framework at our disposal, we will return to the construction of a concrete
-model in Section \ref{sec:concrete-model}.
+% For the sake of reusability and modularity, rather than carry out this
+% construction in the concrete setting developed here, we will instead return to
+% the abstract setting and adjust our definition of model to account for these
+% requirements. We will break the construction into smaller steps and isolate the
+% pieces that require the techniques of SGDT from those that do not. Then with
+% this framework at our disposal, we will return to the construction of a concrete
+% model in Section \ref{sec:concrete-model}.
 % , taking as a starting point the definitions introduced in the current section.

### 2024-07-08 paper-new/concrete-relational-model.tex
[paper] small wording changes

@@
 have not solved the original issue we faced. In a sense, this is true. The model
-of gradual typing that we end up with will not support compositional reasoning.
-However, by decomposing the denotation of term precision in the above manner, we
-can employ compositional reasoning provided we work only with the
-\emph{lock-step error ordering}. Thus, the tentative plan going forward will be
-to carry out the proofs compositionally using the lock-step ordering. We will
-then apply the closure under weak bisimilarity in the denotation of term
-precision given above to handle the aspects involving stepping.
+of gradual typing that we end up with will not support \emph{extensional}
+compositional reasoning. However, by decomposing the denotation of term
+precision in the above manner, we can employ \emph{intensional} compositional
+reasoning by working with the lock-step error ordering. Thus, the tentative plan
+going forward will be to carry out the proofs compositionally using the
+lock-step ordering. We will then apply the closure under weak bisimilarity in
+the denotation of term precision given above to handle the aspects involving
+stepping.
 
@@
 We have seen how we can recover transitive reasoning by working with the
-lock-step error ordering. However, we cannot completely isolate the reasoning
-about steps from the reasoning about the ordering.
+lock-step error ordering. However, it turns out that we cannot completely
+isolate the reasoning about steps from the reasoning about the ordering.
 % Unfortunately, by defining the partial order on the guarded lift monad to be

### 2024-07-08 paper-new/intro.tex
abstract and intro

@@
 \subsection{Gradual Typing and Graduality}
-In modern programming language design, there is a tension between \emph{static} typing
+In programming language design, there is a tension between \emph{static} typing
 and \emph{dynamic} typing disciplines.
@@
 % Sound gradual typing
-Furthermore, the parts of the program that are written in a dynamic style should soundly
-interoperate with the parts that are written in a static style.
-That is, the interaction between the static and dynamic components of the codebase
-should preserve the guarantees made by the static types.
-In particular, while statically-typed code can error at runtime in a gradually-typed language,
-such an error can always be traced back to a dynamically-typed term that
-violated the typing contract imposed by statically typed code.
+Furthermore, the parts of the program that are written in a dynamic
+style should soundly interoperate with the parts that are written in a
+static style.  That is, the interaction between the static and dynamic
+components of the codebase should preserve, to the extent possible,
+the guarantees made by the static types.  In particular, while
+statically-typed code can error at runtime in a gradually-typed
+language, such an error can always be traced back to a
+dynamically-typed term that violated the typing contract imposed by
+statically typed code. Further, static type assertions are sound in
+the static portion, and should enable type-based reasoning and
+optimization.
 
@@
 One of the fundamental theorems for gradually typed languages is
-\emph{dynamic gradual guarantee}, originally defined by Siek, Vitousek, Cimini,
-and Boyland \cite{siek_et_al:LIPIcs:2015:5031}, also called \emph{graduality} \cite{new-ahmed2018},
-by analogy with parametricity.
-Informally, graduality says that going from a dynamic to static style should only allow for the introduction of static or dynamic type errors, and not otherwise change the meaning of the program.
+\emph{graduality}, also known as the \emph{dynamic gradual guarantee}
+\emph{dynamic gradual guarantee}, originally defined by Siek,
+Vitousek, Cimini, and Boyland \cite{siek_et_al:LIPIcs:2015:5031,
+  new-ahmed2018}.
+%
+Informally, graduality says that going from a dynamic to static style should only allow for the introduction of static or dynamic type errors, and not otherwise change the behavior of the program.
 %
 This is a way to capture programmer intuition that increasing type
-precision corresponds to a generalized form assertions at runtime
-only, and so they can trust that prior observed behavior of their
-dynamically typed code will remain unchanged as long as it satisfies
-the new stricter typing discipline.
+precision corresponds to a generalized form of runtime assertions in
+that there are no observable behavioral changes up to the point of the
+first dynamic type error\footnote{once a dynamic type error is raised,
+in languages where the type error can be caught, program behavior may
+then further diverge, but this is typically not modeled in gradual
+calculi.}.
+%
+Fundamentally, this property comes down to the behavior of
+\emph{runtime type casts}\max{TODO: introduce casts more thoroughly as
+  they are important}.
 
@@
 
-Our goal in this work is to provide a reusable semantic framework for
-gradually typed languages that can be used to prove the aforementioned
-properties: graduality and type-based reasoning.
+More concretely, consider a gradually typed language whose only
+effects are gradual type errors and divergence. Then if we fix a
+result type of natural numbers, a whole program semantics is a partial
+function from closed programs to either natural numbers or errors:
+\[ -\Downarrow : \{M \,|\, \cdot \vdash M : \natty \} \rightharpoonup \mathbb{N} \union {\mho} \]
+where $\mho$ is notation for a runtime type error. We write $M
+\Downarrow n$ and $M\Downarrow \mho$ to mean this semantics is defined
+as a number or error, and $M\Uparrow$ to mean the semantics is
+undefined, representing divergence.
+%
+A well-behaved semantics should then satisfy several properties. First, it
+should be adequate: natural number constants should step to
+themselves. Second it should validate type based reasoning. To
+formalize type based reasoning we give a typed equational theory for
+terms of the language $M \cong N$ for when two terms should be
+considered equivalent. Then we want to verify that the big step
+semantics respects this equational theory: if closed programs $M \cong
+N$ are equivalent in the equational theory then they have the same
+semantics, $M \Downarrow n \iff N \downarrow n$ and $M\Uparrow \iff N
+\Uparrow$ and $M \Downarrow \mho \iff N \Downarrow \mho$.
+%
+Lastly, the graduality property is defined by giving an
+\emph{inequational} theory called term precision, where $M \ltdyn N$
+roughly means that $M$ and $N$ have the same type erasure and $M$ has
+at each point in the program a more precise/static type than $N$.
+%
+Then, the graduality property states that if $M \ltdyn N$ are whole
+programs then $M$ must either have the same behavior as $N$ or error:
+Either $M\Downarrow \mho$ or $M \Downarrow n $ and $N \Downarrow n$ or
+$M \Uparrow $ and $N \Uparrow$\footnote{we use a slightly more complex
+definition of this relation in our technical development below that is
+classically equivalent but constructively weaker}.
+
+\subsection{Denotational Semantics in Guarded Domain Theory}
+
+Our goal in this work is to provide an \emph{expressive},
+\emph{reusable}, \emph{compositional} semantic framework for defining
+such well-behaved semantics of programs.
+%
+Our approach to achieving this goal is to provide a compositional
+\emph{denotational semantics}, mapping types to a kind of semantic
+domain, terms to functions and relations such as term precision to
+proofs of semantic relations between the denoted functions.
+%
+Since the denotational constructions are all syntax-independent, the
+constructions we provide may be reused for similar languages. Since it
+is compositional, components can be mixed and matched depending on
+what source language features are present.
+%
+Providing this semantics for gradual typing is inherently complicated
+in that it involves: (1) recursion and recursive types through the
+presence of dynamic types, (2) effects in the form of divergence and
+errors (3) relational models in capturing the graduality
+property. Recursion and recursive types must be handled using some
+flavor of domain theory. Effects can be modeled using monads in the
+style of Moggi, or adjunctions in the style of Levy. Relational
+properties and their verification lead naturally to the use of
+reflexive graph categories or double categories.
+
+The only prior denotational semantics for gradual typing was given by
+New and Licata and is based on \emph{domain theory}
+\cite{newlicata2019}. The fundamental idea is to equip $\omega$-CPOs
+with an additional ``error ordering'' $\ltdyn$ which models the
+graduality ordering, and for casts to arise from
+\emph{embedding-projection pairs}. Then the graduality property
+follows as long as all language constructs can be interpreted using
+constructions that are monotone with respect to the error ordering.
+%
+This framework has the benefit of being compositional, and was
+expressive enough to be extended to model dependently typed gradual
+typing \cite{dependentgradualtyping}.
+%
+However, an approach based on classical domain theory has fundamental
+limitations: domain theory is incapable of modeling certain ``highly
+recursive'' features of programming languages such as dynamic type tag
+generation and higher-order references, which are commonplace in
+real-world gradually typed systems as well as gradual calculi
+\cite{examples-of-gradual-stuff}. Our goal in this work is to develop
+an approach that will eventually scale up to these advanced features,
+though for a first work we focus on adapting their results for a more
+basic semantics.
+
+The main denotational alternative to classical domain theory that can
+successfully model these advanced features is \emph{guarded domain
+theory}, which we adopt in this work. While classical domain theory is
+based on modeling types as ordered sets with certain joins, guarded
+domain theory is based on an entirely different foundations, sometimes
+(ultra)metric spaces but more commonly as ``step-indexed sets'', i.e.,
+objects in the ``topos of trees'', i.e., presheaves on the poset of
+natural numbers. We think of these as modeling a kind of ``sequence of
+approximations'' to the domain being modeled.
+%
+Key to guarded domain theory is that there is a ``later'' operator
+$\triangleright$ on types. Thinking of types as a sequence of
+approximations, the later operator delays the approximation by one
+step. Then the crucial axiom of guarded domain theory is that any
+guarded equation $X \cong F(\triangleright X)$ has a unique
+solution. This allows guarded domain theory to model essentially
+\emph{any} recursive concept, with the caveat that it is guarded by a
+later.
+%
+This caveat is the main source of difficulty in adapting a semantic
+approach based on classical domain theory to guarded domain theory:
+classical domain theory has limitations in what it can model, but it
+provides \emph{exact} solutions to domain equations when it
+applies. When adapting the New-Licata approach to guarded domain
+theory, this means we must work with an \emph{intensional} semantics,
+one where unfolding the dynamic type takes an observable runtime step.
+
+\subsection{Synthetic Guarded Domain Theory}
+
+While guarded domain theory can be presented analytically using
+ultrametric spaces or the topos of trees, in practice it is
+considerably simpler to work \emph{synthetically} by working in a
+non-standard foundation such as guarded type theory where the later
+modality is simply an operation on our basic notion of type, and we
+take as an axiom that guarded domain equations have a (necessarily
+unique) solution. This has the added benefit that providing a
+denotational semantics of gradually typed terms is as simple as an
+semantics of an effectful language using a monad on the category of
+sets. Not only does this make on-paper reasoning about guarded domain
+theory easier, it also enables a simpler avenue to verification in a
+proof assistant. Whereas formalizing analytic guarded domain theory
+would require significant theory of presheaves and making sure that
+all constructions are functors on categories of presheaves,
+formalizing synthetic guarded domain theory can be done by directly
+adding the later modality and the guarded fixed point property
+axiomatically.
+
+\max{TODO: more here}
+
+\subsection{Contributions}
+
+The main contribution of this work is a compositional denotational
+semantics for gradually typed languages that validates $\beta\eta$
+equality and satisfies a graduality theorem. A great deal of the work
+has further been verified in guarded cubical Agda, demonstrating that
+the semantics is readily mechanizable.
+
+\begin{enumerate}
+\item First, we give a simple concrete term semantics where we show
+  how to model the dynamic type as a solution to a guarded domain equation
+
+\item Next, we identify where prior work on classical domain theoretic
+  semantics of gradual typing breaks down when using guarded semantics
+  of recursive types.
+
+\item We develop a key new concept of \emph{syntactic perturbations},
+  which allow us to recover enough extensional reasoning to model the
+  graduality property compositionally
+
+\item We combine this insight together with an abstract categorical
+  model of gradual typing using reflexive graph categories and
+  call-by-push-value to give a compositional construction of our
+  denotational model.
+\item We prove that the resulting denotational model provides a
+  well-behaved semantics as defined above by proving \emph{adequacy},
+  respect for an equational theory and the graduality property.
+\end{enumerate}
+
+The paper is laid out as follows:
+\begin{enumerate}
+\item In Section\ref{sec:syntax} we fix our input language, a fairly
+  typical gradually typed cast calculus.
+\item In Section\ref{sec:concrete-terms} we develop a first
+  denotational semantics in synthetic guarded domain theory that
+  satisfies adequacy and respects the equational theory, but does not
+  validate graduality. We use this to introduce some of our main
+  technical tools: modeling recursive types in guarded type theory and
+  modeling effects using call-by-push-value.
+\item In Section\ref{sec:relational-issues} we show where the
+  New-Licata classical domain theoretic approach fails to adapt
+  cleanly to the guarded setting and explore the difficulties of
+  proving graduality in an intensional model.
+\item In Section\ref{sec:relational-semantics} we
+\item In Section\ref{sec:discussion} we discuss prior work on proving
+  graduality, our partial mechanization of these results and discuss
+  future directions for denotational semantics of gradual typing.
+\end{enumerate}
 

### 2024-07-08 paper-new/paper.tex
mostly writing updates
adjust outline
more intro
remove outdated section
fix a lot of references
PACMPL format for paper
section in paper on concrete term model
[paper] new section on extending the concrete term model to a relational model
[paper] change font size
[paper] move technical background on SGDT; edit section on concrete term model
[paper] remove sections on abstract models from body of paper (TODO: consider incorporating this material into appendix)
abstract and intro

@@
-\documentclass[sigconf,anonymous,review,screen,9pt]{acmart}
+\documentclass[acmsmall,anonymous,review,screen]{acmart}
 \let\Bbbk\relax
@@
   increasingly sophisticated: validity of type based equational
-  reasoning as well as the relational property known as the gradual
-  guarantee or graduality. Many recent works have tackled verifying
-  these properties, but the resulting mathematical developments are
-  highly repetitive and tedious, with few reusable theorems persisting
-  across different developments.
+  reasoning as well as the relational property known as
+  graduality. Many recent works have tackled verifying these
+  properties, but the resulting mathematical developments are highly
+  repetitive and tedious, with few reusable theorems persisting across
+  different developments.
 
-  In this work, we present a new denotational account of gradual
-  typing semantics developed using guarded domain theory. Guarded
-  domain theory combines the expressive power of step-indexed logical
-  relations for modeling recursive features with the modularity and
-  reusability of denotational semantics. Further, recent extensions to
-  cubical Agda mean that synthetic guarded domain theory is readily
-  mechanized in a proof assistant. We demonstrate the feasibility of
-  this approach with a model of gradually typed lambda calculus and
-  prove the validity of beta-eta equality and the graduality theorem
-  for the denotational model. This model should provide the basis for
-  a reusable mathematical theory of gradually typed program semantics.
+  In this work, we present a new denotational semantics for gradual
+  typing developed using guarded domain theory. Guarded domain theory
+  combines the generality of step-indexed logical relations for
+  modeling advanced programming features with the modularity and
+  reusability of denotational semantics. We demonstrate the
+  feasibility of this approach with a model of gradually typed lambda
+  calculus and prove the validity of beta-eta equality and the
+  graduality theorem for the denotational model. This model should
+  provide the basis for a reusable mathematical theory of gradually
+  typed program semantics. Finally, a recent extension to Agda allows
+  synthetic guarded domain theory to be readily mechanized in a proof
+  assistant, and we formalize many of the core theorems of our
+  development.
 %%     We develop a denotational semantics for a simple gradually typed language
@@
 
-\input{technical-background}
+%\input{technical-background}
 
-% \input{syntax}
+\input{syntax}
+
+% \input{extensional-model}
+
+\input{concrete-term-model}
+
+\input{concrete-relational-model}
 
@@
 
-\input{terms}
+%% \input{ordering}
 
-\input{ordering}
+%% \input{graduality}
 
-\input{graduality}
-
-\input{categorical-model}
+% \input{categorical-model}
 
 %\input{denotational-model}
+%\input{terms}
 \input{concrete-model}
@@
 
-% \section{Discussion}\label{sec:discussion}
-
-% \subsection{Synthetic Ordering}
-
-% While the use of synthetic guarded domain theory allows us to very
-% conveniently work with non-well-founded recursive constructions while
-% abstracting away the precise details of step-indexing, we do work with
-% the error ordering in a mostly analytic fashion in that gradual types
-% are interpreted as sets equipped with an ordering relation, and all
-% terms must be proven to be monotone.
-% %
-% It is possible that a combination of synthetic guarded domain theory
-% with \emph{directed} type theory would allow for an a synthetic
-% treatment of the error ordering as well.
-
 \end{document}

### 2024-07-09 paper-new/concrete-model.tex
[paper] revise sections on concrete relational model

@@
 
-In this section, we return to the construction of a concrete model of gradual
-typing begun in Section \ref{sec:towards-relational-model}. We begin by constructing
-a step-1 intensional model, and then apply the abstract constructions outlined
-in the previous section to obtain an extensional model. We begin with some
-definitions:
+In this section, we complete the construction of a relational model of gradual
+typing. To manage the complexity, we carry out the construction in phases,
+beginning with a relational model consisting of partially-ordered sets with a
+notion of bisimilarity and then adding requirements for a structured set of
+delays on each type as well as a notion of representability for relations that
+models the cast rules.
+
+\subsection{Phase One: Predomains and Error Domains}
+
+In the previous section, we introduced the notion of weak bisimilarity as a
+relation on the guarded lift monad. Recall that weak bisimilarity on $\li A$ is
+parameterized by a relation on $A$. Thus, in order to define a compositional
+model for all types, we need to equip the objects (i.e. the posets and error
+domains) a reflexive, symmetric bisimilarity relation. This leads to our first
+new definition:
+
 \begin{definition}
@@
 
-% TODO: later of a predomain
+We will generally write $A$ for both a predomain and its underlying set; if we
+need to emphasize the difference we will write $|A|$ for the underlying set of
+$A$.
+
 Given a predomain $A$, we can form the predomain $\later A$. The underlying set
@@
 \end{definition}
+
+The definition of error domain is similar to that given in the previous section,
+except that now, an error domain also has a notion of bisimilarity. We
+additionally impose a condition relating the $\theta$ morphism to the
+bisimilarity relation. More concretely:
 %
@@
     \item A morphism of predomains $\theta_B \colon \later B \to B$
+    \item For all $x : B$ we have $\theta_B(\nxt\, x) \bisim_B x$
 \end{itemize}
@@
 %
-Morphisms of error domains are morphisms of the underlying predomains that
-preserve the algebraic structure. More formally:
+As before, a morphism of error domains is simply a morphism of the underlying
+predomains that preserves the error element and the $\theta$ map.
+
 %
-\begin{definition}
-Let $B$ and $B'$ be error domains.
-A morphism $\phi : B \wand B'$ is a morphism between the underlying predomains such that:
-\begin{enumerate}
-    \item $\phi(\mho_B) = \mho_{B'}$
-    \item $\phi(\theta_B(\tilde{x})) = \theta_{B'}(\lambda t. \phi(\tilde{x}_t))$
-\end{enumerate}
-\end{definition}
+The definition of monotone relations on predomains is the same as that for
+posets introduced in the previous section, as the bisimilarity relations of the
+predomains are not involved in the definition of monotone relation. Likewise,
+the definition of error domain relations is unchanged.
 %
+The definitions of squares for predomains and error domains are also unchanged
+from the previous section, except of course that the morphisms on the left and
+right, being morphisms of predomains, must also preserve the bisimilarity
+relation.
+
+\begin{comment}
 We define a (monotone) relation on predomains $A$ and $A'$ to be a relation on
@@
 %
-Because of a technical limitation involving the interaction between
+Because of a technical restriction involving the interaction between
 propositional truncation and the later modality, we cannot define the
@@
 underlying predomain relations. Instead, we define composition of error domain
-relations $R$ on $B_1$ and $B_2$ and $S$ on $B_2$ and $B_3$ to be the least
-relation containing $R$ and $S$ that is downward closed under $\le_{B_1}$,
+relations $d$ on $B_1$ and $B_2$ and $d'$ on $B_2$ and $B_3$ to be the least
+relation containing $d$ and $d'$ that is downward closed under $\le_{B_1}$,
 upward-closed under $\le_{B_3}$, respects error, and preserves $\theta$.
@@
     \inferrule*[right = Comp]
-    {b_1 \mathbin{R} b_2 \and b_2 \mathbin{S} b_3}
-    {b_1 \mathbin{R \relcomp S} b_3}
+    {b_1 \mathbin{d} b_2 \and b_2 \mathbin{d'} b_3}
+    {b_1 \mathbin{d \relcomp d'} b_3}
 
     \inferrule*[right = DnClosed]
-    {b_1' \le_{B_1} b_1 \and b_1 \mathbin{R \relcomp S} b_3}
-    {b_1' \mathbin{R \relcomp S} b_3}
+    {b_1' \le_{B_1} b_1 \and b_1 \mathbin{d \relcomp d'} b_3}
+    {b_1' \mathbin{d \relcomp d'} b_3}
 
     \inferrule*[right = UpClosed]
-    {b_1 \mathbin{R \relcomp S} b_3 \and b_3 \le_{B_3} b_3'}
-    {b_1 \mathbin{R \relcomp S} b_3'}
+    {b_1 \mathbin{d \relcomp d'} b_3 \and b_3 \le_{B_3} b_3'}
+    {b_1 \mathbin{d \relcomp d'} b_3'}
 
@@
     { }
-    {\mho_{B_1} \mathbin{R \relcomp S} b_3}
+    {\mho_{B_1} \mathbin{d \relcomp d'} b_3}
 
     \inferrule*[right = PresTheta]
-    {\later_t( \tilde{b_1} \mathbin{R \relcomp S} \tilde{b_3} ) }
-    {\theta_{B_1}(\tilde{b_1}) \mathbin{R \relcomp S} \theta_{B_3}(\tilde{b_3}) }
+    {\later_t( \tilde{b_1} \mathbin{d \relcomp d'} \tilde{b_3} ) }
+    {\theta_{B_1}(\tilde{b_1}) \mathbin{d \relcomp d'} \theta_{B_3}(\tilde{b_3}) }
 \end{mathpar}
@@
 %
-We now describe the squares. Suppose we are given predomains $A_i,
-A_o, A_i'$, and $A_o'$, relations $R_i : A_i \rel A_i'$ and $R_o : A_o \rel
-A_o'$, and morphisms $f : A_i \to A_o, f' : A_i' \to A_o'$ Given a square with
-these morphisms and relations, we say that the square commutes, written $f \le
-f'$, if for all $x \in A_i$ and $x' \in A_i'$ with $x \mathbin{R_i} x'$, we have
-$f(x) \mathbin{R_o} f'(x')$. We make the analogous definition for error domains.
 
-\subsection{Guarded Lift Monad and Free-Forgetful Adjunction}\label{sec:guarded-lift-monad}
+We now describe the squares. Suppose we are given predomains $A_i, A_o, A_i'$,
+and $A_o'$, relations $c_i : A_i \rel A_i'$ and $c_o : A_o \rel A_o'$, and
+morphisms $f : A_i \to A_o, f' : A_i' \to A_o'$ Given a square with these
+morphisms and relations, we say that the square commutes, written $f \le f'$, if
+for all $x \in A_i$ and $x' \in A_i'$ with $(x, x') \in c_i$, we have $(f(x),
+f'(x')) \in c_o$. We make the analogous definition for error domains.
+
+\end{comment}
+
+\subsubsection{Guarded Lift Monad and Free-Forgetful Adjunction}\label{sec:guarded-lift-monad}
 
 % Lift monad
-The guarded error-lift monad $\li$ takes a predomain $A$ to the error domain
-$\li A$. A similar construct was already introduced in Section
-$\ref{sec:lift-monad}$; here we extend it to a monad on \emph{predomains} rather
-than on sets. Given a predomain $A$, the underlying set of $\li A$ is simply the
-guarded lift monad on sets applied to the underlying set of $A$. Recall that
-this has three constructors $\eta : A \to \li A$, $\mho : \li A$, and $\theta :
-\laterhs (\li A) \to \li A$.
+The functor $F$ takes a predomain $A$ to the free error domain on $A$. We first
+define the predomain $UFA$. The underlying set of $UFA$ is defined to be $\li
+|A|$ (the guarded lift monad applied to $|A|$). The ordering relation is the
+lock-step error ordering introduced in the previous section (taking $A = A'$ in
+the definition), and the bisimilarity relation is the weak bisimilarity relation
+on $\li |A|$ also defined in the previous section. With these relations, the
+constructors $\eta$ and $\theta$ of the guarded lift monad are in fact morphisms
+of predomains, i.e., they are monotone and preserve bisimilarity.
 
-The ordering relation $\le_{\li A}$ is defined to be the \emph{lock-step error
-ordering} defined in Section \ref{sec:lock-step-and-weak-bisim}. The
-bisimilarity relation $\bisim_{\li A}$ is defined to be the \emph{weak
-bisimilarity relation} also defined in Section
-\ref{sec:lock-step-and-weak-bisim}. With these relations on $\li
-A$, the constructors $\eta$ and $\theta$ are in fact morphisms of predomains,
-i.e., they are monotone and preserve bisimilarity.
-
-The monad $\li$ decomposes into a free-forgetful adjunction $F \dashv U$. Given
-a predomain $A$, we define the error domain $FA$ whose underlying predomain is
-$\li A$, whose error element is $\mho$, and whose $\theta$ map is the
-constructor $\theta$. The functor $U$ from error domains to predomains simply
-returns the underlying predomain. On morphisms, it returns the underlying
-morphism of predomains.
+We observe that this predomain structure extends to an error domain structure by
+noting that the required error element is given by the constructor $\mho$ and
+the required $\theta$ map is the constructor $\theta$. Lastly, it can be shown
+that the delay morphism $\delta = \theta \circ \nxt$ satisfies $x \bisim
+\delta\, x$ for all $x$ as is required by the definition of error domain.
 
@@
 
-It is easily verified that $\li A$ is the free error- and later-algebra on the
-predomain $A$, so we have that $\li$ is left-adjoint to $U$.
+% The monad $\li$ decomposes into a free-forgetful adjunction $F \dashv U$. Given
+% a predomain $A$, we define the error domain $FA$ whose underlying predomain is
+% $\li A$, whose error element is $\mho$, and whose $\theta$ map is the
+% constructor $\theta$.
+
+% It is easily verified that $\li A$ is the free error- and later-algebra on the
+% predomain $A$, so we have that $\li$ is left-adjoint to $U$.
 
@@
 
-\subsection{Other Constructions}
+\subsubsection{Functors}
+
+The functors $F$, $U$, $\times$, and $\arr$ act on morphisms, relations, and
+squares. Most of these actions have already been described in the previous
+section; the only difference is the added presence of the bisimilarity relation.
+For example, the functor $U$ from error domains to predomains simply returns the
+underlying predomain. On morphisms, it returns the underlying morphism of
+predomains, and on relations and squares it returns the underlying predomain
+relation and predomain square respectively. We define the action of $F$ on a
+relation $c$ to be the lock-step error ordering between $UFA$ and $UFA'$ and
+that it by definition satisfies the additional requirements of being an error
+domain relation between the error domains $FA$ and $FA'$.
+
+\begin{comment}
+
 % internal hom for predomains and error domains
@@
 
-Lastly, given a relation of predomains $R$ between $A$ and $A'$, and a relation
-of error domains $S$ between $B$ and $B'$, we define the relation $R \arr S$
+\end{comment}
+
+\begin{comment}
+Lastly, given a relation of predomains $c$ between $A$ and $A'$, and a relation
+of error domains $d$ between $B$ and $B'$, we define the relation $c \arr d$
 between $A \arr B$ and $A' \arr B'$ in the obvious way, i.e., $f \in A \arr B$
 is related to $g \in A' \arr B'$ iff for all $x \in A$ and $x' \in A'$ with
-$x \mathrel{R} x'$, we have $f(x) \mathrel{S} g(x')$.
+$(x, x') \in c$, we have $(f(x), g(x')) \in Ud$.
 %
@@
 in that it respects error and preserves $\theta$.
+\end{comment}
 
-With all of the above data, we can form a step-1 intensional model of gradual typing
-(See Definition \ref{def:step-1-model}).
+% With all of the above data, we can form a step-1 intensional model of gradual typing
+% (See Definition \ref{def:step-1-model}).
+
+\subsection{Phase 2: Perturbations and Quasi-Representable Relations}
+
+We now introduce the additional definitions needed to complete the construction
+of our model. Recall from Section \ref{sec:adjusting-cast-rules} that in order
+to establish the DnL rule for the downcast of $\iarr$, we needed to adjust the
+rule by inserting a ``delay'' on the right-hand side. A similar adjustment is
+needed for the DnR rule for $\iarr$. Moreover, the need to insert delays impacts
+the semantics of the cast rules for \emph{all} relations, because of the
+functorial nature of casts. That is, the upcast at a function type $c_i \ra c_o$
+involves a downcast in the domain and an upcast in the codomain. This has two
+consequences: first, the squares corresponding to the rules UpL and UpR may
+also require the insertion of a delay. Second, we need to be able to insert
+``higher-order'' delays in a way that follows the structures of the casts.
+
+To accomplish this, we equip every predomain $A$ with a monoid $M_A$ of
+\emph{syntactic perturbations}, as well as a means of \emph{interpreting} these
+perturbations as actual endomorphisms on $A$. Moreover, we want the resulting
+endomorphisms to be extensionally equivalent to the identity function, a
+condition we express by saying that they are weakly bisimilar to the identity
+morphism on $A$. More formally, for every $A$ we require a homomorphism of
+monoids $i_A : M_A \to \{ f : A \to A \mid f \bisim \id_A \}$. We likewise equip
+every error domain $B$ with a monoid $M_B$ and interpretation homomorphism
+$i_B$. These concepts deserve their own definitions: We call the triple $(A,
+M_A, iA)$ a \textbf{value object} as it will be the denotation of value types in
+our final model. Likewise, the triple $(B, M_B, i_B)$ will be called a
+\textbf{computation object}.
+
+We must now impose an additional condition on predomain relations $c$ and error
+domain relations $d$ that specify how they interact with these perturbations. In
+particular, we need to be able to ``push'' and ``pull'' perturbations along
+relations $c$ and $d$. The intuition for this requirement comes from the
+construction of the square corresponding to the original UpL rule from the
+square for the simplified version as was shown in Section
+\ref{sec:modeling-term-precision}. Recall that we horizontally composed the
+simplified UpL for $c$ square with the identity square for $c'$ on the right.
+Now that the square for UpL involves a delay, this construction doesn't work
+unless we can ``push'' the delay on $A_2$ to a delay on $A_3$. That is, given a
+relation $c : A \rel A'$ and a perturbation $m_A \in M_A$, we need to be able to
+turn it into a perturbation $m_{A'} \in M_{A'}$ such that the resulting
+endomorphisms obtained by the respective homomorphisms $i_A$ and $i_{A'}$ form a
+square. This is made formal by the following definition:
+%
+\begin{definition}
+    Let $(A, M_A, i_A)$ and $(A', M_{A'}, and i_{A'})$ be value objects, and let
+    $c : A \rel A'$ be a predomain relation. A \emph{push-pull structure} for
+    $c$ consists of a homomorphism $\push : M_A \to M_A'$ and $\pull : M_A' \to
+    M_A$ such that for all $m_A \in M_A$ there is a square $i_A(m_A) \ltsq{c}{c}
+    i_{A'}(\push\, m_A)$ and for all $m_{A'} \in M_{A'}$ there is a square
+    $i_A(\pull\, m_{A'}) \ltsq{c}{c} i_{A'}(m_{A'})$.
+\end{definition}
+%
+We make the analogous definition for computation objects.
+
+With the notion of perturbation defined, we can now discuss the final aspect of
+the model, namely the representability of relations.
+%
+\begin{definition}[quasi-left-representable relations]
+Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let $c : A
+\rel A'$ be a predomain relation. We say that $c$ is
+\emph{quasi-left-representable} by a predomain morphism $e : A \to A'$ if there
+are perturbations $\delle_c \in M_A$ and $\delre_c \in M_{A'}$ such that the
+following two squares exist: (UpL) $e \ltsq{c}{r(A')} i_{A'}(\delre_c)$, and
+(UpR) $i_A(\delle_c) \ltsq{r(A)}{c} e$.
+\end{definition}
+%
+Observe that this generalizes the previous notion of representability, since
+under that definition the perturbations were always the identity.
+%
+We make the analogous definition of quasi-left-representability for error domain
+relations $d$.
+%
+Likewise, we define quasi-right-representability as follows:
+%
+\begin{definition}
+Let $(B, M_B, i_B)$ and $(B', M_{B'}, i_{B'})$ be computation objects, and let
+$d : B \rel B'$ be an error domain relation. We say that $d$ is
+\emph{quasi-right-representable} by an error domain morphism $p : B' \to B$ if
+there are perturbations $\dellp_d \in M_B$ and $\delrp_d \in M_{B'}$ such that
+the following two squares exist: 
+(DnL) $p \ltsq{r(B')}{d} i_{B'}(\delrp_d)$ and
+(DnR) $i_B(\dellp_d) \ltsq{d}{r(B)} p$.
+\end{definition}
+%
+We make the analogous definition of quasi-right-representability for predomain
+relations $c$.
+
+Although it makes sense to refer to the quasi-representability of an arbitrary
+predomain or error domain relation, quasi-representable relations do not in
+general compose. In order to construct the needed squares for the composition of
+two relations, we require that the relations have push-pull structures.
+
+We can now give the final definition of value and computation relations for our
+model:
+%
+\begin{definition}[value relations]
+A \textbf{value relation} between value objects $(A, M_A, i_A)$ and $(A', M_{A'},
+i_{A'})$ is a predomain relation $c : A \rel A'$ that has a push-pull structure
+is quasi-left-representable by a morphism $e_c : A \to A'$, and is such that
+$Fc$ is quasi-right-representable by an error domain morphism $p_c : FA' \to
+FA$.
+\end{definition}
+
+\begin{definition}[computation relations]
+A \textbf{computation relation} between computation objects $(B, M_B, i_B)$ and $(B',
+M_{B'}, i_{B'})$ is an error domain relation $d$ that has a push-pull structure,
+is quasi-right-representable by a morphism $p_d : B' \to B$, and is such that
+$Ud$ is quasi-left-representable by a morphism $e_d : UB \to UB'$
+\end{definition}
+
+Because we have augmented the definitions of object and relation for our model,
+we need to specify the actions of the functors $F$, $U$, $\times$, and $\arr$ on
+these objects and relations. For instance, the functor $F$ now acts on
+\emph{value objects} $(A, M_A, i_A)$, which means we need to specify the monoid
+$M_{FA}$ and interpretation $i_{FA}$ corresponding to the error domain $FA$. We
+leave these constructions to the appendix, as they are technical and not crucial
+to understanding our main result.
 
 \subsection{The Dynamic Type}
-The predomain representing the dynamic type will be defined using guarded recursion
-as the solution to the equation
-\footnote{In this section, we write $F$ instead of $\li$ so that the notation follows that of the abstract model section.}
+Lastly, we describe the denotation of the dynamic type. The predomain
+representing the dynamic type is defined using guarded recursion as the solution
+to the equation
+% \footnote{In this section, we write $F$ instead of $\li$ so that the notation follows that of the abstract model section.}
 %
@@
 
-For the sake of clarity, we name the ``constructors" $\text{nat}$, $\text{times}$,
-and $\text{fun}$, respectively.
+For the sake of clarity, we name the ``constructors" $\text{nat}$,
+$\text{times}$, and $\text{fun}$, respectively.
 %
-We define $e_\mathbb{N} : \mathbb{N} \to D$ to be the injection into the first
-component of the sum, and $e_\times : D \times D \to D$ to be the injection into
-the second component of the sum, and $e_\to : U(D \arr F D)$ to be the morphism
-$\nxt$ followed by the injection into the third component of the sum.
+We define $e_\mathbb{N} : \mathbb{N} \to D$ to be the $\text{nat}$ constructor,
+$e_\times : D \times D \to D$ to be $\text{times}$, and $e_\to : U(D \arr F D)$
+to be the morphism $\nxt$ followed by $\text{fun}$.
 
@@
 %
-We define a relation $\inat : \mathbb{N} \rel D$ by
-$(n, d) \in \inat$ iff $e_\mathbb{N}(n) \le_D d$.
-We similarly define $\itimes : D \times D \rel D$ by
-$((d_1, d_2), d) \in \itimes$ iff $e_\times(d_1, d_2) \le_D d$,
-and we define $\text{inj}_\to : U(D \arr F D) \rel D$ by
-$(f, d) \in \iarr$ iff $e_\to(f) \le_D d$.
+We define a relation $\inat : \mathbb{N} \rel D$ by $(n, d) \in \inat$ iff
+$e_\mathbb{N}(n) \le_D d$. We similarly define $\itimes : D \times D \rel D$ by
+$((d_1, d_2), d) \in \itimes$ iff $e_\times(d_1, d_2) \le_D d$, and we define
+$\text{inj}_\to : U(D \arr F D) \rel D$ by $(f, d) \in \iarr$ iff $e_\to(f)
+\le_D d$.
 
-Now we define the perturbations for $D$.
-Recall from our construction of a model with perturbations
-(Section \ref{sec:constructing-perturbations}) that for each value
-type $A$ we associate a monoid $P_A$ of perturbations
-and a homomorphism into the monoid of endomorphisms bisimilar to the identity,
-and likewise for computation types.
-We define the perturbations for $D$ via least-fixpoint in the category of monoids as
+Now we define the perturbations for $D$. Recall that for each value type $A$ we
+associate a monoid $P_A$ of perturbations and a homomorphism into the monoid of
+endomorphisms bisimilar to the identity, and likewise for computation types. We
+define the perturbations for $D$ via least-fixpoint in the category of monoids
+as
 %

### 2024-07-09 paper-new/concrete-relational-model.tex
[paper] revise sections on concrete relational model

@@
 the elements they step to are related.}
+%
+Morphisms of error domains are morphisms of the underlying partially-ordered
+sets that preserve the error element and $\theta$ map, as was the case in the
+previous section.
 
+\subsubsection{Modelling Type Precision}
 We model type precision $c : A \ltdyn A'$ as a \emph{monotone relation}, i.e., a
 relation $c$ that is upward-closed under the relation on $A'$ and
-downward-closed under the relation on $A$. The relation on the poset $A$ is
-denoted $r(A)$.
+downward-closed under the relation on $A$. We denote such a relation between $A$
+and $A'$ by $c : A \rel A'$. The relation on the poset $A$ is denoted $r(A)$.
 % This extends to products $c_1 \times c_2$ in the obvious way.
+Composition of relations on predomains is the usual relational composition
+(which is truncated to be propositional).
+
+We likewise define a notion of \emph{error domain relation} between error
+domains $B$ and $B'$. An error domain relation $d : B \rel B'$ is a monotone
+relation on the underlying sets that preserves error and respects $\theta$. More
+formally:
+
+\begin{definition}
+  Let $B$ and $B'$ be error domains. An \emph{error domain relation} between
+  $B$ and $B'$ is a monotone relation $d$ on the underlying partially-ordered sets such that:
+  \begin{enumerate}
+     \item (Respects error): For all $y \in B'$, we have $(\mho_B, y) \in d$.
+     \item (Preserves $\theta$): For all $\tilde{x}$ in $\laterhs B$ and
+     $\tilde{y} \in \laterhs B'$, if $\laterhs_t ((\tilde{x}_t, \tilde{y}_t) \in
+     d)$, then $(\theta_B (\tilde{x}), \theta_{B'} (\tilde{y})) \in d$. 
+  \end{enumerate}
+\end{definition}
+
+Because of a technical restriction involving the interaction between
+propositional truncation and the later modality, we cannot define the
+composition of error domain relations as simply the composition of their
+underlying predomain relations. Instead, we define composition of error domain
+relations $d$ on $B_1$ and $B_2$ and $d'$ on $B_2$ and $B_3$ to be the least
+relation containing $d$ and $d'$ that is downward closed under $\le_{B_1}$,
+upward-closed under $\le_{B_3}$, respects error, and preserves $\theta$.
+Specifically, it is defined inductively by the following rules:
 %
-Likewise, we define a notion of \emph{computation relation} on computation types
-$B$ and $B'$. A computation relation $d : B \rel B'$ is a monotone relation on
-the underlying sets that preserves error and respects $\theta$. That is, $\mho_B
-\mathbin{d} y$ for all $y : B'$ and if $\laterhs_t (\tilde{x}_t \mathbin{d}
-\tilde{y}_t)$, then $\theta_B (\tilde{x}) \mathbin{d} \theta_{B'} (\tilde{y})$.
-The functors $U$ and $F$ introduced in the previous section extend to relations,
-The functor $U$ simply takes the underlying monotone relation of a computation
-relation; we will give the definition of $F$ on relations later in this section.
+\begin{mathpar}
+    \inferrule*[right = Comp]
+    {b_1 \mathbin{d} b_2 \and b_2 \mathbin{d'} b_3}
+    {b_1 \mathbin{d \relcomp d'} b_3}
 
-Before moving on, we introduce the notion of \emph{square}, which specifies
-a relation between two monotone functions. The definition is as follows. Let
-$A_i, A_o, A_i', A_o'$ be partially-ordered sets. Let $c_i : A_i \rel A_i'$ and
-$c_o : A_o \rel A_o'$ be monotone relations, and let $f : A_i \to A_o$ and $g :
-A_i' \to A_o'$ be monotone functions. We say that $f \ltsq{c_i}{c_o} g$ if for
-all $x : A_i$ and $y : A_i'$ with $x \binrel{c_i} y$, we have $f(x) \binrel{c_o}
-g(y)$. We visualize this situation as the following square:
+    \inferrule*[right = DnClosed]
+    {b_1' \le_{B_1} b_1 \and b_1 \mathbin{d \relcomp d'} b_3}
+    {b_1' \mathbin{d \relcomp d'} b_3}
+
+    \inferrule*[right = UpClosed]
+    {b_1 \mathbin{d \relcomp d'} b_3 \and b_3 \le_{B_3} b_3'}
+    {b_1 \mathbin{d \relcomp d'} b_3'}
+
+    \inferrule*[right = PresErr]
+    { }
+    {\mho_{B_1} \mathbin{d \relcomp d'} b_3}
+
+    \inferrule*[right = PresTheta]
+    {\later_t( \tilde{b_1} \mathbin{d \relcomp d'} \tilde{b_3} ) }
+    {\theta_{B_1}(\tilde{b_1}) \mathbin{d \relcomp d'} \theta_{B_3}(\tilde{b_3}) }
+\end{mathpar}
+
+% That is,
+% $(\mho_B, y) \in d$ for all $y : B'$ and if $\laterhs_t ((\tilde{x}_t,
+% \tilde{y}_t) \in d)$, then $(\theta_B (\tilde{x}), \theta_{B'} (\tilde{y})) \in
+% d$.
+
+The functors $U$, $F$, $\times$, and $\arr$ introduced in the previous
+section extend to relations. The functor $U$ simply takes the underlying
+monotone relation of an error domain relation. Given a relation $c_1 : A_1 \rel
+A_1'$ and $c_2 : A_2 \rel A_2'$, the relation $c_1 \times c_2 : (A_1 \times A_2)
+\rel (A_1' \times A_2')$ is defined by $((x, y), (x', y')) \in (c_1 \times c_2)$
+if and only if $(x, x') \in c_1$ and $(y, y') \in c_2$. Given a relation $c : A
+\rel A'$ and an error domain relation $d : B \rel B'$, the relation $c \arr d :
+(A \arr B) \rel (A' \arr B')$ is the error domain relation defined by $(f, g)
+\in (c \arr d)$ if and only if for all $x : A$ and $x' : A'$ with $(x , x') \in
+c$ we have $(f(x), g(x')) \in Ud$. The fact that this relation preserves error
+and respects $\theta$ follows from the corresponding properties for $d$.
+%
+We we will postpone the description of the action of $F$ on relations to later
+in this section.
+
+\subsubsection{Modelling Term Precision}\label{sec:modeling-term-precision}
+To model term precision, we need the notion of \emph{square}, which specifies a
+relation between two monotone functions. The definition is as follows. Let $A_i,
+A_o, A_i', A_o'$ be partially-ordered sets. Let $c_i : A_i \rel A_i'$ and $c_o :
+A_o \rel A_o'$ be monotone relations, and let $f : A_i \to A_o$ and $g : A_i'
+\to A_o'$ be monotone functions. We say that $f \ltsq{c_i}{c_o} g$ if for all $x
+: A_i$ and $y : A_i'$ with $(x, y) \in c_i$, we have $(f(x), g(y)) \in c_o$. We
+visualize this situation as the following square:
 
@@
 \ltsq{r(A_i)}{r(A_o)}$ by the definition of monotone function.
+We likewise make the analogous definition of square for error domains.
 
-The cast rules UpL/UpR/DnL/DnR specify a relationship between the semantics of
+Having defined squares, we now discuss the interpretation of the cast rules
+UpL/UpR/DnL/DnR. These rules specify a relationship between the semantics of
 type precision and the corresponding casts (see Section \ref{sec:GTLC}). In
@@
 The existence of these simpler squares for a relation can be stated without
-reference to other relations. Thus we can require that these squares exist in
+reference to other relations. Thus, we can require that these squares exist in
 the definition of relation without running into the circularity issue we faced
@@
 
-Finally, we model the term precision ordering $\Gamma \ltdyn \Gamma' \vdash M \ltdyn
-N : A \ltdyn A'$ as an ordering relation between the monotone functions denoted
-by $M$ and $N$, as shown below:
+Finally, we use squares to model the term precision ordering $\Gamma \ltdyn
+\Gamma' \vdash M \ltdyn N : A \ltdyn A'$ as an ordering relation between the
+monotone functions denoted by $M$ and $N$:
 
@@
 %
-\[ l_1 \leq l_2 := 
+\[ l_1 \leq^{\bisim} l_2 := 
   \Sigma_{l_1', l_2' \in \li \mathbb{N}}. l_1 \bisim l_1' \ltls l_2' \bisim l_2. \]
@@
 
-\subsection{Adjusting the Cast Rules}
+\subsection{Adjusting the Cast Rules}\label{sec:adjusting-cast-rules}
 
@@
 
-So far, we have been focusing on closed terms of base type, for which the above
-techniques give a semantic interpretation of term precision. We now consider how
-to extend these constructions to \emph{all} types. In particular, in order to
-model higher-order data types (i.e., functions) we need to equip \emph{every}
-semantic object with not only a partial ordering relation but also a
-``bisimilarity'' relation that is reflexive and symmetric. We must similarly
-equip every object with a structured set of delays that can be inserted to
-ensure the appropriate cast rules hold at all higher-order types. For example,
-the upcast for a derivation $c \ra d$ involves the downcast corresponding to $c$
-and an upcast corresponding to $d$. We must be able to insert delays in a
-functorial manner to mimic the structure of the casts. In the next section, we
+The above techniques give a semantic interpretation of term precision for closed
+terms of base type. We now consider how to extend these constructions to
+potentially-open terms of \emph{all} types. In particular, in order to model
+higher-order data types (i.e., functions) we need to equip \emph{every} semantic
+object with not only a partial ordering relation but also a ``bisimilarity''
+relation that is reflexive and symmetric. We must similarly equip every object
+with a structured set of delays that can be inserted and manipulated to ensure
+the appropriate cast rules hold at all higher-order types. For example, the
+upcast for a derivation $c_i \ra c_o$ involves the downcast corresponding to
+$c_i$ and the upcast corresponding to $c_o$. We must be able to insert delays in
+a functorial manner to mimic the structure of the casts. In the next section, we
 will make the necessary definitions and describe the relevant constructions.

### 2024-07-09 paper-new/concrete-term-model.tex
[paper] revise first section on relational model
cast semantics update

@@
 
-\eric{Wording?}
-%
-In the present section, we use the techniques of guarded type theory to give a
-set-theoretic denotational semantics to the terms of the gradually-typed lambda
-calculus introduced in Section \ref{sec:GTLC}. This serves two important
-purposes: First, it acts as a warm-up for employing guarded type theory to
-define a denotational semantics. This acts as a stepping stone into the next
-section, where we extend the techniques employed here to accommodate the type
-and term precision orderings of the gradually-typed lambda calculus. Second,
-constructing this semantic model establishes the validity of the $\beta$ and
-$\eta$ principles for the gradually-typed lambda calculus. 
-%
+As a first step to giving a well-behaved semantics of GTLC, in this
+section we define a simpler semantics that validates all of our
+desired properties except for graduality: that is it is adequate and
+validates $\beta\eta$ equality. Our relational semantics will then be
+a refinement of this simpler design. In the process we will introduce
+some of the technical tools we will use in the refined semantics:
+guarded type theory and call-by-push-value. While these tools can be
+somewhat technical we emphasize that due to the use of synthetic
+guarded domain theory, this initial semantics is essentially just a
+``na\"\i ve'' denotational semantics of an effectful language arising
+from a monad on sets. But since we are working in the non-standard
+foundation of guarded type theory our notion of sets allows for the
+construction of more sophisticated sets and monads than classical
+logic allows.
+
 % In Section \ref{sec:towards-relational-model}, we will discuss how to extend the
@@
 
-We begin with a brief overview of synthetic guarded domain theory, which we will
+\max{TODO: discuss this wording with Eric. Are we working in guarded
+  type theory, SGDT or Ticked Cubical Type Theory in the informal work
+  in the paper?}
+We begin with a brief overview of the background theory that we use in this work.
+
+synthetic guarded domain theory, which we will
 use to construct the model. \emph{Synthetic guarded domain theory}, or SGDT for
@@
 
-For the sake of modularity and ease of description, we will formulate the model
-for the gradually-typed lambda calculus using the \emph{call-by-push-value}
-paradigm \cite{levy99}. In a call-by-push-value model, there is a
-category $\calV$ of \emph{value types} and a category $\calE$ of
-\emph{computation types}. Value types represent static values, while computation
-types represent effectful terms. We denote value types by $A, A', A_1, A_i,
-A_o$, etc, and computation types by $B, B', B_1, B_i, B_o$, etc. There is a pair
-of adjoint functors $F : \calV \to \calE$ and $U : \calE \to \calV$ with $F
-\dashv U$, and a functor $\arr : \calV^{op} \times \calE \to \calE$ such that
-$U(A \arr B) \cong A \To UB$ where $\To$ denotes the internal hom in $\calV$.
+Our denotational model of GTLC must account for the fact that the
+language is call-by-value with non-trivial effects (divergence and
+errors).  To do this, we formulate the model using the categorical
+structures of Levy's \emph{call-by-push-value}
+\cite{levy99}. Call-by-push-value is a categorical model of effects
+that is a refinement of Moggi's monadic semantics
+\cite{moggi-monads}. While Moggi works with a (strong) monad $T$ on a
+category $\mathcal V$ of pure morphisms, call-by-push-value works
+instead with a decomposition of such a monad into an adjunction
+between a category $\mathcal V$ of ``value types'' and pure morphisms
+and a category $\mathcal E$ of ``computation types'' and linear
+morphisms. The monad $T$ on $\calV$ is decomposed into a left adjoint
+$F : \calV \to \calE$ constructing the type $F A$ of computations that
+return $A$ values and and right adjoint $U : \calE \to \calV$
+constructing the type $U B$ of values that are first-class suspended
+computations of type $B$. The monad $T$ can then be reconstituted as
+$T = UF$ but the main advantage is that many constructions on
+effectful programs naturally decompose into constructions on
+value/computation types. Most strikingly, the CBV function type $A
+\rightharpoonup A'$ decomposes into the composition of three
+constructions $U(A \to F A')$ where $\arr : \calV^{op} \times \calE
+\to \calE$ is a kind of mixed kinded function type. This decomposition
+considerably simplifies treatment of the CBV function
+type. Additionally, prior work on gradual typing semantics has shown
+that the casts of gradual typing are naturally formulated in terms of
+pairs of a pure morphism and a linear morphism, as we explain below
+\cite{newlicataahmed}.
 
-We embed our call-by-value gradually-typed cast calculus into a
-call-by-push-value model as follows. The value types will be sets, and
-computation types will be sets with additional algebraic structure corresponding
-to the possible behaviors of a gradually-typed program. One such behavior is
-\emph{failure}: a program may fail at run-time because of a type error. In
-addition to this, a program may fail to terminate. This is possible because the
-dynamic type includes function types and so is recursive. We will discuss the
-dynamic type further in a subsequent section, but for now it suffices to know
-that we can express a diverging computation in our syntax. Therefore, it will be
-necessary to track the ``steps'' taken by a term; we can regard this as a way of
-intensionally modelling \emph{partiality}.
+For our simple denotational semantics, the category $\calV$ is simply
+the category $\Set$ of sets and functions. Then because GTLC is
+call-by-value, all types will be interpreted as objects of this
+category, i.e., sets. More interesting are the computation
+types. These will be interpreted in a category of sets equipped with
+``algebraic''\footnote{the $\theta$ structure is not algebraic in the
+strictest sense since it does not have finite arity, but } structure
+with morphisms being functions that are homomorphic in this structure.
+This algebraic structure needs to model the effects present in our language,
+in this case these are errors as well as taking computational steps
+(potentially diverging by taking computational steps forever).
+%
+We call these algebraic structures simple error domains to distinguish
+them from the error domains we define later to model graduality:
+\begin{definition}[Simple Error Domains]
+  A (simple) error domain $B$ consists of
+  \begin{enumerate}
+  \item A carrier set $UB$
+  \item An element $\mho_{B} : UB$ representing error
+  \item A function $\theta_B : \laterhs UB \to UB$ modeling a computational step
+  \end{enumerate}
+  A homomorphism of error domains $\phi : B \multimap B'$ is a
+  function $\phi : UB \multimap UB'$ that preserves these in that
+  $\phi(\mho_B) = \mho_{B'}$ and $\phi(\theta_{B}(\tilde{x})) =
+  \theta_{B'}(\later (\phi(\tilde{x})))$.  Simple error domains and
+  homomorphisms assemble into a category $\ErrDom$ with a forgetful
+  functor $U : \ErrDom \to \Set$ taking the underlying set and
+  function.
+\end{definition}
+Here is our first point where we utilize guarded type theory: rather
+than simply being a function $UB \to UB$, the ``think'' map $\theta$
+takes an element \emph{later}. This makes a major difference, because
+the structure of a think map combined with the guarded fixpoint
+operator allows us to define recursive elements of $UB$ in that any
+function $f : UB \to UB$ has a ``quasi-fixed point'' $\textrm{qfix}(f)
+= \fix(f \circ \theta_B)$ satisfying the quasi-fixed point property:
+\[ \qfix(f) = f(\delta_B(\qfix f)) \]
+where $\delta_B = \theta_B \circ \nxt$ is a map we call the ``delay''
+map which trivially delays an element now to be available later.  As
+an example, we can define $\Omega_B = \qfix(\id) : UB$, the
+``diverging element'' that ``thinks forever'' in that $\Omega_B =
+\delta(\Omega_B)$. We call this a ``quasi'' fixed point because it is
+\emph{nearly} a fixed point except for the presence of the delay map
+$\delta_B$, which is irrelevant from an extensional point of view
+where we would prefer to ignore differences in the number of steps
+that computations take.
 
-More concretely, the computation types are defined to be sets $B$ that are
-equipped with a distinguished error element $\mho_B$ and a map $\theta_B \colon
-\laterhs B \to B$. We call such sets \emph{error domains}. A \emph{morphism of
-error domains} from $B_1$ to $B_2$ is a function $\phi$ on the underlying sets
-that respects the error and $\theta$ maps, i.e., for which 
-%
-(1) $\phi(\mho_{B_1}) = \mho_{B_2}$ and
-%
-(2) $\phi(\theta_{B_1}(\tilde{x})) = \theta_{B_2}(\later (\phi(\tilde{x})))$
-(using the functorial action of $\later$).
-
-The functor $U : \errdom \to \Set$ simply takes the underlying set of the error
-domain. The functor $F : \Set \to \errdom$ is defined to be the left adjoint to
-$U$ and constructs the \emph{free error domain} on a set $A$.
-%
+\max{move this}
 The functor $\arr : \Set^{op} \times \errdom \to \errdom$ is defined on objects
@@
 
-\subsection{The Lift + Error Monad}\label{sec:lift-monad}
-
-We describe a monad that decomposes into the free-forgetful adjunction
-between sets and error domains. We base the construction on the \emph{guarded
+Then to model effectful programs, we need to construct a left adjoint
+$\li : \Set \to \ErrDom$ to $U$ which constructs the ``free error
+domain'' on a set, which models an effectful CBV computation.
+%% \subsection{The Lift + Error Monad}\label{sec:lift-monad}
+We base the construction on the \emph{guarded
 lift monad} described in \cite{mogelberg-paviotti2016}. Here, we augment the
-guarded lift monad to accommodate the additional effect of failure. For a type
-$A$, we define the \emph{guarded lift monad with failure} $\li A$, which we will
-just call the \emph{guarded lift monad} or simply the \emph{lift monad}.
-%
-\begin{align*}
-  \li A :=\quad
-  &\alt \eta \colon A \to \li A \\
-  &\alt \mho \colon \li A \\
-  &\alt \theta \colon \laterhs (\li A) \to \li A
-\end{align*}
-%
-Unless otherwise mentioned, all constructs involving $\later$ or $\fix$ are
-understood to be with respect to a fixed clock $k$. So for the above, we really
-have for each clock $k$ a type $\li^k A$ with respect to that clock.
-%
-Formally, the lift monad $\li A$ is defined as the solution to the guarded
-recursive type equation
-%
-\[ \li A \cong A + 1\, + \laterhs \li A. \]
+guarded lift monad to accommodate the additional effect of failure.
+\begin{definition}[Free error domain]
+  For a set $A$, we define the (carrier of) \emph{free error domain} $U(\li A)$ as the unique solution to the guarded domain equation:
+  \[ U(\li A) \cong A + 1\, + \laterhs U(\li A). \]
+  For which we use the following notation for the three constructors:
+  \begin{enumerate}
+  \item $\eta \colon A \to U(\li A)$
+  \item $\mho \colon U(\li A)$
+  \item $\theta \colon \laterhs U(\li A) \to U(\li A)$
+  \end{enumerate}
+  Here $\mho, \theta$ provide the error domain structure for $\li A$,
+  and this has the universal property of being free in that any
+  function $f : U(\li A) \to UB$ uniquely extends to a homomorphism
+  $f^\dagger : \li A \multimap B$ satisfying $Uf^\dagger \circ \eta =
+  f$. In other words, the free error monad is left adjoint $\li \dashv
+  U$ to the forgetful functor, with $\eta$ the unit of the adjunction.
+\end{definition}
+%% %
+%% Unless otherwise mentioned, all constructs involving $\later$ or $\fix$ are
+%% understood to be with respect to a fixed clock $k$. So for the above, we really
+%% have for each clock $k$ a type $\li^k A$ with respect to that clock.
+%% %
+%% Formally, the lift monad $\li A$ is defined as the solution to the guarded
+%% recursive type equation
+%% %
 %
@@
 return a value (via $\eta$), (2) raise an error and stop (via $\mho$), or (3)
-take a step (via $\theta$).
+take a computational step (via $\theta$).
 %
-Notice there is a computation $\fix\, \theta$ of type $\li A$. This represents a
-computation that runs forever and never returns a value.
-%
-Since we claimed that $\li A$ is a monad, we need to define the monadic
-operations and show that they respect the monadic laws. The return is just
-$\eta$, and given $f : A \to \li A'$, the monadic extension $\text{ext}(f)
-\colon (\li A \to \li A')$ is defined via guarded recursion by cases on the
-input value of type $\li A$.
+The operation $f^\dagger$ can be defined by guarded recursion in the
+apparent way, and proven to satisfy the freeness property by L\"ob
+induction.
 % It is instructive to give at least one example of a use of guarded recursion, so
@@
 %
-Verifying that the monadic laws hold uses \lob-induction and is straightforward.
-%
-% TODO mention that error domains are algebras of the lift monad?
-We observe that the guarded lift monad applied to a set $A$ is the underlying
-set of the free error domain on $A$ $A$, i.e., we have $\li = UF$ where $F :
-\Set \to \errdom$ and $U : \errdom \to \Set$ and $F \dashv U$. Given a set $A$
-and an error domain $B$, the set of functions $A \to UB$ is naturally isomorphic
-to the set of error domain morphisms $FA \to B$.
+%% Verifying that the monadic laws hold uses \lob-induction and is straightforward.
+
+%% % TODO mention that error domains are algebras of the lift monad?
+%% We observe that the guarded lift monad applied to a set $A$ is the underlying
+%% set of the free error domain on $A$ $A$, i.e., we have $\li = UF$ where $F :
+%% \Set \to \errdom$ and $U : \errdom \to \Set$ and $F \dashv U$. Given a set $A$
+%% and an error domain $B$, the set of functions $A \to UB$ is naturally isomorphic
+%% to the set of error domain morphisms $FA \to B$.
+
+%% Lastly, we define the function $\delta : \li A \to \li A$ by $\delta = \theta_A
+%% \circ \nxt$.
 
@@
 
-\subsection{Modeling the Dynamic Type}\label{sec:dynamic-type}
+The free error domain is essential to modeling cbv computation: a term
+$\Gamma \vdash M : A$ in GTLC will be modeled as a function $M :
+\sem{\Gamma} \to U\li \sem{A}$ from the set denoted by $\Gamma$ to the
+underlying set of the free error domain on the set denoted by $A$,
+just as in a monadic semantics with monad $U\li$.
 
-We now discuss how to model the dynamic type $\dyn$.
-The goal is to define a set $D$ that satisfies the isomorphism
+\subsection{Modeling the type structure}\label{sec:dynamic-type}
+Much of the type structure of GTLC is simple to model: $\nat$ denotes
+the set of natural numbers and $\times$ denotes the cartesian product
+of sets. To model the CBV function type $A \rightharpoonup A'$ we use
+the CBPV decomposition $U(\sem{A} \to \li \sem{A'})$ where $A \to B$
+is the obvious point-wise algebraic structure on the set of functions
+from $A$ to $UB$.
+
+The final type to model is the the dynamic type $\dyn$.  In the style
+of Scott, a classical domain theoretic model for $\dyn$ would be given
+by solving a domain equation
 %
-\[ D \cong \Nat + (D \times D) + (D \to \li D). \]
+\[ D \cong \Nat + (D \times D) + U(D \to \li D). \]
 %
-where the use of $\li D$ in the codomain corresponds to the fact that when a
-value of type $\dyn$ is a function, the function when called may return an error
-or take a step of computation. This equation does not have inductive or
-coinductive solutions. The usual way of dealing with such equations is via
-domain theory, by which we can obtain an exact solution. However, by using
-guarded recursion we can solve the equation and obtain a guarded solution in a
+representing the fact that dynamically typed values can be numbers,
+pairs or CBV functions. This equation does not have inductive or
+coinductive solutions due to the negative occurrence of $D$ in the
+domain of the function type. In classical domain theory this is
+avoided by instead taking the initial algebra in a category of
+embedding-projection pairs, by which we obtain an exact solution to
+this equation. However, in guarded domain theory we instead replace this domain equation with a guarded domain equation.
 process analogous to giving a normal set-theoretic denotational semantics.
@@
 %
-\[ D \cong \Nat + (D \times D)\, + \laterhs (D \to \li D). \]
+\[ D \cong \Nat + (D \times D)\, + \laterhs U(D \to \li D). \]
 %
-Since the negative occurrence of $D$ is guarded under a later, this equation has
-a solution constructed via a combination of least fixed-point and guarded
-recursion. Specifically, consider the parameterized inductive type
+Since the negative occurrence of $D$ is guarded under a later, we can
+solve this equation by a mixture of inductive types and guarded fixed
+points. Specifically, consider the parameterized inductive type
 %
@@
 %
-Now consider the function $f$ defined by
+Then we can construct $D$ as the unique solution to the guarded domain equation
+\[ D \cong D'(\laterhs U(D \to \li D)) \]
+which is a guarded equation in that all occurrences of $D$ on the right hand side are guarded by the $\laterhs$.
 %
-\[ \lambda (\tilde{D} \colon \laterhs \type) . D' (\laterhs_t (\tilde{D}_t \to (\li (\tilde{D}_t)))). \]
+Expanding out the guarded fixed point and least fixed point property
+gives us that $D$ satisfies our desired equation above. \max{give
+  that equation a number}.
 %
-Recall that the tick $t : \tick$ is evidence that time has passed, and since
-$\tilde{D}$ has type $\later \type$, i.e. $\tick \to \type$, then $\tilde{D}_t$
-has type $\type$.
-%
-Finally, define 
-\[ D := \fix f. \]
-%
-Observe that the unfolding equality characterizing the guarded fixpoint implies
-that $D = D' (\later D \to \li D)$, and by the definition of $D'$ as a least
-fixpoint we see that this is equal to $\Nat + (D \times D) + \later (D \to \li
-D)$.
+We then refer to the three injections for numbers, pairs and functions
+as $\inat, \itimes, \iarr$ respectively.
 
@@
 
-We can now give a semantics to the types and terms of the gradually-typed lambda
-calculus. 
+We now extend our semantics of types to a semantics of terms of
+GTLC. As mentioned above, terms $x:A_1,\ldots \vdash M : A$ are
+modeled as effectful functions $\sem{M}: \sem{A_1}\times \cdots \to U\li\sem{A}$
+and values $x:A_1,\ldots \vdash V : A$ are also be modeled as pure
+functions $\sem{V} : \sem{A_1}\times\cdots \to \sem{A}$.
 %
-Much of the semantics is similar to a normal call-by-value denotational
-semantics, so we focus only on the cast semantics. We interpret types as sets;
-the interpretation of types is given in Figure \ref{fig:type-interpretation}.
-Contexts $\Gamma = x_1 \colon A_1, \dots, x_n \colon A_n$ are interpreted as the
-product $\sem{A_1} \times \cdots \times \sem{A_n}$. The semantics of the dynamic
-type $\dyn$ is the set $D$ introduced in Section \ref{sec:dynamic-type}. The
-product type $A \times A'$ is interpreted as the Cartesian product of the
-denotations of $A$ and $A'$. The function type $A \ra A'$ is interpreted as the
-set of functions from $\sem{A}$ to $\li (\sem {A'})$.
+The interpretation of the typical CBV features of GTLC into a CBPV is
+standard \cite{levy}, the only additional part we need to account for
+is the semantics of casts, shown in Figure \ref{fig:term-semantics}.
 %
-The interpretation of a value $\hasty {\Gamma} V A$ is a function from
-$\sem{\Gamma}$ to $\sem{A}$. Likewise, a term $\hasty {\Gamma} M {{A}}$ is
-interpreted as a function from $\sem{\Gamma}$ to $\li \sem{A}$.
+%% Much of the semantics is similar to a normal call-by-value denotational
+%% semantics, so we focus only on the cast semantics. We interpret types as sets;
+%% the interpretation of types is given in Figure \ref{fig:type-interpretation}.
+%% Contexts $\Gamma = x_1 \colon A_1, \dots, x_n \colon A_n$ are interpreted as the
+%% product $\sem{A_1} \times \cdots \times \sem{A_n}$. The semantics of the dynamic
+%% type $\dyn$ is the set $D$ introduced in Section \ref{sec:dynamic-type}. The
+%% product type $A \times A'$ is interpreted as the Cartesian product of the
+%% denotations of $A$ and $A'$. The function type $A \ra A'$ is interpreted as the
+%% set of functions from $\sem{A}$ to $\li (\sem {A'})$.
+%
+%% The interpretation of a value $\hasty {\Gamma} V A$ is a function from
+%% $\sem{\Gamma}$ to $\sem{A}$. Likewise, a term $\hasty {\Gamma} M {{A}}$ is
+%% interpreted as a function from $\sem{\Gamma}$ to $\li \sem{A}$.
+Following the CBPV treatment of gradual typing given by
+New-Licata-Ahmed, an upcast $\upc c$ where $c : A \ltdyn A'$ is
+interpreted as a pure function $\sem{\upc c} :\sem{A} \to \sem{A'}$ whereas a
+downcast $\dnc c$ is interpreted as a homomorphism $\sem{\dnc c} :
+\li\sem{A'} \multimap \li\sem{A}$.
+%% By the universal property of $\li$, this is determined by a function $\sem{A'} \to U\li\sem{A}$ and
+%% We use $\text{ext}$ to go from the former to the latter in the definitions of the downcasts.
+We define the semantics of upcasts and downcasts simultaneously by
+recursion over type precision derivations.
+%
+The reflexivity casts are given by identities, and the transitivity
+casts by composition. The upcasts for products, and the downcasts for
+functions are given by the functorial actions of the type constructors
+on the casts for the subformulae.
+%
+The \emph{up}casts for functions and similarly the \emph{down}casts
+for products, on the other hand, are more complex.
+%
+For these, we use not the ordinary functorial action of type
+constructors, but an action we call the \emph{Kleisli} action.
 
-We now discuss the semantics of the cast terms, shown in Figure
-\ref{fig:term-semantics}. Upcasts are pure, so the upcast ${\upc c}$, where $c :
-A \ltdyn A'$, denotes an ordinary function from $\sem{A}$ to $\sem{A'}$.
-Downcasts are effectful, so a downcast ${\dnc c}$ is interpreted as a morphism
-of error domains between $F\sem{A'}$ and $F\sem{A}$. By definition of the
-adjunction this is equivalent to an ordinary function between $\sem{A'}$ and
-$UF\sem{A}$, i.e., $\sem{A'} \to \li \sem{A}$. We use $\text{ext}$ to go from
-the former to the latter in the definitions of the downcasts.
+The Kleisli action of type constructors can be defined in any CBPV
+model. First, we can define the Kleisli category of value types
+$\calV_k$ to have value types as objects, but have morphisms
+$\calV_k(A,A') = \calE(FA,FA')$. Similarly, we can define a Kleisli
+category of computation types $\calE_k$ to have computation types as
+objects, but have morphisms $\calE_k(B,B') = \calV(UB,UB')$. The
+intuition is that these are the ``effectful'' morphisms between
+value/computation types, whereas the morphisms of $\calV/\calE$ are
+``pure''/``linear''. Then every CBPV type constructor $F,U,\times,\to$
+has in addition to its usual functorial action on pure/linear
+morphisms, an action in each argument on Kleisli morphisms. So for
+instance, the function construction is functorial in pure/linear
+morphisms $\to : \op{\calV} \times \calE \to \calE$ but on Kleisli
+morphisms what we have is for each computation type $B$ a functorial
+action $-\to B : \calV_k^{op} \to \calE_k$ given by $-\tok B$ on
+objects and for each value type a functorial action $A \tok - :
+\calE_k \to \calE_k$. Similarly we get two actions $A \timesk -$ and
+$-\timesk A'$ on Kleisli morphisms between value types. In these two
+cases we do not get a bifunctorial action because the two Kleisli
+actions do not commute past each other, i.e., in general $(A_l'
+\timesk f_r \circ f_l \timesk A_r) \neq (f_l \timesk A_r' \circ A_l
+\timesk f_r)$. The full definitions of the Kleisli actions are
+included in the appendix.
+
+Lastly, we have the upcasts and downcasts for the injections, which
+are the core primitive casts. The upcasts are simply the injections
+themselves, except the function case which must include a $\nxt$ to
+account for the fact that the functions are under a later in the
+dynamic type. Similarly, the downcasts are similar in that on values
+they pattern match on the input and return it if it is of the correct
+type, otherwise erroring. Again, the function case is slightly
+different in that if the input is in the function case, then it is
+actually only a function available later, and so we must insert a
+``think'' in order to return it.
 
@@
 
-\eric{Introduce Kleisli categories and functors}
+%% \eric{Introduce Kleisli categories and functors}
 
-Recall the definition of type precision derivations given in Figure
-\ref{fig:typrec}. The semantics of the up- and downcasts for a type precision
-derivation $c$ are defined by induction on $c$.
-%
-For $r(A)$ the up- and downcasts are simply the identity function. 
-%
-For the composition $c \comp c'$ we compose the cast for $c$ and the cast for
-$c'$ as functions.
-%
-For $c_i \ra c_o$ we apply the casts for $c_i$ and $c_o$ in the domain and
-codomain respectively. More concretely, for the upcast we are given a function
-$V_f : A_i \to \li A_o$ and we must define a function $A_i' \to \li A_o'$. The
-function first downcasts its argument according to $c_i$, resulting in an
-element of $\li A_i$. It then applies $\ext{V_f}{}$ to this value to obtain an
-element of $\li A_o$. Finally, it applies the upcast of $c_o$ via the functorial
-action of the lift monad, obtaining an element of $\li A_o'$ as needed. For the
-downcast from $A_i' \to \li A_o'$ to $A_i \to \li A_o$, the resulting function
-first applies the upcast by $c_i$ to its argument, then applies the original
-function, and finally applies the downcast by $c_o$ to the result.
-%
-Likewise, in the upcast for $c_1 \times c_2$, we apply the cast for $c_1$ on the
-left and the cast for $c_2$ on the right.
+%% Recall the definition of type precision derivations given in Figure
+%% \ref{fig:typrec}. The semantics of the up- and downcasts for a type precision
+%% derivation $c$ are defined by induction on $c$.
+%% %
+%% For $r(A)$ the up- and downcasts are simply the identity function. 
+%% %
+%% For the composition $c \comp c'$ we compose the cast for $c$ and the cast for
+%% $c'$ as functions.
+%% %
+%% For $c_i \ra c_o$ we apply the casts for $c_i$ and $c_o$ in the domain and
+%% codomain respectively. More concretely, for the upcast we are given a function
+%% $V_f : A_i \to \li A_o$ and we must define a function $A_i' \to \li A_o'$. The
+%% function first downcasts its argument according to $c_i$, resulting in an
+%% element of $\li A_i$. It then applies $\ext{V_f}{}$ to this value to obtain an
+%% element of $\li A_o$. Finally, it applies the upcast of $c_o$ via the functorial
+%% action of the lift monad, obtaining an element of $\li A_o'$ as needed. For the
+%% downcast from $A_i' \to \li A_o'$ to $A_i \to \li A_o$, the resulting function
+%% first applies the upcast by $c_i$ to its argument, then applies the original
+%% function, and finally applies the downcast by $c_o$ to the result.
+%% %
+%% Likewise, in the upcast for $c_1 \times c_2$, we apply the cast for $c_1$ on the
+%% left and the cast for $c_2$ on the right.
 
-The ``base cases'' for the casts are $\inat$, $\itimes$, and $\iarr$. Recall
-that $D$ is isomorphic to $\Nat\, + (D \times D)\, + \laterhs (D \to \li D)$
-(say that the sum is left-associative).
-%
-For $\inat$, the upcast is simply $\inl \circ \inl$. The downcast has type $D
-\to \li D$ and performs a case inspection on the sum type. If it is a natural
-number $n$, then we return $\eta\, n$. Otherwise, we return $\mho$, modelling
-the fact that a run-time error occurs.
-%
-For $\itimes$, the upcast is $\inl \circ \inr$, and the downcast performs a case
-inspection analogously to the natural number downcast.
-%
-% TODO explain this further
-For $\iarr$, the upcast is given by $\inr \circ \nxt$. The downcast performs a
-case inspection, and in the $\inr$ case, it uses a $\theta$ in order to gain
-access to the function under the $\later$.
+%% The ``base cases'' for the casts are $\inat$, $\itimes$, and $\iarr$. Recall
+%% that $D$ is isomorphic to $\Nat\, + (D \times D)\, + \laterhs (D \to \li D)$
+%% (say that the sum is left-associative).
+%% %
+%% For $\inat$, the upcast is simply $\inl \circ \inl$. The downcast has type $D
+%% \to \li D$ and performs a case inspection on the sum type. If it is a natural
+%% number $n$, then we return $\eta\, n$. Otherwise, we return $\mho$, modelling
+%% the fact that a run-time error occurs.
+%% %
+%% For $\itimes$, the upcast is $\inl \circ \inr$, and the downcast performs a case
+%% inspection analogously to the natural number downcast.
+%% %
+%% % TODO explain this further
+%% For $\iarr$, the upcast is given by $\inr \circ \nxt$. The downcast performs a
+%% case inspection, and in the $\inr$ case, it uses a $\theta$ in order to gain
+%% access to the function under the $\later$.
 
@@
 % TODO should we include function casts?
+%% \begin{figure*}
+%%   \begin{align*}
+%%     \sem{\nat} &= \Nat \\
+%%     \sem{\dyn} &= D \\
+%%     \sem{A \times A'} &= \sem{A} \times \sem{A'} \\ 
+%%     \sem{A \ra A'} &= \sem{A} \To \li \sem{A'} \\
+%%   \end{align*}
+%%   \caption{Interpretation of types}
+%%   \label{fig:type-interpretation}
+%% \end{figure*}
 
 \begin{figure*}
-  \begin{align*}
-    \sem{\nat} &= \Nat \\
-    \sem{\dyn} &= D \\
-    \sem{A \times A'} &= \sem{A} \times \sem{A'} \\ 
-    \sem{A \ra A'} &= \sem{A} \To \li \sem{A'} \\
-  \end{align*}
-  \caption{Interpretation of types}
-  \label{fig:type-interpretation}
-\end{figure*}
-
-\begin{figure*}
-  \eric{Check this}
+  \max{Can we squeeze this figure very tightly?}
 
@@
     % upcasts
-    \sem{\upc{r(A)}} &= \id \\
+    \sem{\upc{r(A)}} &= \id_A \\
     \sem{\upc{(c \comp c')}} &= \sem{\upc{c'}} \circ \sem{\upc{c}} \\
-    \sem{\upc{(c_i \ra c_o)}} &=
-      \lambda V_f . \text{map } \upc{c_o} \circ \ext{V_f}{} \circ \dnc{c_i} \\
-    \sem{\upc{(c_1 \times c_2)}} &= 
-      \lambda (V_1, V_2) . ((\sem{\upc{c_1} V_1}) , 
-                            (\sem{\upc{c_2} V_2})) \\
-    \sem{\upc{\inat}} &= 
-      \lambda V_n . \inl\, \inl\, V_n \\
-    \sem{\upc{\itimes}} &= 
-      \lambda (V_1, V_2) . \inl\, \inr\, (V_1 , V_2) \\
-    \sem{\upc{\iarr}} &= 
-      \lambda V_f . \inr\, \nxt V_f \\[2ex]
+    \sem{\upc{(c_i \ra c_o)}} &= (\sem{\dnc {c_i}} \tok F\sem{A_o'})\circ (\sem{A_i} \tok UF(\sem{\upc {c_o}}))\\
+    \sem{\upc{(c_1 \times c_2)}} &= \sem{\upc{c_1}} \times \sem{\upc{c_2}}\\
+    \sem{\upc{\inat}} &= \inat\\
+    \sem{\upc{\itimes}} &= \itimes\\
+    \sem{\upc{\iarr}} &= \iarr \circ \nxt\\[2ex]
     %
     % downcasts
-    \sem{\dnc{r(A)}} &= \id \\
+    \sem{\dnc{r(A)}} &= \id_{FA} \\
     \sem{\dnc{(c \comp c')}} &= \sem{\dnc{c}} \circ \sem{\dnc{c'}} \\
-    \sem{\dnc{(c_i \ra c_o)}} &=
-     \text{ext}(\eta(\lambda V_f . \dnc{c_o} \circ V_f \circ \upc{c_i})) \\
+    \sem{\dnc{(c_i \ra c_o)}} &= U(\sem{\upc {c_i}} \to \sem{\dnc {c_o}})\\
     \sem{\dnc{(c_1 \times c_2)}} &= (\dnc{c_1} \timesk A_2) \circ (A_1' \timesk \dnc{c_2}) \\
-    \sem{\dnc{\inat}} &= \text{ext}(
+    \sem{\dnc{\inat}} &= (
       \lambda V_d . \text{case $({V_d})$ of}
-      \{ \inl\, (\inl\, n) \to \eta\, n
-         \alt \text{otherwise} \to \mho \}) \\
-    \sem{\dnc{\itimes}} &= \text{ext}(
+      \{ \inat\,n \to \eta\, n
+         \alt \text{otherwise} \to \mho \})^\dagger \\
+    \sem{\dnc{\itimes}} &= (
       \lambda V_d . \text{case $({V_d})$ of}
-      \{ \inl\, (\inr\, (d_1, d_2)) \to \eta\, (d_1, d_2)
-         \alt \text{otherwise} \to \mho \}) \\
-    \sem{\dnc{\iarr}} &= \text{ext}(
+      \{ \itimes(d_1, d_2) \to \eta\, (d_1, d_2)
+         \alt \text{otherwise} \to \mho \})^\dagger \\
+    \sem{\dnc{\iarr}} &= (
       \lambda V_d . \text{case $({V_d})$ of}
-      \{ \inr\, \tilde{f} \to \theta\, (\lambda t. \eta (\tilde{f}_t))
-         \alt \text{otherwise} \to \mho \}) \\
+      \{ \iarr \tilde{f} \to \theta_t\, (\eta (\tilde{f}_t))
+         \alt \text{otherwise} \to \mho \})^\dagger \\
     % \sem{\ret\, V}       &= \lambda \gamma . \eta\, \sem{V} \\
@@
   \end{align*}
-
-  \caption{Semantics of casts.}
+  \caption{Semantics of casts}
   \label{fig:term-semantics}
 \end{figure*}
+
+\subsection{Extracting a well-behaved big-step semantics}
+
+\max{TODO: Eric needs to add a discussion of adequacy/clock stuff here}

### 2024-07-09 paper-new/intro.tex
fix undef
cast semantics update

@@
 function from closed programs to either natural numbers or errors:
-\[ -\Downarrow : \{M \,|\, \cdot \vdash M : \natty \} \rightharpoonup \mathbb{N} \union {\mho} \]
+\[ -\Downarrow : \{M \,|\, \cdot \vdash M : \nat \} \rightharpoonup \mathbb{N} \cup \{\mho\} \]
 where $\mho$ is notation for a runtime type error. We write $M
@@
 formalize type based reasoning we give a typed equational theory for
-terms of the language $M \cong N$ for when two terms should be
+terms of the language $M \equiv N$ for when two terms should be
 considered equivalent. Then we want to verify that the big step
-semantics respects this equational theory: if closed programs $M \cong
+semantics respects this equational theory: if closed programs $M \equiv
 N$ are equivalent in the equational theory then they have the same
@@
 The only prior denotational semantics for gradual typing was given by
-New and Licata and is based on \emph{domain theory}
-\cite{newlicata2019}. The fundamental idea is to equip $\omega$-CPOs
-with an additional ``error ordering'' $\ltdyn$ which models the
-graduality ordering, and for casts to arise from
+New and Licata and is based on a classical Scott-style \emph{domain
+theory} \cite{newlicata2019}. The fundamental idea is to equip
+$\omega$-CPOs with an additional ``error ordering'' $\ltdyn$ which
+models the graduality ordering, and for casts to arise from
 \emph{embedding-projection pairs}. Then the graduality property
@@
 However, an approach based on classical domain theory has fundamental
-limitations: domain theory is incapable of modeling certain ``highly
-recursive'' features of programming languages such as dynamic type tag
+limitations: domain theory is incapable of modeling certain perversely
+recursive features of programming languages such as dynamic type tag
 generation and higher-order references, which are commonplace in
 real-world gradually typed systems as well as gradual calculi
-\cite{examples-of-gradual-stuff}. Our goal in this work is to develop
-an approach that will eventually scale up to these advanced features,
-though for a first work we focus on adapting their results for a more
-basic semantics.
+\cite{examples-of-gradual-stuff}. Our long-term goal is to develop a
+semantics that will scale up to these advanced features, and so we
+begin with this work by departing from classical domain theory.
 
@@
 (ultra)metric spaces but more commonly as ``step-indexed sets'', i.e.,
-objects in the ``topos of trees'', i.e., presheaves on the poset of
-natural numbers. We think of these as modeling a kind of ``sequence of
-approximations'' to the domain being modeled.
+objects in the \emph{topos of trees}, i.e., presheaves on the poset of
+natural numbers\cite{topos-of-trees}. We think of such a
+$\mathbb{N}$-indexed set as modeling a kind of ``sequence of
+approximations'' to the true type being modeled.
 %
@@
 theory, this means we must work with an \emph{intensional} semantics,
-one where unfolding the dynamic type takes an observable runtime step.
+one where unfolding the dynamic type requires an observable
+computational step: e.g., a function that pattern matches on an
+element of the dynamic type and then returns it unchanged is
+\emph{not} equal to the identity function, because it ``costs'' a step
+to perform the pattern match.
+%
+This leads to the main departure of our semantics from the New-Licata
+approach: while the New-Licata semantics equips $\omega$-CPOs with an
+error ordering, we work with types equipped with not just an error
+ordering but also a \emph{bisimilarity} relation that reflects when
+elements differ only in the number of abstract computational steps
+that they take.
+%
+Because bisimilarity, unlike equality, is not transitive this
+necessitates further changes to the core concepts of our semantics.
 
@@
 non-standard foundation such as guarded type theory where the later
-modality is simply an operation on our basic notion of type, and we
-take as an axiom that guarded domain equations have a (necessarily
-unique) solution. This has the added benefit that providing a
-denotational semantics of gradually typed terms is as simple as an
-semantics of an effectful language using a monad on the category of
-sets. Not only does this make on-paper reasoning about guarded domain
-theory easier, it also enables a simpler avenue to verification in a
-proof assistant. Whereas formalizing analytic guarded domain theory
-would require significant theory of presheaves and making sure that
-all constructions are functors on categories of presheaves,
-formalizing synthetic guarded domain theory can be done by directly
-adding the later modality and the guarded fixed point property
-axiomatically.
-
-\max{TODO: more here}
+modality is taken as a primitive operation on types, and we take as an
+axiom that guarded domain equations have a (necessarily unique)
+solution. This has the added benefit that providing a denotational
+semantics of gradually typed terms is as simple as an semantics of an
+effectful language using a monad on the category of sets. Not only
+does this make on-paper reasoning about guarded domain theory easier,
+it also enables a simpler avenue to verification in a proof
+assistant. Whereas formalizing analytic guarded domain theory would
+require significant theory of presheaves and making sure that all
+constructions are functors on categories of presheaves, formalizing
+synthetic guarded domain theory can be done by directly adding the
+later modality and the guarded fixed point property axiomatically.
+%
+\max{TODO: more here, specifically this is where we should talk about adequacy I think}
 
@@
   how to model the dynamic type as a solution to a guarded domain equation
-
 \item Next, we identify where prior work on classical domain theoretic
@@
   of recursive types.
-
 \item We develop a key new concept of \emph{syntactic perturbations},
@@
   graduality property compositionally
-
 \item We combine this insight together with an abstract categorical

### 2024-07-09 paper-new/concrete-model.tex
work on relational model

@@
 delays on each type as well as a notion of representability for relations that
-models the cast rules.
+models the cast rules in the intensional setting.
 
@@
 
-\subsubsection{Guarded Lift Monad and Free-Forgetful Adjunction}\label{sec:guarded-lift-monad}
+\subsubsection{The Free Error Domain}\label{sec:free-error-domain}
 
@@
 
-\subsection{Phase 2: Perturbations and Quasi-Representable Relations}
+\subsection{Phase Two: Perturbations and Quasi-Representable Relations}
 
@@
 square for the simplified version as was shown in Section
-\ref{sec:modeling-term-precision}. Recall that we horizontally composed the
+\ref{sec:towards-relational-model}. Recall that we horizontally composed the
 simplified UpL for $c$ square with the identity square for $c'$ on the right.
@@
 \begin{definition}
-    Let $(A, M_A, i_A)$ and $(A', M_{A'}, and i_{A'})$ be value objects, and let
+    Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let
     $c : A \rel A'$ be a predomain relation. A \emph{push-pull structure} for
@@
 are perturbations $\delle_c \in M_A$ and $\delre_c \in M_{A'}$ such that the
-following two squares exist: (UpL) $e \ltsq{c}{r(A')} i_{A'}(\delre_c)$, and
-(UpR) $i_A(\delle_c) \ltsq{r(A)}{c} e$.
+following two squares exist: (1) UpL: $e \ltsq{c}{r(A')} i_{A'}(\delre_c)$, and
+(2) UpR: $i_A(\delle_c) \ltsq{r(A)}{c} e$.
 \end{definition}
@@
 the following two squares exist: 
-(DnL) $p \ltsq{r(B')}{d} i_{B'}(\delrp_d)$ and
-(DnR) $i_B(\dellp_d) \ltsq{d}{r(B)} p$.
+(1) DnL: $p \ltsq{r(B')}{d} i_{B'}(\delrp_d)$ and
+(2) DnR: $i_B(\dellp_d) \ltsq{d}{r(B)} p$.
 \end{definition}
@@
 \begin{definition}[value relations]
-A \textbf{value relation} between value objects $(A, M_A, i_A)$ and $(A', M_{A'},
-i_{A'})$ is a predomain relation $c : A \rel A'$ that has a push-pull structure
-is quasi-left-representable by a morphism $e_c : A \to A'$, and is such that
-$Fc$ is quasi-right-representable by an error domain morphism $p_c : FA' \to
-FA$.
+A \textbf{value relation} between value objects $(A, M_A, i_A)$ and $(A',
+M_{A'}, i_{A'})$ consists of a predomain relation $c : A \rel A'$ that has a
+push-pull structure, is quasi-left-representable by a morphism $e_c : A \to A'$,
+and is such that $Fc$ is quasi-right-representable by an error domain morphism
+$p_c : FA' \to FA$.
 \end{definition}
@@
 \begin{definition}[computation relations]
-A \textbf{computation relation} between computation objects $(B, M_B, i_B)$ and $(B',
-M_{B'}, i_{B'})$ is an error domain relation $d$ that has a push-pull structure,
-is quasi-right-representable by a morphism $p_d : B' \to B$, and is such that
-$Ud$ is quasi-left-representable by a morphism $e_d : UB \to UB'$
+A \textbf{computation relation} between computation objects $(B, M_B, i_B)$ and
+$(B', M_{B'}, i_{B'})$ consists of an error domain relation $d$ that has a
+push-pull structure, is quasi-right-representable by a morphism $p_d : B' \to
+B$, and is such that $Ud$ is quasi-left-representable by a morphism $e_d : UB
+\to UB'$
 \end{definition}
 
-Because we have augmented the definitions of object and relation for our model,
-we need to specify the actions of the functors $F$, $U$, $\times$, and $\arr$ on
-these objects and relations. For instance, the functor $F$ now acts on
+The reason for requiring $Fc$ to be quasi-right-representable in the definition
+of value relations is so that we can define the action of the functor $F$ on
+relations. Similarly, we require $Ud$ to be quasi-left-representable in the
+definition of computation relations so that we can define the action of $U$ on
+relations.
+
+It remains to specify the actions of the functors on value and computation
+objects and relations.
+% Because we have augmented the definitions of object and relation for our model,
+% we need to specify the actions of the functors $F$, $U$, $\times$, and $\arr$ on
+% these objects and relations. 
+For instance, the functor $F$ now acts on
 \emph{value objects} $(A, M_A, i_A)$, which means we need to specify the monoid
-$M_{FA}$ and interpretation $i_{FA}$ corresponding to the error domain $FA$. We
-leave these constructions to the appendix, as they are technical and not crucial
-to understanding our main result.
+$M_{FA}$ and interpretation $i_{FA}$ corresponding to the error domain $FA$.
+
+\subsubsection{Actions on Objects}
+We first discuss the action of $\li$ on objects. Given a value object $(A, M_A,
+i_A)$ we define $M_{\li A}$ to be $\mathbb{N} \oplus M_A$, where $\oplus$
+denotes the free product of monoids (the coproduct in the category of monoids).
+The intuition behind needing to include $\mathbb{N}$ comes from the example of
+the downcast for $\iarr$. Recall that we adjusted the squares corresponding to
+the DnL and DnR rules by adding $\delta = \theta \circ \nxt$ on the side
+opposite the downcast. Thus, we need a syntactic perturbation in $M_{\li A}$
+that will be interpreted as $\delta$. We accomplish this by taking the coproduct
+with $\mathbb{N}$.
+%
+Then to interpret the perturbations as endomorphisms, by the universal property
+of the coproduct of monoids it suffices to define two homomorphisms, one from
+$\mathbb{N} \to \morbisimid{\li A}$ and one from $M_A \to \morbisimid{\li A}$.
+The first homomorphism sends the generator $1 \in \mathbb{N}$ to the endomorpism
+$(\delta \circ \eta)^\dagger$, and the second sends $m_A \in M_A$ to
+$\li(i_A(m_A))$ and we observe in both cases that the resulting endomorphisms
+are bisimilar to the identity since the action of $\li$ on morphisms preserves
+bisimilarity.
+
+Now we discuss the action of $U$ on objects. Given a computation object $(B,
+M_B, i_B)$, we define $M_{UB}$ to be $\mathbb{N} \oplus M_B$. The reason for
+requiring $\mathbb{N}$ is related to the Kleisli arrow functor: to establish
+quasi-representability of $U(c \arr d)$ given quasi-representability of $c$ and
+$d$, we will need to turn a perturbation on $FA$ into a perturbation on $U(A
+\arr B)$. Since the perturbations for $FA$ involve $\mathbb{N}$, so must the
+perturbations for $UB$. The interpretation $i_{UB}$ of the perturbations on $UB$
+works in the same manner as that of $FA$.
+
+The action of $\times$ on objects is as follows. Given $(A_1, M_{A_1}, i_{A_1})$
+and $(A_2, M_{A_2}, i_{A_2})$, we define the monoid $M_{A_1 \times A_2} =
+M_{A_1} \oplus M_{A_2}$. To construct the homomorphism $i_{A_1 \times A_2}$, we
+appeal to the universal property and give homomorphisms $M_{A_1} \to
+\morbisimid{A_1 \times A_2}$ and $M_{A_2} \to \morbisimid{A_1 \times A_2}$.
+
+The action of $\arr$ on objects is as follows. Given $(A, M_A, iA)$ and $(B,
+M_B, i_B)$ we define the monoid $M_{A \arr B} = M_A^{op} \oplus M_B$. Then to
+construct the interpretation as endomorphisms it suffices to define a
+homomorphism $M_A^{op} \to \morbisimid{A \arr B}$ and $M_B \to \morbisimid{A
+\arr B}$. The former takes an element $m_A$ and a predomain morphism $f : A \to
+UB$ and returns the composition $f \circ i_A(m_A)$, while the latter is defined
+similarly by post-composition with $i_B$.
+
+\subsubsection{Actions on Relations}
 
@@
 Lastly, we describe the denotation of the dynamic type. The predomain
-representing the dynamic type is defined using guarded recursion as the solution
-to the equation
+representing the dynamic type is defined using a combination of least fixpoint
+and guarded recursion as the solution to the equation
 % \footnote{In this section, we write $F$ instead of $\li$ so that the notation follows that of the abstract model section.}
@@
 Now we define the perturbations for $D$. Recall that for each value type $A$ we
-associate a monoid $P_A$ of perturbations and a homomorphism into the monoid of
+associate a monoid $M_A$ of perturbations and a homomorphism into the monoid of
 endomorphisms bisimilar to the identity, and likewise for computation types. We
@@
 %
-\( P_D \cong (P_{D \times D}) \times P_{U(D \to FD)}. \)
+\( M_D \cong (M_{D \times D}) \times M_{U(D \to FD)}. \)
 %
@@
 %
-\( P_D \cong (P_D \times P_D) \times (\mathbb{N} \times P_D^{op} \times \mathbb{N} \times P_D). \)
+\( M_D \cong (M_D \times M_D) \times (\mathbb{N} \times M_D^{op} \times \mathbb{N} \times M_D). \)
 %
 We now explain how to interpret these perturbations as endomorphisms.
-We define $\ptb_D : P_D \to \{ f : D \to D \mid f \bisim \id \}$ below,
+We define $i_D : M_D \to \{ f : D \to D \mid f \bisim \id \}$ below,
 % via the universal property of the coproduct of monoids, giving a case for each of the generators.
@@
 \begin{align*}
- \ptb_D(p_{\text{times}}, p_{\text{fun}}) &= \lambda d.\text{case $d$ of}  \\
+ i_D(p_{\text{times}}, p_{\text{fun}}) &= \lambda d.\text{case $d$ of}  \\
  &\alt \tnat(m) \mapsto \tnat(m) \\
-    &\alt \ttimes(d_1, d_2) \mapsto {\ttimes(\ptb_{D \times D}(p_\text{times})(d_1, d_2))} \\
-    &\alt \tfun(\tilde{f}) \mapsto {\tfun(\lambda t. \ptb_{U(D \to FD)}()(\tilde{f}_t))}
+    &\alt \ttimes(d_1, d_2) \mapsto {\ttimes(i_{D \times D}(p_\text{times})(d_1, d_2))} \\
+    &\alt \tfun(\tilde{f}) \mapsto {\tfun(\lambda t. i_{U(D \to FD)}(\tilde{f}_t))}
 \end{align*}
@@
     % \[ (d_1, \ptb_D(\delta^K_D)(d_2)) \] instead.
-One can verify that this defines a homomorphism from $P_D \to \{ f : D \to D : f \bisim \id \}$.
-We claim that the three relations $\inat$, $\itimes$, and $\iarr$
-%and their lifted versions 
-satisfy the push-pull property.
-As an illustrative case, we establish the push-pull property for the relation $\iarr$.
-We define $\pull_{\iarr} : P_D \to P_{U(D \arr FD)}$ by \( \pull_{\iarr}(p_{\text{times}}, p_{\text{fun}}) = p_{\text{fun}}, \)
+One can verify that this defines a homomorphism from $M_D \to \{ f : D \to D : f
+\bisim \id \}$. We claim that the three relations $\inat$, $\itimes$, and
+$\iarr$ satisfy the push-pull property. As an illustrative case, we establish
+the push-pull property for the relation $\iarr$. We define $\pull_{\iarr} : M_D
+\to P_{U(D \arr FD)}$ by \( \pull_{\iarr}(p_{\text{times}}, p_{\text{fun}}) =
+p_{\text{fun}}, \)
 % (recall that $P_{U(D \arr FD)} = \mathbb{N} \times P_D^{op} \times \mathbb{N} \times P_D$).
@@
 %
-We define $\push_{\iarr} : P_{U(D \arr FD)} \to P_D$ by \( \push_{\iarr}(p_{\text{fun}}) = (\id, p_{\text{fun}}) \).
+We define $\push_{\iarr} : M_{U(D \arr FD)} \to M_D$ by \( \push_{\iarr}(p_{\text{fun}}) = (\id, p_{\text{fun}}) \).
 %
@@
 
-We next claim that the relations $\inat$, $\itimes$, and $\iarr$ are quasi-left-representable,
-and that their lifts are quasi-right-representable.
-Indeed, since the relations are functional, it is easy to see that they are quasi-left-representable
-where the perturbations are taken to be the identity.
+We next claim that the relations $\inat$, $\itimes$, and $\iarr$ are
+quasi-left-representable, and that their lifts are quasi-right-representable.
+Indeed, since the relations are functional, it is easy to see that they are
+quasi-left-representable where the perturbations are taken to be the identity.
 %
 For quasi-right-representability, the most interesting case is $\li(\iarr)$.
-Defining the projection $p_{\iarr} : FD \to FU(D \to FD)$ is equivalent to defining
-$p' : D \to UFU(D \to FD)$. We define
+Defining the projection $p_{\iarr} : FD \to FU(D \to FD)$ is equivalent to
+defining $p' : D \to UFU(D \to FD)$. We define
 \begin{align*}
@@
 %
-We define $\dellp_D = \delrp_D = \delta_{FD}$.
-Then it is easy to show using the definition of $\iarr$ that the squares for $\dnl$ and $\dnr$ commute.
+We define $\dellp_D = \delrp_D = \delta_{FD}$. Then it is easy to show using the
+definition of $\iarr$ that the squares for $\dnl$ and $\dnr$ commute.
 % Then for $\dnr$, we need to show that if $(f, d) \in \iarr$ then
 %
-It is also straightforward to establish the retraction property for
-each of these three relations. In the case of $\iarr$, we
-have that the property holds up to a delay.
+It is also straightforward to establish the retraction property for each of
+these three relations. In the case of $\iarr$, we have that the property holds
+up to a delay.
 
@@
 
-Now that we have defined an intensional model with an interpretation for the dynamic type, we can apply
-the abstract constructions introduced in Section \ref{sec:extensional-model-construction}.
-Doing so, we obtain an extensional model of gradual typing, where the squares are given by the
-``bisimilarity closure'' of the intensional error ordering.
+\subsection{Defining an Extensional Model}
+
+Lastly, we apply the ``bisimilarity-closure'' construction introduced in the
+previous section as follows: In particular, let $c_i : A_i \rel A_i'$ and $c_o :
+A_o \rel A_o'$.
+%
+% We define an extensional square between morphisms $f : A_i \to A_o$ and $g :
+% A_i' \to A_o'$ to consist of:
+
+% \begin{itemize}
+%   \item A morphism $f' : A_i \to A_o$ with $f \bisim f'$.
+%   \item A morphism $g' : A_i' \to A_o'$ with $g \bisim g'$.
+%   \item A square $f' \ltsq{c_i}{c_o} g'$.
+% \end{itemize}
+%
+We say that $f \ltsqbisim{c^i}{c_o} g$ if there
+exist $f'$ and $g'$ such that
+%
+\[ f \bisim f' \ltsq{c_i}{c_o} g' \bisim g. \]
+%
+We make the analogous construction for the computation squares. If the relations
+$c_i$ and $c_o$ are obvious from the context, we will simply write $f \ltbisim
+g$.
+
+% Now that we have defined an intensional model with an interpretation for the
+% dynamic type, we can apply the abstract constructions introduced in Section
+% \ref{sec:extensional-model-construction}. Doing so, we obtain an extensional
+% model of gradual typing, where the squares are given by the ``bisimilarity
+% closure'' of the intensional error ordering.
 
@@
 
-In this section, we prove an adequacy result for the concrete extensional model of GTT we obtained in the previous section.
-applying the abstract constructions introduced in Section
-\ref{sec:extensional-model-construction} to the concrete model built in the previous section.
-%\ref{sec:concrete-model}.
-
-First we establish some notation. Fix a morphism $f : 1 \to \li \mathbb{N} \cong \li \mathbb{N}$.
-We write $f \da n$ to mean that there exists $m$ such that $f = \delta^m(\eta n)$
+In this section, we prove an adequacy result for the extensional model of
+gradual typing we obtained in the previous section. First we establish some
+notation. Fix a morphism $f : 1 \to \li \mathbb{N} \cong \li \mathbb{N}$. We
+write $f \da n$ to mean that there exists $m$ such that $f = \delta^m(\eta n)$
 and $f \da \mho$ to mean that there exists $m$ such that $f = \delta^m(\mho)$.
 
-Recall that $\ltls$ denotes the relation on value morphisms defined as the bisimilarity-closure
-of the intensional error-ordering on morphisms.
-That is, we have $f \ltls g$ iff there exists $f'$ and $g'$ with
+Recall that $\ltbisim$ denotes the relation on value morphisms defined as the
+bisimilarity-closure of the intensional error-ordering on morphisms. That is, we
+have $f \ltbisim g$ iff there exist $f'$ and $g'$ with
 %
@@
 %
-Here $\le_{\li \mathbb{N}}$ is the lock-step error ordering, and
-$\bisim_{\li \mathbb{N}}$ is weak bisimilarity.
-First observe that in this ordering, the semantics of error is not equivalent to
-the semantics of the diverging term.
-The main result we would like to show is as follows:
-\begin{lemma}
-If $f \ltls g : \li \mathbb{N}$, then:
+Here $\le_{\li \mathbb{N}}$ is the lock-step error ordering, and $\bisim_{\li
+\mathbb{N}}$ is weak bisimilarity. First observe that in this ordering, the
+semantics of error is not equivalent to the semantics of the diverging term. The
+main result we would like to show is as follows: If $f \ltbisim g : \li \mathbb{N}$, then:
 \begin{itemize}
@@
 \end{itemize}
-\end{lemma}
+
 %

### 2024-07-09 paper-new/concrete-relational-model.tex
work on relational model

@@
 
-\subsubsection{Modelling Type Precision}
-We model type precision $c : A \ltdyn A'$ as a \emph{monotone relation}, i.e., a
+We model a type precision relation $c : A \ltdyn A'$ as a \emph{monotone relation}, i.e., a
 relation $c$ that is upward-closed under the relation on $A'$ and

### 2024-07-09 paper-new/towards-relational-model.tex
rename some files, progress on the relational model motivation section

--- 2024-07-09 paper-new/concrete-relational-model.tex
@@
 
-\eric{Review this section}
+Now that we have seen how to model the terms without regards to
+graduality, we next turn to how to enhance this model to provide a
+compositional semantics that additionally satisfies graduality. By a
+compositional semantics, we mean that we want to provide a
+compositional interpretation of \emph{type} and \emph{term precision}
+such that we can extract the graduality relation from our model of
+term precision.
 
-In the previous section, we gave a denotational call-by-push-value model for the
-terms of the gradually-typed lambda calculus using the tools of SGDT. In this
-section, we build on the ideas used in defining a term model with the goal of
-defining a model of the full gradually-typed lambda calculus, i.e., one that has
-constructs that model type and term precision.
+\subsection{A first attempt: Modeling Term Precision with Posets}
 
-Recall that the cast calculus introduced in Section \ref{sec:GTLC} has not only
-terms but also an \emph{inequational theory} with axioms for type precision and
-term precision that constitute an abstract formulation of the graduality
-property. Thus, our goal now is to build a denotational model interpreting these
-axioms.
+In prior work, New and Licata model the term precision relation by
+equipping $\omega$-CPOs with a second ordering $\ltdyn$ they call the
+``error ordering'' such that a term precision relation $M \ltdyn M' :
+r(A)$ is modeled by $\sem{M} \ltdyn \sem{M'}$ in the semantics. Then
+the error ordering on the free error domain $\li A$ is based on the
+graduality property: essentially either $\sem{M}$ errors or both terms
+diverge or both terms terminate with values related by the ordering
+relation on $A$.
 %
+The obvious first try at a guarded semantics that models graduality
+would be to similarly enhance our value types and computation types to
+additionally carry a poset structure. For this value types can simply
+be posets. For computation types, we want our error domains to be
+posets as well. To consider what additional properties we want the
+error domains to have, let's consider the particular case of the free
+error domain $\li A$.
+
+The ordering on $\li A$ should model the graduality property, but how
+to interpret this is not so straightforward. The graduality property
+as stated mentions divergence, but we only work indirectly with
+diverging computations in the form of the abstract steps/think
+maps. The closest direct encoding of the graduality property is an
+ordering we call the \emph{step-insensitive error ordering}, defined
+by guarded recursion in
+Figure~\ref{fig:step-insensitive-error-ordering}.
+%
+Firstly, to model graduality, $\mho$ is the least element. Next, two
+returned values are related just when they are in the ordering $\leq$
+on $A$.
+%
+Two thinking elements are related just when they are related later.
+%
+The interesting cases are then those where one side is thinking and
+the other has completed to either an error or a value.
+%
+Since the graduality property is \emph{extensional}, i.e., oblivious
+to the number of steps taken, it is sensible to say that if one side
+has terminated and the other side is thinking, that we must require
+the thinking side to eventually terminate with a related behavior,
+which is the content of the final $4$ cases of the definition.
+
+\begin{figure*}
+    \begin{align*}
+        &\mho \semltbad l \text{ iff } \top \\
+        %
+        &\eta\, x \semltbad \eta\, y \text{ if } 
+            x \semlt y \\
+        %		
+        &\theta\, \tilde{l} \semltbad \theta\, \tilde{l'} \text{ iff } 
+            \later_t (\tilde{l}_t \semltbad \tilde{l'}_t) \\
+        %	
+        &\theta\, \tilde{l} \semltbad \mho \text{ iff } \exists n. \theta\, \tilde{l} = \delta^n(\mho) \\
+        %	
+        &\theta\, \tilde{l} \semltbad \eta\, y \text{ iff } \exists n. \exists x \leq y.
+            (\theta\, \tilde{l} = \delta^n(\eta\, x)) \\
+        %
+        &\theta\, \tilde{l} \semltbad \mho \text { iff } 
+            \exists n. \theta\, \tilde{l} = \delta^n(\mho) \\
+        %	
+        &\eta\, x \semltbad \theta\, \tilde{l'} \text { iff }
+            \exists n. \exists y \geq x. (\theta\, \tilde{l'} = \delta^n (\eta\, y))
+    \end{align*}
+    \caption{The step-insensitive error ordering.}
+    \label{fig:step-insensitive-error-ordering}
+\end{figure*}
+
+This definition of $\semltbad$ is an entirely reasonable attempt to
+model graduality: for instance it is true that if $l \semltbad l'$ for
+$\li \mathbb{N}$ where $\mathbb{N}$ is the flat poset whose ordering
+is equality, then the bigstep semantics we derive for $l, l'$ do in
+fact satisfy the intended graduality relation.
+%
+However, there is a major issue with this definition: it's not a poset
+at all: it is not anti-symmetric, but more importantly it is
+\emph{not} transitive!
+%
+To see why, we observe the following undesirable property of any
+relation that is transitive and ``step-insensitive'' on one side:
+\begin{theorem}\label{thm:no-go}
+  Let $R$ be a binary relation on the free error domain $U(\li A)$. If
+  $R$ satisfies the following properties
+  \begin{enumerate}
+  \item Transitivity
+  \item $\theta$-congruence: If $\later_t (\tilde{x}_t \binrel{R} \tilde{y}_t)$, then $\theta(\tilde{x}) \binrel{R} \theta(\tilde{y})$.
+  \item Right step-insensitivity: If $x \binrel{R} y$ then $x \binrel{R} \delta y$.
+  \end{enumerate}
+  Then for any $l : U(\li A)$, we have $l R \Omega$. If $R$ is left
+  step-insensitive instead then $\Omega \binrel{R} x$.
+\end{theorem}
+\begin{proof}
+  By L\"ob induction, in the appendix.
+\end{proof}
+\begin{corollary}
+  Let $R$ be a binary relation on $U(\li A)$. If $R$ satisfies
+  transitivity, $\theta$-congruence and left and right
+  step-insensitivity, then $R$ is the total relation: $\forall x, y. x
+  \binrel{R} y$.
+\end{corollary}
+\begin{proof}
+  $x \binrel{R} \Omega$ and $\Omega \binrel{R} y$ by the previous
+  theorem. Then by transitivity $x \binrel R y$.
+\end{proof}
+
+This in turn implies $\semltbad$ is not transitive, since it is a
+$\theta$-congruence and step-insensitive on both sides, but not
+generally trivial: e.g., for the flat poset on $\mathbb N$, $\eta 0 \semltdbad
+\eta 1$ is false.
+%
+This shows definitively that we cannot provide a compositional
+semantic graduality proof based on types as posets if we use this
+step-insensitive ordering on $\li A$.
+
+\subsection{Why transitivity is essential}
+
+At this point we might try to weaken from our initial attempt at
+providing a poset semantics to a semantics where types are equipped
+with merely a reflexive relation, and continue using the
+step-insensitive ordering. However some level of transitivity is
+absolutely essential to providing compositional reasoning, and is used
+pervasively in the New-Licata development.
+%
+The reason that it is essential has to do with the details of the
+graduality proof, so first we expand on how we prove graduality
+compositionally. First of all, we need to extend our compositional
+semantics of types and terms to also a compositional semantics of
+\emph{type precision derivations} and \emph{term precision}.
+%
+Sticking to a poset-based semantics, if our value types are posets, we
+then want our precision derivations $c : A \ltdyn A'$ to be
+\emph{poset relations} on $A,A'$, that is relations between the
+underlying sets of $A,A'$ that are \emph{downward-closed} in $A$ and
+\emph{upper-closed} in $A'$. The obvious example of such a relation is
+the poset ordering itself $\leq_A : A \rel A'$...\max{TODO: Eric pick
+  up from here. Go into the squares for casts and how representability
+  gives us a compositional approach to proving them}
+
+\eric{Read this}
+The lack of transitivity presents a major barrier towards giving a compositional
+semantics to gradual typing: if our relations are not transitive, then we cannot
+compose squares horizontally. But we have argued above that horizontal
+composition is essential for modelling the axioms of gradual typing in a
+syntax-independent manner.
+
+% In particular, to model the UpL/UpR/DnL/DnR rules
+% for casts in the presence of transitivity, it is sufficient to establish the
+% validity of simpler rules that do not ``build in'' composition. We can then use
+% transitivity to derive the original versions of the rules. On the other hand,
+% without transitivity we must instead validate the cast rules in the model ``from
+% scratch''. In fact, we cannot even define the semantic interpretation of type
+% precision in a syntax-independent manner. The issue is that in the definition of
+% relation, the requirement that it be representable now involves quantifying over
+% all other relations, which is circular.
+
+% So, although this approach would suffice for proving graduality, it lacks the
+% compositionality that we seek in a reusable framework for the semantics of
+% gradual typing.
+
+%Because it is convenient to make use of transitive reasoning in proving graduality, ...
+We therefore seek to formulate an ordering relation on $\li A \times \li A'$
+that \emph{is} transitive. The theorem above tells us, however, that any such
+relation must either not be a congruence with respect to $\theta$, or must not
+be closed under delay.
+\footnote{Technically, we only need to give up closure under delay on \emph{one}
+side, as the above theorem requires $R$ to be closed on both sides. We could
+define two separate ``step-semi-sensitive'' relations on $\li A \times \li A'$,
+each closed under delays on the left and right respectively. The denotation of
+term precision would then involve the conjunction of those relations. We do not
+pursue this approach in this paper, because (1) it is not clear if the
+step-semi-sensitive relations would be transitive, and (2) we would need to
+reason about the two ordering relations separately for every pair of terms.}
+%
+We cannot forego the property of being a $\theta$-congruence, as without this we
+would not be able to prove properties of the relation using \lob-induction,
+e.g., that $\ext{f}{}$ is a \emph{monotone} function with respect to the
+ordering.
+%and moreover this would mean that $\theta$ would not be a monotone function.
+
+Thus, we choose to sacrifice closure under delays. That is, we will define an
+ordering relation that requires terms to be in ``lock-step''. In order for two
+computations to be related in this ordering, they must have the same stepping
+behavior (or the left-hand side must be an error). To deal with terms that take
+differing numbers of steps, we then define a separate relation called \emph{weak
+bisimilarity} that relates terms that are extensionally equal and may differ by
+a finite number of delays. Then, the semantics of the error ordering for the
+guarded lift monad will involve a combination of these two relations: the
+``closure'' of the lock-step error ordering under weak bisimilarity on both
+sides.
+% Although the combined relation will not be transitive (for the same reason that
+% $\semltbad$ is not transitive), this...
+This decomposition has the advantage that we can recover some transitive
+reasoning and push much of the reasoning about stepping to the margins of the
+development.
+
+% Doing so, we define a \emph{lock-step} error ordering, where roughly speaking,
+% in order for computations to be related, they must have the same stepping
+% behavior. We then formulate a separate relation, \emph{weak bisimilarity}, that
+% relates computations that are extensionally equal and may differ only in their
+% stepping behavior. % up to a finite number of \delta's
+% Finally, the semantics of term precision will involve a combination of these two
+% relations, a sort of closure of the lock-step ordering under weak bisimilarity
+% on both sides. This decomposition has the advantage that we can recover some
+% transitive reasoning and push the parts involving stepping to the margins of the
+% development.
+
+\subsection{Resolution: Splitting Error Ordering and Bisimilarity}\label{sec:lock-step-and-weak-bisim}
+
+As mentioned, to recover transitive reasoning we will decompose the error
+ordering on the guarded lift monad into two separate relations. The first is the
+\emph{lock-step error-ordering} (also called the ``step-sensitive error
+ordering''), the idea being that two computations $l$ and $l'$ are related if
+they are in lock-step with regard to their intensional behavior, up to $l$
+erroring. This will be the partial ordering with which we equip the guarded lift monad.
+Figure \ref{fig:lock-step-error-ordering} gives the definition.
+
+\begin{figure}
+    \fbox{$l_1 \ltls l_2$}
+    \begin{align*}
+        &\eta\, x \ltls \eta\, y \text{ if } 
+            x \mathbin{R} y \\
+        %		
+        &\mho \ltls l' \\
+        %
+        &\theta\, \tilde{l} \ltls \theta\, \tilde{l'} \text{ if } 
+            \later_t (\tilde{l}_t \ltls \tilde{l'}_t)
+    \end{align*}
+    \caption{The lock-step error ordering.}
+    \label{fig:lock-step-error-ordering}
+\end{figure}
+
+When both sides are $\eta$, then we check that the returned values are related
+in $R$. The error term $\mho$ is below everything. Lastly, if both sides step
+(i.e., are a $\theta$) then we compare the resulting computations one time step
+later.
+%
+It is straightforward to prove using \lob-induction that this relation is
+reflexive, transitive and anti-symmetric given that the underyling relation $R$
+has those properties. The lock-step ordering is therefore the partial ordering
+we will associate with $\li X$.
+%
+We similarly define a heterogeneous version of this ordering between the lifts
+of two different types $X$ and $Y$, parameterized by a relation $R$ between $X$
+and $Y$.
+
+The second relation we define on the guarded lift monad is \emph{weak bisimilarity}.
+% For a type $X$, we define a relation on $\li X$, called ``weak bisimilarity",
+% written $l \bisim l'$. 
+Intuitively, we say $l \bisim l'$ if they are the same ``up to delays''.
+%
+Weak bisimilarity on $\li X$ is parameterized by an underlying relation $R$ on $X$
+and is defined by guarded fixpoint as shown in Figure \ref{fig:weak-bisimilarity}.
+
+\begin{figure}
+    \fbox{$l_1 \bisim l_2$}
+    \begin{align*}
+        &\mho \bisim \mho \\
+      %
+        &\eta\, x \bisim \eta\, y \text{ if } 
+          x \mathbin{R} y \\
+      %		
+        &\theta\, \tilde{x} \bisim \theta\, \tilde{y} \text{ if } 
+          \later_t (\tilde{x}_t \bisim \tilde{y}_t) \\
+      %	
+        &\theta\, \tilde{x} \bisim \mho \text{ if } 
+          \theta\, \tilde{x} = \delta^n(\mho) \text { for some $n$ } \\
+      %	
+        &\theta\, \tilde{x} \bisim \eta\, y \text{ if }
+          (\theta\, \tilde{x} = \delta^n(\eta\, x))
+        \text { for some $n$ and $x : X$ such that $x \mathbin{R} y$ } \\
+      %
+        &\mho \bisim \theta\, \tilde{y} \text { if } 
+          \theta\, \tilde{y} = \delta^n(\mho) \text { for some $n$ } \\
+      %	
+        &\eta\, x \bisim \theta\, \tilde{y} \text { if }
+          (\theta\, \tilde{y} = \delta^n (\eta\, y))
+        \text { for some $n$ and $y : X$ such that $x \mathbin{R} y$ }
+      \end{align*}
+    \caption{The weak bisimilarity relation.}
+    \label{fig:weak-bisimilarity}
+\end{figure}
+
+When both sides are $\eta$, then we ensure that the underlying values are
+bisimilar in the underlying bisimilarity relation on $A$. When one side is a
+$\theta$ and the other is $\eta x$ (i.e., one side steps), we stipulate that the
+$\theta$-term runs to $\eta y$ where $x$ is related to $y$. When one side is
+$\theta$ and the other $\mho$, we require that the $\theta$-term runs to $\mho$.
+If both sides step, then we allow one time step to pass and compare the
+resulting terms. This definition captures the intuition of two terms being
+related up to delays.
+
+It can be shown (by \lob-induction) that weak bisimilarity is reflexive and
+symmetric. We can also show that this relation is \emph{not} transitive. The
+argument is the same as that used to show that the step-insensitive error
+ordering $\semltbad$ described above is not transitive. In particular, we
+observe that weak bisimilarity is a congruence with respect to $\theta$, and we
+can show by $\lob$-induction that it is closed under delay on both sides. Thus,
+we may appeal to the no-go theorem as before.
+% We appeal to the no-go theorem given above
+
+% This issue of non-transitivity will be resolved when we consider the relation's \emph{globalization}.
+
+Now we can define the denotation of term precision for closed terms of type
+$\nat$ as follows:
+%
+\[ l_1 \leq^{\bisim} l_2 := 
+  \Sigma_{l_1', l_2' \in \li \mathbb{N}}. l_1 \bisim l_1' \ltls l_2' \bisim l_2. \]
+
+Importantly, we note that as with the original ordering $\semltbad$ defined at
+the beginning of this section, the relation just defined is not transitive
+(again as a result of the above no-go theorem). It may therefore seem that we
+have not solved the original issue we faced. In a sense, this is true. The model
+of gradual typing that we end up with will not support \emph{extensional}
+compositional reasoning. However, by decomposing the denotation of term
+precision in the above manner, we can employ \emph{intensional} compositional
+reasoning by working with the lock-step error ordering. Thus, the tentative plan
+going forward will be to carry out the proofs compositionally using the
+lock-step ordering. We will then apply the closure under weak bisimilarity in
+the denotation of term precision given above to handle the aspects involving
+stepping.
+
+% ...to carry out the proofs using the lock-step ordering and handle the stepping
+% behavior separately via weak bisimilarity. In the end we combine everything
+% together using the above denotation for term precision
+
+% Indeed, we will need to weaken our abstract notion of categorical model to no
+% longer require transitivity of term precision. On the other hand, we now have
+% the ability to apply transitive reasoning when working with the lock-step
+% ordering. This allows us to carry out complex proofs compositionally, a key goal
+% of our framework.
 
 \subsection{Modeling Type and Term Precision}
-
+\max{TODO: figure out where this stuff goes}
 % We will model type precision $c : A \ltdyn A'$ as a relation between the sets
@@
 ordering for which the error element is the bottom element, and the map
-$\theta_B \laterhs B \to B$ is now required to be monotone. 
-%
-\footnote{This requires us to define a notion of ``later'' for partially-ordered
-sets $X$. The elements are simply ``later''-elements of $X$, and the ordering is
-the obvious one, i.e., two later-elements are related if after one time step,
-the elements they step to are related.}
+$\theta_B : \laterhs B \to B$ is now required to be monotone. 
 %
@@
 
-\begin{figure*}
-    \begin{align*}
-        &\eta\, x \semltbad \eta\, y \text{ if } 
-            x \semlt y \\
-        %		
-        &\mho \semltbad l \\
-        %
-        &\theta\, \tilde{l} \semltbad \theta\, \tilde{l'} \text{ if } 
-            \later_t (\tilde{l}_t \semltbad \tilde{l'}_t) \\
-        %	
-        &\theta\, \tilde{l} \semltbad \mho \text{ if } 
-            \theta\, \tilde{l} = \delta^n(\mho) \text { for some $n$ } \\
-        %	
-        &\theta\, \tilde{l} \semltbad \eta\, y \text{ if }
-            (\theta\, \tilde{l} = \delta^n(\eta\, x))
-        \text { for some $n$ and $x : A$ such that $x \le y$ } \\
-        %
-        &\theta\, \tilde{l} \semltbad \mho \text { if } 
-            \theta\, \tilde{l} = \delta^n(\mho) \text { for some $n$ } \\
-        %	
-        &\eta\, x \semltbad \theta\, \tilde{l'} \text { if }
-            (\theta\, \tilde{l'} = \delta^n (\eta\, y))
-        \text { for some $n$ and $y : A'$ such that $x \le y$ }
-    \end{align*}
-    \caption{The step-insensitive error ordering.}
-    \label{fig:step-insensitive-error-ordering}
-\end{figure*}
-
 Two computations that immediately return $(\eta)$ are related if the returned
@@
 terminate with error.
-
-\subsection{Problem: Failure of Transitivity}
-
-The key issue with the above ordering is it is \emph{not} transitive, as we will
-show below. We begin with a general ``no-go'' theorem about relations defined on
-the guarded lift monad:
-
-\begin{theorem}[No-go theorem for relations on $\li X$]\label{thm:no-go}
-    Let $R$ be a relation on the guarded lift monad satisfying the following
-    three properties:
-    \begin{enumerate}
-        \item Transitivity
-
-        \item Congruence with respect to $\theta$: If $\later_t (\tilde{x}_t
-        \binrel{R} \tilde{y}_t)$, then $\theta(\tilde{x}) \binrel{R}
-        \theta(\tilde{y})$
-
-        \item Closure under delay on both sides: If $x \binrel{R} y$, then $x
-        \binrel{R} \delta y$ and $\delta x  \binrel{R} y$
-    \end{enumerate}
-
-    Then $R$ is trivial, i.e., $x \binrel{R} y$ for all $x : \li X$ and $y : \li
-    Y$.
-
-\end{theorem}
-
-This theorem follows from two ``one-sided'' lemmas where we assume that the
-relation $R$ is closed under delay on the left (resp. right) and show that
-$(\fix\, \theta) \binrel{R} y$ for all $y$ (resp. $x \binrel{R} (\fix\, \theta)$
-for all $x$). These lemmas in turn are proved by \lob-induction.
-
-The theorem implies that the relation $\semltbad$ defined on $\li X \times \li
-Y$ is \emph{not} transitive whenever the underlying relation $\semlt$ on $X$ and
-$Y$ is not trivial, i.e., when there are $x_0$ and $y_0$ that are not related in
-the relation $\semlt$ on $X$ and $Y$. To show this, first observe that
-$\semltbad$ satisfies properties (2) and (3) in the above theorem. Congruence
-with respect to $\theta$ is immediate from the definition, while closure under
-delay can be proved using \lob-induction.
-% TODO the proof of this uses tick irrelevance
-Then since $x_0$ is not related to $y_0$, we have that $\eta\, x_0$ is not
-related to $\eta\, y_0$ in $\semltbad$, i.e., $\semltbad$ is non-trivial. But if
-it were transitive, then by the theorem, it would be trivial, a contradiction.
-Thus, we are able to refute transitivity of $\semltbad$ provided the underlying
-relation is non-trivial.
-
-\eric{Read this}
-The lack of transitivity presents a major barrier towards giving a compositional
-semantics to gradual typing: if our relations are not transitive, then we cannot
-compose squares horizontally. But we have argued above that horizontal
-composition is essential for modelling the axioms of gradual typing in a
-syntax-independent manner.
-
-% In particular, to model the UpL/UpR/DnL/DnR rules
-% for casts in the presence of transitivity, it is sufficient to establish the
-% validity of simpler rules that do not ``build in'' composition. We can then use
-% transitivity to derive the original versions of the rules. On the other hand,
-% without transitivity we must instead validate the cast rules in the model ``from
-% scratch''. In fact, we cannot even define the semantic interpretation of type
-% precision in a syntax-independent manner. The issue is that in the definition of
-% relation, the requirement that it be representable now involves quantifying over
-% all other relations, which is circular.
-
-% So, although this approach would suffice for proving graduality, it lacks the
-% compositionality that we seek in a reusable framework for the semantics of
-% gradual typing.
-
-%Because it is convenient to make use of transitive reasoning in proving graduality, ...
-We therefore seek to formulate an ordering relation on $\li A \times \li A'$
-that \emph{is} transitive. The theorem above tells us, however, that any such
-relation must either not be a congruence with respect to $\theta$, or must not
-be closed under delay.
-\footnote{Technically, we only need to give up closure under delay on \emph{one}
-side, as the above theorem requires $R$ to be closed on both sides. We could
-define two separate ``step-semi-sensitive'' relations on $\li A \times \li A'$,
-each closed under delays on the left and right respectively. The denotation of
-term precision would then involve the conjunction of those relations. We do not
-pursue this approach in this paper, because (1) it is not clear if the
-step-semi-sensitive relations would be transitive, and (2) we would need to
-reason about the two ordering relations separately for every pair of terms.}
-%
-We cannot forego the property of being a $\theta$-congruence, as without this we
-would not be able to prove properties of the relation using \lob-induction,
-e.g., that $\ext{f}{}$ is a \emph{monotone} function with respect to the
-ordering.
-%and moreover this would mean that $\theta$ would not be a monotone function.
-
-Thus, we choose to sacrifice closure under delays. That is, we will define an
-ordering relation that requires terms to be in ``lock-step''. In order for two
-computations to be related in this ordering, they must have the same stepping
-behavior (or the left-hand side must be an error). To deal with terms that take
-differing numbers of steps, we then define a separate relation called \emph{weak
-bisimilarity} that relates terms that are extensionally equal and may differ by
-a finite number of delays. Then, the semantics of the error ordering for the
-guarded lift monad will involve a combination of these two relations: the
-``closure'' of the lock-step error ordering under weak bisimilarity on both
-sides.
-% Although the combined relation will not be transitive (for the same reason that
-% $\semltbad$ is not transitive), this...
-This decomposition has the advantage that we can recover some transitive
-reasoning and push much of the reasoning about stepping to the margins of the
-development.
-
-% Doing so, we define a \emph{lock-step} error ordering, where roughly speaking,
-% in order for computations to be related, they must have the same stepping
-% behavior. We then formulate a separate relation, \emph{weak bisimilarity}, that
-% relates computations that are extensionally equal and may differ only in their
-% stepping behavior. % up to a finite number of \delta's
-% Finally, the semantics of term precision will involve a combination of these two
-% relations, a sort of closure of the lock-step ordering under weak bisimilarity
-% on both sides. This decomposition has the advantage that we can recover some
-% transitive reasoning and push the parts involving stepping to the margins of the
-% development.
-
-\subsection{Solution: Splitting into Two Relations}\label{sec:lock-step-and-weak-bisim}
-
-As mentioned, to recover transitive reasoning we will decompose the error
-ordering on the guarded lift monad into two separate relations. The first is the
-\emph{lock-step error-ordering} (also called the ``step-sensitive error
-ordering''), the idea being that two computations $l$ and $l'$ are related if
-they are in lock-step with regard to their intensional behavior, up to $l$
-erroring. This will be the partial ordering with which we equip the guarded lift monad.
-Figure \ref{fig:lock-step-error-ordering} gives the definition.
-
-\begin{figure}
-    \fbox{$l_1 \ltls l_2$}
-    \begin{align*}
-        &\eta\, x \ltls \eta\, y \text{ if } 
-            x \mathbin{R} y \\
-        %		
-        &\mho \ltls l' \\
-        %
-        &\theta\, \tilde{l} \ltls \theta\, \tilde{l'} \text{ if } 
-            \later_t (\tilde{l}_t \ltls \tilde{l'}_t)
-    \end{align*}
-    \caption{The lock-step error ordering.}
-    \label{fig:lock-step-error-ordering}
-\end{figure}
-
-When both sides are $\eta$, then we check that the returned values are related
-in $R$. The error term $\mho$ is below everything. Lastly, if both sides step
-(i.e., are a $\theta$) then we compare the resulting computations one time step
-later.
-%
-It is straightforward to prove using \lob-induction that this relation is
-reflexive, transitive and anti-symmetric given that the underyling relation $R$
-has those properties. The lock-step ordering is therefore the partial ordering
-we will associate with $\li X$.
-%
-We similarly define a heterogeneous version of this ordering between the lifts
-of two different types $X$ and $Y$, parameterized by a relation $R$ between $X$
-and $Y$.
-
-The second relation we define on the guarded lift monad is \emph{weak bisimilarity}.
-% For a type $X$, we define a relation on $\li X$, called ``weak bisimilarity",
-% written $l \bisim l'$. 
-Intuitively, we say $l \bisim l'$ if they are the same ``up to delays''.
-%
-Weak bisimilarity on $\li X$ is parameterized by an underlying relation $R$ on $X$
-and is defined by guarded fixpoint as shown in Figure \ref{fig:weak-bisimilarity}.
-
-\begin{figure}
-    \fbox{$l_1 \bisim l_2$}
-    \begin{align*}
-        &\mho \bisim \mho \\
-      %
-        &\eta\, x \bisim \eta\, y \text{ if } 
-          x \mathbin{R} y \\
-      %		
-        &\theta\, \tilde{x} \bisim \theta\, \tilde{y} \text{ if } 
-          \later_t (\tilde{x}_t \bisim \tilde{y}_t) \\
-      %	
-        &\theta\, \tilde{x} \bisim \mho \text{ if } 
-          \theta\, \tilde{x} = \delta^n(\mho) \text { for some $n$ } \\
-      %	
-        &\theta\, \tilde{x} \bisim \eta\, y \text{ if }
-          (\theta\, \tilde{x} = \delta^n(\eta\, x))
-        \text { for some $n$ and $x : X$ such that $x \mathbin{R} y$ } \\
-      %
-        &\mho \bisim \theta\, \tilde{y} \text { if } 
-          \theta\, \tilde{y} = \delta^n(\mho) \text { for some $n$ } \\
-      %	
-        &\eta\, x \bisim \theta\, \tilde{y} \text { if }
-          (\theta\, \tilde{y} = \delta^n (\eta\, y))
-        \text { for some $n$ and $y : X$ such that $x \mathbin{R} y$ }
-      \end{align*}
-    \caption{The weak bisimilarity relation.}
-    \label{fig:weak-bisimilarity}
-\end{figure}
-
-When both sides are $\eta$, then we ensure that the underlying values are
-bisimilar in the underlying bisimilarity relation on $A$. When one side is a
-$\theta$ and the other is $\eta x$ (i.e., one side steps), we stipulate that the
-$\theta$-term runs to $\eta y$ where $x$ is related to $y$. When one side is
-$\theta$ and the other $\mho$, we require that the $\theta$-term runs to $\mho$.
-If both sides step, then we allow one time step to pass and compare the
-resulting terms. This definition captures the intuition of two terms being
-related up to delays.
-
-It can be shown (by \lob-induction) that weak bisimilarity is reflexive and
-symmetric. We can also show that this relation is \emph{not} transitive. The
-argument is the same as that used to show that the step-insensitive error
-ordering $\semltbad$ described above is not transitive. In particular, we
-observe that weak bisimilarity is a congruence with respect to $\theta$, and we
-can show by $\lob$-induction that it is closed under delay on both sides. Thus,
-we may appeal to the no-go theorem as before.
-% We appeal to the no-go theorem given above
-
-% This issue of non-transitivity will be resolved when we consider the relation's \emph{globalization}.
-
-Now we can define the denotation of term precision for closed terms of type
-$\nat$ as follows:
-%
-\[ l_1 \leq^{\bisim} l_2 := 
-  \Sigma_{l_1', l_2' \in \li \mathbb{N}}. l_1 \bisim l_1' \ltls l_2' \bisim l_2. \]
-
-Importantly, we note that as with the original ordering $\semltbad$ defined at
-the beginning of this section, the relation just defined is not transitive
-(again as a result of the above no-go theorem). It may therefore seem that we
-have not solved the original issue we faced. In a sense, this is true. The model
-of gradual typing that we end up with will not support \emph{extensional}
-compositional reasoning. However, by decomposing the denotation of term
-precision in the above manner, we can employ \emph{intensional} compositional
-reasoning by working with the lock-step error ordering. Thus, the tentative plan
-going forward will be to carry out the proofs compositionally using the
-lock-step ordering. We will then apply the closure under weak bisimilarity in
-the denotation of term precision given above to handle the aspects involving
-stepping.
-
-% ...to carry out the proofs using the lock-step ordering and handle the stepping
-% behavior separately via weak bisimilarity. In the end we combine everything
-% together using the above denotation for term precision
-
-% Indeed, we will need to weaken our abstract notion of categorical model to no
-% longer require transitivity of term precision. On the other hand, we now have
-% the ability to apply transitive reasoning when working with the lock-step
-% ordering. This allows us to carry out complex proofs compositionally, a key goal
-% of our framework.
 

### 2024-07-09 paper-new/towards-relational-model.tex
fix typo

@@
 $\theta$-congruence and step-insensitive on both sides, but not
-generally trivial: e.g., for the flat poset on $\mathbb N$, $\eta 0 \semltdbad
+generally trivial: e.g., for the flat poset on $\mathbb N$, $\eta 0 \semltbad
 \eta 1$ is false.

### 2024-07-10 paper-new/concrete-term-model.tex
discussion of adequacy for term semantics

@@
 
-\max{TODO: Eric needs to add a discussion of adequacy/clock stuff here}
+% \max{TODO: Eric needs to add a discussion of adequacy/clock stuff here}
+\eric{Check this}
+
+We now define from the above term model a ``big-step'' semantics. More
+concretely, our goal is to define a function 
+$-\Downarrow : \{M \,|\, \cdot \vdash M : \nat \} \rightharpoonup \mathbb{N} + {\mho}$.
+
+To begin, we must first discuss another aspect of guarded type theory,
+\emph{clocks}, originally introduced in \cite{atkey-mcbride2013}. The
+constructions of guarded type theory ($\later$, $\nxt$, and $\fix$) we have been
+using may be indexed by objects called \emph{clocks}. A clock intuitively serves
+as a reference relative to which steps are counted. For instance, given a clock
+$k$ and type $X$, the type $\later^k X$ represents a value of type $X$ one unit
+of time in the future according to clock $k$.
+%
+Up to this point, all uses of guarded constructs have been indexed by a single
+clock $k$. In particular, we have for each clock $k$ a type $\li^k X$. The
+notion of \emph{clock quantification} allows us to pass from the guarded setting
+to the usual set-theoretic world. Clock quantification enables us to encode
+coinductive types using guarded recursion. That is, given a type $A$ with a free
+variable $k : \Clock$, we can form the type $A^{gl} := \forall k. A$. Given a
+functor $F$ that satisfies a technical condition, if $A$ is a guarded recursive
+type satisfying $A \cong F(\later^k A)$, then the type $A^{gl}$ has a final
+$F$-coalgebra structure (see Theorem 4.3 of
+\cite{kristensen-mogelberg-vezzosi2022}).
+
+Consider the functor $F(X) = \mathbb{N} + {\mho} + X$. By the definition of the
+guarded lift, we have that
+% 
+$\li^k \mathbb{N} 
+  \cong \mathbb{N} + {\mho} + \later^k (\li \mathbb{N}) 
+  \cong F(\later \li^k \mathbb{N})$.
+%
+Furthermore, it can be shown that the functor $F$ satisfies the technical
+condition in the statement of the above-mentioned theorem. Thus, the theorem
+implies that the type $\li^{gl} \mathbb{N} := \forall k. \li^k \mathbb{N}$ is a
+final $F$-coalgebra.
+%
+Given a type $Y$, Capretta's \emph{delay monad} \cite{lmcs:2265},
+which we write as $\text{Delay}(Y)$, is defined as the coinductive type
+generated by $\tnow : Y \to \delay(Y)$ and $\tlater : \delay(Y) \to \delay(X)$.
+We have that $\delay (\mathbb{N} + {\mho})$ is a final coalgebra for the functor
+$F$ defined above. This means $\li^{gl} \mathbb{N}$ is isomorphic to
+$\delay(\mathbb{N} + {\mho})$.
+
+We can define a notion of termination in $i$ steps for the Delay monad $d \da^i
+n_?$ where $d : \delay(\mathbb{N} + {\mho})$ and $n_? : \mathbb{N} + {\mho}$. From
+this, can define a partial function $\delay (\mathbb{N} + {\mho})
+\rightharpoonup \mathbb{N} + {\mho}$ by existentially quantifying over $i$
+for which $d \da^i n_?$.
+
+Now consider the function $\{M \,|\, \cdot \vdash M : \nat \} \to \li^{gl}
+\mathbb{N}$ defined by $\lambda M. \lambda k .\sem{M}$. Composing this with the
+above partial function yields the desired partial function $-\Downarrow : \{M
+\,|\, \cdot \vdash M : \nat \} \rightharpoonup \mathbb{N} + {\mho}$. It is
+then straightforward to see that this big-step semantics respects the equational
+theory and that the denotations of the diverging term $\Omega$, the error term
+$\mho$, and a syntactic natural number value $n$ are all distinct.

### 2024-07-10 paper-new/concrete-relational-model.tex
revise sections on relational model

@@
-\section{Towards a Full Relational Model}\label{sec:towards-relational-model}
+\section{Constructing a Concrete Model}\label{sec:concrete-model}
 
-% In the previous section, we gave a denotational semantics to the terms of the
-% gradually-typed lambda calculus using the tools of SGDT. We defined a big-step
-% semantics for closed terms of type $\nat$, as functions from $\mathbb{N} \to
-% ((\mathbb{N} + 1) + 1)$.
+In this section, we complete the construction of a relational model of gradual
+typing. To manage the complexity, we carry out the construction in phases,
+beginning with a relational model consisting of partially-ordered sets with a
+notion of bisimilarity and then adding requirements for a structured set of
+delays on each type as well as a notion of representability for relations that
+models the cast rules in the intensional setting.
 
-\eric{Review this section}
+\subsection{Phase One: Predomains and Error Domains}
 
-In the previous section, we gave a denotational call-by-push-value model for the
-terms of the gradually-typed lambda calculus using the tools of SGDT. In this
-section, we build on the ideas used in defining a term model with the goal of
-defining a model of the full gradually-typed lambda calculus, i.e., one that has
-constructs that model type and term precision.
-
-Recall that the cast calculus introduced in Section \ref{sec:GTLC} has not only
-terms but also an \emph{inequational theory} with axioms for type precision and
-term precision that constitute an abstract formulation of the graduality
-property. Thus, our goal now is to build a denotational model interpreting these
-axioms.
-%
-
-\subsection{Modeling Type and Term Precision}
-
-% We will model type precision $c : A \ltdyn A'$ as a relation between the sets
-% $\sem{A}$ and $\sem{A'}$.
-
-To model term precision, we begin by equipping the denotation of every type with
-an ordering relation. Since term precision is reflexive and transitive, and
-since we identify terms that are equi-precise, we model value types as
-partially-ordered sets and values $\Gamma \vdash V : A$ as \emph{monotone}
-functions. Analogously, every error domain is now equipped with a partial
-ordering for which the error element is the bottom element, and the map
-$\theta_B \laterhs B \to B$ is now required to be monotone. 
-%
-\footnote{This requires us to define a notion of ``later'' for partially-ordered
-sets $X$. The elements are simply ``later''-elements of $X$, and the ordering is
-the obvious one, i.e., two later-elements are related if after one time step,
-the elements they step to are related.}
-%
-Morphisms of error domains are morphisms of the underlying partially-ordered
-sets that preserve the error element and $\theta$ map, as was the case in the
-previous section.
-
-We model a type precision relation $c : A \ltdyn A'$ as a \emph{monotone relation}, i.e., a
-relation $c$ that is upward-closed under the relation on $A'$ and
-downward-closed under the relation on $A$. We denote such a relation between $A$
-and $A'$ by $c : A \rel A'$. The relation on the poset $A$ is denoted $r(A)$.
-% This extends to products $c_1 \times c_2$ in the obvious way.
-Composition of relations on predomains is the usual relational composition
-(which is truncated to be propositional).
-
-We likewise define a notion of \emph{error domain relation} between error
-domains $B$ and $B'$. An error domain relation $d : B \rel B'$ is a monotone
-relation on the underlying sets that preserves error and respects $\theta$. More
-formally:
+In the previous section, we introduced the notion of weak bisimilarity as a
+relation on the guarded lift monad. Recall that weak bisimilarity on $\li A$ is
+parameterized by a relation on $A$. Thus, in order to define a compositional
+model for all types, we need to equip the objects (i.e. the posets and error
+domains) with a reflexive, symmetric bisimilarity relation. This leads to our
+first new definition:
 
 \begin{definition}
-  Let $B$ and $B'$ be error domains. An \emph{error domain relation} between
+A \textbf{predomain} $A$ consists of a set $A$ along with two relations:
+\begin{itemize}
+    \item A partial order $\le_A$.
+    \item A reflexive, symmetric ``bisimilarity'' relation $\bisim_A$.
+\end{itemize}
+\end{definition}
+
+We will generally write $A$ for both a predomain and its underlying set; if we
+need to emphasize the difference we will write $|A|$ for the underlying set of
+$A$.
+
+Given a predomain $A$, we can form the predomain $\later A$. The underlying set
+is $\later |A|$ and the relation is defined in the obvious way, i.e., $\tilde{x}
+\le_{\later A} \tilde{x'}$ iff $\later_t(\tilde{x}_t \le_A \tilde{x'}_t)$.
+Likewise for bisimilarity.
+%
+We also give a predomain structure to the natural numbers $\mathbb{N}$, where
+both the ordering and the bisimilarity relation are equality.
+%
+Morphisms of predomains are functions between the underlying sets that preserve
+the ordering and the bisimilarity relation. More formally:
+%
+\begin{definition}
+Let $A$ and $A'$ be predomains. A \textbf{morphism of predomains} $f : A \to A'$
+is a function between the underlying sets such that for all $x, x'$, if $x \le_A
+x'$, then $f(x) \le f(x')$, and if $x \bisim_A x'$, then $f(x) \bisim_{A'}
+f(x')$.
+\end{definition}
+
+We also equip error domains with a bisimilarity relation, and we impose a
+condition relating the $\theta$ morphism to the bisimilarity relation. More
+concretely:
+%
+\begin{definition}
+An \textbf{error domain} $B$ consists of a predomain $B$ along with the following data:
+\begin{itemize}
+    \item A distinguished ``error'' element $\mho_B \in B$
+    \item A morphism of predomains $\theta_B \colon \later B \to B$
+    \item For all $x : B$ we have $\theta_B(\nxt\, x) \bisim_B x$
+\end{itemize}
+\end{definition}
+%
+For an error domain $B$, we define the predomain morphism $\delta_B := \theta_B
+\circ \nxt$.
+%
+A \textbf{morphism of error domains} is simply a morphism of the underlying
+predomains that preserves the error element and the $\theta$ map.
+
+%
+The definition of relations between predomains is the same as that between
+posets introduced in the previous section, as the bisimilarity relations of the
+predomains are not involved in the definition of monotone relation.
+
+Next we define a notion of \emph{error domain relation} between error domains $B$ and
+$B'$. An error domain relation $d : B \rel B'$ is a relation on the underlying
+posets that preserves error and respects $\theta$. More formally:
+%
+\begin{definition}
+  Let $B$ and $B'$ be error domains. An \textbf{error domain relation} between
   $B$ and $B'$ is a monotone relation $d$ on the underlying partially-ordered sets such that:
@@
 \end{definition}
+% Likewise, the definition of error domain relations is unchanged.
 
+The definitions of squares for predomains are also unchanged from the previous
+section, except that the morphisms on the left and right, are now morphisms of
+predomains. We likewise define a notion of \textbf{error domain square}: Let
+$d_i : B_i \rel B_i'$ and $d_o : B_o \rel B_o'$ be error domain relations, and
+let $\phi : B_i \multimap B_o$ and $\phi' : B_i' \multimap B_o'$ be morphisms of
+error domains. We say that there is an error domain square $\phi \ltsq{d_i}{d_o}
+\phi'$ exactly when there is a predomain square involving the underlying
+predomain morphisms and relations.
+
+\begin{comment}
+We define a (monotone) relation on predomains $A$ and $A'$ to be a relation on
+the underlying sets that is downward-closed under $\le_A$ and upward-closed
+under $\le_{A'}$. More formally:
+%
+\begin{definition}
+Let $A$ and $A'$ be predomains. A \emph{predomain relation} between $A$ and $A'$
+is a relation $R$ between the underlying sets such that:
+\begin{enumerate}
+    \item (Downward closure): For all $x_1, x_2 \in A$ and $y \in A'$,
+    if $x_1 \le_A x_2$ and $x_2 \mathbin{R} y$, then $x_1 \mathbin{R} y$.
+    \item (Upward closure): For all $x \in A$ and $y_1, y_2 \in A'$,
+    if $x \mathbin{R} y_1$ and $y_1 \le_{A'} y_2$, then $x \mathbin{R} y_2$.
+\end{enumerate}
+\end{definition}
+%
+Composition of relations on predomains is the usual relational composition
+(which is truncated to be propositional). Similarly, we define a (monotone)
+relation on error domains to be a relation on the underlying predomains that
+respects error and preserves $\theta$:
+%
+\begin{definition}
+    Let $B$ and $B'$ be error domains. An \emph{error domain relation} between
+    $B$ and $B'$ is a relation $R$ between the underlying predomains such that
+    \begin{enumerate}
+       \item (Respects error): For all $y \in B'$, we have $\mho_B \mathbin{R} y$.
+       \item (Preserves $\theta$): For all $\tilde{x}$ in $\later B$ and $\tilde{y} \in \later B'$,
+       if $\later_t( \tilde{x}_t \mathbin{R} \tilde{y}_t )$ then
+       $\theta_B(\tilde{x}) \mathbin{R} \theta_{B'}(\tilde{y})$.
+    \end{enumerate}
+\end{definition}
+%
 Because of a technical restriction involving the interaction between
@@
 \end{mathpar}
+%
+% We note that this composition has the following universal property.
+%
 
-% That is,
-% $(\mho_B, y) \in d$ for all $y : B'$ and if $\laterhs_t ((\tilde{x}_t,
-% \tilde{y}_t) \in d)$, then $(\theta_B (\tilde{x}), \theta_{B'} (\tilde{y})) \in
-% d$.
+We now describe the squares. Suppose we are given predomains $A_i, A_o, A_i'$,
+and $A_o'$, relations $c_i : A_i \rel A_i'$ and $c_o : A_o \rel A_o'$, and
+morphisms $f : A_i \to A_o, f' : A_i' \to A_o'$ Given a square with these
+morphisms and relations, we say that the square commutes, written $f \le f'$, if
+for all $x \in A_i$ and $x' \in A_i'$ with $(x, x') \in c_i$, we have $(f(x),
+f'(x')) \in c_o$. We make the analogous definition for error domains.
 
-The functors $U$, $F$, $\times$, and $\arr$ introduced in the previous
-section extend to relations. The functor $U$ simply takes the underlying
-monotone relation of an error domain relation. Given a relation $c_1 : A_1 \rel
-A_1'$ and $c_2 : A_2 \rel A_2'$, the relation $c_1 \times c_2 : (A_1 \times A_2)
-\rel (A_1' \times A_2')$ is defined by $((x, y), (x', y')) \in (c_1 \times c_2)$
-if and only if $(x, x') \in c_1$ and $(y, y') \in c_2$. Given a relation $c : A
-\rel A'$ and an error domain relation $d : B \rel B'$, the relation $c \arr d :
-(A \arr B) \rel (A' \arr B')$ is the error domain relation defined by $(f, g)
-\in (c \arr d)$ if and only if for all $x : A$ and $x' : A'$ with $(x , x') \in
-c$ we have $(f(x), g(x')) \in Ud$. The fact that this relation preserves error
-and respects $\theta$ follows from the corresponding properties for $d$.
+\end{comment}
+
+\subsubsection{Functors}\label{sec:free-error-domain}
+
+The functors $U$, $\li$, $\times$, and $\arr$ introduced in the previous section
+extend to predomains, error domains, and their relations. The functor $U$ simply
+takes the underlying poset relation of an error domain relation. Given a
+relation $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$, the relation $c_1
+\times c_2 : (A_1 \times A_2) \rel (A_1' \times A_2')$ is defined by $((x, y),
+(x', y')) \in (c_1 \times c_2)$ if and only if $(x, x') \in c_1$ and $(y, y')
+\in c_2$. Given a relation $c : A \rel A'$ and an error domain relation $d : B
+\rel B'$, the relation $c \arr d : (A \arr B) \rel (A' \arr B')$ is the error
+domain relation defined by $(f, g) \in (c \arr d)$ if and only if for all $x :
+A$ and $x' : A'$ with $(x , x') \in c$ we have $(f(x), g(x')) \in Ud$. The fact
+that this relation preserves error and respects $\theta$ follows from the
+corresponding properties for $d$.
+
+% Free error domain
+The functor $\li$ takes a predomain $A$ to the free error domain on $A$. We first
+define the predomain $U\li A$. The underlying set of $U\li A$ is defined to be $\li
+|A|$ (the guarded lift monad applied to $|A|$). The ordering relation is the
+lock-step error ordering introduced in the previous section (taking $A = A'$ in
+the definition), and the bisimilarity relation is the weak bisimilarity relation
+on $\li |A|$ also defined in the previous section. With these relations, the
+constructors $\eta$ and $\theta$ of the guarded lift monad are in fact morphisms
+of predomains, i.e., they are monotone and preserve bisimilarity.
+
+We observe that this predomain structure extends to an error domain structure by
+noting that the required error element is given by the constructor $\mho$ and
+the required $\theta$ map is the constructor $\theta$. Lastly, it can be shown
+that the delay morphism $\delta = \theta \circ \nxt$ satisfies $x \bisim
+\delta\, x$ for all $x$ as is required by the definition of error domain.
+
+The monadic bind operation $-^\dagger$ takes a predomain morphism $f : A \to UB$
+to an error domain morphism $\li A \to B$; with this we define the action of $\li$ on
+morphisms: Given $f : A \to A'$ we observe that $\eta \circ f : A \to U\li A'$, so
+we define $\li f = (\eta \circ f)^\dagger : \li A \to \li A'$.
+
+Finally, we define the action of $\li$ on a relation $c$ to be the heterogeneous
+lock-step error ordering between $U\li A$ and $U\li A'$ and note that it by
+definition satisfies the additional requirements of being an error domain
+relation between the error domains $\li A$ and $\li A'$.
+
+% The monad $\li$ decomposes into a free-forgetful adjunction $F \dashv U$. Given
+% a predomain $A$, we define the error domain $FA$ whose underlying predomain is
+% $\li A$, whose error element is $\mho$, and whose $\theta$ map is the
+% constructor $\theta$.
+
+% It is easily verified that $\li A$ is the free error- and later-algebra on the
+% predomain $A$, so we have that $\li$ is left-adjoint to $U$.
+
+% i.e., error domain morphisms from $\li A$ to $B$ are in one-to-one correspondence with
+% predomain morphisms from $A$ to $UB$.
+
+% We define $\delta : U(\li A) \To U(\li A)$ by $\delta(x) = \theta(\nxt x)$.
+% We define $\hat{\delta} : \li A \arr \li A$ by $\hat{\delta} = \ext{(\delta \circ \eta)}{}$.
+% Note that by definition of $\text{ext}$, we have that $\hat{\delta}$ is a morphism of error domains.
+
+%\subsubsection{Lock-Step Error Ordering}\label{sec:lock-step}
+
+% We define the action of $\li$ on a relation $R$ between $A$ and $A'$ to be the
+% ``heterogeneous" version of the lock-step error ordering.
+
+% TODO action of \li on commuting squares
+
+\begin{comment}
+
+% internal hom for predomains and error domains
+Given predomains $A$ and $A'$, we can form the predomain of
+predomain morphisms from $A$ to $A'$, denoted $A \To A'$.
+\begin{itemize}
+    % Should we give the definition involving x and x'?
+    \item The ordering is defined by $f \le_{A \To A'} f'$ iff for all
+    $x \in A$, we have $f(x) \le_{A'} f'(x)$.
+    \item The bisimilarity relation is defined by $f \bisim_{A \To A'} f'$ iff
+    for all $x, x' \in A$ with $x \bisim_{A} x'$, we have $f(x) \bisim_{A'} f'(x')$. 
+\end{itemize}
 %
-We we will postpone the description of the action of $F$ on relations to later
-in this section.
+Given $f : A_1' \to A_1$ and $g : A_2 \to A_2'$ we define the predomain morphism
+$f \To g : (A_1 \To A_2) \to (A_1' \To A_2')$ by $\lambda h. \lambda x'. g(h(f(x')))$.
 
-\subsubsection{Modelling Term Precision}\label{sec:modeling-term-precision}
-To model term precision, we need the notion of \emph{square}, which specifies a
-relation between two monotone functions. The definition is as follows. Let $A_i,
-A_o, A_i', A_o'$ be partially-ordered sets. Let $c_i : A_i \rel A_i'$ and $c_o :
-A_o \rel A_o'$ be monotone relations, and let $f : A_i \to A_o$ and $g : A_i'
-\to A_o'$ be monotone functions. We say that $f \ltsq{c_i}{c_o} g$ if for all $x
-: A_i$ and $y : A_i'$ with $(x, y) \in c_i$, we have $(f(x), g(y)) \in c_o$. We
-visualize this situation as the following square:
+% TODO: include this?
+% The monadic extension operation $\ext{\cdot}{} : (A_1 \To U (\li A_2)) \To (\li A_1 \To U(\li A_2))$
+% is a morphism of predomains from $A_1 \To U(\li A_2)$ to $U(\li A_1) \To U(\li A_2)$, i.e.,
+% it preserves the ordering and bisimilarity relations.
 
-% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBX2kiXSxbMCwxLCJBX28iXSxbMSwwLCJBX2knIl0sWzEsMSwiQV9vJyJdLFsyLDMsImciXSxbMCwxLCJmIiwyXSxbMCwyLCJjX2kiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxLDMsImNfbyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
-\[\begin{tikzcd}[ampersand replacement=\&]
-	{A_i} \& {A_i'} \\
-	{A_o} \& {A_o'}
-	\arrow["{c_i}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-	\arrow["f"', from=1-1, to=2-1]
-	\arrow["g", from=1-2, to=2-2]
-	\arrow["{c_o}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-\end{tikzcd}\]
+% Given a predomain $A$ and error domain $B$, we define $A \arr B := A \To UB$.
+We note that $A \To UB$ carries a natural error domain structure
+(in the below, the lambda is a meta-theoretic notation):
+\begin{itemize}
+    \item The error is given by $\lambda x . \mho_B$
+    \item The $\theta$ operation is defined by
+      \[ \theta_{A \To UB}(\tilde{f}) = \lambda x . \theta_B(\lambda t . \tilde{f}_t(x)). \]
+\end{itemize}
 
-We note that for any $c$, there is a ``vertical identity'' square $\id
-\ltsq{c}{c} \id$, as can be seen by unfolding the definition. Similarly, for any
-montone function $f : A_i \to A_o$, there is a ``horizontal identity'' square $f
-\ltsq{r(A_i)}{r(A_o)}$ by the definition of monotone function.
-We likewise make the analogous definition of square for error domains.
+Given a predomain $A$ and error domain $B$, we define
+$A \arr B$ to be the error domain such that $U(A \arr B) = A \To UB$,
+and whose error and $\theta$ operations are as defined above.
+We can define the functorial action of $\arr$ on morphisms
+$f \arr \phi$ in the obvious way.
+%
+It is easily verified that $A \arr B$ is an exponential of $UB$ by $A$
+in the category of predomains and their morphisms.
 
-Having defined squares, we now discuss the interpretation of the cast rules
-UpL/UpR/DnL/DnR. These rules specify a relationship between the semantics of
-type precision and the corresponding casts (see Section \ref{sec:GTLC}). In
-particular, they specify that the relations are \emph{representable} by the
-casts. The UpL and UpR rules refer to upcasts, and since upcasts are pure, these
-rules apply to \emph{value} relations $c$. The DnL and DnR rules refer to
-downcasts, and since downcasts are effectful, these rules apply to
-\emph{computation} relations $d$.
+\end{comment}
 
-As an example, consider the UpL rule. It states that if $c : A_1 \ltdyn A_2$ and
-$c' : A_2 \ltdyn A_3$ and $x : A_1$ is related to $y : A_3$ via the composition
-$c \comp c'$, then the upcast $\upc{c} x$ is related to $y$ via $c'$. This is
-equivalent to the existence of the following square:
+\begin{comment}
+Lastly, given a relation of predomains $c$ between $A$ and $A'$, and a relation
+of error domains $d$ between $B$ and $B'$, we define the relation $c \arr d$
+between $A \arr B$ and $A' \arr B'$ in the obvious way, i.e., $f \in A \arr B$
+is related to $g \in A' \arr B'$ iff for all $x \in A$ and $x' \in A'$ with
+$(x, x') \in c$, we have $(f(x), g(x')) \in Ud$.
+%
+One can verify that this relation is indeed a relation of error domains
+in that it respects error and preserves $\theta$.
+\end{comment}
 
-% https://q.uiver.app/#q=WzAsNSxbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMyw0LCJjJyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMywiXFx1cGN7Y30iLDJdLFsyLDQsIlxcaWQiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
-\[\begin{tikzcd}[ampersand replacement=\&]
-	{A_1} \& {A_2} \& {A_3} \\
-	{A_2} \&\& {A_3}
-	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-	\arrow["{\upc{c}}"', from=1-1, to=2-1]
-	\arrow["{c'}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
-	\arrow["\id", from=1-3, to=2-3]
-	\arrow["{c'}"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
-\end{tikzcd}\]
+\subsection{Phase Two: Perturbations and Quasi-Representable Relations}
 
-The three squares corresponding to the other rules are analogous. Observe that
-although this is really a fact about the relationship between $c$ and $\upc{c}$,
-we must universally quantify over all relations $c'$. This means we cannot
-require the existence of this square as part of the definition of a monotone
-relation, as it is self-referential. This would seem to imply that we cannot
-give a syntax-independent semantic model for gradual typing. However, the key
-observation is that through compositionality, we can \emph{derive} the above
-squares from simpler ones that do not involve composition of relations. For
-example, below is the simpler square corresponding to UpL:
+We now introduce the additional definitions needed to complete the construction
+of our model. Recall from Section \ref{sec:adjusting-cast-rules} that in order
+to establish the DnL rule for the downcast of $\iarr$, we needed to adjust the
+rule by inserting a ``delay'' on the right-hand side. A similar adjustment is
+needed for the DnR rule for $\iarr$. Moreover, the need to insert delays impacts
+the semantics of the cast rules for \emph{all} relations, because of the
+functorial nature of casts. That is, the upcast at a function type $c_i \ra c_o$
+involves a downcast in the domain and an upcast in the codomain. This has two
+consequences: first, the squares corresponding to the rules UpL and UpR may also
+require the insertion of a delay. Second, we need to be able to insert
+``higher-order'' delays in a way that follows the structures of the casts.
 
-% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzIiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzIiXSxbMCwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJyKEFfMikiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxcdXBje2N9IiwyXSxbMiwzLCJcXGlkIl1d
-\[\begin{tikzcd}[ampersand replacement=\&]
-	{A_1} \& {A_2} \\
-	{A_2} \& {A_2}
-	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-	\arrow["{\upc{c}}"', from=1-1, to=2-1]
-	\arrow["\id", from=1-2, to=2-2]
-	\arrow["{r(A_2)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-\end{tikzcd}\]
+To accomplish this, we equip every predomain $A$ with a monoid $M_A$ of
+\emph{syntactic perturbations}, as well as a means of \emph{interpreting} these
+perturbations as actual endomorphisms on $A$. Moreover, we want the resulting
+endomorphisms to be extensionally equivalent to the identity function, a
+condition we express by saying that they are weakly bisimilar to the identity
+morphism on $A$. More formally, for every $A$ we require a homomorphism of
+monoids $i_A : M_A \to \morbisimid{A} := \{ f : A \to A \mid f \bisim \id_A \}$.
+We likewise equip every error domain $B$ with a monoid $M_B$ and interpretation
+homomorphism $i_B : M_B \to \morbisimid{B}$. These concepts deserve their own
+definitions: We call the triple $(A, M_A, iA)$ a \textbf{value object} as it
+will be the denotation of value types in our final model. Likewise, the triple
+$(B, M_B, i_B)$ will be called a \textbf{computation object}.
 
-We can use this to derive the original UpL square by composing this square with
-the identity square for $c'$:
+We must now impose an additional condition on predomain relations $c$ and error
+domain relations $d$ that specify how they interact with these perturbations. In
+particular, we need to be able to ``push'' and ``pull'' perturbations along
+relations $c$ and $d$. The intuition for this requirement comes from the
+construction of the square corresponding to the original UpL rule from the
+square for the simplified version as was shown in Section
+\ref{sec:towards-relational-model}. Recall that we horizontally composed the
+simplified UpL square for $c$ with the identity square for $c'$ on the right.
+Now that the square for UpL involves a perturbation on the right instead of the
+identity, this construction will not work unless we can ``push'' the
+perturbation on $A_2$ to one on $A_3$. That is, given a relation $c : A \rel A'$
+and a perturbation $m_A \in M_A$, we need to be able to turn it into a
+perturbation $m_{A'} \in M_{A'}$ such that the resulting endomorphisms obtained
+by applying the respective homomorphisms $i_A$ and $i_{A'}$ to the perturbations
+form a square. Dually, we will need to ``pull'' a perturbation from right to
+left. This notion is made formal by the following definition:
+%
+\begin{definition}
+    Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let $c
+    : A \rel A'$ be a predomain relation. A \textbf{push-pull structure} for $c$
+    consists of monoid homomorphisms $\push : M_A \to M_A'$ and $\pull : M_A'
+    \to M_A$ such that for all $m_A \in M_A$ there is a square $i_A(m_A)
+    \ltsq{c}{c} i_{A'}(\push\, m_A)$ and for all $m_{A'} \in M_{A'}$ there is a
+    square $i_A(\pull\, m_{A'}) \ltsq{c}{c} i_{A'}(m_{A'})$.
+\end{definition}
+%
+We make the analogous definition for computation objects.
 
-% https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMSwxLCJBXzIiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzMsNSwicihBXzIpIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNSw0LCJjJyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMywiXFx1cGN7Y30iLDJdLFsyLDQsIlxcaWQiXSxbMSw1LCJcXGlkIiwyXV0=
-\[\begin{tikzcd}[ampersand replacement=\&]
+With the notion of perturbation defined, we can now discuss the final aspect of
+the model, namely the representability of relations.
+%
+\begin{definition}[quasi-left-representable relations]
+Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let $c : A
+\rel A'$ be a predomain relation. We say that $c$ is
+\emph{quasi-left-representable} by a predomain morphism $e : A \to A'$ if there
+are perturbations $\delle_c \in M_A$ and $\delre_c \in M_{A'}$ such that the
+following two squares exist: (1) UpL: $e \ltsq{c}{r(A')} i_{A'}(\delre_c)$, and
+(2) UpR: $i_A(\delle_c) \ltsq{r(A)}{c} e$.
+\end{definition}
+%
+Observe that this generalizes the previous notion of representability, since
+under that definition the perturbations were always the identity.
+%
+We make the analogous definition of quasi-left-representability for error domain
+relations $d$.
+%
+Likewise, we define quasi-right-representability as follows:
+%
+\begin{definition}
+Let $(B, M_B, i_B)$ and $(B', M_{B'}, i_{B'})$ be computation objects, and let
+$d : B \rel B'$ be an error domain relation. We say that $d$ is
+\emph{quasi-right-representable} by an error domain morphism $p : B' \to B$ if
+there are perturbations $\dellp_d \in M_B$ and $\delrp_d \in M_{B'}$ such that
+the following two squares exist: 
+(1) DnL: $p \ltsq{r(B')}{d} i_{B'}(\delrp_d)$ and
+(2) DnR: $i_B(\dellp_d) \ltsq{d}{r(B)} p$.
+\end{definition}
+%
+We make the analogous definition of quasi-right-representability for predomain
+relations $c$.
+
+Although it makes sense to refer to the quasi-representability of an arbitrary
+predomain or error domain relation, quasi-representable relations do not in
+general compose. In order to construct the needed squares for the composition of
+two relations, we require that the relations have push-pull structures.
+
+We can now give the final definition of value and computation relations for our
+model:
+%
+\begin{definition}[value relations]
+A \textbf{value relation} between value objects $(A, M_A, i_A)$ and $(A',
+M_{A'}, i_{A'})$ consists of a predomain relation $c : A \rel A'$ that has a
+push-pull structure, is quasi-left-representable by a morphism $e_c : A \to A'$,
+and is such that $\li c$ is quasi-right-representable by an error domain morphism
+$p_c : \li A' \to \li A$.
+\end{definition}
+
+\begin{definition}[computation relations]
+A \textbf{computation relation} between computation objects $(B, M_B, i_B)$ and
+$(B', M_{B'}, i_{B'})$ consists of an error domain relation $d$ that has a
+push-pull structure, is quasi-right-representable by a morphism $p_d : B' \to
+B$, and is such that $Ud$ is quasi-left-representable by a morphism $e_d : UB
+\to UB'$
+\end{definition}
+
+The reason for requiring $\li c$ to be quasi-right-representable in the definition
+of value relations is so that we can define the action of the functor $\li$ on
+relations. Similarly, we require $Ud$ to be quasi-left-representable in the
+definition of computation relations so that we can define the action of $U$ on
+relations.
+
+It remains to specify the actions of the functors on value and computation
+objects and relations.
+% Because we have augmented the definitions of object and relation for our model,
+% we need to specify the actions of the functors $F$, $U$, $\times$, and $\arr$ on
+% these objects and relations. 
+% For instance, the functor $F$ now acts on \emph{value objects} $(A, M_A, i_A)$,
+% which means we need to specify the monoid $M_{FA}$ and interpretation $i_{FA}$
+% corresponding to the error domain $FA$.
+
+\subsubsection{Constructions involving Perturbations}\label{sec:perturbation-constructions}
+
+We first discuss the action of $\li$ on objects. Given a value object $(A, M_A,
+i_A)$ we define $M_{\li A}$ to be $\mathbb{N} \oplus M_A$, where $\oplus$
+denotes the free product of monoids (the coproduct in the category of monoids).
+The intuition behind needing to include $\mathbb{N}$ comes from the example of
+the downcast for $\iarr$. Recall that we adjusted the squares corresponding to
+the DnL and DnR rules by adding $(\delta \circ \eta)^\dagger$ on the side
+opposite the downcast. Thus, we need a syntactic perturbation in $M_{\li A}$
+that will be interpreted as $(\delta \circ \eta)^\dagger$. We accomplish this by
+taking the coproduct with $\mathbb{N}$.
+%
+Then to interpret the perturbations as endomorphisms, by the universal property
+of the coproduct of monoids it suffices to define two homomorphisms, one from
+$\mathbb{N} \to \morbisimid{\li A}$ and one from $M_A \to \morbisimid{\li A}$.
+The first homomorphism sends the generator $1 \in \mathbb{N}$ to the error
+domain endomorpism $(\delta \circ \eta)^\dagger$, and the second sends $m_A \in
+M_A$ to $\li(i_A(m_A))$ and we observe in both cases that the resulting
+endomorphisms are bisimilar to the identity since the action of $\li$ on
+morphisms preserves bisimilarity.
+
+Now we discuss the action of $U$ on objects. Given a computation object $(B,
+M_B, i_B)$, we define $M_{UB}$ to be $\mathbb{N} \oplus M_B$. The reason for
+requiring $\mathbb{N}$ is related to the Kleisli arrow functor: to establish
+quasi-representability of $U(c \arr d)$ given quasi-representability of $c$ and
+$d$, we will need to turn a perturbation on $\li A$ into a perturbation on $U(A
+\arr B)$. Since the perturbations for $\li A$ involve $\mathbb{N}$, so must the
+perturbations for $UB$. The interpretation $i_{UB}$ of the perturbations on $UB$
+works in the same manner as that of $\li A$.
+
+The action of $\times$ on objects is as follows. Given $(A_1, M_{A_1}, i_{A_1})$
+and $(A_2, M_{A_2}, i_{A_2})$, we define the monoid $M_{A_1 \times A_2} =
+M_{A_1} \oplus M_{A_2}$. To construct the homomorphism $i_{A_1 \times A_2}$, we
+appeal to the universal property and give homomorphisms $M_{A_1} \to
+\morbisimid{A_1 \times A_2}$ and $M_{A_2} \to \morbisimid{A_1 \times A_2}$. The
+former applies $i_{A_1}$ in the first component and the identity in the second
+component, while the latter does the reverse with $i_{A_2}$.
+
+The action of $\arr$ on objects is as follows. Given $(A, M_A, iA)$ and $(B,
+M_B, i_B)$ we define the monoid $M_{A \arr B} = M_A^{op} \oplus M_B$. Then to
+construct the interpretation as endomorphisms it suffices to define a
+homomorphism $M_A^{op} \to \morbisimid{A \arr B}$ and $M_B \to \morbisimid{A
+\arr B}$. The former takes an element $m_A$ and a predomain morphism $f : A \to
+UB$ and returns the composition $f \circ i_A(m_A)$, while the latter is defined
+similarly by post-composition with $i_B$.
+
+\subsubsection{Constructions involving Value and Computation Relations}
+
+\eric{What, if anything, should go here? Should we discuss composition of
+value/computation relations? Actions of the functors on relations?}
+
+\eric{TODO: Quasi-equivalence of the relations mentioned in the syntax section.}
+
+\subsection{The Dynamic Type}
+We now describe the denotation of the dynamic type. Analogously to how we
+defined the dynamic type in Section \ref{sec:concrete-term-model}, the predomain
+representing the dynamic type is defined using a combination of least fixpoint
+and guarded recursion as the solution to the equation of \emph{predomains}
+%
+\[ D \cong \mathbb{N}\, + (D \times D)\, + \laterhs U(D \arr \li D). \]
+%
+As before, we name the constructors $\inat$, $\itimes$, and $\iarr$
+respectively, and the upcasts $e_{\nat}$, $e_\times$, and $e_\to$ are the same
+as the constructors except for $e_{\iarr}$ which includes a $\nxt$.
+
+%
+% We define $e_\mathbb{N} : \mathbb{N} \to D$ to be the $\text{nat}$ constructor,
+% $e_\times : D \times D \to D$ to be $\text{times}$, and $e_\to : U(D \arr F D)$
+% to be the morphism $\nxt$ followed by $\text{fun}$.
+
+The ordering $\le_D$ is given by:
+%
+\begin{align*}
+    \inat(n) \le \inat(n') 
+        &\iff n = n' \\
+    \itimes (d_1, d_2) \le \itimes (d_1', d_2')
+        &\iff d_1 \le_D d_2 \text{ and } d_1' \le_D d_2'\\
+    \iarr(\tilde{f}) \le \iarr(\tilde{f'}) 
+        &\iff \later_t(\tilde{f}_t \le \tilde{f'}_t),
+\end{align*}
+%
+and the bisimilarity relation is defined analogously.
+%
+We define three relations involving $D$ as follows. We define $\inat :
+\mathbb{N} \rel D$ by $(n, d) \in \inat$ iff $e_\nat(n) \le_D d$. We similarly
+define $\itimes : D \times D \rel D$ by $((d_1, d_2), d) \in \itimes$ iff
+$e_\times(d_1, d_2) \le_D d$, and we define $\text{inj}_\to : U(D \arr \li D)
+\rel D$ by $(f, d) \in \iarr$ iff $e_\to(f) \le_D d$.
+
+Now we define the perturbations for $D$. Recall that to each predomain $A$ we
+associate a monoid $M_A$ of perturbations and a homomorphism into the monoid of
+endomorphisms bisimilar to the identity, and likewise for error domains. It may
+seem as though we need to define the perturbations for $D$ by guarded recursion,
+but in fact we define them via least-fixpoint in the category of monoids:
+%
+\( M_D \cong (M_{D \times D}) \oplus M_{U(D \to \li D)}. \)
+%
+% Unfolding the definitions, this is the same as
+% %
+% \( M_D \cong (M_D \times M_D) \oplus (\mathbb{N} \times M_D^{op} \times \mathbb{N} \times M_D). \)
+% %
+We now explain how to interpret these perturbations as endomorphisms.
+
+\eric{Check this}
+
+We define $i_D : M_D \to \{ f : D \to D \mid f \bisim \id \}$ by induction as follows:
+%
+\begin{align*}
+  i_D(m_\times) &= \lambda d. \text{case $d$ of }
+    \{ \itimes (d_1, d_2) \To \itimes (i_{D \times D}(m_\times)(d_1, d_2)) \alt x \To x \} \\
+  i_D(m_\to) &= \lambda d. \text{case $d$ of }
+    \{ \iarr (\tilde{f}) \To \iarr (\lambda t. i_{U(D \to \li D)}(m_\to)(\tilde{f}_t)) \alt x \To x \} \\
+\end{align*}
+%
+In the case of a perturbation on $D \times D$, we use the interpretation for the
+perturbations of $D \times D$ as defined earlier in this section, which in turn
+will use $i_D$ inductively. Similarly, for a perturbation on $U(D \to \li D)$ we
+use the interpretation for perturbations on $U(D \to \li D)$.
+
+% \begin{align*}
+%  i_D(m_{\text{times}}, m_{\text{fun}}) &= \lambda d.\text{case $d$ of}  \\
+%  &\alt \tnat(m) \mapsto \tnat(m) \\
+%     &\alt \ttimes(d_1, d_2) \mapsto {\ttimes(i_{D \times D}(p_\text{times})(d_1, d_2))} \\
+%     &\alt \tfun(\tilde{f}) \mapsto {\tfun(\lambda t. i_{U(D \to \li D)}(\tilde{f}_t))}
+% \end{align*}
+
+One can verify that this defines a homomorphism from $M_D \to \{ f : D \to D : f
+\bisim \id \}$. We claim that the three relations $\inat$, $\itimes$, and
+$\iarr$ satisfy the push-pull property. As an illustrative case, we establish
+the push-pull property for the relation $\iarr$. We define $\pull_{\iarr} : M_D
+\to M_{U(D \arr \li D)}$ by giving a homomorphism from $M_{D \times D} \to
+M_{U(D \arr \li D)}$ and from $M_{U(D \to \li D)} \to M_{U(D \arr \li D)}$. The
+former is the trivial homomorphism sending everything to the identity element,
+while the latter is the identity homomorphism.
+%
+Conversely, we define $\push_{\iarr} : M_{U(D \arr \li D)} \to M_D$ as the
+inclusion into the coproduct.
+%
+Showing that the relevant squares exist involving push and pull is
+straightforward.
+
+We next claim that the relations $\inat$, $\itimes$, and $\iarr$ are
+quasi-left-representable, and that their lifts are quasi-right-representable.
+Indeed, it is straightforward to see that they are quasi-left-representable
+where all of the perturbations are taken to be the identity elements of the
+respective monoids.
+%
+For quasi-right-representability, the only nontrivial case is $\li(\iarr)$.
+Recall the defintion of the downcast for $\iarr$ given in Section
+\ref{sec:term-interpretation}.
+
+We know that in order for the DnR and DnL squares for $\li \iarr$ to exist, we must
+insert a delay on the side opposite the downcast. In terms of perturbations,
+this means that we take the perturbations $\dellp$ and $\delrp$ in the
+definition of quasi-right-representability to both be $\inl\, 1$, where $\inl$
+is the injection into the coproduct of monoids. We recall that the monoid of
+perturbations for $\li$ involves a coproduct with $\mathbb{N}$, and that the
+interpretation homomorphism maps the generator $1$ to the endomorphism of error
+domains $(\delta \circ \eta)^\dagger$.
+
+With this choice for the perturbations, it is straightforward to show that the
+DnL and DnR squares exist using the definition of the downcast and the
+definition of the relation $\iarr$.
+
+% We define $\dellp_D = \delrp_D = \delta_{FD}$. Then it is easy to show using the
+% definition of $\iarr$ that the squares for $\dnl$ and $\dnr$ commute.
+% Then for $\dnr$, we need to show that if $(f, d) \in \iarr$ then
+%
+It is also straightforward to establish the retraction property for each of
+these three relations. In the case of $\iarr$, we have that the property holds
+up to a delay.
+
+% \subsection{Obtaining an Extensional Model}
+
+\subsection{Defining an Extensional Model}
+
+To obtain a model for gradual typing that is oblivious to the stepping behavior
+of terms, we apply the ``bisimilarity-closure'' construction introduced in the
+previous section as follows. Let $c_i : A_i \rel A_i'$ and $c_o : A_o \rel
+A_o'$.
+%
+% We define an extensional square between morphisms $f : A_i \to A_o$ and $g :
+% A_i' \to A_o'$ to consist of:
+
+% \begin{itemize}
+%   \item A morphism $f' : A_i \to A_o$ with $f \bisim f'$.
+%   \item A morphism $g' : A_i' \to A_o'$ with $g \bisim g'$.
+%   \item A square $f' \ltsq{c_i}{c_o} g'$.
+% \end{itemize}
+%
+We say that $f \ltsqbisim{c^i}{c_o} g$ if there exist $f'$ and $g'$ such that
+%
+\[ f \bisim f' \ltsq{c_i}{c_o} g' \bisim g. \]
+%
+If the relations $c_i$ and $c_o$ are obvious from the context, we will simply
+write $f \ltbisim g$. We make the analogous construction for the computation
+squares.
+%
+We now verify the existence of the squares for the cast rules. For UpL, the
+square is obtained as follows:
+
+% https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMSwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMCwzLCJlX2MiLDIseyJjdXJ2ZSI6MX1dLFsxLDQsIlxcZGVscmVfYyIsMl0sWzIsNSwiXFxwdXNoX3tjJ30oXFxkZWxyZV9jKSIsMix7ImN1cnZlIjoxfV0sWzAsMSwiYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEsMiwiYyciLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFszLDQsIlxcbGVfe0FfMn0iLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs0LDUsImMnIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMiw1LCJcXGlkIiwwLHsiY3VydmUiOi0xfV0sWzAsMywiZV9jIiwwLHsiY3VydmUiOi0xfV0sWzgsMTMsIlxcYmlzaW0iLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNiwxNCwiXFxiaXNpbSIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
+\[\begin{tikzcd}[ampersand replacement=\&,column sep=large]
 	{A_1} \& {A_2} \& {A_3} \\
@@
 	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-	\arrow["{\upc{c}}"', from=1-1, to=2-1]
+	\arrow[""{name=0, anchor=center, inner sep=0}, "{e_c}"', curve={height=6pt}, from=1-1, to=2-1]
+	\arrow[""{name=1, anchor=center, inner sep=0}, "{e_c}", curve={height=-6pt}, from=1-1, to=2-1]
 	\arrow["{c'}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
-	\arrow["\id"', from=1-2, to=2-2]
-	\arrow["\id", from=1-3, to=2-3]
-	\arrow["{r(A_2)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+	\arrow["{\delre_c}"', from=1-2, to=2-2]
+	\arrow[""{name=2, anchor=center, inner sep=0}, "{\push_{c'}(\delre_c)}"', curve={height=6pt}, from=1-3, to=2-3]
+	\arrow[""{name=3, anchor=center, inner sep=0}, "\id", curve={height=-6pt}, from=1-3, to=2-3]
+	\arrow["{\le_{A_2}}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
 	\arrow["{c'}"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
+	\arrow["\bisim"{description}, draw=none, from=0, to=1]
+	\arrow["\bisim"{description}, draw=none, from=2, to=3]
 \end{tikzcd}\]
 
-The existence of these simpler squares for a relation can be stated without
-reference to other relations. Thus, we can require that these squares exist in
-the definition of relation without running into the circularity issue we faced
-with the original rules.
+Note that the horizontal composition of the squares is defined, as it takes
+place in the intensional setting. We have used the push-pull structure on $c'$
+to turn the perturbation on $A_2$ into a perturbation on $A_3$. Lastly we note
+that the bottom edge of the square is equal to $c'$ by the fact that $c'$ is
+downward closed. The squares for UpR, DnL, and DnR are obtained similarly.
 
-Finally, we use squares to model the term precision ordering $\Gamma \ltdyn
-\Gamma' \vdash M \ltdyn N : A \ltdyn A'$ as an ordering relation between the
-monotone functions denoted by $M$ and $N$:
+% Now that we have defined an intensional model with an interpretation for the
+% dynamic type, we can apply the abstract constructions introduced in Section
+% \ref{sec:extensional-model-construction}. Doing so, we obtain an extensional
+% model of gradual typing, where the squares are given by the ``bisimilarity
+% closure'' of the intensional error ordering.
 
-% https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXEdhbW1hIl0sWzAsMSwiVUYgQSJdLFsxLDAsIlxcR2FtbWEnIl0sWzEsMSwiVUZBJyJdLFsyLDMsIlxcc2Vte059Il0sWzAsMSwiXFxzZW17TX0iLDJdLFswLDIsIlxcR2FtbWFee1xcbHRkeW59IiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJVRmMiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
-\[\begin{tikzcd}[ampersand replacement=\&]
-	\Gamma \& {\Gamma'} \\
-	{UF A} \& {UFA'}
-	\arrow["{\Gamma^{\ltdyn}}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-	\arrow["{\sem{M}}"', from=1-1, to=2-1]
-	\arrow["{\sem{N}}", from=1-2, to=2-2]
-	\arrow["UFc"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-\end{tikzcd}\]
+\eric{Order-equivalence of relations}
 
-where $\Gamma = \{A_1, A_2, \dots\}$ and $\Gamma' = \{A_1' , A_2', \dots\}$ and for
-each $j$, $c_j : A_j \ltdyn A_j'$, and the relation $\Gamma^{\ltdyn}$ is simply
-the product of the $c_i$.
+\subsection{Relational Adequacy}\label{sec:adequacy}
 
-It remains to define $Fc$, i.e., the action of $F$ on relations.
+\eric{TODO: update this in light of the new section on adequacy of the term model.}
+
+In this section, we prove an adequacy result for the extensional model of
+gradual typing we obtained in the previous section. First we establish some
+notation. Fix a morphism $f : 1 \to \li \mathbb{N} \cong \li \mathbb{N}$. We
+write $f \da n$ to mean that there exists $m$ such that $f = \delta^m(\eta n)$
+and $f \da \mho$ to mean that there exists $m$ such that $f = \delta^m(\mho)$.
+
+Recall that $\ltbisim$ denotes the relation on value morphisms defined as the
+bisimilarity-closure of the intensional error-ordering on morphisms. That is, we
+have $f \ltbisim g$ iff there exist $f'$ and $g'$ with
 %
-% However, lifting a relation between $A$ and $A'$ to a relation between $\li A$
-% and $\li A'$ proves problematic if we allow computations that take different
-% numbers of steps to be related. To illustrate the issue, let us define an
-% ordering $\semltbad$ between $\li A$ and $\li A'$; we call this the
-% \emph{step-insensitive error ordering}. 
+\[ f \bisim_{\li \mathbb{N}} f' \le_{\li \mathbb{N}} g' \bisim_{\li \mathbb{N}} g. \]
 %
-We note that the graduality property and the axioms of the inequational theory
-are independent of the intensional stepping behavior of terms, so our ultimate
-notion that interprets term precision will need to be oblivious to stepping as
-well.
+Here $\le_{\li \mathbb{N}}$ is the lock-step error ordering, and $\bisim_{\li
+\mathbb{N}}$ is weak bisimilarity. First observe that in this ordering, the
+semantics of error is not equivalent to the semantics of the diverging term. The
+main result we would like to show is as follows: If $f \ltbisim g : \li \mathbb{N}$, then:
+\begin{itemize}
+  \item If $f \da n$ then $g \da n$.
+  \item If $g \da \mho$ then $f \da \mho$.
+  \item If $g \da n$ then $f \da n$ or $f \da \mho$.
+\end{itemize}
+
 %
-To that end, let us define a \emph{step-insensitive error ordering} $\semltbad$
-between $\li A$ and $\li A'$; The ordering is parameterized by an ordering
-relation $\le$ between $A$ and $A'$. The definition is by guarded recursion and
-is shown in Figure \ref{fig:step-insensitive-error-ordering}. Recall that
-$\delta : \li A \to \li A$ is defined by $\delta = \theta_A \circ \nxt$.
+Unfortunately, this is actually not provable!
+Roughly speaking, the issue is that this is a ``global'' result, and it is not possible
+to prove such results inside of the guarded setting. 
+In particular, if we tried to prove the above result in the guarded
+setting, we would run into a problem where we would have a natural number
+``stuck'' under a $\later$, with no way to get out the underlying number.
+%
+Thus, to prove our adequacy result, we need to leave the guarded setting and pass back
+to the more familiar, set-theoretic world with no internal notion of step-indexing.
+We can do this using a process known as \emph{clock quantification}.
+Recall that all of the constructions we have made in SGDT take place in the context of a clock $k$.
+All of our uses of the later modality and guarded recursion have taken place with respect to this clock.
+For example, recall the definition of the lift monad by guarded recursion.
+% We define the lift monad $\li^k X$ as the guarded fixpoint of $\lambda \tilde{T}. X + 1 + \later^k_t (\tilde{T}_t)$.
+We can view this definition as being parameterized by a clock $k$: $\li^k : \type \to \type$.
+Then for $X$ satisfying a certain technical requirement known as \emph{clock-irrelevance},
+\footnote{A type $X$ is clock-irrelevant if there is an isomorphism $\forall k.X \cong X$.}
+we can define the ``global lift'' monad as $\li^{gl} X := \forall k. \li^k X$.
 
-\begin{figure*}
-    \begin{align*}
-        &\eta\, x \semltbad \eta\, y \text{ if } 
-            x \semlt y \\
+It can be shown that there is an isomorphism between the global lift monad and the
+delay monad of Capretta \cite{lmcs:2265}.
+Recall that, given a type $X$, the delay monad $\text{Delay}(X)$ is defined as the coinductive
+type generated by 
+$\tnow : X \to \delay(X)$ and $\tlater : \delay(X) \to \delay(X)$.
+
+% solution to the equation
+
+% \[ \text{Delay}(X) \cong X + \text{Delay}(X). \]
+
+It can be shown that for a clock-irrelevant type $X$, $\li^{gl} X$ is a final
+coalgebra of the functor $F(Y) = X + 1 + Y$ (For example, this follows from Theorem 4.3 in
+\cite{kristensen-mogelberg-vezzosi2022}.) 
+\footnote{The proof relies on the existence of an operation  
+$\mathsf{force} : \forall k. \later^k A \to \forall k. X$ that
+allows us to eliminate the later operator under a clock quantifier.
+This must be added as an axiom in guarded type theory.}
+Since $\delay(X + 1)$ is also a final coalgebra
+of this functor, then we have $\li^{gl} X \cong \delay(X + 1)$.
+
+Given a predomain $X$ on a clock-irrelevant type, we can define a
+``global'' version of the lock-step error ordering and the
+weak bisimilarity relation on elements of the global lift; the former is defined by
 %
-        &\mho \semltbad l \\
+\( x \le^{gl}_X y := \forall k. x[k] \le y[k], \)
 %
-        &\theta\, \tilde{l} \semltbad \theta\, \tilde{l'} \text{ if } 
-            \later_t (\tilde{l}_t \semltbad \tilde{l'}_t) \\
+and the latter is defined by
 %
-        &\theta\, \tilde{l} \semltbad \mho \text{ if } 
-            \theta\, \tilde{l} = \delta^n(\mho) \text { for some $n$ } \\
+\( x \bisim^{gl}_X y := \forall k. x[k] \bisim y[k]. \)
 %
-        &\theta\, \tilde{l} \semltbad \eta\, y \text{ if }
-            (\theta\, \tilde{l} = \delta^n(\eta\, x))
-        \text { for some $n$ and $x : A$ such that $x \le y$ } \\
-        %
-        &\theta\, \tilde{l} \semltbad \mho \text { if } 
-            \theta\, \tilde{l} = \delta^n(\mho) \text { for some $n$ } \\
-        %	
-        &\eta\, x \semltbad \theta\, \tilde{l'} \text { if }
-            (\theta\, \tilde{l'} = \delta^n (\eta\, y))
-        \text { for some $n$ and $y : A'$ such that $x \le y$ }
-    \end{align*}
-    \caption{The step-insensitive error ordering.}
-    \label{fig:step-insensitive-error-ordering}
-\end{figure*}
-
-Two computations that immediately return $(\eta)$ are related if the returned
-values are related in the underlying ordering. The computation that errors
-$(\mho)$ is the least term in the ordering. If both sides step (i.e., both sides
-are $\theta$), then we allow one time step to pass and compare the resulting
-terms (this is where use the relation defined ``later'' and is why we employ
-guarded recursion to define the relation).
-%
-Lastly, if one side steps and the other immediately returns a value, then in
-order for these terms to be related, the side that steps must terminate with a
-value in some finite number of steps $n$, and that value must be related to the
-value returned by the other side. Likewise, if the LHS steps and the RHS
-immediately errors, then in order to be related, the LHS must eventually
-terminate with error.
-
-\subsection{Problem: Failure of Transitivity}
-
-The key issue with the above ordering is it is \emph{not} transitive, as we will
-show below. We begin with a general ``no-go'' theorem about relations defined on
-the guarded lift monad:
-
-\begin{theorem}[No-go theorem for relations on $\li X$]\label{thm:no-go}
-    Let $R$ be a relation on the guarded lift monad satisfying the following
-    three properties:
-    \begin{enumerate}
-        \item Transitivity
-
-        \item Congruence with respect to $\theta$: If $\later_t (\tilde{x}_t
-        \binrel{R} \tilde{y}_t)$, then $\theta(\tilde{x}) \binrel{R}
-        \theta(\tilde{y})$
-
-        \item Closure under delay on both sides: If $x \binrel{R} y$, then $x
-        \binrel{R} \delta y$ and $\delta x  \binrel{R} y$
-    \end{enumerate}
-
-    Then $R$ is trivial, i.e., $x \binrel{R} y$ for all $x : \li X$ and $y : \li
-    Y$.
-
-\end{theorem}
-
-This theorem follows from two ``one-sided'' lemmas where we assume that the
-relation $R$ is closed under delay on the left (resp. right) and show that
-$(\fix\, \theta) \binrel{R} y$ for all $y$ (resp. $x \binrel{R} (\fix\, \theta)$
-for all $x$). These lemmas in turn are proved by \lob-induction.
-
-The theorem implies that the relation $\semltbad$ defined on $\li X \times \li
-Y$ is \emph{not} transitive whenever the underlying relation $\semlt$ on $X$ and
-$Y$ is not trivial, i.e., when there are $x_0$ and $y_0$ that are not related in
-the relation $\semlt$ on $X$ and $Y$. To show this, first observe that
-$\semltbad$ satisfies properties (2) and (3) in the above theorem. Congruence
-with respect to $\theta$ is immediate from the definition, while closure under
-delay can be proved using \lob-induction.
-% TODO the proof of this uses tick irrelevance
-Then since $x_0$ is not related to $y_0$, we have that $\eta\, x_0$ is not
-related to $\eta\, y_0$ in $\semltbad$, i.e., $\semltbad$ is non-trivial. But if
-it were transitive, then by the theorem, it would be trivial, a contradiction.
-Thus, we are able to refute transitivity of $\semltbad$ provided the underlying
-relation is non-trivial.
-
-\eric{Read this}
-The lack of transitivity presents a major barrier towards giving a compositional
-semantics to gradual typing: if our relations are not transitive, then we cannot
-compose squares horizontally. But we have argued above that horizontal
-composition is essential for modelling the axioms of gradual typing in a
-syntax-independent manner.
-
-% In particular, to model the UpL/UpR/DnL/DnR rules
-% for casts in the presence of transitivity, it is sufficient to establish the
-% validity of simpler rules that do not ``build in'' composition. We can then use
-% transitivity to derive the original versions of the rules. On the other hand,
-% without transitivity we must instead validate the cast rules in the model ``from
-% scratch''. In fact, we cannot even define the semantic interpretation of type
-% precision in a syntax-independent manner. The issue is that in the definition of
-% relation, the requirement that it be representable now involves quantifying over
-% all other relations, which is circular.
-
-% So, although this approach would suffice for proving graduality, it lacks the
-% compositionality that we seek in a reusable framework for the semantics of
-% gradual typing.
-
-%Because it is convenient to make use of transitive reasoning in proving graduality, ...
-We therefore seek to formulate an ordering relation on $\li A \times \li A'$
-that \emph{is} transitive. The theorem above tells us, however, that any such
-relation must either not be a congruence with respect to $\theta$, or must not
-be closed under delay.
-\footnote{Technically, we only need to give up closure under delay on \emph{one}
-side, as the above theorem requires $R$ to be closed on both sides. We could
-define two separate ``step-semi-sensitive'' relations on $\li A \times \li A'$,
-each closed under delays on the left and right respectively. The denotation of
-term precision would then involve the conjunction of those relations. We do not
-pursue this approach in this paper, because (1) it is not clear if the
-step-semi-sensitive relations would be transitive, and (2) we would need to
-reason about the two ordering relations separately for every pair of terms.}
-%
-We cannot forego the property of being a $\theta$-congruence, as without this we
-would not be able to prove properties of the relation using \lob-induction,
-e.g., that $\ext{f}{}$ is a \emph{monotone} function with respect to the
-ordering.
-%and moreover this would mean that $\theta$ would not be a monotone function.
-
-Thus, we choose to sacrifice closure under delays. That is, we will define an
-ordering relation that requires terms to be in ``lock-step''. In order for two
-computations to be related in this ordering, they must have the same stepping
-behavior (or the left-hand side must be an error). To deal with terms that take
-differing numbers of steps, we then define a separate relation called \emph{weak
-bisimilarity} that relates terms that are extensionally equal and may differ by
-a finite number of delays. Then, the semantics of the error ordering for the
-guarded lift monad will involve a combination of these two relations: the
-``closure'' of the lock-step error ordering under weak bisimilarity on both
-sides.
-% Although the combined relation will not be transitive (for the same reason that
-% $\semltbad$ is not transitive), this...
-This decomposition has the advantage that we can recover some transitive
-reasoning and push much of the reasoning about stepping to the margins of the
-development.
-
-% Doing so, we define a \emph{lock-step} error ordering, where roughly speaking,
-% in order for computations to be related, they must have the same stepping
-% behavior. We then formulate a separate relation, \emph{weak bisimilarity}, that
-% relates computations that are extensionally equal and may differ only in their
-% stepping behavior. % up to a finite number of \delta's
-% Finally, the semantics of term precision will involve a combination of these two
-% relations, a sort of closure of the lock-step ordering under weak bisimilarity
-% on both sides. This decomposition has the advantage that we can recover some
-% transitive reasoning and push the parts involving stepping to the margins of the
-% development.
-
-\subsection{Solution: Splitting into Two Relations}\label{sec:lock-step-and-weak-bisim}
-
-As mentioned, to recover transitive reasoning we will decompose the error
-ordering on the guarded lift monad into two separate relations. The first is the
-\emph{lock-step error-ordering} (also called the ``step-sensitive error
-ordering''), the idea being that two computations $l$ and $l'$ are related if
-they are in lock-step with regard to their intensional behavior, up to $l$
-erroring. This will be the partial ordering with which we equip the guarded lift monad.
-Figure \ref{fig:lock-step-error-ordering} gives the definition.
-
-\begin{figure}
-    \fbox{$l_1 \ltls l_2$}
-    \begin{align*}
-        &\eta\, x \ltls \eta\, y \text{ if } 
-            x \mathbin{R} y \\
-        %		
-        &\mho \ltls l' \\
-        %
-        &\theta\, \tilde{l} \ltls \theta\, \tilde{l'} \text{ if } 
-            \later_t (\tilde{l}_t \ltls \tilde{l'}_t)
-    \end{align*}
-    \caption{The lock-step error ordering.}
-    \label{fig:lock-step-error-ordering}
-\end{figure}
-
-When both sides are $\eta$, then we check that the returned values are related
-in $R$. The error term $\mho$ is below everything. Lastly, if both sides step
-(i.e., are a $\theta$) then we compare the resulting computations one time step
-later.
-%
-It is straightforward to prove using \lob-induction that this relation is
-reflexive, transitive and anti-symmetric given that the underyling relation $R$
-has those properties. The lock-step ordering is therefore the partial ordering
-we will associate with $\li X$.
-%
-We similarly define a heterogeneous version of this ordering between the lifts
-of two different types $X$ and $Y$, parameterized by a relation $R$ between $X$
-and $Y$.
-
-The second relation we define on the guarded lift monad is \emph{weak bisimilarity}.
-% For a type $X$, we define a relation on $\li X$, called ``weak bisimilarity",
-% written $l \bisim l'$. 
-Intuitively, we say $l \bisim l'$ if they are the same ``up to delays''.
-%
-Weak bisimilarity on $\li X$ is parameterized by an underlying relation $R$ on $X$
-and is defined by guarded fixpoint as shown in Figure \ref{fig:weak-bisimilarity}.
-
-\begin{figure}
-    \fbox{$l_1 \bisim l_2$}
-    \begin{align*}
-        &\mho \bisim \mho \\
-      %
-        &\eta\, x \bisim \eta\, y \text{ if } 
-          x \mathbin{R} y \\
-      %		
-        &\theta\, \tilde{x} \bisim \theta\, \tilde{y} \text{ if } 
-          \later_t (\tilde{x}_t \bisim \tilde{y}_t) \\
-      %	
-        &\theta\, \tilde{x} \bisim \mho \text{ if } 
-          \theta\, \tilde{x} = \delta^n(\mho) \text { for some $n$ } \\
-      %	
-        &\theta\, \tilde{x} \bisim \eta\, y \text{ if }
-          (\theta\, \tilde{x} = \delta^n(\eta\, x))
-        \text { for some $n$ and $x : X$ such that $x \mathbin{R} y$ } \\
-      %
-        &\mho \bisim \theta\, \tilde{y} \text { if } 
-          \theta\, \tilde{y} = \delta^n(\mho) \text { for some $n$ } \\
-      %	
-        &\eta\, x \bisim \theta\, \tilde{y} \text { if }
-          (\theta\, \tilde{y} = \delta^n (\eta\, y))
-        \text { for some $n$ and $y : X$ such that $x \mathbin{R} y$ }
-      \end{align*}
-    \caption{The weak bisimilarity relation.}
-    \label{fig:weak-bisimilarity}
-\end{figure}
-
-When both sides are $\eta$, then we ensure that the underlying values are
-bisimilar in the underlying bisimilarity relation on $A$. When one side is a
-$\theta$ and the other is $\eta x$ (i.e., one side steps), we stipulate that the
-$\theta$-term runs to $\eta y$ where $x$ is related to $y$. When one side is
-$\theta$ and the other $\mho$, we require that the $\theta$-term runs to $\mho$.
-If both sides step, then we allow one time step to pass and compare the
-resulting terms. This definition captures the intuition of two terms being
-related up to delays.
-
-It can be shown (by \lob-induction) that weak bisimilarity is reflexive and
-symmetric. We can also show that this relation is \emph{not} transitive. The
-argument is the same as that used to show that the step-insensitive error
-ordering $\semltbad$ described above is not transitive. In particular, we
-observe that weak bisimilarity is a congruence with respect to $\theta$, and we
-can show by $\lob$-induction that it is closed under delay on both sides. Thus,
-we may appeal to the no-go theorem as before.
-% We appeal to the no-go theorem given above
-
-% This issue of non-transitivity will be resolved when we consider the relation's \emph{globalization}.
-
-Now we can define the denotation of term precision for closed terms of type
-$\nat$ as follows:
-%
-\[ l_1 \leq^{\bisim} l_2 := 
-  \Sigma_{l_1', l_2' \in \li \mathbb{N}}. l_1 \bisim l_1' \ltls l_2' \bisim l_2. \]
-
-Importantly, we note that as with the original ordering $\semltbad$ defined at
-the beginning of this section, the relation just defined is not transitive
-(again as a result of the above no-go theorem). It may therefore seem that we
-have not solved the original issue we faced. In a sense, this is true. The model
-of gradual typing that we end up with will not support \emph{extensional}
-compositional reasoning. However, by decomposing the denotation of term
-precision in the above manner, we can employ \emph{intensional} compositional
-reasoning by working with the lock-step error ordering. Thus, the tentative plan
-going forward will be to carry out the proofs compositionally using the
-lock-step ordering. We will then apply the closure under weak bisimilarity in
-the denotation of term precision given above to handle the aspects involving
-stepping.
-
-% ...to carry out the proofs using the lock-step ordering and handle the stepping
-% behavior separately via weak bisimilarity. In the end we combine everything
-% together using the above denotation for term precision
-
-% Indeed, we will need to weaken our abstract notion of categorical model to no
-% longer require transitivity of term precision. On the other hand, we now have
-% the ability to apply transitive reasoning when working with the lock-step
-% ordering. This allows us to carry out complex proofs compositionally, a key goal
-% of our framework.
-
-\subsection{Adjusting the Cast Rules}\label{sec:adjusting-cast-rules}
-
-We have seen how we can recover transitive reasoning by working with the
-lock-step error ordering. However, it turns out that we cannot completely
-isolate the reasoning about steps from the reasoning about the ordering.
-% Unfortunately, by defining the partial order on the guarded lift monad to be
-% the lock-step error ordering, we introduce a new issue.
-In particular, the cast rules required for proving graduality no longer hold if
-interpreted using the lock-step ordering. The source of the problem is that the
-downcast from the dynamic type to a function takes a step, i.e., involves a
-$\theta$. Consider the DnL rule shown below:
+On the other hand, we can define coinductively a ``lock-step error ordering"
+relation on $\delay(X + 1)$:
 %
 \begin{mathpar}
-  \inferrule*{c : A \ltdyn A' \and M \ltdyn N : A'}
-             {\dnc{c}{M} \ltdyn N : c}
+  \inferrule*[]
+  { }
+  {\tnow (\inr\, 1) \ledelay d}
+
+  \inferrule*[]
+  {x_1 \le_X x_2}
+  {\tnow (\inl\, x_1) \ledelay \tnow (\inl\, x_2)}
+
+  \inferrule*[]
+  {d_1 \ledelay d_2}
+  {\tlater\, d_1 \ledelay \tlater\, d_2}
 \end{mathpar}
 %
-Suppose that $c$ is $\iarr \colon \dyntodyn \ltdyn\, \dyn$, and recall the
-definition of the semantics of a downcast given in Section
-\ref{sec:term-interpretation}. The DnL rule corresponds to the following square:
+And we similarly define by coinduction a ``weak bisimilarity'' relation on $\delay(X + 1)$, which uses
+a relation $d \Da x_?$ between $\delay(X+1)$ and $X+1$ that is defined as 
+$d \Da x_? := \Sigma_{m \in \mathbb{N}} d = \tlater^m(\tnow\, x_?)$.
+Then weak bisimilarity is defined by the rules
+%
+\begin{mathpar}
+  \inferrule*[]
+  {x_? \bisim_{X + 1} y_?}
+  {\tnow\, x_? \bisimdelay \tnow\, y_? }
 
-% https://q.uiver.app/#q=WzAsNCxbMCwwLCJGRCJdLFswLDEsIkZVKEQgXFxhcnIgRkQpIl0sWzEsMCwiRkQiXSxbMSwxLCJGRCJdLFswLDIsInIoRkQpIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJGXFxpYXJyIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJcXGRuY3tcXGlhcnJ9IiwyXSxbMiwzLCJcXGlkIl1d
-\[\begin{tikzcd}[ampersand replacement=\&]
-	FD \& FD \\
-	{FU(D \arr FD)} \& FD
-	\arrow["{r(FD)}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-	\arrow["{\dnc{\iarr}}"', from=1-1, to=2-1]
-	\arrow["\id", from=1-2, to=2-2]
-	\arrow["F\iarr"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-\end{tikzcd}\]
+  \inferrule*[]
+  {d_1 \Da x_? \and x_? \bisim_{X + 1} y_?}
+  {\tlater\, d_1 \bisimdelay \tnow\, y_? }
 
-The downcast on the left will insert a $\theta$ in the case where the value of
-type $D$ is a later-function $\tilde{f}$.
+  \inferrule*[]
+  {d_2 \Da y_? \and x_? \bisim_{X + 1} y_?}
+  {\tnow\, x_? \bisimdelay \tlater\, d_2}
+
+  \inferrule*[]
+  {d_1 \bisimdelay d_2}
+  {\tlater\, d_1 \bisimdelay \tlater\, d_2 }
+
+  % \inferrule*[]
+  % {d_1 \Da x_? \and d_2 \Da y_? \and x_? \bisim_{X + 1} y_?}
+  % {d_1 \bisimdelay d_2}
+
+  % \inferrule*[]
+  % {d_1 \bisimdelay d_2}
+  % {\tlater d_1 \bisimdelay \tlater d_2 }
+
+\end{mathpar}
 %
-Thus, in order for the downcast to be related to the RHS in the lock-step error
-ordering on $\li (D \to \li D)$, the RHS must be of the form $\theta(\dots)$,
-and moreover, after one time step, the argument to $\theta$ on the LHS must be
-related to the argument of $\theta$ on the RHS. As it stands now, this need not
-be the case. Thus, we conclude that the DnL rule does not in general hold under
-the lock-step error ordering. However, observe that we can remedy this
-particular situation by inserting a ``delay'' on the right, i.e., wrapping the
-RHS in $\delta = \theta \circ \nxt$. Notice that this delay function is weakly
-bisimlar to the identity, and thus it will disappear when we apply the
-bisimilarity-closure construction described above. That is, this delay makes no
-difference in the extensional setting, but its presence is crucial in the
-intensional setting.
+Note the similarity of these definitions to the corresponding guarded definitions.
+By adapting the aforementioned theorem to the setting of inductively-defined relations,
+we can show that both the global lock-step error ordering and the global weak bisimilarity
+admit coinductive definitions. In particular, modulo the above isomorphism
+between $\li^{gl} X$ and $\delay(X+1)$, the global version of the lock-step
+error ordering is equivalent to the lock-step error ordering on $\delay(X + 1)$,
+and likewise, the global version of the weak bisimilarity relation is equivalent to the
+weak bisimilarity relation on $\delay(X + 1)$.
 
-% TODO Maybe show the correct form of the rule in the special case $c = inj-arr$.
+This implies that the global version of the extensional term precision semantics for
+$\li^{gl} X$ agrees with the corresponding notion for $\delay(X + 1)$.
+Then adequacy follows by proving the corresponding
+result for $\delay(X + 1)$ which in turn follows from the definitions of the relations.
 
-\subsection{Extending the Model to Higher-Order Types}
+% We have been writing the type as $\li X$, but it is perhaps more accurate to write it as $\li^k X$ to
+% emphasize that the construction is parameterized by a clock $k$.
 
-% At this point, we could carry out the full construction of a concrete relational
-% model with these additional features.
-
-The above techniques give a semantic interpretation of term precision for closed
-terms of base type. We now consider how to extend these constructions to
-potentially-open terms of \emph{all} types. In particular, in order to model
-higher-order data types (i.e., functions) we need to equip \emph{every} semantic
-object with not only a partial ordering relation but also a ``bisimilarity''
-relation that is reflexive and symmetric. We must similarly equip every object
-with a structured set of delays that can be inserted and manipulated to ensure
-the appropriate cast rules hold at all higher-order types. For example, the
-upcast for a derivation $c_i \ra c_o$ involves the downcast corresponding to
-$c_i$ and the upcast corresponding to $c_o$. We must be able to insert delays in
-a functorial manner to mimic the structure of the casts. In the next section, we
-will make the necessary definitions and describe the relevant constructions.
-
-% For the sake of reusability and modularity, rather than carry out this
-% construction in the concrete setting developed here, we will instead return to
-% the abstract setting and adjust our definition of model to account for these
-% requirements. We will break the construction into smaller steps and isolate the
-% pieces that require the techniques of SGDT from those that do not. Then with
-% this framework at our disposal, we will return to the construction of a concrete
-% model in Section \ref{sec:concrete-model}.
-% , taking as a starting point the definitions introduced in the current section.
-
-% Thus in the next section we define revised notions of a model of
-% gradual typing based on the lessons learned in the present section.
+% Need : nat is clock irrelevant, as well as the inputs and outputs of effects
+% Axioms about forcing clock
+% Adapt prior argument to get that the defining of the global bisim
+% and global lock-step error ordering are coinductive

### 2024-07-10 paper-new/towards-relational-model.tex
revise sections on relational model

@@
 %
-Sticking to a poset-based semantics, if our value types are posets, we
-then want our precision derivations $c : A \ltdyn A'$ to be
-\emph{poset relations} on $A,A'$, that is relations between the
-underlying sets of $A,A'$ that are \emph{downward-closed} in $A$ and
-\emph{upper-closed} in $A'$. The obvious example of such a relation is
-the poset ordering itself $\leq_A : A \rel A'$...\max{TODO: Eric pick
-  up from here. Go into the squares for casts and how representability
-  gives us a compositional approach to proving them}
+Sticking to a poset-based semantics, if our value types are posets, we then want
+our precision derivations $c : A \ltdyn A'$ to be \emph{poset relations} on $A,
+A'$, that is relations between the underlying sets of $A, A'$ that are
+\emph{downward-closed} in $A$ and \emph{upward-closed} in $A'$.  We denote such
+a relation between $A$ and $A'$ by $c : A \rel A'$. The obvious example of a
+poset relation is the poset ordering itself, $\leq_A : A \rel A$.
 
-\eric{Read this}
-The lack of transitivity presents a major barrier towards giving a compositional
-semantics to gradual typing: if our relations are not transitive, then we cannot
-compose squares horizontally. But we have argued above that horizontal
-composition is essential for modelling the axioms of gradual typing in a
-syntax-independent manner.
+% \max{TODO: Eric pick
+%   up from here. Go into the squares for casts and how representability
+%   gives us a compositional approach to proving them}
 
-% In particular, to model the UpL/UpR/DnL/DnR rules
-% for casts in the presence of transitivity, it is sufficient to establish the
-% validity of simpler rules that do not ``build in'' composition. We can then use
-% transitivity to derive the original versions of the rules. On the other hand,
-% without transitivity we must instead validate the cast rules in the model ``from
-% scratch''. In fact, we cannot even define the semantic interpretation of type
-% precision in a syntax-independent manner. The issue is that in the definition of
-% relation, the requirement that it be representable now involves quantifying over
-% all other relations, which is circular.
+\eric{Picked up here; Max, please read this.}
 
-% So, although this approach would suffice for proving graduality, it lacks the
-% compositionality that we seek in a reusable framework for the semantics of
-% gradual typing.
+To model term precision, we use the notion of \emph{square}, which specifies a
+relation between two monotone functions. This has been used to model term
+precision in previous work on double-categorical models for gradual typing
+\cite{new-licata18}. The definition of square is as follows. Let $A_i, A_o,
+A_i', A_o'$ be partially-ordered sets. Let $c_i : A_i \rel A_i'$ and $c_o : A_o
+\rel A_o'$ be poset relations, and let $f : A_i \to A_o$ and $g : A_i' \to A_o'$
+be monotone functions. We say that $f \ltsq{c_i}{c_o} g$ if for all $x : A_i$
+and $y : A_i'$ with $(x, y) \in c_i$, we have $(f(x), g(y)) \in c_o$. We
+visualize this situation as follows:
+%
+% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBX2kiXSxbMCwxLCJBX28iXSxbMSwwLCJBX2knIl0sWzEsMSwiQV9vJyJdLFsyLDMsImciXSxbMCwxLCJmIiwyXSxbMCwyLCJjX2kiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxLDMsImNfbyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+\[\begin{tikzcd}[ampersand replacement=\&]
+  {A_i} \& {A_i'} \\
+  {A_o} \& {A_o'}
+  \arrow["{c_i}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+  \arrow["f"', from=1-1, to=2-1]
+  \arrow["g", from=1-2, to=2-2]
+  \arrow["{c_o}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+\end{tikzcd}\]
+%
 
-%Because it is convenient to make use of transitive reasoning in proving graduality, ...
-We therefore seek to formulate an ordering relation on $\li A \times \li A'$
-that \emph{is} transitive. The theorem above tells us, however, that any such
-relation must either not be a congruence with respect to $\theta$, or must not
-be closed under delay.
-\footnote{Technically, we only need to give up closure under delay on \emph{one}
-side, as the above theorem requires $R$ to be closed on both sides. We could
-define two separate ``step-semi-sensitive'' relations on $\li A \times \li A'$,
-each closed under delays on the left and right respectively. The denotation of
-term precision would then involve the conjunction of those relations. We do not
-pursue this approach in this paper, because (1) it is not clear if the
-step-semi-sensitive relations would be transitive, and (2) we would need to
-reason about the two ordering relations separately for every pair of terms.}
+\eric{Not sure if this belongs here...}
+Squares are used to model the term precision ordering $\Gamma \ltdyn \Gamma'
+\vdash M \ltdyn N : A \ltdyn A'$ as an ordering relation between the monotone
+functions denoted by $M$ and $N$:
+
+% https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXEdhbW1hIl0sWzAsMSwiVVxcbGkgQSJdLFsxLDAsIlxcR2FtbWEnIl0sWzEsMSwiVVxcbGkgQSciXSxbMiwzLCJcXHNlbXtOfSJdLFswLDEsIlxcc2Vte019IiwyXSxbMCwyLCJcXEdhbW1hXntcXGx0ZHlufSIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEsMywiVVxcbGkgYyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+\[\begin{tikzcd}[ampersand replacement=\&]
+	\Gamma \& {\Gamma'} \\
+	{U\li A} \& {U\li A'}
+	\arrow["{\Gamma^{\ltdyn}}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+	\arrow["{\sem{M}}"', from=1-1, to=2-1]
+	\arrow["{\sem{N}}", from=1-2, to=2-2]
+	\arrow["{U\li c}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+\end{tikzcd}\]
+
+where $\Gamma = \{A_1, A_2, \dots\}$ and $\Gamma' = \{A_1' , A_2', \dots\}$ and for
+each $j$, $c_j : A_j \ltdyn A_j'$, and the relation $\Gamma^{\ltdyn}$ is simply
+the product of the $c_i$.
+
+We note that for any $c$, there is a ``vertical identity'' square $\id
+\ltsq{c}{c} \id$, as can be seen by unfolding the definition. Similarly, for any
+montone function $f : A_i \to A_o$, there is a ``horizontal identity'' square $f
+\ltsq{r(A_i)}{r(A_o)}$ by the definition of monotone function.
+We likewise make the analogous definition of square for error domains.
+
+Having defined squares, we now discuss the interpretation of the cast rules UpL
+and UpR \footnote{The cast rules DnL and DnR involve error domain relations and
+squares, which for the sake of this discussion are not relevant.}. These rules
+specify a relationship between the semantics of type precision and the
+corresponding casts (see Section \ref{sec:GTLC}). For example, consider the UpL
+rule. It states that if $c : A_1 \ltdyn A_2$ and $c' : A_2 \ltdyn A_3$ and $x :
+A_1$ is related to $y : A_3$ via the composition $c \comp c'$, then the upcast
+$\upc{c} x$ is related to $y$ via $c'$. This is equivalent to the existence of
+the following square:
+%
+% https://q.uiver.app/#q=WzAsNSxbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMyw0LCJjJyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMywiXFx1cGN7Y30iLDJdLFsyLDQsIlxcaWQiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+\[\begin{tikzcd}[ampersand replacement=\&]
+  {A_1} \& {A_2} \& {A_3} \\
+  {A_2} \&\& {A_3}
+  \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+  \arrow["{\upc{c}}"', from=1-1, to=2-1]
+  \arrow["{c'}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
+  \arrow["\id", from=1-3, to=2-3]
+  \arrow["{c'}"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
+\end{tikzcd}\]
+%
+Observe that although this is really a fact about the relationship between $c$
+and $\upc{c}$, we must universally quantify over all relations $c'$. This means
+we cannot require the existence of this square as part of the definition of a
+monotone relation, as it is self-referential. This would seem to imply that we
+cannot give a syntax-independent semantic model for gradual typing. However, the
+key observation is that through compositionality, we can \emph{derive} the above
+squares from simpler ones that do not involve composition of relations. These
+simpler squares have been used in prior work for giving a compositional
+semantics to gradual typing \cite{new-licata18}. Below is the square
+corresponding to UpL:
+%
+% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzIiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzIiXSxbMCwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJyKEFfMikiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxcdXBje2N9IiwyXSxbMiwzLCJcXGlkIl1d
+\[\begin{tikzcd}[ampersand replacement=\&]
+  {A_1} \& {A_2} \\
+  {A_2} \& {A_2}
+  \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+  \arrow["{\upc{c}}"', from=1-1, to=2-1]
+  \arrow["\id", from=1-2, to=2-2]
+  \arrow["{r(A_2)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+\end{tikzcd}\]
+%
+This square, and the corresponding one related to UpR, together express the
+\emph{representability} of $c$ by the upcast morphism $\upc{c}$, which
+intuitively says that $c$ is determined by the behavior of $\upc{c}$. We can use
+this simpler square to derive the original UpL square by horizontally composing
+with the identity square for $c'$:
+%
+% https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMSwxLCJBXzIiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzMsNSwicihBXzIpIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNSw0LCJjJyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMywiXFx1cGN7Y30iLDJdLFsyLDQsIlxcaWQiXSxbMSw1LCJcXGlkIiwyXV0=
+\[\begin{tikzcd}[ampersand replacement=\&]
+  {A_1} \& {A_2} \& {A_3} \\
+  {A_2} \& {A_2} \& {A_3}
+  \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+  \arrow["{\upc{c}}"', from=1-1, to=2-1]
+  \arrow["{c'}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
+  \arrow["\id"', from=1-2, to=2-2]
+  \arrow["\id", from=1-3, to=2-3]
+  \arrow["{r(A_2)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+  \arrow["{c'}"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
+\end{tikzcd}\]
+%
+The existence of these simpler squares for a relation can be stated without
+reference to other relations, and hence we can require that these squares exist
+in the definition of relation without running into the circularity issue we
+faced with the original rules. This allows us to formulate a notion of model
+that is independent of the particular syntax of any cast calculus.
+
+We now see why compositionality of squares, and hence transitivity of the
+relations on the objects that model types, is an essential property that we
+cannot sacrifice.
+
+\subsection{Resolution: Splitting Error Ordering and Bisimilarity}\label{sec:lock-step-and-weak-bisim}
+
+Given that compositional reasoning is essential, we seek to formulate an
+ordering relation on $\li A$ that is transitive. The theorem above
+tells us, however, that any such relation must either not be a congruence with
+respect to $\theta$, or must not be left- and right-step-insensitive.
+\footnote{Technically, we only need to give up step insensitivity on \emph{one}
+side, as the above theorem requires $R$ to be both left- and
+right-step-insensitive. We could define two separate ``step-semi-sensitive''
+relations on $\li A$, each closed under delays on the left and
+right respectively. The denotation of term precision would then involve the
+conjunction of those relations. We do not pursue this approach in this paper,
+because (1) it is not clear if the step-semi-sensitive relations would be
+transitive, and (2) we would need to reason about the two ordering relations
+separately for every pair of terms.}
 %
@@
 would not be able to prove properties of the relation using \lob-induction,
-e.g., that $\ext{f}{}$ is a \emph{monotone} function with respect to the
-ordering.
+e.g., that the monadic extension of a monotone function $f$ is a \emph{monotone}
+function with respect to the ordering.
 %and moreover this would mean that $\theta$ would not be a monotone function.
 
-Thus, we choose to sacrifice closure under delays. That is, we will define an
-ordering relation that requires terms to be in ``lock-step''. In order for two
-computations to be related in this ordering, they must have the same stepping
-behavior (or the left-hand side must be an error). To deal with terms that take
-differing numbers of steps, we then define a separate relation called \emph{weak
-bisimilarity} that relates terms that are extensionally equal and may differ by
-a finite number of delays. Then, the semantics of the error ordering for the
-guarded lift monad will involve a combination of these two relations: the
-``closure'' of the lock-step error ordering under weak bisimilarity on both
+Thus, we choose to sacrifice left and right step-insensitivity. That is, we will
+define an ordering relation that requires terms to be in ``lock-step''. In order
+for two computations to be related in this ordering, they must have the same
+stepping behavior (or the left-hand side must be an error). To deal with terms
+that take differing numbers of steps, we then define a separate relation called
+\emph{weak bisimilarity} that relates terms that are extensionally equal and may
+differ by a finite number of delays. Then, the semantics of the error ordering
+for the guarded lift monad will involve a combination of these two relations:
+a ``closure'' of the lock-step error ordering under weak bisimilarity on both
 sides.
@@
 
-% Doing so, we define a \emph{lock-step} error ordering, where roughly speaking,
-% in order for computations to be related, they must have the same stepping
-% behavior. We then formulate a separate relation, \emph{weak bisimilarity}, that
-% relates computations that are extensionally equal and may differ only in their
-% stepping behavior. % up to a finite number of \delta's
-% Finally, the semantics of term precision will involve a combination of these two
-% relations, a sort of closure of the lock-step ordering under weak bisimilarity
-% on both sides. This decomposition has the advantage that we can recover some
-% transitive reasoning and push the parts involving stepping to the margins of the
-% development.
-
-\subsection{Resolution: Splitting Error Ordering and Bisimilarity}\label{sec:lock-step-and-weak-bisim}
-
-As mentioned, to recover transitive reasoning we will decompose the error
-ordering on the guarded lift monad into two separate relations. The first is the
-\emph{lock-step error-ordering} (also called the ``step-sensitive error
-ordering''), the idea being that two computations $l$ and $l'$ are related if
-they are in lock-step with regard to their intensional behavior, up to $l$
-erroring. This will be the partial ordering with which we equip the guarded lift monad.
-Figure \ref{fig:lock-step-error-ordering} gives the definition.
+More formally, we define the \emph{lock-step error-ordering}, with the idea
+being that two computations $l$ and $l'$ are related if they are in lock-step
+with regard to their intensional behavior, up to $l$ erroring. Figure
+\ref{fig:lock-step-error-ordering} gives the definition.
 
@@
         &\eta\, x \ltls \eta\, y \text{ if } 
-            x \mathbin{R} y \\
+            x \mathbin{\le_A} y \\
         %		
@@
 When both sides are $\eta$, then we check that the returned values are related
-in $R$. The error term $\mho$ is below everything. Lastly, if both sides step
+in $\le_A$. The error term $\mho$ is the least element. Lastly, if both sides step
 (i.e., are a $\theta$) then we compare the resulting computations one time step
@@
 has those properties. The lock-step ordering is therefore the partial ordering
-we will associate with $\li X$.
+we will associate with $\li A$.
 %
 We similarly define a heterogeneous version of this ordering between the lifts
-of two different types $X$ and $Y$, parameterized by a relation $R$ between $X$
-and $Y$.
+of two different types $A$ and $A'$, parameterized by a poset relation $c : A \rel A'$.
 
@@
 %
-Weak bisimilarity on $\li X$ is parameterized by an underlying relation $R$ on $X$
+Weak bisimilarity on $\li A$ is parameterized by an underlying relation $R$ on $A$
 and is defined by guarded fixpoint as shown in Figure \ref{fig:weak-bisimilarity}.
@@
 
-% This issue of non-transitivity will be resolved when we consider the relation's \emph{globalization}.
-
 Now we can define the denotation of term precision for closed terms of type
@@
 
-% Indeed, we will need to weaken our abstract notion of categorical model to no
-% longer require transitivity of term precision. On the other hand, we now have
-% the ability to apply transitive reasoning when working with the lock-step
-% ordering. This allows us to carry out complex proofs compositionally, a key goal
-% of our framework.
+\subsection{Adjusting the Cast Rules}\label{sec:adjusting-cast-rules}
 
+\eric{We may want to introduce error domain relations and squares at this point,
+since the squares for DnL and DnR involve lift.}
+
+We have seen how we can recover transitive reasoning by working with the
+lock-step error ordering. However, it turns out that we cannot completely
+isolate the reasoning about steps from the reasoning about the ordering.
+% Unfortunately, by defining the partial order on the guarded lift monad to be
+% the lock-step error ordering, we introduce a new issue.
+In particular, the cast rules required for proving graduality no longer hold if
+interpreted using the lock-step ordering. The source of the problem is that the
+downcast from the dynamic type to a function takes a step, i.e., involves a
+$\theta$, and the lock-step ordering requires both sides to be in sync. More
+specifically, consider the DnL rule for $\iarr \colon \dyntodyn \ltdyn\, \dyn$
+shown below:
+%
+\begin{mathpar}
+  \inferrule*{M \ltdyn N : \iarr}
+             {\dnc{\iarr}{M} \ltdyn N : c}
+\end{mathpar}
+%
+This rule corresponds to the following square:
+
+% https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXGxpIEQiXSxbMCwxLCJcXGxpIFUoRCBcXGFyciBcXGxpIEQpIl0sWzEsMCwiXFxsaSBEIl0sWzEsMSwiXFxsaSBEIl0sWzAsMiwiXFxsZXFfe1xcbGkgRH0iLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxLDMsIlxcbGlcXGlhcnIiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxcZG5je1xcaWFycn0iLDJdLFsyLDMsIlxcaWQiXV0=
+\[\begin{tikzcd}[ampersand replacement=\&]
+	{\li D} \& {\li D} \\
+	{\li U(D \arr \li D)} \& {\li D}
+	\arrow["{\leq_{\li D}}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+	\arrow["{\dnc{\iarr}}"', from=1-1, to=2-1]
+	\arrow["\id", from=1-2, to=2-2]
+	\arrow["\li\iarr"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+\end{tikzcd}\]
+%
+where $\li \iarr$ is the heterogeneous version of the lock-step error ordering
+discussed above applied to the relation $\iarr$. We note that the precise
+definition of $\iarr$ as a poset relation is not important for this discussion.
+%
+We now recall the definition of the semantics of a downcast given in Section
+\ref{sec:term-interpretation}. The downcast on the left will insert a $\theta$
+in the case where the value of type $D$ is a later-function $\tilde{f}$.
+%
+Thus, in order for the downcast to be related to the RHS in the lock-step error
+ordering on $\li (D \to \li D)$, the RHS must be of the form $\theta(\dots)$,
+and moreover, after one time step, the argument of $\theta$ on the LHS must be
+related to the argument of $\theta$ on the RHS. As it stands now, this need not
+be the case, e.g., if the RHS is of the form $\eta(\dots)$. Thus, we conclude
+that the DnL rule does not in general hold under the lock-step error ordering.
+However, observe that we can remedy this particular situation by inserting a
+``delay'' on the right, i.e., wrapping the RHS in $\delta^* = (\delta \circ
+\eta)^\dagger$ where $\delta = \theta \circ \nxt$. Notice that this delay
+function is weakly bisimilar to the identity function in that $\delta^*\, x
+\bisim x$ for all $x$. \footnote{This follows from the fact that $\delta$ is
+weakly bisimilar to the identity ($\delta\, x \bisim x$ for all $x$) and that
+$-^\dagger$ preserves weak bisimilarity.} Thus, it will disappear when we apply
+the bisimilarity-closure construction described above. That is, this delay makes
+no difference in the extensional setting, but its presence is crucial in the
+intensional setting.
+
+\subsection{Extending the Model to Higher-Order Types}
+
+% At this point, we could carry out the full construction of a concrete relational
+% model with these additional features.
+
+The above techniques give a semantic interpretation of term precision for closed
+terms of base type. We now consider how to extend these constructions to
+potentially-open terms of \emph{all} types. In particular, in order to model
+higher-order data types (i.e., functions) we need to equip \emph{every} semantic
+object with not only a partial ordering relation but also a ``bisimilarity''
+relation that is reflexive and symmetric. We must similarly equip every object
+with a structured set of delays that can be inserted and manipulated to ensure
+the appropriate cast rules hold at all higher-order types. For example, the
+upcast for a derivation $c_i \ra c_o$ involves the downcast corresponding to
+$c_i$ and the upcast corresponding to $c_o$. We must be able to insert delays in
+a functorial manner to mimic the structure of the casts. In the next section, we
+will make the necessary definitions and describe the relevant constructions.
+
+% For the sake of reusability and modularity, rather than carry out this
+% construction in the concrete setting developed here, we will instead return to
+% the abstract setting and adjust our definition of model to account for these
+% requirements. We will break the construction into smaller steps and isolate the
+% pieces that require the techniques of SGDT from those that do not. Then with
+% this framework at our disposal, we will return to the construction of a concrete
+% model in Section \ref{sec:concrete-model}.
+% , taking as a starting point the definitions introduced in the current section.
+
+% Thus in the next section we define revised notions of a model of
+% gradual typing based on the lessons learned in the present section.
+
+% The lack of transitivity presents a major barrier towards giving a compositional
+% semantics to gradual typing: if our relations are not transitive, then we cannot
+% compose squares horizontally. But we have argued above that horizontal
+% composition is essential for modelling the axioms of gradual typing in a
+% syntax-independent manner.
+
+% In particular, to model the UpL/UpR/DnL/DnR rules
+% for casts in the presence of transitivity, it is sufficient to establish the
+% validity of simpler rules that do not ``build in'' composition. We can then use
+% transitivity to derive the original versions of the rules. On the other hand,
+% without transitivity we must instead validate the cast rules in the model ``from
+% scratch''. In fact, we cannot even define the semantic interpretation of type
+% precision in a syntax-independent manner. The issue is that in the definition of
+% relation, the requirement that it be representable now involves quantifying over
+% all other relations, which is circular.
+
+% So, although this approach would suffice for proving graduality, it lacks the
+% compositionality that we seek in a reusable framework for the semantics of
+% gradual typing.
+
+%Because it is convenient to make use of transitive reasoning in proving graduality, ...
+
+% Doing so, we define a \emph{lock-step} error ordering, where roughly speaking,
+% in order for computations to be related, they must have the same stepping
+% behavior. We then formulate a separate relation, \emph{weak bisimilarity}, that
+% relates computations that are extensionally equal and may differ only in their
+% stepping behavior. % up to a finite number of \delta's
+% Finally, the semantics of term precision will involve a combination of these two
+% relations, a sort of closure of the lock-step ordering under weak bisimilarity
+% on both sides. This decomposition has the advantage that we can recover some
+% transitive reasoning and push the parts involving stepping to the margins of the
+% development.
+
+\begin{comment}
 \subsection{Modeling Type and Term Precision}
@@
 
-We model a type precision relation $c : A \ltdyn A'$ as a \emph{monotone relation}, i.e., a
-relation $c$ that is upward-closed under the relation on $A'$ and
-downward-closed under the relation on $A$. We denote such a relation between $A$
-and $A'$ by $c : A \rel A'$. The relation on the poset $A$ is denoted $r(A)$.
+We model a type precision relation $c : A \ltdyn A'$ as a \emph{monotone
+relation}, i.e., a relation $c$ that is upward-closed under the relation on $A'$
+and downward-closed under the relation on $A$. We denote such a relation between
+$A$ and $A'$ by $c : A \rel A'$. The relation on the poset $A$ is denoted
+$r(A)$.
 % This extends to products $c_1 \times c_2$ in the obvious way.
@@
 
-We likewise define a notion of \emph{error domain relation} between error
-domains $B$ and $B'$. An error domain relation $d : B \rel B'$ is a monotone
-relation on the underlying sets that preserves error and respects $\theta$. More
-formally:
-
-\begin{definition}
-  Let $B$ and $B'$ be error domains. An \emph{error domain relation} between
-  $B$ and $B'$ is a monotone relation $d$ on the underlying partially-ordered sets such that:
-  \begin{enumerate}
-     \item (Respects error): For all $y \in B'$, we have $(\mho_B, y) \in d$.
-     \item (Preserves $\theta$): For all $\tilde{x}$ in $\laterhs B$ and
-     $\tilde{y} \in \laterhs B'$, if $\laterhs_t ((\tilde{x}_t, \tilde{y}_t) \in
-     d)$, then $(\theta_B (\tilde{x}), \theta_{B'} (\tilde{y})) \in d$. 
-  \end{enumerate}
-\end{definition}
-
-Because of a technical restriction involving the interaction between
-propositional truncation and the later modality, we cannot define the
-composition of error domain relations as simply the composition of their
-underlying predomain relations. Instead, we define composition of error domain
-relations $d$ on $B_1$ and $B_2$ and $d'$ on $B_2$ and $B_3$ to be the least
-relation containing $d$ and $d'$ that is downward closed under $\le_{B_1}$,
-upward-closed under $\le_{B_3}$, respects error, and preserves $\theta$.
-Specifically, it is defined inductively by the following rules:
-%
-\begin{mathpar}
-    \inferrule*[right = Comp]
-    {b_1 \mathbin{d} b_2 \and b_2 \mathbin{d'} b_3}
-    {b_1 \mathbin{d \relcomp d'} b_3}
-
-    \inferrule*[right = DnClosed]
-    {b_1' \le_{B_1} b_1 \and b_1 \mathbin{d \relcomp d'} b_3}
-    {b_1' \mathbin{d \relcomp d'} b_3}
-
-    \inferrule*[right = UpClosed]
-    {b_1 \mathbin{d \relcomp d'} b_3 \and b_3 \le_{B_3} b_3'}
-    {b_1 \mathbin{d \relcomp d'} b_3'}
-
-    \inferrule*[right = PresErr]
-    { }
-    {\mho_{B_1} \mathbin{d \relcomp d'} b_3}
-
-    \inferrule*[right = PresTheta]
-    {\later_t( \tilde{b_1} \mathbin{d \relcomp d'} \tilde{b_3} ) }
-    {\theta_{B_1}(\tilde{b_1}) \mathbin{d \relcomp d'} \theta_{B_3}(\tilde{b_3}) }
-\end{mathpar}
-
-% That is,
-% $(\mho_B, y) \in d$ for all $y : B'$ and if $\laterhs_t ((\tilde{x}_t,
-% \tilde{y}_t) \in d)$, then $(\theta_B (\tilde{x}), \theta_{B'} (\tilde{y})) \in
-% d$.
-
-The functors $U$, $F$, $\times$, and $\arr$ introduced in the previous
-section extend to relations. The functor $U$ simply takes the underlying
-monotone relation of an error domain relation. Given a relation $c_1 : A_1 \rel
-A_1'$ and $c_2 : A_2 \rel A_2'$, the relation $c_1 \times c_2 : (A_1 \times A_2)
-\rel (A_1' \times A_2')$ is defined by $((x, y), (x', y')) \in (c_1 \times c_2)$
-if and only if $(x, x') \in c_1$ and $(y, y') \in c_2$. Given a relation $c : A
-\rel A'$ and an error domain relation $d : B \rel B'$, the relation $c \arr d :
-(A \arr B) \rel (A' \arr B')$ is the error domain relation defined by $(f, g)
-\in (c \arr d)$ if and only if for all $x : A$ and $x' : A'$ with $(x , x') \in
-c$ we have $(f(x), g(x')) \in Ud$. The fact that this relation preserves error
-and respects $\theta$ follows from the corresponding properties for $d$.
-%
-We we will postpone the description of the action of $F$ on relations to later
-in this section.
-
 \subsubsection{Modelling Term Precision}\label{sec:modeling-term-precision}
-To model term precision, we need the notion of \emph{square}, which specifies a
-relation between two monotone functions. The definition is as follows. Let $A_i,
-A_o, A_i', A_o'$ be partially-ordered sets. Let $c_i : A_i \rel A_i'$ and $c_o :
-A_o \rel A_o'$ be monotone relations, and let $f : A_i \to A_o$ and $g : A_i'
-\to A_o'$ be monotone functions. We say that $f \ltsq{c_i}{c_o} g$ if for all $x
-: A_i$ and $y : A_i'$ with $(x, y) \in c_i$, we have $(f(x), g(y)) \in c_o$. We
-visualize this situation as the following square:
-
-% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBX2kiXSxbMCwxLCJBX28iXSxbMSwwLCJBX2knIl0sWzEsMSwiQV9vJyJdLFsyLDMsImciXSxbMCwxLCJmIiwyXSxbMCwyLCJjX2kiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxLDMsImNfbyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
-\[\begin{tikzcd}[ampersand replacement=\&]
-	{A_i} \& {A_i'} \\
-	{A_o} \& {A_o'}
-	\arrow["{c_i}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-	\arrow["f"', from=1-1, to=2-1]
-	\arrow["g", from=1-2, to=2-2]
-	\arrow["{c_o}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-\end{tikzcd}\]
-
-We note that for any $c$, there is a ``vertical identity'' square $\id
-\ltsq{c}{c} \id$, as can be seen by unfolding the definition. Similarly, for any
-montone function $f : A_i \to A_o$, there is a ``horizontal identity'' square $f
-\ltsq{r(A_i)}{r(A_o)}$ by the definition of monotone function.
-We likewise make the analogous definition of square for error domains.
-
-Having defined squares, we now discuss the interpretation of the cast rules
-UpL/UpR/DnL/DnR. These rules specify a relationship between the semantics of
-type precision and the corresponding casts (see Section \ref{sec:GTLC}). In
-particular, they specify that the relations are \emph{representable} by the
-casts. The UpL and UpR rules refer to upcasts, and since upcasts are pure, these
-rules apply to \emph{value} relations $c$. The DnL and DnR rules refer to
-downcasts, and since downcasts are effectful, these rules apply to
-\emph{computation} relations $d$.
-
-As an example, consider the UpL rule. It states that if $c : A_1 \ltdyn A_2$ and
-$c' : A_2 \ltdyn A_3$ and $x : A_1$ is related to $y : A_3$ via the composition
-$c \comp c'$, then the upcast $\upc{c} x$ is related to $y$ via $c'$. This is
-equivalent to the existence of the following square:
-
-% https://q.uiver.app/#q=WzAsNSxbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMyw0LCJjJyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMywiXFx1cGN7Y30iLDJdLFsyLDQsIlxcaWQiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
-\[\begin{tikzcd}[ampersand replacement=\&]
-	{A_1} \& {A_2} \& {A_3} \\
-	{A_2} \&\& {A_3}
-	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-	\arrow["{\upc{c}}"', from=1-1, to=2-1]
-	\arrow["{c'}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
-	\arrow["\id", from=1-3, to=2-3]
-	\arrow["{c'}"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
-\end{tikzcd}\]
-
-The three squares corresponding to the other rules are analogous. Observe that
-although this is really a fact about the relationship between $c$ and $\upc{c}$,
-we must universally quantify over all relations $c'$. This means we cannot
-require the existence of this square as part of the definition of a monotone
-relation, as it is self-referential. This would seem to imply that we cannot
-give a syntax-independent semantic model for gradual typing. However, the key
-observation is that through compositionality, we can \emph{derive} the above
-squares from simpler ones that do not involve composition of relations. For
-example, below is the simpler square corresponding to UpL:
-
-% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzIiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzIiXSxbMCwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJyKEFfMikiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxcdXBje2N9IiwyXSxbMiwzLCJcXGlkIl1d
-\[\begin{tikzcd}[ampersand replacement=\&]
-	{A_1} \& {A_2} \\
-	{A_2} \& {A_2}
-	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-	\arrow["{\upc{c}}"', from=1-1, to=2-1]
-	\arrow["\id", from=1-2, to=2-2]
-	\arrow["{r(A_2)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-\end{tikzcd}\]
-
-We can use this to derive the original UpL square by composing this square with
-the identity square for $c'$:
-
-% https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMSwxLCJBXzIiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzMsNSwicihBXzIpIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNSw0LCJjJyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMywiXFx1cGN7Y30iLDJdLFsyLDQsIlxcaWQiXSxbMSw1LCJcXGlkIiwyXV0=
-\[\begin{tikzcd}[ampersand replacement=\&]
-	{A_1} \& {A_2} \& {A_3} \\
-	{A_2} \& {A_2} \& {A_3}
-	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-	\arrow["{\upc{c}}"', from=1-1, to=2-1]
-	\arrow["{c'}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
-	\arrow["\id"', from=1-2, to=2-2]
-	\arrow["\id", from=1-3, to=2-3]
-	\arrow["{r(A_2)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-	\arrow["{c'}"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
-\end{tikzcd}\]
-
-The existence of these simpler squares for a relation can be stated without
-reference to other relations. Thus, we can require that these squares exist in
-the definition of relation without running into the circularity issue we faced
-with the original rules.
-
-Finally, we use squares to model the term precision ordering $\Gamma \ltdyn
-\Gamma' \vdash M \ltdyn N : A \ltdyn A'$ as an ordering relation between the
-monotone functions denoted by $M$ and $N$:
-
-% https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXEdhbW1hIl0sWzAsMSwiVUYgQSJdLFsxLDAsIlxcR2FtbWEnIl0sWzEsMSwiVUZBJyJdLFsyLDMsIlxcc2Vte059Il0sWzAsMSwiXFxzZW17TX0iLDJdLFswLDIsIlxcR2FtbWFee1xcbHRkeW59IiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJVRmMiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
-\[\begin{tikzcd}[ampersand replacement=\&]
-	\Gamma \& {\Gamma'} \\
-	{UF A} \& {UFA'}
-	\arrow["{\Gamma^{\ltdyn}}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-	\arrow["{\sem{M}}"', from=1-1, to=2-1]
-	\arrow["{\sem{N}}", from=1-2, to=2-2]
-	\arrow["UFc"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-\end{tikzcd}\]
-
-where $\Gamma = \{A_1, A_2, \dots\}$ and $\Gamma' = \{A_1' , A_2', \dots\}$ and for
-each $j$, $c_j : A_j \ltdyn A_j'$, and the relation $\Gamma^{\ltdyn}$ is simply
-the product of the $c_i$.
 
@@
 
-\subsection{Adjusting the Cast Rules}\label{sec:adjusting-cast-rules}
-
-We have seen how we can recover transitive reasoning by working with the
-lock-step error ordering. However, it turns out that we cannot completely
-isolate the reasoning about steps from the reasoning about the ordering.
-% Unfortunately, by defining the partial order on the guarded lift monad to be
-% the lock-step error ordering, we introduce a new issue.
-In particular, the cast rules required for proving graduality no longer hold if
-interpreted using the lock-step ordering. The source of the problem is that the
-downcast from the dynamic type to a function takes a step, i.e., involves a
-$\theta$. Consider the DnL rule shown below:
-%
-\begin{mathpar}
-  \inferrule*{c : A \ltdyn A' \and M \ltdyn N : A'}
-             {\dnc{c}{M} \ltdyn N : c}
-\end{mathpar}
-%
-Suppose that $c$ is $\iarr \colon \dyntodyn \ltdyn\, \dyn$, and recall the
-definition of the semantics of a downcast given in Section
-\ref{sec:term-interpretation}. The DnL rule corresponds to the following square:
-
-% https://q.uiver.app/#q=WzAsNCxbMCwwLCJGRCJdLFswLDEsIkZVKEQgXFxhcnIgRkQpIl0sWzEsMCwiRkQiXSxbMSwxLCJGRCJdLFswLDIsInIoRkQpIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJGXFxpYXJyIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJcXGRuY3tcXGlhcnJ9IiwyXSxbMiwzLCJcXGlkIl1d
-\[\begin{tikzcd}[ampersand replacement=\&]
-	FD \& FD \\
-	{FU(D \arr FD)} \& FD
-	\arrow["{r(FD)}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-	\arrow["{\dnc{\iarr}}"', from=1-1, to=2-1]
-	\arrow["\id", from=1-2, to=2-2]
-	\arrow["F\iarr"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-\end{tikzcd}\]
-
-The downcast on the left will insert a $\theta$ in the case where the value of
-type $D$ is a later-function $\tilde{f}$.
-%
-Thus, in order for the downcast to be related to the RHS in the lock-step error
-ordering on $\li (D \to \li D)$, the RHS must be of the form $\theta(\dots)$,
-and moreover, after one time step, the argument to $\theta$ on the LHS must be
-related to the argument of $\theta$ on the RHS. As it stands now, this need not
-be the case. Thus, we conclude that the DnL rule does not in general hold under
-the lock-step error ordering. However, observe that we can remedy this
-particular situation by inserting a ``delay'' on the right, i.e., wrapping the
-RHS in $\delta = \theta \circ \nxt$. Notice that this delay function is weakly
-bisimlar to the identity, and thus it will disappear when we apply the
-bisimilarity-closure construction described above. That is, this delay makes no
-difference in the extensional setting, but its presence is crucial in the
-intensional setting.
-
-% TODO Maybe show the correct form of the rule in the special case $c = inj-arr$.
-
-\subsection{Extending the Model to Higher-Order Types}
-
-% At this point, we could carry out the full construction of a concrete relational
-% model with these additional features.
-
-The above techniques give a semantic interpretation of term precision for closed
-terms of base type. We now consider how to extend these constructions to
-potentially-open terms of \emph{all} types. In particular, in order to model
-higher-order data types (i.e., functions) we need to equip \emph{every} semantic
-object with not only a partial ordering relation but also a ``bisimilarity''
-relation that is reflexive and symmetric. We must similarly equip every object
-with a structured set of delays that can be inserted and manipulated to ensure
-the appropriate cast rules hold at all higher-order types. For example, the
-upcast for a derivation $c_i \ra c_o$ involves the downcast corresponding to
-$c_i$ and the upcast corresponding to $c_o$. We must be able to insert delays in
-a functorial manner to mimic the structure of the casts. In the next section, we
-will make the necessary definitions and describe the relevant constructions.
-
-% For the sake of reusability and modularity, rather than carry out this
-% construction in the concrete setting developed here, we will instead return to
-% the abstract setting and adjust our definition of model to account for these
-% requirements. We will break the construction into smaller steps and isolate the
-% pieces that require the techniques of SGDT from those that do not. Then with
-% this framework at our disposal, we will return to the construction of a concrete
-% model in Section \ref{sec:concrete-model}.
-% , taking as a starting point the definitions introduced in the current section.
-
-% Thus in the next section we define revised notions of a model of
-% gradual typing based on the lessons learned in the present section.
+\end{comment}

### 2024-07-10 paper-new/towards-relational-model.tex
revisions to the relational motivation section

@@
-\section{Towards a Full Relational Model}\label{sec:towards-relational-model}
+\section{Challenges to a Guarded Model of Graduality}\label{sec:towards-relational-model}
 
@@
 equipping $\omega$-CPOs with a second ordering $\ltdyn$ they call the
-``error ordering'' such that a term precision relation $M \ltdyn M' :
-r(A)$ is modeled by $\sem{M} \ltdyn \sem{M'}$ in the semantics. Then
-the error ordering on the free error domain $\li A$ is based on the
-graduality property: essentially either $\sem{M}$ errors or both terms
-diverge or both terms terminate with values related by the ordering
-relation on $A$.
+``error ordering'' such that a term precision relationship at a fixed
+type $M \ltdyn M' : r(A)$ implies the error ordering $\sem{M} \ltdyn
+\sem{M'}$ in the semantics. Then the error ordering on the free error
+domain $\li A$ is based on the graduality property: essentially either
+$\sem{M}$ errors or both terms diverge or both terms terminate with
+values related by the ordering relation on $A$.
 %
@@
 would be to similarly enhance our value types and computation types to
-additionally carry a poset structure. For this value types can simply
-be posets. For computation types, we want our error domains to be
-posets as well. To consider what additional properties we want the
-error domains to have, let's consider the particular case of the free
+additionally carry a poset structure. For most type formers this
+ordering can be defined as lifting poset structures on the
+subformulae, the key design choice is in the ordering on the free
 error domain $\li A$.
@@
 as stated mentions divergence, but we only work indirectly with
-diverging computations in the form of the abstract steps/think
-maps. The closest direct encoding of the graduality property is an
-ordering we call the \emph{step-insensitive error ordering}, defined
-by guarded recursion in
-Figure~\ref{fig:step-insensitive-error-ordering}.
+diverging computations in the form of the think maps. The closest
+direct encoding of the graduality property is an ordering we call the
+\emph{step-insensitive error ordering}, defined by guarded recursion
+in Figure~\ref{fig:step-insensitive-error-ordering}.
 %
@@
         %	
-        &\theta\, \tilde{l} \semltbad \mho \text { iff } 
-            \exists n. \theta\, \tilde{l} = \delta^n(\mho) \\
-        %	
         &\eta\, x \semltbad \theta\, \tilde{l'} \text { iff }
@@
 
-This definition of $\semltbad$ is an entirely reasonable attempt to
-model graduality: for instance it is true that if $l \semltbad l'$ for
-$\li \mathbb{N}$ where $\mathbb{N}$ is the flat poset whose ordering
-is equality, then the bigstep semantics we derive for $l, l'$ do in
-fact satisfy the intended graduality relation.
+This definition of $\semltbad$ is sufficient to model the graduality
+property in that if $l \semltbad l'$ for $l,l' : \li \mathbb{N}$ where
+$\mathbb{N}$ is the flat poset whose ordering is equality, then the
+bigstep semantics we derive for $l, l'$ do in fact satisfy the
+intended graduality relation.
 %
@@
   \end{enumerate}
-  Then for any $l : U(\li A)$, we have $l R \Omega$. If $R$ is left
+  Then for any $l : U(\li A)$, we have $l \binrel{R} \Omega$. If $R$ is left
   step-insensitive instead then $\Omega \binrel{R} x$.
@@
 \begin{proof}
-  By L\"ob induction, in the appendix.
+  By L\"ob induction, in the appendix.\max{Eric make sure this proof is in the appendix}
 \end{proof}
@@
 $\theta$-congruence and step-insensitive on both sides, but not
-generally trivial: e.g., for the flat poset on $\mathbb N$, $\eta 0 \semltbad
+trivial: e.g., for the flat poset on $\mathbb N$, $\eta 0 \semltbad
 \eta 1$ is false.
@@
 step-insensitive ordering. However some level of transitivity is
-absolutely essential to providing compositional reasoning, and is used
-pervasively in the New-Licata development.
+absolutely essential to providing compositional reasoning, and
+transitivity is used pervasively in the prior work by New and Licata.
 %
@@
 %
-Sticking to a poset-based semantics, if our value types are posets, we then want
-our precision derivations $c : A \ltdyn A'$ to be \emph{poset relations} on $A,
-A'$, that is relations between the underlying sets of $A, A'$ that are
-\emph{downward-closed} in $A$ and \emph{upward-closed} in $A'$.  We denote such
-a relation between $A$ and $A'$ by $c : A \rel A'$. The obvious example of a
-poset relation is the poset ordering itself, $\leq_A : A \rel A$.
+Sticking to a poset-based semantics, if our value types are posets, we
+then want our precision derivations $c : A \ltdyn A'$ to be
+\emph{poset relations} on $A, A'$, that is relations between the
+underlying sets of $A, A'$ that are \emph{downward-closed} in $A$ and
+\emph{upward-closed} in $A'$.  We denote such a relation between $A$
+and $A'$ by $c : A \rel A'$. The paradigmatic example of a poset
+relation is the poset ordering itself, $\leq_A : A \rel A$.
 
@@
 
-\eric{Picked up here; Max, please read this.}
-
-To model term precision, we use the notion of \emph{square}, which specifies a
-relation between two monotone functions. This has been used to model term
-precision in previous work on double-categorical models for gradual typing
-\cite{new-licata18}. The definition of square is as follows. Let $A_i, A_o,
-A_i', A_o'$ be partially-ordered sets. Let $c_i : A_i \rel A_i'$ and $c_o : A_o
-\rel A_o'$ be poset relations, and let $f : A_i \to A_o$ and $g : A_i' \to A_o'$
-be monotone functions. We say that $f \ltsq{c_i}{c_o} g$ if for all $x : A_i$
-and $y : A_i'$ with $(x, y) \in c_i$, we have $(f(x), g(y)) \in c_o$. We
-visualize this situation as follows:
+\eric{Picked up here; Max, please read this.}  These relations are
+then used to model term precision. For closed programs $M \ltdyn M' :
+c$ where $c : A \ltdyn A'$ we have that these denote elements of the
+free error domain $\sem{M},\sem{M'} : U\li \sem{A}$ and we model this
+as a relation holding $\sem{M} (U\li\sem{c}) \sem{M'}$ where
+$U\li\sem{c}$ is a relational lifting of $U\circ\li$.
+%
+More generally, $M$ and $M'$ can be \emph{open} terms, in which case
+they denote functions, not just elements. We model that as not simply
+being elements related by a relation but in a typical
+logical-relations style of \emph{preserving} relatedness: if they are
+passed in inputs related by the domain relations then their output is
+related by the codomain relation.
+%
+To capture this preservation of relations property, we use the notion
+of a \emph{square}, which specifies a relation between two monotone
+functions. The definition of square is as follows. Let $A_i,
+A_o, A_i', A_o'$ be partially-ordered sets. Let $c_i : A_i \rel A_i'$
+and $c_o : A_o \rel A_o'$ be poset relations, and let $f : A_i \to
+A_o$ and $g : A_i' \to A_o'$ be monotone functions. We say that $f
+\ltsq{c_i}{c_o} g$ if for all $x : A_i$ and $y : A_i'$ with $(x, y)
+\in c_i$, we have $(f(x), g(y)) \in c_o$. We visualize this situation
+as follows:
 %
@@
 %
+Squares are then used to model the term precision ordering in that
+$\Delta \vdash M \ltdyn N : c$ implies $\sem{M}
+\ltsq{\sem{\Delta}}{U\li\sem{c}} \sem{N}$ where here $\sem{\Delta =
+  x_1:c_1,\ldots}$ is the action of the product on the relations
+$c_1,\ldots$.
 
-\eric{Not sure if this belongs here...}
-Squares are used to model the term precision ordering $\Gamma \ltdyn \Gamma'
-\vdash M \ltdyn N : A \ltdyn A'$ as an ordering relation between the monotone
-functions denoted by $M$ and $N$:
+%% % https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXEdhbW1hIl0sWzAsMSwiVVxcbGkgQSJdLFsxLDAsIlxcR2FtbWEnIl0sWzEsMSwiVVxcbGkgQSciXSxbMiwzLCJcXHNlbXtOfSJdLFswLDEsIlxcc2Vte019IiwyXSxbMCwyLCJcXEdhbW1hXntcXGx0ZHlufSIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEsMywiVVxcbGkgYyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+%% \[\begin{tikzcd}[ampersand replacement=\&]
+%% 	\Gamma \& {\Gamma'} \\
+%% 	{U\li A} \& {U\li A'}
+%% 	\arrow["{\Gamma^{\ltdyn}}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+%% 	\arrow["{\sem{M}}"', from=1-1, to=2-1]
+%% 	\arrow["{\sem{N}}", from=1-2, to=2-2]
+%% 	\arrow["{U\li c}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+%% \end{tikzcd}\]
 
-% https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXEdhbW1hIl0sWzAsMSwiVVxcbGkgQSJdLFsxLDAsIlxcR2FtbWEnIl0sWzEsMSwiVVxcbGkgQSciXSxbMiwzLCJcXHNlbXtOfSJdLFswLDEsIlxcc2Vte019IiwyXSxbMCwyLCJcXEdhbW1hXntcXGx0ZHlufSIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEsMywiVVxcbGkgYyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
-\[\begin{tikzcd}[ampersand replacement=\&]
-	\Gamma \& {\Gamma'} \\
-	{U\li A} \& {U\li A'}
-	\arrow["{\Gamma^{\ltdyn}}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-	\arrow["{\sem{M}}"', from=1-1, to=2-1]
-	\arrow["{\sem{N}}", from=1-2, to=2-2]
-	\arrow["{U\li c}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-\end{tikzcd}\]
-
-where $\Gamma = \{A_1, A_2, \dots\}$ and $\Gamma' = \{A_1' , A_2', \dots\}$ and for
-each $j$, $c_j : A_j \ltdyn A_j'$, and the relation $\Gamma^{\ltdyn}$ is simply
-the product of the $c_i$.
-
+\max{This needs to also talk about relational composition and horizontal pasting squares, since that's transitivity!}
 We note that for any $c$, there is a ``vertical identity'' square $\id
@@
 \ltsq{r(A_i)}{r(A_o)}$ by the definition of monotone function.
-We likewise make the analogous definition of square for error domains.
 
-Having defined squares, we now discuss the interpretation of the cast rules UpL
-and UpR \footnote{The cast rules DnL and DnR involve error domain relations and
-squares, which for the sake of this discussion are not relevant.}. These rules
-specify a relationship between the semantics of type precision and the
-corresponding casts (see Section \ref{sec:GTLC}). For example, consider the UpL
-rule. It states that if $c : A_1 \ltdyn A_2$ and $c' : A_2 \ltdyn A_3$ and $x :
-A_1$ is related to $y : A_3$ via the composition $c \comp c'$, then the upcast
-$\upc{c} x$ is related to $y$ via $c'$. This is equivalent to the existence of
-the following square:
+Now that we have an intended model of term precision in the form of
+squares, we can see what is required to give a compositional proof of
+graduality. Most of the cases of term precision are just congruence
+and follow easily. The core of the proof of the graduality property is
+in proving the validity of the \emph{cast rules} UpL, UpR, DnL and
+DnR.
+%
+These rules specify a relationship between the semantics of type
+precision derivations $c$ and the corresponding casts $\upc c, \dnc c$ (see Section
+\ref{sec:GTLC}).
+%
+To prove graduality compositionally then, we will need to codify in an
+extra property on $\sem{c}$ that the relation $\sem{c}$ has some
+correspondence with the casts $\sem{\upc c}, \sem{\dnc c}$.
+%
+What property should we require?
+
+Consider the UpL rule. It states that if
+$c : A_1 \ltdyn A_2$ and $c' : A_2 \ltdyn A_3$ and $M : A_1$ is
+related to $N : A_3$ via the composition $c \comp c'$, then the upcast
+$\upc{c} M$ is related to $N$ via $c'$. Since the upcasts are pure
+functions in the semantics this can be deduced from the existence of
+the following square:\max{the square doesn't make sense because we
+  haven't discussed horizontal composition of relations or that this
+  notation means horizontal composition of relations}
 %
@@
 \[\begin{tikzcd}[ampersand replacement=\&]
-  {A_1} \& {A_2} \& {A_3} \\
-  {A_2} \&\& {A_3}
-  \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-  \arrow["{\upc{c}}"', from=1-1, to=2-1]
-  \arrow["{c'}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
+  {\sem{A_1}} \& {\sem{A_2}} \& {\sem{A_3}} \\
+  {\sem{A_2}} \&\& {\sem{A_3}}
+  \arrow["\sem{c}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+  \arrow["{\sem{\upc{c}}}"', from=1-1, to=2-1]
+  \arrow["{\sem{c'}}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
   \arrow["\id", from=1-3, to=2-3]
-  \arrow["{c'}"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
+  \arrow["{\sem{c'}}"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
 \end{tikzcd}\]
-%
-Observe that although this is really a fact about the relationship between $c$
-and $\upc{c}$, we must universally quantify over all relations $c'$. This means
-we cannot require the existence of this square as part of the definition of a
-monotone relation, as it is self-referential. This would seem to imply that we
-cannot give a syntax-independent semantic model for gradual typing. However, the
-key observation is that through compositionality, we can \emph{derive} the above
-squares from simpler ones that do not involve composition of relations. These
-simpler squares have been used in prior work for giving a compositional
-semantics to gradual typing \cite{new-licata18}. Below is the square
+While at first look this seems to be specifying a relationship between
+$\sem{c}$ and $\sem{\upc c}$, there is a problem: it also quantifies
+involves an arbitrary other relation $\sem{c'}$! This means we cannot
+require the existence of this square as part of the definition of a
+relation between value types, as it is self-referential. This would
+seem to imply that we cannot give a compositional model for
+graduality. However, New and Licata observed that in the presence of
+transitivity, we can \emph{derive} the above squares from simpler ones
+that do not involve composition of relations. Below is the square
 corresponding to UpL:
-%
 % https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzIiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzIiXSxbMCwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJyKEFfMikiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxcdXBje2N9IiwyXSxbMiwzLCJcXGlkIl1d
@@
 %
-This square, and the corresponding one related to UpR, together express the
-\emph{representability} of $c$ by the upcast morphism $\upc{c}$, which
-intuitively says that $c$ is determined by the behavior of $\upc{c}$. We can use
-this simpler square to derive the original UpL square by horizontally composing
-with the identity square for $c'$:
+This square, with a dual one related to UpR, say that the relation $c$
+is \emph{representable} by the upcast morphism $\upc{c}$, essentially
+that the relation is a kind of \emph{graph} of the function in that $x
+\binrel{\sem c} y$ if and only if $\sem{upc c}(x) \leq_{A_2} y$.
+
+We can use this simpler square to derive the original UpL square as long as we can
+horizontally compose with the identity square for $c'$:
 %
@@
 \end{tikzcd}\]
-%
-The existence of these simpler squares for a relation can be stated without
-reference to other relations, and hence we can require that these squares exist
-in the definition of relation without running into the circularity issue we
-faced with the original rules. This allows us to formulate a notion of model
-that is independent of the particular syntax of any cast calculus.
-
-We now see why compositionality of squares, and hence transitivity of the
-relations on the objects that model types, is an essential property that we
-cannot sacrifice.
+So we see that in order to get a sensible condition that can give us a
+compositional definition of a well-behaved relation, some amount of
+transitive reasoning on squares is required, and so we cannot entirely
+drop posets from our semantics.
 
@@
 
-Given that compositional reasoning is essential, we seek to formulate an
-ordering relation on $\li A$ that is transitive. The theorem above
-tells us, however, that any such relation must either not be a congruence with
-respect to $\theta$, or must not be left- and right-step-insensitive.
-\footnote{Technically, we only need to give up step insensitivity on \emph{one}
-side, as the above theorem requires $R$ to be both left- and
-right-step-insensitive. We could define two separate ``step-semi-sensitive''
-relations on $\li A$, each closed under delays on the left and
-right respectively. The denotation of term precision would then involve the
-conjunction of those relations. We do not pursue this approach in this paper,
-because (1) it is not clear if the step-semi-sensitive relations would be
-transitive, and (2) we would need to reason about the two ordering relations
-separately for every pair of terms.}
+Given that some amount of transitive reasoning is essential for
+modeling type and term precision compositionally, we need to revisit
+our ordering relation on $\li A$ to get one that is
+transitive. Theorem\ref{thm:no-go} tells us, however, that any non-trivial such
+relation must either not be a congruence with respect to $\theta$, or
+must not be left- and right-step-insensitive.
+\footnote{Technically, the theorem only implies that we need to drop
+one of both left- and right-step-insensitivity. We might then attempt
+to work with two separate ``step-semi-sensitive'' relations on $\li
+A$, each closed under delays on the left and right respectively. This
+has some similarities to prior work on logical relations models
+\cite{newahmedlicata,newgiovanninilicata}, but we do not take this
+approach because the relations still do not seem to be provably
+transitive, though we have not been able to prove that they are not
+transitive either.\max{Eric check if this footnote is correct}}
 %
-We cannot forego the property of being a $\theta$-congruence, as without this we
-would not be able to prove properties of the relation using \lob-induction,
-e.g., that the monadic extension of a monotone function $f$ is a \emph{monotone}
-function with respect to the ordering.
-%and moreover this would mean that $\theta$ would not be a monotone function.
-
-Thus, we choose to sacrifice left and right step-insensitivity. That is, we will
-define an ordering relation that requires terms to be in ``lock-step''. In order
-for two computations to be related in this ordering, they must have the same
-stepping behavior (or the left-hand side must be an error). To deal with terms
-that take differing numbers of steps, we then define a separate relation called
-\emph{weak bisimilarity} that relates terms that are extensionally equal and may
-differ by a finite number of delays. Then, the semantics of the error ordering
-for the guarded lift monad will involve a combination of these two relations:
-a ``closure'' of the lock-step error ordering under weak bisimilarity on both
-sides.
-% Although the combined relation will not be transitive (for the same reason that
-% $\semltbad$ is not transitive), this...
-This decomposition has the advantage that we can recover some transitive
-reasoning and push much of the reasoning about stepping to the margins of the
-development.
+We cannot forego the property of being a $\theta$-congruence, as
+without this we would not be able to prove basic properties of the
+relation using \lob-induction, e.g., that the extension $f^\dagger$ is
+\emph{monotone}, which is used pervasively in the proof of graduality.
+%
+Thus, we choose to sacrifice left and right step-insensitivity. That
+is, we will define an ordering relation that requires terms to be in
+``lock-step''. In order for two computations to be related in this
+ordering, they must have the exact same stepping behavior
+(unless/until the left-hand side results in an error).
 
@@
 %
-We similarly define a heterogeneous version of this ordering between the lifts
-of two different types $A$ and $A'$, parameterized by a poset relation $c : A \rel A'$.
+More generally we can define a heterogeneous version of this ordering that lifts
+poset relation $c : A \rel A'$ to a poset relation $\li c : \li A \rel \li A'$.
 
-The second relation we define on the guarded lift monad is \emph{weak bisimilarity}.
-% For a type $X$, we define a relation on $\li X$, called ``weak bisimilarity",
-% written $l \bisim l'$. 
-Intuitively, we say $l \bisim l'$ if they are the same ``up to delays''.
+However, we also cannot completely drop step-\emph{insensitivity} from
+our model. The graduality property itself is not sensitive to steps,
+and in fact there are terms related by term precision that take
+differing numbers of steps.
 %
-Weak bisimilarity on $\li A$ is parameterized by an underlying relation $R$ on $A$
-and is defined by guarded fixpoint as shown in Figure \ref{fig:weak-bisimilarity}.
-
-\begin{figure}
-    \fbox{$l_1 \bisim l_2$}
-    \begin{align*}
-        &\mho \bisim \mho \\
+The offending step arises precisely from the place where our
+definition of $D$ uses a $\later$: the function case.
 %
-        &\eta\, x \bisim \eta\, y \text{ if } 
-          x \mathbin{R} y \\
-      %		
-        &\theta\, \tilde{x} \bisim \theta\, \tilde{y} \text{ if } 
-          \later_t (\tilde{x}_t \bisim \tilde{y}_t) \\
-      %	
-        &\theta\, \tilde{x} \bisim \mho \text{ if } 
-          \theta\, \tilde{x} = \delta^n(\mho) \text { for some $n$ } \\
-      %	
-        &\theta\, \tilde{x} \bisim \eta\, y \text{ if }
-          (\theta\, \tilde{x} = \delta^n(\eta\, x))
-        \text { for some $n$ and $x : X$ such that $x \mathbin{R} y$ } \\
-      %
-        &\mho \bisim \theta\, \tilde{y} \text { if } 
-          \theta\, \tilde{y} = \delta^n(\mho) \text { for some $n$ } \\
-      %	
-        &\eta\, x \bisim \theta\, \tilde{y} \text { if }
-          (\theta\, \tilde{y} = \delta^n (\eta\, y))
-        \text { for some $n$ and $y : X$ such that $x \mathbin{R} y$ }
-      \end{align*}
-    \caption{The weak bisimilarity relation.}
-    \label{fig:weak-bisimilarity}
-\end{figure}
-
-When both sides are $\eta$, then we ensure that the underlying values are
-bisimilar in the underlying bisimilarity relation on $A$. When one side is a
-$\theta$ and the other is $\eta x$ (i.e., one side steps), we stipulate that the
-$\theta$-term runs to $\eta y$ where $x$ is related to $y$. When one side is
-$\theta$ and the other $\mho$, we require that the $\theta$-term runs to $\mho$.
-If both sides step, then we allow one time step to pass and compare the
-resulting terms. This definition captures the intuition of two terms being
-related up to delays.
-
-It can be shown (by \lob-induction) that weak bisimilarity is reflexive and
-symmetric. We can also show that this relation is \emph{not} transitive. The
-argument is the same as that used to show that the step-insensitive error
-ordering $\semltbad$ described above is not transitive. In particular, we
-observe that weak bisimilarity is a congruence with respect to $\theta$, and we
-can show by $\lob$-induction that it is closed under delay on both sides. Thus,
-we may appeal to the no-go theorem as before.
-% We appeal to the no-go theorem given above
-
-Now we can define the denotation of term precision for closed terms of type
-$\nat$ as follows:
-%
-\[ l_1 \leq^{\bisim} l_2 := 
-  \Sigma_{l_1', l_2' \in \li \mathbb{N}}. l_1 \bisim l_1' \ltls l_2' \bisim l_2. \]
-
-Importantly, we note that as with the original ordering $\semltbad$ defined at
-the beginning of this section, the relation just defined is not transitive
-(again as a result of the above no-go theorem). It may therefore seem that we
-have not solved the original issue we faced. In a sense, this is true. The model
-of gradual typing that we end up with will not support \emph{extensional}
-compositional reasoning. However, by decomposing the denotation of term
-precision in the above manner, we can employ \emph{intensional} compositional
-reasoning by working with the lock-step error ordering. Thus, the tentative plan
-going forward will be to carry out the proofs compositionally using the
-lock-step ordering. We will then apply the closure under weak bisimilarity in
-the denotation of term precision given above to handle the aspects involving
-stepping.
-
-% ...to carry out the proofs using the lock-step ordering and handle the stepping
-% behavior separately via weak bisimilarity. In the end we combine everything
-% together using the above denotation for term precision
-
-\subsection{Adjusting the Cast Rules}\label{sec:adjusting-cast-rules}
-
-\eric{We may want to introduce error domain relations and squares at this point,
-since the squares for DnL and DnR involve lift.}
-
-We have seen how we can recover transitive reasoning by working with the
-lock-step error ordering. However, it turns out that we cannot completely
-isolate the reasoning about steps from the reasoning about the ordering.
-% Unfortunately, by defining the partial order on the guarded lift monad to be
-% the lock-step error ordering, we introduce a new issue.
-In particular, the cast rules required for proving graduality no longer hold if
-interpreted using the lock-step ordering. The source of the problem is that the
-downcast from the dynamic type to a function takes a step, i.e., involves a
-$\theta$, and the lock-step ordering requires both sides to be in sync. More
-specifically, consider the DnL rule for $\iarr \colon \dyntodyn \ltdyn\, \dyn$
-shown below:
-%
-\begin{mathpar}
-  \inferrule*{M \ltdyn N : \iarr}
-             {\dnc{\iarr}{M} \ltdyn N : c}
-\end{mathpar}
-%
-This rule corresponds to the following square:
-
+Consider the representability condition corresponding to the DnL rule
+for $\iarr \colon \dyntodyn \ltdyn\, \dyn$.
 % https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXGxpIEQiXSxbMCwxLCJcXGxpIFUoRCBcXGFyciBcXGxpIEQpIl0sWzEsMCwiXFxsaSBEIl0sWzEsMSwiXFxsaSBEIl0sWzAsMiwiXFxsZXFfe1xcbGkgRH0iLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxLDMsIlxcbGlcXGlhcnIiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxcZG5je1xcaWFycn0iLDJdLFsyLDMsIlxcaWQiXV0=
@@
 	\arrow["{\leq_{\li D}}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-	\arrow["{\dnc{\iarr}}"', from=1-1, to=2-1]
+	\arrow["{\sem{\dnc{\iarr}}}"', from=1-1, to=2-1]
 	\arrow["\id", from=1-2, to=2-2]
-	\arrow["\li\iarr"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+	\arrow["\li\sem{\iarr}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
 \end{tikzcd}\]
+For this square to be valid with the lock-step ordering the left and
+right hand side would need to take the same number of steps, at least
+when the left-hand side does not error.
 %
-where $\li \iarr$ is the heterogeneous version of the lock-step error ordering
-discussed above applied to the relation $\iarr$. We note that the precise
-definition of $\iarr$ as a poset relation is not important for this discussion.
+However the right hand side \emph{always} takes $0$ steps, as it is
+the identity function, whereas on the left, our definition of
+$\sem{\dnc{\iarr}}$ had to take an observable step when its input is a
+function: this was inherent to the fact that the function case of the
+dynamic type is guarded by a later.
+
+However, observe that we can remedy this particular situation by
+replacing the $\id$ on the right hand side by an innocuous function
+$\delta^* = (\delta \circ \eta)^\dagger$ that on an input value takes
+a single computational step, but is otherwise the identity
+function. If we were to ignore computational steps, then this function
+\emph{would} be the identity function. We call such a function that is
+the identity function except for the introduction of computational
+steps a \emph{perturbation}.
+
+We formalize this property of being equivalent ``except for steps''
+with a second relation on the free error domain $\li A$: \emph{weak
+bisimilarity}, defined in Figure~\ref{fig:weak-bisim}, which is
+parameterized by a binary relation $\bisim_A$ on $A$
+\cite{guarded-stuff-that-uses-bisim}.
+% For a type $X$, we define a relation on $\li X$, called ``weak bisimilarity",
+% written $l \bisim l'$. 
+Two errors are bisimilar, and when both sides are $\eta$, we ensure
+that the underlying values are bisimilar in the underlying
+bisimilarity relation on $A$. When both sides are thinking, we ensure
+the terms are bisimilar later.  Most importantly, when one side is
+thinking but the other terminates at $\eta x$ (i.e., one side steps),
+we stipulate that the $\theta$-term runs to $\eta y$ where $x$ is
+related to $y$. And similarly, if one side is thinking and the other
+errors, we ensure the thinking side eventually errors.
+\begin{figure}
+    \fbox{$l_1 \bisim l_2$}
+    \begin{align*}
+        \mho \bisim \mho &\text{ iff } \top \\
       %
-We now recall the definition of the semantics of a downcast given in Section
-\ref{sec:term-interpretation}. The downcast on the left will insert a $\theta$
-in the case where the value of type $D$ is a later-function $\tilde{f}$.
+        \eta\, x \bisim \eta\, y &\text{ iff } x \bisim_A y \\
       %		
-Thus, in order for the downcast to be related to the RHS in the lock-step error
-ordering on $\li (D \to \li D)$, the RHS must be of the form $\theta(\dots)$,
-and moreover, after one time step, the argument of $\theta$ on the LHS must be
-related to the argument of $\theta$ on the RHS. As it stands now, this need not
-be the case, e.g., if the RHS is of the form $\eta(\dots)$. Thus, we conclude
-that the DnL rule does not in general hold under the lock-step error ordering.
-However, observe that we can remedy this particular situation by inserting a
-``delay'' on the right, i.e., wrapping the RHS in $\delta^* = (\delta \circ
-\eta)^\dagger$ where $\delta = \theta \circ \nxt$. Notice that this delay
-function is weakly bisimilar to the identity function in that $\delta^*\, x
-\bisim x$ for all $x$. \footnote{This follows from the fact that $\delta$ is
-weakly bisimilar to the identity ($\delta\, x \bisim x$ for all $x$) and that
-$-^\dagger$ preserves weak bisimilarity.} Thus, it will disappear when we apply
-the bisimilarity-closure construction described above. That is, this delay makes
-no difference in the extensional setting, but its presence is crucial in the
-intensional setting.
+        \theta\, \tilde{x} \bisim \theta\, \tilde{y} &\text{ iff } \later_t (\tilde{x}_t \bisim \tilde{y}_t) \\
+      %	
+        \theta\, \tilde{x} \bisim \mho &\text{ iff } \exists n. \theta\, \tilde{x} = \delta^n(\mho)\\
+      %	
+        \theta\, \tilde{x} \bisim \eta\, y &\text{ iff } \exists n. \exists x \bisim_A y.
+          (\theta\, \tilde{x} = \delta^n(\eta\, x))\\
+      %
+        \mho \bisim \theta\, \tilde{y} &\text{ iff } \exists n. \theta\, \tilde{y} = \delta^n(\mho) \\
+      %	
+        \eta\, x \bisim \theta\, \tilde{y} &\text { iff } \exists n. \exists y \bisim_A x. (\theta\, \tilde{y} = \delta^n (\eta\, y))
+      \end{align*}
+    \caption{The weak bisimilarity relation on the free error domain}
+    \label{fig:weak-bisimilarity}
+\end{figure}
 
-\subsection{Extending the Model to Higher-Order Types}
+It can be shown (by \lob-induction) that weak bisimilarity is
+reflexive and symmetric. Since it is non-trivial, a $\theta$
+congruence and step-insensitive on both sides, by
+Theorem~\ref{thm:no-go}, we also know that it is \emph{not}
+transitive.
+%
+We will then require our denotations of types to be not just posets,
+but posets additionally equipped with a reflexive symmetric relation
+$\bisim_A$.
 
-% At this point, we could carry out the full construction of a concrete relational
-% model with these additional features.
+Then we can refine our denotation of term precision $\Delta \vdash M
+\ltdyn N : c$ to mean not that $\sem{M}$ and $\sem{N}$ are necessarily
+in a lock-step error ordering directly, but that they can be
+``synchronized'' to do so, that is that there exist $f \bisim \sem{M}$
+and $g \bisim \sem{N}$ such that $f \ltsq{\sem{\Delta}}{U\li\sem{c}}
+g$.
+%
+Note that this in turn implies that $\sem{M}$ and $\sem{N}$ are
+related in the original \emph{step-insensitive} error ordering that we
+sought to prove!
+%
+We additionally weaken our representability squares so that our
+upcasts and downcasts are not required to be in a lock-step ordering
+with an identity but instead to be in a lock-step ordering with some
+perturbation, that is a function \emph{bisimilar to} the identity. We
+call this weakened form of representability
+\emph{quasi}-representability.
+%
+Then if our relations are quasi-representable we can in fact prove the
+validity of our cast rules.
+%
+The remainder of the proof of graduality then is to show that this
+property of being quasi-representable is itself compositional: that
+all of the constructions we have on type precision preserve the
+property of being quasi-representable.
+%
+While this is true for representability, it is not quite true for
+quasi-representability. The reason is that perturbations are not quite
+as well behaved as \emph{actual} identity functions. To solve this
+final issue, we attach one final piece of information to our types
+$A$: a type of \emph{syntactic} perturbations that represent functions
+bisimilar to the identity, but are presented as data so that we can
+perform operations such as composition and functorial actions on them.
+%
+With this notion of syntactic perturbation, and additional
+requirements that the relations interact well with perturbations, we
+finally achieve our desired result: a compositional,
+syntax-independent proof of graduality.
 
-The above techniques give a semantic interpretation of term precision for closed
-terms of base type. We now consider how to extend these constructions to
-potentially-open terms of \emph{all} types. In particular, in order to model
-higher-order data types (i.e., functions) we need to equip \emph{every} semantic
-object with not only a partial ordering relation but also a ``bisimilarity''
-relation that is reflexive and symmetric. We must similarly equip every object
-with a structured set of delays that can be inserted and manipulated to ensure
-the appropriate cast rules hold at all higher-order types. For example, the
-upcast for a derivation $c_i \ra c_o$ involves the downcast corresponding to
-$c_i$ and the upcast corresponding to $c_o$. We must be able to insert delays in
-a functorial manner to mimic the structure of the casts. In the next section, we
-will make the necessary definitions and describe the relevant constructions.
+%% Importantly, we note that as with the original ordering $\semltbad$ defined at
+%% the beginning of this section, the relation just defined is not transitive
+%% (again as a result of the above no-go theorem). It may therefore seem that we
+%% have not solved the original issue we faced. In a sense, this is true. The model
+%% of gradual typing that we end up with will not support \emph{extensional}
+%% compositional reasoning. However, by decomposing the denotation of term
+%% precision in the above manner, we can employ \emph{intensional} compositional
+%% reasoning by working with the lock-step error ordering. Thus, the tentative plan
+%% going forward will be to carry out the proofs compositionally using the
+%% lock-step ordering. We will then apply the closure under weak bisimilarity in
+%% the denotation of term precision given above to handle the aspects involving
+%% stepping.
+
+%% Notice that this delay
+%% function is weakly bisimilar to the identity function in that $\delta^*\, x
+%% \bisim x$ for all $x$. \footnote{This follows from the fact that $\delta$ is
+%% weakly bisimilar to the identity ($\delta\, x \bisim x$ for all $x$) and that
+%% $-^\dagger$ preserves weak bisimilarity.} Thus, it will disappear when we apply
+%% the bisimilarity-closure construction described above. That is, this delay makes
+%% no difference in the extensional setting, but its presence is crucial in the
+%% intensional setting.
+
+%% The precise definition of $\iarr$ as a poset relation is not important
+%% for this discussion.
+%% %
+
+%% \eric{We may want to introduce error domain relations and squares at this point,
+%% since the squares for DnL and DnR involve lift.}
+%
+%% We now recall the definition of the semantics of a downcast given in Section
+%% \ref{sec:term-interpretation}. The downcast on the left will insert a $\theta$
+%% in the case where the value of type $D$ is a later-function $\tilde{f}$.
+%% %
+%% Thus, in order for the downcast to be related to the RHS in the lock-step error
+%% ordering on $\li (D \to \li D)$, the RHS must be of the form $\theta(\dots)$,
+%% and moreover, after one time step, the argument of $\theta$ on the LHS must be
+%% related to the argument of $\theta$ on the RHS. As it stands now, this need not
+%% be the case, e.g., if the RHS is of the form $\eta(\dots)$. Thus, we conclude
+%% that the DnL rule does not in general hold under the lock-step error ordering.
+
+%% \max{resume here}
+%% To deal with
+%% terms that take differing numbers of steps, we then define a separate
+%% relation called \emph{weak bisimilarity} that relates terms that are
+%% extensionally equal and may differ by a finite number of delays. Then,
+%% the semantics of the error ordering for the guarded lift monad will
+%% involve a combination of these two relations: a ``closure'' of the
+%% lock-step error ordering under weak bisimilarity on both sides.
+%% % Although the combined relation will not be transitive (for the same reason that
+%% % $\semltbad$ is not transitive), this...
+%% This decomposition has the advantage that we can recover some transitive
+%% reasoning and push much of the reasoning about stepping to the margins of the
+%% development.
+
+%% % ...to carry out the proofs using the lock-step ordering and handle the stepping
+%% % behavior separately via weak bisimilarity. In the end we combine everything
+%% % together using the above denotation for term precision
+
+%% \subsection{Extending the Model to Higher-Order Types}
+%% \max{this subsection doesn't really say anything concrete. If we want to talk about perturbations, actually talk about them}
+
+%% % At this point, we could carry out the full construction of a concrete relational
+%% % model with these additional features.
+
+%% The above techniques give a semantic interpretation of term precision for closed
+%% terms of base type. We now consider how to extend these constructions to
+%% potentially-open terms of \emph{all} types. In particular, in order to model
+%% higher-order data types (i.e., functions) we need to equip \emph{every} semantic
+%% object with not only a partial ordering relation but also a ``bisimilarity''
+%% relation that is reflexive and symmetric. We must similarly equip every object
+%% with a structured set of delays that can be inserted and manipulated to ensure
+%% the appropriate cast rules hold at all higher-order types. For example, the
+%% upcast for a derivation $c_i \ra c_o$ involves the downcast corresponding to
+%% $c_i$ and the upcast corresponding to $c_o$. We must be able to insert delays in
+%% a functorial manner to mimic the structure of the casts. In the next section, we
+%% will make the necessary definitions and describe the relevant constructions.
 

### 2024-07-11 paper-new/concrete-relational-model.tex
revisions to the relational motivation section
revise sections on relational model

@@
-\section{Constructing a Concrete Model}\label{sec:concrete-model}
+\section{A Guarded Model of Graduality}\label{sec:concrete-relational-model}
 
-In this section, we complete the construction of a relational model of gradual
-typing. To manage the complexity, we carry out the construction in phases,
-beginning with a relational model consisting of partially-ordered sets with a
-notion of bisimilarity and then adding requirements for a structured set of
-delays on each type as well as a notion of representability for relations that
-models the cast rules in the intensional setting.
+In this section, we describe in detail our model of gradual typing. To manage
+the complexity, we carry out the construction in two phases, beginning with a
+relational model consisting of partially-ordered sets with a notion of
+bisimilarity and then adding requirements for a structured set of syntactic
+perturbations on each type as well as the notion of quasi-representability for
+relations that models the cast rules. We then discuss the construction of the
+dynamic type and the refined denotation of term precision that relates terms
+that can be synchronized to be in lock-step. We conclude by demonstrating the
+adequacy of our model with respect to the graduality property.
 
@@
 relation on the guarded lift monad. Recall that weak bisimilarity on $\li A$ is
-parameterized by a relation on $A$. Thus, in order to define a compositional
-model for all types, we need to equip the objects (i.e. the posets and error
-domains) with a reflexive, symmetric bisimilarity relation. This leads to our
-first new definition:
+parameterized by a relation on $A$. To define a compositional model for all
+types, we equip the posets and error domains with a reflexive, symmetric
+bisimilarity relation. This leads to our first new definition:
 
@@
 
-Next we define a notion of \emph{error domain relation} between error domains $B$ and
-$B'$. An error domain relation $d : B \rel B'$ is a relation on the underlying
-posets that preserves error and respects $\theta$. More formally:
+Next we define an analogous notion of \emph{relation} between error domains $B$
+and $B'$. An error domain relation $d : B \rel B'$ is a relation on the
+underlying posets that preserves error and respects $\theta$. More formally:
 %
@@
 \end{definition}
-% Likewise, the definition of error domain relations is unchanged.
+%
+As with predomain relations, we can compose error domain relations. However,
+because of a technical restriction involving the interaction between
+propositional truncation and the later modality, we cannot define the
+composition of error domain relations as the composition of their underlying
+predomain relations. Instead, we define composition of error domain relations
+$d$ on $B_1$ and $B_2$ and $d'$ on $B_2$ and $B_3$ inductively to be the least
+relation containing $d$ and $d'$ that is downward closed under $\le_{B_1}$,
+upward-closed under $\le_{B_3}$, respects error, and preserves $\theta$.
+Specifically, it is defined inductively by the following rules:
+% %
+% \begin{mathpar}
+%     \inferrule*[right = Comp]
+%     {b_1 \mathbin{d} b_2 \and b_2 \mathbin{d'} b_3}
+%     {b_1 \mathbin{d \relcomp d'} b_3}
+
+%     \inferrule*[right = DnClosed]
+%     {b_1' \le_{B_1} b_1 \and b_1 \mathbin{d \relcomp d'} b_3}
+%     {b_1' \mathbin{d \relcomp d'} b_3}
+
+%     \inferrule*[right = UpClosed]
+%     {b_1 \mathbin{d \relcomp d'} b_3 \and b_3 \le_{B_3} b_3'}
+%     {b_1 \mathbin{d \relcomp d'} b_3'}
+
+%     \inferrule*[right = PresErr]
+%     { }
+%     {\mho_{B_1} \mathbin{d \relcomp d'} b_3}
+
+%     \inferrule*[right = PresTheta]
+%     {\later_t( \tilde{b_1} \mathbin{d \relcomp d'} \tilde{b_3} ) }
+%     {\theta_{B_1}(\tilde{b_1}) \mathbin{d \relcomp d'} \theta_{B_3}(\tilde{b_3}) }
+% \end{mathpar}
+% %
 
 The definitions of squares for predomains are also unchanged from the previous
-section, except that the morphisms on the left and right, are now morphisms of
+section, except that the morphisms on the left and right are now morphisms of
 predomains. We likewise define a notion of \textbf{error domain square}: Let
@@
 
-\begin{comment}
-We define a (monotone) relation on predomains $A$ and $A'$ to be a relation on
-the underlying sets that is downward-closed under $\le_A$ and upward-closed
-under $\le_{A'}$. More formally:
-%
-\begin{definition}
-Let $A$ and $A'$ be predomains. A \emph{predomain relation} between $A$ and $A'$
-is a relation $R$ between the underlying sets such that:
-\begin{enumerate}
-    \item (Downward closure): For all $x_1, x_2 \in A$ and $y \in A'$,
-    if $x_1 \le_A x_2$ and $x_2 \mathbin{R} y$, then $x_1 \mathbin{R} y$.
-    \item (Upward closure): For all $x \in A$ and $y_1, y_2 \in A'$,
-    if $x \mathbin{R} y_1$ and $y_1 \le_{A'} y_2$, then $x \mathbin{R} y_2$.
-\end{enumerate}
-\end{definition}
-%
-Composition of relations on predomains is the usual relational composition
-(which is truncated to be propositional). Similarly, we define a (monotone)
-relation on error domains to be a relation on the underlying predomains that
-respects error and preserves $\theta$:
-%
-\begin{definition}
-    Let $B$ and $B'$ be error domains. An \emph{error domain relation} between
-    $B$ and $B'$ is a relation $R$ between the underlying predomains such that
-    \begin{enumerate}
-       \item (Respects error): For all $y \in B'$, we have $\mho_B \mathbin{R} y$.
-       \item (Preserves $\theta$): For all $\tilde{x}$ in $\later B$ and $\tilde{y} \in \later B'$,
-       if $\later_t( \tilde{x}_t \mathbin{R} \tilde{y}_t )$ then
-       $\theta_B(\tilde{x}) \mathbin{R} \theta_{B'}(\tilde{y})$.
-    \end{enumerate}
-\end{definition}
-%
-Because of a technical restriction involving the interaction between
-propositional truncation and the later modality, we cannot define the
-composition of error domain relations as simply the composition of their
-underlying predomain relations. Instead, we define composition of error domain
-relations $d$ on $B_1$ and $B_2$ and $d'$ on $B_2$ and $B_3$ to be the least
-relation containing $d$ and $d'$ that is downward closed under $\le_{B_1}$,
-upward-closed under $\le_{B_3}$, respects error, and preserves $\theta$.
-Specifically, it is defined inductively by the following rules:
-%
-\begin{mathpar}
-    \inferrule*[right = Comp]
-    {b_1 \mathbin{d} b_2 \and b_2 \mathbin{d'} b_3}
-    {b_1 \mathbin{d \relcomp d'} b_3}
-
-    \inferrule*[right = DnClosed]
-    {b_1' \le_{B_1} b_1 \and b_1 \mathbin{d \relcomp d'} b_3}
-    {b_1' \mathbin{d \relcomp d'} b_3}
-
-    \inferrule*[right = UpClosed]
-    {b_1 \mathbin{d \relcomp d'} b_3 \and b_3 \le_{B_3} b_3'}
-    {b_1 \mathbin{d \relcomp d'} b_3'}
-
-    \inferrule*[right = PresErr]
-    { }
-    {\mho_{B_1} \mathbin{d \relcomp d'} b_3}
-
-    \inferrule*[right = PresTheta]
-    {\later_t( \tilde{b_1} \mathbin{d \relcomp d'} \tilde{b_3} ) }
-    {\theta_{B_1}(\tilde{b_1}) \mathbin{d \relcomp d'} \theta_{B_3}(\tilde{b_3}) }
-\end{mathpar}
-%
-% We note that this composition has the following universal property.
-%
-
-We now describe the squares. Suppose we are given predomains $A_i, A_o, A_i'$,
-and $A_o'$, relations $c_i : A_i \rel A_i'$ and $c_o : A_o \rel A_o'$, and
-morphisms $f : A_i \to A_o, f' : A_i' \to A_o'$ Given a square with these
-morphisms and relations, we say that the square commutes, written $f \le f'$, if
-for all $x \in A_i$ and $x' \in A_i'$ with $(x, x') \in c_i$, we have $(f(x),
-f'(x')) \in c_o$. We make the analogous definition for error domains.
-
-\end{comment}
-
 \subsubsection{Functors}\label{sec:free-error-domain}
@@
 extend to predomains, error domains, and their relations. The functor $U$ simply
-takes the underlying poset relation of an error domain relation. Given a
+takes the underlying predomain relation of an error domain relation. Given a
 relation $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$, the relation $c_1
@@
 of predomains, i.e., they are monotone and preserve bisimilarity.
-
+%
 We observe that this predomain structure extends to an error domain structure by
@@
 the required $\theta$ map is the constructor $\theta$. Lastly, it can be shown
-that the delay morphism $\delta = \theta \circ \nxt$ satisfies $x \bisim
+that the delay morphism $\delta = \theta \circ \nxt : UB \to UB$ satisfies $x \bisim
 \delta\, x$ for all $x$ as is required by the definition of error domain.
@@
 The monadic bind operation $-^\dagger$ takes a predomain morphism $f : A \to UB$
-to an error domain morphism $\li A \to B$; with this we define the action of $\li$ on
-morphisms: Given $f : A \to A'$ we observe that $\eta \circ f : A \to U\li A'$, so
-we define $\li f = (\eta \circ f)^\dagger : \li A \to \li A'$.
+to an error domain morphism $\li A \multimap B$; with this we define the action of $\li$ on
+morphisms: Given $f : A \to A'$ we define $\li f = (\eta \circ f)^\dagger : \li A \to \li A'$.
 
-Finally, we define the action of $\li$ on a relation $c$ to be the heterogeneous
-lock-step error ordering between $U\li A$ and $U\li A'$ and note that it by
-definition satisfies the additional requirements of being an error domain
-relation between the error domains $\li A$ and $\li A'$.
-
-% The monad $\li$ decomposes into a free-forgetful adjunction $F \dashv U$. Given
-% a predomain $A$, we define the error domain $FA$ whose underlying predomain is
-% $\li A$, whose error element is $\mho$, and whose $\theta$ map is the
-% constructor $\theta$.
-
-% It is easily verified that $\li A$ is the free error- and later-algebra on the
-% predomain $A$, so we have that $\li$ is left-adjoint to $U$.
-
-% i.e., error domain morphisms from $\li A$ to $B$ are in one-to-one correspondence with
-% predomain morphisms from $A$ to $UB$.
-
-% We define $\delta : U(\li A) \To U(\li A)$ by $\delta(x) = \theta(\nxt x)$.
-% We define $\hat{\delta} : \li A \arr \li A$ by $\hat{\delta} = \ext{(\delta \circ \eta)}{}$.
-% Note that by definition of $\text{ext}$, we have that $\hat{\delta}$ is a morphism of error domains.
-
-%\subsubsection{Lock-Step Error Ordering}\label{sec:lock-step}
-
-% We define the action of $\li$ on a relation $R$ between $A$ and $A'$ to be the
-% ``heterogeneous" version of the lock-step error ordering.
-
-% TODO action of \li on commuting squares
-
-\begin{comment}
-
-% internal hom for predomains and error domains
-Given predomains $A$ and $A'$, we can form the predomain of
-predomain morphisms from $A$ to $A'$, denoted $A \To A'$.
-\begin{itemize}
-    % Should we give the definition involving x and x'?
-    \item The ordering is defined by $f \le_{A \To A'} f'$ iff for all
-    $x \in A$, we have $f(x) \le_{A'} f'(x)$.
-    \item The bisimilarity relation is defined by $f \bisim_{A \To A'} f'$ iff
-    for all $x, x' \in A$ with $x \bisim_{A} x'$, we have $f(x) \bisim_{A'} f'(x')$. 
-\end{itemize}
-%
-Given $f : A_1' \to A_1$ and $g : A_2 \to A_2'$ we define the predomain morphism
-$f \To g : (A_1 \To A_2) \to (A_1' \To A_2')$ by $\lambda h. \lambda x'. g(h(f(x')))$.
-
-% TODO: include this?
-% The monadic extension operation $\ext{\cdot}{} : (A_1 \To U (\li A_2)) \To (\li A_1 \To U(\li A_2))$
-% is a morphism of predomains from $A_1 \To U(\li A_2)$ to $U(\li A_1) \To U(\li A_2)$, i.e.,
-% it preserves the ordering and bisimilarity relations.
-
-% Given a predomain $A$ and error domain $B$, we define $A \arr B := A \To UB$.
-We note that $A \To UB$ carries a natural error domain structure
-(in the below, the lambda is a meta-theoretic notation):
-\begin{itemize}
-    \item The error is given by $\lambda x . \mho_B$
-    \item The $\theta$ operation is defined by
-      \[ \theta_{A \To UB}(\tilde{f}) = \lambda x . \theta_B(\lambda t . \tilde{f}_t(x)). \]
-\end{itemize}
-
-Given a predomain $A$ and error domain $B$, we define
-$A \arr B$ to be the error domain such that $U(A \arr B) = A \To UB$,
-and whose error and $\theta$ operations are as defined above.
-We can define the functorial action of $\arr$ on morphisms
-$f \arr \phi$ in the obvious way.
-%
-It is easily verified that $A \arr B$ is an exponential of $UB$ by $A$
-in the category of predomains and their morphisms.
-
-\end{comment}
-
-\begin{comment}
-Lastly, given a relation of predomains $c$ between $A$ and $A'$, and a relation
-of error domains $d$ between $B$ and $B'$, we define the relation $c \arr d$
-between $A \arr B$ and $A' \arr B'$ in the obvious way, i.e., $f \in A \arr B$
-is related to $g \in A' \arr B'$ iff for all $x \in A$ and $x' \in A'$ with
-$(x, x') \in c$, we have $(f(x), g(x')) \in Ud$.
-%
-One can verify that this relation is indeed a relation of error domains
-in that it respects error and preserves $\theta$.
-\end{comment}
+Finally, we define the action of $\li$ on a predomain relation $c$ to be the
+heterogeneous lock-step error ordering between $U\li A$ and $U\li A'$ and note
+that it by definition satisfies the additional requirements of being an error
+domain relation between the error domains $\li A$ and $\li A'$.
 
@@
 We now introduce the additional definitions needed to complete the construction
-of our model. Recall from Section \ref{sec:adjusting-cast-rules} that in order
+of our model. Recall from Section \ref{sec:lock-step-and-weak-bisim} that in order
 to establish the DnL rule for the downcast of $\iarr$, we needed to adjust the
@@
 \emph{syntactic perturbations}, as well as a means of \emph{interpreting} these
-perturbations as actual endomorphisms on $A$. Moreover, we want the resulting
-endomorphisms to be extensionally equivalent to the identity function, a
-condition we express by saying that they are weakly bisimilar to the identity
-morphism on $A$. More formally, for every $A$ we require a homomorphism of
-monoids $i_A : M_A \to \morbisimid{A} := \{ f : A \to A \mid f \bisim \id_A \}$.
-We likewise equip every error domain $B$ with a monoid $M_B$ and interpretation
-homomorphism $i_B : M_B \to \morbisimid{B}$. These concepts deserve their own
-definitions: We call the triple $(A, M_A, iA)$ a \textbf{value object} as it
-will be the denotation of value types in our final model. Likewise, the triple
-$(B, M_B, i_B)$ will be called a \textbf{computation object}.
+syntactic perturbations as \emph{semantic} perturbations on $A$, i.e., as
+endomorphisms that are bisimilar to the identity.
+% Moreover, we want the
+% resulting endomorphisms to be extensionally equivalent to the identity function,
+% a condition we express by saying that they are weakly bisimilar to the identity
+% morphism on $A$. 
+More formally, for every $A$ we require a homomorphism of monoids $i_A : M_A \to
+\morbisimid{A} := \{ f : A \to A \mid f \bisim \id_A \}$. We likewise equip
+every error domain $B$ with a monoid $M_B$ and interpretation homomorphism $i_B
+: M_B \to \morbisimid{B}$. This grouping of data deserves its own definition: We
+call the triple $(A, M_A, iA)$ a \textbf{value object} as it will be the
+denotation of value types in our final model. Likewise, the triple $(B, M_B,
+i_B)$ will be called a \textbf{computation object}.
 
-We must now impose an additional condition on predomain relations $c$ and error
+We must impose an additional condition on predomain relations $c$ and error
 domain relations $d$ that specify how they interact with these perturbations. In
@@
 perturbation on $A_2$ to one on $A_3$. That is, given a relation $c : A \rel A'$
-and a perturbation $m_A \in M_A$, we need to be able to turn it into a
-perturbation $m_{A'} \in M_{A'}$ such that the resulting endomorphisms obtained
-by applying the respective homomorphisms $i_A$ and $i_{A'}$ to the perturbations
-form a square. Dually, we will need to ``pull'' a perturbation from right to
-left. This notion is made formal by the following definition:
+and a syntactic perturbation $m_A \in M_A$, we need to be able to turn it into a
+syntactic perturbation $m_{A'} \in M_{A'}$ such that the resulting semantic
+perturbations obtained by applying the respective homomorphisms $i_A$ and
+$i_{A'}$ form a square. Dually, we will need to ``pull'' a perturbation from
+right to left. This notion is made formal by the following definition:
 %
@@
     : A \rel A'$ be a predomain relation. A \textbf{push-pull structure} for $c$
-    consists of monoid homomorphisms $\push : M_A \to M_A'$ and $\pull : M_A'
-    \to M_A$ such that for all $m_A \in M_A$ there is a square $i_A(m_A)
-    \ltsq{c}{c} i_{A'}(\push\, m_A)$ and for all $m_{A'} \in M_{A'}$ there is a
-    square $i_A(\pull\, m_{A'}) \ltsq{c}{c} i_{A'}(m_{A'})$.
+    (with respect to $M_A$ and $M_{A'}$) consists of monoid homomorphisms $\push
+    : M_A \to M_A'$ and $\pull : M_A' \to M_A$ such that for all $m_A \in M_A$
+    there is a square $i_A(m_A) \ltsq{c}{c} i_{A'}(\push\, m_A)$ and for all
+    $m_{A'} \in M_{A'}$ there is a square $i_A(\pull\, m_{A'}) \ltsq{c}{c}
+    i_{A'}(m_{A'})$.
 \end{definition}
@@
 
-With the notion of perturbation defined, we can now discuss the final aspect of
-the model, namely the representability of relations.
+With the notion of syntactic perturbation defined, we can now make formal the
+final aspect of the model: the notion of quasi-represntability of relations.
 %
@@
 are perturbations $\delle_c \in M_A$ and $\delre_c \in M_{A'}$ such that the
-following two squares exist: (1) UpL: $e \ltsq{c}{r(A')} i_{A'}(\delre_c)$, and
-(2) UpR: $i_A(\delle_c) \ltsq{r(A)}{c} e$.
+following two squares exist: (1) $\upl$: $e \ltsq{c}{\le_{A'}} i_{A'}(\delre_c)$, and
+(2) $\upr$: $i_A(\delle_c) \ltsq{\le_{A}}{c} e$.
 \end{definition}
 %
-Observe that this generalizes the previous notion of representability, since
-under that definition the perturbations were always the identity.
+Observe that this weakens the previous notion of representability, since under
+that definition the perturbations were required to be the identity.
 %
@@
 relations $d$.
-%
+
 Likewise, we define quasi-right-representability as follows:
@@
 $d : B \rel B'$ be an error domain relation. We say that $d$ is
-\emph{quasi-right-representable} by an error domain morphism $p : B' \to B$ if
+\emph{quasi-right-representable} by an error domain morphism $p : B' \multimap B$ if
 there are perturbations $\dellp_d \in M_B$ and $\delrp_d \in M_{B'}$ such that
 the following two squares exist: 
-(1) DnL: $p \ltsq{r(B')}{d} i_{B'}(\delrp_d)$ and
-(2) DnR: $i_B(\dellp_d) \ltsq{d}{r(B)} p$.
+(1) $\dnl$: $p \ltsq{\le_{B'}}{d} i_{B'}(\delrp_d)$ and
+(2) $\dnr$: $i_B(\dellp_d) \ltsq{d}{\le_{B}} p$.
 \end{definition}
@@
 
-Although it makes sense to refer to the quasi-representability of an arbitrary
-predomain or error domain relation, quasi-representable relations do not in
-general compose. In order to construct the needed squares for the composition of
-two relations, we require that the relations have push-pull structures.
+Given $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$, it makes sense to refer to the
+quasi-representability of a predomain relation $c : A \rel A'$. However,
+quasi-representable relations do not in general compose: in order to construct
+the needed squares for the composition of two relations, we require that the
+relations have push-pull structures.
 
@@
 
-The reason for requiring $\li c$ to be quasi-right-representable in the definition
-of value relations is so that we can define the action of the functor $\li$ on
-relations. Similarly, we require $Ud$ to be quasi-left-representable in the
-definition of computation relations so that we can define the action of $U$ on
-relations.
+The reason for requiring $\li c$ to be quasi-right-representable in the
+definition of value relations is so that we can define the action of the functor
+$\li$ on relations, taking a value relation to a computation relation. Likewise,
+we require $Ud$ to be quasi-left-representable in the definition of computation
+relations so that we can define the action of $U$ on relations.
 
@@
 We first discuss the action of $\li$ on objects. Given a value object $(A, M_A,
-i_A)$ we define $M_{\li A}$ to be $\mathbb{N} \oplus M_A$, where $\oplus$
-denotes the free product of monoids (the coproduct in the category of monoids).
-The intuition behind needing to include $\mathbb{N}$ comes from the example of
-the downcast for $\iarr$. Recall that we adjusted the squares corresponding to
-the DnL and DnR rules by adding $(\delta \circ \eta)^\dagger$ on the side
-opposite the downcast. Thus, we need a syntactic perturbation in $M_{\li A}$
-that will be interpreted as $(\delta \circ \eta)^\dagger$. We accomplish this by
-taking the coproduct with $\mathbb{N}$.
+i_A)$ we define the syntactic perturbations $M_{\li A}$ to be $\mathbb{N} \oplus
+M_A$, where $\oplus$ denotes the free product of monoids (the coproduct in the
+category of monoids). The intuition behind needing to include $\mathbb{N}$ comes
+from the example of the downcast for $\iarr$. Recall that we adjusted the
+squares corresponding to the DnL and DnR rules by adding $(\delta \circ
+\eta)^\dagger$ on the side opposite the downcast. Thus, we need a syntactic
+perturbation in $M_{\li A}$ that will be interpreted as $(\delta \circ
+\eta)^\dagger$. We accomplish this by taking the coproduct with $\mathbb{N}$.
 %
@@
 value/computation relations? Actions of the functors on relations?}
-
-\eric{TODO: Quasi-equivalence of the relations mentioned in the syntax section.}
 
@@
 %
-% Unfolding the definitions, this is the same as
-% %
-% \( M_D \cong (M_D \times M_D) \oplus (\mathbb{N} \times M_D^{op} \times \mathbb{N} \times M_D). \)
-% %
 We now explain how to interpret these perturbations as endomorphisms.
@@
 
-We know that in order for the DnR and DnL squares for $\li \iarr$ to exist, we must
-insert a delay on the side opposite the downcast. In terms of perturbations,
-this means that we take the perturbations $\dellp$ and $\delrp$ in the
-definition of quasi-right-representability to both be $\inl\, 1$, where $\inl$
-is the injection into the coproduct of monoids. We recall that the monoid of
-perturbations for $\li$ involves a coproduct with $\mathbb{N}$, and that the
-interpretation homomorphism maps the generator $1$ to the endomorphism of error
-domains $(\delta \circ \eta)^\dagger$.
-
+We know that in order for the DnR and DnL squares for $\li \iarr$ to exist, we
+must insert a delay on the side opposite the downcast. In terms of syntactic
+perturbations, this means that we take $\dellp$ and $\delrp$ in the definition
+of quasi-right-representability to both be $\inl\, 1$, where $\inl$ is the
+injection into the coproduct of monoids. We recall that the definition of
+syntactic perturbations for $\li$ involves a coproduct with $\mathbb{N}$, and
+that the interpretation homomorphism maps the generator $1$ to the endomorphism
+of error domains $(\delta \circ \eta)^\dagger$.
+%
 With this choice for the perturbations, it is straightforward to show that the
@@
 
-% We define $\dellp_D = \delrp_D = \delta_{FD}$. Then it is easy to show using the
-% definition of $\iarr$ that the squares for $\dnl$ and $\dnr$ commute.
-% Then for $\dnr$, we need to show that if $(f, d) \in \iarr$ then
-%
 It is also straightforward to establish the retraction property for each of
@@
 
-% \subsection{Obtaining an Extensional Model}
+\subsection{Interpreting Term Precision: Extensional Squares}
 
-\subsection{Defining an Extensional Model}
-
-To obtain a model for gradual typing that is oblivious to the stepping behavior
-of terms, we apply the ``bisimilarity-closure'' construction introduced in the
-previous section as follows. Let $c_i : A_i \rel A_i'$ and $c_o : A_o \rel
-A_o'$.
+As introduced in the previous section, to obtain a model of term precision that
+is oblivious to the stepping behavior of terms, we weaken our notion of square
+to allow for the morphisms on either side to be synchronized to be in lock-step.
+More formally, let $c_i : A_i \rel A_i'$ and $c_o : A_o \rel A_o'$.
 %
@@
 % A_i' \to A_o'$ to consist of:
-
+%
 % \begin{itemize}
@@
 %
-If the relations $c_i$ and $c_o$ are obvious from the context, we will simply
-write $f \ltbisim g$. We make the analogous construction for the computation
-squares.
+We call a square $f \ltsqbisim{c^i}{c_o} g$ an \emph{extensional square}, and we
+will call a square $f \ltsq{c^i}{c_o} g$ an \emph{intensional square}. If the
+relations $c_i$ and $c_o$ are obvious from the context, we will simply write $f
+\ltbisim g$. We make the analogous construction for the computation squares.
 %
 We now verify the existence of the squares for the cast rules. For UpL, the
-square is obtained as follows:
-
+relevant extensional square is obtained as follows:
+%
 % https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMSwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMCwzLCJlX2MiLDIseyJjdXJ2ZSI6MX1dLFsxLDQsIlxcZGVscmVfYyIsMl0sWzIsNSwiXFxwdXNoX3tjJ30oXFxkZWxyZV9jKSIsMix7ImN1cnZlIjoxfV0sWzAsMSwiYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEsMiwiYyciLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFszLDQsIlxcbGVfe0FfMn0iLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs0LDUsImMnIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMiw1LCJcXGlkIiwwLHsiY3VydmUiOi0xfV0sWzAsMywiZV9jIiwwLHsiY3VydmUiOi0xfV0sWzgsMTMsIlxcYmlzaW0iLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNiwxNCwiXFxiaXNpbSIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
@@
 \end{tikzcd}\]
-
-Note that the horizontal composition of the squares is defined, as it takes
-place in the intensional setting. We have used the push-pull structure on $c'$
-to turn the perturbation on $A_2$ into a perturbation on $A_3$. Lastly we note
-that the bottom edge of the square is equal to $c'$ by the fact that $c'$ is
-downward closed. The squares for UpR, DnL, and DnR are obtained similarly.
+%
+% Note that the horizontal composition of the squares is defined, as they are
+% intensional squares.
+We have used the push-pull structure on $c'$ to turn the perturbation on $A_2$
+into a perturbation on $A_3$. Lastly we note that the bottom edge of the square
+is equal to $c'$ by the fact that $c'$ is downward closed. The squares for UpR,
+DnL, and DnR are obtained similarly.
 
@@
 
-\eric{Order-equivalence of relations}
+\eric{TODO: Does this belong here?}
+
+Recall the equivalence rules for type precision given in Section \ref{sec:GTLC},
+for example the rule that $(c_i \ra c_o)(c_i' \ra c_o')\equiv (c_ic_i' \ra
+c_oc_o')$ To validate this rule in the model, we construct two squares $\id
+\ltsqbisim{(c_i \ra c_o)(c_i' \ra c_o')}{(c_ic_i' \ra c_oc_o')} \id$ and $\id
+\ltsqbisim{(c_ic_i' \ra c_oc_o')}{(c_i \ra c_o)(c_i' \ra c_o')} \id$. Since the
+definition of extensional square allows us to specify intermediate morphisms
+that are synchronized in the lock-step error ordering, it will be sufficient to
+construct two intensional squares involving perturbations on both sides. This
+motivates the definition of \emph{quasi-order-equivalence}.
+
+\begin{definition}[quasi-order-equivalence]\label{def:quasi-order-equivalent}
+  Let $c, c' : A \rel A'$. We say that $c$ and $c'$ are
+  \textbf{quasi-order-equivalent}, written $c \qordeq c'$, if there exist
+  perturbations $\delta^l_1, \delta^l_2 \in \pv_A$ and $\delta^r_1, \delta^r_2
+  \in \pv_{A'}$ such that there is a square $\delta^l_1 \ltsq{c}{c'} \delta^r_1$
+  and a square $\delta^l_2 \ltsq{c'}{c} \delta^r_2$.
+\end{definition}
+We make the analogous definition for error domain relations $d, d' : B \rel B'$.
+
+Then we have the following result that shows that the functors
+$U,\li,\times,\to$ are \emph{quasi-functorial} on relations:
+
+\begin{lemma}
+\begin{itemize}
+  \item $U(d \comp d') \qordeq U(d) \comp U(d')$
+  \item $F(c \comp c') \qordeq F(c) \comp F(c')$
+  \item $(c \comp c') \to (d \comp d') \qordeq (c \to d) \comp (c' \to d')$
+  \item $(c_1 \comp c_1') \times (c_2 \comp c_2') \qordeq (c_1 \times c_2) \comp (c_1'\times c_2')$
+\end{itemize}
+\end{lemma}
+\begin{proof} See appendix \eric{TODO: make sure this is in the appendix}
+\end{proof}
 
@@
 
-\eric{TODO: update this in light of the new section on adequacy of the term model.}
-
-In this section, we prove an adequacy result for the extensional model of
-gradual typing we obtained in the previous section. First we establish some
-notation. Fix a morphism $f : 1 \to \li \mathbb{N} \cong \li \mathbb{N}$. We
-write $f \da n$ to mean that there exists $m$ such that $f = \delta^m(\eta n)$
-and $f \da \mho$ to mean that there exists $m$ such that $f = \delta^m(\mho)$.
+In this section, we prove the adequacy of the interpretation of term precision.
+First we establish some notation. Fix a morphism $f : 1 \to \li \mathbb{N} \cong
+\li \mathbb{N}$. We write $f \da n$ to mean that there exists $m$ such that $f =
+\delta^m(\eta n)$ and $f \da \mho$ to mean that there exists $m$ such that $f =
+\delta^m(\mho)$.
 
 Recall that $\ltbisim$ denotes the relation on value morphisms defined as the
-bisimilarity-closure of the intensional error-ordering on morphisms. That is, we
-have $f \ltbisim g$ iff there exist $f'$ and $g'$ with
+closure of the intensional error-ordering on morphisms under weak bisimilarity
+on bboth sides. Specialized to the setting of morphisms from $1 \to U \li
+\mathbb{N}$, we have $f \ltbisim g$ iff there exist $f'$ and $g'$ with
 %
@@
 %
-Here $\le_{\li \mathbb{N}}$ is the lock-step error ordering, and $\bisim_{\li
-\mathbb{N}}$ is weak bisimilarity. First observe that in this ordering, the
-semantics of error is not equivalent to the semantics of the diverging term. The
-main result we would like to show is as follows: If $f \ltbisim g : \li \mathbb{N}$, then:
+
+We first note that in this ordering, the semantics of error is not equivalent to
+the semantics of the diverging term. The main result we would like to show is as
+follows: If $f \ltbisim g : \li \mathbb{N}$, then:
 \begin{itemize}
@@
 %
-Unfortunately, this is actually not provable!
-Roughly speaking, the issue is that this is a ``global'' result, and it is not possible
-to prove such results inside of the guarded setting. 
-In particular, if we tried to prove the above result in the guarded
-setting, we would run into a problem where we would have a natural number
-``stuck'' under a $\later$, with no way to get out the underlying number.
+Unfortunately, this is actually not provable! Roughly speaking, the issue is
+that this is a ``global'' result, and it is not possible to prove such results
+in the guarded setting. In particular, if we tried to prove the above result in
+the guarded setting we would run into a problem where we would have a natural
+number ``stuck'' under a $\later$, with no way to get out the underlying number.
 %
-Thus, to prove our adequacy result, we need to leave the guarded setting and pass back
-to the more familiar, set-theoretic world with no internal notion of step-indexing.
-We can do this using a process known as \emph{clock quantification}.
-Recall that all of the constructions we have made in SGDT take place in the context of a clock $k$.
-All of our uses of the later modality and guarded recursion have taken place with respect to this clock.
-For example, recall the definition of the lift monad by guarded recursion.
-% We define the lift monad $\li^k X$ as the guarded fixpoint of $\lambda \tilde{T}. X + 1 + \later^k_t (\tilde{T}_t)$.
-We can view this definition as being parameterized by a clock $k$: $\li^k : \type \to \type$.
-Then for $X$ satisfying a certain technical requirement known as \emph{clock-irrelevance},
-\footnote{A type $X$ is clock-irrelevant if there is an isomorphism $\forall k.X \cong X$.}
-we can define the ``global lift'' monad as $\li^{gl} X := \forall k. \li^k X$.
+Thus, to prove our adequacy result, we need to pass to leave the guarded
+setting. We have seen in Section \ref{sec:big-step-term-semantics} that clock
+quantification makes this possible.
 
-It can be shown that there is an isomorphism between the global lift monad and the
-delay monad of Capretta \cite{lmcs:2265}.
-Recall that, given a type $X$, the delay monad $\text{Delay}(X)$ is defined as the coinductive
-type generated by 
-$\tnow : X \to \delay(X)$ and $\tlater : \delay(X) \to \delay(X)$.
+Recall that the definitions of our model, e.g., the free error domain, the
+lock-step ordering, and weak bisimilarity, have all been parameterized by a
+clock $k$. We saw in previously that there is an isomorphism between the
+globalization of the free error domain and Capretta's coinductive delay monad:
+$\li^{gl} \mathbb{N} \cong \delay(\mathbb{N} + {\mho})$.
 
-% solution to the equation
-
-% \[ \text{Delay}(X) \cong X + \text{Delay}(X). \]
-
-It can be shown that for a clock-irrelevant type $X$, $\li^{gl} X$ is a final
-coalgebra of the functor $F(Y) = X + 1 + Y$ (For example, this follows from Theorem 4.3 in
-\cite{kristensen-mogelberg-vezzosi2022}.) 
-\footnote{The proof relies on the existence of an operation  
-$\mathsf{force} : \forall k. \later^k A \to \forall k. X$ that
-allows us to eliminate the later operator under a clock quantifier.
-This must be added as an axiom in guarded type theory.}
-Since $\delay(X + 1)$ is also a final coalgebra
-of this functor, then we have $\li^{gl} X \cong \delay(X + 1)$.
-
-Given a predomain $X$ on a clock-irrelevant type, we can define a
-``global'' version of the lock-step error ordering and the
-weak bisimilarity relation on elements of the global lift; the former is defined by
+Now consider the predomain $\mathbb{N}$ with ordering and bisimilarity both the
+equality relation. We can define a ``global'' version of the lock-step error
+ordering and the weak bisimilarity relation on elements of the $\li^{gl}
+\mathbb{N}$; the former is defined by
 %
-\( x \le^{gl}_X y := \forall k. x[k] \le y[k], \)
+\( x \ltls^{gl}_X y := \forall k. x[k] \ltls y[k], \)
 %
@@
 %
-On the other hand, we can define coinductively a ``lock-step error ordering"
-relation on $\delay(X + 1)$:
+On the other hand, we can define coinductively a ``lock-step error ordering''
+relation on $\delay(\mathbb{N} + {\mho})$:
 %
@@
 %
-And we similarly define by coinduction a ``weak bisimilarity'' relation on $\delay(X + 1)$, which uses
-a relation $d \Da x_?$ between $\delay(X+1)$ and $X+1$ that is defined as 
-$d \Da x_? := \Sigma_{m \in \mathbb{N}} d = \tlater^m(\tnow\, x_?)$.
-Then weak bisimilarity is defined by the rules
+And we similarly define by coinduction a ``weak bisimilarity'' relation on
+$\delay(\mathbb{N} + {\mho})$. This uses a relation $d \Da x_?$ between
+$\delay(\mathbb{N} + {\mho})$ and $\mathbb{N} + {\mho}$ that is defined as $d
+\Da n_? := \Sigma_{i \in \mathbb{N}} d = \tlater^i(\tnow\, n_?)$. Then weak
+bisimilarity for the delay monad is defined coinductively by the rules
 %
@@
   \inferrule*[]
-  {x_? \bisim_{X + 1} y_?}
-  {\tnow\, x_? \bisimdelay \tnow\, y_? }
+  {n_? \bisim_{\mathbb{N} + {\mho}} m_?}
+  {\tnow\, n_? \bisimdelay \tnow\, m_? }
 
-  \inferrule*[]
-  {d_1 \Da x_? \and x_? \bisim_{X + 1} y_?}
-  {\tlater\, d_1 \bisimdelay \tnow\, y_? }
+  \inferrule*[leftskip=1.5em]
+  {d_1 \Da n_? \and n_? \bisim_{\mathbb{N} + {\mho}} m_?}
+  {\tlater\, d_1 \bisimdelay \tnow\, m_? }
 
-  \inferrule*[]
-  {d_2 \Da y_? \and x_? \bisim_{X + 1} y_?}
-  {\tnow\, x_? \bisimdelay \tlater\, d_2}
+  \inferrule*[leftskip=1.5em]
+  {d_2 \Da m_? \and n_? \bisim_{\mathbb{N} + {\mho}} m_?}
+  {\tnow\, n_? \bisimdelay \tlater\, d_2}
 
-  \inferrule*[]
+  \inferrule*[leftskip=1.5em]
   {d_1 \bisimdelay d_2}
   {\tlater\, d_1 \bisimdelay \tlater\, d_2 }
-
+  %
   % \inferrule*[]
@@
   % {d_1 \bisimdelay d_2}
-
+  %
   % \inferrule*[]
@@
   % {\tlater d_1 \bisimdelay \tlater d_2 }
-
 \end{mathpar}
 %
-Note the similarity of these definitions to the corresponding guarded definitions.
-By adapting the aforementioned theorem to the setting of inductively-defined relations,
-we can show that both the global lock-step error ordering and the global weak bisimilarity
-admit coinductive definitions. In particular, modulo the above isomorphism
-between $\li^{gl} X$ and $\delay(X+1)$, the global version of the lock-step
-error ordering is equivalent to the lock-step error ordering on $\delay(X + 1)$,
-and likewise, the global version of the weak bisimilarity relation is equivalent to the
-weak bisimilarity relation on $\delay(X + 1)$.
+Note the similarity of these definitions to the corresponding guarded
+definitions. By adapting Theorem 4.3 of \cite{kristensen-mogelberg-vezzosi2022}
+to the setting of inductively-defined relations, we can show that both the
+global lock-step error ordering and the global weak bisimilarity admit
+coinductive definitions. In particular, modulo the above isomorphism between
+$\li^{gl} X$ and $\delay(\mathbb{N} + {\mho})$, the global version of the
+lock-step error ordering is equivalent to the lock-step error ordering on
+$\delay(\mathbb{N} + {\mho})$, and likewise, the global version of the weak
+bisimilarity relation is equivalent to the weak bisimilarity relation on
+$\delay(\mathbb{N} + {\mho})$.
 
-This implies that the global version of the extensional term precision semantics for
-$\li^{gl} X$ agrees with the corresponding notion for $\delay(X + 1)$.
-Then adequacy follows by proving the corresponding
-result for $\delay(X + 1)$ which in turn follows from the definitions of the relations.
+This implies that the global version of the term precision semantics for
+$\li^{gl} \mathbb{N}$ agrees with the corresponding notion for
+$\delay(\mathbb{N} + {\mho})$. Then adequacy follows by proving the
+corresponding result for $\delay(\mathbb{N} + {\mho})$ which in turn follows
+from the definitions of the relations.
 

### 2024-07-11 paper-new/towards-relational-model.tex
revise sections on relational model

@@
 %
-Sticking to a poset-based semantics, if our value types are posets, we
-then want our precision derivations $c : A \ltdyn A'$ to be
-\emph{poset relations} on $A, A'$, that is relations between the
-underlying sets of $A, A'$ that are \emph{downward-closed} in $A$ and
-\emph{upward-closed} in $A'$.  We denote such a relation between $A$
-and $A'$ by $c : A \rel A'$. The paradigmatic example of a poset
-relation is the poset ordering itself, $\leq_A : A \rel A$.
+Sticking to a poset-based semantics, if our value types are posets, we then want
+our precision derivations $c : A \ltdyn A'$ to be \emph{poset relations} on $A,
+A'$, that is relations between the underlying sets of $A, A'$ that are
+\emph{downward-closed} in $A$ and \emph{upward-closed} in $A'$.  We denote such
+a relation between $A$ and $A'$ by $c : A \rel A'$. The paradigmatic example of
+a poset relation is the poset ordering itself, $\leq_A : A \rel A$, which is a
+poset relation because it is transitive.
+%
+Given two poset relations $c : A_1 \rel A_2$ and $c' : A_2 \rel A_3$, we can
+form their \emph{composition} in the usual manner, i.e., $(x, y) \in c$ if and
+only if there is an element $z : A_2$ with $(x, z) \in c$ and $(z, y) \in c'$.
+We write the composition of relations as $c \comp c'$.
 
@@
 
-\eric{Picked up here; Max, please read this.}  These relations are
-then used to model term precision. For closed programs $M \ltdyn M' :
+% \eric{Picked up here; Max, please read this.}
+
+These relations are then used to model term precision. For closed programs $M \ltdyn M' :
 c$ where $c : A \ltdyn A'$ we have that these denote elements of the
@@
 
-\max{This needs to also talk about relational composition and horizontal pasting squares, since that's transitivity!}
+% \max{This needs to also talk about relational composition and horizontal pasting squares, since that's transitivity!}
 We note that for any $c$, there is a ``vertical identity'' square $\id
@@
 montone function $f : A_i \to A_o$, there is a ``horizontal identity'' square $f
-\ltsq{r(A_i)}{r(A_o)}$ by the definition of monotone function.
+\ltsq{\le_{A_i}}{\le_{A_o}}$ by the definition of monotone function.
+%
+We can \emph{compose} squares vertically, i.e., if we have $f \ltsq{c_1}{c_2}
+f'$ and $g \ltsq{c_2}{c_3} g'$ then we obtain a square $g \circ f
+\ltsq{c_1}{c_3} g' \circ f'$. Likewise, we can compose squares horizontally: If
+$f \ltsq{c_i}{c_o} g$ and $g \ltsq{c_i'}{c_o'} h$, then we obtain a square $f
+\ltsq{c_i \comp c_i'}{c_o \comp c_o'} h$. Horizontal composition of squares
+corresponds to transitivity of term precision.
 
@@
 functions in the semantics this can be deduced from the existence of
-the following square:\max{the square doesn't make sense because we
-  haven't discussed horizontal composition of relations or that this
-  notation means horizontal composition of relations}
+the following square, where the relation on the top is $\sem{c} \comp \sem{c'}$:
+% \max{the square doesn't make sense because we
+%   haven't discussed horizontal composition of relations or that this
+%   notation means horizontal composition of relations}
 %
@@
 corresponding to UpL:
-% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzIiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzIiXSxbMCwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJyKEFfMikiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxcdXBje2N9IiwyXSxbMiwzLCJcXGlkIl1d
+% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzIiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzIiXSxbMCwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJcXGxlX3tBXzJ9IiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJcXHVwY3tjfSIsMl0sWzIsMywiXFxpZCJdXQ==
 \[\begin{tikzcd}[ampersand replacement=\&]
@@
 	\arrow["\id", from=1-2, to=2-2]
-  \arrow["{r(A_2)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+	\arrow["{\le_{A_2}}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
 \end{tikzcd}\]
@@
 that the relation is a kind of \emph{graph} of the function in that $x
-\binrel{\sem c} y$ if and only if $\sem{upc c}(x) \leq_{A_2} y$.
+\binrel{\sem c} y$ if and only if $\sem{\upc c}(x) \leq_{A_2} y$.
 
@@
 %
-% https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMSwxLCJBXzIiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzMsNSwicihBXzIpIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNSw0LCJjJyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMywiXFx1cGN7Y30iLDJdLFsyLDQsIlxcaWQiXSxbMSw1LCJcXGlkIiwyXV0=
+% https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMSwxLCJBXzIiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzMsNSwiXFxsZV97QV8yfSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzUsNCwiYyciLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDMsIlxcdXBje2N9IiwyXSxbMiw0LCJcXGlkIl0sWzEsNSwiXFxpZCIsMl1d
 \[\begin{tikzcd}[ampersand replacement=\&]
@@
 	\arrow["\id", from=1-3, to=2-3]
-  \arrow["{r(A_2)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+	\arrow["{\le_{A_2}}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
 	\arrow["{c'}"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
 \end{tikzcd}\]
+%
+Observe that the composition of $\le_{A_2}$ with $c'$ implies
+$c'$, because $c'$ is \emph{downward-closed} under the relation on $A_2$. Thus, we
+obtain the original UpL square by vertically composing the square shown above
+with the square $\id \ltsq{\le_{A_2}\, c'}{c'} \id$ on the bottom.
+
+The actual place where transitivity of the relations on the value types is
+needed in this argument occurs when $A_3 = A_2$ and we take $c'$ to be
+$\le_{A_2}$ (recall that $c'$ is universally quantified, so it can be any
+relation). The fact that $\le_{A_2}$ is downward-closed is equivalent to it
+being transitive. In other words, all relations $\le_A$ on value types must be
+poset relations in order to carry out the above compositional construction.
+
 So we see that in order to get a sensible condition that can give us a
@@
 \footnote{Technically, the theorem only implies that we need to drop
-one of both left- and right-step-insensitivity. We might then attempt
+one of left- or right-step-insensitivity. We might then attempt
 to work with two separate ``step-semi-sensitive'' relations on $\li
@@
 transitive, though we have not been able to prove that they are not
-transitive either.\max{Eric check if this footnote is correct}}
+transitive either.}
+% \max{Eric check if this footnote is correct}}
 %
@@
 with a second relation on the free error domain $\li A$: \emph{weak
-bisimilarity}, defined in Figure~\ref{fig:weak-bisim}, which is
+bisimilarity}, defined in Figure~\ref{fig:weak-bisimilarity}, which is
 parameterized by a binary relation $\bisim_A$ on $A$

### 2024-07-11 paper-new/discussion.tex
clean up intro and move section on prior approaches to discussion

@@
 \section{Discussion}\label{sec:discussion}
+
+\eric{I moved this from the intro. We may need to revise it.}
+
+\subsection{Limitations of Prior Work}
+
+We give an overview of current approaches to proving graduality of
+languages and why they do not meet our criteria of a reusable semantic
+framework.
+
+\subsubsection{From Static to Gradual}
+
+Current approaches to constructing languages that satisfy the
+graduality property include the methods of Abstracting Gradual Typing
+\cite{garcia-clark-tanter2016} and the formal tools of the Gradualizer
+\cite{cimini-siek2016}.  These allow the language developer to start
+with a statically typed language and derive a gradually typed language
+that satisfies the gradual guarantee. The main downside to these
+approaches lies in their inflexibility: since the process in entirely
+mechanical, the language designer must adhere to the predefined
+framework.  Many gradually typed languages do not fit into either
+framework, e.g., Typed Racket \cite{tobin-hochstadt06,
+  tobin-hochstadt08} and the semantics produced is not always the
+desired one.
+%
+Furthermore, while these frameworks do prove graduality of the
+resulting languages, they do not show the correctness of the
+equational theory, which is equally important to sound gradual typing.
+
+\subsubsection{Double Categorical Semantics}
+
+New and Licata \cite{new-licata18} developed an axiomatic account of
+the graduality relation on a call-by-name cast calculus terms and
+showed that the graduality proof could be modeled using semantics in
+certain kinds of \emph{double categories}, categories internal to the
+category of categories. A double category extends a category with a
+second notion of morphism, often a notion of ``relation'' to be paired
+with the notion of functional morphism, as well as a notion of
+functional morphisms preserving relations. In gradual typing the
+notion of relation models type precision and the squares model the
+term precision relation. This approach was influenced by the semantics
+of parametricity using reflexive graph categories
+\cite{ma-reynolds,dunphythesis,reynoldsprogramme}: reflexive graph
+categories are essentially double categories without a notion of
+relational composition. In addition to capturing the notions of type
+and term precision, the double categorical approach allows for a
+\emph{universal property} for casts: upcasts are the \emph{universal}
+way to turn a relation arrow into a function in a forward direction
+and downcasts are the dual universal arrow.  Later, New, Licata and
+Ahmed \cite{new-licata-ahmed2019} extended this axiomatic treatment from
+call-by-name to call-by-value as well by giving an axiomatic theory of
+type and term precision in call-by-push-value. This left implicit any
+connection to a ``double call-by-push-value'', which we make explicit
+in Section~\ref{sec:cbpv}.
+
+With this notion of abstract categorical model in hand, denotational
+semantics is then the work of constructing concrete models that
+exhibit the categorical construction. New and Licata
+\cite{new-licata18} present such a model using categories of
+$\omega$-CPOs, and this model was extended by Lennon-Bertrand,
+Maillard, Tabareau and Tanter to prove graduality of a gradual
+dependently typed calculus $\textrm{CastCIC}^{\mathcal G}$. This
+domain-theoretic approach meets our criteria of being a semantic
+framework for proving graduality, but suffers from the limitations of
+classical domain theory: the inability to model viciously
+self-referential structures such as higher-order extensible state and
+similar features such as runtime-extensible dynamic types. Since these
+features are quite common in dynamically typed languages, we seek a
+new denotational framework that can model these type system features.
+
+The standard alternative to domain theory that scales to essentially
+arbitrary self-referential definitions is \emph{step-indexing} or its
+synthetic form of \emph{guarded recursion}. A series of works
+\cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
+developed step-indexed logical relations models of gradually typed
+languages based on operational semantics. Unlike classical domain
+theory, such step-indexed techniques are capable of modeling
+higher-order store and runtime-extensible dynamic types
+\cite{appelmcallester01,ahmed06,neis09,new-jamner-ahmed19}. However,
+their proof developments are highly repetitive and technical, with
+each development formulating a logical relation from first-principles
+and proving many of the same tedious lemmas without reusable
+mathematical abstractions.
 

### 2024-07-11 paper-new/intro.tex
clean up intro and move section on prior approaches to discussion

@@
 
+In this paper, we develop an adequate denotational semantics that satisfies
+graduality and soundness of the equational theory of cast calculi using the
+techniques of SGDT.
+
 \subsection{Contributions}
@@
 equality and satisfies a graduality theorem. A great deal of the work
-has further been verified in guarded cubical Agda, demonstrating that
-the semantics is readily mechanizable.
+has further been verified in guarded cubical Agda \cite{veltri-vezzosi2020}, 
+demonstrating that the semantics is readily mechanizable.
 
@@
 \begin{enumerate}
-\item In Section\ref{sec:syntax} we fix our input language, a fairly
+\item In Section\ref{sec:GTLC} we fix our input language, a fairly
   typical gradually typed cast calculus.
-\item In Section\ref{sec:concrete-terms} we develop a first
+\item In Section\ref{sec:concrete-term-model} we develop a first
   denotational semantics in synthetic guarded domain theory that
@@
   modeling effects using call-by-push-value.
-\item In Section\ref{sec:relational-issues} we show where the
+\item In Section\ref{sec:towards-relational-model} we show where the
   New-Licata classical domain theoretic approach fails to adapt
@@
   proving graduality in an intensional model.
-\item In Section\ref{sec:relational-semantics} we
+\item In Section\ref{sec:concrete-relational-model} we describe the resulting
+  model in detail. We then prove that the resulting denotational model is
+  \emph{adequate} for the graduality property: a closed term precision $M \ltdyn
+  N : \nat$ has the expected semantics, that $M$ errors or $M$ and $N$ have the
+  same extensional behavior.
 \item In Section\ref{sec:discussion} we discuss prior work on proving
@@
 \end{enumerate}
-
-\subsection{Limitations of Prior Work}
-
-We give an overview of current approaches to proving graduality of
-languages and why they do not meet our criteria of a reusable semantic
-framework.
-
-\subsubsection{From Static to Gradual}
-
-Current approaches to constructing languages that satisfy the
-graduality property include the methods of Abstracting Gradual Typing
-\cite{garcia-clark-tanter2016} and the formal tools of the Gradualizer
-\cite{cimini-siek2016}.  These allow the language developer to start
-with a statically typed language and derive a gradually typed language
-that satisfies the gradual guarantee. The main downside to these
-approaches lies in their inflexibility: since the process in entirely
-mechanical, the language designer must adhere to the predefined
-framework.  Many gradually typed languages do not fit into either
-framework, e.g., Typed Racket \cite{tobin-hochstadt06,
-  tobin-hochstadt08} and the semantics produced is not always the
-desired one.
-%
-Furthermore, while these frameworks do prove graduality of the
-resulting languages, they do not show the correctness of the
-equational theory, which is equally important to sound gradual typing.
-
-%% For example, programmers often refactor their code, and in so doing they rely
-%% implicitly on the validity of the laws in the equational theory.
-%% Similarly, correctness of compiler optimizations rests on the validity of the
-%% corresponding equations from the equational theory. It is therefore important
-%% that the languages that claim to be gradually typed have provably correct
-%% equational theories.
-
-% The approaches are too inflexible... the fact that the resulting semantics are too lazy
-% is a consequence of that inflexibility.
-% The validity of the equational theory captures the programmer's intuitive thinking when they refactor their code
-
-%The downside is that
-%not all gradually typed languages can be derived from these frameworks, and moreover, in both
-%approaches the semantics is derived from the static type system as opposed to the alternative
-%in which the semantics determines the type checking. Without a clear semantic interpretation of type
-%dynamism, it becomes difficult to extend these techniques to new language features such as polymorphism.
-
-% [Eric] I moved the next two paragraphs from the technical background section
-% to here in the intro.
-\subsubsection{Double Categorical Semantics}
-
-New and Licata \cite{new-licata18} developed an axiomatic account of
-the graduality relation on a call-by-name cast calculus terms and
-showed that the graduality proof could be modeled using semantics in
-certain kinds of \emph{double categories}, categories internal to the
-category of categories. A double category extends a category with a
-second notion of morphism, often a notion of ``relation'' to be paired
-with the notion of functional morphism, as well as a notion of
-functional morphisms preserving relations. In gradual typing the
-notion of relation models type precision and the squares model the
-term precision relation. This approach was influenced by the semantics
-of parametricity using reflexive graph categories
-\cite{ma-reynolds,dunphythesis,reynoldsprogramme}: reflexive graph
-categories are essentially double categories without a notion of
-relational composition. In addition to capturing the notions of type
-and term precision, the double categorical approach allows for a
-\emph{universal property} for casts: upcasts are the \emph{universal}
-way to turn a relation arrow into a function in a forward direction
-and downcasts are the dual universal arrow.  Later, New, Licata and
-Ahmed \cite{new-licata-ahmed2019} extended this axiomatic treatment from
-call-by-name to call-by-value as well by giving an axiomatic theory of
-type and term precision in call-by-push-value. This left implicit any
-connection to a ``double call-by-push-value'', which we make explicit
-in Section~\ref{sec:cbpv}.
-
-With this notion of abstract categorical model in hand, denotational
-semantics is then the work of constructing concrete models that
-exhibit the categorical construction. New and Licata
-\cite{new-licata18} present such a model using categories of
-$\omega$-CPOs, and this model was extended by Lennon-Bertrand,
-Maillard, Tabareau and Tanter to prove graduality of a gradual
-dependently typed calculus $\textrm{CastCIC}^{\mathcal G}$. This
-domain-theoretic approach meets our criteria of being a semantic
-framework for proving graduality, but suffers from the limitations of
-classical domain theory: the inability to model viciously
-self-referential structures such as higher-order extensible state and
-similar features such as runtime-extensible dynamic types. Since these
-features are quite common in dynamically typed languages, we seek a
-new denotational framework that can model these type system features.
-
-The standard alternative to domain theory that scales to essentially
-arbitrary self-referential definitions is \emph{step-indexing} or its
-synthetic form of \emph{guarded recursion}. A series of works
-\cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
-developed step-indexed logical relations models of gradually typed
-languages based on operational semantics. Unlike classical domain
-theory, such step-indexed techniques are capable of modeling
-higher-order store and runtime-extensible dynamic types
-\cite{appelmcallester01,ahmed06,neis09,new-jamner-ahmed19}. However,
-their proof developments are highly repetitive and technical, with
-each development formulating a logical relation from first-principles
-and proving many of the same tedious lemmas without reusable
-mathematical abstractions. Our goal in the current work is to extract
-these reusable mathematical principles from these explicit
-step-indexed to make formalization of realistic gradual languages
-tractible.
-
-% Alternative phrasing:
-\begin{comment}
-\subsubsection{Embedding-Projection Pairs}
-
-The series of works by New, Licata, and Ahmed \cite{new-licata18,new-ahmed2018,new-licata-ahmed2019}
-develop an axiomatic account of gradual typing involving \emph{embedding-projection pairs}.
-This allows for a particularly elegant formulation of the gradual guarantee.
-Moreover, their axiomatic account of program equivalence allows for type-based reasoning about program equivalences.
-
-In \cite{new-licata18}, New and Licata construct a denotational model of the axioms of
-gradual typing using tools from classical domain theory. The benefit to this approach is that it is a reusable mathematical theory:
-general semantic theorems about gradual domains can be developed independent of any particular syntax and then reused in many different denotational models.
-Unfortunately, however, it is unclear how to extend these domain-theoretic techniques to incorporate more advanced language features such as higher-order store, a standard feature of realistic gradually typed languages such as Typed
-Racket.
-Thus, if we want a reusable mathematical theory of gradual typing that can scale to realistic programming languages, we need to look to different techniques.
-
-In \cite{new-ahmed2018,new-licata-ahmed2019}, a more traditional operational semantics for gradual typing is derived from the axioms.
-% The axioms are then proven to be sound with respect to this operational semantics by constructing a logical relation.
-A logical relations model is constructed and used to prove that the axioms are sound with respect to the operational semantics.
-The downside of this approach is that each new language requires a different logical relation
-to prove graduality, even though the logical relations for different languages end up sharing many commonalities.
-Furthermore, the logical relations tend to be quite complicated due to a technical requirement known as \emph{step-indexing},
-where the stepping behavior of terms must be accounted for in the logical relation.
-As a result, developments using this approach tend to require vast effort, with the
-corresponding technical reports having 50+ pages of proofs.
-
-% In this approach, a logical relation is constructed and used to show that the equational theory
-% is sound with respect to the operational semantics.
-
-% Additionally, while the axioms of gradual type theory are compositional at a ``global'' level,
-% they do not compose in the step-indexed setting. One of the main goals of the present work
-% is to formulate a composable theory of gradual typing in a setting where the stepping behavior
-% is tracked.
-\end{comment}
-
-An alternative approach, which we investigate in this paper, is provided by
-\emph{synthetic guarded domain
-theory}\cite{birkedal-mogelberg-schwinghammer-stovring2011}, abbreviated as
-SGDT. The techniques of SGDT allow us to internalize the step-indexed reasoning
-normally required in logical relations proofs of graduality, ultimately allowing
-us to specify the logical relation in a manner that looks nearly identical to a
-typical, non-step-indexed logical relation. In fact, guarded domain theory goes
-further, allowing us to define step-indexed \emph{denotational semantics} not
-just step-indexed relations, just as easily as constructing an ordinary
-set-theoretic semantics.
-
-In this paper, we develop an adequate denotational semantics that satisfies
-graduality and soundness of the equational theory of cast calculi using the
-techniques of SGDT.
-
-\eric{This is out of date: the mechanization is now an artifact of the current paper.}
-Our longer-term goal is to mechanize these proofs in a reusable way in the
-Guarded Cubical Agda \cite{veltri-vezzosi2020} proof assistant, thereby
-providing a framework to more easily and conveniently prove that existing
-languages satisfy graduality and have sound equational theories.
-
-Moreover, the aim is for designers of new languages to utilize the framework to
-facilitate the design of new provably-correct gradually-typed languages with
-more complex features.
-
-\subsection{Contributions}
-
-The main contribution of this work is a categorical and denotational semantics
-for gradually typed langauges that models not just the term language but the
-graduality property as well.
-\begin{enumerate}
-\item First, we give a simple abstract categorical model of Gradual Type Theory
-using CBPV double categories.
-\item Next, we modify this semantics to develop reflexive graph- and double
-  categorical models that abstract over the details of step-indexed models, and
-  provide a method for constructing such models.
-\item We instantiate the abstract construction to provide a concrete semantics
-  in informal guarded type theory.
-\item We prove that the resulting denotational model is \emph{adequate} for the
-  graduality property: a closed term precision $M \ltdyn N : \nat$ has the
-  expected semantics, that $M$ errors or $M$ and $N$ have the same extensional
-  behavior.
-\end{enumerate}
-
-% TODO is it okay to refer to our cast calculus as *the* gradually-typed lambda
-% calculus?
-\eric{Where should we put the description of the big-step semantics and adequacy
-of the semantics? I can put it in the outline of the remainder of the paper, or
-I can postpone it until the subsequent section on the syntax of gradual typing.
-Adequacy seems to be better introduced there since at that point we have
-introduced the notion of term precision.}
-
-The remainder of the paper is organized as follows. In Section
-\ref{sec:technical-background}, we provide technical background on synthetic
-guarded domain theory and some important details about the specific variant of
-SGDT we employ as our ambient language.
-%
-In Section \ref{sec:GTLC}, we describe the syntactic theory of a standard cast
-calculus for gradual typing, which we also refer to as the \emph{gradually-typed
-lambda calculus}. We do this to establish a common syntax that will be
-interpreted into the models we subsequently construct.
-%
-In Section \ref{sec:cbpv}, we develop \emph{abstract} categorical
-semantics of gradually typed languages with the goal of organizing our
-construction of denotational models of gradual typing.
-%
-In Sections \ref{sec:concrete-term-model} and
-\ref{sec:towards-relational-model}, we set out to instantiate the abstract
-categorical semantics by constructing a concrete model in a suitable double
-category. First, in Section \ref{sec:concrete-term-model}, we give a concrete
-model for the \emph{terms} of the gradually-typed lambda calculus. This involves
-using the tools of SGDT to solve a domain equation to give the semantics of the
-dynamic type. We explain how to \emph{globalize} this model to obtain a model in
-the usual (non-guarded) dependent type theory. In the end, we obtain a
-\emph{big-step semantics} for closed terms of type $\nat$:
-%
-\[ \sem{\cdot} \colon (\cdot \vdash M : \nat) \to (\mathbb{N} \to (\sem{\nat} + 1)) \] 
-%
-where a closed term $M$ of type $\nat$ denotes a \emph{function} from
-$\mathbb{N}$ to $\sem{\nat} + 1$ which given input $i$ returns $\inl\, m$ if $M$
-terminates with value $m$ in $i$ or fewer steps, and returns $\inr\, *$ if $M$
-fails to terminate in $i$ or fewer steps. This is already a significant result,
-as it serves to verify the $\beta$ and $\eta$ laws of the equational theory.
-%
-Then in Section \ref{sec:towards-relational-model}, we attempt to extend the
-methodology of the previous section to give a denotational semantics that models
-the relational theory of the gradually-typed lambda calculus in addition to the
-terms. Interesting issues arise here arising from the \emph{intensional} nature
-of SGDT, that is, the fact that \emph{steps are observable}. We then develop
-novel solutions to address these issues. We do not complete the construction of
-the concrete relational model in this section. Instead, we choose to reformulate
-the definition of abstract categorical model to take into account the solutions
-to the issues discussed here.
-
-% Interesting issues arise here, and the novel solutions we employ to
-% address these issues inform the definition of the abstract models introduced in
-% the subsequent section.
-
-With the lessons from the concrete setting at our disposal, in Section
-\ref{sec:abstract-models}, we describe revised categorical models of gradual
-typing. We actually introduce two separate notions of model: \emph{intensional},
-where the stepping behavior of terms is observable, and \emph{extensional},
-where we consider terms that differ only in their stepping behavior to be equal.
-The extensional model is what we will ultimately use to interpret the syntax and
-relational axioms of gradually-typed languages. The principal benefit of the
-notion of intensional model is that it admits more compositional reasoning. It
-therefore serves as a convenient intermediate setting that enjoys many of the
-benefits of the usual double-categorical models of gradual typing.
-% We give the definition of intensional model as a sequence of steps each
-% building on the previous. 
-We then outline how to build an extensional model in a sequence of steps beginning
-with a significantly simpler ``input'' model. The construction proceeds in
-steps, first building up an intensional model, and then from there extracting an
-extensional model.
-%We are able to leverage the compositional reasoning afforded by the intensional
-%model to more easily meet the requirements of an extensional model.
-This gives us a more tractable way to construct models.
-Moreover, the abstract definitions and constructions in the definitions make no
-mention of SGDT, so that the places where SGDT techniques are employed are
-conveniently limited to the data passed as input to the model construction.
-%
-%In addition to organizing and facilitating the construction of a model, the abstract constructions are described independently of SGDT
-%
-%we also isolate the parts of the definition the notion of model without any explicit mention of SGDT.
-%
-Then in Section \ref{sec:concrete-model} we return to the task of defining a
-concrete model. Equipped now with the abstract constructions of the previous
-section, we provide only the data needed as an input to the construction. Many
-of the relevant definitions have already been introduced in Section
-\ref{sec:towards-relational-model} and can be provided to the construction
-as-is; the remainder of the necessary data is defined in this section.
-%
-We end the section by sketching a proof of the \emph{adequacy} of our concrete
-model with respect to graduality.
-%
-Finally, in Section \ref{sec:discussion} we conclude with discussion: comparison
-to related work, the mechanization of our concrete model in Guarded Cubical
-Agda, and directions for future research.
-
-%% \subsection{Outline of Remainder of Paper}
-
-%% \max{this is entirely out of date, update later}
-
-%% % In Section \ref{sec:overview}, we give an overview of the gradually-typed lambda
-%% % calculus and the graduality theorem.
-%% %
-%% In Section \ref{sec:technical-background}, we provide technical background on synthetic guarded domain theory.
-%% % 
-%% In Section \ref{sec:gtlc-terms}, we formally introduce the gradually-typed cast calculus
-%% for which we will prove graduality. We give a semantics to the terms using
-%% synthetic guarded domain theory.
-%% % Important here are the notions of syntactic
-%% % type precision and term precision. For reasons we describe below, we
-%% % introduce two related calculi, one in which the stepping behavior of terms is
-%% % implicit (an \emph{extensional} calculus, $\extlc$), and another where this behavior
-%% % is made explicit (an \emph{intensional} calculus, $\intlc$).
-%% %
-
-%% In Section \ref{sec:gtlc-precision}, we define the term precision ordering for
-%% the gradually-typed lambda calculus and describe our approach to assigning a
-%% denotational semantics to this ordering.
-%% This approach builds on the semantics constructed in the previous section,
-%% but extending it to the term ordering presents new challenges.
-
-%% % In Section \ref{sec:domain-theory}, we define several fundamental constructions
-%% % internal to SGDT that will be needed when we give a denotational semantics to
-%% % the gradual lambda calculus.
-%% %This includes the Lift monad, predomains and error domains.
-%% %
-%% % In Section \ref{sec:semantics}, we define the denotational semantics for the
-%% % intensional gradually-typed lambda calculus using the domain theoretic
-%% % constructions defined in the previous section.
-%% %
-%% In Section \ref{sec:graduality}, we outline in more detail the proof of graduality for the
-%% gradual lambda calculus.
-%% %
-%% In Section \ref{sec:discussion}, we discuss the benefits and drawbacks to our approach in comparison
-%% to the traditional step-indexing approach, as well as possibilities for future work.
-
-%% \subsection{Overview of Results}\label{sec:overview}
-
-%% % This section used to be part of the intro.
-%% % \subsection{Proving Graduality in SGDT}
-%% % TODO This section should probably be moved to after the relevant background has been introduced?
-
-%% % TODO introduce the idea of cast calculus and explicit casts?
-
-%% In this paper, we will utilize SGDT techniques to prove graduality for a particularly
-%% simple gradually-typed cast calculus, the gradually-typed lambda calculus.
-%% Before stating the graduality theorem, we introduce some definitions related to gradual typing.
-
-%% % Cast calculi
-%% % TODO move this to earlier?
-%% Gradually typed languages are generally elaborated to a \emph{cast calculus}, in which run-time type checking
-%% that is made explicit. Elaboration inserts \emph{type casts} at the boundaries between static and dynamic code.
-%% In particular, cast insertion happens at the elimination forms of the language (e.g., pattern matching, field reference, etc.).
-%% The original gradually typed language that is elaborated to a cast calculus is called the \emph{surface language}.
-
-%% % Up and down casts
-%% In a cast calculus, there is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that
-%% $A$ is a \emph{more precise} type than $B$.
-%% There a distinguished type $\dyn$, the \emph{dynamic type}, with the property that $A \ltdyn\, \dyn$ for all $A$.
-%% The parts of the code that are dynamically typed will be given type $\dyn$ in the cast calculus.
-%% %
-%% If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
-%% and a term $N$ of type $B$ may be \emph{down}casted to $A$, written $\dn A B N$.
-%% Upcasts always succeed, while downcasts may fail at runtime, resulting in a type error.
-%% Cast calculi have a term $\mho$ representing a run-time type error for any type $A$.
-
-%% % Syntactic term precision
-%% We also have a notion of \emph{syntactic term precision}.
-%% If $A \ltdyn B$, and $M$ and $N$ are terms of type $A$ and $B$ respectively, we write
-%% $M \ltdyn N : A \ltdyn B$ to mean that
-%% $M$ is ``syntactically more precise'' than $N$, or equivalently, $N$ is 
-%% ``more dynamic'' than $M$. Intuitively, this models the situation where $M$ and $N$
-%% behave the same except that $M$ may error more.
-%% Term precision is defined by a set of axioms that capture this intuitive notion.
-%% The specific rules for term precision depend on the specifics of the language, but
-%% many of the rules can be derived from the typing rules in a straightforward way.
-%% % Cast calculi also have a term $\mho$ representing a run-time type error for any type $A$,
-%% % and since
-%% Since $\mho$ always errors, there is a term precision rule stating that $\mho \ltdyn M$ for all $M$.
-
-%% % GTLC
-%% The gradually-typed lambda calculus is the usual simply-typed lambda calculus with a dynamic
-%% type $\dyn$ such that $A \ltdyn\, \dyn$ for all types $A$, as well as upcasts and downcasts
-%% between any types $A$ and $B$ such that $A \ltdyn B$. The complete definition will be provided in
-%% Section \ref{sec:gtlc-precision}.
-
-%% With these definitions, we can state the graduality theorem for the gradually-typed lambda calculus.
-
-%% % \begin{theorem}[Graduality]
-%% %   If $M \ltdyn N : \nat$, then either:
-%% %   \begin{enumerate}
-%% %     \item $M = \mho$
-%% %     \item $M = N = \zro$
-%% %     \item $M = N = \suc n$ for some $n$
-%% %     \item $M$ and $N$ diverge
-%% %   \end{enumerate}
-%% % \end{theorem}
-
-%% \begin{theorem}[Graduality]
-%%   Let $\cdot \vdash M \ltdyn N : \nat$. 
-%%   If $M \Downarrow v_?$ and $N \Downarrow v'_?$, then either $v_? = \mho$, or $v_? = v'_?$.
-%% \end{theorem}
-
-%% Note:
-%% \begin{itemize}
-
-%%   \item $\cdot \Downarrow$ is a relation between terms that is defined such that $M \Downarrow$ means
-%%   that $M$ terminates, either with a run-time error or a value $n$ of type $\nat$.
-
-%%   \item $v_?$ is shorthand for the syntactic representation of a term that is either equal to
-%%   $\mho$, or equal to the syntactic representation of a value $n$ of type $\nat$.
-%% \end{itemize}
-
-%% % We also should be able to show that $\mho$, $\zro$, and $\suc\, N$ are not equal.
-
-%% To prove graduality and validate the equational theory, we construct a model of the types
-%% and terms and show that all of the axioms for term precision and for equality of terms
-%% hold in this model. Modeling the dynamic type presents a challenge in the presence of a
-%% language with functions: we want the dynamic type to represent a sum of all possible types
-%% in the language, so we write down an recursive equation that the semantic object modeling
-%% dynamic type should satisfy. When the language includes function types, this equation involves a
-%% negative occurrence of the variable for which we are solving, and so the equation 
-%% does not have inductive or coinductive solutions.
-%% %
-%% To model the dynamic type, we therefore use guarded recursion to define a suitable
-%% semantic object that satisfies the unfolding isomorphism expected of the dynamic type.
-%% The key is that we do not actually get an exact solution to the equation in the style
-%% of traditional domain theory; rather, we get a ``guarded'' solution that holds ``up to a time step''.
-%% %
-%% That is, we introduce a notion of ``time'' and in the equation for the dynamic type,
-%% we guard the negative occurrences of the variable by a special operator that
-%% specifies that its argument is available ``later''.
-%% %This can be seen as a logic that internalizes the notion of step-indexing.
-%% See Section \ref{sec:technical-background} for more details on guarded type theory.
-
-%% At a high level, the key parts of our proof are as follows:
-
-%% % TODO revise this
-%% \begin{itemize}
-%%   \item Our first step toward proving graduality is to formulate a \emph{step-sensitive},
-%%   or \emph{intensional}, gradual lambda calculus, which we call $\intlc$, in which the
-%%   computation steps taken by a term are made explicit.
-%%   The ``normal'' gradual lambda calculus for which we want to prove graduality will be called the
-%%   \emph{surface}, \emph{step-insensitive}, or \emph{extensional}, gradual lambda calculus,
-%%   denoted $\extlc$.
-
-%%   \item We define a translation from the surface syntax to the intensional syntax, and
-%%   prove a theorem relating the term precision in the surface to term precision in the
-%%   intensional syntax.
-
-%%   \item We define a semantics for the intensional syntax in guarded type theory, for both the
-%%   terms and for the term precision ordering $\ltdyn$.
-
-%% \end{itemize}

### 2024-07-11 paper-new/discussion.tex
update discussion of mechanization

@@
 %
-We have formalized the key aspects in the definition of the concrete model
-discussed in Section \ref{sec:concrete-model}. For instance, we have defined
-types for predomains, error domains, and their morphisms and relations, and we
-used the guarded features to define the guarded lift + error monad and the
-dynamic type.
+We have formalized the major components of the definition of the concrete model
+described in Section \ref{sec:concrete-relational-model}: predomains, error
+domains, their morphisms, relations, and squares, and we used the guarded
+features to define the free error domain, the lock-step error ordering, and weak
+bisimilarity. We also formalized the no-go theorem from Section
+\ref{sec:concrete-term-model}.
 
-Building on these definitions, we formalized the notion of perturbations and
-push-pull structures as well as quasi-representable relations, culminating in
-the definition of value and computation types as predomains (resp. error
-domains) equipped with a monoid of perturbations with an interpretation
-homomorphism into the endomorphisms bisimilar to the identity. We constructed
-the predomain for the dynamic type via mixed induction and guarded recursion,
-defined the three relations $\inat$, $\itimes$, and $\iarr$, and proved that they are
+Building on these definitions, we formalized the notion of semantic
+perturbations and push-pull structures as well as quasi-representable relations,
+culminating in the definition of value and computation types and relations as
+introduced in Section \ref{sec:concrete-relational-model}.
+%
+% culminating in the definition of value and computation types as predomains
+% (resp. error domains) equipped with a monoid of syntactic perturbations with an
+% interpretation homomorphism into the semantic perturbations. 
+%
+We constructed the predomain for the dynamic type via mixed induction and
+guarded recursion, and defined its monoid of perturbations. We also defined the
+relations $\inat$, $\itimes$, and $\iarr$, and proved that they are
 quasi-representable.
@@
 
-Lastly, we have formalized the adequacy result discussed in Section
-\ref{sec:adequacy}. This required us to add axioms about clock quantification as
-well as axioms asserting the clock-irrelevance of booleans and natural numbers,
-since as of this writing these axioms are not built-in to Guarded Cubical Agda.
-The mechanization of adequacy also required us to formalize some essential
-lemmas involving clocks and clock-irrelevance; we may later refactor these as
-part of a ``standard library'' for Guarded Cubical Agda.
+Lastly, we have formalized the big-step term semantics discussed in Section
+\ref{sec:big-step-term-semantics} and the adequacy of the relational model
+discussed in Section \ref{sec:adequacy}. This required us to add axioms about
+clock quantification as well as axioms asserting the \emph{clock-irrelevance} of
+booleans and natural numbers, since as of this writing these axioms are not
+built-in to Guarded Cubical Agda. The mechanization of adequacy also required us
+to formalize some essential lemmas involving clocks and clock-irrelevance; we
+are considering later refactoring these as part of a ``standard library'' for
+Guarded Cubical Agda.
 

### 2024-07-11 paper-new/concrete-relational-model.tex
some small changes to relational model section

@@
 The functors $U$, $\li$, $\times$, and $\arr$ introduced in the previous section
-extend to predomains, error domains, and their relations. The functor $U$ simply
-takes the underlying predomain relation of an error domain relation. Given a
-relation $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$, the relation $c_1
-\times c_2 : (A_1 \times A_2) \rel (A_1' \times A_2')$ is defined by $((x, y),
-(x', y')) \in (c_1 \times c_2)$ if and only if $(x, x') \in c_1$ and $(y, y')
-\in c_2$. Given a relation $c : A \rel A'$ and an error domain relation $d : B
-\rel B'$, the relation $c \arr d : (A \arr B) \rel (A' \arr B')$ is the error
-domain relation defined by $(f, g) \in (c \arr d)$ if and only if for all $x :
-A$ and $x' : A'$ with $(x , x') \in c$ we have $(f(x), g(x')) \in Ud$. The fact
-that this relation preserves error and respects $\theta$ follows from the
-corresponding properties for $d$.
+act on predomains and error domains and also have actions on their morphisms,
+relations, and squares. The functor $U$ simply takes the underlying predomain
+relation of an error domain relation. Given a relation $c_1 : A_1 \rel A_1'$ and
+$c_2 : A_2 \rel A_2'$, the relation $c_1 \times c_2 : (A_1 \times A_2) \rel
+(A_1' \times A_2')$ is defined by $((x, y), (x', y')) \in (c_1 \times c_2)$ if
+and only if $(x, x') \in c_1$ and $(y, y') \in c_2$. Given a relation $c : A
+\rel A'$ and an error domain relation $d : B \rel B'$, the relation $c \arr d :
+(A \arr B) \rel (A' \arr B')$ is the error domain relation defined by $(f, g)
+\in (c \arr d)$ if and only if for all $x : A$ and $x' : A'$ with $(x , x') \in
+c$ we have $(f(x), g(x')) \in Ud$. The fact that this relation preserves error
+and respects $\theta$ follows from the corresponding properties for $d$.
 
@@
 %
-\begin{definition}[quasi-left-representable relations]
+\begin{definition}[quasi-left-representable relations]\label{def:quasi-left-representable}
 Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let $c : A
@@
 %
-\begin{definition}
+\begin{definition}[quasi-right-represntable relations]\label{def:quasi-right-representable}
 Let $(B, M_B, i_B)$ and $(B', M_{B'}, i_{B'})$ be computation objects, and let
@@
 component, while the latter does the reverse with $i_{A_2}$.
+% Could state this using the functorial action of \times
 
@@
 similarly by post-composition with $i_B$.
+% Could state this using the functorial action of \arr
 
@@
 definition of extensional square allows us to specify intermediate morphisms
-that are synchronized in the lock-step error ordering, it will be sufficient to
-construct two intensional squares involving perturbations on both sides. This
-motivates the definition of \emph{quasi-order-equivalence}.
+that are synchronized in the intensional ordering, it will be sufficient to
+construct two intensional squares between the relations, with perturbations on
+both sides. This motivates the definition of \emph{quasi-order-equivalence}.
 
 \begin{definition}[quasi-order-equivalence]\label{def:quasi-order-equivalent}
-  Let $c, c' : A \rel A'$. We say that $c$ and $c'$ are
+  Let $A$ and $A'$ be value objects, and let $c, c' : A \rel A'$ be relations
+  between the underlying predomains. We say that $c$ and $c'$ are
   \textbf{quasi-order-equivalent}, written $c \qordeq c'$, if there exist
-  perturbations $\delta^l_1, \delta^l_2 \in \pv_A$ and $\delta^r_1, \delta^r_2
-  \in \pv_{A'}$ such that there is a square $\delta^l_1 \ltsq{c}{c'} \delta^r_1$
-  and a square $\delta^l_2 \ltsq{c'}{c} \delta^r_2$.
+  syntactic perturbations $\delta^l_1, \delta^l_2 \in M_A$ and $\delta^r_1,
+  \delta^r_2 \in M_{A'}$ such that there is a square $\delta^l_1 \ltsq{c}{c'}
+  \delta^r_1$ and a square $\delta^l_2 \ltsq{c'}{c} \delta^r_2$.
 \end{definition}
-We make the analogous definition for error domain relations $d, d' : B \rel B'$.
+Given computation objects $B$ and $B$, we make the analogous definition for
+relations $d, d' : B \rel B'$ between the underlying error domains.
 
@@
 \begin{lemma}
+The following hold:
 \begin{itemize}
   \item $U(d \comp d') \qordeq U(d) \comp U(d')$
-  \item $F(c \comp c') \qordeq F(c) \comp F(c')$
+  \item $\li(c \comp c') \qordeq \li (c) \comp \li (c')$
   \item $(c \comp c') \to (d \comp d') \qordeq (c \to d) \comp (c' \to d')$
@@
 \end{itemize}
+
 \end{lemma}

### 2024-07-11 paper-new/concrete-relational-model.tex
tweaks and comments

@@
 
-In this section, we describe in detail our model of gradual typing. To manage
-the complexity, we carry out the construction in two phases, beginning with a
-relational model consisting of partially-ordered sets with a notion of
-bisimilarity and then adding requirements for a structured set of syntactic
-perturbations on each type as well as the notion of quasi-representability for
-relations that models the cast rules. We then discuss the construction of the
-dynamic type and the refined denotation of term precision that relates terms
-that can be synchronized to be in lock-step. We conclude by demonstrating the
+Next we describe in detail our compositional model of GTLC that
+satisfies graduality. To manage the complexity, we carry out the
+construction in two phases, beginning with a relational model
+consisting of partially-ordered sets with a notion of bisimilarity and
+then adding requirements for a structured set of syntactic
+perturbations on each type as well as the notion of
+quasi-representability for relations that models the cast rules. We
+then discuss the construction of the dynamic type and the refined
+denotation of term precision that relates terms that can be
+synchronized to be in lock-step. We conclude by demonstrating the
 adequacy of our model with respect to the graduality property.
@@
     \item A reflexive, symmetric ``bisimilarity'' relation $\bisim_A$.
-\end{itemize}
+\end{itemize}\max{shouldn't this be $\ltdyn$?}
 \end{definition}
@@
 
-We also equip error domains with a bisimilarity relation, and we impose a
-condition relating the $\theta$ morphism to the bisimilarity relation. More
-concretely:
+We also enhance our error domains with ordering and bisimilarity
+relation, and we impose a condition relating the $\theta$ morphism to
+the bisimilarity relation. More concretely:
 %
@@
     \item For all $x : B$ we have $\theta_B(\nxt\, x) \bisim_B x$
-\end{itemize}
+\end{itemize}\max{Don't we also need $\forall x. \mho \le x$?}
 \end{definition}
@@
   \begin{enumerate}
-     \item (Respects error): For all $y \in B'$, we have $(\mho_B, y) \in d$.
+     \item (Respects error): For all $y \in B'$, we have $(\mho_B, y) \in d$.\max{can we please use infix notation consistently for relations?}
      \item (Preserves $\theta$): For all $\tilde{x}$ in $\laterhs B$ and
@@
 %
-As with predomain relations, we can compose error domain relations. However,
-because of a technical restriction involving the interaction between
-propositional truncation and the later modality, we cannot define the
-composition of error domain relations as the composition of their underlying
-predomain relations. Instead, we define composition of error domain relations
-$d$ on $B_1$ and $B_2$ and $d'$ on $B_2$ and $B_3$ inductively to be the least
-relation containing $d$ and $d'$ that is downward closed under $\le_{B_1}$,
-upward-closed under $\le_{B_3}$, respects error, and preserves $\theta$.
-Specifically, it is defined inductively by the following rules:
+As with predomain relations, we can compose error domain
+relations. However, because of a technical restriction\max{technical
+  restriction?? isn't is simpler/correct to say that the composition
+  of underlying relations is not an error domain relation?} involving
+the interaction between propositional truncation and the later
+modality, we cannot define the composition of error domain relations
+as the composition of their underlying predomain relations. Instead,
+we define composition of error domain relations $d$ on $B_1$ and $B_2$
+and $d'$ on $B_2$ and $B_3$ inductively to be the least relation
+containing $d$ and $d'$ that is downward closed under $\le_{B_1}$,
+upward-closed under $\le_{B_3}$, respects error, and preserves
+$\theta$.  Specifically, it is defined inductively by the following
+rules:\max{what rules??}
 % %
@@
 \in (c \arr d)$ if and only if for all $x : A$ and $x' : A'$ with $(x , x') \in
-c$ we have $(f(x), g(x')) \in Ud$. The fact that this relation preserves error
+c$ we have $(f(x), g(x')) \in Ud$. The fact that this relation preserves error\max{preserves error?}
 and respects $\theta$ follows from the corresponding properties for $d$.
@@
 % Free error domain
-The functor $\li$ takes a predomain $A$ to the free error domain on $A$. We first
-define the predomain $U\li A$. The underlying set of $U\li A$ is defined to be $\li
-|A|$ (the guarded lift monad applied to $|A|$). The ordering relation is the
-lock-step error ordering introduced in the previous section (taking $A = A'$ in
-the definition), and the bisimilarity relation is the weak bisimilarity relation
-on $\li |A|$ also defined in the previous section. With these relations, the
-constructors $\eta$ and $\theta$ of the guarded lift monad are in fact morphisms
-of predomains, i.e., they are monotone and preserve bisimilarity.
+The functor $\li$ takes a predomain $A$ to the free error domain on
+$A$. We first define the predomain $U\li A$. The underlying set of
+$U\li A$ is defined to be $\li |A|$ (the guarded lift monad applied to
+$|A|$). The ordering relation is the lock-step error ordering
+introduced in the previous section , and the bisimilarity relation is
+the weak bisimilarity relation on $\li |A|$ also defined in the
+previous section. With these relations, the constructors $\eta$ and
+$\theta$ of the guarded lift monad are in fact morphisms of
+predomains, i.e., they are monotone and preserve bisimilarity.
 %
@@
 to an error domain morphism $\li A \multimap B$; with this we define the action of $\li$ on
-morphisms: Given $f : A \to A'$ we define $\li f = (\eta \circ f)^\dagger : \li A \to \li A'$.
+morphisms: Given $f : A \to A'$ we define $\li f = (\eta \circ f)^\dagger : \li A \to \li A'$.\max{what is the point of this sentence?}
 
@@
 denotation of value types in our final model. Likewise, the triple $(B, M_B,
-i_B)$ will be called a \textbf{computation object}.
+i_B)$ will be called a \textbf{computation object}.\max{make this a proper \textit{Definition} and define morphisms as well}
 
+\max{idea: move this after the quasi-representability part and say that this structure is what's needed to make them compose}
 We must impose an additional condition on predomain relations $c$ and error
@@
 \begin{definition}
-    Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let $c
-    : A \rel A'$ be a predomain relation. A \textbf{push-pull structure} for $c$
-    (with respect to $M_A$ and $M_{A'}$) consists of monoid homomorphisms $\push
-    : M_A \to M_A'$ and $\pull : M_A' \to M_A$ such that for all $m_A \in M_A$
-    there is a square $i_A(m_A) \ltsq{c}{c} i_{A'}(\push\, m_A)$ and for all
-    $m_{A'} \in M_{A'}$ there is a square $i_A(\pull\, m_{A'}) \ltsq{c}{c}
-    i_{A'}(m_{A'})$.
+    Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects,
+    and let $c : A \rel A'$ be a predomain relation. A
+    \textbf{push-pull structure} for $c$ (with respect to $M_A$ and
+    $M_{A'}$) consists of monoid homomorphisms $\push : M_A \to M_A'$
+    and $\pull : M_A' \to M_A$ such that for all $m_A \in M_A$ there
+    is a square $i_A(m_A) \ltsq{c}{c} i_{A'}(\push\, m_A)$ and for all
+    $m_{A'} \in M_{A'}$ there is a square $i_A(\pull\, m_{A'})
+    \ltsq{c}{c} i_{A'}(m_{A'})$. A push-pull structure for error
+    domain relations is defined similarly.
 \end{definition}
-%
-We make the analogous definition for computation objects.
 
@@
 (2) $\dnr$: $i_B(\dellp_d) \ltsq{d}{\le_{B}} p$.
+Quasi-right-representability for predomain relations is defined similarly.
 \end{definition}
-%
-We make the analogous definition of quasi-right-representability for predomain
-relations $c$.
 
-Given $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$, it makes sense to refer to the
-quasi-representability of a predomain relation $c : A \rel A'$. However,
-quasi-representable relations do not in general compose: in order to construct
+\max{see above comment about reordering these concepts}
+do not in general compose: in order to construct
 the needed squares for the composition of two relations, we require that the
@@
 \subsubsection{Constructions involving Perturbations}\label{sec:perturbation-constructions}
-
+\max{this section needs prose}
 We first discuss the action of $\li$ on objects. Given a value object $(A, M_A,
@@
 
-The ordering $\le_D$ is given by:
+The ordering $\le_D$ is given by:\max{this is kind of obvious IMO maybe it can go in the appendix}
 %
@@
 One can verify that this defines a homomorphism from $M_D \to \{ f : D \to D : f
-\bisim \id \}$. We claim that the three relations $\inat$, $\itimes$, and
+\bisim \id \}$. We claim\max{why we claim? Just say it's true!} that the three relations $\inat$, $\itimes$, and
 $\iarr$ satisfy the push-pull property. As an illustrative case, we establish
@@
 
-We next claim that the relations $\inat$, $\itimes$, and $\iarr$ are
+We next claim\max{claim} that the relations $\inat$, $\itimes$, and $\iarr$ are
 quasi-left-representable, and that their lifts are quasi-right-representable.
@@
 Recall the defintion of the downcast for $\iarr$ given in Section
-\ref{sec:term-interpretation}.
+\ref{sec:term-interpretation}.\max{why is there a paragraph break here?}
 
@@
 these three relations. In the case of $\iarr$, we have that the property holds
-up to a delay.
+up to a delay.\max{retraction property? has this ever come up anywhere in the paper before? Should it?}
 
@@
 %
-\[ f \bisim_{\li \mathbb{N}} f' \le_{\li \mathbb{N}} g' \bisim_{\li \mathbb{N}} g. \]
+\[ f \bisim_{\li \mathbb{N}} f' \le_{\li \mathbb{N}} g' \bisim_{\li \mathbb{N}} g. \]\max{$\le$? I thought it was $\ltdyn$}
 %
@@
 %
-Unfortunately, this is actually not provable! Roughly speaking, the issue is
-that this is a ``global'' result, and it is not possible to prove such results
-in the guarded setting. In particular, if we tried to prove the above result in
+Unfortunately\max{it's not unfortunate}, this is actually not provable! Roughly speaking, the issue is
+that this is a ``global''\max{what does ``global'' mean?} result, and it is not possible to prove such results
+in the guarded setting\max{isn't everything we do in a ``guarded setting''?}. In particular, if we tried to prove the above result in
 the guarded setting we would run into a problem where we would have a natural
@@
 Thus, to prove our adequacy result, we need to pass to leave the guarded
-setting. We have seen in Section \ref{sec:big-step-term-semantics} that clock
+setting\max{``pass to leave the guarded setting''?}. We have seen in Section \ref{sec:big-step-term-semantics} that clock
 quantification makes this possible.
@@
 Now consider the predomain $\mathbb{N}$ with ordering and bisimilarity both the
-equality relation. We can define a ``global'' version of the lock-step error
+equality relation. We can define a ``global''\max{stop using scare quotes around ``global''} version of the lock-step error
 ordering and the weak bisimilarity relation on elements of the $\li^{gl}
@@
 On the other hand, we can define coinductively a ``lock-step error ordering''
-relation on $\delay(\mathbb{N} + {\mho})$:
+relation on $\delay(\mathbb{N} + {\mho})$:\max{do these intermediate details matter for the main body of the paper?}
 %
@@
 from the definitions of the relations.
+\max{this doesn't actually relate to our notion of big step semantics}
 

### 2024-07-11 paper-new/discussion.tex
tweaks and comments

@@
 with a statically typed language and derive a gradually typed language
-that satisfies the gradual guarantee. The main downside to these
+that satisfies the graduality. The main downside to these
 approaches lies in their inflexibility: since the process in entirely
@@
 However, they do not give a denotational semantics to term precision and it is unclear
-how to prove the gradual guarantee in this setting.
+how to prove the graduality in this setting.
 The work includes a formalization of the syntactic model in Guarded Cubical Agda.
@@
 relations $\inat$, $\itimes$, and $\iarr$, and proved that they are
-quasi-representable.
+quasi-representable.\max{need to say what's not formalized}
 
@@
 are considering later refactoring these as part of a ``standard library'' for
-Guarded Cubical Agda.
+Guarded Cubical Agda.\max{say these axioms are taken from prior work and cite that}
 
@@
 
+\max{shallow}
 Working internally to guarded type theory reduces the overhead of
@@
 \subsection{Synthetic Ordering}
+\max{cut this subsection if we need space}
 A key to managing the complexity of our concrete construction is in
@@
 
+\max{Expand on this, e.g. what some of the challenges might be and what would be reusable}
 In the future, we plan to apply our approach to give a denotational

### 2024-07-11 paper-new/towards-relational-model.tex
tweaks/comments to towards relational model section

@@
 These relations are then used to model term precision. For closed programs $M \ltdyn M' :
-c$ where $c : A \ltdyn A'$ we have that these denote elements of the
-free error domain $\sem{M},\sem{M'} : U\li \sem{A}$ and we model this
-as a relation holding $\sem{M} (U\li\sem{c}) \sem{M'}$ where
+c$ where $c : A \ltdyn A'$, the terms denote elements of the
+free error domain $\sem{M},\sem{M'} : U\li \sem{A}$ and we model the term precision semantically as the relation holding between those denoted elements $\sem{M} \binrel{(U\li\sem{c})} \sem{M'}$ where
 $U\li\sem{c}$ is a relational lifting of $U\circ\li$.
@@
 they denote functions, not just elements. We model that as not simply
-being elements related by a relation but in a typical
-logical-relations style of \emph{preserving} relatedness: if they are
-passed in inputs related by the domain relations then their output is
-related by the codomain relation.
+elements being related but in a typical logical-relations style as
+\emph{preserving} relatedness: if they are passed in inputs related by
+the domain relations then their output is related by the codomain
+relation.
 %
@@
 of a \emph{square}, which specifies a relation between two monotone
-functions. The definition of square is as follows. Let $A_i,
-A_o, A_i', A_o'$ be partially-ordered sets. Let $c_i : A_i \rel A_i'$
-and $c_o : A_o \rel A_o'$ be poset relations, and let $f : A_i \to
-A_o$ and $g : A_i' \to A_o'$ be monotone functions. We say that $f
+functions. The definition of square is as follows. Let $A_i, A_o,
+A_i', A_o'$ be partially-ordered sets. Let $c_i : A_i \rel A_i'$ and
+$c_o : A_o \rel A_o'$ be poset relations, and let $f : A_i \to A_o$
+and $g : A_i' \to A_o'$ be monotone functions. We say that $f
 \ltsq{c_i}{c_o} g$ if for all $x : A_i$ and $y : A_i'$ with $(x, y)
-\in c_i$, we have $(f(x), g(y)) \in c_o$. We visualize this situation
-as follows:
+\in c_i$, we have $(f(x), g(y)) \in c_o$. We call this a \emph{square}
+because we visualize this situation as follows:
 %
@@
 We note that for any $c$, there is a ``vertical identity'' square $\id
-\ltsq{c}{c} \id$, as can be seen by unfolding the definition. Similarly, for any
-montone function $f : A_i \to A_o$, there is a ``horizontal identity'' square $f
-\ltsq{\le_{A_i}}{\le_{A_o}}$ by the definition of monotone function.
+\ltsq{c}{c} \id$, as can be seen by unfolding the
+definition. Similarly, for any montone function $f : A_i \to A_o$,
+there is a ``horizontal identity'' square $f
+\ltsq{\le_{A_i}}{\le_{A_o}} f$ arising from the fact that $f$ is
+monotone.
 %
-We can \emph{compose} squares vertically, i.e., if we have $f \ltsq{c_1}{c_2}
-f'$ and $g \ltsq{c_2}{c_3} g'$ then we obtain a square $g \circ f
-\ltsq{c_1}{c_3} g' \circ f'$. Likewise, we can compose squares horizontally: If
-$f \ltsq{c_i}{c_o} g$ and $g \ltsq{c_i'}{c_o'} h$, then we obtain a square $f
-\ltsq{c_i \comp c_i'}{c_o \comp c_o'} h$. Horizontal composition of squares
-corresponds to transitivity of term precision.
+We can \emph{compose} squares vertically, i.e., if we have $f
+\ltsq{c_1}{c_2} f'$ and $g \ltsq{c_2}{c_3} g'$ then we obtain a square
+$g \circ f \ltsq{c_1}{c_3} g' \circ f'$. Likewise, we can compose
+squares horizontally: If $f \ltsq{c_i}{c_o} g$ and $g
+\ltsq{c_i'}{c_o'} h$, then we obtain a square $f \ltsq{c_i \comp
+  c_i'}{c_o \comp c_o'} h$. Horizontal composition of squares
+corresponds to transitivity of term precision. While we do not dwell
+on categorical abstractions in this work, we note that this structure
+of functions, relations and squares forms a locally thin \emph{double
+category} as used in previous work \cite{newlicata}.
 
@@
 that the relation is a kind of \emph{graph} of the function in that $x
-\binrel{\sem c} y$ if and only if $\sem{\upc c}(x) \leq_{A_2} y$.
+\binrel{\sem c} y$ if and only if $\sem{\upc c}(x) \leq_{A_2} y$ \cite{shulmandoublecategories}.
 

### 2024-07-11 paper-new/appendix.tex
more comments
work on appendix

--- 2024-01-27 paper-new/appendix.tex
@@
 
+\max{there is not definition of a CBPV model...}
 There are two relevant notions of \emph{morphism} of CBPV models:
@@
 
-\section{Details of the Construction of an Extensional Model}
-
-In Section \ref{sec:extensional-model-construction}, we outline the construction
-of an extensional model of gradual typing starting from a step-1 intensional model.
-In this section, we provide the details for each of the constructions mentioned there.
-
-\subsection{Constructing a Model with Perturbations}
-
-The goal of this section is to prove the following lemma:
-
-\begin{lemma}\label{lem:step-1-model-to-step-2-model}
-Let $\mathcal M$ be a \hyperref[def:step-1-model]{step-1 intensional model}.
-
-Then we can construct a \hyperref[def:step-2-model]{step-2 intensional model}.
-\end{lemma}
-
-We begin with a definition and some lemmas that will be useful in the construction
-of the model.
-
-\begin{definition}
-  Let $c : A \rel A'$ be a value relation of $\mathcal M$.
-  Let $P_A$ be a monoid of perturbations on $A$ and $P_{A'}$ a monoid of perturbations on $A'$.
-  A \emph{push-pull structure} $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$ consists of:
-  \begin{itemize}
-    \item A function $\push : P_A \to P_{A'}$ 
-    such that for all $\delta^l \in P_A$ we have $\delta^l \ltdyn_c^c \push(\delta^l)$.
-    \item A function $\pull : P_{A'} \to P_A$
-    such that for all $\delta_r \in P_{A'}$ we have $\pull(\delta^r) \ltdyn_{c}^c \delta^r$.
-  \end{itemize}
-
-  We define a push-pull structure $\pie_d$ for $d : B \rel B'$ with respect to perturbation monoids
-  $P_B$ and $P_{B'}$ in an analogous manner.
-
-  % \eric{do we need that $\delta^*$ is preserved by push and pull?}
-  % We stipulate the additional requirement that if $\delta_A^*$ is in $P_B$ 
-  % (where $\delta_A^*$ is the distinguished perturbation required to be in every $\ef(FA, FA)$),
-  % then $\push(\delta_A^*) = \delta_{A'}^*$ and likewise $\pull(\delta_{A'}^*) = \delta_A^*$.
+\section{Omitted Proofs Section \ref{sec:towards-relational-model}}
+
+\begin{theorem}
+  Let $R$ be a binary relation on the free error domain $U(\li A)$. If
+  $R$ satisfies the following properties
+  \begin{enumerate}
+  \item Transitivity
+  \item $\theta$-congruence: If $\later_t (\tilde{x}_t \binrel{R} \tilde{y}_t)$, then $\theta(\tilde{x}) \binrel{R} \theta(\tilde{y})$.
+  \item Right step-insensitivity: If $x \binrel{R} y$ then $x \binrel{R} \delta y$.
+  \end{enumerate}
+  Then for any $l : U(\li A)$, we have $l \binrel{R} \Omega$. If $R$ is left
+  step-insensitive instead then $\Omega \binrel{R} x$.
+\end{theorem}
+\begin{proof}
+  By \lob-induction: we assume that $\laterhs (l \binrel{R} \Omega)$, and we
+  show $(l \binrel{R} \Omega)$. We have $l \binrel{R} \delta l$ by right
+  step-insensitivity.
+  %
+  By transitivity, it will therefore suffice to show that $\delta l \binrel{R}
+  \Omega$.
+  %
+  Then since $\Omega = \delta \Omega$, it suffices to show $\delta l \binrel{R}
+  \delta\Omega$.
+  %
+  By $\theta$-congruence, it suffices to show that
+  $\later_t [(\nxt\, l)_t \binrel{R} (\nxt\, \Omega)_t]$, i.e., $\later (l \binrel{R} \Omega)$,
+  which is our induction hypothesis.
+
+  The proof when $R$ is left step-insensitive is symmetric.
+\end{proof}
+
+\section{Details of the Relational Model Construction}
+
+In Section \ref{sec:concrete-relational-model} we described our model of gradual
+typing. We omitted several technical proofs involving the well-definedness of
+composition of relations and the actions of functors. We provide those proofs in
+this section.
+
+% We note that our approach in this section will be differ from the body of the
+% paper in two ways:
+
+% First, we will work \emph{abstractly} with a notion of model of gradual typing
+% defined in the setting of an arbitrary double category.
+% %
+% Second, we build up the abstract notion of model in several phases, and layer
+% the constructions in a modular way so that we can isolate the essential parts of
+% the proofs and abstract away unnecessary details about previous parts of the
+% construction.
+
+% The constructions we carry out in this section specialize to the model we
+% describe in the body of the paper. We find, however, that organizing things in
+% this manner makes the arguments easier to follow.
+
+\subsection{Free Composition of Error Domain Relations}
+\begin{definition}\label{def:free-comp-ed-rel}
+  Let $d : B_1 \rel B_2$ and $d' : B_2 \rel B_3$ be error domain relations. We
+  define the composition $dd'$ inductively by the following rules:
+  \begin{mathpar}
+      \inferrule*[right = Comp]
+      {b_1 \mathbin{d} b_2 \and b_2 \mathbin{d'} b_3}
+      {b_1 \mathbin{d \relcomp d'} b_3}
+
+      \inferrule*[right = DnClosed]
+      {b_1' \le_{B_1} b_1 \and b_1 \mathbin{d \relcomp d'} b_3}
+      {b_1' \mathbin{d \relcomp d'} b_3}
+
+      \inferrule*[right = UpClosed]
+      {b_1 \mathbin{d \relcomp d'} b_3 \and b_3 \le_{B_3} b_3'}
+      {b_1 \mathbin{d \relcomp d'} b_3'}
+
+      \inferrule*[right = PresErr]
+      { }
+      {\mho_{B_1} \mathbin{d \relcomp d'} b_3}
+
+      \inferrule*[right = PresTheta]
+      {\later_t( \tilde{b_1} \mathbin{d \relcomp d'} \tilde{b_3} ) }
+      {\theta_{B_1}(\tilde{b_1}) \mathbin{d \relcomp d'} \theta_{B_3}(\tilde{b_3}) }
+  \end{mathpar}
 \end{definition}
+
+\subsection{The Dynamic Type}
+The ordering $\le_D$ on the predomain $D$ is defined by
+\begin{align*}
+  \inat(n) \le \inat(n') 
+      &\iff n = n' \\
+  \itimes (d_1, d_2) \le \itimes (d_1', d_2')
+      &\iff d_1 \le_D d_2 \text{ and } d_1' \le_D d_2'\\
+  \iarr(\tilde{f}) \le \iarr(\tilde{f'}) 
+      &\iff \later_t(\tilde{f}_t \le \tilde{f'}_t),
+\end{align*}
+
+and the bisimilarity relation is analogous.
+
+\subsection{Lemmas about Perturbations}
+
+The goal of this section is to prove the following lemmas:
 
@@
 \begin{lemma}\label{lem:push-pull-comp}
-  Let $c : A \rel A'$ and $c' : A' \rel A''$ be value relations,
-  and let $P_A, P_{A'}$ and $P_{A''}$ be monoids of perturbations.
-  Given a push-pull structure $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$, and
-  and a push-pull structure $\piv_{c'}$ for $c'$ with respect to $P_{A'}$ and $P_{A''}$,
-  we can define a push-pull structure $\piv_{c \comp c'}$ for $c \comp c'$.
-
-  Likewise, we can define a push-pull structure for the composition of computation relations.
+  Let $(A_1, M_{A_1}, i_{A_1})$, $(A_2, M_{A_2}, i_{A_2})$, and $(A_3, M_{A_3},
+  i_{A_3})$ be value objects, and let $c : A_1 \rel A_2$ and $c' : A_2 \rel A_3$
+  be predomain relations.
+
+  Given a push-pull structure $\Pi_c$ for $c$ and $\Pi_{c'}$ for $c'$, we can
+  define a push-pull structure $\Pi_{c \comp c'}$ for $c \comp c'$.
+
+  Likewise, we can define a push-pull structure for the composition of error
+  domain relations.
 \end{lemma}
 \begin{proof}
-  We define $\piv_{c \comp c'}$ as the following push-pull structure:
-
-  \begin{itemize}
-    \item $\push_{c \comp c'} = \push_{c'} \circ \push_{c}$
-    \item $\pull_{c \comp c'} = \pull_{c} \circ \pull_{c'}$
-  \end{itemize}
-
-  We observe that the required squares exist for both push and pull.
-  In particular, for push we have that $\delta^l \ltdyn_c^c \push_c(\delta^l)$
-  using the push property for $c$, and then using the push property for $c'$ we have
-  $\push_c(\delta^l) \ltdyn_{c'}^{c'} \push_{c'}(\push_c(\delta^l))$.
-  We can then compose these squares horizontally to obtain the desired square.
-  The pull property follows similarly.
-
-  The push-pull structure for the composition of computation relations is defined analogously.
+  We define $\piv_{c \comp c'}$ as follows: We first define
+  %
+  $\push_{c \comp c'} = \push_{c'} \circ \push_{c}$ and 
+  $\pull_{c \comp c'} = \pull_{c} \circ \pull_{c'}$.
+  %
+  We then observe that the required squares exist for both push and pull. In
+  particular, for push we have that $i_{A_1}(m_{A_1}) \ltdyn_c^c
+  i_{A_2}(\push_c(m_{A_1}))$ using the push property for $c$, and then using the
+  push property for $c'$ we have $i_{A_2}(\push_c(m_{A_1})) \ltdyn_{c'}^{c'}
+  i_{A_3}(\push_{c'}(\push_c(m_{A_1})))$. We can then compose these squares
+  horizontally to obtain the desired square. The pull property follows
+  similarly.
+
+  The push-pull structure for the composition of computation relations is
+  defined analogously.
 \end{proof}
@@
 \begin{lemma}\label{lem:push-pull-U-F}
-  Let $c : A \rel A'$ and let $P_A$ and $P_{A'}$ be monoids of perturbations.
-  Given a push-pull structure $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$,
-  we can define a push-pull structure $\pie_{Fc}$ for $F(c)$ with respect to
-  $\mathbb{N} \times P_{A}$ and $\mathbb{N} \times P_{A'}$.
-
-  Likewise, from a push-pull structure for $d$ with respect to $P_B$ and $P_{B'}$,
-  we can define a push-pull structure for $Ud$ with respect to $\mathbb{N} \times P_B$
-  and $\mathbb{N} \times P_{B'}$.
+
+  Let $A$ and $A'$ be value objects and let $c : A \rel A'$ be a relation on the
+  underlying predomains. Given a push-pull structure $\Pi_c$ for $c$ we can
+  define a push-pull structure $\Pi_{\li c}$ for $\li c$ with respect to $\li A$
+  and $\li A'$.
+
+  Likewise, let $B$ and $B'$ be computation objects and let $d : B \rel B'$ be
+  an error domain relation. Given a push-pull structure for $d$ with respect to
+  $B$ and $B'$, we can define a push-pull structure for $Ud$ with respect to
+  $UB$ and $UB'$.
 \end{lemma}
@@
 
-  We define $\push_{Fc} : \mathbb{N} \times P_{A} \to \mathbb{N} \times P_{A'}$ by
-
-  \[ \push_{Fc}(n, a) = (n, \push_c(a)). \]
-
-  Then we need to build the following square:
-
-  % https://q.uiver.app/#q=WzAsNCxbMCwwLCJGQSJdLFsxLDAsIkZBJyJdLFswLDEsIkZBIl0sWzEsMSwiRkEnIl0sWzIsMywiRmMiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIkZjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwyLCIoXFxkZWx0YV97RkF9XiopXm4gXFxjaXJjIFxccHRiX0EoYSkiLDJdLFsxLDMsIihcXGRlbHRhX3tGQSd9XiopXm4gXFxjaXJjIFxccHRiX3tBJ30oXFxwdXNoX2MoYSkpIl1d
+  We define $\push_{\li c} : \mathbb{N} \oplus M_{A} \to \mathbb{N} \oplus
+  M_{A'}$ by the universal property of the coproduct of monoids. In particular,
+  it suffices to define a homomorphism of monoids $\mathbb{N} \to \mathbb{N}
+  \oplus M_{A'}$ and $M_{A} \to \mathbb{N} \oplus M_{A'}$.
+
+  The former is simply $\inl$ and the latter is $\inr \circ \push_c$.
+
+  Then to establish the push property, it suffices by the universal property of
+  the coproduct, and the fact that $\mathbb{N}$ is the free monoid on one
+  generator, to show that the following two squares exist:
+
+% https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXGxpIEEiXSxbMSwwLCJcXGxpIEEnIl0sWzAsMSwiXFxsaSBBIl0sWzEsMSwiXFxsaSBBJyJdLFswLDEsIlxcbGkgYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzIsMywiXFxsaSBjIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwyLCJcXGRlbHRhXipfQSIsMl0sWzEsMywiXFxkZWx0YV4qX3tBJ30iXV0=
 \[\begin{tikzcd}[ampersand replacement=\&]
-    FA \& {FA'} \\
-    FA \& {FA'}
-    \arrow["Fc", "\shortmid"{marking}, no head, from=2-1, to=2-2]
-    \arrow["Fc", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-    \arrow["{(\delta_{FA}^*)^n \circ \ptb_A(a)}"', from=1-1, to=2-1]
-    \arrow["{(\delta_{FA'}^*)^n \circ \ptb_{A'}(\push_c(a))}", from=1-2, to=2-2]
+	{\li A} \& {\li A'} \\
+	{\li A} \& {\li A'}
+	\arrow["{\li c}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+	\arrow["{\delta^*_A}"', from=1-1, to=2-1]
+	\arrow["{\delta^*_{A'}}", from=1-2, to=2-2]
+	\arrow["{\li c}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
 \end{tikzcd}\]
 
-  This is obtained by composing the square $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$
-  with the square corresponding to the push property for $B$
-
-  % Given $n \in \mathbb{N}$ we define the push to be $n$.
-  % Then the required square is $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$,
-  % which exists by the definition of a step-1 model.
-  % On the other hand, given $a \in P_{A}$, we define the push to be $\push_c(a)$
-  % and then the required exists by the push property for $c$.
-  % Defining $\pull_{Fc}$ is symmetric.
-
-  The proof for $Ud$ is analogous.
+% https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXGxpIEEiXSxbMCwxLCJcXGxpIEEiXSxbMSwwLCJcXGxpIEEnIl0sWzEsMSwiXFxsaSBBJyJdLFswLDIsIlxcbGkgYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEsMywiXFxsaSBjIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJcXGxpKGlfQShtX0EpKSIsMl0sWzIsMywiXFxsaSAoaV97QSd9KFxccHVzaF9jKG1fQSkpKSJdXQ==
+\[\begin{tikzcd}[ampersand replacement=\&]
+	{\li A} \& {\li A'} \\
+	{\li A} \& {\li A'}
+	\arrow["{\li c}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+	\arrow["{\li(i_A(m_A))}"', from=1-1, to=2-1]
+	\arrow["{\li (i_{A'}(\push_c(m_A)))}", from=1-2, to=2-2]
+	\arrow["{\li c}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+\end{tikzcd}\]
+
+where $\delta* = (\delta \circ \eta)^\dagger$
+
+The first square exists by the fact that the monadic extension operation
+$-^\dagger$ is monotone in its argument, and the fact that there is a
+square $\delta_A \circ \eta_A \ltsq{}{} \delta_{A'} \circ \eta_{A'}$.
+
+The second square exists by the push property for $c$ and the fact that $\li$
+acts on squares.
+
+The proof for the pull property is dual, and the construction of a push-pull
+structure for $Ud$ is analogous.
 \end{proof}
@@
 %%%%%%%%%%%%
-\begin{lemma}\label{lem:push-pull-times}
-  Let $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$, and let $P_{A_1}$,
-  $P_{A_2}$, $P_{A_1'}$ and $P_{A_2'}$ be monoids of perturbations.
-  Given push-pull structures 
-  $\piv_{c_1}$ with respect to $P_{A_1}$ and $P_{A_1'}$, and 
-  $\piv_{c_2}$ with respect to $P_{A_2}$ and $P_{A_2'}$,
-  we can define a push-pull structure $\piv_{c_1 \times c_2}$
-  with respect to $P_{A_1} \times P_{A_2}$ and $P_{A_1'} \times P_{A_2'}$.
-
-  % Likewise, given push-pull structures 
-  % $\pie_{Fc_1}$ with respect to $F(P_{A_1})$ and $F(P_{A_1'})$, and
-  % $\pie_{Fc_2}$ with respect to $F(P_{A_2})$ and $F(P_{A_2'})$,
-  % we can define a push-pull structure $\pie_{F(c_1 \times c_2)}$
-  % with respect to $F(P_{A_1} \times P_{A_2})$ and $F(P_{A_1'} \times P_{A_2'})$.
+\begin{lemma}\label{lem:push-pull-times} Let $A_1$, $A_1'$, $A_2$, and $A_2'$ be
+  value objects. Let $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$ be
+  relations on the underlying predomains. Given push-pull structures $\Pi_{c_1}$
+  for $c_1$ and $\Pi_{c_2}$ for $c_2$ we can define a push-pull structure
+  $\Pi_{c_1 \times c_2}$ for $c_1 \times c_2$.
 \end{lemma}
 \begin{proof}
-We define the push function for $\piv_{c_1 \times c_2}$ by 
-$\push(a_1, a_2) = (\push_{c_1}(a_1), \push_{c_2}(a_2))$ and likewise for $\pull$.
-The push property holds because it holds for each component.
-The pull property is established similarly.
+Recall that the monoid of syntactic perturbations for $A_1 \times A_2$ is
+$M_{A_1} \oplus M_{A_2}$ and the monoid for $A_1' \times A_2'$ is $M_{A_1'}
+\oplus M_{A_2'}$. Thus to define the push homomorphism for $c_1 \times c_2$, it
+suffices by the universal property of the coproduct of monoids to define a
+homomorphism $M_{A_1} \to M_{A_1'} \oplus M_{A_2'}$ and $M_{A_2} \to M_{A_1'}
+\oplus M_{A_2'}$. The former is $\inl \circ \push_{c_1}$, and the latter is
+$\inr \circ \push_{c_2}$. The pull homomorphism is defined similarly.
+
+For the push property, we need to show that there is a square 
+$i_{A_1 \times A_2}(m_1) \ltsq{}{} i_{A_1' \times A_2'}(\push_{c_1 \times c_2}(m_1))$.
+%
+By the universal property of the coproduct, it suffices to consider two cases: a
+perturbation $m_1 \in M_{A_1}$ and a perturbation $m_2 \in M_{A_2}$. In the
+first case, using the definition of the interpretation homomorphism for the
+product, the square becomes $(i_{A_1}(m_1) \times \id_{A_2}) \ltsq{}{}
+(i_{A_1'}(\push_{c_1}(m_1)) \times \id_{A_2'})$.
+%
+Then by the action of $\times$ on squares, it suffices to show that there are
+squares $i_{A_1}(m_1) \ltsq{}{} i_{A_1'}(\push_{c_1}(m_1))$ and $\id_{A_2}
+\ltsq{}{} \id_{A_2'}$. The former is the push property for $c_1$, and the latter
+is an identity square.
+
+The second case, i.e., $m_2 \in M_{A_2}$, is symmetric, and the proof of the
+pull property is dual.
 \end{proof}
@@
 %%%%%%%%%
-\begin{lemma}\label{lem:push-pull-arrow}
-  Let $c : A \rel A'$ and $d : B \rel B'$.
-  Given push-pull structures $\piv_c$ for $c$ and $\pie_d$ for $d$,
-  we can define a push-pull structure $\pie_{c \arr d}$ for $c \arr d$.
-
-  % Given push-pull structures $\pie_{Fc}$ for $Fc$ and $\piv_{Ud}$ for $Ud$,
-  % we can define a push-pull structure $\piv_{U(c \arr d)}$ for $U(c \arr d)$
-  % with respect to the monoids $(P_{FA})^{op} \oplus P_B$
+\begin{lemma}\label{lem:push-pull-arrow} Let $A$ and $A'$ be value objects and
+  $B$ and $B'$ be computation objects. Let $c : A \rel A'$ be a relation on the
+  underlying predomains and $d : B \rel B'$ a relation on the underlying error
+  domains. Given push-pull structures $\Pi_c$ for $c$ and $\Pi_d$ for $d$, we
+  can define a push-pull structure $\Pi_{c \arr d}$ for $c \arr d$.
 \end{lemma}
 \begin{proof}
-  Follows from the functorial action of $\arr$ on perturbations and squares.
-  % and the latter uses the functorial action of $\tok$ on perturbations and squares.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%
-% Model Construction %
-%%%%%%%%%%%%%%%%%%%%%%
-We now proceed with the construction of the model:
-
-    Define a step-2 model $\mathcal M'$ as follows:
-    \begin{itemize}
-      \item Value objects are pairs consisting of:
-      \begin{itemize}
-        \item A value object $A$ in $\vf$.
-        \item A commutative monoid $P_A$ of perturbations
-        and a monoid homomorphism $\ptb_A : P_A \to \{ f \in \vf(A, A) \mid f \bisim \id_A \}$.
-
-      \end{itemize}
-
-      \item Computation objects are pairs consisting of:
-      \begin{itemize}
-        \item A computation object $B$ in $\ef$.
-        \item A commutative monoid $P_B$ of perturbations and a monoid homomorphism
-        $\ptb_B : P_B \to \{ \phi \in \ef(B, B) \mid \phi \bisim \id_B \}$.
-      \end{itemize}
-
-      \item Morphisms are given by morphisms of the underlying objects in $\vf$ and $\ef$, respectively.
-      %, i.e.,
-      % \[ \vf'((A, P_A, \ptb_A, P^K_A, \ptbk_A), (A', P_{A'}, \ptb_{A'}, P^K_{A'}, \ptbk_{A'})) = \vf(A, A') \]
-      %
-      % and likewise for computations.
-
-      \item The objects of $\vsq'$ (i.e., the value relations) are pairs consisting of:
-      \begin{itemize}
-        \item A value relation $c \in \vsq$.
-        \item A push-pull structure $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$.
-        %\item A push-pull structure $\pie_{Fc}$ for $Fc$ with respect to $P^K_A$ and $P^K_{A'}$.
-      \end{itemize}
-
-      The objects of $\esq'$ are defined analogously, i.e.,
-      \begin{itemize}
-        \item A computation relation $d \in \esq$.
-        \item A push-pull structure $\pie_d$ for $d$ with respect to $P_B$ and $P_{B'}$.
-        %\item A push-pull structure $\piv_{Ud}$ for $Ud$ with respect to $P^K_B$ and $P^K_{B'}$.
-      \end{itemize}
-
-      \item The morphisms of $\vsq'$ and $\esq'$ are given by the morphisms of $\vsq$ and $\esq$.
-
-      \item We define composition of relations $(c, \piv_c)$ and $(c', \piv_{c'})$
-      as $(c \comp c', \piv_{c \comp c'})$ where $\piv_{c \comp c'}$ is as in Lemma
-      \ref{lem:push-pull-comp}, and likewise for computation relations.
-
-    \end{itemize}
-
-      % Functors \times, +, F, U, arrow
-      Now we define the actions of the functors:
-      \begin{itemize}
-
-      \item We define $\times$ on objects by
-
-      \[ (A_1, P_{A_1}, \ptb_{A_1}) \times (A_2, P_{A_2}, \ptb_{A_2}) =
-        (A_1 \times A_2, P_{A_1} \times P_{A_2}, \ptb_{A_1 \times A_2}) \]
-
-      where $\ptb_{A_1 \times A_2}(p_1, p_2) = \ptb_{A_1}(p_1) \times \ptb_{A_2}(p_2)$.
-
-      Using Lemma \ref{lem:push-pull-times}, we define $\times$ on relations by
-
-      \[ (c_1, \piv_{c_1}), (c_2 \piv_{c_2}) = 
-        (c_1 \times c_2, \piv_{c_1 \times c_2}). \]
-
-      \item We define $F$ on objects by
-
-      \[ F(A, P_A, \ptb_A) = (FA, \mathbb{N} \times P_A, \ptb_{FA}), \]
-
-      where we define $\ptb_{FA}(n, a) = (\delta_{FA}^*)^n \circ F(\ptb_A(a))$.
-
-      Using Lemma \ref{lem:push-pull-U-F}, we define $F$ on relations by
-
-      \[ F(c, \piv_c) = (Fc, \pie_{Fc}). \]
-
-      \item We define $U$ on objects by
-
-      \[ U(B, P_B, \ptb_B) = (UB, \mathbb{N} \times P_B, \ptb_{UB}), \]
-
-      where we define $\ptb_{UB}(n, b) = (\delta_{UB}^*)^n \circ U(\ptb_B(b))$.
-
-      Using Lemma \ref{lem:push-pull-U-F}, we define $U$ on relations by
-
-      \[ U(d, \pie_d) = (Ud, \piv_{Ud}).  \]
-
-      We define $\arr$ on objects by
-
-      \[ (A, P_A, \ptb_A) \arr (B, P_B, \ptb_B) = 
-        (A \arr B, P_A^{op} \times P_B, \ptb_{A \arr B}), \]
-
-      where we define $\ptb_{A \arr B}(a, b) = \ptb_A(a) \arr \ptb_B(b)$.
-
-      Using Lemma \ref{lem:push-pull-arrow}, we define $\arr$ on relations by
-
-      \[ (c, \piv_c) \arr (d, \pie_d) =
-        (c \arr d, \pie_{c \arr d}). \]
-
-    \end{itemize}
-
-    We define a ``Kleisli arrow'' operation taking a perturbation in $P_{UB}$
-    to a perturbation in $P_{U(A \to B)}$, as follows:
-    Given a perturbation $u = (n, b) \in P_{UB}$ we define $\id \tok u$ to be 
-    $(n, \id_{P_A}, b) \in P_{U(A \to B)}$.
-    We need to show that
-%
-    \[ \ptb_{U(A \arr B)}(\id \tok (n, b)) = \id \tok \ptb_{UB}(n, b). \]
-%
-    That is, we need to show that
-%
-    \[ (\delta_{U(A \arr B)}^*)^n \circ U(\id_{A} \arr \ptb_B(b)) = 
-        \id \tok (\delta_{UB}^*)^n \circ U(\ptb_B(b)) \]
-%
-    This follows from the fact that $\delta_{UB}^*$ commutes with computation morphisms.
-%
-
-    Given a perturbation $q = (n, a) \in P_{FA}$ we define
-    $(n, a) \tok \id = (n, a, \id_{P_B}) \in P_{U(A \to B)}$.
-    We need to show that
-  %
-    \[ \ptb_{U(A \arr B)}((n, a) \tok \id) = \ptb_{FA}(n, a) \tok \id. \]
-  %
-    By similar reasoning to the above, this holds because $\delta_{FA}^*$ commutes with all computation morphisms.
-
-    We can similarly construct a Kleisli product operation on perturbations.
-    Given a perturbation $(n, a_1, a_2) \in P_{F(A_1 \times A_2)}$ we define
-    $(n, a_1, a_2) \timesk \id = (n, a_1, \id_{P_{A_2}})$,
-    and likewise we define $\id \timesk (n, a_1, a_2) = (n, \id_{P_{A_1}}, a_2)$.
+  Similar to the product case above.
+\end{proof}
+
+We next define the actions of the Kleisli arrow and product functors on
+syntactic perturbations.
+
+% ``Kleisli arrow'' action on syntactic perturbations that takes a
+% perturbation in $M_{UB}$ to a perturbation in $M_{U(A \to B)}$, as follows:
+
+\begin{definition}\label{def:kleisli-arrow-perturbations} Let $A$ be a value
+  object and $B$ be a computation object. We define a homomorphism of monoids
+  $id \tok - \colon M_{UB} \to M_{U(A \arr B)}$ via the universal property of the
+  coproduct of monoids by sending $\mathbb{N}$ to the first injection in $M_{U(A
+  \arr B)} = \mathbb{N} \oplus M_A^{op} \oplus M_B$, and sending $M_B$ to the
+  third injection.
+
+  Likewise, we define $- \tok \id \colon M_{\li A}^{op} \to M_{U(A \arr B)}$ as
+  follows. First note that $M_{\li A}^{op} = \mathbb{N} \oplus M_A^{op}$. Then
+  we construct the homomorphism via the universal property, sending $\mathbb{N}$
+  to the first injection $M_A^{op}$ to the second injection.
+\end{definition}
+
+We can similarly construct a Kleisli product operation on perturbations:
+
+\begin{definition}\label{def:kleisli-product-perturbations}
+  Let $A_1$ and $A_2$ be value objects. We define a homomorphism of monoids 
+  $- \timesk \id \colon M_{\li A_1} \to M_{\li (A_1 \times A_2)}$ using the universal property,
+  sending $\mathbb{N}$ to the first injection and $M_{A_1}$ to the second injection.
+
+  Likewise, we define $\id \timesk - \colon M_{\li A_2} \to M_{\li (A_1 \times A_2)}$
+  by sending $\mathbb{N}$ to the first injection and $M_{A_2}$ to the third injection.
+
+  %  Given a perturbation $(n, a_1, a_2) \in $ we define $(n, a_1, a_2)
+  % \timesk \id = (n, a_1, \id_{P_{A_2}})$, and likewise we define $\id \timesk (n,
+  % a_1, a_2) = (n, \id_{P_{A_1}}, a_2)$.
+\end{definition}
+
+It is then easy to verify that the Kleisli arrow action on perturbation is well-defined 
+in that for any $m \in M_{UB}$, we have
+%
+\[ \ptb_{U(A \arr B)}(\id \tok m) = \id \tok i_{UB}(m), \]
+%
+where the LHS is the Kleisli action on perturbation and the RHS is the Kleisli
+action on morphisms. 
+%
+The proof uses the fact that the morphism $\delta$ commutes with all error
+domain morphisms, which is a consequence of the definition of error domain
+morphism.
+%
+The other verification involving the Kleisli arrow action is similar, as are the
+two Kleisli product actions.
+
+% %
+% \[ (\delta_{U(A \arr B)}^*)^n \circ U(\id_{A} \arr \i_B(b)) = 
+%     \id \tok (\delta_{UB}^*)^n \circ U(\ptb_B(b)) \]
+% %
+% This can be checked by unfolding the definition of $\tok$; the proof uses the
+% fact that $\delta_{UB}^*$ commutes with computation morphisms.
+% %
+
+% Given a perturbation $q = (n, a) \in P_{FA}$ we define $(n, a) \tok \id = (n, a,
+% \id_{P_B}) \in P_{U(A \to B)}$. We need to show that
+% %
+% \[ \ptb_{U(A \arr B)}((n, a) \tok \id) = \ptb_{FA}(n, a) \tok \id. \]
+% %
+% By similar reasoning to the above, this holds because $\delta_{FA}^*$ commutes
+% with computation morphisms.
 
@@
 
+% %%%%%%%%%%%%%%%%%%%%%%
+% % Model Construction %
+% %%%%%%%%%%%%%%%%%%%%%%
+% We now proceed with the construction of the model:
+
+%     Define a step-2 model $\mathcal M'$ as follows:
+%     \begin{itemize}
+%       \item Value objects are pairs consisting of:
+%       \begin{itemize}
+%         \item A value object $A$ in $\vf$.
+%         \item A monoid $P_A$ of perturbations
+%         and a monoid homomorphism $\ptb_A : P_A \to \{ f \in \vf(A, A) \mid f \bisim \id_A \}$.
+
+%       \end{itemize}
+
+%       \item Computation objects are pairs consisting of:
+%       \begin{itemize}
+%         \item A computation object $B$ in $\ef$.
+%         \item A monoid $P_B$ of perturbations and a monoid homomorphism
+%         $\ptb_B : P_B \to \{ \phi \in \ef(B, B) \mid \phi \bisim \id_B \}$.
+%       \end{itemize}
+
+%       \item Morphisms are given by morphisms of the underlying objects in $\vf$ and $\ef$, respectively.
+%       %, i.e.,
+%       % \[ \vf'((A, P_A, \ptb_A, P^K_A, \ptbk_A), (A', P_{A'}, \ptb_{A'}, P^K_{A'}, \ptbk_{A'})) = \vf(A, A') \]
+%       %
+%       % and likewise for computations.
+
+%       \item Given objects $(A, P_A, \ptb_A)$ and $(A', P_{A'}, \ptb_{A'})$ a value relation
+%       is a pair consisting of
+%       \begin{itemize}
+%         \item A value relation $c \in \vsq$.
+%         \item A push-pull structure $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$.
+%         %\item A push-pull structure $\pie_{Fc}$ for $Fc$ with respect to $P^K_A$ and $P^K_{A'}$.
+%       \end{itemize}
+
+%       \item Similarly, a computation relation between $(B, P_B, \ptb_B)$ and $(B', P_{B'}, \ptb_{B'})$
+%       consists of
+%       \begin{itemize}
+%         \item A computation relation $d \in \esq$.
+%         \item A push-pull structure $\pie_d$ for $d$ with respect to $P_B$ and $P_{B'}$.
+%         %\item A push-pull structure $\piv_{Ud}$ for $Ud$ with respect to $P^K_B$ and $P^K_{B'}$.
+%       \end{itemize}
+
+%       \item The squares are the same as the squares of the original model $\mathcal M$
+%       %morphisms of $\vsq'$ and $\esq'$ are given by the morphisms of $\vsq$ and $\esq$.
+
+%       \item We define composition of relations $(c, \piv_c)$ and $(c', \piv_{c'})$
+%       as $(c \comp c', \piv_{c \comp c'})$ where $\piv_{c \comp c'}$ is as in Lemma
+%       \ref{lem:push-pull-comp}, and likewise for computation relations.
+
+%     \end{itemize}
+
+%       % Functors \times, +, F, U, arrow
+%       Now we define the actions of the functors:
+%       \begin{itemize}
+
+%       \item We define $\times$ on objects by
+
+%       \[ (A_1, P_{A_1}, \ptb_{A_1}) \times (A_2, P_{A_2}, \ptb_{A_2}) =
+%         (A_1 \times A_2, P_{A_1} \times P_{A_2}, \ptb_{A_1 \times A_2}) \]
+
+%       where $\ptb_{A_1 \times A_2}(p_1, p_2) = \ptb_{A_1}(p_1) \times \ptb_{A_2}(p_2)$.
+
+%       Using Lemma \ref{lem:push-pull-times}, we define $\times$ on relations by
+
+%       \[ (c_1, \piv_{c_1}), (c_2 \piv_{c_2}) = 
+%         (c_1 \times c_2, \piv_{c_1 \times c_2}). \]
+
+%       \item We define $F$ on objects by
+
+%       \[ F(A, P_A, \ptb_A) = (FA, \mathbb{N} \times P_A, \ptb_{FA}), \]
+
+%       where we define $\ptb_{FA}(n, a) = (\delta_{FA}^*)^n \circ F(\ptb_A(a))$.
+
+%       Using Lemma \ref{lem:push-pull-U-F}, we define $F$ on relations by
+
+%       \[ F(c, \piv_c) = (Fc, \pie_{Fc}). \]
+
+%       \item We define $U$ on objects by
+
+%       \[ U(B, P_B, \ptb_B) = (UB, \mathbb{N} \times P_B, \ptb_{UB}), \]
+
+%       where we define $\ptb_{UB}(n, b) = (\delta_{UB}^*)^n \circ U(\ptb_B(b))$.
+
+%       Using Lemma \ref{lem:push-pull-U-F}, we define $U$ on relations by
+
+%       \[ U(d, \pie_d) = (Ud, \piv_{Ud}).  \]
+
+%       We define $\arr$ on objects by
+
+%       \[ (A, P_A, \ptb_A) \arr (B, P_B, \ptb_B) = 
+%         (A \arr B, P_A^{op} \times P_B, \ptb_{A \arr B}), \]
+
+%       where we define $\ptb_{A \arr B}(a, b) = \ptb_A(a) \arr \ptb_B(b)$.
+
+%       Using Lemma \ref{lem:push-pull-arrow}, we define $\arr$ on relations by
+
+%       \[ (c, \piv_c) \arr (d, \pie_d) =
+%         (c \arr d, \pie_{c \arr d}). \]
+
+%     \end{itemize}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\subsection{Constructing a Model with Quasi-Representable Relations}
-
-The goal of this section is to prove the following lemma:
-
-\begin{lemma}\label{lem:step-2-model-to-step-3-model}
-  Let $\mathcal M$ be a \hyperref[def:step-2-model]{step-2 intensional model}.
-
-  Then we can construct a \hyperref[def:step-3-model]{step-3 intensional model}.
-\end{lemma}
-
-Before proceeding with the proof, we begin with a definition.
-
-\begin{definition}[representation structure]
-  Let $c : A \rel A'$ be a value relation. A \emph{left-representation structure}
-  $\rho^L_c$ for $c$ consists of a value morphism $e_c \in \vf(A, A')$ such that
-  $c$ is quasi-left-representable by $e_c$ (see Definition \ref{def:quasi-left-representable}).
-  We similarly define a {right-representation structure} $\rho^R_c$ for $c$ to consist of a morphism
-  $p_c \in \vf{A', A}$ such that $c$ is quasi-right-representable by $p_c$ (see Definition \ref{def:quasi-right-representable}).
+\subsection{Lemmas involving Quasi-Representable Relations}
+
+Before proceeding with the lemmas, we begin with a definition.
+
+\begin{definition}[representation structure]\label{def:representation-structure}
+  Let $A$ and $A'$ be value objects, and let $c : A \rel A'$ be a relation on
+  the underlying predomains. A \emph{left-representation structure} $\rho^L_c$
+  for $c$ consists of a predomain morphism $e_c : A \to A'$ such that $c$ is
+  quasi-left-representable by $e_c$ (see Definition
+  \ref{def:quasi-left-representable}). We similarly define a
+  {right-representation structure} $\rho^R_c$ for $c$ to consist of a predomain
+  morphism $p_c : A' \to A$ such that $c$ is quasi-right-representable by $p_c$
+  (see Definition \ref{def:quasi-right-representable}).
   % (Notice that the direction of the morphism is opposite in the definition of
@@
 
-  Likewise, let $d : B \rel B'$ be a computation relation. Left- and right-representation
-  structures for $d$ are defined in a similar manner, except the representing morphisms
-  are now computation morphisms rather than value morphisms.
-
-  % Likewise, let $d : B \rel B'$ be a computation relation. A \emph{right-representation structure}
-  % $\rho^R_d$ for $d$ consists of a computation morphism $p_d \in \ef(B', B)$
-  % such that $d$ is quasi-right-representable by $p_d$ (see Definition
-  % \ref{def:quasi-right-representable}).
+  Likewise, let $d : B \rel B'$ be a computation relation. Left- and
+  right-representation structures for $d$ are defined similarly, except the
+  representing morphisms are now error domain morphisms rather than predomain
+  morphisms.
 \end{definition}
 
-% TODO establish the notation
-
-Recall the notion of quasi-order-equivalence as defined in Definition \ref{def:quasi-order-equivalent}.
-The following lemma will be useful in showing that two relations are quasi-order-equivalent.
-
-\begin{lemma}\label{lem:left-rep-by-same-morphism}
-  Let $c, c' : A \rel A'$. If $c$ and $c'$ are both quasi-left-representable
-  by the same $f$, then $c \qordeq c'$.
-
-  Dually, if $d$ and $d'$ are both quasi-right-representable by the same $\phi$, then
-  $d \qordeq d'$.
+Recall the notion of quasi-order-equivalence as defined in Definition
+\ref{def:quasi-order-equivalent}. The following lemma will be useful in showing
+that two relations are quasi-order-equivalent.
+
+\begin{lemma}\label{lem:left-rep-by-same-morphism} Let $A$ and $A'$ be value
+  objects, and let $c, c' : A \rel A'$ be relations between the underlying
+  predomains. If $c$ and $c'$ are both quasi-left-representable by the same
+  predomain morphism $f : A \to A'$, then $c \qordeq c'$. If $c$ and $c'$ are
+  both quasi-right-representable by the same predomain morphism $g : A' \to A$
+  then $c \qordeq c'$.
+
+  Dually, let $B$ and $B'$ be computation objects, and let $d, d' : B \rel B'$
+  be relations between the underlying error domains. If $d$ and $d'$ are both
+  quasi-right-representable by the same error domain morphism $\phi : B'
+  \multimap B$, then $d \qordeq d'$. If $d$ and $d'$ are both
+  quasi-left-representable by the same $\phi : B \multimap B'$
 \end{lemma}
 \begin{proof}
+  We show the result for $c$ and $c'$ that are quasi-left-representable by the same $f$;
+  the other proofs are analogous.
+
   By $\upr$ for $c'$, there exists a perturbation $\delle_{c'}$ and a square
@@
 
-  % https://q.uiver.app/#q=WzAsNixbMCwwLCJBIl0sWzEsMCwiQSJdLFsyLDAsIkEnIl0sWzAsMSwiQSJdLFsxLDEsIkEnIl0sWzIsMSwiQSciXSxbMCwxLCJyKEEpIl0sWzEsMiwiYyJdLFszLDQsImMnIiwyXSxbNCw1LCJyKEEnKSIsMl0sWzAsMywiXFxkZWxsZV97Yyd9IiwyXSxbMSw0LCJmIiwyXSxbMiw1LCJcXGRlbHJlX2MiXV0=
+  % https://q.uiver.app/#q=WzAsNixbMCwwLCJBIl0sWzEsMCwiQSJdLFsyLDAsIkEnIl0sWzAsMSwiQSJdLFsxLDEsIkEnIl0sWzIsMSwiQSciXSxbMCwxLCJyKEEpIl0sWzEsMiwiYyJdLFszLDQsImMnIiwyXSxbNCw1LCJyKEEnKSIsMl0sWzAsMywiaV9BKFxcZGVsbGVfe2MnfSkiLDJdLFsxLDQsImYiLDJdLFsyLDUsImlfe0EnfShcXGRlbHJlX2MpIl1d
   \[\begin{tikzcd}[ampersand replacement=\&]
@@
     \arrow["{r(A)}", from=1-1, to=1-2]
+    \arrow["{i_A(\delle_{c'})}"', from=1-1, to=2-1]
     \arrow["c", from=1-2, to=1-3]
+    \arrow["f"', from=1-2, to=2-2]
+    \arrow["{i_{A'}(\delre_c)}", from=1-3, to=2-3]
     \arrow["{c'}"', from=2-1, to=2-2]
     \arrow["{r(A')}"', from=2-2, to=2-3]
-    \arrow["{\delle_{c'}}"', from=1-1, to=2-1]
-    \arrow["f"', from=1-2, to=2-2]
-    \arrow["{\delre_c}", from=1-3, to=2-3]
   \end{tikzcd}\]
 
-  And since $r(A) \comp c = c$ and $c' \comp r(A') = c'$, we are finished.
-
-  The other square (i.e., with $c'$ on top) is constructed in an analogous manner.
-\end{proof}
+  Then since $r(A) \comp c = c$ and $c' \comp r(A') = c'$, we obtain the desired square.
+  %
+  The other square (i.e., with $c'$ on top) is constructed in an analogous
+  manner.
+
+\end{proof}
+
+Next we show that the functors $\li$ and $U$ preserve quasi-representability.
+
+%%%%%%%%%%%%%
+% U and F
+%%%%%%%%%%%%%
+\begin{lemma}\label{lem:representation-U-F}
+  Let $A$ and $A'$ be value objects and let $c : A \rel A'$ be a predomain relation.
+  If $c$ is quasi-left-representable, then $\li c$ is quasi-left-representable.
+  Likewise, if $c$ is quasi-right-representable, then $\li c$ is quasi-right-representable.
+
+  %Then we can define a left-representation structure $\rho^L_{UF(c)}$ for $UF(c)$.
+
+  Similarly, let $B$ and $B'$ be computation objects and let $d : B \rel B'$ be an error domain relation.
+  If $d$ is quasi-right-representable, then $Ud$ is quasi-right-representable.
+
+  % Then we can define a right-representation structure $\rho^R_{FU(d)}$ for $FU(d)$.
+\end{lemma}
+\begin{proof}
+  To show that $\li c$ is quasi-left-representable, we define
+  \begin{itemize}
+    \item $e_{\li c} = \li e_c$
+    \item $\delre_{\li c} = \inr (\delre_c)$ 
+    (recalling that the monoid $M_{\li A'} = \mathbb{N} \oplus M_{A'}$)
+    \item $\delle_{\li c} = \inr (\delle_c)$
+  \end{itemize}
+  We obtain the two squares $\upl$ and $\upr$ using the definition of the
+  interpretation of syntactic perturbations for $\li$ and the functorial action
+  of $\li$ on the corresponding squares for $c$.
+
+  The proof for $Ud$ is analogous.
+\end{proof}
+
+%%%%%%%%%%%%
+% Identity %
+%%%%%%%%%%%%
+
+\eric{Identity relation is quasi-representable}
 
@@
 %%%%%%%%%%%%%%%
+
+We begin by showing that the composition of quasi-representable relations with
+push-pull structures is quasi-representable.
+
 \begin{lemma}\label{lem:representation-comp}
-  In the below, let $c : A \rel A'$ and $c' : A' \rel A''$ and $d : B \rel B'$ and $d' : B' \rel B''$.
-  \begin{enumerate}
-    \item Given left-representation structures $\rho^L_c$ for $c$ and $\rho^L_{c'}$ for $c'$,
-    we can define a left-representation structure for the composition $c \comp c'$
-    \item Given right-representation structures $\rho^R_d$ for $d$ and $\rho^R_{d'}$ for $d'$,
-    we can define a right-representation structure for the composition $d \comp d'$
-    \item Given right-representation structures $\rho^R_{Fc}$ for $Fc$ and $\rho^R_{Fc'}$ for $Fc'$,
-    we can define a right-representation structure $\rho^R_{F(c \comp c')}$ for $F(c \comp c')$.
-    \item Given left-representation structures $\rho^L_{Ud}$ for $Ud$ and $\rho^L_{Ud'}$ for $Ud'$,
-    we can define a left-representation structure $\rho^L_{U(d \comp d')}$ for $U(d \comp d')$.
-  \end{enumerate}
-
+  Let $A$, $A'$, and $A''$ be value objects, and $B$, $B'$, and
+  $B''$ be computation objects. Let $c : A \rel A'$ and $c' : A' \rel A''$ be
+  predomain relations with push-pull structures, and let $d : B \rel B'$ and $d'
+  : B' \rel B''$ be error domain relations with push-pull structures.
+  \begin{enumerate}
+    \item If $c$ and $c'$ are quasi-left- (resp. right)-representable, then $c
+    \comp c'$ is quasi-left- (resp. right)-representable.
+
+    \item If $d$ and $d'$ are quasi-left- (resp. right)-representable, then
+    $d \comp d'$ is quasi-left (resp. right)-representable.
+  \end{enumerate}
 \end{lemma}
 \begin{proof}
-  \begin{enumerate}
-
     % 1.
-    \item We define $\rho^L_{c \comp c'}$ as follows. In the definitions of the perturbations,
-    we make use of the fact that $c$ and $c'$ satisfy the push-pull property.
+    \item To show $c \comp c'$ is quasi-left-representable we define
     \begin{itemize}
       \item $e_{c \comp c'} = e_{c'} \circ e_c$
-      \item $\delre_{c \comp c'} = \delre_{c'} \circ \push_{c'}(\delre_c)$
-      \item $\delle_{c \comp c'} = \pull_c(\delle_{c'}) \circ \delle_c$
+      \item $\delre_{c \comp c'} = \delre_{c'} \cdot \push_{c'}(\delre_c)$ where
+      $\cdot$ denotes multiplication in the monoid $M_A$
+      \item $\delle_{c \comp c'} = \pull_c(\delle_{c'}) \cdot \delle_c$
       \item $\upl$ is the following square: \input{squares/UpL-comp}
 
-      The square $(*)$ exists by the push-pull property for $c'$, and the square $(**)$ exists
-      because $r$ is a unit for horizontal composition, so $r(A') \comp c' = c'$, and so this
-      is simply the identity square $\id_{c'} \in \vsq(c', c')$.
+      The square $(*)$ exists by the push-pull property for $c'$, and the square
+      $(**)$ exists by the downward closure of $c'$.
 
@@
 
-    % 2.
-    \item We define $\rho^R_{d \comp d'}$ as follows:
+    To show $c \comp c'$ is quasi-right-representable we define
     \begin{itemize}
-      \item $p_{d \comp d'} = p_d \circ p_{d'}$
-      \item $\dellp_{d \comp d'} = \dellp_d \circ \pull_d(\dellp_{d'})$
-      \item $\delrp_{d \comp d'} = \push_{d'}(\delrp_d) \circ \delrp_{d'}$
+      \item $p_{c \comp c'} = p_c \circ p_{c'}$
+      \item $\dellp_{c \comp c'} = \dellp_c \cdot \pull_c(\dellp_{c'})$
+      \item $\delrp_{c \comp c'} = \push_{c'}(\delrp_c) \cdot \delrp_{c'}$
       \item $\dnr$ is the following square: \input{squares/DnR-comp}
 
-      Here the square $(*)$ exists by the push-pull property for $d$.
+      Here the square $(*)$ exists by the push-pull property for $c$.
 
@@
 
-    % 3.
-    \item We define $\rho^R_{F(c \comp c')}$ as follows.
-    % First, we observe that by Lemma \ref{lem:left-rep-by-same-morphism},
-    % $F(c \comp c')$ and $Fc \comp Fc'$ are quasi-order-equivalent,
-    % as both are left-represented by $Fe_c' \circ Fe_c$.
-
-    First, we claim that $F(c \comp c')$ and $Fc \comp Fc'$ are
-    both quasi-left-represented by $Fe_{c'} \circ Fe_c$.
-    Indeed, we have by part (1) that $e_c' \circ e_c$ left-represents $c \comp c'$,
-    and then since $F$ preserves perturbations and squares it follows that
-    $F(e_c' \circ e_c) = Fe_{c'} \circ Fe_c$ quasi-left-represents $F(c \comp c')$.
-    On the other hand, we also know that $Fe_c$ quasi-left-represents $Fc$ and $Fe_{c'}$
-    left-represents $Fc'$, so their composition quasi-left-represents $Fc \comp Fc'$.
-    Thus, by Lemma \ref{lem:left-rep-by-same-morphism}, there is a square $\alpha$ of the form
-
-    % https://q.uiver.app/#q=WzAsNSxbMCwwLCJGQSJdLFsyLDAsIkZBJyciXSxbMCwxLCJGQSJdLFsxLDEsIkZBJyJdLFsyLDEsIkZBJyciXSxbMCwxLCJGKGMgXFxjb21wIGMnKSIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzIsMywiRmMiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFszLDQsIkZjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMiwiXFxkZWx0YV5sIiwyXSxbMSw0LCJcXGRlbHRhXnIiXSxbOCw5LCJcXGFscGhhIiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+    % 2.
+    \item Same as (1), replacing predomains/morphisms/relations/squares with error domains.
+\end{proof}
+
+The next two lemmas concern quasi-order-equivalence and the functors $\li$ and
+$U$.
+
+\begin{lemma}\label{lem:Fcc-equiv-FcFc'}
+
+  Let $A$, $A'$, and $A''$ be value objects and let $c : A \rel A'$ and $c' : A'
+  \rel A''$ be predomain relations. Then we have $\li(c \comp c') \bisim \li c \comp \li c'$
+
+\end{lemma}
+\begin{proof}
+  First, we claim that $\li(c \comp c')$ and $\li c \comp \li c'$ are both
+  quasi-left-represented by $\li e_{c'} \circ \li e_c$. Indeed, we have by part
+  (1) of Lemma \ref{lem:representation-comp} that $e_c' \circ e_c$
+  quasi-left-represents $c \comp c'$, and then by Lemma
+  \ref{lem:representation-U-F} we have $F(e_c' \circ e_c) = Fe_{c'} \circ Fe_c$
+  quasi-left-represents $F(c \comp c')$.
+  %
+  On the other hand, we also know that $Fe_c$ quasi-left-represents $Fc$ and
+  $Fe_{c'}$ quasi-left-represents $Fc'$ again by Lemma
+  \ref{lem:representation-U-F}. Then by part (2) of Lemma
+  \ref{lem:representation-comp}, their composition quasi-left-represents $Fc
+  \comp Fc'$.
+
+  The result now follows by Lemma \ref{lem:left-rep-by-same-morphism}.
+\end{proof}
+
+\begin{lemma}\label{lem:Udd-equiv-UdUd'}
+  Let $B$, $B'$, and $B''$ be value objects and let $d : B \rel B'$ and $d' : B'
+  \rel B''$ be error domain relations. Then we have $U(d \comp d') \bisim Ud \comp Ud'$.
+\end{lemma}
+\begin{proof}
+  Analogous to the proof of the previous lemma.
+\end{proof}
+
+Now we combine the above lemmas to show a result about the functors $\li$ and
+$U$ applied to a composition of relations:
+
+\begin{lemma}\label{lem:representation-comp-F-U}
+  Let $A$, $A'$, and $A''$ be value objects, and $B$, $B'$, and
+  $B''$ be computation objects. Let $c : A \rel A'$ and $c' : A' \rel A''$ be
+  predomain relations with push-pull structures, and let $d : B \rel B'$ and $d'
+  : B' \rel B''$ be error domain relations with push-pull structures.
+  \begin{enumerate}
+    \item If $\li c$ and $\li c'$ are quasi-right-representable, then $\li (c
+    \comp c')$ is quasi-right-representable.
+
+    \item If $Ud$ and $Ud'$ are quasi-left-representable, then $U(d \comp d')$
+    is quasi-left-representable.
+
+    % \item Given left-representation structures $\rho^L_c$ for $c$ and $\rho^L_{c'}$ for $c'$,
+    % we can define a left-representation structure for the composition $c \comp c'$
+    % \item Given right-representation structures $\rho^R_d$ for $d$ and $\rho^R_{d'}$ for $d'$,
+    % we can define a right-representation structure for the composition $d \comp d'$
+    % \item Given right-representation structures $\rho^R_{Fc}$ for $Fc$ and $\rho^R_{Fc'}$ for $Fc'$,
+    % we can define a right-representation structure $\rho^R_{F(c \comp c')}$ for $F(c \comp c')$.
+    % \item Given left-representation structures $\rho^L_{Ud}$ for $Ud$ and $\rho^L_{Ud'}$ for $Ud'$,
+    % we can define a left-representation structure $\rho^L_{U(d \comp d')}$ for $U(d \comp d')$.
+  \end{enumerate}
+
+\end{lemma}
+\begin{proof}
+  \begin{enumerate}
+    % 1.
+    \item We show that $\li (c \comp c')$ is quasi-right-representable.
+
+    First, by Lemma \ref{lem:Fcc-equiv-FcFc'} there is a square $\alpha$ of the form
+
+    % https://q.uiver.app/#q=WzAsNSxbMCwwLCJcXGxpIEEiXSxbMiwwLCJcXGxpIEEnJyJdLFswLDEsIlxcbGkgQSJdLFsxLDEsIlxcbGkgQSciXSxbMiwxLCJcXGxpIEEnJyJdLFswLDEsIlxcbGkgKGMgXFxjb21wIGMnKSIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzIsMywiXFxsaSBjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMyw0LCJcXGxpIGMnIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwyLCJpX0EoXFxkZWx0YV5sKSIsMl0sWzEsNCwiaV97QScnfShcXGRlbHRhXnIpIl0sWzgsOSwiXFxhbHBoYSIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
     \[\begin{tikzcd}[ampersand replacement=\&]
-      FA \&\& {FA''} \\
-      FA \& {FA'} \& {FA''}
-      \arrow["{F(c \comp c')}", "\shortmid"{marking}, no head, from=1-1, to=1-3]
-      \arrow["Fc", "\shortmid"{marking}, no head, from=2-1, to=2-2]
-      \arrow["{Fc'}", "\shortmid"{marking}, no head, from=2-2, to=2-3]
-      \arrow[""{name=0, anchor=center, inner sep=0}, "{\delta^l}"', from=1-1, to=2-1]
-      \arrow[""{name=1, anchor=center, inner sep=0}, "{\delta^r}", from=1-3, to=2-3]
+      {\li A} \&\& {\li A''} \\
+      {\li A} \& {\li A'} \& {\li A''}
+      \arrow["{\li (c \comp c')}", "\shortmid"{marking}, no head, from=1-1, to=1-3]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "{i_A(\delta^l)}"', from=1-1, to=2-1]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "{i_{A''}(\delta^r)}", from=1-3, to=2-3]
+      \arrow["{\li c}", "\shortmid"{marking}, no head, from=2-1, to=2-2]
+      \arrow["{\li c'}", "\shortmid"{marking}, no head, from=2-2, to=2-3]
       \arrow["\alpha"{description}, draw=none, from=0, to=1]
     \end{tikzcd}\]
-
-    for some perturbations $\delta^l$ and $\delta^r$.
-    We define the projection $p_{F(c \comp c')}$ to be $p_{Fc \comp Fc'} \circ \delta^r$.
-    We define $\dellp_{F(c \comp c')}$ to be $\dellp_{Fc \comp Fc'} \circ \delta^l$
-
-    Then we can build the square for $\dnr$ by pasting the square $\alpha$ on top of
-    the $\dnr$ square for the composition $Fc \comp Fc'$, as shown below:
+    where $\delta^l$ and $\delta^r$ are syntactic perturbations.
+
+    We define the projection $p_{\li (c \comp c')}$ to be $p_{\li c \comp \li c'} \circ i_{A''}(\delta^r)$.
+    We define $\dellp_{\li (c \comp c')}$ to be $\dellp_{\li c \comp \li c'} \cdot \delta^l$.
+
+    Then we can build the $\dnr$ square by pasting the square $\alpha$ on top of
+    the $\dnr$ square for the composition $\li c \comp \li c'$, as shown below:
+
+    \eric{Need to change F to $\li$}
 
@@
 
-    We define $\delrp_{F(c \comp c')}$ to be $\delrp_{Fc \comp Fc'} \circ \delta^r$.
+    We define $\delrp_{\li (c \comp c')}$ to be $\delrp_{\li c \comp \li c'} \cdot \delta^r$.
     For $\dnl$, we paste the identity square $\delta^r \ltdyn \delta^r$ on top of
-    the $\dnl$ square for the composition $Fc \comp Fc'$, and below that we paste
-    the square $\id \ltdyn_{F(c \comp c')}^{Fc \comp Fc'} \id$ which we get from
-    the lax property of $F$.
+    the $\dnl$ square for the composition $\li c \comp \li c'$, and below that we paste
+    the square $\id \ltdyn_{\li (c \comp c')}^{\li c \comp \li c'} \id$ which we get from
+    the fact that $\li$ is lax.
 
@@
 
-    % 4.
-    \item We define $\rho^L{(U(d \comp d'))}$ in an analogous manner to the above.
+    % 2.
+    \item The proof that $U(d \comp d')$ is quasi-left-representable is analogous.
 
@@
 
-%%%%%%%%%%%%%
-% UF and FU %
-%%%%%%%%%%%%%
-\begin{lemma}\label{lem:representation-UF-FU}
-  Let $c : A \rel A'$, and let $\rho^L_c$ be a left-representation structure for $c$.
-  Then we can define a left-representation structure $\rho^L_{UF(c)}$ for $UF(c)$.
-
-  Similarly, let $d : B \rel B'$ and let $\rho^R_d$ be a right-representation structure for $d$.
-  Then we can define a right-representation structure $\rho^R_{FU(d)}$ for $FU(d)$.
-\end{lemma}
-\begin{proof}
-  We define $\rho^L_{UF(c)}$ as follows:
-  \begin{itemize}
-    \item $e_{UF(c)} = UF(e_c)$
-    \item $\delre_{UF(c)} = UF(\delre_c)$ (which is in the monoid of perturbations of $UF(A')$ because the
-    perturbation monoids are closed under the actions of the functors $F$ and $U$)
-    \item $\delle_{UF(c)} = UF(\delle_c)$
-    \item We get the two commuting squares by the functorial action of $UF$ on the two squares for $c$,
-    i.e., $\upr_{UF(c)} = UF(\upr_c)$
-  \end{itemize}
-
-  We define $\rho^R_{FU(d)}$ in a similar manner.
-\end{proof}
-
 %%%%%%%%%%%%
@@
 \begin{lemma}\label{lem:representation-product}
-  Let $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$.
-  Let $\rho^L_{c_1}$ be a left-representation structure for $c_1$, and
-  let $\rho^L_{c_2}$ be a left-representation structure for $c_2$.
-  Then we can define a left-representation structure for $c_1 \times c_2$.
-
-  Likewise, let $\rho^R_{Fc_1}$ and $\rho^R_{Fc_2}$ be right-representation
-  structures for $Fc_1$ and $Fc_2$ respectively.
-  Then we can deinfe a right-representation structure for $F(c_1 \times c_2)$.
+
+  Let $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$. If $c_1$ and $c_2$ are
+  quasi-left-representable, then $c_1 \times c_2$ is quasi-left-representable.
+  Furthermore, if $\li c_1$ and $\li c_2$ are quasi-right-representable, then so
+  is $\li (c_1 \times c_2)$.
+
+  % Let $\rho^L_{c_1}$ be a left-representation structure for $c_1$, and
+  % let $\rho^L_{c_2}$ be a left-representation structure for $c_2$.
+  % Then we can define a left-representation structure for $c_1 \times c_2$.
+
+  % Likewise, let $\rho^R_{Fc_1}$ and $\rho^R_{Fc_2}$ be right-representation
+  % structures for $Fc_1$ and $Fc_2$ respectively.
+  % Then we can define a right-representation structure for $F(c_1 \times c_2)$.
 \end{lemma}
 \begin{proof}
-  We define $\rho^L_{c_1 \times c_2}$ as follows:
+  To show $c_1 \times c_2$ is quasi-left-representable, we define
   \begin{itemize}
     \item $e_{c_1 \times c_2} = e_{c_1} \times e_{c_2}$
-    \item $\delre_{c_1 \times c_2} = \delre_{c_1} \times \delre_{c_2}$ and
-    likewise for $\delle_{c_1 \times c_2}$
-    \item We get the commuting squares via the functorial action of $\times$
-    on the corresponding squares for $c_1$ and $c_2$.
+    \item $\delre_{c_1 \times c_2} = \inl (\delre_{c_1}) \cdot \inr
+    (\delre_{c_2})$ (recall from Section \ref{sec:perturbation-constructions}
+    that $M_{A_1' \times A_2'}$ is the coproduct $M_{A_1'} \oplus M_{A_2'}$)
+    \item $\delle_{c_1 \times c_2}$ is defined similarly
+    \item The $\upl$ square is constructed using the functorial action of
+    $\times$ and the corresponding squares for $c_1$ and $c_2$:
+    % https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEgXFx0aW1lcyBBXzIiXSxbMCwxLCJBXzEnIFxcdGltZXMgQV8yIl0sWzAsMiwiQV8xJyBcXHRpbWVzIEFfMiciXSxbMSwwLCJBXzEnIFxcdGltZXMgQV8yJyJdLFsxLDEsIkFfMScgXFx0aW1lcyBBXzInIl0sWzEsMiwiQV8xJyBcXHRpbWVzIEFfMiciXSxbMCwzLCJjXzEgXFx0aW1lcyBjXzIiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxLDQsInIoQV8xJykgXFx0aW1lcyBjXzIiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsyLDUsInIoQV8xJykgXFx0aW1lcyByKEFfMicpIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJlX3tjXzF9IFxcdGltZXMgXFxpZCIsMl0sWzEsMiwiXFxpZCBcXHRpbWVzIGVfe2NfMn0iLDJdLFszLDQsImlfe0FfMSd9KFxcZGVscmVfe2NfMX0pIFxcdGltZXMgXFxpZCJdLFs0LDUsIlxcaWQgXFx0aW1lcyBpX3tBXzInfShcXGRlbHJlX3tjXzJ9KSJdLFs5LDExLCJcXHVwbF97Y18xfSBcXHRpbWVzIFxcaWQiLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMTAsMTIsIlxcaWQgXFx0aW1lcyBcXHVwbF97Y18yfSIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
+      \[\begin{tikzcd}[ampersand replacement=\&]
+        {A_1 \times A_2} \& {A_1' \times A_2'} \\
+        {A_1' \times A_2} \& {A_1' \times A_2'} \\
+        {A_1' \times A_2'} \& {A_1' \times A_2'}
+        \arrow["{c_1 \times c_2}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+        \arrow[""{name=0, anchor=center, inner sep=0}, "{e_{c_1} \times \id}"', from=1-1, to=2-1]
+        \arrow[""{name=1, anchor=center, inner sep=0}, "{i_{A_1'}(\delre_{c_1}) \times \id}", from=1-2, to=2-2]
+        \arrow["{r(A_1') \times c_2}", "\shortmid"{marking}, no head, from=2-1, to=2-2]
+        \arrow[""{name=2, anchor=center, inner sep=0}, "{\id \times e_{c_2}}"', from=2-1, to=3-1]
+        \arrow[""{name=3, anchor=center, inner sep=0}, "{\id \times i_{A_2'}(\delre_{c_2})}", from=2-2, to=3-2]
+        \arrow["{r(A_1') \times r(A_2')}"', "\shortmid"{marking}, no head, from=3-1, to=3-2]
+        \arrow["{\upl_{c_1} \times \id}"{description}, draw=none, from=0, to=1]
+        \arrow["{\id \times \upl_{c_2}}"{description}, draw=none, from=2, to=3]
+      \end{tikzcd}\]
+    \item The $\upr$ square is defined similarly
   \end{itemize}
 
-  % TODO check this
-  We define $\rho^R_{F(c_1 \times c_2)}$ as follows:
+  To show that $\li(c_1 \times c_2)$ is quasi-right-representable, we define
   \begin{itemize}
-    \item $p_{F(c_1 \times c_2)} = (p_{Fc_1} \timesk A_2) \circ (A_1' \timesk p_{Fc_2})$
-    \item $\dellp_{F(c_1 \times c_2)} = (\dellp_{Fc_1} \timesk A_2) \circ (A_1 \timesk \dellp_{Fc_2})$
-    \item $\delrp_{F(c_1 \times c_2)} = (\delrp_{Fc_1} \timesk A_2') \circ (A_1' \timesk \delrp_{Fc_2})$
-    \item The commuting squares are obtained via the functorial action of $\timesk$ on the squares for $Fc_1$ and $Fc_2$.
+    \item $p_{\li (c_1 \times c_2)} = (p_{\li c_1} \timesk A_2) \circ (A_1' \timesk p_{\li c_2})$
+    \item $\dellp_{\li (c_1 \times c_2)} = (\dellp_{\li c_1} \timesk \id) \cdot (\id \timesk \dellp_{\li c_2})$ 
+    using the Kleisli product action on syntactic perturbations in Definition \ref{def:kleisli-product-perturbations}
+    \item $\delrp_{\li (c_1 \times c_2)} = (\delrp_{\li c_1} \timesk \id) \cdot (\id \timesk \delrp_{\li c_2})$
+    \item The squares are obtained via the functorial action of $\timesk$ on the squares for $\li c_1$ and $\li c_2$.
   \end{itemize}
 \end{proof}
+
+\eric{Left off here}
 
@@
   Let $c : A \rel A'$ and $d : B \rel B'$.
-  Let $\rho^L_c$ be a left-representation structure for $c$,
-  and let $\rho^R_d$ be a right-representation structure for $d$.
-  Then we can define a right-representation structure for $c \arr d$.
-
-  Likewise, let $\rho^R_{Fc}$ be a right-representation structure for $Fc$,
-  and let $\rho^L_{Ud}$ be a left-representation structure for $Ud$.
-  Then we can define a left-representation structure for $U(c \arr d)$.
+  %
+  If $c$ is quasi-left-representable and $d$ is quasi-right-representable, then
+  $c \arr d$ is quasi-right-representable.
+
+  Likewise, if $\li c$ is quasi-right-representable and $Ud$ is
+  quasi-left-representable, then $U(c \arr d)$ is quasi-left-representable.
+
+  % Let $\rho^L_c$ be a left-representation structure for $c$,
+  % and let $\rho^R_d$ be a right-representation structure for $d$.
+  % Then we can define a right-representation structure for $c \arr d$.
+
+  % Likewise, let $\rho^R_{Fc}$ be a right-representation structure for $Fc$,
+  % and let $\rho^L_{Ud}$ be a left-representation structure for $Ud$.
+  % Then we can define a left-representation structure for $U(c \arr d)$.
 \end{lemma}
 \begin{proof}
-  We define $\rho^R_{c \arr d}$ as follows:
+  To show $c \arr d$ is quasi-right-representable, we define:
   \begin{itemize}
-    \item $p_{c \arr d} = e_c \arr p_d \in \ef(A' \arr B', A \arr B)$ 
-    (using the functorial action of $\arr$ on morphisms).
-    \item $\dellp_{c \arr d} = \delle_c \arr \dellp_d$
-    \item $\delrp_{c \arr d} = \delre_c \arr \delrp_d$
+    \item $p_{c \arr d} = e_c \arr p_d$
+    \item $\dellp_{c \arr d} = \inl(\delle_c) \cdot \inr(\dellp_d)$
+    \item $\delrp_{c \arr d} = \inl(\delre_c) \cdot \inr(\delrp_d)$
     \item The squares $\dnr$ and $\dnl$ are obtained via the functorial action of $\arr$, i.e., we define
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-% Model Construction %
-%%%%%%%%%%%%%%%%%%%%%%
-Now we can give the proof of the main lemma:
-  % Write 
+% We can now show that our notions of value and computation relations compose,
+% which follows from the preceding lemmas.
+
+% We now establish the quasi-order-equivalence for the functors.
+% We already showed that $U(d \comp d') \qordeq U(d)U(d')$ and $F(c \comp c'') \qordeq F(c)F(c')$
+% in the proof of Lemma \ref{lem:representation-comp}.
+% The other two properties
+% $(cc') \to (dd') \qordeq (c \to d)(c' \to d')$ and
+% $(c_1c_1') \times (c_2c_2') \qordeq (c_1 \times c_2)(c_1'\times c_2')$
+% are proved similarly, noting in both cases that the both relations are
+% quasi-represented by the same morphism.
+
+% %%%%%%%%%%%%%%%%%%%%%%
+% % Model Construction %
+% %%%%%%%%%%%%%%%%%%%%%%
+% Now we can give the proof of the main lemma:
+%   % Write 
+%   % %
+%   % \[ \mathcal M = (\vf, \vsq, \ef, \esq, \Ff, \Fsq, \Uf, \Usq, \arrf, \arrsq). \] 
+%   % %
+
+%   We define a step-3 model $\mathcal M'$ as follows:
+%   \begin{itemize}
+%     \item The objects of $\mathcal M'$ are defined to be the same as the objects of $\mathcal M$.
+%     \item The value and computation morphisms in $\mathcal M'$ are the same as those of $\mathcal M$.
+%     \item A value relation is defined to be a tuple $(c, \rho^L_c, \rho^R_{Fc})$ with
+%     \begin{itemize}
+%       \item $c$ a value relation in $\mathcal M$,
+%       \item $\rho^L_c$ a left-representation structure for $c$, and
+%       \item $\rho^R_{Fc}$ a right-representation structure for $Fc$
+%     \end{itemize}
+%     \item Likewise, a computation relation is defined to be a tuple $(d, \rho^R_d, \rho^L_{Ud})$ with
+%     \begin{itemize}
+%       \item $d$ a computation relation in $\mathcal M$,
+%       \item $\rho^R_d$ a right-representation structure for $d$, and
+%       \item $\rho^L_{Ud}$ a left-representation structure for $Ud$.
+%     \end{itemize}
+%     \item Morphisms of value relations (i.e., the value squares) are defined by simply
+%     ignoring the representation structures. That is, a morphism of value relations
+%     $\alpha \in \vsq'((c, \rho^L_c, \rho^R_{Fc}), (c' \rho^L_{c'}, \rho^R_{Fc'}))$ is simply a morphism of value
+%     relations in $\vsq(c, c')$. Likewise for computations.
+%   \end{itemize}
+
+% We define horizontal composition of relations and squares as follows:
+% Let $c : A \rel A'$ and $c' : A' \rel A''$. We define
 % %
-  % \[ \mathcal M = (\vf, \vsq, \ef, \esq, \Ff, \Fsq, \Uf, \Usq, \arrf, \arrsq). \] 
+% \begin{align*} 
+%   ((c, \rho^L_c, &\rho^R_{Fc}) \comp (c', \rho^L_{c'}, \rho^R_{Fc'})) =
+%   (c \comp c', \rho^L_{c \comp c'}, \rho^R_{F(c \comp c')}), 
+% \end{align*}
 % %
-
-  We define a step-3 model $\mathcal M'$ as follows:
-  \begin{itemize}
-    \item The objects of $\mathcal M'$ are defined to be the same as the objects of $\mathcal M$.
-    \item The value and computation morphisms in $\mathcal M'$ are the same as those of $\mathcal M$.
-    \item A value relation is defined to be a tuple $(c, \rho^L_c, \rho^R_{Fc})$ with
-    \begin{itemize}
-      \item $c$ a value relation in $\mathcal M$,
-      \item $\rho^L_c$ a left-representation structure for $c$, and
-      \item $\rho^R_{Fc}$ a right-representation structure for $Fc$
-    \end{itemize}
-    \item Likewise, a computation relation is defined to be a tuple $(d, \rho^R_d, \rho^L_{Ud})$ with
-    \begin{itemize}
-      \item $d$ a computation relation in $\mathcal M$,
-      \item $\rho^R_d$ a right-representation structure for $d$, and
-      \item $\rho^L_{Ud}$ a left-representation structure for $Ud$.
-    \end{itemize}
-    \item Morphisms of value relations (i.e., the value squares) are defined by simply
-    ignoring the representation structures. That is, a morphism of value relations
-    $\alpha \in \vsq'((c, \rho^L_c, \rho^R_{Fc}), (c' \rho^L_{c'}, \rho^R_{Fc'}))$ is simply a morphism of value
-    relations in $\vsq(c, c')$. Likewise for computations.
-  \end{itemize}
-
-We define the functor $m$ which composes relations and squares horizontally as follows:
-Let $c : A \rel A'$ and $c' : A' \rel A''$. We let
-
-\begin{align*} 
-  m ((c, \rho^L_c, \rho^R_{Fc}), (c', \rho^L_{c'}, \rho^R_{Fc'})) = \\
-  (c \comp c', \rho^L_{c \comp c'}, \rho^R_{F(c \comp c')}). 
-\end{align*}
-
-Now we define the functors $F$, $U$, $\times$, and $\arr$.
-On objects, the behavior is the same as the respective functors in $\mathcal M$.
-For relations, we define
-
-\[ \Fsq^{\mathcal M'} (c, \rho^L_c, \rho^R_{Fc}) = 
-  (\Fsq^M c, \rho^R_{Fc}, \rho^L_{UF(c)}), \]
-and
-\[ \Usq^{\mathcal M'} (d, \rho^R_d, \rho^L_{Ud}) = 
-  (\Usq^M d, \rho^L_{Ud}, \rho^R_{FU(d)}), \]
-
-where $\rho^L_{UF(c)}$ and $\rho^R_{FU(d)}$ are as defined in the proof of Lemma
-\ref{lem:representation-UF-FU}.
-
-We define \[ (c_1, \rho^L_{c_1}, \rho^R_{Fc_1}) \times (c_2, \rho^L_{c_2}, \rho^R_{Fc_2}) = 
-             (c_1 \times c_2, \rho^L_{c_1 \times c_2}, \rho^R_{F(c_1 \times c_2)}), \]
-
-where $\rho^L_{c_1 \times c_2}$ and $\rho^R_{F(c_1 \times c_2)}$ are as defined in the
-proof of Lemma \ref{lem:representation-product}.
-
-Lastly, we define
-
-\[ (c, \rho^L_c, \rho^R_{Fc}) \arr (d, \rho^R_d, \rho^L_{Ud}) =
-    (c \arr d, \rho^R_{c \arr d}, \rho^L_{U(c \arr d)}), \]
-
-where $\rho^R_{c \arr d}$ and $\rho^L_{U(c \arr d)}$ are as defined in the proof
-of Lemma \ref{lem:representation-arrow}.
-
-% We define $(c, \rho^L_c) \arr (d, \rho^R_d) = (c \arr d, \rho^R_{c \arr d})$.
+% and
+% %
+% \begin{align*}
+%   ((d, \rho^R_d, &\rho^L_{Ud}) \comp (d', \rho^R_{d'}, \rho^L_{Ud'})) =
+%   (d \comp d', \rho^R_{d \comp d'}, \rho^L_{U(d \comp d')}),
+% \end{align*}
+% %
+% where $\rho^L_{c \comp c'}, \rho^R_{F(c \comp c')}, \rho^R_{d \comp d'},$
+% and $\rho^L_{U(d \comp d')}$ are as defined in Lemma \ref{lem:representation-comp}.
+
+% Now we define the functors $F$, $U$, $\times$, and $\arr$.
+% On objects, the behavior is the same as the respective functors in $\mathcal M$.
+% For relations, we define
+
+% \[ F(c, \rho^L_c, \rho^R_{Fc}) = 
+%   (F c, \rho^R_{Fc}, \rho^L_{UF(c)}), \]
+% and
+% \[ U(d, \rho^R_d, \rho^L_{Ud}) = 
+%   (U d, \rho^L_{Ud}, \rho^R_{FU(d)}), \]
+
+% where $\rho^L_{UF(c)}$ and $\rho^R_{FU(d)}$ are as defined in the proof of Lemma
+% \ref{lem:representation-UF-FU}.
+
+% We define \[ (c_1, \rho^L_{c_1}, \rho^R_{Fc_1}) \times (c_2, \rho^L_{c_2}, \rho^R_{Fc_2}) = 
+%              (c_1 \times c_2, \rho^L_{c_1 \times c_2}, \rho^R_{F(c_1 \times c_2)}), \]
+
+% where $\rho^L_{c_1 \times c_2}$ and $\rho^R_{F(c_1 \times c_2)}$ are as defined in the
+% proof of Lemma \ref{lem:representation-product}.
+
+% Lastly, we define
+
+% \[ (c, \rho^L_c, \rho^R_{Fc}) \arr (d, \rho^R_d, \rho^L_{Ud}) =
+%     (c \arr d, \rho^R_{c \arr d}, \rho^L_{U(c \arr d)}), \]
+
+% where $\rho^R_{c \arr d}$ and $\rho^L_{U(c \arr d)}$ are as defined in the proof
+% of Lemma \ref{lem:representation-arrow}.
+
+% % We define $(c, \rho^L_c) \arr (d, \rho^R_d) = (c \arr d, \rho^R_{c \arr d})$.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Constructing an Extensional Model}\label{sec:extensional-construction-appendix}
-
-We aim to prove the following lemma:
-
-\begin{lemma}\label{lem:step-4-model-to-extensional-model}
-  Let $\mathcal M$ be a \hyperref[def:step-4-model]{step-4 intensional model}.
-  Then we can define an extensional model.
-\end{lemma}
-\begin{proof}
-
-  Recall the extensional model construction outlined in Section \ref{sec:extensional-model-definition}.
-  We first establish the representability properties for this model.
-  We show the left-representability squares; the right-representability squares are dual.
-
-  \begin{itemize}
-    \item We have the square
-
-    % https://q.uiver.app/#q=WzAsNixbMCwwLCJBIl0sWzAsMSwiQSciXSxbNCwxLCJBX3IiXSxbNCwwLCJBX3IiXSxbMiwwLCJBJyJdLFsyLDEsIkEnIl0sWzAsNCwiYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzQsMywiY19yIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJlX2MiLDIseyJjdXJ2ZSI6MX1dLFszLDIsIlxccHVzaF97Y19yfShcXGRlbHJlX2MpIiwyLHsiY3VydmUiOjF9XSxbMSw1LCJyKEEnKSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzQsNSwiXFxkZWxyZV9jIiwyXSxbNSwyLCJjX3IiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsImVfYyIsMCx7ImN1cnZlIjotMX1dLFszLDIsIlxcaWQiLDAseyJjdXJ2ZSI6LTF9XSxbOSwxNCwiXFxiaXNpbSIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs4LDEzLCJcXGJpc2ltIiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzE2LDExLCIiLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJsZXZlbCI6Miwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
-    \[\begin{tikzcd}[ampersand replacement=\&]
-      A \&\& {A'} \&\& {A_r} \\
-      {A'} \&\& {A'} \&\& {A_r}
-      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-3]
-      \arrow["{c_r}", "\shortmid"{marking}, no head, from=1-3, to=1-5]
-      \arrow[""{name=0, anchor=center, inner sep=0}, "{e_c}"', curve={height=6pt}, from=1-1, to=2-1]
-      \arrow[""{name=1, anchor=center, inner sep=0}, "{\push_{c_r}(\delre_c)}"', curve={height=6pt}, from=1-5, to=2-5]
-      \arrow["{r(A')}"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
-      \arrow[""{name=2, anchor=center, inner sep=0}, "{\delre_c}"', from=1-3, to=2-3]
-      \arrow["{c_r}"', "\shortmid"{marking}, no head, from=2-3, to=2-5]
-      \arrow[""{name=3, anchor=center, inner sep=0}, "{e_c}", curve={height=-6pt}, from=1-1, to=2-1]
-      \arrow[""{name=4, anchor=center, inner sep=0}, "\id", curve={height=-6pt}, from=1-5, to=2-5]
-      \arrow["\bisim"{description}, draw=none, from=1, to=4]
-      \arrow[""{name=5, anchor=center, inner sep=0}, "\bisim"{description}, draw=none, from=0, to=3]
-      \arrow[draw=none, from=5, to=2]
-    \end{tikzcd}\]
-
-    \item We have the square
-
-    % https://q.uiver.app/#q=WzAsNixbMCwwLCJBX2wiXSxbMCwxLCJBX2wiXSxbMiwxLCJBIl0sWzQsMSwiQSciXSxbMiwwLCJBIl0sWzQsMCwiQSJdLFswLDQsImNfbCJdLFs0LDUsInIoQSkiXSxbMSwyLCJjX2wiLDJdLFsyLDMsImMiLDJdLFswLDEsIlxcaWQiLDIseyJjdXJ2ZSI6MX1dLFs1LDMsImVfYyIsMix7ImN1cnZlIjoxfV0sWzQsMiwiXFxkZWxsZV9jIl0sWzAsMSwiXFxwdWxsX3tjX2x9KFxcZGVsbGVfYykiLDAseyJjdXJ2ZSI6LTF9XSxbNSwzLCJlX2MiLDAseyJjdXJ2ZSI6LTF9XSxbMTAsMTMsIlxcYmlzaW0iLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMTEsMTQsIlxcYmlzaW0iLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
-    \[\begin{tikzcd}[ampersand replacement=\&]
-      {A_l} \&\& A \&\& A \\
-      {A_l} \&\& A \&\& {A'}
-      \arrow["{c_l}", from=1-1, to=1-3]
-      \arrow["{r(A)}", from=1-3, to=1-5]
-      \arrow["{c_l}"', from=2-1, to=2-3]
-      \arrow["c"', from=2-3, to=2-5]
-      \arrow[""{name=0, anchor=center, inner sep=0}, "\id"', curve={height=6pt}, from=1-1, to=2-1]
-      \arrow[""{name=1, anchor=center, inner sep=0}, "{e_c}"', curve={height=6pt}, from=1-5, to=2-5]
-      \arrow["{\delle_c}", from=1-3, to=2-3]
-      \arrow[""{name=2, anchor=center, inner sep=0}, "{\pull_{c_l}(\delle_c)}", curve={height=-6pt}, from=1-1, to=2-1]
-      \arrow[""{name=3, anchor=center, inner sep=0}, "{e_c}", curve={height=-6pt}, from=1-5, to=2-5]
-      \arrow["\bisim"{description}, draw=none, from=0, to=2]
-      \arrow["\bisim"{description}, draw=none, from=1, to=3]
-    \end{tikzcd}\]
-
-\end{itemize}
-
-We observe that the functoriality of the CBPV connectives on relations up to
-order-equivalence is a direct consequence of the functoriality of the CBPV connectives
-on relations up to quasi-order-equivalence in the step-4 intensional model, since the
-perturbations on both sides of the square are by definition bisimilar to the identity.
-
-Likewise, the retraction properties for the relations
-$\iarr{}: U(D \to F D) \rel D$, $\inat : \nat \rel D$, and $\itimes : D \times D \rel D$
-hold because they held up to delays in the intensional model, and the delays
-disappear in the extensional model construction.
-
-\end{proof}

### 2024-07-11 paper-new/appendix.tex
edits to appendix

@@
 
-Before proceeding with the lemmas, we begin with a definition.
-
-\begin{definition}[representation structure]\label{def:representation-structure}
-  Let $A$ and $A'$ be value objects, and let $c : A \rel A'$ be a relation on
-  the underlying predomains. A \emph{left-representation structure} $\rho^L_c$
-  for $c$ consists of a predomain morphism $e_c : A \to A'$ such that $c$ is
-  quasi-left-representable by $e_c$ (see Definition
-  \ref{def:quasi-left-representable}). We similarly define a
-  {right-representation structure} $\rho^R_c$ for $c$ to consist of a predomain
-  morphism $p_c : A' \to A$ such that $c$ is quasi-right-representable by $p_c$
-  (see Definition \ref{def:quasi-right-representable}).
-  % (Notice that the direction of the morphism is opposite in the definition of
-  % right-representation structure.)
-
-  Likewise, let $d : B \rel B'$ be a computation relation. Left- and
-  right-representation structures for $d$ are defined similarly, except the
-  representing morphisms are now error domain morphisms rather than predomain
-  morphisms.
-\end{definition}
+In this section we prove lemmas about quasi-representability needed for showing
+that our notions of value and computation relation are compositional, and for
+defining the action of the functors $U$, $\li$, $\arr$, and $\times$ on value
+and computation relations.
 
@@
 \end{proof}
+
+Next we show that the identity relation is quasi-representable.
 

### 2024-07-11 paper-new/appendix.tex
fix analysis of CastRight/CastLeft in appendix

@@
 \begin{theorem}
+  \label{thm:thin}
   For any two derivations $c,c' : A \ltdyn A'$ of the same precision
@@
 
-Type precision is a binary relation on typed terms. The original
-gradual guarantee rules are as follows:
+As mentioned in Section~\ref{sec:syntax}, the cast calculus we employ uses a splitting of casts into downcasts and upcasts, whereas gradual surface languages are built out of a single notion of cast
+\[ \inferrule{\Gamma \vdash M : A}{\Gamma \vdash (M :: A') : A' }\]
+The term precision rules for this style of calculus are as follows\cite{siekvitousekciminiboyland}:
 \begin{mathpar}
-  \inferrule
-  {\Gamma^\ltdyn \vdash M \ltdyn M' : c \and
-    c : A \ltdyn A'\and
-    c' : A \ltdyn A_2'
+  \inferrule*[Right=CastRight]
+  {\Delta \vdash M \ltdyn N_1 : c_1 \and
+    c_{1} : A \ltdyn A_1\and
+    c_2 : A \ltdyn A_2
   }
-  {\Gamma^\ltdyn \vdash M \ltdyn (M :: A_2') : c'}
-
-  \inferrule
-  {\Gamma^\ltdyn \vdash M \ltdyn M' : c \and
-    c : A \ltdyn A'\and
-    c' : A_2 \ltdyn A'
+  {\Delta \vdash M \ltdyn (N :: A_2) : c_2}
+
+  \inferrule*[Right=CastLeft]
+  {\Delta \vdash M_1 \ltdyn N : c_1 \and
+    c_1 : A_1 \ltdyn A\and
+    c_2 : A_2 \ltdyn A
   }
-  {\Gamma^\ltdyn \vdash (M :: A_2) \ltdyn M' : c'}
+  {\Delta \vdash (M_1 :: A_2) \ltdyn N : c_2}
 \end{mathpar}
-Where the cast $M :: A_2$ is defined to be
-\[ \dnc {dyn(A_2)}{\upc {dyn(A)} M} \]
-
-These two rules are admissible in our presentation.
-
-For the first rule, we first prove that $\dnc {dyn(A_2)}\upc {dyn(A)} M = \dnc {c'}\upc{c} M$
+New-Ahmed and New-Licata showed that such if the cast $M_1 :: A_2$ for $M_1 : A_1$ is interpreted as
+\[ \dnc {(\dynof {A_2})}{\upc {(\dynof{A_1})} M} \]
+then the CastRight and CastLeft rules are derivable from the UpL/UpR/DnL/DnR rules
+and the \emph{retraction} property of casts: that a downcast after an upcast
+is equal to the identity.
+%
+In the intensional setting, this retraction property needs to be
+weakened: a downcast after an upcast is merely \emph{weakly bisimilar}
+to the identity.
+%
+But this weak bisimilarity is no longer strong enough to show the validity
+of CastRight and CastLeft, due to the lack of transitivity of weak
+bisimilarity.
+
+Fortunately there is an alternative argument that makes CastRight and
+CastLeft valid if we change slightly the interpretation of casts. To
+show this alternative translation we need to first introduce the
+\emph{least upper bound} of types:
+\begin{lemma}
+  The preordered set of types with type precision as ordering has all
+  binary least upper bounds defined as follows:
   \begin{align*}
-  \dnc {dyn(A_2)}\upc {dyn(A)} M
-  &= \dnc {(c \dynof {A'})}\upc{(c'\dynof{A'})} M \tag{All derivations are equal}\\
-  &= \dnc {c}\dnc {\dynof{A'}}\upc{\dynof{A'}}\upc{c'} M\tag{functoriality}\\
-  &= \dnc {c}\upc{c'} M\tag{retraction}\\
+    (A_1 \times A_2) \sqcup (A_1' \times A_2') &= (A_1 \sqcup A_1') \times (A_2 \sqcup A_2')\\
+    (A_1 \ra A_2) \sqcup (A_1' \ra A_2') &= (A_1 \ra A_1') \times (A_2 \ra A_2')\\
+    \nat \sqcup \nat &= \nat\\
+    A \sqcup A' &= \dyn \text{ otherwise }
   \end{align*}
-
-Then the rest follows by the up/dn rules above and the fact that
-precision derivations are all equal.
-
-Thus the following properties are sufficient to provide an extensional
-model of gradual typing without requiring transitivity of term
-precision:
-\begin{enumerate}
-\item Every precision is representable in the above sense,
-\item The association of casts to precision is functorial
-\item Type constructors are covariant functorial with respect to
-  relational identity and composition
-\end{enumerate}
+\end{lemma}
+
+Then for $M_1 : A_1$ we translate $M_1 :: A_2$ to
+\[ \dnc {c^\sqcup_2}\upc{c^\sqcup_1} M \]
+where $c^\sqcup_1 : A_1 \ltdyn A_1\sqcup A_2$ and $c^\sqcup_2 : A_2 \ltdyn
+A_1\sqcup A_2$. In the presence of the strong retraction property, New
+and Ahmed showed that this is equal to the prior interpretation. With
+a weak retraction property, it is still weakly bisimilar to the prior
+interpretation, so for the extensional properties we care about
+ultimately this semantics is equivalent. However our compositional
+arguments do care about this difference, and so we use instead this
+translation that avoids the need for any retraction:
+\begin{theorem}
+  If $M_1 :: A_2$ is interpreted as $\dnc {c^\sqcup_2}\upc{c^\sqcup_1} M$,
+  then CastRight and CastLeft are admissible.
+\end{theorem}
+\begin{proof}
+  We show CastLeft, as this is the one that required retraction in
+  prior work. CastRight follows similarly.
+
+  Assume $M_1 \ltdyn N : c_1$, we seek to prove that
+  \[ \dnc {c^\sqcup_2}\upc{c^\sqcup_1} M \ltdyn N : c_2 \]
+  %
+  Since $A \sqcup A_2$ is the
+  least upper bound, we have there must be a derivation $c^{\sqcup} :
+  A_1\sqcup A_2 \ltdyn A$.
+  %
+  Further by Theorem~\ref{thm:thin}, we have
+  $c^\sqcup_2c^{\sqcup} \equiv c_2$, so it suffices to show
+  \[M_1 \ltdyn \dnc {c^\sqcup_2}\upc{c^\sqcup} N : c^\sqcup_2c^{\sqcup} \]
+  %
+  Then applying DnL and UpL it is sufficient to show
+  \[ M_1 \ltdyn N : c^\sqcup_1c^\sqcup \]
+  But again by Theorem~\ref{thm:thin} we have $c^\sqcup_1c^\sqcup
+  \equiv c_1$ so this follows by our assumption that $M_1 \ltdyn N
+  :c_1$.
+\end{proof}
 

### 2024-07-11 paper-new/appendix.tex
work on appendix

@@
 
-\section{Call-by-push-value}
-
-\subsection{Morphisms of CBPV Models}
-
-\max{there is not definition of a CBPV model...}
-There are two relevant notions of \emph{morphism} of CBPV models:
-\emph{strict} and \emph{lax}.
-% morphisms of CBPV models
-Given call-by-push-value models
-$\mathcal M_1 = (\mathcal V_1, \mathcal E_1, \arr_1, U_1, F_1)$ and
-$\mathcal M_2 = (\mathcal V_2, \mathcal E_2, \arr_2, U_2, F_2)$,
-A \emph{strict} morphism $G$ from $M_1$ to $M_2$ is given by a pair of functors
-$G_{\mathcal{V}}: V_1 \to V_2$ and $G_{\mathcal{E}} : E_1 \to E_2$
-that strictly presere the constructors:
-\begin{enumerate}
-  \item $G_{\mathcal{E}} \circ F_1 = F_2 \circ G_{\mathcal{V}}$
-  \item $G_{\mathcal{V}} \circ U_1 = U_2 \circ G_{\mathcal{E}}$
-  \item $G_{\mathcal{E}}(A \arr_1 B) = G_{\mathcal{V}}(A) \arr_2 G_{\mathcal{E}}(B)$
-  \item $G_{\mathcal{V}}(A_1 \times_1 A_2) = G_{\mathcal{V}}(A_1) \times_2 G_{\mathcal{V}}(A_2)$
-  \item $G_{\mathcal{V}}(1_1) = 1_2$
-\end{enumerate}
-As well as strictly preserving the corresponding universal morphisms
-and coherence isomorphisms.
-
-A lax morphism instead preserves these only up to transformation
-\begin{enumerate}
-  \item $G_{\mathcal{E}} \circ F_1 \Rightarrow F_2 \circ G_{\mathcal{V}}$
-  \item $G_{\mathcal{V}} \circ U_1 \Rightarrow U_2 \circ G_{\mathcal{E}}$
-  \item $G_{\mathcal{E}}(A \arr_1 B) \Rightarrow G_{\mathcal{V}}(A) \arr_2 G_{\mathcal{E}}(B)$
-  \item $G_{\mathcal{V}}(A_1 \times_1 A_2) \Rightarrow G_{\mathcal{V}}(A_1) \times_2 G_{\mathcal{V}}(A_2)$
-  \item $G_{\mathcal{V}}(1_1) \Rightarrow 1_2$
-\end{enumerate}
-Additionally a lax morphism should have a relationship between these
-transformations and the universal morphisms, but we will only consider
-lax morphisms of thin categories, where such conditions hold
-trivially.
-
-\subsection{Kleisli Actions of CBPV Type Constructors}
+\section{Kleisli Actions}
 
@@
 
-% We note that our approach in this section will be differ from the body of the
-% paper in two ways:
-
-% First, we will work \emph{abstractly} with a notion of model of gradual typing
-% defined in the setting of an arbitrary double category.
-% %
-% Second, we build up the abstract notion of model in several phases, and layer
-% the constructions in a modular way so that we can isolate the essential parts of
-% the proofs and abstract away unnecessary details about previous parts of the
-% construction.
-
-% The constructions we carry out in this section specialize to the model we
-% describe in the body of the paper. We find, however, that organizing things in
-% this manner makes the arguments easier to follow.
-
 \subsection{Free Composition of Error Domain Relations}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% quasi-representable by same morphism --> quasi-equivalent
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{lemma}\label{lem:left-rep-by-same-morphism} Let $A$ and $A'$ be value
@@
 
-Next we show that the identity relation is quasi-representable.
-
 %%%%%%%%%%%%
@@
 
-\eric{Identity relation is quasi-representable}
+Next we show that the reflexive relation is quasi-representable.
+
+\begin{lemma}\label{lem:reflexive-rel-quasi-rep}
+  Let $A$ be a value object. Then $r(A)$ is quasi-left-representable and quasi-right-representable.
+  Likewise, $r(B)$ is quasi-left- and quasi-right-representable for any computation object $B$.
+\end{lemma}
+\begin{proof}
+  To show $r(A)$ is quasi-left-representable, we take the embedding to be the
+  identity morphism and the perturbations $\delle = \delre = \id_{M_{A}}$, i.e.,
+  the identtiy of the monoid. Then because $i_A$ is a homomorphism of monoids,
+  it sends the identity element to the identity morphism, so the $\upl$ and
+  $\upr$ squares are just identity squares.
+
+  The same argument shows that $r(A)$ is quasi-right-representable and that
+  $r(B)$ is quasi-left- and quasi-right-representable.
+\end{proof}
 
@@
   Let $A$, $A'$, and $A''$ be value objects and let $c : A \rel A'$ and $c' : A'
-  \rel A''$ be predomain relations. Then we have $\li(c \comp c') \bisim \li c \comp \li c'$
-
+  \rel A''$ be predomain relations. Then we have $\li(c \comp c') \bisim \li c
+  \comp \li c'$.
 \end{lemma}
@@
   quasi-left-represents $c \comp c'$, and then by Lemma
-  \ref{lem:representation-U-F} we have $F(e_c' \circ e_c) = Fe_{c'} \circ Fe_c$
-  quasi-left-represents $F(c \comp c')$.
-  %
-  On the other hand, we also know that $Fe_c$ quasi-left-represents $Fc$ and
-  $Fe_{c'}$ quasi-left-represents $Fc'$ again by Lemma
+  \ref{lem:representation-U-F} we have $\li(e_{c'} \circ e_c) = \li e_{c'} \circ \li e_c$
+  quasi-left-represents $\li (c \comp c')$.
+  %
+  On the other hand, we also know that $\li e_c$ quasi-left-represents $\li c$ and
+  $\li e_{c'}$ quasi-left-represents $\li c'$ again by Lemma
   \ref{lem:representation-U-F}. Then by part (2) of Lemma
-  \ref{lem:representation-comp}, their composition quasi-left-represents $Fc
-  \comp Fc'$.
+  \ref{lem:representation-comp}, their composition quasi-left-represents $\li c
+  \comp \li c'$.
 
@@
 
-    % https://q.uiver.app/#q=WzAsNSxbMCwwLCJcXGxpIEEiXSxbMiwwLCJcXGxpIEEnJyJdLFswLDEsIlxcbGkgQSJdLFsxLDEsIlxcbGkgQSciXSxbMiwxLCJcXGxpIEEnJyJdLFswLDEsIlxcbGkgKGMgXFxjb21wIGMnKSIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzIsMywiXFxsaSBjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMyw0LCJcXGxpIGMnIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwyLCJpX0EoXFxkZWx0YV5sKSIsMl0sWzEsNCwiaV97QScnfShcXGRlbHRhXnIpIl0sWzgsOSwiXFxhbHBoYSIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
+    % https://q.uiver.app/#q=WzAsNSxbMCwwLCJcXGxpIEEiXSxbMiwwLCJcXGxpIEEnJyJdLFswLDEsIlxcbGkgQSJdLFsxLDEsIlxcbGkgQSciXSxbMiwxLCJcXGxpIEEnJyJdLFswLDEsIlxcbGkgKGMgXFxjb21wIGMnKSIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzIsMywiXFxsaSBjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMyw0LCJcXGxpIGMnIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwyLCJpX3tcXGxpIEF9KFxcZGVsdGFebCkiLDJdLFsxLDQsImlfe1xcbGkgQScnfShcXGRlbHRhXnIpIl0sWzgsOSwiXFxhbHBoYSIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
     \[\begin{tikzcd}[ampersand replacement=\&]
@@
       \arrow["{\li (c \comp c')}", "\shortmid"{marking}, no head, from=1-1, to=1-3]
-      \arrow[""{name=0, anchor=center, inner sep=0}, "{i_A(\delta^l)}"', from=1-1, to=2-1]
-      \arrow[""{name=1, anchor=center, inner sep=0}, "{i_{A''}(\delta^r)}", from=1-3, to=2-3]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "{i_{\li A}(\delta^l)}"', from=1-1, to=2-1]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "{i_{\li A''}(\delta^r)}", from=1-3, to=2-3]
       \arrow["{\li c}", "\shortmid"{marking}, no head, from=2-1, to=2-2]
@@
 
-    \eric{Need to change F to $\li$}
-
-    % https://q.uiver.app/#q=WzAsNyxbMCwwLCJGQSJdLFsyLDAsIkZBJyciXSxbMCwxLCJGQSJdLFsxLDEsIkZBJyJdLFsyLDEsIkZBJyciXSxbMCwyLCJGQSJdLFsyLDIsIkZBIl0sWzAsMSwiRihjIFxcY29tcCBjJykiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsyLDMsIkZjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMyw0LCJGYyciLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs1LDYsInIoRkEpIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwyLCJcXGRlbHRhXmwiLDJdLFsyLDUsIlxcZGVsbHBfe0ZjIFxcY29tcCBGYyd9IiwyXSxbMSw0LCJcXGRlbHRhXnIiXSxbNCw2LCJwX3tGYyBcXGNvbXAgRmMnfSJdLFsxMSwxMywiXFxhbHBoYSIsMyx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxMiwxNCwiXFxkbnJfe0ZjIFxcY29tcCBGYyd9IiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+    % https://q.uiver.app/#q=WzAsNyxbMCwwLCJcXGxpIEEiXSxbMiwwLCJcXGxpIEEnJyJdLFswLDEsIlxcbGkgQSJdLFsxLDEsIlxcbGkgQSciXSxbMiwxLCJcXGxpIEEnJyJdLFswLDIsIlxcbGkgQSJdLFsyLDIsIlxcbGkgQSJdLFswLDEsIlxcbGkgKGMgXFxjb21wIGMnKSIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzIsMywiXFxsaSBjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMyw0LCJcXGxpIGMnIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNSw2LCJyKFxcbGkgQSkiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDIsImlfe1xcbGkgQX0oXFxkZWx0YV5sKSIsMl0sWzIsNSwiaV9BKFxcZGVsbHBfe1xcbGkgYyBcXGNvbXAgXFxsaSBjJ30pIiwyXSxbMSw0LCJpX3tcXGxpIEEnJ30oXFxkZWx0YV5yKSJdLFs0LDYsInBfe1xcbGkgYyBcXGNvbXAgXFxsaSBjJ30iXSxbMTIsMTQsIlxcZG5yX3tcXGxpIGMgXFxjb21wIFxcbGkgYyd9IiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzExLDEzLCJcXGFscGhhIiwzLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
     \[\begin{tikzcd}[ampersand replacement=\&]
-      FA \&\& {FA''} \\
-      FA \& {FA'} \& {FA''} \\
-      FA \&\& FA
-      \arrow["{F(c \comp c')}", "\shortmid"{marking}, no head, from=1-1, to=1-3]
-      \arrow["Fc", "\shortmid"{marking}, no head, from=2-1, to=2-2]
-      \arrow["{Fc'}", "\shortmid"{marking}, no head, from=2-2, to=2-3]
-      \arrow["{r(FA)}"', "\shortmid"{marking}, no head, from=3-1, to=3-3]
-      \arrow[""{name=0, anchor=center, inner sep=0}, "{\delta^l}"', from=1-1, to=2-1]
-      \arrow[""{name=1, anchor=center, inner sep=0}, "{\dellp_{Fc \comp Fc'}}"', from=2-1, to=3-1]
-      \arrow[""{name=2, anchor=center, inner sep=0}, "{\delta^r}", from=1-3, to=2-3]
-      \arrow[""{name=3, anchor=center, inner sep=0}, "{p_{Fc \comp Fc'}}", from=2-3, to=3-3]
-      \arrow["\alpha"{marking, allow upside down}, draw=none, from=0, to=2]
-      \arrow["{\dnr_{Fc \comp Fc'}}"{description}, draw=none, from=1, to=3]
+      {\li A} \&\& {\li A''} \\
+      {\li A} \& {\li A'} \& {\li A''} \\
+      {\li A} \&\& {\li A}
+      \arrow["{\li (c \comp c')}", "\shortmid"{marking}, no head, from=1-1, to=1-3]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "{i_{\li A}(\delta^l)}"', from=1-1, to=2-1]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "{i_{\li A''}(\delta^r)}", from=1-3, to=2-3]
+      \arrow["{\li c}", "\shortmid"{marking}, no head, from=2-1, to=2-2]
+      \arrow[""{name=2, anchor=center, inner sep=0}, "{i_A(\dellp_{\li c \comp \li c'})}"', from=2-1, to=3-1]
+      \arrow["{\li c'}", "\shortmid"{marking}, no head, from=2-2, to=2-3]
+      \arrow[""{name=3, anchor=center, inner sep=0}, "{p_{\li c \comp \li c'}}", from=2-3, to=3-3]
+      \arrow["{r(\li A)}"', "\shortmid"{marking}, no head, from=3-1, to=3-3]
+      \arrow["\alpha"{marking, allow upside down}, draw=none, from=0, to=1]
+      \arrow["{\dnr_{\li c \comp \li c'}}"{description}, draw=none, from=2, to=3]
     \end{tikzcd}\]
@@
 
-    % https://q.uiver.app/#q=WzAsOSxbMCwwLCJGQScnIl0sWzIsMCwiRkEnJyJdLFswLDEsIkZBJyciXSxbMiwxLCJGQScnIl0sWzAsMiwiRkEiXSxbMiwyLCJGQScnIl0sWzAsMywiRkEiXSxbMiwzLCJGQScnIl0sWzEsMiwiRkEnIl0sWzAsMSwicihGQScnKSIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzIsMywicihGQScnKSIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzYsNywiRihjIFxcY29tcCBjJykiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDIsIlxcZGVsdGFeciIsMl0sWzIsNCwicF97RmMgXFxjb21wIEZjJ30iLDJdLFs0LDYsIlxcaWQiLDJdLFsxLDMsIlxcZGVsdGFeciJdLFszLDUsIlxcZGVscnBfe0ZjIFxcY29tcCBGYyd9Il0sWzUsNywiXFxpZCJdLFs0LDgsIkZjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbOCw1LCJGYyciLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxMiwxNSwicihcXGRlbHRhXnIpIiwzLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEzLDE2LCJcXGRubF97RmMgXFxjb21wIEZjJ30iLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
+    % https://q.uiver.app/#q=WzAsOSxbMCwwLCJcXGxpIEEnJyJdLFsyLDAsIlxcbGkgQScnIl0sWzAsMSwiXFxsaSBBJyciXSxbMiwxLCJcXGxpIEEnJyJdLFswLDIsIlxcbGkgQSJdLFsyLDIsIlxcbGkgQScnIl0sWzAsMywiXFxsaSBBIl0sWzIsMywiXFxsaSBBJyciXSxbMSwyLCJcXGxpIEEnIl0sWzAsMSwicihcXGxpIEEnJykiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsyLDMsInIoXFxsaSBBJycpIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNiw3LCJcXGxpIChjIFxcY29tcCBjJykiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDIsImlfe0EnJ30oXFxkZWx0YV5yKSIsMl0sWzIsNCwicF97RmMgXFxjb21wIEZjJ30iLDJdLFs0LDYsIlxcaWQiLDJdLFsxLDMsImlfe0EnJ30oXFxkZWx0YV5yKSJdLFszLDUsImlfe0EnJ30oXFxkZWxycF97RmMgXFxjb21wIEZjJ30pIl0sWzUsNywiXFxpZCJdLFs0LDgsIlxcbGkgYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzgsNSwiXFxsaSBjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEyLDE1LCJcXGlkIiwzLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEzLDE2LCJcXGRubF97XFxsaSBjIFxcY29tcCBcXGxpIGMnfSIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
     \[\begin{tikzcd}[ampersand replacement=\&]
-      {FA''} \&\& {FA''} \\
-      {FA''} \&\& {FA''} \\
-      FA \& {FA'} \& {FA''} \\
-      FA \&\& {FA''}
-      \arrow["{r(FA'')}", "\shortmid"{marking}, no head, from=1-1, to=1-3]
-      \arrow["{r(FA'')}", "\shortmid"{marking}, no head, from=2-1, to=2-3]
-      \arrow["{F(c \comp c')}"', "\shortmid"{marking}, no head, from=4-1, to=4-3]
-      \arrow[""{name=0, anchor=center, inner sep=0}, "{\delta^r}"', from=1-1, to=2-1]
-      \arrow[""{name=1, anchor=center, inner sep=0}, "{p_{Fc \comp Fc'}}"', from=2-1, to=3-1]
+      {\li A''} \&\& {\li A''} \\
+      {\li A''} \&\& {\li A''} \\
+      {\li A} \& {\li A'} \& {\li A''} \\
+      {\li A} \&\& {\li A''}
+      \arrow["{r(\li A'')}", "\shortmid"{marking}, no head, from=1-1, to=1-3]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "{i_{A''}(\delta^r)}"', from=1-1, to=2-1]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "{i_{A''}(\delta^r)}", from=1-3, to=2-3]
+      \arrow["{r(\li A'')}", "\shortmid"{marking}, no head, from=2-1, to=2-3]
+      \arrow[""{name=2, anchor=center, inner sep=0}, "{p_{Fc \comp Fc'}}"', from=2-1, to=3-1]
+      \arrow[""{name=3, anchor=center, inner sep=0}, "{i_{A''}(\delrp_{Fc \comp Fc'})}", from=2-3, to=3-3]
+      \arrow["{\li c}", "\shortmid"{marking}, no head, from=3-1, to=3-2]
       \arrow["\id"', from=3-1, to=4-1]
-      \arrow[""{name=2, anchor=center, inner sep=0}, "{\delta^r}", from=1-3, to=2-3]
-      \arrow[""{name=3, anchor=center, inner sep=0}, "{\delrp_{Fc \comp Fc'}}", from=2-3, to=3-3]
+      \arrow["{\li c'}", "\shortmid"{marking}, no head, from=3-2, to=3-3]
       \arrow["\id", from=3-3, to=4-3]
-      \arrow["Fc", "\shortmid"{marking}, no head, from=3-1, to=3-2]
-      \arrow["{Fc'}", "\shortmid"{marking}, no head, from=3-2, to=3-3]
-      \arrow["{r(\delta^r)}"{marking, allow upside down}, draw=none, from=0, to=2]
-      \arrow["{\dnl_{Fc \comp Fc'}}"{description}, draw=none, from=1, to=3]
+      \arrow["{\li (c \comp c')}"', "\shortmid"{marking}, no head, from=4-1, to=4-3]
+      \arrow["\id"{marking, allow upside down}, draw=none, from=0, to=1]
+      \arrow["{\dnl_{\li c \comp \li c'}}"{description}, draw=none, from=2, to=3]
     \end{tikzcd}\]
@@
 \begin{lemma}\label{lem:representation-product} 
-
-  Let $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$. If $c_1$ and $c_2$ are
+  Let $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$. 
+  \begin{enumerate}
+    \item If $c_1$ and $c_2$ are
     quasi-left-representable, then $c_1 \times c_2$ is quasi-left-representable.
-  Furthermore, if $\li c_1$ and $\li c_2$ are quasi-right-representable, then so
+    \item If $\li c_1$ and $\li c_2$ are quasi-right-representable, then so
     is $\li (c_1 \times c_2)$.
+  \end{enumerate}
 
@@
 \begin{proof}
-  To show $c_1 \times c_2$ is quasi-left-representable, we define
+  \begin{enumerate}
+    \item To show $c_1 \times c_2$ is quasi-left-representable, we define
     \begin{itemize}
@@
 
-  To show that $\li(c_1 \times c_2)$ is quasi-right-representable, we define
+    \item To show that $\li(c_1 \times c_2)$ is quasi-right-representable, we define
     \begin{itemize}
@@
       \item $\dellp_{\li (c_1 \times c_2)} = (\dellp_{\li c_1} \timesk \id) \cdot (\id \timesk \dellp_{\li c_2})$ 
-    using the Kleisli product action on syntactic perturbations in Definition \ref{def:kleisli-product-perturbations}
+      using the Kleisli product actions on syntactic perturbations in Definition \ref{def:kleisli-product-perturbations}
       \item $\delrp_{\li (c_1 \times c_2)} = (\delrp_{\li c_1} \timesk \id) \cdot (\id \timesk \delrp_{\li c_2})$
@@
     \end{itemize}
+  \end{enumerate}
 \end{proof}
@@
   %
-  If $c$ is quasi-left-representable and $d$ is quasi-right-representable, then
+  \begin{enumerate}
+    \item If $c$ is quasi-left-representable and $d$ is quasi-right-representable, then
     $c \arr d$ is quasi-right-representable.
 
-  Likewise, if $\li c$ is quasi-right-representable and $Ud$ is
+    \item If $\li c$ is quasi-right-representable and $Ud$ is
     quasi-left-representable, then $U(c \arr d)$ is quasi-left-representable.
+  \end{enumerate}
 
@@
 \begin{proof}
+  \begin{enumerate}
+    \item 
       To show $c \arr d$ is quasi-right-representable, we define:
@@
         \item $\delrp_{c \arr d} = \inl(\delre_c) \cdot \inr(\delrp_d)$
-    \item The squares $\dnr$ and $\dnl$ are obtained via the functorial action of $\arr$, i.e., we define
-
-    \[ \dnr_{c \arr d} = \upr_{c} \arr \dnr_{d} : 
-      (\delle_c \arr \dellp_d) \ltdyn_{r(A \arr B)}^{c \arr d} (e_c \arr p_d),  \]
-
-    and
-
-    \[ \dnl_{c \arr d} = \upl_{c} \arr \dnl_{d}. \]
+        \item The squares $\dnr$ and $\dnl$ are obtained via the functorial action of $\arr$ on squares.
 
@@
     % TODO check this
-  We define $\rho^L_{U(c \arr d)}$ as follows:
-  \begin{itemize}
-    \item $e_{U(c \arr d)} = (p_{Fc} \tok B') \circ (A \tok e_{Ud})$
-    \item $\delre_{U(c \arr d)} = (\delrp_{Fc} \tok B') \circ (A' \tok \delre_{Ud})$
-    \item $\delle_{U(c \arr d)} = (\dellp_{Fc} \tok B) \circ (A \tok \delle_{Ud})$
-    \item The squares $\upl$ and $\upr$ are obtained via the functorial action of $\tok$.
+    \item
+      We show $U(c \arr d)$ is quasi-left-representable as follows:
+      \begin{itemize}
+        \item $e_{U(c \arr d)} = (p_{\li c} \tok B') \circ (A \tok e_{Ud})$
+        \item $\delre_{U(c \arr d)} = (\delrp_{\li c} \tok B') \cdot (A' \tok
+              \delre_{Ud})$ using the Kleisli arrow actions on syntactic
+              perturbations in Definition \ref{def:kleisli-arrow-perturbations}.
+        \item $\delle_{U(c \arr d)} = (\dellp_{\li c} \tok B) \cdot (A \tok \delle_{Ud})$
+        \item The squares $\upl$ and $\upr$ are obtained via the functorial action of $\tok$ on squares.
         For instance, $\upl$ is given by the following square:
@@
     The construction of $\upr$ is similar.
-\end{proof}
-
-% We can now show that our notions of value and computation relations compose,
-% which follows from the preceding lemmas.
+\end{enumerate}
+\end{proof}
+
+\subsection{Model Construction: Composition and Functorial Actions on Relations}
+
+We now show that our notions of value and computation relations compose.
+\begin{definition}[composition of value (resp. computation) relations]\label{def:value-computation-rel-comp}
+  Let $A$, $A'$ and $A''$ be value objects and let $c$ and $c'$ be \emph{value
+  relations} between $A$ and $A'$ and $A'$ and $A''$ respectively. Then we
+  define their composition $cc'$ as follows:
+
+  \begin{itemize}
+    \item The predomain relation is the composition of the underlying predomain
+    relations $cc'$
+    \item The push-pull structure is given by Lemma \ref{lem:push-pull-comp}.
+    \item Quasi-left-representability of $cc'$ follows from Lemma
+    \ref{lem:representation-comp} and the fact that $c$ and $c'$ are
+    quasi-left-representable
+    \item Quasi-right-representability of $\li(cc')$ holds by Lemma
+    \ref{lem:representation-comp-F-U} and the quasi-right-representability of
+    $\li c$ and $\li c'$
+  \end{itemize}
+
+  Likewise, let $B$, $B'$ and $B''$ be computation objects and let $d$ and $d'$ be
+  \emph{computation relations} between $B$ and $B'$ and $B'$ and $B''$
+  respectively. We define their composition $dd'$ as follows:
+
+  \begin{itemize}
+    \item The error domain relation is the composition of the underlying error
+    domain relations $dd'$
+    \item The push-pull structure is given by Lemma \ref{lem:push-pull-comp}.
+    \item Quasi-right-representability of $dd'$ follows from Lemma
+    \ref{lem:representation-comp} and the fact that $d$ and $d'$ are
+    quasi-right-representable
+    \item Quasi-left-representability of $U(dd')$ holds by Lemma
+    \ref{lem:representation-comp-F-U} and the quasi-left-representability of
+    $Ud$ and $Ud'$
+  \end{itemize}
+\end{definition}
+
+We can define the functorial actions of $\li$, $U$, $\times$, and $\arr$ on
+value and computation relations as follows:
+\begin{definition}[functorial actions on relations]\label{def:functorial-actions-on-relations}
+
+  \begin{enumerate}
+
+    \item Let $A$ and $A'$ be value objects and $c : A \rel A'$ a value relation.
+          Then we define the computation relation $\li c$ as follows:
+      \begin{itemize}
+        \item The error domain relation is given by the action of $\li$ on the
+        underlying predomain relation of $c$
+        \item The push-pull structure is given by Lemma \ref{lem:push-pull-U-F}
+        \item Quasi-right-representability of $\li c$ holds because it is part
+        of the definition of the value relation $c$
+        \item Quasi-left-representability of $U \li c$ follows from Lemma
+        \ref{lem:representation-U-F} and the quasi-left-representability of $c$
+      \end{itemize}
+
+    \item Let $B$ and $B'$ be computation objects and $d : B \rel B'$ a computation relation.
+          We define $Ud$ as follows:
+          \begin{itemize}
+            \item The predomain relation is given by the action of $U$ on the
+            error domain relation of $d$
+            \item The push-pull structure is given by Lemma
+            \ref{lem:push-pull-U-F}
+            \item Quasi-left-representability of $Ud$ holds because it is part
+            of the definition of the computation relation $d$
+            \item Quasi-right-representability of $\li(Ud)$ follows from Lemma
+            \ref{lem:representation-U-F} and the quasi-right-representability of
+            $d$
+          \end{itemize}
+
+    \item Let $c_1 : A_1 \rel A_1'$ and $c_2 : A_2 \rel A_2'$ be value relations.
+          We define $c_1 \times c_2$ as follows:
+          \begin{itemize}
+            \item The relation is given by the action of $\times$ on the
+            predomain relations of $c_1$ and $c_2$
+            \item The push-pull structure is given by Lemma
+            \ref{lem:push-pull-times}
+            \item Quasi-left-representability of $c_1 \times c_2$ and
+            quasi-right-representabiity of $\li(c_1 \times c_2)$ are given by
+            Lemma \ref{lem:representation-product}
+          \end{itemize}
+
+    \item Let $c : A \rel A'$ be a value relation and $d : B \rel B'$ be a computation relation.
+          We define $c \arr d$ as follows:
+          \begin{itemize}
+            \item The error domain relation is given by the action of $\arr$ on
+            the underlying relations of $c$ and $d$
+            \item The push-pull structure is given by Lemma
+            \ref{lem:push-pull-arrow}
+            \item Quasi-right-representability of $c \arr d$, and
+                  quasi-left-representability of $U(c \arr d)$ follow from Lemma
+                  \ref{lem:representation-arrow}
+          \end{itemize}
+
+  \end{enumerate}
+\end{definition}
+
+Lastly, we establish the quasi-order-equivalence for the functors.
+
+\begin{lemma}
+  The following hold:
+  \begin{itemize}
+    \item $U(d \comp d') \qordeq U(d) \comp U(d')$
+    \item $\li(c \comp c') \qordeq \li (c) \comp \li (c')$
+    \item $(c \comp c') \to (d \comp d') \qordeq (c \to d) \comp (c' \to d')$
+    \item $(c_1 \comp c_1') \times (c_2 \comp c_2') \qordeq (c_1 \times c_2) \comp (c_1'\times c_2')$
+  \end{itemize}
+\end{lemma}
 

### 2024-07-11 paper-new/appendix.tex
laborious account of functoriality and sq preservation for Kleisli operations

@@
 
-In CBPV models, all the type constructors are interpreted as functors:
+We now define the Kleisli actions of the type constructors. These are the following
 \begin{enumerate}
-\item $\to : \op\calV \times \calE \to \calE$
-\item $\times : \calV \times \calV \to \calV$
-\item $F : \calV \to \calE$
-\item $U : \calE \to \calV$
+\item $- \tok B : \op\PreDom_k \to \ErrDom_k$
+\item $A \tok - : \ErrDom_k \to \ErrDom_k$
+\item $- \timesk A_2 : \PreDom_k \to \PreDom_k$
+\item $A_1 \timesk - : \PreDom_k \to \PreDom_k$
 \end{enumerate}
-That is, they all have functorial actions on \emph{pure} morphisms of
-value types and \emph{linear} morphisms of computation types.
-%
-We use these functorial actions extensively in the construction of
-casts and their corresponding perturbations. But when defining
-downcasts of value types and upcasts of computation types, we
-additionally need a second functorial action of these categories:
-functoriality in \emph{impure} morphisms of value types and
-\emph{non-linear} morphisms of computation types. These notions of
-morphism are given by the \emph{Kleisli} categories $\calVk$ and
-$\calEk$ which have value types and computation types as objects but
-morphisms are defined as
-\[ \calVk(A,A') = \calE(F A, FA')\]
-\[ \calEk(B,B') = \calV(U B, U B')\]
-with composition given by composition in $\calE/\calV$.  That is we
-need to define a second functorial action, that agrees with the above
-on objects for these Kleisli categories:
-\begin{enumerate}
-\item $\tok : \op\calVk \square \calEk \to \calEk$
-\item $\timesk : \calVk \square \calVk \to \calVk$
-\item $\Fk : \calVk \to \calEk$
-\item $\Uk : \calEk \to \calVk$
-\end{enumerate}
-Note that rather than the product of categories we use the ``funny
-tensor product'' $\square$. This is because the action on
-impure/non-linear morphisms for $\tok/\timesk$ do not satisfy ``joint
-functoriality'' but instead only ``separate functoriality'', meaning
-we give rather than an action on morphisms in both categories
-simultaneously instead an action on each argument categories morphisms
-with the object in the other category fixed. The existence of these
-functorial actions for $\tok$ and $\timesk$ is reliant on the
-\emph{strength} of the adjunction. We describe them using the internal
-language of CBPV in order to more easily verify their
-existence/functoriality:
-\begin{enumerate}
-\item For $\tok$ we define for $\phi : \calE(F A,F A')$ and $B \in \calE$ the morphism $\phi \tok B : \calV(U(A' \to B),U(A\to B))$ as
-  \[ t:U(A'\to B) \vdash \phi \tok B = \{ \lambda x. x' \leftarrow \phi\,[\ret x]; ! t x'\} : U(A \to B) \]
-  and for $A \in \calV$ and $f : \calV(UB,UB')$ we define $A \tok f : \calV(U(A \to B),U(A\to B'))$ as
-  \[ t : U(A \to B) \vdash A \tok f = \{ \lambda x. !f[\{ ! t x \}]\} \]
-\item For $\timesk$ we define for $\phi : \calE(F A_1,FA_2)$ and $A' \in \calV$ the morphism $\phi \timesk A_2$ as
-  \[ \bullet : F(A_1\times A_2) \vdash \phi \timesk A_2 = (x_1,x_2) \leftarrow \bullet; x_1' \leftarrow \phi[\ret x_1]; \ret (x_1',x_2) : F(A_1'\times A_2)\]
-  and $A_1 \timesk \phi$ is defined symmetrically.
-\item For $\Uk$ we need to define for $f : \calV(UB,UB')$ a morphism $\Uk f : \calE(FUB,FUB')$. This is simply given by the functorial action of $F$: $\Uk f = F(f)$
-\item Similarly $\Fk \phi = U\phi$
-\end{enumerate}
-
-Functoriality in each argument is easily established, meaning for
-example for the function type is functorial in each argument:
-\begin{enumerate}
-\item $(\phi \circ \phi') \tok B = (\phi' \tok B) \circ (\phi \tok B)$
-\item $\id \tok B = \id$
-\item $A \tok (f \circ f') = (A \tok f) \circ (A \tok f)$
-\item $A \tok \id = \id$
-\end{enumerate}
-
-Finally, note that all of these constructions lift to squares in a
-double CBPV model since the squares themselves form a CBPV model and
-the projection functions preserve CBPV structure. For instance, given a square
-$\alpha : \phi \ltdyn_{F c_o}^{F c_i} \phi'$ and a horizontal morphism $d : B \rel B'$ of appropriate type, we get a square
-\[ \alpha \tok d : \phi \tok B \ltdyn_{U(c_o \to d)}^{U(c_i \to d)} \phi' \tok B' \]
+While these can be defined in any CBPV model, for simplicity we just
+give their definition explicitly for predomains and error domains:
+\begin{definition}{~}
+  Given $\phi : \li A_i \multimap \li A_o$, we define $\phi \tok B : U(A_o \to B) \to U(A_i \to B)$ as
+  \[ (\phi \tok B)(f)(x) = f^\dagger(\phi(\eta(x))) \]
+
+  This is functorial in that $\id \tok B = \id$ and $(\phi \circ
+  \phi') \tok B = (\phi' \tok B) \circ (\phi \tok B)$.
+
+  Further, this preserves squares in that if $\phi \ltsq{\li c_i}{\li c_o} \phi'$ then $\phi \tok B \ltsq{U(c_o \to d)}{U(c_i \to d)} \phi' \tok B'$.
+\end{definition}
+\begin{proof}
+  {~}
+  \begin{itemize}
+  \item For identity
+    \[ (\id \tok B)(f)(x) = f^\dagger(\eta(x)) = f(x) \]
+
+  \item 
+    For composition, expanding definitions it suffices to show
+    \[ f^\dagger(\phi(\phi'(\eta(x)))) = ((\phi \tok B)(f))^\dagger(\phi'(\eta(x))) \]
+    Which follows if we can show
+    \[ ((\phi \tok B)(f))^\dagger = f^\dagger \circ \phi \]
+    By the freeness of $\li -$ it suffices to show it for inputs of the form $\eta y$:
+    \[ ((\phi \tok B)(f))^\dagger(\eta y) = ((\phi \tok B)(f))(y) = f^\dagger(\phi(\eta y))\]
+  \item For squares, we assume $f \binrel{U(c_o\to d)} f'$ and $x
+    \binrel {c_i} x'$ and we need to show $f^\dagger(\phi(x)) \binrel
+    d (f')^\dagger(\phi'(x))$. This follows easily since $f^\dagger
+    \ltsq{\li c_o}{d} f^{\prime\dagger}$.
+  \end{itemize}
+\end{proof}
+
+\begin{definition}
+  Given $f : UB \to UB'$, define $A \tok f : U(A \to B) \to U(A \to B')$ as
+    \[ (A \tok f)(g)(x) = f(g(x)) \]
+
+  This is functorial in that $A \tok \id = \id$ and $A \tok (f \circ
+  f') = (A \tok f) \circ (A \tok f')$.
+
+  Further this preserves squares in that if $f \ltsq{Ud_i}{Ud_o} f'$
+  then $A \tok f \ltsq{U(c \to d_i)}{U(c \to d_o)} A' \tok f'$
+\end{definition}
+\begin{proof}
+  \begin{itemize}
+  \item For identity
+    \[ (A \tok \id)(g)(x) = g(x) \]
+  \item 
+    for composition
+    \[  (A \tok (f \circ f'))(g)(x) = f(f'(g(x))) = (A\tok f)((A \tok f')(g))(x) \]
+  \item 
+   for squares, given $g \binrel{U(c \to d_i)} g'$ and $x \binrel{c}
+   x'$ we have $f(g(x)) \binrel{d_o} f'(g'(x'))$.
+  \end{itemize}
+\end{proof}
+
+\begin{definition}[Kleisli Actions of $\times$]
+  Given $\phi_1 : \li A_1 \multimap \li A_1'$ we define $\phi_1 \timesk A_2 : \li (A_1 \times A_2) \multimap (A_1' \times A_2)$ as the unique extension of the following to a homomorphism:
+  \[ (\phi_1 \timesk A_2)(\eta (x_1,x_2)) = (\lambda x_1'. \eta(x_1',x_2))^\dagger(\phi_1(\eta(x_1))) \]
+  and similarly define
+  \[ (A_1 \timesk \phi_2)(\eta (x_1,x_2)) = (\lambda x_2'. \eta(x_1,x_2'))^\dagger(\phi_2(\eta(x_2)))\]
+\end{definition}
+\begin{proof}
+  We show cases only for $-\timesk A_2$ as the other is symmetric.
+  \begin{itemize}
+  \item For identity it is sufficient to consider inputs of the form $\eta(x_1,x_2)$, then indeed
+    \[ (\id \timesk A_2)(\eta(x_1,x_2)) = (\lambda x_1'. \eta(x_1',x_2))^\dagger(\eta(x_1)) = \eta(x_1,x_2) \]
+  \item For composition, again for pure inputs this reduces to
+    \[ (\lambda x_1'. \eta(x_1',x_2))^\dagger(\phi_1'(\phi_1(\eta(x_1))))
+    = (\phi_1'\timesk A_2)((\lambda x_1'. \eta(x_1',x_2))^\dagger(\phi_1(\eta(x_1)))) \]
+    So it suffices to show
+    \[ (\lambda x_1'. \eta(x_1',x_2))^\dagger \circ \phi_1' = (\phi_1'\timesk A_2) \circ (\lambda x_1'. \eta(x_1',x_2))^\dagger \]
+    Since both sides are homomorphisms out of the free error domain by freeness it suffices to show they are equal for inputs of the form $\eta(x_1')$:
+    \[  (\lambda x_1'. \eta(x_1',x_2))^\dagger(\phi_1'(\eta(x_1')))
+    = (\phi_1'\timesk A_2)(\eta(x_1',x_2))
+    = (\phi_1'\timesk A_2)((\lambda x_1'. \eta(x_1',x_2))^\dagger(\eta(x_1')))
+    \]
+
+  \item Finally, for squares, we need to show if $\phi_l \ltsq{\li
+    c_1}{\li c_1'} \phi_r$ then $\phi_l \timesk A_{2l} \ltsq{\li (c_1
+    \times c_2)}{\li (c_1' \times c_2)} \phi_r \timesk A_{2r}$.
+
+    By the universal property of $\li-$ on relations, it suffices to
+    assume $x_{1l} \binrel{c_1} x_{1r}$ and $x_{2l} \binrel{c_2} x_{2r}$ and prove
+    \[ (\phi_l \timesk A_{2l})(\eta(x_{1l},x_{2l})) \binrel{(\li (c_1' \times c_2))} (\phi_r \timesk A_{2r})(\eta(x_{1r},x_{2r})) \]
+    expanding definitions this is
+    \[ (\lambda x_{1l}'. \eta(x_{1l}',x_{2l}))^\dagger\phi_l(\eta(x_{1l}))
+    \binrel{(\li (c_1' \times c_2))}
+    (\lambda x_{1r}'. \eta(x_{1r}',x_{2r}))^\dagger\phi_r(\eta(x_{1r}))
+    \]
+    By our assumptions we have $\phi_l(\eta(x_{1l}) \binrel {\li c_1'} \phi_r(\eta(x_{1r})$ so since $-^\dagger$ preserves squares it suffices to show
+    \[ (\lambda x_{1l}'. \eta(x_{1l}',x_{2l})) \ltsq{\li c_1}{\li (c_1' \times c_2)} (\lambda x_{1r}'. \eta(x_{1r}',x_{2r})) \]
+    To show this assume $x_{1l}' \binrel{c_1'} x_{1r}'$. Then we have
+    \[ \eta(x_{1l}',x_{2l}) \binrel{\li (c_1' \times c_2)} \eta(x_{1r'},x_{2l})  \]
+    holds.
+  \end{itemize}
+\end{proof}
 

### 2024-07-12 paper-new/appendix.tex
small changes to appendix

@@
 
-  % https://q.uiver.app/#q=WzAsNixbMCwwLCJBIl0sWzEsMCwiQSJdLFsyLDAsIkEnIl0sWzAsMSwiQSJdLFsxLDEsIkEnIl0sWzIsMSwiQSciXSxbMCwxLCJyKEEpIl0sWzEsMiwiYyJdLFszLDQsImMnIiwyXSxbNCw1LCJyKEEnKSIsMl0sWzAsMywiaV9BKFxcZGVsbGVfe2MnfSkiLDJdLFsxLDQsImYiLDJdLFsyLDUsImlfe0EnfShcXGRlbHJlX2MpIl1d
+  % https://q.uiver.app/#q=WzAsNixbMCwwLCJBIl0sWzEsMCwiQSJdLFsyLDAsIkEnIl0sWzAsMSwiQSJdLFsxLDEsIkEnIl0sWzIsMSwiQSciXSxbMCwxLCJyKEEpIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMyw0LCJjJyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzQsNSwicihBJykiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDMsImlfQShcXGRlbGxlX3tjJ30pIiwyXSxbMSw0LCJmIiwyXSxbMiw1LCJpX3tBJ30oXFxkZWxyZV9jKSJdXQ==
   \[\begin{tikzcd}[ampersand replacement=\&]
@@
     A \& {A'} \& {A'}
-    \arrow["{r(A)}", from=1-1, to=1-2]
+    \arrow["{r(A)}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
     \arrow["{i_A(\delle_{c'})}"', from=1-1, to=2-1]
-    \arrow["c", from=1-2, to=1-3]
+    \arrow["c", "\shortmid"{marking}, no head, from=1-2, to=1-3]
     \arrow["f"', from=1-2, to=2-2]
     \arrow["{i_{A'}(\delre_c)}", from=1-3, to=2-3]
-    \arrow["{c'}"', from=2-1, to=2-2]
-    \arrow["{r(A')}"', from=2-2, to=2-3]
+    \arrow["{c'}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+    \arrow["{r(A')}"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
   \end{tikzcd}\]
@@
 
-    % https://q.uiver.app/#q=WzAsOSxbMCwwLCJcXGxpIEEnJyJdLFsyLDAsIlxcbGkgQScnIl0sWzAsMSwiXFxsaSBBJyciXSxbMiwxLCJcXGxpIEEnJyJdLFswLDIsIlxcbGkgQSJdLFsyLDIsIlxcbGkgQScnIl0sWzAsMywiXFxsaSBBIl0sWzIsMywiXFxsaSBBJyciXSxbMSwyLCJcXGxpIEEnIl0sWzAsMSwicihcXGxpIEEnJykiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsyLDMsInIoXFxsaSBBJycpIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNiw3LCJcXGxpIChjIFxcY29tcCBjJykiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDIsImlfe0EnJ30oXFxkZWx0YV5yKSIsMl0sWzIsNCwicF97RmMgXFxjb21wIEZjJ30iLDJdLFs0LDYsIlxcaWQiLDJdLFsxLDMsImlfe0EnJ30oXFxkZWx0YV5yKSJdLFszLDUsImlfe0EnJ30oXFxkZWxycF97RmMgXFxjb21wIEZjJ30pIl0sWzUsNywiXFxpZCJdLFs0LDgsIlxcbGkgYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzgsNSwiXFxsaSBjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEyLDE1LCJcXGlkIiwzLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEzLDE2LCJcXGRubF97XFxsaSBjIFxcY29tcCBcXGxpIGMnfSIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
+    % https://q.uiver.app/#q=WzAsOSxbMCwwLCJcXGxpIEEnJyJdLFsyLDAsIlxcbGkgQScnIl0sWzAsMSwiXFxsaSBBJyciXSxbMiwxLCJcXGxpIEEnJyJdLFswLDIsIlxcbGkgQSJdLFsyLDIsIlxcbGkgQScnIl0sWzAsMywiXFxsaSBBIl0sWzIsMywiXFxsaSBBJyciXSxbMSwyLCJcXGxpIEEnIl0sWzAsMSwicihcXGxpIEEnJykiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsyLDMsInIoXFxsaSBBJycpIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNiw3LCJcXGxpIChjIFxcY29tcCBjJykiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDIsImlfe0EnJ30oXFxkZWx0YV5yKSIsMl0sWzIsNCwicF97XFxsaSBjIFxcY29tcCBcXGxpIGMnfSIsMl0sWzQsNiwiXFxpZCIsMl0sWzEsMywiaV97QScnfShcXGRlbHRhXnIpIl0sWzMsNSwiaV97QScnfShcXGRlbHJwX3tcXGxpIGMgXFxjb21wIFxcbGkgYyd9KSJdLFs1LDcsIlxcaWQiXSxbNCw4LCJcXGxpIGMiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs4LDUsIlxcbGkgYyciLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxMiwxNSwiXFxpZCIsMyx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxMywxNiwiXFxkbmxfe1xcbGkgYyBcXGNvbXAgXFxsaSBjJ30iLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
     \[\begin{tikzcd}[ampersand replacement=\&]
@@
       \arrow["{r(\li A'')}", "\shortmid"{marking}, no head, from=2-1, to=2-3]
-      \arrow[""{name=2, anchor=center, inner sep=0}, "{p_{Fc \comp Fc'}}"', from=2-1, to=3-1]
-      \arrow[""{name=3, anchor=center, inner sep=0}, "{i_{A''}(\delrp_{Fc \comp Fc'})}", from=2-3, to=3-3]
+      \arrow[""{name=2, anchor=center, inner sep=0}, "{p_{\li c \comp \li c'}}"', from=2-1, to=3-1]
+      \arrow[""{name=3, anchor=center, inner sep=0}, "{i_{A''}(\delrp_{\li c \comp \li c'})}", from=2-3, to=3-3]
       \arrow["{\li c}", "\shortmid"{marking}, no head, from=3-1, to=3-2]
@@
 
-\eric{Left off here}
-
 %%%%%%%%%
@@
 
-\begin{lemma}
+\begin{lemma}\label{lem:quasi-order-equiv-functors}
   The following hold:
@@
 \end{lemma}
+\begin{proof}
+  (1) and (2) were already shown in Lemmas \ref{lem:Fcc-equiv-FcFc'} and
+  \ref{lem:Udd-equiv-UdUd'}. (3) follows from observing that both relations are
+  right-represented by the morphism $e_c'e_c \arr p_{d}p_{d'}$ and (4) from the
+  fact that both are left-represented by the morphism $e_{c_1'} e_{c_1} \times
+  e_{c_2'} e_{c_2}$
+\end{proof}
 
@@
 % quasi-represented by the same morphism.
+
+\subsection{Definitions of Error Ordering and Weak Bisimilarity for the Delay Monad}
+\label{sec:relations-on-delay-monad}
+
+Correspondingly, we define a notion of lock-step error ordering and weak
+bisimilarity relation for the coinductive delay monad $\delay(\mathbb{N} + {\mho})$:
+
+The lock-step error ordering is defined coinductively by the following rules:
+
+%
+\begin{mathpar}
+  \inferrule*[]
+  { }
+  {\tnow (\inr\, 1) \ledelay d}
+
+  \inferrule*[]
+  {x_1 \le_X x_2}
+  {\tnow (\inl\, x_1) \ledelay \tnow (\inl\, x_2)}
+
+  \inferrule*[]
+  {d_1 \ledelay d_2}
+  {\tlater\, d_1 \ledelay \tlater\, d_2}
+\end{mathpar}
+%
+And we similarly define by coinduction a ``weak bisimilarity'' relation on
+$\delay(\mathbb{N} + {\mho})$. This uses a relation $d \Da x_?$ between
+$\delay(\mathbb{N} + {\mho})$ and $\mathbb{N} + {\mho}$ that is defined as $d
+\Da n_? := \Sigma_{i \in \mathbb{N}} d = \tlater^i(\tnow\, n_?)$. Then weak
+bisimilarity for the delay monad is defined coinductively by the rules
+%
+\begin{mathpar}
+  \inferrule*[]
+  {n_? \bisim_{\mathbb{N} + {\mho}} m_?}
+  {\tnow\, n_? \bisimdelay \tnow\, m_? }
+
+  \inferrule*[leftskip=1.5em]
+  {d_1 \Da n_? \and n_? \bisim_{\mathbb{N} + {\mho}} m_?}
+  {\tlater\, d_1 \bisimdelay \tnow\, m_? }
+
+  \inferrule*[leftskip=1.5em]
+  {d_2 \Da m_? \and n_? \bisim_{\mathbb{N} + {\mho}} m_?}
+  {\tnow\, n_? \bisimdelay \tlater\, d_2}
+
+  \inferrule*[leftskip=1.5em]
+  {d_1 \bisimdelay d_2}
+  {\tlater\, d_1 \bisimdelay \tlater\, d_2 }
+  %
+  % \inferrule*[]
+  % {d_1 \Da x_? \and d_2 \Da y_? \and x_? \bisim_{X + 1} y_?}
+  % {d_1 \bisimdelay d_2}
+  %
+  % \inferrule*[]
+  % {d_1 \bisimdelay d_2}
+  % {\tlater d_1 \bisimdelay \tlater d_2 }
+\end{mathpar}
 

### 2024-07-12 paper-new/concrete-relational-model.tex
more comments
updates to relational model section: discuss interpretation of syntax and revise section on adequacy

@@
 \begin{itemize}
-    \item A distinguished ``error'' element $\mho_B \in B$
+    \item A distinguished ``error'' element $\mho_B \in B$ such that $\mho \le_B x$ for all $x$
     \item A morphism of predomains $\theta_B \colon \later B \to B$
@@
   \begin{enumerate}
-     \item (Respects error): For all $y \in B'$, we have $(\mho_B, y) \in d$.\max{can we please use infix notation consistently for relations?}
+     \item (Respects error): For all $y \in B'$, we have $\mho_B \binrel{d} y$.\max{can we please use infix notation consistently for relations?}
      \item (Preserves $\theta$): For all $\tilde{x}$ in $\laterhs B$ and
-     $\tilde{y} \in \laterhs B'$, if $\laterhs_t ((\tilde{x}_t, \tilde{y}_t) \in
-     d)$, then $(\theta_B (\tilde{x}), \theta_{B'} (\tilde{y})) \in d$. 
+     $\tilde{y} \in \laterhs B'$, if $\laterhs_t (\tilde{x}_t \binrel{d} \tilde{y}_t)$, then 
+     $(\theta_B (\tilde{x}) \binrel{d} \theta_{B'} (\tilde{y}))$. 
   \end{enumerate}
@@
 %
-As with predomain relations, we can compose error domain
-relations. However, because of a technical restriction\max{technical
-  restriction?? isn't is simpler/correct to say that the composition
-  of underlying relations is not an error domain relation?} involving
-the interaction between propositional truncation and the later
-modality, we cannot define the composition of error domain relations
-as the composition of their underlying predomain relations. Instead,
-we define composition of error domain relations $d$ on $B_1$ and $B_2$
-and $d'$ on $B_2$ and $B_3$ inductively to be the least relation
-containing $d$ and $d'$ that is downward closed under $\le_{B_1}$,
-upward-closed under $\le_{B_3}$, respects error, and preserves
-$\theta$.  Specifically, it is defined inductively by the following
-rules:\max{what rules??}
-% %
-% \begin{mathpar}
-%     \inferrule*[right = Comp]
-%     {b_1 \mathbin{d} b_2 \and b_2 \mathbin{d'} b_3}
-%     {b_1 \mathbin{d \relcomp d'} b_3}
-
-%     \inferrule*[right = DnClosed]
-%     {b_1' \le_{B_1} b_1 \and b_1 \mathbin{d \relcomp d'} b_3}
-%     {b_1' \mathbin{d \relcomp d'} b_3}
-
-%     \inferrule*[right = UpClosed]
-%     {b_1 \mathbin{d \relcomp d'} b_3 \and b_3 \le_{B_3} b_3'}
-%     {b_1 \mathbin{d \relcomp d'} b_3'}
-
-%     \inferrule*[right = PresErr]
-%     { }
-%     {\mho_{B_1} \mathbin{d \relcomp d'} b_3}
-
-%     \inferrule*[right = PresTheta]
-%     {\later_t( \tilde{b_1} \mathbin{d \relcomp d'} \tilde{b_3} ) }
-%     {\theta_{B_1}(\tilde{b_1}) \mathbin{d \relcomp d'} \theta_{B_3}(\tilde{b_3}) }
-% \end{mathpar}
-% %
+As with predomain relations, we can compose error domain relations. However, the
+composition of the underlying predomain relations is not an error domain
+relation.
+% Can explain this further in the appendix
+Instead, we define composition of error domain relations $d$ on $B_1$ and $B_2$
+and $d'$ on $B_2$ and $B_3$ inductively to be the least relation containing $d$
+and $d'$ that is downward closed under $\le_{B_1}$, upward-closed under
+$\le_{B_3}$, respects error, and preserves $\theta$. See Definition
+\ref{def:free-comp-ed-rel} in the appendix for the precise definition.
 
@@
 \in (c \arr d)$ if and only if for all $x : A$ and $x' : A'$ with $(x , x') \in
-c$ we have $(f(x), g(x')) \in Ud$. The fact that this relation preserves error\max{preserves error?}
-and respects $\theta$ follows from the corresponding properties for $d$.
+c$ we have $(f(x), g(x')) \in Ud$. The fact that this relation respects error\max{preserves error?}
+and preserves $\theta$ follows from the corresponding properties for $d$.
 
@@
 % morphism on $A$. 
-More formally, for every $A$ we require a homomorphism of monoids $i_A : M_A \to
-\morbisimid{A} := \{ f : A \to A \mid f \bisim \id_A \}$. We likewise equip
-every error domain $B$ with a monoid $M_B$ and interpretation homomorphism $i_B
-: M_B \to \morbisimid{B}$. This grouping of data deserves its own definition: We
-call the triple $(A, M_A, iA)$ a \textbf{value object} as it will be the
-denotation of value types in our final model. Likewise, the triple $(B, M_B,
-i_B)$ will be called a \textbf{computation object}.\max{make this a proper \textit{Definition} and define morphisms as well}
+More formally:
+\begin{definition}[value and computation objects + morphisms]
+  A \textbf{value object} consists of a predomain $A$, a monoid $M_A$ and a
+  homomorphism of monoids $M_A \to \morbisimid{A} := \{ f : A \to A \mid f
+  \bisim \id_A \}$.
 
-\max{idea: move this after the quasi-representability part and say that this structure is what's needed to make them compose}
-We must impose an additional condition on predomain relations $c$ and error
-domain relations $d$ that specify how they interact with these perturbations. In
-particular, we need to be able to ``push'' and ``pull'' perturbations along
-relations $c$ and $d$. The intuition for this requirement comes from the
-construction of the square corresponding to the original UpL rule from the
-square for the simplified version as was shown in Section
-\ref{sec:towards-relational-model}. Recall that we horizontally composed the
-simplified UpL square for $c$ with the identity square for $c'$ on the right.
-Now that the square for UpL involves a perturbation on the right instead of the
-identity, this construction will not work unless we can ``push'' the
-perturbation on $A_2$ to one on $A_3$. That is, given a relation $c : A \rel A'$
-and a syntactic perturbation $m_A \in M_A$, we need to be able to turn it into a
-syntactic perturbation $m_{A'} \in M_{A'}$ such that the resulting semantic
-perturbations obtained by applying the respective homomorphisms $i_A$ and
-$i_{A'}$ form a square. Dually, we will need to ``pull'' a perturbation from
-right to left. This notion is made formal by the following definition:
-%
-\begin{definition}
-    Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects,
-    and let $c : A \rel A'$ be a predomain relation. A
-    \textbf{push-pull structure} for $c$ (with respect to $M_A$ and
-    $M_{A'}$) consists of monoid homomorphisms $\push : M_A \to M_A'$
-    and $\pull : M_A' \to M_A$ such that for all $m_A \in M_A$ there
-    is a square $i_A(m_A) \ltsq{c}{c} i_{A'}(\push\, m_A)$ and for all
-    $m_{A'} \in M_{A'}$ there is a square $i_A(\pull\, m_{A'})
-    \ltsq{c}{c} i_{A'}(m_{A'})$. A push-pull structure for error
-    domain relations is defined similarly.
+  Likewise, a \textbf{computation object} consists of an error domain $B$, a
+  monoid $M_B$ and a homomorphism of monoids $M_B \to \morbisimid{B}$.
+
+  We define a \textbf{value morphism} to be a morphism of the underlying
+  predomains, and a \textbf{computation morphism} to be a morphism of the
+  underlying error domains.
 \end{definition}
 
-With the notion of syntactic perturbation defined, we can now make formal the
-final aspect of the model: the notion of quasi-represntability of relations.
+% every error domain $B$ with a monoid $M_B$ and interpretation homomorphism $i_B
+% : M_B \to \morbisimid{B}$. This grouping of data deserves its own definition: We
+% call the triple $(A, M_A, iA)$ a \textbf{value object} as it will be the
+% denotation of value types in our final model. Likewise, the triple $(B, M_B,
+% i_B)$ will be called a \textbf{computation object}.\max{make this a proper \textit{Definition} and define morphisms as well}
+
+Having defined syntactic perturbations and value and computation objects, we can
+now make formal the final aspect of the model: the notion of
+quasi-represntability of relations.
 %
@@
 
-\max{see above comment about reordering these concepts}
-do not in general compose: in order to construct
-the needed squares for the composition of two relations, we require that the
-relations have push-pull structures.
+% \max{idea: move this after the quasi-representability part and say that this structure is what's needed to make them compose}
+Given two quasi-representable relations $c$ and $c'$, it is not the case in
+general that $cc'$ is quasi-representable. We need an additional condition that
+specifies how relations interact with perturbations. In particular, we need to
+be able to ``push'' and ``pull'' perturbations along relations $c$ and $d$.
+%
+The intuition for this requirement comes from the construction of the square
+corresponding to the original UpL rule from the square for the simplified
+version as was shown in Section \ref{sec:towards-relational-model}. Recall that
+we horizontally composed the simplified UpL square for $c$ with the identity
+square for $c'$ on the right. Now that the square for UpL involves a
+perturbation on the right instead of the identity, this construction will not
+work unless we can ``push'' the perturbation on $A_2$ to one on $A_3$. That is,
+given a relation $c : A \rel A'$ and a syntactic perturbation $m_A \in M_A$, we
+need to be able to turn it into a syntactic perturbation $m_{A'} \in M_{A'}$
+such that the resulting semantic perturbations obtained by applying the
+respective homomorphisms $i_A$ and $i_{A'}$ form a square. Dually, we will need
+to ``pull'' a perturbation from right to left. This notion is made formal by the
+following definition:
+%
+\begin{definition}
+    Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects,
+    and let $c : A \rel A'$ be a predomain relation. A
+    \textbf{push-pull structure} for $c$ (with respect to $M_A$ and
+    $M_{A'}$) consists of monoid homomorphisms $\push : M_A \to M_A'$
+    and $\pull : M_A' \to M_A$ such that for all $m_A \in M_A$ there
+    is a square $i_A(m_A) \ltsq{c}{c} i_{A'}(\push\, m_A)$ and for all
+    $m_{A'} \in M_{A'}$ there is a square $i_A(\pull\, m_{A'})
+    \ltsq{c}{c} i_{A'}(m_{A'})$. A push-pull structure for error
+    domain relations is defined similarly.
+\end{definition}
 
@@
 
-It remains to specify the actions of the functors on value and computation
-objects and relations.
 % Because we have augmented the definitions of object and relation for our model,
@@
 \max{this section needs prose}
+In this section, we discuss the actions of the functors on value and computation
+objects. Specifically, we define the actions of the functors on the syntactic
+perturbations and their interpretation as semantic perturbations.
+
 We first discuss the action of $\li$ on objects. Given a value object $(A, M_A,
@@
 
+The fact that value and computation relations compose is proved in the appendix
+(Definition \ref{def:value-computation-rel-comp}). The actions of the functors
+$\li$, $U$, $\times$, and $\arr$ are also defined in the appendix (Definition
+\ref{def:functorial-actions-on-relations}).
+
 \subsection{The Dynamic Type}
@@
 as the constructors except for $e_{\iarr}$ which includes a $\nxt$.
-
+%
 %
@@
 % to be the morphism $\nxt$ followed by $\text{fun}$.
+%
+The definition of the ordering $\le_D$ is provided in the appendix for completeness.
 
-The ordering $\le_D$ is given by:\max{this is kind of obvious IMO maybe it can go in the appendix}
+% given by:\max{this is kind of obvious IMO maybe it can go in the appendix}
 %
-\begin{align*}
-    \inat(n) \le \inat(n') 
-        &\iff n = n' \\
-    \itimes (d_1, d_2) \le \itimes (d_1', d_2')
-        &\iff d_1 \le_D d_2 \text{ and } d_1' \le_D d_2'\\
-    \iarr(\tilde{f}) \le \iarr(\tilde{f'}) 
-        &\iff \later_t(\tilde{f}_t \le \tilde{f'}_t),
-\end{align*}
+
 %
-and the bisimilarity relation is defined analogously.
 %
@@
 Now we define the perturbations for $D$. Recall that to each predomain $A$ we
-associate a monoid $M_A$ of perturbations and a homomorphism into the monoid of
-endomorphisms bisimilar to the identity, and likewise for error domains. It may
-seem as though we need to define the perturbations for $D$ by guarded recursion,
-but in fact we define them via least-fixpoint in the category of monoids:
+associate a monoid $M_A$ of syntactic perturbations and a homomorphism into the
+monoid of endomorphisms bisimilar to the identity, and likewise for error
+domains. It may seem as though we need to define the perturbations for $D$ by
+guarded recursion, but in fact we define them via least-fixpoint in the category
+of monoids:
 %
@@
 We now explain how to interpret these perturbations as endomorphisms.
-
-\eric{Check this}
 
@@
 One can verify that this defines a homomorphism from $M_D \to \{ f : D \to D : f
-\bisim \id \}$. We claim\max{why we claim? Just say it's true!} that the three relations $\inat$, $\itimes$, and
+\bisim \id \}$. Then it is straightforward to check that the three relations $\inat$, $\itimes$, and
 $\iarr$ satisfy the push-pull property. As an illustrative case, we establish
@@
 
-We next claim\max{claim} that the relations $\inat$, $\itimes$, and $\iarr$ are
-quasi-left-representable, and that their lifts are quasi-right-representable.
-Indeed, it is straightforward to see that they are quasi-left-representable
-where all of the perturbations are taken to be the identity elements of the
-respective monoids.
+It is easy to see that the relations $\inat$, $\itimes$, and $\iarr$
+are quasi-left-representable, where all perturbations are taken to be the
+identity elements of the respective monoids.
 %
@@
 Recall the defintion of the downcast for $\iarr$ given in Section
-\ref{sec:term-interpretation}.\max{why is there a paragraph break here?}
-
-We know that in order for the DnR and DnL squares for $\li \iarr$ to exist, we
-must insert a delay on the side opposite the downcast. In terms of syntactic
-perturbations, this means that we take $\dellp$ and $\delrp$ in the definition
-of quasi-right-representability to both be $\inl\, 1$, where $\inl$ is the
-injection into the coproduct of monoids. We recall that the definition of
-syntactic perturbations for $\li$ involves a coproduct with $\mathbb{N}$, and
-that the interpretation homomorphism maps the generator $1$ to the endomorphism
-of error domains $(\delta \circ \eta)^\dagger$.
+\ref{sec:term-interpretation}. We know that in order for the DnR and DnL squares
+for $\li \iarr$ to exist, we must insert a delay on the side opposite the
+downcast. In terms of syntactic perturbations, this means that we take $\dellp$
+and $\delrp$ in the definition of quasi-right-representability to both be
+$\inl\, 1$, where $\inl$ is the left injection into the coproduct of monoids. We
+recall that the definition of syntactic perturbations for $\li$ involves a
+coproduct with $\mathbb{N}$, and that the interpretation homomorphism maps the
+generator $1$ to the endomorphism of error domains $(\delta \circ
+\eta)^\dagger$.
 %
@@
 definition of the relation $\iarr$.
-
-It is also straightforward to establish the retraction property for each of
-these three relations. In the case of $\iarr$, we have that the property holds
-up to a delay.\max{retraction property? has this ever come up anywhere in the paper before? Should it?}
 
@@
 
-\eric{TODO: Does this belong here?}
+\max{Missing a subsection on how to interpret the syntax into the semantics, i.e., the actual graduality proof}
+\subsection{Interpreting the Relational Axioms}
 
-Recall the equivalence rules for type precision given in Section \ref{sec:GTLC},
-for example the rule that $(c_i \ra c_o)(c_i' \ra c_o')\equiv (c_ic_i' \ra
-c_oc_o')$ To validate this rule in the model, we construct two squares $\id
-\ltsqbisim{(c_i \ra c_o)(c_i' \ra c_o')}{(c_ic_i' \ra c_oc_o')} \id$ and $\id
+\eric{Check this}
+
+We now turn to the interpretation of the type precision and term precision
+axioms of the cast calculus (Section \ref{sec:GTLC}).
+
+The congruence rules for term precision hold in the model, as all of
+constructions are monotone and preserve weak bisimilarity.
+%
+The interpretation of the cast rules was discussed above.
+%
+
+It remains to verify the equivalence rules for type precision. For example,
+consider the rule $(c_i \ra c_o)(c_i' \ra c_o')\equiv (c_ic_i' \ra c_oc_o')$. To
+validate this rule in the model, we construct two squares $\id \ltsqbisim{(c_i
+\ra c_o)(c_i' \ra c_o')}{(c_ic_i' \ra c_oc_o')} \id$ and $\id
 \ltsqbisim{(c_ic_i' \ra c_oc_o')}{(c_i \ra c_o)(c_i' \ra c_o')} \id$. Since the
@@
   syntactic perturbations $\delta^l_1, \delta^l_2 \in M_A$ and $\delta^r_1,
-  \delta^r_2 \in M_{A'}$ such that there is a square $\delta^l_1 \ltsq{c}{c'}
-  \delta^r_1$ and a square $\delta^l_2 \ltsq{c'}{c} \delta^r_2$.
+  \delta^r_2 \in M_{A'}$ such that there is a square $i_A(\delta^l_1)
+  \ltsq{c}{c'} i_{A'}(\delta^r_1)$ and a square $i_A(\delta^l_2) \ltsq{c'}{c}
+  i_{A'}(\delta^r_2)$. Given computation objects $B$ and $B$, we make the
+  analogous definition for relations $d, d' : B \rel B'$ between the underlying
+  error domains.
 \end{definition}
-Given computation objects $B$ and $B$, we make the analogous definition for
-relations $d, d' : B \rel B'$ between the underlying error domains.
 
-Then we have the following result that shows that the functors
-$U,\li,\times,\to$ are \emph{quasi-functorial} on relations:
-
+Then we have the following quasi-order-equivalences:
+%
 \begin{lemma}
@@
 \end{itemize}
-
 \end{lemma}
-\begin{proof} See appendix \eric{TODO: make sure this is in the appendix}
+\begin{proof}
+  See appendix (Lemma \ref{lem:quasi-order-equiv-functors}).
 \end{proof}
@@
 
+\eric{Check this}
+
 In this section, we prove the adequacy of the interpretation of term precision.
+
+\begin{comment}
 First we establish some notation. Fix a morphism $f : 1 \to \li \mathbb{N} \cong
@@
 %
-
 We first note that in this ordering, the semantics of error is not equivalent to
@@
 \end{itemize}
-
 %
-Unfortunately\max{it's not unfortunate}, this is actually not provable! Roughly speaking, the issue is
-that this is a ``global''\max{what does ``global'' mean?} result, and it is not possible to prove such results
+However, this is actually not provable! Intuitively, the issue is
+that to prove this result we need to the above result is a ``global''\max{what does ``global'' mean?} result, and it is not possible to prove such results
 in the guarded setting\max{isn't everything we do in a ``guarded setting''?}. In particular, if we tried to prove the above result in
@@
 %
-Thus, to prove our adequacy result, we need to pass to leave the guarded
+Thus, to prove our adequacy result, we need to leave the guarded
 setting\max{``pass to leave the guarded setting''?}. We have seen in Section \ref{sec:big-step-term-semantics} that clock
 quantification makes this possible.
+\end{comment}
 
-Recall that the definitions of our model, e.g., the free error domain, the
-lock-step ordering, and weak bisimilarity, have all been parameterized by a
-clock $k$. We saw in previously that there is an isomorphism between the
+Recall the big-step term semantics $- \Downarrow -$ defined in Section
+\ref{sec:big-step-term-semantics}. The result we seek to show is the following:
+% If $f \bisim^{gl} f' \ltls^{gl} g' \bisim^{gl} g$, then
+
+\begin{theorem}[adequacy]
+  If $\cdot \vdash M \ltdyn N : \nat$, then:
+  \begin{itemize}
+    \item If $M \Downarrow n$ then $N \Downarrow n$.
+    \item If $N \Downarrow \mho$ then $M \Downarrow \mho$.
+    \item If $N \Downarrow n$ then $M \Downarrow n$ or $M \Downarrow \mho$.
+  \end{itemize}
+\end{theorem}
+
+We have already established graduality above, which states that syntactic term
+precision implies that the denoted terms are related in the semantic
+interpretation of term precision.
+%
+% Recall as discussed in that section that the definitions of our model are all
+% parameterized by a clock $k$. 
+%
+It therefore remains to show that the denotation of term precision in our model
+$f \ltsqbisim{}{} g$, where $f, g : \li \mathbb{N}$, implies a relationship
+between the partial elements $\mathbb{N} + {\mho}$ corresponding to $f$ and $g$. Recall
+that to define the big-step semantics we used the equivalence of the
 globalization of the free error domain and Capretta's coinductive delay monad:
-$\li^{gl} \mathbb{N} \cong \delay(\mathbb{N} + {\mho})$.
+$\li^{gl} \mathbb{N} \cong \delay(\mathbb{N} + {\mho})$. We now extend this
+correspondence to a relationship between a global version of our denotation of
+term precision and the corresponding relation on the delay monad.
 
-Now consider the predomain $\mathbb{N}$ with ordering and bisimilarity both the
-equality relation. We can define a ``global''\max{stop using scare quotes around ``global''} version of the lock-step error
-ordering and the weak bisimilarity relation on elements of the $\li^{gl}
-\mathbb{N}$; the former is defined by
+% We saw in Section \ref{sec:big-step-term-semantics} that there is an isomorphism
+% between the globalization of the free error domain and Capretta's coinductive
+% delay monad: $\li^{gl} \mathbb{N} \cong \delay(\mathbb{N} + {\mho})$. We now
+% seek to extend this to a relatinoship between a global version of our semantic
+% notion of term precision and the corresponding relation for the delay monad.
+
+To that end, we define a version of our denotation of term precision
+that relates terms of type $\li^{gl} \mathbb{N}$. More concretely, consider the
+predomain $\mathbb{N}$ with ordering and bisimilarity both the equality
+relation. We define a global version of the lock-step error ordering and the
+weak bisimilarity relation on elements of $\li^{gl} \mathbb{N}$; the former is
+defined by
 %
-\( x \ltls^{gl}_X y := \forall k. x[k] \ltls y[k], \)
+\( x \ltls^{gl} y := \forall k. x[k] \ltls y[k], \)
 %
@@
 %
-\( x \bisim^{gl}_X y := \forall k. x[k] \bisim y[k]. \)
+\( x \bisim^{gl} y := \forall k. x[k] \bisim y[k]. \)
 %
-On the other hand, we can define coinductively a ``lock-step error ordering''
-relation on $\delay(\mathbb{N} + {\mho})$:\max{do these intermediate details matter for the main body of the paper?}
+
+Correspondingly, we define a version of the lock-step error ordering and weak
+bisimilarity relation on $\delay(\mathbb{N} + {\mho})$ 
+(see Section \ref{sec:relations-on-delay-monad} in the appendix for the definitions).
+% \max{do these intermediate details matter for the main body of the paper?}
 %
-\begin{mathpar}
-  \inferrule*[]
-  { }
-  {\tnow (\inr\, 1) \ledelay d}
+%
+By adapting Theorem 4.3 of \cite{kristensen-mogelberg-vezzosi2022} to the
+setting of inductively-defined relations, we can show that both the global
+lock-step error ordering and the global weak bisimilarity admit coinductive
+definitions. In particular, modulo the above isomorphism between $\li^{gl} X$
+and $\delay(\mathbb{N} + {\mho})$, the global version of the lock-step error
+ordering is equivalent to the lock-step error ordering on $\delay(\mathbb{N} +
+{\mho})$, and the global version of weak bisimilarity is equivalent to weak
+bisimilarity on $\delay(\mathbb{N} + {\mho})$.
+%
+This implies that the global denotation of the term precision semantics for
+$\li^{gl} \mathbb{N}$ agrees with the corresponding relation $\ltsqbisim{}{}^\text{Delay}$ for $\delay(\mathbb{N} + {\mho})$, i.e.,
+the closure of its lock-step ordering under weak bisimilarity on both sides.
+%
+Then to obtain the desired result, it suffices to show being related in
+$\ltsqbisim{}{}^\text{Delay}$ implies that the corresponding partial elements of
+$\mathbb{N} + {\mho}$ are related in the expected manner. That is, if $d
+\ltsqbisim{}{}^\text{Delay} d'$, then $p(d) \ltdyn^{Partial} p(d')$ where $p$ is
+a partial function from delays into $\mathbb{N} + {\mho}$ defined by existential
+quantification over the number of steps $i$ in which $d$ terminates, and the
+relation $\ltdyn^\text{Partial}$ is defined such that $\mho$ is the least
+element, $\bot \ltdyn^\text{Partial} \bot$, and $n \ltdyn^\text{Partial} n$.
+This implication follows from the definition of $\ltsqbisim{}{}^\text{Delay}$.
 
-  \inferrule*[]
-  {x_1 \le_X x_2}
-  {\tnow (\inl\, x_1) \ledelay \tnow (\inl\, x_2)}
+% Then our result follows by proving the corresponding result for
+% $\delay(\mathbb{N} + {\mho})$ which in turn follows from the definitions of the
+% relations.
 
-  \inferrule*[]
-  {d_1 \ledelay d_2}
-  {\tlater\, d_1 \ledelay \tlater\, d_2}
-\end{mathpar}
-%
-And we similarly define by coinduction a ``weak bisimilarity'' relation on
-$\delay(\mathbb{N} + {\mho})$. This uses a relation $d \Da x_?$ between
-$\delay(\mathbb{N} + {\mho})$ and $\mathbb{N} + {\mho}$ that is defined as $d
-\Da n_? := \Sigma_{i \in \mathbb{N}} d = \tlater^i(\tnow\, n_?)$. Then weak
-bisimilarity for the delay monad is defined coinductively by the rules
-%
-\begin{mathpar}
-  \inferrule*[]
-  {n_? \bisim_{\mathbb{N} + {\mho}} m_?}
-  {\tnow\, n_? \bisimdelay \tnow\, m_? }
-
-  \inferrule*[leftskip=1.5em]
-  {d_1 \Da n_? \and n_? \bisim_{\mathbb{N} + {\mho}} m_?}
-  {\tlater\, d_1 \bisimdelay \tnow\, m_? }
-
-  \inferrule*[leftskip=1.5em]
-  {d_2 \Da m_? \and n_? \bisim_{\mathbb{N} + {\mho}} m_?}
-  {\tnow\, n_? \bisimdelay \tlater\, d_2}
-
-  \inferrule*[leftskip=1.5em]
-  {d_1 \bisimdelay d_2}
-  {\tlater\, d_1 \bisimdelay \tlater\, d_2 }
-  %
-  % \inferrule*[]
-  % {d_1 \Da x_? \and d_2 \Da y_? \and x_? \bisim_{X + 1} y_?}
-  % {d_1 \bisimdelay d_2}
-  %
-  % \inferrule*[]
-  % {d_1 \bisimdelay d_2}
-  % {\tlater d_1 \bisimdelay \tlater d_2 }
-\end{mathpar}
-%
-Note the similarity of these definitions to the corresponding guarded
-definitions. By adapting Theorem 4.3 of \cite{kristensen-mogelberg-vezzosi2022}
-to the setting of inductively-defined relations, we can show that both the
-global lock-step error ordering and the global weak bisimilarity admit
-coinductive definitions. In particular, modulo the above isomorphism between
-$\li^{gl} X$ and $\delay(\mathbb{N} + {\mho})$, the global version of the
-lock-step error ordering is equivalent to the lock-step error ordering on
-$\delay(\mathbb{N} + {\mho})$, and likewise, the global version of the weak
-bisimilarity relation is equivalent to the weak bisimilarity relation on
-$\delay(\mathbb{N} + {\mho})$.
-
-This implies that the global version of the term precision semantics for
-$\li^{gl} \mathbb{N}$ agrees with the corresponding notion for
-$\delay(\mathbb{N} + {\mho})$. Then adequacy follows by proving the
-corresponding result for $\delay(\mathbb{N} + {\mho})$ which in turn follows
-from the definitions of the relations.
-\max{this doesn't actually relate to our notion of big step semantics}
+% \max{this doesn't actually relate to our notion of big step semantics}
 

### 2024-07-12 paper-new/towards-relational-model.tex
change to infix notation for relations

@@
 Given two poset relations $c : A_1 \rel A_2$ and $c' : A_2 \rel A_3$, we can
-form their \emph{composition} in the usual manner, i.e., $(x, y) \in c$ if and
-only if there is an element $z : A_2$ with $(x, z) \in c$ and $(z, y) \in c'$.
-We write the composition of relations as $c \comp c'$.
+form their \emph{composition} in the usual manner, i.e., $x \binrel{c \comp c'} y$ if and
+only if there is an element $z : A_2$ with $x \binrel{c} z$ and $z \binrel{c'} y$.
+%We write the composition of relations as $c \comp c'$.
 
@@
 %
-To capture this preservation of relations property, we use the notion
-of a \emph{square}, which specifies a relation between two monotone
-functions. The definition of square is as follows. Let $A_i, A_o,
-A_i', A_o'$ be partially-ordered sets. Let $c_i : A_i \rel A_i'$ and
-$c_o : A_o \rel A_o'$ be poset relations, and let $f : A_i \to A_o$
-and $g : A_i' \to A_o'$ be monotone functions. We say that $f
-\ltsq{c_i}{c_o} g$ if for all $x : A_i$ and $y : A_i'$ with $(x, y)
-\in c_i$, we have $(f(x), g(y)) \in c_o$. We call this a \emph{square}
-because we visualize this situation as follows:
+To capture this preservation of relations property, we use the notion of a
+\emph{square}, which specifies a relation between two monotone functions. The
+definition of square is as follows. Let $A_i, A_o, A_i', A_o'$ be
+partially-ordered sets. Let $c_i : A_i \rel A_i'$ and $c_o : A_o \rel A_o'$ be
+poset relations, and let $f : A_i \to A_o$ and $g : A_i' \to A_o'$ be monotone
+functions. We say that $f \ltsq{c_i}{c_o} g$ if for all $x : A_i$ and $y : A_i'$
+with $x \binrel{c_i} y$, we have $f(x) \binrel{c_o} g(y)$. We call this a
+\emph{square} because we visualize this situation as follows:
 %
@@
 of functions, relations and squares forms a locally thin \emph{double
-category} as used in previous work \cite{newlicata}.
+category} as used in previous work \cite{new-licata18}.
 

### 2024-07-12 paper-new/concrete-relational-model.tex
tweaks to notation; remove some comments

@@
 \begin{itemize}
-    \item A partial order $\le_A$.
+    \item A partial order $\ltdyn_A$.
     \item A reflexive, symmetric ``bisimilarity'' relation $\bisim_A$.
-\end{itemize}\max{shouldn't this be $\ltdyn$?}
+\end{itemize} %\max{shouldn't this be $\ltdyn$?}
 \end{definition}
@@
 $A$.
+%
+When we refer to the order on a predomain $A$ in isolation we will denote it
+$r(A)$, and to express that two elements are related in $r(A)$ we will
+write $x \ltdyn_A y$.
 
@@
 is $\later |A|$ and the relation is defined in the obvious way, i.e., $\tilde{x}
-\le_{\later A} \tilde{x'}$ iff $\later_t(\tilde{x}_t \le_A \tilde{x'}_t)$.
+\ltdyn_{\later A} \tilde{x'}$ iff $\later_t(\tilde{x}_t \ltdyn_A \tilde{x'}_t)$.
 Likewise for bisimilarity.
@@
 Let $A$ and $A'$ be predomains. A \textbf{morphism of predomains} $f : A \to A'$
-is a function between the underlying sets such that for all $x, x'$, if $x \le_A
-x'$, then $f(x) \le f(x')$, and if $x \bisim_A x'$, then $f(x) \bisim_{A'}
+is a function between the underlying sets such that for all $x, x'$, if $x \ltdyn_A
+x'$, then $f(x) \ltdyn_{A'} f(x')$, and if $x \bisim_A x'$, then $f(x) \bisim_{A'}
 f(x')$.
@@
 \begin{itemize}
-    \item A distinguished ``error'' element $\mho_B \in B$ such that $\mho \le_B x$ for all $x$
+    \item A distinguished ``error'' element $\mho_B \in B$ such that $\mho \ltdyn_B x$ for all $x$
     \item A morphism of predomains $\theta_B \colon \later B \to B$
@@
   \begin{enumerate}
-     \item (Respects error): For all $y \in B'$, we have $\mho_B \binrel{d} y$.\max{can we please use infix notation consistently for relations?}
+     \item (Respects error): For all $y \in B'$, we have $\mho_B \binrel{d} y$. %\max{can we please use infix notation consistently for relations?}
      \item (Preserves $\theta$): For all $\tilde{x}$ in $\laterhs B$ and
@@
 and $d'$ on $B_2$ and $B_3$ inductively to be the least relation containing $d$
-and $d'$ that is downward closed under $\le_{B_1}$, upward-closed under
-$\le_{B_3}$, respects error, and preserves $\theta$. See Definition
+and $d'$ that is downward closed under $\ltdyn_{B_1}$, upward-closed under
+$\ltdyn_{B_3}$, respects error, and preserves $\theta$. See Definition
 \ref{def:free-comp-ed-rel} in the appendix for the precise definition.
@@
 \in (c \arr d)$ if and only if for all $x : A$ and $x' : A'$ with $(x , x') \in
-c$ we have $(f(x), g(x')) \in Ud$. The fact that this relation respects error\max{preserves error?}
+c$ we have $(f(x), g(x')) \in Ud$. The fact that this relation respects error %\max{preserves error?}
 and preserves $\theta$ follows from the corresponding properties for $d$.
@@
 
-The monadic bind operation $-^\dagger$ takes a predomain morphism $f : A \to UB$
-to an error domain morphism $\li A \multimap B$; with this we define the action of $\li$ on
-morphisms: Given $f : A \to A'$ we define $\li f = (\eta \circ f)^\dagger : \li A \to \li A'$.\max{what is the point of this sentence?}
+% The monadic bind operation $-^\dagger$ takes a predomain morphism $f : A \to UB$
+% to an error domain morphism $\li A \multimap B$; with this we define the action of $\li$ on
+% morphisms: Given $f : A \to A'$ we define $\li f = (\eta \circ f)^\dagger : \li A \to \li A'$.
+%\max{what is the point of this sentence?}
 
@@
 are perturbations $\delle_c \in M_A$ and $\delre_c \in M_{A'}$ such that the
-following two squares exist: (1) $\upl$: $e \ltsq{c}{\le_{A'}} i_{A'}(\delre_c)$, and
-(2) $\upr$: $i_A(\delle_c) \ltsq{\le_{A}}{c} e$.
+following two squares exist: (1) $\upl$: $e \ltsq{c}{\ltdyn_{A'}} i_{A'}(\delre_c)$, and
+(2) $\upr$: $i_A(\delle_c) \ltsq{\ltdyn_{A}}{c} e$.
 \end{definition}
@@
 the following two squares exist: 
-(1) $\dnl$: $p \ltsq{\le_{B'}}{d} i_{B'}(\delrp_d)$ and
-(2) $\dnr$: $i_B(\dellp_d) \ltsq{d}{\le_{B}} p$.
+(1) $\dnl$: $p \ltsq{\ltdyn_{B'}}{d} i_{B'}(\delrp_d)$ and
+(2) $\dnr$: $i_B(\dellp_d) \ltsq{d}{\ltdyn_{B}} p$.
 Quasi-right-representability for predomain relations is defined similarly.
@@
 \subsubsection{Constructions involving Perturbations}\label{sec:perturbation-constructions}
-\max{this section needs prose}
+% \max{this section needs prose}
+
+\eric{May be able to move some of this to the appendix}
 In this section, we discuss the actions of the functors on value and computation
@@
 %
-The definition of the ordering $\le_D$ is provided in the appendix for completeness.
+The definition of the ordering $\ltdyn_D$ is provided in the appendix for completeness.
 
@@
 We define three relations involving $D$ as follows. We define $\inat :
-\mathbb{N} \rel D$ by $(n, d) \in \inat$ iff $e_\nat(n) \le_D d$. We similarly
+\mathbb{N} \rel D$ by $(n, d) \in \inat$ iff $e_\nat(n) \ltdyn_D d$. We similarly
 define $\itimes : D \times D \rel D$ by $((d_1, d_2), d) \in \itimes$ iff
-$e_\times(d_1, d_2) \le_D d$, and we define $\text{inj}_\to : U(D \arr \li D)
-\rel D$ by $(f, d) \in \iarr$ iff $e_\to(f) \le_D d$.
+$e_\times(d_1, d_2) \ltdyn_D d$, and we define $\text{inj}_\to : U(D \arr \li D)
+\rel D$ by $(f, d) \in \iarr$ iff $e_\to(f) \ltdyn_D d$.
 
@@
 %
-% https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMSwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMCwzLCJlX2MiLDIseyJjdXJ2ZSI6MX1dLFsxLDQsIlxcZGVscmVfYyIsMl0sWzIsNSwiXFxwdXNoX3tjJ30oXFxkZWxyZV9jKSIsMix7ImN1cnZlIjoxfV0sWzAsMSwiYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEsMiwiYyciLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFszLDQsIlxcbGVfe0FfMn0iLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs0LDUsImMnIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMiw1LCJcXGlkIiwwLHsiY3VydmUiOi0xfV0sWzAsMywiZV9jIiwwLHsiY3VydmUiOi0xfV0sWzgsMTMsIlxcYmlzaW0iLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNiwxNCwiXFxiaXNpbSIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
+% https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMSwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMCwzLCJlX2MiLDIseyJjdXJ2ZSI6MX1dLFsxLDQsIlxcZGVscmVfYyIsMl0sWzIsNSwiXFxwdXNoX3tjJ30oXFxkZWxyZV9jKSIsMix7ImN1cnZlIjoxfV0sWzAsMSwiYyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEsMiwiYyciLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFszLDQsInIoQV8yKSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzQsNSwiYyciLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsyLDUsIlxcaWQiLDAseyJjdXJ2ZSI6LTF9XSxbMCwzLCJlX2MiLDAseyJjdXJ2ZSI6LTF9XSxbOCwxMywiXFxiaXNpbSIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs2LDE0LCJcXGJpc2ltIiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
 \[\begin{tikzcd}[ampersand replacement=\&,column sep=large]
@@
 	\arrow[""{name=3, anchor=center, inner sep=0}, "\id", curve={height=-6pt}, from=1-3, to=2-3]
-	\arrow["{\le_{A_2}}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+	\arrow["{r(A_2)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
 	\arrow["{c'}"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
@@
 
-\max{Missing a subsection on how to interpret the syntax into the semantics, i.e., the actual graduality proof}
+% \max{Missing a subsection on how to interpret the syntax into the semantics, i.e., the actual graduality proof}
 \subsection{Interpreting the Relational Axioms}

### 2024-07-12 paper-new/discussion.tex
discuss what has not been formalized in Agda

@@
 relations $\inat$, $\itimes$, and $\iarr$, and proved that they are
-quasi-representable.\max{need to say what's not formalized}
+quasi-representable.
+%
+We have not yet formalized the constructions involving value and computation
+relations: showing that these relations compose, and defining the actions of the
+functors $\li$, $U$, $\times$, and $\arr$ on the relations. We have also not yet
+formalized the syntax-to-semantics translation.
+
+% \max{need to say what's not formalized}
 
@@
 clock quantification as well as axioms asserting the \emph{clock-irrelevance} of
-booleans and natural numbers, since as of this writing these axioms are not
-built-in to Guarded Cubical Agda. The mechanization of adequacy also required us
-to formalize some essential lemmas involving clocks and clock-irrelevance; we
-are considering later refactoring these as part of a ``standard library'' for
-Guarded Cubical Agda.\max{say these axioms are taken from prior work and cite that}
+booleans and natural numbers since as of this writing these axioms are not
+built-in to Guarded Cubical Agda. These axioms are discussed in prior work on
+guarded type theory \cite{atkey-mcbride2013, kristensen-mogelberg-vezzosi2022}.
+The mechanization of adequacy also required us to formalize some essential
+lemmas involving clocks and clock-irrelevance; we are considering later
+refactoring these as part of a ``standard library'' for Guarded Cubical
+Agda.
+% \max{say these axioms are taken from prior work and cite that}
 

### 2024-07-12 paper-new/intro.tex
tweaks to intro and syntax sections

@@
 In this paper, we develop an adequate denotational semantics that satisfies
-graduality and soundness of the equational theory of cast calculi using the
-techniques of SGDT.
+graduality and soundness of the equational theory of cast calculi using
+synthetic guarded domain theory.
 
@@
 \item First, we give a simple concrete term semantics where we show
-  how to model the dynamic type as a solution to a guarded domain equation
+  how to model the dynamic type as a solution to a guarded domain equation.
 \item Next, we identify where prior work on classical domain theoretic
@@
   which allow us to recover enough extensional reasoning to model the
-  graduality property compositionally
+  graduality property compositionally.
 \item We combine this insight together with an abstract categorical
@@
 \begin{enumerate}
-\item In Section\ref{sec:GTLC} we fix our input language, a fairly
+\item In Section \ref{sec:GTLC} we fix our input language, a fairly
   typical gradually typed cast calculus.
-\item In Section\ref{sec:concrete-term-model} we develop a first
+\item In Section \ref{sec:concrete-term-model} we develop a first
   denotational semantics in synthetic guarded domain theory that
@@
   modeling effects using call-by-push-value.
-\item In Section\ref{sec:towards-relational-model} we show where the
+\item In Section \ref{sec:towards-relational-model} we show where the
   New-Licata classical domain theoretic approach fails to adapt
@@
   proving graduality in an intensional model.
-\item In Section\ref{sec:concrete-relational-model} we describe the resulting
+\item In Section \ref{sec:concrete-relational-model} we describe the resulting
   model in detail. We then prove that the resulting denotational model is
   \emph{adequate} for the graduality property: a closed term precision $M \ltdyn
-  N : \nat$ has the expected semantics, that $M$ errors or $M$ and $N$ have the
+  N : \nat$ has the expected semantics, i.e., that $M$ errors or $M$ and $N$ have the
   same extensional behavior.
-\item In Section\ref{sec:discussion} we discuss prior work on proving
+\item In Section \ref{sec:discussion} we discuss prior work on proving
   graduality, our partial mechanization of these results and discuss

### 2024-07-12 paper-new/appendix.tex
some discussion

@@
 
-As mentioned in Section~\ref{sec:syntax}, the cast calculus we employ uses a splitting of casts into downcasts and upcasts, whereas gradual surface languages are built out of a single notion of cast
+As mentioned in Section~\ref{sec:syntax}, the cast calculus we
+separates gradual type casts into downcasts and upcasts, whereas many
+gradual calculi are built out of a single notion of cast
 \[ \inferrule{\Gamma \vdash M : A}{\Gamma \vdash (M :: A') : A' }\]
@@
   \end{mathpar}
+
+  This has the universal property that $\phi \ltsq{d \relcomp d'}{d''}
+  \phi'$ if and only if for any $b \binrel{d} b'$ and $b' \binrel{d'}
+  b''$ that $\phi(b) \binrel{d''} \phi'(b'')$.
 \end{definition}

### 2024-07-12 paper-new/discussion.tex
some discussion

@@
 call-by-name to call-by-value as well by giving an axiomatic theory of
-type and term precision in call-by-push-value. This left implicit any
-connection to a ``double call-by-push-value'', which we make explicit
-in Section~\ref{sec:cbpv}.
+type and term precision in call-by-push-value.
 
-With this notion of abstract categorical model in hand, denotational
-semantics is then the work of constructing concrete models that
-exhibit the categorical construction. New and Licata
-\cite{new-licata18} present such a model using categories of
-$\omega$-CPOs, and this model was extended by Lennon-Bertrand,
-Maillard, Tabareau and Tanter to prove graduality of a gradual
-dependently typed calculus $\textrm{CastCIC}^{\mathcal G}$. This
-domain-theoretic approach meets our criteria of being a semantic
-framework for proving graduality, but suffers from the limitations of
-classical domain theory: the inability to model viciously
-self-referential structures such as higher-order extensible state and
-similar features such as runtime-extensible dynamic types. Since these
-features are quite common in dynamically typed languages, we seek a
-new denotational framework that can model these type system features.
+%% With this notion of abstract categorical model in hand, denotational
+%% semantics is then the work of constructing concrete models that
+%% exhibit the categorical construction. New and Licata
+%% \cite{new-licata18} present such a model using categories of
+%% $\omega$-CPOs, and this model was extended by Lennon-Bertrand,
+%% Maillard, Tabareau and Tanter to prove graduality of a gradual
+%% dependently typed calculus $\textrm{CastCIC}^{\mathcal G}$. This
+%% domain-theoretic approach meets our criteria of being a semantic
+%% framework for proving graduality, but suffers from the limitations of
+%% classical domain theory: the inability to model viciously
+%% self-referential structures such as higher-order extensible state and
+%% similar features such as runtime-extensible dynamic types. Since these
+%% features are quite common in dynamically typed languages, we seek a
+%% new denotational framework that can model these type system features.
 
@@
 and proving many of the same tedious lemmas without reusable
-mathematical abstractions.
+mathematical abstractions. This is addressed somewhat by Siek and Chen
+\cite{siek-chen2021}, who give a proof in Agda of graduality for an
+operational semantics who use a \emph{guarded logic of propositions}
+shallowly embedded in Agda. The guarded logic simplifies the treatment
+of step-indexed logical relations, but the approach is still
+fundamentally operational, and so the main lemmas of the work are
+still tied to the particular operational syntactic calculus being
+modeled. A further advantage of the denotational approach is that it
+easily validates equational reasoning, not just graduality, and it is
+completely independent of any particular syntax of gradual typing.
 
-\subsection{Related Work}
 % Discuss Joey Eremondi's thesis on gradual dependent types
@@
 % Discuss Jeremy Siek's work on graduality in Agda
-Siek and Chen \cite{siek-chen2021} give a proof in Agda of graduality
-for an operational semantics. While they do not use the Guarded
-Cubical extension, they do use a guarded logic of propositions
-shallowly embedded in Agda. Our denotational approach requires full
-guarded type theory not just guarded logic. An advantage of the
-denotational approach is that it easily validates equational
-reasoning, not just graduality, and it is completely independent of
-any particular syntax of gradual typing.
 
@@
 
-\subsection{Comparison to Explicit Step-Indexing}
+\subsection{Comparison to Embedding-projection pairs}
 
-\max{shallow}
+A line of work by New, Licata and Ahmed proves graduality by
+interpreting type precision $c : A \ltdyn A'$ as an
+\emph{embedding-projection pair}, that is a pure function $e : A \to
+A'$ and a possibly erroring projection $p : \li A' \multimap \li A$
+such that $p \circ e = \id$ and $e \circ p \ltdyn \id$
+\cite{newahmed,newlicata,newlicataahmed}. At first glance, our model
+of precision looks somewhat different: it is a \emph{relation}
+$\sem{c} : A \relto A'$ with a quasi-representability structure that
+gives us something like $e$ and $p$ in an embedding-projection
+pair. The relationship between these models is that \emph{if} the
+relation were \emph{truly} representable rather than only
+quasi-representable we would have that $e$ and $p$ form a \emph{Galois
+connection} $e \circ p \ltdyn \id$ and $\id \ltdyn p \circ
+e$. However, we have dropped the stronger property of retraction from
+our analysis in this work. With true representability, the relation
+$c$ is \emph{uniquely determined} by the embedding $e$, but since we
+only have quasi-representability, we need to keep the relation around
+explicitly. So the extra complexity of managing explicit relations is
+a cost of the intensional reasoning that guarded type theory
+introduces.
+
+\subsection{Comparison to Logical Relations Models}
 Working internally to guarded type theory reduces the overhead of

### 2024-07-12 paper-new/discussion.tex
fix an error

@@
 of precision looks somewhat different: it is a \emph{relation}
-$\sem{c} : A \relto A'$ with a quasi-representability structure that
+$\sem{c} : A \rel A'$ with a quasi-representability structure that
 gives us something like $e$ and $p$ in an embedding-projection

### 2024-07-12 paper-new/intro.tex
revisions to intro

@@
 One of the fundamental theorems for gradually typed languages is
-\emph{graduality}, also known as the \emph{dynamic gradual guarantee}
-\emph{dynamic gradual guarantee}, originally defined by Siek,
-Vitousek, Cimini, and Boyland \cite{siek_et_al:LIPIcs:2015:5031,
-  new-ahmed2018}.
+\emph{graduality}, also known as the \emph{dynamic gradual guarantee},
+originally defined by Siek, Vitousek, Cimini, and Boyland
+\cite{siek_et_al:LIPIcs:2015:5031, new-ahmed2018}.
 %
-Informally, graduality says that going from a dynamic to static style should only allow for the introduction of static or dynamic type errors, and not otherwise change the behavior of the program.
+Informally, graduality says that going from a dynamic to static style should
+only allow for the introduction of static or dynamic type errors, and not
+otherwise change the behavior of the program.
 %
@@
 A well-behaved semantics should then satisfy several properties. First, it
-should be adequate: natural number constants should step to
+should be \emph{adequate}: natural number constants should step to
 themselves. Second it should validate type based reasoning. To
@@
 New and Licata and is based on a classical Scott-style \emph{domain
-theory} \cite{newlicata2019}. The fundamental idea is to equip
+theory} \cite{new-licata18}. The fundamental idea is to equip
 $\omega$-CPOs with an additional ``error ordering'' $\ltdyn$ which
@@
 expressive enough to be extended to model dependently typed gradual
-typing \cite{dependentgradualtyping}.
+typing \cite{10.1145/3495528}.
 %
@@
 objects in the \emph{topos of trees}, i.e., presheaves on the poset of
-natural numbers\cite{topos-of-trees}. We think of such a
+natural numbers\cite{birkedal-mogelberg-schwinghammer-stovring2011}. We think of such a
 $\mathbb{N}$-indexed set as modeling a kind of ``sequence of
@@
 solution. This has the added benefit that providing a denotational
-semantics of gradually typed terms is as simple as an semantics of an
+semantics of gradually typed terms is as simple as a semantics of an
 effectful language using a monad on the category of sets. Not only
@@
 %
-\max{TODO: more here, specifically this is where we should talk about adequacy I think}
+% \max{TODO: more here, specifically this is where we should talk about adequacy I think}
 
-In this paper, we develop an adequate denotational semantics that satisfies
-graduality and soundness of the equational theory of cast calculi using
-synthetic guarded domain theory.
+Because the solutions we obtain working in guarded domain theory are constructed
+as a sequence of increasingly better approximations, when we want to establish a
+property that holds in the limit, we must reason about all finite
+approximations. In the setting of the topos of trees, this manifests as the need
+to consider \emph{global elements} of the presheaves, i.e., a family of elements
+$x_i \in X_i$ compatible with the restriction maps. For example, consider the
+set of programs that may take a computational step or return unit. In the topos
+of trees, we can model this set as the indexed family of sets $X_n =
+\{0,1,\dots,n-1, \bot\}$ for $n \ge 1$, where $i$ denotes a program that steps
+$i$ times and then returns, and $\bot$ denotes a program that fails to terminate
+in $n-1$ steps or fewer. For any fixed $n$, the set $X_n$ fails to distinguish a
+program that takes $n$ steps and then terminates from a program that never
+terminates. On the other hand, if we instead consider the denotation of a
+program to be a global element of the presheaf $\{X_n\}$ -- a family of
+elements $x_i \in X_i$ for all $i \in \mathbb{N}$ -- then the denotation of the
+diverging program will be distinct from that of a program that terminates,
+regardless of the number of steps it takes.
+
+In the setting of synthetic guarded domain theory, this same idea applies. To
+construct a global solution we must use an additional construct, e.g., the
+$\square$ modality \cite{10.1007/978-3-319-08918-8_8}, whose semantics in the
+topos of trees model is to compute the global elements of a presheaf. A related
+approach involves objects known as \emph{clocks}, whereby
+\emph{clock-quantification} \cite{atkey-mcbride2013} provides a means to obtain
+a global solution.
+
+This has important ramifications for defining an adequate denotational semantics
+in guarded domain theory, as we seek to do in this paper: we do not want our
+denotational semantics to conflate a diverging program with a program that fails
+to terminate in sufficiently few steps. Thus, establishing adequacy of a
+semantics in guarded domain theory will require a means of globalizing,
+a point we will return to in Section \ref{sec:big-step-term-semantics}.
+
+% For example, when establishing the adequacy of a semantics defined using
+% guarded type theory, we must reason about all finite approximations.
+
+% In this paper, we develop an adequate denotational semantics that satisfies
+% graduality and soundness of the equational theory of cast calculi using
+% synthetic guarded domain theory.
 
@@
 equality and satisfies a graduality theorem. A great deal of the work
-has further been verified in guarded cubical Agda \cite{veltri-vezzosi2020}, 
+has further been verified in Guarded Cubical Agda \cite{veltri-vezzosi2020}, 
 demonstrating that the semantics is readily mechanizable.

### 2024-07-12 paper-new/concrete-term-model.tex
revisions to the relational motivation section
revise sections on relational model
revise the background on guarded type theory in the term model section

@@
-\section{A Concrete Term Model}\label{sec:concrete-term-model}
+\section{A Denotational Semantics for Types and Terms}\label{sec:concrete-term-model}
 
@@
 
-\max{TODO: discuss this wording with Eric. Are we working in guarded
-  type theory, SGDT or Ticked Cubical Type Theory in the informal work
-  in the paper?}
+% \max{TODO: discuss this wording with Eric. Are we working in guarded
+%   type theory, SGDT or Ticked Cubical Type Theory in the informal work
+%   in the paper?}
 We begin with a brief overview of the background theory that we use in this work.
-
-synthetic guarded domain theory, which we will
-use to construct the model. \emph{Synthetic guarded domain theory}, or SGDT for
+\emph{Synthetic guarded domain theory}, or SGDT for
 short, is an axiomatic framework in which we can reason about non-well-founded
@@
 
+The constructions we carry out in this paper will take place in a guarded type
+theory with ticks.
 \subsection{A Call-by-push-value Model}
@@
 that is a refinement of Moggi's monadic semantics
-\cite{moggi-monads}. While Moggi works with a (strong) monad $T$ on a
+\cite{MOGGI199155}. While Moggi works with a (strong) monad $T$ on a
 category $\mathcal V$ of pure morphisms, call-by-push-value works
@@
 pairs of a pure morphism and a linear morphism, as we explain below
-\cite{newlicataahmed}.
+\cite{new-licata-ahmed2019}.
 
@@
   A homomorphism of error domains $\phi : B \multimap B'$ is a
-  function $\phi : UB \multimap UB'$ that preserves these in that
+  function $\phi : UB \to UB'$ that preserves these in that
   $\phi(\mho_B) = \mho_{B'}$ and $\phi(\theta_{B}(\tilde{x})) =
@@
 
-The free error domain is essential to modeling cbv computation: a term
+The free error domain is essential to modeling CBV computation: a term
 $\Gamma \vdash M : A$ in GTLC will be modeled as a function $M :
@@
 %
-\[ D \cong \Nat + (D \times D) + U(D \to \li D). \]
+\[ D \cong \Nat + (D \times D) + U(D \to \li D), \]
 %
@@
 this equation. However, in guarded domain theory we instead replace this domain equation with a guarded domain equation.
-process analogous to giving a normal set-theoretic denotational semantics.
 Consider the following similar-looking equation:
 %
-\[ D \cong \Nat + (D \times D)\, + \laterhs U(D \to \li D). \]
+\begin{equation}\label{eq:dyn}
+D \cong \Nat + (D \times D)\, + \laterhs U(D \to \li D).
+\end{equation}
 %
@@
 Expanding out the guarded fixed point and least fixed point property
-gives us that $D$ satisfies our desired equation above. \max{give
-  that equation a number}.
+gives us that $D$ satisfies Equation $\ref{eq:dyn}$. 
+% \max{give that equation a number}.
 %
@@
 \timesk f_r)$. The full definitions of the Kleisli actions are
-included in the appendix.
+included in Appendix \ref{sec:kleisli-actions}.
 
-Lastly, we have the upcasts and downcasts for the injections, which
+Lastly, we have the upcasts and downcasts for the injections into the dynamic type, which
 are the core primitive casts. The upcasts are simply the injections
@@
 account for the fact that the functions are under a later in the
-dynamic type. Similarly, the downcasts are similar in that on values
+dynamic type. The downcasts are similar in that on values
 they pattern match on the input and return it if it is of the correct
@@
 
-\subsection{Extracting a well-behaved big-step semantics}
+\subsection{Extracting a well-behaved big-step semantics}\label{sec:big-step-term-semantics}
 
@@
 implies that the type $\li^{gl} \mathbb{N} := \forall k. \li^k \mathbb{N}$ is a
-final $F$-coalgebra.
+final $F$-coalgebra. In fact, $\li^{gl}$ is isomorphic to a more familiar coninductive type.
 %
@@
 \mathbb{N}$ defined by $\lambda M. \lambda k .\sem{M}$. Composing this with the
-above partial function yields the desired partial function $-\Downarrow : \{M
+above partial function yields the desired partial function $-\Downarrow \colon \{M
 \,|\, \cdot \vdash M : \nat \} \rightharpoonup \mathbb{N} + {\mho}$. It is

### 2024-07-12 paper-new/towards-relational-model.tex
small fixes in towards a relational model section

@@
 ordering can be defined as lifting poset structures on the
-subformulae, the key design choice is in the ordering on the free
+subformulae; the key design choice is in the ordering on the free
 error domain $\li A$.
@@
 $\mathbb{N}$ is the flat poset whose ordering is equality, then the
-bigstep semantics we derive for $l, l'$ do in fact satisfy the
+big-step semantics we derive for $l, l'$ does in fact satisfy the
 intended graduality relation.

### 2024-07-12 paper-new/appendix.tex
small changes in appendix

@@
 
-\section{Kleisli Actions}
+\section{Kleisli Actions}\label{sec:kleisli-actions}
 
@@
 
-Recall the notion of quasi-order-equivalence as defined in Definition
+Recall the notion of quasi-equivalence of relations as defined in Definition
 \ref{def:quasi-order-equivalent}. The following lemma will be useful in showing
-that two relations are quasi-order-equivalent.
+that two relations are quasi-equivalent.
 
@@
 
-The next two lemmas concern quasi-order-equivalence and the functors $\li$ and
+The next two lemmas concern quasi-equivalence and the functors $\li$ and
 $U$.
@@
 
-Lastly, we establish the quasi-order-equivalence for the functors.
+Lastly, we prove quasi-equivalence of relations involving composition and the
+functors $U$, $\li$, $\to$ and $\times$.
 
@@
 
-Correspondingly, we define a notion of lock-step error ordering and weak
-bisimilarity relation for the coinductive delay monad $\delay(\mathbb{N} + {\mho})$:
+We define a notion of lock-step error ordering and weak bisimilarity relation
+for the coinductive delay monad $\delay(\mathbb{N} + {\mho})$:
 

### 2024-07-12 paper-new/discussion.tex
revisions to discussion section

@@
 
-\eric{I moved this from the intro. We may need to revise it.}
+% \eric{I moved this from the intro. We may need to revise it.}
 
@@
 
-Current approaches to constructing languages that satisfy the
-graduality property include the methods of Abstracting Gradual Typing
-\cite{garcia-clark-tanter2016} and the formal tools of the Gradualizer
-\cite{cimini-siek2016}.  These allow the language developer to start
-with a statically typed language and derive a gradually typed language
-that satisfies the graduality. The main downside to these
-approaches lies in their inflexibility: since the process in entirely
-mechanical, the language designer must adhere to the predefined
-framework.  Many gradually typed languages do not fit into either
-framework, e.g., Typed Racket \cite{tobin-hochstadt06,
-  tobin-hochstadt08} and the semantics produced is not always the
-desired one.
+The methods of Abstracting Gradual Typing \cite{garcia-clark-tanter2016} and the
+formal tools of the Gradualizer \cite{cimini-siek2016} provide a way to derive
+from a statically-typed language a gradually-typed language languages that
+satisfies graduality by construction. The main downside to these approaches lies
+in their inflexibility: since the process in entirely mechanical, the language
+designer must adhere to the predefined framework.  Many gradually typed
+languages do not fit into either framework, e.g., Typed Racket
+\cite{tobin-hochstadt06, tobin-hochstadt08} and the semantics produced is not
+always the desired one.
 %
-Furthermore, while these frameworks do prove graduality of the
-resulting languages, they do not show the correctness of the
-equational theory, which is equally important to sound gradual typing.
+Furthermore, while these frameworks do prove graduality of the resulting
+languages, they do not show the correctness of the equational theory, which is
+equally important to sound gradual typing.
 
@@
 
-\subsection{Synthetic Ordering}
-\max{cut this subsection if we need space}
-A key to managing the complexity of our concrete construction is in
-using a \emph{synthetic} approach to step-indexing rather than working
-analytically with presheaves. This has helped immensely in our ongoing
-mechanization in cubical Agda as it sidesteps the need to formalize
-these constructions internally. 
-%
-However, there are other aspects of the model, the bisimilarity and
-the monotonicity, which are treated analytically and are similarly
-tedious.
-%
-It may be possible to utilize further synthetic techniques to reduce
-this burden as well, and have all type intrinsically carry a notion of
-bisimilarity and ordering relation, and all constructions to
-automatically preserve them.
-%
-A synthetic approach to ordering is common in (non-guarded) synthetic
-domain theory and has also been used for synthetic reasoning for cost
-models \cite{fiore_1997,GrodinNSH24}.
+% \subsection{Synthetic Ordering}
+% \max{cut this subsection if we need space}
+% A key to managing the complexity of our concrete construction is in
+% using a \emph{synthetic} approach to step-indexing rather than working
+% analytically with presheaves. This has helped immensely in our ongoing
+% mechanization in cubical Agda as it sidesteps the need to formalize
+% these constructions internally. 
+% %
+% However, there are other aspects of the model, the bisimilarity and
+% the monotonicity, which are treated analytically and are similarly
+% tedious.
+% %
+% It may be possible to utilize further synthetic techniques to reduce
+% this burden as well, and have all type intrinsically carry a notion of
+% bisimilarity and ordering relation, and all constructions to
+% automatically preserve them.
+% %
+% A synthetic approach to ordering is common in (non-guarded) synthetic
+% domain theory and has also been used for synthetic reasoning for cost
+% models \cite{fiore_1997,GrodinNSH24}.
 
@@
 
-\max{Expand on this, e.g. what some of the challenges might be and what would be reusable}
-In the future, we plan to apply our approach to give a denotational
-semantics for languages that feature higher-order state or
-runtime-extensible dynamic typing
-\cite{DBLP:journals/corr/abs-2210-02169} as well as richer type
-disciplines such as gradual dependent types and effect systems.
+% \max{Expand on this, e.g. what some of the challenges might be and what would be reusable}
+% In the future, we plan to apply our approach to give a denotational
+% semantics for languages that feature higher-order state or
+% runtime-extensible dynamic typing
+% \cite{DBLP:journals/corr/abs-2210-02169} as well as richer type
+% disciplines such as gradual dependent types and effect systems.
+
+% \max{Expand on this, e.g. what some of the challenges might be and what would be reusable}
+In the future, we plan to apply our approach to give a denotational semantics
+for languages that feature higher-order state or runtime-extensible dynamic
+typing \cite{DBLP:journals/corr/abs-2210-02169} as well as richer type
+disciplines such as gradual dependent types and effect systems. This will
+involve adapting prior operational models based on step-indexing (e.g.,
+\cite{new-giovannini-licata-2022}) to the denotational setting and using guarded
+type theory to obtain solutions to guarded domain equations as we did in this
+work for the denotation of the dynamic type. The generality of our techniques
+should allow for many of the constructions to be reused across different
+languages. For instance, the free error domain construction can be easily
+extended with additional cases to model effects besides error and stepping.
+Additionally, we aim to complete our Agda formalization and evolve it into a
+reusable framework for mechanized denotational semantics of gradually-typed
+languages.
 

### 2024-07-12 paper-new/concrete-relational-model.tex
concrete relational model section re-org

@@
 Next we describe in detail our compositional model of GTLC that
-satisfies graduality. To manage the complexity, we carry out the
-construction in two phases, beginning with a relational model
-consisting of partially-ordered sets with a notion of bisimilarity and
-then adding requirements for a structured set of syntactic
-perturbations on each type as well as the notion of
-quasi-representability for relations that models the cast rules. We
-then discuss the construction of the dynamic type and the refined
-denotation of term precision that relates terms that can be
-synchronized to be in lock-step. We conclude by demonstrating the
-adequacy of our model with respect to the graduality property.
+satisfies graduality. The definitions proceed in two phases: in the
+first phase we define our objects, functions, relations and squares to
+model the error ordering and bisimilarity, and then we add the extra
+structure of perturbations and quasi-representability to our objects
+and relations to ensure that we can model the cast rules.
+%
+Once these definitions are complete we show how to interpret the
+types, type precision, terms and term precision compositionally in our
+model, and finally we conclude by extracting a big step semantics from
+this model that satisfies the graduality theorem.
 
@@
 
-In the previous section, we introduced the notion of weak bisimilarity as a
-relation on the guarded lift monad. Recall that weak bisimilarity on $\li A$ is
-parameterized by a relation on $A$. To define a compositional model for all
-types, we equip the posets and error domains with a reflexive, symmetric
-bisimilarity relation. This leads to our first new definition:
-
+As discussed in the previous section our model of types must now come
+with both an error ordering and a bisimilarity relation.
 \begin{definition}
@@
     \item A reflexive, symmetric ``bisimilarity'' relation $\bisim_A$.
-\end{itemize} %\max{shouldn't this be $\ltdyn$?}
+\end{itemize}
+A \textbf{morphism of predomains} $f : A \to A'$ is a function between
+the underlying sets that is both \emph{monotone} (if
+$x \ltdyn_A x'$, then $f(x) \ltdyn_{A'} f(x')$), and
+\emph{extensional} (if $x \bisim_A x'$, then $f(x) \bisim_{A'} f(x')$).
 \end{definition}
@@
 %
-When we refer to the order on a predomain $A$ in isolation we will denote it
-$r(A)$, and to express that two elements are related in $r(A)$ we will
-write $x \ltdyn_A y$.
-
-Given a predomain $A$, we can form the predomain $\later A$. The underlying set
-is $\later |A|$ and the relation is defined in the obvious way, i.e., $\tilde{x}
-\ltdyn_{\later A} \tilde{x'}$ iff $\later_t(\tilde{x}_t \ltdyn_A \tilde{x'}_t)$.
-Likewise for bisimilarity.
+Given a predomain $A$, we can form the predomain $\later A$. The
+underlying set is $\later |A|$ and the ordering is given by $\tilde{x}
+\ltdyn_{\later A} \tilde{x'}$ iff $\later_t(\tilde{x}_t \ltdyn_A
+\tilde{x'}_t)$. Likewise for bisimilarity.
 %
 We also give a predomain structure to the natural numbers $\mathbb{N}$, where
-both the ordering and the bisimilarity relation are equality.
-%
-Morphisms of predomains are functions between the underlying sets that preserve
-the ordering and the bisimilarity relation. More formally:
-%
-\begin{definition}
-Let $A$ and $A'$ be predomains. A \textbf{morphism of predomains} $f : A \to A'$
-is a function between the underlying sets such that for all $x, x'$, if $x \ltdyn_A
-x'$, then $f(x) \ltdyn_{A'} f(x')$, and if $x \bisim_A x'$, then $f(x) \bisim_{A'}
-f(x')$.
-\end{definition}
+both the ordering and the bisimilarity relation are equality.\max{why natural numbers here? I don't think we need it. check and come back}
 
-We also enhance our error domains with ordering and bisimilarity
-relation, and we impose a condition relating the $\theta$ morphism to
-the bisimilarity relation. More concretely:
+We similarly enhance our error domains with ordering and bisimilarity
+relations, but we further need to impose conditions on the algebraic
+structure: the error element should be the least element (so that it
+models graduality), the think map should be monotone and extensional
+and the bisimilarity relation should relate $\delta$ to the identity
+(so that it models bisimilarity).
 %
@@
     \item For all $x : B$ we have $\theta_B(\nxt\, x) \bisim_B x$
-\end{itemize}\max{Don't we also need $\forall x. \mho \le x$?}
+\end{itemize}
+A \textbf{morphism of error domains} is a morphism of the underlying
+predomains that preserves the error element and the $\theta$ map.
 \end{definition}
-%
 For an error domain $B$, we define the predomain morphism $\delta_B := \theta_B
 \circ \nxt$.
-%
-A \textbf{morphism of error domains} is simply a morphism of the underlying
-predomains that preserves the error element and the $\theta$ map.
 
-%
-The definition of relations between predomains is the same as that between
-posets introduced in the previous section, as the bisimilarity relations of the
-predomains are not involved in the definition of monotone relation.
+A \emph{predomain relation} $c : A \rel A'$ is a monotone relation as
+defined in the previous section (no condition is required for
+bisimilarity). The ``reflexive'' predomain relation is written $r(A)$
+and is given by the ordering $\ltdyn_A$. The relational composition of
+two predomains is also monotone, and $r(A)$ is the identity for this
+composition.
 
-Next we define an analogous notion of \emph{relation} between error domains $B$
-and $B'$. An error domain relation $d : B \rel B'$ is a relation on the
-underlying posets that preserves error and respects $\theta$. More formally:
-%
+Error domain relations are given by predomain relations that are
+additionally a congruence with respect to the algebraic structures
+$\theta$ and $\mho$.
 \begin{definition}
-  Let $B$ and $B'$ be error domains. An \textbf{error domain relation} between
-  $B$ and $B'$ is a monotone relation $d$ on the underlying partially-ordered sets such that:
+  An \textbf{error domain relation} $d : B \rel B'$ is a monotone relation $d$ on the underlying predomains satisfying
   \begin{enumerate}
-     \item (Respects error): For all $y \in B'$, we have $\mho_B \binrel{d} y$. %\max{can we please use infix notation consistently for relations?}
-     \item (Preserves $\theta$): For all $\tilde{x}$ in $\laterhs B$ and
+     \item ($\mho$ congruence): $\mho_B \binrel{d} \mho_{B'}$.
+     \item ($\theta$ congruence): For all $\tilde{x}$ in $\laterhs B$ and
      $\tilde{y} \in \laterhs B'$, if $\laterhs_t (\tilde{x}_t \binrel{d} \tilde{y}_t)$, then 
@@
 \end{definition}
-%
-As with predomain relations, we can compose error domain relations. However, the
-composition of the underlying predomain relations is not an error domain
-relation.
-% Can explain this further in the appendix
-Instead, we define composition of error domain relations $d$ on $B_1$ and $B_2$
-and $d'$ on $B_2$ and $B_3$ inductively to be the least relation containing $d$
-and $d'$ that is downward closed under $\ltdyn_{B_1}$, upward-closed under
-$\ltdyn_{B_3}$, respects error, and preserves $\theta$. See Definition
-\ref{def:free-comp-ed-rel} in the appendix for the precise definition.
+Again we have a reflexive error domain relation $r(B)$ given by the
+ordering. The relational composition of two error domain relations in
+general is not an error domain relation ($\theta$ congruence in
+particular is not preserved), but we instead use an alternative ``free
+composition'' $d \odot d'$ that is inductively generated from the
+relational composition. Details of this construction are in the
+appendix (Definition \ref{def:free-comp-ed-rel}).
 
-The definitions of squares for predomains are also unchanged from the previous
-section, except that the morphisms on the left and right are now morphisms of
-predomains. We likewise define a notion of \textbf{error domain square}: Let
-$d_i : B_i \rel B_i'$ and $d_o : B_o \rel B_o'$ be error domain relations, and
-let $\phi : B_i \multimap B_o$ and $\phi' : B_i' \multimap B_o'$ be morphisms of
-error domains. We say that there is an error domain square $\phi \ltsq{d_i}{d_o}
-\phi'$ exactly when there is a predomain square involving the underlying
-predomain morphisms and relations.
+We also will use the notion of a square $f \ltsq{c_i}{c_o} g$ between
+predomain morphisms as well as squares $\phi \ltsq{d_i}{d_o} \psi$
+between error domain morphisms, which are defined similarly, with no
+extra interaction with the algebraic structure.
 
-\subsubsection{Functors}\label{sec:free-error-domain}
+In our denotational model of GTLC, we map every type $A$ to a
+predomain $\sem{A}$. These constructions are given by the same
+constructions as in Section~\ref{sec:tm-model}, but now enriched to
+act also on orderings and bisimilarity. For products this is the
+obvious lifting, and for functions we use the CBPV decomposition $U(A
+\to \li A')$ where the only one that is not clear is $\li A'$.
+\begin{definition}
+  The free error domain $\li A$ has as the underlying set the free
+  simple error domain with ordering given by the lock-step error
+  ordering and bisimilarity by weak bisimilarity both generated by the
+  ordering/bisimilarity on $A$. This is free in the same sense as the
+  free simple error domain that it is left adjoint to $U$.
+\end{definition}
 
-The functors $U$, $\li$, $\times$, and $\arr$ introduced in the previous section
-act on predomains and error domains and also have actions on their morphisms,
-relations, and squares. The functor $U$ simply takes the underlying predomain
-relation of an error domain relation. Given a relation $c_1 : A_1 \rel A_1'$ and
-$c_2 : A_2 \rel A_2'$, the relation $c_1 \times c_2 : (A_1 \times A_2) \rel
-(A_1' \times A_2')$ is defined by $((x, y), (x', y')) \in (c_1 \times c_2)$ if
-and only if $(x, x') \in c_1$ and $(y, y') \in c_2$. Given a relation $c : A
-\rel A'$ and an error domain relation $d : B \rel B'$, the relation $c \arr d :
-(A \arr B) \rel (A' \arr B')$ is the error domain relation defined by $(f, g)
-\in (c \arr d)$ if and only if for all $x : A$ and $x' : A'$ with $(x , x') \in
-c$ we have $(f(x), g(x')) \in Ud$. The fact that this relation respects error %\max{preserves error?}
-and preserves $\theta$ follows from the corresponding properties for $d$.
+%% \subsubsection{Functors}\label{sec:free-error-domain}
 
-% Free error domain
-The functor $\li$ takes a predomain $A$ to the free error domain on
-$A$. We first define the predomain $U\li A$. The underlying set of
-$U\li A$ is defined to be $\li |A|$ (the guarded lift monad applied to
-$|A|$). The ordering relation is the lock-step error ordering
-introduced in the previous section , and the bisimilarity relation is
-the weak bisimilarity relation on $\li |A|$ also defined in the
-previous section. With these relations, the constructors $\eta$ and
-$\theta$ of the guarded lift monad are in fact morphisms of
-predomains, i.e., they are monotone and preserve bisimilarity.
-%
-We observe that this predomain structure extends to an error domain structure by
-noting that the required error element is given by the constructor $\mho$ and
-the required $\theta$ map is the constructor $\theta$. Lastly, it can be shown
-that the delay morphism $\delta = \theta \circ \nxt : UB \to UB$ satisfies $x \bisim
-\delta\, x$ for all $x$ as is required by the definition of error domain.
+%% The functors $U$, $\li$, $\times$, and $\arr$ introduced in the previous section
+%% act on predomains and error domains and also have actions on their morphisms,
+%% relations, and squares. The functor $U$ simply takes the underlying predomain
+%% relation of an error domain relation. Given a relation $c_1 : A_1 \rel A_1'$ and
+%% $c_2 : A_2 \rel A_2'$, the relation $c_1 \times c_2 : (A_1 \times A_2) \rel
+%% (A_1' \times A_2')$ is defined by $((x, y), (x', y')) \in (c_1 \times c_2)$ if
+%% and only if $(x, x') \in c_1$ and $(y, y') \in c_2$. Given a relation $c : A
+%% \rel A'$ and an error domain relation $d : B \rel B'$, the relation $c \arr d :
+%% (A \arr B) \rel (A' \arr B')$ is the error domain relation defined by $(f, g)
+%% \in (c \arr d)$ if and only if for all $x : A$ and $x' : A'$ with $(x , x') \in
+%% c$ we have $(f(x), g(x')) \in Ud$. The fact that this relation respects error %\max{preserves error?}
+%% and preserves $\theta$ follows from the corresponding properties for $d$.
+
+%% % Free error domain
+%% The functor $\li$ takes a predomain $A$ to the free error domain on
+%% $A$. We first define the predomain $U\li A$. The underlying set of
+%% $U\li A$ is defined to be $\li |A|$ (the guarded lift monad applied to
+%% $|A|$). The ordering relation is the lock-step error ordering
+%% introduced in the previous section , and the bisimilarity relation is
+%% the weak bisimilarity relation on $\li |A|$ also defined in the
+%% previous section. With these relations, the constructors $\eta$ and
+%% $\theta$ of the guarded lift monad are in fact morphisms of
+%% predomains, i.e., they are monotone and preserve bisimilarity.
+%% %
+%% We observe that this predomain structure extends to an error domain structure by
+%% noting that the required error element is given by the constructor $\mho$ and
+%% the required $\theta$ map is the constructor $\theta$. Lastly, it can be shown
+%% that the delay morphism $\delta = \theta \circ \nxt : UB \to UB$ satisfies $x \bisim
+%% \delta\, x$ for all $x$ as is required by the definition of error domain.
 
@@
 
-Finally, we define the action of $\li$ on a predomain relation $c$ to be the
-heterogeneous lock-step error ordering between $U\li A$ and $U\li A'$ and note
-that it by definition satisfies the additional requirements of being an error
-domain relation between the error domains $\li A$ and $\li A'$.
+Additionally, all of these core constructions $\times, \to, U, \li$
+act on heterogeneous relations as well in an analogous way.
+
+The most interesting type to interpret is the dynamic
+type. The beauty of guarded domain theory is that once we have define
+$\laterhs$ for predomains, we can solve the same domain equation as in
+Section~\ref{sec:tm-model} changed only in that we interpret it as a
+domain equation on predomains rather than sets:
+%
+\[ D \cong \mathbb{N}\, + (D \times D)\, + \laterhs U(D \arr \li D) \]
+The complete definition of the ordering $\ltdyn_D$ is provided in the appendix
+for completeness.
+
+To interpret term precision, we define three relations involving $D$
+as follows. We define $\inat : \mathbb{N} \rel D$ by $(n, d) \in
+\inat$ iff $e_\nat(n) \ltdyn_D d$. We similarly define $\itimes : D
+\times D \rel D$ by $((d_1, d_2), d) \in \itimes$ iff $e_\times(d_1,
+d_2) \ltdyn_D d$, and we define $\text{inj}_\to : U(D \arr \li D) \rel
+D$ by $(f, d) \in \iarr$ iff $e_\to(f) \ltdyn_D d$.
 
@@
 
-To accomplish this, we equip every predomain $A$ with a monoid $M_A$ of
-\emph{syntactic perturbations}, as well as a means of \emph{interpreting} these
-syntactic perturbations as \emph{semantic} perturbations on $A$, i.e., as
-endomorphisms that are bisimilar to the identity.
-% Moreover, we want the
-% resulting endomorphisms to be extensionally equivalent to the identity function,
-% a condition we express by saying that they are weakly bisimilar to the identity
-% morphism on $A$. 
-More formally:
+To accomplish this, we equip the predomains $A$ in our semantics with
+a monoid $M_A$ of \emph{syntactic perturbations}, as well as a means
+of \emph{interpreting} these syntactic perturbations as
+\emph{semantic} perturbations on $A$, i.e., as endomorphisms that are
+bisimilar to the identity. We call such an object a \emph{value
+object} and this will serve as the final definition of a denotation of
+a type in our semantics. We define computation objects to similarly be
+error domains equipped with a notion of syntactic perturbation:
 \begin{definition}[value and computation objects + morphisms]
@@
 
+For ordinary representability, the relation is \emph{uniqueley
+determined} by any morphism that left- or right-represents it: if two
+relations are left-represented by the same morphism, then they are
+equal. But for quasi-representability, this is no longer the case due
+to the presence of perturbations. Instead we have the following weaker property:
+\begin{definition}[quasi-equivalence]\label{def:quasi-equivalent}
+  Let $A$ and $A'$ be value objects, and let $c, c' : A \rel A'$ be relations
+  between the underlying predomains. We say that $c$ and $c'$ are
+  \textbf{quasi-equivalent}, written $c \qordeq c'$, if there exist
+  syntactic perturbations $\delta^l_1, \delta^l_2 \in M_A$ and $\delta^r_1,
+  \delta^r_2 \in M_{A'}$ such that there is a square $i_A(\delta^l_1)
+  \ltsq{c}{c'} i_{A'}(\delta^r_1)$ and a square $i_A(\delta^l_2) \ltsq{c'}{c}
+  i_{A'}(\delta^r_2)$. Given computation objects $B$ and $B$, we make the
+  analogous definition for relations $d, d' : B \rel B'$ between the underlying
+  error domains.
+\end{definition}
+
 % \max{idea: move this after the quasi-representability part and say that this structure is what's needed to make them compose}
@@
 We can now give the final definition of value and computation relations for our
-model:
+model, which serve as the denotation of our type precision derivations:
 %
@@
 $p_c : \li A' \to \li A$.
-\end{definition}
 
-\begin{definition}[computation relations]
 A \textbf{computation relation} between computation objects $(B, M_B, i_B)$ and
@@
 
-The reason for requiring $\li c$ to be quasi-right-representable in the
-definition of value relations is so that we can define the action of the functor
-$\li$ on relations, taking a value relation to a computation relation. Likewise,
-we require $Ud$ to be quasi-left-representable in the definition of computation
-relations so that we can define the action of $U$ on relations.
+It is essential in the definition of a value relation that we have not
+just that $c$ is quasi-left-representable but also that $\li c$ is
+quasi-right-representable so that we can define the action of the
+functor $\li$ on relations, taking a value relation to a computation
+relation. Likewise, we require $Ud$ to be quasi-left-representable in
+the definition of computation relations so that we can define the
+action of $U$ on relations.
 
@@
 
-\subsubsection{Constructions involving Perturbations}\label{sec:perturbation-constructions}
+To model the term precision in a way that is oblivious to the stepping
+behavior of terms, we need to weaken the notion of square to allow for the
+morphisms on either side to be ``out of sync'' rather than in lock step.
+\begin{definition}
+  Let $c_i : A_i \rel A_i'$, $c_o : A_o \rel A_o'$ $f : A_i \to A_o$
+  and $g : A_i' \to A_o'$.  An extensional square $f
+  \ltsqbisim{c_i}{c_o} g$ holds if there \emph{exist} $f'
+  \bisimsq{A_i}{A_o} f$ and $g' \bisimsq{A_i'}{A_o'} g$ such that $f' \ltsq{c_i}{c_o} g'$.
+\end{definition}
+Here $f' \bisimsq{A_i}{A_o} f$ is the natural extension of
+bisimilarity to morphisms: given bisimilar inputs the functions have
+bisimilar outputs.
+
+\subsection{Constructing the denotational semantics}
+
+With these concepts now defined, we can give an overview of our
+denotational model:
+\begin{theorem}
+  \begin{itemize}
+  \item For every type $A$ we define a value type $\sem{A}$, and define $\sem{x_1:A_1,\ldots} = \sem{A_1}\times\cdots$
+  \item For every term $\Gamma \vdash M : A$, we define a value morphism $\sem{M} : \sem{\Gamma} \to U\li\sem{A}$.
+  \item If $M = M'$ are equal according to our equational theory, then $\sem{M} = \sem{M'}$
+  \item For every type precision derivation $c : A \ltdyn A'$ we
+    define a value relation $\sem{c} : \sem{A} \rel \sem{A'}$.
+  \item For every type precision equivalence $c \equiv c'$ we have a
+    quasi-order-equivalence of relations $\sem{c} \bisim \sem{c'}$.
+  \item For every term precision $\Delta \vdash M \ltdyn M' : c$ we
+    have an \emph{extensional} square
+    $\sem{M}\ltsqbisim{\sem{\Delta}}{U\li\sem{c}} \sem{M'}$
+  \end{itemize}
+\end{theorem}
+
+For the term semantics $\sem{M}$, nothing really differs from the term
+semantics in Section~\ref{sec:tm-semantics}. The only difference is
+that the definitions of $\sem{\upc c}$ and $\sem{\dnc c}$ can be
+defined as the morphisms that left- and right-quasi-represent $\sem
+c$, but upon inspection this definition agrees with the one given
+before. Similarly, the validity of the equational theory follows
+easily: $\beta\eta$ holds in our model. We next elaborate on each of
+the remaining components.
+
+\subsubsection{Interpreting types}\label{sec:perturbation-constructions}
 % \max{this section needs prose}
 
-\eric{May be able to move some of this to the appendix}
-In this section, we discuss the actions of the functors on value and computation
-objects. Specifically, we define the actions of the functors on the syntactic
-perturbations and their interpretation as semantic perturbations.
+\eric{May be able to move some of this to the appendix} For our value
+type semantics, we already have an interpretation as predomains, we
+describe how we enhance each of the constructions we use
+$U,\li,\to,\times,D$ with a notion of syntactic perturbation.
 
-We first discuss the action of $\li$ on objects. Given a value object $(A, M_A,
-i_A)$ we define the syntactic perturbations $M_{\li A}$ to be $\mathbb{N} \oplus
-M_A$, where $\oplus$ denotes the free product of monoids (the coproduct in the
-category of monoids). The intuition behind needing to include $\mathbb{N}$ comes
-from the example of the downcast for $\iarr$. Recall that we adjusted the
-squares corresponding to the DnL and DnR rules by adding $(\delta \circ
-\eta)^\dagger$ on the side opposite the downcast. Thus, we need a syntactic
-perturbation in $M_{\li A}$ that will be interpreted as $(\delta \circ
-\eta)^\dagger$. We accomplish this by taking the coproduct with $\mathbb{N}$.
+First consider the free error domain $\li-$. Given a value object
+$(A, M_A, i_A)$ we define the syntactic perturbations $M_{\li A}$ to
+be $\mathbb{N} \oplus M_A$, where $\oplus$ denotes the free product of
+monoids (the coproduct in the category of monoids). The intuition
+behind needing to include $\mathbb{N}$ comes from the example of the
+downcast for $\iarr$. Recall that we adjusted the squares
+corresponding to the DnL and DnR rules by adding $(\delta \circ
+\eta)^\dagger$ on the side opposite the downcast. Thus, we need a
+syntactic perturbation in $M_{\li A}$ that will be interpreted as
+$(\delta \circ \eta)^\dagger$. We accomplish this by taking the
+coproduct with $\mathbb{N}$.
 %
@@
 
-\subsubsection{Constructions involving Value and Computation Relations}
-
-\eric{What, if anything, should go here? Should we discuss composition of
-value/computation relations? Actions of the functors on relations?}
-
-The fact that value and computation relations compose is proved in the appendix
-(Definition \ref{def:value-computation-rel-comp}). The actions of the functors
-$\li$, $U$, $\times$, and $\arr$ are also defined in the appendix (Definition
-\ref{def:functorial-actions-on-relations}).
-
-\subsection{The Dynamic Type}
-We now describe the denotation of the dynamic type. Analogously to how we
-defined the dynamic type in Section \ref{sec:concrete-term-model}, the predomain
-representing the dynamic type is defined using a combination of least fixpoint
-and guarded recursion as the solution to the equation of \emph{predomains}
-%
-\[ D \cong \mathbb{N}\, + (D \times D)\, + \laterhs U(D \arr \li D). \]
-%
-As before, we name the constructors $\inat$, $\itimes$, and $\iarr$
-respectively, and the upcasts $e_{\nat}$, $e_\times$, and $e_\to$ are the same
-as the constructors except for $e_{\iarr}$ which includes a $\nxt$.
-%
-%
-% We define $e_\mathbb{N} : \mathbb{N} \to D$ to be the $\text{nat}$ constructor,
-% $e_\times : D \times D \to D$ to be $\text{times}$, and $e_\to : U(D \arr F D)$
-% to be the morphism $\nxt$ followed by $\text{fun}$.
-%
-The definition of the ordering $\ltdyn_D$ is provided in the appendix for completeness.
-
-% given by:\max{this is kind of obvious IMO maybe it can go in the appendix}
-%
-
-%
-%
-We define three relations involving $D$ as follows. We define $\inat :
-\mathbb{N} \rel D$ by $(n, d) \in \inat$ iff $e_\nat(n) \ltdyn_D d$. We similarly
-define $\itimes : D \times D \rel D$ by $((d_1, d_2), d) \in \itimes$ iff
-$e_\times(d_1, d_2) \ltdyn_D d$, and we define $\text{inj}_\to : U(D \arr \li D)
-\rel D$ by $(f, d) \in \iarr$ iff $e_\to(f) \ltdyn_D d$.
-
+Finally, the most interesting type is of course the dynamic type. 
 Now we define the perturbations for $D$. Recall that to each predomain $A$ we
@@
 use the interpretation for perturbations on $U(D \to \li D)$.
+One can verify that this defines a homomorphism from $M_D \to \{ f : D \to D : f
+\bisim \id \}$.
 
@@
 
-One can verify that this defines a homomorphism from $M_D \to \{ f : D \to D : f
-\bisim \id \}$. Then it is straightforward to check that the three relations $\inat$, $\itimes$, and
-$\iarr$ satisfy the push-pull property. As an illustrative case, we establish
-the push-pull property for the relation $\iarr$. We define $\pull_{\iarr} : M_D
-\to M_{U(D \arr \li D)}$ by giving a homomorphism from $M_{D \times D} \to
-M_{U(D \arr \li D)}$ and from $M_{U(D \to \li D)} \to M_{U(D \arr \li D)}$. The
-former is the trivial homomorphism sending everything to the identity element,
-while the latter is the identity homomorphism.
+\subsubsection{Interpreting type precision}
+
+Next we need to confirm that our interpretation of type precision
+derivations as predomain relations indeed extends to value relations,
+i.e., that the relations are quasi-representable and that they satisfy
+the push-pull property for perturbations. We include most of these technical
+verifications in the appendix (Definition
+\ref{def:value-computation-rel-comp},
+\ref{def:functorial-actions-on-relations}).
+
+We go into more detail on the three relations involving the dynamic
+type $\inat$, $\itimes$, and $\iarr$. As an illustrative case, we
+establish the push-pull property for the relation $\iarr$. We define
+$\pull_{\iarr} : M_D \to M_{U(D \arr \li D)}$ by giving a homomorphism
+from $M_{D \times D} \to M_{U(D \arr \li D)}$ and from $M_{U(D \to \li
+  D)} \to M_{U(D \arr \li D)}$. The former is the trivial homomorphism
+sending everything to the identity element, while the latter is the
+identity homomorphism.
 %
@@
 
-\subsection{Interpreting Term Precision: Extensional Squares}
+\subsubsection{Interpreting Type Precision Equivalence}
 
-As introduced in the previous section, to obtain a model of term precision that
-is oblivious to the stepping behavior of terms, we weaken our notion of square
-to allow for the morphisms on either side to be synchronized to be in lock-step.
-More formally, let $c_i : A_i \rel A_i'$ and $c_o : A_o \rel A_o'$.
+To model type precision equivalence we need to show the following quasi-equivalences:
 %
-% We define an extensional square between morphisms $f : A_i \to A_o$ and $g :
-% A_i' \to A_o'$ to consist of:
+\begin{lemma}
+The following hold:
+\begin{itemize}
+  \item $U(d \comp d') \qordeq U(d) \comp U(d')$
+  \item $\li(c \comp c') \qordeq \li (c) \comp \li (c')$
+  \item $(c \comp c') \to (d \comp d') \qordeq (c \to d) \comp (c' \to d')$
+  \item $(c_1 \comp c_1') \times (c_2 \comp c_2') \qordeq (c_1 \times c_2) \comp (c_1'\times c_2')$
+\end{itemize}
+\end{lemma}
+\begin{proof}
+  Details in the appendix (Lemma \ref{lem:quasi-order-equiv-functors}).
+\end{proof}
+
+\subsubsection{Interpreting Term Precision: Extensional Squares}
+
+Finally, we establish the core result of the semantics: interpreting
+term precision as extensional squares. Fortunately the hard work is
+now complete and the term precision rules follow from our
+compositional constructions.
 %
-% \begin{itemize}
-%   \item A morphism $f' : A_i \to A_o$ with $f \bisim f'$.
-%   \item A morphism $g' : A_i' \to A_o'$ with $g \bisim g'$.
-%   \item A square $f' \ltsq{c_i}{c_o} g'$.
-% \end{itemize}
+The congruence rules of term precision are modeled easily: congruence
+squares hold both for bisimilarity as well as error ordering, and so
+by combination we also have these for extensional squares.
 %
-We say that $f \ltsqbisim{c^i}{c_o} g$ if there exist $f'$ and $g'$ such that
-%
-\[ f \bisim f' \ltsq{c_i}{c_o} g' \bisim g. \]
-%
-We call a square $f \ltsqbisim{c^i}{c_o} g$ an \emph{extensional square}, and we
-will call a square $f \ltsq{c^i}{c_o} g$ an \emph{intensional square}. If the
-relations $c_i$ and $c_o$ are obvious from the context, we will simply write $f
-\ltbisim g$. We make the analogous construction for the computation squares.
-%
-We now verify the existence of the squares for the cast rules. For UpL, the
+The type precision equivalence rule follows because $c \equiv c'$
+implies quasi-equivalence, and that quasi-equivalence implies that
+$\id \ltsqbisim{\sem{c}}{\sem{c'}} \id$.
+
+Most crucially, we verify the existence of the squares for the cast rules. For UpL, the
 relevant extensional square is obtained as follows:
@@
 is equal to $c'$ by the fact that $c'$ is downward closed. The squares for UpR,
-DnL, and DnR are obtained similarly.
+DnL, and DnR are obtained analogously.
 
@@
 % closure'' of the intensional error ordering.
-
-% \max{Missing a subsection on how to interpret the syntax into the semantics, i.e., the actual graduality proof}
-\subsection{Interpreting the Relational Axioms}
-
-\eric{Check this}
-
-We now turn to the interpretation of the type precision and term precision
-axioms of the cast calculus (Section \ref{sec:GTLC}).
-
-The congruence rules for term precision hold in the model, as all of
-constructions are monotone and preserve weak bisimilarity.
-%
-The interpretation of the cast rules was discussed above.
-%
-
-It remains to verify the equivalence rules for type precision. For example,
-consider the rule $(c_i \ra c_o)(c_i' \ra c_o')\equiv (c_ic_i' \ra c_oc_o')$. To
-validate this rule in the model, we construct two squares $\id \ltsqbisim{(c_i
-\ra c_o)(c_i' \ra c_o')}{(c_ic_i' \ra c_oc_o')} \id$ and $\id
-\ltsqbisim{(c_ic_i' \ra c_oc_o')}{(c_i \ra c_o)(c_i' \ra c_o')} \id$. Since the
-definition of extensional square allows us to specify intermediate morphisms
-that are synchronized in the intensional ordering, it will be sufficient to
-construct two intensional squares between the relations, with perturbations on
-both sides. This motivates the definition of \emph{quasi-order-equivalence}.
-
-\begin{definition}[quasi-order-equivalence]\label{def:quasi-order-equivalent}
-  Let $A$ and $A'$ be value objects, and let $c, c' : A \rel A'$ be relations
-  between the underlying predomains. We say that $c$ and $c'$ are
-  \textbf{quasi-order-equivalent}, written $c \qordeq c'$, if there exist
-  syntactic perturbations $\delta^l_1, \delta^l_2 \in M_A$ and $\delta^r_1,
-  \delta^r_2 \in M_{A'}$ such that there is a square $i_A(\delta^l_1)
-  \ltsq{c}{c'} i_{A'}(\delta^r_1)$ and a square $i_A(\delta^l_2) \ltsq{c'}{c}
-  i_{A'}(\delta^r_2)$. Given computation objects $B$ and $B$, we make the
-  analogous definition for relations $d, d' : B \rel B'$ between the underlying
-  error domains.
-\end{definition}
-
-Then we have the following quasi-order-equivalences:
-%
-\begin{lemma}
-The following hold:
-\begin{itemize}
-  \item $U(d \comp d') \qordeq U(d) \comp U(d')$
-  \item $\li(c \comp c') \qordeq \li (c) \comp \li (c')$
-  \item $(c \comp c') \to (d \comp d') \qordeq (c \to d) \comp (c' \to d')$
-  \item $(c_1 \comp c_1') \times (c_2 \comp c_2') \qordeq (c_1 \times c_2) \comp (c_1'\times c_2')$
-\end{itemize}
-\end{lemma}
-\begin{proof}
-  See appendix (Lemma \ref{lem:quasi-order-equiv-functors}).
-\end{proof}
 

### 2024-07-12 paper-new/concrete-relational-model.tex
tweak adequacy prose to flow better

@@
 
-\eric{Check this}
-
-In this section, we prove the adequacy of the interpretation of term precision.
-
+Now that we have established our denotational model, we can, as in
+Section~\ref{sec:big-step-term-semantics}, extract a big step semantics. Our final
+goal then is to prove that the extensional squares in our model imply
+the graduality property for this final big step semantics.
 \begin{comment}
@@
 \end{comment}
-
-Recall the big-step term semantics $- \Downarrow -$ defined in Section
-\ref{sec:big-step-term-semantics}. The result we seek to show is the following:
+The result we seek to show is the following:
 % If $f \bisim^{gl} f' \ltls^{gl} g' \bisim^{gl} g$, then
-
 \begin{theorem}[adequacy]
@@
 \end{theorem}
-
-We have already established graduality above, which states that syntactic term
-precision implies that the denoted terms are related in the semantic
-interpretation of term precision.
 %

### 2024-07-12 paper-new/concrete-term-model.tex
remove definition of the arrow functor from term model section
barely any space saving

@@
 There is a \emph{guarded fixpoint} operator
-%
-\[
-  \fix : \forall T, (\later T \to T) \to T.
-\]
-%
+\[ \fix : \forall T, (\later T \to T) \to T. \]
 That is, to construct a term of type $T$, it suffices to assume that we have
@@
 For the sake of brevity, we will also write tick application as $M_t$.
-
 The constructions we carry out in this paper will take place in a guarded type
@@
 
-\max{move this}
-The functor $\arr : \Set^{op} \times \errdom \to \errdom$ is defined on objects
-as follows. Given a set $A$ and error domain $B$, $A \arr B$ is the error domain
-whose underlying set is the set of functions $A \to UB$. The error element is
-the constant error function, i.e., $\lambda x. \mho_B$. The map $\theta :
-(\laterhs (A \to UB)) \to (A \to UB)$ is defined by 
-$\theta(\tilde{f}) = \lambda x . \theta_B (\tilde{f}_t)$.
-%
-Given a morphism $f : A_o \to A_i$ and $\phi : B_i \to B_o$, the morphism $f
-\arr \phi : (A_i \arr B_i) \to (A_o \arr B_o)$ is given by pre-composing by $f$
-and post-composing by $\phi$.
+% \max{move this}
+% The functor $\arr : \Set^{op} \times \errdom \to \errdom$ is defined on objects
+% as follows. Given a set $A$ and error domain $B$, $A \arr B$ is the error domain
+% whose underlying set is the set of functions $A \to UB$. The error element is
+% the constant error function, i.e., $\lambda x. \mho_B$. The map $\theta :
+% (\laterhs (A \to UB)) \to (A \to UB)$ is defined by 
+% $\theta(\tilde{f}) = \lambda x . \theta_B (\tilde{f}_t)$.
+% %
+% Given a morphism $f : A_o \to A_i$ and $\phi : B_i \to B_o$, the morphism $f
+% \arr \phi : (A_i \arr B_i) \to (A_o \arr B_o)$ is given by pre-composing by $f$
+% and post-composing by $\phi$.
 
@@
 %% \end{figure*}
-
 \begin{figure*}
-  \max{Can we squeeze this figure very tightly?}
-
-  % TODO check these
+  \begin{minipage}{0.3\textwidth}
+    \begin{footnotesize}
   \begin{align*}
+    % upcasts
+    \sem{\upc{r(A)}} &= \id_A \\
+    \sem{\upc{(c \comp c')}} &= \sem{\upc{c'}} \circ \sem{\upc{c}} \\
+    \sem{\upc{(c_i \ra c_o)}} &= (\sem{\dnc {c_i}} \tok \li\sem{A_o'})\circ (\sem{A_i} \tok U\li(\sem{\upc {c_o}}))\\
+    \sem{\upc{(c_1 \times c_2)}} &= \sem{\upc{c_1}} \times \sem{\upc{c_2}}\\
+    \sem{\upc{\inat}} &= \inat\\
+    \sem{\upc{\itimes}} &= \itimes\\
+    \sem{\upc{\iarr}} &= \iarr \circ \nxt\\
     % \sem{\zro}         &= \lambda \gamma . 0 \\
@@
     % \sem{\err_B}       &= \lambda \gamma . \mho \\
-    %
-    % upcasts
-    \sem{\upc{r(A)}} &= \id_A \\
-    \sem{\upc{(c \comp c')}} &= \sem{\upc{c'}} \circ \sem{\upc{c}} \\
-    \sem{\upc{(c_i \ra c_o)}} &= (\sem{\dnc {c_i}} \tok F\sem{A_o'})\circ (\sem{A_i} \tok UF(\sem{\upc {c_o}}))\\
-    \sem{\upc{(c_1 \times c_2)}} &= \sem{\upc{c_1}} \times \sem{\upc{c_2}}\\
-    \sem{\upc{\inat}} &= \inat\\
-    \sem{\upc{\itimes}} &= \itimes\\
-    \sem{\upc{\iarr}} &= \iarr \circ \nxt\\[2ex]
     %
@@
   \end{align*}
+    \end{footnotesize}
+  \end{minipage}
   \caption{Semantics of casts}
@@
 % \max{TODO: Eric needs to add a discussion of adequacy/clock stuff here}
-\eric{Check this}
+% \eric{Check this}
 

### 2024-07-12 paper-new/towards-relational-model.tex
remove comment
barely any space saving

@@
 which is the content of the final $4$ cases of the definition.
-
 \begin{figure*}
       \begin{align*}
-        &\mho \semltbad l \text{ iff } \top \\
+        \mho \semltbad l &\text{ iff } \top \\
         %
-        &\eta\, x \semltbad \eta\, y \text{ if } 
+        \eta\, x \semltbad \eta\, y &\text{ iff } 
             x \semlt y \\
         %		
-        &\theta\, \tilde{l} \semltbad \theta\, \tilde{l'} \text{ iff } 
+        \theta\, \tilde{l} \semltbad \theta\, \tilde{l'} &\text{ iff } 
             \later_t (\tilde{l}_t \semltbad \tilde{l'}_t) \\
         %	
-        &\theta\, \tilde{l} \semltbad \mho \text{ iff } \exists n. \theta\, \tilde{l} = \delta^n(\mho) \\
+        \theta\, \tilde{l} \semltbad \mho &\text{ iff } \exists n. \theta\, \tilde{l} = \delta^n(\mho) \\
         %	
-        &\theta\, \tilde{l} \semltbad \eta\, y \text{ iff } \exists n. \exists x \leq y.
+        \theta\, \tilde{l} \semltbad \eta\, y &\text{ iff } \exists n. \exists x \leq y.
             (\theta\, \tilde{l} = \delta^n(\eta\, x)) \\
         %	
-        &\eta\, x \semltbad \theta\, \tilde{l'} \text { iff }
+        \eta\, x \semltbad \theta\, \tilde{l'} &\text { iff }
             \exists n. \exists y \geq x. (\theta\, \tilde{l'} = \delta^n (\eta\, y))
     \end{align*}
-    \caption{The step-insensitive error ordering.}
+    \caption{Step-insensitive error ordering}
     \label{fig:step-insensitive-error-ordering}
@@
 \begin{proof}
-  By L\"ob induction, in the appendix.\max{Eric make sure this proof is in the appendix}
+  By L\"ob induction, in the appendix.
 \end{proof}
@@
 with regard to their intensional behavior, up to $l$ erroring. Figure
-\ref{fig:lock-step-error-ordering} gives the definition.
+\ref{fig:lock-step-error-ordering} gives the definition of this relation.
 
 \begin{figure}
+  \begin{minipage}{0.5\textwidth}
     \fbox{$l_1 \ltls l_2$}
@@
             \later_t (\tilde{l}_t \ltls \tilde{l'}_t)
+  \end{align*}\end{minipage}\begin{minipage}{0.5\textwidth}
+\fbox{$l_1 \bisim l_2$}
+    \begin{align*}
+        \mho \bisim \mho &\text{ iff } \top \\
+      %
+        \eta\, x \bisim \eta\, y &\text{ iff } x \bisim_A y \\
+      %		
+        \theta\, \tilde{x} \bisim \theta\, \tilde{y} &\text{ iff } \later_t (\tilde{x}_t \bisim \tilde{y}_t) \\
+      %	
+        \theta\, \tilde{x} \bisim \mho &\text{ iff } \exists n. \theta\, \tilde{x} = \delta^n(\mho)\\
+      %	
+        \theta\, \tilde{x} \bisim \eta\, y &\text{ iff } \exists n. \exists x \bisim_A y.
+          (\theta\, \tilde{x} = \delta^n(\eta\, x))\\
+      %
+        \mho \bisim \theta\, \tilde{y} &\text{ iff } \exists n. \theta\, \tilde{y} = \delta^n(\mho) \\
+      %	
+        \eta\, x \bisim \theta\, \tilde{y} &\text { iff } \exists n. \exists y \bisim_A x. (\theta\, \tilde{y} = \delta^n (\eta\, y))
       \end{align*}
-    \caption{The lock-step error ordering.}
+  \end{minipage}
+    \caption{lock-step error ordering and weak bisimilarity}
     \label{fig:lock-step-error-ordering}
@@
 Consider the representability condition corresponding to the DnL rule
-for $\iarr \colon \dyntodyn \ltdyn\, \dyn$.
-% https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXGxpIEQiXSxbMCwxLCJcXGxpIFUoRCBcXGFyciBcXGxpIEQpIl0sWzEsMCwiXFxsaSBEIl0sWzEsMSwiXFxsaSBEIl0sWzAsMiwiXFxsZXFfe1xcbGkgRH0iLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxLDMsIlxcbGlcXGlhcnIiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxcZG5je1xcaWFycn0iLDJdLFsyLDMsIlxcaWQiXV0=
-\[\begin{tikzcd}[ampersand replacement=\&]
-	{\li D} \& {\li D} \\
-	{\li U(D \arr \li D)} \& {\li D}
-	\arrow["{\leq_{\li D}}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-	\arrow["{\sem{\dnc{\iarr}}}"', from=1-1, to=2-1]
-	\arrow["\id", from=1-2, to=2-2]
-	\arrow["\li\sem{\iarr}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-\end{tikzcd}\]
+for $\iarr \colon \dyntodyn \ltdyn\, \dyn$. This is the square $\sem{\dnc{\iarr}} \ltsq{r({\li D})}{\li\sem{\iarr}} \id$.
+%% % https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXGxpIEQiXSxbMCwxLCJcXGxpIFUoRCBcXGFyciBcXGxpIEQpIl0sWzEsMCwiXFxsaSBEIl0sWzEsMSwiXFxsaSBEIl0sWzAsMiwiXFxsZXFfe1xcbGkgRH0iLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxLDMsIlxcbGlcXGlhcnIiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxcZG5je1xcaWFycn0iLDJdLFsyLDMsIlxcaWQiXV0=
+%% \[\begin{tikzcd}[ampersand replacement=\&]
+%% 	{\li D} \& {\li D} \\
+%% 	{\li U(D \arr \li D)} \& {\li D}
+%% 	\arrow["{\leq_{\li D}}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+%% 	\arrow["{\sem{\dnc{\iarr}}}"', from=1-1, to=2-1]
+%% 	\arrow["\id", from=1-2, to=2-2]
+%% 	\arrow["\li\sem{\iarr}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+%% \end{tikzcd}\]
 For this square to be valid with the lock-step ordering the left and
@@
 with a second relation on the free error domain $\li A$: \emph{weak
-bisimilarity}, defined in Figure~\ref{fig:weak-bisimilarity}, which is
+bisimilarity}, defined in Figure~\ref{fig:lock-step-error-ordering}, which is
 parameterized by a binary relation $\bisim_A$ on $A$
@@
 errors, we ensure the thinking side eventually errors.
-\begin{figure}
-    \fbox{$l_1 \bisim l_2$}
-    \begin{align*}
-        \mho \bisim \mho &\text{ iff } \top \\
-      %
-        \eta\, x \bisim \eta\, y &\text{ iff } x \bisim_A y \\
-      %		
-        \theta\, \tilde{x} \bisim \theta\, \tilde{y} &\text{ iff } \later_t (\tilde{x}_t \bisim \tilde{y}_t) \\
-      %	
-        \theta\, \tilde{x} \bisim \mho &\text{ iff } \exists n. \theta\, \tilde{x} = \delta^n(\mho)\\
-      %	
-        \theta\, \tilde{x} \bisim \eta\, y &\text{ iff } \exists n. \exists x \bisim_A y.
-          (\theta\, \tilde{x} = \delta^n(\eta\, x))\\
-      %
-        \mho \bisim \theta\, \tilde{y} &\text{ iff } \exists n. \theta\, \tilde{y} = \delta^n(\mho) \\
-      %	
-        \eta\, x \bisim \theta\, \tilde{y} &\text { iff } \exists n. \exists y \bisim_A x. (\theta\, \tilde{y} = \delta^n (\eta\, y))
-      \end{align*}
-    \caption{The weak bisimilarity relation on the free error domain}
-    \label{fig:weak-bisimilarity}
-\end{figure}
 

### 2024-07-12 paper-new/discussion.tex
revisions to discussion section

@@
 
-\subsection{Limitations of Prior Work}
+\subsection{Related Work}
 
-We give an overview of current approaches to proving graduality of
-languages and why they do not meet our criteria of a reusable semantic
-framework.
+We give an overview of current approaches to proving graduality and discuss
+their limitations where appropriate.
 
@@
 formal tools of the Gradualizer \cite{cimini-siek2016} provide a way to derive
-from a statically-typed language a gradually-typed language languages that
-satisfies graduality by construction. The main downside to these approaches lies
-in their inflexibility: since the process in entirely mechanical, the language
-designer must adhere to the predefined framework.  Many gradually typed
-languages do not fit into either framework, e.g., Typed Racket
-\cite{tobin-hochstadt06, tobin-hochstadt08} and the semantics produced is not
-always the desired one.
+from a statically-typed language a language that satisfies graduality by
+construction. The main downside to these approaches lies in their inflexibility:
+since the process in entirely mechanical, the language designer must adhere to
+the predefined framework.  Many gradually typed languages do not fit into either
+framework, e.g., Typed Racket \cite{tobin-hochstadt06, tobin-hochstadt08} and
+the semantics produced is not always the desired one.
 %
@@
 
-\subsubsection{Double Categorical Semantics}
+\subsubsection{Double-Categorical Semantics}
 
-New and Licata \cite{new-licata18} developed an axiomatic account of
-the graduality relation on a call-by-name cast calculus terms and
-showed that the graduality proof could be modeled using semantics in
-certain kinds of \emph{double categories}, categories internal to the
-category of categories. A double category extends a category with a
-second notion of morphism, often a notion of ``relation'' to be paired
-with the notion of functional morphism, as well as a notion of
-functional morphisms preserving relations. In gradual typing the
-notion of relation models type precision and the squares model the
-term precision relation. This approach was influenced by the semantics
-of parametricity using reflexive graph categories
-\cite{ma-reynolds,dunphythesis,reynoldsprogramme}: reflexive graph
-categories are essentially double categories without a notion of
-relational composition. In addition to capturing the notions of type
-and term precision, the double categorical approach allows for a
-\emph{universal property} for casts: upcasts are the \emph{universal}
-way to turn a relation arrow into a function in a forward direction
-and downcasts are the dual universal arrow.  Later, New, Licata and
-Ahmed \cite{new-licata-ahmed2019} extended this axiomatic treatment from
-call-by-name to call-by-value as well by giving an axiomatic theory of
-type and term precision in call-by-push-value.
+% New and Licata \cite{new-licata18} developed an axiomatic account of
+% the graduality relation on a call-by-name cast calculus terms and
+% showed that the graduality proof could be modeled using semantics in
+% certain kinds of \emph{double categories}, categories internal to the
+% category of categories. A double category extends a category with a
+% second notion of morphism, often a notion of ``relation'' to be paired
+% with the notion of functional morphism, as well as a notion of
+% functional morphisms preserving relations. In gradual typing the
+% notion of relation models type precision and the squares model the
+% term precision relation. This approach was influenced by the semantics
+% of parametricity using reflexive graph categories
+% \cite{ma-reynolds,dunphythesis,reynoldsprogramme}: reflexive graph
+% categories are essentially double categories without a notion of
+% relational composition. In addition to capturing the notions of type
+% and term precision, the double categorical approach allows for a
+% \emph{universal property} for casts: upcasts are the \emph{universal}
+% way to turn a relation arrow into a function in a forward direction
+% and downcasts are the dual universal arrow.
+
+New and Licata \cite{new-licata18} developed an axiomatic account of the
+graduality relation on a call-by-name cast calculus and showed that the
+graduality proof could be modeled using semantics in certain kinds of double
+categories.
+%
+Later, New, Licata and Ahmed \cite{new-licata-ahmed2019} extended this axiomatic
+treatment from call-by-name to call-by-value as well by giving an axiomatic
+theory of type and term precision in call-by-push-value. The double-categorical
+framework serves as a foundation on which we base our model construction in this
+paper. However, because of the lack of transitivity of the relation modeling
+term precision, our model is not a double category but instead a reflexive graph
+category, that is, a double category without a notion of relational composition.
 
@@
 
+\subsubsection{Embedding-Projection Pairs}
+
+A line of work by New, Licata and Ahmed proves graduality by
+interpreting type precision $c : A \ltdyn A'$ as an
+\emph{embedding-projection pair}, that is a pure function $e : A \to
+A'$ and a possibly erroring projection $p : \li A' \multimap \li A$
+such that $p \circ e = \id$ and $e \circ p \ltdyn \id$
+\cite{new-ahmed2018,new-licata18,new-licata-ahmed2019}. At first glance, our model
+of precision looks somewhat different: it is a \emph{relation}
+$\sem{c} : A \rel A'$ with a quasi-representability structure that
+gives us something like $e$ and $p$ in an embedding-projection
+pair. The relationship between these models is that \emph{if} the
+relation were \emph{truly} representable rather than only
+quasi-representable we would have that $e$ and $p$ form a \emph{Galois
+connection} $e \circ p \ltdyn \id$ and $\id \ltdyn p \circ
+e$. However, we have dropped the stronger property of retraction from
+our analysis in this work. With true representability, the relation
+$c$ is \emph{uniquely determined} by the embedding $e$, but since we
+only have quasi-representability, we need to keep the relation around
+explicitly. So the extra complexity of managing explicit relations is
+a cost of the intensional reasoning that guarded type theory
+introduces.
+
+\subsubsection{Step-Indexed Logical Relations Models}
 The standard alternative to domain theory that scales to essentially
@@
 and proving many of the same tedious lemmas without reusable
-mathematical abstractions. This is addressed somewhat by Siek and Chen
+mathematical abstractions.
+%
+% Additionally, the logical relations constructed to prove graduality in prior
+% work \cite{new-ahmed2018,new-licata-ahmed2019,new-giovannini-licata-2022} suffer
+% from technical complications of requiring two separate expression relations, one
+% that counts steps on the left and the other on the right, and there is no
+% analogue of this in our approach. However, using two expression relations allows
+% some but not all transitive reasoning of term precision to be recovered. In the
+% future we aim to explore if this approach is feasible in guarded semantics.
+%
+This is addressed somewhat by Siek and Chen
 \cite{siek-chen2021}, who give a proof in Agda of graduality for an
@@
 
+\subsubsection{Gradual Dependent Types}
 % Discuss Joey Eremondi's thesis on gradual dependent types
@@
 However, they do not give a denotational semantics to term precision and it is unclear
-how to prove the graduality in this setting.
+how to prove graduality in this setting.
 The work includes a formalization of the syntactic model in Guarded Cubical Agda.
@@
 
-\subsection{Comparison to Embedding-projection pairs}
-
-A line of work by New, Licata and Ahmed proves graduality by
-interpreting type precision $c : A \ltdyn A'$ as an
-\emph{embedding-projection pair}, that is a pure function $e : A \to
-A'$ and a possibly erroring projection $p : \li A' \multimap \li A$
-such that $p \circ e = \id$ and $e \circ p \ltdyn \id$
-\cite{newahmed,newlicata,newlicataahmed}. At first glance, our model
-of precision looks somewhat different: it is a \emph{relation}
-$\sem{c} : A \rel A'$ with a quasi-representability structure that
-gives us something like $e$ and $p$ in an embedding-projection
-pair. The relationship between these models is that \emph{if} the
-relation were \emph{truly} representable rather than only
-quasi-representable we would have that $e$ and $p$ form a \emph{Galois
-connection} $e \circ p \ltdyn \id$ and $\id \ltdyn p \circ
-e$. However, we have dropped the stronger property of retraction from
-our analysis in this work. With true representability, the relation
-$c$ is \emph{uniquely determined} by the embedding $e$, but since we
-only have quasi-representability, we need to keep the relation around
-explicitly. So the extra complexity of managing explicit relations is
-a cost of the intensional reasoning that guarded type theory
-introduces.
-
-\subsection{Comparison to Logical Relations Models}
-Working internally to guarded type theory reduces the overhead of
-needing to carry around the step-indices in the proofs as is required
-when using explicit step-indexing. Additionally, the logical relations
-constructed to prove graduality in prior work \cite{new-ahmed2018,new-licata-ahmed2019,new-giovannini-licata-2022} suffer
-from technical complications of requiring two separate expression
-relations, one that counts steps on the left and the other on the
-right, and there is no analogue of this in our approach. However,
-using two expression relations allows some but not all transitive
-reasoning of term precision to be recovered. In the future we aim to
-explore if this approach is feasible in guarded semantics.
-
 % \subsection{Synthetic Ordering}

### 2024-07-12 paper-new/towards-relational-model.tex
add some missing citations

@@
 has some similarities to prior work on logical relations models
-\cite{newahmedlicata,newgiovanninilicata}, but we do not take this
+\cite{new-licata-ahmed2019,new-giovannini-licata-2022}, but we do not take this
 approach because the relations still do not seem to be provably

### 2024-07-12 paper-new/concrete-relational-model.tex
space

@@
 \tilde{x'}_t)$. Likewise for bisimilarity.
-%
-We also give a predomain structure to the natural numbers $\mathbb{N}$, where
-both the ordering and the bisimilarity relation are equality.\max{why natural numbers here? I don't think we need it. check and come back}
 
@@
   i_D(m_\to) &= \lambda d. \text{case $d$ of }
-    \{ \iarr (\tilde{f}) \To \iarr (\lambda t. i_{U(D \to \li D)}(m_\to)(\tilde{f}_t)) \alt x \To x \} \\
+    \{ \iarr (\tilde{f}) \To \iarr (\lambda t. i_{U(D \to \li D)}(m_\to)(\tilde{f}_t)) \alt x \To x \}
 \end{align*}
@@
 
-To model type precision equivalence we need to show the following quasi-equivalences:
-%
-\begin{lemma}
-The following hold:
-\begin{itemize}
-  \item $U(d \comp d') \qordeq U(d) \comp U(d')$
-  \item $\li(c \comp c') \qordeq \li (c) \comp \li (c')$
-  \item $(c \comp c') \to (d \comp d') \qordeq (c \to d) \comp (c' \to d')$
-  \item $(c_1 \comp c_1') \times (c_2 \comp c_2') \qordeq (c_1 \times c_2) \comp (c_1'\times c_2')$
-\end{itemize}
-\end{lemma}
-\begin{proof}
-  Details in the appendix (Lemma \ref{lem:quasi-order-equiv-functors}).
-\end{proof}
+The proofs that the precision equivalence imply quasi-equivalence of
+relations are in the appendix (Lemma
+\ref{lem:quasi-order-equiv-functors}).
 

### 2024-07-12 paper-new/intro.tex
casts sentence

@@
 Fundamentally, this property comes down to the behavior of
-\emph{runtime type casts}\max{TODO: introduce casts more thoroughly as
-  they are important}.
+\emph{runtime type casts}, which implement these generalized runtime
+assertions.
 
@@
 generation and higher-order references, which are commonplace in
-real-world gradually typed systems as well as gradual calculi
-\cite{examples-of-gradual-stuff}. Our long-term goal is to develop a
-semantics that will scale up to these advanced features, and so we
-begin with this work by departing from classical domain theory.
+real-world gradually typed systems as well as gradual calculi. Our
+long-term goal is to develop a semantics that will scale up to these
+advanced features, and so we begin with this work by departing from
+classical domain theory.
 

### 2024-07-12 paper-new/appendix.tex
squeezing

@@
 
-As mentioned in Section~\ref{sec:syntax}, the cast calculus we
+As mentioned in Section~\ref{sec:GTLC}, the cast calculus
 separates gradual type casts into downcasts and upcasts, whereas many
@@
 \[ \inferrule{\Gamma \vdash M : A}{\Gamma \vdash (M :: A') : A' }\]
-The term precision rules for this style of calculus are as follows\cite{siekvitousekciminiboyland}:
+The term precision rules for this style of calculus are as follows\cite{siek_et_al:LIPIcs:2015:5031}:
 \begin{mathpar}
@@
 
+\subsection{Functorial Actions on Syntactic Perturbations}
+
+\begin{itemize}
+  % \li
+  \item Then to interpret the perturbations as endomorphisms, by the universal property
+  of the coproduct of monoids it suffices to define two homomorphisms, one from
+  $\mathbb{N} \to \morbisimid{\li A}$ and one from $M_A \to \morbisimid{\li A}$.
+  The first homomorphism sends the generator $1 \in \mathbb{N}$ to the error
+  domain endomorpism $(\delta \circ \eta)^\dagger$, and the second sends $m_A \in
+  M_A$ to $\li(i_A(m_A))$ and we observe in both cases that the resulting
+  endomorphisms are bisimilar to the identity since the action of $\li$ on
+  morphisms preserves bisimilarity.
+
+  \item To construct the homomorphism $i_{A_1 \times A_2}$, we
+  appeal to the universal property and give homomorphisms $M_{A_1} \to
+  \morbisimid{A_1 \times A_2}$ and $M_{A_2} \to \morbisimid{A_1 \times A_2}$. The
+  former applies $i_{A_1}$ in the first component and the identity in the second
+  component, while the latter does the reverse with $i_{A_2}$.
+
+  \item To
+  construct the interpretation as endomorphisms $i_{A \arr B}$ it suffices to define a
+  homomorphism $M_A^{op} \to \morbisimid{A \arr B}$ and $M_B \to \morbisimid{A
+  \arr B}$. The former takes an element $m_A$ and a predomain morphism $f : A \to
+  UB$ and returns the composition $f \circ i_A(m_A)$, while the latter is defined
+  similarly by post-composition with $i_B$.
+
+\end{itemize}
+
 \subsection{Lemmas about Perturbations}
@@
 Recall the notion of quasi-equivalence of relations as defined in Definition
-\ref{def:quasi-order-equivalent}. The following lemma will be useful in showing
+\ref{def:quasi-equivalent}. The following lemma will be useful in showing
 that two relations are quasi-equivalent.

### 2024-07-12 paper-new/concrete-relational-model.tex
squeezing

@@
 predomain $\sem{A}$. These constructions are given by the same
-constructions as in Section~\ref{sec:tm-model}, but now enriched to
+constructions as in Section~\ref{sec:concrete-term-model}, but now enriched to
 act also on orderings and bisimilarity. For products this is the
@@
 $\laterhs$ for predomains, we can solve the same domain equation as in
-Section~\ref{sec:tm-model} changed only in that we interpret it as a
+Section~\ref{sec:concrete-term-model} changed only in that we interpret it as a
 domain equation on predomains rather than sets:
@@
 For the term semantics $\sem{M}$, nothing really differs from the term
-semantics in Section~\ref{sec:tm-semantics}. The only difference is
+semantics in Section~\ref{sec:concrete-term-model}. The only difference is
 that the definitions of $\sem{\upc c}$ and $\sem{\dnc c}$ can be
@@
 
-\eric{May be able to move some of this to the appendix} For our value
-type semantics, we already have an interpretation as predomains, we
-describe how we enhance each of the constructions we use
-$U,\li,\to,\times,D$ with a notion of syntactic perturbation.
+% \eric{May be able to move some of this to the appendix}
+For our value type semantics, we already have an interpretation as predomains,
+we describe how we enhance each of the constructions we use $U,\li,\to,\times,D$
+with a notion of syntactic perturbation.
 
@@
 %
-Then to interpret the perturbations as endomorphisms, by the universal property
-of the coproduct of monoids it suffices to define two homomorphisms, one from
-$\mathbb{N} \to \morbisimid{\li A}$ and one from $M_A \to \morbisimid{\li A}$.
-The first homomorphism sends the generator $1 \in \mathbb{N}$ to the error
-domain endomorpism $(\delta \circ \eta)^\dagger$, and the second sends $m_A \in
-M_A$ to $\li(i_A(m_A))$ and we observe in both cases that the resulting
-endomorphisms are bisimilar to the identity since the action of $\li$ on
-morphisms preserves bisimilarity.
+The interpretation of the perturbations as endomorphisms is defined using the
+universal property of the coproduct of monoids (see appendix for details).
 
@@
 and $(A_2, M_{A_2}, i_{A_2})$, we define the monoid $M_{A_1 \times A_2} =
-M_{A_1} \oplus M_{A_2}$. To construct the homomorphism $i_{A_1 \times A_2}$, we
-appeal to the universal property and give homomorphisms $M_{A_1} \to
-\morbisimid{A_1 \times A_2}$ and $M_{A_2} \to \morbisimid{A_1 \times A_2}$. The
-former applies $i_{A_1}$ in the first component and the identity in the second
-component, while the latter does the reverse with $i_{A_2}$.
+M_{A_1} \oplus M_{A_2}$. The interpretation is defined using the universal property
+of the coproduct (see appendix for the details).
 % Could state this using the functorial action of \times
@@
 The action of $\arr$ on objects is as follows. Given $(A, M_A, iA)$ and $(B,
-M_B, i_B)$ we define the monoid $M_{A \arr B} = M_A^{op} \oplus M_B$. Then to
-construct the interpretation as endomorphisms it suffices to define a
-homomorphism $M_A^{op} \to \morbisimid{A \arr B}$ and $M_B \to \morbisimid{A
-\arr B}$. The former takes an element $m_A$ and a predomain morphism $f : A \to
-UB$ and returns the composition $f \circ i_A(m_A)$, while the latter is defined
-similarly by post-composition with $i_B$.
+M_B, i_B)$ we define the monoid $M_{A \arr B} = M_A^{op} \oplus M_B$ and the interpretation
+is defined analogously to that of $\times$.
 % Could state this using the functorial action of \arr

### 2024-07-12 paper-new/concrete-term-model.tex
squeezing

@@
 The interpretation of the typical CBV features of GTLC into a CBPV is
-standard \cite{levy}, the only additional part we need to account for
+standard \cite{levy99}, the only additional part we need to account for
 is the semantics of casts, shown in Figure \ref{fig:term-semantics}.

### 2024-07-12 paper-new/towards-relational-model.tex
squeezing

@@
 Firstly, to model graduality, $\mho$ is the least element. Next, two
-returned values are related just when they are in the ordering $\leq$
+returned values are related just when they are in the ordering $\ltdyn$
 on $A$.
@@
         %	
-        \theta\, \tilde{l} \semltbad \eta\, y &\text{ iff } \exists n. \exists x \leq y.
+        \theta\, \tilde{l} \semltbad \eta\, y &\text{ iff } \exists n. \exists x \ltdyn y.
             (\theta\, \tilde{l} = \delta^n(\eta\, x)) \\
@@
 a relation between $A$ and $A'$ by $c : A \rel A'$. The paradigmatic example of
-a poset relation is the poset ordering itself, $\leq_A : A \rel A$, which is a
+a poset relation is the poset ordering itself, $\ltdyn_A : A \rel A$, which is a
 poset relation because it is transitive.
@@
 that the relation is a kind of \emph{graph} of the function in that $x
-\binrel{\sem c} y$ if and only if $\sem{\upc c}(x) \leq_{A_2} y$ \cite{shulmandoublecategories}.
+\binrel{\sem c} y$ if and only if $\sem{\upc c}(x) \leq_{A_2} y$ \cite{shulman2007}.
 
@@
 %
-% https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMSwxLCJBXzIiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzMsNSwiXFxsZV97QV8yfSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzUsNCwiYyciLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDMsIlxcdXBje2N9IiwyXSxbMiw0LCJcXGlkIl0sWzEsNSwiXFxpZCIsMl1d
+% https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMSwxLCJBXzIiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzMsNSwiXFxsdGR5bl97QV8yfSIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzUsNCwiYyciLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDMsIlxcdXBje2N9IiwyXSxbMiw0LCJcXGlkIl0sWzEsNSwiXFxpZCIsMl1d
 \[\begin{tikzcd}[ampersand replacement=\&]
@@
 	\arrow["\id", from=1-3, to=2-3]
-	\arrow["{\le_{A_2}}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+	\arrow["{\ltdyn_{A_2}}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
 	\arrow["{c'}"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
@@
 %
-Observe that the composition of $\le_{A_2}$ with $c'$ implies
-$c'$, because $c'$ is \emph{downward-closed} under the relation on $A_2$. Thus, we
-obtain the original UpL square by vertically composing the square shown above
-with the square $\id \ltsq{\le_{A_2}\, c'}{c'} \id$ on the bottom.
+Then to complete the proof, we observe that the composition of $\ltdyn_{A_2}$ with $c'$ is equal to
+$c'$, because $c'$ is \emph{downward-closed} under the relation on $A_2$.
+% Thus, we
+% obtain the original UpL square by vertically composing the square shown above
+% with the square $\id \ltsq{\le_{A_2}\, c'}{c'} \id$ on the bottom.
 
@@
 needed in this argument occurs when $A_3 = A_2$ and we take $c'$ to be
-$\le_{A_2}$ (recall that $c'$ is universally quantified, so it can be any
-relation). The fact that $\le_{A_2}$ is downward-closed is equivalent to it
-being transitive. In other words, all relations $\le_A$ on value types must be
+$\ltdyn_{A_2}$ (recall that $c'$ is universally quantified, so it can be any
+relation). The fact that $\ltdyn_{A_2}$ is downward-closed is equivalent to it
+being transitive. In other words, the relations $\ltdyn_A$ on value types must be
 poset relations in order to carry out the above compositional construction.
@@
         &\eta\, x \ltls \eta\, y \text{ if } 
-            x \mathbin{\le_A} y \\
+            x \mathbin{\ltdyn_A} y \\
         %		
@@
 When both sides are $\eta$, then we check that the returned values are related
-in $\le_A$. The error term $\mho$ is the least element. Lastly, if both sides step
+in $\ltdyn_A$. The error term $\mho$ is the least element. Lastly, if both sides step
 (i.e., are a $\theta$) then we compare the resulting computations one time step
@@
 parameterized by a binary relation $\bisim_A$ on $A$
-\cite{guarded-stuff-that-uses-bisim}.
+\cite{mogelberg-paviotti2016}.
 % For a type $X$, we define a relation on $\li X$, called ``weak bisimilarity",

### 2024-07-12 paper-new/discussion.tex
ez squeeze

@@
 
-\subsubsection{From Static to Gradual}
+%%\subsubsection{From Static to Gradual}
 
@@
 
-\subsubsection{Double-Categorical Semantics}
+%\subsubsection{Double-Categorical Semantics}
 
@@
 
-\subsubsection{Embedding-Projection Pairs}
+%\subsubsection{Embedding-Projection Pairs}
 
@@
 
-\subsubsection{Step-Indexed Logical Relations Models}
+%\subsubsection{Step-Indexed Logical Relations Models}
 The standard alternative to domain theory that scales to essentially
@@
 
-\subsubsection{Gradual Dependent Types}
+%\subsubsection{Gradual Dependent Types}
 % Discuss Joey Eremondi's thesis on gradual dependent types

### 2024-07-12 paper-new/discussion.tex
25 pages

@@
 \cite{new-ahmed2018,new-licata18,new-licata-ahmed2019}. At first glance, our model
-of precision looks somewhat different: it is a \emph{relation}
+of precision looks different: it is a \emph{relation}
 $\sem{c} : A \rel A'$ with a quasi-representability structure that
@@
 pair. The relationship between these models is that \emph{if} the
-relation were \emph{truly} representable rather than only
-quasi-representable we would have that $e$ and $p$ form a \emph{Galois
+relation were \emph{truly} representable we would have that $e$ and $p$ form a \emph{Galois
 connection} $e \circ p \ltdyn \id$ and $\id \ltdyn p \circ
@@
 %
-We have formalized the major components of the definition of the concrete model
+We formalized the major components of the definition of the concrete model
 described in Section \ref{sec:concrete-relational-model}: predomains, error
-domains, their morphisms, relations, and squares, and we used the guarded
+domains, morphisms, relations, squares, using the guarded
 features to define the free error domain, the lock-step error ordering, and weak

### 2024-10-01 paper-new/appendix.tex
check in small changes to paper

@@
 
-\begin{itemize}
-  % \li
-  \item Then to interpret the perturbations as endomorphisms, by the universal property
+We provide details on the definition of the homomorphisms specifying the
+interpretation of perturbations corresponding for the functors $\li$, $U$,
+$\times$, and $\arr$.
+
+\begin{itemize}
+  \item To define the homomorphism $i_{\li A}$, by the universal property
   of the coproduct of monoids it suffices to define two homomorphisms, one from
@@
 
-  \item To construct the homomorphism $i_{A_1 \times A_2}$, we
+  \item To define the homomorphism $i_{A_1 \times A_2}$, we
   appeal to the universal property and give homomorphisms $M_{A_1} \to
@@
 
-  \item To
-  construct the interpretation as endomorphisms $i_{A \arr B}$ it suffices to define a
+  \item To define the homomorphism $i_{A \arr B}$ it suffices to define a
   homomorphism $M_A^{op} \to \morbisimid{A \arr B}$ and $M_B \to \morbisimid{A

### 2024-10-01 paper-new/concrete-term-model.tex
check in small changes to paper

@@
 section we define a simpler semantics that validates all of our
-desired properties except for graduality: that is it is adequate and
+desired properties except for graduality: that is, it is adequate and
 validates $\beta\eta$ equality. Our relational semantics will then be
@@
   U$ to the forgetful functor, with $\eta$ the unit of the adjunction.
+  %
+  We will write $\theta_t(\dots)$ to mean $\theta (\lambda t. \dots)$.
 \end{definition}

### 2024-10-01 paper-new/discussion.tex
check in small changes to paper

@@
 paper. However, because of the lack of transitivity of the relation modeling
-term precision, our model is not a double category but instead a reflexive graph
-category, that is, a double category without a notion of relational composition.
+term precision, our model is not a double category because we cannot compose
+squares horizontally.
 

### 2024-10-01 paper-new/intro.tex
check in small changes to paper

@@
 $x_i \in X_i$ compatible with the restriction maps. For example, consider the
-set of programs that may take a computational step or return unit. In the topos
-of trees, we can model this set as the indexed family of sets $X_n =
+set of programs that may take a computational step or return unit.\footnote{This example is adapted from \cite{mogelberg-paviotti2016}.} In the
+topos of trees, we can model this set as the indexed family of sets $X_n =
 \{0,1,\dots,n-1, \bot\}$ for $n \ge 1$, where $i$ denotes a program that steps
@@
 terminates. On the other hand, if we instead consider the denotation of a
-program to be a global element of the presheaf $\{X_n\}$ -- a family of
-elements $x_i \in X_i$ for all $i \in \mathbb{N}$ -- then the denotation of the
-diverging program will be distinct from that of a program that terminates,
-regardless of the number of steps it takes.
+program to be a global element of the presheaf $\{X_n\}$ -- a family of elements
+$x_i \in X_i$ for all $i \in \mathbb{N}$ -- then the denotation of the diverging
+program will be distinct from that of a program that terminates, regardless of
+the number of steps it takes.
 

### 2024-10-01 paper-new/towards-relational-model.tex
check in small changes to paper

@@
 \ltsq{c}{c} \id$, as can be seen by unfolding the
-definition. Similarly, for any montone function $f : A_i \to A_o$,
+definition. Similarly, for any monotone function $f : A_i \to A_o$,
 there is a ``horizontal identity'' square $f
@@
 It is straightforward to prove using \lob-induction that this relation is
-reflexive, transitive and anti-symmetric given that the underyling relation $R$
+reflexive, transitive and anti-symmetric given that the underlying relation $R$
 has those properties. The lock-step ordering is therefore the partial ordering

### 2024-10-04 paper-new/intro.tex
some fixes to paper based on reviews

--- 2024-07-12 paper-new/intro.tex
@@
 N$ are equivalent in the equational theory then they have the same
-semantics, $M \Downarrow n \iff N \downarrow n$ and $M\Uparrow \iff N
+semantics, $M \Downarrow n \iff N \Downarrow n$ and $M\Uparrow \iff N
 \Uparrow$ and $M \Downarrow \mho \iff N \Downarrow \mho$.
@@
 (ultra)metric spaces but more commonly as ``step-indexed sets'', i.e.,
-objects in the \emph{topos of trees}, i.e., presheaves on the poset of
-natural numbers\cite{birkedal-mogelberg-schwinghammer-stovring2011}. We think of such a
-$\mathbb{N}$-indexed set as modeling a kind of ``sequence of
-approximations'' to the true type being modeled.
+objects in the \emph{topos of trees} \cite{birkedal-mogelberg-schwinghammer-stovring2011}. 
+Such an object consists of a family $\{X_n\}_{n \in \mathbb{N}}$ of sets
+along with restriction functions $r_n : X_{n+1} \to X_n$ for all $n$.
+(These are simply presheaves on the poset of natural numbers, for the
+reader familiar with this terminology.)
+We think of such a $\mathbb{N}$-indexed set as modeling a kind of 
+``sequence of approximations'' to the true type being modeled.
 %
@@
 %
-Because bisimilarity, unlike equality, is not transitive this
-necessitates further changes to the core concepts of our semantics.
+% ERIC: This could be merged into the summary at the end of the section
+The reason for requiring these two separate notions is that in the setting of
+guarded domain theory, it is not possible to carry out transitive reasoning
+while being oblivious to the number of steps the elements take. We turn this
+statement into a formal theorem in \ref{sec:towards-relational-model}. To overcome
+this limitation, we develop further changes to the core concepts of our
+semantics as discussed in that section.
+
+% Because bisimilarity, unlike equality, is not transitive in the guarded
+% setting, this necessitates further changes to the core
+% concepts of our semantics. These challenges are discussed in Section \ref{sec:towards-relational-model}.
 
@@
 solution. This has the added benefit that providing a denotational
-semantics of gradually typed terms is as simple as a semantics of an
+semantics of gradually typed terms is as simple as defining the semantics of an
 effectful language using a monad on the category of sets. Not only
@@
 assistant. Whereas formalizing analytic guarded domain theory would
-require significant theory of presheaves and making sure that all
+require a significant theory of presheaves and making sure that all
 constructions are functors on categories of presheaves, formalizing
@@
 $x_i \in X_i$ compatible with the restriction maps. For example, consider the
-set of programs that may take a computational step or return unit. In the topos
-of trees, we can model this set as the indexed family of sets $X_n =
-\{0,1,\dots,n-1, \bot\}$ for $n \ge 1$, where $i$ denotes a program that steps
+set of programs that may take a computational step or return unit.\footnote{This example is adapted from \cite{mogelberg-paviotti2016}.} In the
+topos of trees, we can model this set as the indexed family of sets $\{X_n\}_{n \in \mathbb{N}}$
+where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1, \bot\}$. Here, $i$ denotes a program that steps
 $i$ times and then returns, and $\bot$ denotes a program that fails to terminate
@@
 terminates. On the other hand, if we instead consider the denotation of a
-program to be a global element of the presheaf $\{X_n\}$ -- a family of
-elements $x_i \in X_i$ for all $i \in \mathbb{N}$ -- then the denotation of the
-diverging program will be distinct from that of a program that terminates,
-regardless of the number of steps it takes.
+program to be a global element of the presheaf $\{X_n\}_{n \in \mathbb{N}}$ -- a family of elements
+$x_i \in X_i$ for all $i \in \mathbb{N}$ -- then the denotation of the diverging
+program will be distinct from that of a program that terminates, regardless of
+the number of steps it takes.
 
@@
 demonstrating that the semantics is readily mechanizable.
+We provide an overview of the mechanization effort in Section \ref{sec:mechanization}.
 
@@
   proving graduality in an intensional model.
+  %
+  % ... including the reason why we need to introduce weak bisimilarity.
+  %
+  The key technical issue is that reasoning principles that are simultaneously
+  extensional \emph{and} compositional are not compatible with guarded domain
+  theory (here, extensional means relating elements that differ only in the
+  number of computational steps they take). We make this statement precise with
+  a no-go theorem (Theorem \ref{thm:no-go}). As a result, we split the
+  graduality ordering into two relations: a strong error ordering in which
+  elements must take the same number of steps, and a weak bisimilarity relation
+  where elements differing only in their stepping behavior are related.
+
 \item In Section \ref{sec:concrete-relational-model} we describe the resulting

### 2024-10-22 paper-new/concrete-relational-model.tex
ez squeeze
check in small changes to paper
paper revisions

@@
 
-A \emph{predomain relation} $c : A \rel A'$ is a monotone relation as
-defined in the previous section (no condition is required for
-bisimilarity). The ``reflexive'' predomain relation is written $r(A)$
-and is given by the ordering $\ltdyn_A$. The relational composition of
-two predomains is also monotone, and $r(A)$ is the identity for this
-composition.
+A \emph{predomain relation} $c : A \rel A'$ is a relation between $A$ and $A'$
+that is upward closed in $A'$ and downward closed in $A$, as defined in the
+previous section (no condition is required for bisimilarity). The ``reflexive''
+predomain relation is written $r(A)$ and is given by the ordering $\ltdyn_A$.
+The relational composition of two predomain relations is also a predomain
+relation, and $r(A)$ is the identity for this composition.
 
@@
 bisimilar to the identity. We call such an object a \emph{value
-object} and this will serve as the final definition of a denotation of
-a type in our semantics. We define computation objects to similarly be
+object}, and this will serve as the final definition of a denotation of
+a type in our semantics. Similarly, we define computation objects to be
 error domains equipped with a notion of syntactic perturbation:
@@
   A \textbf{value object} consists of a predomain $A$, a monoid $M_A$ and a
-  homomorphism of monoids $M_A \to \morbisimid{A} := \{ f : A \to A \mid f
+  homomorphism of monoids $i_A \colon M_A \to \morbisimid{A} := \{ f : A \to A \mid f
   \bisim \id_A \}$.
@@
   Likewise, a \textbf{computation object} consists of an error domain $B$, a
-  monoid $M_B$ and a homomorphism of monoids $M_B \to \morbisimid{B}$.
+  monoid $M_B$ and a homomorphism of monoids $i_B \colon M_B \to \morbisimid{B}$.
 
-  We define a \textbf{value morphism} to be a morphism of the underlying
+  We define a \textbf{value morphism} to be simply a morphism of the underlying
   predomains, and a \textbf{computation morphism} to be a morphism of the
@@
 \end{definition}
+
+\begin{remark}
+  The reason the set of perturbations needs to carry a monoid structure is that
+  we will need to produce a syntactic perturbation corresponding to the identity
+  relations $r(A)$, as well as one for the composition of relations. The monoid
+  structure gives us what we need to produce these elements. Likewise, the
+  reason that the interpretations as semantic perturbations must be
+  \emph{homomorphisms} of monoids is that we need a syntactic perturbation that
+  is interpreted as the identity semantic perturbation, and likewise given two
+  syntactic perturbations $\delta_1$ and $\delta_2$ we need a syntactic
+  pertubration $\delta_{1,2}$ that is interpreted as the composition of the
+  semantic perturbations $i_A(\delta_1)$ and $i_A(\delta_2)$.
+\end{remark}
+
+\begin{remark}
+  It might seem strange that in defining the morphisms of value and computation
+  objects, there is no requirement involving the monoids. For example, one could
+  imagine defining a morphism between value objects $(A, M_A, i_A)$ and $(A',
+  M_{A'}, i_{A'})$ to be a predomain morphism $f \colon A \to A'$ along with a
+  monoid homomorphism $h \colon M_A \to M_{A'}$ such that $f$ and $h$ satisfied
+  a compatibility condition with respect to the interpretations $i_A$ and
+  $i_{A'}$. However, with this definition of morphism, the category of value
+  types is no longer Cartesian closed and is not a model of call-by-push-value.
+  This is unfortunate, as it would simplify some aspects of the model construction.
+\end{remark}
 
@@
 morphisms on either side to be ``out of sync'' rather than in lock step.
+In the below, $f' \bisimsq{A_i}{A_o} f$ is the natural extension of
+bisimilarity to morphisms: given bisimilar inputs, the functions have
+bisimilar outputs.
 \begin{definition}
@@
 \end{definition}
-Here $f' \bisimsq{A_i}{A_o} f$ is the natural extension of
-bisimilarity to morphisms: given bisimilar inputs the functions have
-bisimilar outputs.
 
@@
 \begin{theorem}
+  \,
   \begin{itemize}
@@
 % \eric{May be able to move some of this to the appendix}
-For our value type semantics, we already have an interpretation as predomains,
-we describe how we enhance each of the constructions we use $U,\li,\to,\times,D$
-with a notion of syntactic perturbation.
+For our value type semantics, we already have an interpretation as predomains;
+we describe how we enhance each of the constructions we use ($U,\li,\to,\times,D$)
+with a monoid of syntactic perturbation and a means of interpreting them as semantic
+perturbations.
 
@@
 The interpretation of the perturbations as endomorphisms is defined using the
-universal property of the coproduct of monoids (see appendix for details).
+universal property of the coproduct of monoids. Specifically, to define the
+homomorphism $i_{\li A}$, by the universal property of the coproduct of monoids
+it suffices to define two homomorphisms, one from $\mathbb{N} \to
+\morbisimid{\li A}$ and one from $M_A \to \morbisimid{\li A}$. The first
+homomorphism sends the generator $1 \in \mathbb{N}$ to the error domain
+endomorpism $(\delta \circ \eta)^\dagger$, and the second sends $m_A \in M_A$ to
+$\li(i_A(m_A))$. We observe in both cases that the resulting endomorphisms are
+bisimilar to the identity, since in the first case, $-^\dagger$ preserves
+bisimilarity, and in the second case, the action of $\li$ on morphisms preserves
+bisimilarity.
 
-Now we discuss the action of $U$ on objects. Given a computation object $(B,
-M_B, i_B)$, we define $M_{UB}$ to be $\mathbb{N} \oplus M_B$. The reason for
-requiring $\mathbb{N}$ is related to the Kleisli arrow functor: to establish
-quasi-representability of $U(c \arr d)$ given quasi-representability of $c$ and
-$d$, we will need to turn a perturbation on $\li A$ into a perturbation on $U(A
-\arr B)$. Since the perturbations for $\li A$ involve $\mathbb{N}$, so must the
-perturbations for $UB$. The interpretation $i_{UB}$ of the perturbations on $UB$
-works in the same manner as that of $\li A$.
+Now we define the action of $U$ on objects. Given a computation object $(B, M_B,
+i_B)$, we define $M_{UB}$ to be $\mathbb{N} \oplus M_B$. The reason for
+requiring $\mathbb{N}$ is related to the Kleisli action of the arrow functor: to
+establish quasi-representability of $U(c \arr d)$ given quasi-representability
+of $c$ and $d$, we will need to turn a perturbation on $\li A$ into a
+perturbation on $U(A \arr B)$. Since the perturbations for $\li A$ involve
+$\mathbb{N}$, so must the perturbations for $UB$. The interpretation $i_{UB}$ of
+the perturbations on $UB$ works in an analogous manner to that of $\li A$: we send
+the generator $1 \in \mathbb{N}$ to the delay morphism $\delta_B \colon UB \to
+UB$, and we send a perturbation $m_B \in M_B$ to $U(i_B(m_B))$.
 
@@
 and $(A_2, M_{A_2}, i_{A_2})$, we define the monoid $M_{A_1 \times A_2} =
-M_{A_1} \oplus M_{A_2}$. The interpretation is defined using the universal property
-of the coproduct (see appendix for the details).
+M_{A_1} \oplus M_{A_2}$. The interpretation $i_{A_1 \times A_2}$ is defined
+using the universal property of the coproduct of monoids. That is, we define
+homomorphisms $M_{A_1} \to \morbisimid{A_1 \times A_2}$ and $M_{A_2} \to
+\morbisimid{A_1 \times A_2}$. The former takes $m_1 \in M_{A_1}$ to the morphism
+$(i_{A_1}(m_1), \id)$, while the latter takes $m_2 \in M_{A_2}$ to the morphism
+$(\id, i_{A_2}(m_2))$.
 % Could state this using the functorial action of \times
@@
 Finally, the most interesting type is of course the dynamic type. 
-Now we define the perturbations for $D$. Recall that to each predomain $A$ we
-associate a monoid $M_A$ of syntactic perturbations and a homomorphism into the
-monoid of endomorphisms bisimilar to the identity, and likewise for error
-domains. It may seem as though we need to define the perturbations for $D$ by
+It may seem as though we need to define the perturbations for $D$ by
 guarded recursion, but in fact we define them via least-fixpoint in the category
@@
 definition of the relation $\iarr$.
-
-\subsubsection{Interpreting Type Precision Equivalence}
 

### 2024-10-22 paper-new/concrete-term-model.tex
paper revisions

@@
 access to such a term ``later'' and use that to help us build a term ``now''.
-This operator satisfies the axiom that $\fix f = f (\nxt (\fix f))$. In
-particular, this axiom applies to propositions $P : \texttt{Prop}$; proving a
-statement in this manner is known as $\lob$-induction.
+This operator satisfies the axiom that $\fix f = f (\nxt (\fix f))$.
+%
+In particular, $\fix$ applies when type $T$ is instantiated to a proposition 
+$P : \texttt{Prop}$; in that case, it corresponds to $\lob$-induction.
 
@@
 term $M[t'] : A[t'/t]$. However, there is an important restriction on when we
-are allowed to apply ticks. In order to apply $M$ to tick $t$, $M$ must be
+are allowed to apply ticks: To apply $M$ to tick $t$, we require $M$ to be
 well-typed in the prefix of the context occurring before the tick $t$. That is,
@@
 We now define from the above term model a ``big-step'' semantics. More
-concretely, our goal is to define a function 
+concretely, our goal is to define a partial function 
 $-\Downarrow : \{M \,|\, \cdot \vdash M : \nat \} \rightharpoonup \mathbb{N} + {\mho}$.
@@
 coinductive types using guarded recursion. That is, given a type $A$ with a free
-variable $k : \Clock$, we can form the type $A^{gl} := \forall k. A$. Given a
+variable $k : \Clock$, we can form the type $A^{gl} := \forall k. A$ (here, $gl$ stands for ``global''). Given a
 functor $F$ that satisfies a technical condition, if $A$ is a guarded recursive
@@
 
-We can define a notion of termination in $i$ steps for the Delay monad $d \da^i
-n_?$ where $d : \delay(\mathbb{N} + {\mho})$ and $n_? : \mathbb{N} + {\mho}$. From
-this, can define a partial function $\delay (\mathbb{N} + {\mho})
-\rightharpoonup \mathbb{N} + {\mho}$ by existentially quantifying over $i$
-for which $d \da^i n_?$.
+Given $d : \delay(\mathbb{N} + {\mho})$, we define a notion of termination in
+$i$ steps, written $d \da^i$. We define $d \da^i$ inductively by the rules
+$(\tnow\, n_?) \da^0$ and $(\tlater\, d) \da^{i+1}$ if $d \da^i$ (here, $n_? :
+\mathbb{N} + {\mho}$). From this definition, it is clear that if there is an $i$
+such that $d \da^i$, then this $i$ is unique. Furthermore, from the proof that
+$d$ terminates in $i$ steps, we can extract the result with which $d$
+terminates. This is defined by a straightforward induction on the proof that $d \da^i$.
 
-Now consider the function $\{M \,|\, \cdot \vdash M : \nat \} \to \li^{gl}
-\mathbb{N}$ defined by $\lambda M. \lambda k .\sem{M}$. Composing this with the
-above partial function yields the desired partial function $-\Downarrow \colon \{M
-\,|\, \cdot \vdash M : \nat \} \rightharpoonup \mathbb{N} + {\mho}$. It is
-then straightforward to see that this big-step semantics respects the equational
-theory and that the denotations of the diverging term $\Omega$, the error term
-$\mho$, and a syntactic natural number value $n$ are all distinct.
+Using this definition of termination, we define a partial function 
+%
+$-\Downarrow_\delay \colon \delay (\mathbb{N} + {\mho}) \rightharpoonup \mathbb{N} + {\mho}$
+%
+by existentially quantifying over $i$ for which $d \da^i n_?$. That is, we define
+$d \Downarrow_\delay n_? := \exists i. d \da^i n_?$.
+
+Now consider the global version of the term semantics, $\sem{\cdot}^{gl} \colon
+\{M \,|\, \cdot \vdash M : \nat \} \to \li^{gl} \mathbb{N}$, defined by
+$\sem{M}^{gl} := \lambda (k : \Clock) .\sem{M}$. Composing this with the partial
+function $\Downarrow_\delay$ yields the desired partial function
+%
+$-\Downarrow \colon \{M \,|\, \cdot \vdash M : \nat \} \rightharpoonup \mathbb{N} + {\mho}$.
+%
+It is then straightforward to see that this big-step semantics respects the
+equational theory and that the denotations of the diverging term $\Omega$, the
+error term $\mho$, and a syntactic natural number value $n$ are all distinct.

### 2024-10-22 paper-new/discussion.tex
some fixes to paper based on reviews
paper revisions

@@
 \section{Discussion}\label{sec:discussion}
-
-% \eric{I moved this from the intro. We may need to revise it.}
 
@@
 the predefined framework.  Many gradually typed languages do not fit into either
-framework, e.g., Typed Racket \cite{tobin-hochstadt06, tobin-hochstadt08} and
+framework, e.g., Typed Racket \cite{tobin-hochstadt06, tobin-hochstadt08}, and
 the semantics produced is not always the desired one.
@@
 
+% \subsubsection{Graduality via Embedding-Projection Pairs}
+
+A line of work by New, Licata and Ahmed
+\cite{new-ahmed2018,new-licata18,new-licata-ahmed2019} develops an axiomatic
+account of gradual typing and proves graduality by interpreting type precision
+$c : A \ltdyn A'$ as an \emph{embedding-projection pair}, that is, a pure
+embedding function $e : A \to A'$ and a possibly erroring projection $p : \li A'
+\multimap \li A$ such that $p \circ e = \id$ and $e \circ p \ltdyn \id$. At
+first glance, our model of precision looks different: it is a \emph{relation}
+$\sem{c} : A \rel A'$ with a quasi-representability structure that gives us
+something close to $e$ and $p$ in an embedding-projection pair. The relationship
+between these models is that \emph{if} the relation were \emph{truly}
+representable we would have that $e$ and $p$ form a \emph{Galois connection} $e
+\circ p \ltdyn \id$ and $\id \ltdyn p \circ e$. However, we have dropped the
+stronger property of retraction from our analysis in this work. With true
+representability, the relation $c$ is \emph{uniquely determined} by the
+embedding $e$, but since we only have quasi-representability, we need to keep
+the relation around explicitly. So the extra complexity of managing explicit
+relations is a cost of the intensional reasoning that guarded type theory
+introduces.
+
+% Later, New, Licata and Ahmed \cite{new-licata-ahmed2019} extended this axiomatic
+% treatment from call-by-name to call-by-value as well by giving an axiomatic
+% theory of type and term precision in call-by-push-value.
+
+% \subsubsection{Denotational and Operational Approaches}
+
+The above axiomatic account of gradual typing is independent of the choice of
+operational versus denotational semantics; we give an overview of the two
+approaches below.
+
 %\subsubsection{Double-Categorical Semantics}
 
-% New and Licata \cite{new-licata18} developed an axiomatic account of
-% the graduality relation on a call-by-name cast calculus terms and
-% showed that the graduality proof could be modeled using semantics in
-% certain kinds of \emph{double categories}, categories internal to the
-% category of categories. A double category extends a category with a
-% second notion of morphism, often a notion of ``relation'' to be paired
-% with the notion of functional morphism, as well as a notion of
-% functional morphisms preserving relations. In gradual typing the
-% notion of relation models type precision and the squares model the
-% term precision relation. This approach was influenced by the semantics
-% of parametricity using reflexive graph categories
-% \cite{ma-reynolds,dunphythesis,reynoldsprogramme}: reflexive graph
-% categories are essentially double categories without a notion of
-% relational composition. In addition to capturing the notions of type
-% and term precision, the double categorical approach allows for a
-% \emph{universal property} for casts: upcasts are the \emph{universal}
-% way to turn a relation arrow into a function in a forward direction
-% and downcasts are the dual universal arrow.
+On the denotational side, New and Licata \cite{new-licata18} showed that the
+graduality proof could be modeled using semantics in certain kinds of double
+categories, which are categories internal to the category of categories. A
+double category extends a category with a second notion of morphism, often a
+notion of ``relation'' to be paired with the notion of functional morphism, as
+well as a notion of functional morphisms preserving relations. In gradual
+typing, the notion of relation models type precision and the squares model the
+term precision relation.
+%
+% This approach was influenced by the semantics of parametricity using reflexive
+% graph categories \cite{ma-reynolds,dunphythesis,reynoldsprogramme}: reflexive
+% graph categories are essentially double categories without a notion of
+% relational composition. In addition to capturing the notions of type and term
+% precision, the double categorical approach allows for a \emph{universal
+% property} for casts: upcasts are the \emph{universal} way to turn a relation
+% arrow into a function in a forward direction and downcasts are the dual
+% universal arrow.
+%
+The double-categorical framework serves as a foundation on which we base our
+model construction in this paper. However, our model is not actually a double
+category: the lack of transitivity of bisimilarity means that we cannot compose
+extensional squares horizontally.
 
-New and Licata \cite{new-licata18} developed an axiomatic account of the
-graduality relation on a call-by-name cast calculus and showed that the
-graduality proof could be modeled using semantics in certain kinds of double
-categories.
-%
-Later, New, Licata and Ahmed \cite{new-licata-ahmed2019} extended this axiomatic
-treatment from call-by-name to call-by-value as well by giving an axiomatic
-theory of type and term precision in call-by-push-value. The double-categorical
-framework serves as a foundation on which we base our model construction in this
-paper. However, because of the lack of transitivity of the relation modeling
-term precision, our model is not a double category because we cannot compose
-squares horizontally.
+With the notion of abstract categorical model for gradual typing in hand,
+denotational semantics is then the work of constructing concrete models that
+exhibit the categorical construction. New and Licata \cite{new-licata18} present
+such a model using categories of $\omega$-CPOs, and this model was extended by
+Lennon-Bertrand, Maillard, Tabareau and Tanter \cite{gradualizing-cic} to prove
+graduality of a gradual dependently typed calculus $\textrm{CastCIC}^{\mathcal
+G}$. This domain-theoretic approach meets our criterion of being a semantic
+framework for proving graduality, but suffers from the limitations of classical
+domain theory: the inability to model viciously self-referential structures such
+as higher-order extensible state and similar features such as runtime-extensible
+dynamic types.
 
-%% With this notion of abstract categorical model in hand, denotational
-%% semantics is then the work of constructing concrete models that
-%% exhibit the categorical construction. New and Licata
-%% \cite{new-licata18} present such a model using categories of
-%% $\omega$-CPOs, and this model was extended by Lennon-Bertrand,
-%% Maillard, Tabareau and Tanter to prove graduality of a gradual
-%% dependently typed calculus $\textrm{CastCIC}^{\mathcal G}$. This
-%% domain-theoretic approach meets our criteria of being a semantic
-%% framework for proving graduality, but suffers from the limitations of
-%% classical domain theory: the inability to model viciously
-%% self-referential structures such as higher-order extensible state and
-%% similar features such as runtime-extensible dynamic types. Since these
-%% features are quite common in dynamically typed languages, we seek a
-%% new denotational framework that can model these type system features.
-
-%\subsubsection{Embedding-Projection Pairs}
-
-A line of work by New, Licata and Ahmed proves graduality by
-interpreting type precision $c : A \ltdyn A'$ as an
-\emph{embedding-projection pair}, that is a pure function $e : A \to
-A'$ and a possibly erroring projection $p : \li A' \multimap \li A$
-such that $p \circ e = \id$ and $e \circ p \ltdyn \id$
-\cite{new-ahmed2018,new-licata18,new-licata-ahmed2019}. At first glance, our model
-of precision looks different: it is a \emph{relation}
-$\sem{c} : A \rel A'$ with a quasi-representability structure that
-gives us something like $e$ and $p$ in an embedding-projection
-pair. The relationship between these models is that \emph{if} the
-relation were \emph{truly} representable we would have that $e$ and $p$ form a \emph{Galois
-connection} $e \circ p \ltdyn \id$ and $\id \ltdyn p \circ
-e$. However, we have dropped the stronger property of retraction from
-our analysis in this work. With true representability, the relation
-$c$ is \emph{uniquely determined} by the embedding $e$, but since we
-only have quasi-representability, we need to keep the relation around
-explicitly. So the extra complexity of managing explicit relations is
-a cost of the intensional reasoning that guarded type theory
-introduces.
+% Since these features are quite common in dynamically typed languages, we have
+% developed in this work a new denotational framework that can be extended to
+% model these type system features.
 
 %\subsubsection{Step-Indexed Logical Relations Models}
-The standard alternative to domain theory that scales to essentially
-arbitrary self-referential definitions is \emph{step-indexing} or its
-synthetic form of \emph{guarded recursion}. A series of works
+% The standard alternative to domain theory that scales to essentially
+% arbitrary self-referential definitions is \emph{step-indexing} or its
+% synthetic form of \emph{guarded recursion}.
+
+On the operational side, a series of works
 \cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
-developed step-indexed logical relations models of gradually typed
-languages based on operational semantics. Unlike classical domain
-theory, such step-indexed techniques are capable of modeling
+developed step-indexed logical relations models of gradually typed languages.
+Unlike classical domain theory, such step-indexed techniques can scale to
+essentially arbitrary self-referential definitions, which means they can model
 higher-order store and runtime-extensible dynamic types
-\cite{appelmcallester01,ahmed06,neis09,new-jamner-ahmed19}. However,
-their proof developments are highly repetitive and technical, with
-each development formulating a logical relation from first-principles
-and proving many of the same tedious lemmas without reusable
-mathematical abstractions.
+\cite{appelmcallester01,ahmed06,neis09,new-jamner-ahmed19}. However, as is
+common with operational approaches, their proof developments are highly
+repetitive and technical, with each development formulating a logical relation
+from first-principles and proving many of the same tedious lemmas without
+reusable mathematical abstractions.
+%
 %
@@
 %
-This is addressed somewhat by Siek and Chen
-\cite{siek-chen2021}, who give a proof in Agda of graduality for an
-operational semantics who use a \emph{guarded logic of propositions}
-shallowly embedded in Agda. The guarded logic simplifies the treatment
-of step-indexed logical relations, but the approach is still
-fundamentally operational, and so the main lemmas of the work are
-still tied to the particular operational syntactic calculus being
-modeled. A further advantage of the denotational approach is that it
-easily validates equational reasoning, not just graduality, and it is
-completely independent of any particular syntax of gradual typing.
+This is addressed somewhat by Siek and Chen \cite{siek-chen2021}, who give a
+proof in Agda of graduality for an operational semantics using a \emph{guarded
+logic of propositions} shallowly embedded in Agda. The guarded logic simplifies
+the treatment of step-indexed logical relations, but the approach is still
+fundamentally operational, and so the main lemmas of the work are still tied to
+the particular operational syntactic calculus being modeled.
+% A further advantage of the denotational approach is that it easily validates
+% equational reasoning, not just graduality, and it is completely independent of
+% any particular syntax of gradual typing.
 
-%\subsubsection{Gradual Dependent Types}
-% Discuss Joey Eremondi's thesis on gradual dependent types
+Our work combines the benefits of both the denotational and step-indexed
+approaches. We take as a starting point the denotational approach of New and
+Licata, but we work with guarded type theory (a synthetic form of step-indexing)
+rather than classical domain theory. In the guarded setting, the compositional
+theory of New-Licata breaks down: when accounting for steps, the casts are no
+longer embedding-projection pairs, and we cannot carry out compositional
+reasoning on terms when those terms differ by an arbitrary unknown number of
+steps. Our work provides a novel refinement to their theory that allows for some
+compositional reasoning to be recovered even in the guarded setting.
+
+% \subsubsection{Gradual Dependent Types}
 Eremondi \cite{Eremondi_2023} uses guarded type theory to
@@
 how to prove graduality in this setting.
-The work includes a formalization of the syntactic model in Guarded Cubical Agda.
+Their work includes a formalization of the syntactic model in Guarded Cubical Agda.
 
-% Discuss Jeremy Siek's work on graduality in Agda
+\subsection{Comparison of Denotational and Operational Approaches}
+\eric{Some of this could be folded into the previous section, but this was one of the
+big-picture questions from one of the reviewers.}
 
-\subsection{Mechanization}
+Denotational methods have several benefits compared to operational approaches,
+and this carries over to the setting of gradual typing. First, denotational
+methods are compositional and reusable, and allow for the use of existing
+mathematical constructs and theorems, e.g., partial orderings, monads, etc,
+while operational methods tend to require a significant amount of boilerplate
+work to be done from scratch in each new development.
+%
+As a specific example of the compositional nature of our approach, the treatment
+of the cast rules in our work is more compositional than in previous work using
+operational semantics. Recall from Section \ref{sec:towards-relational-model}
+that the cast rules needed for the proof of graduality build in composition of
+type precision derivations. Rather than proving these from scratch in the model,
+we are able to take as primitive simpler versions of the cast rules whose
+validity in the model is easier to establish. Then from these simpler rules, we
+derive the original ones using compositional reasoning.
+
+A further advantage of the denotational approach is that it is completely
+independent of any particular syntax of gradual typing. This allows for reuse
+across multiple languages and makes it particularly straightforward to
+accommodate additions to a language: adding support for a new type amounts to
+defining a new object and extending the dynamic type accordingly. It also may be
+possible to model alternative cast semantics by changing the definition of the
+dynamic type and some type casts.
+%
+In contrast, operational methods are not as readily extensible, generally
+requiring adding cases to the logical relations and the inductive proofs. In a
+mechanized metatheory, this manifests as a ``copy-paste'' reusability rather than
+the true code reuse one obtains with a denotational semantics.
+%
+An additional benefit to the denotational approach is that it is trivial to
+establish the validity of the $\beta$ and $\eta$ principles, because they are
+equalities in the semantics, whereas they require tedious manual proof in the
+operational approach.
+
+Finally, while we advocate for a denotational approach, the overall structure of
+our semantics could be applied to an operational logical relations proof. For
+example, there could be a step-indexed logical relation for strong error
+ordering and one for weak bisimilarity, corresponding to the fact that objects
+in our denotational model have an error ordering and a bisimilarity relation.
+Then the precision rules for casts could be similarly validated using
+perturbations as we did in this work.
+
+\subsection{Implications for Gradual Typing Research}
+
+While we have focused on a single semantics in this work, we expect that our
+development can be adapted to prove graduality and (more limited) equational
+reasoning principles for other cast semantics.
+%
+In addition, while we have focused on proving extensional properties in this
+work, the intensional nature of our semantics can be viewed as a form of
+\emph{cost semantics}, where a runtime cost is incurred from inspecting the
+dynamic type. This could be used to verify that cast optimizations such as space
+efficient implementations \cite{herman-tomb-flanagan-2010} are not only
+extensionally correct, but also have lower quantitative cost. We leave this as
+future work.
+
+\subsection{Mechanization}\label{sec:mechanization}
 % Discuss Guarded Cubical Agda and mechanization efforts
@@
 quasi-representable.
-%
-We have not yet formalized the constructions involving value and computation
-relations: showing that these relations compose, and defining the actions of the
-functors $\li$, $U$, $\times$, and $\arr$ on the relations. We have also not yet
-formalized the syntax-to-semantics translation.
+
+% We have not yet formalized the constructions involving value and computation
+% relations: showing that these relations compose, and defining the actions of the
+% functors $\li$, $U$, $\times$, and $\arr$ on the relations. We have also not yet
+% formalized the syntax-to-semantics translation.
 
@@
 % \max{say these axioms are taken from prior work and cite that}
+
+The remaining formalization work is the following:
+\begin{enumerate}
+    \item Showing that the functors $\times$ and $\arr$ preserve
+    quasi-representability, which requires tedious reasoning about the Kleisli actions of
+    $\times$ and $\arr$. These results are proved in the Appendix.
+    \item Verifying the rules in the model corresponding to the equations for
+    type precision derivations (see the bottom of Figure \ref{fig:gtlc-syntax}).
+    \item Defining the syntax-to-semantics translation.
+\end{enumerate}
 
@@
 In the future, we plan to apply our approach to give a denotational semantics
-for languages that feature higher-order state or runtime-extensible dynamic
+to gradually-typed languages that feature higher-order state or runtime-extensible dynamic
 typing \cite{DBLP:journals/corr/abs-2210-02169} as well as richer type

### 2024-10-22 paper-new/intro.tex
paper revisions

@@
 In programming language design, there is a tension between \emph{static} typing
-and \emph{dynamic} typing disciplines.
-With static typing, the code is type-checked at compile time, while in dynamic typing,
-the type checking is deferred to run-time. Both approaches have benefits: with static 
-typing, the programmer is assured that if the program passes the type-checker, their
-program is free of type errors, and moreover, soundness of the equational theory implies
-that program optimizations are valid.
-Meanwhile, dynamic typing allows the programmer to rapidly prototype
-their application code without needing to commit to fixed type signatures for their
-functions. Most languages choose between static or dynamic typing and as a result, programmers that initially write their code in a dynamically typed language
-in order to benefit from faster prototyping and development time need to rewrite
-some or all of their codebase in a static language if they would like to receive the benefits of static
-typing once their codebase has matured.
+and \emph{dynamic} typing disciplines. With static typing, the code is
+type-checked at compile time, while in dynamic typing, the type checking is
+deferred to run-time. Both approaches have benefits and excel in different
+scenarios, with static typing offering compile-time assurance of a program's
+type safety and type-based reasoning principles that justify program
+optimizations, and dynamic typing allowing for rapid prototyping of a codebase
+without committing to fixed type signatures.
+%
+Most languages choose between static or dynamic typing and as a result,
+programmers that initially write their code in a dynamically typed language need
+to rewrite some or all of their codebase in a static language if they would like
+to receive the benefits of static typing once their codebase has matured.
 
-\emph{Gradually-typed languages} \cite{siek-taha06, tobin-hochstadt06} seek to resolve this tension
-by allowing for both static and dynamic typing disciplines to be used in the same codebase,
-and by supporting smooth interoperability between statically-typed and dynamically-typed code.
-This flexibility allows programmers to begin their projects in a dynamic style and
-enjoy the benefits of dynamic typing related to rapid prototyping and easy modification
-while the codebase ``solidifies''. Over time, as parts of the code become more mature
-and the programmer is more certain of what the types should be, the code can be
-\emph{gradually} migrated to a statically typed style without needing to
-rewrite the project in a completely different language.
+\emph{Gradually-typed languages} \cite{siek-taha06, tobin-hochstadt06} seek to
+resolve this tension by allowing for both static and dynamic typing disciplines
+to be used in the same codebase. These languages support smooth interoperability
+between statically-typed and dynamically-typed styles, allowing the programmer to
+begin with fully dynamically-typed code and \emph{gradually} migrate portions of the
+codebase to a statically typed style without needing to rewrite the project in a
+completely different language.
 
@@
 
+\eric{This paragraph could be deleted}
 In order for this to work as expected, gradually-typed languages should allow for
@@
 
+\eric{One of the reviewers said this should be moved later in the intro, but I think it makes sense here.}
 More concretely, consider a gradually typed language whose only
@@
 should be \emph{adequate}: natural number constants should step to
-themselves. Second it should validate type based reasoning. To
-formalize type based reasoning we give a typed equational theory for
+themselves. Second, it should validate type based reasoning. To
+formalize type based reasoning, we give a typed equational theory for
 terms of the language $M \equiv N$ for when two terms should be
@@
 expressive enough to be extended to model dependently typed gradual
-typing \cite{10.1145/3495528}.
+typing \cite{gradualizing-cic}.
 %
@@
 generation and higher-order references, which are commonplace in
-real-world gradually typed systems as well as gradual calculi. Our
-long-term goal is to develop a semantics that will scale up to these
+real-world gradually typed systems as well as gradual calculi.
+
+Our long-term goal is to develop a semantics that will scale up to these
 advanced features, and so we begin with this work by departing from
@@
 guarded domain theory, it is not possible to carry out transitive reasoning
-while being oblivious to the number of steps the elements take. We turn this
-statement into a formal theorem in \ref{sec:towards-relational-model}. To overcome
-this limitation, we develop further changes to the core concepts of our
-semantics as discussed in that section.
+while being oblivious to the number of steps the elements take. We make this
+statement precise in \ref{sec:towards-relational-model}. To overcome this
+limitation, we develop further changes to the core concepts of our semantics as
+discussed in that section.
 
@@
 
-\subsection{Contributions}
+\subsection{Contributions and Outline}
 
@@
 demonstrating that the semantics is readily mechanizable.
-We provide an overview of the mechanization effort in Section \ref{sec:mechanization}.
+We provide an overview of the mechanization effort, and what remains to be formalized,
+in Section \ref{sec:mechanization}.
 
+\begin{comment}
 \begin{enumerate}
@@
 \end{enumerate}
+\end{comment}
 
@@
   typical gradually typed cast calculus.
-\item In Section \ref{sec:concrete-term-model} we develop a first
-  denotational semantics in synthetic guarded domain theory that
-  satisfies adequacy and respects the equational theory, but does not
-  validate graduality. We use this to introduce some of our main
-  technical tools: modeling recursive types in guarded type theory and
-  modeling effects using call-by-push-value.
+\item In Section \ref{sec:concrete-term-model} we develop a denotational
+  semantics in synthetic guarded domain theory for the \emph{terms} of the
+  gradual lambda calculus. The model validates the equational theory, but it
+  does not prove graduality as it does not model type or term precision. We use
+  this to introduce some of our main technical tools: modeling recursive types
+  in guarded type theory and modeling effects using call-by-push-value. After
+  defining the model, we use the technique of globalization to extract a
+  well-behaved big-step term semantics satisfying adequacy.
 \item In Section \ref{sec:towards-relational-model} we show where the
@@
   number of computational steps they take). We make this statement precise with
-  a no-go theorem (Theorem \ref{thm:no-go}). As a result, we split the
-  graduality ordering into two relations: a strong error ordering in which
-  elements must take the same number of steps, and a weak bisimilarity relation
-  where elements differing only in their stepping behavior are related.
+  a no-go theorem (Theorem \ref{thm:no-go}). In light of this result, to recover
+  some amount of compositional reasoning we model the graduality ordering using
+  a combination of two relations: a lock-step error ordering in which elements
+  must take the same number of steps, and a weak bisimilarity relation where
+  elements differing only in their stepping behavior are related.
 
-\item In Section \ref{sec:concrete-relational-model} we describe the resulting
-  model in detail. We then prove that the resulting denotational model is
+  % As a result, we split the
+  % graduality ordering into two relations: a strong error ordering in which
+  % elements must take the same number of steps, and a weak bisimilarity relation
+  % where elements differing only in their stepping behavior are related.
+
+\item In Section \ref{sec:concrete-relational-model} we describe the
+  construction of the model in detail. Key here is a notion of \emph{syntactic
+  perturbations}, which allow us to encode the steps taken by a term in a form
+  that we can manipulate. This allows us to recover enough extensional reasoning
+  to model the graduality property compositionally.
+  % Key here is a notion of \emph{syntactic
+  % perturbation}, which give us a type-directed way of imposing delays on terms.
+  The resulting model validates the axioms for type and term precision specified
+  in Section \ref{sec:GTLC}. The final step is to prove that the model is
   \emph{adequate} for the graduality property: a closed term precision $M \ltdyn
-  N : \nat$ has the expected semantics, i.e., that $M$ errors or $M$ and $N$ have the
-  same extensional behavior.
+  N : \nat$ has the expected semantics, i.e., that $M$ errors or $M$ and $N$
+  have the same extensional behavior. We do so by extending the globalization
+  techniques used in defining the big-step term model to account for the
+  interpretation of term precision.
 \item In Section \ref{sec:discussion} we discuss prior work on proving
-  graduality, our partial mechanization of these results and discuss
+  graduality, the partial mechanization of our results in Agda, and
   future directions for denotational semantics of gradual typing.

### 2024-10-22 paper-new/towards-relational-model.tex
paper revisions

@@
 the thinking side to eventually terminate with a related behavior,
-which is the content of the final $4$ cases of the definition.
+which is the content of the final three cases of the definition.
 \begin{figure*}
@@
         \eta\, x \semltbad \eta\, y &\text{ iff } 
-            x \semlt y \\
+            x \ltdyn y \\
         %		
@@
         \eta\, x \semltbad \theta\, \tilde{l'} &\text { iff }
-            \exists n. \exists y \geq x. (\theta\, \tilde{l'} = \delta^n (\eta\, y))
+            \exists n. \exists y \gtdyn x. (\theta\, \tilde{l'} = \delta^n (\eta\, y))
     \end{align*}
@@
 with merely a reflexive relation, and continue using the
-step-insensitive ordering. However some level of transitivity is
+step-insensitive ordering. However, some level of transitivity is
 absolutely essential to providing compositional reasoning, and
@@
 %
-The reason that it is essential has to do with the details of the
-graduality proof, so first we expand on how we prove graduality
-compositionally. First of all, we need to extend our compositional
-semantics of types and terms to also a compositional semantics of
-\emph{type precision derivations} and \emph{term precision}.
+The reason that it is essential becomes clear by examining the details of the
+graduality proof, so first we expand on how we prove graduality compositionally.
+First, we need to extend our compositional semantics of types and terms to a
+compositional semantics of \emph{type precision derivations} and \emph{term
+precision}.
 %
-Sticking to a poset-based semantics, if our value types are posets, we then want
-our precision derivations $c : A \ltdyn A'$ to be \emph{poset relations} on $A,
-A'$, that is relations between the underlying sets of $A, A'$ that are
-\emph{downward-closed} in $A$ and \emph{upward-closed} in $A'$. We denote such
-a relation between $A$ and $A'$ by $c : A \rel A'$. The paradigmatic example of
-a poset relation is the poset ordering itself, $\ltdyn_A : A \rel A$, which is a
-poset relation because it is transitive.
+Sticking to a poset-based semantics, if our value types are interpreted as
+posets, then we want our precision derivations $c : A \ltdyn A'$ to be
+interpreted as a \emph{relation} between $A$ and $A'$. It is also natural to
+ask that the relation $c$ interact with the orderings on $A$ and $A'$. More
+specifically, if $x' \ltdyn_{A} x$ and $x \mathrel{c} y$, then $x' \mathrel{c}
+y$. Similarly, if $x \mathrel{c} y$ and $y \ltdyn_{A'} y'$, then $x \mathrel{c}
+y'$. We summarize this requirement by saying $c$ is \emph{downward-closed} in
+$A$ and $\emph{upward-closed}$ in $A'$. We will see below why this requirement
+is necessary for the graduality proof.
+
+% ...we then want our precision derivations $c : A \ltdyn A'$ to be \emph{poset relations} on $A,
+% A'$, that is, relations between the underlying sets of $A, A'$ that are
+% \emph{downward-closed} in $A$ and \emph{upward-closed} in $A'$.
+% The former means that if $x' \ltdyn_{A} x$ and $x \mathrel{c} y$, then $x' \mathrel{c} y$,
+% while the latter says that if $x \mathrel{c} y$ and $y \ltdyn_{A'} y'$, then $x \mathrel{c} y'$.
+
+We call such a relation between $A$ and $A'$ a \emph{poset relation}, and we
+write $c : A \rel A'$. The paradigmatic example of a poset relation is the poset
+ordering itself, $\ltdyn_A : A \rel A$, which is a poset relation because it is
+transitive.
+
 %
@@
 c$ where $c : A \ltdyn A'$, the terms denote elements of the
-free error domain $\sem{M},\sem{M'} : U\li \sem{A}$ and we model the term precision semantically as the relation holding between those denoted elements $\sem{M} \binrel{(U\li\sem{c})} \sem{M'}$ where
+free error domain $\sem{M},\sem{M'} : U\li \sem{A}$, and we model the term precision semantically as the relation holding between those denoted elements $\sem{M} \binrel{(U\li\sem{c})} \sem{M'}$ where
 $U\li\sem{c}$ is a relational lifting of $U\circ\li$.
@@
 transitivity, we can \emph{derive} the above squares from simpler ones
-that do not involve composition of relations. Below is the square
-corresponding to UpL:
-% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzIiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzIiXSxbMCwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJcXGxlX3tBXzJ9IiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJcXHVwY3tjfSIsMl0sWzIsMywiXFxpZCJdXQ==
-\[\begin{tikzcd}[ampersand replacement=\&]
+that do not involve composition of relations. Below are the simpler squares
+corresponding to UpL, UpR, DnL, and DnR:
+\begin{center}
+  \begin{tabular}{ c | c | c | c} 
+    % UpL
+    % https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzIiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzIiXSxbMCwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJcXGx0ZHluX3tBXzJ9IiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJcXHVwY3tjfSIsMl0sWzIsMywiXFxpZCJdLFs2LDcsIlxcdGV4dHtVcEx9IiwzLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+    \begin{tikzcd}[ampersand replacement=\&]
       {A_1} \& {A_2} \\
@@
       \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
-	\arrow["{\upc{c}}"', from=1-1, to=2-1]
-	\arrow["\id", from=1-2, to=2-2]
-	\arrow["{\le_{A_2}}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
-\end{tikzcd}\]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "{\upc{c}}"', from=1-1, to=2-1]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "\id", from=1-2, to=2-2]
+      \arrow["{\ltdyn_{A_2}}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+      \arrow["{\text{UpL}}"{marking, allow upside down}, draw=none, from=0, to=1]
+  \end{tikzcd} &
     %
-This square, with a dual one related to UpR, say that the relation $c$
-is \emph{representable} by the upcast morphism $\upc{c}$, essentially
-that the relation is a kind of \emph{graph} of the function in that $x
-\binrel{\sem c} y$ if and only if $\sem{\upc c}(x) \leq_{A_2} y$ \cite{shulman2007}.
+    % UpR
+    % https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzEiXSxbMSwwLCJBXzEiXSxbMSwxLCJBXzIiXSxbMCwyLCJcXGx0ZHluX3tBXzF9IiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJjIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJcXGlkIiwyXSxbMiwzLCJcXHVwYyBjIl0sWzYsNywiXFx0ZXh0e1VwUn0iLDMseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
+    \begin{tikzcd}[ampersand replacement=\&]
+      {A_1} \& {A_1} \\
+      {A_1} \& {A_2}
+      \arrow["{\ltdyn_{A_1}}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "\id"', from=1-1, to=2-1]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "{\upc c}", from=1-2, to=2-2]
+      \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+      \arrow["{\text{UpR}}"{marking, allow upside down}, draw=none, from=0, to=1]
+    \end{tikzcd} &
+    %
+    %
+    %
+    % DnL
+    % https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzIiXSxbMCwxLCJBXzEiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzIiXSxbMCwyLCJcXGx0ZHluX3tBXzJ9IiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJjIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJcXGRuYyBjIiwyXSxbMiwzLCJcXGlkIl0sWzYsNywiXFx0ZXh0e0RuTH0iLDMseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
+    \begin{tikzcd}[ampersand replacement=\&]
+      {A_2} \& {A_2} \\
+      {A_1} \& {A_2}
+      \arrow["{\ltdyn_{A_2}}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "{\dnc c}"', from=1-1, to=2-1]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "\id", from=1-2, to=2-2]
+      \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+      \arrow["{\text{DnL}}"{marking, allow upside down}, draw=none, from=0, to=1]
+    \end{tikzcd} &
+    %
+    % DnR
+    % https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzEiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzEiXSxbMCwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJcXGx0ZHluX3tBXzF9IiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJcXGlkIiwyXSxbMiwzLCJcXGRuYyBjIl0sWzYsNywiXFx0ZXh0e0RuUn0iLDMseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
+    \begin{tikzcd}[ampersand replacement=\&]
+      {A_1} \& {A_2} \\
+      {A_1} \& {A_1}
+      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "\id"', from=1-1, to=2-1]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "{\dnc c}", from=1-2, to=2-2]
+      \arrow["{\ltdyn_{A_1}}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+      \arrow["{\text{DnR}}"{marking, allow upside down}, draw=none, from=0, to=1]
+    \end{tikzcd}
+  \end{tabular}
+\end{center}
+%
+These squares say that the relation $c$ is \emph{representable} by the upcast
+morphism $\upc{c}$, essentially that the relation is a kind of \emph{graph} of
+the function in that $x \binrel{\sem c} y$ if and only if $\sem{\upc c}(x)
+\leq_{A_2} y$ \cite{shulman2007}.
 
-We can use this simpler square to derive the original UpL square as long as we can
-horizontally compose with the identity square for $c'$:
+We can use the simpler UpL square to derive the original UpL square by
+horizontally composing with the identity square for $c'$:
 %
@@
 $c'$, because $c'$ is \emph{downward-closed} under the relation on $A_2$.
+An analogous argument shows that we can derive the original UpR square from the simpler version of UpR,
+this time using the fact that poset relations are \emph{upward-closed}.
 % Thus, we
@@
 
-The actual place where transitivity of the relations on the value types is
-needed in this argument occurs when $A_3 = A_2$ and we take $c'$ to be
-$\ltdyn_{A_2}$ (recall that $c'$ is universally quantified, so it can be any
-relation). The fact that $\ltdyn_{A_2}$ is downward-closed is equivalent to it
-being transitive. In other words, the relations $\ltdyn_A$ on value types must be
-poset relations in order to carry out the above compositional construction.
+To recap, we have shown that in order to carry out the graduality proof
+compositionally, we need that the relations are closed under the ordering on
+each side. In particular, this implies that the ordering relations $\ltdyn_{A}$
+must be transitive. Thus, we cannot entirely drop posets from our semantics.
 
-So we see that in order to get a sensible condition that can give us a
-compositional definition of a well-behaved relation, some amount of
-transitive reasoning on squares is required, and so we cannot entirely
-drop posets from our semantics.
+% The actual place where transitivity of the relations on the value types is
+% needed in this argument occurs when $A_3 = A_2$ and we take $c'$ to be
+% $\ltdyn_{A_2}$ (recall that $c'$ is universally quantified, so it can be any
+% relation). The fact that $\ltdyn_{A_2}$ is downward-closed is equivalent to it
+% being transitive. In other words, the relations $\ltdyn_A$ on value types must be
+% poset relations in order to carry out the above compositional construction.
+
+% So we see that in order to get a sensible condition that can give us a
+% compositional definition of a well-behaved relation, some amount of
+% transitive reasoning on squares is required, and so we cannot entirely
+% drop posets from our semantics.
 
@@
 our ordering relation on $\li A$ to get one that is
-transitive. Theorem\ref{thm:no-go} tells us, however, that any non-trivial such
+transitive. Theorem \ref{thm:no-go} tells us, however, that any non-trivial such
 relation must either not be a congruence with respect to $\theta$, or

### 2024-10-22 paper-new/intro.tex
some changes ot the intro and some comments

@@
 \eric{This paragraph could be deleted}
-In order for this to work as expected, gradually-typed languages should allow for
-different parts of the codebase to be in different places along the spectrum from
-dynamic to static, and allow for those different parts to interact with one another.
-Moreover, gradually-typed languages should support the smooth migration from
-dynamic typing to static typing, in that the programmer can initially leave off the
-typing annotations and provide them later without altering the meaning of the program.
-% Sound gradual typing
-Furthermore, the parts of the program that are written in a dynamic
-style should soundly interoperate with the parts that are written in a
-static style.  That is, the interaction between the static and dynamic
-components of the codebase should preserve, to the extent possible,
-the guarantees made by the static types.  In particular, while
-statically-typed code can error at runtime in a gradually-typed
-language, such an error can always be traced back to a
-dynamically-typed term that violated the typing contract imposed by
-statically typed code. Further, static type assertions are sound in
-the static portion, and should enable type-based reasoning and
-optimization.
+%% In order for this to work as expected, gradually-typed languages should allow for
+%% different parts of the codebase to be in different places along the spectrum from
+%% dynamic to static, and allow for those different parts to interact with one another.
+%% Moreover, gradually-typed languages should support the smooth migration from
+%% dynamic typing to static typing, in that the programmer can initially leave off the
+%% typing annotations and provide them later without altering the meaning of the program.
+%% % Sound gradual typing
+%% Furthermore, the parts of the program that are written in a dynamic
+%% style should soundly interoperate with the parts that are written in a
+%% static style.  That is, the interaction between the static and dynamic
+%% components of the codebase should preserve, to the extent possible,
+%% the guarantees made by the static types.  In particular, while
+%% statically-typed code can error at runtime in a gradually-typed
+%% language, such an error can always be traced back to a
+%% dynamically-typed term that violated the typing contract imposed by
+%% statically typed code. Further, static type assertions are sound in
+%% the static portion, and should enable type-based reasoning and
+%% optimization.
 
@@
 %
-Informally, graduality says that going from a dynamic to static style should
-only allow for the introduction of static or dynamic type errors, and not
-otherwise change the behavior of the program.
+Informally, graduality says that migrating code from dynamical to
+statical typing should only allow for the introduction of static or
+dynamic type errors, and not otherwise change the behavior of the
+program.
 %
@@
 should be \emph{adequate}: natural number constants should step to
-themselves. Second, it should validate type based reasoning. To
+themselves $n \Downarrow n$. Second, it should validate type based reasoning. To
 formalize type based reasoning, we give a typed equational theory for
@@
 \emph{reusable}, \emph{compositional} semantic framework for defining
-such well-behaved semantics of programs.
+such well-behaved semantics of gradually typed programs.
 %
@@
 what source language features are present.
-%
-Providing this semantics for gradual typing is inherently complicated
-in that it involves: (1) recursion and recursive types through the
+
+Providing a semantics for gradual typing is inherently complicated in
+that it involves: (1) recursion and recursive types through the
 presence of dynamic types, (2) effects in the form of divergence and
@@
 flavor of domain theory. Effects can be modeled using monads in the
-style of Moggi, or adjunctions in the style of Levy. Relational
-properties and their verification lead naturally to the use of
-reflexive graph categories or double categories.
+style of Moggi, or adjunctions in the style of
+Levy\cite{moggi,levy}. Relational properties and their verification
+lead naturally to the use of reflexive graph categories or double
+categories\cite{reflgraphcats,doublecats}.
 
@@
 %
+\max{we should make clear that we don't yet model these perverse recursive features}
 However, an approach based on classical domain theory has fundamental
@@
 real-world gradually typed systems as well as gradual calculi.
-
+%
 Our long-term goal is to develop a semantics that will scale up to these
@@
 The main denotational alternative to classical domain theory that can
-successfully model these advanced features is \emph{guarded domain
-theory}, which we adopt in this work. While classical domain theory is
+successfully model these advanced features is \emph{guarded} domain
+theory, which we adopt in this work. While classical domain theory is
 based on modeling types as ordered sets with certain joins, guarded
@@
 (ultra)metric spaces but more commonly as ``step-indexed sets'', i.e.,
-objects in the \emph{topos of trees} \cite{birkedal-mogelberg-schwinghammer-stovring2011}. 
-Such an object consists of a family $\{X_n\}_{n \in \mathbb{N}}$ of sets
-along with restriction functions $r_n : X_{n+1} \to X_n$ for all $n$.
-(These are simply presheaves on the poset of natural numbers, for the
-reader familiar with this terminology.)
-We think of such a $\mathbb{N}$-indexed set as modeling a kind of 
-``sequence of approximations'' to the true type being modeled.
+objects in the \emph{topos of trees}
+\cite{birkedal-mogelberg-schwinghammer-stovring2011}.  Such an object
+consists of a family $\{X_n\}_{n \in \mathbb{N}}$ of sets along with
+restriction functions $r_n : X_{n+1} \to X_n$ for all $n$.  (in
+category theoretic terminology, these are presheaves on the poset of
+natural numbers.)  We think a $\mathbb{N}$-indexed set as an infinite
+sequence of increasingly precise approximations to the true type being
+modeled.
 %
-Key to guarded domain theory is that there is a ``later'' operator
-$\triangleright$ on types. Thinking of types as a sequence of
-approximations, the later operator delays the approximation by one
-step. Then the crucial axiom of guarded domain theory is that any
-guarded equation $X \cong F(\triangleright X)$ has a unique
-solution. This allows guarded domain theory to model essentially
-\emph{any} recursive concept, with the caveat that it is guarded by a
-later.
-%
-This caveat is the main source of difficulty in adapting a semantic
-approach based on classical domain theory to guarded domain theory:
-classical domain theory has limitations in what it can model, but it
-provides \emph{exact} solutions to domain equations when it
-applies. When adapting the New-Licata approach to guarded domain
-theory, this means we must work with an \emph{intensional} semantics,
-one where unfolding the dynamic type requires an observable
-computational step: e.g., a function that pattern matches on an
-element of the dynamic type and then returns it unchanged is
-\emph{not} equal to the identity function, because it ``costs'' a step
-to perform the pattern match.
+Key to guarded domain theory is that there is an operator
+$\triangleright$ on step-indexed sets called ``later''. In terms of
+sequences of approximations, the later operator delays the
+approximation by one step. Then the crucial axiom of guarded domain
+theory is that any guarded domain equation $X \cong F(\triangleright
+X)$ has a unique solution. This allows guarded domain theory to model
+essentially \emph{any} recursive concept, with the caveat that the
+recursion is \emph{guarded} by a later.
+
+Since guarded domain theory only provides solutions to guarded domain
+equations, there is no systematic way to convert a classical
+domain-theoretic semantics to a guarded one.  Classical domain theory
+has limitations in what it can model, but it provides \emph{exact}
+solutions to domain equations when it applies. When adapting the
+New-Licata approach to guarded domain theory, the presence of later in
+the semantics makes it \emph{intensional}, unfolding the dynamic type
+requires an observable computational step. For example, a function
+that pattern matches on an element of the dynamic type and then
+returns it unchanged is \emph{not} equal to the identity function,
+because it ``costs'' a step to perform the pattern match.
 %
@@
 considerably simpler to work \emph{synthetically} by working in a
-non-standard foundation such as guarded type theory where the later
-modality is taken as a primitive operation on types, and we take as an
-axiom that guarded domain equations have a (necessarily unique)
-solution. This has the added benefit that providing a denotational
-semantics of gradually typed terms is as simple as defining the semantics of an
-effectful language using a monad on the category of sets. Not only
-does this make on-paper reasoning about guarded domain theory easier,
-it also enables a simpler avenue to verification in a proof
-assistant. Whereas formalizing analytic guarded domain theory would
-require a significant theory of presheaves and making sure that all
-constructions are functors on categories of presheaves, formalizing
-synthetic guarded domain theory can be done by directly adding the
-later modality and the guarded fixed point property axiomatically.
+non-standard foundational system such as \emph{guarded type
+theory}. In guarded type theory later is taken as a primitive
+operation on types, and we take as an axiom that guarded domain
+equations have a (necessarily unique) solution. The benefit of this
+synthetic approach is that when working in the non-standard
+foundation, we don't need to model an object language type as a
+step-indexed set, but instead simply as a set, and object-language
+terms can be modeled in the Kleisli category of a simple monad defined
+using guarded recursion. Not only does this make on-paper reasoning
+about guarded domain theory easier, it also enables a simpler avenue
+to verification in a proof assistant. Whereas formalizing analytic
+guarded domain theory would require a significant theory of presheaves
+and making sure that all constructions are functors on categories of
+presheaves, formalizing synthetic guarded domain theory can be done by
+directly adding the later modality and the guarded fixed point
+property axiomatically.
 %
@@
 
+\max{it's confusing that this is the synthetic section but we switch back to external reasoning. Maybe move this to the previous section and revisit it here?}
 Because the solutions we obtain working in guarded domain theory are constructed
@@
 
+\max{shouldn't we say we use clock quantification?}
 In the setting of synthetic guarded domain theory, this same idea applies. To
@@
 
+\max{this next paragraph is a bit too mysterious}
 This has important ramifications for defining an adequate denotational semantics
@@
 The paper is laid out as follows:
+\max{I think there's too much detail in this outline, especially parts 4 and 5. Can we instead expand those into the Intro?}
 \begin{enumerate}
@@
   typical gradually typed cast calculus.
-\item In Section \ref{sec:concrete-term-model} we develop a denotational
-  semantics in synthetic guarded domain theory for the \emph{terms} of the
-  gradual lambda calculus. The model validates the equational theory, but it
-  does not prove graduality as it does not model type or term precision. We use
-  this to introduce some of our main technical tools: modeling recursive types
-  in guarded type theory and modeling effects using call-by-push-value. After
-  defining the model, we use the technique of globalization to extract a
-  well-behaved big-step term semantics satisfying adequacy.
+\item In Section \ref{sec:concrete-term-model} we develop a
+  denotational semantics in synthetic guarded domain theory for the
+  \emph{terms} of the gradual lambda calculus.  The model is adeqauate
+  and validates the equational theory, but it does not satisfy
+  graduality. We use this to introduce some of our main technical
+  tools: modeling recursive types in guarded type theory and modeling
+  effects using call-by-push-value.
 \item In Section \ref{sec:towards-relational-model} we show where the

### 2024-10-22 paper-new/paper.tex
rename some files, progress on the relational model motivation section
paper revisions
some changes ot the intro and some comments

@@
 \usepackage{array}
+
+\newtheorem{remark}{Remark}
 
@@
   metatheoretic properties of gradually typed languages have become
-  increasingly sophisticated: validity of type based equational
+  increasingly sophisticated: validity of type-based equational
   reasoning as well as the relational property known as
@@
   reusability of denotational semantics. We demonstrate the
-  feasibility of this approach with a model of gradually typed lambda
-  calculus and prove the validity of beta-eta equality and the
-  graduality theorem for the denotational model. This model should
-  provide the basis for a reusable mathematical theory of gradually
-  typed program semantics. Finally, a recent extension to Agda allows
-  synthetic guarded domain theory to be readily mechanized in a proof
-  assistant, and we formalize many of the core theorems of our
-  development.
+  feasibility of this approach with a model of a simple gradually
+  typed lambda calculus and prove the validity of beta-eta equality
+  and the graduality theorem for the denotational model. This model
+  should provide the basis for a reusable mathematical theory of
+  gradually typed program semantics. Finally, we have mechanized most
+  of the core theorems of our development in Guarded Cubical Agda, a
+  recent extension of Agda with support for the guarded recursive
+  constructions we use.
 %%     We develop a denotational semantics for a simple gradually typed language
@@
 
-\input{concrete-relational-model}
+\input{towards-relational-model}
 
@@
 %\input{terms}
-\input{concrete-model}
+\input{concrete-relational-model}
 

### 2024-10-23 paper-new/concrete-relational-model.tex
more paper revisions

@@
 
-\subsection{Constructing the denotational semantics}
+\subsection{Constructing the Denotational Semantics}\label{sec:relational-model-construction}
 
@@
 denotational model:
-\begin{theorem}
-  \,
 \begin{itemize}
@@
   \item For every type precision equivalence $c \equiv c'$ we have a
-    quasi-order-equivalence of relations $\sem{c} \bisim \sem{c'}$.
+    quasi-equivalence of relations $\sem{c} \bisim \sem{c'}$.
   \item For every term precision $\Delta \vdash M \ltdyn M' : c$ we
@@
 \end{itemize}
-\end{theorem}
 
@@
 It may seem as though we need to define the perturbations for $D$ by
-guarded recursion, but in fact we define them via least-fixpoint in the category
-of monoids:
+guarded recursion, but in fact we define them via least-fixpoint\footnote{
+  The reason for defining the perturbations inductively is technical and has to
+  do with the fact that we needed to define the perturbations corresponding to the
+  predomain $\later A$ to be the perturbations of $A$, \emph{not} later of the
+  perturbations of $A$. There may be a way to make this work by changing some
+  definitions, but we leave this as future work}
+in the category of monoids:
 %
@@
 We now explain how to interpret these perturbations as endomorphisms.
-
 We define $i_D : M_D \to \{ f : D \to D \mid f \bisim \id \}$ by induction as follows:
@@
 the push-pull property for perturbations. We include most of these technical
-verifications in the appendix (Definition
-\ref{def:value-computation-rel-comp},
+verifications in the appendix (Definitions
+\ref{def:value-computation-rel-comp} and
 \ref{def:functorial-actions-on-relations}).
@@
 For quasi-right-representability, the only nontrivial case is $\li(\iarr)$.
-Recall the defintion of the downcast for $\iarr$ given in Section
+Recall the definition of the downcast for $\iarr$ given in Section
 \ref{sec:term-interpretation}. We know that in order for the DnR and DnL squares
@@
 
-The proofs that the precision equivalence imply quasi-equivalence of
-relations are in the appendix (Lemma
-\ref{lem:quasi-order-equiv-functors}).
+Lastly, the interpretation of the type precision equivalence rules $c \equiv c'$
+as quasi-equivalence of the denoted relations $\sem{c} \bisim \sem{c'}$ is
+justified by Lemma \ref{lem:quasi-order-equiv-functors} in the Appendix.
+
+% The type precision equivalence rules follow because $c \equiv c'$
+% implies quasi-equivalence, and that quasi-equivalence implies that
+% $\id \ltsqbisim{\sem{c}}{\sem{c'}} \id$.
 
@@
 Finally, we establish the core result of the semantics: interpreting
-term precision as extensional squares. Fortunately the hard work is
-now complete and the term precision rules follow from our
+term precision as extensional squares. Fortunately, the hard work is
+now complete, and the term precision rules follow from our
 compositional constructions.
 %
-The congruence rules of term precision are modeled easily: congruence
-squares hold both for bisimilarity as well as error ordering, and so
-by combination we also have these for extensional squares.
+In particular, the congruence rules of term precision are modeled easily: congruence
+squares exist both for bisimilarity and for the error ordering, and so
+we also have these for extensional squares.
 %
-The type precision equivalence rule follows because $c \equiv c'$
-implies quasi-equivalence, and that quasi-equivalence implies that
-$\id \ltsqbisim{\sem{c}}{\sem{c'}} \id$.
-
-Most crucially, we verify the existence of the squares for the cast rules. For UpL, the
+Thus, it remains to verify the existence of the squares for the cast rules. For UpL, the
 relevant extensional square is obtained as follows:

### 2024-10-23 paper-new/discussion.tex
more paper revisions

@@
 introduces. 
+%
+The line of work by New, Licata, and Ahmed involved both denotational methods
+and operational logical relations approaches.
+% The New-Licata-Ahmed axiomatic approach to gradual typing can be
+% applied to both operational and denotational semantics.
 
@@
 % \subsubsection{Denotational and Operational Approaches}
-
-The above axiomatic account of gradual typing is independent of the choice of
-operational versus denotational semantics; we give an overview of the two
-approaches below.
 
@@
 Licata, but we work with guarded type theory (a synthetic form of step-indexing)
-rather than classical domain theory. In the guarded setting, the compositional
-theory of New-Licata breaks down: when accounting for steps, the casts are no
-longer embedding-projection pairs, and we cannot carry out compositional
-reasoning on terms when those terms differ by an arbitrary unknown number of
-steps. Our work provides a novel refinement to their theory that allows for some
+rather than classical domain theory. As we have seen, in the guarded setting the compositional
+theory of New-Licata breaks down. 
+% when accounting for steps, the casts are no
+% longer embedding-projection pairs, and we cannot carry out compositional
+% reasoning on terms when those terms differ by an arbitrary unknown number of
+% steps. 
+Our work provides a novel refinement to their theory that allows for some
 compositional reasoning to be recovered even in the guarded setting.
@@
 \subsection{Implications for Gradual Typing Research}
-
-While we have focused on a single semantics in this work, we expect that our
-development can be adapted to prove graduality and (more limited) equational
-reasoning principles for other cast semantics.
-%
-In addition, while we have focused on proving extensional properties in this
-work, the intensional nature of our semantics can be viewed as a form of
-\emph{cost semantics}, where a runtime cost is incurred from inspecting the
-dynamic type. This could be used to verify that cast optimizations such as space
-efficient implementations \cite{herman-tomb-flanagan-2010} are not only
-extensionally correct, but also have lower quantitative cost. We leave this as
-future work.
 
@@
     type precision derivations (see the bottom of Figure \ref{fig:gtlc-syntax}).
+    These are also included in the Appendix.
     \item Defining the syntax-to-semantics translation.
@@
 
-% \max{Expand on this, e.g. what some of the challenges might be and what would be reusable}
-In the future, we plan to apply our approach to give a denotational semantics
-to gradually-typed languages that feature higher-order state or runtime-extensible dynamic
-typing \cite{DBLP:journals/corr/abs-2210-02169} as well as richer type
-disciplines such as gradual dependent types and effect systems. This will
-involve adapting prior operational models based on step-indexing (e.g.,
+Our immediate next step is to apply our approach to give a denotational
+semantics to gradually-typed languages with advanced type systems including
+higher-order state or runtime-extensible dynamic typing
+\cite{DBLP:journals/corr/abs-2210-02169}, as well as richer type disciplines
+such as gradual dependent types and effect systems. This will involve adapting
+prior operational models based on step-indexing (e.g.,
 \cite{new-giovannini-licata-2022}) to the denotational setting and using guarded
@@
 extended with additional cases to model effects besides error and stepping.
-Additionally, we aim to complete our Agda formalization and evolve it into a
+We also aim to complete our Agda formalization and evolve it into a
 reusable framework for mechanized denotational semantics of gradually-typed
 languages.
+
+Additionally, while we have focused on a single semantics in this work, there are a
+variety of other cast calculi that satisfy graduality but for which one or more
+equational axioms fail to hold. % TODO cite
+We expect that our development can be adapted to other cast semantics to prove
+graduality and equational reasoning principles (to the extent that the
+principles hold in those cast semantics).
+
+We also plan to use our framework to investigate the performance of
+gradually-typed languages. Although we have focused on proving extensional
+properties in this work, the intensional nature of our semantics can be viewed
+as a form of \emph{cost semantics}, where a runtime cost is incurred from
+inspecting the dynamic type. This could be used to verify that cast
+optimizations such as space efficient implementations
+\cite{herman-tomb-flanagan-2010} are not only extensionally correct, but also
+have lower quantitative cost. We leave this as future work.
+
+% \max{Expand on this, e.g. what some of the challenges might be and what would be reusable}
+% In the future, we plan to apply our approach to give a denotational semantics
+% to gradually-typed languages that feature higher-order state or runtime-extensible dynamic
+% typing \cite{DBLP:journals/corr/abs-2210-02169} as well as richer type
+% disciplines such as gradual dependent types and effect systems. This will
+% involve adapting prior operational models based on step-indexing (e.g.,
+% \cite{new-giovannini-licata-2022}) to the denotational setting and using guarded
+% type theory to obtain solutions to guarded domain equations as we did in this
+% work for the denotation of the dynamic type. The generality of our techniques
+% should allow for many of the constructions to be reused across different
+% languages. For instance, the free error domain construction can be easily
+% extended with additional cases to model effects besides error and stepping.
+% Additionally, we aim to complete our Agda formalization and evolve it into a
+% reusable framework for mechanized denotational semantics of gradually-typed
+% languages.
 

### 2024-10-23 paper-new/intro.tex
more paper revisions

@@
 
-\eric{This paragraph could be deleted}
+% \eric{This paragraph could be deleted}
 %% In order for this to work as expected, gradually-typed languages should allow for
@@
 %
-Informally, graduality says that migrating code from dynamical to
-statical typing should only allow for the introduction of static or
+Informally, graduality says that migrating code from dynamic to
+static typing should only allow for the introduction of static or
 dynamic type errors, and not otherwise change the behavior of the
@@
 
-\eric{One of the reviewers said this should be moved later in the intro, but I think it makes sense here.}
-More concretely, consider a gradually typed language whose only
-effects are gradual type errors and divergence. Then if we fix a
-result type of natural numbers, a whole program semantics is a partial
-function from closed programs to either natural numbers or errors:
-\[ -\Downarrow : \{M \,|\, \cdot \vdash M : \nat \} \rightharpoonup \mathbb{N} \cup \{\mho\} \]
-where $\mho$ is notation for a runtime type error. We write $M
-\Downarrow n$ and $M\Downarrow \mho$ to mean this semantics is defined
-as a number or error, and $M\Uparrow$ to mean the semantics is
-undefined, representing divergence.
-%
-A well-behaved semantics should then satisfy several properties. First, it
-should be \emph{adequate}: natural number constants should step to
-themselves $n \Downarrow n$. Second, it should validate type based reasoning. To
-formalize type based reasoning, we give a typed equational theory for
-terms of the language $M \equiv N$ for when two terms should be
-considered equivalent. Then we want to verify that the big step
-semantics respects this equational theory: if closed programs $M \equiv
-N$ are equivalent in the equational theory then they have the same
-semantics, $M \Downarrow n \iff N \Downarrow n$ and $M\Uparrow \iff N
-\Uparrow$ and $M \Downarrow \mho \iff N \Downarrow \mho$.
-%
-Lastly, the graduality property is defined by giving an
-\emph{inequational} theory called term precision, where $M \ltdyn N$
-roughly means that $M$ and $N$ have the same type erasure and $M$ has
-at each point in the program a more precise/static type than $N$.
-%
-Then, the graduality property states that if $M \ltdyn N$ are whole
-programs then $M$ must either have the same behavior as $N$ or error:
-Either $M\Downarrow \mho$ or $M \Downarrow n $ and $N \Downarrow n$ or
-$M \Uparrow $ and $N \Uparrow$\footnote{we use a slightly more complex
-definition of this relation in our technical development below that is
-classically equivalent but constructively weaker}.
+% moved the remaining paragraphs to a new section later in the intro
 
@@
 category theoretic terminology, these are presheaves on the poset of
-natural numbers.)  We think a $\mathbb{N}$-indexed set as an infinite
+natural numbers.)  We think of a $\mathbb{N}$-indexed set as an infinite
 sequence of increasingly precise approximations to the true type being
@@
 essentially \emph{any} recursive concept, with the caveat that the
-recursion is \emph{guarded} by a later.
+recursion is \emph{guarded} by a use of the later operator.
 
-Since guarded domain theory only provides solutions to guarded domain
-equations, there is no systematic way to convert a classical
-domain-theoretic semantics to a guarded one.  Classical domain theory
-has limitations in what it can model, but it provides \emph{exact}
-solutions to domain equations when it applies. When adapting the
-New-Licata approach to guarded domain theory, the presence of later in
-the semantics makes it \emph{intensional}, unfolding the dynamic type
-requires an observable computational step. For example, a function
-that pattern matches on an element of the dynamic type and then
-returns it unchanged is \emph{not} equal to the identity function,
-because it ``costs'' a step to perform the pattern match.
-%
-This leads to the main departure of our semantics from the New-Licata
-approach: while the New-Licata semantics equips $\omega$-CPOs with an
-error ordering, we work with types equipped with not just an error
-ordering but also a \emph{bisimilarity} relation that reflects when
-elements differ only in the number of abstract computational steps
-that they take.
-%
-% ERIC: This could be merged into the summary at the end of the section
-The reason for requiring these two separate notions is that in the setting of
-guarded domain theory, it is not possible to carry out transitive reasoning
-while being oblivious to the number of steps the elements take. We make this
-statement precise in \ref{sec:towards-relational-model}. To overcome this
-limitation, we develop further changes to the core concepts of our semantics as
-discussed in that section.
+%\max{it's confusing that this is the synthetic section but we switch back to external reasoning. Maybe move this to the previous section and revisit it here?}
+Because the solutions we obtain working in guarded domain theory are constructed
+as a sequence of increasingly better approximations, when we want to establish a
+property that holds in the limit, we must reason about all finite
+approximations. In the setting of the topos of trees, this manifests as the need
+to consider \emph{global elements} of the presheaves, i.e., a family of elements
+$x_i \in X_i$ compatible with the restriction maps. For example, consider the
+set of programs that may take a computational step or return unit.\footnote{This example is adapted from \cite{mogelberg-paviotti2016}.} In the
+topos of trees, we can model this set as the indexed family of sets $\{X_n\}_{n \in \mathbb{N}}$
+where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1, \bot\}$. Here, $i$ denotes a program that steps
+$i$ times and then returns, and $\bot$ denotes a program that fails to terminate
+in $n-1$ steps or fewer. For any fixed $n$, the set $X_n$ fails to distinguish a
+program that takes $n$ steps and then terminates from a program that never
+terminates. On the other hand, if we instead consider the denotation of a
+program to be a global element of the presheaf $\{X_n\}_{n \in \mathbb{N}}$ -- a family of elements
+$x_i \in X_i$ for all $i \in \mathbb{N}$ -- then the denotation of the diverging
+program will be distinct from that of a program that terminates, regardless of
+the number of steps it takes.
 
-% Because bisimilarity, unlike equality, is not transitive in the guarded
-% setting, this necessitates further changes to the core
-% concepts of our semantics. These challenges are discussed in Section \ref{sec:towards-relational-model}.
+% \max{this next paragraph is a bit too mysterious}
+% This has important ramifications for defining an adequate denotational semantics
+% in guarded domain theory, as we seek to do in this paper: we do not want our
+% denotational semantics to conflate a diverging program with a program that fails
+% to terminate in sufficiently few steps. Thus, establishing adequacy of a
+% semantics in guarded domain theory will require a means of globalizing,
+% a point we will return to in Section \ref{sec:big-step-term-semantics}.
 
@@
 % \max{TODO: more here, specifically this is where we should talk about adequacy I think}
+%
+In this paper, we work informally in a guarded type theory which is described in
+more detail in Section \ref{sec:guarded-type-theory}.
+%
+As in the setting of analytic guarded domain theory, in the synthetic setting it
+is also possible to construct global solutions. The specific approach we use in
+this work is that of \emph{clocks} and \emph{clock-quantification}
+\cite{atkey-mcbride2013}.
 
-\max{it's confusing that this is the synthetic section but we switch back to external reasoning. Maybe move this to the previous section and revisit it here?}
-Because the solutions we obtain working in guarded domain theory are constructed
-as a sequence of increasingly better approximations, when we want to establish a
-property that holds in the limit, we must reason about all finite
-approximations. In the setting of the topos of trees, this manifests as the need
-to consider \emph{global elements} of the presheaves, i.e., a family of elements
-$x_i \in X_i$ compatible with the restriction maps. For example, consider the
-set of programs that may take a computational step or return unit.\footnote{This example is adapted from \cite{mogelberg-paviotti2016}.} In the
-topos of trees, we can model this set as the indexed family of sets $\{X_n\}_{n \in \mathbb{N}}$
-where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1, \bot\}$. Here, $i$ denotes a program that steps
-$i$ times and then returns, and $\bot$ denotes a program that fails to terminate
-in $n-1$ steps or fewer. For any fixed $n$, the set $X_n$ fails to distinguish a
-program that takes $n$ steps and then terminates from a program that never
-terminates. On the other hand, if we instead consider the denotation of a
-program to be a global element of the presheaf $\{X_n\}_{n \in \mathbb{N}}$ -- a family of elements
-$x_i \in X_i$ for all $i \in \mathbb{N}$ -- then the denotation of the diverging
-program will be distinct from that of a program that terminates, regardless of
-the number of steps it takes.
+% \max{shouldn't we say we use clock quantification?}
+% In the setting of synthetic guarded domain theory, there is an analogue of
+% constructing global solutions. To construct a global solution we must use an
+% additional construct, e.g., the $\square$ modality
+% \cite{10.1007/978-3-319-08918-8_8}, whose semantics in the topos of trees model
+% is to compute the global elements of a presheaf. A related approach involves
+% objects known as \emph{clocks}, whereby \emph{clock-quantification}
+% \cite{atkey-mcbride2013} provides a means to obtain a global solution.
 
-\max{shouldn't we say we use clock quantification?}
-In the setting of synthetic guarded domain theory, this same idea applies. To
-construct a global solution we must use an additional construct, e.g., the
-$\square$ modality \cite{10.1007/978-3-319-08918-8_8}, whose semantics in the
-topos of trees model is to compute the global elements of a presheaf. A related
-approach involves objects known as \emph{clocks}, whereby
-\emph{clock-quantification} \cite{atkey-mcbride2013} provides a means to obtain
-a global solution.
+\subsection{Adapting the New-Licata Model to the Guarded Setting}
 
-\max{this next paragraph is a bit too mysterious}
-This has important ramifications for defining an adequate denotational semantics
-in guarded domain theory, as we seek to do in this paper: we do not want our
-denotational semantics to conflate a diverging program with a program that fails
-to terminate in sufficiently few steps. Thus, establishing adequacy of a
-semantics in guarded domain theory will require a means of globalizing,
-a point we will return to in Section \ref{sec:big-step-term-semantics}.
+Since guarded domain theory only provides solutions to guarded domain equations,
+there is no systematic way to convert a classical domain-theoretic semantics to
+a guarded one.  Classical domain theory has limitations in what it can model,
+but it provides \emph{exact} solutions to domain equations when it applies. When
+adapting the New-Licata approach to guarded domain theory, the presence of later
+in the semantics makes it \emph{intensional}: unfolding the dynamic type
+requires an observable computational step. For example, a function that pattern
+matches on an element of the dynamic type and then returns it unchanged is
+\emph{not} equal to the identity function, because it ``costs'' a step to
+perform the pattern match.
+
+% - The essential role of transitivity
+% - We can't have transitivity + extensional reasoning
+% - Solution: split the ordering into two, and introduce perturbations
+%   to recover some amount of transitive reasoning
+
+The intensional nature of guarded domain theory leads to the main departure of
+our semantics from the New-Licata approach. New and Licata observed that
+in order to even \emph{formulate} the model for gradual typing in a
+syntax-independent manner, some access to transitive reasoning is essential.
+%
+However, in the setting of guarded domain theory, it is not possible to carry
+out reasoning that is simultaneously transitive and \emph{extensional}, where
+the latter means being insensitive to the number of steps the elements take. We
+make this statement precise in a no-go theorem (Theorem \ref{thm:no-go}). This fundamental
+limitation means that the natural adaptation of the New-Licata semantics to the
+guarded setting, where types are interpreted as sets with a
+\emph{step-insensitive} error ordering, cannot work, because that ordering is
+necessarily \emph{not} transitive.
+
+To recover some amount of compositional reasoning, we model the graduality
+ordering using a combination of two relations: a \emph{lock-step error ordering}
+in which elements must take the same number of steps, and a \emph{weak
+bisimilarity relation} that relates elements differing only in the number of
+computational steps that they take. The lock-step ordering \emph{is} transitive,
+but it is not sufficient to model the New-Licata proof. The problem is that in
+the guarded setting, the denotation of a downcast from the dynamic type involves
+a computational step, and by the functorial nature of casts this implies that
+the denotations of arbitrary casts may involve computational steps. As a result,
+the embedding-projection characterization of casts in the New-Licata model does
+not hold in the guarded setting, at least not up to equality.
+%
+However, while the casts may involve delays, they are otherwise extensionally
+equivalent to embedding-projection pairs. Thus, one might try to use weak
+bisimilarity to account for the delays, viewing casts as embedding-projection
+pairs ``up to weak bisimilarity''. Unfortunately, this does not work: since weak
+bisimilarity is \emph{not} transitive (which follows also from our no-go
+result), we are back to the original issue of lack of transitive reasoning.
+
+% However, since the casts involve delays but are otherwise extensionally
+% equivalent to embedding-projection pairs, one might try to use weak bisimilarity
+% to account for the delays. That is, one could try characterizing the casts as
+% embedding-projection pairs ``up to weak bisimilarity''. Unfortunately, this does
+% not work: since weak bisimilarity is \emph{not} transitive (which follows also
+% from our no-go result), we are back to the original issue of lack of
+% transitive reasoning.
+
+It would seem that at this point, we are stuck: we need compositional reasoning,
+but we also need to be able to account for the steps taken by casts. And now we
+come to the key technical contribution of this work. The crucial observation is
+that, although the denotations of casts involve delays, the specific ways in
+which those delays arise follow a specific, type-directed pattern. Hence, we can
+track the delay behavior in a type-directed manner, equipping each type with a
+structured set of what we call \emph{syntactic perturbations}. These can then be
+used to explicitly \emph{synchronize} elements to ensure that they are in lock-step.
+%
+In the end, we have come to a resolution of the issue: the lock-step error
+ordering where we employ syntactic perturbations to synchronize elements is
+still transitive, while simultaneously being able to relate terms that involve
+an explicit, known pattern of computational steps.
+%
+We discuss these issues in more detail in Section
+\ref{sec:towards-relational-model} and give a detailed description of the
+resulting model in Section \ref{sec:concrete-relational-model}.
+
+% Perturbations allow us to recover enough extensional reasoning to model the
+% graduality property compositionally.
+
+% weak bisimilarity: elements differing only in their stepping behavior are related.
+
+% while the New-Licata semantics
+% equips $\omega$-CPOs with an error ordering, we work with types equipped with
+% not just an error ordering but also a \emph{bisimilarity} relation that reflects
+% when elements differ only in the number of abstract computational steps that
+% they take. The reason for requiring these two separate notions is that in the
+% setting of guarded domain theory, it is not possible to carry out transitive
+% reasoning while being oblivious to the number of steps the elements take. We
+% make this statement precise in \ref{sec:towards-relational-model}. To overcome
+% this limitation, we develop further changes to the core concepts of our
+% semantics as discussed in that section.
+
+% Because bisimilarity, unlike equality, is not transitive in the guarded
+% setting, this necessitates further changes to the core
+% concepts of our semantics. These challenges are discussed in Section \ref{sec:towards-relational-model}.
+
+\subsection{The Graduality Proof}
+
+Proving graduality by constructing a denotational model in guarded type theory
+consists of two main steps. First, there is the construction of the model
+itself. This gives us an interpretation of the terms as well as the syntactic
+axioms that model the graduality property. However, because this model is
+defined in the guarded domain theory, while graduality is a statement about a
+set-theoretic semantics, it is not a priori clear that the guarded model
+validates the graduality property.
+%
+What we need is to show that the guarded model induces a sensible set-theoretic
+semantics in ordinary mathematics. We cannot hope to derive such a semantics for
+\emph{all} types (e.g., the dynamic type), but for the subset of closed terms of
+base type, we should be able to extract a well-behaved semantics for which the
+graduality property holds.
+
+% The next step is to ensure that the semantics actually validates the graduality
+% property. That is, we can extract from this guarded model a well-behaved
+% semantics in ordinary mathematics For the terms, this means that closed terms of
+% base type should evaluate to themselves.
+
+% Defining a model of gradual typing in guarded type theory is not enough to prove
+% graduality. We must then extract a well-behaved ``big-step'' semantics for terms
+% of base type and show that the denotation of term precision has the expected
+% behavior. More specifically, we want a semantics in which the denotation of
+% error is different from the denotation of a value, which is in turn different
+% from the denotation of the diverging computation. Recall that by working in
+% guarded domain theory, our definitions are constructed as a sequence of
+% approximations. What we actually want is a statement that holds ``in the
+% limit''.
+
+More concretely, consider a gradually typed language whose only effects are
+gradual type errors and divergence. If we fix a result type of natural numbers,
+a whole program semantics is a partial function from closed programs to either
+natural numbers or errors:
+%
+\[ -\Downarrow : \{M \,|\, \cdot \vdash M : \nat \} \rightharpoonup \mathbb{N}
+\cup \{\mho\} \] 
+%
+where $\mho$ is notation for a runtime type error. We write $M \Downarrow n$ and
+$M\Downarrow \mho$ to mean this semantics is defined as a number or error, and
+$M\Uparrow$ to mean the semantics is undefined, representing divergence.
+%
+A well-behaved semantics should then satisfy several properties. First, it
+should be \emph{adequate}: natural number constants should evaluate to
+themselves: $n \Downarrow n$. Second, it should validate type based reasoning.
+To formalize type based reasoning, we give a typed equational theory for terms
+of the language $M \equiv N$ for when two terms should be considered equivalent.
+Then we want to verify that the big step semantics respects this equational
+theory: if closed programs $M \equiv N$ are equivalent in the equational theory
+then they have the same semantics, $M \Downarrow n \iff N \Downarrow n$ and
+$M\Uparrow \iff N \Uparrow$ and $M \Downarrow \mho \iff N \Downarrow \mho$.
+
+Lastly, the semantics should be adequate for the graduality property. Graduality
+is typically axiomatized by giving an \emph{inequational} theory called term
+precision, where $M \ltdyn N$ roughly means that $M$ and $N$ have the same type
+erasure and $M$ has at each point in the program a more precise/static type than
+$N$. Then adequacy says that if $M$ and $N$ are whole programs returning type
+$\nat$ and $M \ltdyn N$, then the denotations of $M$ and $N$ as big-step terms
+should be related in the expected way: either $M$ errors, or $M$ and $N$ have
+the same extensional behavior. That is, either $M\Downarrow \mho$ or $M
+\Downarrow n $ and $N \Downarrow n$ or $M \Uparrow $ and $N
+\Uparrow$\footnote{we use a slightly more complex definition of this relation in
+our technical development below that is classically equivalent but
+constructively weaker}.
+
+To define the big-step semantics, we apply clock-quantification to the guarded
+model to define a global semantics for closed terms of base type where a term
+$M$ is mapped to an element of the coinductive delay monad of Capretta
+\cite{lmcs:2265}. We compose this map with a partial function on elements of the
+delay monad that returns the result with which given element terminates, or is
+undefined if it does not terminate. This gives us the desired partial function
+from closed programs to results. We describe this process in more detail in Section
+\ref{sec:big-step-term-semantics}.
+%
+We then make the analogous construction for the relation that denotes the term
+precision ordering, showing that a term precision ordering $M \ltdyn N$ implies
+the corresponding ordering on the partial functions denoted by $M$ and $N$. This
+construction uses as an intermediate step an analogue of the graduality relation
+for the coinductive delay monad. The details of the proof are given in Section
+\ref{sec:adequacy}.
 
@@
 
-The main contribution of this work is a compositional denotational
-semantics for gradually typed languages that validates $\beta\eta$
-equality and satisfies a graduality theorem. A great deal of the work
-has further been verified in Guarded Cubical Agda \cite{veltri-vezzosi2020}, 
-demonstrating that the semantics is readily mechanizable.
-We provide an overview of the mechanization effort, and what remains to be formalized,
-in Section \ref{sec:mechanization}.
+The main contribution of this work is a compositional denotational semantics in
+guarded type theory for a simple gradually typed language that validates
+$\beta\eta$ equality and satisfies a graduality theorem.
+%
+% In particular, the notion of syntactic perturbations stands out as a key
+% technical contribution and is the most significant and novel feature of our
+% model that allowed us to successfully adapt the denotational approach of New and
+% Licata to the guarded setting.
+% In particular, the notion of syntactic perturbations stands out as a key feature
+% of our model, and is our most significant technical contribution allowing us to
+% successfully adapt the denotational approach of New and Licata to the guarded
+% setting.
+The goal we have kept in mind while designing our model is to handle
+gradually-typed languages with advanced features that classical domain theory in
+incapable of modeling. We do not address these features in this work but leave
+them instead to immediate future work, as handling even the simplest
+gradually-typed language already poses interesting challenges.
+%
+The notion of syntactic perturbation stands out as a key aspect of our model
+and our most significant technical contribution, allowing us to successfully
+adapt the denotational approach of New and Licata to the guarded setting.
+%
+Most of our work has further been verified in Guarded Cubical Agda
+\cite{veltri-vezzosi2020}, demonstrating that the semantics is readily
+mechanizable. We provide an overview of the mechanization effort, and what
+remains to be formalized, in Section \ref{sec:mechanization}.
+%
+
+% Syntactic perturbations 
+%
+% The notion of \emph{syntactic perturbations} allow us to encode the steps
+% taken by a term in a form that we can manipulate. This allows us to recover
+% enough extensional reasoning to model the graduality property compositionally.
+
+% Syntactic perturbations give us a type-directed way of imposing delays on terms.
 
@@
 The paper is laid out as follows:
-\max{I think there's too much detail in this outline, especially parts 4 and 5. Can we instead expand those into the Intro?}
+% \max{I think there's too much detail in this outline, especially parts 4 and 5. Can we instead expand those into the Intro?}
 \begin{enumerate}
@@
   effects using call-by-push-value.
-\item In Section \ref{sec:towards-relational-model} we show where the
-  New-Licata classical domain theoretic approach fails to adapt
-  cleanly to the guarded setting and explore the difficulties of
-  proving graduality in an intensional model.
-  %
-  % ... including the reason why we need to introduce weak bisimilarity.
-  %
-  The key technical issue is that reasoning principles that are simultaneously
-  extensional \emph{and} compositional are not compatible with guarded domain
-  theory (here, extensional means relating elements that differ only in the
-  number of computational steps they take). We make this statement precise with
-  a no-go theorem (Theorem \ref{thm:no-go}). In light of this result, to recover
-  some amount of compositional reasoning we model the graduality ordering using
-  a combination of two relations: a lock-step error ordering in which elements
-  must take the same number of steps, and a weak bisimilarity relation where
-  elements differing only in their stepping behavior are related.
-
-  % As a result, we split the
-  % graduality ordering into two relations: a strong error ordering in which
-  % elements must take the same number of steps, and a weak bisimilarity relation
-  % where elements differing only in their stepping behavior are related.
-
+\item In Section \ref{sec:towards-relational-model} we show where the New-Licata
+  classical domain theoretic approach fails to adapt cleanly to the guarded
+  setting and explore the difficulties of proving graduality in an intensional
+  model. We prove the no-go theorem about extensional, transitive relations,
+  introduce the lock-step error ordering and weak bisimilarity relation, and
+  motivate the need for perturbations.
 \item In Section \ref{sec:concrete-relational-model} we describe the
-  construction of the model in detail. Key here is a notion of \emph{syntactic
-  perturbations}, which allow us to encode the steps taken by a term in a form
-  that we can manipulate. This allows us to recover enough extensional reasoning
-  to model the graduality property compositionally.
-  % Key here is a notion of \emph{syntactic
-  % perturbation}, which give us a type-directed way of imposing delays on terms.
-  The resulting model validates the axioms for type and term precision specified
-  in Section \ref{sec:GTLC}. The final step is to prove that the model is
-  \emph{adequate} for the graduality property: a closed term precision $M \ltdyn
-  N : \nat$ has the expected semantics, i.e., that $M$ errors or $M$ and $N$
-  have the same extensional behavior. We do so by extending the globalization
-  techniques used in defining the big-step term model to account for the
-  interpretation of term precision.
+  construction of the model in detail, and discuss the
+  translation of the syntax and axioms of the gradually typed cast calculus into the model.
+  Lastly, we prove that the model is adequate for the graduality property.
+  % The resulting model validates the axioms for type and term precision specified
+  % in Section \ref{sec:GTLC}. The final step is to prove that the model is
+  % \emph{adequate} for the graduality property: a closed term precision $M \ltdyn
+  % N : \nat$ has the expected semantics, i.e., that $M$ errors or $M$ and $N$
+  % have the same extensional behavior. We do so by extending the globalization
+  % techniques used in defining the big-step term model to account for the
+  % interpretation of term precision.
 \item In Section \ref{sec:discussion} we discuss prior work on proving

### 2024-10-24 paper-new/intro.tex
revise intro

@@
 \max{we should make clear that we don't yet model these perverse recursive features}
+\eric{In the contributions section, I explain that we don't yet model these features}
 However, an approach based on classical domain theory has fundamental
@@
 
-%\max{it's confusing that this is the synthetic section but we switch back to external reasoning. Maybe move this to the previous section and revisit it here?}
-Because the solutions we obtain working in guarded domain theory are constructed
-as a sequence of increasingly better approximations, when we want to establish a
-property that holds in the limit, we must reason about all finite
-approximations. In the setting of the topos of trees, this manifests as the need
-to consider \emph{global elements} of the presheaves, i.e., a family of elements
-$x_i \in X_i$ compatible with the restriction maps. For example, consider the
-set of programs that may take a computational step or return unit.\footnote{This example is adapted from \cite{mogelberg-paviotti2016}.} In the
-topos of trees, we can model this set as the indexed family of sets $\{X_n\}_{n \in \mathbb{N}}$
-where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1, \bot\}$. Here, $i$ denotes a program that steps
-$i$ times and then returns, and $\bot$ denotes a program that fails to terminate
-in $n-1$ steps or fewer. For any fixed $n$, the set $X_n$ fails to distinguish a
-program that takes $n$ steps and then terminates from a program that never
-terminates. On the other hand, if we instead consider the denotation of a
-program to be a global element of the presheaf $\{X_n\}_{n \in \mathbb{N}}$ -- a family of elements
-$x_i \in X_i$ for all $i \in \mathbb{N}$ -- then the denotation of the diverging
+% The theorems we want are about the whole sequence of approximations
+% whereas working with presheaves is about each finite approximation
+% The statement of graduality is in terms of global elements
+\eric{This next paragraph could be moved to the section on adequacy, but since
+the example is in the analytic setting it also makes sense to keep it here. I'm
+not sure which option makes more sense.} While the definitions in guarded domain
+theory are constructed as sequences of approximations, results about semantics
+of programs, e.g., graduality, are actually statements concerning entire
+\emph{sequences} of approximations. As a simple example, consider the set of
+programs that may take a computational step or return unit.\footnote{This
+example is adapted from \cite{mogelberg-paviotti2016}.} In the topos of trees,
+we can model the set of these programs as the indexed family of sets $\{X_n\}_{n
+\in \mathbb{N}}$ where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1,
+\bot\}$. Here, $i$ denotes a program that steps $i$ times and then returns, and
+$\bot$ denotes a program that fails to terminate in $n-1$ steps or fewer. For
+any fixed $n$, the set $X_n$ fails to distinguish a program that takes $n$ steps
+and then terminates from a program that never terminates. On the other hand, if
+we define the denotation of a program to be a \emph{sequence} of elements $x_i
+\in X_i$ for all $i \in \mathbb{N}$, then the denotation of the diverging
 program will be distinct from that of a program that terminates, regardless of
 the number of steps it takes.
+
+% Because the solutions we obtain working in guarded domain theory are constructed
+% as a sequence of increasingly better approximations, when we want to establish a
+% property that holds in the limit, we must reason about all finite
+% approximations. In the setting of the topos of trees, this manifests as the need
+% to consider \emph{global elements} of the presheaves, i.e., a family of elements
+% $x_i \in X_i$ compatible with the restriction maps.
 
@@
 %
-As in the setting of analytic guarded domain theory, in the synthetic setting it
-is also possible to construct global solutions. The specific approach we use in
-this work is that of \emph{clocks} and \emph{clock-quantification}
-\cite{atkey-mcbride2013}.
+% As in the setting of analytic guarded domain theory, in the synthetic setting it
+% is also possible to construct global solutions. The specific approach we use in
+% this work is that of \emph{clocks} and \emph{clock-quantification}
+% \cite{atkey-mcbride2013}.
 
@@
 \subsection{Adapting the New-Licata Model to the Guarded Setting}
+
+% Working in a step-counting model, but graduality is independent of steps
+% Casts take stpes, so we need to reason up to weak bisimilarity, but it is not transitive
+% New-Licata freely uses transitivity, but we can't adapt their proof methodology because it uses transitivity pervasively
+% Thus the old proofs from New-Licata that use transitivity don't apply.
+% Work with a combination of weak bisimilarity and step-counting reasoning
+% Need a tiny bit of weak bisimilarity which are perturbations, explicit synchronizations that we manipulate syntactically
 
@@
 The intensional nature of guarded domain theory leads to the main departure of
-our semantics from the New-Licata approach. New and Licata observed that
-in order to even \emph{formulate} the model for gradual typing in a
-syntax-independent manner, some access to transitive reasoning is essential.
-%
-However, in the setting of guarded domain theory, it is not possible to carry
-out reasoning that is simultaneously transitive and \emph{extensional}, where
-the latter means being insensitive to the number of steps the elements take. We
-make this statement precise in a no-go theorem (Theorem \ref{thm:no-go}). This fundamental
-limitation means that the natural adaptation of the New-Licata semantics to the
-guarded setting, where types are interpreted as sets with a
-\emph{step-insensitive} error ordering, cannot work, because that ordering is
-necessarily \emph{not} transitive.
+our semantics from the New-Licata approach. Because casts involve computational
+steps, the graduality property must be insensitive to the steps taken by terms.
+This means that the model must allow for reasoning \emph{up to weak
+bisimilarity}, where two terms are weakly bisimilar if they differ only in their
+number of computational steps. However, the weak bisimilarity relation is not
+transitive in the guarded setting, which follows from a no-go theorem we
+establish (Theorem \ref{thm:no-go}). The New-Licata proofs freely use
+transitivity, and we argue in Section \ref{sec:towards-relational-model} that
+some amount of transitive reasoning is nessecary for defining a
+syntax-independent model of gradual typing. As a result, the lack of
+transitivity of weak bisimilarity presents a challenge in adapting the
+New-Licata model to the guarded setting. Our solution is to work with a
+combination of weak bisimilarity and step-sensitive reasoning, where the latter
+notion \emph{is} transitive. To deal with the fact that casts take computational
+steps, we introduce the novel concept of \emph{syntactic perturbations}, which
+are explicit synchronizations that we manipulate syntactically. The combination
+of lock-step reasoning with perturbations is transitive and is able to account
+for the step-insensitive nature of the graduality property.
 
-To recover some amount of compositional reasoning, we model the graduality
-ordering using a combination of two relations: a \emph{lock-step error ordering}
-in which elements must take the same number of steps, and a \emph{weak
-bisimilarity relation} that relates elements differing only in the number of
-computational steps that they take. The lock-step ordering \emph{is} transitive,
-but it is not sufficient to model the New-Licata proof. The problem is that in
-the guarded setting, the denotation of a downcast from the dynamic type involves
-a computational step, and by the functorial nature of casts this implies that
-the denotations of arbitrary casts may involve computational steps. As a result,
-the embedding-projection characterization of casts in the New-Licata model does
-not hold in the guarded setting, at least not up to equality.
-%
-However, while the casts may involve delays, they are otherwise extensionally
-equivalent to embedding-projection pairs. Thus, one might try to use weak
-bisimilarity to account for the delays, viewing casts as embedding-projection
-pairs ``up to weak bisimilarity''. Unfortunately, this does not work: since weak
-bisimilarity is \emph{not} transitive (which follows also from our no-go
-result), we are back to the original issue of lack of transitive reasoning.
+% These can then be used to explicitly \emph{synchronize} elements to ensure
+% that they are in lock-step.
+% %
+% In the end, we have come to a resolution of the issue: the lock-step error
+% ordering where we employ syntactic perturbations to synchronize elements is
+% still transitive, while simultaneously being able to relate terms that involve
+% an explicit, known pattern of computational steps.
 
-% However, since the casts involve delays but are otherwise extensionally
-% equivalent to embedding-projection pairs, one might try to use weak bisimilarity
-% to account for the delays. That is, one could try characterizing the casts as
-% embedding-projection pairs ``up to weak bisimilarity''. Unfortunately, this does
-% not work: since weak bisimilarity is \emph{not} transitive (which follows also
-% from our no-go result), we are back to the original issue of lack of
-% transitive reasoning.
+\subsection{Adequacy}
 
-It would seem that at this point, we are stuck: we need compositional reasoning,
-but we also need to be able to account for the steps taken by casts. And now we
-come to the key technical contribution of this work. The crucial observation is
-that, although the denotations of casts involve delays, the specific ways in
-which those delays arise follow a specific, type-directed pattern. Hence, we can
-track the delay behavior in a type-directed manner, equipping each type with a
-structured set of what we call \emph{syntactic perturbations}. These can then be
-used to explicitly \emph{synchronize} elements to ensure that they are in lock-step.
-%
-In the end, we have come to a resolution of the issue: the lock-step error
-ordering where we employ syntactic perturbations to synchronize elements is
-still transitive, while simultaneously being able to relate terms that involve
-an explicit, known pattern of computational steps.
-%
-We discuss these issues in more detail in Section
-\ref{sec:towards-relational-model} and give a detailed description of the
-resulting model in Section \ref{sec:concrete-relational-model}.
-
-% Perturbations allow us to recover enough extensional reasoning to model the
-% graduality property compositionally.
-
-% weak bisimilarity: elements differing only in their stepping behavior are related.
-
-% while the New-Licata semantics
-% equips $\omega$-CPOs with an error ordering, we work with types equipped with
-% not just an error ordering but also a \emph{bisimilarity} relation that reflects
-% when elements differ only in the number of abstract computational steps that
-% they take. The reason for requiring these two separate notions is that in the
-% setting of guarded domain theory, it is not possible to carry out transitive
-% reasoning while being oblivious to the number of steps the elements take. We
-% make this statement precise in \ref{sec:towards-relational-model}. To overcome
-% this limitation, we develop further changes to the core concepts of our
-% semantics as discussed in that section.
-
-% Because bisimilarity, unlike equality, is not transitive in the guarded
-% setting, this necessitates further changes to the core
-% concepts of our semantics. These challenges are discussed in Section \ref{sec:towards-relational-model}.
-
-\subsection{The Graduality Proof}
-
-Proving graduality by constructing a denotational model in guarded type theory
-consists of two main steps. First, there is the construction of the model
-itself. This gives us an interpretation of the terms as well as the syntactic
-axioms that model the graduality property. However, because this model is
-defined in the guarded domain theory, while graduality is a statement about a
-set-theoretic semantics, it is not a priori clear that the guarded model
-validates the graduality property.
+Our approach to proving graduality by constructing a denotational model in
+guarded type theory consists of two main steps. First, there is the construction
+of the model itself. This gives us an interpretation of the terms as well as the
+axioms that model the graduality property. However, since this model is defined
+in guarded domain theory, the denotations of terms are constructed as sequences of
+approximations and our graduality result is actually a statement about each
+finite approximation. The actual graduality property should be a statement about
+entire sequences of approximations.
 %
@@
 
-% The next step is to ensure that the semantics actually validates the graduality
-% property. That is, we can extract from this guarded model a well-behaved
-% semantics in ordinary mathematics For the terms, this means that closed terms of
-% base type should evaluate to themselves.
-
-% Defining a model of gradual typing in guarded type theory is not enough to prove
-% graduality. We must then extract a well-behaved ``big-step'' semantics for terms
-% of base type and show that the denotation of term precision has the expected
-% behavior. More specifically, we want a semantics in which the denotation of
-% error is different from the denotation of a value, which is in turn different
-% from the denotation of the diverging computation. Recall that by working in
-% guarded domain theory, our definitions are constructed as a sequence of
-% approximations. What we actually want is a statement that holds ``in the
-% limit''.
-
 More concretely, consider a gradually typed language whose only effects are
-gradual type errors and divergence. If we fix a result type of natural numbers,
-a whole program semantics is a partial function from closed programs to either
-natural numbers or errors:
+gradual type errors and divergence (errors arise from failing casts; divergence
+arises because we can use the dynamic type to encode untyped lambda calculus;
+see Section \ref{sec:GTLC}). If we fix a result type of natural numbers, a
+``big-step'' semantics is a partial function from closed programs to either natural
+numbers or errors:
 %
@@
 themselves: $n \Downarrow n$. Second, it should validate type based reasoning.
-To formalize type based reasoning, we give a typed equational theory for terms
-of the language $M \equiv N$ for when two terms should be considered equivalent.
+To formalize type based reasoning, languages typically have an equational theory
+$M \equiv N$ specifying when two terms should be considered equivalent.
 Then we want to verify that the big step semantics respects this equational
@@
 $M\Uparrow \iff N \Uparrow$ and $M \Downarrow \mho \iff N \Downarrow \mho$.
+
+To define the big-step semantics, we apply the technique of
+\emph{clock-quantification} \cite{atkey-mcbride2013}, which is the synthetic
+analogue of computing global elements in analytic guarded domain theory. This
+gives us an interpretation of closed programs of base type as partial functions.
+We describe the process in more detail in Section
+\ref{sec:big-step-term-semantics}. The resulting term semantics is adequate, and
+furthemore it validates the equational theory, since equivalent terms in the
+equational theory denote equal terms in the big-step semantics.
 
@@
 constructively weaker}.
-
-To define the big-step semantics, we apply clock-quantification to the guarded
-model to define a global semantics for closed terms of base type where a term
-$M$ is mapped to an element of the coinductive delay monad of Capretta
-\cite{lmcs:2265}. We compose this map with a partial function on elements of the
-delay monad that returns the result with which given element terminates, or is
-undefined if it does not terminate. This gives us the desired partial function
-from closed programs to results. We describe this process in more detail in Section
-\ref{sec:big-step-term-semantics}.
 %
-We then make the analogous construction for the relation that denotes the term
-precision ordering, showing that a term precision ordering $M \ltdyn N$ implies
-the corresponding ordering on the partial functions denoted by $M$ and $N$. This
-construction uses as an intermediate step an analogue of the graduality relation
-for the coinductive delay monad. The details of the proof are given in Section
-\ref{sec:adequacy}.
+To prove that the semantics is adequate for graduality, we again apply clock
+quantification, this time to the relation that denotes the term precision
+ordering. We show that a term precision ordering $M \ltdyn N$ implies the
+corresponding ordering on the partial functions denoted by $M$ and $N$. The
+details of the proof are given in Section \ref{sec:adequacy}.
 
@@
 %
-The notion of syntactic perturbation stands out as a key aspect of our model
-and our most significant technical contribution, allowing us to successfully
-adapt the denotational approach of New and Licata to the guarded setting.
+Within our semantics, the notion of syntactic perturbation is our most
+significant technical contribution, allowing us to successfully adapt the
+denotational approach of New and Licata to the guarded setting.
 %

### 2024-10-24 paper-new/concrete-relational-model.tex
some smaller revisions

@@
 \ltdyn_{\later A} \tilde{x'}$ iff $\later_t(\tilde{x}_t \ltdyn_A
-\tilde{x'}_t)$. Likewise for bisimilarity.
+\tilde{x'}_t)$, and likewise for bisimilarity.
 
@@
 the remaining components.
+% Interpretations of functions are different: one is arbitrary fucntions, the other is monotone functions that preserve bisimilarity
+% Interpretation of the projection is different since we add a delay
 
@@
 
-% The type precision equivalence rules follow because $c \equiv c'$
-% implies quasi-equivalence, and that quasi-equivalence implies that
-% $\id \ltsqbisim{\sem{c}}{\sem{c'}} \id$.
-
 \subsubsection{Interpreting Term Precision: Extensional Squares}
@@
 we also have these for extensional squares.
+%
+The rule involving error holds because $\mho$ is the least element in the
+ordering on the free error domain. 
+%
+Next, the rule involving type-precision equivalence follows because $c \equiv c'$
+implies quasi-equivalence of the denoted relations, and this in turn implies that
+$\id \ltsqbisim{\sem{c}}{\sem{c'}} \id$.
 %

### 2024-10-24 paper-new/concrete-term-model.tex
more paper revisions
some smaller revisions

@@
 
-\subsection{Guarded Type Theory}
+\subsection{Guarded Type Theory}\label{sec:guarded-type-theory}
 
@@
 $F$ defined above. This means $\li^{gl} \mathbb{N}$ is isomorphic to
-$\delay(\mathbb{N} + {\mho})$.
+$\delay(\mathbb{N} + {\mho})$. Denote this isomorphism by $\alpha$.
 
@@
 %
-$-\Downarrow_\delay \colon \delay (\mathbb{N} + {\mho}) \rightharpoonup \mathbb{N} + {\mho}$
+$f \colon \delay (\mathbb{N} + {\mho}) \rightharpoonup \mathbb{N} + {\mho}$
 %
 by existentially quantifying over $i$ for which $d \da^i n_?$. That is, we define
-$d \Downarrow_\delay n_? := \exists i. d \da^i n_?$.
+$f(d) := n_?$ if $d \da^i n_?$ for some $i$, and undefined otherwise.
 
@@
 \{M \,|\, \cdot \vdash M : \nat \} \to \li^{gl} \mathbb{N}$, defined by
-$\sem{M}^{gl} := \lambda (k : \Clock) .\sem{M}$. Composing this with the partial
-function $\Downarrow_\delay$ yields the desired partial function
+$\sem{M}^{gl} := \lambda (k : \Clock) .\sem{M}$. Composing this with the
+isomorphism $\alpha$ and the partial function $f$ yields the desired partial
+function
 %

### 2024-10-24 paper-new/discussion.tex
some smaller revisions

@@
 
-\subsection{Implications for Gradual Typing Research}
-
 \subsection{Mechanization}\label{sec:mechanization}
@@
     These are also included in the Appendix.
-    \item Defining the syntax-to-semantics translation.
+    \item Formalizing the syntax-to-semantics translation.
 \end{enumerate}
@@
 languages. For instance, the free error domain construction can be easily
-extended with additional cases to model effects besides error and stepping.
+extended with additional cases to model effects besides error and stepping. % TODO cite guarded interaction trees
 We also aim to complete our Agda formalization and evolve it into a
@@
 variety of other cast calculi that satisfy graduality but for which one or more
-equational axioms fail to hold. % TODO cite
+equational axioms fail to hold. \eric{Max, please add examples}
 We expect that our development can be adapted to other cast semantics to prove
@@
 \cite{herman-tomb-flanagan-2010} are not only extensionally correct, but also
-have lower quantitative cost. We leave this as future work.
+have lower quantitative cost.
 

### 2024-10-24 paper-new/towards-relational-model.tex
some smaller revisions

@@
 We call such a relation between $A$ and $A'$ a \emph{poset relation}, and we
-write $c : A \rel A'$. The paradigmatic example of a poset relation is the poset
-ordering itself, $\ltdyn_A : A \rel A$, which is a poset relation because it is
-transitive.
+write $c : A \rel A'$. The paradigmatic example of a poset relation is given by
+the ordering $\ltdyn_A$. We call this the reflexive poset relation and
+denote it by $r(A) : A \rel A$. This is a poset relation because the ordering
+$\ltdyn_A$ is transitive.
+
+% The paradigmatic example of a poset relation is the ``reflexive'' relation,
+% which is given by the ordering $\ltdyn_A$ and denoted by $r(A)$.
 
@@
 
-These relations are then used to model term precision. For closed programs $M \ltdyn M' :
-c$ where $c : A \ltdyn A'$, the terms denote elements of the
-free error domain $\sem{M},\sem{M'} : U\li \sem{A}$, and we model the term precision semantically as the relation holding between those denoted elements $\sem{M} \binrel{(U\li\sem{c})} \sem{M'}$ where
-$U\li\sem{c}$ is a relational lifting of $U\circ\li$.
+These relations are then used to model term precision. For closed programs $M
+\ltdyn M' : c$ where $c : A \ltdyn A'$, the terms denote elements of the free
+error domain $\sem{M},\sem{M'} : U\li \sem{A}$, and we model the term precision
+semantically as the relation holding between those denoted elements $\sem{M}
+\binrel{(U\li\sem{c})} \sem{M'}$. Here $U\li\sem{c}$ is a relational lifting of
+$U\circ\li$, which extends $\sem{c}$ to account for error and computational
+steps.
 %
@@
 there is a ``horizontal identity'' square $f
-\ltsq{\le_{A_i}}{\le_{A_o}} f$ arising from the fact that $f$ is
+\ltsq{r(A_i)}{r(A_o)} f$ arising from the fact that $f$ is
 monotone.
@@
     % UpL
-    % https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzIiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzIiXSxbMCwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJcXGx0ZHluX3tBXzJ9IiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJcXHVwY3tjfSIsMl0sWzIsMywiXFxpZCJdLFs2LDcsIlxcdGV4dHtVcEx9IiwzLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+    % https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzIiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzIiXSxbMCwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJyKEFfMikiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxcdXBje2N9IiwyXSxbMiwzLCJcXGlkIl0sWzYsNywiXFx0ZXh0e1VwTH0iLDMseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
     \begin{tikzcd}[ampersand replacement=\&]
@@
       \arrow[""{name=1, anchor=center, inner sep=0}, "\id", from=1-2, to=2-2]
-      \arrow["{\ltdyn_{A_2}}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+      \arrow["{r(A_2)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
       \arrow["{\text{UpL}}"{marking, allow upside down}, draw=none, from=0, to=1]
@@
     % UpR
-    % https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzEiXSxbMSwwLCJBXzEiXSxbMSwxLCJBXzIiXSxbMCwyLCJcXGx0ZHluX3tBXzF9IiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJjIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJcXGlkIiwyXSxbMiwzLCJcXHVwYyBjIl0sWzYsNywiXFx0ZXh0e1VwUn0iLDMseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
+    % https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzEiXSxbMSwwLCJBXzEiXSxbMSwxLCJBXzIiXSxbMCwyLCJyKEFfMSkiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxLDMsImMiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxcaWQiLDJdLFsyLDMsIlxcdXBjIGMiXSxbNiw3LCJcXHRleHR7VXBSfSIsMyx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
     \begin{tikzcd}[ampersand replacement=\&]
@@
       {A_1} \& {A_2}
-      \arrow["{\ltdyn_{A_1}}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+      \arrow["{r(A_1)}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
       \arrow[""{name=0, anchor=center, inner sep=0}, "\id"', from=1-1, to=2-1]
@@
     % DnL
-    % https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzIiXSxbMCwxLCJBXzEiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzIiXSxbMCwyLCJcXGx0ZHluX3tBXzJ9IiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJjIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJcXGRuYyBjIiwyXSxbMiwzLCJcXGlkIl0sWzYsNywiXFx0ZXh0e0RuTH0iLDMseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
+    % https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzIiXSxbMCwxLCJBXzEiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzIiXSxbMCwyLCJyKEFfMikiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxLDMsImMiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxcZG5jIGMiLDJdLFsyLDMsIlxcaWQiXSxbNiw3LCJcXHRleHR7RG5MfSIsMyx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
     \begin{tikzcd}[ampersand replacement=\&]
@@
       {A_1} \& {A_2}
-      \arrow["{\ltdyn_{A_2}}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
+      \arrow["{r(A_2)}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
       \arrow[""{name=0, anchor=center, inner sep=0}, "{\dnc c}"', from=1-1, to=2-1]
@@
     % DnR
-    % https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzEiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzEiXSxbMCwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJcXGx0ZHluX3tBXzF9IiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMCwxLCJcXGlkIiwyXSxbMiwzLCJcXGRuYyBjIl0sWzYsNywiXFx0ZXh0e0RuUn0iLDMseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
+    % https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzEiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzEiXSxbMCwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJyKEFfMSkiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxcaWQiLDJdLFsyLDMsIlxcZG5jIGMiXSxbNiw3LCJcXHRleHR7RG5SfSIsMyx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
     \begin{tikzcd}[ampersand replacement=\&]
@@
       \arrow[""{name=1, anchor=center, inner sep=0}, "{\dnc c}", from=1-2, to=2-2]
-      \arrow["{\ltdyn_{A_1}}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
+      \arrow["{r(A_1)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
       \arrow["{\text{DnR}}"{marking, allow upside down}, draw=none, from=0, to=1]
@@
 It is straightforward to prove using \lob-induction that this relation is
-reflexive, transitive and anti-symmetric given that the underlying relation $R$
+reflexive, transitive and anti-symmetric given that the underlying relation $\ltdyn_A$
 has those properties. The lock-step ordering is therefore the partial ordering
@@
 poset relation $c : A \rel A'$ to a poset relation $\li c : \li A \rel \li A'$.
+The definition is nearly identical, except that in the case when both sides are $\eta$ we
+check that the returned values are related in $c$.
 

### 2024-10-24 paper-new/concrete-relational-model.tex
small revisions

@@
 The complete definition of the ordering $\ltdyn_D$ is provided in the appendix
-for completeness.
+for completeness. We refer to the three injections as $e_\nat$, $e_\times$,
+and $e_\to$, respectively.
 
-To interpret term precision, we define three relations involving $D$
+To interpret type precision, we define three relations involving $D$
 as follows. We define $\inat : \mathbb{N} \rel D$ by $(n, d) \in
@@
 d_2) \ltdyn_D d$, and we define $\text{inj}_\to : U(D \arr \li D) \rel
-D$ by $(f, d) \in \iarr$ iff $e_\to(f) \ltdyn_D d$.
+D$ by $(f, d) \in \iarr$ iff $(e_\to(f) \circ \nxt) \ltdyn_D d$.
 
@@
 of our model. Recall from Section \ref{sec:lock-step-and-weak-bisim} that in order
-to establish the DnL rule for the downcast of $\iarr$, we needed to adjust the
+to establish the $\dnl$ rule for the downcast of $\iarr$, we needed to adjust the
 rule by inserting a ``delay'' on the right-hand side. A similar adjustment is
-needed for the DnR rule for $\iarr$. Moreover, the need to insert delays impacts
+needed for the $\dnr$ rule for $\iarr$. Moreover, the need to insert delays impacts
 the semantics of the cast rules for \emph{all} relations, because of the
@@
 involves a downcast in the domain and an upcast in the codomain. This has two
-consequences: first, the squares corresponding to the rules UpL and UpR may also
+consequences: first, the squares corresponding to the rules $\upl$ and $\upr$ may also
 require the insertion of a delay. Second, we need to be able to insert
@@
 now make formal the final aspect of the model: the notion of
-quasi-represntability of relations.
+quasi-representability of relations.
 %
@@
 %
-\begin{definition}[quasi-right-represntable relations]\label{def:quasi-right-representable}
+\begin{definition}[quasi-right-representable relations]\label{def:quasi-right-representable}
 Let $(B, M_B, i_B)$ and $(B', M_{B'}, i_{B'})$ be computation objects, and let
@@
 
-For ordinary representability, the relation is \emph{uniqueley
+For ordinary representability, the relation is \emph{uniquely
 determined} by any morphism that left- or right-represents it: if two
@@
 The intuition for this requirement comes from the construction of the square
-corresponding to the original UpL rule from the square for the simplified
+corresponding to the original $\upl$ rule from the square for the simplified
 version as was shown in Section \ref{sec:towards-relational-model}. Recall that
-we horizontally composed the simplified UpL square for $c$ with the identity
-square for $c'$ on the right. Now that the square for UpL involves a
+we horizontally composed the simplified $\upl$ square for $c$ with the identity
+square for $c'$ on the right. Now that the square for $\upl$ involves a
 perturbation on the right instead of the identity, this construction will not
@@
 
-For the term semantics $\sem{M}$, nothing really differs from the term
-semantics in Section~\ref{sec:concrete-term-model}. The only difference is
-that the definitions of $\sem{\upc c}$ and $\sem{\dnc c}$ can be
-defined as the morphisms that left- and right-quasi-represent $\sem
-c$, but upon inspection this definition agrees with the one given
-before. Similarly, the validity of the equational theory follows
-easily: $\beta\eta$ holds in our model. We next elaborate on each of
-the remaining components.
+The term semantics $\sem{M}$ is nearly identical to the term semantics in
+Section~\ref{sec:concrete-term-model}. One difference is that the interpretation
+of the function type $A_i \ra A_o$ in the term semantics of Section
+\ref{sec:concrete-term-model} involves \emph{arbitrary functions}, while the
+interpretation of the function type in the term semantics defined here involves
+only those functions that are morphisms of predomains. Additionally, the
+semantics of casts is slightly different. In the semantics of Section
+\ref{sec:concrete-term-model}, the projection corresponding to the type
+precision $cc'$ is simply the composition of the projections. In the new
+semantics, the corresponding projection involves a perturbation. The reason for
+this is technical and is a consequence of the construction of the composition of
+value/computation relations. (The relevant projection is defined in the proof of
+Lemma \ref{lem:representation-comp-F-U} in the Appendix.) However, the two
+projections are extensionally equivalent since they are the same up to a
+perturbation.
+
+ % nothing really differs from the term semantics in Section~\ref{sec:concrete-term-model}. The only difference is
+% One difference is that the definitions of $\sem{\upc c}$ and $\sem{\dnc c}$ can be
+% defined as the morphisms that left- and right-quasi-represent $\sem
+% c$, but upon inspection this definition agrees with the one given
+% before.
+
+The validity of the equational theory follows easily: $\beta\eta$ holds in our
+model. In the remainder of this section, we elaborate on each of the remaining
+components of the model.
 % Interpretations of functions are different: one is arbitrary fucntions, the other is monotone functions that preserve bisimilarity
@@
 
-First consider the free error domain $\li-$. Given a value object
-$(A, M_A, i_A)$ we define the syntactic perturbations $M_{\li A}$ to
-be $\mathbb{N} \oplus M_A$, where $\oplus$ denotes the free product of
-monoids (the coproduct in the category of monoids). The intuition
-behind needing to include $\mathbb{N}$ comes from the example of the
-downcast for $\iarr$. Recall that we adjusted the squares
-corresponding to the DnL and DnR rules by adding $(\delta \circ
-\eta)^\dagger$ on the side opposite the downcast. Thus, we need a
-syntactic perturbation in $M_{\li A}$ that will be interpreted as
-$(\delta \circ \eta)^\dagger$. We accomplish this by taking the
-coproduct with $\mathbb{N}$.
+First consider the free error domain $\li-$. Given a value object $(A, M_A,
+i_A)$ we define the syntactic perturbations $M_{\li A}$ to be $\mathbb{N} \oplus
+M_A$, where $\oplus$ denotes the free product of monoids (the coproduct in the
+category of monoids). The intuition behind needing to include $\mathbb{N}$ comes
+from the example of the downcast for $\iarr$. Recall that we adjusted the
+squares corresponding to the $\dnl$ and $\dnr$ rules by adding $(\delta \circ
+\eta)^\dagger$ on the side opposite the downcast. Thus, we need a syntactic
+perturbation in $M_{\li A}$ that will be interpreted as $(\delta \circ
+\eta)^\dagger$. We accomplish this by taking the coproduct with $\mathbb{N}$.
+Since $\mathbb{N}$ is the free monoid on one generator, defining a homomorphism
+from $\mathbb{N}$ to $\morbisimid{\li A}$ is equivalent to a choice of an
+element in $\morbisimid{\li A}$.
 %
@@
 The action of $\arr$ on objects is as follows. Given $(A, M_A, iA)$ and $(B,
-M_B, i_B)$ we define the monoid $M_{A \arr B} = M_A^{op} \oplus M_B$ and the interpretation
-is defined analogously to that of $\times$.
+M_B, i_B)$ we define the monoid $M_{A \arr B} = M_A^{op} \oplus M_B$. (Here,
+$M^{op}$ denotes the monoid with the same elements but with the order of
+multiplication reversed, i.e., $x \cdot_{M^{op}} y := y \cdot_M x$.) To define
+the interpretation $i_{A \arr B}$, it suffices to define a homomorphism
+$M_A^{op} \to \morbisimid{A \arr B}$ and $M_B \to \morbisimid{A \arr B}$. The
+former takes an element $m_A$ and a predomain morphism $f : A \to UB$ and
+returns the composition $f \circ i_A(m_A)$, while the latter is defined
+similarly by post-composition with $i_B$.
 % Could state this using the functorial action of \arr
@@
 Recall the definition of the downcast for $\iarr$ given in Section
-\ref{sec:term-interpretation}. We know that in order for the DnR and DnL squares
+\ref{sec:term-interpretation}. We know that in order for the $\dnr$ and $\dnl$ squares
 for $\li \iarr$ to exist, we must insert a delay on the side opposite the
@@
 With this choice for the perturbations, it is straightforward to show that the
-DnL and DnR squares exist using the definition of the downcast and the
+$\dnl$ and $\dnr$ squares exist using the definition of the downcast and the
 definition of the relation $\iarr$.
@@
 %
-The rule involving error holds because $\mho$ is the least element in the
+The \textsc{ErrBot} rule holds because $\mho$ is the least element in the
 ordering on the free error domain. 
 %
-Next, the rule involving type-precision equivalence follows because $c \equiv c'$
+Next, the \textsc{EquivTyPrec} rule follows because $c \equiv c'$
 implies quasi-equivalence of the denoted relations, and this in turn implies that
@@
 %
-Thus, it remains to verify the existence of the squares for the cast rules. For UpL, the
+Lastly, it remains to verify the existence of the squares for the cast rules. For $\upl$, the
 relevant extensional square is obtained as follows:
@@
 into a perturbation on $A_3$. Lastly we note that the bottom edge of the square
-is equal to $c'$ by the fact that $c'$ is downward closed. The squares for UpR,
-DnL, and DnR are obtained analogously.
+is equal to $c'$ by the fact that $c'$ is downward closed. The squares for $\upr$,
+$\dnl$, and $\dnr$ are obtained analogously.
 

### 2024-10-24 paper-new/concrete-term-model.tex
small revisions

@@
   A homomorphism of error domains $\phi : B \multimap B'$ is a
-  function $\phi : UB \to UB'$ that preserves these in that
+  function $\phi : UB \to UB'$ that preserves $\mho$ and $\theta$, i.e.,
   $\phi(\mho_B) = \mho_{B'}$ and $\phi(\theta_{B}(\tilde{x})) =
@@
 ``diverging element'' that ``thinks forever'' in that $\Omega_B =
-\delta(\Omega_B)$. We call this a ``quasi'' fixed point because it is
+\delta_B(\Omega_B)$. We call this a ``quasi'' fixed point because it is
 \emph{nearly} a fixed point except for the presence of the delay map
@@
 modeled as effectful functions $\sem{M}: \sem{A_1}\times \cdots \to U\li\sem{A}$
-and values $x:A_1,\ldots \vdash V : A$ are also be modeled as pure
+and values $x:A_1,\ldots \vdash V : A$ are also modeled as pure
 functions $\sem{V} : \sem{A_1}\times\cdots \to \sem{A}$.
 %
-The interpretation of the typical CBV features of GTLC into a CBPV is
-standard \cite{levy99}, the only additional part we need to account for
+The interpretation of the typical CBV features of GTLC into a model of CBPV is
+standard \cite{levy99}; the only additional part we need to account for
 is the semantics of casts, shown in Figure \ref{fig:term-semantics}.
@@
     % upcasts
-    \sem{\upc{r(A)}} &= \id_A \\
+    \sem{\upc{r(A)}} &= \id_{\sem{A}} \\
     \sem{\upc{(c \comp c')}} &= \sem{\upc{c'}} \circ \sem{\upc{c}} \\
@@
     % downcasts
-    \sem{\dnc{r(A)}} &= \id_{FA} \\
+    \sem{\dnc{r(A)}} &= \id_{\li \sem{A}} \\
     \sem{\dnc{(c \comp c')}} &= \sem{\dnc{c}} \circ \sem{\dnc{c'}} \\
@@
 \mathbb{N} + {\mho}$). From this definition, it is clear that if there is an $i$
-such that $d \da^i$, then this $i$ is unique. Furthermore, from the proof that
-$d$ terminates in $i$ steps, we can extract the result with which $d$
-terminates. This is defined by a straightforward induction on the proof that $d \da^i$.
-
+such that $d \da^i$, then this $i$ is unique. 
+% Furthermore, from the proof that
+% $d$ terminates in $i$ steps, we can extract the result with which $d$
+% terminates. This is defined by a straightforward induction on the proof that $d \da^i$.
 Using this definition of termination, we define a partial function 
@@
 %
-by existentially quantifying over $i$ for which $d \da^i n_?$. That is, we define
-$f(d) := n_?$ if $d \da^i n_?$ for some $i$, and undefined otherwise.
+by existentially quantifying over $i$ for which $d \da^i n_?$. That is, we
+define $f(d) := n_?$ if $d \da^i n_?$ for some (necessarily unique) $i$, and $f$
+is undefined otherwise.
 

### 2024-10-24 paper-new/discussion.tex
small revisions

@@
 bisimilarity. We also formalized the no-go theorem from Section
-\ref{sec:concrete-term-model}.
+\ref{sec:towards-relational-model}.
 
@@
 languages. For instance, the free error domain construction can be easily
-extended with additional cases to model effects besides error and stepping. % TODO cite guarded interaction trees
+extended to model effects besides error and stepping
+\cite{guarded-interaction-trees, probabilistic-fpc-guarded}.
 We also aim to complete our Agda formalization and evolve it into a

### 2024-10-24 paper-new/towards-relational-model.tex
small revisions

@@
 
-At this point we might try to weaken from our initial attempt at
+At this point we might try to weaken our initial attempt at
 providing a poset semantics to a semantics where types are equipped
@@
 and follow easily. The core of the proof of the graduality property is
-in proving the validity of the \emph{cast rules} UpL, UpR, DnL and
-DnR.
+in proving the validity of the \emph{cast rules} $\upl$, $\upr$, $\dnl$, and
+$\dnr$.
 %
@@
 
-Consider the UpL rule. It states that if
+Consider the $\upl$ rule. It states that if
 $c : A_1 \ltdyn A_2$ and $c' : A_2 \ltdyn A_3$ and $M : A_1$ is
@@
 $\sem{c}$ and $\sem{\upc c}$, there is a problem: it also quantifies
-involves an arbitrary other relation $\sem{c'}$! This means we cannot
+over an arbitrary other relation $\sem{c'}$! This means we cannot
 require the existence of this square as part of the definition of a
@@
 that do not involve composition of relations. Below are the simpler squares
-corresponding to UpL, UpR, DnL, and DnR:
+corresponding to $\upl$, $\upr$, $\dnl$, and $\dnr$:
 \begin{center}
@@
 
-We can use the simpler UpL square to derive the original UpL square by
+We can use the simpler $\upl$ square to derive the original $\upl$ square by
 horizontally composing with the identity square for $c'$:
@@
 $c'$, because $c'$ is \emph{downward-closed} under the relation on $A_2$.
-An analogous argument shows that we can derive the original UpR square from the simpler version of UpR,
+An analogous argument shows that we can derive the original $\upr$ square from the simpler version of $\upr$,
 this time using the fact that poset relations are \emph{upward-closed}.
@@
 each side. In particular, this implies that the ordering relations $\ltdyn_{A}$
-must be transitive. Thus, we cannot entirely drop posets from our semantics.
+must be transitive, so we cannot entirely drop posets from our semantics.
 
@@
 %
-Consider the representability condition corresponding to the DnL rule
+Consider the representability condition corresponding to the $\dnl$ rule
 for $\iarr \colon \dyntodyn \ltdyn\, \dyn$. This is the square $\sem{\dnc{\iarr}} \ltsq{r({\li D})}{\li\sem{\iarr}} \id$.
@@
 %
-However the right hand side \emph{always} takes $0$ steps, as it is
+However, the right hand side \emph{always} takes $0$ steps, as it is
 the identity function, whereas on the left, our definition of
@@
 
-However, observe that we can remedy this particular situation by
+Observe that we can remedy this particular situation by
 replacing the $\id$ on the right hand side by an innocuous function

### 2024-10-24 paper-new/intro.tex
small tweaks

@@
 
-\emph{Gradually-typed languages} \cite{siek-taha06, tobin-hochstadt06} seek to
+\emph{Gradually typed languages} \cite{siek-taha06, tobin-hochstadt06} seek to
 resolve this tension by allowing for both static and dynamic typing disciplines
@@
 Additionally, gradually typed languages should offer some of the
-benefits of static typing. While classical type soundness, that
+benefits of static typing. While standard type soundness, that
 well-typed programs are free from runtime errors, is not compatible
-with runtime type errors, we can instead focus on \emph{type-based
-reasoning}. For instance, while dynamically typed $\lambda$ calculi
-only satisfy $\beta$ equality for their type formers, statically typed
-$\lambda$ calculi additionally satisfy type-dependent $\eta$
-properties that ensure that functions are determined by their behavior
-under application and that pattern matching on data types
-is safe and exhaustive.
+with runtime type errors, it is possible instead to prove that
+gradually typed languages validate \emph{type-based reasoning}. For
+instance, while dynamically typed $\lambda$ calculi only satisfy
+$\beta$ equality for their type formers, statically typed $\lambda$
+calculi additionally satisfy type-dependent $\eta$ properties that
+ensure that functions are determined by their behavior under
+application and that pattern matching on data types is safe and
+exhaustive. A gradually typed calculus that validates these
+type-dependent $\eta$ laws then provides some of the type-based
+reasoning that dynamic languages lack.
 
@@
 %
-\max{we should make clear that we don't yet model these perverse recursive features}
-\eric{In the contributions section, I explain that we don't yet model these features}
 However, an approach based on classical domain theory has fundamental
@@
 %
-Our long-term goal is to develop a semantics that will scale up to these
-advanced features, and so we begin with this work by departing from
-classical domain theory.
+Our long-term goal is to develop a denotational approach that can
+scale up to these advanced features, and so we must abandon classical
+domain theory as the foundation in order to make progress. In this
+work, we provide first steps towards this goal by adapting work on a
+simple gradually typed lambda calculus to \emph{guarded domain
+theory}. Our goal is for this model to scale to dynamic type tag
+generation and higher-order references in future work.
 
-The main denotational alternative to classical domain theory that can
-successfully model these advanced features is \emph{guarded} domain
-theory, which we adopt in this work. While classical domain theory is
-based on modeling types as ordered sets with certain joins, guarded
-domain theory is based on an entirely different foundations, sometimes
-(ultra)metric spaces but more commonly as ``step-indexed sets'', i.e.,
-objects in the \emph{topos of trees}
+\emph{Guarded} domain theory is the main denotational alternative to
+classical domain theory that can successfully model these advanced
+features. While classical domain theory is based on modeling types as
+ordered sets with certain joins, guarded domain theory is based on an
+entirely different foundations, sometimes (ultra)metric spaces but
+more commonly as ``step-indexed sets'', i.e., objects in the
+\emph{topos of trees}
 \cite{birkedal-mogelberg-schwinghammer-stovring2011}.  Such an object
@@
 category theoretic terminology, these are presheaves on the poset of
-natural numbers.)  We think of a $\mathbb{N}$-indexed set as an infinite
-sequence of increasingly precise approximations to the true type being
-modeled.
+natural numbers.)  We think of a $\mathbb{N}$-indexed set as an
+infinite sequence of increasingly precise approximations to the true
+type being modeled.
 %
@@
 \in X_i$ for all $i \in \mathbb{N}$, then the denotation of the diverging
+
+%% \max{what is the point you are trying to convey in this paragraph?}
+%% Because the solutions we obtain working in guarded domain theory are constructed
+%% as a sequence of increasingly better approximations, when we want to establish a
+%% property that holds in the limit, we must reason about all finite
+%% approximations. In the setting of the topos of trees, this manifests as the need
+%% to consider \emph{global elements} of the presheaves, i.e., a family of elements
+%% $x_i \in X_i$ compatible with the restriction maps. For example, consider the
+%% set of programs that may take a computational step or return unit.\footnote{This example is adapted from \cite{mogelberg-paviotti2016}.} In the
+%% topos of trees, we can model this set as the indexed family of sets $\{X_n\}_{n \in \mathbb{N}}$
+%% where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1, \bot\}$. Here, $i$ denotes a program that steps
+%% $i$ times and then returns, and $\bot$ denotes a program that fails to terminate
+%% in $n-1$ steps or fewer. For any fixed $n$, the set $X_n$ fails to distinguish a
+%% program that takes $n$ steps and then terminates from a program that never
+%% terminates. On the other hand, if we instead consider the denotation of a
+%% program to be a global element of the presheaf $\{X_n\}_{n \in \mathbb{N}}$ -- a family of elements
+%% $x_i \in X_i$ for all $i \in \mathbb{N}$ -- then the denotation of the diverging
+%% >>>>>>> Stashed changes
 program will be distinct from that of a program that terminates, regardless of

### 2024-10-24 paper-new/discussion.tex
discussion stuff

@@
 
-Additionally, while we have focused on a single semantics in this work, there are a
-variety of other cast calculi that satisfy graduality but for which one or more
-equational axioms fail to hold. \eric{Max, please add examples}
-We expect that our development can be adapted to other cast semantics to prove
-graduality and equational reasoning principles (to the extent that the
-principles hold in those cast semantics).
+This work has focused on the ``Natural'' semantics of casts , which
+validate the full call-by-value $\eta$ laws for types, but other cast
+semantics have been proposed such as eager
+(\cite{herman-tomb-flanagan-2010}) and transient (\cite{transient})
+cast semantics which trade off certain $\eta$ equalities for other
+benefits such as early error detection or reduced runtime
+overhead\cite{deepandshallowtypes,newlicataahmed19}. It would be a
+good test of the generality of our semantic framework to see if it can
+model these alternative semantics by adjusting the semantics of types
+and casts, and providing proofs of weakened $\eta$ principles.
 
-We also plan to use our framework to investigate the performance of
-gradually-typed languages. Although we have focused on proving extensional
-properties in this work, the intensional nature of our semantics can be viewed
-as a form of \emph{cost semantics}, where a runtime cost is incurred from
-inspecting the dynamic type. This could be used to verify that cast
-optimizations such as space efficient implementations
-\cite{herman-tomb-flanagan-2010} are not only extensionally correct, but also
-have lower quantitative cost.
+Additionally, while our focus in this work has been on reasoning up to
+weak bisimilarity, the presence of explicit step counting in the model
+could be viewed as a form of \emph{cost semantics}, where a runtime
+cost is incurred from inspecting the dynamic type. This could possibly
+be used to verify that cast optimizations such as space efficient
+implementations \cite{herman-tomb-flanagan-2010} are not only
+extensionally correct, but have a lower abstract cost.
 

### 2024-10-24 paper-new/concrete-relational-model.tex
revise intro; edit the two remarks in the relational model section

@@
 \begin{remark}
-  The reason the set of perturbations needs to carry a monoid structure is that
-  we will need to produce a syntactic perturbation corresponding to the identity
-  relations $r(A)$, as well as one for the composition of relations. The monoid
-  structure gives us what we need to produce these elements. Likewise, the
-  reason that the interpretations as semantic perturbations must be
-  \emph{homomorphisms} of monoids is that we need a syntactic perturbation that
-  is interpreted as the identity semantic perturbation, and likewise given two
-  syntactic perturbations $\delta_1$ and $\delta_2$ we need a syntactic
-  pertubration $\delta_{1,2}$ that is interpreted as the composition of the
-  semantic perturbations $i_A(\delta_1)$ and $i_A(\delta_2)$.
+  The reason why the set of syntactic perturbations needs to carry a monoid
+  structure and why the interpretation needs to be a homomorphism of monoids is
+  related to the construction of the identity relations and composition of
+  relations on value/computation types. As part of this construction, we will
+  need to produce a syntactic perturbation corresponding to the identity
+  predomain/error domain relation, as well as corresponding to the composition
+  of relations. The monoid structure gives us what we need to produce these
+  perturbations.
+  % we will need to produce a syntactic perturbation corresponding to the identity
+  % relations $r(A)$, as well as one for the composition of relations. The monoid
+  % structure gives us what we need to produce these elements. Likewise, the
+  % reason that the interpretations as semantic perturbations must be
+  % \emph{homomorphisms} of monoids is that we need a syntactic perturbation that
+  % is interpreted as the identity semantic perturbation, and likewise given two
+  % syntactic perturbations $\delta_1$ and $\delta_2$ we need a syntactic
+  % pertubration $\delta_{1,2}$ that is interpreted as the composition of the
+  % semantic perturbations $i_A(\delta_1)$ and $i_A(\delta_2)$.
+  % These properties are used to define the identity and composition of value and computation relations
 \end{remark}
 
+% It doesn't work: not all types have such an action, e.g., the function type, because of the definition of the perturbations
+% The fact that we cannot define the morphisms this way is unfortunate, since doing so would simplify
+% some aspects of the model construction.
 \begin{remark}
@@
   M_{A'}, i_{A'})$ to be a predomain morphism $f \colon A \to A'$ along with a
-  monoid homomorphism $h \colon M_A \to M_{A'}$ such that $f$ and $h$ satisfied
-  a compatibility condition with respect to the interpretations $i_A$ and
-  $i_{A'}$. However, with this definition of morphism, the category of value
-  types is no longer Cartesian closed and is not a model of call-by-push-value.
-  This is unfortunate, as it would simplify some aspects of the model construction.
+  monoid homomorphism $h \colon M_A \to M_{A'}$ such that $f$ and $h$ satisfy a
+  compatibility condition with respect to the interpretations $i_A$ and
+  $i_{A'}$. However, with this definition we cannot interpret the function type,
+  since the morphisms defined in this way do not themselves form a value type. 
+  % This is due to the way that the syntactic perturbations are defined for the function type.
+  %
+  The fact that we cannot define the morphisms in this way is unfortunate, since
+  doing so would simplify some aspects of the model construction.
+
+  % However, with this definition of morphism, the category of value
+  % types is no longer Cartesian closed and is not a model of call-by-push-value.
+  % This is unfortunate, as it would simplify some aspects of the model construction.
 \end{remark}
@@
 %
-The \textsc{ErrBot} rule holds because $\mho$ is the least element in the
-ordering on the free error domain. 
+The \textsc{ErrBot} rule holds because $\mho$ is the least element in the semantics of the . 
 %

### 2024-10-24 paper-new/intro.tex
revise intro; edit the two remarks in the relational model section

@@
 However, an approach based on classical domain theory has fundamental
-limitations: domain theory is incapable of modeling certain perversely
+limitations: domain theory appears incapable of modeling certain perversely
 recursive features of programming languages such as dynamic type tag
-generation and higher-order references, which are commonplace in
+generation and higher-order references \cite{Birkedal-Stovring-Thamsborg-2009}, which are commonplace in
 real-world gradually typed systems as well as gradual calculi.
@@
 % The statement of graduality is in terms of global elements
-\eric{This next paragraph could be moved to the section on adequacy, but since
-the example is in the analytic setting it also makes sense to keep it here. I'm
-not sure which option makes more sense.} While the definitions in guarded domain
-theory are constructed as sequences of approximations, results about semantics
-of programs, e.g., graduality, are actually statements concerning entire
-\emph{sequences} of approximations. As a simple example, consider the set of
-programs that may take a computational step or return unit.\footnote{This
-example is adapted from \cite{mogelberg-paviotti2016}.} In the topos of trees,
-we can model the set of these programs as the indexed family of sets $\{X_n\}_{n
-\in \mathbb{N}}$ where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1,
-\bot\}$. Here, $i$ denotes a program that steps $i$ times and then returns, and
-$\bot$ denotes a program that fails to terminate in $n-1$ steps or fewer. For
-any fixed $n$, the set $X_n$ fails to distinguish a program that takes $n$ steps
-and then terminates from a program that never terminates. On the other hand, if
-we define the denotation of a program to be a \emph{sequence} of elements $x_i
-\in X_i$ for all $i \in \mathbb{N}$, then the denotation of the diverging
-
-%% \max{what is the point you are trying to convey in this paragraph?}
-%% Because the solutions we obtain working in guarded domain theory are constructed
-%% as a sequence of increasingly better approximations, when we want to establish a
-%% property that holds in the limit, we must reason about all finite
-%% approximations. In the setting of the topos of trees, this manifests as the need
-%% to consider \emph{global elements} of the presheaves, i.e., a family of elements
-%% $x_i \in X_i$ compatible with the restriction maps. For example, consider the
-%% set of programs that may take a computational step or return unit.\footnote{This example is adapted from \cite{mogelberg-paviotti2016}.} In the
-%% topos of trees, we can model this set as the indexed family of sets $\{X_n\}_{n \in \mathbb{N}}$
-%% where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1, \bot\}$. Here, $i$ denotes a program that steps
-%% $i$ times and then returns, and $\bot$ denotes a program that fails to terminate
-%% in $n-1$ steps or fewer. For any fixed $n$, the set $X_n$ fails to distinguish a
-%% program that takes $n$ steps and then terminates from a program that never
-%% terminates. On the other hand, if we instead consider the denotation of a
-%% program to be a global element of the presheaf $\{X_n\}_{n \in \mathbb{N}}$ -- a family of elements
-%% $x_i \in X_i$ for all $i \in \mathbb{N}$ -- then the denotation of the diverging
-%% >>>>>>> Stashed changes
-program will be distinct from that of a program that terminates, regardless of
-the number of steps it takes.
-
-% Because the solutions we obtain working in guarded domain theory are constructed
-% as a sequence of increasingly better approximations, when we want to establish a
-% property that holds in the limit, we must reason about all finite
-% approximations. In the setting of the topos of trees, this manifests as the need
-% to consider \emph{global elements} of the presheaves, i.e., a family of elements
-% $x_i \in X_i$ compatible with the restriction maps.
-
-% \max{this next paragraph is a bit too mysterious}
-% This has important ramifications for defining an adequate denotational semantics
-% in guarded domain theory, as we seek to do in this paper: we do not want our
-% denotational semantics to conflate a diverging program with a program that fails
-% to terminate in sufficiently few steps. Thus, establishing adequacy of a
-% semantics in guarded domain theory will require a means of globalizing,
-% a point we will return to in Section \ref{sec:big-step-term-semantics}.
+% \eric{This next paragraph could be moved to the section on adequacy, but since
+% the example is in the analytic setting it also makes sense to keep it here. I'm
+% not sure which option makes more sense.} While the definitions in guarded domain
+% theory are constructed as sequences of approximations, results about semantics
+% of programs, e.g., graduality, are actually statements concerning entire
+% \emph{sequences} of approximations. As a simple example, consider the set of
+% programs that may take a computational step or return unit.\footnote{This
+% example is adapted from \cite{mogelberg-paviotti2016}.} In the topos of trees,
+% we can model the set of these programs as the indexed family of sets $\{X_n\}_{n
+% \in \mathbb{N}}$ where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1,
+% \bot\}$. Here, $i$ denotes a program that steps $i$ times and then returns, and
+% $\bot$ denotes a program that fails to terminate in $n-1$ steps or fewer. For
+% any fixed $n$, the set $X_n$ fails to distinguish a program that takes $n$ steps
+% and then terminates from a program that never terminates. On the other hand, if
+% we define the denotation of a program to be a \emph{sequence} of elements $x_i
+% \in X_i$ for all $i \in \mathbb{N}$, then the denotation of the diverging
+% program will be distinct from that of a program that terminates, regardless of
+% the number of steps it takes.
 
@@
 more detail in Section \ref{sec:guarded-type-theory}.
+
+% The move to synthetic setting is useful, but now you're stuck in the guarded setting
+The simplicity afforded by working in the synthetic setting comes with a
+drawback: by changing our ambient foundations so that all types represent
+step-indexed sets, it is no longer possible to define a type that denotes an
+ordinary set. Roughly speaking, what is needed is an ``escape hatch'' to leave
+the guarded setting and make definitions that are interpreted as constructions
+in ordinary mathematics. This is typically accomplished by adding a modality to
+the guarded type theory. The specific approach we take in this paper is the
+technique of \emph{clock-quantification} \cite{atkey-mcbride2013}. In
+particular, the guarded type theory in which we work has a notion of clocks that
+index all of our guarded constructions. This is discussed in more detail in
+Section \ref{sec:big-step-term-semantics}.
+
 %
@@
 
-Our approach to proving graduality by constructing a denotational model in
-guarded type theory consists of two main steps. First, there is the construction
-of the model itself. This gives us an interpretation of the terms as well as the
-axioms that model the graduality property. However, since this model is defined
-in guarded domain theory, the denotations of terms are constructed as sequences of
-approximations and our graduality result is actually a statement about each
-finite approximation. The actual graduality property should be a statement about
-entire sequences of approximations.
-%
-What we need is to show that the guarded model induces a sensible set-theoretic
-semantics in ordinary mathematics. We cannot hope to derive such a semantics for
-\emph{all} types (e.g., the dynamic type), but for the subset of closed terms of
-base type, we should be able to extract a well-behaved semantics for which the
-graduality property holds.
+% Our approach to proving graduality by constructing a denotational model in
+% guarded type theory consists of two main steps. First, there is the construction
+% of the model itself. This gives us an interpretation of the terms as well as the
+% axioms that model the graduality property.
+
+Having defined the model of gradual typing in guarded type theory, we need to
+show that the guarded model induces a sensible set-theoretic semantics in
+ordinary mathematics. We cannot hope to derive such a semantics for \emph{all}
+types (e.g., the dynamic type), but for the subset of closed terms of base type,
+we should be able to extract a well-behaved semantics for which the graduality
+property holds.
 
@@
 
-To define the big-step semantics, we apply the technique of
-\emph{clock-quantification} \cite{atkey-mcbride2013}, which is the synthetic
-analogue of computing global elements in analytic guarded domain theory. This
-gives us an interpretation of closed programs of base type as partial functions.
+To define the big-step semantics, we apply clock quantification to obtain
+an interpretation of closed programs of base type as partial functions.
 We describe the process in more detail in Section
@@
 % setting.
-The goal we have kept in mind while designing our model is to handle
-gradually-typed languages with advanced features that classical domain theory in
-incapable of modeling. We do not address these features in this work but leave
-them instead to immediate future work, as handling even the simplest
-gradually-typed language already poses interesting challenges.
 %
@@
   setting and explore the difficulties of proving graduality in an intensional
-  model. We prove the no-go theorem about extensional, transitive relations,
+  model. We prove a no-go theorem about extensional, transitive relations,
   introduce the lock-step error ordering and weak bisimilarity relation, and

### 2024-10-24 paper-new/concrete-relational-model.tex
edit the explanation of the perturbations for dyn

@@
 It may seem as though we need to define the perturbations for $D$ by
-guarded recursion, but in fact we define them via least-fixpoint\footnote{
-  The reason for defining the perturbations inductively is technical and has to
-  do with the fact that we needed to define the perturbations corresponding to the
-  predomain $\later A$ to be the perturbations of $A$, \emph{not} later of the
-  perturbations of $A$. There may be a way to make this work by changing some
-  definitions, but we leave this as future work}
+guarded recursion, but in fact we define them via least-fixpoint
 in the category of monoids:
@@
 \( M_D \cong (M_{D \times D}) \oplus M_{U(D \to \li D)}. \)
+%
+\begin{remark}
+  The reason for defining the perturbations inductively is technical and has to
+  do with the fact that the perturbations corresponding to the predomain $\later
+  A$ need to be the same as the perturbations of $A$, \emph{not} later of the
+  perturbations of $A$. There may be a way to make this work by changing some
+  definitions; we leave this to future investigation.
+\end{remark}
 %

### 2024-10-24 paper-new/discussion.tex
fix broken citations

@@
 benefits such as early error detection or reduced runtime
-overhead\cite{deepandshallowtypes,newlicataahmed19}. It would be a
+overhead\cite{deepandshallowtypes,new-licata-ahmed2019}. It would be a
 good test of the generality of our semantic framework to see if it can

### 2024-10-24 paper-new/intro.tex
fix broken citations

@@
 style of Moggi, or adjunctions in the style of
-Levy\cite{moggi,levy}. Relational properties and their verification
+Levy\cite{MOGGI199155,39155,levy99,levystacks}. Relational properties and their verification
 lead naturally to the use of reflexive graph categories or double
-categories\cite{reflgraphcats,doublecats}.
+categories\cite{dunphythesis,new-licata18}.
 

### 2024-10-24 paper-new/concrete-relational-model.tex
changes we worked on together

@@
 \begin{remark}
-  The reason why the set of syntactic perturbations needs to carry a monoid
-  structure and why the interpretation needs to be a homomorphism of monoids is
-  related to the construction of the identity relations and composition of
-  relations on value/computation types. As part of this construction, we will
-  need to produce a syntactic perturbation corresponding to the identity
-  predomain/error domain relation, as well as corresponding to the composition
-  of relations. The monoid structure gives us what we need to produce these
-  perturbations.
+  Syntactic perturbations are equipped with a monoid structure in
+  order to support identity and composition relations on value and
+  computation types. That is, the identity relation requires an
+  identity perturbation and composition of relations requires a
+  composition operation on perturbations.
+  %% The reason why the set of syntactic perturbations needs to carry a monoid
+  %% structure and why the interpretation needs to be a homomorphism of monoids is
+  %% related to the construction of the identity relations and composition of
+  %% relations on value/computation types. As part of this construction, we will
+  %% need to produce a syntactic perturbation corresponding to the identity
+  %% predomain/error domain relation, as well as corresponding to the composition
+  %% of relations. The monoid structure gives us what we need to produce these
+  %% perturbations.
   % we will need to produce a syntactic perturbation corresponding to the identity
@@
 \begin{remark}
-  It might seem strange that in defining the morphisms of value and computation
-  objects, there is no requirement involving the monoids. For example, one could
-  imagine defining a morphism between value objects $(A, M_A, i_A)$ and $(A',
-  M_{A'}, i_{A'})$ to be a predomain morphism $f \colon A \to A'$ along with a
-  monoid homomorphism $h \colon M_A \to M_{A'}$ such that $f$ and $h$ satisfy a
-  compatibility condition with respect to the interpretations $i_A$ and
-  $i_{A'}$. However, with this definition we cannot interpret the function type,
-  since the morphisms defined in this way do not themselves form a value type. 
+  It might seem strange that we equip objects with a monoid of
+  perturbations but we do not require morphisms to respect this
+  structure in any way. A natural attempt would be defining a morphism
+  between value objects $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ to
+  be a predomain morphism $f \colon A \to A'$ along with a monoid
+  homomorphism $h \colon M_A \to M_{A'}$ such that $f$ and $h$ satisfy
+  a compatibility condition with respect to the interpretations $i_A$
+  and $i_{A'}$. However, this condition simply fails to hold for all
+  terms, in particular the property is preserved by $\lambda$
+  abstraction.
+  %% with this definition we cannot interpret the
+  %% function type, since the morphisms defined in this way do not
+  %% themselves form a value type.
   % This is due to the way that the syntactic perturbations are defined for the function type.
   %
-  The fact that we cannot define the morphisms in this way is unfortunate, since
-  doing so would simplify some aspects of the model construction.
+  %% The fact that we cannot define the morphisms in this way is unfortunate, since
+  %% doing so would simplify some aspects of the model construction.
 
@@
 \begin{remark}
-  The reason for defining the perturbations inductively is technical and has to
-  do with the fact that the perturbations corresponding to the predomain $\later
-  A$ need to be the same as the perturbations of $A$, \emph{not} later of the
-  perturbations of $A$. There may be a way to make this work by changing some
-  definitions; we leave this to future investigation.
+  Here it is crucial to define the perturbations of the dynamic type
+  \emph{inductively} rather than by guarded recursion. If we were to
+  define the perturbations by guarded recursion, then it would be
+  impossible to define the pull homomorphism for the $\iarr$ relation.
 \end{remark}

### 2024-10-24 paper-new/intro.tex
changes we worked on together

@@
 
-\subsection{Synthetic Guarded Domain Theory}
-
 While guarded domain theory can be presented analytically using
@@
 more detail in Section \ref{sec:guarded-type-theory}.
-
-% The move to synthetic setting is useful, but now you're stuck in the guarded setting
-The simplicity afforded by working in the synthetic setting comes with a
-drawback: by changing our ambient foundations so that all types represent
-step-indexed sets, it is no longer possible to define a type that denotes an
-ordinary set. Roughly speaking, what is needed is an ``escape hatch'' to leave
-the guarded setting and make definitions that are interpreted as constructions
-in ordinary mathematics. This is typically accomplished by adding a modality to
-the guarded type theory. The specific approach we take in this paper is the
-technique of \emph{clock-quantification} \cite{atkey-mcbride2013}. In
-particular, the guarded type theory in which we work has a notion of clocks that
-index all of our guarded constructions. This is discussed in more detail in
-Section \ref{sec:big-step-term-semantics}.
 
@@
 $M\Uparrow \iff N \Uparrow$ and $M \Downarrow \mho \iff N \Downarrow \mho$.
+
+% The move to synthetic setting is useful, but now you're stuck in the guarded setting
+We have to take care when defining such a big-step semantics in
+guarded type theory, as a naive definition of termination is
+incorrect: we can prove using guarded recursion that an infinite loop
+terminates! This is because the definition of termination inside
+guarded logic has a different meaning, essentially meaning that for
+every finite number $n$, either the program terminates in $< n$ steps
+or it takes $n+1$ steps, which is equivalent to a statement of
+progress rather than termination. For this reason, we need a kind of
+``escape hatch'' in our ambient type theory to leave the guarded
+setting and make definitions that are interpreted as constructions in
+ordinary mathematics. We incorporate this by using
+\emph{clock-quantification}\max{Eric should provide a citation for Clocked cubical here since that's close to our actual approach} \cite{atkey-mcbride2013} in our type
+theory. In particular, the guarded type theory in which we work has a
+notion of clocks that index all of our guarded constructions. This is
+discussed in more detail in Section \ref{sec:big-step-term-semantics}.
 

### 2024-10-25 paper-new/concrete-relational-model.tex
move definition of the free composition of error domain relations out of appendix

@@
 particular is not preserved), but we instead use an alternative ``free
-composition'' $d \odot d'$ that is inductively generated from the
-relational composition. Details of this construction are in the
-appendix (Definition \ref{def:free-comp-ed-rel}).
+composition'' $d \odot d' : B \rel B''$ that is inductively generated from the
+relational composition. The rules are given in Figure \ref{fig:free-comp-ed-rel}.
+% the appendix (Definition \ref{def:free-comp-ed-rel}).
 
-We also will use the notion of a square $f \ltsq{c_i}{c_o} g$ between
-predomain morphisms as well as squares $\phi \ltsq{d_i}{d_o} \psi$
-between error domain morphisms, which are defined similarly, with no
-extra interaction with the algebraic structure.
+\begin{figure}
+  \begin{mathpar}
+    \inferrule
+    {x \mathbin{d} y \and y \mathbin{d'} z}
+    {x \mathbin{(d \relcomp d')} z}
+
+    \inferrule
+    {x' \ltdyn_{B_1} x \and x \mathbin{(d \relcomp d')} z}
+    {x' \mathbin{(d \relcomp d')} z}
+
+    \inferrule
+    {x \mathbin{(d \relcomp d')} z \and z \ltdyn_{B_3} z'}
+    {x \mathbin{(d \relcomp d')} z'}
+
+    \inferrule
+    { }
+    {\mho_{B_1} \mathbin{(d \relcomp d')} z}
+
+    \inferrule
+    {\later_t[ \tilde{x}_t \mathbin{(d \relcomp d')} \tilde{z}_t ] }
+    {\theta_{B_1}(\tilde{x}) \mathbin{(d \relcomp d')} \theta_{B_3}(\tilde{z}) }
+  \end{mathpar}
+  \caption{The free composition of error domain relations.}
+  \label{fig:free-comp-ed-rel}
+\end{figure}
+
+As in the previous section, we also have the notion of a square $f
+\ltsq{c_i}{c_o} g$ between predomain morphisms, which is defined to mean that
+for all $x \binrel{c_i} y$, we have $f(x) \binrel{c_o} g(y)$. Likewise, we have
+squares $\phi \ltsq{d_i}{d_o} \psi$ between error domain morphisms, which are
+defined similarly, with no extra interaction with the algebraic structure.
+%
+The free composition of error domain relations has the universal property that
+$\phi \ltsq{d \relcomp d'}{d''} \phi'$ if and only if for any $x \binrel{d} y$
+and $y \binrel{d'} z$, we have $\phi(x) \binrel{d''} \phi'(z)$.
 

### 2024-10-25 paper-new/concrete-relational-model.tex
clarify the use of clocks in the relational adequacy section

@@
 
-To that end, we define a version of our denotation of term precision
+To that end, we define, using clock quantification, a version of our denotation of term precision
 that relates terms of type $\li^{gl} \mathbb{N}$. More concretely, consider the
@@
 %
+In these two definitions, we apply the clock $k$ to the elements $x$ and $y$ of
+type $\li^{gl} \mathbb{N} := \forall (k : \Clock). \li^k \mathbb{N}$.
 

### 2024-10-25 paper-new/intro.tex
small change in intro

@@
 simple gradually typed lambda calculus to \emph{guarded domain
-theory}. Our goal is for this model to scale to dynamic type tag
+theory}, a task that already involves nontrivial issues.
+Our goal is for this model to scale to dynamic type tag
 generation and higher-order references in future work.

### 2024-10-25 paper-new/concrete-term-model.tex
fix the footnote

@@
 types. These will be interpreted in a category of sets equipped with
-``algebraic''\footnote{the $\theta$ structure is not algebraic in the
-strictest sense since it does not have finite arity, but } structure
-with morphisms being functions that are homomorphic in this structure.
-This algebraic structure needs to model the effects present in our language,
-in this case these are errors as well as taking computational steps
-(potentially diverging by taking computational steps forever).
+algebraic\footnote{the $\theta$ structure is not algebraic in the
+strictest sense since it does not have finite arity} structure with
+morphisms being functions that are homomorphic in this structure.
+This algebraic structure needs to model the effects present in our
+language, in this case these are errors as well as taking
+computational steps (potentially diverging by taking computational
+steps forever).
 %

### 2024-10-25 paper-new/concrete-term-model.tex
fix the footnote better

@@
 types. These will be interpreted in a category of sets equipped with
-algebraic\footnote{the $\theta$ structure is not algebraic in the
-strictest sense since it does not have finite arity} structure with
+algebraic\footnote{The $\theta$ structure is not algebraic in the
+strictest sense since it does not have finite arity.} structure with
 morphisms being functions that are homomorphic in this structure.

### 2024-10-25 paper-new/concrete-relational-model.tex
small wording change in concrete relational model section

@@
 precision $cc'$ is simply the composition of the projections. In the new
-semantics, the corresponding projection involves a perturbation. The reason for
-this is technical and is a consequence of the construction of the composition of
-value/computation relations. (The relevant projection is defined in the proof of
-Lemma \ref{lem:representation-comp-F-U} in the Appendix.) However, the two
-projections are extensionally equivalent since they are the same up to a
-perturbation.
+semantics, the corresponding projection involves a perturbation. This is a
+consequence of the construction of the composition of value/computation
+relations (the relevant projection is defined in the proof of Lemma
+\ref{lem:representation-comp-F-U} in the Appendix). However, the two projections
+are extensionally equivalent since they are the same up to a perturbation.
 

### 2024-10-25 paper-new/appendix.tex
paper revisions
latest changes

--- 2024-07-12 paper-new/appendix.tex
@@
-\section{Gradual Typing Syntax}
+\section{Gradual Typing Syntax}\label{sec:appendix-gtlc-syntax}
 
@@
 
-\subsection{Free Composition of Error Domain Relations}
-\begin{definition}\label{def:free-comp-ed-rel}
-  Let $d : B_1 \rel B_2$ and $d' : B_2 \rel B_3$ be error domain relations. We
-  define the composition $dd'$ inductively by the following rules:
-  \begin{mathpar}
-      \inferrule*[right = Comp]
-      {b_1 \mathbin{d} b_2 \and b_2 \mathbin{d'} b_3}
-      {b_1 \mathbin{d \relcomp d'} b_3}
-
-      \inferrule*[right = DnClosed]
-      {b_1' \le_{B_1} b_1 \and b_1 \mathbin{d \relcomp d'} b_3}
-      {b_1' \mathbin{d \relcomp d'} b_3}
-
-      \inferrule*[right = UpClosed]
-      {b_1 \mathbin{d \relcomp d'} b_3 \and b_3 \le_{B_3} b_3'}
-      {b_1 \mathbin{d \relcomp d'} b_3'}
-
-      \inferrule*[right = PresErr]
-      { }
-      {\mho_{B_1} \mathbin{d \relcomp d'} b_3}
-
-      \inferrule*[right = PresTheta]
-      {\later_t( \tilde{b_1} \mathbin{d \relcomp d'} \tilde{b_3} ) }
-      {\theta_{B_1}(\tilde{b_1}) \mathbin{d \relcomp d'} \theta_{B_3}(\tilde{b_3}) }
-  \end{mathpar}
-
-  This has the universal property that $\phi \ltsq{d \relcomp d'}{d''}
-  \phi'$ if and only if for any $b \binrel{d} b'$ and $b' \binrel{d'}
-  b''$ that $\phi(b) \binrel{d''} \phi'(b'')$.
-\end{definition}
+% \subsection{Free Composition of Error Domain Relations}
+% \begin{definition}\label{def:free-comp-ed-rel}
+%   Let $d : B_1 \rel B_2$ and $d' : B_2 \rel B_3$ be error domain relations. We
+%   define the composition $dd'$ inductively by the following rules:
+%   \begin{mathpar}
+%       \inferrule*[right = Comp]
+%       {b_1 \mathbin{d} b_2 \and b_2 \mathbin{d'} b_3}
+%       {b_1 \mathbin{d \relcomp d'} b_3}
+
+%       \inferrule*[right = DnClosed]
+%       {b_1' \le_{B_1} b_1 \and b_1 \mathbin{d \relcomp d'} b_3}
+%       {b_1' \mathbin{d \relcomp d'} b_3}
+
+%       \inferrule*[right = UpClosed]
+%       {b_1 \mathbin{d \relcomp d'} b_3 \and b_3 \le_{B_3} b_3'}
+%       {b_1 \mathbin{d \relcomp d'} b_3'}
+
+%       \inferrule*[right = PresErr]
+%       { }
+%       {\mho_{B_1} \mathbin{d \relcomp d'} b_3}
+
+%       \inferrule*[right = PresTheta]
+%       {\later_t( \tilde{b_1} \mathbin{d \relcomp d'} \tilde{b_3} ) }
+%       {\theta_{B_1}(\tilde{b_1}) \mathbin{d \relcomp d'} \theta_{B_3}(\tilde{b_3}) }
+%   \end{mathpar}
+
+%   This has the universal property that $\phi \ltsq{d \relcomp d'}{d''}
+%   \phi'$ if and only if for any $b \binrel{d} b'$ and $b' \binrel{d'}
+%   b''$ that $\phi(b) \binrel{d''} \phi'(b'')$.
+% \end{definition}
 
@@
 
+% \subsection{Functorial Actions on Syntactic Perturbations}
+
+% We provide details on the definition of the homomorphisms specifying the
+% interpretation of perturbations corresponding for the functors $\li$, $U$,
+% $\times$, and $\arr$.
+
+% \begin{itemize}
+%   \item To define the homomorphism $i_{\li A}$, by the universal property
+%   of the coproduct of monoids it suffices to define two homomorphisms, one from
+%   $\mathbb{N} \to \morbisimid{\li A}$ and one from $M_A \to \morbisimid{\li A}$.
+%   The first homomorphism sends the generator $1 \in \mathbb{N}$ to the error
+%   domain endomorpism $(\delta \circ \eta)^\dagger$, and the second sends $m_A \in
+%   M_A$ to $\li(i_A(m_A))$ and we observe in both cases that the resulting
+%   endomorphisms are bisimilar to the identity since the action of $\li$ on
+%   morphisms preserves bisimilarity.
+
+%   \item To define the homomorphism $i_{A_1 \times A_2}$, we
+%   appeal to the universal property and give homomorphisms $M_{A_1} \to
+%   \morbisimid{A_1 \times A_2}$ and $M_{A_2} \to \morbisimid{A_1 \times A_2}$. The
+%   former applies $i_{A_1}$ in the first component and the identity in the second
+%   component, while the latter does the reverse with $i_{A_2}$.
+
+%   \item To define the homomorphism $i_{A \arr B}$ it suffices to define a
+%   homomorphism $M_A^{op} \to \morbisimid{A \arr B}$ and $M_B \to \morbisimid{A
+%   \arr B}$. The former takes an element $m_A$ and a predomain morphism $f : A \to
+%   UB$ and returns the composition $f \circ i_A(m_A)$, while the latter is defined
+%   similarly by post-composition with $i_B$.
+
+% \end{itemize}
+
 \subsection{Lemmas about Perturbations}

### 2024-11-10 paper-new/concrete-term-model.tex
small edits

@@
 $F$ defined above. This means $\li^{gl} \mathbb{N}$ is isomorphic to
-$\delay(\mathbb{N} + {\mho})$. Denote this isomorphism by $\alpha$.
+$\delay(\mathbb{N} + {\mho})$. We write $\alpha$ to denote this isomorphism.
 
 Given $d : \delay(\mathbb{N} + {\mho})$, we define a notion of termination in
-$i$ steps, written $d \da^i$. We define $d \da^i$ inductively by the rules
-$(\tnow\, n_?) \da^0$ and $(\tlater\, d) \da^{i+1}$ if $d \da^i$ (here, $n_? :
-\mathbb{N} + {\mho}$). From this definition, it is clear that if there is an $i$
-such that $d \da^i$, then this $i$ is unique. 
+$i$ steps with result $n_? \in \mathbb{N} + {\mho}$, written $d \da^i n_?$.
+We define $d \da^i n_?$ inductively by the rules
+$(\tnow\, n_?) \da^0 n_?$ and $(\tlater\, d) \da^{i+1} n_?$ if $d \da^i n_?$.
+From this definition, it is clear that if there is an $i$
+such that $d \da^i n_?$, then this $i$ is unique. 
 % Furthermore, from the proof that

### 2024-11-10 paper-new/intro.tex
add citation for Clocked Cubical Type Theory in the intro
small edits

--- 2024-10-24 paper-new/intro.tex
@@
 work, we provide first steps towards this goal by adapting work on a
-simple gradually typed lambda calculus to \emph{guarded domain
-theory}. Our goal is for this model to scale to dynamic type tag
+simple gradually typed lambda calculus to the setting of \emph{guarded domain
+theory}, a task that already involves nontrivial issues.
+Our goal is for this model to scale to dynamic type tag
 generation and higher-order references in future work.
 
-\emph{Guarded} domain theory is the main denotational alternative to
-classical domain theory that can successfully model these advanced
-features. While classical domain theory is based on modeling types as
+Guarded domain theory is the main denotational alternative to
+classical domain theory that can successfully model the advanced
+features we aim to support. While classical domain theory is based on modeling types as
 ordered sets with certain joins, guarded domain theory is based on an
@@
 ordinary mathematics. We incorporate this by using
-\emph{clock-quantification}\max{Eric should provide a citation for Clocked cubical here since that's close to our actual approach} \cite{atkey-mcbride2013} in our type
-theory. In particular, the guarded type theory in which we work has a
-notion of clocks that index all of our guarded constructions. This is
+\emph{clock-quantification} \cite{atkey-mcbride2013, kristensen-mogelberg-vezzosi2022}
+% \max{Eric should provide a citation for Clocked cubical here since that's close to our actual approach}
+in our type theory. In particular, the guarded type theory in which we work has
+a notion of clocks that index all of our guarded constructions. This is
 discussed in more detail in Section \ref{sec:big-step-term-semantics}.

### 2024-11-12 paper-new/concrete-relational-model.tex
some small writing tweaks to big-step term semantics and relational adequacy sections

@@
 %
-Then to obtain the desired result, it suffices to show being related in
+Then to obtain the desired result, it suffices to show that being related in
 $\ltsqbisim{}{}^\text{Delay}$ implies that the corresponding partial elements of
-$\mathbb{N} + {\mho}$ are related in the expected manner. That is, if $d
-\ltsqbisim{}{}^\text{Delay} d'$, then $p(d) \ltdyn^{Partial} p(d')$ where $p$ is
-a partial function from delays into $\mathbb{N} + {\mho}$ defined by existential
-quantification over the number of steps $i$ in which $d$ terminates, and the
-relation $\ltdyn^\text{Partial}$ is defined such that $\mho$ is the least
-element, $\bot \ltdyn^\text{Partial} \bot$, and $n \ltdyn^\text{Partial} n$.
-This implication follows from the definition of $\ltsqbisim{}{}^\text{Delay}$.
+$\mathbb{N} + {\mho}$ are related in the expected manner. Specifically, recall
+from Section \ref{sec:big-step-term-semantics} that $-\Downarrow^\text{Delay}$
+is a partial function from delays to $\mathbb{N} + {\mho}$ defined by
+existential quantification over the number of steps in which the argument terminates.
+We want to show that if $d \ltsqbisim{}{}^\text{Delay} d'$, then 
+%
+$(d \Downarrow^\text{Delay}) \ltdyn^\text{Partial}\, (d' \Downarrow^\text{Delay})$,
+%
+where the relation $\ltdyn^\text{Partial}$ is defined such that $\mho$ is the
+least element, $\bot \ltdyn^\text{Partial} \bot$, and $n \ltdyn^\text{Partial}
+n$. This implication follows from the definition of
+$\ltsqbisim{}{}^\text{Delay}$.
 

### 2024-11-12 paper-new/concrete-term-model.tex
small fixes to cast semantics figure
some small writing tweaks to big-step term semantics and relational adequacy sections

@@
     \sem{\dnc{(c_i \ra c_o)}} &= U(\sem{\upc {c_i}} \to \sem{\dnc {c_o}})\\
-    \sem{\dnc{(c_1 \times c_2)}} &= (\dnc{c_1} \timesk A_2) \circ (A_1' \timesk \dnc{c_2}) \\
+    \sem{\dnc{(c_1 \times c_2)}} &= (\sem{\dnc{c_1}} \timesk \sem{A_2}) \circ (\sem{A_1'} \timesk \sem{\dnc{c_2}}) \\
     \sem{\dnc{\inat}} &= (
-      \lambda V_d . \text{case $({V_d})$ of}
+      \lambda V_d . \text{case $({V_d})$ of }
       \{ \inat\,n \to \eta\, n
@@
     \sem{\dnc{\itimes}} &= (
-      \lambda V_d . \text{case $({V_d})$ of}
+      \lambda V_d . \text{case $({V_d})$ of }
       \{ \itimes(d_1, d_2) \to \eta\, (d_1, d_2)
@@
     \sem{\dnc{\iarr}} &= (
-      \lambda V_d . \text{case $({V_d})$ of}
+      \lambda V_d . \text{case $({V_d})$ of }
       \{ \iarr \tilde{f} \to \theta_t\, (\eta (\tilde{f}_t))
@@
 which we write as $\text{Delay}(Y)$, is defined as the coinductive type
-generated by $\tnow : Y \to \delay(Y)$ and $\tlater : \delay(Y) \to \delay(X)$.
+generated by $\tnow : Y \to \delay(Y)$ and $\tlater : \delay(Y) \to \delay(Y)$.
 We have that $\delay (\mathbb{N} + {\mho})$ is a final coalgebra for the functor
@@
 %
-$f \colon \delay (\mathbb{N} + {\mho}) \rightharpoonup \mathbb{N} + {\mho}$
+$-\Downarrow^\text{Delay} \colon \delay (\mathbb{N} + {\mho}) \rightharpoonup \mathbb{N} + {\mho}$
 %
 by existentially quantifying over $i$ for which $d \da^i n_?$. That is, we
-define $f(d) := n_?$ if $d \da^i n_?$ for some (necessarily unique) $i$, and $f$
-is undefined otherwise.
+define $d \Downarrow^\text{Delay} \,= n_?$ if $d \da^i n_?$ for some (necessarily
+unique) $i$, and $d \Downarrow^\text{Delay}$ is undefined otherwise.
 
@@
 $\sem{M}^{gl} := \lambda (k : \Clock) .\sem{M}$. Composing this with the
-isomorphism $\alpha$ and the partial function $f$ yields the desired partial
+isomorphism $\alpha : \li^{gl} \mathbb{N} \cong \delay(\mathbb{N} + {\mho})$ and
+the partial function $-\Downarrow^\text{Delay}$ yields the desired partial
 function

### 2024-11-13 paper-new/concrete-relational-model.tex
a couple more minor edits

@@
 \rel A'$ be a predomain relation. We say that $c$ is
-\emph{quasi-left-representable} by a predomain morphism $e : A \to A'$ if there
+\emph{quasi-left-representable} by a predomain morphism $e_c : A \to A'$ if there
 are perturbations $\delle_c \in M_A$ and $\delre_c \in M_{A'}$ such that the
-following two squares exist: (1) $\upl$: $e \ltsq{c}{\ltdyn_{A'}} i_{A'}(\delre_c)$, and
-(2) $\upr$: $i_A(\delle_c) \ltsq{\ltdyn_{A}}{c} e$.
+following two squares exist: 
+(1) $\upl$: $e_c \ltsq{c}{r(A')} i_{A'}(\delre_c)$, and
+(2) $\upr$: $i_A(\delle_c) \ltsq{r(A)}{c} e_c$.
+% If $c$ is quasi-left-representable by $e_c$, we call $e_c$ an \emph{embedding} morphism.
+
+Quasi-left-representability for error domain relations is defined analogously.
 \end{definition}
@@
 %
-We make the analogous definition of quasi-left-representability for error domain
-relations $d$.
+% We make the analogous definition of quasi-left-representability for error domain
+% relations $d$.
 
@@
 $d : B \rel B'$ be an error domain relation. We say that $d$ is
-\emph{quasi-right-representable} by an error domain morphism $p : B' \multimap B$ if
+\emph{quasi-right-representable} by an error domain morphism $p_d : B' \multimap B$ if
 there are perturbations $\dellp_d \in M_B$ and $\delrp_d \in M_{B'}$ such that
 the following two squares exist: 
-(1) $\dnl$: $p \ltsq{\ltdyn_{B'}}{d} i_{B'}(\delrp_d)$ and
-(2) $\dnr$: $i_B(\dellp_d) \ltsq{d}{\ltdyn_{B}} p$.
+(1) $\dnl$: $p_d \ltsq{r(B')}{d} i_{B'}(\delrp_d)$, and
+(2) $\dnr$: $i_B(\dellp_d) \ltsq{d}{r(B)} p_d$.
+% If $d$ is quasi-right-representable by $p_d$, we call $p_d$ a \emph{projection} morphism.
+
 Quasi-right-representability for predomain relations is defined similarly.

### 2024-11-13 paper-new/intro.tex
a couple more minor edits

@@
 transitivity, and we argue in Section \ref{sec:towards-relational-model} that
-some amount of transitive reasoning is nessecary for defining a
+some amount of transitive reasoning is necessary for defining a
 syntax-independent model of gradual typing. As a result, the lack of

### 2024-11-18 paper-new/intro.tex
small wording change in intro

@@
 are explicit synchronizations that we manipulate syntactically. The combination
-of lock-step reasoning with perturbations is transitive and is able to account
+of step-sensitive reasoning with perturbations is transitive and is able to account
 for the step-insensitive nature of the graduality property.

### 2024-11-20 paper-new/concrete-term-model.tex
small wording changes in section on big-step term semantics

@@
 
-To begin, we must first discuss another aspect of guarded type theory,
-\emph{clocks}, originally introduced in \cite{atkey-mcbride2013}. The
+To begin, we must first discuss another aspect of guarded type theory. The
 constructions of guarded type theory ($\later$, $\nxt$, and $\fix$) we have been
-using may be indexed by objects called \emph{clocks}. A clock intuitively serves
+using may be indexed by objects called \emph{clocks} \cite{atkey-mcbride2013}. A clock intuitively serves
 as a reference relative to which steps are counted. For instance, given a clock
@@
 coinductive types using guarded recursion. That is, given a type $A$ with a free
-variable $k : \Clock$, we can form the type $A^{gl} := \forall k. A$ (here, $gl$ stands for ``global''). Given a
-functor $F$ that satisfies a technical condition, if $A$ is a guarded recursive
-type satisfying $A \cong F(\later^k A)$, then the type $A^{gl}$ has a final
-$F$-coalgebra structure (see Theorem 4.3 of
+variable $k : \Clock$, we can form the type $A^{gl} := \forall k. A$ (here, $gl$ stands for ``global'').
+%
+A functor $F \colon \mathsf{Type} \to \mathsf{Type}$ is said to \emph{commute
+with clock quantification} if for all $X : \Clock \to \mathsf{Type}$, the
+canonical map $F(\forall k. X_k) \to \forall k. F(X_k)$ is an equivalence \cite{kristensen-mogelberg-vezzosi2022},
+where we write $X_k$ for the application of clock $k$ to $X$.
+%
+Given a functor $F$ that commutes with clock quantification, if $A$ is a guarded
+recursive type satisfying $A \cong F(\later^k A)$, then the type $A^{gl}$ has a
+final $F$-coalgebra structure (see Theorem 4.3 of
 \cite{kristensen-mogelberg-vezzosi2022}).
@@
 %
-Furthermore, it can be shown that the functor $F$ satisfies the technical
-condition in the statement of the above-mentioned theorem. Thus, the theorem
-implies that the type $\li^{gl} \mathbb{N} := \forall k. \li^k \mathbb{N}$ is a
-final $F$-coalgebra. In fact, $\li^{gl}$ is isomorphic to a more familiar coninductive type.
+Furthermore, it can be shown that the functor $F$ commutes with clock
+quantification. Thus, the type $\li^{gl} \mathbb{N} := \forall k. \li^k
+\mathbb{N}$ is a final $F$-coalgebra. In fact, $\li^{gl}$ is isomorphic to a
+more familiar coninductive type.
 %

### 2024-11-20 paper-new/concrete-relational-model.tex
some small tweaks

@@
 %
-By adapting Theorem 4.3 of \cite{kristensen-mogelberg-vezzosi2022} to the
+By adapting Theorem 4.3 of \cite{kristensen-mogelberg-vezzosi2022} \max{citations are not nouns} to the
 setting of inductively-defined relations, we can show that both the global

### 2024-11-20 paper-new/towards-relational-model.tex
add brief sketch of proof of no-go theorem to the body of the paper
some small tweaks

@@
 \begin{proof}
-  By L\"ob induction, in the appendix.
+  By L\"ob induction: we assume that $l$ is related to $\Omega$ later, which implies that
+  $\theta (\nxt\, l) \binrel{R} \theta (\nxt\, \Omega)$. Then observe that 
+  $l \binrel{R} \theta (\nxt\, l) \binrel{R} \theta (\nxt\, \Omega) = \Omega$.
+  See the appendix for more details.
 \end{proof}
@@
 providing a poset semantics to a semantics where types are equipped
-with merely a reflexive relation, and continue using the
+with merely a reflexive relation and continue using the
 step-insensitive ordering. However, some level of transitivity is
@@
 
-We call such a relation between $A$ and $A'$ a \emph{poset relation}, and we
-write $c : A \rel A'$. The paradigmatic example of a poset relation is given by
-the ordering $\ltdyn_A$. We call this the reflexive poset relation and
-denote it by $r(A) : A \rel A$. This is a poset relation because the ordering
-$\ltdyn_A$ is transitive.
-
+We call such a relation between $A$ and $A'$ a \emph{poset relation},
+and we write $c : A \rel A'$ to mean $c$ is a poset relation downward
+closed in $A$ and upper closed in A'. The paradigmatic example of a
+poset relation is given by the ordering $\ltdyn_A$. We call this the
+reflexive poset relation and denote it by $r(A) : A \rel A$. This is a
+poset relation because the ordering $\ltdyn_A$ is transitive.
 % The paradigmatic example of a poset relation is the ``reflexive'' relation,
 % which is given by the ordering $\ltdyn_A$ and denoted by $r(A)$.
-
 %
@@
 
-These relations are then used to model term precision. For closed programs $M
-\ltdyn M' : c$ where $c : A \ltdyn A'$, the terms denote elements of the free
-error domain $\sem{M},\sem{M'} : U\li \sem{A}$, and we model the term precision
-semantically as the relation holding between those denoted elements $\sem{M}
-\binrel{(U\li\sem{c})} \sem{M'}$. Here $U\li\sem{c}$ is a relational lifting of
-$U\circ\li$, which extends $\sem{c}$ to account for error and computational
-steps.
+Poset relations are then used in the semantics of \emph{term}
+precision. For closed programs $M \ltdyn M' : c$ where $c : A \ltdyn
+A'$, the terms denote elements of the free error domain
+$\sem{M},\sem{M'} : U\li \sem{A}$, and we model the term precision
+semantically as the relation holding between those denoted elements
+$\sem{M} \binrel{(U\li\sem{c})} \sem{M'}$. Here $U\li\sem{c}$ is a
+relational lifting of $U\circ\li$, which extends $\sem{c}$ to account
+for error and computational steps.
 %

### 2024-11-21 paper-new/concrete-term-model.tex
rename "linear morphism" to "homomorphism"

@@
 between a category $\mathcal V$ of ``value types'' and pure morphisms
-and a category $\mathcal E$ of ``computation types'' and linear
-morphisms. The monad $T$ on $\calV$ is decomposed into a left adjoint
+and a category $\mathcal E$ of ``computation types'' and homomorphisms.
+The monad $T$ on $\calV$ is decomposed into a left adjoint
 $F : \calV \to \calE$ constructing the type $F A$ of computations that
-return $A$ values and and right adjoint $U : \calE \to \calV$
+return $A$ values and right adjoint $U : \calE \to \calV$
 constructing the type $U B$ of values that are first-class suspended
 computations of type $B$. The monad $T$ can then be reconstituted as
-$T = UF$ but the main advantage is that many constructions on
+$T = UF$, but the main advantage is that many constructions on
 effectful programs naturally decompose into constructions on
@@
 that the casts of gradual typing are naturally formulated in terms of
-pairs of a pure morphism and a linear morphism, as we explain below
+pairs of a pure morphism and a homomorphism, as we explain below
 \cite{new-licata-ahmed2019}.
@@
 intuition is that these are the ``effectful'' morphisms between
-value/computation types, whereas the morphisms of $\calV/\calE$ are
-``pure''/``linear''. Then every CBPV type constructor $F,U,\times,\to$
-has in addition to its usual functorial action on pure/linear
-morphisms, an action in each argument on Kleisli morphisms. So for
-instance, the function construction is functorial in pure/linear
-morphisms $\to : \op{\calV} \times \calE \to \calE$ but on Kleisli
+value/computation types, whereas the morphisms of $\calV$ and $\calE$ are
+pure morphisms and homomorphisms, respectively. Then every CBPV type constructor $F,U,\times,\to$
+has in addition to its usual functorial action on pure morphisms/homomorphisms
+an action in each argument on Kleisli morphisms. So for
+instance, the function construction is functorial in pure morphisms/homomorphisms
+$\to : \op{\calV} \times \calE \to \calE$ but on Kleisli
 morphisms what we have is for each computation type $B$ a functorial

### 2024-11-21 paper-new/concrete-relational-model.tex
rename "extensional" to "bisimilarity-preserving"

@@
 $x \ltdyn_A x'$, then $f(x) \ltdyn_{A'} f(x')$), and
-\emph{extensional} (if $x \bisim_A x'$, then $f(x) \bisim_{A'} f(x')$).
+\emph{bisimilarity-preserving} (if $x \bisim_A x'$, then $f(x) \bisim_{A'} f(x')$).
 \end{definition}
@@
 structure: the error element should be the least element (so that it
-models graduality), the think map should be monotone and extensional
+models graduality), the think map should be monotone and bisimilarity-preserving,
 and the bisimilarity relation should relate $\delta$ to the identity

### 2024-11-21 paper-new/concrete-relational-model.tex
remove \odot from notation for the free composition of error domain relations

@@
 particular is not preserved), but we instead use an alternative ``free
-composition'' $d \odot d' : B \rel B''$ that is inductively generated from the
+composition'' $dd' : B \rel B''$ that is inductively generated from the
 relational composition. The rules are given in Figure \ref{fig:free-comp-ed-rel}.
@@
     {x \mathbin{d} y \and y \mathbin{d'} z}
-    {x \mathbin{(d \relcomp d')} z}
+    {x \mathbin{(dd')} z}
 
     \inferrule
-    {x' \ltdyn_{B_1} x \and x \mathbin{(d \relcomp d')} z}
-    {x' \mathbin{(d \relcomp d')} z}
+    {x' \ltdyn_{B_1} x \and x \mathbin{(dd')} z}
+    {x' \mathbin{(dd')} z}
 
     \inferrule
-    {x \mathbin{(d \relcomp d')} z \and z \ltdyn_{B_3} z'}
-    {x \mathbin{(d \relcomp d')} z'}
+    {x \mathbin{(dd')} z \and z \ltdyn_{B_3} z'}
+    {x \mathbin{(dd')} z'}
 
@@
     { }
-    {\mho_{B_1} \mathbin{(d \relcomp d')} z}
+    {\mho_{B_1} \mathbin{(dd')} z}
 
     \inferrule
-    {\later_t[ \tilde{x}_t \mathbin{(d \relcomp d')} \tilde{z}_t ] }
-    {\theta_{B_1}(\tilde{x}) \mathbin{(d \relcomp d')} \theta_{B_3}(\tilde{z}) }
+    {\later_t[ \tilde{x}_t \mathbin{(dd')} \tilde{z}_t ] }
+    {(\theta_{B_1}\, \tilde{x}) \mathbin{(dd')} (\theta_{B_3}\, \tilde{z}) }
   \end{mathpar}
@@
 The free composition of error domain relations has the universal property that
-$\phi \ltsq{d \relcomp d'}{d''} \phi'$ if and only if for any $x \binrel{d} y$
+$\phi \ltsq{dd'}{d''} \phi'$ if and only if for any $x \binrel{d} y$
 and $y \binrel{d'} z$, we have $\phi(x) \binrel{d''} \phi'(z)$.

### 2024-11-21 paper-new/concrete-relational-model.tex
small writing tweaks

@@
 %
-By adapting Theorem 4.3 of \cite{kristensen-mogelberg-vezzosi2022} \max{citations are not nouns} to the
+By adapting Theorem 4.3 of Kristensen et al. \cite{kristensen-mogelberg-vezzosi2022} to the
 setting of inductively-defined relations, we can show that both the global

### 2024-11-21 paper-new/concrete-term-model.tex
small writing tweaks

@@
 variable $k : \Clock$, we can form the type $A^{gl} := \forall k. A$ (here, $gl$ stands for ``global'').
-%
+
 A functor $F \colon \mathsf{Type} \to \mathsf{Type}$ is said to \emph{commute
@@
 recursive type satisfying $A \cong F(\later^k A)$, then the type $A^{gl}$ has a
-final $F$-coalgebra structure (see Theorem 4.3 of
-\cite{kristensen-mogelberg-vezzosi2022}).
+final $F$-coalgebra structure (see Theorem 4.3 of Kristensen et al. \cite{kristensen-mogelberg-vezzosi2022}).
 

### 2024-11-21 paper-new/towards-relational-model.tex
small writing tweaks

@@
 and we write $c : A \rel A'$ to mean $c$ is a poset relation downward
-closed in $A$ and upper closed in A'. The paradigmatic example of a
+closed in $A$ and upward closed in A'. The paradigmatic example of a
 poset relation is given by the ordering $\ltdyn_A$. We call this the

### 2024-11-21 paper-new/paper.tex
more paper revisions
prepare paper.tex for diff
latest changes
small edits
add orcid for Max and Tingting
add Eric's orcid
add POPL-specific metadata

@@
-\documentclass[acmsmall,anonymous,review,screen]{acmart}
+\documentclass[acmsmall]{acmart}
 \let\Bbbk\relax
@@
 
+\theoremstyle{remark}
 \newtheorem{remark}{Remark}
+\newenvironment{DIFnomarkup}{}{}
+\pagestyle{empty}
+\settopmatter{printfolios=false}
+
+%% CCS Concepts
+\begin{CCSXML}
+  <ccs2012>
+     <concept>
+         <concept_id>10003752.10010124.10010131.10010133</concept_id>
+         <concept_desc>Theory of computation~Denotational semantics</concept_desc>
+         <concept_significance>500</concept_significance>
+         </concept>
+     <concept>
+         <concept_id>10011007.10011006.10011039.10011311</concept_id>
+         <concept_desc>Software and its engineering~Semantics</concept_desc>
+         <concept_significance>300</concept_significance>
+         </concept>
+     <concept>
+         <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
+         <concept_desc>Software and its engineering~Functional languages</concept_desc>
+         <concept_significance>300</concept_significance>
+         </concept>
+   </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Theory of computation~Denotational semantics}
+\ccsdesc[300]{Software and its engineering~Semantics}
+\ccsdesc[300]{Software and its engineering~Functional languages}
+
+%% Additional Key Words and Phrases
+\keywords{Gradual Typing, Denotational Semantics, Synthetic Guarded Domain Theory, Guarded Type Theory}
+
+%%% The following is specific to POPL '25 and the paper
+%%% 'Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory'
+%%% by Eric Giovannini, Tingting Ding, and Max S. New.
+%%%
+\setcopyright{rightsretained}
+\acmDOI{10.1145/3704863}
+\acmYear{2025}
+\acmJournal{PACMPL}
+\acmVolume{9}
+\acmNumber{POPL}
+\acmArticle{27}
+\acmMonth{1}
+\received{2024-07-11}
+\received[accepted]{2024-11-07}
 
@@
 %% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2018}
-\acmYear{2018}
-\acmDOI{10.1145/1122445.1122456}
+
+% \setcopyright{acmcopyright}
+% \copyrightyear{2018}
+% \acmYear{2018}
+% \acmDOI{10.1145/1122445.1122456}
 
 %% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
-  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
-\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-  June 03--05, 2018, Woodstock, NY}
-\acmPrice{15.00}
-\acmISBN{978-1-4503-XXXX-X/18/06}
+% \acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
+%   Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
+% \acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
+%   June 03--05, 2018, Woodstock, NY}
+% \acmPrice{15.00}
+% \acmISBN{978-1-4503-XXXX-X/18/06}
 
@@
 \email{ericgio@umich.edu}
+\orcid{0009-0003-6871-1714}
 
@@
 \email{tingtind@umich.edu}
+\orcid{0009-0000-5676-1886}
 
@@
 \email{maxsnew@umich.edu}
+\orcid{0000-0001-8141-195X}
 
@@
   constructions we use.
-%%     We develop a denotational semantics for a simple gradually typed language
-%%     that is adequate and proves the graduality theorem.
-%%     %
-%%     The denotational semantics is constructed using \emph{synthetic
-%%     guarded domain theory} working in a type theory with a later
-%%     modality and clock quantification.
-%%     %
-%%     This provides a remarkably simple presentation of the semantics,
-%%     where gradual types are interpreted as ordinary types in our ambient
-%%     type theory equipped with an ordinary preorder structure to model
-%%     the error ordering.
-%%     %
-%%     This avoids the complexities of classical domain-theoretic models
-%%     (New and Licata) or logical relations models using explicit
-%%     step-indexing (New and Ahmed).
-%%     %
-%%     In particular, we avoid a major technical complexity of New and
-%%     Ahmed that requires two logical relations to prove the graduality
-%%     theorem.
-
-%%     By working synthetically we can treat the domains in which gradual
-%%     types are interpreted as if they were ordinary sets. This allows us
-%%     to give a ``na\"ive'' presentation of gradual typing where each
-%%     gradual type is modeled as a well-behaved subset of the universal
-%%     domain used to model the dynamic type, and type precision is modeled
-%%     as simply a subset relation.
-%%     %
 \end{abstract}
@@
 \bibliographystyle{ACM-Reference-Format}
+\begin{DIFnomarkup}
 \bibliography{references}
+\end{DIFnomarkup}
 
+\newpage
 \appendix

### 2024-11-22 paper-new/concrete-relational-model.tex
fix third requirement in definition of error domain to match the Agda development
add citation info for extended version to bibliography; replace each reference to the appendix with a citation of extended version

@@
     \item A morphism of predomains $\theta_B \colon \later B \to B$
-    \item For all $x : B$ we have $\theta_B(\nxt\, x) \bisim_B x$
+    \item For all $x, y : B$ with $x \bisim_B y$, we have $\theta_B(\nxt\, x) \bisim_B y$
 \end{itemize}
@@
 precision $cc'$ is simply the composition of the projections. In the new
-semantics, the corresponding projection involves a perturbation. This is a
-consequence of the construction of the composition of value/computation
-relations (the relevant projection is defined in the proof of Lemma
-\ref{lem:representation-comp-F-U} in the Appendix). However, the two projections
-are extensionally equivalent since they are the same up to a perturbation.
+semantics, the corresponding projection involves a perturbation.  
+%
+% EXTENDED VERSION
+% This is a consequence of the construction of the composition of
+% value/computation relations (the relevant projection is defined in the proof
+% of Lemma \ref{lem:representation-comp-F-U} in the Appendix).
+%
+% CONFERENCE VERSION
+This is a consequence of the construction of the composition of
+value/computation relations (the relevant projection is defined in the appendix
+of the extended version of this paper \cite{gtt-sgdt-extended-version}).
+%
+However, the two projections are extensionally equivalent since they are the
+same up to a perturbation.
 
@@
 i.e., that the relations are quasi-representable and that they satisfy
-the push-pull property for perturbations. We include most of these technical
-verifications in the appendix (Definitions
-\ref{def:value-computation-rel-comp} and
-\ref{def:functorial-actions-on-relations}).
+the push-pull property for perturbations.
+%
+% EXTENDED VERSION
+% We include most of these technical verifications in the appendix (Definitions
+% \ref{def:value-computation-rel-comp} and
+% \ref{def:functorial-actions-on-relations}).
+%
+% CONFERENCE VERSION
+We include most of these technical verifications in the appendix of the extended
+version \cite{gtt-sgdt-extended-version}.
 
@@
 
+% EXTENDED VERSION
+% Lastly, the interpretation of the type precision equivalence rules $c \equiv c'$
+% as quasi-equivalence of the denoted relations $\sem{c} \bisim \sem{c'}$ is
+% justified by Lemma \ref{lem:quasi-order-equiv-functors} in the Appendix.
+%
+% CONFERENCE VERSION
 Lastly, the interpretation of the type precision equivalence rules $c \equiv c'$
 as quasi-equivalence of the denoted relations $\sem{c} \bisim \sem{c'}$ is
-justified by Lemma \ref{lem:quasi-order-equiv-functors} in the Appendix.
+justified in the appendix of the extended version
+\cite{gtt-sgdt-extended-version}.
 
@@
 
+% EXTENDED VERSION
+% Correspondingly, we define a version of the lock-step error ordering and weak
+% bisimilarity relation on $\delay(\mathbb{N} + {\mho})$ (see Section
+% \ref{sec:relations-on-delay-monad} in the appendix for the definitions).
+%
+% CONFERENCE VERSION
 Correspondingly, we define a version of the lock-step error ordering and weak
-bisimilarity relation on $\delay(\mathbb{N} + {\mho})$ 
-(see Section \ref{sec:relations-on-delay-monad} in the appendix for the definitions).
+bisimilarity relation on $\delay(\mathbb{N} + {\mho})$ (see the appendix of the
+extended version \cite{gtt-sgdt-extended-version} for the
+definitions).
+%
 % \max{do these intermediate details matter for the main body of the paper?}

### 2024-11-22 paper-new/concrete-relational-model.tex
spellcheck

@@
   % syntactic perturbations $\delta_1$ and $\delta_2$ we need a syntactic
-  % pertubration $\delta_{1,2}$ that is interpreted as the composition of the
+  % perturbation $\delta_{1,2}$ that is interpreted as the composition of the
   % semantic perturbations $i_A(\delta_1)$ and $i_A(\delta_2)$.
@@
 components of the model.
-% Interpretations of functions are different: one is arbitrary fucntions, the other is monotone functions that preserve bisimilarity
+% Interpretations of functions are different: one is arbitrary functions, the other is monotone functions that preserve bisimilarity
 % Interpretation of the projection is different since we add a delay
@@
 homomorphism sends the generator $1 \in \mathbb{N}$ to the error domain
-endomorpism $(\delta \circ \eta)^\dagger$, and the second sends $m_A \in M_A$ to
+endomorphism $(\delta \circ \eta)^\dagger$, and the second sends $m_A \in M_A$ to
 $\li(i_A(m_A))$. We observe in both cases that the resulting endomorphisms are

### 2024-11-22 paper-new/concrete-term-model.tex
add citation info for extended version to bibliography; replace each reference to the appendix with a citation of extended version
spellcheck

@@
 Much of the type structure of GTLC is simple to model: $\nat$ denotes
-the set of natural numbers and $\times$ denotes the cartesian product
+the set of natural numbers and $\times$ denotes the Cartesian product
 of sets. To model the CBV function type $A \rightharpoonup A'$ we use
@@
 \timesk f_r \circ f_l \timesk A_r) \neq (f_l \timesk A_r' \circ A_l
-\timesk f_r)$. The full definitions of the Kleisli actions are
-included in Appendix \ref{sec:kleisli-actions}.
+\timesk f_r)$.
+%
+% EXTENDED VERSION 
+% The full definitions of the Kleisli actions are included in Appendix
+% \ref{sec:kleisli-actions}.
+%
+% CONFERENCE VERSION
+%
+The full definitions of the Kleisli actions are included in the appendix of the
+extended version of this paper \cite{gtt-sgdt-extended-version}.
 
@@
 %% \to \li D$ and performs a case inspection on the sum type. If it is a natural
-%% number $n$, then we return $\eta\, n$. Otherwise, we return $\mho$, modelling
+%% number $n$, then we return $\eta\, n$. Otherwise, we return $\mho$, modeling
 %% the fact that a run-time error occurs.
@@
 \mathbb{N}$ is a final $F$-coalgebra. In fact, $\li^{gl}$ is isomorphic to a
-more familiar coninductive type.
+more familiar coinductive type.
 %

### 2024-11-22 paper-new/discussion.tex
remove comment in discussion
update citation
spellcheck

@@
 \subsection{Comparison of Denotational and Operational Approaches}
-\eric{Some of this could be folded into the previous section, but this was one of the
-big-picture questions from one of the reviewers.}
+% \eric{Some of this could be folded into the previous section, but this was one of the
+% big-picture questions from one of the reviewers.}
 
@@
 which requires us to provide each piece of supporting constructor. This step 
-and futher steps towards to the model construction are still 
+and further steps towards to the model construction are still 
 work-in-progress, but once it's finished, we will provide a complete 
@@
 extended to model effects besides error and stepping
-\cite{guarded-interaction-trees, probabilistic-fpc-guarded}.
+\cite{guarded-interaction-trees, recursion-probabilistic-choice-guarded}.
 We also aim to complete our Agda formalization and evolve it into a

### 2024-11-22 paper-new/intro.tex
mention the extended version at the end of the intro
spellcheck

@@
 \ref{sec:big-step-term-semantics}. The resulting term semantics is adequate, and
-furthemore it validates the equational theory, since equivalent terms in the
+furthermore it validates the equational theory, since equivalent terms in the
 equational theory denote equal terms in the big-step semantics.
@@
   denotational semantics in synthetic guarded domain theory for the
-  \emph{terms} of the gradual lambda calculus.  The model is adeqauate
+  \emph{terms} of the gradual lambda calculus.  The model is adequate
   and validates the equational theory, but it does not satisfy
@@
 \end{enumerate}
+
+\textbf{Extended version}: An extended version of this paper is available
+\cite{gtt-sgdt-extended-version}. It includes an appendix containing definitions
+and proofs elided from this version of the paper.

### 2025-01-07 paper-new/concrete-relational-model.tex
check in paper changes as of camera-ready submission on 11/22/24

@@
 \tilde{x'}_t)$, and likewise for bisimilarity.
+%
+% Given predomains $A_i$ and $A_o$ and morphisms $f, g : A_i \to A_o$, we define
+% $f \bisimsq{A_i}{A_o} g$ to be the natural extension of bisimilarity to
+% morphisms: given bisimilar inputs, the morphisms have bisimilar outputs.
 
@@
     \item For all $x, y : B$ with $x \bisim_B y$, we have $\theta_B(\nxt\, x) \bisim_B y$
+    % \item We have $\theta_B \circ \nxt \bisimsq{B}{B} \id$
 \end{itemize}
@@
 \[ D \cong \mathbb{N}\, + (D \times D)\, + \laterhs U(D \arr \li D) \]
+%
+% EXTENDED VERSION
+% The complete definition of the ordering $\ltdyn_D$ is provided in the appendix
+% for completeness.
+%
+% CONFERENCE VERSION
 The complete definition of the ordering $\ltdyn_D$ is provided in the appendix
-for completeness. We refer to the three injections as $e_\nat$, $e_\times$,
+of the extended version of this paper \cite{gtt-sgdt-extended-version}.
+%
+We refer to the three injections as $e_\nat$, $e_\times$,
 and $e_\to$, respectively.
@@
 error domains equipped with a notion of syntactic perturbation:
-\begin{definition}[value and computation objects + morphisms]
+\begin{definition}
   A \textbf{value object} consists of a predomain $A$, a monoid $M_A$ and a
@@
 %
-\begin{definition}[quasi-left-representable relations]\label{def:quasi-left-representable}
+\begin{definition}\label{def:quasi-left-representable}
 Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let $c : A
 \rel A'$ be a predomain relation. We say that $c$ is
-\emph{quasi-left-representable} by a predomain morphism $e_c : A \to A'$ if there
+\textbf{quasi-left-representable} by a predomain morphism $e_c : A \to A'$ if there
 are perturbations $\delle_c \in M_A$ and $\delre_c \in M_{A'}$ such that the
@@
 %
-\begin{definition}[quasi-right-representable relations]\label{def:quasi-right-representable}
+\begin{definition}\label{def:quasi-right-representable}
 Let $(B, M_B, i_B)$ and $(B', M_{B'}, i_{B'})$ be computation objects, and let
 $d : B \rel B'$ be an error domain relation. We say that $d$ is
-\emph{quasi-right-representable} by an error domain morphism $p_d : B' \multimap B$ if
+\textbf{quasi-right-representable} by an error domain morphism $p_d : B' \multimap B$ if
 there are perturbations $\dellp_d \in M_B$ and $\delrp_d \in M_{B'}$ such that
@@
 to the presence of perturbations. Instead we have the following weaker property:
-\begin{definition}[quasi-equivalence]\label{def:quasi-equivalent}
+\begin{definition}\label{def:quasi-equivalent}
   Let $A$ and $A'$ be value objects, and let $c, c' : A \rel A'$ be relations
@@
 %
-\begin{definition}[value relations]
+\begin{definition}
 A \textbf{value relation} between value objects $(A, M_A, i_A)$ and $(A',
@@
   Let $c_i : A_i \rel A_i'$, $c_o : A_o \rel A_o'$ $f : A_i \to A_o$
-  and $g : A_i' \to A_o'$.  An extensional square $f
+  and $g : A_i' \to A_o'$.  An \textbf{extensional square} $f
   \ltsqbisim{c_i}{c_o} g$ holds if there \emph{exist} $f'
@@
 value/computation relations (the relevant projection is defined in the appendix
-of the extended version of this paper \cite{gtt-sgdt-extended-version}).
+of the extended version \cite{gtt-sgdt-extended-version}).
 %

### 2025-01-07 paper-new/concrete-term-model.tex
check in paper changes as of camera-ready submission on 11/22/24

@@
 them from the error domains we define later to model graduality:
-\begin{definition}[Simple Error Domains]
+\begin{definition}
   A (simple) error domain $B$ consists of
@@
 guarded lift monad to accommodate the additional effect of failure.
-\begin{definition}[Free error domain]
+\begin{definition}
   For a set $A$, we define the (carrier of) \emph{free error domain} $U(\li A)$ as the unique solution to the guarded domain equation:

### 2025-01-07 paper-new/discussion.tex
cite our artifact's zenodo doi
check in paper changes as of camera-ready submission on 11/22/24

@@
 developed a partial formalization of our results in Guarded Cubical Agda
-\cite{veltri-vezzosi2020}.
+\cite{veltri-vezzosi2020}; the formalization is available online \cite{artifact}.
 %
@@
 
-The remaining formalization work is the following:
+As of this writing, the remaining formalization work is the following:
 \begin{enumerate}
@@
 benefits such as early error detection or reduced runtime
-overhead\cite{deepandshallowtypes,new-licata-ahmed2019}. It would be a
+overhead \cite{deepandshallowtypes,new-licata-ahmed2019}. It would be a
 good test of the generality of our semantic framework to see if it can

### 2025-01-07 paper-new/intro.tex
cite our artifact's zenodo doi
check in paper changes as of camera-ready submission on 11/22/24

@@
 style of Moggi, or adjunctions in the style of
-Levy\cite{MOGGI199155,39155,levy99,levystacks}. Relational properties and their verification
+Levy \cite{MOGGI199155,39155,levy99,levystacks}. Relational properties and their verification
 lead naturally to the use of reflexive graph categories or double
-categories\cite{dunphythesis,new-licata18}.
+categories \cite{dunphythesis,new-licata18}.
 
@@
 \cite{veltri-vezzosi2020}, demonstrating that the semantics is readily
-mechanizable. We provide an overview of the mechanization effort, and what
+mechanizable. The Agda formalization is available online \cite{artifact}.
+We provide an overview of the mechanization effort, and what
 remains to be formalized, in Section \ref{sec:mechanization}.

### 2025-01-07 paper-new/paper.tex
remove two of the formatting commands from paper.tex
check in paper changes as of camera-ready submission on 11/22/24

@@
 
+% \theoremstyle{remark} 
+% \newtheorem{theorem}{Theorem}
+% \newtheorem{corollary}{Corollary}
 \theoremstyle{remark}
@@
 \newenvironment{DIFnomarkup}{}{}
-\pagestyle{empty}
-\settopmatter{printfolios=false}
+% \pagestyle{empty}
+% \settopmatter{printfolios=false}
 
@@
          <concept_desc>Software and its engineering~Semantics</concept_desc>
-         <concept_significance>300</concept_significance>
+         <concept_significance>500</concept_significance>
          </concept>
@@
 \ccsdesc[500]{Theory of computation~Denotational semantics}
-\ccsdesc[300]{Software and its engineering~Semantics}
+\ccsdesc[500]{Software and its engineering~Semantics}
 \ccsdesc[300]{Software and its engineering~Functional languages}
@@
 
-\newpage
-\appendix
-\input{appendix}
+% \newpage
+% \appendix
+% \input{appendix}
 
